### 2016-08-03 written report/2levelmixture.Rnw
First

\documentclass[12pt]{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}

\graphicspath{{images/}}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\begin{document}

\section{Variational Bayes - Normal Mixture Model}

We have $iid$ observations $y_i$ genrated by a two level Normal Mixture Model with means $\mu_1 \mbox{ and } \mu_2$ and known variance $1$, so

$$p(y_i | \mu_1, \mu_2, k_i) = \mathcal{N}(\mu_1, 1)^{k_i} \mathcal{N}(\mu_2, 1)^{1-k_i}.$$

where the latent variable $k_i = 1$ if $y_i$ is drawn from $\mathcal{N}(\mu_1, 1)$ and $k_i = 0$ otherwise.

Further, $\bold{k}$ is modelled as $iid$ Bernoulli with parameter $\pi$, so
$$p(k_i | \pi) = \pi^{k_i} (1-\pi)^{1-k_i}. $$

Introducing the priors $p(\pi) \sim U(0, 1)$ and $p(\mu_1, \mu_2) \propto 1$, the joint distribution becomes
\begin{eqnarray}
\label{1}
p(y, k, \mu_1, \mu_2, \pi) & = & \prod^{n}_{i=1} p(y_i | k_i, \mu_1, \mu_2, \pi) p(k_i | \pi) p(\pi) p(\mu_1 \mu_2) \nonumber \\
& \propto & \prod^{n}_{i=1} \left(\frac{1}{\sqrt{2\pi}} \exp \left\{\frac{-(y_i-\mu_1)^2}{2}\right\}\right)^{k_i} \left(\frac{1}{\sqrt{2\pi}} \exp \left\{\frac{-(y_i-\mu_2)^2}{2}\right\}\right)^{1-k_i} \nonumber \\
& \times & \pi^{k_i} (1-\pi)^{1-k_i} \nonumber \\
\ln(p(y, k, \mu_1, \mu_2, \pi)) & = & \sum_{i=1}^{N} \left[ \ln\left(\exp \left\{\frac{-(y_i-\mu_1)^2}{2}\right\}^{k_i}\right)\right] + \sum_{i=1}^{N} \left[ \ln\left(\exp \left\{\frac{-(y_i-\mu_2)^2}{2}\right\}^{1-k_i}\right)\right] \nonumber \\
& + & \sum_{i=1}^{N} k_i \ln(\pi) + \sum_{i=1}^{N} (1-k_i) \ln(1-\pi) \nonumber \\
& = & \sum_{i=1}^{N} \left[ k_{i} \frac{-(y_i-\mu_1)^2}{2}\right] + \sum_{i=1}^{N} \left[ (1-k_{i}) \frac{-(y_i-\mu_2)^2}{2}\right] \nonumber \\
& + & \sum_{i=1}^{N} k_i \ln(\pi) + \sum_{i=1}^{N} (1-k_i) \ln(1-\pi) + c.
\end{eqnarray}

We can take the variational approximation factorisation $q(k_{1:n}, \mu_1, \mu_2, \pi) = \prod_{i=1}^{n} q(k_i)q(\mu_1)q(\mu_2)q(\pi)$, which implies independence of $k_i, k_j$ for $i \neq j$:

It can be shown that the factorisable distribution that minmises the KL Divergence between $q(\theta)$ and $p(\theta|y)$ satisfies

\begin{equation}
\label{2}
q_{i} \propto \exp (\mathbb{E}_{q_{j \neq i}} (\ln(p(y, x, \theta))))
\end{equation}

for all $q_{i}$, where $y$ is the observed data, $x$ is a latent variable and $\theta$ is a vector of unknown parameters.

Substituting (\ref{1}) into (\ref{2}) yields

\begin{eqnarray*}
\ln(q(\pi)) & = & \mathbb{E}_{k_{1:n}} \left[ \sum_{i=1}^{n} k_i \ln(\pi) + (1-k_i) \ln(1-\pi) + c \right] \\
& = & \sum_{i=1}^{n} \mathbb{E}(k_i) \ln (\pi) + (n- \sum_{i=1}^{n}\mathbb{E}(k_i)) \ln(1-\pi) + c \\
& = & \ln(\pi^{\sum_{i=1}^{n}\mathbb{E}(k_i)}(1-\pi)^{n-\sum_{i=1}^{n}\mathbb{E}(k_i)}) + c\\
q(\pi) & \propto & \pi^{\sum_{i=1}^{n}\mathbb{E}(k_i)}(1-\pi)^{n-\sum_{i=1}^{n}\mathbb{E}(k_i)} 
\end{eqnarray*}

Recognizing the kernel of a Beta distribution, we see that $q(\pi) \sim  \mathcal{B} (\alpha = \sum_{i=1}^{n} \mathbb{E}(k_i)+1, \beta = n-\sum_{i=1}^{n}\mathbb{E}(k_i) + 1)$. Continuing, we can find

\begin{eqnarray*}
\ln(q(\mu_1)) & = & \mathbb{E}_{k_{1:n}} \sum_{i=1}^{n} -k_i \frac{(y_i-\mu_1)^2}{2} + c  \\
& = & -\frac{1}{2} \left(\sum_{i=1}^{n}\mathbb{E}(k_i) (y_i-\mu_{1})^2\right) + c \\ 
& = & -\frac{1}{2}\left(\sum_{i=1}^{n}\mathbb{E}(k_i) ((y_i - \tilde{y}_1) + (\tilde{y}_1 - \mu_1))^2 \right)+ c \\
& = & -\frac{1}{2}\left(\sum_{i=1}^{n}\mathbb{E}(k_i) ((y_i - \tilde{y}_1)^2 + (\tilde{y}_1 - \mu_1)^2 - 2(y_i - \tilde{y}_1)(\tilde{y}_1 - \mu_1))\right) + c.
\end{eqnarray*}

Where
$$\tilde{y}_1 = \frac{\sum_{i=1}^{n} \mathbb{E}(k_i)y_i}{\sum_{i=1}^{n} \mathbb{E}(k_i)}.$$

Note that
$$\sum_{i=1}^{n}\mathbb{E}(k_i)(y_i - \tilde{y}_1) = \sum_{i=1}^{n}\mathbb{E}(k_i)\left(y_i - \frac{\sum_{i=1}^{n} \mathbb{E}(k_i)y_i}{\sum_{i=1}^{n} \mathbb{E}(k_i)}\right) = 0,$$

hence 
\begin{eqnarray*}
\ln(q(\mu_1)) & = & -\frac{\sum_{i=1}^{n}\mathbb{E}(k_i) (\tilde{y}_1-\mu_{1})^2}{2} + c.
\end{eqnarray*}

Recognizing the kernel of a Gaussian distribution, we can see that $q(\mu_1) \sim \mathcal{N} (\bar{\mu}_{1} = \tilde{y}_1, \lambda_{1} = (\sum_{i=1}^{n}\mathbb{E}(k_i)^{-1}).$
Similarly, $q(\mu_2) \sim \mathcal{N} (\bar{\mu}_{2} = \tilde{y}_2, \lambda_{2} = \sum_{i=1}^{n}\mathbb{E}(1-k_i)^{-1}))$ with
$$\tilde{y_2} = \frac{\sum_{i=1}^{n} \mathbb{E}(1-k_i)y_i}{\sum_{i=1}^{n} \mathbb{E}(1-k_i)}.$$

Through independence, all $q(k_i)$ have the same form,

\begin{eqnarray*}
\ln(q(k_i)) & = & \mathbb{E}_{\mu_1, \mu_2, \pi} \left[ k_{i} \frac{-(y_i-\mu_1)^2}{2} + (1-k_{i}) \frac{-(y_i-\mu_2)^2}{2} + k_i \ln(\pi) + (1-k_i) \ln(1-\pi) + c \right]\\
& = & k_i\frac{\mathbb{E}_{\mu_1} -(y_i-\mu_1)^2}{2}  + (1-k_i) \frac{\mathbb{E}_{\mu_2} -(y_i-\mu_2)^2}{2}  + k_i \mathbb{E}_\pi \ln(\pi) + (1-k_i) \mathbb{E}_\pi \ln(1-\pi) + c \\
& = & k_i \frac{2\tilde{\pi}_1-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} + (1-k_i) \frac{2\tilde{\pi}_2-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} + c\\
q(k_i) & \propto & \exp \left\{\frac{2\tilde{\pi}_1-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} \right\}^{k_i} \exp \left\{\frac{2\tilde{\pi}_2-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} \right\}^{1-k_i}
\end{eqnarray*}

The quantity $\tilde{\pi}_1 = \mathbb{E}_\pi \ln(\pi) = \psi(\alpha)-\psi(\alpha+\beta)$, and $\tilde{\pi}_2 = \mathbb{E}_\pi \ln(1-\pi) = \psi(\beta)-\psi(\alpha+\beta)$, where $\psi(\cdot)$ is the digamma function (Archambeau and Verleysen 2007).

Each $k_i$ has a Bernoulli distribution with parameters $p_i = \exp \left\{\frac{2\tilde{\pi}_1-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} \right\}, \mbox{ and } q_i = \exp \left\{\frac{2\tilde{\pi}_2-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} \right\}.$

This gives us the update rules for the Variational Bayes iterations:

\begin{eqnarray*}
\alpha & = & \sum_{i=1}^{n} \frac{p_i}{p_i+q_i} + 1 \\
\beta & = & \sum_{i=1}^{n} \frac{q_i}{p_i+q_i} + 1 \\
\bar{\mu}_1 & = &  \frac{\sum_{i=1}^{n} y_i p_i /(p_i+q_i)}{\sum_{i=1}^{n} p_i /(p_i+q_i)} \\
\lambda_1 & = & \left(\sum_{i=1}^{n} \frac{p_i}{p_i+q_i}\right)^{-1} \\
\bar{\mu}_2 & = &  \frac{\sum_{i=1}^{n} y_i q_i /(p_i+q_i)}{\sum_{i=1}^{n} q_i /(p_i+q_i)} \\
\lambda_2 & = & \left(\sum_{i=1}^{n} \frac{q_i}{p_i+q_i}\right)^{-1} \\
p_i & = & \exp \left\{\frac{2(\psi(\alpha)-\psi(\alpha+\beta))-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} \right\} \\
q_i & = & \exp \left\{\frac{2(\psi(\beta)-\psi(\alpha+\beta))-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} \right\}
\end{eqnarray*}

<<vbayes, echo=FALSE>>=
library(ggplot2)
#True Values
set.seed(12)
mu1 <- 3
mu2 <- 6
pi <- 0.6
T <- 250
y1 <- rnorm(T*pi, mu1, 1)
y2 <- rnorm(T*(1-pi), mu2, 1)
y <- c(y1, y2)

#Setup
bmu1 <- vector(length=100)
bmu2 <- bmu1
l1 <- bmu1
l2 <- bmu1
a <- bmu1
b <- bmu1
p <- matrix(0, nrow=100, ncol=T)
q <- p

#Initial values
p0 <- runif(T)
q0 <- runif(T)

#Loop
a[1] = sum(p0/(p0+q0))+1
b[1] = sum(q0/(p0+q0))+1
bmu1[1] = sum(y*p0/(p0+q0))/sum(p0/(p0+q0))
l1[1] = sum(p0/(p0+q0))^(-1)
bmu2[1] = sum(y*q0/(p0+q0))/sum(q0/(p0+q0))
l2[1] = sum(q0/(p0+q0))^(-1)
for(j in 1:T){
  p[1,j] = exp((2*(digamma(a[1])-digamma(a[1]+b[1]))-((y[j]-bmu1[1])^2+l1[1]))/2)
  q[1,j] = exp((2*(digamma(b[1])-digamma(a[1]+b[1]))-((y[j]-bmu2[1])^2+l2[1]))/2)
}
for(i in 2:100) {
  a[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
  b[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
  bmu1[i] = sum(y*p[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))
  l1[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
  bmu2[i] = sum(y*q[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))
  l2[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
  for(j in 1:T){
    p[i,j] = exp((2*(digamma(a[i])-digamma(a[i]+b[i]))-((y[j]-bmu1[i])^2+l1[i]))/2)
    q[i,j] = exp((2*(digamma(b[i])-digamma(a[i]+b[i]))-((y[j]-bmu2[i])^2+l2[i]))/2)
  }
}
supmu1 <- seq(2, 4, 0.01)
denmu1 <- dnorm(supmu1, mean=bmu1[100], sd=sqrt(l1[100]))
supmu2 <- seq(5, 7, 0.01)
denmu2 <- dnorm(supmu2, mean=bmu2[100], sd=sqrt(l2[100]))
suppi <- seq(0, 1, 0.01)
denpi <- dbeta(suppi, a[100], b[100])
qdist <- data.frame(support = c(supmu1, supmu2, suppi), density = c(denmu1, denmu2, denpi), parameter = c(rep("Mu 1", 201), rep("Mu 2", 201), rep("Pi", 101)))
truev <- data.frame(true = c(mu1, mu2, pi), parameter=c("Mu 1", "Mu 2", "Pi"))
ggplot() + geom_line(data=qdist, aes(x=support, y = density)) + geom_vline(data=truev, aes(xintercept=true), colour="red") +
  facet_wrap(~parameter, scales="free") + theme(axis.ticks.y=element_blank(), axis.text.y=element_blank()) + theme_bw()
@
250 draws were simulated with parameters $\mu_1 = 3, \mu_2 = 6, \pi=0.6$ and th e variational algorithm was ran. After manually correcting mislabeling, $y_i$ was allocated to distribution 1 if $p_i > q_i$ and to distribution 2 if $p_i < q_i$, resulting in the successful classification of 235/250 draws. A more trivial decision rule to allocate $y_i$ to distribution 1 if $y_i < \bar{y}$ and to distribution 2 if $y_i > \bar{y}$ successfully classified 234/250 draws.

<<vbayes2, echo=FALSE>>=
library(ggplot2)
#True Values
set.seed(141242)
mu1 <- 5.5
mu2 <- 6
pi <- 0.6
T <- 250
y1 <- rnorm(T*pi, mu1, 1)
y2 <- rnorm(T*(1-pi), mu2, 1)
y <- c(y1, y2)

#Setup
bmu1 <- vector(length=100)
bmu2 <- bmu1
l1 <- bmu1
l2 <- bmu1
a <- bmu1
b <- bmu1
p <- matrix(0, nrow=100, ncol=T)
q <- p

#Initial values
p0 <- runif(T)
q0 <- runif(T)

#Loop
a[1] = sum(p0/(p0+q0))+1
b[1] = sum(q0/(p0+q0))+1
bmu1[1] = sum(y*p0/(p0+q0))/sum(p0/(p0+q0))
l1[1] = sum(p0/(p0+q0))^(-1)
bmu2[1] = sum(y*q0/(p0+q0))/sum(q0/(p0+q0))
l2[1] = sum(q0/(p0+q0))^(-1)
for(j in 1:T){
  p[1,j] = exp((2*(digamma(a[1])-digamma(a[1]+b[1]))-((y[j]-bmu1[1])^2+l1[1]))/2)
  q[1,j] = exp((2*(digamma(b[1])-digamma(a[1]+b[1]))-((y[j]-bmu2[1])^2+l2[1]))/2)
}
for(i in 2:100) {
  a[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
  b[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
  bmu1[i] = sum(y*p[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))
  l1[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
  bmu2[i] = sum(y*q[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))
  l2[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
  for(j in 1:T){
    p[i,j] = exp((2*(digamma(a[i])-digamma(a[i]+b[i]))-((y[j]-bmu1[i])^2+l1[i]))/2)
    q[i,j] = exp((2*(digamma(b[i])-digamma(a[i]+b[i]))-((y[j]-bmu2[i])^2+l2[i]))/2)
  }
}
supmu1 <- seq(5, 7, 0.01)
denmu1 <- dnorm(supmu1, mean=bmu1[100], sd=sqrt(l1[100]))
supmu2 <- seq(5, 7, 0.01)
denmu2 <- dnorm(supmu2, mean=bmu2[100], sd=sqrt(l2[100]))
suppi <- seq(0, 1, 0.01)
denpi <- dbeta(suppi, a[100], b[100])
qdist <- data.frame(support = c(supmu1, supmu2, suppi), density = c(denmu1, denmu2, denpi), parameter = c(rep("Mu 1", 201), rep("Mu 2", 201), rep("Pi", 101)))
truev <- data.frame(true = c(mu1, mu2, pi), parameter=c("Mu 1", "Mu 2", "Pi"))
ggplot() + geom_line(data=qdist, aes(x=support, y = density)) + geom_vline(data=truev, aes(xintercept=true), colour="red") +
  facet_wrap(~parameter, scales="free") + theme(axis.ticks.y=element_blank(), axis.text.y=element_blank()) + theme_bw()
@
250 draws were simulated with parameters $\mu_1 = 5.5, \mu_2 = 6, \pi=0.6$ to try and force an overlap in the data. $y_i$ was allocated to distribution 1 if $p_i > q_i$ and to distribution 2 if $p_i < q_i$, resulting in the successful classification of 158/250 draws. The trivial decision rule had identical classifications.

\end{document}

### 2016-08-03 written report/2levelmixture.tex
First

--- 2016-08-03 written report/2levelmixture.Rnw
@@
-\documentclass[12pt]{article}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage{natbib}
@@
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 The quantity $\tilde{\pi}_1 = \mathbb{E}_\pi \ln(\pi) = \psi(\alpha)-\psi(\alpha+\beta)$, and $\tilde{\pi}_2 = \mathbb{E}_\pi \ln(1-\pi) = \psi(\beta)-\psi(\alpha+\beta)$, where $\psi(\cdot)$ is the digamma function (Archambeau and Verleysen 2007).
+We also expanded the quadratic term and substitued in $(y_i^2 - \bar{\mu}_j)^2 + \lambda_j$ for $\mathbb{E} (y_i-\mu_j)^2$.
 
@@
 \begin{eqnarray*}
-\alpha & = & \sum_{i=1}^{n} \frac{p_i}{p_i+q_i} + 1 \\
-\beta & = & \sum_{i=1}^{n} \frac{q_i}{p_i+q_i} + 1 \\
+a & = & \sum_{i=1}^{n} \frac{p_i}{p_i+q_i} + 1 \\
+b & = & \sum_{i=1}^{n} \frac{q_i}{p_i+q_i} + 1 \\
 \bar{\mu}_1 & = &  \frac{\sum_{i=1}^{n} y_i p_i /(p_i+q_i)}{\sum_{i=1}^{n} p_i /(p_i+q_i)} \\
@@
 \lambda_2 & = & \left(\sum_{i=1}^{n} \frac{q_i}{p_i+q_i}\right)^{-1} \\
-p_i & = & \exp \left\{\frac{2(\psi(\alpha)-\psi(\alpha+\beta))-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} \right\} \\
-q_i & = & \exp \left\{\frac{2(\psi(\beta)-\psi(\alpha+\beta))-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} \right\}
+p_i & = & \exp \left\{\frac{2(\psi(a)-\psi(a+b))-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} \right\} \\
+q_i & = & \exp \left\{\frac{2(\psi(b)-\psi(a+b))-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} \right\}
 \end{eqnarray*}
 
-<<vbayes, echo=FALSE>>=
-library(ggplot2)
-#True Values
-set.seed(12)
-mu1 <- 3
-mu2 <- 6
-pi <- 0.6
-T <- 250
-y1 <- rnorm(T*pi, mu1, 1)
-y2 <- rnorm(T*(1-pi), mu2, 1)
-y <- c(y1, y2)
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/vbayes-1}
 
-#Setup
-bmu1 <- vector(length=100)
-bmu2 <- bmu1
-l1 <- bmu1
-l2 <- bmu1
-a <- bmu1
-b <- bmu1
-p <- matrix(0, nrow=100, ncol=T)
-q <- p
-
-#Initial values
-p0 <- runif(T)
-q0 <- runif(T)
-
-#Loop
-a[1] = sum(p0/(p0+q0))+1
-b[1] = sum(q0/(p0+q0))+1
-bmu1[1] = sum(y*p0/(p0+q0))/sum(p0/(p0+q0))
-l1[1] = sum(p0/(p0+q0))^(-1)
-bmu2[1] = sum(y*q0/(p0+q0))/sum(q0/(p0+q0))
-l2[1] = sum(q0/(p0+q0))^(-1)
-for(j in 1:T){
-  p[1,j] = exp((2*(digamma(a[1])-digamma(a[1]+b[1]))-((y[j]-bmu1[1])^2+l1[1]))/2)
-  q[1,j] = exp((2*(digamma(b[1])-digamma(a[1]+b[1]))-((y[j]-bmu2[1])^2+l2[1]))/2)
-}
-for(i in 2:100) {
-  a[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
-  b[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
-  bmu1[i] = sum(y*p[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))
-  l1[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
-  bmu2[i] = sum(y*q[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))
-  l2[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
-  for(j in 1:T){
-    p[i,j] = exp((2*(digamma(a[i])-digamma(a[i]+b[i]))-((y[j]-bmu1[i])^2+l1[i]))/2)
-    q[i,j] = exp((2*(digamma(b[i])-digamma(a[i]+b[i]))-((y[j]-bmu2[i])^2+l2[i]))/2)
-  }
-}
-supmu1 <- seq(2, 4, 0.01)
-denmu1 <- dnorm(supmu1, mean=bmu1[100], sd=sqrt(l1[100]))
-supmu2 <- seq(5, 7, 0.01)
-denmu2 <- dnorm(supmu2, mean=bmu2[100], sd=sqrt(l2[100]))
-suppi <- seq(0, 1, 0.01)
-denpi <- dbeta(suppi, a[100], b[100])
-qdist <- data.frame(support = c(supmu1, supmu2, suppi), density = c(denmu1, denmu2, denpi), parameter = c(rep("Mu 1", 201), rep("Mu 2", 201), rep("Pi", 101)))
-truev <- data.frame(true = c(mu1, mu2, pi), parameter=c("Mu 1", "Mu 2", "Pi"))
-ggplot() + geom_line(data=qdist, aes(x=support, y = density)) + geom_vline(data=truev, aes(xintercept=true), colour="red") +
-  facet_wrap(~parameter, scales="free") + theme(axis.ticks.y=element_blank(), axis.text.y=element_blank()) + theme_bw()
-@
+\end{knitrout}
 250 draws were simulated with parameters $\mu_1 = 3, \mu_2 = 6, \pi=0.6$ and th e variational algorithm was ran.
@@
 
-<<vbayes2, echo=FALSE>>=
-library(ggplot2)
-#True Values
-set.seed(141242)
-mu1 <- 5.5
-mu2 <- 6
-pi <- 0.6
-T <- 250
-y1 <- rnorm(T*pi, mu1, 1)
-y2 <- rnorm(T*(1-pi), mu2, 1)
-y <- c(y1, y2)
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/vbayes2-1}
 
-#Setup
-bmu1 <- vector(length=100)
-bmu2 <- bmu1
-l1 <- bmu1
-l2 <- bmu1
-a <- bmu1
-b <- bmu1
-p <- matrix(0, nrow=100, ncol=T)
-q <- p
-
-#Initial values
-p0 <- runif(T)
-q0 <- runif(T)
-
-#Loop
-a[1] = sum(p0/(p0+q0))+1
-b[1] = sum(q0/(p0+q0))+1
-bmu1[1] = sum(y*p0/(p0+q0))/sum(p0/(p0+q0))
-l1[1] = sum(p0/(p0+q0))^(-1)
-bmu2[1] = sum(y*q0/(p0+q0))/sum(q0/(p0+q0))
-l2[1] = sum(q0/(p0+q0))^(-1)
-for(j in 1:T){
-  p[1,j] = exp((2*(digamma(a[1])-digamma(a[1]+b[1]))-((y[j]-bmu1[1])^2+l1[1]))/2)
-  q[1,j] = exp((2*(digamma(b[1])-digamma(a[1]+b[1]))-((y[j]-bmu2[1])^2+l2[1]))/2)
-}
-for(i in 2:100) {
-  a[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
-  b[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))+1
-  bmu1[i] = sum(y*p[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))
-  l1[i] = sum(p[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
-  bmu2[i] = sum(y*q[(i-1),]/(p[(i-1),]+q[(i-1),]))/sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))
-  l2[i] = sum(q[(i-1),]/(p[(i-1),]+q[(i-1),]))^(-1)
-  for(j in 1:T){
-    p[i,j] = exp((2*(digamma(a[i])-digamma(a[i]+b[i]))-((y[j]-bmu1[i])^2+l1[i]))/2)
-    q[i,j] = exp((2*(digamma(b[i])-digamma(a[i]+b[i]))-((y[j]-bmu2[i])^2+l2[i]))/2)
-  }
-}
-supmu1 <- seq(5, 7, 0.01)
-denmu1 <- dnorm(supmu1, mean=bmu1[100], sd=sqrt(l1[100]))
-supmu2 <- seq(5, 7, 0.01)
-denmu2 <- dnorm(supmu2, mean=bmu2[100], sd=sqrt(l2[100]))
-suppi <- seq(0, 1, 0.01)
-denpi <- dbeta(suppi, a[100], b[100])
-qdist <- data.frame(support = c(supmu1, supmu2, suppi), density = c(denmu1, denmu2, denpi), parameter = c(rep("Mu 1", 201), rep("Mu 2", 201), rep("Pi", 101)))
-truev <- data.frame(true = c(mu1, mu2, pi), parameter=c("Mu 1", "Mu 2", "Pi"))
-ggplot() + geom_line(data=qdist, aes(x=support, y = density)) + geom_vline(data=truev, aes(xintercept=true), colour="red") +
-  facet_wrap(~parameter, scales="free") + theme(axis.ticks.y=element_blank(), axis.text.y=element_blank()) + theme_bw()
-@
+\end{knitrout}
 250 draws were simulated with parameters $\mu_1 = 5.5, \mu_2 = 6, \pi=0.6$ to try and force an overlap in the data.

### 2016-08-03 written report/Variational_Bayes_derivation.tex
First

\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}

\graphicspath{{images/}}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\title{Derivation of Variational Bayes}

\section{The Evidence Lower Bound}

We are interested in finding the distribution $q(\theta)$ that best approximates the true posterior $p(\theta|y)$, with the best approximation taken as the $q(\theta)$ that minimises the Kullback-Leibler divergence with the posterior, defined as

$$KL[q(\theta)|p(\theta|y)] = \int q(\theta) \ln \left( \frac{q(\theta)}{p(\theta|y)}\right) d\theta.$$

Beginning with the unknown constant $p(y)$, we have that

\begin{eqnarray}
\ln(p(y)) & = & \int_{\theta} q(\theta) \ln(p(y)) d\theta \nonumber \\
& = & \int_{\theta} q(\theta) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \right) d\theta \nonumber \\
& = & \int_{\theta} q(\theta) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \frac{q(\theta)}{q(\theta)}\right) d\theta \nonumber \\
& = & \int_{\theta} q(\theta) \ln  \left( \frac{q(\theta)}{p(\theta | y)}\right) + \ln \left( \frac{p(y, \theta)}{q(\theta)} \right) d\theta \nonumber \\
& = & \int_{\theta} q(\theta) \ln  \left( \frac{q(\theta)}{p(\theta | y)}\right) d\theta + \int_{\theta} q(\theta) \ln \left( \frac{p(y, \theta)}{q(\theta)} \right) d\theta \nonumber \\
& = & KL [q(\theta) || p(\theta | y)] + F(q, y) \nonumber
\end{eqnarray}

where 
\begin{eqnarray*}
F(q, y) & = & \int_{\theta} q(\theta) \ln \left( \frac{p(y, \theta)}{q(\theta)} \right) d\theta \\
& = &  \int_{\theta} q(\theta) \ln (p(y, \theta)) d\theta -  \int_{\theta} q(\theta) \ln (q(\theta)) d\theta.
\end{eqnarray*}

As $\ln(p(y))$ is constant, the distribution $q(\theta)$ that minimises the KL divergence can be found by maximising $F(q, y)$, which is called the Evidence Lower Bound (ELBO, or $\mathcal{L}(q, y)$) in the machine learning literature. We are yet to make any assumptions about the distribution of $q(\theta)$.

\section{The Mean Field Assumption}
The mean-field assumption is that $q(\theta)$ can be factorised as $q(\theta) = \prod_{i} q_{i}(\theta_{i})$, where each $i$ may be a scalar or vector.
The mean-field derivations follows, assuming that $q(\theta) = q_{1}(\theta_{1}) q_{2}(\theta_{2})$.

\begin{eqnarray}
F(q, y) & = &\int_{\theta} q_{1}q_{2} \ln (p(y, \theta)) d\theta - \int_{\theta} q_{1}q_{2} \ln (q_{1}q_{2}) d\theta \nonumber \\
& = & \int_{\theta} q_{1}q_{2} ( \ln (p(y, \theta)) - \ln(q_{1})) d\theta - \int_{\theta} q_{1}q_{2} \ln (q_{2}) d\theta \\
& = & \int_{\theta_{1}} q_{1} \left( \int_{\theta_{2}} q_{2} \ln (p(y, \theta )) d\theta_{2}- \ln(q_{1}) \right) d\theta_{1} - \int_{\theta_{1}} q_{1} \int_{\theta_{2}} q_{2} \ln(q_{2}) d\theta_{2} d\theta_{1} \nonumber \\
& = & \int_{\theta_{1}} q_{1} \ln \left( \frac{\exp(  \mathbb{E}_{q_2} [\ln(p(y,\theta))])}{q_{1}} \right) d\theta_{1} + c \\
& = & -KL (q_{1} || \exp( \mathbb{E}_{q_2} [\ln(p(y,\theta))])) + c \nonumber
\end{eqnarray}

(1) uses the fact that
$$ \int_{\theta_{2}} q_{2} \ln(q_{1}) d\theta_{2} = \ln(q_{1}),$$

and (2) uses the fact that
$$ \int_{\theta_{1}} q_{1} \int_{\theta_{2}} q_{2} \ln(q_{2}) d\theta_{2} d\theta_{1} = \mathbb{E}_{q_1}\left[  \int_{\theta_{2}} q_{2} \ln(q_{2}) d\theta_{2}\right] =  \int_{\theta_{2}} q_{2} \ln(q_{2}) d\theta_{2}, $$

a constant term with respect to $q_{1}$.

Maximisation with respect to $q_{1}$ is simple due to the KL divergence term, as we can minimise 
$$KL (q_{1} || \exp(E_{q_{2}} (\ln(p(y,\theta)))))$$
by setting 
$$q_{1} \propto \exp(E_{q_{2}} (\ln(p(y,\theta))))$$

for every $i$ in $\theta$.

\section{Copula Variational Bayes}

If we attach a copula function to our approximation, so that $q(\theta) = q_1(\theta_1) q_2(\theta_2) c(Q_1(\theta_1), Q_2(\theta_2)$, we get the Evidence Lower Bound of
\begin{eqnarray*}
F(q, y) & = & \int_{\theta} q_{1}q_{2}c(Q_1, Q_2) \ln (p(y, \theta)) d\theta - \int_{\theta} q_{1}q_{2}c(Q_1, Q_2) \ln (q_{1}q_{2}c(Q_1, Q_2)) d\theta \\
& = & \int_{\theta_1} q_{1}c(Q_1, Q_2) \left( \int_{\theta_2} q_{2}\ln (p(y, \theta))d\theta_2 - \ln(q_{1}) \right) d\theta_1 - \int_{\theta} q_{1}q_{2}c(Q_1, Q_2) \ln (q_{2}c(Q_1, Q_2)) d\theta \\
& = & \int_{\theta_{1}} q_{1}c(Q_1, Q_2) \ln \left( \frac{\exp(  \mathbb{E}_{q_2} [\ln(p(y,\theta))])}{q_{1}} \right) d\theta_{1} + f(q_{1}, q_{2}).
\end{eqnarray*}

This derivation runs into two problems: the first term is no longer a KL divergence, and the second term is no longer constant with respect to $q_{1}$.
If, between the first and second lines we moved $q_{1}q_{2}c(Q_1, Q_2) \ln(q_{1}c(Q_1, Q_2))$ from the second term to the first, instead of $q_{1}q_{2}c(Q_1, Q_2) \ln(q_{1})$ we could attempt to the match the denominator with the $ q_{1}c(Q_1, Q_2)$ term and make something in the form of a KL divergence.

We would have had
\begin{eqnarray*}
F(q, y) & = & \int_{\theta_1} q_{1}c(Q_1, Q_2) \left( \int_{\theta_2} q_{2}\ln (p(y, \theta))d\theta_2 - \int_{\theta_{2}} q_{2} \ln(q_{1}c(Q_1, Q_2)) d\theta_{2}\right) d\theta_1 \\
& - & \int_{\theta} q_{1}q_{2}c(Q_1, Q_2) \ln (q_{2}) d\theta
\end{eqnarray*}
and still would not get the KL divergence term as the result used in (1) would no longer hold as
$$ \int_{\theta_{2}} q_{2} \ln(q_{1}c(Q_1, Q_2)) d\theta_{2} \neq \ln(q_{1}c(Q_1, Q_2)).$$

Even if we could do this, the second term would be
\begin{eqnarray*}
\int_{\theta} q_{1}q_{2}c(Q_1, Q_2) \ln (q_{2}) d\theta & = &  \int_{\theta_1} q_{1} \int_{\theta_2} q_{2}c(Q_1, Q_2) \ln (q_{2}) d\theta_2 d\theta_1 \\
& = & \mathbb{E}_{q_1} \left [\int_{\theta_2} q_{2}c(Q_1, Q_2) \ln (q_{2}) d\theta_2 \right]
\end{eqnarray*}
which is not constant with respect to $q_{1}$.

Clearly the mean field approach to the derivation can not be directly adapted.

\end{document}

### 2016-08-03 written report/ar1_predictive.Rnw
First

\documentclass{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage{wrapfig,float}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage[section]{placeins}
\usepackage{afterpage}
\usepackage{bbm}
\begin{document}
\SweaveOpts{concordance=TRUE}

Note: Code uses a stretched beta, latex math still uses regular beta.

We have $T$ observations from an AR(1) model:

$$y_{t} = \mu + \psi y_{t-1}+ \epsilon_{t},$$

Where $\epsilon_{t}$ is $iid \hspace{4pt} \mathcal{N}(0,\sigma^2).$

We use $y_1 \sim \mathcal{N}(0, \tau^2)$, which results in the vector $y_{1:T}$ having the likelihood function

\begin{eqnarray*}
\label{likelihood}
L(\mu, \psi, \sigma | y_{1:T}, \tau) & = & \mathcal{N}(0, \tau) \prod_{t=2}^{T} \mathcal{N}(\mu + \psi y_{t-1}, \sigma^2) \nonumber \\
& = & \frac{1}{(2\pi)^T} \tau^{-1} \exp \left\{ \frac{-y_1^2}{2\tau^2}\right\} \sigma^{-(T-1)} \exp \left\{ \frac{-\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1}))}{2\sigma^2}\right\}  
\end{eqnarray*}

We restrict $\psi$ to $(0, 1)$ and use the priors:
\begin{eqnarray*}
\label{priors}
p(\mu) & \propto & 1 \\
p(\sigma^2) & \propto & \sigma^{-2} \\
p(\psi) & \sim & \mathcal{B}(p1, p2).
\end{eqnarray*}

We get a joint distribution
\begin{eqnarray*}
\label{joint}
p(y_{1:T}, \mu, \psi, \sigma^2) & \propto &  \tau^{-1} \sigma^{-(T-1)}  \exp \left\{ \frac{-y_1^2}{2\tau^2}\right\} \exp \left\{ \frac{-\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1}))^2}{2\sigma^2}\right\} \psi^{p1-1}(1-\psi)^{p2-1}
\end{eqnarray*}

However, we are interested in the forecast distribution of $y_{T+1}$, where $p(y_{T+1} | \mu, \psi, \sigma^2, y_T) \sim \mathcal{N}(\mu + \psi y_T, \sigma^2).$

Incorporating this into the above results in
\begin{eqnarray*}
\label{jointpred}
p(y_{1:T+1}, \mu, \psi, \sigma^2) & \propto &  \tau^{-1} \sigma^{-(T-2)}  \exp \left\{ \frac{-y_1^2}{2\tau^2}\right\} \exp \left\{ \frac{-\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1}))^2}{2\sigma^2}\right\} \\
& \times &  \psi^{p1-1}(1-\psi)^{p2-1} \exp \left\{ \frac{-(y_{T+1}-(\mu+\psi y_T))^2}{2\sigma^2} \right\} \\
\ln(p(y_{1:T+1}, \mu, \psi, \sigma^2)) & = & -\ln(\tau) - (T-2) \ln(\sigma) - \frac{y_1^2}{2\tau^2} - \frac{\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1}))^2}{2\sigma^2} \\
& - &  \frac{(y_{T+1}-(\mu+\psi y_T))^2}{2\sigma^2} + (p_1-1) \ln(\psi) + (p_2-1)\ln(1-\psi).
\end{eqnarray*}

We use the factorisation $q(\mu, \psi, \sigma^2, y_{T+1}) = q(\mu)q(\psi)q(\sigma^2)q(y_{T+1})$ in the variational bayes approximation.

It can be shown that the approximations that minimise the KL Divergence between the true posterior and the approximation is of the form
$$q_{\theta_i} \propto \exp(E_{q_{\theta_{\setminus i}}} (\ln(p(y,\theta)))) \mbox{ for i = 1,2, \dots}$$
where the notation $\theta_{\setminus i}$ refers to all $\theta$ except $\theta_i$.

Proceeding by taking the expectation of the log joint distribution with respect to all terms but $\mu$, and ignoring all terms that do not depend on $\mu$ we have

\begin{eqnarray*}
\label{qmu}
\ln(q(\mu)) & = & \mathbb{E}_{\psi, \sigma^2, y_{T+1}} \left[ \frac{-1}{2\sigma^2} \left(\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1})) + (y_{T+1}-(\mu+\psi y_T))^2 \right) \right] + c \\
& = &  \mathbb{E}_{\sigma^2} \left[ \frac{-1}{2\sigma^2} \right] \mathbb{E}_{\psi, y_{T+1}} \Bigg[ \sum^{T}_{t=2} (y_t^2 + (\mu+\psi y_{t-1})^2 - 2 y_t (\mu+\psi y_{t-1})) + \\
& + & y_{T+1}^2 +  (\mu+\psi y_T)^2 - 2 y_{T+1}(\mu+\psi y_T)\Bigg] + c \\
& = & -T\mathbb{E}_{\sigma^2} \left[ \frac{1}{2\sigma^2} \right] \left(\mu^2 -2 \mu \frac{\sum^{T}_{t=2}y_t + \mathbb{E}_{y_{T+1}}[y_{T+1}] - \mathbb{E}_{\psi} [\psi] \sum^{T+1}_{t=2}y_{t-1}}{T} \right) \\
& = & -T\mathbb{E}_{\sigma^2} \left[ \frac{1}{2\sigma^2} \right] (\mu - \bar{\mu})^2 + c
\end{eqnarray*}

We recognize the kernel of a Gaussian Distribution, and hence $q(\mu) \sim \mathcal{N}(\bar{\mu}, \lambda = (T\mathbb{E}_{\sigma^2} [\sigma^{-2} ])^{-1})$ with
$$\bar{\mu} = \frac{\sum^{T}_{t=2}y_t + \mathbb{E}_{y_{T+1}}[y_{T+1}] - \mathbb{E}_{\psi} [\psi ]\sum^{T+1}_{t=2}y_{t-1}}{T}.$$

The derivation for $q(\psi)$ follows on as

\begin{eqnarray*}
\label{qpsi}
\ln(q(\psi)) & = &  \mathbb{E}_{\mu, \sigma^2, y_{T+1}} \left[ \frac{-1}{2\sigma^2} \left(\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1}))^2 + (y_{T+1}-(\mu+\psi y_T))^2 \right) \right] \\
& +  & (p_1-1) \ln(\psi) + (p_2-1)\ln(1-\psi) + c \\
& = & \mathbb{E}_{\sigma^2} \left[ \frac{-1}{2\sigma^2} \right] \mathbb{E}_{\mu, y_{T+1}} \Bigg[ \sum^{T}_{t=2} (y_t^2 + (\mu+\psi y_{t-1})^2 - 2 y_t (\mu+\psi y_{t-1})) + \\
& + & y_{T+1}^2 +  (\mu+\psi y_T)^2 - 2 y_{T+1}(\mu+\psi y_T)\Bigg] + (p_1-1) \ln(\psi) + (p_2-1)\ln(1-\psi) + c \\
& = & \mathbb{E}_{\sigma^2} \left[ \frac{-1}{2\sigma^2} \right] \left(\psi^2 \sum^{T+1}_{t=2} y_{t-1}^2 - 2\psi \left( \sum^{T}_{t=2} y_t y_{t-1} + \mathbb{E}_{y_{T+1}} [y_{T+1}]y_T - \bar{\mu}\sum^{T+1}_{t=2} y_{t-1} \right) \right) \\
& + & (p_1-1) (p_1-1) \ln(\psi) + (p_2-1)\ln(1-\psi) + c \\
& = & \sum^{T+1}_{t=2} y_{t-1}^2  \mathbb{E}_{\sigma^2} \left[ \frac{-1}{2\sigma^2} \right] \left(\psi - \bar{\psi} \right)^2 + (p_1-1) \ln(\psi) + (p_2-1)\ln(1-\psi) + c
\end{eqnarray*}

With $$\bar{\psi} = \frac{\left(\sum^{T}_{t=2} y_t y_{t-1} + \mathbb{E}_{y_{T+1}} [y_{T+1}]y_T - \bar{\mu} \sum^{T+1}_{t=2} y_{t-1} \right)}{\sum^{T+1}_{t=2} y_{t-1}^2}$$

As we chose a non-conjugate prior distribution, this doesn't have the form of an exponential family distribution and we will replace it with a Gaussian Laplace approximation, $\tilde{q}(\psi) \sim \mathcal{N}(\hat{\psi}, \gamma = -H(\hat{\psi})^{-1})$ where $\hat{\psi}$ is the MAP estimate of $\ln(q(\psi))$ and $H(\hat{\psi})$ is the second derivative of $\ln(q(\psi))$ evaluated at $\hat{\psi}$.

Moving onto $q(\sigma^2)$,

\begin{eqnarray*}
\label{qsig}
\ln(q(\sigma^2)) & = & \mathbb{E}_{\mu, \psi, y_{T+1}} \left[  (T-2) \ln(\sigma) - \frac{\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1}))^2 + (y_{T+1}-(\mu+\psi y_T))^2}{2\sigma^2}  \right] \\
& = & (T-2) \ln(\sigma) - \frac{1}{2\sigma^2} \mathbb{E}_{\mu, \psi, y_{T+1}} \left[\sum^{T}_{t=2}(y_t-(\mu+\psi y_{t-1}))^2 + (y_{T+1}-(\mu+\psi y_T))^2 \right] \\
& = & (T-2) \ln(\sigma) - \frac{1}{2\sigma^2} \mathbb{E}_{\mu, \psi, y_{T+1}} \Bigg[ \sum^{T}_{t=2} (y_t^2 + \mu^2 + \psi^2 y_{t-1}^2 + 2 \mu \psi y_{t-1} - 2 y_t (\mu+\psi y_{t-1})) + \\
& + & y_{T+1}^2 +  \mu^2 + \psi^2 y_{T}^2 + 2 \mu \psi y_{T} - 2 y_{T+1}(\mu+\psi y_T)\Bigg] \\
& = & (T-2) \ln(\sigma) - \frac{1}{2\sigma^2} \Bigg[ \sum^{T}_{t=2} \left(y_t^2  - 2 y_t (\bar{\mu} + \hat{\psi} y_{t-1})\right) + T [\bar{\mu}^2 + \lambda ]  \\
& + & \sum^{T+1}_{t=2} \left([\hat{\psi}^2 + \gamma ] y_{t-1}^2 + 2 \bar{\mu} \hat{\psi} y_{t-1} \right) + \mathbb{E}_{y_{T+1}} [y_{T+1}^2 ]  - 2 \mathbb{E}_{y_{T+1}} [y_{T+1} ](\bar{\mu} + \hat{\psi} y_T)\Bigg]
\end{eqnarray*}

We can recognize the kernel of an $Inv. Gamma ( \mbox{shape } (a) = (T-2)/2, \mbox{scale } (b) = (T-2)\bar{\sigma}^2 /2 )$, with

\begin{eqnarray*}
\label{sigbar}
\bar{\sigma}^2 & = & \frac{1}{T-2} \Bigg[ \sum^{T}_{t=2} \left(y_t^2  - 2 y_t (\bar{\mu} + \hat{\psi} y_{t-1})\right) + T [\bar{\mu}^2 + \lambda ] + \sum^{T+1}_{t=2} \left([\hat{\psi}^2 + \gamma ] y_{t-1}^2 + 2 \bar{\mu} \hat{\psi} y_{t-1} \right) \\
& + & \mathbb{E}_{y_{T+1}} [y_{T+1}^2 ]  - 2 \mathbb{E}_{y_{T+1}} [y_{T+1} ](\bar{\mu} + \hat{\psi} y_T) \Bigg]
\end{eqnarray*}

Finally the forecast distribution,

\begin{eqnarray*}
\label{T+1}
\ln(q(y_{T+1})) & = & \mathbb{E}_{\mu, \psi, \sigma^2} \left[ \frac{(y_{T+1}-(\mu+\psi y_T))^2}{2\sigma^2} \right] + c \\
& = & - (a/b)  \mathbb{E}_{\mu, \psi} \left[ y_{T+1}^2+(\mu+\psi y_T)^2 - 2 y_{T+1}(\mu+\psi y_T) \right] + c\\
& = & - (a/b) \left(Y_{T+1} - (\bar{\mu} + \hat{\psi}Y_{T}) \right) + c
\end{eqnarray*}

This is the kernel of a $\mathcal{N}\left(\bar{y}_{T+1} = \bar{\mu} + \hat{\psi}Y_{T}, \delta = b/a \right)$ distribution.

The resulting update equations are:

\begin{eqnarray*}
\label{update}
\bar{\mu} & = & \frac{1}{T} \left(\sum^{T}_{t=2}y_t + \bar{y}_{T+1} - \hat{\psi}\sum^{T+1}_{t=2}y_{t-1} \right) \\
\lambda & = & \frac{b}{aT} \\
\bar{\psi} & = &\frac{\left(\sum^{T}_{t=2} y_t y_{t-1} + \bar{y}_{T+1}y_T - \bar{\mu} \sum^{T+1}_{t=2} y_{t-1} \right)}{\sum^{T+1}_{t=2} y_{t-1}^2} \\
\hat{\psi} & = & \arg \max_{\psi} \left\{  \frac{-b\sum^{T+1}_{t=2} y_{t-1}^2}{a} \left(\psi - \bar{\psi} \right)^2 + (p_1-1) \ln(\psi) + (p_2-1)\ln(1-\psi) \right\} \\
\gamma & = & -\left(\frac{d^2 \ln(q(\psi))}{d \psi^2} \vline_{\psi = \hat{\psi}}\right)^{-1} \\
a & = & \frac{T-2}{2} \\
b & = & \frac{1}{2} \Bigg[\sum^{T}_{t=2}\left(y_t^2-2y_t(\bar{\mu}+\hat{\psi}y_{t-1})\right) + T(\bar{\mu}^2 + \lambda) \\
& + &  \sum^{T+1}_{t=2}\left((\hat{\psi}^2+\gamma)y_{t-1}^2 + 2\bar{\mu}\hat{\psi}y_{t-1}\right) + \bar{y}_{T+1}^2 + \delta - 2 \bar{y}_{T+1} (\bar{\mu} + \hat{\psi}y_T) \Bigg] \\
\bar{Y}_{T+1} & = & \bar{\mu} + \hat{\psi}y_T \\
\delta & = & \frac{b}{a}
\end{eqnarray*}

An AR(1) model was simulated with $T=100, \mu = 3, \psi = 0.5, \sigma^2 = 1, p1 = 2$ and $p2 = 3$. The algorithm ran for 500 iterations and converged after approximately 200. The algorithm took 0.204 seconds to run in R and resulted in a $\mathcal{N}(3.06, 0.96)$ distribution for $q(Y_{T+1})$.

<<forecast, echo=FALSE>>=
##One step variational forecast
library(Rcpp)
library(tidyr)
library(ggplot2)
library(dplyr)
library(RcppArmadillo)
sourceCpp("AR1c.cpp")
library(truncnorm)
library(flexmix)

set.seed(100)
T <- 100
tmu <- 3
tpsi <- 0.5
tsig <- 1
y <- tmu + arima.sim(list(ar=c(tpsi)), T, sd=tsig)

#Empty vectors
N <- 500
barmu = vector(length=N)
lambda = vector(length=N)
hatpsi = vector(length=N)
gamma = vector(length=N)
a = vector(length=N)
b = vector(length=N)
yhat = vector(length=N)
d = vector(length=N)

b0 = 1
a0 = 1
yhat0 = 1
psi0 = 0.8
gamma0 = 1
barmu0 = 1
lambda0 = 1
d0 = 1
p1 = 2
p2 = 3

#psi kernel function
psi.k <- function(psi, y, a, b, yhat, mu, p1, p2) {
  psibar = (sum(y[2:T]*y[1:T-1])+yhat*y[T]-mu*sum(y))/sum(y^2)
  out = sum(y^2)*(-b/a)*(psi-psibar)^2+(p1-1)*log(psi)+(p2-1)*log(1-psi)
  return(out)
}
h <- 1e-08
psi.hessian <- function(psi, y, a, b, p1, p2) {
  const = sum(y^2)*(-b/a)*2
  out = const-(p1-1)/(psi+1)^2-(p2-1)/(1-psi)^2
  return(out)
}

#Iteration 1
barmu[1] = 1/T*(sum(y[2:T]) + yhat0 - psi0*sum(y))
lambda[1] = b0/(a0*T)
hatpsi[1] = optimise(psi.k, c(0, 1), y, a0, b0, yhat0, barmu0, p1, p2, maximum=TRUE)$maximum
gamma[1] = -1/(psi.hessian(psi0, y, a0, b0, p1, p2))
a = (T-2)/2
b[1] = 1/2*(sum(y[2:T]^2-2*y[2:T]*(barmu0+psi0*y[1:(T-1)])) + T*(barmu0^2+lambda0) +
            sum((psi0^2+gamma0)*y[1:T]^2+2*barmu0*psi0*y[1:T]) + yhat0^2+d0-2*yhat0*(barmu0+psi0*y[T]))
yhat[1] = barmu0 + psi0*y[T]
d[1] = b0/a0

#Loop
for(i in 2:N){
  barmu[i] = 1/T*(sum(y[2:T]) + yhat[(i-1)] - hatpsi[(i-1)]*sum(y))
  lambda[i] = b[(i-1)]/(a*T)
  hatpsi[i] = optimise(psi.k, c(0, 1), y, a, b[(i-1)], yhat[(i-1)], barmu[(i-1)], p1, p2, maximum=TRUE)$maximum
  gamma[i] = -1/(psi.hessian(hatpsi[(i-1)], y, a, b[(i-1)], p1, p2))
  b[i] = 1/2*(sum(y[2:T]^2-2*y[2:T]*(barmu[(i-1)]+hatpsi[(i-1)]*y[1:(T-1)])) + T*(barmu[(i-1)]^2+lambda[(i-1)]) +
                sum((hatpsi[(i-1)]^2+gamma[(i-1)])*y[1:T]^2+2*barmu[(i-1)]*hatpsi[(i-1)]*y[1:T]) +
                yhat[(i-1)]^2+d[(i-1)]-2*yhat[(i-1)]*(barmu[(i-1)]+hatpsi[(i-1)]*y[T]))
  yhat[i] = barmu[(i-1)] + hatpsi[(i-1)]*y[T]
  d[i] = b[(i-1)]/a
}
support <- seq(-1, 8, 0.005)
dens <- dnorm(support, yhat[N], sqrt(d[N]))
@

A Metropolis Hastings MCMC scheme was set up using a random walk candidate distribution for $\psi$ to draw from $p(\mu | y_{1:T}), p(\sigma^2 | y_{1:T})$ and $p(\psi | y_{1:T})$. These draws were then used to find the Rao-Blackwellised distribution of $p(y_{T+1} | y_{1:T})$.
The chain had 1000 burn in draws followed by 50000 kept draws and took 6.736 seconds to complete the chain and forecast distribution.
<<mcmc, echo=FALSE>>=
##MCMC
B <- 1000
K <- 50000
chain <- matrix(0, ncol=3, nrow=(B+K))
mu0 = 1
sig0 = 1
psi0 = 0.8

psi.kmcmc <- function(psi, mu, sigma2, y, p1, p2) {
  out = -1/(2*sigma2)*sum((y[2:T]-(mu+psi*y[1:(T-1)]))^2) + (p1-1)*log((1+psi)/2) + (p2-1)*log((1-psi)/2)
  return(exp(out))
}

#Starting iteration
mubar = sum(y[2:T]-chain[1,2]*(y[1:(T-1)]))/T
chain[1,1] = rnorm(1, mubar, sqrt(sig0/T))
old = psi0
candidate = rtruncnorm(1, -1, 1, old, 0.15)
alpha = psi.kmcmc(candidate, chain[1,1], sig0, y, p1, p2)/dtruncnorm(candidate, -1, 1, old, 0.15)/
        (psi.kmcmc(old, chain[1,1], sig0, y, p1, p2)/dtruncnorm(old, -1, 1, candidate, 0.15))
alpha = min(alpha, 1)
test = runif(1)
if(alpha > test) {
  chain[1,2] = candidate
} else {
  chain[1,2] = old
}

chain[1,3] = 1/rgamma(1, shape=(T-1)/2, rate=sum((y[2:T]-(chain[1,1]+chain[1,2]*y[1:(T-1)]))^2)/2)

#Loop
for(i in 2:(B+K)) {
  mubar = sum(y[2:T]-chain[(i-1),2]*(y[1:(T-1)]))/T
  chain[i,1] = rnorm(1, mubar, sqrt(chain[(i-1),3]/T))
  old = chain[(i-1),2]
  candidate = rtruncnorm(1, -1, 1, old, 0.15)
  alpha = psi.kmcmc(candidate, chain[1,1], sig0, y, p1, p2)/dtruncnorm(candidate, -1, 1, old, 0.15)/
         (psi.kmcmc(old, chain[1,1], sig0, y, p1, p2)/dtruncnorm(old, -1, 1, candidate, 0.15))
  alpha = min(alpha, 1)
  test = runif(1)
  if(alpha > test) {
    chain[i,2] = candidate
  } else {
    chain[i,2] = old
  }
  chain[i,3] = 1/rgamma(1, shape=(T-1)/2, rate=sum((y[2:T]-(chain[i,1]+chain[i,2]*y[1:(T-1)]))^2)/2)
}

ypred = data.frame(support = seq(-1, 8, 0.005))
ypred$density = rao(chain[1001:51000,], ypred$support, 50000, y[T])
@
\begin{figure}[hb]
\centering
<<plots, echo=FALSE, fig=TRUE>>=
densities <- cbind(q = dens, p = ypred$density)
ggplot() + geom_line(data=data.frame(sup=support, dens=dens), aes(x=sup, y=dens), colour="blue") + theme_bw() + 
  geom_line(data=data.frame(s = support, d = ypred$density), aes(x=s, y=d), colour="red") + labs(x="Support", y="Density")
@
\caption{Variational approximation for $q(y_{T+1}|y_{1:T})$ (blue) and the MCMC density for $p(y_{T+1}|y_{1:T})$ (red). 
The KL Divergence from q to p is 0.0049, and from p to q is 0.0056.}
\label{densities}
\end{figure}

\end{document}

### 2016-08-03 written report/ar1_predictive.tex
First

--- 2016-08-03 written report/ar1_predictive.Rnw
@@
 \usepackage{bbm}
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
-
-Note: Code uses a stretched beta, latex math still uses regular beta.
+\input{ar1_predictive-concordance}
 
@@
 
-<<forecast, echo=FALSE>>=
-##One step variational forecast
-library(Rcpp)
-library(tidyr)
-library(ggplot2)
-library(dplyr)
-library(RcppArmadillo)
-sourceCpp("AR1c.cpp")
-library(truncnorm)
-library(flexmix)
+A Metropolis Hastings MCMC scheme was set up using a random walk candidate distribution for $\psi$ to draw from $p(\mu | y_{1:T}), p(\sigma^2 | y_{1:T})$ and $p(\psi | y_{1:T})$.
+These draws were then sampled to find the distribution of $p(y_{T+1} | y_{1:T})$.
+The chain had 1000 burn in draws followed by 50000 kept draws and took 6.736 seconds to complete the chain and forecast distribution.
 
-set.seed(100)
-T <- 100
-tmu <- 3
-tpsi <- 0.5
-tsig <- 1
-y <- tmu + arima.sim(list(ar=c(tpsi)), T, sd=tsig)
-
-#Empty vectors
-N <- 500
-barmu = vector(length=N)
-lambda = vector(length=N)
-hatpsi = vector(length=N)
-gamma = vector(length=N)
-a = vector(length=N)
-b = vector(length=N)
-yhat = vector(length=N)
-d = vector(length=N)
-
-b0 = 1
-a0 = 1
-yhat0 = 1
-psi0 = 0.8
-gamma0 = 1
-barmu0 = 1
-lambda0 = 1
-d0 = 1
-p1 = 2
-p2 = 3
-
-#psi kernel function
-psi.k <- function(psi, y, a, b, yhat, mu, p1, p2) {
-  psibar = (sum(y[2:T]*y[1:T-1])+yhat*y[T]-mu*sum(y))/sum(y^2)
-  out = sum(y^2)*(-b/a)*(psi-psibar)^2+(p1-1)*log(psi)+(p2-1)*log(1-psi)
-  return(out)
-}
-h <- 1e-08
-psi.hessian <- function(psi, y, a, b, p1, p2) {
-  const = sum(y^2)*(-b/a)*2
-  out = const-(p1-1)/(psi+1)^2-(p2-1)/(1-psi)^2
-  return(out)
-}
-
-#Iteration 1
-barmu[1] = 1/T*(sum(y[2:T]) + yhat0 - psi0*sum(y))
-lambda[1] = b0/(a0*T)
-hatpsi[1] = optimise(psi.k, c(0, 1), y, a0, b0, yhat0, barmu0, p1, p2, maximum=TRUE)$maximum
-gamma[1] = -1/(psi.hessian(psi0, y, a0, b0, p1, p2))
-a = (T-2)/2
-b[1] = 1/2*(sum(y[2:T]^2-2*y[2:T]*(barmu0+psi0*y[1:(T-1)])) + T*(barmu0^2+lambda0) +
-            sum((psi0^2+gamma0)*y[1:T]^2+2*barmu0*psi0*y[1:T]) + yhat0^2+d0-2*yhat0*(barmu0+psi0*y[T]))
-yhat[1] = barmu0 + psi0*y[T]
-d[1] = b0/a0
-
-#Loop
-for(i in 2:N){
-  barmu[i] = 1/T*(sum(y[2:T]) + yhat[(i-1)] - hatpsi[(i-1)]*sum(y))
-  lambda[i] = b[(i-1)]/(a*T)
-  hatpsi[i] = optimise(psi.k, c(0, 1), y, a, b[(i-1)], yhat[(i-1)], barmu[(i-1)], p1, p2, maximum=TRUE)$maximum
-  gamma[i] = -1/(psi.hessian(hatpsi[(i-1)], y, a, b[(i-1)], p1, p2))
-  b[i] = 1/2*(sum(y[2:T]^2-2*y[2:T]*(barmu[(i-1)]+hatpsi[(i-1)]*y[1:(T-1)])) + T*(barmu[(i-1)]^2+lambda[(i-1)]) +
-                sum((hatpsi[(i-1)]^2+gamma[(i-1)])*y[1:T]^2+2*barmu[(i-1)]*hatpsi[(i-1)]*y[1:T]) +
-                yhat[(i-1)]^2+d[(i-1)]-2*yhat[(i-1)]*(barmu[(i-1)]+hatpsi[(i-1)]*y[T]))
-  yhat[i] = barmu[(i-1)] + hatpsi[(i-1)]*y[T]
-  d[i] = b[(i-1)]/a
-}
-support <- seq(-1, 8, 0.005)
-dens <- dnorm(support, yhat[N], sqrt(d[N]))
-@
-
-A Metropolis Hastings MCMC scheme was set up using a random walk candidate distribution for $\psi$ to draw from $p(\mu | y_{1:T}), p(\sigma^2 | y_{1:T})$ and $p(\psi | y_{1:T})$.
-These draws were then used to find the Rao-Blackwellised distribution of $p(y_{T+1} | y_{1:T})$.
-The chain had 1000 burn in draws followed by 50000 kept draws and took 6.736 seconds to complete the chain and forecast distribution.
-<<mcmc, echo=FALSE>>=
-##MCMC
-B <- 1000
-K <- 50000
-chain <- matrix(0, ncol=3, nrow=(B+K))
-mu0 = 1
-sig0 = 1
-psi0 = 0.8
-
-psi.kmcmc <- function(psi, mu, sigma2, y, p1, p2) {
-  out = -1/(2*sigma2)*sum((y[2:T]-(mu+psi*y[1:(T-1)]))^2) + (p1-1)*log((1+psi)/2) + (p2-1)*log((1-psi)/2)
-  return(exp(out))
-}
-
-#Starting iteration
-mubar = sum(y[2:T]-chain[1,2]*(y[1:(T-1)]))/T
-chain[1,1] = rnorm(1, mubar, sqrt(sig0/T))
-old = psi0
-candidate = rtruncnorm(1, -1, 1, old, 0.15)
-alpha = psi.kmcmc(candidate, chain[1,1], sig0, y, p1, p2)/dtruncnorm(candidate, -1, 1, old, 0.15)/
-        (psi.kmcmc(old, chain[1,1], sig0, y, p1, p2)/dtruncnorm(old, -1, 1, candidate, 0.15))
-alpha = min(alpha, 1)
-test = runif(1)
-if(alpha > test) {
-  chain[1,2] = candidate
-} else {
-  chain[1,2] = old
-}
-
-chain[1,3] = 1/rgamma(1, shape=(T-1)/2, rate=sum((y[2:T]-(chain[1,1]+chain[1,2]*y[1:(T-1)]))^2)/2)
-
-#Loop
-for(i in 2:(B+K)) {
-  mubar = sum(y[2:T]-chain[(i-1),2]*(y[1:(T-1)]))/T
-  chain[i,1] = rnorm(1, mubar, sqrt(chain[(i-1),3]/T))
-  old = chain[(i-1),2]
-  candidate = rtruncnorm(1, -1, 1, old, 0.15)
-  alpha = psi.kmcmc(candidate, chain[1,1], sig0, y, p1, p2)/dtruncnorm(candidate, -1, 1, old, 0.15)/
-         (psi.kmcmc(old, chain[1,1], sig0, y, p1, p2)/dtruncnorm(old, -1, 1, candidate, 0.15))
-  alpha = min(alpha, 1)
-  test = runif(1)
-  if(alpha > test) {
-    chain[i,2] = candidate
-  } else {
-    chain[i,2] = old
-  }
-  chain[i,3] = 1/rgamma(1, shape=(T-1)/2, rate=sum((y[2:T]-(chain[i,1]+chain[i,2]*y[1:(T-1)]))^2)/2)
-}
-
-ypred = data.frame(support = seq(-1, 8, 0.005))
-ypred$density = rao(chain[1001:51000,], ypred$support, 50000, y[T])
-@
 \begin{figure}[hb]
 \centering
-<<plots, echo=FALSE, fig=TRUE>>=
-densities <- cbind(q = dens, p = ypred$density)
-ggplot() + geom_line(data=data.frame(sup=support, dens=dens), aes(x=sup, y=dens), colour="blue") + theme_bw() + 
-  geom_line(data=data.frame(s = support, d = ypred$density), aes(x=s, y=d), colour="red") + labs(x="Support", y="Density")
-@
+\includegraphics{ar1_predictive-plots}
 \caption{Variational approximation for $q(y_{T+1}|y_{1:T})$ (blue) and the MCMC density for $p(y_{T+1}|y_{1:T})$ (red). 
-The KL Divergence from q to p is 0.0049, and from p to q is 0.0056.}
+The KL Divergence from q to p is 0.015, and from p to q is 0.018.}
 \label{densities}

### 2016-08-03 written report/gradient_ascent.Rnw
First

\documentclass[12pt]{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}

\graphicspath{{images/}}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\begin{document}
\SweaveOpts{concordance=TRUE}

\section{Gradient Ascent}

Say we want to maximise an arbitary function $F(x, y)$, where the partial derivatives $\delta(F(x, y)) / \delta x$ and $\delta(F(x, y)) / \delta y$ can not be found analytically. Then gradient ascent is a simple search algorithm that can be used to find a local maximum.

Gradient ascent has a simple mechanism, start at an arbitary point, calculate numerical partial derivatives for each variable then step in the direction of that derivative. Each step is
$$\left[ \begin{array}{c}
x^{(t)} \\
y^{(t)} \end{array} \right] 
= 
  \left[ \begin{array}{c}
           x^{(t-1)} \\
           y^{(t-1)}  \end{array} \right] 
+ p_t
  \left[ \begin{array}{c}
         \frac{\delta F(x, y)}{\delta x} \vline_{(x, y) = (x^{(t-1)}, y^{(t-1)})} \\
         \frac{\delta F(x, y)}{\delta y} \vline_{(x, y) = (x^{(t-1)}, y^{(t-1)})} \end{array} \right]
$$

and the algorithm is guaranteed to converge to a local minimum with

\begin{eqnarray*}
\sum_{t=1}^{\infty} p_t & = & \infty \\
\sum_{t=1}^{\infty} p_t^2 & < & \infty.
\end{eqnarray*}

$p_t = t^{-a}, 0.5 < a \leq 1$ satisfies this condition.

Intuitively, if the function is increasing, the algorithm steps forwards, and if it is decreasing the algorithm steps backwards. Smaller and smaller steps allow it to zig zag between either side of the maximum as it converges.

\section{Stochastic Gradient Ascent}

In the derivation for variational bayes, we had
$$\ln(p(y)) = KL [q(\theta) || p(\theta | y)] + \mathcal{L}(q, y),$$

where the Evidence Lower Bound (ELBO) is
\begin{equation}
\label{ELBO}
\mathcal{L}(q, y) = \int_{\theta} q(\theta) \ln \left( \frac{p(y, \theta)}{q(\theta)} \right) d\theta.
\end{equation}

As $KL[q(\theta) || p(\theta | y)]$ is non-negative, maximising $\mathcal{L}(q, y)$ with respect to a distribution  $q(\theta)$ is equivalent to minimising the KL Divergence between $p(\theta | y)$ and $q(\theta)$.
If $q(\theta)$ factorises into $\prod_{i} q(\theta_i)$, and the likelihood and prior form exponential family conjugate pairs, then variational bayes maximises $\mathcal{L}(q, y)$ with the solutions

$$q(\theta_{i}) \propto \exp(E_{\setminus i} (ln(p(y,\theta)))) \mbox{ for all } i.$$

However, (\ref{ELBO}) makes no assumption about the form of $q(\theta)$, and gradient ascent can be used to find the maximum of $\mathcal{L}(q, y)$ for a much more flexible choice of $q$.

With $\lambda$ as the vector of hyperparameters in the approximate distribution $q(\theta | \lambda)$, we require the partial derivative of $\mathcal{L}(q, y)$ with respect to $\lambda$. Paisley, Blei and Jordan (2012) introduce the stochastic gradient ascent approach, assuming the conditions to exchange the order of differentiation and integration hold.

\begin{eqnarray}
\nabla_{\lambda} \mathcal{L} & = & \nabla_{\lambda} \int_{\theta} q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
& = &  \int_{\theta} \nabla_{\lambda}[  q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)] d\theta \nonumber \\
& = &  \int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta)) - \ln(q(\theta | \lambda))] d\theta + \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
& = & -  \int_{\theta} \nabla_{\lambda}  [\ln(q(\theta | \lambda)] q(\theta | \lambda) d\theta +  \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
& = &-  \int_{\theta} \frac{\nabla_{\lambda}q(\theta | \lambda)}{q(\theta | \lambda)} q(\theta | \lambda) d\theta +   \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
& = & -  \nabla_{\lambda}\int_{\theta} q(\theta | \lambda) d\theta +  \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
& = &  \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber
\end{eqnarray}

This expression can be further simplified by the identity $\nabla_{\lambda} [q(\theta | \lambda)] =  \nabla_{\lambda} [\ln(q(\theta | \lambda))]q(\theta | \lambda)$.

\begin{eqnarray}
\label{gradient}
\nabla_{\lambda} \mathcal{L} & = & \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
& = &   \int_{\theta} \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)q(\theta | \lambda) d\theta \nonumber \\
& = & E_q [  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)].
\end{eqnarray}

We can compute a noisy but unbiased estimate of (\ref{gradient}) by

\begin{equation}
\label{mc}
\nabla_{\lambda}\mathcal{L} \approx \frac{1}{N}\sum_{n=1}^{N} \nabla_{\lambda} [\ln(q(\theta_n | \lambda))] (\ln (p(y, \theta_n)) - \ln(q(\theta_n | \lambda)))
\end{equation}

with $\theta_n \sim q(\theta | \lambda)$.

To use stochastic gradient ascent to maximise $\mathcal{L}$ we just need to be able to evaluate the log-joint density and each log-approximate density, as well as calculate partial derivatives of each log-approximate density within a Monte-Carlo approximation.

\section{Variance Reduction}

The Monte-Carlo approximation is unbiased but can have a large variance if a small amount of draws is chosen. We can reduce the variance by instead sampling another variable with the same expected value but a reduced variance.

For the scalar $\theta$ case, consider the function

\begin{equation}
\label{control}
\hat{f}(\theta) = f(\theta) - a(g(\theta)-E[g(\theta)])
\end{equation}

Paisley, Blei and Jordan (2012) call the function $g$ a control variate. Note that
$$E[\hat{f}(\theta)] = E[f(\theta)]$$
and 
\begin{equation}
\label{var}
Var(\hat{f}) = Var(f) + a^2 Var(g) - 2a Cov(f, g)
\end{equation}

The variance of $\hat{f}$ is minimised at

$$a^* = \frac{Cov(f, g)}{Var(g)}$$

A similar approach holds for a vector $\theta$.

Ranganath, Gerrish and Blei (2014) suggest the following function $g$

$$g(\theta) = \nabla_{\lambda} [\ln(q(\theta_n | \lambda))]$$

as it has a high covariance with (\ref{gradient}) and thus efficiently reduces variance in (\ref{var}).

If the joint and approximating distributions factorise, a Rao-Blackwellisation approach can be used to ignore the parts of the function that do not depend on the $\theta_i$ of interest in that step of the algorithm and further reduce variance. More details are in Ranganath, Gerrish and Blei (2014).

\section{Copula Variational Inference}

Using stochastic gradient ascent, we are no longer restricted to the mean-field assumption $q(\theta) = \prod_i q(\theta_i)$, and Tran, Blei and Airoldi (2015) suggest a copula model,

\begin{equation}
\label{copula}
q(\theta | \lambda, \eta) = \prod_{i} q(\theta_i|\lambda) c(Q(\theta_1|\lambda), \dots , Q(\theta_D|\lambda) |\eta) \mbox{ for } i = 1, \dots D.
\end{equation}

The copula function is factorised into a sequence of pairs $c(Q(\theta_i |\lambda), Q(\theta_j | \lambda) | \eta)$ using a vine copula and a stochastic gradient ascent algorithm alternates between maximising $\mathcal{L}$ with respect to the variational parameters $\lambda$ and the copula parameters $\eta$. 
The algorithm is not particularly fast to converge but allows and extremely flexible approximation with strong predictive abilities. If data was streaming, and the model was just incrementing $q(\theta_t)$ to $q(\theta_{t+1})$ repeatedly it may be faster.

\end{document}

### 2016-08-03 written report/variationalwriteup.Rnw
First

\documentclass{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\usepackage{hyperref}
\begin{document}
\SweaveOpts{concordance=TRUE}

We have the model

$$y_{t} = X^{'}_{t}\beta + \epsilon_{t},$$

where $Y$ is a $T \times 1$ random vector, $X$ is a $k \times T$ matrix of covariates, $\beta$ is a $k \times 1$ matrix of coefficients and $\epsilon_{t}$ is iid N$(0,\sigma^2).$

We are interested in the posterior density

$$p(\sigma^2, \beta | y) = \frac{p(y | \sigma^2, \beta)p(\sigma^2, \beta)}{p(y)}$$

Taking $p(\sigma^2, \beta) \propto \frac{1}{\sigma^2}$,

\begin{eqnarray}
\label{joint}
p(\sigma^2, \beta | y) & \propto & \prod_{t=1}^{T} \frac{1}{\sigma} exp \left\{\frac{-1}{2 \sigma^2} (y_{t}-X^{'}_{t}\beta)^2 \right\} \frac{1}{\sigma^2} \nonumber \\
& = & \sigma^{-(T+2)} exp \left\{ \frac{-1}{2\sigma^2} \sum_{t=1}^{T} (y_{t}-X^{'}_{t}\beta)^2 \right\} \nonumber \\
& = & \sigma^{-(T+2)} exp \left\{ \frac{-1}{2\sigma^2}(T-k)\hat{\sigma}^2 \right\} exp \left\{ \frac{-1}{2\sigma^2} (\beta-\hat{\beta})'X'X(\beta-\hat{\beta})\right\} 
\end{eqnarray}

Where $\hat{\beta} = (X'X)^{-1}X'Y$, the OLS estimator of $\beta$ and $\hat{\sigma}^2 = (y-X\hat{\beta})'(Y-X\hat{\beta})/(T-k)$ is an unbiases estimator of $\sigma^2$.

We approximate $p(\beta, \sigma^2 | y)$ with $Q(\beta, \sigma^2) = Q(\beta)Q(\sigma^2)$, the factorisable distribution that minimises the Kullback - Leibler divergence between $p(\beta, \sigma^2 | y)$ and $Q(\beta, \sigma^2)$. It can be shown that the distributions $Q(\beta)$ and $Q(\sigma^2)$ that minimise this are:

\begin{equation}
\label{Qb}
ln(Q(\beta)) = E_{\sigma^2}\left[ ln(y, \sigma^2, \beta))\right] + C
\end{equation}

and

\begin{equation}
\label{Qs}
ln(Q(\sigma^2)) = E_{\beta}\left[ ln(p(y, \sigma^2, \beta))\right] + C
\end{equation}

Substituting (\ref{joint}) into (\ref{Qb}) and ignoring the terms that do not depend on $\beta$, we get

$$ln(Q(\beta)) = E_{\sigma^2}\left[ \frac{-1}{2\sigma^2} (\beta-\hat{\beta})'X'X(\beta-\hat{\beta}) \right] + C$$.

Recognising the kernel of a normal distribution, we see that

$$Q(\beta) \sim N(\mu = \hat{\beta}, \lambda = (X'X)^{-1} E[\sigma^{-2}]^{-1})$$

Similarly substituting (\ref{joint}) into (\ref{Qs}) results in

\begin{eqnarray}
ln(Q(\sigma^2)) & = &  -(T+2)ln(\sigma) - \frac{1}{2\sigma^2} \left( (T-k)\hat{\sigma}^2 + E_{\beta}[(\beta-\hat{\beta})'X'X(\beta-\hat{\beta})] \right) + C \nonumber \\
& = & -(T+2)ln(\sigma) - \frac{1}{2\sigma^2} \left( (T-k)\hat{\sigma}^2 + tr(X'X Var(\beta)) \right) + C \nonumber
\end{eqnarray}

Recoginising the kernel of an inverse gamma distribution, we see that

$$Q(\sigma^2) \sim Inv.Gamma ( shape = (T+1)/2 = a, scale = \frac{(T-k)\hat{\sigma}^2 + tr(X'X \lambda)}{2} = b)$$

Therefore $1/\sigma^2 \sim Gamma(shape = a, rate = b)$ so $E[\sigma^{-2}] = a/b$, and substituting this into $\lambda$ gives us:

$$\lambda = (X'X)^{-1} b/a $$

and so
\begin{eqnarray}
b & = & \frac{(T-k)\hat{\sigma}^2 + tr(X'X (X'X)^{-1})b/a}{2} \nonumber \\
& = & \frac{(T-k)\hat{\sigma}^2}{2} + \frac{kb}{T+1} \nonumber \\
& = & \frac{(T-k)\hat{\sigma}^2}{2(1-k/(T+1))} \nonumber
\end{eqnarray}

Figure \ref{ex} shows the results of variational bayes compared to the exact posterior for the linear regression: $$y_{i} = b_{1} x_{1i} + b_{2} x_{2i} + \epsilon_{i}$$

\begin{figure}[htbp]
  \includegraphics[height=8cm,keepaspectratio]{varbayes.pdf}
  \caption{Variational Bayes compared to the exact posterior. It took three iterations for the parameters to converge.}
  \label{ex}
\end{figure}

\end{document}

### 2016-08-03 AR1C.R

@@ -1,0 +1,88 @@
+##AR(1) Model
+library(Rcpp)
+library(tidyr)
+library(ggplot2)
+library(dplyr)
+library(RcppArmadillo)
+library(microbenchmark)
+library(MASS)
+library(parallel)
+sourceCpp("AR1c.cpp")
+
+T <- 50
+J <- 10
+sigma <- 2
+mu <- 2
+rho <- 0.5
+
+y <- genAR1(T+J, c(mu, rho, sigma))
+
+
+X <- as.matrix(cbind(rep(1,59), y[1:59]))
+InvXX <- solve(t(X)%*%X)
+b <- InvXX%*%t(X)%*%y[2:60]                   
+v <- T+J-3
+
+#exact posterior
+sigsqexact <- t(y[2:60]-X%*%b)%*%(y[2:60]-X%*%b)/(T+J-3)
+sigmaexact <- 1/sqrt(gen_gamma(2000, v/2, 1/(sigsqexact*v/2)))
+sigtrans <- transformsigma(sigmaexact, InvXX)
+betaexact <- t(apply(sigtrans, 3, gen_mvrnorm, n=1, mu=b))
+exact <- data.frame(mu = betaexact[,1], psi = betaexact[,2], sigma = sigmaexact, method="Exact")
+
+
+
+#The ABC Algorithm
+
+index.bottom.N = function(x, N){
+  o = order(x, na.last=FALSE, decreasing=TRUE)
+  o.length = length(o)
+  o[((o.length-N+1):o.length)]
+}
+
+abcpar <- function(theta, cores, n, T, yss) { ##Parallelisation returns a 20% performance boost, 4 cores seems optimal
+  theta.list <- list()
+  n <- nrow(theta)
+  for(i in 1:cores) { #split theta into a list, one for each core to reduce parallelisation overhead
+    theta.list[[i]] <- theta[((i-1)*(n/cores)+1):(i*(n/cores)),]
+  }
+  distl <- unlist(mclapply(theta.list, ABC_Loop, n/cores, T, yss=yss, mc.cores=cores)) #run the mclapply on each element on list
+  return(distl)
+} 
+#ABC_Loop function does all the work
+
+
+yss <- sum_stats(y[51:60]) #Calculate summarry statistics of the updated data
+InvXXprior <- solve(t(X[1:49,])%*%X[1:49,]) #run through exact inference on the original data to get a prior
[…]

### 2016-08-03 AR1c.cpp

@@ -1,0 +1,116 @@
+// [[Rcpp::depends(RcppArmadillo)]]
+// [[Rcpp::plugins(cpp11)]]
+
+#include <RcppArmadillo.h>
+#include <math.h>
+#include <algorithm>
+#include <queue>
+#include <vector>
+#include <functional>   
+using namespace Rcpp;
+using namespace arma;
+using namespace std;
+
+
+// [[Rcpp::export]]
+vec genAR1(int n, rowvec theta) {
+  double mu = theta[0];
+  double rho = theta[1];
+  double sigma = theta[2];
+  vec eps(n);
+  eps = randn<vec>(n);
+  vec out(n);
+  out[0] = mu + rho*(0-mu) + sigma*eps[0];
+  for(int i = 1; i < n; ++i){
+    out[i] = mu + rho*(out[i-1]-mu) + sigma*eps[i];
+  }
+  return out;
+}
+
+// [[Rcpp::export]]
+vec gen_gamma(int n, int a, double scale) {
+  //NumericVector unif;
+  vec out(n);
+  for(int i = 0; i < n; ++i){
+    vec unif = randu<vec>(a);
+    out[i] = -scale * sum(log(unif));
+  }
+  return out;
+}
+
+// [[Rcpp::export]]
+arma::mat gen_mvrnorm(int n, arma::vec mu, arma::mat sigma) {
+  int ncols = sigma.n_cols;
+  arma::mat Y = arma::randn(n, ncols);
+  return arma::repmat(mu, 1, n).t() + Y * arma::chol(sigma);
+}
+
+// [[Rcpp::export]]
+double autocor(vec x, int lag) {
+  int T = x.size();  
+  double out;
+  for(int i = 0; i < T-lag; ++i) {
+    out += x[i+lag]*x[i];
+  }
+  return out/T;
+}
+
+// [[Rcpp::export]]
+vec sum_stats(vec x) { //Take three summary statistics of a vector
+  vec out(3);
+  out[0] = mean(x);
+  out[1] = var(x);
+  out[2] = autocor(x, 1);
+  return out;
+}
+
+// [[Rcpp::export]]
+cube transformsigma(vec x, mat Inv) { //Multiply each element of the Inverse matrix by each element of x
+  int n = x.size();
+  cube out(2,2,n);
+  for(int i = 0; i < n; ++i){
+    out(0,0,i) = Inv(0,0)*x[i];
+    out(0,1,i) = Inv(0,1)*x[i];
[…]

### 2016-08-03 generic regression variational.R

@@ -1,0 +1,46 @@
+##Generic regression
+library(tidyr)
+library(ggplot2)
+library(dplyr)
+library(MASS)
+
+T <- 250
+sigma <- 0.75
+b1 <- 1.3
+b2 <- 2.5
+eps <- rnorm(T)
+x1 <- rnorm(T)
+x2 <- rnorm(T)
+y <- b1*x1+b2*x2+sigma*eps
+
+X <- cbind(x1, x2)
+XX <- t(X)%*%X
+b <- solve(XX)%*%t(X)%*%y
+
+#Variational
+
+k <- 2
+sigsqhat <- t(y-X%*%b)%*%(y-X%*%b)/(T-k)
+
+scale <- rep(0, 25)
+scale[1] <- (T-k)*sigsqhat/2
+for(i in 2:25){
+  scale[i] <- (T-k)*sigsqhat/2 + scale[(i-1)]/T
+}
+betavar2 <- mvrnorm(2000, b, solve(t(X)%*%X)*scale[25]*2/T)
+sigmavar2 <- 1/(rgamma(2000, T/2, scale[25]))
+variational <- data.frame(b1 = betavar2[,1], b2 = betavar2[,2], sigma.squared = sigmavar2, method="Variational")
+
+#Exact
+
+sigmaexact <- 1/rgamma(2000, (T-k)/2, sigsqhat*(T-k)/2)
+sigtransfex <- array(sapply(sigmaexact, function(x) x*solve(XX)), dim=c(2,2,2000))
+betaexact <- t(apply(sigtransfex, 3, mvrnorm, n=1, mu=b))
+
+exact <- data.frame(b1 = betaexact[,1], b2 = betaexact[,2], sigma.squared = sigmaexact, method="Exact")
+
+combined <- rbind(exact, variational)
+combl <- gather(combined, theta, density, -method)
+ggplot(data=combl, aes(x=density, colour=method)) + geom_density() + facet_wrap(~theta, scales="free")
+
+

### 2016-08-03 heirarchical.R

@@ -1,0 +1,87 @@
+library(tidyr)
+library(coda)
+library(Rcpp)
+library(RcppArmadillo)
+library(microbenchmark)
+library(dplyr)
+library(ggplot2)
+sourceCpp("heirarchical.cpp")
+
+#Drawing data
+t.mubar = 3.5
+t.tau2 = 1
+J = 100
+N = 1000
+t.muj = rnorm(J, t.mubar, sqrt(t.tau2))
+y = sapply(t.muj, rnorm, n = N, sd = 1)
+ybar = apply(y, 2, mean)
+
+#Hyperpriors
+mubarbar = 2
+lambda = 10
+a = 0.0001
+b = 0.0001
+
+#Gibbs setup
+M = 5000 #retained
+B = 5000 #burn in
+draws = matrix(0, ncol=J+2, nrow=M+B) #1-100: muj, 101 = mubar, 102 = tau2
+#Initials
+draws[1,] = 1
+
+#Gibbs Chain
+draws = gibbschain(y, M, B, mubarbar, lambda, a, b)
+results = mcmc(draws[(B+1):(B+M),])
+qplot(draws[(B+1):(M+B), 15], draws[(B+1):(M+B), 102]) + labs(x="Mu bar", y="Tau squared")
+
+#Variational setup
+n = 10
+cv = varfit(y, n, mubarbar, lambda, a, b)
+colnames(cv) = c(1:100, "lstar", "mubar", "lambdabar", "alpha", "beta")
+cv = as.data.frame(cv)
+
+#Plotting results
+support = seq(2.9, 4.1, 0.01)
+vdens = dnorm(support, cv$mubar[n], sqrt(cv$lambdabar[n]))
+mcdens = dnorm(support, mean(draws[(B+1):(M+B), 101]), sd(draws[(B+1):(M+B), 101]))
+p.mubar = data.frame(variational = vdens, mcmc = mcdens, support = support)
+p.mubar = gather(p.mubar, key=method, value=density, -support)
+ggplot(data=p.mubar, aes(y=density, x=support, colour=method))+geom_line()
+
+
+#Draw new data
+muj.new = rnorm(10000, t.mubar, sqrt(t.tau2))
+ynew = sapply(muj.new, rnorm, n = 1, sd = 1)
+
+#Calculate log score as an expectation over the posterior
+samp = sample((B+1):(M+B), 1000)
+theta.s = draws[samp, 101:102]
+
+#To use apply with a vector argument
[…]

### 2016-08-03 kalman.R

@@ -1,0 +1,7 @@
+library(FKF)
+library(gridExtra)
+library(Rcpp)
+library(RcppArmadillo)
+library(microbenchmark)
+library(ggplot2)
+sourceCpp("kalman.cpp")

### 2016-08-03 mcmc.cpp

@@ -1,0 +1,97 @@
+// [[Rcpp::depends(RcppArmadillo)]]
+// [[Rcpp::plugins(cpp11)]]
+
+#include <RcppArmadillo.h>
+#include <math.h>
+#include <algorithm>
+#include <queue>
+#include <vector>
+#include <functional>   
+#include <cmath>
+using namespace Rcpp;
+using namespace arma;
+using namespace std;
+
+// [[Rcpp::export]]
+double logpkernelfC(vec y_in, double mu_in, double sig_in){
+  int n = y_in.size();
+  double logkernel = (-51)*log(sig_in);
+  vec terms(n);
+  for(int i = 0; i < n; ++i){
+    terms[i] = -0.5*(6)*log(1+(pow((y_in[i]-mu_in)/sig_in, 2.0)/3));
+  }
+  return(logkernel + sum(terms));
+}
+
+// [[Rcpp::export]]
+double IGfC(double sigin, double vin, double sigsqhatin) {
+  double out1 = 2/tgamma(vin/2);
+  double out2 = pow(vin*sigsqhatin/2, vin/2);
+  double out3 = exp(-vin*sigsqhatin/(2*pow(sigin, 2.0)));
+  double out4 = pow(sigin, vin+1);
+  return(out1*out2*out3/out4);
+}
+
+
+// [[Rcpp::export]]
+double mu_MHfC (double mu_last, double sig_in, vec y_in) {
+  double ybar = mean(y_in);
+  double sy = sqrt(var(y_in));
+  double mu_c = ybar + sy/sqrt(50)*randn<vec>(1)[1];
+  double logalpha;
+  logalpha = logpkernelfC(y_in, mu_c, sig_in);
+  logalpha -= logpkernelfC(y_in, mu_last, sig_in);
+  logalpha -= log(R::dnorm(mu_c, ybar, sy/sqrt(50), FALSE));
+  logalpha += log(R::dnorm(mu_last, ybar, sy/sqrt(50), FALSE));
+  double alpha = min(1.0, exp(logalpha));
+  double u = randu<vec>(1)[0];
+    if(u <= alpha)
+      return mu_c;
+    else
+      return mu_last;
+}
+
+// [[Rcpp::export]]
+vec gen_gamma(int n, int a, double scale) {
+  //NumericVector unif;
+  vec out(n);
[…]

### 2016-08-03 mcmcR.r

@@ -1,0 +1,92 @@
+library(gridExtra)
+library(Rcpp)
+library(RcppArmadillo)
+library(microbenchmark)
+library(ggplot2)
+sourceCpp("mcmc.cpp")
+
+#Generate Data
+set.seed(753115)
+psi <- 5
+tmu <- 5
+tsig <- 1
+y <- tmu + tsig*sqrt((psi-2)/psi)*rt(50, psi)
+y <- as.data.frame(y)
+ggplot(data=y, aes(x=y, y=..density..)) + geom_histogram(binwidth=0.5)
+
+logpkernel.f = function(y_in, mu_in, sig_in){
+  logkernel = -(50+1)*log(sig_in)
+  terms = -0.5*(5+1)*log(1+(((y_in - mu_in)/sig_in)^2)/(5-2))
+  out = logkernel+sum(terms)
+  return(out)
+}
+
+#Alg 1 stuff
+IG.f = function(sigin, vin, sigsqhatin){
+  out = (2/gamma(vin/2))*((vin*sigsqhatin/2)^(vin/2))*exp(-vin*sigsqhatin/(2*sigin^2))/(sigin^(vin+1))
+  return(out)
+}
+
+mu_MH.f = function(mu_last, sig_in, y_in){
+  ybar = mean(y_in)
+  sy = sd(y_in) 
+  mu_c = rnorm(1,mean=ybar,sd=(sy/sqrt(50))) 
+  logalpha = logpkernel.f(y_in, mu_c, sig_in) 
+  logalpha = logalpha - logpkernel.f(y_in, mu_last, sig_in)
+  logalpha = logalpha - log(dnorm(mu_c, mean = ybar, sd = (sy/sqrt(50)))) 
+  logalpha = logalpha + log(dnorm(mu_last, mean = ybar, sd = (sy/sqrt(50))))
+  alpha = min(1, exp(logalpha))
+  if(alpha==1){
+    out = mu_c
+  } else if (runif(1,0,1)<=alpha){
+    out = mu_c
+  } else {
+    out = mu_last
+  }
+  return(out)
+}
+
+sig_MH.f = function(mu_in, sig_last, y_in){
+  sighatsq = sum((y_in - mu_in)^2)/50
+  sig_shape = 25
+  sig_scale = 50*sighatsq/2
+  sig_c = 1/sqrt(rgamma(1,shape=sig_shape,rate = sig_scale)) 
[…]

### 2016-09-22 AR2again.Rnw
updates

\documentclass{article}

\begin{document}
\SweaveOpts{concordance=TRUE}
%\SweaveOpts{concordance=TRUE}

\section{The AR2 Model}

Consider a mean zero AR(2) model conditioned on $y_1$ and $y_2$.

$$y_t = \phi_1 * y_{t-1} + \phi_2 * y_{t-2} + \epsilon_t \mbox{ for } t = 3,\dots,T$$,

where $\epsilon$ are iid $\mathcal{N}(0, \sigma^2)$.

Using the independent priors:

\begin{eqnarray*}
p(\phi_1) & \sim & \mathcal{N}(\bar{phi_1}, \tau_{1}^2) \\
p(\phi_2) & \sim & \mathcal{N}(\bar{phi_2}, \tau_{2}^2) \\
p(\sigma^2) & \sim & \mathcal{IG} (\mbox{shape} = \alpha, \mbox{ rate} = \beta),
\end{eqnarray*}
we have a model with zero stationary restrictions that can be easily sampled through Gibbs MCMC or approximated by a range of variational techniques.

There are five approaches considered:

\begin{itemize}
\item Gibbs MCMC
\item Mean field Variational Bayes
\item Stochastic Variational Bayes
\item Stochastic Variational Bayes with Control Variates
\item Re-parameterised Stochastic Variational Bayes
\end{itemize}

Simulation proceeds with $T = 100$, $\phi_1 = 0.7$, $\phi_2 = 0.15$ and $\sigma^2 = 2$. Each approach estimates the same approximating distribution, $q(\theta) = BVN((\phi_1, \phi_2)' | \mu, \Sigma) IG(\sigma^2 | \alpha, \beta)$ with estimated parameters:

\begin{itemize}
\item $\mu_1$: The mean of $\phi_1$
\item $\mu_2$: The mean of $\phi_2$
\item $L_{11}, L_{21}, L_{22}$, the three components of the lower triangular decomposition of $\Sigma$
\item $\alpha, \beta$, the shape and rate parameters of $\sigma^2$.
\end{itemize}

The diagonal of $L$ is not forced to be positive so the decomposition is not invariant to sign changes.

Collectively, we have $\lambda = (\mu_1, \mu_2, L_{11}, L_{21}, L_{22}, \alpha, \beta)'$.

\begin{figure}
<<MCMC>>=
setwd("/home/nltom2/Desktop/Confirmation")
library(ggplot2)
library(GGally)
thin <- read.csv("AR2thin.csv")[,-1]
ggpairs(thin)
@
\caption{Result of MCMC with 120,000 draws. The first 20\% were discarded then the remaining draws were thinned by a factor of 20, which was required to remove any dependence. The remaining effective sample size for each parameter is 5000.
\end{figure}

<<summary>>=
library(knitr)
library(mvtnorm)
library(reshape)
summary = matrix( c(0.78, 0.16, 0.10, -0.09, 0.05, 48.9, 77.3, -76.7, 2.6,
                    0.83, 0.10, 0.04, 0, 0.05, 50, 79.7, -77.7, "1.3 x 10^{-6}", 
                    0.63, 0.27, 0.09, -0.12, 0.05, 45.6, 81.4, -78.3, 51.1, 
                    0.78, 0.15, 0.1, -0.09, 0.05, 49.3, 78, -76.7, 4.0,
                    0.77, 0.15, 0.09, -0.09, 0.05, 48.5, 77.6, -76.9, 1.3) , byrow = TRUE, ncol = 9)

colnames(summary) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)*")
rownames(summary) = c("Method of Moments", "Mean Field", "SVB", "SVB CV", "SVB RP")
kable(summary)
@
The original stochastic variational bayes algorithm converged to a different local maximum, possibly due to the high variance in its gradient estimates.

Note that all of the variational algorithms are currently unparallelised, and the stochastic versions in particular have a huge potential for faster performance.

\end{document}

### 2016-09-22 AR2again.tex
updates

--- 2016-09-22 AR2again.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{AR2again-concordance}
 %\SweaveOpts{concordance=TRUE}
@@
 \begin{figure}
-<<MCMC>>=
-setwd("/home/nltom2/Desktop/Confirmation")
-library(ggplot2)
-library(GGally)
-thin <- read.csv("AR2thin.csv")[,-1]
-ggpairs(thin)
-@
+\begin{Schunk}
+\begin{Sinput}
+> setwd("/home/nltom2/Desktop/Confirmation")
+> library(ggplot2)
+> library(GGally)
+> thin <- read.csv("AR2thin.csv")[,-1]
+> ggpairs(thin)
+\end{Sinput}
+\end{Schunk}
 \caption{Result of MCMC with 120,000 draws.
@@
 
-<<summary>>=
-library(knitr)
-library(mvtnorm)
-library(reshape)
-summary = matrix( c(0.78, 0.16, 0.10, -0.09, 0.05, 48.9, 77.3, -76.7, 2.6,
-                    0.83, 0.10, 0.04, 0, 0.05, 50, 79.7, -77.7, "1.3 x 10^{-6}", 
-                    0.63, 0.27, 0.09, -0.12, 0.05, 45.6, 81.4, -78.3, 51.1, 
-                    0.78, 0.15, 0.1, -0.09, 0.05, 49.3, 78, -76.7, 4.0,
-                    0.77, 0.15, 0.09, -0.09, 0.05, 48.5, 77.6, -76.9, 1.3) , byrow = TRUE, ncol = 9)
-
-colnames(summary) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)*")
-rownames(summary) = c("Method of Moments", "Mean Field", "SVB", "SVB CV", "SVB RP")
-kable(summary)
-@
+\begin{Schunk}
+\begin{Sinput}
+> library(knitr)
+> library(mvtnorm)
+> library(reshape)
+> summary = matrix( c(0.78, 0.16, 0.10, -0.09, 0.05, 48.9, 77.3, -76.7, 2.6,
++                     0.83, 0.10, 0.04, 0, 0.05, 50, 79.7, -77.7, "1.3 x 10^{-6}", 
++                     0.63, 0.27, 0.09, -0.12, 0.05, 45.6, 81.4, -78.3, 51.1, 
++                     0.78, 0.15, 0.1, -0.09, 0.05, 49.3, 78, -76.7, 4.0,
++                     0.77, 0.15, 0.09, -0.09, 0.05, 48.5, 77.6, -76.9, 1.3) , byrow = TRUE, ncol = 9)
+> colnames(summary) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)*")
+> rownames(summary) = c("Method of Moments", "Mean Field", "SVB", "SVB CV", "SVB RP")
+> kable(summary)
+\end{Sinput}
+\begin{Soutput}
+|                  |mu1  |mu2  |L11  |L21   |L22  |alpha |beta |L(theta) |time (seconds)* |
+|:-----------------|:----|:----|:----|:-----|:----|:-----|:----|:--------|:---------------|
+|Method of Moments |0.78 |0.16 |0.1  |-0.09 |0.05 |48.9  |77.3 |-76.7    |2.6             |
+|Mean Field        |0.83 |0.1  |0.04 |0     |0.05 |50    |79.7 |-77.7    |1.3 x 10^{-6}   |
+|SVB               |0.63 |0.27 |0.09 |-0.12 |0.05 |45.6  |81.4 |-78.3    |51.1            |
+|SVB CV            |0.78 |0.15 |0.1  |-0.09 |0.05 |49.3  |78   |-76.7    |4               |
+|SVB RP            |0.77 |0.15 |0.09 |-0.09 |0.05 |48.5  |77.6 |-76.9    |1.3             |
+\end{Soutput}
+\end{Schunk}
 The original stochastic variational bayes algorithm converged to a different local maximum, possibly due to the high variance in its gradient estimates.

### 2016-09-22 written report/2levelmixture.tex
updates

@@
-\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
+\documentclass[12pt]{article}
 \usepackage{natbib}
@@
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
+\usepackage{Sweave}
 \begin{document}
+\input{2levelmixture-concordance}
 
@@
 The quantity $\tilde{\pi}_1 = \mathbb{E}_\pi \ln(\pi) = \psi(\alpha)-\psi(\alpha+\beta)$, and $\tilde{\pi}_2 = \mathbb{E}_\pi \ln(1-\pi) = \psi(\beta)-\psi(\alpha+\beta)$, where $\psi(\cdot)$ is the digamma function (Archambeau and Verleysen 2007).
-We also expanded the quadratic term and substitued in $(y_i^2 - \bar{\mu}_j)^2 + \lambda_j$ for $\mathbb{E} (y_i-\mu_j)^2$.
 
@@
 \begin{eqnarray*}
-a & = & \sum_{i=1}^{n} \frac{p_i}{p_i+q_i} + 1 \\
-b & = & \sum_{i=1}^{n} \frac{q_i}{p_i+q_i} + 1 \\
+\alpha & = & \sum_{i=1}^{n} \frac{p_i}{p_i+q_i} + 1 \\
+\beta & = & \sum_{i=1}^{n} \frac{q_i}{p_i+q_i} + 1 \\
 \bar{\mu}_1 & = &  \frac{\sum_{i=1}^{n} y_i p_i /(p_i+q_i)}{\sum_{i=1}^{n} p_i /(p_i+q_i)} \\
@@
 \lambda_2 & = & \left(\sum_{i=1}^{n} \frac{q_i}{p_i+q_i}\right)^{-1} \\
-p_i & = & \exp \left\{\frac{2(\psi(a)-\psi(a+b))-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} \right\} \\
-q_i & = & \exp \left\{\frac{2(\psi(b)-\psi(a+b))-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} \right\}
+p_i & = & \exp \left\{\frac{2(\psi(\alpha)-\psi(\alpha+\beta))-((y_i - \bar{\mu}_1)^2 + \lambda_1)}{2} \right\} \\
+q_i & = & \exp \left\{\frac{2(\psi(\beta)-\psi(\alpha+\beta))-((y_i - \bar{\mu}_2)^2 + \lambda_2)}{2} \right\}
 \end{eqnarray*}
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/vbayes-1}
-
-\end{knitrout}
 250 draws were simulated with parameters $\mu_1 = 3, \mu_2 = 6, \pi=0.6$ and th e variational algorithm was ran.
@@
 
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/vbayes2-1}
-
-\end{knitrout}
 250 draws were simulated with parameters $\mu_1 = 5.5, \mu_2 = 6, \pi=0.6$ to try and force an overlap in the data.

### 2016-09-22 written report/AR2.Rnw
updates

\documentclass{article}
\usepackage{amsmath}
\usepackage[]{algorithm2e}

\makeatletter
\def\BState{\State\hskip-\ALG@thistlm}
\makeatother

\begin{document}
\SweaveOpts{concordance=TRUE}
\section{The AR2 Model}

Consider a mean zero AR(2) model conditioned on $y_1$ and $y_2$.

$$y_t = \phi_1 * y_{t-1} + \phi_2 * y_{t-2} + \epsilon_t \mbox{ for } t = 3,\dots,T$$,

where $\epsilon$ are iid $\mathcal{N}(0, \sigma^2)$.

Using the independent priors:

\begin{eqnarray*}
p(\phi_1) & \sim & \mathcal{N}(\bar{phi_1}, \tau_{1}^2) \\
p(\phi_2) & \sim & \mathcal{N}(\bar{phi_2}, \tau_{2}^2) \\
p(\sigma^2) & \sim & \mathcal{IG} (\mbox{shape} = \alpha, \mbox{ rate} = \beta),
\end{eqnarray*}
we have a model with zero stationary restrictions that can be easily sampled through Gibbs MCMC or approximated by a range of variational techniques.

There are four approaches considered:

\begin{itemize}
\item MCMC
\item Mean field Variational Bayes
\item Copula Variational Bayes
\item A re-parameterised Copula Variational Bayes
\end{itemize}

\section{Summary}

Each approach uses the same model, $q(\theta) = BVN((\phi_1, \phi_2)' | \mu, \Sigma) IG(\sigma^2 | \alpha, \beta)$ with estimated parameters
\begin{itemize}
\item $\mu_1$: The mean of $\phi_1$
\item $\mu_2$: The mean of $\phi_2$
\item $L_{11}, L_{21}, L_{22}$, the three components of the lower triangular decomposition of $\Sigma$
\item $\alpha, \beta$, the shape and rate parameters of $\sigma^2$.
\end{itemize}

The diagonal of $L$ is not forced to be positive so the decomposition is not invariant to sign changes.

Collectively, we have $\lambda = (\mu_1, \mu_2, L_{11}, L_{21}, L_{22}, \alpha, \beta)'$.

<<summary, Echo = FALSE, Eval = TRUE>>=
library(knitr)
library(mvtnorm)
library(reshape)
summary = matrix( c(0.37, -0.30, 0.10, -0.03, 0.10, 49.2, 78.4, -77.5, 7.4,
                    0.37, -0.29, 0.09, 0, 0.10, 49.5, 79.8, -77.4, "1.3 x 10^{-6}", 
                    0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7, -77.2, 59.5, 
                    0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9, -77.4, 1.2) , byrow = TRUE, ncol = 9)

colnames(summary) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)*")
rownames(summary) = c("Method of Moments", "Mean Field", "Copula 1", "Copula 2")
kable(summary)
@
\*Note that all of the variational algorithms are currently unparallelised, and the copula versions in particular have a huge potential for faster performance.

\section{MCMC}

Taking fifty thousand draws and discarding the first ten thousand, the effective sample sizes for each parameter is:
\begin{itemize}
\item $\phi_1 : 4500$
\item $\phi_2 : 4500$
\item $\sigma^2: 35000$
\end{itemize}

Retaining only every tenth draws returns samples with effectively no auto-correlation. This thinned out sample is plotted below.

\begin{figure}
<<MCMC, Echo = FALSE>>=
setwd("/home/nltom2/Desktop/Code/Approximate-Updating")
library(ggplot2)
library(GGally)
thin <- read.csv("AR2thin.csv")[,-1]
ggpairs(thin)
@
\end{figure}

There is obvious eliptical dependence between $\phi_1$ and $\phi_2$, but not between either $\phi$ parameter and $\sigma^2$. The best fitting copula by BIC is gaussian, with the parameter $\rho = -0.29$.

<<copula, Echo = FALSE>>=

copest =data.frame(copula = c("gaussian", "t"), rho = c(-0.29, -0.29), df = c("NA", 112.5), BIC = c(-167.8, -161.2))
kable(copest)
@

For consistency with the variational approaches, the method of moments is used to fit a parametric approximation of $q(\theta) = BVN((\phi_1, \phi_2)' | \mu, \Sigma) IG(\sigma^2 | \alpha, \beta)$. This form is justified by the mean field approximation returning gaussian marginals for $\phi_1$ and $\phi_2$, and an inverse gamma marginal for $\sigma^2$. Combining this with the gaussian copula returns the bivariate normal distribution.

<<MM, Echo = FALSE>>=
MM = matrix( c(0.37, -0.30, 0.10, -0.03, 0.10, 49.2, 78.4, -77.5, 7.4), ncol = 9)
colnames(MM) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
kable(MM)
@

\section{Mean Field Variational Bayes}
The next step is to perform a mean-field variational bayes, which returns the optimal distribution families in a factorisable posterior approximaiton. 
These were found to be gaussian for $\phi_1$ and $\phi_2$, and inverse gamma for $\sigma^2$. The parameterisation for these are similar to the conditional distributions used in Gibbs sampling, with dependence on other parameter draws replaced with the corresponding expectations. The hyperparameters of these distributions depend on each other through the expectations and the algorithm iterates between these until convergence.

\begin{algorithm}[H]
 \SetKwInOut{Input}{Log Joint Density}
 \KwResult{Mean Field Approximation}
 Initialise $\lambda$ randomly\;
 \While{Not converged}{
  \For{$i =  1$ \KwTo $7$}{
      Hold $\lambda_{j \neq i}$ fixed\;
      Maximise $\lambda_{i}$ \;
     }
 }
 \caption{Mean Field Variational Bayes}
\end{algorithm}

Mean Field Variational Bayes converged within five iterations, and the results are summarised below

<<MFSummary, Echo = FALSE>>=
MFsum = matrix(c(0.37, -0.29, 0.09, 0, 0.10, 49.5, 79.8, -77.4, "1.3 x 10^{-6}"), ncol = 9)
colnames(MFsum) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
kable(MFsum)
@

\begin{figure}
<<meanfield, Echo = FALSE>>=
library(pscl)
library(gridExtra)
mf <- read.csv("AR2mf.csv")[,-1]
astar = 50
xp1 = seq(-0.1, 1, length.out=1000)
p1dens = dnorm(xp1, mf[10, 2], sqrt(mf[10, 3]))
p1dat = data.frame(cbind(xp1, p1dens))
p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)

xp2 = seq(-0.8, 0.4, length.out=1000)
p2dens = dnorm(xp2, mf[10, 4], sqrt(mf[10, 5]))
p2dat = data.frame(cbind(xp2, p2dens))
p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)

xs = seq(0.5, 3.5, length.out = 1000)
sdens = densigamma(xs, astar, mf[10, 1])
sigdat = data.frame(cbind(xs, sdens))
sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)

grid.arrange(p1plot, p2plot, sigplot, ncol = 2)
@
\end{figure}

\section{Copula Variational Bayes 1}

Remembering the Variational Bayes seeks to minimised the Evidence Lower Bound,

$$L(\theta, \lambda) = E_{q} \left[ q_{\theta}(\log(p(\theta, y)) - \log(q(\theta))) \right]$$

We can take gradients as
\begin{eqnarray}
\nabla_{\lambda} L(\theta, \lambda) & = & E_{q} \left[ \nabla_{\lambda}[q_{\theta}(\log(p(\theta, y)) - \log(q(\theta)))] \right] \\
& = & E_{q} \left[ \nabla_{\lambda}[\log(q_{\theta})](\log(p(\theta, y)) - \log(q(\theta))) \right] \nonumber \\
& \approx & 1/S \sum_{s=1}^{S} \nabla_{\lambda}[\log(q_{\theta^s})](\log(p(\theta^s, y)) - \log(q(\theta^s))). \nonumber
\end{eqnarray}

This results in the Stochastic Gradient Ascent algorithm for Variational Bayes:

\begin{algorithm}[H]
 \SetKwInOut{Input}{Log Joint Density, Approximation family q}
 \KwResult{Variational Approximation}
 Initialise $\lambda$ to Methods of Moments Estimator\;
 \While{Not converged}{
  Simulate $\theta_s$ for $s = 1, \dots S$ from $q(\theta^{t-1})$
  \For{$i =  1$ \KwTo $7$}{
      Hold $\lambda_{j \neq i}$ fixed\;
      Calculate $\nabla_{\lambda_i}^{t} =  1/S\sum_{s=1}^{S} \nabla_{\lambda}[\log(q_{\theta^s})](\log(p(\theta^s, y)) - \log(q(\theta^s)))$
     }
  Set $\lambda^{t} = \lambda^{t-1} + p_t \nabla_{\lambda}^{t}$
  Set $t = t + 1$
 }
 \caption{Stochastic Gradient Ascent Algorithm 1}
\end{algorithm}

The Monte Carlo estimator has a huge variance and there were often draws of the same partial derivates between $-200,000$ and $200,000$ for $\alpha$ and $\beta$. Setting $S > 200$ was required, however the algorithm was extremely sensitive to initial conditions, so the Method of Moments estimators were used. Often $\alpha$ and $\beta$ would take negative values so the natural log of these parameters was used instead.

Ranganath et al. 2014 reccomends setting 
$$p_t = \eta diag(G_t)^{-1/2}$$

for some $\eta$, where $G_t$ is the sum of the outer products of partial derivatives for $k = 1, \dots, t$.

With $\eta = 0.1$ and $S = 200$, the algorithm took 244 iterations until the change in $L(\theta, \lambda)$ was less than $0.001$.

\begin{figure}
<<SGA1, Echo = FALSE>>=
SGA1 = matrix( c(0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7, -77.2, 59.5), ncol = 9)
colnames(SGA1) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
kable(SGA1)
L2 =  c(0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7)
L = matrix(c(L2[3:4], 0, L2[5]), 2)
Sigma = L %*% t(L)

xp1 = seq(-0.1, 1, length.out=1000)
p1dens = dnorm(xp1, L2[1], sqrt(Sigma[1, 1]))
p1dat = data.frame(cbind(xp1, p1dens))
p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)

xp2 = seq(-0.8, 0.4, length.out=1000)
p2dens = dnorm(xp2, L2[2], sqrt(Sigma[2, 2]))
p2dat = data.frame(cbind(xp2, p2dens))
p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)

xs = seq(0.5, 3.5, length.out = 1000)
sdens = densigamma(xs, L2[6], L2[7])
sigdat = data.frame(cbind(xs, sdens))
sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)

x1 = seq(-1, 1, length.out=1000)
x2 = seq(-1, 1, length.out=1000)
z = matrix(0, length(x1), length(x2))
for (i in 1:length(x1)) {
  d1 = x1
  d2= x2[i]
  z[,i] = dmvnorm(cbind(d1,d2), L2[1:2], Sigma)
}
colnames(z) = x1
rownames(z) = x2
# reshape the data
dat.p1p2 <- melt(z)
colnames(dat.p1p2) = c("Phi1", "Phi2", "value")

p1p2 <- ggplot(dat.p1p2, aes(x=Phi1, y=Phi2, z = value)) + geom_contour()+ coord_equal()
grid.arrange(p1plot, p2plot, sigplot, p1p2)
@
\end{figure}
\section{Copula Variational Bayes 2}

If we can transform $\theta = f(\epsilon, \lambda)$ where $\epsilon$ is parameter free, the derivative in (1) simplifies.

\begin{eqnarray}
\nabla_{\lambda} L(\theta, \lambda) & = & E_{q} \left[ [q_{\epsilon}\nabla_{\lambda}[(\log(p(\theta, y)) - \log(q(\theta)))] \right] \nonumber \\
& \approx & 1/S \sum_{s=1}^{S} \nabla_{\lambda}(\log(p(\theta^s, y)) - \log(q(\theta^s))) \nonumber
\end{eqnarray}

We can use $(\theta_1, \theta_2)' = \mu + L c(\epsilon_1, \epsilon_2)'$ and $\theta_3 = Q^{-1}(\epsilon_3 | \alpha, \beta)$ where $(\epsilon_1, \epsilon_2) \sim N(0, I)$ and $\epsilon_3 \sim U(0, 1)$

\begin{algorithm}[H]
 \SetKwInOut{Input}{Log Joint Density, Approximation family q}
 \KwResult{Variational Approximation}
 Initialise $\lambda$ to Methods of Moments Estimator\;
 \While{Not converged}{
  Simulate $\epsilon$ for $s = 1, \dots S$ from $N(0, I)$ and $U(0, 1)$
  Transform $\theta^s = f(\epsilon^s, \lambda^{t-1})$
  \For{$i =  1$ \KwTo $7$}{
      Hold $\lambda_{j \neq i}$ fixed\;
      Calculate $\nabla_{\lambda_i}^{t} =  1/S\sum_{s=1}^{S} \nabla_{\lambda}[(\log(p(\theta^s, y)) - \log(q(\theta^s)))]$
     }
  Set $\lambda^{t} = \lambda^{t-1} + p_t \nabla_{\lambda}^{t}$
  Set $t = t + 1$
 }
 \caption{Stochastic Gradient Ascent Algorithm 2}
\end{algorithm}

The resulting Monte Carlo draws from this algorithm have a much smaller variance and much less numerical instability than algorithm 1. $S$ can be set to equal 1, massively speeding up the calculations. Further, it is unneccesary to transform $\alpha$ and $\beta$, and most parameters and invariant to their starting values. $\alpha$ and $\beta$ are not invariant to their starting conditions, but after setting the other parameters to their converged values, it was found that $L(\theta, \lambda)$ is extremely flat for a wide range of $\alpha$ and $\beta$ so the gradients are near zero.

The algorithm converged within a tolerance of $0.001$ after 504 draws.

\begin{figure}
<<SGA2, Echo = FALSE>>=
SGA2 = matrix( c(0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9, -77.4, 1.2), ncol = 9)
colnames(SGA2) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
kable(SGA2)
L2 =  c(0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9)
L = matrix(c(L2[3:4], 0, L2[5]), 2)
Sigma = L %*% t(L)

xp1 = seq(-0.1, 1, length.out=1000)
p1dens = dnorm(xp1, L2[1], sqrt(Sigma[1, 1]))
p1dat = data.frame(cbind(xp1, p1dens))
p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)

xp2 = seq(-0.8, 0.4, length.out=1000)
p2dens = dnorm(xp2, L2[2], sqrt(Sigma[2, 2]))
p2dat = data.frame(cbind(xp2, p2dens))
p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)

xs = seq(0.5, 3.5, length.out = 1000)
sdens = densigamma(xs, L2[6], L2[7])
sigdat = data.frame(cbind(xs, sdens))
sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)

x1 = seq(-1, 1, length.out=1000)
x2 = seq(-1, 1, length.out=1000)
z = matrix(0, length(x1), length(x2))
for (i in 1:length(x1)) {
  d1 = x1
  d2= x2[i]
  z[,i] = dmvnorm(cbind(d1,d2), L2[1:2], Sigma)
}
colnames(z) = x1
rownames(z) = x2
# reshape the data
dat.p1p2 <- melt(z)
colnames(dat.p1p2) = c("Phi1", "Phi2", "value")

p1p2 <- ggplot(dat.p1p2, aes(x=Phi1, y=Phi2, z = value)) + geom_contour()+ coord_equal()
grid.arrange(p1plot, p2plot, sigplot, p1p2)
@
\end{figure}
\end{document}

### 2016-09-22 written report/AR2.tex
updates

--- 2016-09-22 written report/AR2.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{AR2-concordance}
 \section{The AR2 Model}
@@
 
-<<summary, Echo = FALSE, Eval = TRUE>>=
-library(knitr)
-library(mvtnorm)
-library(reshape)
-summary = matrix( c(0.37, -0.30, 0.10, -0.03, 0.10, 49.2, 78.4, -77.5, 7.4,
-                    0.37, -0.29, 0.09, 0, 0.10, 49.5, 79.8, -77.4, "1.3 x 10^{-6}", 
-                    0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7, -77.2, 59.5, 
-                    0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9, -77.4, 1.2) , byrow = TRUE, ncol = 9)
-
-colnames(summary) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)*")
-rownames(summary) = c("Method of Moments", "Mean Field", "Copula 1", "Copula 2")
-kable(summary)
-@
+\begin{Schunk}
+\begin{Sinput}
+> library(knitr)
+> library(mvtnorm)
+> library(reshape)
+> summary = matrix( c(0.37, -0.30, 0.10, -0.03, 0.10, 49.2, 78.4, -77.5, 7.4,
++                     0.37, -0.29, 0.09, 0, 0.10, 49.5, 79.8, -77.4, "1.3 x 10^{-6}", 
++                     0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7, -77.2, 59.5, 
++                     0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9, -77.4, 1.2) , byrow = TRUE, ncol = 9)
+> colnames(summary) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)*")
+> rownames(summary) = c("Method of Moments", "Mean Field", "Copula 1", "Copula 2")
+> kable(summary)
+\end{Sinput}
+\begin{Soutput}
+|                  |mu1  |mu2   |L11  |L21   |L22  |alpha |beta |L(theta) |time (seconds)* |
+|:-----------------|:----|:-----|:----|:-----|:----|:-----|:----|:--------|:---------------|
+|Method of Moments |0.37 |-0.3  |0.1  |-0.03 |0.1  |49.2  |78.4 |-77.5    |7.4             |
+|Mean Field        |0.37 |-0.29 |0.09 |0     |0.1  |49.5  |79.8 |-77.4    |1.3 x 10^{-6}   |
+|Copula 1          |0.32 |-0.32 |0.11 |-0.01 |0.12 |46.4  |78.7 |-77.2    |59.5            |
+|Copula 2          |0.38 |-0.3  |-0.1 |0.03  |-0.1 |48.7  |78.9 |-77.4    |1.2             |
+\end{Soutput}
+\end{Schunk}
 \*Note that all of the variational algorithms are currently unparallelised, and the copula versions in particular have a huge potential for faster performance.
@@
 \begin{figure}
-<<MCMC, Echo = FALSE>>=
-setwd("/home/nltom2/Desktop/Code/Approximate-Updating")
-library(ggplot2)
-library(GGally)
-thin <- read.csv("AR2thin.csv")[,-1]
-ggpairs(thin)
-@
+\begin{Schunk}
+\begin{Sinput}
+> setwd("/home/nltom2/Desktop/Code/Approximate-Updating")
+> library(ggplot2)
+> library(GGally)
+> thin <- read.csv("AR2thin.csv")[,-1]
+> ggpairs(thin)
+\end{Sinput}
+\end{Schunk}
 \end{figure}
@@
 
-<<copula, Echo = FALSE>>=
-
-copest =data.frame(copula = c("gaussian", "t"), rho = c(-0.29, -0.29), df = c("NA", 112.5), BIC = c(-167.8, -161.2))
-kable(copest)
-@
+\begin{Schunk}
+\begin{Sinput}
+> copest =data.frame(copula = c("gaussian", "t"), rho = c(-0.29, -0.29), df = c("NA", 112.5), BIC = c(-167.8, -161.2))
+> kable(copest)
+\end{Sinput}
+\begin{Soutput}
+|copula   |   rho|df    |    BIC|
+|:--------|-----:|:-----|------:|
+|gaussian | -0.29|NA    | -167.8|
+|t        | -0.29|112.5 | -161.2|
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<MM, Echo = FALSE>>=
-MM = matrix( c(0.37, -0.30, 0.10, -0.03, 0.10, 49.2, 78.4, -77.5, 7.4), ncol = 9)
-colnames(MM) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
-kable(MM)
-@
+\begin{Schunk}
+\begin{Sinput}
+> MM = matrix( c(0.37, -0.30, 0.10, -0.03, 0.10, 49.2, 78.4, -77.5, 7.4), ncol = 9)
+> colnames(MM) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
+> kable(MM)
+\end{Sinput}
+\begin{Soutput}
+|  mu1|  mu2| L11|   L21| L22| alpha| beta| L(theta)| time (seconds)|
+|----:|----:|---:|-----:|---:|-----:|----:|--------:|--------------:|
+| 0.37| -0.3| 0.1| -0.03| 0.1|  49.2| 78.4|    -77.5|            7.4|
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<MFSummary, Echo = FALSE>>=
-MFsum = matrix(c(0.37, -0.29, 0.09, 0, 0.10, 49.5, 79.8, -77.4, "1.3 x 10^{-6}"), ncol = 9)
-colnames(MFsum) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
-kable(MFsum)
-@
+\begin{Schunk}
+\begin{Sinput}
+> MFsum = matrix(c(0.37, -0.29, 0.09, 0, 0.10, 49.5, 79.8, -77.4, "1.3 x 10^{-6}"), ncol = 9)
+> colnames(MFsum) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
+> kable(MFsum)
+\end{Sinput}
+\begin{Soutput}
+|mu1  |mu2   |L11  |L21 |L22 |alpha |beta |L(theta) |time (seconds) |
+|:----|:-----|:----|:---|:---|:-----|:----|:--------|:--------------|
+|0.37 |-0.29 |0.09 |0   |0.1 |49.5  |79.8 |-77.4    |1.3 x 10^{-6}  |
+\end{Soutput}
+\end{Schunk}
 
 \begin{figure}
-<<meanfield, Echo = FALSE>>=
-library(pscl)
-library(gridExtra)
-mf <- read.csv("AR2mf.csv")[,-1]
-astar = 50
-xp1 = seq(-0.1, 1, length.out=1000)
-p1dens = dnorm(xp1, mf[10, 2], sqrt(mf[10, 3]))
-p1dat = data.frame(cbind(xp1, p1dens))
-p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)
-
-xp2 = seq(-0.8, 0.4, length.out=1000)
-p2dens = dnorm(xp2, mf[10, 4], sqrt(mf[10, 5]))
-p2dat = data.frame(cbind(xp2, p2dens))
-p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)
-
-xs = seq(0.5, 3.5, length.out = 1000)
-sdens = densigamma(xs, astar, mf[10, 1])
-sigdat = data.frame(cbind(xs, sdens))
-sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)
-
-grid.arrange(p1plot, p2plot, sigplot, ncol = 2)
-@
+\begin{Schunk}
+\begin{Sinput}
+> library(pscl)
+> library(gridExtra)
+> mf <- read.csv("AR2mf.csv")[,-1]
+> astar = 50
+> xp1 = seq(-0.1, 1, length.out=1000)
+> p1dens = dnorm(xp1, mf[10, 2], sqrt(mf[10, 3]))
+> p1dat = data.frame(cbind(xp1, p1dens))
+> p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)
+> xp2 = seq(-0.8, 0.4, length.out=1000)
+> p2dens = dnorm(xp2, mf[10, 4], sqrt(mf[10, 5]))
+> p2dat = data.frame(cbind(xp2, p2dens))
+> p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)
+> xs = seq(0.5, 3.5, length.out = 1000)
+> sdens = densigamma(xs, astar, mf[10, 1])
+> sigdat = data.frame(cbind(xs, sdens))
+> sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)
+> grid.arrange(p1plot, p2plot, sigplot, ncol = 2)
+\end{Sinput}
+\end{Schunk}
 \end{figure}
@@
 \begin{figure}
-<<SGA1, Echo = FALSE>>=
-SGA1 = matrix( c(0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7, -77.2, 59.5), ncol = 9)
-colnames(SGA1) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
-kable(SGA1)
-L2 =  c(0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7)
-L = matrix(c(L2[3:4], 0, L2[5]), 2)
-Sigma = L %*% t(L)
-
-xp1 = seq(-0.1, 1, length.out=1000)
-p1dens = dnorm(xp1, L2[1], sqrt(Sigma[1, 1]))
-p1dat = data.frame(cbind(xp1, p1dens))
-p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)
-
-xp2 = seq(-0.8, 0.4, length.out=1000)
-p2dens = dnorm(xp2, L2[2], sqrt(Sigma[2, 2]))
-p2dat = data.frame(cbind(xp2, p2dens))
-p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)
-
-xs = seq(0.5, 3.5, length.out = 1000)
-sdens = densigamma(xs, L2[6], L2[7])
-sigdat = data.frame(cbind(xs, sdens))
-sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)
-
-x1 = seq(-1, 1, length.out=1000)
-x2 = seq(-1, 1, length.out=1000)
-z = matrix(0, length(x1), length(x2))
-for (i in 1:length(x1)) {
-  d1 = x1
-  d2= x2[i]
-  z[,i] = dmvnorm(cbind(d1,d2), L2[1:2], Sigma)
-}
-colnames(z) = x1
-rownames(z) = x2
-# reshape the data
-dat.p1p2 <- melt(z)
-colnames(dat.p1p2) = c("Phi1", "Phi2", "value")
-
-p1p2 <- ggplot(dat.p1p2, aes(x=Phi1, y=Phi2, z = value)) + geom_contour()+ coord_equal()
-grid.arrange(p1plot, p2plot, sigplot, p1p2)
-@
+\begin{Schunk}
+\begin{Sinput}
+> SGA1 = matrix( c(0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7, -77.2, 59.5), ncol = 9)
+> colnames(SGA1) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
+> kable(SGA1)
+\end{Sinput}
+\begin{Soutput}
+|  mu1|   mu2|  L11|   L21|  L22| alpha| beta| L(theta)| time (seconds)|
+|----:|-----:|----:|-----:|----:|-----:|----:|--------:|--------------:|
+| 0.32| -0.32| 0.11| -0.01| 0.12|  46.4| 78.7|    -77.2|           59.5|
+\end{Soutput}
+\begin{Sinput}
+> L2 =  c(0.32, -0.32, 0.11, -0.01, 0.12, 46.4, 78.7)
+> L = matrix(c(L2[3:4], 0, L2[5]), 2)
+> Sigma = L %*% t(L)
+> xp1 = seq(-0.1, 1, length.out=1000)
+> p1dens = dnorm(xp1, L2[1], sqrt(Sigma[1, 1]))
+> p1dat = data.frame(cbind(xp1, p1dens))
+> p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)
+> xp2 = seq(-0.8, 0.4, length.out=1000)
+> p2dens = dnorm(xp2, L2[2], sqrt(Sigma[2, 2]))
+> p2dat = data.frame(cbind(xp2, p2dens))
+> p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)
+> xs = seq(0.5, 3.5, length.out = 1000)
+> sdens = densigamma(xs, L2[6], L2[7])
+> sigdat = data.frame(cbind(xs, sdens))
+> sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)
+> x1 = seq(-1, 1, length.out=1000)
+> x2 = seq(-1, 1, length.out=1000)
+> z = matrix(0, length(x1), length(x2))
+> for (i in 1:length(x1)) {
++   d1 = x1
++   d2= x2[i]
++   z[,i] = dmvnorm(cbind(d1,d2), L2[1:2], Sigma)
++ }
+> colnames(z) = x1
+> rownames(z) = x2
+> # reshape the data
+> dat.p1p2 <- melt(z)
+> colnames(dat.p1p2) = c("Phi1", "Phi2", "value")
+> p1p2 <- ggplot(dat.p1p2, aes(x=Phi1, y=Phi2, z = value)) + geom_contour()+ coord_equal()
+> grid.arrange(p1plot, p2plot, sigplot, p1p2)
+\end{Sinput}
+\end{Schunk}
 \end{figure}
@@
 \begin{figure}
-<<SGA2, Echo = FALSE>>=
-SGA2 = matrix( c(0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9, -77.4, 1.2), ncol = 9)
-colnames(SGA2) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
-kable(SGA2)
-L2 =  c(0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9)
-L = matrix(c(L2[3:4], 0, L2[5]), 2)
-Sigma = L %*% t(L)
-
-xp1 = seq(-0.1, 1, length.out=1000)
-p1dens = dnorm(xp1, L2[1], sqrt(Sigma[1, 1]))
-p1dat = data.frame(cbind(xp1, p1dens))
-p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)
-
-xp2 = seq(-0.8, 0.4, length.out=1000)
-p2dens = dnorm(xp2, L2[2], sqrt(Sigma[2, 2]))
-p2dat = data.frame(cbind(xp2, p2dens))
-p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)
-
-xs = seq(0.5, 3.5, length.out = 1000)
-sdens = densigamma(xs, L2[6], L2[7])
-sigdat = data.frame(cbind(xs, sdens))
-sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)
-
-x1 = seq(-1, 1, length.out=1000)
-x2 = seq(-1, 1, length.out=1000)
-z = matrix(0, length(x1), length(x2))
-for (i in 1:length(x1)) {
-  d1 = x1
-  d2= x2[i]
-  z[,i] = dmvnorm(cbind(d1,d2), L2[1:2], Sigma)
-}
-colnames(z) = x1
-rownames(z) = x2
-# reshape the data
-dat.p1p2 <- melt(z)
-colnames(dat.p1p2) = c("Phi1", "Phi2", "value")
-
-p1p2 <- ggplot(dat.p1p2, aes(x=Phi1, y=Phi2, z = value)) + geom_contour()+ coord_equal()
-grid.arrange(p1plot, p2plot, sigplot, p1p2)
-@
+\begin{Schunk}
+\begin{Sinput}
+> SGA2 = matrix( c(0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9, -77.4, 1.2), ncol = 9)
+> colnames(SGA2) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L(theta)", "time (seconds)")
+> kable(SGA2)
+\end{Sinput}
+\begin{Soutput}
+|  mu1|  mu2|  L11|  L21|  L22| alpha| beta| L(theta)| time (seconds)|
+|----:|----:|----:|----:|----:|-----:|----:|--------:|--------------:|
+| 0.38| -0.3| -0.1| 0.03| -0.1|  48.7| 78.9|    -77.4|            1.2|
+\end{Soutput}
+\begin{Sinput}
+> L2 =  c(0.38, -0.30, -0.1, 0.03, -0.10, 48.7, 78.9)
+> L = matrix(c(L2[3:4], 0, L2[5]), 2)
+> Sigma = L %*% t(L)
+> xp1 = seq(-0.1, 1, length.out=1000)
+> p1dens = dnorm(xp1, L2[1], sqrt(Sigma[1, 1]))
+> p1dat = data.frame(cbind(xp1, p1dens))
+> p1plot = ggplot(p1dat, aes(x=xp1, y=p1dens)) + geom_line() + labs(x="Phi 1", y=NULL)
+> xp2 = seq(-0.8, 0.4, length.out=1000)
+> p2dens = dnorm(xp2, L2[2], sqrt(Sigma[2, 2]))
+> p2dat = data.frame(cbind(xp2, p2dens))
+> p2plot = ggplot(p2dat, aes(x=xp2, y=p2dens)) + geom_line() + labs(x="Phi 2", y=NULL)
+> xs = seq(0.5, 3.5, length.out = 1000)
+> sdens = densigamma(xs, L2[6], L2[7])
+> sigdat = data.frame(cbind(xs, sdens))
+> sigplot = ggplot(sigdat, aes(x=xs, y=sdens)) + geom_line() + labs(x="Sigma Squared", y=NULL)
+> x1 = seq(-1, 1, length.out=1000)
+> x2 = seq(-1, 1, length.out=1000)
+> z = matrix(0, length(x1), length(x2))
+> for (i in 1:length(x1)) {
++   d1 = x1
++   d2= x2[i]
++   z[,i] = dmvnorm(cbind(d1,d2), L2[1:2], Sigma)
++ }
+> colnames(z) = x1
+> rownames(z) = x2
+> # reshape the data
+> dat.p1p2 <- melt(z)
+> colnames(dat.p1p2) = c("Phi1", "Phi2", "value")
+> p1p2 <- ggplot(dat.p1p2, aes(x=Phi1, y=Phi2, z = value)) + geom_contour()+ coord_equal()
+> grid.arrange(p1plot, p2plot, sigplot, p1p2)
+\end{Sinput}
+\end{Schunk}
 \end{figure}

### 2016-09-22 written report/ar1_predictive.Rnw
updates

@@
 library(RcppArmadillo)
-sourceCpp("AR1c.cpp")
+#sourceCpp("AR1c.cpp")
 library(truncnorm)

### 2016-09-22 written report/ar1_predictive.tex
updates

@@
 \input{ar1_predictive-concordance}
+
+Note: Code uses a stretched beta, latex math still uses regular beta.
 
@@
 The algorithm took 0.204 seconds to run in R and resulted in a $\mathcal{N}(3.06, 0.96)$ distribution for $q(Y_{T+1})$.
-
-A Metropolis Hastings MCMC scheme was set up using a random walk candidate distribution for $\psi$ to draw from $p(\mu | y_{1:T}), p(\sigma^2 | y_{1:T})$ and $p(\psi | y_{1:T})$.
-These draws were then sampled to find the distribution of $p(y_{T+1} | y_{1:T})$.
-The chain had 1000 burn in draws followed by 50000 kept draws and took 6.736 seconds to complete the chain and forecast distribution.
-
-\begin{figure}[hb]
-\centering
-\includegraphics{ar1_predictive-plots}
-\caption{Variational approximation for $q(y_{T+1}|y_{1:T})$ (blue) and the MCMC density for $p(y_{T+1}|y_{1:T})$ (red). 
-The KL Divergence from q to p is 0.015, and from p to q is 0.018.}
-\label{densities}
-\end{figure}
-
-\end{document}

### 2016-09-22 AR1_twoparam.R

@@ -1,0 +1,77 @@
+library(GGally)
+library(coda)
+library(ggplot2)
+library(gridExtra)
+library(reshape)
+library(mvtnorm)
+
+T = 1000    #Length of y
+N = 15000   #MCMC draws
+
+#True parameters
+sigma2 <- 1
+phi <- 0.97
+
+#Generate data
+y = rnorm(1, 0, sqrt(sigma2/(1-phi^2)))
+for(i in 2:T){
+  y = c(y, phi*y[(i-1)] + rnorm(1, 0, sqrt(sigma2)))
+}
+
+#Uninformative prior
+p1 = 1
+p2 = 1
+a = 1
+b = 1
+
+#constant
+shape = T/2 + a
+
+#lphi kernel
+lphiden = function(draw, sig2) {
+  t1 = 1/2*log(1-draw^2) -(1-draw^2)*y[1]^2/(2*sig2)  #The normal exponential for y1
+  t2 = -sum(((y[2:T]-draw*y[1:(T-1)])^2))/(2*sig2)    #The normal exponentials for the rest of the y's
+  t3 = log(1+draw)*(p1-1) + log(1-draw)*(p2-1)   #The stretched beta prior
+  return(t1+t2+t3)
+}
+
+theta = matrix(0, ncol = 2, nrow = N)
+theta[1, ] = c(0.5, 1)
+accept = 0
+
+for(i in 2:N){
+  #phi MH step
+  #phihat = sum(y[2:T]*y[1:(T-1)])/sum(y[1:(T-1)]^2)
+  #vphi = theta[(i-1),3]/sum(y[1:(T-1)]^2)
+  #phidraw = rnorm(1, phihat, sqrt(vphi))
+  phidraw = rnorm(1, theta[(i-1),1], 0.1)     
+  
+  if(phidraw > -1 & phidraw < 1) {                       #within stationary conditions
+    mh.candidate = lphiden(phidraw, theta[(i-1), 2])     #symmetrical candidate so only need to evaluate the true density components
+    mh.prev = lphiden(theta[(i-1), 1], theta[(i-1), 2])  #using log density
+    alpha = min(1, exp(mh.candidate-mh.prev))
+    u = runif(1)
+    if(u <= alpha) {             #accept draw
+      theta[i, 1] = phidraw
+      accept = accept + 1
+    } else {                    #reject draw
+      theta[i, 1] = theta[(i-1), 1]
+    }
[…]

### 2017-01-02 Confirmation/AR2_example.Rnw
Xmas break updates

\documentclass{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}

\begin{document}

\section{The AR2 Model}

Consider a mean zero AR(2) model conditioned on $y_1$ and $y_2$.

$$y_t = \phi_1 * y_{t-1} + \phi_2 * y_{t-2} + \epsilon_t \mbox{ for } t = 3,\dots,T$$,

where $\epsilon$ are iid $\mathcal{N}(0, \sigma^2)$.

Using the independent priors:

\begin{align}
p(\phi_1) &\sim \mathcal{N}(\bar{\phi_1}, \tau_{1}^2) \nonumber \\
p(\phi_2) &\sim \mathcal{N}(\bar{\phi_2}, \tau_{2}^2) \nonumber \\
p(\sigma^2) &\sim \mathcal{IG} (\mbox{shape} = \alpha, \mbox{ rate} = \beta),
\end{align}
we have a model with zero stationary restrictions that can be easily sampled through Gibbs MCMC or approximated by a range of variational techniques.

There are five approaches considered:

\begin{itemize}
\item Gibbs MCMC
\item Mean field Variational Bayes
\item Stochastic Variational Bayes
\item Stochastic Variational Bayes with Control Variates
\item Re-parameterised Stochastic Variational Bayes
\end{itemize}

Simulation proceeds with $T = 100$, $\phi_1 = 0.7$, $\phi_2 = 0.15$ and $\sigma^2 = 2$. Each approach estimates the same approximating distribution, $q(\theta) = \mathcal{BVN}((\phi_1, \phi_2)' | \mu, \Sigma) \mathcal{IG}(\sigma^2 | \alpha, \beta)$ with estimated parameters:

\begin{itemize}
\item $\mu_1$: The mean of $\phi_1$,
\item $\mu_2$: The mean of $\phi_2$,
\item $L_{11}, L_{21}, L_{22}$, the three components of $L$, the lower triangular decomposition of $\Sigma$,
\item $\alpha, \beta$, the shape and rate parameters of an Inverse Gamma distribution for $\sigma^2$.
\end{itemize}

The diagonal of $L$ is not forced to be positive so the decomposition is not invariant to sign changes.

Collectively, we have $\lambda = (\mu_1, \mu_2, L_{11}, L_{21}, L_{22}, \alpha, \beta)'$.

The reparameterised SVB algorithm drew $(\epsilon_1, \epsilon_2)' \sim \mathcal{N}(0, I)$ and $\epsilon_3 \sim \mathcal{U}(0, 1)$, then made the transformations:

\begin{align}
(\theta_1, \theta_2)' &= (\mu_1, \mu_2)' + L (\epsilon_1, \epsilon_2)' \nonumber \\
\theta_3 &= Q^{-1}_{IG}(\epsilon_3| \alpha, \beta) \nonumber
\end{align}

where $Q^{-1}_{IG}$ is the inverse CDF of an Inverse Gamma distribution with shape $\alpha$ and scale $\beta$.

We compare the converged values of each algorithm to the Methods of Moments (MM) estimator of the approximating distribution using the MCMC draws as data.
\vspace{3mm}
\begin{table}[hb]
\centering
<<summary, echo = FALSE>>=
library(knitr)
library(mvtnorm)
library(reshape)
summary = matrix( c(0.78, 0.16, 0.10, -0.09, 0.05, 48.9, 77.3, -76.7, 2.6,
                    0.83, 0.10, 0.04, 0, 0.05, 50, 79.7, -77.7, "1.3 x 10^{-6}", 
                    0.63, 0.27, 0.09, -0.12, 0.05, 45.6, 81.4, -78.3, 51.1, 
                    0.78, 0.15, 0.1, -0.09, 0.05, 49.3, 78, -76.7, 4.0,
                    0.77, 0.15, 0.09, -0.09, 0.05, 48.5, 77.6, -76.9, 1.3) , byrow = TRUE, ncol = 9)

colnames(summary) = c(paste0(expression(mu), 1), paste0(expression(mu), 2), "L11", "L21", "L22", expression(alpha), expression(beta), "L", "time (seconds)*")
rownames(summary) = c("MCMC + MM", "Mean Field", "SVB", "SVB CV", "SVB RP")
kable(summary)
@
\caption{$\lambda$ estimates, maximised $\mathcal{L}(q(\theta)|\lambda), y)$ and runtime for each algorithm. The Mean Field algorithm is the fastest, but has not captured the posterior variance, while the original Stochastic algorithm has converged to a different local maximum than the others. MCMC with the Method of Moments and both reduced variance VB algorithms have captured almost the same distribution, with re-parameterised Stochastic VB as the fastest. All algorithms are run unparallelised, so the VB algorithms have potential for further speed increases.}
\end{table}

sga <- rmvnorm(4000, mean = c(0.77, 0.15), sigma = (L %*% t(L)))
sga <- cbind(sga, 1/rgamma(4000, shape = 48.5, rate = 77.6))
\begin{figure}[h]
<<MCMC, echo= FALSE, fig.width = 7, fig.height = 6>>=
setwd("/home/nltom2/Desktop/Confirmation")
library(ggplot2)
library(GGally)
thin <- read.csv("AR2thin.csv")[,-1]
colnames(thin) <- c("Phi_1", "Phi_2", "Sigma_Squared")
ggpairs(thin) + theme_bw()
@
\caption{Result of MCMC with 120,000 draws. The first 20\% were discarded then the remaining draws were thinned by a factor of 20, which was required to remove any dependence. The remaining effective sample size for each parameter is 5000.}
\end{figure}

\end{document}

### 2017-01-02 Confirmation/VB.Rnw
Xmas break updates

\documentclass{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\numberwithin{equation}{section}

\begin{document}

\tableofcontents

\section{Bayesian Inference}

\subsection{Bayes' Rule}

In Bayesian statistics, we are interested in making inferences about an unknown parameter $\theta$ using what is known as the posterior distribution, $p(\theta | y)$, where $y$ is our observed data. Bayes' rule provides a form for this distribution,

\begin{equation}
\label{BR1}
p(\theta | y) = \frac{p(y | \theta) p(\theta)}{p(y)},
\end{equation}

which shows that the posterior can be found by taking the product of the likelihood, $p(y | \theta)$ and the prior distribution, $p(\theta)$, and normalising with a constant $p(y)$. In practice $p(y)$ is unknown and can be found from

\begin{equation}
\label{BR2}
p(y) = \int_\theta p(y | \theta) p(\theta) d \theta.
\end{equation}

Finding the value of $p(y)$ is often the most difficult part of working with Bayesian statistics. In some simple models, this integration can be solved analytically, but for most models of interest it is intractable and we can only work with the un-normalised version

\begin{equation}
\label{BR3}
p(\theta | y)^* = p(y | \theta) p(\theta).
\end{equation}

This work will concern itself with two of the methods used in the case where (\ref{BR2}) is intractable, exact methods such as MCMC, which will be discussed later in this section as well as an approximation method known as Variational Bayes which appears in section two.

\subsection{Exact Bayesian Computation}

\textbf{1. Numerical Integration}\newline

It might be tempting to numerically approximate (\ref{BR2}) by evaluting the integrand over a grid and applying methods such as the Trapezoid Rule or Adaptive Quadrature. These may be suitable for integration in a low dimension, but if we have a grid with $G$ points in one dimension, then a two dimensional integration requires $G^2$ points, and a three dimensional integration requires $G^3$ points. In general, the computational time increases exponentially with the dimension of $\theta$ in numerical integration. In almost all cases of interest, $\theta$ is a high-dimensional vector which makes numerical integration prohibitively expensive. This is known as the curse of dimensionality, where the difficulty of the problem increases quickly with the dimension of $\theta$ and requires us to use more sophisticated methods.\newline

\textbf{2. Importance Sampling} \newline

Candidate Density
Accept Reject
Good candidate is hard but useful
VB as a candidate \newline

\textbf{3. Gibbs Sampling}\newline

Breaks down into simple problems
Becomes linear instead of exponential
Exact sampling \newline

\textbf{4. Metropolis Hastings within Gibbs}\newline

Gibbs only works exactly on some problems
Approximation within Gibbs
Also can benefit from VB

\section{Variational Bayes}

\subsection{Estimation as an optimisation problem}

In the event that our dataset is too large, or our model is too complex, for MCMC to converge within a reasonable time frame we must resort to approximation. One important question that might be asked then is \textit{`How do I measure how good my approximation is?'}. If two distributions come from the same family, one might consider the difference in the parameters, but it is trivial to come up with distributions with slight parameter differences and vastly different regions of probability mass. Consider the difference between a $\mathcal{N}(0, 0.1)$ and a $\mathcal{N}(0.5, 0.1)$ distribution, the change in mean is only $0.5$ but there is little overlap in the densities. Further, large values for the variance allow large changes in mean with very little change in the resulting densities. If the distributions come from different families it is almost impossible to use parameters to measure the degree of closeness between two distributions. Instead we must move from the parameter space to the distribution space, where one measure of the difference between distributions $p(\theta)$ and $q(\theta)$ is the Kullback-Leibler (KL) divergence \citep{Kullback1951}, defined by

\begin{equation}
\label{KL-def}
KL[q(\theta)||p(\theta)] = \int q(\theta) \ln \left( \frac{q(\theta)}{p(\theta)}\right) d\theta.
\end{equation}

KL divergence is a non-negative, assymetric measure of the difference between $p(\theta)$ and $q(\theta)$ that will equal zero if and only if $p(\theta) = q(\theta)$ almost everywhere. It has origins in information theory, and can be interpreted as: \textit{`Given that I know $p(\theta)$ for some $\theta \in \Theta$, how much extra information is required, on average, to know the value of $q(\theta)$?'} There are examples of the literature of approximations with other measures of divergence, such as \citet{Minka2001} introducting Expectation Propogation, which minimises the reverse measure $KL[p(\theta)||q(\theta)]$, which was extended to Power-EP in \citet{Minka2004}, which aims to minimisea the more general $\alpha-\mbox{divergence}$ \citep{Amari1985}.
\vspace{3mm}

\begin{figure}[hb]
<<klnorm, echo = FALSE, fig.width = 6, fig.height = 2>>=
library(ggplot2)
xseq = seq(-0.5, 1, 0.01)
w1 = dnorm(xseq, 0, 0.1)
w2 = dnorm(xseq, 0.5, 0.1)
ggplot() + geom_line(data=NULL, aes(x=xseq, y = w1), colour = "red") + geom_line(data=NULL, aes(x=xseq, y = w2), colour = "green") + labs(x="Support", y = "Density") + theme_bw()
@
\caption{Red: A Normal(0, 0.1) distribution. Green: A Normal(0.5, 0.1) distribution.}
\end{figure}

\newpage

Variational Bayes aims to find the approximating distribution, $q(\theta | \lambda)$, that minimises $KL[q(\theta | \lambda)||p(\theta |y)]$. This approximating distribution is parameterised by a $\lambda$ vector, each element of which may be a function of $y$, the hyperparameters of the prior distribution, and even of other elements in $\lambda$. The optimal approximation, in the sense of having the minimum KL divergence with the posterior, can be derived by beginning with the unknown constant, $\ln(p(y))$.
\begin{align}
\ln(p(y)) &= \int_{\theta} q(\theta|\lambda) \ln(p(y)) d\theta \nonumber \\
&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \right) d\theta \nonumber \\
&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \frac{q(\theta|\lambda)}{q(\theta|\lambda)}\right) d\theta \nonumber \\
&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) + \ln \left( \frac{p(y, \theta)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) d\theta + \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
&= KL [q(\theta||\lambda) || p(\theta | y)] + \mathcal{L}(q(\theta | \lambda), y) \nonumber
\end{align}

where

\begin{align}
\mathcal{L}(q(\theta|\lambda), y) &= \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta|\lambda)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
&=  \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta. \label{ELBO}
\end{align}

Hence we have that
\begin{equation}
\label{VBKL}
KL [q(\theta|\lambda) || p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q(\theta | \lambda), y).
\end{equation}

From (\ref{VBKL}) it is clear that by maximising $\mathcal{L}(q(\theta | \lambda), y)$ with respect to $q(\theta|\lambda)$ the KL divergence between the unknown true posterior $p(\theta|y)$ and the approximating distribution $q(\theta|\lambda)$ will be minimised.
Hence the Variational Bayes solution will be to maximise the Evidence Lower Bound (ELBO), $\mathcal{L}(q(\theta|\lambda), y)$, in (\ref{ELBO}).

\subsection{Mean Field Variational Bayes}

Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and uses the assumption that the approximation dististribution factorises, 
\begin{equation}
\label{mf1}
q(\theta|\lambda) = \prod_i q_i(\theta_i | \lambda_i).
\end{equation}

This assumption is known as the Mean Field assumption and is has been widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}). Each component $\theta_i$ may be a scalar or a vector, and $\lambda_i$ is the component of the $\lambda$ vector that parameterises the relevant factor $q_i(\theta_i |\lambda_i)$. From here, we will use the shorthand notation that $q_i = q_i(\theta_i|\lambda_i)$ and $q_{\setminus i} = \prod_{j\neq i}q_j$. Maximising the ELBO with respect to $q_i$ is analytically involved but computationally simple, as (\ref{ELBO}) can be expressed as a function of $q_i$ only and then maximised with respect to each $q_i$ individually.

\begin{align}
\mathcal{L}(q_i) &= \int_{\theta} q_i q_{\setminus i} \ln (p(y, \theta)) d\theta - \int_{\theta} q_{i}q_{\setminus i} \ln (q_{i}q_{\setminus i}) d\theta \nonumber \\
&= \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} ( \ln (p(y, \theta)) - \ln(q_{i})) d\theta_i d\theta_{\setminus i} - \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} \ln (q_{\setminus i}) d\theta_i d\theta_{\setminus i}  \nonumber \\
&= \int_{\theta_{i}} q_{i} \left( \int_{\theta_{\setminus i}} q_{\setminus i} \ln (p(y, \theta )) d\theta_{\setminus i} - \ln(q_i) \right) d\theta_i \nonumber \\
&- \int_{\theta_{\setminus i}} q_{\setminus i} \ln(q_{\setminus i}) d\theta_{\setminus i} \label{mf2}\\
&= \int_{\theta_{1}} q_{i} \ln \left( \frac{\exp(  \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])}{q_i} \right) d\theta_i + c \nonumber \\
&= -KL (q_i || \exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])) + c \label{mf2.2}
\end{align}

as the rightmost component of (\ref{mf2}) is constant with respect to $q_i$.

We can see that by holding $q_{\setminus_i}$ fixed, we can maximise the ELBO with respect to $q_i$ by minimising the KL Divergence between $q_i$ and $\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])$ in (\ref{mf2.2}). This is easily done by setting

\begin{equation}
\label{mf3}
q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))]).
\end{equation}

If the likelihood and prior for $\theta_i$ form an exponential family conjugate pair, then this maximisation proceeds by keeping the distributional family constant and simply updating the parameters $\lambda_i$. Otherwise, we can find the optimal distributional family for any $q_i$ by taking the log-joint density and then ignoring any term that doesn't depend on $\theta_i$. Next, take the expectation of any remaining $\theta_{j \neq i}$ and then match to the kernel of a known distribution. If it does not match a known distribution, we may need to make a further approximation $\tilde{q_i(\theta_i|\lambda_i)}$ which has a recognizable distribution. One method as used in \citet{Friston2006} uses a Laplace approximation for the otherwise unrecognizable $q_i(\theta_i | \lambda_i)$ with reasonable results.

These steps are almost the same as those used to find a conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, with dependence on other parameters replaced by their expectations. Hence, the optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family. Deriving the updating algorithm for MFVB takes more analytical effort than finding the conditionals for a Gibbs MCMC scheme, as we must then find the values of each $\lambda_i$ implied by (\ref{mf3}), which are often a function of $\lambda_{j \neq i}$ through the expectations.
\vspace{5mm}

Given these optimal distributional families, an algorithm that cycles through all of the $\lambda_i$ and sets each to its maximising value is known as a coordiante ascent algorithm. The algorithm must continuously iterate between each parameter until $\mathcal{L}(q(\theta | \lambda), y)$ converges with some pre-defined threshold noting that the $\lambda$ may depend on each other. If $\lambda$ has dimension $k$, the algorithm follows below.

\vspace{2mm}

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density}
 \KwResult{Mean Field Approximation}
 Initialise $\lambda$ randomly\;
 \While{Not converged}{
  \For{$i =  1$ \KwTo $k$}{
      Hold $\lambda_{j \neq i}$ fixed\;
      Use (\ref{mf3}) to match $q(\theta_i|\lambda_i)$ to a known distribution\;
      Update $\lambda_i$ using the most recent values of $\lambda_{j \neq i}$\;
     }
 }
 \caption{Coordinate Ascent for MFVB}
  \label{alg:algorithm1}
\end{algorithm}

\subsection{Stochastic Gradient Ascent}

In many cases, the posterior is too complex to be captured by a factorising approximation and we desire our distribution $q(\theta | \lambda)$ to capture dependence between parameters. In this case, the easy maximisation in MFVB is unavailable, and we must resort to what is called Stochastic Variational Bayes (SVB) using a gradient ascent algorithm developed by \citet{Paisley2012} and \citet{Hoffman2013}.

To maximise the function $\mathcal{L}(q(\theta | \lambda), y)$ we can take the derivative of $\mathcal{L}(q(\theta | \lambda), y)$ with respect to $\lambda$ and use the updating step:

\begin{equation}
\label{SGA1}
\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m-1)}), y),
\end{equation}

where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m-1)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m-1)}), y)$ with respect to each element of $\lambda$. This update requires initial values $\lambda^{(0)}$ and some sequence $\rho^{(m)}$. If $\rho^{(m)}$ is chosen to satisfy the following conditions, it is a Robbins-Monro sequence and the algorithm is guaranteed to converge to a local maximum.

\begin{align}
\lim_{m \xrightarrow[]{} \infty} \rho^{(m)} &= 0 \nonumber \\
\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \nonumber \\
\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty. \nonumber
\end{align}

Whilst a global maximum is desired, the ELBO can have a problem specific shape that makes finding the global maximum extremely difficult, as we do not know how many stationary points exist. The dimension of the ELBO is the dimension of the $\lambda$ vector, which is often much greater than the dimension of the parameters $\theta$ so a grid search is suspect to the curse of dimensionality. To alleiviate this problem, we can start the algorithm at a range of initial values choose the converged value with the highest value of the ELBO.

SVB does not provide the family of the the optimal approximating distribution, unlike under the mean field assumption. To run SVB we may need to try many approximating distributions and then choose the $q(\theta | \lambda)$ that has the highest ELBO, and hence lowest KL divergence to the true posterior. We must restrict the approximation to distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable. In this case, we can calculate derivatives of the ELBO.

\begin{align}
\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \nabla_{\lambda} \int_{\theta} q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
&=   \int_{\theta} \nabla_{\lambda}[  q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)] d\theta \label{SGA2} \\
&=  \int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta)) - \ln(q(\theta | \lambda))] d\theta \nonumber \\ 
&+ \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \label{SGA3}
\end{align}

Note that the first integral on the right hand side of (\ref{SGA3}) equals 0 as

\begin{equation}
\int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta))] d\theta = 0
\end{equation}

and

\begin{align}
\int_{\theta} - q(\theta | \lambda) \nabla_{\lambda}[ \ln(q(\theta | \lambda))] d\theta &= - \int_{\theta} \frac{\nabla_{\lambda}q(\theta | \lambda)}{q(\theta | \lambda)} q(\theta | \lambda) d\theta \nonumber \\
&=  -  \nabla_{\lambda}\int_{\theta} q(\theta | \lambda) d\theta \nonumber \\
&=  0 \nonumber
\end{align}

Then we continue from (\ref{SGA3}) with

\begin{align}
\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
&=   \int_{\theta} q(\theta | \lambda)  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)d\theta \nonumber \\
&= E_q \left(  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) \right). \label{SGA4}
\end{align}

We can compute a Monte-Carlo estimate of (\ref{SGA4}) by

\begin{equation}
\label{SGA5}
\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda)) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)))
\end{equation}

where $\theta_s$, $s = 1, \dots S$, are simulated from the most recent iteration of  $q(\theta | \lambda)$.

As the distribution $q(\theta | \lambda)$ is specified by the user, the main restriction on the use of SVB is that the log-joint density $\ln(p(y, \theta))$ is able to be evaluated. Running an SVB algorithm involves very little analytical effort unlike MFVB, most of the difficulty is calculating the score function of the approximating distribution,
$\nabla_{\lambda} [\ln(q(\theta_s | \lambda))]$. But as this is part of the approximation and not the true model the code to run this could potentially be re-used in many different models.

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density, Approximation family q}
 \KwResult{Variational Approximation}
 Initialise $\lambda$\;
 \While{Not converged}{
  Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
  \For{$i =  1$ \KwTo $k$}{
      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
     }
  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
  Set $m = m + 1$\;
 }
 \caption{Gradient Ascent for SVB}
  \label{alg:algorithm2}
\end{algorithm}

\subsection{SVB Extensions}

\textbf{1. AdaGrad} \newline

If the gradient is small, we want the learning rate, $\rho_t$, to be large, and if the gradient is large, we want the learning rate to be small. AdaGrad \citep{Duchi2011} is a method of setting the learning rate to dynamically adapt as the values of the calculated gradients changes. Further, AdaGrad allows the step size of each element of $\lambda$ to move independently of the other elements of $\lambda$.

If $\nabla_{\lambda}^{(m)}$ is the vector of gradients at iteration $m$, define

\begin{equation}
G^{(m)} = \sum_{i = 1}^{m} \nabla_{\lambda}^{i} (\nabla_{\lambda}^{i})^T,
\end{equation}

the sum of the outer products of the gradients up to iteration $m$. Then set

\begin{equation}
\label{adagrad}
\rho^{(m)} = \eta \mbox{diag}(G^{(m)})^{-1/2}
\end{equation}

with some preset tuning parameter $\eta$. If we have $k$ elements in $\lambda$, $G^{(m)}$ is a $k \times k$ matrix so we have one element in $\rho^{(m)}$ for each element of $\lambda$. As we only need the diagonal of $G^{(m)}$, this may be calculated in linear time. \newline

\textbf{2. Reparameterisation} \newline

The reparameterisation trick was popularised following \citet{Kingma2013} and allows the gradient of the ELBO to be simplified.

Consider a parameter free auxillary distribution $q(\epsilon)$ and differentiable transformation $f(\cdot,\cdot)$ such that $\theta = f(\epsilon, \lambda)$. Examples include a location-scale transformation from a standard normal or an inverse-CDF transform from a uniform$(0, 1)$ variable. Note that

$$q_\theta(\theta | \lambda) = q_\epsilon(\epsilon) \left| \frac{d\epsilon}{d\theta} \right| $$

where the parameters that govern the transform $f$ are the same $\lambda$ parameters as in $q(\theta | \lambda)$. Now (\ref{SGA2}) becomes

\begin{align}
\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\epsilon} \nabla_{\lambda}[  q(\epsilon) \left( \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda) | \lambda) \right)] d\epsilon \nonumber \\
&=  \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda)| \lambda) \right] d\epsilon \label{RP1} \\
&= \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda)) \right] d\epsilon\nonumber
\end{align}

as the second part on the right hand side of (\ref{RP1}) is a score function with expectation equal to zero. This can be estimated using

\begin{equation}
\label{RP2}
 \mathcal{L}(q(\theta | \lambda), y) \approx  \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon_s, \lambda))) \right].
\end{equation}

with $\epsilon_s \sim q(\epsilon)$.

\citet{Kingma2015} explore how this reparameterised version can have orders of magnitude lower variance than the estimator in (\ref{SGA5}).

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density, Approximation family q}
 \KwResult{Variational Approximation}
 Initialise $\lambda$\;
 \While{Not converged}{
  Simulate $\epsilon^s$ for $s = 1, \dots S$\;
  \For{$i =  1$ \KwTo $k$}{
      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(p(f(\epsilon^s, \lambda^{(m-1)}), y))]$
     }
  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda}^{t}$\;
  Set $m = m + 1$\;
 }
 \caption{Gradient Ascent for re-parameterised SVB}
  \label{alg:algorithm3}
\end{algorithm}

$\mbox{ }$\newline \textbf{3. Control Variates} \newline

If reparameterisation is too difficult or otherwise unavailable, \citet{Paisly2012} introduce the use use of control variates to estimate the gradient with reduced variance.

Define

\begin{equation}
\label{CV1}
g(\theta, \lambda, y) =  \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda))),
\end{equation}

the function estimated in (\ref{SGA5}), then for some other function $h$, define

\begin{equation}
\label{CV2}
\hat{g}(\theta, \lambda, y) = g(\theta, \lambda, y) - a(h - \mathbb{E}(h)).
\end{equation}

Then we have that

\begin{equation}
\mathbb{E}(\hat{g}) = \mathbb{E}(g)
\end{equation}

and

\begin{equation}
\label{CV3}
\mbox{Var}(\hat{g}) = \mbox{Var}(g) + a^2 \mbox{Var}(h) - 2a\mbox{Cov}(g, h). 
\end{equation}

Solving the polynomial (\ref{CV3}) in $a$ shows that the variance of $\hat{g}$ is minimised by

\begin{equation}
\label{CV4}
\hat{a} = \mbox{Cov}(g, h)/\mbox{Var}(h).
\end{equation}

Substituting (\ref{CV4}) into (\ref{CV3}) yields

\begin{equation}
\label{CV5}
\mbox{Var}(\hat{g}) = \mbox{Var}(g) - \mbox{Cov}(g, h)^2/\mbox{Var}(h).
\end{equation}

We need to choose some function $h$ that has a large covariance with $g$. \citet{Ranganath2014} suggests that the an easy option is

\begin{equation}
\label{CV6}
h(\theta, \lambda) = \nabla_{\lambda} [\ln(q(\theta_s | \lambda))].
\end{equation}

As $h$ is a score function, it has an expectation of zero and our estimator becomes

\begin{equation}
\label{CV7}
\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)) - \hat{a}),
\end{equation}

where $\hat{a}$ is estimated from the sample variance and covariance of a subset of the $S$ Monte-Carlo draws.

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density, Approximation family q}
 \KwResult{Variational Approximation}
 Initialise $\lambda$\;
 \While{Not converged}{
  Simulate $\theta_s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
  \For{$i =  1$ \KwTo $k$}{
      Calculate $g_{\lambda_i} = 1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
      Calculate $h_{\lambda_i} = 1/S\sum_{s=1}^{S}\nabla_{\lambda_i}[\log(q(\theta^s |\lambda^{(m-1)}))]$\;
      Use a subset of $S$ to estimate $\hat{a} = \mbox{Cov}(h, g)/\mbox{Var}(g)$\;
      Calculate $\nabla_{\lambda_i} = g_{\lambda_i} - \hat{a} h_{\lambda_i}$\;
     }
  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
  Set $m = m + 1$\;
 }
 \caption{Gradient Ascent for SVB with control variates}
  \label{alg:algorithm4}
\end{algorithm}

\subsection{AR2 model example}

Consider $T$ observations from an AR(2) model conditioned on $y_1$ and $y_2$,

\begin{equation}
\label{AR2}
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t \mbox{ for } t = 3,\dots,T,
\end{equation}

where $\epsilon$ are iid $\mathcal{N}(0, \sigma^2)$.

Using the independent priors:

\begin{align}
p(\phi_1) &\sim \mathcal{N}(\bar{\phi_1}, \tau_{1}) \nonumber \\
p(\phi_2) &\sim \mathcal{N}(\bar{\phi_2}, \tau_{2}) \nonumber \\
p(\sigma^2) &\sim \mathcal{IG} (\mbox{shape} = a, \mbox{ scale} = b), \nonumber
\end{align}

we have a model with zero stationary restrictions which can easily be sampled through a Gibbs MCMC scheme. This exact posterior can be used in comparision with the results of our variational approximations.

The Gibbs conditional distributions are derived in the appendix and are

\begin{align}
p(\phi_1 | \phi_2, \sigma^2, y_{1:T}) &\propto \mathcal{N}(\mu_1^G, \gamma_1^G) \nonumber \\
p(\phi_2 | \phi_1, \sigma^2, y_{1:T}) &\propto \mathcal{N}(\mu_2^G, \gamma_2^G) \nonumber \\
p(\sigma^2 | \phi_1, \phi_2, y_{1:T}) &\propto \mathcal{IG}(shape = \alpha^G,  scale = \beta^G), \nonumber
\end{align}

where

\begin{align}
\mu_1^G &= \frac{\tau_1 (\sum_{t=3}^{T} y_t y_{t-1} - \phi_2 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \sigma^2 \bar{\phi_1}}
          {\tau_1 \sum_{t=3}^{T} y_{t-1}^2 + \sigma^2} \nonumber \\
\gamma_1^G &= \frac{\sigma^2 \tau_1}{\tau_1 \sum_{t=3}^{T} y_{t-1}^2 + \sigma^2} \nonumber \\
\mu_2^G &= \frac{\tau_2 (\sum_{t=3}^{T} y_t y_{t-2} - \phi_1 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \sigma^2 \bar{\phi_2}}
          {\tau_2 \sum_{t=3}^{T} y_{t-2}^2 + \sigma^2} \nonumber \\
\gamma_2^G &= \frac{\sigma^2 \tau_2}{\tau_2 \sum_{t=3}^{T} y_{t-2}^2 + \sigma^2} \nonumber \\
\alpha^G &= a + T/2 - 1 \nonumber \\
\beta^G &= b + 1/2 \sum_{t=3}^{T} \left((y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right).  \nonumber
\end{align}

100 data points are simulated from the model in (\ref{AR2}) with true values of $\phi_1 = 0.8, \phi_2 = 0.15$ and $\sigma^2 = 2$. These values were chosen to induce a strong dependence between the $\phi$ parameters as it is close to the non-stationary region of $\phi_1 + \phi_2 > 1$. The resulting posterior is sampled with 100,000 draws from Gibbs MCMC, discarding the first 20\% of all draws as a burn in. The resulting draws were strongly dependent so they were thinned by a factor of 20 to result in 4000 effectively independent draws. The total runtime for MCMC was 2.6 seconds. The dependence between $\phi_1$ and $\phi_2$ is apparent in figure (\ref{AR2MCMC}), so a SGA algorithm that captures this dependence should be a better approximation than a MFVB algorithm that assumes independence.

\begin{figure}
\label{AR2MCMC}
\centering
<<AR2MCMC, echo = FALSE, fig.height = 4, fig.width = 4>>=
library(ggplot2)
library(gridExtra)
draws <- read.csv("AR2thin.csv")[,-1]
colnames(draws) = c("Phi1", "Phi2", "SigmaSquared")
m1 <- ggplot(data=draws, aes(x=Phi1)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(phi[1]))
m2 <-  ggplot(data=draws, aes(x=Phi2)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(phi[2]))
m3 <-  ggplot(data=draws, aes(x=SigmaSquared)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(sigma^2))
j1 <- ggplot(data=draws, aes(x=Phi1, y = Phi2)) + geom_density2d() + theme_bw() + xlab(expression(phi[1])) + ylab(expression(phi[2]))
j2 <- ggplot(data=draws, aes(x=Phi1, y = SigmaSquared)) + geom_density2d() + theme_bw() + xlab(expression(phi[1])) + ylab(expression(sigma^2))
j3 <- ggplot(data=draws, aes(x=Phi2, y = SigmaSquared)) + geom_density2d() + theme_bw() + xlab(expression(phi[2])) + ylab(expression(sigma^2))
grid.arrange(m1, m2, m3, j1, j2, j3, ncol = 3)
@
\caption{Marginal and bivariate density plots of the MCMC draws. 100,000 draws per parameter were sampled from (\ref{AR2}) via Gibbs MCMC then reduced to 4000 each. Note the strong dependence between the two $\phi$ parameters.}
\end{figure}

The derivation of the MFVB approximation is also in the appendix and each marginal has the same form as the corresponding true conditional distribution above,
\begin{align}
q(\phi_1 | \lambda) &\propto \mathcal{N}(\mu_1^{MF}, \gamma_1^{MF}) \nonumber \\
q(\phi_2 | \lambda) &\propto \mathcal{N}(\mu_2^{MF}, \gamma_2^{MF}) \nonumber \\
q(\sigma^2 | \lambda) &\propto \mathcal{IG}(shape = \alpha^{MF},  scale = \beta^{MF}), \nonumber
\end{align}

where the parameterising vector $\lambda = (\mu_1^{MF}, \gamma_1^{MF}, \mu_2^{MF}, \gamma_2^{MF}, \alpha^{MF}, \beta^{MF})'$.

The MVFB derivation yields the following updating equations for $\lambda$,

\begin{align}
\mu_1^{MF} &= \frac{\alpha^{MF}/\beta^{MF} (\sum_{t=3}^{T} y_t y_{t-1} - \mu_2 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \tau_1^{-1} \bar{\phi_1}}{\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-1}^2  +  \tau_1^{-1}} \nonumber \\
\gamma_1^{MF} &= \left(\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-1}^2  +  \tau_1^{-1} \right)^{-1} \nonumber \\
\mu_2^{MF} &= \frac{\alpha^{MF}/\beta^{MF} (\sum_{t=3}^{T} y_t y_{t-2} - \mu_1 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \tau_2^{-1} \bar{\phi_2}}{\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-2}^2  +  \tau_2^{-1}} \nonumber \\
\gamma_1^{MF} &= \left(\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-2}^2  +  \tau_2^{-1} \right)^{-1} \nonumber \\
\alpha^{MF} &=  a + T/2 - 1  \nonumber \\
\beta^{MF} &= b + 1/2 \sum_{t=3}^{T} \left((y_t^2 - ((\mu_1^{MF})^2 + \gamma^{MF}_1) y_{t-1}^2 - ((\mu_2^{MF})^2 + \gamma_2^{MF}) y_{t-2}^2\right) \nonumber \\
&+ 2 \sum_{t=3}^{T} \left(\mu^{MF}_1 \mu^{MF}_2 y_{t-1}y_{t-2} - \mu^{MF}_1 y_{t} y_{t-1} - \mu^{MF}_2 y_t y_{t-2}\right) \nonumber.
\end{align}

The forms of each component are similar to the conditionals derived for Gibbs sampling, with changes coming from linearising the expressions before taking expectations. Notice that all components of $\lambda$ except $\alpha^{MF}$ depend on each other through the expectations in (\ref{mf3}).

The SGA algorithm requires the specification of the approximating family so we make the following assumptions:
\begin{itemize}
\item The optimal marginal distribution for the dependent SGA approximation is the same as the independent MFVB approximation.
\item There is no dependence between either $\phi$ parameter and $\sigma^2$.
\item There is eliptical dependence between $\phi_1$ and $\phi_2$.
\end{itemize}

To determine the form of dependence between $\phi_1$ and $\phi_2$ we fit elliptical copulas to the MCMC draws and determined that the gaussian copula is the best fit by BIC. Gaussian marginals for each $\phi$ and a gaussian copula leads to $\phi_1$ and $\phi_2$ having a bivariate gaussian distribution.

Hence we have the following form for the SGA approximations,

\begin{align}
q(\phi_1, \phi_2) &\sim \mathcal{N}(\mu, \Sigma) \nonumber \\
q(\sigma^2) &\sim \mathcal{IG}(\alpha, \beta). \nonumber
\end{align}

As each parameter in $\Sigma$ can move independently in the SGA algorithm, we instead use the lower triangluar decomposition $L L^T = \Sigma$. This results in the vector $\lambda = (\mu_1, \mu_2, L_{11}, L_{21}, L_{22}, \alpha, \beta)'$, where

\begin{itemize}
\item $\mu_1$: The mean of $\phi_1$,
\item $\mu_2$: The mean of $\phi_2$,
\item $L_{11}, L_{21}, L_{22}$, the three components of $L$, the lower triangular decomposition of $\Sigma$,
\item $\alpha, \beta$, the shape and rate parameters of an Inverse Gamma distribution for $\sigma^2$.
\end{itemize}

The diagonal of $L$ is not forced to be positive so the decomposition is not invariant to sign changes.

We compare three forms of the SGA algorithm, the original algorithm described by \ref{alg:algorithm2}, the reparameterised version described by \ref{alg:algorithm3} and the control variate version described by \ref{alg:algorithm4}. Each version is implemented with $\rho^(m)$ using AdaGrad as in (\ref{adagrad}.

The reparameterised SGA algorithm requires an auxillay distribution $q(\epsilon)$ and transformation. We use $(\epsilon_1, \epsilon_2)' \sim \mathcal{N}(0, I)$ and $\epsilon_3 \sim \mathcal{U}(0, 1)$. Then make the transformation

\begin{align}
\phi_1 &= \mu_1 + L_11 \epsilon_1 \nonumber \\
\phi_2 &= \mu_2 + L_21 \epsilon_1 + L_22 \epsilon_2 \nonumber \\
\sigma^2 &= Q^{-1}(\epsilon_3, \alpha, \beta) \nonumber
\end{align}

where $Q^{-1}(\epsilon_3, \alpha, \beta)$ is the Inverse Gamma distribution inverse cdf function with shape $\alpha$ and scale $\beta$.

We summarise each algorithm below in table \ref{table} by its estimated parameters, converged value of the ELBO and runtime. Mean Field algorithm runs extremely quickly but can not estimate the dependence between paramters so the converged ELBO is lower than the alternatives, except for the original SGA algorithm. This estimator is extremely noisy and required a value of $S$ of $200$, which slowed computation but it still converged to a lower optima than the variance reduced algorithms. The reparameterised algorithm, SGA-RP, could be run with $S = 5$ leading to its extremely fast runtime. The control variate algorithm, SGA-CV, was run with $S = 50$, and used $20$ draws per iteration to estimate $\hat{a}$.

\begin{table}
\centering
\label{AR2results}
<<AR2summary, echo = FALSE>>=
library(knitr)
summary = matrix(c(0.83, 0.10, 0.04, 0, 0.05, 50, 79.7, -77.7, "1.3 x 10^{-6}", 
                 0.63, 0.27, 0.09, -0.12, 0.05, 45.6, 81.4, -78.3, 51.1, 
                 0.78, 0.15, 0.1, -0.09, 0.05, 49.3, 78, -76.7, 4.0,
                 0.77, 0.15, 0.09, -0.09, 0.05, 48.5, 77.6, -76.9, 1.3), byrow = TRUE, ncol = 9)
colnames(summary) = c("mu1", "mu2", "L11", "L21", "L22", "alpha", "beta", "ELBO", "Runtime (seconds)")
rownames(summary) = c("Mean Field", "SGA", "SGA-RP", "SGA-CV")
kable(summary)
@
\caption{Parameter estimates, maximised ELBO and runtime of all Variational Bayes algorithms.}
\end{table}

\section{Approximate Forecast Updating}
\subsection{Motivation}
-Slow convergence of MCMC
-Bad scaling to large dimensional models and datasets
-Time/Computational Constraints
-Motivation from short forecast windows
\subsection{Model Selection}
-Overnight/Intermittent MCMC
-Infer model structure from draws
-Have easy parametric form of posterior to work with

\bibliographystyle{asa}
\bibliography{references}

\section{Appendix}
\subsection{AR2 Model}
\end{document}

### 2017-01-02 Confirmation/VB.tex
Xmas break updates

--- 2017-01-02 Confirmation/VB.Rnw
@@
-\documentclass{article}
+\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 \usepackage{natbib}
@@
 \numberwithin{equation}{section}
-
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 \begin{figure}[hb]
-<<klnorm, echo = FALSE, fig.width = 6, fig.height = 2>>=
-library(ggplot2)
-xseq = seq(-0.5, 1, 0.01)
-w1 = dnorm(xseq, 0, 0.1)
-w2 = dnorm(xseq, 0.5, 0.1)
-ggplot() + geom_line(data=NULL, aes(x=xseq, y = w1), colour = "red") + geom_line(data=NULL, aes(x=xseq, y = w2), colour = "green") + labs(x="Support", y = "Density") + theme_bw()
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/klnorm-1}
+
+\end{knitrout}
 \caption{Red: A Normal(0, 0.1) distribution. Green: A Normal(0.5, 0.1) distribution.}
@@
 \centering
-<<AR2MCMC, echo = FALSE, fig.height = 4, fig.width = 4>>=
-library(ggplot2)
-library(gridExtra)
-draws <- read.csv("AR2thin.csv")[,-1]
-colnames(draws) = c("Phi1", "Phi2", "SigmaSquared")
-m1 <- ggplot(data=draws, aes(x=Phi1)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(phi[1]))
-m2 <-  ggplot(data=draws, aes(x=Phi2)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(phi[2]))
-m3 <-  ggplot(data=draws, aes(x=SigmaSquared)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(sigma^2))
-j1 <- ggplot(data=draws, aes(x=Phi1, y = Phi2)) + geom_density2d() + theme_bw() + xlab(expression(phi[1])) + ylab(expression(phi[2]))
-j2 <- ggplot(data=draws, aes(x=Phi1, y = SigmaSquared)) + geom_density2d() + theme_bw() + xlab(expression(phi[1])) + ylab(expression(sigma^2))
-j3 <- ggplot(data=draws, aes(x=Phi2, y = SigmaSquared)) + geom_density2d() + theme_bw() + xlab(expression(phi[2])) + ylab(expression(sigma^2))
-grid.arrange(m1, m2, m3, j1, j2, j3, ncol = 3)
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\includegraphics[width=\maxwidth]{figure/AR2MCMC-1}
+
+\end{knitrout}
 \caption{Marginal and bivariate density plots of the MCMC draws.
@@
 \label{AR2results}
-<<AR2summary, echo = FALSE>>=
-library(knitr)
-summary = matrix(c(0.83, 0.10, 0.04, 0, 0.05, 50, 79.7, -77.7, "1.3 x 10^{-6}", 
-                 0.63, 0.27, 0.09, -0.12, 0.05, 45.6, 81.4, -78.3, 51.1, 
-                 0.78, 0.15, 0.1, -0.09, 0.05, 49.3, 78, -76.7, 4.0,
-                 0.77, 0.15, 0.09, -0.09, 0.05, 48.5, 77.6, -76.9, 1.3), byrow = TRUE, ncol = 9)
-colnames(summary) = c("mu1", "mu2", "L11", "L21", "L22", "alpha", "beta", "ELBO", "Runtime (seconds)")
-rownames(summary) = c("Mean Field", "SGA", "SGA-RP", "SGA-CV")
-kable(summary)
-@
+\begin{knitrout}
+\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
+\begin{tabular}{l|l|l|l|l|l|l|l|l|l}
+\hline
+  & mu1 & mu2 & L11 & L21 & L22 & alpha & beta & ELBO & Runtime (seconds)\\
+\hline
+Mean Field & 0.83 & 0.1 & 0.04 & 0 & 0.05 & 50 & 79.7 & -77.7 & 1.3 x 10\textasciicircum{}\{-6\}\\
+\hline
+SGA & 0.63 & 0.27 & 0.09 & -0.12 & 0.05 & 45.6 & 81.4 & -78.3 & 51.1\\
+\hline
+SGA-RP & 0.78 & 0.15 & 0.1 & -0.09 & 0.05 & 49.3 & 78 & -76.7 & 4\\
+\hline
+SGA-CV & 0.77 & 0.15 & 0.09 & -0.09 & 0.05 & 48.5 & 77.6 & -76.9 & 1.3\\
+\hline
+\end{tabular}
+
+\end{knitrout}
 \caption{Parameter estimates, maximised ELBO and runtime of all Variational Bayes algorithms.}

### 2017-01-02 Confirmation/bayes.Rnw
Xmas break updates

--- 2017-01-02 Confirmation/VB.Rnw
@@
 \usepackage{array}
-\usepackage[hidelinks]{hyperref}
+\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
 \usepackage[font=small,skip=5pt]{caption}
@@
 \usepackage{amsmath}
-\usepackage{dsfont}
 \usepackage[]{algorithm2e}
@@
 \begin{document}
-
-\tableofcontents
 
@@
 
-\section{Variational Bayes}
-
-\subsection{Estimation as an optimisation problem}
-
-In the event that our dataset is too large, or our model is too complex, for MCMC to converge within a reasonable time frame we must resort to approximation.
-One important question that might be asked then is \textit{`How do I measure how good my approximation is?'}.
-If two distributions come from the same family, one might consider the difference in the parameters, but it is trivial to come up with distributions with slight parameter differences and vastly different regions of probability mass.
-Consider the difference between a $\mathcal{N}(0, 0.1)$ and a $\mathcal{N}(0.5, 0.1)$ distribution, the change in mean is only $0.5$ but there is little overlap in the densities.
-Further, large values for the variance allow large changes in mean with very little change in the resulting densities.
-If the distributions come from different families it is almost impossible to use parameters to measure the degree of closeness between two distributions.
-Instead we must move from the parameter space to the distribution space, where one measure of the difference between distributions $p(\theta)$ and $q(\theta)$ is the Kullback-Leibler (KL) divergence \citep{Kullback1951}, defined by
-
-\begin{equation}
-\label{KL-def}
-KL[q(\theta)||p(\theta)] = \int q(\theta) \ln \left( \frac{q(\theta)}{p(\theta)}\right) d\theta.
-\end{equation}
-
-KL divergence is a non-negative, assymetric measure of the difference between $p(\theta)$ and $q(\theta)$ that will equal zero if and only if $p(\theta) = q(\theta)$ almost everywhere.
-It has origins in information theory, and can be interpreted as: \textit{`Given that I know $p(\theta)$ for some $\theta \in \Theta$, how much extra information is required, on average, to know the value of $q(\theta)$?'} There are examples of the literature of approximations with other measures of divergence, such as \citet{Minka2001} introducting Expectation Propogation, which minimises the reverse measure $KL[p(\theta)||q(\theta)]$, which was extended to Power-EP in \citet{Minka2004}, which aims to minimisea the more general $\alpha-\mbox{divergence}$ \citep{Amari1985}.
-\vspace{3mm}
-
-\begin{figure}[hb]
-<<klnorm, echo = FALSE, fig.width = 6, fig.height = 2>>=
-library(ggplot2)
-xseq = seq(-0.5, 1, 0.01)
-w1 = dnorm(xseq, 0, 0.1)
-w2 = dnorm(xseq, 0.5, 0.1)
-ggplot() + geom_line(data=NULL, aes(x=xseq, y = w1), colour = "red") + geom_line(data=NULL, aes(x=xseq, y = w2), colour = "green") + labs(x="Support", y = "Density") + theme_bw()
-@
-\caption{Red: A Normal(0, 0.1) distribution. Green: A Normal(0.5, 0.1) distribution.}
-\end{figure}
-
-\newpage
-
-Variational Bayes aims to find the approximating distribution, $q(\theta | \lambda)$, that minimises $KL[q(\theta | \lambda)||p(\theta |y)]$.
-This approximating distribution is parameterised by a $\lambda$ vector, each element of which may be a function of $y$, the hyperparameters of the prior distribution, and even of other elements in $\lambda$.
-The optimal approximation, in the sense of having the minimum KL divergence with the posterior, can be derived by beginning with the unknown constant, $\ln(p(y))$.
-\begin{align}
-\ln(p(y)) &= \int_{\theta} q(\theta|\lambda) \ln(p(y)) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \right) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \frac{q(\theta|\lambda)}{q(\theta|\lambda)}\right) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) + \ln \left( \frac{p(y, \theta)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) d\theta + \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
-&= KL [q(\theta||\lambda) || p(\theta | y)] + \mathcal{L}(q(\theta | \lambda), y) \nonumber
-\end{align}
-
-where
-
-\begin{align}
-\mathcal{L}(q(\theta|\lambda), y) &= \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta|\lambda)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
-&=  \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta. \label{ELBO}
-\end{align}
-
-Hence we have that
-\begin{equation}
-\label{VBKL}
-KL [q(\theta|\lambda) || p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q(\theta | \lambda), y).
-\end{equation}
-
-From (\ref{VBKL}) it is clear that by maximising $\mathcal{L}(q(\theta | \lambda), y)$ with respect to $q(\theta|\lambda)$ the KL divergence between the unknown true posterior $p(\theta|y)$ and the approximating distribution $q(\theta|\lambda)$ will be minimised.
-Hence the Variational Bayes solution will be to maximise the Evidence Lower Bound (ELBO), $\mathcal{L}(q(\theta|\lambda), y)$, in (\ref{ELBO}).
-
-\subsection{Mean Field Variational Bayes}
-
-Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and uses the assumption that the approximation dististribution factorises, 
-\begin{equation}
-\label{mf1}
-q(\theta|\lambda) = \prod_i q_i(\theta_i | \lambda_i).
-\end{equation}
-
-This assumption is known as the Mean Field assumption and is has been widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}).
-Each component $\theta_i$ may be a scalar or a vector, and $\lambda_i$ is the component of the $\lambda$ vector that parameterises the relevant factor $q_i(\theta_i |\lambda_i)$.
-From here, we will use the shorthand notation that $q_i = q_i(\theta_i|\lambda_i)$ and $q_{\setminus i} = \prod_{j\neq i}q_j$.
-Maximising the ELBO with respect to $q_i$ is analytically involved but computationally simple, as (\ref{ELBO}) can be expressed as a function of $q_i$ only and then maximised with respect to each $q_i$ individually.
-
-\begin{align}
-\mathcal{L}(q_i) &= \int_{\theta} q_i q_{\setminus i} \ln (p(y, \theta)) d\theta - \int_{\theta} q_{i}q_{\setminus i} \ln (q_{i}q_{\setminus i}) d\theta \nonumber \\
-&= \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} ( \ln (p(y, \theta)) - \ln(q_{i})) d\theta_i d\theta_{\setminus i} - \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} \ln (q_{\setminus i}) d\theta_i d\theta_{\setminus i}  \nonumber \\
-&= \int_{\theta_{i}} q_{i} \left( \int_{\theta_{\setminus i}} q_{\setminus i} \ln (p(y, \theta )) d\theta_{\setminus i} - \ln(q_i) \right) d\theta_i \nonumber \\
-&- \int_{\theta_{\setminus i}} q_{\setminus i} \ln(q_{\setminus i}) d\theta_{\setminus i} \label{mf2}\\
-&= \int_{\theta_{1}} q_{i} \ln \left( \frac{\exp(  \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])}{q_i} \right) d\theta_i + c \nonumber \\
-&= -KL (q_i || \exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])) + c \label{mf2.2}
-\end{align}
-
-as the rightmost component of (\ref{mf2}) is constant with respect to $q_i$.
-
-We can see that by holding $q_{\setminus_i}$ fixed, we can maximise the ELBO with respect to $q_i$ by minimising the KL Divergence between $q_i$ and $\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])$ in (\ref{mf2.2}).
-This is easily done by setting
-
-\begin{equation}
-\label{mf3}
-q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))]).
-\end{equation}
-
-If the likelihood and prior for $\theta_i$ form an exponential family conjugate pair, then this maximisation proceeds by keeping the distributional family constant and simply updating the parameters $\lambda_i$.
-Otherwise, we can find the optimal distributional family for any $q_i$ by taking the log-joint density and then ignoring any term that doesn't depend on $\theta_i$.
-Next, take the expectation of any remaining $\theta_{j \neq i}$ and then match to the kernel of a known distribution.
-If it does not match a known distribution, we may need to make a further approximation $\tilde{q_i(\theta_i|\lambda_i)}$ which has a recognizable distribution.
-One method as used in \citet{Friston2006} uses a Laplace approximation for the otherwise unrecognizable $q_i(\theta_i | \lambda_i)$ with reasonable results.
-
-These steps are almost the same as those used to find a conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, with dependence on other parameters replaced by their expectations.
-Hence, the optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family.
-Deriving the updating algorithm for MFVB takes more analytical effort than finding the conditionals for a Gibbs MCMC scheme, as we must then find the values of each $\lambda_i$ implied by (\ref{mf3}), which are often a function of $\lambda_{j \neq i}$ through the expectations.
-\vspace{5mm}
-
-Given these optimal distributional families, an algorithm that cycles through all of the $\lambda_i$ and sets each to its maximising value is known as a coordiante ascent algorithm.
-The algorithm must continuously iterate between each parameter until $\mathcal{L}(q(\theta | \lambda), y)$ converges with some pre-defined threshold noting that the $\lambda$ may depend on each other.
-If $\lambda$ has dimension $k$, the algorithm follows below.
-
-\vspace{2mm}
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density}
- \KwResult{Mean Field Approximation}
- Initialise $\lambda$ randomly\;
- \While{Not converged}{
-  \For{$i =  1$ \KwTo $k$}{
-      Hold $\lambda_{j \neq i}$ fixed\;
-      Use (\ref{mf3}) to match $q(\theta_i|\lambda_i)$ to a known distribution\;
-      Update $\lambda_i$ using the most recent values of $\lambda_{j \neq i}$\;
-     }
- }
- \caption{Coordinate Ascent for MFVB}
-  \label{alg:algorithm1}
-\end{algorithm}
-
-\subsection{Stochastic Gradient Ascent}
-
-In many cases, the posterior is too complex to be captured by a factorising approximation and we desire our distribution $q(\theta | \lambda)$ to capture dependence between parameters.
-In this case, the easy maximisation in MFVB is unavailable, and we must resort to what is called Stochastic Variational Bayes (SVB) using a gradient ascent algorithm developed by \citet{Paisley2012} and \citet{Hoffman2013}.
-
-To maximise the function $\mathcal{L}(q(\theta | \lambda), y)$ we can take the derivative of $\mathcal{L}(q(\theta | \lambda), y)$ with respect to $\lambda$ and use the updating step:
-
-\begin{equation}
-\label{SGA1}
-\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m-1)}), y),
-\end{equation}
-
-where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m-1)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m-1)}), y)$ with respect to each element of $\lambda$.
-This update requires initial values $\lambda^{(0)}$ and some sequence $\rho^{(m)}$.
-If $\rho^{(m)}$ is chosen to satisfy the following conditions, it is a Robbins-Monro sequence and the algorithm is guaranteed to converge to a local maximum.
-
-\begin{align}
-\lim_{m \xrightarrow[]{} \infty} \rho^{(m)} &= 0 \nonumber \\
-\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \nonumber \\
-\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty. \nonumber
-\end{align}
-
-Whilst a global maximum is desired, the ELBO can have a problem specific shape that makes finding the global maximum extremely difficult, as we do not know how many stationary points exist.
-The dimension of the ELBO is the dimension of the $\lambda$ vector, which is often much greater than the dimension of the parameters $\theta$ so a grid search is suspect to the curse of dimensionality.
-To alleiviate this problem, we can start the algorithm at a range of initial values choose the converged value with the highest value of the ELBO.
-
-SVB does not provide the family of the the optimal approximating distribution, unlike under the mean field assumption.
-To run SVB we may need to try many approximating distributions and then choose the $q(\theta | \lambda)$ that has the highest ELBO, and hence lowest KL divergence to the true posterior.
-We must restrict the approximation to distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable.
-In this case, we can calculate derivatives of the ELBO.
-
-\begin{align}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \nabla_{\lambda} \int_{\theta} q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
-&=   \int_{\theta} \nabla_{\lambda}[  q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)] d\theta \label{SGA2} \\
-&=  \int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta)) - \ln(q(\theta | \lambda))] d\theta \nonumber \\ 
-&+ \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \label{SGA3}
-\end{align}
-
-Note that the first integral on the right hand side of (\ref{SGA3}) equals 0 as
-
-\begin{equation}
-\int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta))] d\theta = 0
-\end{equation}
-
-and
-
-\begin{align}
-\int_{\theta} - q(\theta | \lambda) \nabla_{\lambda}[ \ln(q(\theta | \lambda))] d\theta &= - \int_{\theta} \frac{\nabla_{\lambda}q(\theta | \lambda)}{q(\theta | \lambda)} q(\theta | \lambda) d\theta \nonumber \\
-&=  -  \nabla_{\lambda}\int_{\theta} q(\theta | \lambda) d\theta \nonumber \\
-&=  0 \nonumber
-\end{align}
-
-Then we continue from (\ref{SGA3}) with
-
-\begin{align}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
-&=   \int_{\theta} q(\theta | \lambda)  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)d\theta \nonumber \\
-&= E_q \left(  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) \right). \label{SGA4}
-\end{align}
-
-We can compute a Monte-Carlo estimate of (\ref{SGA4}) by
-
-\begin{equation}
-\label{SGA5}
-\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda)) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)))
-\end{equation}
-
-where $\theta_s$, $s = 1, \dots S$, are simulated from the most recent iteration of  $q(\theta | \lambda)$.
-
-As the distribution $q(\theta | \lambda)$ is specified by the user, the main restriction on the use of SVB is that the log-joint density $\ln(p(y, \theta))$ is able to be evaluated.
-Running an SVB algorithm involves very little analytical effort unlike MFVB, most of the difficulty is calculating the score function of the approximating distribution,
-$\nabla_{\lambda} [\ln(q(\theta_s | \lambda))]$. But as this is part of the approximation and not the true model the code to run this could potentially be re-used in many different models.
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
-     }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
-  Set $m = m + 1$\;
- }
- \caption{Gradient Ascent for SVB}
-  \label{alg:algorithm2}
-\end{algorithm}
-
-\subsection{SVB Extensions}
-
-\textbf{1. AdaGrad} \newline
-
-If the gradient is small, we want the learning rate, $\rho_t$, to be large, and if the gradient is large, we want the learning rate to be small.
-AdaGrad \citep{Duchi2011} is a method of setting the learning rate to dynamically adapt as the values of the calculated gradients changes.
-Further, AdaGrad allows the step size of each element of $\lambda$ to move independently of the other elements of $\lambda$.
-
-If $\nabla_{\lambda}^{(m)}$ is the vector of gradients at iteration $m$, define
-
-\begin{equation}
-G^{(m)} = \sum_{i = 1}^{m} \nabla_{\lambda}^{i} (\nabla_{\lambda}^{i})^T,
-\end{equation}
-
-the sum of the outer products of the gradients up to iteration $m$. Then set
-
-\begin{equation}
-\label{adagrad}
-\rho^{(m)} = \eta \mbox{diag}(G^{(m)})^{-1/2}
-\end{equation}
-
-with some preset tuning parameter $\eta$.
-If we have $k$ elements in $\lambda$, $G^{(m)}$ is a $k \times k$ matrix so we have one element in $\rho^{(m)}$ for each element of $\lambda$.
-As we only need the diagonal of $G^{(m)}$, this may be calculated in linear time.
-\newline
-
-\textbf{2. Reparameterisation} \newline
-
-The reparameterisation trick was popularised following \citet{Kingma2013} and allows the gradient of the ELBO to be simplified.
-
-Consider a parameter free auxillary distribution $q(\epsilon)$ and differentiable transformation $f(\cdot,\cdot)$ such that $\theta = f(\epsilon, \lambda)$.
-Examples include a location-scale transformation from a standard normal or an inverse-CDF transform from a uniform$(0, 1)$ variable.
-Note that
-
-$$q_\theta(\theta | \lambda) = q_\epsilon(\epsilon) \left| \frac{d\epsilon}{d\theta} \right| $$
-
-where the parameters that govern the transform $f$ are the same $\lambda$ parameters as in $q(\theta | \lambda)$. Now (\ref{SGA2}) becomes
-
-\begin{align}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\epsilon} \nabla_{\lambda}[  q(\epsilon) \left( \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda) | \lambda) \right)] d\epsilon \nonumber \\
-&=  \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda)| \lambda) \right] d\epsilon \label{RP1} \\
-&= \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda)) \right] d\epsilon\nonumber
-\end{align}
-
-as the second part on the right hand side of (\ref{RP1}) is a score function with expectation equal to zero. This can be estimated using
-
-\begin{equation}
-\label{RP2}
- \mathcal{L}(q(\theta | \lambda), y) \approx  \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon_s, \lambda))) \right].
-\end{equation}
-
-with $\epsilon_s \sim q(\epsilon)$.
-
-\citet{Kingma2015} explore how this reparameterised version can have orders of magnitude lower variance than the estimator in (\ref{SGA5}).
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\epsilon^s$ for $s = 1, \dots S$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(p(f(\epsilon^s, \lambda^{(m-1)}), y))]$
-     }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda}^{t}$\;
-  Set $m = m + 1$\;
- }
- \caption{Gradient Ascent for re-parameterised SVB}
-  \label{alg:algorithm3}
-\end{algorithm}
-
-$\mbox{ }$\newline \textbf{3. Control Variates} \newline
-
-If reparameterisation is too difficult or otherwise unavailable, \citet{Paisly2012} introduce the use use of control variates to estimate the gradient with reduced variance.
-
-Define
-
-\begin{equation}
-\label{CV1}
-g(\theta, \lambda, y) =  \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda))),
-\end{equation}
-
-the function estimated in (\ref{SGA5}), then for some other function $h$, define
-
-\begin{equation}
-\label{CV2}
-\hat{g}(\theta, \lambda, y) = g(\theta, \lambda, y) - a(h - \mathbb{E}(h)).
-\end{equation}
-
-Then we have that
-
-\begin{equation}
-\mathbb{E}(\hat{g}) = \mathbb{E}(g)
-\end{equation}
-
-and
-
-\begin{equation}
-\label{CV3}
-\mbox{Var}(\hat{g}) = \mbox{Var}(g) + a^2 \mbox{Var}(h) - 2a\mbox{Cov}(g, h). 
-\end{equation}
-
-Solving the polynomial (\ref{CV3}) in $a$ shows that the variance of $\hat{g}$ is minimised by
-
-\begin{equation}
-\label{CV4}
-\hat{a} = \mbox{Cov}(g, h)/\mbox{Var}(h).
-\end{equation}
-
-Substituting (\ref{CV4}) into (\ref{CV3}) yields
-
-\begin{equation}
-\label{CV5}
-\mbox{Var}(\hat{g}) = \mbox{Var}(g) - \mbox{Cov}(g, h)^2/\mbox{Var}(h).
-\end{equation}
-
-We need to choose some function $h$ that has a large covariance with $g$. \citet{Ranganath2014} suggests that the an easy option is
-
-\begin{equation}
-\label{CV6}
-h(\theta, \lambda) = \nabla_{\lambda} [\ln(q(\theta_s | \lambda))].
-\end{equation}
-
-As $h$ is a score function, it has an expectation of zero and our estimator becomes
-
-\begin{equation}
-\label{CV7}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)) - \hat{a}),
-\end{equation}
-
-where $\hat{a}$ is estimated from the sample variance and covariance of a subset of the $S$ Monte-Carlo draws.
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\theta_s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $g_{\lambda_i} = 1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
-      Calculate $h_{\lambda_i} = 1/S\sum_{s=1}^{S}\nabla_{\lambda_i}[\log(q(\theta^s |\lambda^{(m-1)}))]$\;
-      Use a subset of $S$ to estimate $\hat{a} = \mbox{Cov}(h, g)/\mbox{Var}(g)$\;
-      Calculate $\nabla_{\lambda_i} = g_{\lambda_i} - \hat{a} h_{\lambda_i}$\;
-     }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
-  Set $m = m + 1$\;
- }
- \caption{Gradient Ascent for SVB with control variates}
-  \label{alg:algorithm4}
-\end{algorithm}
-
-\subsection{AR2 model example}
-
-Consider $T$ observations from an AR(2) model conditioned on $y_1$ and $y_2$,
-
-\begin{equation}
-\label{AR2}
-y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t \mbox{ for } t = 3,\dots,T,
-\end{equation}
-
-where $\epsilon$ are iid $\mathcal{N}(0, \sigma^2)$.
-
-Using the independent priors:
-
-\begin{align}
-p(\phi_1) &\sim \mathcal{N}(\bar{\phi_1}, \tau_{1}) \nonumber \\
-p(\phi_2) &\sim \mathcal{N}(\bar{\phi_2}, \tau_{2}) \nonumber \\
-p(\sigma^2) &\sim \mathcal{IG} (\mbox{shape} = a, \mbox{ scale} = b), \nonumber
-\end{align}
-
-we have a model with zero stationary restrictions which can easily be sampled through a Gibbs MCMC scheme.
-This exact posterior can be used in comparision with the results of our variational approximations.
-
-The Gibbs conditional distributions are derived in the appendix and are
-
-\begin{align}
-p(\phi_1 | \phi_2, \sigma^2, y_{1:T}) &\propto \mathcal{N}(\mu_1^G, \gamma_1^G) \nonumber \\
-p(\phi_2 | \phi_1, \sigma^2, y_{1:T}) &\propto \mathcal{N}(\mu_2^G, \gamma_2^G) \nonumber \\
-p(\sigma^2 | \phi_1, \phi_2, y_{1:T}) &\propto \mathcal{IG}(shape = \alpha^G,  scale = \beta^G), \nonumber
-\end{align}
-
-where
-
-\begin{align}
-\mu_1^G &= \frac{\tau_1 (\sum_{t=3}^{T} y_t y_{t-1} - \phi_2 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \sigma^2 \bar{\phi_1}}
-          {\tau_1 \sum_{t=3}^{T} y_{t-1}^2 + \sigma^2} \nonumber \\
-\gamma_1^G &= \frac{\sigma^2 \tau_1}{\tau_1 \sum_{t=3}^{T} y_{t-1}^2 + \sigma^2} \nonumber \\
-\mu_2^G &= \frac{\tau_2 (\sum_{t=3}^{T} y_t y_{t-2} - \phi_1 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \sigma^2 \bar{\phi_2}}
-          {\tau_2 \sum_{t=3}^{T} y_{t-2}^2 + \sigma^2} \nonumber \\
-\gamma_2^G &= \frac{\sigma^2 \tau_2}{\tau_2 \sum_{t=3}^{T} y_{t-2}^2 + \sigma^2} \nonumber \\
-\alpha^G &= a + T/2 - 1 \nonumber \\
-\beta^G &= b + 1/2 \sum_{t=3}^{T} \left((y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right).  \nonumber
-\end{align}
-
-100 data points are simulated from the model in (\ref{AR2}) with true values of $\phi_1 = 0.8, \phi_2 = 0.15$ and $\sigma^2 = 2$.
-These values were chosen to induce a strong dependence between the $\phi$ parameters as it is close to the non-stationary region of $\phi_1 + \phi_2 > 1$.
-The resulting posterior is sampled with 100,000 draws from Gibbs MCMC, discarding the first 20\% of all draws as a burn in.
-The resulting draws were strongly dependent so they were thinned by a factor of 20 to result in 4000 effectively independent draws.
-The total runtime for MCMC was 2.6 seconds.
-The dependence between $\phi_1$ and $\phi_2$ is apparent in figure (\ref{AR2MCMC}), so a SGA algorithm that captures this dependence should be a better approximation than a MFVB algorithm that assumes independence.
-
-\begin{figure}
-\label{AR2MCMC}
-\centering
-<<AR2MCMC, echo = FALSE, fig.height = 4, fig.width = 4>>=
-library(ggplot2)
-library(gridExtra)
-draws <- read.csv("AR2thin.csv")[,-1]
-colnames(draws) = c("Phi1", "Phi2", "SigmaSquared")
-m1 <- ggplot(data=draws, aes(x=Phi1)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(phi[1]))
-m2 <-  ggplot(data=draws, aes(x=Phi2)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(phi[2]))
-m3 <-  ggplot(data=draws, aes(x=SigmaSquared)) + geom_density() + theme_bw() + ylab(NULL) + xlab(expression(sigma^2))
-j1 <- ggplot(data=draws, aes(x=Phi1, y = Phi2)) + geom_density2d() + theme_bw() + xlab(expression(phi[1])) + ylab(expression(phi[2]))
-j2 <- ggplot(data=draws, aes(x=Phi1, y = SigmaSquared)) + geom_density2d() + theme_bw() + xlab(expression(phi[1])) + ylab(expression(sigma^2))
-j3 <- ggplot(data=draws, aes(x=Phi2, y = SigmaSquared)) + geom_density2d() + theme_bw() + xlab(expression(phi[2])) + ylab(expression(sigma^2))
-grid.arrange(m1, m2, m3, j1, j2, j3, ncol = 3)
-@
-\caption{Marginal and bivariate density plots of the MCMC draws.
-100,000 draws per parameter were sampled from (\ref{AR2}) via Gibbs MCMC then reduced to 4000 each.
-Note the strong dependence between the two $\phi$ parameters.}
-\end{figure}
-
-The derivation of the MFVB approximation is also in the appendix and each marginal has the same form as the corresponding true conditional distribution above,
-\begin{align}
-q(\phi_1 | \lambda) &\propto \mathcal{N}(\mu_1^{MF}, \gamma_1^{MF}) \nonumber \\
-q(\phi_2 | \lambda) &\propto \mathcal{N}(\mu_2^{MF}, \gamma_2^{MF}) \nonumber \\
-q(\sigma^2 | \lambda) &\propto \mathcal{IG}(shape = \alpha^{MF},  scale = \beta^{MF}), \nonumber
-\end{align}
-
-where the parameterising vector $\lambda = (\mu_1^{MF}, \gamma_1^{MF}, \mu_2^{MF}, \gamma_2^{MF}, \alpha^{MF}, \beta^{MF})'$.
-
-The MVFB derivation yields the following updating equations for $\lambda$,
-
-\begin{align}
-\mu_1^{MF} &= \frac{\alpha^{MF}/\beta^{MF} (\sum_{t=3}^{T} y_t y_{t-1} - \mu_2 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \tau_1^{-1} \bar{\phi_1}}{\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-1}^2  +  \tau_1^{-1}} \nonumber \\
-\gamma_1^{MF} &= \left(\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-1}^2  +  \tau_1^{-1} \right)^{-1} \nonumber \\
-\mu_2^{MF} &= \frac{\alpha^{MF}/\beta^{MF} (\sum_{t=3}^{T} y_t y_{t-2} - \mu_1 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \tau_2^{-1} \bar{\phi_2}}{\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-2}^2  +  \tau_2^{-1}} \nonumber \\
-\gamma_1^{MF} &= \left(\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-2}^2  +  \tau_2^{-1} \right)^{-1} \nonumber \\
-\alpha^{MF} &=  a + T/2 - 1  \nonumber \\
-\beta^{MF} &= b + 1/2 \sum_{t=3}^{T} \left((y_t^2 - ((\mu_1^{MF})^2 + \gamma^{MF}_1) y_{t-1}^2 - ((\mu_2^{MF})^2 + \gamma_2^{MF}) y_{t-2}^2\right) \nonumber \\
-&+ 2 \sum_{t=3}^{T} \left(\mu^{MF}_1 \mu^{MF}_2 y_{t-1}y_{t-2} - \mu^{MF}_1 y_{t} y_{t-1} - \mu^{MF}_2 y_t y_{t-2}\right) \nonumber.
-\end{align}
-
-The forms of each component are similar to the conditionals derived for Gibbs sampling, with changes coming from linearising the expressions before taking expectations.
-Notice that all components of $\lambda$ except $\alpha^{MF}$ depend on each other through the expectations in (\ref{mf3}).
-
-The SGA algorithm requires the specification of the approximating family so we make the following assumptions:
-\begin{itemize}
-\item The optimal marginal distribution for the dependent SGA approximation is the same as the independent MFVB approximation.
-\item There is no dependence between either $\phi$ parameter and $\sigma^2$.
-\item There is eliptical dependence between $\phi_1$ and $\phi_2$.
-\end{itemize}
-
-To determine the form of dependence between $\phi_1$ and $\phi_2$ we fit elliptical copulas to the MCMC draws and determined that the gaussian copula is the best fit by BIC.
-Gaussian marginals for each $\phi$ and a gaussian copula leads to $\phi_1$ and $\phi_2$ having a bivariate gaussian distribution.
-
-Hence we have the following form for the SGA approximations,
-
-\begin{align}
-q(\phi_1, \phi_2) &\sim \mathcal{N}(\mu, \Sigma) \nonumber \\
-q(\sigma^2) &\sim \mathcal{IG}(\alpha, \beta). \nonumber
-\end{align}
-
-As each parameter in $\Sigma$ can move independently in the SGA algorithm, we instead use the lower triangluar decomposition $L L^T = \Sigma$.
-This results in the vector $\lambda = (\mu_1, \mu_2, L_{11}, L_{21}, L_{22}, \alpha, \beta)'$, where
-
-\begin{itemize}
-\item $\mu_1$: The mean of $\phi_1$,
-\item $\mu_2$: The mean of $\phi_2$,
-\item $L_{11}, L_{21}, L_{22}$, the three components of $L$, the lower triangular decomposition of $\Sigma$,
-\item $\alpha, \beta$, the shape and rate parameters of an Inverse Gamma distribution for $\sigma^2$.
-\end{itemize}
-
-The diagonal of $L$ is not forced to be positive so the decomposition is not invariant to sign changes.
-
-We compare three forms of the SGA algorithm, the original algorithm described by \ref{alg:algorithm2}, the reparameterised version described by \ref{alg:algorithm3} and the control variate version described by \ref{alg:algorithm4}.
-Each version is implemented with $\rho^(m)$ using AdaGrad as in (\ref{adagrad}.
-
-The reparameterised SGA algorithm requires an auxillay distribution $q(\epsilon)$ and transformation.
-We use $(\epsilon_1, \epsilon_2)' \sim \mathcal{N}(0, I)$ and $\epsilon_3 \sim \mathcal{U}(0, 1)$.
-Then make the transformation
-
-\begin{align}
-\phi_1 &= \mu_1 + L_11 \epsilon_1 \nonumber \\
-\phi_2 &= \mu_2 + L_21 \epsilon_1 + L_22 \epsilon_2 \nonumber \\
-\sigma^2 &= Q^{-1}(\epsilon_3, \alpha, \beta) \nonumber
-\end{align}
-
-where $Q^{-1}(\epsilon_3, \alpha, \beta)$ is the Inverse Gamma distribution inverse cdf function with shape $\alpha$ and scale $\beta$.
-
-We summarise each algorithm below in table \ref{table} by its estimated parameters, converged value of the ELBO and runtime.
-Mean Field algorithm runs extremely quickly but can not estimate the dependence between paramters so the converged ELBO is lower than the alternatives, except for the original SGA algorithm.
-This estimator is extremely noisy and required a value of $S$ of $200$, which slowed computation but it still converged to a lower optima than the variance reduced algorithms.
-The reparameterised algorithm, SGA-RP, could be run with $S = 5$ leading to its extremely fast runtime.
-The control variate algorithm, SGA-CV, was run with $S = 50$, and used $20$ draws per iteration to estimate $\hat{a}$.
-
-\begin{table}
-\centering
-\label{AR2results}
-<<AR2summary, echo = FALSE>>=
-library(knitr)
-summary = matrix(c(0.83, 0.10, 0.04, 0, 0.05, 50, 79.7, -77.7, "1.3 x 10^{-6}", 
-                 0.63, 0.27, 0.09, -0.12, 0.05, 45.6, 81.4, -78.3, 51.1, 
-                 0.78, 0.15, 0.1, -0.09, 0.05, 49.3, 78, -76.7, 4.0,
-                 0.77, 0.15, 0.09, -0.09, 0.05, 48.5, 77.6, -76.9, 1.3), byrow = TRUE, ncol = 9)
-colnames(summary) = c("mu1", "mu2", "L11", "L21", "L22", "alpha", "beta", "ELBO", "Runtime (seconds)")
-rownames(summary) = c("Mean Field", "SGA", "SGA-RP", "SGA-CV")
-kable(summary)
-@
-\caption{Parameter estimates, maximised ELBO and runtime of all Variational Bayes algorithms.}
-\end{table}
-
-\section{Approximate Forecast Updating}
-\subsection{Motivation}
--Slow convergence of MCMC
--Bad scaling to large dimensional models and datasets
--Time/Computational Constraints
--Motivation from short forecast windows
-\subsection{Model Selection}
--Overnight/Intermittent MCMC
--Infer model structure from draws
--Have easy parametric form of posterior to work with
-
-\bibliographystyle{asa}
-\bibliography{references}
-
-\section{Appendix}
-\subsection{AR2 Model}
 \end{document}

### 2017-01-02 Confirmation/confirmation_outline.Rnw
Xmas break updates

\documentclass{article}

\begin{document}

\tableofcontents

\section{Bayesian Inference}
\subsection{Bayes Rule}
-General Intro to Bayesian Methodology
-Denominator as an integral
-Often no analytical solutions
\subsection{Exact Bayesian Computation}
-Numerical Integration
-Importance Sampling
-MCMC
\subsection{Variational Bayes}
-KL Divergence as a transformation to distribution space
-VB Derivation
\subsection{The Mean Field Assumption}
-Mean Field Derivation
-Gives optimal family of approximating distributions
-Lots of analytical work
-Very Fast
\subsection{Stochastic Gradient Ascent}
-SGA Derivation
-Step Size
-Flexibility 
-Little analytical work
-Needs a pre-defined approximating distribution family
\subsection{Variance Reduction Techniques}
-Rao Blackwellisation
-Control Variates
-Reparameterisation
\subsection{AR2 model example}

\section{Electricity Load Forecasts}
\subsection{Motivation}
-Real data has five minute updates
-Density is of interest due to rare price spikes, so more interested in the chance load is above a certain point
-Slow convergence of MCMC / Bad scaling to large dimensional models and datasets
\subsection{Exponential Smoothing}
-Bayesian Version
-Difficulty with smoothing parameters in large models
-Infer model structure from MCMC draws where possible
\subsection{Variational Bayes Implementation}
-Posterior Distributions should not change much as more data is observed, so we can keep approximating distribution family constant
-But we might want a model that does change to make updates worthwhile - Markov Switching mechanism to be added?
\subsection{Forecasting}
-Compare to MCMC when possible

\end{document}

### 2017-01-02 Confirmation/confirmation_outline.tex
Xmas break updates

--- 2017-01-02 Confirmation/VB.tex
@@
 %% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\usepackage{natbib}
+\usepackage{color}
+\usepackage[dvipsnames,svgnames*]{xcolor}
+\usepackage{array}
+\usepackage[hidelinks]{hyperref}
+\usepackage[font=small,skip=5pt]{caption}
+\usepackage[aboveskip=2pt]{subcaption}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{dsfont}
+\usepackage[]{algorithm2e}
+\usepackage{amsthm}
+\usepackage{amsfonts}
+\usepackage{url}
+\usepackage{ulem}
+\usepackage{afterpage}
+\usepackage{bbm}
+\numberwithin{equation}{section}
 \makeatletter
@@
 \usepackage{alltt}
-\usepackage{natbib}
-\usepackage{color}
-\usepackage[dvipsnames,svgnames*]{xcolor}
-\usepackage{array}
-\usepackage[hidelinks]{hyperref}
-\usepackage[font=small,skip=5pt]{caption}
-\usepackage[aboveskip=2pt]{subcaption}
-\usepackage{graphicx}
-\usepackage{amsmath}
-\usepackage{dsfont}
-\usepackage[]{algorithm2e}
-\usepackage{amsthm}
-\usepackage{amsfonts}
-\usepackage{url}
-\usepackage{ulem}
-\usepackage{afterpage}
-\usepackage{bbm}
-\numberwithin{equation}{section}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
@@
 \tableofcontents
+\section{Introduction}
+
+For many time-series applications a point forecast of the mean of the next observation in a series, $y_{T+1}$, can be easily obtained with frequentist methods, conditioning on the forecaster's observed series, $y_{1:T}$.
+Often the forecast is supplemented with a prediction interval backed by asymptotic theory, however there is a growing demand in the literature for density forecasts (find citations, probably Hyndman) which are typically much harder to obtain.
+
+As a contrast to frequentism, the Bayesian methodology implicitly provides the entire probability density for variables of interest, in this case $p(y_{T+1} | y_{1:T})$.
+However the Bayesian approach requires solving integrals of the form
+
+\begin{equation}
+\label{forecastintegral}
+\int_\theta p(y_{T+1} | \theta) p(\theta | y_{1:T}) d \theta,
+\end{equation}
+
+where $p(\theta | y_{1:T})$ is known as the posterior distribution, with
+
+\begin{equation}
+\label{posterior}
+ p(\theta | y_{1:T}) = \frac{p(y_{1:T}|\theta)p(\theta)}{\int_\theta p(y_{1:T}|\theta)p(\theta) d\theta}
+\end{equation}
+
+and $p(\theta)$ is some pre-specified prior distribution.
+
+These often cannot be solved analytically, while numeric integration is computationally infeasible when the dimension of $\theta$ is large.
+To address this problem there is a wide range of techniques used to approximate the solution to the integral in (\ref{forecastintegral}), such as Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
+These approximations have an implicit trade-off: The better approximations are compuationaly intensive, and the forecaster must decide how much computation time and approximation error is acceptable.
+The focus of this research is in situtations where the time budget is too small for MCMC to be reliable, and uses VB as an alternative.
+This situtation is common in time-series forecasting, which must update the posterior distribution for each new data point observed via
+
+\begin{equation}
+\label{posteriorupdate}
+p(\theta | y_{1:T+1}) = \frac{p(y_{T+1} | \theta) p(\theta | y_{1:T})}{\int_\theta p(y_{T+1} | \theta) p(\theta | y_{1:T}) d\theta}.
+\end{equation}
 
 \section{Bayesian Inference}
+\subsection{Exact Bayesian Computation}
+While it is technically an approximation method, MCMC algorithms result in what is often called an exact computation of the posterior.
+A Gibbs based MCMC iteratively samples from the conditional distributions
 
-\subsection{Bayes' Rule}
+\begin{align}
+&p(\theta_1 | \theta_2, \dots, \theta_p, y_{1:T}) \nonumber \\
+&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_p, y_{1:T}) \nonumber \\
+&\vdots \nonumber \\
+&p(\theta_p | \theta_1, \dots, \theta_{p-1}, y_{1:T}) \nonumber
+\end{align}
 
-In Bayesian statistics, we are interested in making inferences about an unknown parameter $\theta$ using what is known as the posterior distribution, $p(\theta | y)$, where $y$ is our observed data.
-Bayes' rule provides a form for this distribution,
+where $p$ is the dimension of $\theta$.
+With enough iterations, the error in the approximation converges to zero and the algorithm can be ran for as much time as the forecaster desires to reduce error to a desired level.
+However, in the first iteration we must set arbitary starting values for each of $\theta_2, \dots, \theta_p$ introducing a large amount of error in the early iterations.
+To avoid this error MCMC generally must be run for a large number of iterations and these early samples are discarded.
+The computation time per iteration and speed of convergence is extremely problem specific.
+
+We illustrate this with an AR(2), a simple time series model used in the remainder of this section. The AR(2) is described by
 
 \begin{equation}
-\label{BR1}
-p(\theta | y) = \frac{p(y | \theta) p(\theta)}{p(y)},
+\label{AR2}
+y_t = \mu + \rho_1 y_{t-1} + \rho_2 y_{t-2} + \epsilon_t
 \end{equation}
 
-which shows that the posterior can be found by taking the product of the likelihood, $p(y | \theta)$ and the prior distribution, $p(\theta)$, and normalising with a constant $p(y)$.
-In practice $p(y)$ is unknown and can be found from
+where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. We collect the unknown parameters as $\theta = (\mu, \rho_1, \rho_2, \sigma^2)'$ and set priors as
 
-\begin{equation}
-\label{BR2}
-p(y) = \int_\theta p(y | \theta) p(\theta) d \theta.
-\end{equation}
+\begin{align}
+p(\mu) &\propto 1 \nonumber \\
+p(\sigma^2) &\propto \sigma^{-2} \nonumber
+\end{align}
 
-Finding the value of $p(y)$ is often the most difficult part of working with Bayesian statistics.
-In some simple models, this integration can be solved analytically, but for most models of interest it is intractable and we can only work with the un-normalised version
+add something for $\rho$ to enforce stationarity - maybe switch to an AR(1) or ARMA(1,1).
 
-\begin{equation}
-\label{BR3}
-p(\theta | y)^* = p(y | \theta) p(\theta).
-\end{equation}
+Describe conditional distributions, possible MH step, result in forecast distribution for $y_{T+1}$
 
-This work will concern itself with two of the methods used in the case where (\ref{BR2}) is intractable, exact methods such as MCMC, which will be discussed later in this section as well as an approximation method known as Variational Bayes which appears in section two.
+\subsection{Variational Bayes}
 
-\subsection{Exact Bayesian Computation}
-
-\textbf{1. Numerical Integration}\newline
-
-It might be tempting to numerically approximate (\ref{BR2}) by evaluting the integrand over a grid and applying methods such as the Trapezoid Rule or Adaptive Quadrature.
-These may be suitable for integration in a low dimension, but if we have a grid with $G$ points in one dimension, then a two dimensional integration requires $G^2$ points, and a three dimensional integration requires $G^3$ points.
-In general, the computational time increases exponentially with the dimension of $\theta$ in numerical integration.
-In almost all cases of interest, $\theta$ is a high-dimensional vector which makes numerical integration prohibitively expensive.
-This is known as the curse of dimensionality, where the difficulty of the problem increases quickly with the dimension of $\theta$ and requires us to use more sophisticated methods.\newline
-
-\textbf{2. Importance Sampling} \newline
-
-Candidate Density
-Accept Reject
-Good candidate is hard but useful
-VB as a candidate \newline
-
-\textbf{3. Gibbs Sampling}\newline
-
-Breaks down into simple problems
-Becomes linear instead of exponential
-Exact sampling \newline
-
-\textbf{4. Metropolis Hastings within Gibbs}\newline
-
-Gibbs only works exactly on some problems
-Approximation within Gibbs
-Also can benefit from VB
-
-\section{Variational Bayes}
-
-\subsection{Estimation as an optimisation problem}
-
-In the event that our dataset is too large, or our model is too complex, for MCMC to converge within a reasonable time frame we must resort to approximation.
-One important question that might be asked then is \textit{`How do I measure how good my approximation is?'}.
-If two distributions come from the same family, one might consider the difference in the parameters, but it is trivial to come up with distributions with slight parameter differences and vastly different regions of probability mass.
-Consider the difference between a $\mathcal{N}(0, 0.1)$ and a $\mathcal{N}(0.5, 0.1)$ distribution, the change in mean is only $0.5$ but there is little overlap in the densities.
-Further, large values for the variance allow large changes in mean with very little change in the resulting densities.
-If the distributions come from different families it is almost impossible to use parameters to measure the degree of closeness between two distributions.
-Instead we must move from the parameter space to the distribution space, where one measure of the difference between distributions $p(\theta)$ and $q(\theta)$ is the Kullback-Leibler (KL) divergence \citep{Kullback1951}, defined by
+Variational Bayes introduces some approximating distribution $q(\theta | \lambda)$ and aims to choose the family $q$ and set of parameters $\lambda$ so that $q(\theta | \lambda$ is as close as possible to the true posterior $p(\theta | y)$.
+It does this by minimising the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ to $p(\theta | y)$.
+The KL divergence is defined by
 
@@
 \label{KL-def}
-KL[q(\theta)||p(\theta)] = \int q(\theta) \ln \left( \frac{q(\theta)}{p(\theta)}\right) d\theta.
+KL[q(\theta | \lambda)||p(\theta | y)] = \int q(\theta | \lambda) \ln \left( \frac{q(\theta | \lambda)}{p(\theta | y)}\right) d\theta.
 \end{equation}
 
-KL divergence is a non-negative, assymetric measure of the difference between $p(\theta)$ and $q(\theta)$ that will equal zero if and only if $p(\theta) = q(\theta)$ almost everywhere.
-It has origins in information theory, and can be interpreted as: \textit{`Given that I know $p(\theta)$ for some $\theta \in \Theta$, how much extra information is required, on average, to know the value of $q(\theta)$?'} There are examples of the literature of approximations with other measures of divergence, such as \citet{Minka2001} introducting Expectation Propogation, which minimises the reverse measure $KL[p(\theta)||q(\theta)]$, which was extended to Power-EP in \citet{Minka2004}, which aims to minimisea the more general $\alpha-\mbox{divergence}$ \citep{Amari1985}.
-\vspace{3mm}
+and is a non-negative, assymetric measure of the difference between $p(\theta | y)$ and $q(\theta | \lambda)$ that will equal zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}.
+It has origins in information theory, and can be interpreted as: \textit{`Given that I know $p(\theta | y)$ for some $\theta \in \Theta$, how much extra information is required, on average, to know the value of $q(\theta | \lambda)$?'} There are examples of the literature of approximations with other measures of divergence, such as \citet{Minka2001} introducting Expectation Propogation (EP), which minimises the reverse measure $KL[p(\theta | y)||q(\theta |\lambda)]$, which was extended to Power-EP in \citet{Minka2004}, which aims to minimise the more general $\alpha-\mbox{divergence}$ \citep{Amari1985}.
+It is shown in \citet{Bishop2006} (note: find original proof) that minimising $KL[p(\theta | y)||q(\theta | \lambda)]$ used in in EP is equivalent to the MLE of $q(\theta | \lambda)$ given a sample of $\theta$.
+However, we can write $KL[q(\theta | \lambda)||p(\theta | y)]$ as
 
-\begin{figure}[hb]
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/klnorm-1}
-
-\end{knitrout}
-\caption{Red: A Normal(0, 0.1) distribution. Green: A Normal(0.5, 0.1) distribution.}
-\end{figure}
-
-\newpage
-
-Variational Bayes aims to find the approximating distribution, $q(\theta | \lambda)$, that minimises $KL[q(\theta | \lambda)||p(\theta |y)]$.
-This approximating distribution is parameterised by a $\lambda$ vector, each element of which may be a function of $y$, the hyperparameters of the prior distribution, and even of other elements in $\lambda$.
-The optimal approximation, in the sense of having the minimum KL divergence with the posterior, can be derived by beginning with the unknown constant, $\ln(p(y))$.
-\begin{align}
-\ln(p(y)) &= \int_{\theta} q(\theta|\lambda) \ln(p(y)) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \right) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \frac{q(\theta|\lambda)}{q(\theta|\lambda)}\right) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) + \ln \left( \frac{p(y, \theta)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
-&= \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) d\theta + \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
-&= KL [q(\theta||\lambda) || p(\theta | y)] + \mathcal{L}(q(\theta | \lambda), y) \nonumber
-\end{align}
-
-where
-
-\begin{align}
-\mathcal{L}(q(\theta|\lambda), y) &= \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta|\lambda)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
-&=  \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta. \label{ELBO}
-\end{align}
-
-Hence we have that
 \begin{equation}
-\label{VBKL}
-KL [q(\theta|\lambda) || p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q(\theta | \lambda), y).
+\label{KL-ELBO}
+KL[q(\theta | \lambda)||p(\theta | y)] = \ln(y) - \mathcal{L}(q, y)
 \end{equation}
 
-From (\ref{VBKL}) it is clear that by maximising $\mathcal{L}(q(\theta | \lambda), y)$ with respect to $q(\theta|\lambda)$ the KL divergence between the unknown true posterior $p(\theta|y)$ and the approximating distribution $q(\theta|\lambda)$ will be minimised.
-Hence the Variational Bayes solution will be to maximise the Evidence Lower Bound (ELBO), $\mathcal{L}(q(\theta|\lambda), y)$, in (\ref{ELBO}).
+where $\mathcal{L}(q, y)$ is known as the Evidence Lower BOund (ELBO), defined by
 
-\subsection{Mean Field Variational Bayes}
+\begin{equation}
+\label{ELBO}
+\mathcal{L}(q, y) = \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta.
+\end{equation}
+
+From (\ref{ELBO}) it is clear that maximising $\mathcal{L}(q, y)$ with respect to $q$ is equivalent to minimising (\ref{KL-def}).
+Maximising the ELBO is much more convenient than minimising either form of the KL Divergence, and has lead to Variational Bayes been used much more widely in the literature than alternatives such as EP.
+
+\subsubsection{The Mean Field Assumption}
 
@@
 If it does not match a known distribution, we may need to make a further approximation $\tilde{q_i(\theta_i|\lambda_i)}$ which has a recognizable distribution.
-One method as used in \citet{Friston2006} uses a Laplace approximation for the otherwise unrecognizable $q_i(\theta_i | \lambda_i)$ with reasonable results.
+One method as used in \citet{Friston2006} uses a Laplace approximation to and substitute in a Gaussian distribution for an otherwise unrecognizable $q_i(\theta_i | \lambda_i)$.
 
@@
 Hence, the optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family.
-Deriving the updating algorithm for MFVB takes more analytical effort than finding the conditionals for a Gibbs MCMC scheme, as we must then find the values of each $\lambda_i$ implied by (\ref{mf3}), which are often a function of $\lambda_{j \neq i}$ through the expectations.
+Hence deriving the updating algorithm for MFVB takes about the same analytical effort as finding the conditionals for a Gibbs MCMC scheme, and the secondary Laplace approximation will occur in parameters where a Metropolis-Hastings-within-Gibbs step must be used.
 \vspace{5mm}
 
-Given these optimal distributional families, an algorithm that cycles through all of the $\lambda_i$ and sets each to its maximising value is known as a coordiante ascent algorithm.
-The algorithm must continuously iterate between each parameter until $\mathcal{L}(q(\theta | \lambda), y)$ converges with some pre-defined threshold noting that the $\lambda$ may depend on each other.
-If $\lambda$ has dimension $k$, the algorithm follows below.
+Given these optimal distributional families, each $\lambda_i$ can be found as a function of the data and $\lambda_{j \neq i}$.
+This dependence requires an algorithm that continuously iterates between each $\lambda_i$ and sets it to its maximising value until $\mathcal{L}(q(\theta | \lambda), y)$ converges with some pre-defined threshold.
+This is known as a coordinate ascent algorithm and follows below using $k$ as the dimension of $\lambda$.
 
@@
  Initialise $\lambda$ randomly\;
+ Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution\;
  \While{Not converged}{
@@
       Hold $\lambda_{j \neq i}$ fixed\;
-      Use (\ref{mf3}) to match $q(\theta_i|\lambda_i)$ to a known distribution\;
-      Update $\lambda_i$ using the most recent values of $\lambda_{j \neq i}$\;
+      Update $\lambda_i$ using $y$ and the most recent values of $\lambda_{j \neq i}$\;
      }
@@
 
-\subsection{Stochastic Gradient Ascent}
+\subsubsection{Stochastic Variational Bayes}
+\emph{Note the change to subsubsection}
 
-In many cases, the posterior is too complex to be captured by a factorising approximation and we desire our distribution $q(\theta | \lambda)$ to capture dependence between parameters.
-In this case, the easy maximisation in MFVB is unavailable, and we must resort to what is called Stochastic Variational Bayes (SVB) using a gradient ascent algorithm developed by \citet{Paisley2012} and \citet{Hoffman2013}.
+In many cases, the posterior is too complex to be captured for a factorising approximation to be reasonabl as we desire our distribution $q(\theta | \lambda)$ to capture dependence between parameters.
+In this case, the easy maximisation in MFVB is unavailable, and we must resort to what is called Stochastic Variational Bayes (SVB) using a gradient ascent algorithm developed by \citet{Paisley2012} and \citet{Ranganash2014}.
 
@@
 
-\subsection{SVB Extensions}
+\subsection{Variance Reduction Techniques} \emph{This is general stuff, not just for VB, so can be a subsection}
+-Rao Blackwellisation
+-Control Variates
+-Reparameterisation
+\subsection{AR2 model example (Revisited using VB)}
 
-\textbf{1. AdaGrad} \newline
+\section{Electricity Load Forecasts}
+\subsection{Motivation}
+\emph{Will you include a subsection showing an overview of the actual electricity load data you have??}
+Real data has five minute updates
+Density is of interest due to rare price spikes
+Slow convergence of MCMC / Bad scaling to large dimensional models and datasets
+\subsection{Exponential Smoothing}
+Bayesian Version
+Difficulty with smoothing parameters in large models
+Infer model structure from draws where possible
+\subsection{Variational Bayes Implementation}
+Posterior Distributions should not change much as more data is observed, so we can keep approximating distribution family constant
+But we might want a model that does change to make updates worthwhile - Markov Switching mechanism to be added?
+\subsection{Forecasting}
+compare to MCMC when possible
 
-If the gradient is small, we want the learning rate, $\rho_t$, to be large, and if the gradient is large, we want the learning rate to be small.
-AdaGrad \citep{Duchi2011} is a method of setting the learning rate to dynamically adapt as the values of the calculated gradients changes.
-Further, AdaGrad allows the step size of each element of $\lambda$ to move independently of the other elements of $\lambda$.
-
-If $\nabla_{\lambda}^{(m)}$ is the vector of gradients at iteration $m$, define
-
-\begin{equation}
-G^{(m)} = \sum_{i = 1}^{m} \nabla_{\lambda}^{i} (\nabla_{\lambda}^{i})^T,
-\end{equation}
-
-the sum of the outer products of the gradients up to iteration $m$. Then set
-
-\begin{equation}
-\label{adagrad}
-\rho^{(m)} = \eta \mbox{diag}(G^{(m)})^{-1/2}
-\end{equation}
-
-with some preset tuning parameter $\eta$.
-If we have $k$ elements in $\lambda$, $G^{(m)}$ is a $k \times k$ matrix so we have one element in $\rho^{(m)}$ for each element of $\lambda$.
-As we only need the diagonal of $G^{(m)}$, this may be calculated in linear time.
-\newline
-
-\textbf{2. Reparameterisation} \newline
-
-The reparameterisation trick was popularised following \citet{Kingma2013} and allows the gradient of the ELBO to be simplified.
-
-Consider a parameter free auxillary distribution $q(\epsilon)$ and differentiable transformation $f(\cdot,\cdot)$ such that $\theta = f(\epsilon, \lambda)$.
-Examples include a location-scale transformation from a standard normal or an inverse-CDF transform from a uniform$(0, 1)$ variable.
-Note that
-
-$$q_\theta(\theta | \lambda) = q_\epsilon(\epsilon) \left| \frac{d\epsilon}{d\theta} \right| $$
-
-where the parameters that govern the transform $f$ are the same $\lambda$ parameters as in $q(\theta | \lambda)$. Now (\ref{SGA2}) becomes
-
-\begin{align}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\epsilon} \nabla_{\lambda}[  q(\epsilon) \left( \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda) | \lambda) \right)] d\epsilon \nonumber \\
-&=  \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda)| \lambda) \right] d\epsilon \label{RP1} \\
-&= \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda)) \right] d\epsilon\nonumber
-\end{align}
-
-as the second part on the right hand side of (\ref{RP1}) is a score function with expectation equal to zero. This can be estimated using
-
-\begin{equation}
-\label{RP2}
- \mathcal{L}(q(\theta | \lambda), y) \approx  \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon_s, \lambda))) \right].
-\end{equation}
-
-with $\epsilon_s \sim q(\epsilon)$.
-
-\citet{Kingma2015} explore how this reparameterised version can have orders of magnitude lower variance than the estimator in (\ref{SGA5}).
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\epsilon^s$ for $s = 1, \dots S$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(p(f(\epsilon^s, \lambda^{(m-1)}), y))]$
-     }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda}^{t}$\;
-  Set $m = m + 1$\;
- }
- \caption{Gradient Ascent for re-parameterised SVB}
-  \label{alg:algorithm3}
-\end{algorithm}
-
-$\mbox{ }$\newline \textbf{3. Control Variates} \newline
-
-If reparameterisation is too difficult or otherwise unavailable, \citet{Paisly2012} introduce the use use of control variates to estimate the gradient with reduced variance.
-
-Define
-
-\begin{equation}
-\label{CV1}
-g(\theta, \lambda, y) =  \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda))),
-\end{equation}
-
-the function estimated in (\ref{SGA5}), then for some other function $h$, define
-
-\begin{equation}
-\label{CV2}
-\hat{g}(\theta, \lambda, y) = g(\theta, \lambda, y) - a(h - \mathbb{E}(h)).
-\end{equation}
-
-Then we have that
-
-\begin{equation}
-\mathbb{E}(\hat{g}) = \mathbb{E}(g)
-\end{equation}
-
-and
-
-\begin{equation}
-\label{CV3}
-\mbox{Var}(\hat{g}) = \mbox{Var}(g) + a^2 \mbox{Var}(h) - 2a\mbox{Cov}(g, h). 
-\end{equation}
-
-Solving the polynomial (\ref{CV3}) in $a$ shows that the variance of $\hat{g}$ is minimised by
-
-\begin{equation}
-\label{CV4}
-\hat{a} = \mbox{Cov}(g, h)/\mbox{Var}(h).
-\end{equation}
-
-Substituting (\ref{CV4}) into (\ref{CV3}) yields
-
-\begin{equation}
-\label{CV5}
-\mbox{Var}(\hat{g}) = \mbox{Var}(g) - \mbox{Cov}(g, h)^2/\mbox{Var}(h).
-\end{equation}
-
-We need to choose some function $h$ that has a large covariance with $g$. \citet{Ranganath2014} suggests that the an easy option is
-
-\begin{equation}
-\label{CV6}
-h(\theta, \lambda) = \nabla_{\lambda} [\ln(q(\theta_s | \lambda))].
-\end{equation}
-
-As $h$ is a score function, it has an expectation of zero and our estimator becomes
-
-\begin{equation}
-\label{CV7}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)) - \hat{a}),
-\end{equation}
-
-where $\hat{a}$ is estimated from the sample variance and covariance of a subset of the $S$ Monte-Carlo draws.
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\theta_s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $g_{\lambda_i} = 1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
-      Calculate $h_{\lambda_i} = 1/S\sum_{s=1}^{S}\nabla_{\lambda_i}[\log(q(\theta^s |\lambda^{(m-1)}))]$\;
-      Use a subset of $S$ to estimate $\hat{a} = \mbox{Cov}(h, g)/\mbox{Var}(g)$\;
-      Calculate $\nabla_{\lambda_i} = g_{\lambda_i} - \hat{a} h_{\lambda_i}$\;
-     }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
-  Set $m = m + 1$\;
- }
- \caption{Gradient Ascent for SVB with control variates}
-  \label{alg:algorithm4}
-\end{algorithm}
-
-\subsection{AR2 model example}
-
-Consider $T$ observations from an AR(2) model conditioned on $y_1$ and $y_2$,
-
-\begin{equation}
-\label{AR2}
-y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t \mbox{ for } t = 3,\dots,T,
-\end{equation}
-
-where $\epsilon$ are iid $\mathcal{N}(0, \sigma^2)$.
-
-Using the independent priors:
-
-\begin{align}
-p(\phi_1) &\sim \mathcal{N}(\bar{\phi_1}, \tau_{1}) \nonumber \\
-p(\phi_2) &\sim \mathcal{N}(\bar{\phi_2}, \tau_{2}) \nonumber \\
-p(\sigma^2) &\sim \mathcal{IG} (\mbox{shape} = a, \mbox{ scale} = b), \nonumber
-\end{align}
-
-we have a model with zero stationary restrictions which can easily be sampled through a Gibbs MCMC scheme.
-This exact posterior can be used in comparision with the results of our variational approximations.
-
-The Gibbs conditional distributions are derived in the appendix and are
-
-\begin{align}
-p(\phi_1 | \phi_2, \sigma^2, y_{1:T}) &\propto \mathcal{N}(\mu_1^G, \gamma_1^G) \nonumber \\
-p(\phi_2 | \phi_1, \sigma^2, y_{1:T}) &\propto \mathcal{N}(\mu_2^G, \gamma_2^G) \nonumber \\
-p(\sigma^2 | \phi_1, \phi_2, y_{1:T}) &\propto \mathcal{IG}(shape = \alpha^G,  scale = \beta^G), \nonumber
-\end{align}
-
-where
-
-\begin{align}
-\mu_1^G &= \frac{\tau_1 (\sum_{t=3}^{T} y_t y_{t-1} - \phi_2 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \sigma^2 \bar{\phi_1}}
-          {\tau_1 \sum_{t=3}^{T} y_{t-1}^2 + \sigma^2} \nonumber \\
-\gamma_1^G &= \frac{\sigma^2 \tau_1}{\tau_1 \sum_{t=3}^{T} y_{t-1}^2 + \sigma^2} \nonumber \\
-\mu_2^G &= \frac{\tau_2 (\sum_{t=3}^{T} y_t y_{t-2} - \phi_1 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \sigma^2 \bar{\phi_2}}
-          {\tau_2 \sum_{t=3}^{T} y_{t-2}^2 + \sigma^2} \nonumber \\
-\gamma_2^G &= \frac{\sigma^2 \tau_2}{\tau_2 \sum_{t=3}^{T} y_{t-2}^2 + \sigma^2} \nonumber \\
-\alpha^G &= a + T/2 - 1 \nonumber \\
-\beta^G &= b + 1/2 \sum_{t=3}^{T} \left((y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right).  \nonumber
-\end{align}
-
-100 data points are simulated from the model in (\ref{AR2}) with true values of $\phi_1 = 0.8, \phi_2 = 0.15$ and $\sigma^2 = 2$.
-These values were chosen to induce a strong dependence between the $\phi$ parameters as it is close to the non-stationary region of $\phi_1 + \phi_2 > 1$.
-The resulting posterior is sampled with 100,000 draws from Gibbs MCMC, discarding the first 20\% of all draws as a burn in.
-The resulting draws were strongly dependent so they were thinned by a factor of 20 to result in 4000 effectively independent draws.
-The total runtime for MCMC was 2.6 seconds.
-The dependence between $\phi_1$ and $\phi_2$ is apparent in figure (\ref{AR2MCMC}), so a SGA algorithm that captures this dependence should be a better approximation than a MFVB algorithm that assumes independence.
-
-\begin{figure}
-\label{AR2MCMC}
-\centering
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\includegraphics[width=\maxwidth]{figure/AR2MCMC-1}
-
-\end{knitrout}
-\caption{Marginal and bivariate density plots of the MCMC draws.
-100,000 draws per parameter were sampled from (\ref{AR2}) via Gibbs MCMC then reduced to 4000 each.
-Note the strong dependence between the two $\phi$ parameters.}
-\end{figure}
-
-The derivation of the MFVB approximation is also in the appendix and each marginal has the same form as the corresponding true conditional distribution above,
-\begin{align}
-q(\phi_1 | \lambda) &\propto \mathcal{N}(\mu_1^{MF}, \gamma_1^{MF}) \nonumber \\
-q(\phi_2 | \lambda) &\propto \mathcal{N}(\mu_2^{MF}, \gamma_2^{MF}) \nonumber \\
-q(\sigma^2 | \lambda) &\propto \mathcal{IG}(shape = \alpha^{MF},  scale = \beta^{MF}), \nonumber
-\end{align}
-
-where the parameterising vector $\lambda = (\mu_1^{MF}, \gamma_1^{MF}, \mu_2^{MF}, \gamma_2^{MF}, \alpha^{MF}, \beta^{MF})'$.
-
-The MVFB derivation yields the following updating equations for $\lambda$,
-
-\begin{align}
-\mu_1^{MF} &= \frac{\alpha^{MF}/\beta^{MF} (\sum_{t=3}^{T} y_t y_{t-1} - \mu_2 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \tau_1^{-1} \bar{\phi_1}}{\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-1}^2  +  \tau_1^{-1}} \nonumber \\
-\gamma_1^{MF} &= \left(\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-1}^2  +  \tau_1^{-1} \right)^{-1} \nonumber \\
-\mu_2^{MF} &= \frac{\alpha^{MF}/\beta^{MF} (\sum_{t=3}^{T} y_t y_{t-2} - \mu_1 \sum_{t=3}^{T} y_{t-1} y_{t-2}) + \tau_2^{-1} \bar{\phi_2}}{\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-2}^2  +  \tau_2^{-1}} \nonumber \\
-\gamma_1^{MF} &= \left(\alpha^{MF}/\beta^{MF} \sum_{t=3}^{T} y_{t-2}^2  +  \tau_2^{-1} \right)^{-1} \nonumber \\
-\alpha^{MF} &=  a + T/2 - 1  \nonumber \\
-\beta^{MF} &= b + 1/2 \sum_{t=3}^{T} \left((y_t^2 - ((\mu_1^{MF})^2 + \gamma^{MF}_1) y_{t-1}^2 - ((\mu_2^{MF})^2 + \gamma_2^{MF}) y_{t-2}^2\right) \nonumber \\
-&+ 2 \sum_{t=3}^{T} \left(\mu^{MF}_1 \mu^{MF}_2 y_{t-1}y_{t-2} - \mu^{MF}_1 y_{t} y_{t-1} - \mu^{MF}_2 y_t y_{t-2}\right) \nonumber.
-\end{align}
-
-The forms of each component are similar to the conditionals derived for Gibbs sampling, with changes coming from linearising the expressions before taking expectations.
-Notice that all components of $\lambda$ except $\alpha^{MF}$ depend on each other through the expectations in (\ref{mf3}).
-
-The SGA algorithm requires the specification of the approximating family so we make the following assumptions:
-\begin{itemize}
-\item The optimal marginal distribution for the dependent SGA approximation is the same as the independent MFVB approximation.
-\item There is no dependence between either $\phi$ parameter and $\sigma^2$.
-\item There is eliptical dependence between $\phi_1$ and $\phi_2$.
-\end{itemize}
-
-To determine the form of dependence between $\phi_1$ and $\phi_2$ we fit elliptical copulas to the MCMC draws and determined that the gaussian copula is the best fit by BIC.
-Gaussian marginals for each $\phi$ and a gaussian copula leads to $\phi_1$ and $\phi_2$ having a bivariate gaussian distribution.
-
-Hence we have the following form for the SGA approximations,
-
-\begin{align}
-q(\phi_1, \phi_2) &\sim \mathcal{N}(\mu, \Sigma) \nonumber \\
-q(\sigma^2) &\sim \mathcal{IG}(\alpha, \beta). \nonumber
-\end{align}
-
-As each parameter in $\Sigma$ can move independently in the SGA algorithm, we instead use the lower triangluar decomposition $L L^T = \Sigma$.
-This results in the vector $\lambda = (\mu_1, \mu_2, L_{11}, L_{21}, L_{22}, \alpha, \beta)'$, where
-
-\begin{itemize}
-\item $\mu_1$: The mean of $\phi_1$,
-\item $\mu_2$: The mean of $\phi_2$,
-\item $L_{11}, L_{21}, L_{22}$, the three components of $L$, the lower triangular decomposition of $\Sigma$,
-\item $\alpha, \beta$, the shape and rate parameters of an Inverse Gamma distribution for $\sigma^2$.
-\end{itemize}
-
-The diagonal of $L$ is not forced to be positive so the decomposition is not invariant to sign changes.
-
-We compare three forms of the SGA algorithm, the original algorithm described by \ref{alg:algorithm2}, the reparameterised version described by \ref{alg:algorithm3} and the control variate version described by \ref{alg:algorithm4}.
-Each version is implemented with $\rho^(m)$ using AdaGrad as in (\ref{adagrad}.
-
-The reparameterised SGA algorithm requires an auxillay distribution $q(\epsilon)$ and transformation.
-We use $(\epsilon_1, \epsilon_2)' \sim \mathcal{N}(0, I)$ and $\epsilon_3 \sim \mathcal{U}(0, 1)$.
-Then make the transformation
-
-\begin{align}
-\phi_1 &= \mu_1 + L_11 \epsilon_1 \nonumber \\
-\phi_2 &= \mu_2 + L_21 \epsilon_1 + L_22 \epsilon_2 \nonumber \\
-\sigma^2 &= Q^{-1}(\epsilon_3, \alpha, \beta) \nonumber
-\end{align}
-
-where $Q^{-1}(\epsilon_3, \alpha, \beta)$ is the Inverse Gamma distribution inverse cdf function with shape $\alpha$ and scale $\beta$.
-
-We summarise each algorithm below in table \ref{table} by its estimated parameters, converged value of the ELBO and runtime.
-Mean Field algorithm runs extremely quickly but can not estimate the dependence between paramters so the converged ELBO is lower than the alternatives, except for the original SGA algorithm.
-This estimator is extremely noisy and required a value of $S$ of $200$, which slowed computation but it still converged to a lower optima than the variance reduced algorithms.
-The reparameterised algorithm, SGA-RP, could be run with $S = 5$ leading to its extremely fast runtime.
-The control variate algorithm, SGA-CV, was run with $S = 50$, and used $20$ draws per iteration to estimate $\hat{a}$.
-
-\begin{table}
-\centering
-\label{AR2results}
-\begin{knitrout}
-\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}
-\begin{tabular}{l|l|l|l|l|l|l|l|l|l}
-\hline
-  & mu1 & mu2 & L11 & L21 & L22 & alpha & beta & ELBO & Runtime (seconds)\\
-\hline
-Mean Field & 0.83 & 0.1 & 0.04 & 0 & 0.05 & 50 & 79.7 & -77.7 & 1.3 x 10\textasciicircum{}\{-6\}\\
-\hline
-SGA & 0.63 & 0.27 & 0.09 & -0.12 & 0.05 & 45.6 & 81.4 & -78.3 & 51.1\\
-\hline
-SGA-RP & 0.78 & 0.15 & 0.1 & -0.09 & 0.05 & 49.3 & 78 & -76.7 & 4\\
-\hline
-SGA-CV & 0.77 & 0.15 & 0.09 & -0.09 & 0.05 & 48.5 & 77.6 & -76.9 & 1.3\\
-\hline
-\end{tabular}
-
-\end{knitrout}
-\caption{Parameter estimates, maximised ELBO and runtime of all Variational Bayes algorithms.}
-\end{table}
-
-\section{Approximate Forecast Updating}
-\subsection{Motivation}
--Slow convergence of MCMC
--Bad scaling to large dimensional models and datasets
--Time/Computational Constraints
--Motivation from short forecast windows
-\subsection{Model Selection}
--Overnight/Intermittent MCMC
--Infer model structure from draws
--Have easy parametric form of posterior to work with
+\section{Timeline}
 
@@
 
-\section{Appendix}
-\subsection{AR2 Model}
 \end{document}
+\grid

### 2017-01-02 Confirmation/derivations.Rnw
Xmas break updates

\documentclass{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}

\begin{document}

\title{Derivations}

\section{Variational Bayes}

First we define the KL Divergence between distributions $q(\theta)$ and $p(\theta)$ as

$$KL[q(\theta)|p(\theta)] = \int q(\theta) \ln \left( \frac{q(\theta)}{p(\theta)}\right) d\theta.$$

This is a non-negative function that takes on a value of $0$ if and only if $q(\theta) = p(\theta)$.

Then, using the unknown constant $\ln(p(y))$, we have

\begin{eqnarray}
\ln(p(y)) & = & \int_{\theta} q(\theta|\lambda) \ln(p(y)) d\theta \nonumber \\
& = & \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \right) d\theta \nonumber \\
& = & \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{p(y, \theta)}{p(\theta | y)} \frac{q(\theta|\lambda)}{q(\theta|\lambda)}\right) d\theta \nonumber \\
& = & \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) + \ln \left( \frac{p(y, \theta}{q(\theta|\lambda)} \right) d\theta \nonumber \\
& = & \int_{\theta} q(\theta|\lambda) \ln  \left( \frac{q(\theta|\lambda)}{p(\theta | y)}\right) d\theta + \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta)}{q(\theta|\lambda)} \right) d\theta \nonumber \\
& = & KL [q(\theta|\lambda) || p(\theta | y)] + \mathcal{L}(q(\theta | \lambda), y) \nonumber
\end{eqnarray}

where 
\begin{eqnarray*}
\mathcal{L}(q(\theta|\lambda), y) & = & \int_{\theta} q(\theta|\lambda) \ln \left( \frac{p(y, \theta|\lambda)}{q(\theta|\lambda)} \right) d\theta \\
& = &  \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta.
\end{eqnarray*}

Hence we have that
$$ KL [q(\theta|\lambda) || p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q(\theta | \lambda), y)$$

and clearly maximising $\mathcal{L}(q(\theta | \lambda), y)$ with respect to $q(\theta|\lambda)$ will minimise the KL divergence between the unknown true posterior $p(\theta|y)$ and an approximating distribution $q(\theta|\lambda)$.

\section{The Mean Field Assumption}

If our approximating distribution $q(\theta|\lambda)$ factorises as

$$q(\theta|\lambda) = \prod_i q(\theta_i | \lambda_i)$$,

where each component $\theta_i$ may be a scalar or a vector, and using the notation $q_i = q(\theta_i|\lambda_i)$ and $q_{\setminus i} = \prod_{j\neq i}q_j$, we have:

\begin{eqnarray}
\label{mf}
\mathcal{L}(q(\theta | \lambda), y) & = &\int_{\theta} q_i q_{\setminus i} \ln (p(y, \theta)) d\theta - \int_{\theta} q_{i}q_{\setminus i} \ln (q_{i}q_{\setminus i}) d\theta \nonumber \\
& = & \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} ( \ln (p(y, \theta)) - \ln(q_{i})) d\theta_i d\theta_{\setminus i} - \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} \ln (q_{\setminus i}) d\theta_i d\theta_{\setminus i}  \nonumber \\
& = & \int_{\theta_{i}} q_{i} \left( \int_{\theta_{\setminus i}} q_{\setminus i} \ln (p(y, \theta )) d\theta_{\setminus i} - \ln(q_i) \right) d\theta_i \nonumber \\
&- & \int_{\theta_i} q_{i} \int_{\theta_{\setminus i}} q_{\setminus i} \ln(q_{\setminus i}) d\theta_{\setminus i} d\theta_i \nonumber \\
& = & \int_{\theta_{1}} q_{i} \ln \left( \frac{\exp(  \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])}{q_i} \right) d\theta_i - \int_{\theta_{\setminus i}} q_{\setminus i} \ln(q_{\setminus i}) d\theta_{\setminus i} \\
& = & -KL (q_i || \exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])) + c \nonumber
\end{eqnarray}

as the rightmost component of (\ref{mf}) is constant with respect to $q_i$.

We can see that, holding $q_{\setminus_i}$ fixed, we can maximise $\mathcal{L}(q(\theta | \lambda))$ with respect to $q_i$ by minimising the term KL Divergence between $q_i$ and $\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])$ by setting

\begin{equation}
\label{mf2}
q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))]).
\end{equation}

The distributional family for any $q_i$ in (\ref{mf2}) that minimises the KL Divergence between the true posterior and the appoximation is found by taking the log-joint density, ignoring any term that doesn't depend on $\theta_i$, taking expectations of any remaining $\theta_{j \neq i}$ and then exponentiating. These steps are the same that can be used to find a conditional distribution $p(\theta_i | \theta_{j \neq i})$, with dependence on other parameters replaced by their expectations. Hence, the distributional family for any $\theta_i$ in Mean-Field Variational Bayes is the exact same distributional family as $p(\theta_i | \theta_{j \neq i})$ as used in Gibbs MCMC schemes.
\vspace{5mm}

Given these distributional families, an algorithm that cycles through the $\lambda_i$ and sets each to the value given by (\ref{mf2}) is known as a coordiante ascent algorithm. Typically the expectations make each $\lambda_i$ a function of $\lambda_{j \neq i}$ and the algorithm must continuously iterate between each parameter until $\mathcal{L}(q(\theta | \lambda), y)$ converges with some pre-defined threshold.

\section{Stochastic Gradient Ascent}

An alternative method to maximise the function $\mathcal{L}(q(\theta | \lambda), y)$ is through a gradient ascent algorithm, where, for $t = 1, 2, \dots$,

$$\lambda_t = \lambda_{t-1} + \rho_t \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda_{t-1}), y),$$

given some initial value $\lambda_0$ for some sequence $\rho_t$ that satisfies the Robbins-Monro conditions,

\begin{eqnarray*}
\sum_{t=1}^{\infty} \rho_t & = & \infty \\
\sum_{t=1}^{\infty} \rho^2_t & < & \infty.
\end{eqnarray*}

This algorithm does not require $q(\theta | \lambda)$ to factorise and so can be applied to distributions $q(\theta | \lambda)$ that allow parameter dependence. However, it does not provide the family of the the optimal approximating distribution, unlike under the mean field assumption. Restricting ourselves by the assumption that we can interchange the order of differentation with respect to $\lambda$ and integration with respect to $\theta$, we can calculate derivatives of $\mathcal{L}(q(\theta | \lambda), y)$ as follows:

\begin{align}
\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \nabla_{\lambda} \int_{\theta} q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
&=   \int_{\theta} \nabla_{\lambda}[  q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)] d\theta \label{grad1} \\
&=  \int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta)) - \ln(q(\theta | \lambda))] d\theta \nonumber \\ 
&+ \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \label{grad2} \\
&=  \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
&=   \int_{\theta} \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)q(\theta | \lambda) d\theta \nonumber \\
&= E_q [  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)] \label{grad3}
\end{align}

Note that the first integral on the right hand side of (\ref{grad2}) equals 0 as

\begin{eqnarray*}
\int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta))] d\theta & = & 0
\end{eqnarray*}

and

\begin{eqnarray*}
\int_{\theta} - q(\theta | \lambda) \nabla_{\lambda}[ \ln(q(\theta | \lambda))] d\theta & = & - \int_{\theta} \frac{\nabla_{\lambda}q(\theta | \lambda)}{q(\theta | \lambda)} q(\theta | \lambda) d\theta \\
& = & -  \nabla_{\lambda}\int_{\theta} q(\theta | \lambda) d\theta \\
& = & 0
\end{eqnarray*}

We can compute a Monte-Carlo estimate of (\ref{grad3}) by

\begin{equation}
\label{mc}
\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda)) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)))
\end{equation}

with $\theta_s \sim q(\theta | \lambda)$.

As the distribution $q(\theta | \lambda)$ is specified by the user, the main restriction on the use of a stochastic gradient ascent algorithm is that the log-joint density $\ln(p(y, \theta))$ is able to be evaluated.

\section{Variance Reduction}

The Monte-Carlo estimate in (\ref{mc}) often has far too large a variance to useful without small step sizes $\rho_t$ and a large number of draws $S$. These significantly slow down the algorithm, and a lower variance estimator is more useful.

\subsection{Reparamatisation}

Consider a unit free distribution $q(\epsilon)$ and detetministic transform $f(\cdot,\cdot)$ such that $\theta = f(\epsilon, \lambda)$. Examples include a location-scale transformation from a standard gaussian or an inverse-CDF transform from a uniform$(0, 1)$ variable. Note that

$$q_\theta(\theta | \lambda) = q_\epsilon(\epsilon) \left| \frac{d\epsilon}{d\theta} \right| $$

where the parameters that govern the transform $f$ are the same $\lambda$ parameters as in $q(\theta | \lambda)$.

and (\ref{grad1}) becomes

\begin{align}
\label{grad4}
\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\epsilon} \nabla_{\lambda}[  q(\epsilon) \left( \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda) | \lambda) \right)] d\theta \nonumber \\
&=  \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[  ( \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda)| \lambda)) \right] d\theta \nonumber \\
&\approx  \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} \left[  ( \ln (p(y, f(\epsilon_s, \lambda))) - \ln(q(f(\epsilon_s, \lambda)| \lambda)) \right].
\end{align}

with $\theta_s \sim q(\epsilon)$.

This reparameterised version often has orders of magnitude lower variance than the estimator in (\ref{mc}).

\subsection{Control Variates}

If reparameterisation is too difficult or otherwise unavailable, we can use control variates to estimate our function $g$ with reduced variance.

With

\begin{equation}
\label{grad5}
g(\theta, \lambda, y) =  \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)))
\end{equation}

and for some choice of $h$, define

\begin{equation}
\label{controlv}
\hat{g}(\theta, \lambda, y) = g(\theta, \lambda, y) - a(h(\theta, \lambda, y) - \mathbb{E}(h(\theta, \lambda, y))).
\end{equation}

Then we have that

\begin{equation}
\label{controlv2}
Var(\hat{g}) = Var(g) + a^2 Var(h) - 2aCov(g, h).
\end{equation}

Solving the polynomial in $a$ shows that the variance of $\hat{g}$ is minimised by

\begin{equation}
\label{controlv3}
\hat{a} = Cov(g, h)/Var(h).
\end{equation}

We need to choose some function $h$ that has a high covariance with (\ref{grad5}), so choose

\begin{equation}
\label{controlv4}
h = \nabla_{\lambda} [\ln(q(\theta_s | \lambda))]
\end{equation}

As $h$ is a score function, it has an expectation of zero and our estimator becomes

\begin{equation}
\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)) + \hat{a})
\end{equation}

with $\hat{a}$ estimated from the corresponding sample statistics from a subset of the $S$ Monte-Carlo draws.

\end{document}

### 2017-01-02 Confirmation/paper.Rnw
Xmas break updates

\documentclass[12pt]{article}
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[colorlinks=TRUE, linkcolor=blue]{hyperref}
\usepackage{wrapfig,float}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage[section]{placeins}
\usepackage{afterpage}
\usepackage{bbm}

\graphicspath{{images/}}
\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}

\newtheorem{thm}{Theorem}[section]
\newtheorem{dfn}{Definition}[section]
\newtheorem{cor}{Corollary}[thm]
\newtheorem{con}{Conjecture}[thm]

% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

\begin{document}
%\SweaveOpts{concordance=TRUE}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\title{\bf What is the Best Way to Display Two Time Series to Read Association?}

\if0\blind
{
\author{Nathaniel Tomasetti and Dianne Cook\\
Department of Econometrics and Business Statistics, Monash University\\
}
  \maketitle
} \fi
\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf What is the Best Way to Display Two Time Series to Read Association?}
\end{center}
  \medskip
} \fi

\bigskip

\begin{abstract}
Visual inference in EDA is prone to type 1 errors from the over-interpretation of randomness \citep{Daniel, Diaconis}. Two competing plot designs, the scatter plot and overlaid line graph are both popular in the analysis of time series data. Lineups \citep{Majumder:2013, Wickham} allow the visual inference power of a graphic display to be evaluated, and were used to compare the plot designs. We collected data on the detection rate of correlated pairs of AR(1) simulations, the time required and the confidence of the selection for 1600 lineup evaluations. The results show that the scatter plot is both the faster and more powerful plot design, despite its inability to display the time dimension. 
\end{abstract}

\noindent
{\it Keywords:}  data visualisation, statistical graphics, visual inference, power comparison, scatter plot, line graph, visual analytics, time series, temporal data, finance, econometrics
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!
\newpage

\tableofcontents

<<setup, echo=FALSE, results='hide', message=FALSE, error=FALSE, warning=FALSE>>=
library(ggplot2)
library(mnormt)
library(tidyr)
library(knitr)
library(nullabor)
library(dplyr)
library(mFilter)
library(grid)
library(gridExtra)
library(stringr)
#library(readr)
#library(lubridate)
# Convert these to read_csv at some point
correl <- read.csv("data/correlations.csv", stringsAsFactors = FALSE) ##Correlations of every plot in every lineup
full <- read.csv("data/full.csv", row.names=1, stringsAsFactors = FALSE) ##MTurk evaluations
prediction <- read.csv("data/predfit.csv") ##fitted data from SAS
time.fit <- read.csv("data/time.csv") ##coefficients for time model from SAS
powercoef <- read.csv("data/PowerEstimates.csv") ##coefficients for power model from SAS

full <- full %>% mutate(design = ifelse(test_param ==1, "Overlaid Lines", "Scatterplot"),
                         r = round(correlation, 1), #remove slight variations in correlation caused by simulation
                         rsgn = ifelse(r >= 0, 1, 0),
                         rabs = abs(r),
                         alpha = log(0.05/0.95), #an offset used to fit the model
                         attempts = sapply(strsplit(as.character(full$response_0), ","), length), 
                         ##count number of selections in response_0 column
                         weights = 1/attempts)

full$subj_id <- factor(full$nick_name, labels=1:212)
y <- data.frame(table(full$subj_id))
keep <- y$Var1[y$Freq>9]
full.sub <- filter(full, subj_id %in% keep) 
full.sub <- Reduce(rbind, by(full.sub, full.sub$subj_id, head, n=10))
##Remove subjects with less than 10 evaluations. If more than 10, keep only first 10.

full <- full %>%  transform(response_0= strsplit(as.character(response_0), ",")) %>% unnest(response_0) %>%
  ##split multiple selections in response_0 into different rows
  mutate(correct = ifelse(response_0 == actual_location, 1, 0), #ensure only the true detection in mult. selections has a 1
         score = ifelse(correct==1, (20-attempts)/19, 0)) %>% 
  select(subj_id, correct, score, r, rsgn, rabs, smoothed, n, design, log_time, response_0,
         actual_location, pic_id, choice_reason, attempts, weights, alpha, conf_level)

full.sub <- full.sub %>% transform(response_0= strsplit(as.character(response_0), ",")) %>% unnest(response_0) %>%
  mutate(correct = ifelse(response_0 == actual_location, 1, 0), 
         score = ifelse(correct==1, (20-attempts)/19, 0)) %>%
  select(subj_id, correct, score, r, rsgn, rabs, smoothed, n, design, log_time, response_0,
         actual_location, pic_id, choice_reason, attempts, weights, alpha, conf_level)

bylineup <- summarise(group_by(full, pic_id, n, smoothed, design, r, actual_location), total=sum(score), correct=sum(correct), evals=sum(weights))
bylineup <- mutate(bylineup, percent = total/evals)
bylineup.sub <- summarise(group_by(full.sub, pic_id), total=sum(score), evals=sum(weights))
bylineup.sub <- mutate(bylineup.sub, percent = total/evals)

##Summarise data by lineup

#sasdata <- select(full.sub, rabs, n, alpha, rsgn, smoothed, design, subj_id, weights, correct, log_time)
#sasdata <- mutate(sasdata, rabs = as.numeric(rabs),
#                   subj_id = as.numeric(subj_id),
#                   design = ifelse(design=="Scatterplot", 1, 0))
#write.csv(sasdata, "sas/sasdata.csv", row.names = FALSE)
#The code that is exported to fit the model into SAS, see code/fitting model.sas

#pred <- data.frame(expand.grid(rabs=seq(0, 0.9, 0.01), t=c(12, 24, 48, 96), alpha=log(.05/0.95), rsgn=c(0,1),
#                                    smoothed=c(0,1), design=c(0,1),
#                                      subj_id = as.numeric(as.character(unique(full.sub$subj_id)))))
#pred <- subset(pred, smoothed==0 | smoothed == 1 & t ==48 | smoothed ==1 & t ==96)
#write.csv(pred, "sas/predgrid.csv", row.names=FALSE)
#Code exported to SAS to predict new values

colnames(prediction) <- c("rabs", "t", "alpha", "rsgn", "smoothed", "design", "subj_id", "weights", "detected", "log_time", "RandomEf", "FixedEf", "residual")
prediction <- mutate(prediction, r = ifelse(rsgn==1, rabs, -rabs),
                     design = ifelse(design==0, "Overlaid Lines", "Scatterplot"), 
                     design = ifelse(smoothed==1, paste0(design, " (Smoothed)"), design),  
                     diff = round(RandomEf - FixedEf,5))
##Data from SAS

gen_true_data <- function(n, r, smoothed = FALSE) {
  d <- data.frame(rmnorm(n, c(0, 0), matrix(c(1, r, r, 1), 2, 2))) 
  d$X1 <- as.vector(arima.sim(list(ar=n/100), n, innov=d$X1))
  d$X2 <- as.vector(arima.sim(list(ar=n/100), n, innov=d$X2))
  if (smoothed) {
    d$X1 <- as.vector(hpfilter(d$X1, freq=1, type="lambda", drift = FALSE)[[2]])
    d$X2 <- as.vector(hpfilter(d$X2, freq=1, type="lambda", drift = FALSE)[[2]])
    d <- as.data.frame(d)
  }
  d$X1 <- scale(d$X1)
  d$X2 <- scale(d$X2)
  d$t <- 1:n
  return(d)
}
##Generates a pair of data, length n, correlation r

gen_null <- function(n, m=20, smoothed = FALSE){
  nd <- NULL
  for(i in 1:(m-1)) {
    d <- data.frame(X1 = as.vector(arima.sim(list(ar=n/100), n)),
                    X2 = as.vector(arima.sim(list(ar=n/100), n)))
    if (smoothed) {
      d$X1 <- as.vector(hpfilter(d$X1, freq=1, type="lambda", drift = FALSE)[[2]])
      d$X2 <- as.vector(hpfilter(d$X2, freq=1, type="lambda", drift = FALSE)[[2]])
      d <- as.data.frame(d)
    }
    d$X1 <- scale(d$X1)
    d$X2 <- scale(d$X2)
    nd <- rbind(nd, d)
  } 
  nd$.n <- rep(1:(m-1), each = n) 
  nd$t <- 1:n
  return(nd)
}
##Generates m-1 pairs of uncorrelated data, length n
@

\section{Introduction}

%In order to work with data, it first must be understood. Statistical inference requires hypotheses to be established prior to data collection, but often data is collected first.
%This is especially so today, for vast databases that have been assembled in the big data era, that now need the data scientist to unravel the meaning of the numbers. 
%Without preset hypotheses to test, the power of statistical inference is impotent, and without hypotheses the data analyst can stumble blindly trying to build up models of structure in the data. 
%To understand data requires good visualisation. This idea was formed as early as the 18th century, when William Playfair institutionalized the then revolutionary idea of graphing government and economic data. 
%Far easier than reading tables of numbers, these ideas were powerful, and by providing the basic building blocks for plotting statistical data, his graphic designs became a conduit for the communication of otherwise complex information \citep{Sachs}. Since then, advances in computing power have allowed statistical graphics to flourish, spearheaded by \citet{Tukey} in 1965 into the new domain of exploratory data analysis (EDA) \citep{Friendly}. 
%EDA can be thought of as a well understood \citep{Cleveland, Vanderplas} set of tools and techniques required to visualise information, to physically see what the data contains. In particular, it incorporates a free roaming approach, where the analyst is able to explore structure to find whatever relationships and structure exists within. 
%Critically, the analyst does not have to have any pre-conceived ideas or hypotheses -- they are not specifically looking for any one particular thing. EDA emphasises letting the data inform us and can lead to the discovery of otherwise unexpected relationships, many of which may seem to become completely obvious after discovery. 
%With the knowledge provided by EDA, ideas are generated about what relationships between variables may potentially exist. This then enables the analyst to use these new hypotheses upon which to apply classical inference and rigorously conduct tests with new data.
%EDA is also related to the field of model diagnostics (MD), where a model can be continuously refined through the visualisation of its fit, its residuals, and the interactions with its variables. Both EDA and MD follow the same framework: Visualise the data, look for patterns that suggest an underlying relationship, and if one is found implement it into the model then continue the exploration of the data. 
% Section 2 describes the visual inference protocols which put EDA more firmly into the rigorous framework of statistical inference.

Temporal dependence is present in many macroeconomic variables \citep{Nelson, Perron} and in financial data \citep{Sher, Turtle}. It is a common, but untested, belief that the inclusion of time in the graphics will increase the detail of information displayed and allow relationships to be examined more comprehensively, so many analysts utilise the overlaid line graph to examine their data (Figure \ref{nzd}). In the field of statistics, the recommendation for exploring association visually is to use a scatterplot. There is substantial literature supporting this \citep{Cleveland, Harrison, Li}.  However, it is still common that when the two variables are temporal they are displayed as an overlaid time series (line graphs). In this paper we have utilized new methods for rigorously comparing plot design, using visual inference, to investigate which is the best display if reading association is important.

\begin{figure}[htbp]
  \centering
    \begin{subfigure}[htbp]{0.45\textwidth}
    <<nzd, error=FALSE, message=FALSE, warning=FALSE, echo=FALSE>>=
    rates <- select(read.csv("data/rates.csv"), date, AUD, CNY, NZD)[36:69,]
    rates <- mutate(rates, AUD2=(AUD-mean(AUD))/sd(AUD), CNY2=(CNY-mean(CNY))/sd(CNY)) #scaling
    ggplot(data=rates, aes(x=AUD, y=NZD)) + geom_point()
    @
    \end{subfigure}
    \begin{subfigure}[htbp]{0.45\textwidth}
    <<nzd2, dependson='nzd', error=FALSE, message=FALSE, warning=FALSE, echo=FALSE>>=
    rates_l <- gather(rates, currency, value, -date)
    ggplot(data=subset(rates_l, currency != "CNY" & currency != "CNY2" & currency != "AUD2"), aes(x=as.Date(date), y=value, colour=currency)) + geom_line() + theme(legend.position="bottom")
    @
    \end{subfigure}
  \caption{Cross-rates for Australian and NZ dollars against US dollar, from Mar 30-May 2, 2015, shown using a scatterplot (left) and overlaid line graph (right). From which plot is correlation more accurately read? Point-wise correlation between the two series, ignoring autodependence, is 0.64.}
  \label{nzd}
\end{figure}

\begin{figure}[htbp]
  \centering
  <<cny, dependson='nzd2', out.width='0.45\\textwidth', error=FALSE, message=FALSE, warning=FALSE, echo=FALSE>>=
   ggplot(data=subset(rates_l, currency != "NZD" & currency != "CNY" & currency != "AUD"), aes(x=as.Date(date), y=value, colour=currency)) + geom_line() + theme(legend.position="bottom")
  @
  \caption{Cross-rates for Australian dollar and Chinese yuan against the United States dollar from Mar 30-May 2, 2015, shown as overlaid line graph. (Both sets of values were standardized to bring them to the same scale for plotting.) What would you report the correlation between the two series to be?}
  \label{cny}
\end{figure}

To determine on an appropriate display requires awareness of cognitive principles such as the Gestalt law of common fate \citep{Wertheimer}. People have a tendency to `see' a positive association between two objects that only briefly move together over time, leading them to believe, perhaps erroneously, that they are linked and share a `common fate'. To the contrary, \citet{Lee} suggests that people may better perceive a relationship between objects that move with temporal synchrony in accordance with the Gestalt Law, implying that overlaid line graphs, that utilize a time axis, may outperform the graph types that do not, such as the scatter plot. The magical thinking ideas described in \citet{Diaconis} would counter \citet{Lee}'s findings. Magical thinking means that humans have a strong desire to find patterns, even imaginary, or mirages, which may lead to a false discovery of positive correlation from a small period of synchrony. Figure~\ref{cny} highlights the interaction between `magical thinking' and the Gestalt Law, the observer may be drawn to the period from April 8 to April 22, and may falsely believe that the two currencies are strongly tied. However, over the entire series the linear correlation between the two currencies is -0.12, indicating there is no actual relationship.

The law then, has two effects. If positive correlation exists in temporal data, the observer may be drawn to it faster from an overlaid line graph display. If positive correlation does not exist, or only exists ephemerally, the observer may be fooled by the overlaid line graphs into thinking there is strong association. This latter effect is a problem.

In practice we have lots of choices in how to present information in exploratory analyses or to communicate information. If linear correlation is the only aspect of the data of interest, we might only calculate the correlation between two series. However, if correlation is not the appropriate statistic, because the association is more complex, or if there are outliers, we can typically only discover this by making pictures. Learning how people read plots of data is an important component of data analysis. If information is to be communicated in graphical form it is good to know what we can confidently expect the audience to infer.

There have been a few studies related to plotting temporal data.  \citet{Robbins} writes a blog and consults on plotting financial data. She is critical of the overlaid line graph, because it is misleading if the purpose is to read the difference between the two series. Her arguments date back a long time to Pfari, as reported in \citet{Tufte:1986} and \citet{friendly}. It is well-established that we cannot read differences between two series accurately, that the difference should be plotted separately if that is the information of interest. \citet{Vanderplas} examines a related problem, dubbed the ``sine illusion'', where the human eye is is fooled into erroroneously thinking that variation is larger at the peaks and troughs of seasonal time series. \citet{Vanderplas} developed a mathematical correction to the illusion.  \citet{Javed} examined the perception of many different sorts of features in multiple time series, and found that there was no optimal plot design for reading all of the different features tested. They did not examine perception of association between two series, though.

Reading correlation between two sets of non-temporal context numbers has been examined in many papers. In \citet{Harrison} different graphical displays of two sets of numbers are ranked of their effectiveness by Weber's Law \citep{webers} which provides a relationship between perception and the ``effect size'' in the data. For example, two series with correlation 0.9 should be proportionately easier to distinguish from noise than two series with correlation 0.3. The visual display should reflect this difference is ease. Early papers \citep{Cleveland1982,lane1985} showed that people read association from a scatterplot differently than they would linearly interpret a correlation coefficient.

%As these problems are unique to the line graph, they argue that the scatter plot, which does not have any ability to display time, is the superior choice. However, in practice many data analysts are split between the two major alternatives for temporal data. The overlaid line graph is justified by its ability to present more information to the analyst; but many argue that it is this extra information that is misleading, and revert to the scatter plot for its strong non-temporal performance.

%\begin{figure}[htbp]
%  \centering
%  \includegraphics[width=0.5\textwidth]{murders}
%  \caption{\citep{Robbins} Murders in New York State (Orange) and New York City (Blue). The difference between the two represents murders outside of the city. What size would you judge the difference to be in 1994?}
%  \label{murder}
%\end{figure}

%There has been other research on the perception of temporal displays, such as in \citet{Javed}, but they do not examine perception of association. Javed et al.  examined optimal ways to visualise local features, such as, which variable had the highest value at a given point in time, and global features, such as, comparing the size of the overall slope of many different variables. The finding was that a different graphical layout is optimal for each type of task. However, they did not test for the perception of correlation, which can be treated as both a local and a global feature within the graph. The slope of each variable must be compared at a point in time, and they must have a persistent relationship across at least the majority of the series. However there is a quandary, the visual features that improve local tasks are unsuited to global tasks and vice versa; so we must now find a form that is well suited to assess both types of tasks simultaneously.

This paper rigorously examines the perception of association between two time series in two different displays, and with an additional inquiry into whether smoothing of series affects the perception, using the lineup protocol. Section 2 describes the lineup protocol, and the background to its use for comparing the power of plot designs for different purposes. Section 3 explains the experimental design.  Section 4 contains the results of the experiment and Section 5 discusses the implications of the findings.

\section{The lineup protocol for comparing plot designs}

It has long been thought that EDA and statistical inference were worlds apart, but this is no longer true. \citet{Buja} and \citet{Majumder:2013} bridges the chasm. Framing a data plot as a test statistic, which when compared to plots of null data, places EDA into the statistical inference recipe. Underlying any data plot, there is an implicit null hypothesis that there is ``no pattern''. Particular types of plots may regulate what no  pattern means. For example, a scatter plot is used to explore for association between the two variables, so the implicit null hypothesis is that there is no association between the two variables. The alternative hypothesis is that there is some association, although it is not required to specify precisely the type of association. The departure from the null might be a single outlier, or few outliers, a non-linear trend, or clusters. To test the null hypothesis the data plot is placed in a lineup with plots of null data, data generated from a scenario of no pattern. For the scatterplot this can be achieved by permuting the values on one of the variables to break any association between the two variables. An impartial viewer is asked to select the plot that is the most different from the others. If they select the data plot as the most different, this counts as a rejection of the null hypothesis, and follow-up questions on why they made their choice may determine the type of departure from the null. With multiple independent observers a $p$-value can be calculated to measure the statistical significance of the departure from the null. This process is called the ``lineup protocol''.

The lineup protocol equips exploratory visual methods with a rigorous procedure that enables statistical inference. This is important because visualization remains essential for data analysis because human eyes can detect patterns which would not be detected mathematically. The lineup protocol provides the calibration of the human eyes to prevent us from erroneously jumping to conclusions about magic patterns. Prior to the protocol description, we can find preliminary ideas along these lines, for example \citet{Daniel} recommends looking at plots of data from a null distribution for exploring two factor design experimental data.  His paper provides 40 pages of plots from the null! He encourages data analysts to understand the patterns that might be seen in samples data drawn from populations without inherent structure. \citet{Buja} has named this the Rorschach protocol.)

%which the human eye detects as more different in this data plot than in any of the nulls.
%The plot of the data is placed in a lineup of plots of null data, data generated assuming that the null hypothesis is true. If an impartial observer asked to select the plot that is most different from the rest, does select the data plot this suggests there exists a pattern that is not the result of chance, and the null hypothesis is rejected. 
%For a scatter plot, this departure from the null, might be a single outlier, or few outliers, a non-linear pattern, or clusters, which the human eye detects as more different in this data plot than in any of the nulls. This is the reason why visualisation remains important today, human eyes can detect patterns which would not be detected mathematically.  But eyes need calibration, which the lineup protocol provides.
%On its own, with a single plot, because of random sampling in collecting data, it is easy for the analyst to imagine a pattern when no real structure exists in the population. Visual skills of an observer in EDA is prone to Type I error, where the null hypothesis is rejected when it is actually true, caused by the inherently random formation of patterns when visualised which are attributed to structure rather than chance. \citet{Daniel} warns against this, by providing 40 pages of plots from the null distribution, he encourages data analysts to understand the patterns that can be created from data without inherent structure. (This is what \citet{Buja} call the Rorschach protocol.) By being aware of what can appear in this type of data, the analyst should be more wary of claiming any visual feature they find is a true relationship. But this is not enough, even seasoned data analysts can be misled.  \citet{Diaconis} introduced the notion of 'magical thinking' which argues that people commonly suffer from the over-interpretation of randomness, particularly if it matches a pre-conception. If it suits their particular bias, an analyst may make false discoveries of structure and false rejections of the null hypothesis. Whilst using both can complement each other, the treatment of Type I error has led EDA and inferential statistics to be considered as very disparate pursuits. The mathematical rigor that governs classical inference relies on a framework that acknowledges and controls for the rate of Type I error.

%With the general lack of inference present in exploratory data analysis, how do we attempt to control error in hypothesis testing? In order to use visualisation effectively, we will get the most benefit if the type of graphic display chosen is best suited to the task at hand, be it EDA, MD, or even the presentation of results.
%We then need to find the statistical power of graphics, the ability of a graphical display to convey information on the structure of the data within. This can be found by using the lineup protocol developed in 2009 by \citet{Buja}, which is easily implemented with the nullabor R package.
%They created a statistically rigorous framework with properties explored primarily in \citet{Majumder:2014} and further in \citet{Hofmann:2014} that allows us to conduct these hypothesis tests with visual inference, thus to some degree allowing the conjoining of EDA with classical inferential statistics.
%The lineup protocol is inspired in part by the police lineup \citep{Wickham}. We place the plot of the 'true' data generated with some underlying structure (The criminal) in amongst plots of data that were generated from a null distribution (The innocent people). If the real data plot is selected by an uninvolved observer as being most different from the other plots there is evidence that the structure of that data has led to a significant difference in the plot (That the criminal is sufficiently different from the innocents). 
%This constitutes a rejection of the null hypothesis. If one of the null plots is chosen instead then either the plot design did not have the sufficient power to display the true relationship (A Type 2 error in EDA), or the null plot exhibited a strong relationship that was generated randomly and an analyst that saw this plot and decided that the data had some structure would’ve committed a Type 1 error.

% First the p-value calculation for a lineup, multiple observers

The lineup protocol specifies that the data plot be placed in a randomly selected location in a field of $(m-1)$ null plots.

Let $p_{i}$ be the probability that plot $i$ is chosen from the lineup. The choice will depend on the plot design, $d$, and the signal strength, $q_{i}$, of the structure displayed in plot relative to the other plots in the lineup. This can be defined as a function $f_{i,d} (q_{1},\dots,q_{20})$\citep{Hofmann:2014}. If the data plot is detected, it indicates that the plot design could convey the desired information about the underlying structure and that the data plot has a higher signal strength than the null plots.

If the data plot has no pattern, that is, the null hypothesis is true, there is a chance of making a Type I error. For a lineup of $m$ plots, the probability of selecting any plot is $1/m$, setting the Type I error rate, $\alpha$, of the hypothesis test to equal $1/m$.  To control our error rate, we could increase change $m$, this could produce a large cognitive burden on one observer, having to sort through many plots. An alternative approach is to recruit more observers, $K$, to evaluate a lineups, then the probability that plot $i$ is selected by $x$ observers under the null hypothesis follows a binomial distribution \citep{Majumder:2013}:

\begin{equation}
\label{pvalue}
  \centering
  p-value = P(X \geq x | H_{0}) = 1 - B_{n, 1/m}(x-1)
\end{equation}

The plot design that has a human observer selecting the data plot the most often will thus minimise both Type I error and Type II error in the hypothesis testing.

If all $K$ observers select the same plot out of a lineup of twenty, we result in a p-value as small as $1/m^K$ . We used $m= 20$, giving us a significance level (and Type 1 error rate) of $\alpha = 0.05$. The plot of 'true' data is placed randomly amongst 19 null plots to form the lineup.

% Multiple picks

If a subject selects multiple plots from the lineup, the p-value has a Bernoulli distribution with a probability $p_{j} = s/m$, where that observer made $s$ selections out of a lineup with $m$ plots. In the event that a lineup has observers making different numbers of selections, we have a binomial distribution with different success probabilities, and a Poisson Binomial distribution must be used to calculate the p-value demonstrated by \citet{Follett}. The probability mass function of a Poisson Binomial distribution can be written as:

\begin{equation} 
\label{pvalue2}
\centering
P(X \geq x) = 1 - P(X < (x - 1)) = 1 - \sum^{x-1}_{i=0} \left\{\sum_{A \in T_{x}} \prod_{j \in A} p_{j} \prod_{j \in A^{c}} (1-p_{j})\right\}
\end{equation}

Where $T_{x}$ is all possible sets of $x$ trials out of $K$ resulting in a success, $A \in T_{x}$ is a specific set of $x$ trials resulting in a success and $A^{c}$ is the complement to $A$, the set of $K-x$ trials resulting in a failure. This can easily be calculated using the poibin package for R.

% Then power calculation for comparing designs

\begin{figure}[htbp]
  \centering
<<lineup.example, dependson='setup', fig.width=6, fig.height=6, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
set.seed(1234)
a <- gen_true_data(24, 0.95)
td <- gather(a, variable, value, c(X1, X2))
b <- gen_null(24)
nd_l <- gather(b, variable, value, c(X1, X2))
pos <- 7
td <- data.frame(.n=rep(pos, nrow(td)), td)
lg <- nd_l$.n < pos
nd_l1 <- data.frame(.n=nd_l$.n[lg], t=nd_l$t[lg], variable=nd_l$variable[lg], 
                    value=nd_l$value[lg])
nd_l2 <- data.frame(.n=nd_l$.n[!lg], t=nd_l$t[!lg], variable=nd_l$variable[!lg], 
                    value=nd_l$value[!lg])
nd_l2$.n <- nd_l2$.n + 1
d <- rbind(nd_l1, td, nd_l2)

ggplot(data=d, aes(x=t, y=value, colour=variable)) + geom_line() + facet_wrap(~.n, ncol=4, scales="free_y")  +
    theme_bw() + theme(axis.title.x = element_blank(),
                       axis.title.y = element_blank(),
                       axis.text.x = element_blank(),
                       axis.text.y = element_blank(),
                       axis.ticks = element_blank(),
                       legend.position="none")
@
  \caption{Lineup of overlaid line graphs of size $m = 20$. In which plot are the two series most strongly associated?}
  \label{lineup:example}
\end{figure}

Essentially, the more time our subjects select the true plot out of the lineup, the more confident we are that that plot was visually significantly different to the nulls; and that the type of graphic involved has the power to display association. The lineup effectively allows us to conduct a hypothesis test on visual features, with the competing hypotheses:

\begin{itemize}
  \item $H_{0}$ : There is no association between the two series (evidence for $H_{0}$: the associated data is indistinguishable from the nulls).
  \item $H_{1}$: There is an association between the two series (evidence against $H_{0}$: the associated data can be distinguished from the nulls).
\end{itemize}

As the power of a statistical test is defined as the probability of rejecting $H_{0}$ when it is false, the power of a lineup test is viewed as the probability of detecting the true plot. We approximate the power as $\hat\pi = x/K$, where $x$ observers out of $K$ detected the true plot out of the lineup.
We can them estimate the power difference of competing lineups, $\hat\pi_{1} - \hat\pi_{2}$. An $\alpha \times 100\%$ confidence interval is calculated as \citep{Hofmann:2012}:

\begin{equation} \label{propci}
  \centering
  \hat\pi_{1} - \hat\pi_{2}\pm t_{1-\alpha,2K-1}\sqrt{\hat\pi_{1} (1-\hat\pi_{1}) /K_{1} + \hat\pi_{2} (1-\hat\pi_{2}) /K_{2} }
\end{equation}

Some individuals may have a better natural ability at detecting a correlation pattern in the graphics, but by having each participant viewing multiple lineups and each lineup being viewed by multiple different participants, this can be controlled via a random effect variable in the model. We recruited participants through Amazon's Mechanical Turk, where the lineup protocol has been applied to a number of problems in prior papers \citep{Heer, Majumder:2013, Wickham}. Amazon’s Mechanical Turk \citep{MTurk} is a labour crowd-sourcing platform developed to give easy access to workers with basic tasks paid in line with the United States minimum wage. MTurk can be used to recruit subjects to read a variety of graphics and report on particular visual tasks. The time taken to decide, confidence in the decision and the reason for that decision can also be recorded. \citet{Heer} use MTurk to replicate previous findings in graphical perception from \citet{Cleveland} and find that it is a valid method of data collection, once controls to eliminate anyone 'gaming' the system by randomly selecting answers to minimise time spent working are implemented. Further details of our use of Amazons’ Mechanical Turk is included in Section 3. The protocol has successfully been used to measure statistical power as a means to determine plot type superiority in \citet{Hofmann:2012}, and this work follows that approach.

\section{Experimental design}

For there to be meaningful information in the time dimension, the data must have autocorrelation. To fulfil this requirement, each plot has a pair of standardised AR(1) models generated by:

$$Y_{1,t} = \beta \times Y_{1,t-1} + e_{1,t}$$

$$Y_{2,t} = \beta \times Y_{2,t-1} + e_{2,t}$$

With $t \in \left\{12, 24, 48, 96\right\}$ and $e_{t}$ generated from a bivariate normal distribution:

$$e_{t} \sim \mathcal{N} \left( \mu = \left| \begin{array}{c}0 \\ 0 \end{array} \right| , \hspace{4mm} \Sigma = \left| \begin{array}{cc}1 & \rho \\ \rho & 1\end{array} \right| \right)$$

The variation in the covariance parameter allows for control in the correlation in the pairs of data. The null data has no underlying structure, so uncorrelated pairs are produced by setting $\rho = 0$. True data pairs are generated with $\rho \in \pm \{0.3, 0.5, 0.7, 0.9\}$. Pilot testing found that the rate of selecting the true data plot fell to approximately $5\%$, the $1/m$ rate of randomly selecting the true plot, for pairs with correlation below $\rho = 0.3$. Further to this, relationships weaker than this point are often of little interest to analysts. Hence we found it unnecessary to test a more full range of correlations. Simulating data does not produce correlations exactly equal to $\rho$, so simulated data pairs were accepted as real data if pairwise correlation between the two series is within 0.015 of the desired value of $\rho$. The null pairs often had correlation generated spuriously through the simulation process, which was particularly problematic for small sample data. Setting $\beta = min(0.5,100t)$ reduced spurious correlation to manageable levels.
For particularly large time series, the AR(1) model can create particularly 'jagged' lines that may be unrealistic for many actual applications. To reduce this, additional sets of data were created with a Hodrick-Prescott filter to capture the trend of the model and remove the noise in the cycle component. This filter creates a trend by penalising both errors in smoothness and in fit. The trend of each series $y_{t}$, $\tau_{t}$ is defined by the equation:

\begin{equation}
\centering
\label{hpfilter}
\min_{\tau}(\sum_{t=1}^{T}(y_{t}-\tau_{t})^2 + \lambda \sum_{t=2}^{T-1}[(\tau_{t+1}-\tau_{t})-(\tau_{t}-\tau_{t-1})]^2)
\end{equation}

Setting $\lambda = 1$ produced a smoothing that is more consistent with real world time series applications and did not introduce excessive spurious null correlation. However, for $t = \{12, 24\}$, smoothing becomes unfeasible without introducing extreme null correlation, often with an absolute value greater than 0.8. Smoothing was only used for $t = \{48, 96\}$. Each real data set was placed with a collection of 19 null plots and produced as both a scatter plot and an overlaid line graph.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{smootha}
<<smoothing, dependson='setup', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
set.seed(123)
uns <- gen_true_data(96, 0.8)
uns_l <- gather(uns, variable, value, -t)
set.seed(123)
sm <- gen_true_data(96, 0.8, smoothed = TRUE)
sm_l <- gather(sm, variable, value, -t)
ggplot(data=uns_l, aes(x=t, y=value, colour=variable)) + geom_line() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
\begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{smoothb}
<<smoothing2, dependson='smoothing', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
ggplot(data=uns, aes(x=X1, y=X2)) + geom_point() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
\begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{smoothc}
<<smoothing3, dependson='smoothing', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
ggplot(data=sm_l, aes(x=t, y=value, colour=variable)) + geom_line() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
\begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{smoothd}
<<smoothing4, dependson='smoothing', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
ggplot(data=sm, aes(x=X1, y=X2)) + geom_point() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
  \caption{Plots illustrating the experimental design. The same AR(1) data both unsmoothed (\subref{smootha} and \subref{smoothb}) and smoothed (\subref{smoothc} and \subref{smoothd}) for both plot designs, with t = 96 and correlation $\rho$ = 0.86}
  \label{smoothing}
\end{figure}

The lineups were generated with three basic factors:

\begin{itemize}
  \item Factor 1: Plot design, levels = scatter plot, overlaid time series.
  \item Factor 2: Sample size, t, levels = $\{12, 24, 48, 96\}$.
  \item Factor 3: True plot correlation, $\rho$, levels = $\pm \{0.3, 0.5, 0.7, 0.9\}$.
\end{itemize}

Additionally, there was a fourth factor for $t = \{48, 96\}$:

\begin{itemize}
  \item Factor 4: Smoothed, levels = yes, no.
\end{itemize}

Each combination of factors was replicated in three different lineups, giving a total of 288 lineups (3 replications x 2 plots x 8 correlations x 6 sample size and smoothness combinations). 
The position of the actual data plot in the lineup was randomized. The order that the factors are presented to a subject were also randomized, and each subject saw 5 scatter plots and 5 overlaid lines at a range of all absolute correlation levels. Each subject did not see the same data more than once. Subjects were asked to select the plot in the lineup that has the strongest association, with basic examples of negatively and positively correlated plots. They also answered with reasons for choosing the plot, and the confidence that they have that this really is the plot showing the strongest correlation. Each subject evaluated 10 lineups, plus a further two simple trials to eliminate people attempted to game the system. If the true plot was not detected in the trial, the subject could not evaluate the remaining lineups. If it was detected, the trial data was discarded and the subject moved onto the rest of the lineups. Earlier work by \citet{Majumder:2014} has found that repeated evaluations from a subject does not increase their ability to detect the true plot from a lineup, however random effects were used to model the dependence between different evaluations from the same subject. We recruited subjects from Amazon's Mechanical Turk service and ensured that each lineup was viewed by multiple subjects. In total there were 2091 lineup evaluations, however to better control correlation between subject’s individual visual ability we removed data from subjects with less than ten individual evaluations. Some subjects chose to participate in the experiment more than once, so we removed all data from any repeated trials leaving only the original ten evaluations. We were left with 1600 evaluations by 160 subjects.

Subjects were allowed to make multiple plot selections from each lineup evaluated, additional selections on lineup $i$ by subject $j$ are penalised according to a score defined by:

\begin{equation}
\label{score}
{score}_{ij} =\frac{m-s_{ij}}{m-1} \mathbbm{I}_{ij}
\end{equation}

where $\mathbbm{I}_{ij} = 1$ if the true plot is detected, $0$ otherwise, $s_{ij}$ is the number of selections made and $m = 20$ is the number of plots in the lineup.

However, for the model fit to power, multiple selections are seperated and given a weighting equal to to inverse of the number of selections made in that lineup evaluation. This increases our data from 1600 to 1762 evaluations.

\begin{figure}[htbp]
  \centering
  \begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{designa}
<<designs, dependson='setup', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
set.seed(101)
a <- gen_true_data(24, 0.95)
b <- gen_true_data(24, -0.95)
a_l <- gather(a, variable, value, -t)
b_l <- gather(b, variable, value, -t)
ggplot(data=a_l, aes(x=t, y=value, colour=variable)) + geom_line() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
\begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{designb}
<<designs2, dependson='designs', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
ggplot(data=b_l, aes(x=t, y=value, colour=variable)) + geom_line() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
\begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{designc}
<<designs3, dependson='designs', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
ggplot(data=a, aes(x=X1, y=X2)) + geom_point() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
\begin{subfigure}[t]{0.24\textwidth}
\subcaption{}\label{designd}
<<designs4, dependson='designs', fig.width=4, fig.height=4, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
ggplot(data=b, aes(x=X1, y=X2)) + geom_point() + theme(aspect.ratio=1, axis.title.x = element_blank(), axis.title.y = element_blank(),
          axis.text.x = element_blank(), axis.text.y = element_blank(), axis.ticks = element_blank(), legend.position="none")
@
\end{subfigure}
\caption{Plots illustrating the experimental design: the two different plot types of the same data with $t=24$, $\rho= 0.946$ (\subref{designa} and \subref{designc}), $\rho= -0.943$ (\subref{designb} and \subref{designd}).}
\label{designs}
\end{figure}

The web site that the Turk workers accessed to complete the task can be seen at http://104.236.245.153:8080/mahbub/turk18/index.html.  You can also try out the tasks, but your results will not be recorded. (The experiment cost \$210 to run.)

\section{Results}

\subsection{Detection, time and confidence}

\begin{figure}[htbp]
 \centering
<<results, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, warning=FALSE, error=FALSE, message=FALSE>>=
ggplot(data=bylineup, aes(x=r, y=percent, colour=design)) + geom_smooth(se=F) + geom_point() + facet_grid(n~smoothed) + 
  scale_colour_discrete(name = "Plot Design") + labs(y="Detection rate", x="Correlation") +
  theme(legend.position="bottom")
@
  \caption{Response accuracy for unsmoothed data, broken by unsmoothed (left), smoothed (right) and sample size. The dots are data points, the lines represent a naive model average fit. Both plot designs have similar performance for positively correlated data, however the overlaid line graphs are much weaker at displaying negative correlation.}
    \label{results}
\end{figure}

Figure~\ref{results} shows the raw results of the survey and Figure~\ref{powerdiff} shows the difference in power with a joint $95\%$ confidence interval calculated according to  formula~\ref{propci}. Values above zero indicate that the scatter plot had more power to detect the true plot than the overlaid lines.
Performance between the two plot designs appears to be similar for positive correlation, with the scatter plot power gradually improving relative to the overlaid lines as sample size increases. The scatter plot appears to be the superior choice as it retains its power with negative correlation, where the overlaid lines performance suffered. This is not surprising, as positive and negative correlation appear as symmetric graphs in the scatter plot but the two characteristics appear very differently in the overlaid line graph (Figure~\ref{designs}). The right side shows the performance benefits of smoothing overlaid line time series data. Much of the 'noise' in the time series is removed and studying the overall trend shows a greatly improved ability to successfully detect the true data plot when it is negatively correlated. The scatter plot still appears to be the superior plot, but much of the gap between the two is reduced for positive correlation.

\begin{figure}[htbp]
  \centering
<<powerdiff, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
prop <- select(bylineup, smoothed, n, evals, total, r, design)
prop <- summarise(group_by(prop, smoothed, n, r), l.attempt=sum(ifelse(design=="Overlaid Lines", evals, 0)),
                                                  l.detect=sum(ifelse(design=="Overlaid Lines", total, 0)),
                                                  s.attempt=sum(ifelse(design=="Scatterplot", evals, 0)),
                                                  s.detect=sum(ifelse(design=="Scatterplot", total, 0)))
alpha <- 1-(1-0.95^(1/48))/2 
##8 correlations * 6 design = 48 boxplots, so this the students t quantile for the joint 95% CI
prop <- mutate(prop, l.power = l.detect/l.attempt, 
               s.power = s.detect/s.attempt,
               df = s.attempt + l.attempt - 1, 
               difference=s.power-l.power, 
               t = qt(alpha, df),
               se = sqrt(s.power*(1-s.power)/s.attempt+l.power*(1-l.power)/l.attempt),
               upper = difference + t*se,
               lower = difference - t*se)
ggplot(data=prop) + geom_errorbar(aes(x=r, ymax=upper, ymin=lower)) + geom_point(aes(x=r, y=difference)) +
        geom_hline(aes(yintercept=0, colour="red")) + labs(x="Correlation", y="Power difference") + facet_grid(n~smoothed)
@
 \caption{The difference in estimated power (scatter plot power minus overlaid line power) to reveal the true plot by smoothness and sample size with a joint 95\% confidence interval. The red line indicates the point where both plot designs are equally powerful The largest differences appear in negative correlation, where the overlaid line graph's power drops substantially.}
 \label{powerdiff}
\end{figure}

\begin{figure}[htbp]
  \centering
<<samplesize, dependson='setup', fig.width=6, fig.height=8, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
ggplot(data=subset(bylineup, smoothed==0), aes(x=r, y=percent, colour=factor(n))) + geom_smooth(se=F) + geom_point() + facet_grid(design~.) + scale_colour_discrete(name = "Sample size") +  labs(y="Detection rate", x="Correlation") + theme(legend.position="bottom")
@
 \caption{Replotting figure~\ref{results} to highlight the effects of changes in sample size with unsmoothed data. The scatter plot appears to universally benefit from increased sample sizes but the overlaid lines have a mixed benefit, with improvements only in positive correlation.}
 \label{samplesize}
\end{figure}

Figure~\ref{samplesize} demonstrates the effects of changing sample size for unsmoothed data. The scatter plot has increased power to detect correlation almost universally for any increase. Sample size influences the overlaid lines differently, they appears to benefit for positively correlated true plots, but it does not improve the weak performance in negative correlation.
Figure~\ref{time} and figure~\ref{confidence} give a breakdown of the interactions between the response variables: plot detection, time and confidence. Figure~\ref{time} is the histogram of the time taken to record a response, split by plot design and sign of correlation and true plot detection. The time data was highly positively skewed with a maximum of 1760 seconds so the variable was log-transformed. We found that the subject could identify the true plot out of the overlaid line graph 33.6\% of the time and it took on average $e^{3.442}$ = 31.2 seconds to make a selection. Successful detections were slightly faster than unsuccessful decision, taking on average 28.7 seconds to select the true plot and 32.6 seconds if a null was selected. The scatter plots performed much better, with the true plot selected 52.9\% of the time, with an average of 22.7 seconds required to make a decision. If the true plot was detected it took only 18.6 seconds, and 29.3 seconds for a null plot selection.

\begin{figure}[htbp]
  \centering
<<time, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
time <- select(full, design, log_time, correct, rsgn)
time$rsgn <- factor(time$rsgn, labels=c("r -","r +"))
tm <- summarise(group_by(time, correct, design, rsgn), m=mean(log_time))
t1 <- ggplot(data=subset(time, correct==0), aes(x=log_time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==0), aes(xintercept = m), color="red") +
  facet_grid(design~rsgn) + labs(title="Did not detect", x="Log(Time) Taken", y=element_blank())  
t2 <- ggplot(data=subset(time, correct==1), aes(x=log_time)) + geom_histogram(binwidth=0.5) + 
  geom_vline(data=subset(tm, correct==1), aes(xintercept = m), color="red") +
  facet_grid(design~rsgn) + labs(title="Detected", x="Log(Time) Taken", y=element_blank())  
grid.arrange(t1, t2, ncol=2)
@
  \caption{Histograms of log time taken to make a decision. r+ and r- indicate the sign of true plot correlation. The vertical red line is the average time of the particular group. On average scatter plots are faster and the true plot is detected more often, with r-.}
    \label{time} 
\end{figure}

Subjects were asked to give a number from 1 to 5 to rate how confident they were in their selection (1 - most confident, 5 - least confident, average = 3.34). There were no statistically significant differences between plot designs, however subjects were slightly less confident on true plot detections (by $0.25 \pm  0.16, p=0.002$) and every one percent extra time required to make a decision increased confidence (by $0.19 \pm 0.07, p<0.001$). One possible explanation is that subjects often focused on a null plot and selected it confidently. This would suggest that the null plots often produced visually interesting features subjects assumed were caused by the association between plots.
Subjects spending longer on a plot immediately draws two hypotheses:

\begin{enumerate}
  \item The lineup was more difficult and it took longer to determine which plot was best, reducing confidence.
  \item The subject selected a plot relatively easily but spent extra time going over its details and making comparisons to alternative plots, increasing confidence.
\end{enumerate}

The result suggests that the latter hypothesis is more likely to be correct. We also found a negative relationship between plot detection and time spent on a lineup, supporting the major relationship found: quick decisions are less confident but more likely to successfully detect the true plot.

\begin{figure}[htbp]
  \centering
<<confidence, dependson='setup', fig.width=6, fig.height=6, out.width='0.7\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
conf <- select(full, design, conf_level, correct, rsgn)
conf$rsgn <- factor(conf$rsgn, labels=c("r -","r +"))
cm <- summarise(group_by(conf, correct, design, rsgn), m=mean(conf_level))
c1 <- ggplot(data=subset(conf, correct==0), aes(x=conf_level)) + geom_histogram(binwidth=1) + 
  geom_vline(data=subset(cm, correct==0), aes(xintercept = m), color="red") +
  facet_grid(design~rsgn) + labs(title="Did not detect", x="Confidence", y=element_blank())  
c2 <- ggplot(data=subset(conf, correct==1), aes(x=conf_level)) + geom_histogram(binwidth=1) + 
  geom_vline(data=subset(cm, correct==1), aes(xintercept = m), color="red") +
  facet_grid(design~rsgn) + labs(title="Detected", x="Confidence", y=element_blank())  
grid.arrange(c1, c2, ncol=2)
@
  \caption{Histograms of reported confidence levels. 1 - Low, 5 - High. Detection had a minor detrimental impact on confidence, whilst time taken increased it. There was no statistically significant difference on confidence level between plot designs.}
  \label{confidence} 
\end{figure}

%Next we analysed the individual ability of each of our subjects. Figure~\ref{confidence2} plots the accuracy of each subject on the y axis versus the average confidence on the x axis. While we expected that subjects with greater visual skills would report lower values of confidence, there does not appear to be any relationship between the two. The accuracy is normally distributed (Jarque-Bera p value = 0.418), with a mean of 42.5\% and a standard deviation of 17.8\%. The strongest subject successfully identified the true plot 91\% of the time, whilst the weakest was never able to identify the true plot. As all subjects were shown a balanced mix of true plot correlations, the extreme difference in individual visual ability is striking. Visual ability will be controlled via a subject specific random effect in the models in the next section.

Figure~\ref{confidence2} compares the reported confidence for each selection against the true plot correlation. The results are mixed with no consistent increase or decrease in confidence for changes in correlation across different plot designs and sample size. Strong correlation appears to decrease confidence more often than it is increased.

\begin{figure}[htbp]
  \centering
<<confidence2, dependson='setup', fig.width=6, fig.height=6, out.width='0.9\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
conf2 <- select(full, conf_level, r, smoothed, n, design) #c(7, 15
qplot(r, 6-conf_level, data=conf2, alpha=I(0.1)) + 
  geom_smooth(method="loess") + 
  facet_grid(n~design+smoothed) +
  labs(x = "Correlation", y = "Confidence (5=High, 1=Low)")
@
  \caption{Reported confidence against true plot correlation. Darker points indicate a confidence level selected more regularly. Strong correlation is more often associated with a lower reported confidence.}
 \label{confidence2}
\end{figure}

\subsection{Selection analysis}

There are two major reasons why a subject may fail to detect the true plot out of a lineup. Either the plot design was unable to display correlation for that particular combination of factors (and the analyst would commit a Type II error by not rejecting the null hypothesis), or a null plot that had a greater correlation than the true plot was selected instead (and the analyst would commit a Type I error by rejecting the null hypothesis). We next can analyse these kinds of errors with the pin plots found in Figure~\ref{pins}. The frequency of selection of each plot in each lineup, either true or null, is represented by a pin, placed at the value of the correlation between the two variables shown in that plot. The correlation of the selected plot is on the x-axis and the frequency of selection is plotted on the y axis. The correlation of the simulated (true) data ($\rho \in \pm \{0.3, 0.5, 0.7, 0.9\}$ in the plot, the sample size ($t \in \{12, 24, 48, 96\}$), and the plot design (unsmoothed/smoothed, scatterplot/lineplot) is used to facet the plots. The blue pins indicate the true plots and the red pins indicate the null plots. If the plot design has the power to display correlation, we would expect the highest correlation plot to be selected, leading to a Type I error where a plot from the null distribution spuriously had high correlation. We find that this is generally the case at higher sample sizes, but at $t = 12$, the selected plos have a wider range of correlations, indicating a general inability of the plot designs to differentiate betwen high and low correlaiton, a type II error. With the catter plot, for the scatter plot, whenever there was a plot with a very strong correlation, usually in the form of the true plot, it was overwhelmingly selected. For the overlaid lines this mostly occurred only for plots with strong positive correlation, which is consistent with its general lack of power to reveal negative relationships in data.

\begin{figure}[htbp]
  \centering
%  \begin{subfigure}[htbp]{1\textwidth}
%\subcaption{}\label{pin1}
<<pins, dependson='setup', fig.height=4, fig.width=7, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
count.setup <- select(full, pic_id, response_0, actual_location, design, smoothed, n)
count.setup$response_0 <- as.character(count.setup$response_0)
counts <- data.frame(table(count.setup$pic_id, count.setup$response_0))
counts$Var1 <- as.numeric(as.character(counts$Var1))
counts$Var2 <- as.numeric(as.character(counts$Var2))
colnames(counts) <- c("lineup", "plot", "count")
pin <- merge(counts, bylineup, by.x="lineup", by.y="pic_id")
pin <- select(pin, lineup, plot, count, smoothed, actual_location, n, design, r, evals)
pin <- mutate(pin, r = factor(r),
              true = ifelse(plot==actual_location, "True plot", "Null plot"), 
              percent = count/evals,
              base = 0,
              correlation = diag(as.matrix(correl[pin$lineup, (pin$plot+2)])))
pin$design <- factor(pin$design, levels=c("Scatterplot", "Overlaid Lines"), labels=c("scat","lines"))
pin$smoothed <- factor(pin$smoothed, levels=c("0", "1"), labels=c("N","Y"))
#ggplot(data=filter(pin, smoothed==0 & design == "Scatterplot")) + 
#         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
#         geom_point(aes(x=correlation, y=percent, colour=true))  + facet_grid(n~r) +
#         scale_x_continuous(breaks=c(-1,0,1)) + scale_y_continuous(breaks=c(0,0.5,1)) +
#         labs(y="Proportion",x="") + theme(legend.position="none")
@
%\end{subfigure}
%\begin{subfigure}[htbp]{1\textwidth}
%\subcaption{Scatterplots}\label{pin2}
<<pin2, dependson='pins', fig.height=8, fig.width=7, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
ggplot(data=subset(pin)) + 
         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
         geom_point(aes(x=correlation, y=percent, colour=true)) + facet_grid(design+n+smoothed~r) + 
         scale_x_continuous(breaks=c(-1,-0.5,0,0.5,1), labels=c("1","-","0","+","1")) + 
         scale_y_continuous(breaks=c(0,0.5,1)) +
         labs(y="Proportion Selected",x="Correlation between series") + theme(legend.position="none")
#ggplot(data=subset(pin, design == "Scatterplot")) + 
#         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
#         geom_point(aes(x=correlation, y=percent, colour=true)) + facet_grid(n+smoothed~r) + 
#         scale_x_continuous(breaks=c(-1,0,1)) + scale_y_continuous(breaks=c(0,0.5,1)) +
#         labs(y="Proportion",x="Correlation") + theme(legend.position="none")
#ggplot(data=subset(pin, smoothed==1 & design == "Scatterplot")) + 
#         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
#         geom_point(aes(x=correlation, y=percent, colour=true)) + facet_grid(n~r) + 
#         scale_x_continuous(breaks=c(-1,0,1)) + scale_y_continuous(breaks=c(0,0.5,1)) +
#         labs(y="Proportion",x="") + theme(legend.position="none")
@
%\end{subfigure}
%\begin{subfigure}[htbp]{1\textwidth}
%\subcaption{}\label{pin3}
<<pin3, dependson='pin', fig.height=5, fig.width=7, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
#ggplot(data=subset(pin, smoothed==0 & design == "Overlaid Lines")) +
#         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
#         geom_point(aes(x=correlation, y=percent, colour=true)) + facet_grid(n~r) +         
#         scale_x_continuous(breaks=c(-1,0,1)) + scale_y_continuous(breaks=c(0,0.5,1)) +
#         labs(y="Proportion",x="") + theme(legend.position="none")
@
%\end{subfigure}
%\begin{subfigure}[htbp]{1\textwidth}
%\subcaption{Overlaid line graphs}\label{pin4}
<<pin4, dependson='smoothing', fig.height=5, fig.width=7, out.width='0.9\\textwidth', echo=FALSE, warning=FALSE, message=FALSE, error=FALSE>>=
#ggplot(data=subset(pin, design == "Overlaid Lines")) + 
#         geom_errorbar(aes(x=correlation, ymin=base, ymax=percent, colour=true)) +
#         geom_point(aes(x=correlation, y=percent, colour=true)) + facet_grid(n+smoothed~r) + 
#         scale_x_continuous(breaks=c(-1,0,1)) + scale_y_continuous(breaks=c(0,0.5,1))+
#         labs(y="Proportion",x="Correlation") + theme(legend.position="none")
@
%\end{subfigure}
  \caption{Pin plots showing subject selections for each plot in all lineups for all factor levels in the experiment, with duplicates of the same factor combination together. Proportion of times a plot was selected is on the y-axis and correlation between the pair of time series on the x-axis. Blue indicates a true data plot and red indicates a null plot. Generally, subjects selected the plot from the lineup with the strongest association in larger sample sizes. For overlaid lines, subjects tended to more often pick the plot showing the strongest positive association, even if the relationship in the true plot was negative.}   
    %: (\subref{pin1}) unsmoothed scatterplots, (\subref{pin2}) smoothed scatterplots, (\subref{pin3}) unsmoothed overlaid lines, (\subref{pin4}) smoothed overlaid lines. On the x-axis is the correlation of the plot (there typically is spurious correlation in the null plots), and on the y-axis is the proportion of times a plot was chosen. Each plot contains data from the three lineups with the requisite factor combination. Red pins indicate null plots and blue indicates the true plot, occasionally the true plot does not seem to appear as it has a proportion of zero and is visually blocked by the red pins. Strongly correlated plot seem to be selected regardless of if the relationship is real or spurious.}
  \label{pins}
\end{figure}

Occasionally there was an anomaly where a single null plot was selected with a high frequency despite having a weak correlation, as the null distribution can generate visually interesting patterns that draw the analyst’s attention. We asked subjects for reasons behind their selections and found that choosing on the basis that “the points in the plot were close together”, or had “a unique pattern” more often led to an unsuccessful detection. Conversely, decisions based on “matching the peaks and troughs” of the overlaid lines or “seeing a line formed in the dots” of the scatter plot were more likely to be successful in identification of the true plot. However, these true detections are, on average, made less confidently than unsuccessful evaluations, which indicates that decisions made on the basis of closeness or patterns are more convincing of an association than the more successful reasons.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{anom1}
  \caption{A lineup of unsmoothed scatter plots with a sample size of 24. The true plot is located in position 4 with a correlation of -0.491, and was selected by one out of eleven observers (p = 0.33). The plot in position 6 was selected by nine of these eleven observers (p < 0.001) despite having a correlation of -0.037. Subjects most often cited the clustering effect shown as the reason for their selection, an example of the over-interpretation of randomness.}
    \label{anom1}
\end{figure}

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.9\textwidth]{anom2}
  \caption{A lineup of unsmoothed overlaid lines with a sample size of 12 and true plot correlation of -0.707 in position 8, which was not detected in five evaluations. Instead, the null plot in position 13 was selected four times, (p –value < 0.001) despite having a correlation of only 0.225. When asked for a reason of their selections, subjects typically answered that the lines were close together, another misleading visual artefact of the random data as the subjects ignore the eventual separation in the lines, similar to in figure ~\ref{cny}.} 
  \label{anom2}
\end{figure}

Figures~\ref{anom1} and \ref{anom2} show examples of this in some lineups where a null plot with an unusually strong pattern led to selection of a null over the data plot. In Figure~\ref{anom1}, the true data plot is in position 4, but 9 out of 11 subjects chose plot 6. Plot 6 is unusual, most of the points are scattered around the vertical median, but there are two outlying points, one with a high y-value and one with a low y-value.  Other concurrent work has suggested that outliers are pre-attentive patterns and this plot selection by the observers is consistent with that. Figure~\ref{anom2} above is another example of where subjects chose a null plot that had smaller correlation than the data plot. Four out of five evaluations selected the null plot in position 13, instead of the data plot in position 8. Plot 13 has a correlation of 0.225, smaller (in absolute value) than the -0.707 of plot 8. This selection gives evidence towards Wertheimer’s gestalt law of common fate, as subjects may be selecting this plot focused on the joint path of the two series early in Plot 13 and ignored the contradictory evidence of no relationship as they diverged later in the plot similar to the series in figure ~\ref{cny}.  It also confounded by the seeming preference to find positive correlation over negative correlation, and we suspect that if one of the series in plot 8 had the values flipped, then the strong positive correlation may have been noticed.

\subsection{Modelling graphical power}

A mixed effects logit model is used separately for scatter plot and overlaid line evaluations. The tested model is:

\begin{eqnarray} 
\label{modelpower}
Power_{ijklt} &=& \psi \{ \beta_{0} + \beta_{1} x_{1,i}  + \beta_{2} x_{2,j}  + \beta_{3} x_{1,i} \* x_{2,j} + \beta_{4} x_{3,k} + \nonumber \\
 & & \beta_{5} x_{4,t} + x_{5,l}(\beta_{6} + \beta_{7} x_{1,i}  + \beta_{8} x_{2,j}  + \nonumber \\
 & & \beta_{9} x_{1,i} \* x_{2,j} + \beta_{10} x_{3,k} + \beta_{11} x_{4,t})  + \gamma z_{s} + \epsilon_{ijklt} \} \nonumber
\end{eqnarray}

Where $\psi$ is the logit link function, $\psi(x) = \frac{e^{x}}{1+e^{x}}$. Additionally:

\begin{itemize}
\item $x_{1,i}$ is the absolute value of the true plot correlation, $\rho$, with $i \in \{0.3, 0.5, 0.7, 0.9\}$,
\item $x_{2,j}$ is a dummy variable for the sign of the true plot correlation, with $j = 0$ for negative correlation, 
\item $x_{3,k}$ is a dummy variable for the smoothing procedure, with $k = 0$ for unsmoothed data, 
\item $x_{4,t}$ is the sample size, with $t \in \{12, 24, 48, 96\}$ ,
\item $x_{5,l}$ si a dummy variable for the plot design, with $l = 0$ for the overlaid lines type,
\item $z_{s}$ is an observer specific random effect, with $s = 1, 2, \dots , 160$.
\end{itemize}

It is assumed that $\epsilon_{ijkt} \sim \mathcal{N} (0, \Sigma_{ijkt})$ and $\gamma \sim \mathcal{N} (0, \delta)$.

If the true plot correlation were equal to $0$, it should be indistinguishable from the null plots and the detection rate would fall to $1/20$, the Type I error rate. An offset is included in the model to incorporate this effect.

If a subject made multiple selections from the same lineup, each selection is treated as a separate data point with a probability weighting equal to the inverse of the number of selections.

\begin{table}[htp]
\caption{Weighted Mixed Effects Logit Model for the predicted power of both plot designs. Sample size for scatter plot evaluations: 859, overlaid lines: 903, total: 1762. Default plot design is overlaid lines. The model is fitted using the GLIMMIX procedure in SAS.}
\label{tab1}
\centering
<<powerfit, dependson='setup', results='asis', echo=FALSE, message=FALSE, error=FALSE, warning=FALSE>>=
powercoef <- powercoef[,-1]
rownames(powercoef) <- c("Intercept", "abs(r)", "sign(r)", "abs(r)*sign(r)", "smoothed", "t", "design", "design*abs(r)", "design*sign(r)", "design*abs(r)*sign(r)", "design*smoothed", "design*t")
colnames(powercoef) <- c("Estimate", "Standard Error", "df", "t-value", "p-value")
kable(powercoef, format="latex", digits=2)
@
\end{table}

Table \ref{tab1} summarises the model estimates. It can be confusing the interpret the model estimates, and at first glance it appears the negative plot coefficient implies that overlaid lines are substantially more powerful than the scatter plot, but this is misleading and interpreting the full model involves combining the estimates for the different interactions with $\rho$. $\beta_{2}$ and $\beta_{8}$ model the power difference between low and high correlations, while $\beta_{3}, \beta_{4}, \beta_{9}$ and $\beta_{10}$ model any increase in power appearing only in positive correlation. 
Taking these into account, the scatter plot shows superior performance relative to the overlaid lines at higher values of abs($\rho$). It needs to be restated that we only tested correlations at an absolute value $\in \{0.3, 0.5, 0.7, 0.9\}$, so the results below an absolute value of $0.3$ are unreliable and may be artifacts of regression smoothing methods.

%The correlation:sign interaction shows that the scatter plot has close to symmetric performance for positive (correlation:sign + correlation) and negative (-correlation) correlation while the overlaid line graph is substantially weaker for negative values of correlation.
Smoothing out the excess 'noise' from a time series had a minimal effect on scatter plots, demonstrated by $\beta_{4} + \beta_{10}$, but it did have a significant positive impact on the power of overlaid line graphs as smoothing induces a large visual difference (See Figure~\ref{smoothing}. Increasing the sample size, $t$, improved the power of the scatter plot but it was not statistically significant for the overlaid line graph possibly due to the negative correlation results interfering with the modeling (See Figure~\ref{samplesize}).

Figure~\ref{powermodel} shows the fitted unrestricted model, with power predictions (y-axis) plotted against correlation (x-axis) for each of the four sample sizes tested. The thicker lines represent the fixed effect fit, or the average performance across all tested individuals whilst the thinner lines represent the random effects fit, the variation in power that can be attributed to an individual's visual skill. \citet{Majumder:2014} show that this random visual skill is not impacted by demographics such as age and gender; whilst education increases the power of a graphic display by only approximately one percent, not a practically significant amount. We did not include these characteristics of the individual in the modelling due to this lack of significance. Figure~\ref{powermodel} especially highlights the weak performance of the overlaid lines and the benefits of smoothing and increased sample sizes.

\begin{figure}[htbp]
 \centering
<<powermodel, dependson='setup', fig.height=8, fig.width=8, out.width='0.9\\textwidth', echo=FALSE, error=FALSE, message=FALSE, warning=FALSE>>=
predplot <- prediction %>% filter(is.na(detected)) %>%
            select(r, t, rsgn, smoothed, design, subj_id, RandomEf, FixedEf) %>%
            mutate(RandomEf = exp(RandomEf)/(1+exp(RandomEf)), 
                   FixedEf = exp(FixedEf)/(1+exp(FixedEf)))
##Logit link function transform

qplot(r, RandomEf, data=predplot, color=design, 
      group=interaction(subj_id, design), geom="line", alpha=I(0.1)) + 
  scale_color_manual("Plot Design", values=c("#33AA33", "#9977EE", "#FF9933", "#00DDAA")) + 
  xlab("Correlation") + ylab("Estimated Power") +
  geom_line(aes(y=FixedEf), size=2, alpha=1) + theme_bw() + facet_wrap(~t, ncol = 2) + theme(aspect.ratio=1.2)
@
  \caption{Predicted power fit from the mixed effects logit power model. The thicker lines shows the fixed effects prediction while the thinner lines show the range of random effects, the observer's visual ability. The weakness of the overlaid line graph in negative correlation and gain from increasing scatter plot sample size or smoothing overlaid lines are highlighted.}
  \label{powermodel}
\end{figure}

\iffalse
This unrestricted model used in to test the symmetry restriction for the scatter plot design was written earlier as:

$$Power_{UR,ijkt} = \psi\left\{ \beta_{0} + \beta_{1} x_{1,i}  + \beta_{2} x_{2,j}  + \beta_{3} x_{3,k} + \beta_{4} x_{4,t} + \beta_{5} x_{1,i} \* x_{2,j} + \gamma z_{s} + \epsilon_{ijkt} \right\}$$

Testing if the scatter plot displays positive and negative correlation equally is equivalent to testing the hypothesis:

\begin{itemize}
  \item $H_{0}: \beta_{2} = \beta_{5} = 0$
  \item $H_{1}:$ Either equality does not hold.
\end{itemize}

The restricted model is therefore:

$$Power_{R,ijkt} = \psi\left\{ \beta_{0} + \beta_{1} x_{1,i} + \beta_{3} x_{3,k} + \beta_{4} x_{4,t} + \gamma z_{s} + \epsilon_{ijkt} \right\}$$

The unrestricted model has a log-likelihood value of -346.9, where the restricted model has a log-likelihood of -355.8. The test statistic is therefore, with a null distributed as a $\chi^{2}_{2}$ variable. The null hypothesis is rejected (p = 0.02). We find that the scatter plot has asymmetric performance over negative and positive values of correlation. This may be a result of unwittingly introducing bias into the methodology, the first scatter plot shown to subjects in the instruction set always had strong positive correlation. 
A similar setup rejects the symmetric performance of the overlaid line graphs with a log likelihood ratio test statistic of 105.30 (p-value < 0.001). Such an extreme result is due to a systematic difference in the overlaid line graph and cannot be attributed simply to bias.
\fi

\subsection{Modelling time taken}

The time taken to evaluate a lineup is recorded and log transformed to reduce the impact of several large ($> 500$ seconds) outliers.

\begin{eqnarray} 
\label{timemodel}
ln(time_{ijklt}) &=& \{ \beta_{0} + \beta_{1} x_{1,i}  + \beta_{2} x_{2,j}  + \beta_{3} x_{1,i} \* x_{2,j} + \beta_{4} x_{3,k} + \nonumber \\
 & & \beta_{5} x_{4,t} + x_{5,l}(\beta_{6} + \beta_{7} x_{1,i}  + \beta_{8} x_{2,j}  + \nonumber \\
 & & \beta_{9} x_{1,i} \* x_{2,j} + \beta_{10} x_{3,k} + \beta_{11} x_{4,t})  + \gamma z_{s} + \epsilon_{ijklt} \} \nonumber 
\end{eqnarray}

Other than the removal of the link function, the structure of the model identical to that in Equation~\ref{modelpower}.

\begin{table}[htp]
\caption{Weighted Mixed Effects Linear Model for time taken to make a selection, default plot style is overlaid lines. Fitted with the MIXED procedure in SAS.}
\label{tab2}
\centering
<<timefit, dependson='setup', results='asis', echo=FALSE, message=FALSE, error=FALSE, warning=FALSE>>=
time.fit <- time.fit[c(1, 5, 6, 2, 3, 4, 7, 8, 12, 13, 9, 10, 11, 14),] 
##separate into main effects then interaction effects in same order
rownames(time.fit) <- c("Intercept", "Smoothed", "t", "abs(r)", "sign(r)", "abs(r)*sign(r)", "detected", "design", "design*Smoothed", "design*t", "design*abs(r)", "design*sign(r)", "design*abs(r)*sign(r)", "design*detected")
time.fit$Effect <- NULL
colnames(time.fit) <- c("Estimate", "standard error", "df", "t-value", "p-value")
kable(time.fit[1:14,], format="latex", digits=2)
@
\end{table}

Table~\ref{tab2} shows a similar story for time as Table~\ref{tab1} did for power, that the overlaid lines have better performance close to zero correlation, but the scatter plot is much faster at high correlations. Postivie correlation reduces time for both plot designs, but only the scatter plot sees improvements in negative correlation, time to detect is flat accross negative correlations for the overlaid line graph. Neither smoothness nor sample size has an impact on the time for either plot design, despite having statistically significant impacts on power. 
%An overlaid line graph having positive correlation increases the time required, eliminating the benefits from the almost significant plot design term. This result seems counter-intuitive as the weak performance of the overlaid lines in negative correlation implies that negative would require more time than positive, but the interactions between plot design, correlation and sign result in there being almost no decrease in time for strongly negatively correlated line plots.

\section{Conclusion}

Power
-In low sample sizes, both designs have similarly weak performance - 
-Relationships between t=12 time series is dubious at best, requires $\rho$ > 0.7 to have more than 50\% power to see.
-In higher sample sizes, the scatterplot improves dramatically while the line plot remains mostly unchanged
-Positive correlation is very similar between the two designs
-The line plot has much weaker performance in negative correlation
-If possible, smooth the lines to improve performance - noisy data hurts visual performance
-If negative association is expected, flip one line to make it positive

Time
-The scatterplot is faster than the lines, particularly at high correlation
-Smoothing and sample size have no real effect

Confidence
-People could be distracted by spurious patterns in the null plots
-These were typically very confident selections
-The data analyst needs to understand what kinds of patterns can be generated randomly

Overall
-Despite hiding trends etc., the seemingly simplistic scatter plot seems to be better suited to these tasks.

\iffalse
This paper showed that the scatterplot is better than the overlaid lines for displaying two time series when the purpose is to examine the association between the two series. This is especially true if correlation between the series is negative. This comparison was made possible by the use of the lineup protocol for comparing plot designs. The power of two plot designs for reading correlation between variables was modelled using a logistic regression incorporating subject’s individual visual ability as a random effect. We found that the scatterplot, in most situations is both more powerful and quicker to process than the overlaid line graph. The overlaid line graph appeared to have stronger performance at extremely low correlations, but this work did not test absolute values of correlation below 0.3. Other non-linear associations may exist in data that are not measured by correlation, and it may be of interest to compare plot power with a broader range of true plot features.
Investigation of the individual selections of the lineup evaluations finds some results in-line with Wertheimer’s Gestalt Law of common fate, as demonstrated in Figure 15. The law suggested that people would find relationships between two lines that briefly move together, regardless of whether the relationship actually existed. Whilst this is shown, the overlaid line graph did not have a corresponding increase in confidence compared to the scatter plot; so research into similar confidence-boosting visual anomalies that can occur in scatter plots would be necessary to further investigate the law. Diaconis's Magical Thinking was supported by the finding on the relationships between time, confidence and true plot detection. A more difficult true plot may be selected rarely and require a longer to make a decision; resulting in decreased subject confidence. However, we found that time increases confidence but reduces the detection rate. Highly confident null selections suggests that subjects may have been distracted by strong, spurious patterns in the null plots.
If a data analyst is required to use overlaid line graphs in their work, substantial improvements to power and hence readability can be made by smoothing time series data if the sample size is long enough for these techniques to become feasible.
\fi

\section{Acknowledgements}

The authors gratefully acknowledge funding from the National Science Foundation Grant DMS \#1007697.

The models used for this paper were generated using SAS software, Version 9 of the SAS System for Unix, with the MIXED and GLIMMIX procedures. Copyright \copyright\ 2015 SAS Institute Inc. SAS and all other SAS Institute Inc. product or service names are registered trademarks or trademarks of SAS Institute Inc., Cary, NC, USA.

Further analysis was carried out using \citet{R} version 3.1.2 and \citet{Rstudio} version 0.99.484 using the packages: dplyr, ggplot2, grid, gridExtra, gridSVG, knitr, maps, mFilter, mnormt, nullabor, poibin, stringr and tidyr.
%\citet{dplyr}, \citet{ggplot2}, \citet{grid}, \citet{gridExtra}, \citet{gridSVG}, \citet{knitr}, \citet{maps}, \citet{mFilter}, \citet{mnormt}, \citet{nullabor}, \citet{poibin}, \citet{stringr} and \citet{tidyr}.

All code required to reproduce this paper can be found at https://github.com/NTomasetti/correlation.

\bibliographystyle{asa}
\bibliography{references}

\end{document}

### 2017-01-03 Confirmation/confirmation_outline.tex
Confirmation work

@@
 
-add something for $\rho$ to enforce stationarity - maybe switch to an AR(1) or ARMA(1,1).
+could make $\rho$ a constant or add a uniform unit circle prior to polynomial roots to enforce stationarity - or maybe switch to an AR(1) or ARMA(1,1).
 
@@
 Maximising the ELBO with respect to $q_i$ is analytically involved but computationally simple, as (\ref{ELBO}) can be expressed as a function of $q_i$ only and then maximised with respect to each $q_i$ individually.
-
-\begin{align}
-\mathcal{L}(q_i) &= \int_{\theta} q_i q_{\setminus i} \ln (p(y, \theta)) d\theta - \int_{\theta} q_{i}q_{\setminus i} \ln (q_{i}q_{\setminus i}) d\theta \nonumber \\
-&= \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} ( \ln (p(y, \theta)) - \ln(q_{i})) d\theta_i d\theta_{\setminus i} - \int_{\theta_i}\int_{\theta_{\setminus i}} q_{i}q_{\setminus i} \ln (q_{\setminus i}) d\theta_i d\theta_{\setminus i}  \nonumber \\
-&= \int_{\theta_{i}} q_{i} \left( \int_{\theta_{\setminus i}} q_{\setminus i} \ln (p(y, \theta )) d\theta_{\setminus i} - \ln(q_i) \right) d\theta_i \nonumber \\
-&- \int_{\theta_{\setminus i}} q_{\setminus i} \ln(q_{\setminus i}) d\theta_{\setminus i} \label{mf2}\\
-&= \int_{\theta_{1}} q_{i} \ln \left( \frac{\exp(  \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])}{q_i} \right) d\theta_i + c \nonumber \\
-&= -KL (q_i || \exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])) + c \label{mf2.2}
-\end{align}
-
-as the rightmost component of (\ref{mf2}) is constant with respect to $q_i$.
-
-We can see that by holding $q_{\setminus_i}$ fixed, we can maximise the ELBO with respect to $q_i$ by minimising the KL Divergence between $q_i$ and $\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])$ in (\ref{mf2.2}).
-This is easily done by setting
+\citet{Attias1999} shows that
 
 \begin{equation}
-\label{mf3}
-q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))]).
+\label{mf2}
+q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])
 \end{equation}
 
-If the likelihood and prior for $\theta_i$ form an exponential family conjugate pair, then this maximisation proceeds by keeping the distributional family constant and simply updating the parameters $\lambda_i$.
-Otherwise, we can find the optimal distributional family for any $q_i$ by taking the log-joint density and then ignoring any term that doesn't depend on $\theta_i$.
-Next, take the expectation of any remaining $\theta_{j \neq i}$ and then match to the kernel of a known distribution.
+and maximisation can proceed by matching (\ref{mf2}) to a known distribution.
+In the case that the likelihood and prior for $\theta_i$ form an exponential family conjugate pair, then the distributional family can be kept constant and all that is required is an update the parameters $\lambda_i$.
 If it does not match a known distribution, we may need to make a further approximation $\tilde{q_i(\theta_i|\lambda_i)}$ which has a recognizable distribution.
@@
 
-These steps are almost the same as those used to find a conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, with dependence on other parameters replaced by their expectations.
+Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | y, \theta_{\setminus i}$ in a Gibbs MCMC scheme, with dependence on other parameters replaced by their expectations.
 Hence, the optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family.
-Hence deriving the updating algorithm for MFVB takes about the same analytical effort as finding the conditionals for a Gibbs MCMC scheme, and the secondary Laplace approximation will occur in parameters where a Metropolis-Hastings-within-Gibbs step must be used.
+The secondary Laplace approximation is analogous to a Metropolis-Hastings-within-Gibbs step in MCMC, as a way to handle unrecognisable distributions.
 \vspace{5mm}
 
-Given these optimal distributional families, each $\lambda_i$ can be found as a function of the data and $\lambda_{j \neq i}$.
-This dependence requires an algorithm that continuously iterates between each $\lambda_i$ and sets it to its maximising value until $\mathcal{L}(q(\theta | \lambda), y)$ converges with some pre-defined threshold.
-This is known as a coordinate ascent algorithm and follows below using $k$ as the dimension of $\lambda$.
+Given these optimal distributional families, a mean field updating equation for each $\lambda_i$ can be found as a function of the data and other $\lambda_{j \neq i}$.
+This dependence requires an algorithm that continuously iterates between each $\lambda_i$ and sets it to its maximising value until $\mathcal{L}(q(\theta | \lambda), y)$ converges within some pre-defined threshold.
+This is known as a coordinate ascent algorithm and follows below, where $k$ as the dimension of $\lambda$.
 
@@
 \subsubsection{Stochastic Variational Bayes}
-\emph{Note the change to subsubsection}
 
@@
 \label{SGA1}
-\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m-1)}), y),
+\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m)}), y),
 \end{equation}
 
-where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m-1)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m-1)}), y)$ with respect to each element of $\lambda$.
-This update requires initial values $\lambda^{(0)}$ and some sequence $\rho^{(m)}$.
+where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ with respect to each element of $\lambda$.
+This update requires initial values $\lambda^{(0)}$ and some learning rate sequence $\rho^{(m)}$.
 If $\rho^{(m)}$ is chosen to satisfy the following conditions, it is a Robbins-Monro sequence and the algorithm is guaranteed to converge to a local maximum.
@@
 We must restrict the approximation to distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable.
-In this case, we can calculate derivatives of the ELBO.
-
-\begin{align}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \nabla_{\lambda} \int_{\theta} q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
-&=   \int_{\theta} \nabla_{\lambda}[  q(\theta | \lambda) \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)] d\theta \label{SGA2} \\
-&=  \int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta)) - \ln(q(\theta | \lambda))] d\theta \nonumber \\ 
-&+ \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \label{SGA3}
-\end{align}
-
-Note that the first integral on the right hand side of (\ref{SGA3}) equals 0 as
+In this case \citet{Ranganath2014} shows that a Monte Carlo estimate of the derivative of the ELBO can be taken by
 
 \begin{equation}
-\int_{\theta}  q(\theta | \lambda) \nabla_{\lambda}[( \ln (p(y, \theta))] d\theta = 0
+\label{SGA2}
+\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda^{(m)}))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda^{(m)})))
 \end{equation}
 
-and
+where $s = 1, \dots, S$ indicates simulations from $q(\theta | \lambda^{(m)})$.
 
-\begin{align}
-\int_{\theta} - q(\theta | \lambda) \nabla_{\lambda}[ \ln(q(\theta | \lambda))] d\theta &= - \int_{\theta} \frac{\nabla_{\lambda}q(\theta | \lambda)}{q(\theta | \lambda)} q(\theta | \lambda) d\theta \nonumber \\
-&=  -  \nabla_{\lambda}\int_{\theta} q(\theta | \lambda) d\theta \nonumber \\
-&=  0 \nonumber
-\end{align}
+As the distribution $q(\theta | \lambda)$ is specified by the user, the main restriction on the use of SVB is that the log-joint density $\ln(p(y, \theta))$ is able to be evaluated.
 
-Then we continue from (\ref{SGA3}) with
+\citet{Duchi2011} introduces the AdaGrad algorithm which can be implemented within SVB to control $\rho^{(m)}$.
+AdaGrad allows each $\lambda_i$ to have an independent $\rho^{(m)}_i$ that is inversely proportional to the gradient.
 
-\begin{align}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\theta} \nabla_{\lambda} [q(\theta | \lambda)] \left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) d\theta \nonumber \\
-&=   \int_{\theta} q(\theta | \lambda)  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right)d\theta \nonumber \\
-&= E_q \left(  \nabla_{\lambda} [\ln(q(\theta | \lambda))]\left( \ln (p(y, \theta)) - \ln(q(\theta | \lambda)) \right) \right). \label{SGA4}
-\end{align}
-
-We can compute a Monte-Carlo estimate of (\ref{SGA4}) by
+Let
 
 \begin{equation}
-\label{SGA5}
-\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda)) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)))
+\label{SGA3}
+G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\theta | \lambda^{(m)})\right)^2,
 \end{equation}
 
-where $\theta_s$, $s = 1, \dots S$, are simulated from the most recent iteration of  $q(\theta | \lambda)$.
+then each component's learning rate is defined as
 
-As the distribution $q(\theta | \lambda)$ is specified by the user, the main restriction on the use of SVB is that the log-joint density $\ln(p(y, \theta))$ is able to be evaluated.
-Running an SVB algorithm involves very little analytical effort unlike MFVB, most of the difficulty is calculating the score function of the approximating distribution,
-$\nabla_{\lambda} [\ln(q(\theta_s | \lambda))]$. But as this is part of the approximation and not the true model the code to run this could potentially be re-used in many different models.
+\begin{equation}
+\label{SGA4}
+\rho^{(m)}_i = \eta \left(G_i^{(m)}\right)^{-1/2}
+\end{equation}
+
+for some tuning parameter $\eta$.
+
+The resulting Stochastic Gradient Ascent algorithm proceeds below.
 
@@
  \While{Not converged}{
-  Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
+  Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m)})$\;
   \For{$i =  1$ \KwTo $k$}{
-      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
+      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m)})))$\;
+      Update $G_i^{(m)} = G_i^{(m-1)} + \left(\nabla_{\lambda_i}\mathcal{L}(q(\theta | \lambda^{(m)})\right)^2$\;
+      Calculate $\rho^{(m)}_i =  \left(G_i^{(m)}\right)^{-1/2}$\;
      }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
+  Set $\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)}  \nabla_{\lambda}$\;
   Set $m = m + 1$\;
  }
- \caption{Gradient Ascent for SVB}
+ \caption{Stochastic Gradient Ascent for SVB}
   \label{alg:algorithm2}
@@
 
-\subsection{Variance Reduction Techniques} \emph{This is general stuff, not just for VB, so can be a subsection}
+\subsection{Variance Reduction Techniques}
+
+For many models the estimator in (\ref{SGA2}) has a large variance, so a more efficient estimator will allow a lower value of $S$ and hence reduced computation time.
+\subsubsection{Rao Blackwellisation}
+
+\citet{Ranganath2014} considers the use of Rao Blackwellisation \citep{Casella1996}
 -Rao Blackwellisation

### 2017-01-05 Confirmation/confirmation.tex
Mostly wrote more on exponential smoothing

--- 2017-01-03 Confirmation/confirmation_outline.tex
@@
 \usepackage{url}
+\usepackage{wasysym}
 \usepackage{ulem}
@@
 These approximations have an implicit trade-off: The better approximations are compuationaly intensive, and the forecaster must decide how much computation time and approximation error is acceptable.
-The focus of this research is in situtations where the time budget is too small for MCMC to be reliable, and uses VB as an alternative.
-This situtation is common in time-series forecasting, which must update the posterior distribution for each new data point observed via
+The focus of this research is in situtations where MCMC is feasible in a moderate amount of time, but is infeasible in time period between two observations.
+To move $p(y_{T+1} | y_{1:T})$ to $p(y_{T+2} | y_{1:T+1})$ requires updating the posterior distribution after observation of each new data point observed via
 
@@
 \end{equation}
+
+However, if the data is based on a market such as with stock returns the market will close overnight and MCMC can be ran.
 
@@
 \label{AR2}
-y_t = \mu + \rho_1 y_{t-1} + \rho_2 y_{t-2} + \epsilon_t
+y_t - \mu = \rho_1 (y_{t-1} - \mu) + \rho_2 (y_{t-2} - \mu) + \epsilon_t
 \end{equation}
 
-where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. We collect the unknown parameters as $\theta = (\mu, \rho_1, \rho_2, \sigma^2)'$ and set priors as
+where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. We collect the unknown parameters as $\theta = (\mu, \sigma^2, \rho_1, \rho_2)'$ and set priors as
 
@@
 p(\mu) &\propto 1 \nonumber \\
-p(\sigma^2) &\propto \sigma^{-2} \nonumber
+p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
+\mbox{Some kind of }&\rho \mbox{ prior}
 \end{align}
 
-could make $\rho$ a constant or add a uniform unit circle prior to polynomial roots to enforce stationarity - or maybe switch to an AR(1) or ARMA(1,1).
+as this prior on $\rho_1$ and $\rho_2$ is uniform over the triangle that defines the AR(2) stationary region.
 
@@
 
-\subsubsection{The Mean Field Assumption}
+\subsubsection{Mean Field Variational Bayes}
 
-Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and uses the assumption that the approximation dististribution factorises, 
+Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the search for an optimal approximating dististribution to the set of factorisable distributions
+
 \begin{equation}
@@
 
-This assumption is known as the Mean Field assumption and is has been widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}).
+This technique is widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}), particularly in exponential family models \citep{Wainwright2008}.
 Each component $\theta_i$ may be a scalar or a vector, and $\lambda_i$ is the component of the $\lambda$ vector that parameterises the relevant factor $q_i(\theta_i |\lambda_i)$.
@@
 In many cases, the posterior is too complex to be captured for a factorising approximation to be reasonabl as we desire our distribution $q(\theta | \lambda)$ to capture dependence between parameters.
-In this case, the easy maximisation in MFVB is unavailable, and we must resort to what is called Stochastic Variational Bayes (SVB) using a gradient ascent algorithm developed by \citet{Paisley2012} and \citet{Ranganash2014}.
+In this case, the easy maximisation in MFVB is unavailable, and we must resort to what is called Stochastic Variational Bayes (SVB) using a gradient ascent algorithm developed by \citet{Paisley2012} and \citet{Ranganath2014}.
 
@@
 \begin{align}
-\lim_{m \xrightarrow[]{} \infty} \rho^{(m)} &= 0 \nonumber \\
-\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \nonumber \\
-\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty. \nonumber
+\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \\
+\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty.
 \end{align}
@@
 For many models the estimator in (\ref{SGA2}) has a large variance, so a more efficient estimator will allow a lower value of $S$ and hence reduced computation time.
-\subsubsection{Rao Blackwellisation}
+\subsubsection{Control Variates}
+\citet{Paisley2012} considers the use of control variates to augment the estimator of the gradient. Define
 
-\citet{Ranganath2014} considers the use of Rao Blackwellisation \citep{Casella1996}
--Rao Blackwellisation
--Control Variates
--Reparameterisation
+\begin{equation}
+\label{CV1}
+g(\theta, \lambda, y) =  \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda))),
+\end{equation}
+
+the function estimated in (\ref{SGA2}), then for some other function $h$, define
+
+\begin{equation}
+\label{CV2}
+\hat{g}(\theta, \lambda, y) = g(\theta, \lambda, y) - a(h - \mathbb{E}(h)).
+\end{equation}
+
+Then both $g(\theta, \lambda, y)$ and $\hat{g}(\theta, \lambda, y)$ have the same expectation and
+
+\begin{equation}
+\label{CV3}
+\mbox{Var}(\hat{g}) = \mbox{Var}(g) + a^2 \mbox{Var}(h) - 2a\mbox{Cov}(g, h). 
+\end{equation}
+
+Solving the polynomial (\ref{CV3}) in $a$ shows that the variance of $\hat{g}$ is minimised by
+
+\begin{equation}
+\label{CV4}
+\hat{a} = \mbox{Cov}(g, h)/\mbox{Var}(h).
+\end{equation}
+
+Substituting (\ref{CV4}) into (\ref{CV3}) yields
+
+\begin{equation}
+\label{CV5}
+\mbox{Var}(\hat{g}) = \mbox{Var}(g) - \mbox{Cov}(g, h)^2/\mbox{Var}(h).
+\end{equation}
+
+We need to choose some function $h$ that has a large covariance with $g$. \citet{Ranganath2014} suggests that the easy option of
+
+\begin{equation}
+\label{CV6}
+h(\theta, \lambda) = \nabla_{\lambda} [\ln(q(\theta_s | \lambda))].
+\end{equation}
+
+As $h$ is a score function, it has an expectation of zero and our estimator becomes
+
+\begin{equation}
+\label{CV7}
+\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)) - \hat{a}),
+\end{equation}
+
+where $\hat{a}$ is estimated from the sample variance and covariance of a subset of the $S$ Monte-Carlo draws.
+
+\begin{algorithm}[H]
+ \SetKwInOut{Input}{Input}
+ \Input{Log Joint Density, Approximation family q}
+ \KwResult{Variational Approximation}
+ Initialise $\lambda$\;
+ \While{Not converged}{
+  Simulate $\theta_s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
+  \For{$i =  1$ \KwTo $k$}{
+      Calculate $g_{\lambda_i} = 1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
+      Calculate $h_{\lambda_i} = 1/S\sum_{s=1}^{S}\nabla_{\lambda_i}[\log(q(\theta^s |\lambda^{(m-1)}))]$\;
+      Use a subset of $S$ to estimate $\hat{a} = \mbox{Cov}(h, g)/\mbox{Var}(g)$\;
+      Calculate $\nabla_{\lambda_i} = g_{\lambda_i} - \hat{a} h_{\lambda_i}$\;
+     }
+  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
+  Set $m = m + 1$\;
+ }
+ \caption{Gradient Ascent for SVB with control variates}
+  \label{alg:algorithm3}
+\end{algorithm}
+
+\subsubsection{Reparameterisation}
+
+The reparameterisation trick was popularised following \citet{Kingma2013} and allows the gradient of the ELBO to be simplified.
+
+Consider a parameter free auxillary distribution $q(\epsilon)$ and differentiable transformation $f(\cdot,\cdot)$ such that $\theta = f(\epsilon, \lambda)$.
+Examples include a location-scale transformation from a standard normal or an inverse-CDF transform from a uniform$(0, 1)$ variable.
+Note that
+
+$$q_\theta(\theta | \lambda) = q_\epsilon(\epsilon) \left| \frac{d\epsilon}{d\theta} \right| $$
+
+where the parameters that govern the transform $f$ are the same $\lambda$ parameters as in $q(\theta | \lambda)$. Now (\ref{SGA2}) becomes
+
+\begin{align}
+\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\epsilon} \nabla_{\lambda}[  q(\epsilon) \left( \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda) | \lambda) \right)] d\epsilon \nonumber \\
+&=  \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda)| \lambda) \right] d\epsilon \label{RP1} \\
+&= \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda)) \right] d\epsilon\nonumber
+\end{align}
+
+as the second part on the right hand side of (\ref{RP1}) is a score function with expectation equal to zero. This can be estimated using
+
+\begin{equation}
+\label{RP2}
+ \mathcal{L}(q(\theta | \lambda), y) \approx  \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon_s, \lambda))) \right].
+\end{equation}
+
+with $\epsilon_s \sim q(\epsilon)$.
+
+\citet{Kingma2015} explore how this reparameterised version can have orders of magnitude lower variance than the estimator in (\ref{SGA2}).
+
+\begin{algorithm}[H]
+ \SetKwInOut{Input}{Input}
+ \Input{Log Joint Density, Approximation family q}
+ \KwResult{Variational Approximation}
+ Initialise $\lambda$\;
+ \While{Not converged}{
+  Simulate $\epsilon^s$ for $s = 1, \dots S$\;
+  \For{$i =  1$ \KwTo $k$}{
+      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(p(f(\epsilon^s, \lambda^{(m-1)}), y))]$
+     }
+  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda}^{t}$\;
+  Set $m = m + 1$\;
+ }
+ \caption{Gradient Ascent for re-parameterised SVB}
+  \label{alg:algorithm4}
+\end{algorithm}
+
 \subsection{AR2 model example (Revisited using VB)}
+
+As the use of SVB does not provide an optimal approximating distribution like in MFVB, we instead analyse the MCMC results and fit a parametric distribution to the MCMC samples.
+\citet{Tran2015} extends SVB for use in approximations with a vine copula dependency structure, so we adopt their approach.
+
+As fitting a distribution to a sample from the posterior by MLE is equivalent to minimising the reverse of the KL dependence, we use the best fitting set of marginal distributions augmented by a vine copula by AIC (or BIC or something else?) and make the assumption that the family $q(\theta | \lambda)$ that minimises the reverse KL, and is optimal in EP, is the same family that is optimal for VB.
+
+\begin{list}
+\item State optimal families
+\item Fit via SVB
+\item Compare forecast accuracy to MCMC 
+\item Maybe use a lot of one step ahead forecasts with continuous VB updates compared to MCMC results with no parameter updates (but still bringing in new data so it's using one step ahead).
+\end{list}
 
@@
 \subsection{Motivation}
-\emph{Will you include a subsection showing an overview of the actual electricity load data you have??}
-Real data has five minute updates
-Density is of interest due to rare price spikes
-Slow convergence of MCMC / Bad scaling to large dimensional models and datasets
+\begin{list}
+\item Describe and plot Data
+\item Real data has five minute updates
+\item Density is of interest due to rare price spikes - show price data/supply curve etc to demonstrate
+\end{list}
+
 \subsection{Exponential Smoothing}
-Bayesian Version
-Difficulty with smoothing parameters in large models
-Infer model structure from draws where possible
+
+\citet{Taylor2003} explores the seasonal properties of electricity demand and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}), and later verifies that this approach is superior to other common time-series models for very short-term load forecasting \citep{Taylor2008}.
+
+\begin{align}
+y_t &= l_{t-1} + d_{t-m_1} + w_{t-m_2} + e_t \label{ds-hw1} \\
+l_t &= \alpha (y_t - d_{t-m_1} - w_{t-m_2}) + (1 - \alpha)l_{t-1} \label{ds-hw2}\\
+d_t &= \delta (y_t - l_{t-1} - w_{t-m_2}) + (1 - \delta)d_{t-m_1} \label{ds-hw3} \\
+w_t &= \omega (y_t - l_{t-1} - d_{t-m_1}) + (1 - \omega)w_{t-m_2} \label{ds-hw4}
+\end{align}
+
+with $\alpha, \delta, \omega \in (0, 1)$. This can be rewritten as a single source of error state-space model \citep{Snyder1985}
+
+\begin{align}
+y_t &= l_{t-1} + d_{t-m_1} + w_{t-m_2} + e_t \label{ds-hw-ssoe1} \\
+l_t &= l_{t-1} + \alpha e_t \label{ds-hw-ssoe2} \\
+d_t &= d_{t-m_1} + \delta e_t \label{ds-hw-ssoe3} \\
+w_t &= w_{t-m_2} + \omega e_t \label{ds-hw-ssoe4}. 
+\end{align}
+
+\citet{Forbes2000} describes a Bayesian method to sample the posterior distribution of the variance, smoothing parameters and initial states for $l_t, d_t$ and $w_t$ when $e_t \sim \mathcal{N}(0, \sigma^2)$.
+This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}) so we reparameterise using (\ref{ds-hw-rp1}) - (\ref{ds-hw-rp4}) to avoid this problem.
+
+\begin{align}
+y_t &= l_{t-1} - \sum_{i = 1}^{m_1 - 1}d_{t-i} - \sum_{i = 1}^{m_2 - 1}w_{t-i} + e_t \label{ds-hw-rp1} \\
+l_t &= l_{t-1} + \alpha e_t \label{ds-hw-rp2} \\
+d_t &= - \sum_{i = 1}^{m_1 - 1}d_{t-i} + \delta e_t \label{ds-hw-rp3} \\
+w_t &= - \sum_{i = 1}^{m_2 - 1}w_{t-i} + \omega e_t \label{ds-hw-rp4}.
+\end{align}
+
+The resulting unknown parameters are $\theta = (\alpha, \delta, \omega)', \sigma^2$ and $b_0$, the $m_1 + m_2 - 1$ length vector of the initial states of $l_0, d_0, \dots, d_{-(m_1 - 2)}, w_0, \dots, w_{-(m_2 - 2)}$.
+\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$,
+
+\begin{equation}
+\label{exp-sm-marginal}
+p(\theta | y_{1:T}) \propto \left| \widetilde{X}' \widetilde{X} \right|^{-1/2} \tilde{s}^{-(T-(m_1 + m_2 - 1))} p(\theta),
+\end{equation}
+
+where $\widetilde{X}$ and $\tilde{s}$ are functions of $\theta$ and $y$.
+The problems with using this method in our model are two-fold: repeating numerical integrion each time a new data point is observed for a three dimensional distribution is computationally difficult, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large.
+
+\begin{list}
+\item Derive some MCMC algorithm using $\theta$ conditional
+\item Infer model structure from draws if we can run this
+\item Or pick as flexible as possible distribution for marginals (normal mixture?) and figure out some vine structure to run VB without MCMC
+\end{list}
+
 \subsection{Variational Bayes Implementation}
-Posterior Distributions should not change much as more data is observed, so we can keep approximating distribution family constant
-But we might want a model that does change to make updates worthwhile - Markov Switching mechanism to be added?
+\begin{list}
+\item Posterior Distributions hopefully do not change much as more data is observed, so we can keep approximating distribution family constant
+\item But we might want a model that does change to make updates worthwhile
+\item Ideal model has constant distributional family with changing parameter values
+\item Maybe via Markov Switching mechanism to be added as probability of being in a certain state should change.
+\end{list}
 \subsection{Forecasting}

### 2017-01-05 Confirmation/confirmation.tex
Intro updated

@@
 For many time-series applications a point forecast of the mean of the next observation in a series, $y_{T+1}$, can be easily obtained with frequentist methods, conditioning on the forecaster's observed series, $y_{1:T}$.
-Often the forecast is supplemented with a prediction interval backed by asymptotic theory, however there is a growing demand in the literature for density forecasts (find citations, probably Hyndman) which are typically much harder to obtain.
+Often the forecast is supplemented with a prediction interval backed by asymptotic theory, however there is a growing demand in the literature for density forecasts which are typically much harder to obtain, see \citet{Gneiting2014} for a review.
 
-As a contrast to frequentism, the Bayesian methodology implicitly provides the entire probability density for variables of interest, in this case $p(y_{T+1} | y_{1:T})$.
+As a contrast to frequentism, the Bayesian methodology has been used extensively in forecasting \citep{Geweke2006} and is of interest as the entire probability density for $p(y_{T+1} | y_{1:T})$ is provided implicitly.
 However the Bayesian approach requires solving integrals of the form
@@
 
-\begin{list}
+\begin{itemize}
 \item State optimal families
@@
 \item Compare forecast accuracy to MCMC 
-\item Maybe use a lot of one step ahead forecasts with continuous VB updates compared to MCMC results with no parameter updates (but still bringing in new data so it's using one step ahead).
-\end{list}
+\item Maybe use a lot of one step ahead forecasts with continuous VB updates compared to MCMC results with no parameter updates 
+\end{itemize}
 
@@
 \subsection{Motivation}
-\begin{list}
+\begin{itemize}
 \item Describe and plot Data
@@
 \item Density is of interest due to rare price spikes - show price data/supply curve etc to demonstrate
-\end{list}
+\item Some general background on load forecasting somewhere here
+\end{itemize}
 
@@
 
-\citet{Taylor2003} explores the seasonal properties of electricity demand and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}), and later verifies that this approach is superior to other common time-series models for very short-term load forecasting \citep{Taylor2008}.
+\citet{Taylor2003} explores the seasonal properties of electricity demand and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}), and in \citet{Taylor2008} verifies that this approach is superior to other common time-series models for very short-term load forecasting \citep{Taylor2008}.
 
@@
 
-with $\alpha, \delta, \omega \in (0, 1)$. This can be rewritten as a single source of error state-space model \citep{Snyder1985}
+where $m_1$ and $m_2$ are the lengths of the daily and weekly cycle, restricting the smoothing parameters $\alpha, \delta, \omega$ to lie in $(0, 1)$.
+This can be rewritten as a single source of error state-space model \citep{Snyder1985},
 
@@
 
-\citet{Forbes2000} describes a Bayesian method to sample the posterior distribution of the variance, smoothing parameters and initial states for $l_t, d_t$ and $w_t$ when $e_t \sim \mathcal{N}(0, \sigma^2)$.
-This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}) so we reparameterise using (\ref{ds-hw-rp1}) - (\ref{ds-hw-rp4}) to avoid this problem.
+\citet{Forbes2000} describes a Bayesian method to sample the posterior distribution  when $e_t \sim \mathcal{N}(0, \sigma^2)$.
+This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}) so we reparameterise the seasonality using (\ref{ds-hw-rp1}) - (\ref{ds-hw-rp4}) to avoid this problem.
 
@@
 
-\begin{list}
+\begin{itemize}
 \item Derive some MCMC algorithm using $\theta$ conditional
@@
 \item Or pick as flexible as possible distribution for marginals (normal mixture?) and figure out some vine structure to run VB without MCMC
-\end{list}
+\end{itemize}
 
 \subsection{Variational Bayes Implementation}
-\begin{list}
+\begin{itemize}
 \item Posterior Distributions hopefully do not change much as more data is observed, so we can keep approximating distribution family constant
@@
 \item Maybe via Markov Switching mechanism to be added as probability of being in a certain state should change.
-\end{list}
+\end{itemize}
 \subsection{Forecasting}
-compare to MCMC when possible
+
+\begin{itemize}
+\item Compare to MCMC when possible
+\end{itemize}
 
 \section{Timeline}
+\begin{itemize}
+\item Put something here eventually
+\end{itemize}
 

### 2017-01-30 Confirmation/confirmation_V2.tex
confirmation 2nd draft

--- 2017-01-05 Confirmation/confirmation.tex
@@
 %% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+
+%2multibyte Version: 5.50.0.2960 CodePage: 1252
+\documentclass[12pt,a4paper]{article}%
+\usepackage{amsfonts}
+\usepackage{amssymb}
+\usepackage[centertags]{amsmath}
+\usepackage{graphicx}%
 \usepackage{natbib}
@@
 \usepackage[aboveskip=2pt]{subcaption}
-\usepackage{graphicx}
 \usepackage{amsmath}
-\usepackage{dsfont}
 \usepackage[]{algorithm2e}
 \usepackage{amsthm}
-\usepackage{amsfonts}
 \usepackage{url}
@@
 \usepackage{bbm}
-\numberwithin{equation}{section}
-\makeatletter
-\def\maxwidth{ %
-  \ifdim\Gin@nat@width>\linewidth
-    \linewidth
-  \else
-    \Gin@nat@width
-  \fi
-}
-\makeatother
-
-\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
-\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
-\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
-\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
-\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
-\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
-\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
-\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
-\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
-\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-
-\usepackage{framed}
-\makeatletter
-\newenvironment{kframe}{%
- \def\at@end@of@kframe{}%
- \ifinner\ifhmode%
-  \def\at@end@of@kframe{\end{minipage}}%
-  \begin{minipage}{\columnwidth}%
- \fi\fi%
- \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
- \colorbox{shadecolor}{##1}\hskip-\fboxsep
-     % There is no \\@totalrightmargin, so:
-     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
- \MakeFramed {\advance\hsize-\width
-   \@totalleftmargin\z@ \linewidth\hsize
-   \@setminipage}}%
- {\par\unskip\endMakeFramed%
- \at@end@of@kframe}
-\makeatother
-
-\definecolor{shadecolor}{rgb}{.97, .97, .97}
-\definecolor{messagecolor}{rgb}{0, 0, 0}
-\definecolor{warningcolor}{rgb}{1, 0, 1}
-\definecolor{errorcolor}{rgb}{1, 0, 0}
-\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
-
-\usepackage{alltt}
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\setcounter{MaxMatrixCols}{30}
+\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
+\newtheorem{theorem}{Theorem}
+\newtheorem{acknowledgement}[theorem]{Acknowledgement}
+\newtheorem{algorithm}[theorem]{Algorithm}
+\newtheorem{axiom}[theorem]{Axiom}
+\newtheorem{case}[theorem]{Case}
+\newtheorem{claim}[theorem]{Claim}
+\newtheorem{conclusion}[theorem]{Conclusion}
+\newtheorem{condition}[theorem]{Condition}
+\newtheorem{conjecture}[theorem]{Conjecture}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{criterion}[theorem]{Criterion}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+\newtheorem{exercise}[theorem]{Exercise}
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{notation}[theorem]{Notation}
+\newtheorem{problem}[theorem]{Problem}
+\newtheorem{proposition}[theorem]{Proposition}
+\newtheorem{remark}[theorem]{Remark}
+\newtheorem{solution}[theorem]{Solution}
+\newtheorem{summary}[theorem]{Summary}
+\newenvironment{proof}[1][Proof]{\textbf{#1.} }{\  \rule{0.5em}{0.5em}}
+\setlength{\topmargin}{0in}
+\setlength{\oddsidemargin}{0.1in}
+\setlength{\evensidemargin}{0.1in}
+\setlength{\textwidth}{6.5in}
+\setlength{\textheight}{8.25in}
+\title{Real Time Variational Density Forecasts}
+\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
 \begin{document}
 
+\maketitle
 \tableofcontents
@@
 
-As a contrast to frequentism, the Bayesian methodology has been used extensively in forecasting \citep{Geweke2006} and is of interest as the entire probability density for $p(y_{T+1} | y_{1:T})$ is provided implicitly.
-However the Bayesian approach requires solving integrals of the form
+As a contrast to frequentism, the Bayesian methodology has been used extensively in forecasting \citep{Geweke2006} and is it can capture the error components from parameter estimation and modelling error.
+However calculation of the Bayesian posterior distribution, $p(\theta | y)$ requires solving complex integrals that often do not have an analytical solution, in addition, if the the dimension of the parameter vector $\theta$ is large then numerical integration will be computationally infeasible.
 
-\begin{equation}
-\label{forecastintegral}
-\int_\theta p(y_{T+1} | \theta) p(\theta | y_{1:T}) d \theta,
-\end{equation}
+To address this problem a wide range of techniques are available, such as Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
+The forecasters decision of technique to use is a compromise to implicit trade-off: MCMC captures the posterior distribution accurately but is compuationaly intensive, while other methods such as VB tend to be faster but are inaccurate.
+One time series of interest, electricity demand data, operates on five minute interval in Victoria, too short for MCMC to be a feasible option in a moderately complex model.
+The focus of this research is in time series that operate in real-time, where the observation of each data point requires the model parameters to be updated before a density forecast of the next data point can be made.
+Variational Bayes is used as an alternative computation strategy, and the trade off between the statistical and computational efficiency will be explored using data on the electricity demand of Victoria for the 2012-2015 period.
 
-where $p(\theta | y_{1:T})$ is known as the posterior distribution, with
+\section{Bayesian Inference}
+
+Exact Bayesian inference in the context of forecasting uses the predictive density of $Y_{T+1}$ conditioned on the observed series, $Y_{t}, t = 1, \dots T$, which is denoted by $y_{1:T}$.
+The first step requires the posterior distribution of the model parameters $\theta$, p(\theta | y), where
 
@@
 \label{posterior}
- p(\theta | y_{1:T}) = \frac{p(y_{1:T}|\theta)p(\theta)}{\int_\theta p(y_{1:T}|\theta)p(\theta) d\theta}
+ p(\theta | y_{1:T}) = \frac{p(y_{1:T}|\theta)p(\theta)}{\int_\theta p(y_{1:T}|\theta)p(\theta) d\theta}.
 \end{equation}
 
-and $p(\theta)$ is some pre-specified prior distribution.
+Generally the solution to (\ref{posterior}) is unavailable. This section will offer two alternatives, MCMC and VB.
+MCMC is used to create a sample from $p(\theta | y)$, and any function of interest from $p(\theta | y$ can be found with the relevant function applied to the sample, while VB replaces $p(\theta | y)$ with an approximation $q(\theta | \lambda)$, choosing some family $q$ and approximation parameters $\lambda$.
 
-These often cannot be solved analytically, while numeric integration is computationally infeasible when the dimension of $\theta$ is large.
-To address this problem there is a wide range of techniques used to approximate the solution to the integral in (\ref{forecastintegral}), such as Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
-These approximations have an implicit trade-off: The better approximations are compuationaly intensive, and the forecaster must decide how much computation time and approximation error is acceptable.
-The focus of this research is in situtations where MCMC is feasible in a moderate amount of time, but is infeasible in time period between two observations.
-To move $p(y_{T+1} | y_{1:T})$ to $p(y_{T+2} | y_{1:T+1})$ requires updating the posterior distribution after observation of each new data point observed via
+\subsection{Markov Chain Monte Carlo}
 
-\begin{equation}
-\label{posteriorupdate}
-p(\theta | y_{1:T+1}) = \frac{p(y_{T+1} | \theta) p(\theta | y_{1:T})}{\int_\theta p(y_{T+1} | \theta) p(\theta | y_{1:T}) d\theta}.
-\end{equation}
-
-However, if the data is based on a market such as with stock returns the market will close overnight and MCMC can be ran.
-
-\section{Bayesian Inference}
-\subsection{Exact Bayesian Computation}
-While it is technically an approximation method, MCMC algorithms result in what is often called an exact computation of the posterior.
-A Gibbs based MCMC iteratively samples from the conditional distributions
+There are many types of MCMC algorithm, the most common being a Gibbs sampler, which iteratively samples from a $p$ dimensional $\theta$ via the conditional distributions
 
@@
 &\vdots \nonumber \\
-&p(\theta_p | \theta_1, \dots, \theta_{p-1}, y_{1:T}) \nonumber
+&p(\theta_p | \theta_1, \dots, \theta_{p-1}, y_{1:T}). \nonumber
 \end{align}
 
-where $p$ is the dimension of $\theta$.
-With enough iterations, the error in the approximation converges to zero and the algorithm can be ran for as much time as the forecaster desires to reduce error to a desired level.
-However, in the first iteration we must set arbitary starting values for each of $\theta_2, \dots, \theta_p$ introducing a large amount of error in the early iterations.
-To avoid this error MCMC generally must be run for a large number of iterations and these early samples are discarded.
-The computation time per iteration and speed of convergence is extremely problem specific.
+With enough iterations, these samples converge in distribution to $p(\theta | y)$ and can be used for inference.
+The computation time per iteration and number of iterations required for convergence is problem specific but typically increases with model complexity.
 
-We illustrate this with an AR(2), a simple time series model used in the remainder of this section. The AR(2) is described by
+To illustrate issues involved consider a simple auto-regressive time series model which will be used in the remainder of this section. This model is described by
 
@@
 \label{AR2}
-y_t - \mu = \rho_1 (y_{t-1} - \mu) + \rho_2 (y_{t-2} - \mu) + \epsilon_t
+y_t = \rho_1 y_{t-1} + \rho_2 y_{t-2} + \epsilon_t
 \end{equation}
 
-where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. We collect the unknown parameters as $\theta = (\mu, \sigma^2, \rho_1, \rho_2)'$ and set priors as
+where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. The likelihood of parameters $\theta = (\rho_1, \rho_2m \sigma^2)'$ is given by
+
+\begin{align}
+\label{likelihood}
+L(\theta | y_{1:T}) &= p(y_1 | \theta) p(y_2 | y_1, \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
+&= \frac{1}{(2\pi)^{T/2}\gamma(0)^{1/2}\Gamma{0}^{1/2}\sigma^{(T-2)/2}} \exp \left\{ \frac{-y_1^2}{2 \gamma(0)} \right\} 
+\exp \left\{\frac{-(y_2 - \gamma(1)/\gamma(2)y_1)^2}{2 \Gamma(0)} \right} \exp \left\{ \frac{-\sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2}{2\sigma^2} \right\}
+\end{align}
+
+where $\gamma(k)$ is the $k-th$ autocorrelation and $\Gamma(1) = \gamma(0) - \gamma(1)^2 / \gamma(0)$.
 

### 2017-01-30 Confirmation/confirmation_V2.tex
conf writing

@@
 
-For many time-series applications a point forecast of the mean of the next observation in a series, $y_{T+1}$, can be easily obtained with frequentist methods, conditioning on the forecaster's observed series, $y_{1:T}$.
-Often the forecast is supplemented with a prediction interval backed by asymptotic theory, however there is a growing demand in the literature for density forecasts which are typically much harder to obtain, see \citet{Gneiting2014} for a review.
+Electricity prices have been subject to an extensive forecasting literature, as detailed in \citet{Weron2014}, and can be extremely volatile.
+For the 2012-2015 period in Victoria, the mean price was X/KWH but on X occaisions the price spiked upwards several orders of magnitude to upwards of $1,000/KWH, reaching as high as $10,000/KWH on X date.
+For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period is more important than the mean value.
+To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it can average over parameter and model uncertainty that is often ignored in frequentist methods.
+See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general.
 
-As a contrast to frequentism, the Bayesian methodology has been used extensively in forecasting \citep{Geweke2006} and is it can capture the error components from parameter estimation and modelling error.
-However calculation of the Bayesian posterior distribution, $p(\theta | y)$ requires solving complex integrals that often do not have an analytical solution, in addition, if the the dimension of the parameter vector $\theta$ is large then numerical integration will be computationally infeasible.
+Bayesian methods are computationally complex as they require integration over the set of unknown parameters, which must be large in a model that can capture the strong seasonal patterns in short term electricity load \citep{Taylor2003}, which is the main driver for electricity price and the focus of this work.
+To avoid this problem there are a range of techniques available to obtain the predictive distribution which do not involve integration, notably Markov Chain Monte Carlo (MCMC) methods.
+While MCMC has very strong convergence properties, computation is slow for even moderately complex models, which is problematic as Electricity demand in Victoria operates on a five minute interval.
+The predictive density for the next period's electricity load may not be available before it is observed.
 
-To address this problem a wide range of techniques are available, such as Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
-The forecasters decision of technique to use is a compromise to implicit trade-off: MCMC captures the posterior distribution accurately but is compuationaly intensive, while other methods such as VB tend to be faster but are inaccurate.
-One time series of interest, electricity demand data, operates on five minute interval in Victoria, too short for MCMC to be a feasible option in a moderately complex model.
-The focus of this research is in time series that operate in real-time, where the observation of each data point requires the model parameters to be updated before a density forecast of the next data point can be made.
-Variational Bayes is used as an alternative computation strategy, and the trade off between the statistical and computational efficiency will be explored using data on the electricity demand of Victoria for the 2012-2015 period.
+Alternatives to MCMC include Variational Bayes (VB) and Expectation Propogation (EP), which aim to substitute the true posterior distribution for a vector of unknown parameters $\theta$, denoted by $p(\theta | y), with some approximate distribution denoted by $q(\theta | \lambda)$.
+Here $\lambda$ is a set of auxillary parameters associated with the approximation and not the model.
+For example, if $q(\theta | \lambda)$ is Gaussian $\lambda$ would be the mean and variance that parameterise $q$.
+Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2012-2015 period in Victoria.
+The advantages in computation time for VB will be compared to the loss of statisitcal accuracy associated with the use of an approximation of the posterior instead of the truth.
 
@@
 
-Exact Bayesian inference in the context of forecasting uses the predictive density of $Y_{T+1}$ conditioned on the observed series, $Y_{t}, t = 1, \dots T$, which is denoted by $y_{1:T}$.
-The first step requires the posterior distribution of the model parameters $\theta$, p(\theta | y), where
+Exact Bayesian inference in the context of forecasting uses the predictive density of $Y_{T+1}$ conditioned on the observed series, $y_{t}, t = 1, \dots T$, which is denoted by $y_{1:T}$.
+The first step requires the posterior distribution of the model parameters $\theta$, $p(\theta | y)$, where
 
@@
 Generally the solution to (\ref{posterior}) is unavailable. This section will offer two alternatives, MCMC and VB.
-MCMC is used to create a sample from $p(\theta | y)$, and any function of interest from $p(\theta | y$ can be found with the relevant function applied to the sample, while VB replaces $p(\theta | y)$ with an approximation $q(\theta | \lambda)$, choosing some family $q$ and approximation parameters $\lambda$.
+MCMC is used to create a sample from $p(\theta | y)$, and any function of interest from $p(\theta | y$ can be found with the relevant function applied to the sample, while VB replaces $p(\theta | y)$ with an approximation $q(\theta | \lambda)$, choosing some family $q$ and auxillary parameters $\lambda$ to minimise the Kullback-Leibler divergence from $q(\theta | \lambda)$ to $p(\theta | y)$.
 
@@
 
-There are many types of MCMC algorithm, the most common being a Gibbs sampler, which iteratively samples from a $p$ dimensional $\theta$ via the conditional distributions
+There are many types of MCMC algorithm, the most common being a Gibbs sampler which iteratively samples from a $k$ dimensional parameter vector $\theta$ via the conditional distributions
 
 \begin{align}
-&p(\theta_1 | \theta_2, \dots, \theta_p, y_{1:T}) \nonumber \\
-&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_p, y_{1:T}) \nonumber \\
+&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
+&p(\theta_2 | \theta_1, \theta_3, \dots, \thetakp, y_{1:T}) \nonumber \\
 &\vdots \nonumber \\
-&p(\theta_p | \theta_1, \dots, \theta_{p-1}, y_{1:T}). \nonumber
+&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
 \end{align}
@@
 \label{likelihood}
-L(\theta | y_{1:T}) &= p(y_1 | \theta) p(y_2 | y_1, \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
-&= \frac{1}{(2\pi)^{T/2}\gamma(0)^{1/2}\Gamma{0}^{1/2}\sigma^{(T-2)/2}} \exp \left\{ \frac{-y_1^2}{2 \gamma(0)} \right\} 
-\exp \left\{\frac{-(y_2 - \gamma(1)/\gamma(2)y_1)^2}{2 \Gamma(0)} \right} \exp \left\{ \frac{-\sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2}{2\sigma^2} \right\}
+L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
+&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} \right\} + \frac{1}{\sigma^2} 
+\sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
 \end{align}
 
-where $\gamma(k)$ is the $k-th$ autocorrelation and $\Gamma(1) = \gamma(0) - \gamma(1)^2 / \gamma(0)$.
+where
+
+\begin{equation}
+\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1), \gamma(0) \end{array} \right],\
+\end{equation}
+
+$\gamma(k)$ is the $k-th$ autocorrelation and $\textbf{y} = (y_1, y_2)'$.
+
+The priors
 
 \begin{align}
-p(\mu) &\propto 1 \nonumber \\
 p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
-\mbox{Some kind of }&\rho \mbox{ prior}
+p(\rho_1, \rho_2) &\propto \mathbb{I}(\rho_2 > -1)\mathbb{I}(\rho_2 < 1 + \rho_1) \mathbb{I}(\rho_2 < 1 - \rho_1) \nonumber
 \end{align}
 
-as this prior on $\rho_1$ and $\rho_2$ is uniform over the triangle that defines the AR(2) stationary region.
+are used, where $\mathbb{I}$ is the indicator and $p(\rho_1, \rho_2)$ defines the AR(2) stationary region.
 
@@
 
-Variational Bayes introduces some approximating distribution $q(\theta | \lambda)$ and aims to choose the family $q$ and set of parameters $\lambda$ so that $q(\theta | \lambda$ is as close as possible to the true posterior $p(\theta | y)$.
+Variational Bayes introduces some approximating distribution $q(\theta | \lambda)$ and aims to choose the family $q$ and set of auxillary parameters $\lambda$ so that $q(\theta | \lambda$ is as close as possible to the true posterior $p(\theta | y)$.
 It does this by minimising the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ to $p(\theta | y)$.
@@
 and is a non-negative, assymetric measure of the difference between $p(\theta | y)$ and $q(\theta | \lambda)$ that will equal zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}.
-It has origins in information theory, and can be interpreted as: \textit{`Given that I know $p(\theta | y)$ for some $\theta \in \Theta$, how much extra information is required, on average, to know the value of $q(\theta | \lambda)$?'} There are examples of the literature of approximations with other measures of divergence, such as \citet{Minka2001} introducting Expectation Propogation (EP), which minimises the reverse measure $KL[p(\theta | y)||q(\theta |\lambda)]$, which was extended to Power-EP in \citet{Minka2004}, which aims to minimise the more general $\alpha-\mbox{divergence}$ \citep{Amari1985}.
-It is shown in \citet{Bishop2006} (note: find original proof) that minimising $KL[p(\theta | y)||q(\theta | \lambda)]$ used in in EP is equivalent to the MLE of $q(\theta | \lambda)$ given a sample of $\theta$.
-However, we can write $KL[q(\theta | \lambda)||p(\theta | y)]$ as
+$KL[q(\theta | \lambda)||p(\theta | y)]$ can be re-written as
 
@@
 From (\ref{ELBO}) it is clear that maximising $\mathcal{L}(q, y)$ with respect to $q$ is equivalent to minimising (\ref{KL-def}).
-Maximising the ELBO is much more convenient than minimising either form of the KL Divergence, and has lead to Variational Bayes been used much more widely in the literature than alternatives such as EP.
+Maximising the ELBO is much more convenient than minimising either form of the KL Divergence, and has lead to VB being used much more widely in the literature than alternatives such as EP.
 
@@
 In the case that the likelihood and prior for $\theta_i$ form an exponential family conjugate pair, then the distributional family can be kept constant and all that is required is an update the parameters $\lambda_i$.
-If it does not match a known distribution, we may need to make a further approximation $\tilde{q_i(\theta_i|\lambda_i)}$ which has a recognizable distribution.
-One method as used in \citet{Friston2006} uses a Laplace approximation to and substitute in a Gaussian distribution for an otherwise unrecognizable $q_i(\theta_i | \lambda_i)$.
+
+For illustration, consider data $y_i, i = 1, \dots, n$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution where $p(\mu) \sim \mathcal{N}(\gamma, \lambda)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$.
+It can be shown that
+
+\begin{equation}
+\label{mf3}
+q(\mu) \propto \exp \left\{ \frac{-(n\lambda + \mathbb{E}(\sigma^{-2}))}{2\lambda\mathbb{E}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}(\sigma^{-2})\gamma + \lambda \sum_{i=1}^{n} y_i}{n \lambda + \mathbb{E}(\sigma^{-2})} \right) \right\}
+\end{equation}
+
+and
+
+\begin{equation}
+\label{mf4}
+q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1) \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + \mathbb{E}(\mu^2) - 2 y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
+
+It is evident that $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ and $q(\sigma^2 | \lambda) \sim Inv.Gamma( \tilde_{\alpha}, \tilde{\beta})$.
+The parameters $\lambda = (\tilde{\gamma}, \tilde{\lambda}, \tilde_{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations,
+
+\begin{align}
+
+\tilde{\gamma} &= \frac{\alpba/\beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta} \label{mf5} \\ 
+\tilde{\lambda} &= \frac{n \alpha /\beta}{n \lambda + \alpha / \beta} \\
+\tilde{\alpha} &= \frac{n}{2} + \alpha \\
+\tilde{\beta} &= \frac{1}{2} \left(\sum_{i=1}^{n} y_i^2 + n(\tilde{\gamma}^2 + \tilde{\lambda}) - 2 \sum_{i=1}^{n} y_i \tilde{\gamma} \right) + \beta. \label{mf6}
+\end{align}.
+
+As there is a circular dependence in these equations an algorithm that cycles through each auxillary parameter until convergence to some pre-specified threshold is required.
+
+If (\ref{mf2})  does not match a known distribution, we may need to make a further approximation substituting in another $\tilde{q_i(\theta_i|\lambda_i)}.
+One method as used in \citet{Friston2006} uses a Laplace approximation to and substitute in a Gaussian distribution for these unrecognizable $q_i(\theta_i | \lambda_i)$.
 
 Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | y, \theta_{\setminus i}$ in a Gibbs MCMC scheme, with dependence on other parameters replaced by their expectations.
-Hence, the optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family.
-The secondary Laplace approximation is analogous to a Metropolis-Hastings-within-Gibbs step in MCMC, as a way to handle unrecognisable distributions.
+The optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family.
+The secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable distributions.
 \vspace{5mm}
 
-Given these optimal distributional families, a mean field updating equation for each $\lambda_i$ can be found as a function of the data and other $\lambda_{j \neq i}$.
-This dependence requires an algorithm that continuously iterates between each $\lambda_i$ and sets it to its maximising value until $\mathcal{L}(q(\theta | \lambda), y)$ converges within some pre-defined threshold.
-This is known as a coordinate ascent algorithm and follows below, where $k$ as the dimension of $\lambda$.
+The required algorithm is known as a coordinate ascent algorithm, and is described below where $\lambda$ is $k$ dimensional.
 
@@
  \KwResult{Mean Field Approximation}
- Initialise $\lambda$ randomly\;
- Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution\;
+ Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution, or substitute a further approximation if neccesary.\;
+ Derive the set of mean field equations such as in (\ref{mf5}) - (\ref{mf6}).\;
+ Initialise $\lambda$ randomly. \;
  \While{Not converged}{
@@
       Hold $\lambda_{j \neq i}$ fixed\;
-      Update $\lambda_i$ using $y$ and the most recent values of $\lambda_{j \neq i}$\;
+      Update $\lambda_i$ using the relevant mean field equation.
      }

### 2017-02-01 Confirmation/confirmation_V2.tex
Second Draft

@@
-\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[12pt,a4paper]{article}%
+\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 %2multibyte Version: 5.50.0.2960 CodePage: 1252
-\documentclass[12pt,a4paper]{article}%
 \usepackage{amsfonts}
@@
 \newtheorem{acknowledgement}[theorem]{Acknowledgement}
-\newtheorem{algorithm}[theorem]{Algorithm}
 \newtheorem{axiom}[theorem]{Axiom}
@@
 \newtheorem{summary}[theorem]{Summary}
-\newenvironment{proof}[1][Proof]{\textbf{#1.} }{\  \rule{0.5em}{0.5em}}
 \setlength{\topmargin}{0in}
@@
 \setlength{\textheight}{8.25in}
+\numberwithin{equation}{section}
 \title{Real Time Variational Density Forecasts}
@@
 
-Alternatives to MCMC include Variational Bayes (VB) and Expectation Propogation (EP), which aim to substitute the true posterior distribution for a vector of unknown parameters $\theta$, denoted by $p(\theta | y), with some approximate distribution denoted by $q(\theta | \lambda)$.
+Alternatives to MCMC include Variational Bayes (VB) and Expectation Propogation (EP), which aim to substitute the true posterior distribution for a vector of unknown parameters $\theta$, denoted by $p(\theta | y)$, with some approximate distribution denoted by $q(\theta | \lambda)$.
 Here $\lambda$ is a set of auxillary parameters associated with the approximation and not the model.
@@
 &p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
-&p(\theta_2 | \theta_1, \theta_3, \dots, \thetakp, y_{1:T}) \nonumber \\
+&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
 &\vdots \nonumber \\
@@
 L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
-&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} \right\} + \frac{1}{\sigma^2} 
-\sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
+&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
 \end{align}
@@
 
-Describe conditional distributions, possible MH step, result in forecast distribution for $y_{T+1}$
+The conditional distribution $p(\sigma^2 | y, \phi_1, \phi_2)$ is Inverse Gamma, while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step with truncated normal random walk proposals.
+The truncation ensures that draws are only taken from the AR(2) stationary region.
+
+Note: Will make the picture later
+\begin{figure}
+\label{MCMC}
+%\includegraphics{MCMC.png}
+\caption{MCMC draws for the AR(2) model with $T = 150, \phi_1 = 0.75, \phi_2 = 0.2$ and $\sigma^2 = 1$.
+Approximately 24\% of both $\phi$ parameters were accepted by the Metropolis Hastings step, then the first 50,000 draws were discarded and the remainder thinned to 1 in 300 to make dependence between draws negligible.}
+\end{figure}
 
@@
 
-Variational Bayes introduces some approximating distribution $q(\theta | \lambda)$ and aims to choose the family $q$ and set of auxillary parameters $\lambda$ so that $q(\theta | \lambda$ is as close as possible to the true posterior $p(\theta | y)$.
-It does this by minimising the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ to $p(\theta | y)$.
+A faster alternative to MCMC is Variational Bayes, which introduces an approximating distribution $q(\theta | \lambda)$ choosing the functional form $q$ and given auxillary parameters $\lambda$ so that  misimises the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ the true posterior $p(\theta | y)$.
 The KL divergence is defined by
@@
 
-and is a non-negative, assymetric measure of the difference between $p(\theta | y)$ and $q(\theta | \lambda)$ that will equal zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}.
-$KL[q(\theta | \lambda)||p(\theta | y)]$ can be re-written as
+The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\theta | y)$ and $q(\theta | \lambda)$ that will equal zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}.
+Note that $KL[q(\theta | \lambda)||p(\theta | y)]$ can be expressed as
 
@@
 \label{KL-ELBO}
-KL[q(\theta | \lambda)||p(\theta | y)] = \ln(y) - \mathcal{L}(q, y)
+KL[q(\theta | \lambda)||p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q, y)
 \end{equation}
 
-where $\mathcal{L}(q, y)$ is known as the Evidence Lower BOund (ELBO), defined by
+where $\mathcal{L}(q, y)$ is known as the Evidence Lower Bound (ELBO), defined by
 
@@
 
-From (\ref{ELBO}) it is clear that maximising $\mathcal{L}(q, y)$ with respect to $q$ is equivalent to minimising (\ref{KL-def}).
-Maximising the ELBO is much more convenient than minimising either form of the KL Divergence, and has lead to VB being used much more widely in the literature than alternatives such as EP.
+Since $\ln(p(y))$ is constant given $y$, maximising (\ref{KL-ELBO}) with respect to $q$ is equivalent to minimising (\ref{KL-def}).
+Maximising the ELBO is much more convenient than minimising the KL Divergence, which can be seen in the following sections based on the two major implementations of Varitaional Bayes: Mean Field Variational Bayes and Stochastic Variational Bayes.
+Each of these take advantage of the functional form of the ELBO and offer computationally efficient algorithms to find a $q(\theta | \lambda)$ that is optimal within a particular clas of distributions.
 
@@
 
-Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the search for an optimal approximating dististribution to the set of factorisable distributions
+Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the search for an optimal approximating dististribution to the class of factorisable distributions
 
@@
 \label{mf1}
-q(\theta|\lambda) = \prod_i q_i(\theta_i | \lambda_i).
+q(\theta|\lambda) = \prod_i^k q_i(\theta_i | \lambda_i).
 \end{equation}
 
-This technique is widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}), particularly in exponential family models \citep{Wainwright2008}.
-Each component $\theta_i$ may be a scalar or a vector, and $\lambda_i$ is the component of the $\lambda$ vector that parameterises the relevant factor $q_i(\theta_i |\lambda_i)$.
-From here, we will use the shorthand notation that $q_i = q_i(\theta_i|\lambda_i)$ and $q_{\setminus i} = \prod_{j\neq i}q_j$.
-Maximising the ELBO with respect to $q_i$ is analytically involved but computationally simple, as (\ref{ELBO}) can be expressed as a function of $q_i$ only and then maximised with respect to each $q_i$ individually.
-\citet{Attias1999} shows that
+Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars.
+Each $\theta_i$ has an associated vector $\lambda_i$ which may be of a higher dimension than $\theta_i$.
+$\lambda_i$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\lambda_i)$.
+
+MFVB is widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}), particularly in exponential family models \citep{Wainwright2008}.
+Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the computation is simple.
+MFVB expresses (\ref{ELBO}) as a function of one $q_i$ only and then maximises with respect to that $q_i$.
+The final algorithm requires this to be repeated for each $i$ in $1, \dots, k$.
+Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that
 
@@
 \end{equation}
-
-and maximisation can proceed by matching (\ref{mf2}) to a known distribution.
-In the case that the likelihood and prior for $\theta_i$ form an exponential family conjugate pair, then the distributional family can be kept constant and all that is required is an update the parameters $\lambda_i$.
 
@@
 \label{mf4}
-q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1) \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + \mathbb{E}(\mu^2) - 2 y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
+q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + \mathbb{E}(\mu^2) - 2 y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
+\end{equation}
 
-It is evident that $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ and $q(\sigma^2 | \lambda) \sim Inv.Gamma( \tilde_{\alpha}, \tilde{\beta})$.
-The parameters $\lambda = (\tilde{\gamma}, \tilde{\lambda}, \tilde_{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations,
+It is evident that $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ and $q(\sigma^2 | \lambda) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$.
+The parameters $\lambda = (\tilde{\gamma}, \tilde{\lambda}, \tilde{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations:
 
 \begin{align}
-
-\tilde{\gamma} &= \frac{\alpba/\beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta} \label{mf5} \\ 
-\tilde{\lambda} &= \frac{n \alpha /\beta}{n \lambda + \alpha / \beta} \\
+\tilde{\gamma} &= \frac{\alpha / \beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta} \label{mf5} \\ 
+\tilde{\lambda} &= \frac{n \alpha / \beta}{n \lambda + \alpha / \beta} \\
 \tilde{\alpha} &= \frac{n}{2} + \alpha \\
@@
 
-If (\ref{mf2})  does not match a known distribution, we may need to make a further approximation substituting in another $\tilde{q_i(\theta_i|\lambda_i)}.
-One method as used in \citet{Friston2006} uses a Laplace approximation to and substitute in a Gaussian distribution for these unrecognizable $q_i(\theta_i | \lambda_i)$.
+In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One method in \citet{Friston2006} uses a Laplace approximation to substitute a Gaussian distribution for these unrecognizable $q_i(\theta_i | \lambda_i)$, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used in (\ref{mf2}).
 
@@
 
-The required algorithm is known as a coordinate ascent algorithm, and is described below where $\lambda$ is $k$ dimensional.
+The required algorithm is known as a coordinate ascent algorithm, and is described below where $\lambda$ is a $p$ dimensional vector.
 
@@
 
-In many cases, the posterior is too complex to be captured for a factorising approximation to be reasonabl as we desire our distribution $q(\theta | \lambda)$ to capture dependence between parameters.
-In this case, the easy maximisation in MFVB is unavailable, and we must resort to what is called Stochastic Variational Bayes (SVB) using a gradient ascent algorithm developed by \citet{Paisley2012} and \citet{Ranganath2014}.
+Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure (\ref{MCMC}) shows strong dependence between $\phi_1$ and $\phi_2$.
+Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable.
+\citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, resulting in what is known as Stochastic Variational Bayes (SVB).
 
-To maximise the function $\mathcal{L}(q(\theta | \lambda), y)$ we can take the derivative of $\mathcal{L}(q(\theta | \lambda), y)$ with respect to $\lambda$ and use the updating step:
+The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\theta | \lambda), y)$ in (\ref{ELBO}) with respect to $\lambda$ and the following updating step is applied until the ELBO converges to a maximum.
 
@@
 where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ with respect to each element of $\lambda$.
-This update requires initial values $\lambda^{(0)}$ and some learning rate sequence $\rho^{(m)}$.
-If $\rho^{(m)}$ is chosen to satisfy the following conditions, it is a Robbins-Monro sequence and the algorithm is guaranteed to converge to a local maximum.
+This update requires some initial values $\lambda^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
+If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.
 
@@
 
-Whilst a global maximum is desired, the ELBO can have a problem specific shape that makes finding the global maximum extremely difficult, as we do not know how many stationary points exist.
-The dimension of the ELBO is the dimension of the $\lambda$ vector, which is often much greater than the dimension of the parameters $\theta$ so a grid search is suspect to the curse of dimensionality.
-To alleiviate this problem, we can start the algorithm at a range of initial values choose the converged value with the highest value of the ELBO.
+Whilst a global maximum is desired, the ELBO is a high dimensional problem specific function that makes finding the global maximum extremely difficult and often only a local maximum can be found.
+One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior
 
-SVB does not provide the family of the the optimal approximating distribution, unlike under the mean field assumption.
-To run SVB we may need to try many approximating distributions and then choose the $q(\theta | \lambda)$ that has the highest ELBO, and hence lowest KL divergence to the true posterior.
-We must restrict the approximation to distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable.
-In this case \citet{Ranganath2014} shows that a Monte Carlo estimate of the derivative of the ELBO can be taken by
+SVB can find only the optimal values for $\lambda$ that are optimal for a distributional family $q$ that is specified by the user.
+To run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO.
+$q(\theta | \lambda)$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable, and SVB can only be applied to models where the log likelihood for $y$ is able to be evaluated.
+Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by
 
@@
 
-As the distribution $q(\theta | \lambda)$ is specified by the user, the main restriction on the use of SVB is that the log-joint density $\ln(p(y, \theta))$ is able to be evaluated.
-
-\citet{Duchi2011} introduces the AdaGrad algorithm which can be implemented within SVB to control $\rho^{(m)}$.
-AdaGrad allows each $\lambda_i$ to have an independent $\rho^{(m)}_i$ that is inversely proportional to the gradient.
+\citet{Duchi2011} introduced the AdaGrad algorithm which can be implemented within SVB to control $\rho^{(m)}$.
+AdaGrad allows each $\lambda_i$ to have an independent $\rho^{(m)}_i$ that is inversely proportional to the gradient, so $\lambda$ takes bigger steps in flat regions and smaller steps in steep regions.
 
@@
 
-The resulting Stochastic Gradient Ascent algorithm proceeds below.
+The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\lambda$ vector.
 
@@
   Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m)})$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m)})))$\;
-      Update $G_i^{(m)} = G_i^{(m-1)} + \left(\nabla_{\lambda_i}\mathcal{L}(q(\theta | \lambda^{(m)})\right)^2$\;
-      Calculate $\rho^{(m)}_i =  \left(G_i^{(m)}\right)^{-1/2}$\;
+  \For{$i =  1$ \KwTo $p$}{
+      Calculate $\nabla_{\lambda_i}$ from (\ref{SGA2})\;
+      Update $G_i^{(m)}$ and $\rho^{(m)}_i$ from (\ref{SGA3}) and (\ref{SGA4})\;
       }
-  Set $\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)}  \nabla_{\lambda}$\;
+  Update $\lambda^{(m+1)}$ from (\ref{SGA1})\;
   Set $m = m + 1$\;
@@
 
-\subsection{Variance Reduction Techniques}
+\subsection{Vine Copulas}
 
-For many models the estimator in (\ref{SGA2}) has a large variance, so a more efficient estimator will allow a lower value of $S$ and hence reduced computation time.
-\subsubsection{Control Variates}
-\citet{Paisley2012} considers the use of control variates to augment the estimator of the gradient. Define
+Armed with an algorithm to find the Variational Bayes optimal paramerters $\lambda$ for a distribution $q$, there is a requirement to have a distribution $q$ that is itself optimal.
+Copulas are a flexible tool for constructing these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions.
+\citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,
 
 \begin{equation}
-\label{CV1}
-g(\theta, \lambda, y) =  \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda))),
+\label{vc1}
+p(\theta_1, \dots, \theta_k) = p(\theta_1) \dots p(\theta_k) c(P(\theta_1), \dots, P(\theta_k))
 \end{equation}
 
-the function estimated in (\ref{SGA2}), then for some other function $h$, define
+where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.
+
+\citet{Joe1994} introduces the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas in a similar way to the factorisation of a joint distribution into a marginal and conditional distribution, for example
 
 \begin{equation}
-\label{CV2}
-\hat{g}(\theta, \lambda, y) = g(\theta, \lambda, y) - a(h - \mathbb{E}(h)).
+\label{vc2}
+c(P(\theta_1), P(\theta_2), P(\theta_3)) = c(P(\theta_1), P(\theta_2)) \cdot c(P(\theta_1), P(\theta_3)) \cdot c(P(\theta_2), P(\theta_3) | P(\theta_1)).
 \end{equation}
 
-Then both $g(\theta, \lambda, y)$ and $\hat{g}(\theta, \lambda, y)$ have the same expectation and
+A copula for a $k$ dimensional $\theta$ can be replaced with $k(k-1)/2$ bivariate copulas arranged in a vine copula, greatly increasing the amount of flexibility available to fit a distribution.
+However there are $O(k!)$ ways to factorise the full copula and choose a family for each bivariate copula, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
+\citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\theta$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.
 
-\begin{equation}
-\label{CV3}
-\mbox{Var}(\hat{g}) = \mbox{Var}(g) + a^2 \mbox{Var}(h) - 2a\mbox{Cov}(g, h). 
-\end{equation}
+Given a sample of $\theta$, \citet{Diimann2013} provide an algorithm to select the best vine factorisation by maximising Kendell's Tau, and the best family for each bivariate copula by maximising an information criterion such as AIC.
+In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date.
+The approach used in this thesis is to run an MCMC algorithm, thin the resulting draws until they are effectively independent, choose an optimal family for each marginal distribution by AIC, then run (Vine copula algorithm here) to infer an optimal vine copula.
+This results in an approximating distribution $q(\theta | \lambda)$ that can then be used in an SVB algorithm to update $q(\theta | \lambda)$ as more observations become available.
+MCMC can be ran simultaenously and the resulting sample can be used to check if the distribution $q$ should be changed.
 
-Solving the polynomial (\ref{CV3}) in $a$ shows that the variance of $\hat{g}$ is minimised by
+\subsection{Applying Stochastic Variational Bayes}
 
-\begin{equation}
-\label{CV4}
-\hat{a} = \mbox{Cov}(g, h)/\mbox{Var}(h).
-\end{equation}
-
-Substituting (\ref{CV4}) into (\ref{CV3}) yields
-
-\begin{equation}
-\label{CV5}
-\mbox{Var}(\hat{g}) = \mbox{Var}(g) - \mbox{Cov}(g, h)^2/\mbox{Var}(h).
-\end{equation}
-
-We need to choose some function $h$ that has a large covariance with $g$. \citet{Ranganath2014} suggests that the easy option of
-
-\begin{equation}
-\label{CV6}
-h(\theta, \lambda) = \nabla_{\lambda} [\ln(q(\theta_s | \lambda))].
-\end{equation}
-
-As $h$ is a score function, it has an expectation of zero and our estimator becomes
-
-\begin{equation}
-\label{CV7}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda)) - \hat{a}),
-\end{equation}
-
-where $\hat{a}$ is estimated from the sample variance and covariance of a subset of the $S$ Monte-Carlo draws.
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\theta_s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m-1)})$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $g_{\lambda_i} = 1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(q(\theta^s|\lambda^{(m-1)})](\log(p(\theta^s, y)) - \log(q(\theta^s |\lambda^{(m-1)})))$\;
-      Calculate $h_{\lambda_i} = 1/S\sum_{s=1}^{S}\nabla_{\lambda_i}[\log(q(\theta^s |\lambda^{(m-1)}))]$\;
-      Use a subset of $S$ to estimate $\hat{a} = \mbox{Cov}(h, g)/\mbox{Var}(g)$\;
-      Calculate $\nabla_{\lambda_i} = g_{\lambda_i} - \hat{a} h_{\lambda_i}$\;
-     }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)}  \nabla_{\lambda}$\;
-  Set $m = m + 1$\;
- }
- \caption{Gradient Ascent for SVB with control variates}
-  \label{alg:algorithm3}
-\end{algorithm}
-
-\subsubsection{Reparameterisation}
-
-The reparameterisation trick was popularised following \citet{Kingma2013} and allows the gradient of the ELBO to be simplified.
-
-Consider a parameter free auxillary distribution $q(\epsilon)$ and differentiable transformation $f(\cdot,\cdot)$ such that $\theta = f(\epsilon, \lambda)$.
-Examples include a location-scale transformation from a standard normal or an inverse-CDF transform from a uniform$(0, 1)$ variable.
-Note that
-
-$$q_\theta(\theta | \lambda) = q_\epsilon(\epsilon) \left| \frac{d\epsilon}{d\theta} \right| $$
-
-where the parameters that govern the transform $f$ are the same $\lambda$ parameters as in $q(\theta | \lambda)$. Now (\ref{SGA2}) becomes
-
-\begin{align}
-\nabla_{\lambda} \mathcal{L}(q(\theta | \lambda), y) &=  \int_{\epsilon} \nabla_{\lambda}[  q(\epsilon) \left( \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda) | \lambda) \right)] d\epsilon \nonumber \\
-&=  \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda))) - \ln(q(f(\epsilon, \lambda)| \lambda) \right] d\epsilon \label{RP1} \\
-&= \int_{\epsilon}  q(\epsilon) \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon, \lambda)) \right] d\epsilon\nonumber
-\end{align}
-
-as the second part on the right hand side of (\ref{RP1}) is a score function with expectation equal to zero. This can be estimated using
-
-\begin{equation}
-\label{RP2}
- \mathcal{L}(q(\theta | \lambda), y) \approx  \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} \left[ \ln (p(y, f(\epsilon_s, \lambda))) \right].
-\end{equation}
-
-with $\epsilon_s \sim q(\epsilon)$.
-
-\citet{Kingma2015} explore how this reparameterised version can have orders of magnitude lower variance than the estimator in (\ref{SGA2}).
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\epsilon^s$ for $s = 1, \dots S$\;
-  \For{$i =  1$ \KwTo $k$}{
-      Calculate $\nabla_{\lambda_i} =  1/S\sum_{s=1}^{S} \nabla_{\lambda_i}[\log(p(f(\epsilon^s, \lambda^{(m-1)}), y))]$
-     }
-  Set $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m)} \nabla_{\lambda}^{t}$\;
-  Set $m = m + 1$\;
- }
- \caption{Gradient Ascent for re-parameterised SVB}
-  \label{alg:algorithm4}
-\end{algorithm}
-
-\subsection{AR2 model example (Revisited using VB)}
-
-As the use of SVB does not provide an optimal approximating distribution like in MFVB, we instead analyse the MCMC results and fit a parametric distribution to the MCMC samples.
-\citet{Tran2015} extends SVB for use in approximations with a vine copula dependency structure, so we adopt their approach.
-
-As fitting a distribution to a sample from the posterior by MLE is equivalent to minimising the reverse of the KL dependence, we use the best fitting set of marginal distributions augmented by a vine copula by AIC (or BIC or something else?) and make the assumption that the family $q(\theta | \lambda)$ that minimises the reverse KL, and is optimal in EP, is the same family that is optimal for VB.
+Using the approach described in section 2.3, an application of VC algorirhtm to the MCMC draws obtained in section 2.1 found the following distribution to be optimal by AIC.
 
 \begin{itemize}
-\item State optimal families
-\item Fit via SVB
-\item Compare forecast accuracy to MCMC 
-\item Maybe use a lot of one step ahead forecasts with continuous VB updates compared to MCMC results with no parameter updates 
+\item $q(\phi_1)$ - Gaussian marginal
+\item $q(\phi_2)$ - Gaussian marginal
+\item $q(\sigma^2)$ - Inverse Gamma marginal
+\item $c(Q(\phi_1), Q(\phi_2))$ - Gaussian copula
+\item $c(Q(\phi_1), Q(\sigma^2))$ - Independent
+\item $c(Q(\phi_2), Q(\sigma^2) | Q(\phi_1))$ - Independent
 \end{itemize}
+
+Figure (\ref{VBfit}) shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.
+
+\begin{figure}[h]
+\label{VBfit}
+\centering
+\includegraphics[scale = 0.5]{VBfit.png}
+\caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model described in section 2.1.
+SVB resulted in an almost identical posterior to MCMC while taking X less time to run.
+(working from home, will check runtime at uni)}
+\end{figure}
 
@@
 \subsection{Motivation}
+
+Note: Don't have data on hand, will do this section shortly
+
 \begin{itemize}
-\item Describe and plot Data
-\item Real data has five minute updates
-\item Density is of interest due to rare price spikes - show price data/supply curve etc to demonstrate
-\item Some general background on load forecasting somewhere here
+\item Describe and plot load Data
+\item Plot price data to emphasize price spikes
+\item Follow up the intro to talk a bit about load forecasting
 \end{itemize}
@@
 
-\citet{Taylor2003} explores the seasonal properties of electricity demand and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}), and in \citet{Taylor2008} verifies that this approach is superior to other common time-series models for very short-term load forecasting \citep{Taylor2008}.
+\citet{Taylor2003} explores the seasonal properties of electricity demand and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}), and later in \citet{Taylor2008} verifies that this approach is superior to other common time-series models for very short-term load forecasting \citep{Taylor2008}.
 
@@
 
-\citet{Forbes2000} describes a Bayesian method to sample the posterior distribution  when $e_t \sim \mathcal{N}(0, \sigma^2)$.
-This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}) so we reparameterise the seasonality using (\ref{ds-hw-rp1}) - (\ref{ds-hw-rp4}) to avoid this problem.
+The unknown parameters are $\theta = (\alpha, \delta, \omega)', \sigma^2$ and $b_0 = (l_0, d_0, \dots, d_{-(m_1 - 1)}, w_0, \dots, w_{-(m_2 - 1)})'$ a $m_1 + m_2 + 1$ length vector of the initial states of the latent variables.
+\citet{Forbes2000} derives the marginal distribution $p(\theta | y)$ and notes that, conditioned on a draw of these parameters, the remaining unknown parameters can be sampled as a Bayesian linear regression problem.
+This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}), but restricting the latent parameters via
+
+\begin{align}
+d_t &= - \sum_{i=1}^{m_1-1} d_{t-i} \\
+w_t &= - \sum_{i=1}^{m_2-1} w_{t-i}
+\end{align}
+
+avoids this problem. The resulting model is described by
 
@@
 
-The resulting unknown parameters are $\theta = (\alpha, \delta, \omega)', \sigma^2$ and $b_0$, the $m_1 + m_2 - 1$ length vector of the initial states of $l_0, d_0, \dots, d_{-(m_1 - 2)}, w_0, \dots, w_{-(m_2 - 2)}$.
 \citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$,
@@
 where $\widetilde{X}$ and $\tilde{s}$ are functions of $\theta$ and $y$.
-The problems with using this method in our model are two-fold: repeating numerical integrion each time a new data point is observed for a three dimensional distribution is computationally difficult, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large.
+The problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible difficult, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large.
 
 \begin{itemize}
-\item Derive some MCMC algorithm using $\theta$ conditional
-\item Infer model structure from draws if we can run this
-\item Or pick as flexible as possible distribution for marginals (normal mixture?) and figure out some vine structure to run VB without MCMC
-\end{itemize}
-
-\subsection{Variational Bayes Implementation}
-\begin{itemize}
-\item Posterior Distributions hopefully do not change much as more data is observed, so we can keep approximating distribution family constant
-\item But we might want a model that does change to make updates worthwhile
-\item Ideal model has constant distributional family with changing parameter values
-\item Maybe via Markov Switching mechanism to be added as probability of being in a certain state should change.
-\end{itemize}
-\subsection{Forecasting}
-
-\begin{itemize}
-\item Compare to MCMC when possible
+\item Should try and get an exponential smoothing MCMC working using Metropolis-Hastings for $\theta$.
+Log densities help with the $T$ being too large problem but the grid is too large for numerical integration to be a good idea.
 \end{itemize}

### 2017-02-03 Confirmation/confirmation_V2.tex
electricity section started

@@
 
-Electricity prices have been subject to an extensive forecasting literature, as detailed in \citet{Weron2014}, and can be extremely volatile.
-For the 2012-2015 period in Victoria, the mean price was X/KWH but on X occaisions the price spiked upwards several orders of magnitude to upwards of $1,000/KWH, reaching as high as $10,000/KWH on X date.
+Electricity prices have been subject to an extensive forecasting literature and is known to be extremely volatile.
+For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was $42.06, with 99\% of oberservations less than \$100/mWh, but on 39 occaisions the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29 2012.
 For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period is more important than the mean value.
@@
 For example, if $q(\theta | \lambda)$ is Gaussian $\lambda$ would be the mean and variance that parameterise $q$.
-Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2012-2015 period in Victoria.
+Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2011-2015 period in Victoria.
 The advantages in computation time for VB will be compared to the loss of statisitcal accuracy associated with the use of an approximation of the posterior instead of the truth.
@@
 
-Note: Will make the picture later
-\begin{figure}
-\label{MCMC}
-%\includegraphics{MCMC.png}
+\begin{figure}[h]
+\label{MCMCplot}
+\centering
+\includegraphics[scale = 0.5]{MCMC.png}
 \caption{MCMC draws for the AR(2) model with $T = 150, \phi_1 = 0.75, \phi_2 = 0.2$ and $\sigma^2 = 1$.
-Approximately 24\% of both $\phi$ parameters were accepted by the Metropolis Hastings step, then the first 50,000 draws were discarded and the remainder thinned to 1 in 300 to make dependence between draws negligible.}
+Approximately 24\% of both $\phi$ parameters were accepted by the Metropolis Hastings step, then the first 10,000 draws were discarded and the remainder thinned to 1 in 200 to make dependence between draws negligible.}
 \end{figure}
@@
 
-Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure (\ref{MCMC}) shows strong dependence between $\phi_1$ and $\phi_2$.
+Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure (\ref{MCMCplot}) shows strong dependence between $\phi_1$ and $\phi_2$.
 Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable.
@@
 
-Note: Don't have data on hand, will do this section shortly
+The literature for electricity forecasting is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review.
+Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016} being notable examples, which used boot strapping and quantile regression respectively to generate forecast densities.
+Both of these approaches create forecast densities that are conditioned on the model and estimated parameters being correct, while the Bayesian approach used in this thesis can average over parameter and model uncertainty leading to a more accurate forecast density.
+This thesis also aims to re-estimate the model after each data point is observed to fully take advantage of any extra information on changes in parameters and thus further increase the accuracy of the forecast density.
 

### 2017-02-03 Confirmation/confirmation_V2.tex
electricity section finished

@@
 Electricity prices have been subject to an extensive forecasting literature and is known to be extremely volatile.
-For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was $42.06, with 99\% of oberservations less than \$100/mWh, but on 39 occaisions the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29 2012.
+For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of oberservations less than \$100/mWh, but on 39 occaisions the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29, 2012.
 For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period is more important than the mean value.
@@
 
-Alternatives to MCMC include Variational Bayes (VB) and Expectation Propogation (EP), which aim to substitute the true posterior distribution for a vector of unknown parameters $\theta$, denoted by $p(\theta | y)$, with some approximate distribution denoted by $q(\theta | \lambda)$.
-Here $\lambda$ is a set of auxillary parameters associated with the approximation and not the model.
-For example, if $q(\theta | \lambda)$ is Gaussian $\lambda$ would be the mean and variance that parameterise $q$.
+Alternatives to MCMC include Variational Bayes (VB), which aim to replace the unknown true posterior distribution for a vector of unknown parameters $\theta$ with a parametric approximation.
 Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2011-2015 period in Victoria.
@@
 Generally the solution to (\ref{posterior}) is unavailable. This section will offer two alternatives, MCMC and VB.
-MCMC is used to create a sample from $p(\theta | y)$, and any function of interest from $p(\theta | y$ can be found with the relevant function applied to the sample, while VB replaces $p(\theta | y)$ with an approximation $q(\theta | \lambda)$, choosing some family $q$ and auxillary parameters $\lambda$ to minimise the Kullback-Leibler divergence from $q(\theta | \lambda)$ to $p(\theta | y)$.
+MCMC is used to create a sample from $p(\theta | y)$, and any function of interest from $p(\theta | y)$ can be found with the relevant function applied to the sample, while VB replaces $p(\theta | y)$ with an approximation $q(\theta | \lambda)$, where $\lambda$ is a vector of auxillary parameters associated with the approximation and not the model.
+For example, if $q(\theta | \lambda)$ is Gaussian $\lambda$ would be the mean and variance of $\theta$.
+Variational Bayes aims to choose the distribution $q(\theta | \lambda)$ such that the Kullback-Leibler divergence from $q(\theta | \lambda)$ to $p(\theta | y)$ is minimised.
 
@@
 \begin{figure}[h]
-\label{MCMCplot}
 \centering
-\includegraphics[scale = 0.5]{MCMC.png}
+\includegraphics[scale = 0.6]{MCMC.png}
 \caption{MCMC draws for the AR(2) model with $T = 150, \phi_1 = 0.75, \phi_2 = 0.2$ and $\sigma^2 = 1$.
 Approximately 24\% of both $\phi$ parameters were accepted by the Metropolis Hastings step, then the first 10,000 draws were discarded and the remainder thinned to 1 in 200 to make dependence between draws negligible.}
+\label{MCMCplot}
 \end{figure}
@@
 
-Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the search for an optimal approximating dististribution to the class of factorisable distributions
+Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the class of distributions Variational Bayes selects from to the set of factorisable distributions,
 
@@
 
-For illustration, consider data $y_i, i = 1, \dots, n$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution where $p(\mu) \sim \mathcal{N}(\gamma, \lambda)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$.
+For illustration, consider data $y_i$ for $i = 1, \dots, n$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution with priors $p(\mu) \sim \mathcal{N}(\gamma, \lambda)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$.
 It can be shown that
@@
 \label{mf3}
-q(\mu) \propto \exp \left\{ \frac{-(n\lambda + \mathbb{E}(\sigma^{-2}))}{2\lambda\mathbb{E}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}(\sigma^{-2})\gamma + \lambda \sum_{i=1}^{n} y_i}{n \lambda + \mathbb{E}(\sigma^{-2})} \right) \right\}
+q(\mu) \propto \exp \left\{ \frac{-(n\lambda + \mathbb{E}(\sigma^{-2}))}{2\lambda\mathbb{E}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}(\sigma^{-2})\gamma + \lambda \sum_{i=1}^{n} y_i}{n \lambda + \mathbb{E}(\sigma^{-2})} \right)^2 \right\}
 \end{equation}
@@
 \label{mf4}
-q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + \mathbb{E}(\mu^2) - 2 y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
+q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + n\mathbb{E}(\mu^2) - 2\sum_{i=1}^{n} y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
 \end{equation}
@@
 
-In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One method in \citet{Friston2006} uses a Laplace approximation to substitute a Gaussian distribution for these unrecognizable $q_i(\theta_i | \lambda_i)$, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used in (\ref{mf2}).
+In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One method in \citet{Friston2006} uses a Laplace approximation to substitute a Gaussian distribution for these unrecognizable $q_i(\theta_i | \lambda_i)$, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).
 
@@
  \KwResult{Mean Field Approximation}
- Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution, or substitute a further approximation if neccesary.\;
- Derive the set of mean field equations such as in (\ref{mf5}) - (\ref{mf6}).\;
+ Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution or substitute a further approximation if neccesary.\;
+ Derive the set of mean field equations, such as (\ref{mf5}) - (\ref{mf6}).\;
  Initialise $\lambda$ randomly. \;
@@
 
-Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure (\ref{MCMCplot}) shows strong dependence between $\phi_1$ and $\phi_2$.
+Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure \ref{MCMCplot} shows strong dependence between $\phi_1$ and $\phi_2$.
 Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable.
@@
 
-Figure (\ref{VBfit}) shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.
+Figure \ref{VBfit} shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.
 
 \begin{figure}[h]
-\label{VBfit}
 \centering
@@
 \caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model described in section 2.1.
-SVB resulted in an almost identical posterior to MCMC while taking X less time to run.
-(working from home, will check runtime at uni)}
+SVB resulted in an almost identical posterior to MCMC, requiring around one hundredth of the runtime for MCMC.}
+\label{VBfit}
 \end{figure}
@@
 
-\begin{itemize}
-\item Describe and plot load Data
-\item Plot price data to emphasize price spikes
-\item Follow up the intro to talk a bit about load forecasting
-\end{itemize}
+Half hourly electricity load data for Victoria, Australia has been collected for the 2011-2015 period, and is plotted in Figure \ref{loadplot}.
+The time series displays strong seasonal characteristics, with a yearly, day of week, and time of day pattern.
+Electricity load is strongly dependent on temperature, with high load experienced on both very cold and very hot days.
+In summer periods, load is often low but shows extreme volatility, while load slowly rises then falls in winter and displays significantly less volatility.
+
+The electricity market in Victoria operates by market participants sending electricity bid and offers for each five minute period to  AEMO, a centralised market operator.
+AEMO then determines the dispatch price and load for that five minute period, and determines the final spot price participants pay for each half-hour period by averaging the six five minute dispath prices in that half-hour.
+Market participants are allowed to revise bids and offers at any point before dispatch.
+The electricity supply curve is strongly J shaped, as coal generators are able to provide a large load cheaply but if demand exceeds this level the price spikes as other generators must be brought online expensively.
+These characteristics of the electricity market leads to a large amount of volatility in prices demonstrated by Figure \ref{rrpplot}.
+The seasonal effects of the load data are also present in prices, but these effects are dominated by irregular spikes as high as \$9,974/mWh, compared to a four year mean of \$42/mWh.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale = 0.5]{load_timeseries.png}
+\caption{Half hourly load data measured in megawatt hours for Victoria from January 1 2011 to December 31 2015.
+The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent sine curve in winter.
+Note: Replace with daily average to make the pattern clearer?}
+\label{loadplot}
+\end{figure}.
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale = 0.5]{RRP_timeseries.png}
+\caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015.
+The price shows a strong time of week and time of day seasonal effect in the low price periods, and has irregular spikes.}
+\label{rrpplot}
+\end{figure}.
 
@@
 
-\citet{Taylor2003} explores the seasonal properties of electricity demand and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}), and later in \citet{Taylor2008} verifies that this approach is superior to other common time-series models for very short-term load forecasting \citep{Taylor2008}.
+\citet{Taylor2003} explores the seasonal properties of minute-by-minute electricity demand in the United Kingdom and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}).
+Yearly effects are omitted as their inclusion requires a large number of latent states to be estimated.
+This model is later verified in \citet{Taylor2008} as superior to other common time-series models for very short-term load forecasting.
 

### 2017-02-19 Confirmation/confirmation_V2.tex
tex updates

@@
 
-Electricity prices have been subject to an extensive forecasting literature and is known to be extremely volatile.
+Electricity prices have been subject to an extensive forecasting literature and are known to be extremely volatile.
 For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of oberservations less than \$100/mWh, but on 39 occaisions the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29, 2012.
-For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period is more important than the mean value.
-To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it can average over parameter and model uncertainty that is often ignored in frequentist methods.
+For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period can be more important than the mean value.
+To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it can average over parameter and model uncertainty in a relatively straightforward manner, which can be difficult in frequentist methods.
 See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general.
 
-Bayesian methods are computationally complex as they require integration over the set of unknown parameters, which must be large in a model that can capture the strong seasonal patterns in short term electricity load \citep{Taylor2003}, which is the main driver for electricity price and the focus of this work.
-To avoid this problem there are a range of techniques available to obtain the predictive distribution which do not involve integration, notably Markov Chain Monte Carlo (MCMC) methods.
-While MCMC has very strong convergence properties, computation is slow for even moderately complex models, which is problematic as Electricity demand in Victoria operates on a five minute interval.
+Bayesian methods are computationally complex as they require integration over the set of unknown parameters, which is often of a high dimension in a model that can capture the strong seasonal patterns in short term electricity load \citep{Taylor2003} that is a major driver for changes in electricity load as well as price.
+To avoid this problem there are a range of techniques available to obtain the predictive distribution which do not involve slow grid based numerical integration, notably Markov Chain Monte Carlo (MCMC) methods.
+While MCMC has very strong convergence properties, computation is slow for even moderately complex models, which is problematic as electricity demand in Victoria operates on a five minute interval.
 The predictive density for the next period's electricity load may not be available before it is observed.
 
-Alternatives to MCMC include Variational Bayes (VB), which aim to replace the unknown true posterior distribution for a vector of unknown parameters $\theta$ with a parametric approximation.
+Alternatives to MCMC include Variational Bayes (VB), which aim to replace the unknown true posterior distribution for a vector of unknown parameters $\boldsymbol{\theta}$ with a parametric approximation.
 Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2011-2015 period in Victoria.
-The advantages in computation time for VB will be compared to the loss of statisitcal accuracy associated with the use of an approximation of the posterior instead of the truth.
+The advantages in computation time for VB will be compared to the loss of statistical accuracy associated with the use of an approximation of the posterior instead of the truth.
 
@@
 Exact Bayesian inference in the context of forecasting uses the predictive density of $Y_{T+1}$ conditioned on the observed series, $y_{t}, t = 1, \dots T$, which is denoted by $y_{1:T}$.
-The first step requires the posterior distribution of the model parameters $\theta$, $p(\theta | y)$, where
+The first step requires the posterior distribution of the model parameters $\boldsymbol{\theta}$, $p(\boldsymbol{\theta} | y_{1:T})$, where
 
@@
 \label{posterior}
- p(\theta | y_{1:T}) = \frac{p(y_{1:T}|\theta)p(\theta)}{\int_\theta p(y_{1:T}|\theta)p(\theta) d\theta}.
+ p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
 \end{equation}
@@
 Generally the solution to (\ref{posterior}) is unavailable. This section will offer two alternatives, MCMC and VB.
-MCMC is used to create a sample from $p(\theta | y)$, and any function of interest from $p(\theta | y)$ can be found with the relevant function applied to the sample, while VB replaces $p(\theta | y)$ with an approximation $q(\theta | \lambda)$, where $\lambda$ is a vector of auxillary parameters associated with the approximation and not the model.
-For example, if $q(\theta | \lambda)$ is Gaussian $\lambda$ would be the mean and variance of $\theta$.
-Variational Bayes aims to choose the distribution $q(\theta | \lambda)$ such that the Kullback-Leibler divergence from $q(\theta | \lambda)$ to $p(\theta | y)$ is minimised.
+MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, and any function of interest from $p(\boldsymbol{\theta} | y_{1:T})$ can be found with the relevant function applied to the sample, while VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxillary parameters associated with the approximation and may depend on the observations $y_{1:T}$.
+For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian $\boldsymbol{\lambda}$ would be the mean and variance of $\boldsymbol{\theta}$.
+Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
 
@@
 
-There are many types of MCMC algorithm, the most common being a Gibbs sampler which iteratively samples from a $k$ dimensional parameter vector $\theta$ via the conditional distributions
+There are many types of MCMC algorithm, one of the most common of these is the Gibbs sampler, which iteratively samples through each element of a $k$ dimensional parameter vector $\boldsymbol{\theta}$ via the conditional distributions
 
@@
 
-With enough iterations, these samples converge in distribution to $p(\theta | y)$ and can be used for inference.
+With enough iterations, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$ and can be used for inference.
 The computation time per iteration and number of iterations required for convergence is problem specific but typically increases with model complexity.
@@
 \label{AR2}
-y_t = \rho_1 y_{t-1} + \rho_2 y_{t-2} + \epsilon_t
+y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
 \end{equation}
 
-where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. The likelihood of parameters $\theta = (\rho_1, \rho_2m \sigma^2)'$ is given by
+where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$ is given by
 
@@
 \label{likelihood}
-L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
+L(\boldsymbol{\theta} | y_{1:T}) &= p(y_1, y_2 | \boldsymbol{\theta}) \prod_{t=3}^{T}p(y_t | y_{1:t-1},\boldsymbol{\theta}) \nonumber \\
 &= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
@@
 \begin{equation}
-\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1), \gamma(0) \end{array} \right],\
+\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
 \end{equation}
@@
 p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
-p(\rho_1, \rho_2) &\propto \mathbb{I}(\rho_2 > -1)\mathbb{I}(\rho_2 < 1 + \rho_1) \mathbb{I}(\rho_2 < 1 - \rho_1) \nonumber
+p(\phi_1, \phi_2) &\propto \mathbb{I}(\phi_2 > -1)\mathbb{I}(\phi_2 < 1 + \phi_1) \mathbb{I}(\phi_2 < 1 - \phi_1) \nonumber
 \end{align}
 
-are used, where $\mathbb{I}$ is the indicator and $p(\rho_1, \rho_2)$ defines the AR(2) stationary region.
+are used, where $\mathbb{I}$ is the indicator function that equals one when the condition in the brackets is true and zero otherwise.
+This prior for $p(\phi_1, \phi_2)$ defines the AR(2) stationary region and is independent to $p(\sigma^2)$.
 
-The conditional distribution $p(\sigma^2 | y, \phi_1, \phi_2)$ is Inverse Gamma, while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step with truncated normal random walk proposals.
+The conditional distribution $p(\sigma^2 | y_{1:T}, \phi_1, \phi_2)$ is Inverse Gamma, while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step with truncated normal random walk proposals.
 The truncation ensures that draws are only taken from the AR(2) stationary region.
@@
 
-A faster alternative to MCMC is Variational Bayes, which introduces an approximating distribution $q(\theta | \lambda)$ choosing the functional form $q$ and given auxillary parameters $\lambda$ so that  misimises the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ the true posterior $p(\theta | y)$.
+A faster alternative to MCMC is Variational Bayes, which introduces an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ choosing the functional form $q$ and given auxillary parameters $\boldsymbol{\lambda}$ so that the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
 The KL divergence is defined by
@@
 \label{KL-def}
-KL[q(\theta | \lambda)||p(\theta | y)] = \int q(\theta | \lambda) \ln \left( \frac{q(\theta | \lambda)}{p(\theta | y)}\right) d\theta.
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T}] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
 \end{equation}
 
-The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\theta | y)$ and $q(\theta | \lambda)$ that will equal zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}.
-Note that $KL[q(\theta | \lambda)||p(\theta | y)]$ can be expressed as
+The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that will equal zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}.
+Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
 
@@
 \label{KL-ELBO}
-KL[q(\theta | \lambda)||p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q, y)
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q, y_{1:T})
 \end{equation}
 
-where $\mathcal{L}(q, y)$ is known as the Evidence Lower Bound (ELBO), defined by
+where $\mathcal{L}(q, y_{1:T})$ is known as the Evidence Lower Bound (ELBO), defined by
 
@@
 \label{ELBO}
-\mathcal{L}(q, y) = \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta.
+\mathcal{L}(q, y_{1:T}) = \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
 \end{equation}
 
-Since $\ln(p(y))$ is constant given $y$, maximising (\ref{KL-ELBO}) with respect to $q$ is equivalent to minimising (\ref{KL-def}).
+Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}).
 Maximising the ELBO is much more convenient than minimising the KL Divergence, which can be seen in the following sections based on the two major implementations of Varitaional Bayes: Mean Field Variational Bayes and Stochastic Variational Bayes.
-Each of these take advantage of the functional form of the ELBO and offer computationally efficient algorithms to find a $q(\theta | \lambda)$ that is optimal within a particular clas of distributions.
+Each of these take advantage of the functional form of the ELBO and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.
 
@@
 \label{mf1}
-q(\theta|\lambda) = \prod_i^k q_i(\theta_i | \lambda_i).
+q(\theta|\lambda) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
 \end{equation}
@@
 Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars.
-Each $\theta_i$ has an associated vector $\lambda_i$ which may be of a higher dimension than $\theta_i$.
-$\lambda_i$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\lambda_i)$.
+Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may be of a different dimension to $\theta_i$.
+$\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.
 
-MFVB is widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}), particularly in exponential family models \citep{Wainwright2008}.
+MFVB is widely used as it greatly simplifies maximisation of the ELBO, particularly in exponential family models (\citealp{Jordan1999}; \citealp{Ghahramani2000}, \citealp{Wainwright2008}).
 Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the computation is simple.
@@
 \label{mf2}
-q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])
+q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))])
 \end{equation}
 
-For illustration, consider data $y_i$ for $i = 1, \dots, n$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution with priors $p(\mu) \sim \mathcal{N}(\gamma, \lambda)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$.
+For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution with priors $p(\mu) \sim \mathcal{N}(\gamma, \tau)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$.
 It can be shown that
@@
 \label{mf3}
-q(\mu) \propto \exp \left\{ \frac{-(n\lambda + \mathbb{E}(\sigma^{-2}))}{2\lambda\mathbb{E}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}(\sigma^{-2})\gamma + \lambda \sum_{i=1}^{n} y_i}{n \lambda + \mathbb{E}(\sigma^{-2})} \right)^2 \right\}
+q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T\tau + \mathbb{E}_{q(\sigma^2)}(\sigma^{-2}))}{2\tau\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\gamma + \lambda \sum_{t=1}^{T} y_t}{T \tau + \mathbb{E}_{q(\sigma^2 )}(\sigma^{-2})} \right)^2 \right\}
 \end{equation}
@@
 \label{mf4}
-q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + n\mathbb{E}(\mu^2) - 2\sum_{i=1}^{n} y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
+q(\sigma^2 | \boldsymbol{\lambda}_2) \propto \sigma^{-2(T/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{t=1}^{T}y_t^2 + n\mathbb{E}_{q(\mu)}(\mu^2) - 2\sum_{t=1}^{T} y_t \mathbb{E}_{q(\mu)}(\mu)) - \beta}{\sigma^2} \right\}.
 \end{equation}
 
-It is evident that $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ and $q(\sigma^2 | \lambda) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$.
-The parameters $\lambda = (\tilde{\gamma}, \tilde{\lambda}, \tilde{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations:
+It is evident that $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$.
+The parameters $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations:
 
 \begin{align}
-\tilde{\gamma} &= \frac{\alpha / \beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta} \label{mf5} \\ 
-\tilde{\lambda} &= \frac{n \alpha / \beta}{n \lambda + \alpha / \beta} \\
-\tilde{\alpha} &= \frac{n}{2} + \alpha \\
-\tilde{\beta} &= \frac{1}{2} \left(\sum_{i=1}^{n} y_i^2 + n(\tilde{\gamma}^2 + \tilde{\lambda}) - 2 \sum_{i=1}^{n} y_i \tilde{\gamma} \right) + \beta. \label{mf6}
+\tilde{\gamma} &= \frac{\tilde{\alpha} / \tilde{\beta} \gamma + \tau \sum_{t=1}^{T} y_t} {T \lambda + \tilde{\alpha} / \tilde{\beta}} \label{mf5} \\ 
+\tilde{\tau} &= \frac{T \tilde{\alpha} / \tilde{\beta}}{T \tau + \tilde{\alpha} / \tilde{\beta}} \label{mf6} \\
+\tilde{\alpha} &= \frac{T}{2} + \alpha  \label{mf7} \\
+\tilde{\beta} &= \frac{1}{2} \left(\sum_{t=1}^{T} y_t^2 + T(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \sum_{t=1}^{T} y_t \tilde{\gamma} \right) + \beta. \label{mf8}
 \end{align}.
@@
 
-In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One method in \citet{Friston2006} uses a Laplace approximation to substitute a Gaussian distribution for these unrecognizable $q_i(\theta_i | \lambda_i)$, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).
+In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ One method in \citet{Friston2006} uses a Laplace approximation for these unrecognizable $q_i(\theta_i |\boldsymbol{\lambda_i})$ by matching the mode and inverse hessian matrix to a gaussian distribution, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).
 
-Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | y, \theta_{\setminus i}$ in a Gibbs MCMC scheme, with dependence on other parameters replaced by their expectations.
-The optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$, if it exists in a recognisable form such as the exponential family.
+Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | \boldsymbol{\theta}_{j \neq i}, y_{1:T})$ in a Gibbs MCMC scheme, with dependence on other parameters replaced by their expectations.
+The optimal approximating family for $\theta_i$, $q_i(\theta_i |\boldsymbol{\lambda_i})$, will come from the same distributional family as the conditional distribution $p(\theta_i | \boldsymbol{\theta}_{j \neq i}, y_{1:T})$, if it exists in a recognisable form such as the exponential family.
 The secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable distributions.
@@
 
-The required algorithm is known as a coordinate ascent algorithm, and is described below where $\lambda$ is a $p$ dimensional vector.
+The required algorithm is known as a coordinate ascent algorithm, and is described below for the model discussed above
 
@@
  \KwResult{Mean Field Approximation}
- Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution or substitute a further approximation if neccesary.\;
- Derive the set of mean field equations, such as (\ref{mf5}) - (\ref{mf6}).\;
- Initialise $\lambda$ randomly. \;
- \While{Not converged}{
-  \For{$i =  1$ \KwTo $k$}{
-      Hold $\lambda_{j \neq i}$ fixed\;
-      Update $\lambda_i$ using the relevant mean field equation.
-     }
+ Use (\ref{mf3}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
+ Anaytically derive the set of mean field equations in \ref{mf5}) - (\ref{mf8})\;
+ Initialise $\boldsymbol{\lambda}^{(1)}$ randomly.\;
+ Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
+ Set $\tilde{\alpha}$ using \ref{mf7}.\;
+ \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
+  Set $\tilde{\gamma}^{(m)}$ using \ref{mf5}.\;
+  Set $\tilde{\tau}^{(m)}$ using \ref{mf6}.\;
+  Set $\tilde{\beta}^{(m)}$ using \ref{mf8}.\;
+  Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$.\;
+  Set $m = m + 1$.\;
  }
@@
 
-The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\theta | \lambda), y)$ in (\ref{ELBO}) with respect to $\lambda$ and the following updating step is applied until the ELBO converges to a maximum.
+The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ in (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ and the following updating step is applied until the ELBO converges to a maximum.
 
@@
 \label{SGA1}
-\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m)}), y),
+\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}),
 \end{equation}
 
-where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ with respect to each element of $\lambda$.
-This update requires some initial values $\lambda^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
+where $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$.
+This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
 If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.
@@
 \begin{align}
-\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \\
-\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty.
+&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
+&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
 \end{align}
 
-Whilst a global maximum is desired, the ELBO is a high dimensional problem specific function that makes finding the global maximum extremely difficult and often only a local maximum can be found.
-One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior
+Whilst a global maximum is desired, Variational Bayes theory does not offer any insight to the shape of the ELBO so determining if a local maximum is a global maximum is difficult.
+One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior.
 
-SVB can find only the optimal values for $\lambda$ that are optimal for a distributional family $q$ that is specified by the user.
-To run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO.
-$q(\theta | \lambda)$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable, and SVB can only be applied to models where the log likelihood for $y$ is able to be evaluated.
+SVB can find only the optimal values for $\boldsymbol{\lambda}$ for a user specified distributional family $q$, and to run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO.
+$q$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\boldsymbol{\lambda}$ and integration with respect to $\boldsymbol{\theta}$ are interchangable, and SVB can only be applied to models where the log likelihood for $y_{1:T}$ is able to be evaluated.
 Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by
@@
 \label{SGA2}
-\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda^{(m)}))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda^{(m)})))
+\nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)}))] (\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)})))
 \end{equation}
 
-where $s = 1, \dots, S$ indicates simulations from $q(\theta | \lambda^{(m)})$.
+where $s = 1, \dots, S$ indicates simulations from $q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})$.
 
@@
 \label{SGA3}
-G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\theta | \lambda^{(m)})\right)^2,
+G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(j)}), y_{1:T})\right)^2,
 \end{equation}
@@
 
-The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\lambda$ vector.
+The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\boldsymbol{\lambda}$ vector.
+
+\vspace{2mm}
 
@@
  \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m)})$\;
+ Initialise $\boldsymbol{\lambda}^{(1)}$\;
+ Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
+ \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
+  Simulate $\boldsymbol{\theta}^s$ for $s = 1, \dots S$ from $q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})$\;
   \For{$i =  1$ \KwTo $p$}{
@@
       }
-  Update $\lambda^{(m+1)}$ from (\ref{SGA1})\;
+  Update $\boldsymbol{\lambda}^{(m+1)}$ from (\ref{SGA1})\;
+  Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$\;
   Set $m = m + 1$\;
@@
 
-Armed with an algorithm to find the Variational Bayes optimal paramerters $\lambda$ for a distribution $q$, there is a requirement to have a distribution $q$ that is itself optimal.
+Armed with an algorithm to find the Variational Bayes optimal parameters $\boldsymbol{\lambda}$ for a distribution $q$, there is a requirement to have a distribution $q$ that is itself optimal.
 Copulas are a flexible tool for constructing these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions.
@@
 However there are $O(k!)$ ways to factorise the full copula and choose a family for each bivariate copula, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
-\citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\theta$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.
+\citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\boldsymbol{\theta}$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.
 
-Given a sample of $\theta$, \citet{Diimann2013} provide an algorithm to select the best vine factorisation by maximising Kendell's Tau, and the best family for each bivariate copula by maximising an information criterion such as AIC.
+Given a sample of $\boldsymbol{\theta}$, \citet{Dssmann2013} provide an algorithm to select the best vine factorisation by maximising Kendall's Tau, and the best family for each bivariate copula by maximising an information criterion such as AIC.
 In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date.
 The approach used in this thesis is to run an MCMC algorithm, thin the resulting draws until they are effectively independent, choose an optimal family for each marginal distribution by AIC, then run (Vine copula algorithm here) to infer an optimal vine copula.
-This results in an approximating distribution $q(\theta | \lambda)$ that can then be used in an SVB algorithm to update $q(\theta | \lambda)$ as more observations become available.
+This results in an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that can then be used in an SVB algorithm to update $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ as more observations become available.
 MCMC can be ran simultaenously and the resulting sample can be used to check if the distribution $q$ should be changed.
@@
 \begin{itemize}
-\item $q(\phi_1)$ - Gaussian marginal
-\item $q(\phi_2)$ - Gaussian marginal
-\item $q(\sigma^2)$ - Inverse Gamma marginal
-\item $c(Q(\phi_1), Q(\phi_2))$ - Gaussian copula
+\item $q(\phi_1 | \boldsymbol{\lambda}_1)$ - Gaussian marginal
+\item $q(\phi_2| \boldsymbol{\lambda}_2)$ - Gaussian marginal
+\item $q(\sigma^2| \boldsymbol{\lambda}_3)$ - Inverse Gamma marginal
+\item $c(Q(\phi_1), Q(\phi_2) | \boldsymbol{\lambda}_4)$ - Gaussian copula
 \item $c(Q(\phi_1), Q(\sigma^2))$ - Independent
@@
 
-The literature for electricity forecasting is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review.
+The literature for electricity forecasting for both load and price is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review.
 Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016} being notable examples, which used boot strapping and quantile regression respectively to generate forecast densities.
@@
 Half hourly electricity load data for Victoria, Australia has been collected for the 2011-2015 period, and is plotted in Figure \ref{loadplot}.
-The time series displays strong seasonal characteristics, with a yearly, day of week, and time of day pattern.
+The time series displays strong seasonal characteristics, with yearly, day of week, and time of day patterns.
 Electricity load is strongly dependent on temperature, with high load experienced on both very cold and very hot days.
@@
 Market participants are allowed to revise bids and offers at any point before dispatch.
-The electricity supply curve is strongly J shaped, as coal generators are able to provide a large load cheaply but if demand exceeds this level the price spikes as other generators must be brought online expensively.
+The electricity supply curve is strongly hockey stick shaped, as renewables and coal generators are able to provide a large load cheaply but if demand exceeds this level the price spikes as other generators must be brought online expensively.
 These characteristics of the electricity market leads to a large amount of volatility in prices demonstrated by Figure \ref{rrpplot}.
@@
 \centering
-\includegraphics[scale = 0.5]{load_timeseries.png}
+\includegraphics{load_timeseries.png}
 \caption{Half hourly load data measured in megawatt hours for Victoria from January 1 2011 to December 31 2015.
-The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent sine curve in winter.
-Note: Replace with daily average to make the pattern clearer?}
+The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent sine curve in winter.}
 \label{loadplot}
@@
 \centering
-\includegraphics[scale = 0.5]{RRP_timeseries.png}
+\includegraphics{RRP_timeseries.png}
 \caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015.
@@
 \citet{Taylor2003} explores the seasonal properties of minute-by-minute electricity demand in the United Kingdom and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}).
-Yearly effects are omitted as their inclusion requires a large number of latent states to be estimated.
+Taylor omits yearly effects as their inclusion requires a large number of latent states to be estimated.
 This model is later verified in \citet{Taylor2008} as superior to other common time-series models for very short-term load forecasting.
@@
 
-where $m_1$ and $m_2$ are the lengths of the daily and weekly cycle, restricting the smoothing parameters $\alpha, \delta, \omega$ to lie in $(0, 1)$.
+where $m_1$ and $m_2$ are the lengths of the daily and weekly cycle, and the smoothing parameters $\alpha, \delta, \omega$ are also restricted to to lie in $(0, 1)^3$.
 This can be rewritten as a single source of error state-space model \citep{Snyder1985},
@@
 
-The unknown parameters are $\theta = (\alpha, \delta, \omega)', \sigma^2$ and $b_0 = (l_0, d_0, \dots, d_{-(m_1 - 1)}, w_0, \dots, w_{-(m_2 - 1)})'$ a $m_1 + m_2 + 1$ length vector of the initial states of the latent variables.
-\citet{Forbes2000} derives the marginal distribution $p(\theta | y)$ and notes that, conditioned on a draw of these parameters, the remaining unknown parameters can be sampled as a Bayesian linear regression problem.
-This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}), but restricting the latent parameters via
+The unknown parameters are $\boldsymbol{\theta} = (\alpha, \delta, \omega)', \sigma^2$ and $\boldsymbol{b}_0 = (l_0, d_0, \dots, d_{-(m_1 - 1)}, w_0, \dots, w_{-(m_2 - 1)})'$ a $k = m_1 + m_2 + 1$ length vector of the initial states of the latent variables.
+
+\citet{Forbes2000} provides a transformation to the model that allows a straight-forward Bayesian analysis in small samples, noting that the state space model can be expressed as
+
+\begin{align}
+y_t &= \textbf{x}' \boldsymbol{b}_t + e_t \label{ss_me} \\
+\textbf{b}_t &= \textbf{T} \textbf{b}_{t-1} + \tilde{\boldsymbol{\theta}} e_t \label{ss_te}.
+\end{align}
+
+Using $D = T - \tilde{\boldsymbol{\theta}} \textbf{x}'$, the transition equation in (\ref{ss_te}) can be written as
+
+\begin{equation}
+b_t = D b_{t-1} + \tilde{\boldsymbol{\theta}} y_t \label{ss_te_2}.
+\end{equation}
+
+Setting $\bar{b}_0$ = 0, the recursion in (\ref{ss_te_2}) allows the transformed observations $\tilde{y}_t$ to be obtained via
+
+\begin{equation}
+\tilde{y}_t = y_t - x'\bar{b}_{t-1} \label{ss_rp_y}.
+\end{equation}
+
+Furthermore, transformed regression variables $\tilde{\textbf{x}}_t$ can be obtained from $\tilde{\textbf{x}}_t = D\tilde{\textbf{x}}_{t-1}$ with $\tilde{\textbf{x}}_0 = \textbf{x}$.
+Collecting these as $\widetilde{X}' = (\tilde{\textbf{x}}_1, \tilde{\textbf{x}}_2, \dots, \tilde{\textbf{x}}_T)$ and $\widetilde{Y}' = (\tilde{y}_1, \tilde{y}_2, \dots, \tilde{y}_T)$, and assuming the errors $e_t$ are Gaussian, the model likelihood is
+
+\begin{equation}
+L(\boldsymbol{\theta}, \sigma^2, \boldsymbol{b}_0 | y_{1:T}) \propto \sigma^{-T} \exp \left\{ \frac{-1}{2\sigma^2}(\widetilde{Y} - \widetilde{X}\boldsymbol{b}_0)'(\widetilde{Y} - \widetilde{X}\boldsymbol{b}_0) \right\}.
+\end{equation}
+
+Hence once conditioned on a draw of $\boldsymbol{\theta}$, the remaining unknown parameters can be sampled as a Bayesian linear regression problem.
+
+\citet{Forbes2000} provides the marginal distribution of $\theta$,
+
+\begin{equation}
+\label{exp-sm-marginal}
+p(\boldsymbol{\theta} | y_{1:T}) \propto \left| \widetilde{X}' \widetilde{X} \right|^{-1/2} \tilde{s}^{-(T-k)} p(\theta),
+\end{equation}
+
+where $\tilde{s}^2 = (\widetilde{Y} - \widetilde{X}\hat{\boldsymbol{b}}_0)'(\widetilde{Y} - \widetilde{X}\hat{\boldsymbol{b}}_0) / (T - (m_1 + m_2 + 1))$ is the sum of squared errors for $\hat{\boldsymbol{b}}_0 = (\widetilde{X}'\widetilde{X})^{-1}\widetilde{X}'\widetilde{Y}$.
+With the seasonality parameterisation in (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}), the matrix $\widetilde{X}' \widetilde{X}$ is singular, but restricting the latent variables via
 
@@
 
-\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$,
+and $\boldsymbol{b}_0 = (l_0, d_0, \dots, d_{-(m_1 - 2)}, w_, \dots, w_{-(m_2 - 2)})'$ is a length $k = m_1 + m_2 - 1$ vector.
 
-\begin{equation}
-\label{exp-sm-marginal}
-p(\theta | y_{1:T}) \propto \left| \widetilde{X}' \widetilde{X} \right|^{-1/2} \tilde{s}^{-(T-(m_1 + m_2 - 1))} p(\theta),
-\end{equation}
+\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$ in (\ref{exp-sm_marginal}), however the problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large, as is the case with high frequency electricity load data.
+An efficient MCMC algorithm using the conditional distribution of $p(\theta | y_{1:T}, \sigma^2, b_0)$ may be required to sample the full posterior distribution $p(\theta, \sigma^2, b_0 | y_{1:T}$ and obtain an approximating distribution $q(\theta, \sigma^2, b_0 | \lambda)$ with a reasonable goodness of fit.
+From this point, Stochastic Variational Bayes will be used to update the posterior distribution as more data is observed to $p(\theta, \sigma^2, b_0 | y_{1:T+J}), J = 1, 2, \dots$ and provide the predictive density $p(y_{T+J+1} | y_{T+J})$.
 
-where $\widetilde{X}$ and $\tilde{s}$ are functions of $\theta$ and $y$.
-The problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible difficult, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large.
+\section{Discussion}
 
-\begin{itemize}
-\item Should try and get an exponential smoothing MCMC working using Metropolis-Hastings for $\theta$.
-Log densities help with the $T$ being too large problem but the grid is too large for numerical integration to be a good idea.
-\end{itemize}
+Put main points and issues to investigate here
+
+- When parameters change, models with non-simple filtering methods to obtain changes in states. ie variable parameter state space models.
 
 \section{Timeline}
-\begin{itemize}
-\item Put something here eventually
-\end{itemize}
+
+Put this in!
 

### 2017-02-19 Confirmation/state_space_r.R

@@ -209,9 +209,7 @@ SGA = function(y, lambda, meanX, varX, s, M, threshold, eta, Vine, n){
   epsilon = RVineSim(n, Vine)
   LBnew = ELBO(Vine, lambda, meanX, varX, y, epsilon, n)
   diff = threshold + 1
-  Par1 = matrix(0, T + 5, T + 5)
-  Par2 = matrix(0, T + 5, T + 5)
-  
+
   while(diff > threshold | iter < 10){
     if(iter >= M){
       break

### 2017-02-19 Confirmation/statespace.R

@@ -132,3 +132,8 @@ fitdist(thetaKeep[,2], "norm", method = "mle")$bic
 log(1000)*3 - 2*fit.st(thetaKeep[,2])$ll.max
 log(1000)*8 - 2*normalmixEM(thetaKeep[,2], k = 3)$loglik
 log(1000)*5 - 2*normalmixEM(thetaKeep[,2], k = 2)$loglik
+
+
+Vine <- readRDS("Vine.rds")
+table(Vine$family[101:105,])
+table(Vine$family[101,1:100])

### 2017-02-20 Confirmation/confirmation_V2.tex
report work

@@
 \centering
-\includegraphics[scale = 0.6]{MCMC.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{MCMC.png}
 \caption{MCMC draws for the AR(2) model with $T = 150, \phi_1 = 0.75, \phi_2 = 0.2$ and $\sigma^2 = 1$.
@@
 
-\citet{Joe1994} introduces the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas in a similar way to the factorisation of a joint distribution into a marginal and conditional distribution, for example
+This idea is extended by the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas in a similar way to the factorisation of a joint distribution into a marginal and conditional distribution, see \citet{Joe2014} and references within.
+An example of this factorisation is
 
@@
 A copula for a $k$ dimensional $\theta$ can be replaced with $k(k-1)/2$ bivariate copulas arranged in a vine copula, greatly increasing the amount of flexibility available to fit a distribution.
-However there are $O(k!)$ ways to factorise the full copula and choose a family for each bivariate copula, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
+However factorising the copula and choosing a family for each bivariate copula takes exponential time in the number of parameters, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
 \citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\boldsymbol{\theta}$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.
 
-Given a sample of $\boldsymbol{\theta}$, \citet{Dssmann2013} provide an algorithm to select the best vine factorisation by maximising Kendall's Tau, and the best family for each bivariate copula by maximising an information criterion such as AIC.
+Given a sample of $\boldsymbol{\theta}$, \citet{Dssmann2013} provide an algorithm to select the best vine factorisation by maximising Kendall's Tau, and the best family for each bivariate copula by minimising an information criterion such as AIC.
 In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date.
@@
 \centering
-\includegraphics[scale = 0.5]{VBfit.png}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{VBfit.png}
 \caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model described in section 2.1.
@@
 \centering
-\includegraphics{load_timeseries.png}
+\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{load_timeseries.png}
 \caption{Half hourly load data measured in megawatt hours for Victoria from January 1 2011 to December 31 2015.
-The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent sine curve in winter.}
+The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent curve in winter.}
 \label{loadplot}
-\end{figure}.
+\end{figure}
 
@@
 \centering
-\includegraphics{RRP_timeseries.png}
+\includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{RRP_timeseries.png}
 \caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015.
-The price shows a strong time of week and time of day seasonal effect in the low price periods, and has irregular spikes.}
+While the price has time of week and time of day seasonal effects these features are dominated by irregular spikes.}
 \label{rrpplot}
-\end{figure}.
+\end{figure}
 

### 2017-02-20 Confirmation/confirmation_V2.tex
reference fixes

@@
 
-In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ One method in \citet{Friston2006} uses a Laplace approximation for these unrecognizable $q_i(\theta_i |\boldsymbol{\lambda_i})$ by matching the mode and inverse hessian matrix to a gaussian distribution, though the only requirement is that the substitute has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).
+In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ \citet{Friston2006} and \citet{Wang2013} suggests the use of a Gaussian distribution through either a Laplace approximation or a Delta Method transform, though the only requirement is that the substitute distribution has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).
 
@@
 
-Given a sample of $\boldsymbol{\theta}$, \citet{Dssmann2013} provide an algorithm to select the best vine factorisation by maximising Kendall's Tau, and the best family for each bivariate copula by minimising an information criterion such as AIC.
+Given a sample of $\boldsymbol{\theta}$, \citet{Dissmann2013} provide an algorithm to select the best vine factorisation by maximising Kendall's Tau, and the best family for each bivariate copula by minimising an information criterion such as AIC.
 In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date.

### 2017-02-21 Confirmation/confirmation_V2.tex
Revised draft

@@
 \tableofcontents
-\section{Introduction}
+\section{Introduction} \label{sec:Intro}
 
 Electricity prices have been subject to an extensive forecasting literature and are known to be extremely volatile.
-For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of oberservations less than \$100/mWh, but on 39 occaisions the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29, 2012.
+For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of observations less than \$100 per mWh, however on several occasions the price spiked to greater than $10,000\%$ of the mean value.
 For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period can be more important than the mean value.
-To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it can average over parameter and model uncertainty in a relatively straightforward manner, which can be difficult in frequentist methods.
+To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it accounts for both parameter and model uncertainty in a relatively straightforward manner, which can be difficult in frequentist methods.
 See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general.
 
-Bayesian methods are computationally complex as they require integration over the set of unknown parameters, which is often of a high dimension in a model that can capture the strong seasonal patterns in short term electricity load \citep{Taylor2003} that is a major driver for changes in electricity load as well as price.
-To avoid this problem there are a range of techniques available to obtain the predictive distribution which do not involve slow grid based numerical integration, notably Markov Chain Monte Carlo (MCMC) methods.
-While MCMC has very strong convergence properties, computation is slow for even moderately complex models, which is problematic as electricity demand in Victoria operates on a five minute interval.
-The predictive density for the next period's electricity load may not be available before it is observed.
+Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables.
+This set is quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}.
+While analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods.
+While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models.
+This is problematic for when attempting to forecast electricity load in Victoria, when realizations occur at five minute intervals, yhe predictive density for the next period's electricity load may not be available before it is observed.
 
-Alternatives to MCMC include Variational Bayes (VB), which aim to replace the unknown true posterior distribution for a vector of unknown parameters $\boldsymbol{\theta}$ with a parametric approximation.
-Variational Bayes will be explored as an alternative computation strategy to MCMC in the context of short term electricity load forecasting using data from the 2011-2015 period in Victoria.
-The advantages in computation time for VB will be compared to the loss of statistical accuracy associated with the use of an approximation of the posterior instead of the truth.
+Alternatives to MCMC include Variational Bayes (VB) methods, where the aim is to replace the unknown true posterior distribution for a vector of unknowns with a parametric approximation.
+The Variational Bayes approach will be explored in the thesis, to determine if it is a viable alternative computation strategy to MCMC, in the context of modelling and short term forecasting of Victorian electricity load.
+In particular, we seek to understand the benefits gained in terms of the computational time required to obtain a VB approximate forecast distribution against the loss in statistical accuracy associated with the use of an approximate posterior forecast distribution.
 
-\section{Bayesian Inference}
+The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of a simple time series model.
+Section \ref{sec:ElecLoad} will introduce a planned emperical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period.
+Section \ref{sec:Discussion} outlines the major areas this thesis will make a contribution to and Section \ref{sec:Timeline} outlines the planned timeline of these contributions.
 
-Exact Bayesian inference in the context of forecasting uses the predictive density of $Y_{T+1}$ conditioned on the observed series, $y_{t}, t = 1, \dots T$, which is denoted by $y_{1:T}$.
-The first step requires the posterior distribution of the model parameters $\boldsymbol{\theta}$, $p(\boldsymbol{\theta} | y_{1:T})$, where
+\section{Bayesian Inference} \label{sec:BayesInf}
 
+To facilitate the discussion, consideration is given to models where the unknowns may be summarized in a (finite) $k-$dimensional vector, denoted by $\boldsymbol{\theta}.$ In this setting, given an observed time series, denoted by $y_{1:T}= \{y_{t}, t = 1, \dots T\}$, the Bayesian forecast distribution associated with time $t=T+1$ is characterized by the conditional density
+\begin{equation}
+\label{predictive}
+p(y_{T+1} | y_{1:T}) =\int p(y_{T+1}|y_{1:T},\boldsymbol{\theta}) p(\boldsymbol{\theta}|y_{1:T}) d\boldsymbol{\theta},
+\end{equation}
+where $p(\boldsymbol{\theta}|y_{1:T})$ denotes the posterior density for $\boldsymbol{\theta}$, given by
 \begin{equation}
@@
 
-Generally the solution to (\ref{posterior}) is unavailable. This section will offer two alternatives, MCMC and VB.
-MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, and any function of interest from $p(\boldsymbol{\theta} | y_{1:T})$ can be found with the relevant function applied to the sample, while VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxillary parameters associated with the approximation and may depend on the observations $y_{1:T}$.
-For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian $\boldsymbol{\lambda}$ would be the mean and variance of $\boldsymbol{\theta}$.
+Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}) will be unavailable due to the complex functional form of the distributions involved. 
+In this section two alternative methods for computing the desired posterior distribution in (\ref{posterior}),  MCMC and VB, will be reviewed.
+In brief, MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, with any function of $\boldsymbol{\theta}$ that is desired estimated from that sample.
+In contrast, VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation, denoted by $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxiliary parameters associated with the approximation and may depend on the observations $y_{1:T}$.
+For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian, then $\boldsymbol{\lambda}$ would denote the mean and variance parameters that characeterise this normal distribution for $\boldsymbol{\theta}$.
 Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
+To obtain the predictive distribution in (\ref{predictive}), $y_{T+1}$ can be treated as an unknown variable in $\boldsymbol{\theta}$ and hence directly computed by either MCMC and VB.
+Alternatively, once the posterior distribution in (\ref{posterior}) is available, (\ref{predictive}) can be found by simulation.
 
-\subsection{Markov Chain Monte Carlo}
+\subsection{Markov Chain Monte Carlo} \label{sec:MCMC}
 
-There are many types of MCMC algorithm, one of the most common of these is the Gibbs sampler, which iteratively samples through each element of a $k$ dimensional parameter vector $\boldsymbol{\theta}$ via the conditional distributions
-
+There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler.
+The Gibbs sampler algorithm iteratively samples the components of the $k-$dimensional parameter vector $\boldsymbol{\theta}$ via each of the so-called full conditional distributions as follows,
 \begin{align}
@@
 \end{align}
+Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$ and can be used for forecasting and inference about the unknown parameters.
+The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
+On occasion, the full conditional distribution can not be recognised and a Metropolis-Hastings-within-Gibbs step can be used, see \citet{Gilks1995} for details.
 
-With enough iterations, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$ and can be used for inference.
-The computation time per iteration and number of iterations required for convergence is problem specific but typically increases with model complexity.
-
-To illustrate issues involved consider a simple auto-regressive time series model which will be used in the remainder of this section. This model is described by
-
-\begin{equation}
-\label{AR2}
-y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
-\end{equation}
-
-where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$. The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$ is given by
+For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution with prior distributions given by $\mu | \sigma^2 \sim \mathcal{N}(\gamma, \sigma^2/\tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$.
+It can be shown that
 
 \begin{align}
-\label{likelihood}
-L(\boldsymbol{\theta} | y_{1:T}) &= p(y_1, y_2 | \boldsymbol{\theta}) \prod_{t=3}^{T}p(y_t | y_{1:t-1},\boldsymbol{\theta}) \nonumber \\
-&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
+\mu | \sigma^2, y_{1:T} &\sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
+\sigma^2 | y_{1:T} &\sim Inv.Gamma  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2}
 \end{align}
+where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$.
+Posterior sampling can proceed by first simulating a draw of $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}).
 
-where
+\subsection{Variational Bayes} \label{sec:VB}
 
-\begin{equation}
-\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
-\end{equation}
-
-$\gamma(k)$ is the $k-th$ autocorrelation and $\textbf{y} = (y_1, y_2)'$.
-
-The priors
-
-\begin{align}
-p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
-p(\phi_1, \phi_2) &\propto \mathbb{I}(\phi_2 > -1)\mathbb{I}(\phi_2 < 1 + \phi_1) \mathbb{I}(\phi_2 < 1 - \phi_1) \nonumber
-\end{align}
-
-are used, where $\mathbb{I}$ is the indicator function that equals one when the condition in the brackets is true and zero otherwise.
-This prior for $p(\phi_1, \phi_2)$ defines the AR(2) stationary region and is independent to $p(\sigma^2)$.
-
-The conditional distribution $p(\sigma^2 | y_{1:T}, \phi_1, \phi_2)$ is Inverse Gamma, while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step with truncated normal random walk proposals.
-The truncation ensures that draws are only taken from the AR(2) stationary region.
-
-\begin{figure}[h]
-\centering
-\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{MCMC.png}
-\caption{MCMC draws for the AR(2) model with $T = 150, \phi_1 = 0.75, \phi_2 = 0.2$ and $\sigma^2 = 1$.
-Approximately 24\% of both $\phi$ parameters were accepted by the Metropolis Hastings step, then the first 10,000 draws were discarded and the remainder thinned to 1 in 200 to make dependence between draws negligible.}
-\label{MCMCplot}
-\end{figure}
-
-\subsection{Variational Bayes}
-
-A faster alternative to MCMC is Variational Bayes, which introduces an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ choosing the functional form $q$ and given auxillary parameters $\boldsymbol{\lambda}$ so that the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
+As discussed in Section \ref{sec:Intro}, a faster (albeit approximate) alternative to MCMC is VB.
+This approach introduces an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ with the choice of the functional form $q$ and subsequent auxiliary parameter vector $\boldsymbol{\lambda}$ determined so that the Kullback-Leibler (KL) divergence \citep{Kullback1951} from$q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
 The KL divergence is defined by
-
 \begin{equation}
 \label{KL-def}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T}] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T}] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
 \end{equation}
-
-The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that will equal zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}.
-Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
-
+The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$  that will be equal to zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}.
+Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
 \begin{equation}
 \label{KL-ELBO}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})||p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q, y_{1:T})
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q, y_{1:T})
 \end{equation}
-
-where $\mathcal{L}(q, y_{1:T})$ is known as the Evidence Lower Bound (ELBO), defined by
-
+where $\mathcal{L}(q, y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q, y_{1:T})$ is defined by
 \begin{equation}
@@
 Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}).
-Maximising the ELBO is much more convenient than minimising the KL Divergence, which can be seen in the following sections based on the two major implementations of Varitaional Bayes: Mean Field Variational Bayes and Stochastic Variational Bayes.
-Each of these take advantage of the functional form of the ELBO and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.
+Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Varitaional Bayes, namely Mean Field Variational Bayes and Stochastic Variational Bayes.
+Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.
+of distributions.
 
-\subsubsection{Mean Field Variational Bayes}
+\subsubsection{Mean Field Variational Bayes} \label{sec:mfvb}
 
-Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the class of distributions Variational Bayes selects from to the set of factorisable distributions,
-
+Mean Field Variational Bayes (MFVB) has origins in physics \citep{Chandler1987} and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Beal2003}.
+MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
 \begin{equation}
 \label{mf1}
-q(\theta|\lambda) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
+q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
 \end{equation}
@@
 
-MFVB is widely used as it greatly simplifies maximisation of the ELBO, particularly in exponential family models (\citealp{Jordan1999}; \citealp{Ghahramani2000}, \citealp{Wainwright2008}).
-Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the computation is simple.
-MFVB expresses (\ref{ELBO}) as a function of one $q_i$ only and then maximises with respect to that $q_i$.
-The final algorithm requires this to be repeated for each $i$ in $1, \dots, k$.
-Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that
-
+MFVB is widely used as it greatly simplifies maximisation of the ELBO, and has a straight-forward implementation in exponential family models \citep{Wainwright2008}.
+Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme.
+MFVB factorises (\ref{ELBO}) into a product of $k$ seperate components that each contain only one $q_i$, and then applies an algorithm to maximise each of these factors with respect the associated $q_i$.
+Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
 \begin{equation}
 \label{mf2}
-q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))])
+q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
 \end{equation}
 
-For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated from a $\mathcal{N}(\mu, \sigma^2)$ distribution with priors $p(\mu) \sim \mathcal{N}(\gamma, \tau)$ and $p(\sigma^2) \sim Inv.Gamma(\alpha, \beta)$.
+We illustrate the implementation of MFVB with the model considered in Section (\ref{sec:MCMC}), where $y_t$ for $t = 1, \dots, T$ is generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / \tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$.
 It can be shown that
-
 \begin{equation}
@@
 \end{equation}
-
 and
-
 \begin{equation}
@@
 
-It is evident that $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$.
-The parameters $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ can be found by the set of mean field equations:
+As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q(\mu |  \boldsymbol{\lambda}_1)$ and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ are the same family as the prior distributions, and are denoted by $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$.
+The vector $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ is given by
+\begin{align}
+\tilde{\gamma} &= \frac{\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\gamma + \lambda \sum_{t=1}^{T} y_t}{T \tau + \mathbb{E}_{q(\sigma^2 )}(\sigma^{-2})}  \label{mf5} \\ 
+\tilde{\tau} &= \frac{\tau\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{(T\tau + \mathbb{E}_{q(\sigma^2)}(\sigma^{-2}))} \label{mf6} \\
+\tilde{\alpha} &= T/2 + \alpha  \label{mf7} \\
+\tilde{\beta} &= 1/2\left(\sum_{t=1}^{T}y_t^2 + n\mathbb{E}_{q(\mu)}(\mu^2) - 2\sum_{t=1}^{T} y_t \mathbb{E}_{q(\mu)}(\mu))\right) + \beta. \label{mf8}
+\end{align}.
+
+The known values of the expectations in (\ref{mf5}) - (\ref{mf8}) for these distributions can be substituted into these expressions to derive the set of mean field equations:
 
 \begin{align}
-\tilde{\gamma} &= \frac{\tilde{\alpha} / \tilde{\beta} \gamma + \tau \sum_{t=1}^{T} y_t} {T \lambda + \tilde{\alpha} / \tilde{\beta}} \label{mf5} \\ 
-\tilde{\tau} &= \frac{T \tilde{\alpha} / \tilde{\beta}}{T \tau + \tilde{\alpha} / \tilde{\beta}} \label{mf6} \\
-\tilde{\alpha} &= \frac{T}{2} + \alpha  \label{mf7} \\
-\tilde{\beta} &= \frac{1}{2} \left(\sum_{t=1}^{T} y_t^2 + T(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \sum_{t=1}^{T} y_t \tilde{\gamma} \right) + \beta. \label{mf8}
-\end{align}.
+\tilde{\gamma} &= \frac{\tilde{\alpha} / \tilde{\beta} \gamma + \tau \sum_{t=1}^{T} y_t} {T \lambda + \tilde{\alpha} / \tilde{\beta}} \label{mf9} \\ 
+\tilde{\tau} &= \frac{T \tilde{\alpha} / \tilde{\beta}}{T \tau + \tilde{\alpha} / \tilde{\beta}} \label{mf10} \\
+\tilde{\beta} &= \frac{1}{2} \left(\sum_{t=1}^{T} y_t^2 + T(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \sum_{t=1}^{T} y_t \tilde{\gamma} \right) + \beta. \label{mf11}
+\end{align}
+$\tilde{\alpha}$ is omitted from the mean field equations as its value does not depend on other $\boldsymbol{\lambda}$ parameters.
+(\ref{mf9}) - (\ref{mf11}) contain a circular dependence in these equations an algorithm that iteratively cycles through these equations until the change in the ELBO is less than some pre-specified constant $\epsilon$.
+This algorithm is known as a coordiant ascent algorithm and is described by Algorithm (\ref{alg:algorithm1}), at each step it will fix all but one parameter and select the value of the omitted parameter which maximises the ELBO.
 
-As there is a circular dependence in these equations an algorithm that cycles through each auxillary parameter until convergence to some pre-specified threshold is required.
+The set of mean field equations are only obtainable in the event that a prior that is conjugate to the model likelihood is used, restricting MFVB to exponential family models.
+In this case the optimal distributional family $q_i$ is the same as the prior, analagous to the relationship between a prior-posterior conjugate pair.
+In the event that a non-conjugate prior is used, the form of the distribution in (\ref{mf2}) is unrecognisable, a further approximation may be used by substituting in another distribution to replace the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006}, a Delta Method inspired transform \citet{Wang2013}, though the only requirement is that the substitute distribution has moments that are tractable functions of its parameters to be used for other parameters in (\ref{mf2}).
+This secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable full conditional distributions.
 
-In the event that (\ref{mf2}) is from an unrecognisable distribution a further approximation can be used, substituting in another distribution for the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ \citet{Friston2006} and \citet{Wang2013} suggests the use of a Gaussian distribution through either a Laplace approximation or a Delta Method transform, though the only requirement is that the substitute distribution has an expectation that is a tractable function of its parameters to be used for other parameters in (\ref{mf2}).
-
-Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | \boldsymbol{\theta}_{j \neq i}, y_{1:T})$ in a Gibbs MCMC scheme, with dependence on other parameters replaced by their expectations.
-The optimal approximating family for $\theta_i$, $q_i(\theta_i |\boldsymbol{\lambda_i})$, will come from the same distributional family as the conditional distribution $p(\theta_i | \boldsymbol{\theta}_{j \neq i}, y_{1:T})$, if it exists in a recognisable form such as the exponential family.
-The secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable distributions.
 \vspace{5mm}
-
-The required algorithm is known as a coordinate ascent algorithm, and is described below for the model discussed above
-
-\vspace{2mm}
-
 \begin{algorithm}[H]
@@
  Use (\ref{mf3}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
- Anaytically derive the set of mean field equations in \ref{mf5}) - (\ref{mf8})\;
+ Anaytically derive the set of mean field equations in \ref{mf9}) - (\ref{mf11})\;
  Initialise $\boldsymbol{\lambda}^{(1)}$ randomly.\;
@@
  \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
-  Set $\tilde{\gamma}^{(m)}$ using \ref{mf5}.\;
-  Set $\tilde{\tau}^{(m)}$ using \ref{mf6}.\;
-  Set $\tilde{\beta}^{(m)}$ using \ref{mf8}.\;
+  Set $\tilde{\gamma}^{(m)}$ using \ref{mf9}.\;
+  Set $\tilde{\tau}^{(m)}$ using \ref{mf10}.\;
+  Set $\tilde{\beta}^{(m)}$ using \ref{m11}.\;
   Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$.\;
@@
 
-Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure \ref{MCMCplot} shows strong dependence between $\phi_1$ and $\phi_2$.
+The requirement for an exponential family model, as well as restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure \ref{fig:MCMCplot} shows strong dependence between $\phi_1$ and $\phi_2$.
 Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable.
@@
 
+To illustrate issues involved, consider {\bf the second order} auto-regressive time series {model, denoted by AR(2) and described by}
+% no need to say more than this here.
+% But actually, I think you should just do the very simple Normal Inverse Gamma model here, since that has the parallel with VB below.
+Then in your example later on, you can do the full AR(2) example.
+I think this will help a lot.
+
+\begin{equation}
+\label{AR2}
+y_t = \rho_1 y_{t-1} + \rho_2 y_{t-2} + \epsilon_t
+\end{equation}
+
+where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$.
+The likelihood of parameters {\bf $\theta = (\rho_1, \rho_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\theta$, viewed as a function of $\theta$,)} is given by
+
+\begin{align}
+\label{likelihood}
+L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
+&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
+\end{align}
+
+where
+
+\begin{equation}
+\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
+\end{equation}
+% Changed the matrices above to align properly
+$\gamma(k)$ is the $k-th$ autocorrelation and $\textbf{y} = (y_1, y_2)'$.
+% You also need to mention what $p(y_1, y_2 | \theta)$ is, and perhaps something about how $p(y_t | y_{1:t-1}, \theta)$ is available from the assumptions regarding the distribution of $\epsilon_t$.
+% Also, I don't like the use of vector notation (for $y=(y_1,y_2)$ )and subscripts for the remaining $y$'s in the likelihood. You can just break the equation over two lines if needed. 
+% Finally, perhaps you could say something about where $\gamma_0$ and $\gamma_1$ come from, or at least provide a reference for the reader to go and find out more about this model so you don't have to go into too much detail here.
+
+{\bf We assume that the prior distribution, is given by }
+% remember Tas's comment about stating that the joint prior is assumed to be the product of independent marginal priors
+
+\begin{align}
+p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
+p(\rho_1, \rho_2) &\propto \mathbb{I}(\rho_2 > -1)\mathbb{I}(\rho_2 < 1 + \rho_1) \mathbb{I}(\rho_2 < 1 - \rho_1) \nonumber
+\end{align}
+
+are used, where $\mathbb{I}$ is the indicator and $p(\rho_1, \rho_2)$ defines the AR(2) stationary region.
+
+{\bf Upon examination of the form of the product of the likelihood function in (\ref{likelihood}) and the prior density in (),}
+	%need to be able to reference the prior too
+{\bf it can be seen that the full} conditional distribution $p(\sigma^2 | y, \phi_1, \phi_2)$ is {\bf an} Inverse Gamma {\bf distribution},
+% can you say which one? Be sure it is clear how to interpret the parameters of this distribution too.
+ while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step 
+% Since you are actually using an M-H step, you should have told the reader what this is back up when you introduced the Gibbs sampler.
+with truncated {\bf bivariate} %?
+normal random walk {proposal distribution}. The truncation ensures that draws {\bf of $(\rho_1,\rho_2)$} %just to be clear
+are only taken from the AR(2) stationary region.
+
+%Any graph that you put in to the document needs to be clearly explained.
+Why discard any?
+What does 'thinned' mean, and why did you do it?
+etc....
+Better to put the details of the MCMC in a paragraph, and then point to the figure, with only the essential details left in the caption.
+% Also, ideally you would be able to put the Greek symbols in the margins of the graphs, rather than 'Sigmasquared', for example...
+
+\begin{figure}[h]
+\centering
+\includegraphics[scale = 0.6]{MCMC.png}
+\caption{MCMC draws for the AR(2) model with $T = 150, \phi_1 = 0.75, \phi_2 = 0.2$ and $\sigma^2 = 1$.
+Approximately 24\% of both $\phi$ parameters were accepted by the Metropolis Hastings step, then the first 10,000 draws were discarded and the remainder thinned to 1 in 200 to make dependence between draws negligible.}
+\label{fig:MCMCplot}
+\end{figure}
+
 Using the approach described in section 2.3, an application of VC algorirhtm to the MCMC draws obtained in section 2.1 found the following distribution to be optimal by AIC.
@@
 
-Figure \ref{VBfit} shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.
+Figure \ref{fig:VBfit} shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.
 
@@
 SVB resulted in an almost identical posterior to MCMC, requiring around one hundredth of the runtime for MCMC.}
-\label{VBfit}
+\label{fig:VBfit}
 \end{figure}
 
-\section{Electricity Load Forecasts}
+\section{Electricity Load Forecasts} \label{sec:ElecLoad}
 \subsection{Motivation}
@@
 
-Half hourly electricity load data for Victoria, Australia has been collected for the 2011-2015 period, and is plotted in Figure \ref{loadplot}.
+Half hourly electricity load data for Victoria, Australia has been collected for the 2011-2015 period, and is plotted in Figure \ref{fig:loadplot}.
 The time series displays strong seasonal characteristics, with yearly, day of week, and time of day patterns.
@@
 The electricity supply curve is strongly hockey stick shaped, as renewables and coal generators are able to provide a large load cheaply but if demand exceeds this level the price spikes as other generators must be brought online expensively.
-These characteristics of the electricity market leads to a large amount of volatility in prices demonstrated by Figure \ref{rrpplot}.
+These characteristics of the electricity market leads to a large amount of volatility in prices demonstrated by Figure \ref{fig:rrpplot}.
 The seasonal effects of the load data are also present in prices, but these effects are dominated by irregular spikes as high as \$9,974/mWh, compared to a four year mean of \$42/mWh.
@@
 The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent curve in winter.}
-\label{loadplot}
+\label{fig:loadplot}
 \end{figure}
@@
 While the price has time of week and time of day seasonal effects these features are dominated by irregular spikes.}
-\label{rrpplot}
+\label{fig:rrpplot}
 \end{figure}
@@
 
-\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$ in (\ref{exp-sm_marginal}), however the problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large, as is the case with high frequency electricity load data.
+\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$ in (\ref{exp-sm-marginal}), however the problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large, as is the case with high frequency electricity load data.
 An efficient MCMC algorithm using the conditional distribution of $p(\theta | y_{1:T}, \sigma^2, b_0)$ may be required to sample the full posterior distribution $p(\theta, \sigma^2, b_0 | y_{1:T}$ and obtain an approximating distribution $q(\theta, \sigma^2, b_0 | \lambda)$ with a reasonable goodness of fit.
@@
 
-\section{Discussion}
+\section{Discussion} \label{sec:Discussion}
 
@@
 
-\section{Timeline}
+\section{Timeline} \label{sec:Timeline}
 

### 2017-02-21 Confirmation/confirmation_V2_CF.tex
Revised draft

\documentclass[12pt,a4paper]{article}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\title{Real Time Variational Density Forecasts}
\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
\begin{document}

\maketitle
\tableofcontents
\section{Introduction} \label{sec:Intro}
%label added to section for referencing later

Electricity prices have been subject to an extensive forecasting literature and is known to be extremely volatile. For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of {\bf observations} less than \$100 {\bf per} % Try to minimize or not use shorthand mathematical notation in the introduction. 
mWh but on 39 {\bf occasions} the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29, 2012. 
% This is very precise information, but I think for an introduction it is better to try to keep it a bit more `big picture'. Perhaps even just writing in percentage terms will have more impact, rather than requiring the reader to work out if this difference is big or not. Then also you wouldn't need to units of measure in quite so many places. I'd even prefer not to have the abbreviation `mWh' here, but this is less critical. } 
For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period is more important than the mean value. To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it {\bf accounts for both } parameter and model uncertainty {\bf often} ignored {\bf by} frequentist methods. 
% I agree with Tas that it is better not to be so defensive at the outset. Better to justify that positives of Bayesian methods - in particular the use of prior information and the ability to handle complex, parameter-driven (state space) models. 
See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general. 
% Sorry but I don't know why the references are not working for me!

% It might be useful to say something in the above about the connection between the predictive distribution, and the posterior distribution. Or just talk about distributions here, and then the connection to the forecast density can come in the next section.

Bayesian methods are computationally complex as they require integration over the set of {\bf unknowns, whether parameters or latent stochastic variables. This set is typically quite large for models used to model short term electricity load and price as there are} strong seasonal patterns {\bf in the data \citep{Taylor2003}.} {\bf While analytical integration is typically not feasible to obtain the predictive distribution for such models, however} there are a range of {\bf numerical and simulation-based} techniques available, {\bf for this purpose,} notably Markov Chain Monte Carlo (MCMC) methods. While {\bf it is often possible to demonstrate theoretical convergence of MCMC algorithms,} computation {\bf may be} slow for even moderately complex models. {\bf This} is problematic {\bf for when attempting to forecast} Electricity demand 
%demand? or price? or load? decide which you are going to talk about here, and only talk about that for now - check throughout.
in Victoria, {when realizations occur at} five minute {\bf intervals}. The predictive density for the next period's electricity load may not be available before it is observed.

Alternatives to MCMC include Variational Bayes (VB) {\bf methods, where the aim is} to replace the unknown true posterior distribution for a vector of {\bf unknowns}
% you can reduce to just $\theta$ below - but keep it general in the introduction
with a parametric approximation. {\bf The} Variational Bayes {\bf approach} will be explored {\bf in the thesis, to determine if it is a viable} alternative computation strategy to MCMC,
% comma added
in the context of {\bf modelling and short term forecasting of Victorian electricity load.}
% Unless you have something interesting to say about the time period - which could come later - there isn't any real need to mention the data period at this point. It could come in the last paragraph, however. 
{\bf In particular, we seek to understand the benefits gained in terms of the computational time required to obtain a VB approximate forecast distribution against the loss in statistical} 
% I think you need to use the spell checker!!
accuracy associated with the use of an {approximate posterior forecast distribution.} 
%

% You need a paragraph here detailing the outline of the rest of the paper, section by section.

\section{Bayesian Inference}

{\bf To facilitate the discussion, consideration is given to models where the unknowns may be summarized in a (finite) $k-$dimensional vector, denoted by $\theta.$ In this setting, given an observed time series, denoted by $y_{1:T}= \{y_{t}, t = 1, \dots T\}$, the Bayesian forecast distribution associated with time $t=T+1$ is characterized by the conditional density }
\begin{equation}
\label{predictive}
p(y_{T+1} | y_{1:T}) =\int p(y_{T+1}|y_{1:T},\theta) p(\theta|y_{1:T}) d\theta,
\end{equation}
{\bf where $p(\theta|y_{1:T})$ denotes the posterior density for $\theta$, given by}

\begin{equation}
\label{posterior}
 p(\theta | y_{1:T}) = \frac{p(y_{1:T}|\theta)p(\theta)}{\int_\theta p(y_{1:T}|\theta)p(\theta) d\theta}.
\end{equation}
% You can feel free to modify the above, but be sure to define all notation that you use.

% I think you need to expand this section a bit more - think about what you really need to say, and why.
Generally the {\bf analytical} solution to (\ref{posterior}), 
% comma added
{\bf and hence to (\ref{predictive}) will be unavailable.} 	% Can you say why? 
{\bf In this} section {\bf two alternative methods for computing the desired posterior distribution in (\ref{posterior}),  MCMC and VB, will be reviewed.}
% Since neither method is new, you are not `offering' them, just reviewing them.
% Can you extend this discussion to explain how to use MCMC and VB for (\ref{\predictive})?
{\bf In brief,} MCMC is used to create a sample from $p(\theta | y)$, {\bf with} any function {\bf of $\theta$ that is desired estimated from that sample. In contrast,} VB replaces $p(\theta | y)$ with an approximation,
%comma inserted
{\bf denoted by} $q(\theta | \lambda)$, where $\lambda$ is a vector of {\bf auxiliary} parameters associated with the approximation and not the model. 
% surely these $\lambda$ have something to do with the model??? Perhaps end the sentence with `approximation'. Also need to emphasize that $q(\theta|\lambda)$ is also a function of the data $y_{1:T}$.
For example, if $q(\theta | \lambda)$ is Gaussian,
%comma added
{\bf then} $\lambda$ would {\bf denote} the mean and variance {\bf parameters that characeterize this normal distribution for} $\theta$. Variational Bayes aims to choose the distribution $q(\theta | \lambda)$ such that the Kullback-Leibler divergence from $q(\theta | \lambda)$ to $p(\theta | y)$ is minimised.
% I have used `z' (as in `minimized') in my edits, not `s' (as in `minimised'), but you decide which you prefer and be consistent.
%So obviously now the values of the mean and variance will in fact depend on the data, and the Gaussian model will be selected due following consideration of at least a portion of the original, more complex model (via this minimization). So yes, the approximation and its parameters will, in some sense at least, depend on the model!

\subsection{Markov Chain Monte Carlo}

There are many types of MCMC {\bf algorithms, with arguably the simplest and most commonly used one being the} Gibbs sampler. {The Gibbs sampler algorithm} iteratively samples {\bf the components of the} $k$ dimensional parameter vector $\theta$ via {\bf each of} the {\bf so-called full} conditional distributions {\bf as follows,}
% I'm not sure you need the extra space above each equation - have not changed elsewhere but please consider
\begin{align}
&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
% when you put a space after the equation, LaTeX thinks you are starting a new paragraph. Depending on the style you are using, this may generate a tab at the start of the next line, which you may not want - please be mindful of this, I have not corrected throughout.
{\bf Under mild regularity conditions (see, e.g., Tierney, 1994, Annals of Statistics) and with} enough iterations {\bf of the Markov chain that results from the Gibbs sampler,} these samples converge in distribution to $p(\theta | y)$ and can be used for inference. 
% The phrase `and can be used for inference' is a bit vague...
The computation time {\bf for each} iteration and {\bf the overall} number of iterations required for {\bf to accruately summarize the posterior distribution} is problem specific but typically increases with model complexity. 
% what do we know about model complexity? Can we just say it is $k$? Or do you want to highlight something more?

\subsection{Variational Bayes}

{\bf As discussed in Section \ref{sec:Intro}, a faster (albeit approximate)} alternative to MCMC is {\bf VB. This approach} introduces an approximating distribution {\bf $q(\theta | \lambda),$ with the choice of the functional form $q$ and subsequent auxiliary parameter $\lambda$ determined so that} the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ the true posterior $p(\theta | y)$ {\bf is minimised.} The KL divergence is defined by

\begin{equation}
\label{KL-def}
KL[q(\theta | \lambda) \hspace{.1cm}||\hspace{.1cm}p(\theta | y)] = \int q(\theta | \lambda) \ln \left( \frac{q(\theta | \lambda)}{p(\theta | y)}\right) d\theta.
\end{equation}
% no new para. Note it might be useful to add some space before and after the double vertical lines - to make the equation above easier to read. I have done but see what you think...I haven't changed elsewhere.
The KL divergence is a non-negative, asymetric %spelling
 measure of the discrepancy between $p(\theta | y)$ and $q(\theta | \lambda)$ that will {\bf be equal to} zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}. Note that $KL[q(\theta | \lambda)||p(\theta | y)]$ can be expressed as

\begin{equation}
\label{KL-ELBO}
KL[q(\theta | \lambda)||p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q, y)
\end{equation}

where $\mathcal{L}(q, y)$ is {\bf referred to} as the Evidence Lower Bound (ELBO), {\bf and is} defined by
% Can you give some intuition for why this is called the ELBO?
\begin{equation}
\label{ELBO}
\mathcal{L}(q, y) = \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta.
\end{equation}

Since $\ln(p(y))$ is constant given $y$, maximising (\ref{KL-ELBO}) with respect to $q$ is equivalent to minimising (\ref{KL-def}). Maximising the ELBO is much more convenient than minimising the KL Divergence, {\bf as will} be shown in the following sections {\bf in the context of the} two major implementations of Varitaional %spelling
{\bf Bayes, namely} Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these {\bf implementations} take advantage of the functional form of the ELBO, %comma inserted
and offer computationally efficient algorithms to find a $q(\theta | \lambda)$ that is optimal within a particular clas %spelling
of distributions.

\subsubsection{Mean Field Variational Bayes}

Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the class of distributions {\bf from which to select $q(\theta|\lambda)$} to the set of factorisable distributions,

\begin{equation}
\label{mf1}
q(\theta|\lambda) = \prod_i^k q_i(\theta_i | \lambda_i).
\end{equation}
% Again, Tas's comment here about not re-using $k$ for this
Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars. Each $\theta_i$ has an associated vector $\lambda_i$ which may be of a higher dimension than $\theta_i$. $\lambda_i$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\lambda_i)$.

MFVB is widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}), particularly in exponential family models \citep{Wainwright2008}.  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the computation is simple. 
%It is not clear at this point what you mean by 'the computation'. Is it that the identification process is the same, but rather than sampling from the identified distribution, as in Gibbs sampling, here you will find the value of the parameter $\lambda_i$ to maximise the identified distribution (for fixed $y_{1:T}$ and fixed other parameters in $\theta$ other than $\theta_i$).
MFVB expresses (\ref{ELBO}) as a function of one $q_i$ only 
% This doesn't quite make sense to me. It must be that MFVB assumes that (\ref{ELBO}) can first be represented as some kind of product of similar expressions, one for each $i$, and then that the maximisation of (\ref{ELBO}) can be determined by maximising each of these expressions. We talked about this a bit last week - perhaps you can tease this out a bit more. Somehow the iteration comes in so as to achieve this maximum.
The final algorithm requires this {\bf process} to be repeated for each $i$ in $1, \dots, k$, {\bf until ...}.
%Some kind of convergence criterion is needed.
Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that

\begin{equation}
\label{mf2}
q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])
\end{equation}
% The sentence is missing a full stop, and also I'm not sure how it connects to everything else?
{\bf To illustrate the MFVB approach,} consider data $y_i$ for $i = 1, \dots, n$ generated {\bf independently} from a $\mathcal{N}(\mu, \sigma^2)$ distribution with {\bf independent marginal prior distributions given by} %It is $\mu$ that has a distribution, not $p(\mu)$ - changed here
$\mu \sim \mathcal{N}(\gamma, \lambda)$ and $\sigma^2 \sim Inv.Gamma(\alpha, \beta)$. It can be shown that
% I think you need some $\lambdas$ below? Also would be worthwhile to  explain or show why these are the correct MFVB distributions...or at least be able to say something about it during your presentation.

\begin{equation}
\label{mf3}
q(\mu) \propto \exp \left\{ \frac{-(n\lambda + \mathbb{E}(\sigma^{-2}))}{2\lambda\mathbb{E}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}(\sigma^{-2})\gamma + \lambda \sum_{i=1}^{n} y_i}{n \lambda + \mathbb{E}(\sigma^{-2})} \right)^2 \right\}
\end{equation}

and

\begin{equation}
\label{mf4}
q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + n\mathbb{E}(\mu^2) - 2\sum_{i=1}^{n} y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
\end{equation}

It is evident that $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ and $q(\sigma^2 | \lambda) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$. The parameters $\lambda = (\tilde{\gamma}, \tilde{\lambda}, \tilde{\alpha}, \tilde{\beta})'$ can be found 
% how? I think you are just looking at the equations and `identifying' these guys, but it would be usefult to explain this. At least you have to mention something about how to evaluation, e.g. $E(\sigma^{-2})$ in $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ to get that $\tilde{\gamma} = \frac{\alpha / \beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta}$. Same for the other parameters too..
by the set of mean field equations:

\begin{align}
\tilde{\gamma} &= \frac{\alpha / \beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta} \label{mf5} \\ 
\tilde{\lambda} &= \frac{n \alpha / \beta}{n \lambda + \alpha / \beta} \\
\tilde{\alpha} &= \frac{n}{2} + \alpha \\
\tilde{\beta} &= \frac{1}{2} \left(\sum_{i=1}^{n} y_i^2 + n(\tilde{\gamma}^2 + \tilde{\lambda}) - 2 \sum_{i=1}^{n} y_i \tilde{\gamma} \right) + \beta. \label{mf6}
\end{align}.

As there is a circular dependence in these equations an algorithm that cycles through each auxillary %spelling
parameter until convergence to some pre-specified threshold %need to give some definition of this
is required.

%Paragraph moved below

Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | y, \theta_{\setminus i}),$ %parenthesis and comma added, but I think this is different notation to what you used before - needs to be the same!
{\bf for example} in a Gibbs {\bf sampling} MCMC scheme, with dependence on other parameters replaced by their expectations. The optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$. % full stop added. Isn't this true, regardless of whether we can recognize the form or not? 
if it exists in a recognisable form such as the exponential family. % have left this for you to resolve...

% moved paragraph to here
In the event that {\bf the form of the distribution in} (\ref{mf2}) is {\bf unrecognisable,} a further approximation {\bf may} be used {\bf by} substituting in another distribution {\bf to replace} the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One {\bf approach, given by \citet{Friston2006},} uses a Laplace {\bf approximation, substituting} a Gaussian distribution for {\bf the} unrecognizable $q_i(\theta_i | \lambda_i)$. {\bf The} only requirement {\bf for this approach} is that the substitute {\bf have} an expectation 
% expectations? don't we need both of the first two moments? 
that is a tractable function 
% that are tractable functions? 
of {\bf the} parameters to be used for other parameters in (\ref{mf2}). 
{\bf This} secondary level of {\bf approximation} is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable distributions. % I really like this analogy, and it will help your audience understand what is going on. But to make it work, you should have explained MHwithinGibbs earlier.

% Moved revised sentence down to here. But now you'll need to work this in with the following paragraph!
In the event that {\bf the form of the distribution in} (\ref{mf2}) is {\bf unrecognisable,} a further approximation {\bf may} be used {\bf by} substituting in another distribution {\bf to replace} the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One {\bf approach, given by \citet{Friston2006},} uses a Laplace {\bf approximation, substituting} a Gaussian distribution for {\bf the} unrecognizable $q_i(\theta_i | \lambda_i)$. {\bf The} only requirement {\bf for this approach} is that the substitute {\bf have} an expectation 
% expectations? don't we need both of the first two moments? 
that is a tractable function 
% that are tractable functions? 
of {\bf the} parameters to be used for other parameters in (\ref{mf2}).

\vspace{5mm}

% You have not really fully explained how this algorithm below fits with what is above?? A single sentence might be sufficient.
{\bf Once the approximation has been identified, the algorithm required to find the ELBO is} known as a coordinate ascent algorithm. %full stop
{\bf This algorithm} is described below {\bf in the case} where $\lambda$ is a $p$ dimensional vector. %why change from $k$ to $p$? need to keep consistent notation throughout. 
%We have discussed these algorithms already, and I have nothing more to add now

\vspace{2mm}

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density}
 \KwResult{Mean Field Approximation}
 Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution or substitute a further approximation if neccesary.\;
 Derive the set of mean field equations, such as (\ref{mf5}) - (\ref{mf6}).\;
 Initialise $\lambda$ randomly. \;
 \While{Not converged}{
  \For{$i =  1$ \KwTo $k$}{
      Hold $\lambda_{j \neq i}$ fixed\;
      Update $\lambda_i$ using the relevant mean field equation.
     }
 }
 \caption{Coordinate Ascent for MFVB}
  \label{alg:algorithm1}
\end{algorithm}

\subsubsection{Stochastic Variational Bayes}

Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure \ref{MCMCplot} shows strong dependence between $\phi_1$ and $\phi_2$. Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable. \citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, resulting in what is known as Stochastic Variational Bayes (SVB).

The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\theta | \lambda), y)$ in (\ref{ELBO}) with respect to $\lambda$ and the following updating step is applied until the ELBO converges to a maximum.

\begin{equation}
\label{SGA1}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m)}), y),
\end{equation}

where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ with respect to each element of $\lambda$. This update requires some initial values $\lambda^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.

\begin{align}
\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \\
\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty.
\end{align}

Whilst a global maximum is desired, the ELBO is a high dimensional problem specific function that makes finding the global maximum extremely difficult and often only a local maximum can be found. One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior

SVB can find only the optimal values for $\lambda$ that are optimal for a distributional family $q$ that is specified by the user. To run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO. $q(\theta | \lambda)$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable, and SVB can only be applied to models where the log likelihood for $y$ is able to be evaluated. Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by

\begin{equation}
\label{SGA2}
\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda^{(m)}))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda^{(m)})))
\end{equation}

where $s = 1, \dots, S$ indicates simulations from $q(\theta | \lambda^{(m)})$.

\citet{Duchi2011} introduced the AdaGrad algorithm which can be implemented within SVB to control $\rho^{(m)}$. AdaGrad allows each $\lambda_i$ to have an independent $\rho^{(m)}_i$ that is inversely proportional to the gradient, so $\lambda$ takes bigger steps in flat regions and smaller steps in steep regions.

Let

\begin{equation}
\label{SGA3}
G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\theta | \lambda^{(m)})\right)^2,
\end{equation}

then each component's learning rate is defined as

\begin{equation}
\label{SGA4}
\rho^{(m)}_i = \eta \left(G_i^{(m)}\right)^{-1/2}
\end{equation}

for some tuning parameter $\eta$.

The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\lambda$ vector.

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Log Joint Density, Approximation family q}
 \KwResult{Variational Approximation}
 Initialise $\lambda$\;
 \While{Not converged}{
  Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m)})$\;
  \For{$i =  1$ \KwTo $p$}{
      Calculate $\nabla_{\lambda_i}$ from (\ref{SGA2})\;
      Update $G_i^{(m)}$ and $\rho^{(m)}_i$ from (\ref{SGA3}) and (\ref{SGA4})\;
      }
  Update $\lambda^{(m+1)}$ from (\ref{SGA1})\;
  Set $m = m + 1$\;
 }
 \caption{Stochastic Gradient Ascent for SVB}
  \label{alg:algorithm2}
\end{algorithm}

\subsection{Vine Copulas}
% I won't add too many more to Tas's comments, except to remind you to take the sorts of critiques I have made above and try to apply them here - be clear about what you are saying, don't leave out so many words and use that spell checker :-)

Armed with an algorithm to find the Variational Bayes optimal paramerters $\lambda$ for a distribution $q$, there is a requirement to have a distribution $q$ that is itself optimal. Copulas are a flexible tool for constructing these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions. \citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,

\begin{equation}
\label{vc1}
p(\theta_1, \dots, \theta_k) = p(\theta_1) \dots p(\theta_k) c(P(\theta_1), \dots, P(\theta_k))
\end{equation}

where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.

\citet{Joe1994} introduces the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas in a similar way to the factorisation of a joint distribution into a marginal and conditional distribution, for example

\begin{equation}
\label{vc2}
c(P(\theta_1), P(\theta_2), P(\theta_3)) = c(P(\theta_1), P(\theta_2)) \cdot c(P(\theta_1), P(\theta_3)) \cdot c(P(\theta_2), P(\theta_3) | P(\theta_1)).
\end{equation}

A copula for a $k$ dimensional $\theta$ can be replaced with $k(k-1)/2$ bivariate copulas arranged in a vine copula, greatly increasing the amount of flexibility available to fit a distribution. However there are $O(k!)$ ways to factorise the full copula and choose a family for each bivariate copula, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult. \citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\theta$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.

Given a sample of $\theta$, \citet{Diimann2013} provide an algorithm to select the best vine factorisation by maximising Kendell's Tau, and the best family for each bivariate copula by maximising an information criterion such as AIC. In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date. The approach used in this thesis is to run an MCMC algorithm, thin the resulting draws until they are effectively independent, choose an optimal family for each marginal distribution by AIC, then run (Vine copula algorithm here) to infer an optimal vine copula. This results in an approximating distribution $q(\theta | \lambda)$ that can then be used in an SVB algorithm to update $q(\theta | \lambda)$ as more observations become available. MCMC can be ran simultaenously and the resulting sample can be used to check if the distribution $q$ should be changed.

\subsection{Applying Stochastic Variational Bayes}

Using the approach described in {\bf Section} 2.3,
% When referring to a specific section, use a capital letter. When referring generally to `this section' or `later sections', etc, use lower case.
an application of VC algorirhtm to the MCMC draws obtained in {\bf Section} 2.1 % You can check the rest
found the following distribution to be optimal, {\bf according to the standard Akaike Information Criterion (AIC)}.

\begin{itemize}
\item $q(\phi_1)$ - Gaussian marginal
\item $q(\phi_2)$ - Gaussian marginal
\item $q(\sigma^2)$ - Inverse Gamma marginal
\item $c(Q(\phi_1), Q(\phi_2))$ - Gaussian copula
\item $c(Q(\phi_1), Q(\sigma^2))$ - Independent
\item $c(Q(\phi_2), Q(\sigma^2) | Q(\phi_1))$ - Independent
\end{itemize}

Figure \ref{VBfit} shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.

\begin{figure}[h]
\centering
\includegraphics[scale = 0.5]{VBfit.png}
\caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model described in section 2.1. SVB resulted in an almost identical posterior to MCMC, requiring around one hundredth of the runtime for MCMC.}
\label{VBfit}
\end{figure}

\section{Electricity Load Forecasts}
\subsection{Motivation}

The literature for electricity forecasting is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review. Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016} 
% one of the disadvantages of the automatic citations is that the formatting isn't quite as I would expect. Perhaps a different bibliographical style would be better. In Econometrics, we typically list the full set of authors names the first time a paper having 3 (sometimes 4) or more authors, and then if cited again use the `et al.' convention. Also, the `et al.' should be in italics. But this is minor for now...
being notable examples, which used boot strapping and quantile regression respectively to generate forecast densities. Both of these approaches create forecast densities that are conditioned on the model and estimated parameters being correct, while the Bayesian approach used in this thesis can average over parameter and model uncertainty leading to a more accurate forecast density. This thesis also aims to re-estimate the model after each data point is observed to fully take advantage of any extra information on changes in parameters and thus further increase the accuracy of the forecast density.

Half hourly electricity load data for Victoria, Australia has been collected % You need to report where you got the data from. You didn't collect it yourself, which is how it reads!
 for the 2011-2015 period, and is plotted in Figure \ref{loadplot}. The time series displays strong seasonal characteristics, with a yearly, day of week, and time of day pattern. Electricity load is strongly dependent on temperature, with high load experienced on both very cold and very hot days. In summer periods, load is often low but shows extreme volatility, while load slowly rises then falls in winter and displays significantly less volatility.

The electricity market in Victoria operates by market participants sending electricity bid and offers for each five minute period to  AEMO, a centralised market operator. AEMO then determines the dispatch price and load for that five minute period, and determines the final spot price participants pay for each half-hour period by averaging the six five minute dispath% spelling
 prices in that half-hour. Market participants are allowed to revise bids and offers at any point before dispatch. The electricity supply curve is strongly J shaped, as coal generators are able to provide a large load cheaply but if demand exceeds this level the price spikes as other generators must be brought online expensively. These characteristics of the electricity market leads to a large amount of volatility in prices demonstrated by Figure \ref{rrpplot}. The seasonal effects of the load data are also present in prices, but these effects are dominated by irregular spikes as high as \$9,974/mWh, compared to a four year mean of \$42/mWh.

\begin{figure}[h]
\centering
\includegraphics[scale = 0.5]{load_timeseries.png}
\caption{Half hourly load data measured in megawatt hours for Victoria from January 1 2011 to December 31 2015. The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent sine curve in winter. Note: Replace with daily average to make the pattern clearer?}
\label{loadplot}
\end{figure}.

\begin{figure}[h]
\centering
\includegraphics[scale = 0.5]{RRP_timeseries.png}
\caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015. The price shows a strong time of week and time of day seasonal effect in the low price periods, and has irregular spikes.}
\label{rrpplot}
\end{figure}.

\subsection{Exponential Smoothing}

\citet{Taylor2003} explores the seasonal properties of minute-by-minute electricity demand in the United Kingdom and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}). Yearly effects are omitted as their inclusion requires a large number of latent states to be estimated. This model is later verified 
% Verified is a very strong word. I would say these authors have provided empirical support that the model is superior
in \citet{Taylor2008} as superior to other common time-series models for very short-term load forecasting.

\begin{align}
y_t &= l_{t-1} + d_{t-m_1} + w_{t-m_2} + e_t \label{ds-hw1} \\
l_t &= \alpha (y_t - d_{t-m_1} - w_{t-m_2}) + (1 - \alpha)l_{t-1} \label{ds-hw2}\\
d_t &= \delta (y_t - l_{t-1} - w_{t-m_2}) + (1 - \delta)d_{t-m_1} \label{ds-hw3} \\
w_t &= \omega (y_t - l_{t-1} - d_{t-m_1}) + (1 - \omega)w_{t-m_2} \label{ds-hw4}
\end{align}

where $m_1$ and $m_2$ are the lengths of the daily and weekly cycle, restricting the smoothing parameters $\alpha, \delta, \omega$ to lie in $(0, 1)$. This can be rewritten as a single source of error state-space model, {\bf as per \citep{Snyder1985},  as follows}

\begin{align}
y_t &= l_{t-1} + d_{t-m_1} + w_{t-m_2} + e_t \label{ds-hw-ssoe1} \\
l_t &= l_{t-1} + \alpha e_t \label{ds-hw-ssoe2} \\
d_t &= d_{t-m_1} + \delta e_t \label{ds-hw-ssoe3} \\
w_t &= w_{t-m_2} + \omega e_t \label{ds-hw-ssoe4}. 
\end{align}

% Now I think you should really explain what `single source of error' means, and why it is useful (it means you have the likelihood in closed form, except for the initializing state variables and parameters - which together would make up your $\theta$ vector).  Also you can identify what the `state' at time $t$ is
The unknown parameters are $\theta = (\alpha, \delta, \omega)', \sigma^2$ and $b_0 = (l_0, d_0, \dots, d_{-(m_1 - 1)}, w_0, \dots, w_{-(m_2 - 1)})'$ a $m_1 + m_2 + 1$ length vector of the initial states of the latent variables. 
\citet{Forbes2000} derives the marginal distribution $p(\theta | y)$ and notes that, conditioned on a draw of these parameters, the remaining unknown parameters can be sampled as a Bayesian linear regression problem. This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}), but restricting the latent parameters via

\begin{align}
d_t &= - \sum_{i=1}^{m_1-1} d_{t-i} \\
w_t &= - \sum_{i=1}^{m_2-1} w_{t-i}
\end{align}

avoids this problem. The resulting model is described by

\begin{align}
y_t &= l_{t-1} - \sum_{i = 1}^{m_1 - 1}d_{t-i} - \sum_{i = 1}^{m_2 - 1}w_{t-i} + e_t \label{ds-hw-rp1} \\
l_t &= l_{t-1} + \alpha e_t \label{ds-hw-rp2} \\
d_t &= - \sum_{i = 1}^{m_1 - 1}d_{t-i} + \delta e_t \label{ds-hw-rp3} \\
w_t &= - \sum_{i = 1}^{m_2 - 1}w_{t-i} + \omega e_t \label{ds-hw-rp4}.
\end{align}

\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$,

\begin{equation}
\label{exp-sm-marginal}
p(\theta | y_{1:T}) \propto \left| \widetilde{X}' \widetilde{X} \right|^{-1/2} \tilde{s}^{-(T-(m_1 + m_2 - 1))} p(\theta),
\end{equation}

where $\widetilde{X}$ and $\tilde{s}$ are functions of $\theta$ and $y$. The problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible difficult, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large.

\begin{itemize}
\item Should try and get an exponential smoothing MCMC working using Metropolis-Hastings for $\theta$. Log densities help with the $T$ being too large problem but the grid is too large for numerical integration to be a good idea.
\end{itemize}

% You need a concluding paragraph, in addition to the timeline below.
\section{Timeline}
\begin{itemize}
\item Put something here eventually
\end{itemize}

\bibliographystyle{asa}
\bibliography{references}

\end{document}
\grid

### 2017-02-21 Confirmation/confirmation_V2.tex
Substantial rewrite to report

@@
 \usepackage{ulem}
+\usepackage{tikz}
+%\usetikzlibrary{bayesnet}
 \usepackage{afterpage}
@@
 \numberwithin{equation}{section}
-\title{Real Time Variational Density Forecasts}
+\title{Real Time Variational Density Forecasting}
 \author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
@@
 While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models.
-This is problematic for when attempting to forecast electricity load in Victoria, when realizations occur at five minute intervals, yhe predictive density for the next period's electricity load may not be available before it is observed.
+This is problematic for when attempting to forecast electricity load in Victoria, when realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.
 
@@
 
-The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of a simple time series model.
+The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of two simple models.
 Section \ref{sec:ElecLoad} will introduce a planned emperical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period.
-Section \ref{sec:Discussion} outlines the major areas this thesis will make a contribution to and Section \ref{sec:Timeline} outlines the planned timeline of these contributions.
+Section \ref{sec:Discussion} outlines the major areas to which this thesis will make a contribution, and Section \ref{sec:Timeline} outlines the planned timeline for these contributions.
 
@@
 \label{posterior}
- p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
+ p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta})}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
 \end{equation}
@@
 Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
-To obtain the predictive distribution in (\ref{predictive}), $y_{T+1}$ can be treated as an unknown variable in $\boldsymbol{\theta}$ and hence directly computed by either MCMC and VB.
-Alternatively, once the posterior distribution in (\ref{posterior}) is available, (\ref{predictive}) can be found by simulation.
+Once the posterior distribution in (\ref{posterior}) is available, the predictive density forecast (\ref{predictive}) can be found by simulation.
 
@@
 \begin{align}
-\mu | \sigma^2, y_{1:T} &\sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
-\sigma^2 | y_{1:T} &\sim Inv.Gamma  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2}
+&\mu | \sigma^2, y_{1:T} \sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
+&\sigma^2 | y_{1:T} \sim Inv.Gamma  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2}
 \end{align}
 where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$.
-Posterior sampling can proceed by first simulating a draw of $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}).
+Posterior sampling can proceed by first simulating $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}).
+It should be noted that as we are simulating from $p(\sigma^2 | y)$ rather than $p(\sigma^2 | \mu, y)$, there is no Markov dependency between iterations and this is a Monte Carlo sampler rather than an MCMC.
 
@@
 Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.
-of distributions.
 
@@
 Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme.
-MFVB factorises (\ref{ELBO}) into a product of $k$ seperate components that each contain only one $q_i$, and then applies an algorithm to maximise each of these factors with respect the associated $q_i$.
+MFVB factorises (\ref{ELBO}) into a product of $k$ seperate components that each contain only one $q_i$, and then maximises each of these factors with respect to the associated $q_i$.
 Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
@@
 \label{mf3}
-q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T\tau + \mathbb{E}_{q(\sigma^2)}(\sigma^{-2}))}{2\tau\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\gamma + \lambda \sum_{t=1}^{T} y_t}{T \tau + \mathbb{E}_{q(\sigma^2 )}(\sigma^{-2})} \right)^2 \right\}
+q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T + \tau) \mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{2} \left( \mu - \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau} \right)^2 \right\}
 \end{equation}
 and
-\begin{equation}
+\begin{align}
 \label{mf4}
-q(\sigma^2 | \boldsymbol{\lambda}_2) \propto \sigma^{-2(T/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{t=1}^{T}y_t^2 + n\mathbb{E}_{q(\mu)}(\mu^2) - 2\sum_{t=1}^{T} y_t \mathbb{E}_{q(\mu)}(\mu)) - \beta}{\sigma^2} \right\}.
-\end{equation}
+q(\sigma^2 | \boldsymbol{\lambda}_2) &\propto \sigma^{-2((T+1)/2 + \alpha + 1)} \nonumber \\
+\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
+\end{align}
 
@@
 \begin{align}
-\tilde{\gamma} &= \frac{\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\gamma + \lambda \sum_{t=1}^{T} y_t}{T \tau + \mathbb{E}_{q(\sigma^2 )}(\sigma^{-2})}  \label{mf5} \\ 
-\tilde{\tau} &= \frac{\tau\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{(T\tau + \mathbb{E}_{q(\sigma^2)}(\sigma^{-2}))} \label{mf6} \\
-\tilde{\alpha} &= T/2 + \alpha  \label{mf7} \\
-\tilde{\beta} &= 1/2\left(\sum_{t=1}^{T}y_t^2 + n\mathbb{E}_{q(\mu)}(\mu^2) - 2\sum_{t=1}^{T} y_t \mathbb{E}_{q(\mu)}(\mu))\right) + \beta. \label{mf8}
+\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
+\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\right )^{-1} \label{mf6} \\
+\tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
+\tilde{\beta} &= \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right). \label{mf8}
 \end{align}.
@@
 The known values of the expectations in (\ref{mf5}) - (\ref{mf8}) for these distributions can be substituted into these expressions to derive the set of mean field equations:
+\begin{align}
+\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
+\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
+\end{align}
+$\tilde{\gamma}$ and $\tilde{\alpha}$ are omitted from the mean field equations as their values do not depend on other $\boldsymbol{\lambda}$ parameters.
 
-\begin{align}
-\tilde{\gamma} &= \frac{\tilde{\alpha} / \tilde{\beta} \gamma + \tau \sum_{t=1}^{T} y_t} {T \lambda + \tilde{\alpha} / \tilde{\beta}} \label{mf9} \\ 
-\tilde{\tau} &= \frac{T \tilde{\alpha} / \tilde{\beta}}{T \tau + \tilde{\alpha} / \tilde{\beta}} \label{mf10} \\
-\tilde{\beta} &= \frac{1}{2} \left(\sum_{t=1}^{T} y_t^2 + T(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \sum_{t=1}^{T} y_t \tilde{\gamma} \right) + \beta. \label{mf11}
-\end{align}
-$\tilde{\alpha}$ is omitted from the mean field equations as its value does not depend on other $\boldsymbol{\lambda}$ parameters.
-(\ref{mf9}) - (\ref{mf11}) contain a circular dependence in these equations an algorithm that iteratively cycles through these equations until the change in the ELBO is less than some pre-specified constant $\epsilon$.
-This algorithm is known as a coordiant ascent algorithm and is described by Algorithm (\ref{alg:algorithm1}), at each step it will fix all but one parameter and select the value of the omitted parameter which maximises the ELBO.
+As (\ref{mf9}) and (\ref{mf10}) contain a circular dependence, fitting the MFVB approximation requires an algorithm that cycles between these equations until the change in the ELBO is less than some pre-specified small constant $\epsilon$.
+This algorithm is known as a coordiant ascent, and is described by Algorithm (\ref{alg:algorithm1}).
+At each step coordinate ascent will fix one parameter and then maximise the ELBO with respect to the other parameter, so that changes in parameters results in a non-decreasing sequence of ELBO values.
+Coordinate ascent converges only when neither parameter can be changed independently to increase the ELBO by more than $\epsilon$, and hence reaches a local maximum.
+The current theory for Variational Bayes does not offer any insight to the shape of the ELBO so determining if the algorithm converges to a local or global maximum is difficult.
+One option to alleiviate this problem is to start the algorithm at a range of initial values and choose parameter values with the maximum converged ELBO, and hence lowest KL divergence to the true posterior.
+
+Figure (\ref{fig:norminvg}) compares the simulated posterior densities from Section (\ref{sec:MCMC}), (red) with the approximation from Mean Field Variational Bayes (blue) and the analytically available true posterior marginal densities (black).
+For both parameters all three densities are indistinguishable.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{norminvg}
+\caption{Marginal posterior densities for the model described in Section (\ref{sec:MCMC}).
+The true density is in black, the Monte Carlo simulation is red and the MFVB approximation is blue.
+Both MCMC and MFVB fit this density extremely well.}
+\label{fig:norminvg}
+\end{figure}
 
@@
 In this case the optimal distributional family $q_i$ is the same as the prior, analagous to the relationship between a prior-posterior conjugate pair.
-In the event that a non-conjugate prior is used, the form of the distribution in (\ref{mf2}) is unrecognisable, a further approximation may be used by substituting in another distribution to replace the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006}, a Delta Method inspired transform \citet{Wang2013}, though the only requirement is that the substitute distribution has moments that are tractable functions of its parameters to be used for other parameters in (\ref{mf2}).
+In the event that a non-conjugate prior is used, the form of the distribution in (\ref{mf2}) is unrecognisable, a further approximation may be used by substituting in another distribution to replace the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006} or a Delta Method inspired transform \citet{Wang2013}; though the only requirement is that the substitute distribution has moments that are tractable functions of its parameters to be used for other parameters in (\ref{mf2}).
 This secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable full conditional distributions.
@@
  \KwResult{Mean Field Approximation}
- Use (\ref{mf3}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
- Anaytically derive the set of mean field equations in \ref{mf9}) - (\ref{mf11})\;
+ Use (\ref{mf2}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
+ Anaytically derive the set of mean field equations in (\ref{mf9}) - (\ref{mf10})\;
  Initialise $\boldsymbol{\lambda}^{(1)}$ randomly.\;
  Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
+ Set $\tilde{\gamma}$ using \ref{mf5}.\;
  Set $\tilde{\alpha}$ using \ref{mf7}.\;
  \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
-  Set $\tilde{\gamma}^{(m)}$ using \ref{mf9}.\;
-  Set $\tilde{\tau}^{(m)}$ using \ref{mf10}.\;
-  Set $\tilde{\beta}^{(m)}$ using \ref{m11}.\;
+  Set $\tilde{\tau}^{(m)}$ using \ref{mf9}.\;
+  Set $\tilde{\beta}^{(m)}$ using \ref{mf10}.\;
   Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$.\;
@@
 
-The requirement for an exponential family model, as well as restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure \ref{fig:MCMCplot} shows strong dependence between $\phi_1$ and $\phi_2$.
-Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable.
-\citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, resulting in what is known as Stochastic Variational Bayes (SVB).
+The requirement for an exponential family model without introducing further approximations, as well as restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications.
+\citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, which can maximise the ELBO for non-factorisable functional forms of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
+We refer to the result as Stochastic Variational Bayes (SVB).
 
-The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ in (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ and the following updating step is applied until the ELBO converges to a maximum.
-
+The application of gradient ascent in SVB iteratively takes the derivative of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ in (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ and the following updating step is applied until change in the ELBO is less than $\epsilon$.
 \begin{equation}
-\label{SGA1}
+\label{SVB1}
 \boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}),
 \end{equation}
-
 where $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$.
@@
 If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.
-
 \begin{align}
@@
 
-Whilst a global maximum is desired, Variational Bayes theory does not offer any insight to the shape of the ELBO so determining if a local maximum is a global maximum is difficult.
-One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior.
-
 SVB can find only the optimal values for $\boldsymbol{\lambda}$ for a user specified distributional family $q$, and to run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO.
@@
 Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by
-
 \begin{equation}
-\label{SGA2}
+\label{SVB2}
 \nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)}))] (\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)})))
 \end{equation}
-
 where $s = 1, \dots, S$ indicates simulations from $q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})$.
@@
 Let 
-
 \begin{equation}
-\label{SGA3}
+\label{SVB3}
 G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(j)}), y_{1:T})\right)^2,
 \end{equation}
-
 then each component's learning rate is defined as
-
 \begin{equation}
-\label{SGA4}
+\label{SVB4}
 \rho^{(m)}_i = \eta \left(G_i^{(m)}\right)^{-1/2}
 \end{equation}
-
 for some tuning parameter $\eta$.
 
-The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\boldsymbol{\lambda}$ vector.
+The resulting Stochastic Gradient Ascent algorithm proceeds below in Algorithm (\ref{alg:algorithm2}) with a $p$ dimensional $\boldsymbol{\lambda}$ vector.
 
@@
   \For{$i =  1$ \KwTo $p$}{
-      Calculate $\nabla_{\lambda_i}$ from (\ref{SGA2})\;
-      Update $G_i^{(m)}$ and $\rho^{(m)}_i$ from (\ref{SGA3}) and (\ref{SGA4})\;
+      Calculate $\nabla_{\lambda_i}$ from (\ref{SVB2})\;
+      Update $G_i^{(m)}$ and $\rho^{(m)}_i$ from (\ref{SVB3}) and (\ref{SVB4})\;
       }
-  Update $\boldsymbol{\lambda}^{(m+1)}$ from (\ref{SGA1})\;
+  Update $\boldsymbol{\lambda}^{(m+1)}$ from (\ref{SVB1})\;
   Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$\;
@@
 
-\subsection{Vine Copulas}
+\subsection{Vine Copulas} \label{sec:vines}
 
@@
 \citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,
-
 \begin{equation}
@@
 \end{equation}
-
 where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.
 
-This idea is extended by the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas in a similar way to the factorisation of a joint distribution into a marginal and conditional distribution, see \citet{Joe2014} and references within.
-An example of this factorisation is
+This idea is extended by the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas, see \citet{Joe2014} and references within.
+A vine copula is represented by a series of trees, where the top-most tree is a graph that contains each variable as a node.
+The edges in each tree then form the nodes in the tree below, which are connect by an edge only if the associated edges in the previous tree connected to the same node.
+This sequence is continued or a $k$ dimensional $\boldsymbol{\theta}$ until the $(k-1)'th$ tree is reached that contains only one edge.
+Each of the $k(k-1)/2$ edges represents a bivariate copula between the two unique parameters in the connected nodes, conditioned on every parameter appearing in both nodes.
+Refer to Figure (\ref{fig:vinecop}) for an example.
 
-\begin{equation}
-\label{vc2}
-c(P(\theta_1), P(\theta_2), P(\theta_3)) = c(P(\theta_1), P(\theta_2)) \cdot c(P(\theta_1), P(\theta_3)) \cdot c(P(\theta_2), P(\theta_3) | P(\theta_1)).
-\end{equation}
+\iffalse
+\begin{figure}[h]
+\centering
+\tikz{ %
+    \node[latent] (theta1) {$\theta_1$} ; %
+    \node[latent, right = of theta1] (theta2) {$\theta_2}$} ; %
+    \node[latent, above = of theta2] (theta3) {$\theta_3$} ; %
+    \node[latent, below = of theta2] (theta4) {$\theta_4$} ; %
+    \edge{theta1} {theta2} ; %
+    \edge{theta2} {theta3} ; %
+    \edge{theta2} {theta4} ; %
 
-A copula for a $k$ dimensional $\theta$ can be replaced with $k(k-1)/2$ bivariate copulas arranged in a vine copula, greatly increasing the amount of flexibility available to fit a distribution.
-However factorising the copula and choosing a family for each bivariate copula takes exponential time in the number of parameters, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
+    \node[latent, below = of theta4] (theta2theta3) {$\theta_2 \theta_3$} ; %
+    \node[latent, left = of theta2theta3] (theta1theta2) {$\theta_1 \theta_2$} ; %
+    \node[latent, right = of theta2theta3] (theta2theta4) {$\theta_2 \theta_4$} ; %
+    \edge{theta2theta3} {theta1theta2} ; %
+    \edge{theta2theta3} {theta2theta4} ; %
+
+    \node[latent, below = of theta1theta2] (t1t3t2) {$\theta_1 \theta_3 | \theta_2$} ; %
+    \node[latent, right = of t1t3t2] (t3t4t2) {$\theta_3 \theta_4 | \theta_2$} ; %
+    \edge{t1t3t2} {t3t4t2} ;%
+}
+\caption{An example of a Vine Copula over four variables.
+The edges in the top tree contains unconditional bivariate copulas between the parameters on the connected nodes.
+The second tree contains the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node.
+The edge in the final tree contains the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.
+\label{fig:vinecop}
+\end{figure}
+\fi
+
+The resulting vine copula greatly increases the amount of flexibility available to model dependency in a distribution, but the cardinality of the set of possible vine structures and choices of family for each bivariate copula is exponential in the number of parameters, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
 \citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\boldsymbol{\theta}$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.
 
-Given a sample of $\boldsymbol{\theta}$, \citet{Dissmann2013} provide an algorithm to select the best vine factorisation by maximising Kendall's Tau, and the best family for each bivariate copula by minimising an information criterion such as AIC.
+Given a sample of $\boldsymbol{\theta}$, Di{\ss}mann's algorithm \citep{Dissmann2013} can be used to select the best vine factorisation by selecting a maximum spanning tree for the first tree that maximises the sum of Kendall's Tau for each pair of nodes connected by an edge.
+The algorithm then selects the best family for each of the first tree's bivariate copula by minimising an information criterion such as AIC.
+Edges and copula families in subsequent trees are selected sequentially.
 In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date.
@@
 
-\subsection{Applying Stochastic Variational Bayes}
+\subsection{Time Series Application}
 
-To illustrate issues involved, consider {\bf the second order} auto-regressive time series {model, denoted by AR(2) and described by}
-% no need to say more than this here.
-% But actually, I think you should just do the very simple Normal Inverse Gamma model here, since that has the parallel with VB below.
-Then in your example later on, you can do the full AR(2) example.
-I think this will help a lot.
-
+Consider the second order auto-regressive time series model, denoted by AR(2) and described by
 \begin{equation}
 \label{AR2}
-y_t = \rho_1 y_{t-1} + \rho_2 y_{t-2} + \epsilon_t
+y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
 \end{equation}
-
-where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$.
-The likelihood of parameters {\bf $\theta = (\rho_1, \rho_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\theta$, viewed as a function of $\theta$,)} is given by
-
+where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$ for $t = 3, \dots, T$. The first two observations, $y_1$ and $y_2$ are distributed according to 
+\begin{equation}
+\label{AR2initial}
+y_1, y_2, \sim \mathcal{N}(\boldsymbol{0}, \Sigma)
+\end{equation}
+where 
+\begin{equation}
+\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
+\end{equation}
+and $\gamma(k)$ is the $k-th$ autocorrelation for an AR(2) model.
+The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\boldsymbol{\theta}$, viewed as a function of $\boldsymbol{\theta}$, is given by
 \begin{align}
@@
 L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
-&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( \textbf{y}' \Sigma^{-1} \textbf{y} + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}
+&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( (y_1, y_2)' \Sigma^{-1} (y_1, y_2) + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
 \end{align}
 
-where
+We assume that the prior distribution, is given by
+\begin{equation}
+p(\sigma^2, \phi_1, \phi_2) \propto \sigma^{-2} \mathbb{I}(\phi_2 > -1)\mathbb{I}(\phi_2 < 1 + \phi_1) \mathbb{I}(\phi_2 < 1 - \phi_1), 
+\label{ar2prior}
+\end{equation}
+where $\mathbb{I}$ is the indicator function that equals one if the condition in the brackets is true and zero otherwise.
+This prior has positive mass if and only if a pair $(\phi_1, \phi_2)$ is within the AR(2) stationary region.
+Upon examination of the form of the product of the likelihood function in (\ref{likelihood}) and the prior density in (\ref{ar2prior}) it can be seen that the full conditional distribution $p(\sigma^2 | y, \phi_1, \phi_2)$ is an Inverse Gamma distribution, while the conditionals for both $\phi$ parameters are intractable but can be sampled using a Metropolis-Hastings step with truncated bivariate normal random walk proposal distribution.
+The truncation ensures that draws of $(\phi_1,\phi_2)$ are only taken from the AR(2) stationary region.
 
-\begin{equation}
-\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
-\end{equation}
-% Changed the matrices above to align properly
-$\gamma(k)$ is the $k-th$ autocorrelation and $\textbf{y} = (y_1, y_2)'$.
-% You also need to mention what $p(y_1, y_2 | \theta)$ is, and perhaps something about how $p(y_t | y_{1:t-1}, \theta)$ is available from the assumptions regarding the distribution of $\epsilon_t$.
-% Also, I don't like the use of vector notation (for $y=(y_1,y_2)$ )and subscripts for the remaining $y$'s in the likelihood. You can just break the equation over two lines if needed. 
-% Finally, perhaps you could say something about where $\gamma_0$ and $\gamma_1$ come from, or at least provide a reference for the reader to go and find out more about this model so you don't have to go into too much detail here.
-
-{\bf We assume that the prior distribution, is given by }
-% remember Tas's comment about stating that the joint prior is assumed to be the product of independent marginal priors
-
-\begin{align}
-p(\sigma^2) &\propto \sigma^{-2} \nonumber \\
-p(\rho_1, \rho_2) &\propto \mathbb{I}(\rho_2 > -1)\mathbb{I}(\rho_2 < 1 + \rho_1) \mathbb{I}(\rho_2 < 1 - \rho_1) \nonumber
-\end{align}
-
-are used, where $\mathbb{I}$ is the indicator and $p(\rho_1, \rho_2)$ defines the AR(2) stationary region.
-
-{\bf Upon examination of the form of the product of the likelihood function in (\ref{likelihood}) and the prior density in (),}
-	%need to be able to reference the prior too
-{\bf it can be seen that the full} conditional distribution $p(\sigma^2 | y, \phi_1, \phi_2)$ is {\bf an} Inverse Gamma {\bf distribution},
-% can you say which one? Be sure it is clear how to interpret the parameters of this distribution too.
- while the conditionals for both $\phi$ parameters are intractable and sampled using a Metropolis-Hastings step 
-% Since you are actually using an M-H step, you should have told the reader what this is back up when you introduced the Gibbs sampler.
-with truncated {\bf bivariate} %?
-normal random walk {proposal distribution}. The truncation ensures that draws {\bf of $(\rho_1,\rho_2)$} %just to be clear
-are only taken from the AR(2) stationary region.
-
-%Any graph that you put in to the document needs to be clearly explained.
-Why discard any?
-What does 'thinned' mean, and why did you do it?
-etc....
-Better to put the details of the MCMC in a paragraph, and then point to the figure, with only the essential details left in the caption.
-% Also, ideally you would be able to put the Greek symbols in the margins of the graphs, rather than 'Sigmasquared', for example...
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale = 0.6]{MCMC.png}
-\caption{MCMC draws for the AR(2) model with $T = 150, \phi_1 = 0.75, \phi_2 = 0.2$ and $\sigma^2 = 1$.
-Approximately 24\% of both $\phi$ parameters were accepted by the Metropolis Hastings step, then the first 10,000 draws were discarded and the remainder thinned to 1 in 200 to make dependence between draws negligible.}
-\label{fig:MCMCplot}
-\end{figure}
-
-Using the approach described in section 2.3, an application of VC algorirhtm to the MCMC draws obtained in section 2.1 found the following distribution to be optimal by AIC.
-
+Data was simulated with $T = 150, \sigma^2 = 1, \phi_1 = 0.7$ and $\phi_2 = 0.2$, then the MCMC algorithm described above is applied.
+Marginal distributions are fit to each parameter in $\boldsymbol{\theta}$ by minimising AIC, and an application of Di{\ss}mann's algorithm to the draws found an optimal Vine Copula structure.
+The functional form of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to be used in SVB is:
 \begin{itemize}
@@
 
-Figure \ref{fig:VBfit} shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.
+Figure \ref{fig:VBfit} shows the fit of Stochastic Variational Bayes (blue) to the MCMC (red) samples.
+SVB appears to have selected values for $\boldsymbol{\lambda}$ that closely match the MCMC posterior distribution.
 
@@
 \includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{VBfit.png}
-\caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model described in section 2.1.
-SVB resulted in an almost identical posterior to MCMC, requiring around one hundredth of the runtime for MCMC.}
+\caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model.
+SVB resulted in an almost identical posterior to MCMC, and required approximately one hundredth of the runtime of MCMC.}
 \label{fig:VBfit}
@@
 
-Half hourly electricity load data for Victoria, Australia has been collected for the 2011-2015 period, and is plotted in Figure \ref{fig:loadplot}.
+Half hourly electricity load data for Victoria, Australia is provided by the Australian Energy Market Operator (AEMO) for the 2011-2015 period, and is plotted in Figure \ref{fig:loadplot}.
 The time series displays strong seasonal characteristics, with yearly, day of week, and time of day patterns.
@@
 
-The electricity market in Victoria operates by market participants sending electricity bid and offers for each five minute period to  AEMO, a centralised market operator.
-AEMO then determines the dispatch price and load for that five minute period, and determines the final spot price participants pay for each half-hour period by averaging the six five minute dispath prices in that half-hour.
+The electricity market in Victoria operates by market participants sending electricity bid and offers for each five minute period to AEMO, which then determines the dispatch price and load for that five minute period.
+At the end of each half-hour block the previous six dispatch prices are averaged to obtain the final spot price that participants trade at.
 Market participants are allowed to revise bids and offers at any point before dispatch.
@@
 
-\citet{Taylor2003} explores the seasonal properties of minute-by-minute electricity demand in the United Kingdom and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}).
-Taylor omits yearly effects as their inclusion requires a large number of latent states to be estimated.
-This model is later verified in \citet{Taylor2008} as superior to other common time-series models for very short-term load forecasting.
-
+\citet{Taylor2003} explores the seasonal properties of minute-by-minute electricity demand in the United Kingdom and introduces the double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}).
+Yearly seasonal effects are omitted as their inclusion requires a large number of latent states to be estimated, however \citet{Taylor2008} later provides empirical support that this model is superior to other common time-series models such as SARIMA based models that include a yearly seasonal effect.
+The double seasonal Holt-Winters exponential smoothing model is described by
 \begin{align}
@@
 \end{align}
-
 where $m_1$ and $m_2$ are the lengths of the daily and weekly cycle, and the smoothing parameters $\alpha, \delta, \omega$ are also restricted to to lie in $(0, 1)^3$.
@@
 \end{align}
-
-The unknown parameters are $\boldsymbol{\theta} = (\alpha, \delta, \omega)', \sigma^2$ and $\boldsymbol{b}_0 = (l_0, d_0, \dots, d_{-(m_1 - 1)}, w_0, \dots, w_{-(m_2 - 1)})'$ a $k = m_1 + m_2 + 1$ length vector of the initial states of the latent variables.
+The unknown parameters are $\boldsymbol{\theta} = (\alpha, \delta, \omega)', \sigma^2$ and $\textbf{b}_0 = (l_0, d_0, \dots, d_{-(m_1 - 1)}, w_0, \dots, w_{-(m_2 - 1)})'$ a $k = m_1 + m_2 + 1$ length vector of the initial states of the latent variables.
+The single source of error state-space model refers to the random error $e_t$ being shared across the data each latent variable, and allows the likelihood to be expressed as a closed form function of the unknown parameters.
 
 \citet{Forbes2000} provides a transformation to the model that allows a straight-forward Bayesian analysis in small samples, noting that the state space model can be expressed as
-
 \begin{align}
-y_t &= \textbf{x}' \boldsymbol{b}_t + e_t \label{ss_me} \\
+y_t &= \textbf{x}' \textbf{b}_t + e_t \label{ss_me} \\
 \textbf{b}_t &= \textbf{T} \textbf{b}_{t-1} + \tilde{\boldsymbol{\theta}} e_t \label{ss_te}.
@@
 Using $D = T - \tilde{\boldsymbol{\theta}} \textbf{x}'$, the transition equation in (\ref{ss_te}) can be written as 
-
 \begin{equation}
-b_t = D b_{t-1} + \tilde{\boldsymbol{\theta}} y_t \label{ss_te_2}.
+\textbf{b}_t = D \textbf{b}_{t-1} + \tilde{\boldsymbol{\theta}} y_t \label{ss_te_2}.
 \end{equation}
 
-Setting $\bar{b}_0$ = 0, the recursion in (\ref{ss_te_2}) allows the transformed observations $\tilde{y}_t$ to be obtained via
-
+Setting $\bar{\textbf{b}}_0 = 0$, the recursion in (\ref{ss_te_2}) allows the transformed observations $\tilde{y}_t$ to be obtained via
 \begin{equation}
-\tilde{y}_t = y_t - x'\bar{b}_{t-1} \label{ss_rp_y}.
+\tilde{y}_t = y_t - x'\bar{\textbf{b}}_{t-1} \label{ss_rp_y}.
 \end{equation}
@@
 Collecting these as $\widetilde{X}' = (\tilde{\textbf{x}}_1, \tilde{\textbf{x}}_2, \dots, \tilde{\textbf{x}}_T)$ and $\widetilde{Y}' = (\tilde{y}_1, \tilde{y}_2, \dots, \tilde{y}_T)$, and assuming the errors $e_t$ are Gaussian, the model likelihood is
-
 \begin{equation}
@@
 \end{equation}
-
 Hence once conditioned on a draw of $\boldsymbol{\theta}$, the remaining unknown parameters can be sampled as a Bayesian linear regression problem.
 
-\citet{Forbes2000} provides the marginal distribution of $\theta$,
-
+\citet{Forbes2000} provides the marginal distribution of $\boldsymbol{\theta}$,
 \begin{equation}
 \label{exp-sm-marginal}
-p(\boldsymbol{\theta} | y_{1:T}) \propto \left| \widetilde{X}' \widetilde{X} \right|^{-1/2} \tilde{s}^{-(T-k)} p(\theta),
+p(\boldsymbol{\theta} | y_{1:T}) \propto \left| \widetilde{X}' \widetilde{X} \right|^{-1/2} \tilde{s}^{-(T-k)} p(\boldsymbol{\theta}),
 \end{equation}
-
-where $\tilde{s}^2 = (\widetilde{Y} - \widetilde{X}\hat{\boldsymbol{b}}_0)'(\widetilde{Y} - \widetilde{X}\hat{\boldsymbol{b}}_0) / (T - (m_1 + m_2 + 1))$ is the sum of squared errors for $\hat{\boldsymbol{b}}_0 = (\widetilde{X}'\widetilde{X})^{-1}\widetilde{X}'\widetilde{Y}$.
+where $\tilde{s}^2 = (\widetilde{Y} - \widetilde{X}\hat{\textbf{b}}_0)'(\widetilde{Y} - \widetilde{X}\hat{\textbf{b}}_0) / (T - (m_1 + m_2 + 1))$ is the sum of squared errors for $\hat{\textbf{b}}_0 = (\widetilde{X}'\widetilde{X})^{-1}\widetilde{X}'\widetilde{Y}$.
 With the seasonality parameterisation in (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}), the matrix $\widetilde{X}' \widetilde{X}$ is singular, but restricting the latent variables via
-
 \begin{align}
@@
 \end{align}
-
 avoids this problem. The resulting model is described by
-
 \begin{align}
@@
 \end{align}
+and $\textbf{b}_0 = (l_0, d_0, \dots, d_{-(m_1 - 2)}, w_, \dots, w_{-(m_2 - 2)})'$ is a length $k = m_1 + m_2 - 1$ vector.
 
-and $\boldsymbol{b}_0 = (l_0, d_0, \dots, d_{-(m_1 - 2)}, w_, \dots, w_{-(m_2 - 2)})'$ is a length $k = m_1 + m_2 - 1$ vector.
-
-\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$ in (\ref{exp-sm-marginal}), however the problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large, as is the case with high frequency electricity load data.
-An efficient MCMC algorithm using the conditional distribution of $p(\theta | y_{1:T}, \sigma^2, b_0)$ may be required to sample the full posterior distribution $p(\theta, \sigma^2, b_0 | y_{1:T}$ and obtain an approximating distribution $q(\theta, \sigma^2, b_0 | \lambda)$ with a reasonable goodness of fit.
-From this point, Stochastic Variational Bayes will be used to update the posterior distribution as more data is observed to $p(\theta, \sigma^2, b_0 | y_{1:T+J}), J = 1, 2, \dots$ and provide the predictive density $p(y_{T+J+1} | y_{T+J})$.
+\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\boldsymbol{\theta}$ in (\ref{exp-sm-marginal}), however the problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible, and the $T$ in the exponent makes most evalutions of $p(\boldsymbol{\theta}  | y_{1:T})$ computationally zero when $T$ is large, as is the case with high frequency electricity load data.
+An efficient MCMC algorithm that scales to large datasets with models using multiple seasonality parameters will be required to sample the full posterior distribution $p(\boldsymbol{\theta} , \sigma^2, b_0 | y_{1:T}$ and obtain an approximating distribution $q(\boldsymbol{\theta}, \sigma^2, b_0 | \lambda)$ with a reasonable goodness of fit.
+From this point, Stochastic Variational Bayes will be find an approximation to the updated posterior distribution as more data is observed, and provide the predictive density $q(y_{T+J+1} | y_{T+J})$.
 
@@
 
-Put main points and issues to investigate here
+One of the major aims of this thesis is to provide a methodology to forecast predictive densities with minimal calculation required to update parameter estimates as more data is observed, so an accurate forecast can be made even on a very short time-frame.
+However the short time-frame and high frequency of observating data implies that there is already a large number of past observations available, so any the marginal information from an additional observation on the parameters may be minimal.
 
-- When parameters change, models with non-simple filtering methods to obtain changes in states. ie variable parameter state space models.
+The benefits of rapidly updating parameters will be increased in situations where there are major changes in the data generating process over time.
+One example is a state-space model where information on the next latent state is increased significantly after observing the most recent data point.
+However, unless the model has either a non-linearity or non-Gaussian error, the Kalman filter can be applied to obtain an accurate predictive density without difficult computation.
+Volatility forecasting for stock returns is a potentially useful application as data is observed on a short time-frame, and the stochastic volatility state-space models commonly used do not admit the use of the Kalman filter.
+
+A contribution of this thesis is the use of information generated by MCMC to develop a Vine Copula used by the approximating distribution in Stochastic Variational Bayes, but updating state-space models by SVB requires an approximation for latent states associated with newly observed data that have not been included in the most recent MCMC sample.
+In this case, the dependence for the new latent state must be extrapolated from the copula structure used for previous latent states, which may be sub-optimal.
+The robustness of SVB approximations to mis-specified paramater dependence should be investigated.
+
+Finally, Variational Bayes only becomes a viable method for forecast densities if the statistical error associated with using an approximate distribution is less than the statistical error associated with forecasting based on the MCMC generated posterior distribution that is not conditioned on the most recent data.
 
@@
 
-Put this in!
+\begin{tabular}{|l|l|}
+\hline
+  June 2017 & $\cdot$ Derive an efficient MCMC algorithm for exponential smoothing and \\
+  & apply SVB for electricity load forecasting. \\
+  & $\cdot$ Similarly apply SVB to competing models such as SARIMA. \\
+  & $\cdot$ Compare forecast accuracy to MCMC with and without updates to \\
+  & parameters. \\ \hline
+  December 2017 & $\cdot$ Investigate state-space models, including time-varying parameters, \\
+  & non-linearities or non-Gaussian errors. \\
+  & $\cdot$ Apply this to Stochastic Volatility models. \\
+  & $\cdot$ Extend model with multivariate data and Markov Switching mechanisms.\\
+  & $\cdot$ Compare results to Particle Filter based methods. \\  \hline
+  June 2018 & $\cdot$Investigate impact of sub-optimal approximating distributions  \\
+  & such as mis-specified copulas for new latent states. \\
+  & $\cdot$ Investigate other ways to construct the approximating distribution  \\
+  & either with or without a posterior sample being available. \\  \hline
+  December 2018 & $\cdot$ Allow extra time for updates in the active VB literature  \\
+  & to develop, and adapt them to the benefit my research. \\  \hline
+  May 2019 & $\cdot$ Put seperate parts together and polish the final thesis for submission. \\
+\hline
+\end{tabular}
 

### 2017-02-21 Confirmation/confirmation_example.R

@@ -159,12 +159,12 @@ p1dens = mutate(data.frame(support = seq(0.35, 1.15, length.out = 200)), density
 p2dens = mutate(data.frame(support = seq(-0.1, 0.55, length.out = 200)), density = dnorm(support, lambda[2], sqrt(Sigma[2,2])))
 sig2dens = mutate(data.frame(support = seq(0.501, 2, length.out = 150)), density = densigamma(support, lambda[6], lambda[7]))
 
-plot1 = ggplot() + geom_density(data = thetaKeep, aes(Phi1, y = ..density..)) + theme_bw()
-  #geom_line(data = p1dens, aes(support, density), colour = "blue") + theme_bw()
-plot2 = ggplot() + geom_density(data = thetaKeep, aes(Phi2, y = ..density..)) + theme_bw()
-  #geom_line(data = p2dens, aes(support, density), colour = "blue") + theme_bw()
-plot3 = ggplot() + geom_density(data = thetaKeep, aes(SigmaSquared, y = ..density..)) + theme_bw()
- # geom_line(data = sig2dens, aes(support, density), colour = "blue") + theme_bw()
+plot1 = ggplot() + geom_density(data = thetaKeep, aes(Phi1, y = ..density..), colour = "red") + 
+  geom_line(data = p1dens, aes(support, density), colour = "blue") + theme_bw() + labs(x = expression(phi[1]), y = "Density")
+plot2 = ggplot() + geom_density(data = thetaKeep, aes(Phi2, y = ..density..),  colour = "red") +
+  geom_line(data = p2dens, aes(support, density), colour = "blue") + theme_bw()+ labs(x = expression(phi[1]), y = "Density")
+plot3 = ggplot() + geom_density(data = thetaKeep, aes(SigmaSquared, y = ..density..),  colour = "red") + 
+  geom_line(data = sig2dens, aes(support, density), colour = "blue") + theme_bw() + labs(x = expression(sigma^2), y = "Density")
 
 p1p2 = expand.grid(sup1 = seq(0.35, 1.15, length.out = 200), sup2 = seq(-0.1, 0.55, length.out = 200))
 p1p2$density = apply(p1p2, 1, mvtnorm::dmvnorm, mean = lambda[1:2], sigma = Sigma) 
[…]

### 2017-02-22 Confirmation/confirmation_V2.tex
final tweaks before draft submission

@@
 \end{align}
-Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$ and can be used for forecasting and inference about the unknown parameters.
+Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$.
+Samples taken before the MCMC converges ot the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
 The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
@@
 
-Figure (\ref{fig:norminvg}) compares the simulated posterior densities from Section (\ref{sec:MCMC}), (red) with the approximation from Mean Field Variational Bayes (blue) and the analytically available true posterior marginal densities (black).
+Data was simulated according to the Gaussian model introduced in Section (\ref{sec:MCMC}) with $T = 100, \mu = 2 and \sigma^2 = 1$.
+The hyperparameters were set to $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$.
+Figure (\ref{fig:norminvg}) compares the analytically available true posterior marginal densities for both $\mu$ and $\sigma^2$ (black) with the densities generated by the Monte Carlo sampler (red) and the approximation from Mean Field Variational Bayes (blue).
 For both parameters all three densities are indistinguishable.
@@
 \centering
-\includegraphics[width=0.5\linewidth,height=\textheight,keepaspectratio]{norminvg}
+\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{norminvg}
 \caption{Marginal posterior densities for the model described in Section (\ref{sec:MCMC}).
-The true density is in black, the Monte Carlo simulation is red and the MFVB approximation is blue.
+The true density is in black, the Monte Carlo sampler is red and the MFVB approximation is blue.
 Both MCMC and MFVB fit this density extremely well.}
@@
 A vine copula is represented by a series of trees, where the top-most tree is a graph that contains each variable as a node.
-The edges in each tree then form the nodes in the tree below, which are connect by an edge only if the associated edges in the previous tree connected to the same node.
+The edges in each tree then form the nodes in the tree below, which are connected by an edge only if the associated edges in the previous tree shared the same node.
 This sequence is continued or a $k$ dimensional $\boldsymbol{\theta}$ until the $(k-1)'th$ tree is reached that contains only one edge.
@@
 
-\iffalse
 \begin{figure}[h]
 \centering
-\tikz{ %
-    \node[latent] (theta1) {$\theta_1$} ; %
-    \node[latent, right = of theta1] (theta2) {$\theta_2}$} ; %
-    \node[latent, above = of theta2] (theta3) {$\theta_3$} ; %
-    \node[latent, below = of theta2] (theta4) {$\theta_4$} ; %
-    \edge{theta1} {theta2} ; %
-    \edge{theta2} {theta3} ; %
-    \edge{theta2} {theta4} ; %
-
-    \node[latent, below = of theta4] (theta2theta3) {$\theta_2 \theta_3$} ; %
-    \node[latent, left = of theta2theta3] (theta1theta2) {$\theta_1 \theta_2$} ; %
-    \node[latent, right = of theta2theta3] (theta2theta4) {$\theta_2 \theta_4$} ; %
-    \edge{theta2theta3} {theta1theta2} ; %
-    \edge{theta2theta3} {theta2theta4} ; %
-
-    \node[latent, below = of theta1theta2] (t1t3t2) {$\theta_1 \theta_3 | \theta_2$} ; %
-    \node[latent, right = of t1t3t2] (t3t4t2) {$\theta_3 \theta_4 | \theta_2$} ; %
-    \edge{t1t3t2} {t3t4t2} ;%
-}
+\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
 \caption{An example of a Vine Copula over four variables.
-The edges in the top tree contains unconditional bivariate copulas between the parameters on the connected nodes.
-The second tree contains the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node.
-The edge in the final tree contains the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.
+The edges in the top tree represent unconditional bivariate copulas between the parameters on the connected nodes.
+The second tree edges represent the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node.
+The edge in the final tree represents the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.}
 \label{fig:vinecop}
 \end{figure}
-\fi
 
@@
 L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
-&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} \left( (y_1, y_2)' \Sigma^{-1} (y_1, y_2) + \frac{1}{\sigma^2} \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
+&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} (y_1, y_2) \Sigma^{-1} (y_1, y_2)' \right\} \nonumber \\
+&\times \exp \left\{ \frac{-1}{2\sigma^2} \left( \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
 \end{align}
@@
 
-Finally, Variational Bayes only becomes a viable method for forecast densities if the statistical error associated with using an approximate distribution is less than the statistical error associated with forecasting based on the MCMC generated posterior distribution that is not conditioned on the most recent data.
+Finally, Variational Bayes only becomes a viable method for density forecasts if the statistical error associated with using an approximate distribution is less than the statistical error associated with forecasting based on an MCMC generated posterior distribution that is not conditioned on the most recent data.
+The investigation of this trade-off is of central importance to the thesis.
 

### 2017-02-22 Confirmation/normig.R

@@ -19,7 +19,7 @@ T = 100
 y = rnorm(T, mu, sig2)
 
 gamma = 0
-tau = 10
+tau = 1
 alpha = 1
 beta = 1
 
@@ -33,8 +33,8 @@ for(i in 1:100000){
   muMC[i] = rnorm(1, (tau * gamma + T * mean(y))/(tau+T), sqrt(sigMC[i] / (tau + T)))
 }
 
-sigsup = seq(0.8, 2.5, 0.01)
-musup = seq(1.5, 2.3, 0.01)
+sigsup = seq(0.5, 2.2, 0.01)
+musup = seq(1.6, 2.6, 0.01)
 
 sigtrue = densigamma(sigsup, alpha + T/2, vbarbar*sbarsq/2)
 mutrue = rep(0, length(musup))

### 2017-02-27 Confirmation/confirmation_V2.tex
intro changes

@@
 \usepackage{ulem}
-\usepackage{tikz}
-%\usetikzlibrary{bayesnet}
 \usepackage{afterpage}
@@
 \title{Real Time Variational Density Forecasting}
-\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
+\author{Nathaniel Tomasetti}
 \begin{document}
@@
 Bayesian methods are computationally complex as they require integration over the set of unknowns, whether parameters or latent stochastic variables.
-This set is quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}.
-While analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods.
+This set is typically quite large for models used to model short term electricity load and price as there are strong seasonal patterns in the data \citep{Taylor2003}.
+Analytical integration is typically not feasible to obtain the predictive distribution for such models, however there are a range of numerical and simulation-based techniques available for this purpose, notably Markov Chain Monte Carlo (MCMC) methods.
 While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models.
-This is problematic for when attempting to forecast electricity load in Victoria, when realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.
+This is problematic when attempting to forecast electricity load in Victoria, as realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.
 
@@
 
+The literature on Variational Bayes is dominated by two major implementations: Mean Field Variational Bayes (MFVB) and what we call Stochastic Variational Bayes (SVB).
+For either implementation we emphasise that there are two seperate tasks: Choose a functional form for the approximating distribution, and then optimise the parameters of this distribution to minimise a well defined divergence between the approximation and the true posterior.
+
+Mean Field Variational Bayes (see \citet{Bishop2006} for an overview) can be applied to exponetial family models with conjugate priors, and restricts the class of approximating distributions to the factorisable family.
+MFVB provides both the optimal form of the approximating distribution within this class, and a computationally simple coordinate ascent algorithm is available to optimise parameters.
+Stochastic Variational Bayes (see \citet{Paisley2012} and \citet{Ranganath2014}) lifts both the restrictions for an exponential family model and the use of an approximating distribution that is factorisable.
+While parameter optimisation has a well developed gradient ascent algorithm, the determination of the functional form to be used is an active area of research.
+
+We explore the use of a copula to augment the approximating distribution with a flexible dependence structure, which has seen success in \citet{Tran2015} and \citet{Ng2016}.
+The time series context of our implementation implies that MCMC samples for out of date samples are available, and we emphasise the use of this information to guide the construction of a vine copula so that the approximation has a dependence structure similiar to the true posterior.
+
 The discussion of MCMC and VB will be continued in Section \ref{sec:BayesInf}, with the methods contrasted in the context of two simple models.
-Section \ref{sec:ElecLoad} will introduce a planned emperical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period.
+Section \ref{sec:ElecLoad} will introduce a planned empirical application of VB to Victorian electricity load using an exponential smoothing model and data from the 2012-2015 period.
 Section \ref{sec:Discussion} outlines the major areas to which this thesis will make a contribution, and Section \ref{sec:Timeline} outlines the planned timeline for these contributions.
@@
 \label{posterior}
- p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta})}{\int_{\boldsymbol{\theta}}p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
+ p(\boldsymbol{\theta} | y_{1:T}) = \frac{p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta})}{\int p(y_{1:T}|\boldsymbol{\theta})p(\boldsymbol{\theta}) d\boldsymbol{\theta}}.
 \end{equation}
@@
 In brief, MCMC is used to create a sample from $p(\boldsymbol{\theta} | y_{1:T})$, with any function of $\boldsymbol{\theta}$ that is desired estimated from that sample.
-In contrast, VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation, denoted by $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxiliary parameters associated with the approximation and may depend on the observations $y_{1:T}$.
-For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian, then $\boldsymbol{\lambda}$ would denote the mean and variance parameters that characeterise this normal distribution for $\boldsymbol{\theta}$.
-Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ such that the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
+In contrast, VB replaces $p(\boldsymbol{\theta} | y_{1:T})$ with an approximation, denoted by $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, where $\boldsymbol{\lambda}$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $y_{1:T}$.
+For example, if $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is Gaussian, then $\boldsymbol{\lambda}$ would denote the mean and variance parameters that characterise this distribution for $\boldsymbol{\theta}$.
+Variational Bayes aims to choose the distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ within a restricted class such that the a measure of divergence, typically the Kullback-Leibler divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$, is minimised.
 Once the posterior distribution in (\ref{posterior}) is available, the predictive density forecast (\ref{predictive}) can be found by simulation.
@@
 Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\boldsymbol{\theta} | y_{1:T})$.
-Samples taken before the MCMC converges ot the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
+Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
 The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
-On occasion, the full conditional distribution can not be recognised and a Metropolis-Hastings-within-Gibbs step can be used, see \citet{Gilks1995} for details.
+On occasion, the full conditional distribution cannot be recognised and a Metropolis-Hastings-within-Gibbs step can be used, see \citet{Gilks1995} for details.
 
-For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution with prior distributions given by $\mu | \sigma^2 \sim \mathcal{N}(\gamma, \sigma^2/\tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$.
+For illustration, consider data $y_t$ for $t = 1, \dots, T$ generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution with prior distributions given by $\mu | \sigma^2 \sim \mathcal{N}(\gamma, \sigma^2/\tau)$ and $\sigma^2 \sim IG(\mbox{shape} = \alpha, \mbox{scale =}\beta)$, where $IG$ repesents the inverse of a Gamma distribution.
 It can be shown that
@@
 &\mu | \sigma^2, y_{1:T} \sim \mathcal{N} \left(\frac{\tau\gamma + T\bar{y}}{\tau + T}, \frac{\sigma^2}{\tau  + T} \right) \label{posmu} \\
-&\sigma^2 | y_{1:T} \sim Inv.Gamma  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2}
+&\sigma^2 | \mu, y_{1:T} \sim IG  \left( \mbox{shape} = \alpha + T/2, \mbox{scale} = \beta + \frac{1}{2}\sum_{t=1}^{T}y_t^2 + \tau\gamma^2 - \frac{(\tau\gamma + T\bar{y})^2}{\tau + T} \right) \label{possig2}
 \end{align}
-where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$.
-Posterior sampling can proceed by first simulating $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}).
-It should be noted that as we are simulating from $p(\sigma^2 | y)$ rather than $p(\sigma^2 | \mu, y)$, there is no Markov dependency between iterations and this is a Monte Carlo sampler rather than an MCMC.
+where $\bar{y} = 1/T \sum_{t=1}^{T}y_t$. Posterior sampling can proceed by first simulating $\sigma^2$ from (\ref{possig2}) and conditioning on this draw to simulate $\mu$ from (\ref{posmu}).
 
@@
 As discussed in Section \ref{sec:Intro}, a faster (albeit approximate) alternative to MCMC is VB.
-This approach introduces an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ with the choice of the functional form $q$ and subsequent auxiliary parameter vector $\boldsymbol{\lambda}$ determined so that the Kullback-Leibler (KL) divergence \citep{Kullback1951} from$q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$ is minimised.
+VB is an umbrella term for a broad collection of algorithms that introduce an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ with the choice of the functional form $q$ and subsequent auxiliary parameter vector $\boldsymbol{\lambda}$ determined so that an error function is minimised.
+The classical choice for this error function is the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$.
 The KL divergence is defined by
@@
 \label{KL-def}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T}] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
 \end{equation}
@@
 \label{KL-ELBO}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q, y_{1:T})
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})
 \end{equation}
-where $\mathcal{L}(q, y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q, y_{1:T})$ is defined by
+where $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.
+$\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
 \begin{equation}
 \label{ELBO}
-\mathcal{L}(q, y_{1:T}) = \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int_{\boldsymbol{\theta}} q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
+\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
 \end{equation}
@@
 Since $\ln(p(y_{1:T}))$ is constant with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$, maximising (\ref{ELBO}) with respect to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ is equivalent to minimising (\ref{KL-def}).
-Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Varitaional Bayes, namely Mean Field Variational Bayes and Stochastic Variational Bayes.
+Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Variational Bayes we consider, Mean Field Variational Bayes and Stochastic Variational Bayes.
 Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.
+The statistical properties of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ in MFVB and SVB are largely driven by the choice of the KL divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ as an error function, and there is significant interest in other error functions, such as the KL divergence from $p(\boldsymbol{\theta} | y_{1:T})$ to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ \citep{Minka2001} and Stein's discrepency (\citealp{Liu2016}; \citealp{Ranganath2016b}).
 
@@
 
-Mean Field Variational Bayes (MFVB) has origins in physics \citep{Chandler1987} and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Beal2003}.
+Mean Field Variational Bayes (MFVB) has origins in physics (\citealp{Amari1985}; \citealp{Chandler1987}) and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Ghahramani2000}.
 MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
@@
 Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars.
-Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may be of a different dimension to $\theta_i$.
+Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may have different dimension to $\theta_i$.
 $\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.
@@
 Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme.
-MFVB factorises (\ref{ELBO}) into a product of $k$ seperate components that each contain only one $q_i$, and then maximises each of these factors with respect to the associated $q_i$.
+MFVB factorises (\ref{ELBO}) into a product of $k$ separate components that each contain only one $q_i$, and then maximises each of these factors with respect to the associated $q_i$.
 Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
@@
 
-We illustrate the implementation of MFVB with the model considered in Section (\ref{sec:MCMC}), where $y_t$ for $t = 1, \dots, T$ is generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / \tau)$ and $\sigma^2 \sim Inv.Gamma(\mbox{shape} = \alpha, \mbox{scale =}\beta)$.
+We illustrate the implementation of MFVB with the model considered in Section (\ref{sec:MCMC}), where $y_t$ for $t = 1, \dots, T$ is generated independently from a $\mathcal{N}(\mu, \sigma^2)$ distribution, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / \tau)$ and $\sigma^2 \sim IG(\mbox{shape} = \alpha, \mbox{scale =}\beta)$.
 It can be shown that
@@
 q(\sigma^2 | \boldsymbol{\lambda}_2) &\propto \sigma^{-2((T+1)/2 + \alpha + 1)} \nonumber \\
-\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
+\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)}(\mu^2) - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
 \end{align}
 
-As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q(\mu |  \boldsymbol{\lambda}_1)$ and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ are the same family as the prior distributions, and are denoted by $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$.
+As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q(\mu |  \boldsymbol{\lambda}_1)$ and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ are the same family as the prior distributions, and are denoted by $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim IG(\tilde{\alpha}, \tilde{\beta})$.
 The vector $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$ is given by
@@
 
-The known values of the expectations in (\ref{mf5}) - (\ref{mf8}) for these distributions can be substituted into these expressions to derive the set of mean field equations:
+The expectations in (\ref{mf5}) - (\ref{mf8}) are available in a closed form for these distributions, and these are substituted in to derive the set of mean field equations:
 \begin{align}
@@
 As (\ref{mf9}) and (\ref{mf10}) contain a circular dependence, fitting the MFVB approximation requires an algorithm that cycles between these equations until the change in the ELBO is less than some pre-specified small constant $\epsilon$.
-This algorithm is known as a coordiant ascent, and is described by Algorithm (\ref{alg:algorithm1}).
+This algorithm is known as a coordinate ascent, and is described by Algorithm (\ref{alg:algorithm1}).
 At each step coordinate ascent will fix one parameter and then maximise the ELBO with respect to the other parameter, so that changes in parameters results in a non-decreasing sequence of ELBO values.
@@
 The current theory for Variational Bayes does not offer any insight to the shape of the ELBO so determining if the algorithm converges to a local or global maximum is difficult.
-One option to alleiviate this problem is to start the algorithm at a range of initial values and choose parameter values with the maximum converged ELBO, and hence lowest KL divergence to the true posterior.
-
-Data was simulated according to the Gaussian model introduced in Section (\ref{sec:MCMC}) with $T = 100, \mu = 2 and \sigma^2 = 1$.
+One option to alleviate this problem is to start the algorithm at a range of initial values and choose parameter values with the maximum converged ELBO, and hence lowest KL divergence to the true posterior.
+\vspace{2mm}
+\begin{algorithm}[H]
+ \SetKwInOut{Input}{Input}
+ \Input{Log Joint Density}
+ \KwResult{Mean Field Approximation}
+ Use (\ref{mf2}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
+ Anaytically derive the set of mean field equations in (\ref{mf9}) - (\ref{mf10})\;
+ Set $\tilde{\gamma}$ using \ref{mf5}.\;
+ Set $\tilde{\alpha}$ using \ref{mf7}.\;
+ Initialise $\tilde{\tau}^{(1)}$ and $\tilde{\beta}^{(1)}$ randomly.\;
+ Evaluate $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
+ \While{$\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(m)} - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(m-1)} > \epsilon$}{
+  Set $\tilde{\tau}^{(m)}$ using \ref{mf9}.\;
+  Set $\tilde{\beta}^{(m)}$ using \ref{mf10}.\;
+  Evaluate $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(m)}$.\;
+  Set $m = m + 1$.\;
+ }
+ \caption{Coordinate Ascent for MFVB}
+  \label{alg:algorithm1}
+\end{algorithm}
+\vspace{2mm}
+Data was simulated according to the Gaussian model introduced in Section (\ref{sec:MCMC}) with $T = 100, \mu = 2$ and $\sigma^2 = 1$.
 The hyperparameters were set to $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$.
@@
 \includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{norminvg}
-\caption{Marginal posterior densities for the model described in Section (\ref{sec:MCMC}).
-The true density is in black, the Monte Carlo sampler is red and the MFVB approximation is blue.
-Both MCMC and MFVB fit this density extremely well.}
+\caption{Marginal posterior densities for the model described in Section (\ref{sec:MCMC}). The true density is in black, the Monte Carlo sampler is red and the MFVB approximation is blue.}
 \label{fig:norminvg}
@@
 
-The set of mean field equations are only obtainable in the event that a prior that is conjugate to the model likelihood is used, restricting MFVB to exponential family models.
+The set of mean field equations are only obtainable in the event that a prior that is conjugate to the model likelihood is used, restricting MFVB to exponential family models \citep{Beal2003}.
 In this case the optimal distributional family $q_i$ is the same as the prior, analagous to the relationship between a prior-posterior conjugate pair.
-In the event that a non-conjugate prior is used, the form of the distribution in (\ref{mf2}) is unrecognisable, a further approximation may be used by substituting in another distribution to replace the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006} or a Delta Method inspired transform \citet{Wang2013}; though the only requirement is that the substitute distribution has moments that are tractable functions of its parameters to be used for other parameters in (\ref{mf2}).
-This secondary level of approximations is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable full conditional distributions.
-
-\vspace{5mm}
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density}
- \KwResult{Mean Field Approximation}
- Use (\ref{mf2}) to match $q(\mu|\boldsymbol{\lambda}_1)$ to a Gaussian distributon and $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to an Inverse Gamma distribution\;
- Anaytically derive the set of mean field equations in (\ref{mf9}) - (\ref{mf10})\;
- Initialise $\boldsymbol{\lambda}^{(1)}$ randomly.\;
- Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
- Set $\tilde{\gamma}$ using \ref{mf5}.\;
- Set $\tilde{\alpha}$ using \ref{mf7}.\;
- \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
-  Set $\tilde{\tau}^{(m)}$ using \ref{mf9}.\;
-  Set $\tilde{\beta}^{(m)}$ using \ref{mf10}.\;
-  Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$.\;
-  Set $m = m + 1$.\;
- }
- \caption{Coordinate Ascent for MFVB}
-  \label{alg:algorithm1}
-\end{algorithm}
+In the event that a non-conjugate prior is used, the form of the distribution in (\ref{mf2}) is unrecognisable, a further approximation may be used by substituting in another distribution to replace the problematic $q_i(\theta_i |\boldsymbol{\lambda_i}).$ A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006} or a Delta Method inspired transform \citep{Wang2013}.
+This secondary level of approximation is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable full conditional distributions.
 
@@
 
-The requirement for an exponential family model without introducing further approximations, as well as restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications.
-\citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, which can maximise the ELBO for non-factorisable functional forms of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
+The requirement for an exponential family model without introducing further approximations, as well as restricting the approximating distribution to the factorisable family may be unsatisfactory in many applications.
+\citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, which selects optimal $\boldsymbol{\lambda}$ for a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from the class of distributions for which the score function can be evaluated, and the order of differentation of the ELBO with respect to $\boldsymbol{\lambda}$ and integration with respect to $\boldsymbol{\theta}$ are interchangable.
+The only model restriction is that the log likelihood for $y_{1:T}$ is able to be evaluated.
 We refer to the result as Stochastic Variational Bayes (SVB).
@@
 
-SVB can find only the optimal values for $\boldsymbol{\lambda}$ for a user specified distributional family $q$, and to run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO.
-$q$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\boldsymbol{\lambda}$ and integration with respect to $\boldsymbol{\theta}$ are interchangable, and SVB can only be applied to models where the log likelihood for $y_{1:T}$ is able to be evaluated.
-Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by
+\citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
 \begin{equation}
@@
 
-\citet{Duchi2011} introduced the AdaGrad algorithm which can be implemented within SVB to control $\rho^{(m)}$.
+\citet{Kingma2014} introduces the re-parameterisation trick, where an auxillary noise variable $\boldsymbol{\epsilon}$ and differentiable transformation $f(\cdot, \cdot)$ are introduced such that $\boldsymbol{\theta} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}$.
+Examples include a location scale transformation from a standard normal $\epsilon$ or an inverse-CDF transformation from a uniform$(0, 1) \epsilon$.
+In this case (\ref{SVB2}) can be replaced with a lower variance estimator,
+\begin{equation}
+\label{reparameterise}
+\nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \left( \ln (p(y_{1:T}, f(\boldsymbol{\epsilon}_s, \boldsymbol{\lambda}^{(m)})) - \ln(q(f(\boldsymbol{\epsilon}_s, \boldsymbol{\lambda}^{(m)})) \right)
+\end{equation}
+where $s = 1, \dots, S$ now indicates simulations from $q(\boldsymbol{\epsilon})$.
+This transformation removes the restriction that $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ belong to a distribution where the score function can be evaluated.
+
+\citet{Duchi2011} developed the AdaGrad algorithm which can be implemented within SVB to control $\rho^{(m)}$.
 AdaGrad allows each $\lambda_i$ to have an independent $\rho^{(m)}_i$ that is inversely proportional to the gradient, so $\lambda$ takes bigger steps in flat regions and smaller steps in steep regions.
@@
 
-The resulting Stochastic Gradient Ascent algorithm proceeds below in Algorithm (\ref{alg:algorithm2}) with a $p$ dimensional $\boldsymbol{\lambda}$ vector.
+The resulting Stochastic Gradient Ascent algorithm without reparameterisation proceeds below in Algorithm (\ref{alg:algorithm2}) with a $p$ dimensional $\boldsymbol{\lambda}$ vector.
 
@@
  Initialise $\boldsymbol{\lambda}^{(1)}$\;
- Evaluate $\mathcal{L}(q, y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
- \While{$\mathcal{L}(q, y_{1:T})^{(m)} - \mathcal{L}(q, y_{1:T})^{(m-1)} > \epsilon$}{
+ Evaluate $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(1)}$ using $\boldsymbol{\lambda}^{(1)}$.\;
+ \While{$\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(m)} - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(m-1)} > \epsilon$}{
   Simulate $\boldsymbol{\theta}^s$ for $s = 1, \dots S$ from $q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})$\;
@@
   Update $\boldsymbol{\lambda}^{(m+1)}$ from (\ref{SVB1})\;
-  Evaluate $\mathcal{L}(q, y_{1:T})^{(m)}$\;
+  Evaluate $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})^{(m)}$\;
   Set $m = m + 1$\;
@@
 
-\subsection{Vine Copulas} \label{sec:vines}
+\subsubsection{Selection of the Approximating Distribution} \label{sec:selection}
 
-Armed with an algorithm to find the Variational Bayes optimal parameters $\boldsymbol{\lambda}$ for a distribution $q$, there is a requirement to have a distribution $q$ that is itself optimal.
-Copulas are a flexible tool for constructing these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions.
+Armed with an algorithm to find the Variational Bayes optimal parameters $\boldsymbol{\lambda}$ for a distribution $q$, there is a requirement to select a distribution $q$ that well approximates the true posterior.
+If reparameterised aproximating distribution is used to calculate the gradients in (\ref{reparameterise}) a function $f(\cdot, \cdot)$ that satisfies
+
+\begin{equation}
+\label{optimalf}
+f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}) = p^{-1}(\boldsymbol{\epsilon})
+\end{equation}
+
+almost everywhere, where $\epsilon \sim U(0, 1)$ and $p^{-1}$ denotes the inverse CDF of the true posterior, will yeild $q(\boldsymbol{\theta} | \boldsymbol{\lambda} = p(\boldsymbol{\theta} | y_{1:T}$ and thus minimise the KL divergence.
+Attempts to model a suitably flexible function $f(\cdot, \cdot)$ include heirarchical models \citep{Ranganath2016a}, neural networks (\citealp{Kingma2014}; \citealp{Rezende2014}; \citealp{Rezende2015}), and Gaussian processes \citep{Tran2016}.
+
+These approaches introduce a large number of parameters that would considerably slow computation of updates, so we follow \citet{Tran2015} and use copulas to flexibly model approximations to the true posterior distribution.
+The copula modelling class provides a flexible way to construct these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions.
 \citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,
@@
 
-This idea is extended by the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas, see \citet{Joe2014} and references within.
+The flexibility of a copula is extended by the vine copula, a structure that permits the factorisation of a high-dimensional copula into a set of bivariate copulas, see \citet{Joe2014} and references within.
 A vine copula is represented by a series of trees, where the top-most tree is a graph that contains each variable as a node.
 The edges in each tree then form the nodes in the tree below, which are connected by an edge only if the associated edges in the previous tree shared the same node.
-This sequence is continued or a $k$ dimensional $\boldsymbol{\theta}$ until the $(k-1)'th$ tree is reached that contains only one edge.
+This sequence is continued for a $k$ dimensional $\boldsymbol{\theta}$ until the $(k-1)'th$ tree is reached that contains only one edge.
 Each of the $k(k-1)/2$ edges represents a bivariate copula between the two unique parameters in the connected nodes, conditioned on every parameter appearing in both nodes.
 Refer to Figure (\ref{fig:vinecop}) for an example.
-
+\vspace{3mm}
 \begin{figure}[h]
@@
 \includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
+\vspace{2mm}
 \caption{An example of a Vine Copula over four variables.
@@
 
-The resulting vine copula greatly increases the amount of flexibility available to model dependency in a distribution, but the cardinality of the set of possible vine structures and choices of family for each bivariate copula is exponential in the number of parameters, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
-\citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\boldsymbol{\theta}$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.
+The flexibility in selecting a factorisation and set of bivariate copula familys in vine copula greatly increases the ability to model dependency in a distribution, but the cardinality of the set of possible vine structures and choices of family for each bivariate copula grows factorially with the number of parameters, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
 
@@
 In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date.
-The approach used in this thesis is to run an MCMC algorithm, thin the resulting draws until they are effectively independent, choose an optimal family for each marginal distribution by AIC, then run (Vine copula algorithm here) to infer an optimal vine copula.
-This results in an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that can then be used in an SVB algorithm to update $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ as more observations become available.
-MCMC can be ran simultaenously and the resulting sample can be used to check if the distribution $q$ should be changed.
+The approach used in this thesis is to run an MCMC algorithm, thin the resulting draws until they are effectively independent, choose an optimal family for each marginal distribution by AIC, then run Di{\ss}mann's algorithm to infer a vine copula structure that fits the data well.
+This vine copula is used to construct an approximating distribution $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that can then be used in an SVB algorithm to update $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ as more observations become available.
+MCMC can be ran simultaneously and the resulting sample can be used to check if the distribution $q$ should be changed.
 
@@
 
+In this section the techniques described throughout Section \ref{sec:BayesInf} are applied to a simple, but widely used, time-series model. 
 Consider the second order auto-regressive time series model, denoted by AR(2) and described by 
@@
 where 
-\begin{equation}
-\Sigma = \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right],\
-\end{equation}
-and $\gamma(k)$ is the $k-th$ autocorrelation for an AR(2) model.
+\begin{align}
+\Sigma &= \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right] \nonumber \\
+\gamma(0) &= \sigma^2 \frac{1-\phi_2}{(1+\phi_2)((1-\phi_2)^2 - \phi_1^2)} \nonumber \\
+\gamma(1) &= \sigma^2 \frac{\phi_1}{(1+\phi_2)((1-\phi_2)^2 - \phi_1^2)} \nonumber.
+\end{align}
+
 The likelihood of parameters $\boldsymbol{\theta} = (\phi_1, \phi_2, \sigma^2)'$, that is, the density of the observations $y_{1:T}$, given $\boldsymbol{\theta}$, viewed as a function of $\boldsymbol{\theta}$, is given by
@@
 This prior has positive mass if and only if a pair $(\phi_1, \phi_2)$ is within the AR(2) stationary region.
+Note that this prior is not conjugate due to the form of $\phi_1$ and $\phi_2$ in $\Sigma$, and thus MFVB cannot be applied.
 Upon examination of the form of the product of the likelihood function in (\ref{likelihood}) and the prior density in (\ref{ar2prior}) it can be seen that the full conditional distribution $p(\sigma^2 | y, \phi_1, \phi_2)$ is an Inverse Gamma distribution, while the conditionals for both $\phi$ parameters are intractable but can be sampled using a Metropolis-Hastings step with truncated bivariate normal random walk proposal distribution.
@@
 
-Figure \ref{fig:VBfit} shows the fit of Stochastic Variational Bayes (blue) to the MCMC (red) samples.
-SVB appears to have selected values for $\boldsymbol{\lambda}$ that closely match the MCMC posterior distribution.
+Figure \ref{fig:VBfit} displays the marginal and bivariate posterior distributions that result from each of the Stochastic Variational Bayes (blue) and the MCMC (red) approaches.
+In the case of MCMC, a kernel estimate obtained from the relevant posterior sample is displayed, whereas for SVB the form of the relevant vine copula marginal determines the shape of the marginal posterior.
+As is clear from the figure, SVB appears to have selected values for $\boldsymbol{\lambda}$ that closely match the MCMC posterior distribution.
+SVB required approximately one hundredth of the computation time required for MCMC.
 
@@
 \includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{VBfit.png}
-\caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model.
-SVB resulted in an almost identical posterior to MCMC, and required approximately one hundredth of the runtime of MCMC.}
+\caption{The fit of an SVB algorithm (blue) compared to MCMC (red) for the AR(2) model.}
 \label{fig:VBfit}
@@
 Half hourly electricity load data for Victoria, Australia is provided by the Australian Energy Market Operator (AEMO) for the 2011-2015 period, and is plotted in Figure \ref{fig:loadplot}.
-The time series displays strong seasonal characteristics, with yearly, day of week, and time of day patterns.
+The time series displays strong seasonal characteristics, with time of year, day of week, and time of day patterns.
+This yearly pattern is characterised by a low median load with high volatility in summer and a more consistent curve in winter.
 Electricity load is strongly dependent on temperature, with high load experienced on both very cold and very hot days.
@@
 
-The electricity market in Victoria operates by market participants sending electricity bid and offers for each five minute period to AEMO, which then determines the dispatch price and load for that five minute period.
-At the end of each half-hour block the previous six dispatch prices are averaged to obtain the final spot price that participants trade at.
+The wholesale electricity market in Victoria operates by market participants sending electricity bid and offers for each five minute period to AEMO, which then matches electricity generators with retailers to determine the dispatch load and associated price.
+At the end of each half-hour block the previous six five-minute dispatch prices are averaged to obtain the final spot price that participants trade at.
 Market participants are allowed to revise bids and offers at any point before dispatch.
@@
 These characteristics of the electricity market leads to a large amount of volatility in prices demonstrated by Figure \ref{fig:rrpplot}.
-The seasonal effects of the load data are also present in prices, but these effects are dominated by irregular spikes as high as \$9,974/mWh, compared to a four year mean of \$42/mWh.
+The seasonal effects of the load data are also present in prices, but these effects are dominated by irregular spikes as high as \$9,974/mWh, compared to a four year mean price of \$42/mWh.
 
@@
 \includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{load_timeseries.png}
-\caption{Half hourly load data measured in megawatt hours for Victoria from January 1 2011 to December 31 2015.
-The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent curve in winter.}
+\caption{Half hourly load data measured in megawatt hours for Victoria from January 1 2011 to December 31 2015.}
 \label{fig:loadplot}
@@
 \includegraphics[width=\linewidth,height=\textheight,keepaspectratio]{RRP_timeseries.png}
-\caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015.
-While the price has time of week and time of day seasonal effects these features are dominated by irregular spikes.}
+\caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015.}
 \label{fig:rrpplot}
@@
 y_t &= \textbf{x}' \textbf{b}_t + e_t \label{ss_me} \\
-\textbf{b}_t &= \textbf{T} \textbf{b}_{t-1} + \tilde{\boldsymbol{\theta}} e_t \label{ss_te}.
+\textbf{b}_t &= T \textbf{b}_{t-1} + \tilde{\boldsymbol{\theta}} e_{t-1} \label{ss_te}.
 \end{align}
@@
 
-\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\boldsymbol{\theta}$ in (\ref{exp-sm-marginal}), however the problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible, and the $T$ in the exponent makes most evalutions of $p(\boldsymbol{\theta}  | y_{1:T})$ computationally zero when $T$ is large, as is the case with high frequency electricity load data.
-An efficient MCMC algorithm that scales to large datasets with models using multiple seasonality parameters will be required to sample the full posterior distribution $p(\boldsymbol{\theta} , \sigma^2, b_0 | y_{1:T}$ and obtain an approximating distribution $q(\boldsymbol{\theta}, \sigma^2, b_0 | \lambda)$ with a reasonable goodness of fit.
+\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\boldsymbol{\theta}$ in (\ref{exp-sm-marginal}), however the problems with using this method in our model are two-fold: repeating a three dimensional numerical integration each time a new data point is observed within the five minute window is not feasible, and the $T$ in the exponent makes most evaluations of $p(\boldsymbol{\theta}  | y_{1:T})$ computationally zero when $T$ is large, as is the case with high frequency electricity load data.
+An efficient MCMC algorithm that scales to large datasets with models using multiple seasonality parameters will be required to sample the full posterior distribution $p(\boldsymbol{\theta} , \sigma^2, b_0 | y_{1:T})$ and obtain an approximating distribution $q(\boldsymbol{\theta}, \sigma^2, b_0 | \boldsymbol{\lambda})$ with a reasonable goodness of fit.
 From this point, Stochastic Variational Bayes will be find an approximation to the updated posterior distribution as more data is observed, and provide the predictive density $q(y_{T+J+1} | y_{T+J})$.
@@
 In this case, the dependence for the new latent state must be extrapolated from the copula structure used for previous latent states, which may be sub-optimal.
-The robustness of SVB approximations to mis-specified paramater dependence should be investigated.
+The robustness of SVB approximations to choices of $q$ that badly fit the true posterior should be investigated.
+Development of the theory of choosing the distribution $q$ is currently extremely active, a methodology that relies on the form of the model likelihood and prior rather than MCMC samples to determine the form of a well-fitting approximation would be of interest.
 
-Finally, Variational Bayes only becomes a viable method for density forecasts if the statistical error associated with using an approximate distribution is less than the statistical error associated with forecasting based on an MCMC generated posterior distribution that is not conditioned on the most recent data.
-The investigation of this trade-off is of central importance to the thesis.
+Finally, there is a trade-off between the predictive accuracy of MCMC and the ability to update Variational Bayes posterior distributions.
+The statistical error associated with using an approximate distribution has not been investigated in the literature and we shall contribute by comparing this error in forecasting with VB to the statistical error associated with forecasting based on an MCMC generated posterior distribution that is not conditioned on the most recent data.
 
@@
   & to develop, and adapt them to the benefit my research. \\  \hline
-  May 2019 & $\cdot$ Put seperate parts together and polish the final thesis for submission. \\
+  May 2019 & $\cdot$ Put separate parts together and polish the final thesis for submission. \\
 \hline
@@
 \grid
+\grid
+\grid
+\grid
+\grid

### 2017-02-27 Confirmation/confirmation_example.R

@@ -25,7 +25,7 @@ gamma0 = sigma2 * (1-phi2) / ((1+phi2)*((1-phi2)^2 - phi1^2))
 gamma1 = sigma2 * phi1 / ((1+phi2)*((1-phi2)^2 - phi1^2))
 Sigma = matrix(c(gamma0, gamma1, gamma1, gamma0), 2)
 
-y[1:2] = rmvnorm(1, c(0, 0), Sigma)
+y = rmvnorm(1000, c(0, 0), Sigma)
 for(t in 3:(T+J)){
   y[t] = phi1*y[t-1] + phi2*y[t-2] + rnorm(1, sd = sqrt(sigma2))
 }

### 2017-03-01 Confirmation/confirmation_V2.tex
Started work on slides

@@
 For either implementation we emphasise that there are two seperate tasks: Choose a functional form for the approximating distribution, and then optimise the parameters of this distribution to minimise a well defined divergence between the approximation and the true posterior.
+A recent review of Variational Bayes is provided by \citet{Blei2017}.
 
@@
 \label{mf3}
-q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T + \tau) \mathbb{E}_{q(\sigma^2)}(\sigma^{-2})}{2} \left( \mu - \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau} \right)^2 \right\}
+q(\mu | \boldsymbol{\lambda}_1) \propto \exp \left\{ \frac{-(T + \tau) \mathbb{E}_{q(\sigma^2)}[\sigma^{-2}]}{2} \left( \mu - \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau} \right)^2 \right\}
 \end{equation}
@@
 q(\sigma^2 | \boldsymbol{\lambda}_2) &\propto \sigma^{-2((T+1)/2 + \alpha + 1)} \nonumber \\
-\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)}(\mu^2) - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
+\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)}[\mu^2] - 2 \mathbb{E}_{q(\mu)}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau \right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}.
 \end{align}
@@
 \tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
-\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}(\sigma^{-2})\right )^{-1} \label{mf6} \\
+\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}[\sigma^{-2}]\right )^{-1} \label{mf6} \\
 \tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
-\tilde{\beta} &= \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)} - 2 \mathbb{E}_{q(\mu)}(\mu)(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right). \label{mf8}
+\tilde{\beta} &= \beta + 1/2\left((\tau + T)\mathbb{E}_{q(\mu)}[\mu^2 ]- 2 \mathbb{E}_{q(\mu)}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right). \label{mf8}
 \end{align}.
@@
 \tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
-\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
+\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
 \end{align}
@@
 almost everywhere, where $\epsilon \sim U(0, 1)$ and $p^{-1}$ denotes the inverse CDF of the true posterior, will yeild $q(\boldsymbol{\theta} | \boldsymbol{\lambda} = p(\boldsymbol{\theta} | y_{1:T}$ and thus minimise the KL divergence.
-Attempts to model a suitably flexible function $f(\cdot, \cdot)$ include heirarchical models \citep{Ranganath2016a}, neural networks (\citealp{Kingma2014}; \citealp{Rezende2014}; \citealp{Rezende2015}), and Gaussian processes \citep{Tran2016}.
+Attempts to model a suitably flexible function $f(\cdot, \cdot)$ include heirarchical models \citep{Ranganath2016a}, neural networks (\citealp{Kingma2014}; \citealp{Rezende2014}; \citealp{Rezende2015}; \citealp{Kingma2016}), and Gaussian processes \citep{Tran2016}.
 
@@
 \grid
+\grid
+\grid

### 2017-03-01 Confirmation/presentations.Rnw
Started work on slides

\documentclass{beamer}

\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}

\usepackage{alltt}
\usepackage{natbib}
\usepackage{array}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{dsfont}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\usepackage{tikz}
\usepackage{amssymb}
\usepackage{bm}

\usetheme{Warsaw}
\defbeamertemplate*{footline}{shadow theme}
{%
  \leavevmode%
  \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fil,rightskip=.3cm]{author in head/foot}%
    \usebeamerfont{author in head/foot}\insertframenumber\,/\,\inserttotalframenumber\hfill\insertshortauthor
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
    \usebeamerfont{title in head/foot}\insertshorttitle%
  \end{beamercolorbox}}%
  \vskip0pt%
}

\setbeamertemplate{headline}{%
\leavevmode%
  \hbox{%
    \begin{beamercolorbox}[wd=\paperwidth,ht=2.5ex,dp=1.125ex]{palette quaternary}%
    \insertsectionnavigationhorizontal{\paperwidth}{\hskip0pt plus1filll}{\hskip0pt plus1filll}
    \end{beamercolorbox}%
  }
}

\title[Real-Time Variational Density Forecasts]{Real-Time Variational Density Forecasts}
\author[Nathaniel Tomasetti]{Nathaniel Tomasetti}
\date{ }
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\begin{frame}
\titlepage
\centering
Supervised by Catherine Forbes and Anastasios Panagiotelis
\end{frame}

\begin{frame}
\section{Motivation}
Motivation - Electricity 
Motivation - MCMC/MFVB/SVB
Motivation - Contributions
\end{frame}

\begin{frame}
\section{Markov Chain Monte Carlo}
\frametitle{Markov Chain Monte Carlo}

\begin{itemize}
\item Markov Chain Monte Carlo (MCMC) is a common method to sample from the true posterior.
\item Adding more iterations allows any statistical error to become arbitarily small.
\item Computation for complex models can be extremely slow
\item A predictive density for the next data point may not be available before it is observed
\end{itemize}

\end{frame}
\begin{frame}
\frametitle{Deriving Gibbs Full Conditional Distributions}

Let $y_i \sim \mathcal{N}(\mu, \sigma^2) \mbox{ for } i = 1, 2, \dots, N$, \newline $p(\sigma^2) \sim IG(\mbox{shape} = \alpha, \mbox{scale} = \beta)$ and $p(\mu | \sigma) \sim \mathcal{N}(\gamma, \sigma^2/\tau)$.

Gibbs MCMC requires the distribution $p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y})$.

\begin{itemize}
\item Step One: Write the full log joint distribution
\end{itemize}

\begin{align}
\log(p(\textbf{y}, \mu, \sigma^2)) &=  -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} \nonumber \\
&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2}  + c \nonumber 
\end{align}

\end{frame}
\begin{frame}
\frametitle{Deriving Gibbs Full Conditional Distributions}
\begin{itemize}
\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, \textbf{y}))$
\begin{equation}
\log(p(\mu | \textbf{y}, \sigma^2)) =  - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c \nonumber
\end{equation}
\item Step Three: Condition on the previous draw of $\sigma^2$ to form $\log(p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y}))$
\begin{equation}
\log(p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y})) = - \frac{\sum_{i}(y_i - \mu_{(i)})^2}{2\sigma_{(i-1)}^2} - \frac{\tau(\mu_{(i)} - \gamma)^2}{2\sigma_{(i-1)}^2}  + c \nonumber
\end{equation}
\item Step Four: Manipulate equation until a distribution kernel is recognised. If unrecognised, add a Metropolis-Hastings step.\
\item Step Five: Repeat for $p(\sigma^2_{(i)} | \mu_{(i-1)}, \textbf{y})$
\end{itemize}
\end{frame}

\begin{frame}
\section{Variational Bayes}
\frametitle{Variational Bayes}
\begin{itemize}
\item Key Idea: Replace the intractable $p(\boldsymbol{\theta} | \textbf{y})$ with a tractable approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
\item Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions that minimises a well defined divergence between $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ and $p(\boldsymbol{\theta} | \textbf{y})$.

\item The classic choice is the Kullback-Leibler divergence \citep{Kullback1951} from $q$ to $p$,
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | \textbf{y})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |\textbf{y})}\right) d\boldsymbol{\theta}.
\end{equation}
This is non-negative and equals zero if and only if $p(\boldsymbol{\theta} | \textbf{y}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}.
\end{itemize}
\end{frame}
\begin{frame}
\frametitle{The Evidence Lower Bound}
Note that
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} |\textbf{y})] = \ln(p(\textbf{y})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), \textbf{y}),
\end{equation}
where
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), \textbf{y}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(\textbf{y},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
\end{equation}
is refered to as the Evidence Lower Bound (ELBO).
\begin{itemize}
\item Maximising the ELBO with respect to $q \iff$ Minimising the KL Divergence with respect to $q$.
\item There are two tasks: 
  \begin{enumerate}
  \item Choose a functional form for $q$. 
  \item Optimise parameters $\boldsymbol{\lambda}$. 
  \end{enumerate}
\end{itemize}

\end{frame}

\begin{frame}
\subsection{Mean Field Variational Bayes}
\frametitle{Mean Field Variational Bayes}
Background - MFVB
\end{frame}

\begin{frame}
\frametitle{Deriving $q$ distributions}

Steps to find $q(\mu | \boldsymbol{\lambda}_1)$ are similar to finding $p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y})$

\begin{itemize}
\item Step One: Write the full log joint distribution
\begin{align}
\log(p(\textbf{y}, \mu, \sigma^2)) &= -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} + c \nonumber \\
&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2} + c \nonumber 
\end{align}
\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, \textbf{y}))$
\begin{equation}
\log(p(\mu | \textbf{y}, \sigma^2)) = - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c \nonumber
\end{equation}
\end{itemize}

\end{frame}
\begin{frame}
\frametitle{Deriving $q$ distributions}

\begin{itemize}
\item Step Three: Take the expectation with respect to $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to form $\log(q(\mu | \boldsymbol{\lambda}_1))$
\begin{equation}
\log(q(\mu | \boldsymbol{\lambda}_1)) = - 1/2 \mathbb{E}_{q(\sigma^2)} [\sigma^{-2}] \left( \sum_{i}(y_i - \mu)^2 + \tau(\mu - \gamma)^2 \right)  + c \nonumber
\end{equation}
\item Step Four: Manipulate equation until a distribution kernel is recognised. If unrecognised, add a secondary approximation.
\item Step Five: Repeat for $q(\sigma^2 | \boldsymbol{\lambda}_2)$.
\item In this case $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim IG(\tilde{\alpha}, \tilde{\beta})$.
\end{itemize}

\end{frame}
\begin{frame}
\frametitle{Deriving $q$ distributions}

$\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$, where
\begin{align}
\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{mf5} \\ 
\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q(\sigma^2)}[\sigma^{-2}]\right )^{-1} \label{mf6} \\
\tilde{\alpha} &= (T+1)/2 + \alpha  \label{mf7} \\
\tilde{\beta} &= \beta + 1/2\bigg((\tau + T)\mathbb{E}_{q(\mu)}[\mu^2 ]  \nonumber \\ 
&- 2 \mathbb{E}_{q(\mu)}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \bigg). \label{mf8}
\end{align}.

\end{frame}
\begin{frame}
\frametitle{Deriving $q$ distributions}

Substitute in expectations to (\ref{mf6}) and (\ref{mf8}):

\begin{align}
\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
\end{align}

The circular dependence in (\ref{mf9}) and (\ref{mf10}) and can be resolved with a coordinate ascent algorithm.

\end{frame}
\begin{frame}
\frametitle{Fit to simulated data}

$100$ data points were simualted with $\mu = 2$ and $\sigma^2 = 1$ and both MCMC and MFVB are compared. Hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$.

\begin{figure}[h]
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{norminvg}
\caption{Marginal posterior densities for the Normal Inverse Gamma model. The true density is in black, the Monte Carlo sampler is red and the MFVB approximation is blue. MFVB converged in same time as required to take three MCMC samples of each parameter.}
\label{fig:norminvg}
\end{figure}

\end{frame}
\begin{frame}
\subsection{Stochastic Variational Bayes}
\frametitle{Stochastic Variational Bayes}

\begin{itemize}
\item \citet{Paisley2012} and \citet{Ranganath2014} adapted a gradient ascent algorithm for use in VB.
\item The class of approximating distributions that can be used is broadened.
\item Models no longer have to belong to the exponential family.
\item We require Monte-Carlo estimates of 
\begin{align}
\label{SVB}
\nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) &\approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda})] \nonumber \\
&\times \big(\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}))\big) 
\end{align}
where $\boldsymbol{\theta}_s$ is simulated from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ for $s = 1, 2, \dots, S$. 
\end{itemize}

\end{frame}
\begin{frame}
\frametitle{Gradient Ascent}

Make steps of the form
\begin{equation}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} f(\lambda^{(m)})
\end{equation}
until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$.

\begin{figure}[h]
\centering
<<gradascent, out.width='0.4\\textwidth', error=FALSE, message=FALSE, warning=FALSE, echo=FALSE>>=
x = rep(0, 20)
x[1] = -2.5
for(i in 1:19){
  deriv = -x[i]*dnorm(x[i])
  x[i+1] = x[i] + 23 * deriv / i
}
y2 = dnorm(x)
library(ggplot2)
xsup = seq(-3, 1.5, 0.01)
y = dnorm(xsup, 0, 1)

ggplot() + geom_line(aes(xsup, y)) + geom_path(aes(x, y2), colour = "red") + labs(y = "Gaussian Density", x = expression(lambda)) + theme_bw()
@
\caption{Red steps show the path a variable $\lambda^{(m)}$ took while maximising a standard normal distribution}
\label{fig:GradAscent}
\end{figure}

\end{frame}

\begin{frame}
\frametitle{The reparameterisation trick}
\end{frame}

\begin{frame}
\section{Copula Modelling}
Background - Copulas
\end{frame}

\begin{frame}
Application - AR2
Application - Updating rule

Empirical - Background
Empirical - Model
\end{frame}

\begin{frame}
\bibliographystyle{asa}
\bibliography{references}
\end{frame}

\end{document}

### 2017-03-01 Confirmation/presentations.Rnw
updates

@@
 \usepackage{bm}
+\setbeamertemplate{caption}[numbered]
 
@@
 Motivation - Electricity 
-Motivation - MCMC/MFVB/SVB
-Motivation - Contributions
+\end{frame}
+
+\begin{frame}
+\section{Bayesian Computation Techniques}
+A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
+\begin{itemize}
+\item Converges to the true posterior.
+\item Can be very slow, and is not parallelisable.
+\item Scales badly to large datasets.
+\end{itemize}
+Interest is developing in Bayesian approximations, such as Variational Bayes (VB). VB:
+\begin{itemize}
+\item Is typically orders of magnitude faster than MCMC.
+\item Scales very well to large datasets with subsampling and parallelisation.
+\item Has no theoretical guarantees to be close to the truth.
+\end{itemize}
 \end{frame}
@@
 \frametitle{Markov Chain Monte Carlo}
+\begin{itemize}
+\item A Gibbs MCMC sampler iteratively samples the full conditional distributions:
+\begin{align}
+&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
+&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
+&\vdots \nonumber \\
+&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
+\end{align}
+\item Requires convergence, which can be extremely slow.
+\item A predictive density for the next data point may not be available before it is observed.
+\end{itemize}
+\end{frame}
 
-\begin{itemize}
-\item Markov Chain Monte Carlo (MCMC) is a common method to sample from the true posterior.
-\item Adding more iterations allows any statistical error to become arbitarily small.
-\item Computation for complex models can be extremely slow
-\item A predictive density for the next data point may not be available before it is observed
-\end{itemize}
-
-\end{frame}
-\begin{frame}
-\frametitle{Deriving Gibbs Full Conditional Distributions}
-
-Let $y_i \sim \mathcal{N}(\mu, \sigma^2) \mbox{ for } i = 1, 2, \dots, N$, \newline $p(\sigma^2) \sim IG(\mbox{shape} = \alpha, \mbox{scale} = \beta)$ and $p(\mu | \sigma) \sim \mathcal{N}(\gamma, \sigma^2/\tau)$.
-
-Gibbs MCMC requires the distribution $p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y})$.
-
-\begin{itemize}
-\item Step One: Write the full log joint distribution
-\end{itemize}
-
-\begin{align}
-\log(p(\textbf{y}, \mu, \sigma^2)) &=  -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} \nonumber \\
-&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2}  + c \nonumber 
-\end{align}
-
-\end{frame}
 \begin{frame}
@@
 \begin{itemize}
-\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, \textbf{y}))$
+\item Let $y_t \overset{iid}{\sim} \mathcal{N}(\mu, \sigma^2) \mbox{ for } t = 1, 2, \dots, T$. 
+\item Use priors $p(\sigma^2) \sim IG(\mbox{shape} = \alpha, \mbox{scale} = \beta), p(\mu | \sigma) \sim \mathcal{N}(\gamma, \sigma^2/\tau)$.
+\item Gibbs MCMC requires the distribution $p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T})$. 
+\item Step One: Write the full log joint distribution,
+\end{itemize}
+\begin{align}
+\log(p(y_{1:T}, \mu, \sigma^2)) &=  -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} \nonumber \\
+&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2}  + c. \nonumber 
+\end{align}
+\end{frame}
+
+\begin{frame}
+\frametitle{Deriving Gibbs Full Conditional Distributions}
+\begin{itemize}
+\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, y_{1:T}))$,
 \begin{equation}
-\log(p(\mu | \textbf{y}, \sigma^2)) =  - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c \nonumber
+\log(p(\mu | y_{1:T}, \sigma^2)) =  - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c. \nonumber
 \end{equation}
-\item Step Three: Condition on the previous draw of $\sigma^2$ to form $\log(p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y}))$
+\item Step Three: Condition on the previous draw of $\sigma^2$ to form $\log(p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T}))$,
 \begin{equation}
-\log(p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y})) = - \frac{\sum_{i}(y_i - \mu_{(i)})^2}{2\sigma_{(i-1)}^2} - \frac{\tau(\mu_{(i)} - \gamma)^2}{2\sigma_{(i-1)}^2}  + c \nonumber
+\log(p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T})) = - \frac{\sum_{t}(y_t - \mu_{(i)})^2}{2\sigma_{(i)}^2} - \frac{\tau(\mu_{(i)} - \gamma)^2}{2\sigma_{(i)}^2}  + c. \nonumber
 \end{equation}
 \item Step Four: Manipulate equation until a distribution kernel is recognised. If unrecognised, add a Metropolis-Hastings step.\
-\item Step Five: Repeat for $p(\sigma^2_{(i)} | \mu_{(i-1)}, \textbf{y})$
+\item Step Five: Repeat for $p(\sigma^2_{(i)} | \mu_{(i-1)}, y_{1:T}).$
 \end{itemize}
@@
 \begin{itemize}
-\item Key Idea: Replace the intractable $p(\boldsymbol{\theta} | \textbf{y})$ with a tractable approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
-\item Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions that minimises a well defined divergence between $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ and $p(\boldsymbol{\theta} | \textbf{y})$.
-
-\item The classic choice is the Kullback-Leibler divergence \citep{Kullback1951} from $q$ to $p$,
+\item Key Idea: Replace the intractable $p(\boldsymbol{\theta} | y_{1:T})$ with a tractable approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
+\item How: Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions that minimises a well defined divergence between $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ and $p(\boldsymbol{\theta} | y_{1:T})$.
+\item The classic choice is the Kullback-Leibler divergence from $q$ to $p$,
 \begin{equation}
 \label{KL-def}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | \textbf{y})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |\textbf{y})}\right) d\boldsymbol{\theta}.
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
 \end{equation}
-This is non-negative and equals zero if and only if $p(\boldsymbol{\theta} | \textbf{y}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}.
+This is non-negative and equals zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere.
 \end{itemize}
 \end{frame}
+
 \begin{frame}
 \frametitle{The Evidence Lower Bound}
-Note that
+Rewrite (\ref{KL-def}) as
 \begin{equation}
 \label{KL-ELBO}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} |\textbf{y})] = \ln(p(\textbf{y})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), \textbf{y}),
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} |y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}),
 \end{equation}
@@
 \label{ELBO}
-\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), \textbf{y}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(\textbf{y},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
+\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
 \end{equation}
@@
   \item Choose a functional form for $q$. 
-  \item Optimise parameters $\boldsymbol{\lambda}$. 
+  \item Optimise auxillary parameter vector $\boldsymbol{\lambda}$. 
   \end{enumerate}
 \end{itemize}
-
 \end{frame}
@@
 \begin{frame}
-\subsection{Mean Field Variational Bayes}
 \frametitle{Mean Field Variational Bayes}
-Background - MFVB
+\begin{itemize}
+\item Early Variational Bayes implementation pioneered by \citet{Jordan1999} and \citet{Ghahramani2000}.
+\item Restricts $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the factorisable family,
+\begin{equation}
+\label{mf1}
+q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i=1}^k q_i(\theta_i | \boldsymbol{\lambda_i}).
+\end{equation}
+\item Restricts the model to the exponential family with conjugate priors.
+\item Provides the optimal distribution for each $q_i(\theta_i | \boldsymbol{\lambda_i})$.
+\item Maximise the ELBO by setting each factor to
+\begin{equation}
+\label{mf2}
+q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{\prod_{j \neq i} q_i(\theta_i | \boldsymbol{\lambda_i})} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
+\end{equation}
+\end{itemize}
 \end{frame}
@@
 \frametitle{Deriving $q$ distributions}
-
-Steps to find $q(\mu | \boldsymbol{\lambda}_1)$ are similar to finding $p(\mu_{(i)} | \sigma^2_{(i-1)}, \textbf{y})$
-
+(\ref{mf2}) is applied to the Normal-Inverse Gamma model to find $q(\mu | \boldsymbol{\lambda}_1)$.
 \begin{itemize}
-\item Step One: Write the full log joint distribution
+\item Step One: Write the full log joint distribution,
 \begin{align}
-\log(p(\textbf{y}, \mu, \sigma^2)) &= -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} + c \nonumber \\
-&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2} + c \nonumber 
+\log(p(y_{1:T}, \mu, \sigma^2)) &= -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} \nonumber \\
+&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2} + c. \nonumber 
 \end{align}
-\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, \textbf{y}))$
+\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, y_{1:T})),$
 \begin{equation}
-\log(p(\mu | \textbf{y}, \sigma^2)) = - \frac{\sum_{i}(y_i - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c \nonumber
+\log(p(\mu | y_{1:T}, \sigma^2)) = - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c. \nonumber
 \end{equation}
 \end{itemize}
+\end{frame}
 
-\end{frame}
 \begin{frame}
 \frametitle{Deriving $q$ distributions}
-
 \begin{itemize}
-\item Step Three: Take the expectation with respect to $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to form $\log(q(\mu | \boldsymbol{\lambda}_1))$
+\item Step Three: Take the expectation with respect to $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to form $\log(q(\mu | \boldsymbol{\lambda}_1)),$
 \begin{equation}
-\log(q(\mu | \boldsymbol{\lambda}_1)) = - 1/2 \mathbb{E}_{q(\sigma^2)} [\sigma^{-2}] \left( \sum_{i}(y_i - \mu)^2 + \tau(\mu - \gamma)^2 \right)  + c \nonumber
+\log(q(\mu | \boldsymbol{\lambda}_1)) = - 1/2 \mathbb{E}_{q(\sigma^2)} [\sigma^{-2}] \left( \sum_{t}(y_t - \mu)^2 + \tau(\mu - \gamma)^2 \right)  + c. \nonumber
 \end{equation}
@@
 \end{itemize}
+\end{frame}
 
-\end{frame}
 \begin{frame}
 \frametitle{Deriving $q$ distributions}
-
 $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$, where
@@
 &- 2 \mathbb{E}_{q(\mu)}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \bigg). \label{mf8}
-\end{align}.
+\end{align}
+\end{frame}
 
-\end{frame}
 \begin{frame}
 \frametitle{Deriving $q$ distributions}
-
 Substitute in expectations to (\ref{mf6}) and (\ref{mf8}):
-
 \begin{align}
@@
 \end{align}
+The circular dependence in (\ref{mf9}) and (\ref{mf10}) and can be resolved with a coordinate ascent algorithm.
+\end{frame}
 
-The circular dependence in (\ref{mf9}) and (\ref{mf10}) and can be resolved with a coordinate ascent algorithm.
-
-\end{frame}
 \begin{frame}
 \frametitle{Fit to simulated data}
-
-$100$ data points were simualted with $\mu = 2$ and $\sigma^2 = 1$ and both MCMC and MFVB are compared. Hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$.
-
+\begin{itemize}
+\item Marginal posterior densities for the Normal Inverse Gamma model with $N = 100, \mu = 2$ and $\sigma^2 = 1$; hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. 
+\end{itemize}
 \begin{figure}[h]
@@
 \includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{norminvg}
-\caption{Marginal posterior densities for the Normal Inverse Gamma model.
-The true density is in black, the Monte Carlo sampler is red and the MFVB approximation is blue.
-MFVB converged in same time as required to take three MCMC samples of each parameter.}
+\caption{The analytical true density (black), the Monte Carlo (red) and the MFVB approximation (blue). MFVB converged in same time as required to take three MCMC samples of each parameter.}
 \label{fig:norminvg}
 \end{figure}
+\end{frame}
 
-\end{frame}
 \begin{frame}
-\subsection{Stochastic Variational Bayes}
 \frametitle{Stochastic Variational Bayes}
-
 \begin{itemize}
-\item \citet{Paisley2012} and \citet{Ranganath2014} adapted a gradient ascent algorithm for use in VB.
-\item The class of approximating distributions that can be used is broadened.
-\item Models no longer have to belong to the exponential family.
-\item We require Monte-Carlo estimates of 
+\item \citet{Paisley2012} and \citet{Ranganath2014} adapted a gradient ascent algorithm to maximise the ELBO.
+\item Gradient ascent returns the optimal $\boldsymbol{\lambda}$ for a specified functional form $q$.
+\item Only requires Monte-Carlo estimates of $\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$,
 \begin{align}
 \label{SVB}
-\nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) &\approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda})] \nonumber \\
+\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) &\approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda})] \nonumber \\
 &\times \big(\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}))\big) 
@@
 \end{itemize}
+\end{frame}
 
-\end{frame}
 \begin{frame}
 \frametitle{Gradient Ascent}
-
 Make steps of the form
@@
 until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$.
-
 \begin{figure}[h]
@@
 <<gradascent, out.width='0.4\\textwidth', error=FALSE, message=FALSE, warning=FALSE, echo=FALSE>>=
+library(ggplot2)
 x = rep(0, 20)
@@
 @
-\caption{Red steps show the path a variable $\lambda^{(m)}$ took while maximising a standard normal distribution}
+\caption{Red steps show the path a variable $\lambda^{(m)}$ took while finding the maximum of a standard normal distribution.}
 \label{fig:GradAscent}
 \end{figure}
-
 \end{frame}
@@
 \begin{frame}
-\frametitle{The reparameterisation trick}
+\section{Copulas}
+\frametitle{Vine Copulas}
+\begin{itemize}
+\item Vine copulas allow distributions to be constructed where a flexible dependence structure is fit independently to the margins.
+\vspace{3mm}
+\item Problem: There are $O(k!)$ ways to construct a vine copula for a $k-$dimensional parameter vector $\boldsymbol{\theta}$.
+\item Solution: Di{\ss}mann's algorithm generates a vine copula that fits a sample of data reasonably well.
+\item Implementation: Run MCMC and use Di{\ss}mann's algorithm to find a good vine copula to use as $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
+Run SVB to update this approximation as more data comes in.
+\end{itemize}
 \end{frame}
@@
 \begin{frame}
-\section{Copula Modelling}
-Background - Copulas
+\frametitle{AR2 Model Fit}
+\begin{itemize}
+\item Simulate an AR(2) with: $T = 150, \phi_1 = 0.7, \phi_2 = 0.2$, and $\sigma^2 = 1$.
+\item Run Metropolis-Hastings-within-Gibbs MCMC, select $q$ marginals by BIC from draws, apply Di{\ss}mann's algorithm.
+\item Result: $q(\phi_1, \phi_2) = \mbox{BVN}(\boldsymbol{\mu}, \Sigma)$ and $q(\sigma^2) = IG(\alpha, \beta)$. 
+\end{itemize}
+\begin{figure}[h]
+\centering
+\includegraphics[scale = 0.4]{VBfit.png}
+\caption{The fit of an SVB algorithm (blue) compared to MCMC (red) for the AR(2) model. SVB converged in same time as required to take 3000 MCMC samples of each parameter.}
+\label{fig:VBfit}
+\end{figure}
 \end{frame}
@@
 \begin{frame}
-Application - AR2
-Application - Updating rule
+\section{Empirical Application}
+\frametitle{Electricity Load Forecasting}
+\end{frame}
 
-Empirical - Background
-Empirical - Model
+\begin{frame}
+\frametitle{Exponential Smoothing}
+\end{frame}
+
+\begin{frame}
+\section{Discussion}
+\frametitle{Discussion}
+\begin{itemize}
+\item Updating is worthwhile if the new data brings significant information on the parameters.
+State-space models are of interest as observation specific latent variables are more influenced by new data than global parameters.
+\item Existing online updating algorithms assume data is $iid$. We aim to develop an updating algorithm for dependent data.
+\item Statistical properties of the approximation are unknown and depend on the divergence measure used.
+\item The accuracy/computation time trade-off between MCMC and VB has not been thoroughly investigated.
+\item Our time-series context allows MCMC draws to guide the construction of $q$. We have no information on latent variables associated with new data and we must devise a way to include these in $q$.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Timeline}
+\begin{itemize}
+\item Mid 2017: Complete electricity load forecasting application.
+\item End of 2017: Develop dependent online updating, apply to Stochastic Volatiltiy model with dependent latent variables.
+\item Compare approximations to the exact posterior to understand the efficiency trade-off.
+\item Mid 2018: Investigate statistical properties of other divergence measures such as Stein's discrepancy.
+\item End of 2018: Allow extra time as the literature is evolving rapidly and new research opportunities constantly appear.
+\item Mid 2019: Finalise thesis for submission.
+\end{itemize}
 \end{frame}

### 2017-03-02 Confirmation/presentations.Rnw
not too productive today

@@
 \section{Motivation}
-Motivation - Electricity 
+\frametitle{Motivation}
+\begin{itemize}
+\item There is an increasing demand for forecasts to be of the entire predictive density of the variable of interest.
+\item Many time series of interest can operate on a very short time-frame.
+\item We aim to provide density forecasts for these high frequency time-series by updating posterior and predictive density estimates as each data point is observed.
+\item The algorithm use should scale to multivariate observations driven by complex models with large datasets.
+\end{itemize}
 \end{frame}
@@
 \begin{frame}
-\section{Bayesian Computation Techniques}
+\frametitle{Bayesian Computation Techniques}
 A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
 \begin{itemize}
-\item Converges to the true posterior.
-\item Can be very slow, and is not parallelisable.
+\item Converges in distribution to the posterior density with enough iterations.
+\item Can be very slow, and is hard to parallelise.
 \item Scales badly to large datasets.
@@
 \item Scales very well to large datasets with subsampling and parallelisation.
-\item Has no theoretical guarantees to be close to the truth.
+\item Has no theoretical guarantee to converge to the posterior.
 \end{itemize}
@@
 \end{equation}
-This is non-negative and equals zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere.
+\item (\ref{KL-def}) is non-negative and equals zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere.
 \end{itemize}
@@
 \frametitle{The Evidence Lower Bound}
-Rewrite (\ref{KL-def}) as
+\begin{itemize}
+\item Rewrite (\ref{KL-def}) as
 \begin{equation}
@@
 \label{ELBO}
-\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
+\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \mathbb{E}_{q} \left( \ln (p(y_{1:T},\boldsymbol{\theta})) - \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) \right).
 \end{equation}
-is refered to as the Evidence Lower Bound (ELBO).
-\begin{itemize}
+\item $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is refered to as the Evidence Lower Bound (ELBO).
 \item Maximising the ELBO with respect to $q \iff$ Minimising the KL Divergence with respect to $q$.
@@
 \frametitle{Deriving $q$ distributions}
-(\ref{mf2}) is applied to the Normal-Inverse Gamma model to find $q(\mu | \boldsymbol{\lambda}_1)$.
 \begin{itemize}
+\item (\ref{mf2}) is applied to the Normal-Inverse Gamma model to find $q(\mu | \boldsymbol{\lambda}_1)$.
 \item Step One: Write the full log joint distribution,
@@
 \frametitle{Deriving $q$ distributions}
-$\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$, where
+\begin{itemize}
+\item The auxillary parameter vector is $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$, which maximises the ELBO when
 \begin{align}
@@
 \end{align}
+\end{itemize}
 \end{frame}
@@
 \frametitle{Deriving $q$ distributions}
-Substitute in expectations to (\ref{mf6}) and (\ref{mf8}):
+\begin{itemize}
+\item Substitute in expectations to (\ref{mf6}) and (\ref{mf8}):
 \begin{align}
@@
 \end{align}
-The circular dependence in (\ref{mf9}) and (\ref{mf10}) and can be resolved with a coordinate ascent algorithm.
+\item The circular dependence in (\ref{mf9}) and (\ref{mf10}) and can be resolved with a coordinate ascent algorithm.
+\end{itemize}
 \end{frame}
@@
 \frametitle{Gradient Ascent}
-Make steps of the form
+\begin{itemize}
+\item Make steps of the form
 \begin{equation}
@@
 until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$.
+\end{itemize}
 \begin{figure}[h]
@@
 \frametitle{Electricity Load Forecasting}
+Motivation - Electricity 
 \end{frame}
@@
 \frametitle{Exponential Smoothing}
+\begin{itemize}
+\item \citet{Taylor2008} provides evidence that a double-seasonal Holt-Winters exponential smoothing model performs better than common alternatives for very short term electricity load forecasting.
+\item \citet{Forbes2000} provides a Bayesian approach to estimating exponential smoothing models by Monte Carlo simulation.
+\item We slightly modify the Holt-Winters model to work within the Bayesian approach.
+\item Applying SVB will allow electricity load density forecasts to be made at a very short time-frame.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Double Seasonal Holt-Winters Exponential Smoothing}
+\begin{itemize}
+\item We use the modified double seasonal Holt-Winters exponential smoothing model
+\begin{align}
+y_t &= l_{t-1} - \sum_{i = 1}^{m_1 - 1}d_{t-i} - \sum_{i = 1}^{m_2 - 1}w_{t-i} + e_t \label{ds-hw-rp1} \\
+l_t &= l_{t-1} + \alpha e_t \label{ds-hw-rp2} \\
+d_t &= - \sum_{i = 1}^{m_1 - 1}d_{t-i} + \delta e_t \label{ds-hw-rp3} \\
+w_t &= - \sum_{i = 1}^{m_2 - 1}w_{t-i} + \omega e_t \label{ds-hw-rp4}.
+\end{align}
+where $m_1$ is the length of the daily cycle, and $m_2$ is the length of the weekly cycle.
+\item (\ref{ds-hw-rp3}) and (\ref{ds-hw-rp4}) restrict the seasonality but allow us to avoid inverting a singular matrix in the Bayesian approach.
+\end{itemize}
 \end{frame}
@@
 \item Existing online updating algorithms assume data is $iid$. We aim to develop an updating algorithm for dependent data.
+\item Our time-series context allows MCMC draws to guide the construction of $q$. We have no information on latent variables associated with new data and we must devise a way to include these in $q$.
 \item Statistical properties of the approximation are unknown and depend on the divergence measure used.
 \item The accuracy/computation time trade-off between MCMC and VB has not been thoroughly investigated.
-\item Our time-series context allows MCMC draws to guide the construction of $q$. We have no information on latent variables associated with new data and we must devise a way to include these in $q$.
 \end{itemize}

### 2017-03-03 Confirmation/confirmation_V2.tex
updates
more presentation changes

@@
 
-Mean Field Variational Bayes (MFVB) has origins in physics (\citealp{Amari1985}; \citealp{Chandler1987}) and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Ghahramani2000}.
+Mean Field Variational Bayes (MFVB) has origins in physics \citep{Peterson1987} and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Ghahramani2000}.
 MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
@@
 \label{SVB2}
-\nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)}))] (\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)})))
+\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)}))] (\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)})))
 \end{equation}
@@
 \label{reparameterise}
-\nabla_{\lambda}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \left( \ln (p(y_{1:T}, f(\boldsymbol{\epsilon}_s, \boldsymbol{\lambda}^{(m)})) - \ln(q(f(\boldsymbol{\epsilon}_s, \boldsymbol{\lambda}^{(m)})) \right)
+\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \left( \ln (p(y_{1:T}, f(\boldsymbol{\epsilon}_s, \boldsymbol{\lambda}^{(m)})) - \ln(q(f(\boldsymbol{\epsilon}_s, \boldsymbol{\lambda}^{(m)})) \right)
 \end{equation}

### 2017-03-03 Confirmation/presentations.tex
more presentation changes

--- 2017-03-02 Confirmation/presentations.Rnw
@@
-\documentclass{beamer}
+\documentclass{beamer}\usepackage[]{graphicx}\usepackage[]{color}
+%% maxwidth is the original width if it is less than linewidth
+%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\makeatletter
+\def\maxwidth{ %
+  \ifdim\Gin@nat@width>\linewidth
+    \linewidth
+  \else
+    \Gin@nat@width
+  \fi
+}
+\makeatother
+
+\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
+\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
+\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
+\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
+\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
+\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
+\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
+\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
+\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
+\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
+
+\usepackage{framed}
+\makeatletter
+\newenvironment{kframe}{%
+ \def\at@end@of@kframe{}%
+ \ifinner\ifhmode%
+  \def\at@end@of@kframe{\end{minipage}}%
+  \begin{minipage}{\columnwidth}%
+ \fi\fi%
+ \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
+ \colorbox{shadecolor}{##1}\hskip-\fboxsep
+     % There is no \\@totalrightmargin, so:
+     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
+ \MakeFramed {\advance\hsize-\width
+   \@totalleftmargin\z@ \linewidth\hsize
+   \@setminipage}}%
+ {\par\unskip\endMakeFramed%
+ \at@end@of@kframe}
+\makeatother
+
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
+\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX
+
+\usepackage{alltt}
 
@@
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \begin{document}
@@
 \begin{itemize}
-\item There is an increasing demand for forecasts to be of the entire predictive density of the variable of interest.
+\item There is an increasing demand for forecasts of the entire predictive density of the variable of interest.
 \item Many time series of interest can operate on a very short time-frame.
 \item We aim to provide density forecasts for these high frequency time-series by updating posterior and predictive density estimates as each data point is observed.
-\item The algorithm use should scale to multivariate observations driven by complex models with large datasets.
+\item The algorithm used should scale to multivariate observations driven by complex models with large datasets.
 \end{itemize}
@@
 \end{equation}
-until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$.
+until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$, with $\rho^{(m)}, m = 1, 2, \dots$ that satisfies the \citet{Robbins1951} conditions.
 \end{itemize}
@@
 \centering
-<<gradascent, out.width='0.4\\textwidth', error=FALSE, message=FALSE, warning=FALSE, echo=FALSE>>=
-library(ggplot2)
-x = rep(0, 20)
-x[1] = -2.5
-for(i in 1:19){
-  deriv = -x[i]*dnorm(x[i])
-  x[i+1] = x[i] + 23 * deriv / i
-}
-y2 = dnorm(x)
-library(ggplot2)
-xsup = seq(-3, 1.5, 0.01)
-y = dnorm(xsup, 0, 1)
-
-ggplot() + geom_line(aes(xsup, y)) + geom_path(aes(x, y2), colour = "red") + labs(y = "Gaussian Density", x = expression(lambda)) + theme_bw()
-@
+\includegraphics[scale = 0.4]{gradientascent}
 \caption{Red steps show the path a variable $\lambda^{(m)}$ took while finding the maximum of a standard normal distribution.}
@@
 \frametitle{Electricity Load Forecasting}
-Motivation - Electricity 
+\begin{itemize}
+\item Victorian electricity prices are extremely volatile, and are updated every five minutes.
+\item Prices spike electricity load crosses a particular threshold.
+\item If we can provide density forecasts for load, we can predict how likely it is for prices to spike in the next period.
+\begin{figure}[h]
+\centering
+\includegraphics[scale = 0.3]{RRP_timeseries.png}
+\caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015.}
+\label{fig:rrpplot}
+\end{figure}
+\end{itemize}
 \end{frame}

### 2017-03-05 Confirmation/presentations.Rnw
presentation tweaks

@@
 \begin{itemize}
-\item Converges in distribution to the posterior density with enough iterations.
-\item Can be very slow, and is hard to parallelise.
+\item Converges to the posterior density.
+\item Can be very slow.
 \item Scales badly to large datasets.
@@
 \begin{itemize}
-\item Is typically orders of magnitude faster than MCMC.
-\item Scales very well to large datasets with subsampling and parallelisation.
-\item Has no theoretical guarantee to converge to the posterior.
+\item Significantly faster than MCMC.
+\item Scales to large datasets.
+\item No theoretical guarantee to converge to the posterior.
 \end{itemize}
@@
 \end{align}
-\item Requires convergence, which can be extremely slow.
+\item Requires many samples to converge, which can be extremely slow.
 \item A predictive density for the next data point may not be available before it is observed.

### 2017-03-05 Confirmation/presentations.tex
presentation tweaks

@@
 
+\newcounter{saveenumi}
+\newcommand{\seti}{\setcounter{saveenumi}{\value{enumi}}}
+\newcommand{\conti}{\setcounter{enumi}{\value{saveenumi}}}
+\resetcounteronoverlays{saveenumi}
+
 \title[Real-Time Variational Density Forecasts]{Real-Time Variational Density Forecasts}
@@
 Supervised by Catherine Forbes and Anastasios Panagiotelis
+\end{frame}
+
+\begin{frame}
+\tableofcontents
 \end{frame}
@@
 \frametitle{Bayesian Computation Techniques}
-A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
 \begin{itemize}
-\item Converges in distribution to the posterior density with enough iterations.
-\item Can be very slow, and is hard to parallelise.
+\item A popular technique is to sample the posterior distribution with Markov Chain Monte Carlo (MCMC). MCMC:
+\begin{enumerate}
+\item Converges to the posterior density.
+\item Can be very slow.
 \item Scales badly to large datasets.
-\end{itemize}
-Interest is developing in Bayesian approximations, such as Variational Bayes (VB). VB:
-\begin{itemize}
-\item Is typically orders of magnitude faster than MCMC.
-\item Scales very well to large datasets with subsampling and parallelisation.
+\end{enumerate}
+\item Interest is rapidly developing in approximations, such as Variational Bayes (VB). VB:
+\begin{enumerate}
+\item Is significantly faster than MCMC.
+\item Scales to large datasets.
 \item Has no theoretical guarantee to converge to the posterior.
+\end{enumerate}
 \end{itemize}
@@
 \end{align}
-\item Requires convergence, which can be extremely slow.
-\item A predictive density for the next data point may not be available before it is observed.
+\item MCMC often requires a large number of samples.
+\item The predictive density may be unavailable before more data is observed.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Deriving Gibbs Full Conditional Distributions}
+\frametitle{Gibbs Full Conditional Distributions}
 \begin{itemize}
-\item Let $y_t \overset{iid}{\sim} \mathcal{N}(\mu, \sigma^2) \mbox{ for } t = 1, 2, \dots, T$. 
-\item Use priors $p(\sigma^2) \sim IG(\mbox{shape} = \alpha, \mbox{scale} = \beta), p(\mu | \sigma) \sim \mathcal{N}(\gamma, \sigma^2/\tau)$.
-\item Gibbs MCMC requires the distribution $p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T})$. 
-\item Step One: Write the full log joint distribution,
-\end{itemize}
-\begin{align}
-\log(p(y_{1:T}, \mu, \sigma^2)) &=  -(T/2 + \alpha + 2)\log(\sigma^2) - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} \nonumber \\
-&- \frac{\tau(\mu - \gamma)^2}{2\sigma^2} - \frac{\beta}{\sigma^2}  + c. \nonumber 
-\end{align}
-\end{frame}
-
-\begin{frame}
-\frametitle{Deriving Gibbs Full Conditional Distributions}
-\begin{itemize}
-\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, y_{1:T}))$,
-\begin{equation}
-\log(p(\mu | y_{1:T}, \sigma^2)) =  - \frac{\sum_{t}(y_t - \mu)^2}{2\sigma^2} - \frac{\tau(\mu - \gamma)^2}{2\sigma^2} + c. \nonumber
-\end{equation}
-\item Step Three: Condition on the previous draw of $\sigma^2$ to form $\log(p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T}))$,
-\begin{equation}
-\log(p(\mu_{(i)} | \sigma^2_{(i)}, y_{1:T})) = - \frac{\sum_{t}(y_t - \mu_{(i)})^2}{2\sigma_{(i)}^2} - \frac{\tau(\mu_{(i)} - \gamma)^2}{2\sigma_{(i)}^2}  + c. \nonumber
-\end{equation}
-\item Step Four: Manipulate equation until a distribution kernel is recognised. If unrecognised, add a Metropolis-Hastings step.\
-\item Step Five: Repeat for $p(\sigma^2_{(i)} | \mu_{(i-1)}, y_{1:T}).$
+\item Let $y_t \overset{iid}{\sim} \mathcal{N}\left(\mu, \sigma^2\right) \mbox{ for } t = 1, 2, \dots, T$. 
+\item Use priors
+\begin{enumerate}
+\item $\mu | \sigma \sim \mathcal{N}\left(\gamma, \sigma^2/\tau\right)$
+\item $\sigma^2 \sim IG\left(\mbox{shape} = \alpha, \mbox{scale} = \beta\right)$.
+\end{enumerate}
+\vspace{2mm}
+\item Gibbs MCMC samples the full conditional distributions,
+\begin{enumerate}
+\item $\mu | \sigma^2, y_{1:T} \sim \mathcal{N}\left(\bar{\gamma}, \sigma^2 / \bar{\tau}\right)$
+\item $\sigma^2 | \mu, y_{1:T} \sim IG \left(\mbox{shape} = \bar{\alpha}, \mbox{scale} = \bar{\beta} \right)$
+\end{enumerate}
 \end{itemize}
@@
 \item Key Idea: Replace the intractable $p(\boldsymbol{\theta} | y_{1:T})$ with a tractable approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
-\item How: Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions that minimises a well defined divergence between $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ and $p(\boldsymbol{\theta} | y_{1:T})$.
-\item The classic choice is the Kullback-Leibler divergence from $q$ to $p$,
+\item How: Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions is closest to $p(\boldsymbol{\theta} | y_{1:T})$ by some metric. 
+\item Typically minimise the Kullback-Leibler divergence from $q$ to $p$,
+
 \begin{equation}
 \label{KL-def}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}.
+KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \mathbb{E}_q  \left[ \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) \right]
 \end{equation}
-\item (\ref{KL-def}) is non-negative and equals zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere.
 \end{itemize}
@@
 \label{ELBO}
-\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \mathbb{E}_{q} \left( \ln (p(y_{1:T},\boldsymbol{\theta})) - \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) \right).
+\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \mathbb{E}_{q} \left[ \ln (p(y_{1:T},\boldsymbol{\theta})) - \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) \right].
 \end{equation}
-\item $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is refered to as the Evidence Lower Bound (ELBO).
-\item Maximising the ELBO with respect to $q \iff$ Minimising the KL Divergence with respect to $q$.
+\item $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is called the Evidence Lower Bound (ELBO).
+\item Maximising the ELBO $\iff$ Minimising the KL Divergence.
 \item There are two tasks: 
@@
 \begin{frame}
-\frametitle{Mean Field Variational Bayes}
+\frametitle{Mean Field Variational Bayes - MFVB}
 \begin{itemize}
-\item Early Variational Bayes implementation pioneered by \citet{Jordan1999} and \citet{Ghahramani2000}.
+\item Early Variational Bayes implementation pioneered by Jordan \textit{et al.} (1999) and Ghahramani and Beal (2000).
 \item Restricts $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the factorisable family,
@@
 \end{equation}
-\item Restricts the model to the exponential family with conjugate priors.
-\item Provides the optimal distribution for each $q_i(\theta_i | \boldsymbol{\lambda_i})$.
-\item Maximise the ELBO by setting each factor to
+\item Maximise the ELBO by setting each marginal to
 \begin{equation}
 \label{mf2}
-q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{\prod_{j \neq i} q_i(\theta_i | \boldsymbol{\lambda_i})} [\ln(p(y_{1:T},\boldsymbol{\theta}))]).
+q_i(\theta_i |\boldsymbol{\lambda_i}) \propto\exp( \mathbb{E}_{q_{j \neq i}}[\ln(p(y_{1:T},\boldsymbol{\theta}))]).
 \end{equation}
+\item Solution only in closed form for exponential family models with limited prior choice.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Deriving $q$ distributions}
+\frametitle{$q$ distributions for the Normal-Inverse Gamma model}
 \begin{itemize}
 \item (\ref{mf2}) is applied to the Normal-Inverse Gamma model to find $q(\mu | \boldsymbol{\lambda}_1)$.
-\item Step One: Write the full log joint distribution,
+\end{itemize}
+\begin{enumerate}
+\item Write 
 \begin{align}
@@
 \end{align}
-\item Step Two: Ignore components that do not depend on $\mu$ to form $\log(p(\mu | \sigma^2, y_{1:T})),$
+\item Ignore components without $\mu$
 \begin{equation}
@@
 \end{equation}
-\end{itemize}
+\seti
+\end{enumerate}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Deriving $q$ distributions}
-\begin{itemize}
-\item Step Three: Take the expectation with respect to $q(\sigma^2 | \boldsymbol{\lambda}_2)$ to form $\log(q(\mu | \boldsymbol{\lambda}_1)),$
+\frametitle{$q$ distributions for the Normal-Inverse Gamma model}
+\begin{enumerate}
+\conti
+\item Take the expectation with respect to $q(\sigma^2 | \boldsymbol{\lambda}_2)$
 \begin{equation}
@@
 \end{equation}
-\item Step Four: Manipulate equation until a distribution kernel is recognised. If unrecognised, add a secondary approximation.
-\item Step Five: Repeat for $q(\sigma^2 | \boldsymbol{\lambda}_2)$.
+\item Manipulate until a distribution kernel is recognised. 
+\item If unrecognised, add a secondary approximation.
+\item Repeat for $q(\sigma^2 | \boldsymbol{\lambda}_2)$.
+\end{enumerate}
+\begin{itemize}
 \item In this case $q(\mu | \boldsymbol{\lambda}_1) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q(\sigma^2 | \boldsymbol{\lambda}_2) \sim IG(\tilde{\alpha}, \tilde{\beta})$.
@@
 \begin{frame}
-\frametitle{Deriving $q$ distributions}
+\frametitle{$\boldsymbol{\lambda}$ for the Normal-Inverse Gamma model}
 \begin{itemize}
-\item The auxillary parameter vector is $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$, which maximises the ELBO when
+\item The ELBO maximising choice of $\boldsymbol{\lambda} = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})$ is:
 \begin{align}
@@
 \begin{frame}
-\frametitle{Deriving $q$ distributions}
+\frametitle{$\boldsymbol{\lambda}$ for the Normal-Inverse Gamma model}
 \begin{itemize}
@@
 \tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
-\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}(\sum_{t=1}^{T}y_t + \tau) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
+\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau \right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
 \end{align}
-\item The circular dependence in (\ref{mf9}) and (\ref{mf10}) and can be resolved with a coordinate ascent algorithm.
+\item Circular dependence in (\ref{mf9}) and (\ref{mf10}) 
+\item Requires a coordinate ascent algorithm.
 \end{itemize}
@@
 \begin{itemize}
-\item Marginal posterior densities for the Normal Inverse Gamma model with $N = 100, \mu = 2$ and $\sigma^2 = 1$; hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. 
+\item Marginal posterior densities for the Normal Inverse Gamma model with $T = 100, \mu = 2$ and $\sigma^2 = 1$.
+\item Hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. 
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Stochastic Variational Bayes}
+\frametitle{Stochastic Variational Bayes - SVB}
 \begin{itemize}
-\item \citet{Paisley2012} and \citet{Ranganath2014} adapted a gradient ascent algorithm to maximise the ELBO.
+\item Paisley \textit{et al.} (2012) and Ranganath \textit{et al.} (2014) adapted a gradient ascent algorithm to maximise the ELBO.
 \item Gradient ascent returns the optimal $\boldsymbol{\lambda}$ for a specified functional form $q$.
@@
 \begin{equation}
-\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} f(\lambda^{(m)})
+\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} f(\lambda) \bigg|_{\lambda = \lambda^{(m)}}
 \end{equation}
-until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$, with $\rho^{(m)}, m = 1, 2, \dots$ that satisfies the \citet{Robbins1951} conditions.
+until $| f(\lambda^{(m+1)}) - f(\lambda^{(m)}) | < \epsilon$, with $\rho^{(m)}, m = 1, 2, \dots$ that satisfies the Robbins and Munro (1951) conditions.
 \end{itemize}
@@
 \begin{frame}
+\frametitle{Choosing the $q$ distribution}
+\begin{itemize}
+\item Finding a good functional form for $q$ is an open problem.
+\item Approaches include heirarchical distributions and neural network transformations.
+\item $q$ on needs to be parsimonous to scale.
+\item We augment $q$ with a copula to model dependence separately to the margins.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
 \section{Copulas}
@@
 \begin{itemize}
-\item Vine copulas allow distributions to be constructed where a flexible dependence structure is fit independently to the margins.
+\item Vine copulas increase the flexiblity of copula modelling.
 \vspace{3mm}
@@
 \begin{frame}
-\frametitle{Exponential Smoothing}
-\begin{itemize}
-\item \citet{Taylor2008} provides evidence that a double-seasonal Holt-Winters exponential smoothing model performs better than common alternatives for very short term electricity load forecasting.
-\item \citet{Forbes2000} provides a Bayesian approach to estimating exponential smoothing models by Monte Carlo simulation.
-\item We slightly modify the Holt-Winters model to work within the Bayesian approach.
-\item Applying SVB will allow electricity load density forecasts to be made at a very short time-frame.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Double Seasonal Holt-Winters Exponential Smoothing}
-\begin{itemize}
-\item We use the modified double seasonal Holt-Winters exponential smoothing model
-\begin{align}
-y_t &= l_{t-1} - \sum_{i = 1}^{m_1 - 1}d_{t-i} - \sum_{i = 1}^{m_2 - 1}w_{t-i} + e_t \label{ds-hw-rp1} \\
-l_t &= l_{t-1} + \alpha e_t \label{ds-hw-rp2} \\
-d_t &= - \sum_{i = 1}^{m_1 - 1}d_{t-i} + \delta e_t \label{ds-hw-rp3} \\
-w_t &= - \sum_{i = 1}^{m_2 - 1}w_{t-i} + \omega e_t \label{ds-hw-rp4}.
-\end{align}
-where $m_1$ is the length of the daily cycle, and $m_2$ is the length of the weekly cycle.
-\item (\ref{ds-hw-rp3}) and (\ref{ds-hw-rp4}) restrict the seasonality but allow us to avoid inverting a singular matrix in the Bayesian approach.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
 \section{Discussion}
@@
 
-\begin{frame}
-\bibliographystyle{asa}
-\bibliography{references}
-\end{frame}
-
 \end{document}

### 2017-03-13 Confirmation/confirmation_V2.tex
Python work

@@
 
-Electricity prices have been subject to an extensive forecasting literature and are known to be extremely volatile.
+Electricity prices have been the subject of an extensive forecasting literature and are known to be extremely volatile.
 For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of observations less than \$100 per mWh, however on several occasions the price spiked to greater than $10,000\%$ of the mean value.
@@
 To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it accounts for both parameter and model uncertainty in a relatively straightforward manner, which can be difficult in frequentist methods.
-See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general.
+See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for a discussion of density forecasting in general.
 
@@
 While it is often possible to demonstrate theoretical convergence of MCMC algorithms, computation may be slow for even moderately complex models.
-This is problematic when attempting to forecast electricity load in Victoria, as realizations occur at five minute intervals, the predictive density for the next period's electricity load may not be available before it is observed.
+This is problematic when attempting to forecast electricity load in Victoria, as realizations occur at five minute intervals, making it difficult to compute the predictive density for the next period's electricity load within the time frame.
 
@@
 Mean Field Variational Bayes (see \citet{Bishop2006} for an overview) can be applied to exponetial family models with conjugate priors, and restricts the class of approximating distributions to the factorisable family.
-MFVB provides both the optimal form of the approximating distribution within this class, and a computationally simple coordinate ascent algorithm is available to optimise parameters.
+MFVB provides both the optimal form of the approximating distribution within this class, and a computationally simple coordinate ascent algorithm to optimise parameters.
 Stochastic Variational Bayes (see \citet{Paisley2012} and \citet{Ranganath2014}) lifts both the restrictions for an exponential family model and the use of an approximating distribution that is factorisable.
@@
 We explore the use of a copula to augment the approximating distribution with a flexible dependence structure, which has seen success in \citet{Tran2015} and \citet{Ng2016}.
-The time series context of our implementation implies that MCMC samples for out of date samples are available, and we emphasise the use of this information to guide the construction of a vine copula so that the approximation has a dependence structure similiar to the true posterior.
+The time series context of our implementation implies that MCMC samples for out of date samples are available, and we exploit this information in the construction of a vine copula so that the approximation has a dependence structure similiar to the true posterior.
 
@@
 \end{equation}
-The KL divergence is a non-negative, asymetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$  that will be equal to zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}.
+The KL divergence is a non-negative, asymmetric measure of the discrepancy between $p(\boldsymbol{\theta} | y_{1:T})$ and $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$  that will be equal to zero if and only if $p(\boldsymbol{\theta} | y_{1:T}) = q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ almost everywhere \citep{Bishop2006}.
 Note that $KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})]$ can be expressed as
@@
 Maximising the ELBO is much more convenient than minimising the KL Divergence, as will be shown in the following sections in the context of the two major implementations of Variational Bayes we consider, Mean Field Variational Bayes and Stochastic Variational Bayes.
-Each of these implementations take advantage of the functional form of the ELBO, and offer computationally efficient algorithms to find a $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.
+Each of these implementations takes advantage of the functional form of the ELBO, and offers a computationally efficient algorithm to for finding $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is optimal within a particular class of distributions.
 The statistical properties of $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ in MFVB and SVB are largely driven by the choice of the KL divergence from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to $p(\boldsymbol{\theta} | y_{1:T})$ as an error function, and there is significant interest in other error functions, such as the KL divergence from $p(\boldsymbol{\theta} | y_{1:T})$ to $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ \citep{Minka2001} and Stein's discrepency (\citealp{Liu2016}; \citealp{Ranganath2016b}).
@@
 
-Mean Field Variational Bayes (MFVB) has origins in physics \citep{Peterson1987} and has since developed a large literature in machine learning, notably developed in \citet{Jordan1999} and \citet{Ghahramani2000}.
+Mean Field Variational Bayes (MFVB) has its origin in physics \citep{Peterson1987} and has since developed a large literature in machine learning, notably \citet{Jordan1999} and \citet{Ghahramani2000}.
 MFVB restricts the class of distributions from which to select $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ to the set of factorisable distributions,
@@
 
-Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars.
-Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ which may have different dimension to $\theta_i$.
+The $k$ components $\theta_i$ may be a vector, but in most implementations are scalars.
+Each $\theta_i$ has an associated vector $\boldsymbol{\lambda_i}$ that may have a different dimension from $\theta_i$.
 $\boldsymbol{\lambda_i}$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\boldsymbol{\lambda_i})$.
@@
 
-The expectations in (\ref{mf5}) - (\ref{mf8}) are available in a closed form for these distributions, and these are substituted in to derive the set of mean field equations:
+The expectations in (\ref{mf5}) - (\ref{mf8}) are available in a closed form for these distributions, and these are substituted into (\ref{mf6}) and (\ref{mf8}) to yield the set of mean field equations:
 \begin{align}
 \tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{mf9} \\
-\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right)\label{mf10}
+\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right).\label{mf10}
 \end{align}
-$\tilde{\gamma}$ and $\tilde{\alpha}$ are omitted from the mean field equations as their values do not depend on other $\boldsymbol{\lambda}$ parameters.
 
@@
 
-almost everywhere, where $\epsilon \sim U(0, 1)$ and $p^{-1}$ denotes the inverse CDF of the true posterior, will yeild $q(\boldsymbol{\theta} | \boldsymbol{\lambda} = p(\boldsymbol{\theta} | y_{1:T}$ and thus minimise the KL divergence.
+almost everywhere, where $\epsilon \sim U(0, 1)$ and $p^{-1}$ denotes the inverse CDF of the true posterior, will yield $q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = p(\boldsymbol{\theta} | y_{1:T})$ and thus minimise the KL divergence.
 Attempts to model a suitably flexible function $f(\cdot, \cdot)$ include heirarchical models \citep{Ranganath2016a}, neural networks (\citealp{Kingma2014}; \citealp{Rezende2014}; \citealp{Rezende2015}; \citealp{Kingma2016}), and Gaussian processes \citep{Tran2016}.
@@
 These approaches introduce a large number of parameters that would considerably slow computation of updates, so we follow \citet{Tran2015} and use copulas to flexibly model approximations to the true posterior distribution.
-The copula modelling class provides a flexible way to construct these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions.
+The copula modelling class provides a flexible way to construct these distributions, as it allows the dependence structure between parameters to be fit independently from the marginal distributions.
 \citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,
@@
 L(\theta | y_{1:T}) &= p(y_1, y_2 | \theta) \prod_{t=3}^{T}p(y_t | y_{1:t-1}, \theta) \nonumber \\
-&= \frac{1}{(2\pi)} |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} (y_1, y_2) \Sigma^{-1} (y_1, y_2)' \right\} \nonumber \\
+&\propto |\Sigma|^{-1/2} \sigma^{-(T-2)} \exp \left\{ \frac{-1}{2} (y_1, y_2) \Sigma^{-1} (y_1, y_2)' \right\} \nonumber \\
 &\times \exp \left\{ \frac{-1}{2\sigma^2} \left( \sum_{t=3}^{T}(y_t - \phi_1 y_{t-1} - \phi_2 y_{t-2})^2 \right) \right\}.
@@
 The literature for electricity forecasting for both load and price is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review.
-Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016} being notable examples, which used boot strapping and quantile regression respectively to generate forecast densities.
+Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016} being notable examples, which used bootstrapping and quantile regression respectively to generate forecast densities.
 Both of these approaches create forecast densities that are conditioned on the model and estimated parameters being correct, while the Bayesian approach used in this thesis can average over parameter and model uncertainty leading to a more accurate forecast density.

### 2017-03-13 Python/SSM.py

@@ -1,0 +1,26 @@
+#!/usr/bin/env python
+
+#import edward as ed
+#from edward.models import Normal
+import matplotlib.pyplot as plt
+#import matplotlib.gridspec as gridspec
+import numpy as np
+#import os
+import tensorflow as tf
+import numpy.random as rn
+
+T = 50
+y = np.empty(T)
+x = np.empty(T+1)
+
+xmu = 0
+xsd = 1
+ymu = 2
+xar = 0.5
+ysd = 1
+
+x[0] = rn.normal(xmu, xsd)
+for t in range(T):
+    x[t+1] = rn.normal(xmu + xar*x[t], xsd)
+    y[t] = rn.normal(x[t+1], ysd)
+

### 2017-03-13 Python/ar2.py

@@ -1,0 +1,1 @@
+T = 100 #number of data points

### 2017-03-13 Python/dlmnist.py

@@ -1,0 +1,3 @@
+from tensorflow.examples.tutorials.mnist import input_data
+
+mnist = input_data.read_data_sets("MNIST_data", one_hot=True)

### 2017-03-13 Python/mnistexample.py

@@ -1,0 +1,100 @@
+#!/usr/bin/env python
+"""Generative adversarial network for MNIST (Goodfellow et al., 2014).
+"""
+from __future__ import absolute_import
+from __future__ import print_function
+from __future__ import division
+
+import edward as ed
+import matplotlib.pyplot as plt
+import matplotlib.gridspec as gridspec
+import numpy as np
+import os
+import tensorflow as tf
+
+from edward.models import Uniform
+from tensorflow.contrib import slim
+from tensorflow.examples.tutorials.mnist import input_data
+
+
+def generative_network(z):
+  h1 = slim.fully_connected(z, 128, activation_fn=tf.nn.relu)
+  x = slim.fully_connected(h1, 784, activation_fn=tf.sigmoid)
+  return x
+
+
+def discriminative_network(x):
+  """Outputs probability in logits."""
+  h1 = slim.fully_connected(x, 128, activation_fn=tf.nn.relu)
+  logit = slim.fully_connected(h1, 1, activation_fn=None)
+  return logit
+
+
+def plot(samples):
+  fig = plt.figure(figsize=(4, 4))
+  gs = gridspec.GridSpec(4, 4)
+  gs.update(wspace=0.05, hspace=0.05)
+
+  for i, sample in enumerate(samples):
+    ax = plt.subplot(gs[i])
+    plt.axis('off')
+    ax.set_xticklabels([])
+    ax.set_yticklabels([])
+    ax.set_aspect('equal')
+    plt.imshow(sample.reshape(28, 28), cmap='Greys_r')
+
+  return fig
+
+
+ed.set_seed(42)
+
+M = 128  # batch size during training
+d = 100  # latent dimension
+
+DATA_DIR = "MNIST_data"
+IMG_DIR = "img"
+
+if not os.path.exists(DATA_DIR):
+  os.makedirs(DATA_DIR)
+if not os.path.exists(IMG_DIR):
+  os.makedirs(IMG_DIR)
+print("starting mnist")
+# DATA. MNIST batches are fed at training time.
+mnist = input_data.read_data_sets(DATA_DIR, one_hot=True)
+x_ph = tf.placeholder(tf.float32, [M, 784])
+
+print("mnist finished")
+# MODEL
+with tf.variable_scope("Gen"):
+  z = Uniform(a=tf.zeros([M, d]) - 1.0, b=tf.ones([M, d]))
+  x = generative_network(z)
+
+# INFERENCE
[…]

### 2017-03-13 Python/test.py

@@ -1,0 +1,7 @@
+import os
+
+print("Path at terminal when executing this file")
+print(os.getcwd() + "\n")
+
+print("This file path, relative to os.getcwd()")
+print(__file__ + "\n")

### 2017-03-16 Confirmation/presentations.tex
please work please please please

@@
 \item We aim to provide density forecasts for these high frequency time-series by updating posterior and predictive density estimates as each data point is observed.
-\item The algorithm used should scale to multivariate observations driven by complex models with large datasets.
+\item The approach used should scale to multivariate time-series driven by complex models with large datasets.
 \end{itemize}
@@
 \begin{itemize}
-\item Let $y_t \overset{iid}{\sim} \mathcal{N}\left(\mu, \sigma^2\right) \mbox{ for } t = 1, 2, \dots, T$. 
+\item Consider the Normal-Inverse Gamma model, where $y_t \overset{iid}{\sim} \mathcal{N}\left(\mu, \sigma^2\right) \mbox{ for } t = 1, 2, \dots, T$. 
 \item Use priors
@@
 \item Key Idea: Replace the intractable $p(\boldsymbol{\theta} | y_{1:T})$ with a tractable approximation $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
-\item How: Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ from a restricted class of distributions is closest to $p(\boldsymbol{\theta} | y_{1:T})$ by some metric. 
+\item How: Choose the $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ that is the closest member of a restricted class to $p(\boldsymbol{\theta} | y_{1:T})$ by some metric. 
 \item Typically minimise the Kullback-Leibler divergence from $q$ to $p$,
@@
 \begin{itemize}
-\item Marginal posterior densities for the Normal Inverse Gamma model with $T = 100, \mu = 2$ and $\sigma^2 = 1$.
+\item Marginal posterior densities for the Normal-Inverse Gamma model with $T = 100, \mu = 2$ and $\sigma^2 = 1$.
 \item Hyperparameters are $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. 
@@
 \begin{frame}
-\frametitle{Choosing the $q$ distribution}
+\section{Copulas}
+\frametitle{Choice of approximating functional form}
 \begin{itemize}
 \item Finding a good functional form for $q$ is an open problem.
-\item Approaches include heirarchical distributions and neural network transformations.
-\item $q$ on needs to be parsimonous to scale.
+\item $q$ should approximate the posterior well.
+\item $q$ needs to be parsimonous to scale.
 \item We augment $q$ with a copula to model dependence separately to the margins.
+\item From Sklar (1951), 
+\begin{equation}
+q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \prod_{i = 1}^{k} q(\theta_i | \boldsymbol{\lambda}_i) C(Q(\theta_1), \dots, Q(\theta_k | \boldsymbol{\lambda})),
+\end{equation}
+where $C$ is a copula, $q$ is a pdf and $Q$ is a cdf.
 \end{itemize}
@@
 \begin{frame}
-\section{Copulas}
 \frametitle{Vine Copulas}
 \begin{itemize}
-\item Vine copulas increase the flexiblity of copula modelling.
+\item Vine copulas factorise a copula to increase flexibility.
 \vspace{3mm}
@@
 \item Solution: Di{\ss}mann's algorithm generates a vine copula that fits a sample of data reasonably well.
-\item Implementation: Run MCMC and use Di{\ss}mann's algorithm to find a good vine copula to use as $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$.
-Run SVB to update this approximation as more data comes in.
+\item Implementation: 
+\begin{enumerate}
+\item Run MCMC and use Di{\ss}mann's algorithm to find a good vine copula to use in $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$. 
+\item Use model knowledge to force conditional independence where appropriate.
+\item Run SVB to update this approximation as more data comes in.
+\end{enumerate}
 \end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Vine Copula for a Four Dimensional $\boldsymbol{\theta}$}
+\begin{figure}[h]
+\centering
+\includegraphics[scale = 0.4]{vines}
+\label{fig:vines}
+\caption{A Vine Copula split into three trees. Edges in trees represent bivariate copulas, nodes in trees represent edges in the previous tree.}
+\end{figure}
 \end{frame}
@@
 \item Updating is worthwhile if the new data brings significant information on the parameters.
-State-space models are of interest as observation specific latent variables are more influenced by new data than global parameters.
 \item Existing online updating algorithms assume data is $iid$. We aim to develop an updating algorithm for dependent data.
-\item Our time-series context allows MCMC draws to guide the construction of $q$. We have no information on latent variables associated with new data and we must devise a way to include these in $q$.
-\item Statistical properties of the approximation are unknown and depend on the divergence measure used.
+\item Our time-series context allows MCMC draws to guide the construction of $q$. New data can be associated with variables not included in these draws.
+\item Statistical properties of the approximation are unknown, except that they depend on the divergence measure used.
 \item The accuracy/computation time trade-off between MCMC and VB has not been thoroughly investigated.
@@
 \frametitle{Timeline}
-\begin{itemize}
-\item Mid 2017: Complete electricity load forecasting application.
-\item End of 2017: Develop dependent online updating, apply to Stochastic Volatiltiy model with dependent latent variables.
-\item Compare approximations to the exact posterior to understand the efficiency trade-off.
-\item Mid 2018: Investigate statistical properties of other divergence measures such as Stein's discrepancy.
-\item End of 2018: Allow extra time as the literature is evolving rapidly and new research opportunities constantly appear.
-\item Mid 2019: Finalise thesis for submission.
-\end{itemize}
+\begin{tabular}{| l | l |}
+\hline
+ Mid 2017: & Complete electricity load forecasting application. \\
+ \hline
+ End of 2017: & Develop dependent online updating, \\
+ & apply to Stochastic Volatiltiy model with dependent \\
+ & latent variables. \\
+ & Compare approximations to the exact posterior to \\
+ &understand the efficiency trade-off. \\
+ \hline
+ Mid 2018: & Investigate statistical properties of other divergence \\
+ & measures such as Stein's discrepancy. \\
+ \hline
+ End of 2018: & Allow extra time as the literature is evolving rapidly \\
+ & and new research opportunities constantly appear. \\
+ \hline
+ Mid 2019: & Finalise thesis for submission.\\
+\hline
+\end{tabular}
 \end{frame}

### 2017-03-16 DLM/dlm_frvb.R

@@ -1,0 +1,200 @@
+##Functions for fitting a full rank Gaussian via gradient ascent
+logigamma = function(x, alpha, beta){
+  alpha * log(beta) - log(gamma(alpha)) - (alpha + 1)*log(x) - beta/x
+}
+
+logjoint = function(y, params){
+  theta = params[1:4] #first two params are log(sigma2)
+  x = params[5:(T+5)]
+  T = length(y)
+  logprior = log(1/2) + dnorm(theta[4], 0, sqrt(10), log = TRUE) -
+    theta[1] - 1 / exp(theta[1]) - theta[2] - 1 / exp(theta[2])
+  logstates = dnorm(x[1], 0, sqrt(exp(theta[2])/(1-theta[3]^2)), log = TRUE) + 
+    sum(dnorm(x[2:(T+1)], theta[3]*x[1:T], sqrt(exp(theta[2])), log = TRUE))
+  logy = sum(dnorm(y, x[2:(T+1)] + theta[4], sqrt(exp(theta[1])), log = TRUE))
+  return(logprior + logstates + logy)
+}
+
+logq = function(x, mean, L){
+  cons = -length(x)/2 * log(2*pi) + sum(log(diag(L)))
+  kernel = -1/2 * t(x - mean) %*% solve(L %*% t(L)) %*% (x - mean)
+  cons + kernel
+}
+
+simul = function(mean, L){
+  standardNormal = rnorm(length(mean))
+  normal = mean + L %*% standardNormal
+  while(normal[3] > 0.95 | normal[3] < -0.95){
+    standardNormal = rnorm(length(mean))
+    normal = mean + L %*% standardNormal
+  }
+  output = list(z = standardNormal, normal = normal)
+  return(output)
+}
+
+ELBO = function(y, mean, L, n = 250){
+  eval = 0
+  sigma = L %*% t(L)
+  for(i in 1:n){
+    sims = simul(mean, L)
+    lj = logjoint(y, sims$normal)
+    qcons = -length(x)/2 * log(2*pi) + sum(log(diag(L)))
+    qkernel = -1/2 * t(x - mean) %*% solve(L %*% t(L)) %*% (x - mean)
+    eval = eval + lj - (qcons + qkernel)
+  }
+  return(eval/n)
+}
+
+elboMeanDeriv = function(y, simulation, mean, L, i){
+  #Calculates the analytical derivative of the ELBO for a given simulated value of the parameters
[…]

### 2017-03-26 DLM/outline.tex
DLM additions

--- 2017-02-21 Confirmation/confirmation_V2_CF.tex
@@
 \numberwithin{equation}{section}
-\title{Real Time Variational Density Forecasts}
-\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
 \begin{document}
 
-\maketitle
-\tableofcontents
-\section{Introduction} \label{sec:Intro}
-%label added to section for referencing later
+\section{Introduction}
 
-Electricity prices have been subject to an extensive forecasting literature and is known to be extremely volatile.
-For the 2011-2015 period in Victoria the mean price per megawatt hour (mWh) was \$42.06, with 99\% of {\bf observations} less than \$100 {\bf per} % Try to minimize or not use shorthand mathematical notation in the introduction.
-mWh but on 39 {\bf occasions} the price spiked greater than \$1,000/mWh, reaching as high as \$9,974/mWh on November 29, 2012. 
-% This is very precise information, but I think for an introduction it is better to try to keep it a bit more `big picture'.
-Perhaps even just writing in percentage terms will have more impact, rather than requiring the reader to work out if this difference is big or not.
-Then also you wouldn't need to units of measure in quite so many places.
-I'd even prefer not to have the abbreviation `mWh' here, but this is less critical.
-}
-For this reason market participants are increasingly interested in a predictive density forecast, as the probability that the price will rise above a particular threshold in the next time period is more important than the mean value.
-To calculate a suitably accurate predictive density the Bayesian methodology is to be adopted in this thesis, as it {\bf accounts for both } parameter and model uncertainty {\bf often} ignored {\bf by} frequentist methods.
-% I agree with Tas that it is better not to be so defensive at the outset.
-Better to justify that positives of Bayesian methods - in particular the use of prior information and the ability to handle complex, parameter-driven (state space) models.
-See \citet{Geweke2006} for a review on Bayesian forecasting and \citet{Gneiting2014} for density forecasting in general. 
-% Sorry but I don't know why the references are not working for me!
+\begin{itemize}
+\item Compare MCMC methods with variational methods for forecasting a Dynamic Linear Model and a Stochastic Volatility Model. 
+\item Use a range of different updating methods to adjust forecasts as new data is observed.
+\item Compare compuation time and forecast accuracy for all approaches.
+\end{itemize}
 
-% It might be useful to say something in the above about the connection between the predictive distribution, and the posterior distribution.
-Or just talk about distributions here, and then the connection to the forecast density can come in the next section.
+\section{Approaches Considered}
 
-Bayesian methods are computationally complex as they require integration over the set of {\bf unknowns, whether parameters or latent stochastic variables.
-This set is typically quite large for models used to model short term electricity load and price as there are} strong seasonal patterns {\bf in the data \citep{Taylor2003}.} {\bf While analytical integration is typically not feasible to obtain the predictive distribution for such models, however} there are a range of {\bf numerical and simulation-based} techniques available, {\bf for this purpose,} notably Markov Chain Monte Carlo (MCMC) methods.
-While {\bf it is often possible to demonstrate theoretical convergence of MCMC algorithms,} computation {\bf may be} slow for even moderately complex models.
-{\bf This} is problematic {\bf for when attempting to forecast} Electricity demand
-%demand? or price? or load? decide which you are going to talk about here, and only talk about that for now - check throughout.
-in Victoria, {when realizations occur at} five minute {\bf intervals}. The predictive density for the next period's electricity load may not be available before it is observed.
+\textbf{MCMC Methods}
+\begin{itemize}
+\item Gibbs + Metropolis-Hastings MCMC.
+\item Hamiltonian Monte Carlo.
+\item Particle Filtering / QMLE / Extended Kalman Filter / Whatever else is used for SV.
+\end{itemize}
+\textbf{Updating for MCMC}
+\begin{itemize}
+\item Full updating - rerun MCMC after each new observation for best possible result.
+\item No updating - just use multi step ahead forecasts via filtering.
+\end{itemize}
+\textbf{Variational Methods}
+\begin{itemize}
+\item Independent Gaussian - Transform parameters to real line eg. $\log(\sigma^2)$.
+\item Full Rank Gaussian with real line transformation.
+\item Sampling informed distributions - Copula model from Di{\ss}mann's algorithm.
+\item Neural Network transform of $N(\boldsymbol{0}, I)$ posterior.
+\item Stochastic Recurrent Neural Networks - SV Model only.
+\item Alternative divergence measures eg. Expectation Propogation.
+\end{itemize}
+\textbf{Updating for VB}
+\begin{itemize}
+\item Update all parameters and latent variables.
+\item Update global parameters and latent variables for new data - Hold previous latent variables fixed.
+\item Do not update, just use filtering and multi step ahead forecasts.
+\end{itemize}
 
-Alternatives to MCMC include Variational Bayes (VB) {\bf methods, where the aim is} to replace the unknown true posterior distribution for a vector of {\bf unknowns}
-% you can reduce to just $\theta$ below - but keep it general in the introduction
-with a parametric approximation. {\bf The} Variational Bayes {\bf approach} will be explored {\bf in the thesis, to determine if it is a viable} alternative computation strategy to MCMC,
-% comma added
-in the context of {\bf modelling and short term forecasting of Victorian electricity load.}
-% Unless you have something interesting to say about the time period - which could come later - there isn't any real need to mention the data period at this point.
-It could come in the last paragraph, however.
-{\bf In particular, we seek to understand the benefits gained in terms of the computational time required to obtain a VB approximate forecast distribution against the loss in statistical} 
-% I think you need to use the spell checker!!
-accuracy associated with the use of an {approximate posterior forecast distribution.} 
-%
+\section{Simulation Results}
 
-% You need a paragraph here detailing the outline of the rest of the paper, section by section.
+Simulate each estimation/update combination many times, calculate cumulative scores of new observations as well as compuation time for initial fit and each update.
 
-\section{Bayesian Inference}
-
-{\bf To facilitate the discussion, consideration is given to models where the unknowns may be summarized in a (finite) $k-$dimensional vector, denoted by $\theta.$ In this setting, given an observed time series, denoted by $y_{1:T}= \{y_{t}, t = 1, \dots T\}$, the Bayesian forecast distribution associated with time $t=T+1$ is characterized by the conditional density }
-\begin{equation}
-\label{predictive}
-p(y_{T+1} | y_{1:T}) =\int p(y_{T+1}|y_{1:T},\theta) p(\theta|y_{1:T}) d\theta,
-\end{equation}
-{\bf where $p(\theta|y_{1:T})$ denotes the posterior density for $\theta$, given by}
-
-\begin{equation}
-\label{posterior}
- p(\theta | y_{1:T}) = \frac{p(y_{1:T}|\theta)p(\theta)}{\int_\theta p(y_{1:T}|\theta)p(\theta) d\theta}.
-\end{equation}
-% You can feel free to modify the above, but be sure to define all notation that you use.
-
-% I think you need to expand this section a bit more - think about what you really need to say, and why.
-Generally the {\bf analytical} solution to (\ref{posterior}), 
-% comma added
-{\bf and hence to (\ref{predictive}) will be unavailable.} 	% Can you say why? 
-{\bf In this} section {\bf two alternative methods for computing the desired posterior distribution in (\ref{posterior}),  MCMC and VB, will be reviewed.}
-% Since neither method is new, you are not `offering' them, just reviewing them.
-% Can you extend this discussion to explain how to use MCMC and VB for (\ref{\predictive})?
-{\bf In brief,} MCMC is used to create a sample from $p(\theta | y)$, {\bf with} any function {\bf of $\theta$ that is desired estimated from that sample.
-In contrast,} VB replaces $p(\theta | y)$ with an approximation,
-%comma inserted
-{\bf denoted by} $q(\theta | \lambda)$, where $\lambda$ is a vector of {\bf auxiliary} parameters associated with the approximation and not the model. 
-% surely these $\lambda$ have something to do with the model??? Perhaps end the sentence with `approximation'. Also need to emphasize that $q(\theta|\lambda)$ is also a function of the data $y_{1:T}$.
-For example, if $q(\theta | \lambda)$ is Gaussian,
-%comma added
-{\bf then} $\lambda$ would {\bf denote} the mean and variance {\bf parameters that characeterize this normal distribution for} $\theta$.
-Variational Bayes aims to choose the distribution $q(\theta | \lambda)$ such that the Kullback-Leibler divergence from $q(\theta | \lambda)$ to $p(\theta | y)$ is minimised.
-% I have used `z' (as in `minimized') in my edits, not `s' (as in `minimised'), but you decide which you prefer and be consistent.
-%So obviously now the values of the mean and variance will in fact depend on the data, and the Gaussian model will be selected due following consideration of at least a portion of the original, more complex model (via this minimization).
-So yes, the approximation and its parameters will, in some sense at least, depend on the model!
-
-\subsection{Markov Chain Monte Carlo}
-
-There are many types of MCMC {\bf algorithms, with arguably the simplest and most commonly used one being the} Gibbs sampler.
-{The Gibbs sampler algorithm} iteratively samples {\bf the components of the} $k$ dimensional parameter vector $\theta$ via {\bf each of} the {\bf so-called full} conditional distributions {\bf as follows,}
-% I'm not sure you need the extra space above each equation - have not changed elsewhere but please consider
-\begin{align}
-&p(\theta_1 | \theta_2, \dots, \theta_k, y_{1:T}) \nonumber \\
-&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, y_{1:T}) \nonumber \\
-&\vdots \nonumber \\
-&p(\theta_k | \theta_1, \dots, \theta_{k-1}, y_{1:T}). \nonumber
-\end{align}
-% when you put a space after the equation, LaTeX thinks you are starting a new paragraph.
-Depending on the style you are using, this may generate a tab at the start of the next line, which you may not want - please be mindful of this, I have not corrected throughout.
-{\bf Under mild regularity conditions (see, e.g., Tierney, 1994, Annals of Statistics) and with} enough iterations {\bf of the Markov chain that results from the Gibbs sampler,} these samples converge in distribution to $p(\theta | y)$ and can be used for inference.
-% The phrase `and can be used for inference' is a bit vague...
-The computation time {\bf for each} iteration and {\bf the overall} number of iterations required for {\bf to accruately summarize the posterior distribution} is problem specific but typically increases with model complexity.
-% what do we know about model complexity? Can we just say it is $k$? Or do you want to highlight something more?
-
-\subsection{Variational Bayes}
-
-{\bf As discussed in Section \ref{sec:Intro}, a faster (albeit approximate)} alternative to MCMC is {\bf VB.
-This approach} introduces an approximating distribution {\bf $q(\theta | \lambda),$ with the choice of the functional form $q$ and subsequent auxiliary parameter $\lambda$ determined so that} the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q(\theta | \lambda)$ the true posterior $p(\theta | y)$ {\bf is minimised.} The KL divergence is defined by
-
-\begin{equation}
-\label{KL-def}
-KL[q(\theta | \lambda) \hspace{.1cm}||\hspace{.1cm}p(\theta | y)] = \int q(\theta | \lambda) \ln \left( \frac{q(\theta | \lambda)}{p(\theta | y)}\right) d\theta.
-\end{equation}
-% no new para.
-Note it might be useful to add some space before and after the double vertical lines - to make the equation above easier to read.
-I have done but see what you think...I haven't changed elsewhere.
-The KL divergence is a non-negative, asymetric %spelling
- measure of the discrepancy between $p(\theta | y)$ and $q(\theta | \lambda)$ that will {\bf be equal to} zero if and only if $p(\theta | y) = q(\theta | \lambda)$ almost everywhere \citep{Bishop2006}.
-Note that $KL[q(\theta | \lambda)||p(\theta | y)]$ can be expressed as
-
-\begin{equation}
-\label{KL-ELBO}
-KL[q(\theta | \lambda)||p(\theta | y)] = \ln(p(y)) - \mathcal{L}(q, y)
-\end{equation}
-
-where $\mathcal{L}(q, y)$ is {\bf referred to} as the Evidence Lower Bound (ELBO), {\bf and is} defined by
-% Can you give some intuition for why this is called the ELBO?
-\begin{equation}
-\label{ELBO}
-\mathcal{L}(q, y) = \int_{\theta} q(\theta|\lambda) \ln (p(y, \theta|\lambda)) d\theta -  \int_{\theta} q(\theta|\lambda) \ln (q(\theta|\lambda)) d\theta.
-\end{equation}
-
-Since $\ln(p(y))$ is constant given $y$, maximising (\ref{KL-ELBO}) with respect to $q$ is equivalent to minimising (\ref{KL-def}).
-Maximising the ELBO is much more convenient than minimising the KL Divergence, {\bf as will} be shown in the following sections {\bf in the context of the} two major implementations of Varitaional %spelling
-{\bf Bayes, namely} Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these {\bf implementations} take advantage of the functional form of the ELBO, %comma inserted
-and offer computationally efficient algorithms to find a $q(\theta | \lambda)$ that is optimal within a particular clas %spelling
-of distributions.
-
-\subsubsection{Mean Field Variational Bayes}
-
-Mean Field Variational Bayes (MFVB) has origins in the physics literature \citep{Chandler1987} and restricts the class of distributions {\bf from which to select $q(\theta|\lambda)$} to the set of factorisable distributions,
-
-\begin{equation}
-\label{mf1}
-q(\theta|\lambda) = \prod_i^k q_i(\theta_i | \lambda_i).
-\end{equation}
-% Again, Tas's comment here about not re-using $k$ for this
-Each of the $k$ components $\theta_i$ may be a vector, but in most implementations are scalars.
-Each $\theta_i$ has an associated vector $\lambda_i$ which may be of a higher dimension than $\theta_i$.
-$\lambda_i$ is an auxillary parameter vector for the relevant factor $q_i$, which will be used in this section as shorthand notation for the distribution $q_i(\theta_i |\lambda_i)$.
-
-MFVB is widely used as it greatly simplifies maximisation of the ELBO (\citealp{Jordan1999}; \citealp{Bishop2006}), particularly in exponential family models \citep{Wainwright2008}.
-Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the computation is simple.
-%It is not clear at this point what you mean by 'the computation'.
-Is it that the identification process is the same, but rather than sampling from the identified distribution, as in Gibbs sampling, here you will find the value of the parameter $\lambda_i$ to maximise the identified distribution (for fixed $y_{1:T}$ and fixed other parameters in $\theta$ other than $\theta_i$).
-MFVB expresses (\ref{ELBO}) as a function of one $q_i$ only 
-% This doesn't quite make sense to me.
-It must be that MFVB assumes that (\ref{ELBO}) can first be represented as some kind of product of similar expressions, one for each $i$, and then that the maximisation of (\ref{ELBO}) can be determined by maximising each of these expressions.
-We talked about this a bit last week - perhaps you can tease this out a bit more.
-Somehow the iteration comes in so as to achieve this maximum.
-The final algorithm requires this {\bf process} to be repeated for each $i$ in $1, \dots, k$, {\bf until ...}.
-%Some kind of convergence criterion is needed.
-Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that
-
-\begin{equation}
-\label{mf2}
-q(\theta_i | \lambda_i) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y,\theta))])
-\end{equation}
-% The sentence is missing a full stop, and also I'm not sure how it connects to everything else?
-{\bf To illustrate the MFVB approach,} consider data $y_i$ for $i = 1, \dots, n$ generated {\bf independently} from a $\mathcal{N}(\mu, \sigma^2)$ distribution with {\bf independent marginal prior distributions given by} %It is $\mu$ that has a distribution, not $p(\mu)$ - changed here
-$\mu \sim \mathcal{N}(\gamma, \lambda)$ and $\sigma^2 \sim Inv.Gamma(\alpha, \beta)$. It can be shown that
-% I think you need some $\lambdas$ below?
-Also would be worthwhile to  explain or show why these are the correct MFVB distributions...or at least be able to say something about it during your presentation.
-
-\begin{equation}
-\label{mf3}
-q(\mu) \propto \exp \left\{ \frac{-(n\lambda + \mathbb{E}(\sigma^{-2}))}{2\lambda\mathbb{E}(\sigma^{-2})} \left( \mu - \frac{\mathbb{E}(\sigma^{-2})\gamma + \lambda \sum_{i=1}^{n} y_i}{n \lambda + \mathbb{E}(\sigma^{-2})} \right)^2 \right\}
-\end{equation}
-
-and
-
-\begin{equation}
-\label{mf4}
-q(\sigma^2) \propto \sigma^{-2(n/2 + \alpha + 1)} \exp \left\{ \frac{ -1/2(\sum_{i=1}^{n}y_i^2 + n\mathbb{E}(\mu^2) - 2\sum_{i=1}^{n} y_i \mathbb{E}(\mu)) - \beta}{\sigma^2} \right\}.
-\end{equation}
-
-It is evident that $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ and $q(\sigma^2 | \lambda) \sim Inv.Gamma(\tilde{\alpha}, \tilde{\beta})$.
-The parameters $\lambda = (\tilde{\gamma}, \tilde{\lambda}, \tilde{\alpha}, \tilde{\beta})'$ can be found
-% how?
-I think you are just looking at the equations and `identifying' these guys, but it would be usefult to explain this.
-At least you have to mention something about how to evaluation, e.g.
-$E(\sigma^{-2})$ in $q(\mu | \lambda) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\lambda})$ to get that $\tilde{\gamma} = \frac{\alpha / \beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta}$.
-Same for the other parameters too..
-by the set of mean field equations:
+\subsection{Dynamic Linear Model}
 
 \begin{align}
-\tilde{\gamma} &= \frac{\alpha / \beta \gamma + \lambda \sum_{i=1}^{n} y_i} {n \lambda + \alpha / \beta} \label{mf5} \\ 
-\tilde{\lambda} &= \frac{n \alpha / \beta}{n \lambda + \alpha / \beta} \\
-\tilde{\alpha} &= \frac{n}{2} + \alpha \\
-\tilde{\beta} &= \frac{1}{2} \left(\sum_{i=1}^{n} y_i^2 + n(\tilde{\gamma}^2 + \tilde{\lambda}) - 2 \sum_{i=1}^{n} y_i \tilde{\gamma} \right) + \beta. \label{mf6}
-\end{align}.
+y_t &= \mu + \beta x_t + \sigma_y \epsilon_t \\
+x_t &= \gamma + \phi x_{t-1} + \sigma_x \nu_t
+\end{align}
 
-As there is a circular dependence in these equations an algorithm that cycles through each auxillary %spelling
-parameter until convergence to some pre-specified threshold %need to give some definition of this
-is required.
+$\epsilon_t$ and $\nu_t \sim Z$
 
-%Paragraph moved below
-
-Matching distributions in this way is a very similar method to finding the posterior conditional distribution, $p(\theta_i | y, \theta_{\setminus i}),$ %parenthesis and comma added, but I think this is different notation to what you used before - needs to be the same!
-{\bf for example} in a Gibbs {\bf sampling} MCMC scheme, with dependence on other parameters replaced by their expectations.
-The optimal approximating family for $\theta_i$, $q(\theta_i | \lambda_i)$, will come from the same distributional family as the conditional distribution $p(\theta_i | \theta_{j \neq i}, y)$.
-% full stop added.
-Isn't this true, regardless of whether we can recognize the form or not?
-if it exists in a recognisable form such as the exponential family. % have left this for you to resolve...
-
-% moved paragraph to here
-In the event that {\bf the form of the distribution in} (\ref{mf2}) is {\bf unrecognisable,} a further approximation {\bf may} be used {\bf by} substituting in another distribution {\bf to replace} the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One {\bf approach, given by \citet{Friston2006},} uses a Laplace {\bf approximation, substituting} a Gaussian distribution for {\bf the} unrecognizable $q_i(\theta_i | \lambda_i)$.
-{\bf The} only requirement {\bf for this approach} is that the substitute {\bf have} an expectation
-% expectations? don't we need both of the first two moments? 
-that is a tractable function 
-% that are tractable functions? 
-of {\bf the} parameters to be used for other parameters in (\ref{mf2}). 
-{\bf This} secondary level of {\bf approximation} is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable distributions.
-% I really like this analogy, and it will help your audience understand what is going on.
-But to make it work, you should have explained MHwithinGibbs earlier.
-
-% Moved revised sentence down to here. But now you'll need to work this in with the following paragraph!
-In the event that {\bf the form of the distribution in} (\ref{mf2}) is {\bf unrecognisable,} a further approximation {\bf may} be used {\bf by} substituting in another distribution {\bf to replace} the problematic $\tilde{q_i(\theta_i|\lambda_i)}.$ One {\bf approach, given by \citet{Friston2006},} uses a Laplace {\bf approximation, substituting} a Gaussian distribution for {\bf the} unrecognizable $q_i(\theta_i | \lambda_i)$.
-{\bf The} only requirement {\bf for this approach} is that the substitute {\bf have} an expectation
-% expectations? don't we need both of the first two moments? 
-that is a tractable function 
-% that are tractable functions? 
-of {\bf the} parameters to be used for other parameters in (\ref{mf2}).
-
-\vspace{5mm}
-
-% You have not really fully explained how this algorithm below fits with what is above?? A single sentence might be sufficient.
-{\bf Once the approximation has been identified, the algorithm required to find the ELBO is} known as a coordinate ascent algorithm. %full stop
-{\bf This algorithm} is described below {\bf in the case} where $\lambda$ is a $p$ dimensional vector. %why change from $k$ to $p$? need to keep consistent notation throughout. 
-%We have discussed these algorithms already, and I have nothing more to add now
-
-\vspace{2mm}
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density}
- \KwResult{Mean Field Approximation}
- Use (\ref{mf3}) to match each $q(\theta_i|\lambda_i)$ to a tractable distribution or substitute a further approximation if neccesary.\;
- Derive the set of mean field equations, such as (\ref{mf5}) - (\ref{mf6}).\;
- Initialise $\lambda$ randomly. \;
- \While{Not converged}{
-  \For{$i =  1$ \KwTo $k$}{
-      Hold $\lambda_{j \neq i}$ fixed\;
-      Update $\lambda_i$ using the relevant mean field equation.
-     }
- }
- \caption{Coordinate Ascent for MFVB}
-  \label{alg:algorithm1}
-\end{algorithm}
-
-\subsubsection{Stochastic Variational Bayes}
-
-Restricting the approximating distribution to a factorisable family may be unsatisfactory in many applications where the dependence between parameters is important, as in the AR(2) model where Figure \ref{MCMCplot} shows strong dependence between $\phi_1$ and $\phi_2$.
-Without factorisation the simple method to maximise the ELBO as in MFVB is unavailable.
-\citet{Paisley2012} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, resulting in what is known as Stochastic Variational Bayes (SVB).
-
-The algorithm for SVB iteratively takes the  derivative of $\mathcal{L}(q(\theta | \lambda), y)$ in (\ref{ELBO}) with respect to $\lambda$ and the following updating step is applied until the ELBO converges to a maximum.
-
-\begin{equation}
-\label{SGA1}
-\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \nabla_{\lambda} \mathcal{L}(q(\theta | \lambda^{(m)}), y),
-\end{equation}
-
-where $\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ is the vector of partial derivatives of $\mathcal{L}(q(\theta | \lambda^{(m)}), y)$ with respect to each element of $\lambda$.
-This update requires some initial values $\lambda^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
-If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.
+\subsection{Stochastic Volatility Model}
 
 \begin{align}
-\sum_{m=1}^{\infty} \rho^{(m)} &=  \infty \\
-\sum_{m=1}^{\infty} (\rho^{(m)})^2 &<  \infty.
+y_t &= \exp\left(\alpha_t/2\right) \epsilon_t \\
+\alpha_t &= \gamma + \phi \alpha_{t-1} + \sigma \nu_t
 \end{align}
 
-Whilst a global maximum is desired, the ELBO is a high dimensional problem specific function that makes finding the global maximum extremely difficult and often only a local maximum can be found.
-One option to alleiviate this problem is to start the algorithm at a range of initial values choose the converged value with the highest maximised ELBO and hence lowest KL divergence to the true posterior
+$\epsilon_t$ and $\nu_t \sim Z$.
+Alternatively transform to another representaiton if that works better for filtering/QMLE etc.
 
-SVB can find only the optimal values for $\lambda$ that are optimal for a distributional family $q$ that is specified by the user.
-To run SVB many choices of $q$ may be used and one selected on the basis of the highest ELBO.
-$q(\theta | \lambda)$ is restricted to the family of distributions that satisfy the condition that the order of differentation of the ELBO with respect to $\lambda$ and integration with respect to $\theta$ are interchangable, and SVB can only be applied to models where the log likelihood for $y$ is able to be evaluated.
-Given these conditions \citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be taken by
-
-\begin{equation}
-\label{SGA2}
-\nabla_{\lambda}\mathcal{L}(q(\theta | \lambda^{(m)}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\lambda} [\ln(q(\theta_s | \lambda^{(m)}))] (\ln (p(y, \theta_s)) - \ln(q(\theta_s | \lambda^{(m)})))
-\end{equation}
-
-where $s = 1, \dots, S$ indicates simulations from $q(\theta | \lambda^{(m)})$.
-
-\citet{Duchi2011} introduced the AdaGrad algorithm which can be implemented within SVB to control $\rho^{(m)}$.
-AdaGrad allows each $\lambda_i$ to have an independent $\rho^{(m)}_i$ that is inversely proportional to the gradient, so $\lambda$ takes bigger steps in flat regions and smaller steps in steep regions.
-
-Let
-
-\begin{equation}
-\label{SGA3}
-G_i^{(m)} = \sum_{j = 1}^{m} \left(\nabla_{\lambda_i}\mathcal{L}(q(\theta | \lambda^{(m)})\right)^2,
-\end{equation}
-
-then each component's learning rate is defined as
-
-\begin{equation}
-\label{SGA4}
-\rho^{(m)}_i = \eta \left(G_i^{(m)}\right)^{-1/2}
-\end{equation}
-
-for some tuning parameter $\eta$.
-
-The resulting Stochastic Gradient Ascent algorithm proceeds below with a $p$ dimensional $\lambda$ vector.
-
-\begin{algorithm}[H]
- \SetKwInOut{Input}{Input}
- \Input{Log Joint Density, Approximation family q}
- \KwResult{Variational Approximation}
- Initialise $\lambda$\;
- \While{Not converged}{
-  Simulate $\theta^s$ for $s = 1, \dots S$ from $q(\theta|\lambda^{(m)})$\;
-  \For{$i =  1$ \KwTo $p$}{
-      Calculate $\nabla_{\lambda_i}$ from (\ref{SGA2})\;
-      Update $G_i^{(m)}$ and $\rho^{(m)}_i$ from (\ref{SGA3}) and (\ref{SGA4})\;
-      }
-  Update $\lambda^{(m+1)}$ from (\ref{SGA1})\;
-  Set $m = m + 1$\;
- }
- \caption{Stochastic Gradient Ascent for SVB}
-  \label{alg:algorithm2}
-\end{algorithm}
-
-\subsection{Vine Copulas}
-% I won't add too many more to Tas's comments, except to remind you to take the sorts of critiques I have made above and try to apply them here - be clear about what you are saying, don't leave out so many words and use that spell checker :-)
-
-Armed with an algorithm to find the Variational Bayes optimal paramerters $\lambda$ for a distribution $q$, there is a requirement to have a distribution $q$ that is itself optimal.
-Copulas are a flexible tool for constructing these distributions, as they allow the dependence structure between parameters to be fit independently from the marginal distributions.
-\citet{Sklar1959} proves that any joint probability distribution can be written as the product of the marginals and a copula function,
-
-\begin{equation}
-\label{vc1}
-p(\theta_1, \dots, \theta_k) = p(\theta_1) \dots p(\theta_k) c(P(\theta_1), \dots, P(\theta_k))
-\end{equation}
-
-where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.
-
-\citet{Joe1994} introduces the vine copula, a technique to factorise a high-dimensional copula into a set of bivariate copulas in a similar way to the factorisation of a joint distribution into a marginal and conditional distribution, for example
-
-\begin{equation}
-\label{vc2}
-c(P(\theta_1), P(\theta_2), P(\theta_3)) = c(P(\theta_1), P(\theta_2)) \cdot c(P(\theta_1), P(\theta_3)) \cdot c(P(\theta_2), P(\theta_3) | P(\theta_1)).
-\end{equation}
-
-A copula for a $k$ dimensional $\theta$ can be replaced with $k(k-1)/2$ bivariate copulas arranged in a vine copula, greatly increasing the amount of flexibility available to fit a distribution.
-However there are $O(k!)$ ways to factorise the full copula and choose a family for each bivariate copula, so without detailed problem specific information the application of a vine copula to Stochastic Variational Bayes is difficult.
-\citet{Tran2015} implements Variational Bayes with a vine copula in what they call Copula Variational Bayes, but the dimension of $\theta$ was small in that case allowing an exhaustive search for the vine copula that maximised the ELBO.
-
-Given a sample of $\theta$, \citet{Diimann2013} provide an algorithm to select the best vine factorisation by maximising Kendell's Tau, and the best family for each bivariate copula by maximising an information criterion such as AIC.
-In the real-time forecasting models used for electricity load demand, MCMC samples of the posterior distribution are available, though often immediately out of date.
-The approach used in this thesis is to run an MCMC algorithm, thin the resulting draws until they are effectively independent, choose an optimal family for each marginal distribution by AIC, then run (Vine copula algorithm here) to infer an optimal vine copula.
-This results in an approximating distribution $q(\theta | \lambda)$ that can then be used in an SVB algorithm to update $q(\theta | \lambda)$ as more observations become available.
-MCMC can be ran simultaenously and the resulting sample can be used to check if the distribution $q$ should be changed.
-
-\subsection{Applying Stochastic Variational Bayes}
-
-Using the approach described in {\bf Section} 2.3,
-% When referring to a specific section, use a capital letter. When referring generally to `this section' or `later sections', etc, use lower case.
-an application of VC algorirhtm to the MCMC draws obtained in {\bf Section} 2.1 % You can check the rest
-found the following distribution to be optimal, {\bf according to the standard Akaike Information Criterion (AIC)}.
-
+\section{Empirical Application}
 \begin{itemize}
-\item $q(\phi_1)$ - Gaussian marginal
-\item $q(\phi_2)$ - Gaussian marginal
-\item $q(\sigma^2)$ - Inverse Gamma marginal
-\item $c(Q(\phi_1), Q(\phi_2))$ - Gaussian copula
-\item $c(Q(\phi_1), Q(\sigma^2))$ - Independent
-\item $c(Q(\phi_2), Q(\sigma^2) | Q(\phi_1))$ - Independent
-\end{itemize}
-
-Figure \ref{VBfit} shows that Stochastic Variational Bayes (blue) fit the MCMC (red) sample very closely.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale = 0.5]{VBfit.png}
-\caption{The fit of an SVB algorithm (red) compared to MCMC (red) for the AR(2) model described in section 2.1.
-SVB resulted in an almost identical posterior to MCMC, requiring around one hundredth of the runtime for MCMC.}
-\label{VBfit}
-\end{figure}
-
-\section{Electricity Load Forecasts}
-\subsection{Motivation}
-
-The literature for electricity forecasting is wide, and contains models based on many different fields, such as game theory, time-series modelling and neural networks, see \citet{Weron2014} for a recent review.
-Density forecasting for short term electricity load has had less attention, with \citet{Fan2012} and \citet{He2016}
-% one of the disadvantages of the automatic citations is that the formatting isn't quite as I would expect.
-Perhaps a different bibliographical style would be better.
-In Econometrics, we typically list the full set of authors names the first time a paper having 3 (sometimes 4) or more authors, and then if cited again use the `et al.' convention.
-Also, the `et al.' should be in italics.
-But this is minor for now...
-being notable examples, which used boot strapping and quantile regression respectively to generate forecast densities.
-Both of these approaches create forecast densities that are conditioned on the model and estimated parameters being correct, while the Bayesian approach used in this thesis can average over parameter and model uncertainty leading to a more accurate forecast density.
-This thesis also aims to re-estimate the model after each data point is observed to fully take advantage of any extra information on changes in parameters and thus further increase the accuracy of the forecast density.
-
-Half hourly electricity load data for Victoria, Australia has been collected % You need to report where you got the data from. You didn't collect it yourself, which is how it reads!
- for the 2011-2015 period, and is plotted in Figure \ref{loadplot}.
-The time series displays strong seasonal characteristics, with a yearly, day of week, and time of day pattern.
-Electricity load is strongly dependent on temperature, with high load experienced on both very cold and very hot days.
-In summer periods, load is often low but shows extreme volatility, while load slowly rises then falls in winter and displays significantly less volatility.
-
-The electricity market in Victoria operates by market participants sending electricity bid and offers for each five minute period to  AEMO, a centralised market operator.
-AEMO then determines the dispatch price and load for that five minute period, and determines the final spot price participants pay for each half-hour period by averaging the six five minute dispath% spelling
- prices in that half-hour.
-Market participants are allowed to revise bids and offers at any point before dispatch.
-The electricity supply curve is strongly J shaped, as coal generators are able to provide a large load cheaply but if demand exceeds this level the price spikes as other generators must be brought online expensively.
-These characteristics of the electricity market leads to a large amount of volatility in prices demonstrated by Figure \ref{rrpplot}.
-The seasonal effects of the load data are also present in prices, but these effects are dominated by irregular spikes as high as \$9,974/mWh, compared to a four year mean of \$42/mWh.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale = 0.5]{load_timeseries.png}
-\caption{Half hourly load data measured in megawatt hours for Victoria from January 1 2011 to December 31 2015.
-The yearly pattern is characterised by a low median load with regular spikes in summer and a more consistent sine curve in winter.
-Note: Replace with daily average to make the pattern clearer?}
-\label{loadplot}
-\end{figure}.
-
-\begin{figure}[h]
-\centering
-\includegraphics[scale = 0.5]{RRP_timeseries.png}
-\caption{Half hourly price data for Victoria from January 1 2011 to December 31 2015.
-The price shows a strong time of week and time of day seasonal effect in the low price periods, and has irregular spikes.}
-\label{rrpplot}
-\end{figure}.
-
-\subsection{Exponential Smoothing}
-
-\citet{Taylor2003} explores the seasonal properties of minute-by-minute electricity demand in the United Kingdom and introduces a double seasonal Holt-Winters exponential smoothing model with daily and weekly effects described by (\ref{ds-hw1})-(\ref{ds-hw4}).
-Yearly effects are omitted as their inclusion requires a large number of latent states to be estimated.
-This model is later verified
-% Verified is a very strong word. I would say these authors have provided empirical support that the model is superior
-in \citet{Taylor2008} as superior to other common time-series models for very short-term load forecasting.
-
-\begin{align}
-y_t &= l_{t-1} + d_{t-m_1} + w_{t-m_2} + e_t \label{ds-hw1} \\
-l_t &= \alpha (y_t - d_{t-m_1} - w_{t-m_2}) + (1 - \alpha)l_{t-1} \label{ds-hw2}\\
-d_t &= \delta (y_t - l_{t-1} - w_{t-m_2}) + (1 - \delta)d_{t-m_1} \label{ds-hw3} \\
-w_t &= \omega (y_t - l_{t-1} - d_{t-m_1}) + (1 - \omega)w_{t-m_2} \label{ds-hw4}
-\end{align}
-
-where $m_1$ and $m_2$ are the lengths of the daily and weekly cycle, restricting the smoothing parameters $\alpha, \delta, \omega$ to lie in $(0, 1)$.
-This can be rewritten as a single source of error state-space model, {\bf as per \citep{Snyder1985},  as follows}
-
-\begin{align}
-y_t &= l_{t-1} + d_{t-m_1} + w_{t-m_2} + e_t \label{ds-hw-ssoe1} \\
-l_t &= l_{t-1} + \alpha e_t \label{ds-hw-ssoe2} \\
-d_t &= d_{t-m_1} + \delta e_t \label{ds-hw-ssoe3} \\
-w_t &= w_{t-m_2} + \omega e_t \label{ds-hw-ssoe4}. 
-\end{align}
-
-% Now I think you should really explain what `single source of error' means, and why it is useful (it means you have the likelihood in closed form, except for the initializing state variables and parameters - which together would make up your $\theta$ vector).
-Also you can identify what the `state' at time $t$ is
-The unknown parameters are $\theta = (\alpha, \delta, \omega)', \sigma^2$ and $b_0 = (l_0, d_0, \dots, d_{-(m_1 - 1)}, w_0, \dots, w_{-(m_2 - 1)})'$ a $m_1 + m_2 + 1$ length vector of the initial states of the latent variables.
-\citet{Forbes2000} derives the marginal distribution $p(\theta | y)$ and notes that, conditioned on a draw of these parameters, the remaining unknown parameters can be sampled as a Bayesian linear regression problem.
-This method involves inverting a matrix that is singular using the parameterisation described by (\ref{ds-hw-ssoe1}) - (\ref{ds-hw-ssoe4}), but restricting the latent parameters via
-
-\begin{align}
-d_t &= - \sum_{i=1}^{m_1-1} d_{t-i} \\
-w_t &= - \sum_{i=1}^{m_2-1} w_{t-i}
-\end{align}
-
-avoids this problem. The resulting model is described by
-
-\begin{align}
-y_t &= l_{t-1} - \sum_{i = 1}^{m_1 - 1}d_{t-i} - \sum_{i = 1}^{m_2 - 1}w_{t-i} + e_t \label{ds-hw-rp1} \\
-l_t &= l_{t-1} + \alpha e_t \label{ds-hw-rp2} \\
-d_t &= - \sum_{i = 1}^{m_1 - 1}d_{t-i} + \delta e_t \label{ds-hw-rp3} \\
-w_t &= - \sum_{i = 1}^{m_2 - 1}w_{t-i} + \omega e_t \label{ds-hw-rp4}.
-\end{align}
-
-\citet{Forbes2000} recommends numerical integration of the marginal posterior density of $\theta$,
-
-\begin{equation}
-\label{exp-sm-marginal}
-p(\theta | y_{1:T}) \propto \left| \widetilde{X}' \widetilde{X} \right|^{-1/2} \tilde{s}^{-(T-(m_1 + m_2 - 1))} p(\theta),
-\end{equation}
-
-where $\widetilde{X}$ and $\tilde{s}$ are functions of $\theta$ and $y$.
-The problems with using this method in our model are two-fold: repeating a three dimensional numerical integrion each time a new data point is observed within the five minute window is not feasible difficult, and the $T$ in the exponent makes most evalutions of $p(\theta | y_{1:T})$ computationally zero when $T$ is large.
-
-\begin{itemize}
-\item Should try and get an exponential smoothing MCMC working using Metropolis-Hastings for $\theta$.
-Log densities help with the $T$ being too large problem but the grid is too large for numerical integration to be a good idea.
-\end{itemize}
-
-% You need a concluding paragraph, in addition to the timeline below.
-\section{Timeline}
-\begin{itemize}
-\item Put something here eventually
+\item DLM: Find something
+\item SV: Intraday stock prices? 
+\item Possibly have a multivariate $\textbf{y}_t$
 \end{itemize}
@@
 
+\section{Discussion}
+
 \end{document}
-\grid

### 2017-03-26 DLM/DLM_SGA_eval.cpp

@@ -1,0 +1,72 @@
+// [[Rcpp::depends(RcppArmadillo)]]
+// [[Rcpp::plugins(cpp11)]]
+
+#include <RcppArmadillo.h>
+#include <math.h>
+using namespace Rcpp;
+using namespace arma;
+using namespace std;
+
+double normDens(double x, double mu, double sigmaSq){
+  return 1 / sqrt(2*3.14159*sigmaSq) * exp(-(x - mu)*(x - mu) / (2*sigmaSq));
+}
+
+// Update VB and take logscore of next observed y J many times
+// Repeat for a range of max update iterations - see how cheap the update needs to be
+// To add: code for restricted updates, will require the SGA algorithm itself to update
+// Should add (or a new function): code for full rank approximation
+// Major change is simulation with dependencies - could simulate whole vector, 
+// or extract relevant parts from covariance matrix - requires an extra TXT matrix multiplication and 5x5 chol decomp
+// [[Rcpp::export]]
+mat ForecastEvalMF(vec y, vec ySupport, vec maxIter, int T, int J, string updates){
+  List initialFit = DLM_SGA(y[T]); // fix this
+  int M = 100;
+  int P = ySupport.n_elem;
+  int N = maxIter.n_elem;
+  vec output(J, N);
+  
+  if(updates == "full"){
+    for(int i = 0; i < N; ++i){
+      for(int t = T; t < T+J; ++t){
+        // Update the VB fit
+        if(t > T){
+          List updatedFit = DLM_SGA(y[T+t], maxIter[i]);
+        }
+        // Draw posterior samples
+        vec phi = phimean + phisd * randn<vec>(M);
+        vec mu = mumean + musd * randn<vec>(M);
+        vec sigmaSqX = exp(sigXmean + sigXsd * randn<vec>(M));
+        vec sigmaSqY =  exp(sigYmean + sigYsd * randn<vec>(M));
+        vec xT = xTmean + xTsd * randn<vec>(M);
+        vec xFuture(M);
+        vec forecastDensity(P, fill::zeros);
+      
+        // Forecast future X value, use to form forecast density for future Y
+        for(int m = 0; m < M; ++m){
+        xFuture[m] = phi[m]*xT[m] + sqrt(sigmqSqX[m])*randn<vec>(1)[0]; 
+        for(int p = 0; p < P; ++p){
+          forecastDensity[p] += normDens(ySupport[p], xFuture[m] + mu[m], sqrt(sigmaSqY[m]));
+        }
+      }
[…]

### 2017-03-26 DLM/test.cpp

@@ -1,0 +1,37 @@
+// [[Rcpp::depends(RcppArmadillo)]]
+// [[Rcpp::plugins(cpp11)]]
+
+#include <RcppArmadillo.h>
+#include <math.h>
+using namespace Rcpp;
+using namespace arma;
+using namespace std;
+
+
+// [[Rcpp::export]]
+vec fullsimul(mat L, vec mu, int T){
+  vec epsilon = randn<vec>(T);
+  vec output(T);
+  for(int i = 0; i < T; ++i){
+    output[i] = mu[i];
+    for(int j = 0; j <= i; ++j){
+      output[i] += L(i, j) * epsilon[j];
+    }
+  }
+  return output;
+}
+
+// [[Rcpp::export]]
+vec partialsimul(mat L, vec mu, int T){
+  double xtsd = sqrt(L.submat(T-1, 4, T-1, T-1) * L.submat(T-1, 4, T-1, T-1).t())[0];
+  vec epsilon = randn<vec>(5);
+  vec output(5);
+  for(int i = 0; i < 4; ++i){
+    output[i] = mu[i];
+    for(int j = 0; j <= i; ++j){
+      output[i] += L(i, j) * epsilon[j];
+    }
+  }
+  output[4] = mu[T] + L(T-1, 0)*epsilon[0] + L(T-1, 1)*epsilon [1] + L(T-1, 2)*epsilon[2] + L(T-1, 3)*epsilon[3] + xtsd*epsilon[4];
+  return output;
+}

### 2017-03-30 DLM/writeup.tex
writeup plus SGA fixes

\documentclass[12pt,a4paper]{article}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage[]{algorithm2e}
\usepackage{amsthm}
\usepackage{url}
\usepackage{wasysym}
\usepackage{ulem}
\usepackage{afterpage}
\usepackage{bbm}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}

\begin{document}

\section{Introduction}

Consider a high frequency time series with observed values $y_{1}, \dots, y_{T}$, denoted by $y_{1:T}$, with a forecaster interested in density forecasts for the unknown future values $y_{T+1:T+J}$. Many models of interest contain a set of global parameters $\boldsymbol{\theta}$ and a observation specific dependent latent variables $\textbf{x}_{1:T}$ such that $\textbf{x}_t$ is dependent on $\textbf{x}_{t-1}$. In this case, forecasts of $y_{T+J} | y_{T}$ where $J > 1$ depend on estimates of the posterior distribution of $\textbf{x}_{T+1:T+J} | \theta, y_{1:T}, \textbf{x}_{1:T}$, which can be greatly improved by observation of $y_{T+1:T+J-1}$. With certain restrictions on the model, updated estimates of the latent states after observing more data can be easily obtained via filtering techniques such as the Kalman filter, however in many cases these restrictions are unsuitable and a computationally costly re-estimation of parameters and latent states must be undertaken.

One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution. MCMC can have slow convergence rates and generally does not admit parallel processing so that updating the forecast distribution of $y_{T+J} | y_{1:T}$ to $y_{T+J} | y_{1:T+J-1}$ after observation of $y_{T+1}, \dots, y_{T+J-1}$ may not be available before $y_{T+J}$ is observed.

This paper explores the use of Variational Bayes to approximate the posterior distribution of several common time-series models with dependent latent states, and compares the increased speed of estimation with the decreased statistical accuracy associated with using an approximation to the posterior distribution. We compare several different functional forms of the approximating distribution that are commonly used in the machine learning literature and explore the use of MCMC samples to inform the design of the functional form.

\section{Variational Bayes}

Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T}$ and some approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.

We follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$. The KL divergence is defined by
\begin{equation}
\label{KL-def}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}
\end{equation}

and can alternatively be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})
\end{equation}
where $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
\end{equation}

and maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}). (\ref{ELBO}) can be maximised by gradient ascent, where we iteratively apply the following update step until (\ref{ELBO}) converges within some pre-specified tolerance:
\begin{equation}
\label{GradAscent}
\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
\end{equation}
$\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$. This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum (Robbins and Munro, 1951).
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}

Ranganath, Gerrish and Blei (2014) showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
\begin{equation}
\label{ScoreDeriv}
\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)}))] (\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)})))
\end{equation}
where $s = 1, \dots, S$ indicates simulations from $q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})$.

The estimator in (\ref{ScoreDeriv}) has a large variance which requires a large value of $S$, slowing computation. The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), which introduces parameter free noise variables $\boldsymbol{\epsilon}$ and an invertable transform $f$ such that

\begin{equation}
\label{Reparam}
\boldsymbol{\theta} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}.
\end{equation}

Examples of $f$ include a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable. In this case

\begin{equation}
\label{ReparamDist}
\log(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})) = \log(|J|) + \log(q(\boldsymbol{\epsilon}))
\end{equation}

where $J$ is the Jacobian Matrix of the transformation $f$, a function of $\boldsymbol{\lambda}$ and $\boldsymbol{\epsilon}$. The derivative of (\ref{ELBO}) can be estimated by

\begin{equation}
\label{ReparamDeriv}
\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s))) - \nabla_{\boldsymbol{\lambda}}\log(|J(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s)|),
\end{equation}
where simulations of $\boldsymbol{\theta}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of this estimator is often orders of magnitude lower than in (\ref{ScoreDeriv}) and in many cases $S=1$ is sufficient.

\section{The Dynamic Linear Model}

We begin with the Dynamic Linear Model, which fulfills the neccesary conditions for the Kalman filter to be applicable: (\ref{DLM:Measure}) and (\ref{DLM:Transition}) are linear with gaussian errors.

\begin{align}
y_t &= \mu + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
\end{align}

where $\epsilon_t$ and $\nu_t$ are independently drawn from $\mathcal{N}(0, 1).$

We use the priors

\begin{align}
\mu &\sim \mathcal{N}(\bar{\mu}, \tau) \\
\phi &\sim U(-1, 1) \\
\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
\end{align}

where $\mathcal{IG}(\mbox{shape}, \mbox{ scale})$ denotes the inverse of a Gamma (shape, rate) distribution.

We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\mu, \phi, \log(\sigma^2_y), \log(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$:
\begin{enumerate}
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \textbf{V})$, where $\textbf{V}$ is a diagonal matrix.
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \boldsymbol{\Sigma})$, where $\boldsymbol{\Sigma}$ allows non-zero off-diagonal terms.
\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$ is an Inverse Autoregressive flow (Kingma, Salimans, Jozefowicz, Chen, Sutskever and Welling, 2016).
\item  $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \prod \left(q(\theta_i) q(x_{it})\right) c(\boldsymbol{\theta}, \textbf{x})$ where the form of the copula $c$ and marginals are informed by an out of date MCMC sample.
\end{enumerate}
The first three functional forms are transformation of standard normal noise variables while the fourth is a transformation of uniform noise variables.

We use two approaches for updating forecasts with MCMC,
\begin{enumerate}
\item Use the Kalman filter to update $\textbf{x}_{T+1:T+J}$ as $y_{T+1:T+J}$ is observed using $\boldsymbol{\theta}$ from $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T}$
\item Run the full MCMC for $\boldsymbol{\theta}$ and $\textbf{x}_{1:T+J}$ after observing each additional $y$.
\end{enumerate}
The second strategy has a high computational cost but is required to give us the true forecast distribution for comparision with other methods.

We use three approaches for updating forecasts with VB,
\begin{enumerate}
\item Use the Kalman filter to update $\textbf{x}_{T+1:T+J}$ as $y_{T+1:T+J}$ is observed using $\boldsymbol{\theta}$ from $q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})$.
\item Hold elements of $\boldsymbol{\lambda}$ for $x_{1:T+h-1}$ fixed and maximise (\ref{ELBO}) with respect to the elements of $\boldsymbol{\lambda}$ for $\boldsymbol{\theta}$ and $x_{T+h}$ after observation of $y_{T+h}$.
\item Maximise (\ref{ELBO}) with respect to all elements of $\boldsymbol{\lambda}$
\end{enumerate}

\end{document}

### 2017-03-31 DLM/writeup.tex
writeup changes

@@
 \label{Reparam}
-\boldsymbol{\theta} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}.
+\boldsymbol{\theta} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
 \end{equation}
@@
 \label{ReparamDist}
-\log(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})) = \log(|J|) + \log(q(\boldsymbol{\epsilon}))
+\log(q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \log(|J|) + \log(q(\boldsymbol{\epsilon}))
 \end{equation}
 
-where $J$ is the Jacobian Matrix of the transformation $f$, a function of $\boldsymbol{\lambda}$ and $\boldsymbol{\epsilon}$. The derivative of (\ref{ELBO}) can be estimated by
+where $J$ is the Jacobian Matrix of the transformation $f$, which may depend on $\boldsymbol{\lambda}$ and $\boldsymbol{\epsilon}$. The derivative of (\ref{ELBO}) can be estimated by
 
@@
 where simulations of $\boldsymbol{\theta}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
-The variance of this estimator is often orders of magnitude lower than in (\ref{ScoreDeriv}) and in many cases $S=1$ is sufficient.
+The variance of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}), often $S=1$ is sufficient.
 
@@
 \begin{align}
-y_t &= \mu + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
+y_t &= \gamma + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
 x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
@@
 \begin{align}
-\mu &\sim \mathcal{N}(\bar{\mu}, \tau) \\
-\phi &\sim U(-1, 1) \\
+\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
+\phi &\sim \mathcal{U}(-1, 1) \\
 \sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
@@
 
-We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\mu, \phi, \log(\sigma^2_y), \log(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$:
+We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\gamma, \phi, \log(\sigma^2_y), \log(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$:
 \begin{enumerate}
@@
 
+\section{Appendix}
+
+\subsection{Variational Approximations for the Dynamic Linear Model}
+
+The joint distribution $p(y_{1:T}, \boldsymbol{\theta}, x_{1:T})$ for the Dynamic Linear Model is given by
+
+\begin{align}
+p(y_{1:T}, \boldsymbol{\theta}, x_{1:T}) &= \prod_{t=1}^{T} \left( \mathcal{N}(y_t | x_t + \gamma, \sigma^2_y) \mathcal{N}(x_t | \phi x_{t-1}, \sigma^2_x) \right) \mathcal{N}(x_0 | 0, \sigma^2_x (1 - \phi^2)^{-1}) \nonumber \\
+&\mathcal{N}(\gamma | \bar{\gamma}, \sigma^2_{\gamma}) \mathcal{U}(\phi | -1, 1) \mathcal{IG}(\sigma^2_y | \alpha_y, \beta_y)\mathcal{IG}(\sigma^2_x | \alpha_x, \beta_x) 
+\end{align}
+with 
+\begin{align}
+\log(p(y_{1:T}, \boldsymbol{\theta}, x_{1:T})) &= -(T/2 + \alpha_y + 1) \log(\sigma_y^2) -(T/2 + \alpha_x + 3/2) \log(\sigma_x^2) \nonumber \\
+&- \frac{\sum_{t=1}^{T}(y_t - \gamma - x_t)^2}{2 \sigma^2_y} - \frac{x_0^2(1 -\phi^2) + \sum_{t=1}^{T}(x_t - \phi x_{t-1})^2}{2 \sigma^2_x} \nonumber \\
+&- \frac{(\gamma - \bar{\gamma})^2}{2 \sigma^2_{\gamma}} - \frac{\beta_y}{\sigma^2_y} - \frac{\beta_x}{\sigma^2_x} + c \label{logjoint}. 
+\end{align}
+
+We use Kingma and Welling's (2014) reparameterised approach to the variational approximation in all of our applications.
+
+\subsubsection{The Multivariate Normal Approximations}
+
+The distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ for the first two approaches is defined by the following transformation, where $\epsilon_{1}, \dots, \epsilon_{T+5}$ have independent standard normal distributions.
+\begin{align}
+\sigma^2_y &= \exp(\mu_1 + L_{11} \epsilon_1)  \label{transform1} \\
+\sigma^2_x &= \exp(\mu_2 + L_{21} \epsilon_1 + L_{22} \epsilon_2) \label{transform2} \\
+\phi &= \mu_3 + L_{31} \epsilon_1 + L_{32} \epsilon_2 + L_{33} \epsilon_3 \label{transform3} \\
+\mu &= \mu_4 + L_{41} \epsilon_1 + L_{42} \epsilon_2 + L_{43} \epsilon_3 + L_{44} \epsilon_4 \label{transform4} \\
+x_t &= \mu_{t+5} + \sum_{i=1}^{t+5} L_{t+5, i} \epsilon_i \mbox{ for } t = 0, \dots, T \label{transform5} 
+\end{align}
+
+The approximation is parameterised by $\boldsymbol{\lambda} = \{ \boldsymbol{\mu}, \textbf{L} \}$ where $\textbf{L}$ is the lower triangular decomposition of the Multivariate Normal variance matrices $\textbf{V}$ and $\boldsymbol{\Sigma}$.
+In the case where the diagonal matrix $\textbf{V}$ is used, restrict $L_{ij} = 0$ for $i \neq j$.
+The log determinant of the Jacobian is given by
+\begin{equation}
+\label{logdetjac}
+\log(|J|) = - \left(\sum_{i = 1}^{i = T+5} \log(L_{ii}) + \mu_1 + L_{11} \epsilon_1 + \mu_2 + L_{21} \epsilon_1 + L_{22} \epsilon_2\right).  
+\end{equation}
+
+Derivatives of (\ref{logjoint}) with respect to $\boldsymbol{\theta}$ and $x_{0:T}$, as well as derivatives of (\ref{transform1}) - (\ref{logdetjac}) with respect to $\boldsymbol{\mu}$ and $\textbf{L}$ are easily obtained for use in estimating the gradient of the ELBO.
+
 \end{document}

### 2017-04-03 DLM/writeup.tex
Writeup and SGA fixes

@@
 
-Consider a high frequency time series with observed values $y_{1}, \dots, y_{T}$, denoted by $y_{1:T}$, with a forecaster interested in density forecasts for the unknown future values $y_{T+1:T+J}$.
-Many models of interest contain a set of global parameters $\boldsymbol{\theta}$ and a observation specific dependent latent variables $\textbf{x}_{1:T}$ such that $\textbf{x}_t$ is dependent on $\textbf{x}_{t-1}$.
-In this case, forecasts of $y_{T+J} | y_{T}$ where $J > 1$ depend on estimates of the posterior distribution of $\textbf{x}_{T+1:T+J} | \theta, y_{1:T}, \textbf{x}_{1:T}$, which can be greatly improved by observation of $y_{T+1:T+J-1}$.
-With certain restrictions on the model, updated estimates of the latent states after observing more data can be easily obtained via filtering techniques such as the Kalman filter, however in many cases these restrictions are unsuitable and a computationally costly re-estimation of parameters and latent states must be undertaken.
+Consider a high frequency time series with observed values $y_{1}, \dots, y_{T}$, collectively denoted by $y_{1:T}$, with a forecaster interested in the density forecast of the unknown future value $y_{T+S+h}$, where $S \geq 1$.
+Assuming that a forecast of $p(y_{T+S+h} | y_{1:T})$ is available, uncertainty in the forecast can be reduced by observating and conditioning on the realisations of $y_{T+1:T+S}$ to produce $p(y_{T+S+h} | y_{T+S})$.
+
+Many models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $\textbf{x}_{1:T}$ such that $\textbf{x}_t$ is assumed to follow a Markovian structure.
+In this case $y_{T+S+h}$ depends on the latent variables $\textbf{x}_{T+1:T+S+h}$, and the forecast uncertainty in $y_{T+S+h}$ can be reduced by first incorporating information contained in realisations of $y_{T+1:T+S}$ to form the posterior densities of $\textbf{x}_{T+1:T+S+h} | \textbf{x}_{1:T}, \boldsymbol{\theta}, y_{1:T+S}$.
+With certain restrictions on the model, the additional information can be easily included via filtering techniques such as the Kalman filter, however in many cases these restrictions are unsuitable and a computationally costly re-estimation of posterior densities of parameters and latent states must be undertaken.
 
 One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution.
-MCMC can have slow convergence rates and generally does not admit parallel processing so that updating the forecast distribution of $y_{T+J} | y_{1:T}$ to $y_{T+J} | y_{1:T+J-1}$ after observation of $y_{T+1}, \dots, y_{T+J-1}$ may not be available before $y_{T+J}$ is observed.
+MCMC can have slow convergence rates and generally does not admit parallel processing so that updating the forecast distribution of $y_{T+S+h} | y_{1:T}$ to $y_{T+S+h} | y_{1:T+S}$ after observation of $y_{T+1}, \dots, y_{T+S}$ may not be available before $y_{T+S+h}$ is observed.
 
-This paper explores the use of Variational Bayes to approximate the posterior distribution of several common time-series models with dependent latent states, and compares the increased speed of estimation with the decreased statistical accuracy associated with using an approximation to the posterior distribution.
+This paper explores the use of Variational Bayes to approximate the posterior distribution of several common time-series models with dependent latent states, and compares the increased speed of computation with the decreased statistical accuracy associated with using an approximation to the posterior distribution.
 We compare several different functional forms of the approximating distribution that are commonly used in the machine learning literature and explore the use of MCMC samples to inform the design of the functional form.
@@
 
-We follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta} | \boldsymbol{\lambda})$ the true posterior $p(\boldsymbol{\theta} | y_{1:T})$.
+We follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, \textbf{x}_{1:T}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})$.
 The KL divergence is defined by
@@
 \label{KL-def}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta} |y_{1:T})}\right) d\boldsymbol{\theta}
+KL[q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})] = \int q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, \textbf{x}_{1:T} |y_{1:T})}\right) d\boldsymbol{\theta} d\textbf{x}_{1:T}
 \end{equation}
-
 and can alternatively be expressed as
@@
 \label{KL-ELBO}
-KL[q(\boldsymbol{\theta} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})
+KL[q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T})
 \end{equation}
-where $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.
-$\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
+where $\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.
+$\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
 \begin{equation}
 \label{ELBO}
-\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (p(y_{1:T},\boldsymbol{\theta})) d\boldsymbol{\theta} -  \int q(\boldsymbol{\theta} | \boldsymbol{\lambda}) \ln (q(\boldsymbol{\theta} | \boldsymbol{\lambda})) d\boldsymbol{\theta}.
+\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T},\boldsymbol{\theta}, \textbf{x}_{1:T})}{q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}d\textbf{x}_{1:T},
 \end{equation}
-
-and maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).
-(\ref{ELBO}) can be maximised by gradient ascent, where we iteratively apply the following update step until (\ref{ELBO}) converges within some pre-specified tolerance:
+and as $\ln(p(y_{1:T}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).
+Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where we iteratively apply the following update step until (\ref{ELBO}) converges within some pre-specified tolerance:
 \begin{equation}
 \label{GradAscent}
-\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
+\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
 \end{equation}
-$\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$.
+$\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$.
 This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
@@
 \label{ScoreDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)}))] (\ln (p(y_{1:T}, \boldsymbol{\theta}_s)) - \ln(q(\boldsymbol{\theta}_s | \boldsymbol{\lambda}^{(m)})))
+\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s, \textbf{x}_{1:T, s} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_s, \textbf{x}_{1:T, s})}{q(\boldsymbol{\theta}_s, \textbf{x}_{1:T, s} | \boldsymbol{\lambda}^{(m)})} \right)
 \end{equation}
-where $s = 1, \dots, S$ indicates simulations from $q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)})$.
+where $s = 1, \dots, S$ indicates simulations from $q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)})$. 
+The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $S$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.
 
-The estimator in (\ref{ScoreDeriv}) has a large variance which requires a large value of $S$, slowing computation.
-The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), which introduces parameter free noise variables $\boldsymbol{\epsilon}$ and an invertable transform $f$ such that
-
+The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), which introduces parameter free noise variables $\boldsymbol{\epsilon}$ and a differentiable transform $f$ such that
 \begin{equation}
 \label{Reparam}
-\boldsymbol{\theta} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
+\boldsymbol{\theta}, \textbf{x}_{1:T} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
 \end{equation}
-
-Examples of $f$ include a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable. In this case
-
+Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable.
+Transforming the variables using
+\begin{equation}
+\label{changevar}
+\int q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}) g(\boldsymbol{\theta}, \textbf{x}_{1:T})  d\boldsymbol{\theta}d\textbf{x}_{1:T} = \int q(\boldsymbol{\epsilon}) g ( f (\epsilon, \boldsymbol{\lambda})) d\boldsymbol{\epsilon})
+\end{equation}
+and noting that
 \begin{equation}
 \label{ReparamDist}
-\log(q(\boldsymbol{\theta} | \boldsymbol{\lambda}) = \log(|J|) + \log(q(\boldsymbol{\epsilon}))
+\ln(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})) = \ln(q(\boldsymbol{\epsilon})) - \ln(|J|),
 \end{equation}
+where $J$ is the Jacobian Matrix of the transformation $f$, allows (\ref{ELBO}) to be rewritten as
+\begin{align}
+\label{reparamELBO}
+\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T}) &=  \int  q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda}))}{q( f (\epsilon, \boldsymbol{\lambda}) | \boldsymbol{\lambda})} \right) d\boldsymbol{\epsilon} \nonumber \\
+&= \int  q(\boldsymbol{\epsilon}) \big( \ln (p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda})) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) \big) d\boldsymbol{\epsilon} \nonumber \\
+&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T}).
+\end{align}
 
-where $J$ is the Jacobian Matrix of the transformation $f$, which may depend on $\boldsymbol{\lambda}$ and $\boldsymbol{\epsilon}$. The derivative of (\ref{ELBO}) can be estimated by
-
+The derivative of (\ref{reparamELBO}) can be estimated by
 \begin{equation}
 \label{ReparamDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s))) - \nabla_{\boldsymbol{\lambda}}\log(|J(\boldsymbol{\lambda}, \boldsymbol{\epsilon}_s)|),
+\nabla_{\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_s))) + \nabla_{\boldsymbol{\lambda}}\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_s)|),
 \end{equation}
-where simulations of $\boldsymbol{\theta}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
-The variance of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}), often $S=1$ is sufficient.
+where simulations of $\boldsymbol{\theta}$ and $\textbf{x}_{1:t}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
+The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $S = 1$ provides a precise enough estimate of the gradient for fast convergence.
+
+\section{MCMC Informed Variational Bayes}
 
@@
 
-We begin with the Dynamic Linear Model, which fulfills the neccesary conditions for the Kalman filter to be applicable: (\ref{DLM:Measure}) and (\ref{DLM:Transition}) are linear with gaussian errors.
+We begin with the Dynamic Linear Model, which fulfills the neccesary conditions for the Kalman filter to be applicable: (\ref{DLM:Measure}) and (\ref{DLM:Transition}) are linear with Gaussian errors.
 
@@
 \end{align}
-
-where $\epsilon_t$ and $\nu_t$ are independently drawn from $\mathcal{N}(0, 1).$
+where $\epsilon_t$ and $\nu_t$ are independent $\mathcal{N}(0, 1)$ random variables.
 
@@
 \end{align}
-
 where $\mathcal{IG}(\mbox{shape}, \mbox{ scale})$ denotes the inverse of a Gamma (shape, rate) distribution.
 
-We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\gamma, \phi, \log(\sigma^2_y), \log(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$:
+We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})$:
 \begin{enumerate}
@@
 
-We use two approaches for updating forecasts with MCMC,
+As additional data $y_{T+1:T+J-1}$ are observed, we can incorporate this information to improve forecasts of the future value of $y_{T+J}$. We use two MCMC based approaches for updating this forecast,
 \begin{enumerate}
-\item Use the Kalman filter to update $\textbf{x}_{T+1:T+J}$ as $y_{T+1:T+J}$ is observed using $\boldsymbol{\theta}$ from $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T}$
+\item Use the Kalman filter to update $\textbf{x}_{T+1:T+J}$ as $y_{T+1:T+J}$ is observed using $\boldsymbol{\theta}$ from $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})$.
+This method is computationally simple, but does not allow the posterior distribution of $\boldsymbol{\theta}$ to be conditioned on new observations.
 \item Run the full MCMC for $\boldsymbol{\theta}$ and $\textbf{x}_{1:T+J}$ after observing each additional $y$.
+This strategy has a high computational cost, but does condition $\boldsymbol{\theta}$ on $y_{T+1:T+J-1}$, and results in the forecast distribution with the least uncertainty.
 \end{enumerate}
-The second strategy has a high computational cost but is required to give us the true forecast distribution for comparision with other methods.
 
-We use three approaches for updating forecasts with VB,
+We use three approaches for updating forecasts with VB, listed in increasing computational complexity:
 \begin{enumerate}
@@
 \end{enumerate}
+
+Forecast densities of VB and MCMC can be compared by a numeric calculation of the Kullback-Leibler divergence of samples from both $q$ and $p$, and different updating strategies can be compared by the log-score of a held out $y_{T+J}$ value.
 
@@
 \begin{align}
-p(y_{1:T}, \boldsymbol{\theta}, x_{1:T}) &= \prod_{t=1}^{T} \left( \mathcal{N}(y_t | x_t + \gamma, \sigma^2_y) \mathcal{N}(x_t | \phi x_{t-1}, \sigma^2_x) \right) \mathcal{N}(x_0 | 0, \sigma^2_x (1 - \phi^2)^{-1}) \nonumber \\
+p(y_{1:T}, \boldsymbol{\theta}, x_{1:T}) &= \prod_{t=1}^{T} \bigg( \mathcal{N}(y_t | x_t + \gamma, \sigma^2_y) \mathcal{N}(x_t | \phi x_{t-1}, \sigma^2_x) \bigg) \mathcal{N}(x_0 | 0, \sigma^2_x (1 - \phi^2)^{-1}) \nonumber \\
 &\mathcal{N}(\gamma | \bar{\gamma}, \sigma^2_{\gamma}) \mathcal{U}(\phi | -1, 1) \mathcal{IG}(\sigma^2_y | \alpha_y, \beta_y)\mathcal{IG}(\sigma^2_x | \alpha_x, \beta_x) 
@@
 \begin{align}
-\log(p(y_{1:T}, \boldsymbol{\theta}, x_{1:T})) &= -(T/2 + \alpha_y + 1) \log(\sigma_y^2) -(T/2 + \alpha_x + 3/2) \log(\sigma_x^2) \nonumber \\
+\ln(p(y_{1:T}, \boldsymbol{\theta}, x_{1:T})) &= -(T/2 + \alpha_y + 1) \ln(\sigma_y^2) -(T/2 + \alpha_x + 3/2) \ln(\sigma_x^2) \nonumber \\
 &- \frac{\sum_{t=1}^{T}(y_t - \gamma - x_t)^2}{2 \sigma^2_y} - \frac{x_0^2(1 -\phi^2) + \sum_{t=1}^{T}(x_t - \phi x_{t-1})^2}{2 \sigma^2_x} \nonumber \\
@@
 
-We use Kingma and Welling's (2014) reparameterised approach to the variational approximation in all of our applications.
+We use Kingma and Welling's (2014) reparameterised approach to the variational approximation in all of our applications, with the transforms detailed in Sections 4.1.1 - 4.1.3.
 
@@
 \label{logdetjac}
-\log(|J|) = - \left(\sum_{i = 1}^{i = T+5} \log(L_{ii}) + \mu_1 + L_{11} \epsilon_1 + \mu_2 + L_{21} \epsilon_1 + L_{22} \epsilon_2\right).  
+\ln(|J|) = \left(\sum_{i = 1}^{i = T+5} \ln(L_{ii}) + \mu_1 + L_{11} \epsilon_1 + \mu_2 + L_{21} \epsilon_1 + L_{22} \epsilon_2\right).  
 \end{equation}
 
-Derivatives of (\ref{logjoint}) with respect to $\boldsymbol{\theta}$ and $x_{0:T}$, as well as derivatives of (\ref{transform1}) - (\ref{logdetjac}) with respect to $\boldsymbol{\mu}$ and $\textbf{L}$ are easily obtained for use in estimating the gradient of the ELBO.
+Derivatives of (\ref{logjoint}) with respect to $\boldsymbol{\theta}$ and $\textbf{x}_{0:T}$, as well as derivatives of (\ref{transform1}) - (\ref{logdetjac}) with respect to $\boldsymbol{\mu}$ and $\textbf{L}$ are easily obtained for use in estimating the gradient of the ELBO.
 

### 2017-04-03 DLM/DLM_SGA_FR.cpp

@@ -13,27 +13,23 @@ using namespace std;
 // Find out how to add distribution code as a header to reduce amount of code in one file
 // Make phi have a truncated normal marginal so I can fix the x0 prior variance 
 // Can repeatedly sample from the MVN until we get a draw in (-1, 1) but I'd like to avoid while loops when the mean of phi can go way outside (-1, 1). Could restrict the mean?
-// Find out why lambda values it goes all over the place. Mean of Mu seems to be going to the right place but everything else does it's own thing.
-// Find out how to avoid dynamic casting -> makes it easier to implement new models but if I'm sticking with DLM's for a while it doesn't really matter
+// Find out how to avoid dynamic casting to make it easier to implement new models, but if I'm sticking with DLM's for a while this is not a problem yet
 
 class Distribution{
   // This is the base class for the distributions, and exists so I can have Normal and Inverse Gamma pointers in the same array
-  public:
-  // These functions are only really here so I can call the member function versions
+public:
+  // This function is only here so I can call any of the subclass versions.
   // The code won't compile unless I have a Distribution member function and the subclass member function.
-  // There should be a way around this, I can use dynamic cast to tell the compiler I want to explicitly use a normal sampler for example
+  // There should be a way around this, I can use dynamic cast to tell the compiler I want to explicitly use a normal logdens for example
   // But without the dynamic casting it doesn't know what version of the distribution I want to use, so I need something here
   // It gets overwritten by the subclass version anyway.
   // Mostly I just haven't figured out enough c++ to get this working without dynamic casts
-  virtual void set_values(int, int) {};
   virtual double logdens(double x) {return -99.99;};
-  virtual vec sample (int n_samples) {return zeros<vec>(n_samples);};
 };
 
 class Normal: public Distribution{
   // This just provides methods for the single variate normal distribution
-  public:
[…]

### 2017-04-27 DLM/writeup.tex
Added the SVM model

@@
 \usepackage{amsthm}
+\usepackage{tikz}
+\usetikzlibrary{bayesnet}
 \usepackage{url}
@@
 
-Many models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $\textbf{x}_{1:T}$ such that $\textbf{x}_t$ is assumed to follow a Markovian structure.
-In this case $y_{T+S+h}$ depends on the latent variables $\textbf{x}_{T+1:T+S+h}$, and the forecast uncertainty in $y_{T+S+h}$ can be reduced by first incorporating information contained in realisations of $y_{T+1:T+S}$ to form the posterior densities of $\textbf{x}_{T+1:T+S+h} | \textbf{x}_{1:T}, \boldsymbol{\theta}, y_{1:T+S}$.
+Many models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{1:T}$ such that $x_t$ is assumed to follow a Markovian structure.
+In this case $y_{T+S+h}$ depends on the latent variables $x_{T+1:T+S+h}$, and the forecast uncertainty in $y_{T+S+h}$ can be reduced by first incorporating information contained in realisations of $y_{T+1:T+S}$ to form the posterior densities of $x_{T+1:T+S+h} | x_{1:T}, \boldsymbol{\theta}, y_{1:T+S}$.
 With certain restrictions on the model, the additional information can be easily included via filtering techniques such as the Kalman filter, however in many cases these restrictions are unsuitable and a computationally costly re-estimation of posterior densities of parameters and latent states must be undertaken.
@@
 MCMC can have slow convergence rates and generally does not admit parallel processing so that updating the forecast distribution of $y_{T+S+h} | y_{1:T}$ to $y_{T+S+h} | y_{1:T+S}$ after observation of $y_{T+1}, \dots, y_{T+S}$ may not be available before $y_{T+S+h}$ is observed.
+In this paper we assume that an MCMC will converging before time period $T + S + h$ will be required to begin, at the latest, by time $T$ and hence $y_{T}$ is the last observation that can be included in the MCMC.
 
-This paper explores the use of Variational Bayes to approximate the posterior distribution of several common time-series models with dependent latent states, and compares the increased speed of computation with the decreased statistical accuracy associated with using an approximation to the posterior distribution.
-We compare several different functional forms of the approximating distribution that are commonly used in the machine learning literature and explore the use of MCMC samples to inform the design of the functional form.
+Increasing the speed of estimation will allow more data up to time $T + S$ to be included in the forecast distribution, and this paper explores the use of Variational Bayes to approximate the posterior distribution of several common time-series models with dependent latent states.
+The the increase in computation speed and resulting increase in observations included in a forecast is contrasted with the decreased statistical accuracy associated with using an approximation to the posterior distribution.
+We compare several different functional forms of the approximating distribution that are commonly used in the machine learning literature and explore the use of previous MCMC samples to inform the design of the functional form.
 
@@
 
-Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T}$ and some approximating distribution $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.
+Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T}$ and some approximating distribution $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.
 
-We follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, \textbf{x}_{1:T}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})$.
+We follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, x_{1:T}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$.
 The KL divergence is defined by
@@
 \label{KL-def}
-KL[q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})] = \int q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, \textbf{x}_{1:T} |y_{1:T})}\right) d\boldsymbol{\theta} d\textbf{x}_{1:T}
+KL[q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})] = \int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T} |y_{1:T})}\right) d\boldsymbol{\theta} dx_{1:T}
 \end{equation}
@@
 \label{KL-ELBO}
-KL[q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T})
+KL[q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})
 \end{equation}
-where $\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.
-$\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
+where $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.
+$\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
 \begin{equation}
 \label{ELBO}
-\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T},\boldsymbol{\theta}, \textbf{x}_{1:T})}{q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}d\textbf{x}_{1:T},
+\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T},\boldsymbol{\theta}, x_{1:T})}{q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T},
 \end{equation}
 and as $\ln(p(y_{1:T}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).
+
 Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where we iteratively apply the following update step until (\ref{ELBO}) converges within some pre-specified tolerance:
@@
 \label{GradAscent}
-\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
+\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
 \end{equation}
-$\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$.
+$\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$.
 This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
@@
 \label{ScoreDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_s, \textbf{x}_{1:T, s} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_s, \textbf{x}_{1:T, s})}{q(\boldsymbol{\theta}_s, \textbf{x}_{1:T, s} | \boldsymbol{\lambda}^{(m)})} \right)
+\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_i, x_{1:T, i})}{q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)})} \right)
 \end{equation}
-where $s = 1, \dots, S$ indicates simulations from $q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}^{(m)})$. 
-The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $S$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.
+where $i = 1, \dots, N$ indicates simulations from $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)})$. 
+The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $N$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.
 
@@
 \label{Reparam}
-\boldsymbol{\theta}, \textbf{x}_{1:T} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
+\boldsymbol{\theta}, x_{1:T} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
 \end{equation}
@@
 \label{changevar}
-\int q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}) g(\boldsymbol{\theta}, \textbf{x}_{1:T})  d\boldsymbol{\theta}d\textbf{x}_{1:T} = \int q(\boldsymbol{\epsilon}) g ( f (\epsilon, \boldsymbol{\lambda})) d\boldsymbol{\epsilon})
+\int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) g(\boldsymbol{\theta}, x_{1:T})  d\boldsymbol{\theta}dx_{1:T} = \int q(\boldsymbol{\epsilon}) g ( f (\epsilon, \boldsymbol{\lambda})) d\boldsymbol{\epsilon})
 \end{equation}
@@
 \label{ReparamDist}
-\ln(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})) = \ln(q(\boldsymbol{\epsilon})) - \ln(|J|),
+\ln(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})) = \ln(q(\boldsymbol{\epsilon})) - \ln(|J|),
 \end{equation}
@@
 \label{reparamELBO}
-\mathcal{L}(q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda}), y_{1:T}) &=  \int  q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda}))}{q( f (\epsilon, \boldsymbol{\lambda}) | \boldsymbol{\lambda})} \right) d\boldsymbol{\epsilon} \nonumber \\
+\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) &=  \int  q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda}))}{q( f (\epsilon, \boldsymbol{\lambda}) | \boldsymbol{\lambda})} \right) d\boldsymbol{\epsilon} \nonumber \\
 &= \int  q(\boldsymbol{\epsilon}) \big( \ln (p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda})) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) \big) d\boldsymbol{\epsilon} \nonumber \\
@@
 \label{ReparamDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T}) \approx \frac{1}{S}\sum_{s=1}^{S} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_s))) + \nabla_{\boldsymbol{\lambda}}\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_s)|),
+\nabla_{\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i))) + \nabla_{\boldsymbol{\lambda}}\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|),
 \end{equation}
-where simulations of $\boldsymbol{\theta}$ and $\textbf{x}_{1:t}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
-The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $S = 1$ provides a precise enough estimate of the gradient for fast convergence.
+where simulations of $\boldsymbol{\theta}$ and $x_{1:t}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
+The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $N = 1$ provides a precise enough estimate of the gradient for fast convergence.
 
 \section{MCMC Informed Variational Bayes}
+
+A key decision in Variational Bayes is the choice of the functional form of the approximating distribution $q$.
+Much of the literature focuses on a generic form of $q$ that can flexibly be applied to many models.
+However the time-series nature of our application implies that MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$, which provides useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ for use in the approximate predictive density, $q(y_{T+S+h} | y_{1:T+S})$.
+There are two stages to this construction:
+\begin{enumerate}
+\item Find a well fitting functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ using the MCMC results.
+\item Extrapolate the form of $q$ used for $x_{1:T}$ to $x_{1:T+S}$.
+\end{enumerate}
+Treating the MCMC samples as 'data', the functional form of $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ can be approached as a traditional distribution fitting problem.
+Optimal marginal distributions are selected for each parameter by AIC, while a vine copula modelling approach is adopted for the dependence structure following Tran, Blei and Airoldi (2015).
+Sklar (1959) proves that any joint probability distribution with $k$ variables can be written as the product of the marginals and a copula function,
+\begin{equation}
+\label{vc1}
+p(\omega_1, \dots, \omega_k) = p(\omega_1) \dots p(\omega_k) c(P(\omega_1), \dots, P(\omega_k))
+\end{equation}
+where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.
+A vine copula is a structure to factorise this copula, increasing the flexibility given to the modeller, and is defined by a set of trees $\mathcal{V} = \{T_1, \dots, T_{k-1} \}$, where
+\begin{enumerate}
+\item $T_1$ has $k$ nodes corresponding to the $k$ variables, with $k-1$ edges.
+\item $T_t, t = 2, \dots, p-1$ has nodes corresponding to the edges in tree $T_{t-1}$
+\item Two nodes in tree $T_t$ are joined by an edge only if the associated edges in $T_{t-1}$ share a node. 
+\end{enumerate}
+Each edge in the vine corresponds to a bivariate copula conditioned on any variables that are shared between nodes, and the product of each of these $k(k-1)/2$ bivariate copulas forms the complete copula.
+Figure (\ref{fig:vinecop}) provides an example for a four dimensional $\theta$.
+\begin{figure}[h]
+\centering
+\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
+\vspace{2mm}
+\caption{An example of a Vine Copula over four variables.
+The edges in the top tree represent unconditional bivariate copulas between the parameters on the connected nodes.
+The second tree edges represent the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node.
+The edge in the final tree represents the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.}
+\label{fig:vinecop}
+\end{figure}
+
+The Markovian structure of the latent states in the models used provides additional information that can be used in the structure of the vine.
+Denoting the number of elements in $\boldsymbol{\theta}$ with $k$, allow the first $k$ trees to contain edges that correspond to copulas between different elements of $\boldsymbol{\theta}$, and an element of $\boldsymbol{\theta}$ and $x_t$.
+The $k+1'th$ tree edges correspond to the copulas $c(x_0, x_1 | \boldsymbol{\theta}, c(x_1, x_2 | \boldsymbol{\theta}), \dots, c(x_{T-1}, x_T | \boldsymbol{\theta}$ and each further tree contains only bivariate copulas between $x_t$ and $x_{t+j}$ conditioned on $x_{t+l}$, where $j > l \geq 1$.
+Markovian latent states implies that $x_t$ and $x_{t+2}$ are conditionally independent given $x_{t+1}$, so each copula in tree $k+2$ onwards can be set to the independence copula, reducing the number of copula parameters in the vine to be linear in $T$ instead of quadratic.
+To allow the vine structure to be easily extended to new latent variables $x_{T+1:T+S}$, each $x_t$ node should have edges to the same $\boldsymbol{\theta}$ node.
+Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) can be used to select a well fitting vine structure subject to these restrctions.
 
@@
 
-We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x) \} \in \mathbb{R}^4$ and consider four different functional forms for the approximating distribution $q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})$:
+We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x) \} \in \mathbb{R}^4$ and consider three different functional forms for the approximating distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$:
 \begin{enumerate}
-\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \textbf{V})$, where $\textbf{V}$ is a diagonal matrix.
-\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \boldsymbol{\Sigma})$, where $\boldsymbol{\Sigma}$ allows non-zero off-diagonal terms.
-\item $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda})$ is an Inverse Autoregressive flow (Kingma, Salimans, Jozefowicz, Chen, Sutskever and Welling, 2016).
-\item  $q(\boldsymbol{\theta}, \textbf{x} | \boldsymbol{\lambda}) = \prod \left(q(\theta_i) q(x_{it})\right) c(\boldsymbol{\theta}, \textbf{x})$ where the form of the copula $c$ and marginals are informed by an out of date MCMC sample.
+\item $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \textbf{V})$, where $\textbf{V}$ is a diagonal matrix.
+\item $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \boldsymbol{\Sigma})$, where $\boldsymbol{\Sigma}$ allows non-zero off-diagonal terms.
+\item  $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}) = \prod \left(q(\theta_i) q(x_{it})\right) c(\boldsymbol{\theta}, x)$ where the form of the copula $c$ and marginals are informed by an out of date MCMC sample.
 \end{enumerate}
-The first three functional forms are transformation of standard normal noise variables while the fourth is a transformation of uniform noise variables.
+The first two functional forms are transformation of standard normal noise variables while the third is a transformation of uniform noise variables.
 
-As additional data $y_{T+1:T+J-1}$ are observed, we can incorporate this information to improve forecasts of the future value of $y_{T+J}$. We use two MCMC based approaches for updating this forecast,
+We use three MCMC based approaches for forecasting, listed in increasing computational complexity:
 \begin{enumerate}
-\item Use the Kalman filter to update $\textbf{x}_{T+1:T+J}$ as $y_{T+1:T+J}$ is observed using $\boldsymbol{\theta}$ from $p(\boldsymbol{\theta}, \textbf{x}_{1:T} | y_{1:T})$.
-This method is computationally simple, but does not allow the posterior distribution of $\boldsymbol{\theta}$ to be conditioned on new observations.
-\item Run the full MCMC for $\boldsymbol{\theta}$ and $\textbf{x}_{1:T+J}$ after observing each additional $y$.
-This strategy has a high computational cost, but does condition $\boldsymbol{\theta}$ on $y_{T+1:T+J-1}$, and results in the forecast distribution with the least uncertainty.
+\item Run the full MCMC to obtain $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$ and perform an $S+h$ step ahead forecast of $p(y_{T+S+h} | y_{1:T})$, marginalising out the variables $\boldsymbol{\theta}$ and $x_{1:T}$ using draws from $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$.
+\item Using the MCMC from approach 1, use the Kalman filter to update the distribution of $x_{T+1:T+S+h}$ after $y_{T+1:T+S}$ is observed to obtain the $h$ step ahead forecast of $p(y_{T+S+h} | y_{1:T+S})$.
+This method does not condition the posterior distribution of $\boldsymbol{\theta}$ on new observations.
+\item Run the full MCMC to obtain $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$ and perform a $h$ step ahead forecast of $p(y_{T+S+h} | y_{1:T+S})$, marginalising out the variables $\boldsymbol{\theta}$ and $x_{1:T+S}$.
+We assume that the computational time to run this MCMC after observation of $y_{T+S}$ is greater than the time period between $y_{T+S}$ and $y_{T+S+h}$, but this provides a benchmark for comparision with Variational Bayes methods.
 \end{enumerate}
 
-We use three approaches for updating forecasts with VB, listed in increasing computational complexity:
+We use three similar approaches for VB:
 \begin{enumerate}
-\item Use the Kalman filter to update $\textbf{x}_{T+1:T+J}$ as $y_{T+1:T+J}$ is observed using $\boldsymbol{\theta}$ from $q(\boldsymbol{\theta}, \textbf{x}_{1:T} | \boldsymbol{\lambda})$.
-\item Hold elements of $\boldsymbol{\lambda}$ for $x_{1:T+h-1}$ fixed and maximise (\ref{ELBO}) with respect to the elements of $\boldsymbol{\lambda}$ for $\boldsymbol{\theta}$ and $x_{T+h}$ after observation of $y_{T+h}$.
-\item Maximise (\ref{ELBO}) with respect to all elements of $\boldsymbol{\lambda}$
+\item Optimise the distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ after observation of $y_{1:T}$ and perform an $S+h$ step ahead forecast of $q(y_{T+S+h} | y_{1:T})$, marginalising out the variables $\boldsymbol{\theta}$ and $x_{1:T}$, using draws from $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$.
+\item Using the distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ from approach 1, use the Kalman filter to update the distribution of $x_{T+1:T+S+h}$ after $y_{T+1:T+S}$ is observed to obtain the $h$ step ahead forecast of $q(y_{T+S+h} | y_{1:T+S})$.
+\item Begin with $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ from approach 1, and update to  $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ after observation of $y_{T+1:T+S}$.
+We perform this by holding the elements of $\boldsymbol{\lambda}$ that correspond to the distribution of $x_{1:T}$ fixed, and maximise (\ref{ELBO}) with respect to the elements of $\boldsymbol{\lambda}$ for $\boldsymbol{\theta}$ and $x_{T+1:T+S}$.
 \end{enumerate}
 
-Forecast densities of VB and MCMC can be compared by a numeric calculation of the Kullback-Leibler divergence of samples from both $q$ and $p$, and different updating strategies can be compared by the log-score of a held out $y_{T+J}$ value.
+Forecast densities of VB and MCMC can be compared by a numeric calculation of the Kullback-Leibler divergence of samples from both $q$ and $p$, and different updating strategies can be compared by the log-score of a held out $y_{T+S+h}$ value.
 
+With $T = 50, S = 10$ and $h = 1$, data is simulated from the dynamic linear model described by (\ref{DLM:Measure}) and (\ref{DLM:Transition}) with true parameter values of $\mu = 2, \phi = 0.5, \sigma^2_y = 1$ and $\sigma^2_x = 1$, and hyperparameter values of $ \bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1,$ and $\beta_x = 1$.
+The simulated data values are shown in Figure (\ref{fig:DLMy}), and the resulting forecast distributions are shown in Figure (\ref{fig:DLMpreddens}).
+The Kullback Leibler divergence from the VB based predictive density $q$ to the MCMC based predictive density $p$ of $y_{T+S+h}$ is calculated in Table (\ref{Tab:KLdiv}).
+
+\begin{figure}[h]
+\centering
+\includegraphics{yDLM}
+\caption{Data $y_{1}, \dots, y_{T+S+h}$ simulated from a dynamic linear model with $T = 50, S = 10$ and $h = 1$.
+The black component indicates the observations up to time $T$, which are available for use in an MCMC forecast of $y_{T+S+h}$.
+The red component indicates the observations from time $T+1$ to $T+S$, which cannot be included in an MCMC forecast.
+The blue component indicates $y_{T+S+h}$.}
+\label{fig:DLMy}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics{PredDensDLM}
+\caption{The predictive density of $Y_{T+S+h}$ for each forecasting approach using $50,000$ MCMC samples and the first two Variational Bayes functional forms.
+Both Variational Bayes methods seems to have closely captured the MCMC based density in each approach.
+\textit{To add: Copulas}}
+\label{fig:DLMpreddens}
+\end{figure}
+
+\begin{table}
+\centering
+\begin{tabular}{| l | c c c |}
+  \hline
+  & $S+h$ step & $h$ step - Filter & $h$ step - MCMC \\
+  \hline
+  Diagonal & $-9.35 \times 10^{-9}$ & $-1.89 \times 10^{-9}$ & $\boldsymbol{-6.56 \times 10^{-9}} $ \\
+  Non-Diagonal & $\boldsymbol{-6.19 \times 10^{-8}}$ & $\boldsymbol{-3.05 \times 10^{-9}}$ &$-3.69 \times 10^{-9} $ \\
+  \hline
+\end{tabular}
+\caption{The Kullback Leibler divergence between densities in Figure (\ref{fig:DLMpreddens}) from $q$ to $p$.
+Bold indicates the variational approximation with the lowest divergence in each case.
+The added flexibility in allowing non-zero covariances in the Non-Diagonal functional form appears to improve the fit to the MCMC results in two of the forecasting approaches.}
+\label{Tab:KLdiv}
+\end{table}
+
+\newpage
 \section{Appendix}
@@
 \mu &= \mu_4 + L_{41} \epsilon_1 + L_{42} \epsilon_2 + L_{43} \epsilon_3 + L_{44} \epsilon_4 \label{transform4} \\
-x_t &= \mu_{t+5} + \sum_{i=1}^{t+5} L_{t+5, i} \epsilon_i \mbox{ for } t = 0, \dots, T \label{transform5} 
+x_t &= \mu_{t+5} + \sum_{j=1}^{t+5} L_{t+5, j} \epsilon_j \mbox{ for } t = 0, \dots, T \label{transform5} 
 \end{align}
@@
 
-Derivatives of (\ref{logjoint}) with respect to $\boldsymbol{\theta}$ and $\textbf{x}_{0:T}$, as well as derivatives of (\ref{transform1}) - (\ref{logdetjac}) with respect to $\boldsymbol{\mu}$ and $\textbf{L}$ are easily obtained for use in estimating the gradient of the ELBO.
+Derivatives of (\ref{logjoint}) with respect to $\boldsymbol{\theta}$ and $x_{0:T}$, as well as derivatives of (\ref{transform1}) - (\ref{logdetjac}) with respect to $\boldsymbol{\mu}$ and $\textbf{L}$ are easily obtained for use in estimating the gradient of the ELBO.
+
+\subsubsection{Copula Approximation}
+
+Di{\ss}mann's algorithm was applied to MCMC draws of $\boldsymbol{\theta} = \{ \sigma^2_y, \sigma^2_x, \phi, \gamma \}$ and $x_{0:T}$. The first tree was estimated via:
+\begin{enumerate}
+\item Map $\boldsymbol{\theta}, x_{0:T}$ to $(0, 1)^{T+5}$ by taking the empirical marginal CDF of each variable.
+\item Calculate the absoluate value of Kendall's Tau, $|\tau_{ij}|$ for  $i, j = 1, \dots, T+5$ between each pair.
+\item Find the maximum spanning tree by Kruskal's algorithm with edge weights given by $|\tau_{ij}|$, subject to the restrictions that no edge may be selected between two $x_t$ variables and each $x_t$ must have an edge to the same element of $\theta$.
+This tree forms $T_1$.
+\end{enumerate}
+The tree $T_1$ that was selected only permitted one vine factorisation for trees $T_2, \dots, T_k$ so a further application of Di{\ss}mann's algorithm was unneccesary.
+The full vine copula selected is shown in Figure (\ref{fig:dlmvinecop})
+
+\begin{figure}[h]
+\centering
+\tikz{ %
+    \node[latent] (x) {$x_{0:T}$} ; %
+    \node[latent, right = of x] (gamma) {$\gamma$} ; %
+    \node[latent, right = of gamma] (sigy) {$\sigma^2_y$} ; %
+    \node[latent, right = of sigy] (sigx) {$\sigma^2_x$} ; %
+    \node[latent, right = of sigx] (phi) {$\phi$} ; %
+    \edge [-] {x} {gamma} ; %
+    \edge [-] {gamma} {sigy} ; %
+    \edge [-] {sigy} {sigx} ; %
+    \edge [-] {sigx} {phi} ; %
+
+    \node[latent, below = of x] (xgamma) {$x_{0:T}, \gamma$} ; %
+    \node[latent, right = of xgamma] (gammasigy) {$\gamma, \sigma^2_y$} ; %
+    \node[latent, right = of gammasigy] (sigysigx) {$\sigma^2_y, \sigma^2_x$} ; %
+    \node[latent, right = of sigysigx] (sigxphi) {$\sigma^2_x, \phi$} ; %
+    \edge [-] {xgamma} {gammasigy} ; %
+    \edge [-] {gammasigy} {sigysigx} ; %
+    \edge [-] {sigysigx} {sigxphi} ; %
+
+    \node[latent, below = of xgamma] (xsigy) {$x_{0:T}, \sigma^2_y | \gamma$} ; %
+    \node[latent, right = of xsigy] (gammasigx) {$\gamma, \sigma^2_x | \sigma^2_y$} ; %
+    \node[latent, right = of gammasigx] (sigyphi) {$\sigma^2_y, \phi | \sigma^2_x$} ; %
+    \edge [-] {xsigy} {gammasigx} ; %
+    \edge [-] {gammasigx} {sigyphi} ; %
+
+    \node[latent, below = of xsigy] (xsigx) {$x_{0:T}, \sigma^2_x | \sigma^2_y, \gamma$} ; %
+    \node[latent, right = of xsigx] (gammaphi) {$\gamma, \phi | \sigma^2_x, \sigma^2_y$} ; %
+    \edge [-] {xsigx} {gammaphi} ; %
+
+    \node[latent, below = of xsigx] (x1) {$x_1, \phi | \gamma, \sigma^2_x, \sigma^2_y$} ; %
+    \node[latent, left = of x1] (x0) {$x_0, \phi | \gamma, \sigma^2_x, \sigma^2_y$} ; %
+    \node[latent, right = of x1] (dots) {$\dots$} ; %
+    \node[latent, right = of dots] (xT1) {$x_{T-1}, \phi | \gamma, \sigma^2_x, \sigma^2_y$} ; %
+    \node[latent, right = of xT1] (xT) {$x_T, \phi | \gamma, \sigma^2_x, \sigma^2_y$} ; %
+    \edge [-] {x0} {x1} ; %
+    \edge [-] {x1} {dots} ; %
+    \edge [-] {dots} {xT1} ; %
+    \edge [-] {xT1} {xT} ; %
+
+    \node[latent, below = of x1] (x0x1) {$x_0, x_1 | \boldsymbol{\theta}$} ; %
+    \node[latent, right = of x0x1] (x1x2) {$x_1, x_2 | \boldsymbol{\theta}$} ; %
+    \node[latent, right = of x1x2] (dots2) {$\dots$} ; %
+    \node[latent, right = of dots2] (xT1xT) {$x_{T-1}, x_T | \boldsymbol{\theta}$} ; %
+    \edge [-] {x0x1} {x1x2} ; %
+    \edge [-] {x1x2} {dots2} ; %
+    \edge [-] {dots2} {xT1xT} ; %
+}
+\caption{The vine copula used for the Dynamic Linear Model. Each node containing $x_{0:T}$ denotes $T+1$ individual nodes connected to the same neighbouring node.}
+\label{fig:dlmvinecop}
+\end{figure}
+
+\subsubsection{Algorithm Details}
+
+Gradient Ascent is performed with the Adam optimiser from Kingma and Ba (2015).
+Adam uses estimates of the first two moments of the gradient, and is described in Algorithm (\ref{alg:adam}).
+We use $\alpha = 0.1, \beta_1 = 0.9, \beta_2 = 0.999$ and $\epsilon = 10^{-8}$.
+
+\begin{algorithm}[H]
+ \SetKwInOut{Input}{Input}
+ \Input{Method to estimate gradients of $f(\lambda), \alpha, \beta_1, \beta_2, \epsilon$}
+ \KwResult{Optimal Parameters $\lambda$}
+ Set $m_0 = 0$\;
+ Set $v_0 = 0$\;
+ Set $t = 0$\;
+ \While{not converged}{
+  $t = t + 1$
+  $g_t = \nabla_{\lambda}f(\lambda)$ \;
+  $m_t = \beta_1 m_{t-1} + (1-\beta_1) g_t$ \;
+  $v_t = \beta_2 v_{t-1} + (1-\beta_2) v_t$ \;
+  $\hat{m}_t = m_t / (1 - \beta_1^t)$ \;
+  $\hat{v}_t = v_t / (1 - \beta_2^t)$ \;
+  $\lambda = \lambda + \alpha \cdot \hat{m}_t / (\sqrt{\hat{v}_t} + \epsilon)$
+  }
+ \caption{Adam Optimiser}
+  \label{alg:adam}
+\end{algorithm}
+
+Increases in $N$, the number of simulations of the gradient of the ELBO per iteration, should increase the accuracy of the estimate and thus reduce the number of iterations required for convergence at a cost to compuation time per iteration.
+Figure (\ref{fig:RuntimeELBO}) plots the converged value of the ELBO against algorithm run time for a range of values of $N$.
+On this basis, we select $N = 5$ for the diagonal approximation and $N = 100$ for the non-diagonal approximation.
+
+\begin{figure}[h]
+\centering
+\includegraphics{RuntimevELBO}
+\caption{Converged ELBO value versus total runtime for a range of values of $N$ and approximation functional forms.}
+\label{fig:RuntimeELBO}
+\end{figure}
 

### 2017-04-27 DLM/SVM.R

@@ -1,0 +1,23 @@
+library(Rcpp)
+library(RcppArmadillo)
+library(ggplot2)
+sourceCpp("SVM_SGA.cpp")
+
+T = 50
+phi = 0.7
+gamma = 1
+sigmaSq = 1
+
+a0 = rnorm(1, 0, sqrt(sigmaSq / (1-phi^2)))
+alpha = rep(0, T)
+y = rep(0, T)
+for(t in 1:T){
+  if(t == 1){
+    alpha[t] = gamma + phi*a0 + sqrt(sigmaSq)*rnorm(1)
+  } else {
+    alpha[t] = gamma + phi*alpha[t-1] + sqrt(sigmaSq)*rnorm(1)
+  }
+  y[t] = alpha[t] + log(rnorm(1)^2)
+}
+
+fit = SVM_SGA(y, 10, 25, 5, rep(0, T+4), diag(0.1, T+4))

### 2017-05-06 MIVB/writeup.tex
Cleanup

--- 2017-04-27 DLM/writeup.tex
@@
 \section{Introduction}
+\label{Intro}
 
 Consider a high frequency time series with observed values $y_{1}, \dots, y_{T}$, collectively denoted by $y_{1:T}$, with a forecaster interested in the density forecast of the unknown future value $y_{T+S+h}$, where $S \geq 1$.
-Assuming that a forecast of $p(y_{T+S+h} | y_{1:T})$ is available, uncertainty in the forecast can be reduced by observating and conditioning on the realisations of $y_{T+1:T+S}$ to produce $p(y_{T+S+h} | y_{T+S})$.
+Assuming that a forecast of $p(y_{T+S+h} | y_{1:T})$ is available, uncertainty in the forecast can be reduced by observing and conditioning on the realisations of $y_{T+1:T+S}$ to produce $p(y_{T+S+h} | y_{1:T+S})$.
 
@@
 MCMC can have slow convergence rates and generally does not admit parallel processing so that updating the forecast distribution of $y_{T+S+h} | y_{1:T}$ to $y_{T+S+h} | y_{1:T+S}$ after observation of $y_{T+1}, \dots, y_{T+S}$ may not be available before $y_{T+S+h}$ is observed.
-In this paper we assume that an MCMC will converging before time period $T + S + h$ will be required to begin, at the latest, by time $T$ and hence $y_{T}$ is the last observation that can be included in the MCMC.
+In this paper we assume that an MCMC that converges before time period $T + S + h$ will be required to begin, at the latest, by time $T$ and hence $y_{T}$ is the last observation that can be included in the MCMC.
 
@@
 \section{Variational Bayes}
+\label{VB}
 
-Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T}$ and some approximating distribution $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.
+Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$ and some approximating distribution $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.
 
@@
 \label{GradAscent}
-\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
+\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \frac{\delta}{\delta\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
 \end{equation}
-$\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$.
+$\delta/\delta\boldsymbol{\lambda}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$.
 This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
@@
 \label{ScoreDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_i, x_{1:T, i})}{q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)})} \right)
+\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_i, x_{1:T, i})}{q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)})} \right)
 \end{equation}
@@
 Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable.
-Transforming the variables using
+Applying the transformation to $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ results in
 \begin{equation}
 \label{changevar}
-\int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) g(\boldsymbol{\theta}, x_{1:T})  d\boldsymbol{\theta}dx_{1:T} = \int q(\boldsymbol{\epsilon}) g ( f (\epsilon, \boldsymbol{\lambda})) d\boldsymbol{\epsilon})
+q(\boldsymbol{\epsilon} = q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
 \end{equation}
-and noting that
-\begin{equation}
-\label{ReparamDist}
-\ln(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})) = \ln(q(\boldsymbol{\epsilon})) - \ln(|J|),
-\end{equation}
-where $J$ is the Jacobian Matrix of the transformation $f$, allows (\ref{ELBO}) to be rewritten as
+where $J = d\boldsymbol{\epsilon}/d\boldsymbol{\theta}dx_{1:T}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
 \begin{align}
-\label{reparamELBO}
-\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) &=  \int  q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda}))}{q( f (\epsilon, \boldsymbol{\lambda}) | \boldsymbol{\lambda})} \right) d\boldsymbol{\epsilon} \nonumber \\
-&= \int  q(\boldsymbol{\epsilon}) \big( \ln (p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda})) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) \big) d\boldsymbol{\epsilon} \nonumber \\
-&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T}).
+\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) &= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T}}d\boldsymbol{\theta}dx_{1:T} \nonumber \\
+&= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
+&= \int q(\boldsymbol{\epsilon}) \ln (p(y_{1:T},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
+&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T}). \label{reparamELBO}
 \end{align}
@@
 \label{ReparamDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i))) + \nabla_{\boldsymbol{\lambda}}\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|),
+\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)))\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
 \end{equation}
@@
 \section{MCMC Informed Variational Bayes}
+\label{CopulaVB}
 
@@
 Much of the literature focuses on a generic form of $q$ that can flexibly be applied to many models.
-However the time-series nature of our application implies that MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$, which provides useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ for use in the approximate predictive density, $q(y_{T+S+h} | y_{1:T+S})$.
+However the time-series nature of our application implies that MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$, which provide useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ for use in the approximate predictive density, $q(y_{T+S+h} | y_{1:T+S})$.
 There are two stages to this construction:
@@
 \end{enumerate}
-Treating the MCMC samples as 'data', the functional form of $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ can be approached as a traditional distribution fitting problem.
-Optimal marginal distributions are selected for each parameter by AIC, while a vine copula modelling approach is adopted for the dependence structure following Tran, Blei and Airoldi (2015).
+Treating the MCMC samples as `data', the functional form of $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ can be approached as a traditional distribution fitting problem.
+Optimal marginal distributions are selected from a pool of suitable candidate for each parameter by AIC, while a vine copula modelling approach is adopted for the dependence structure following Tran, Blei and Airoldi (2015).
 Sklar (1959) proves that any joint probability distribution with $k$ variables can be written as the product of the marginals and a copula function,
@@
 The Markovian structure of the latent states in the models used provides additional information that can be used in the structure of the vine.
-Denoting the number of elements in $\boldsymbol{\theta}$ with $k$, allow the first $k$ trees to contain edges that correspond to copulas between different elements of $\boldsymbol{\theta}$, and an element of $\boldsymbol{\theta}$ and $x_t$.
-The $k+1'th$ tree edges correspond to the copulas $c(x_0, x_1 | \boldsymbol{\theta}, c(x_1, x_2 | \boldsymbol{\theta}), \dots, c(x_{T-1}, x_T | \boldsymbol{\theta}$ and each further tree contains only bivariate copulas between $x_t$ and $x_{t+j}$ conditioned on $x_{t+l}$, where $j > l \geq 1$.
-Markovian latent states implies that $x_t$ and $x_{t+2}$ are conditionally independent given $x_{t+1}$, so each copula in tree $k+2$ onwards can be set to the independence copula, reducing the number of copula parameters in the vine to be linear in $T$ instead of quadratic.
-To allow the vine structure to be easily extended to new latent variables $x_{T+1:T+S}$, each $x_t$ node should have edges to the same $\boldsymbol{\theta}$ node.
-Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) can be used to select a well fitting vine structure subject to these restrctions.
+Denoting the number of elements in $\boldsymbol{\theta}$ by $k$, the vine is restricted so that the first $k$ trees do not contain contain edges that correspond to copulas between two different elements of $x_{1:T}$, and that the $k+1'th$ tree edges correspond to the copulas $c(x_0, x_1 | \boldsymbol{\theta}), c(x_1, x_2 | \boldsymbol{\theta}), \dots, c(x_{T-1}, x_T | \boldsymbol{\theta})$.
+This enforces a structure where each further tree contains only bivariate copulas between $x_t$ and $x_{t+j}$ conditioned on $x_{t+l}$, where $j > l \geq 1$.
+The Markovian latent states implies that $x_t$ and $x_{t+j}$ are conditionally independent given $x_{t+l}$ for all $t$, so each copula in tree $k+2$ onwards can be set to the independence copula, reducing the number of copula parameters included in $\boldsymbol{\lambda}$ to be linear in $T$ instead of quadratic.
+For convenience in adding new variables $x_{T+1:T+S}$ to the vine, we further restrict the vine so that the position and copula family for any $x_t$ does not depend on $t$.
+Each $x_t$ node has a single edge to the same $\boldsymbol{\theta}$ node in the first $k$ trees with a single copula family chosen for $c(x_t, \theta)$ copulas appearing in the same tree.
+The algorithm used to select the vine is adapted from Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) and appears below in Algorithm \ref{alg:copula}.
+
+\begin{algorithm}[H]
+ \SetKwInOut{Input}{Input}
+ \Input{Observations of $\boldsymbol{\theta}, x_{0:T}$}
+ \KwResult{Distribution functional form for use in VB}
+ \For {each parameter in $\boldsymbol{\theta}, x_{0:T}$}{
+ Select a marginal distribution by AIC\;
+ Transform observations to $(0, 1)$ psuedo-observations by the Inverse CDF\;
+ }
+ Calculate the empirical Kendall's tau $\tau^{\theta}_{i, j}$ for each pair $\theta_i$, $\theta_j$ \;
+ Select the maximum spanning tree of the $\boldsymbol{\theta}$ subgraph with edge weights given by $\tau^{\theta}_{i, j}$. Add these edges to $T_1$.\;
+ For each edge \{i, j\} in $T_1$, select a copula and estimate any parameters by MLE.
+Select the copula that minimises AIC as $\hat{c}_{ij}$ and transform psudeo-observations with $\hat{F}_{i|j}(\theta_i|\theta_j)$ and $\hat{F}_{j|i}(\theta_j|\theta_i)$\;
+ Calculate the empirical Kendall's tau $\tau^{x}_{i, t}$ for each pair $\theta_i$ and $x_t$ \;
+ Find $i^* = \arg \overset{\max}{i} \sum_t \tau^{x}_{i, t}$ and add edges between each $x_t$ and $\theta_{i^*}$ to $T_1$.\;
+ For each copula family estimate any parameters in $c(x_t, \theta_i^*)$ by MLE for $t = 0, \dots, T$.
+Select the copula family that minimises the sum over $t$ of AIC as $\hat{c}_{it}$ and transform psuedo-observations with $\hat{F}_{i|t}(\theta_i|x_t)$ and $\hat{F}_{t|i}(x_j|\theta_i)$\;
+ \For {$l = 2, \dots, k$} {
+  Repeat Kendall's tau calculations, as well as edge and copula family selection for edges corresponding to conditional copulas $c(i, j | D)$ and $c(i, t | D)$ where $D$ is the conditioning set implied by previous trees.\;
+  Transform psuedo-observations with $\hat{F}_{i|j, D}(\theta_i | \theta_j, \theta_D)$, $\hat{F}_{j, i D} (\theta_j | \theta_i, \theta_D)$, $\hat{F}_{i|t, D}(\theta_i | x_t, \theta_D)$ and $\hat{F}_{t|i, D}(x_t | \theta_j, \theta_D)$ for use in the next tree.
+ }
+ Tree $k+1$ has edges $\{t, t+1 | \theta_1, \dots, \theta_k\}$, and for each edge and copula family, estimate copula parameters by MLE, and select the family that minimises the sum over $t$ of AIC.
+Add this copula as $\hat{c}_{t, t+1 | \theta_1, \dots, \theta_k}$.
+\;
+ Form $q(\boldsymbol{\theta}, x_{0:T} | \boldsymbol{\lambda})$ from selected marginals and copulas.
+ \caption{Restricted Di{\ss}mann's Algorithm}
+  \label{alg:copula}
+\end{algorithm}
 
 \section{The Dynamic Linear Model}
+\label{DLM}
 
@@
 We use the priors
-
 \begin{align}
@@
   \hline
-  Diagonal & $-9.35 \times 10^{-9}$ & $-1.89 \times 10^{-9}$ & $\boldsymbol{-6.56 \times 10^{-9}} $ \\
-  Non-Diagonal & $\boldsymbol{-6.19 \times 10^{-8}}$ & $\boldsymbol{-3.05 \times 10^{-9}}$ &$-3.69 \times 10^{-9} $ \\
+  Diagonal & $-1.72 \times 10^{-8}$ & $-3.40 \times 10^{-8}$ & $\boldsymbol{-8.31 \times 10^{-8}} $ \\
+  Non-Diagonal & $\boldsymbol{-2.34 \times 10^{-7}}$ & $\boldsymbol{-4.05 \times 10^{-8}}$ &$-2.42 \times 10^{-7} $ \\
   \hline
@@
 Bold indicates the variational approximation with the lowest divergence in each case.
-The added flexibility in allowing non-zero covariances in the Non-Diagonal functional form appears to improve the fit to the MCMC results in two of the forecasting approaches.}
+The added flexibility in allowing non-zero covariances in the Non-Diagonal functional form appears to improve the fit to the MCMC results in each of the forecasting approaches.}
 \label{Tab:KLdiv}
 \end{table}
+
+\section{The Stochastic Volatility Model}
+\label{SVM}
+
+This section considers the application of Variational Bayes for a Stochastic Volatility Model, a state space model with non-linearities that is commonly used for forecasting stock returns.
+The Stochastic Volatility model is described by
+\begin{align}
+y_t &= \sqrt{h_t} \upsilon_t \label{SVM:Measure} \\
+\ln (h_{t+1}) &= \gamma + \phi \ln (h_t) + \sigma \eta_t \label{SVM:Transition}
+\end{align}
+where $\upsilon_t$ and $\eta_t$ are independent $\mathcal{N}(0, 1)$ random variables.
+The transformations $y_t^* = \ln(y_t^2)$, $\upsilon_t^* = \ln(\upsilon_t^2)$ and $x_t = \ln(h_t)$ are applied to give a linear form of (\ref{SVM:Measure}) and (\ref{SVM:Transition}),
+\begin{align}
+y_t^* &= \alpha_t + \upsilon_t^* \label{SVM:MeasureL} \\
+x_{t+1} &= \gamma + \phi x_t + \sigma \eta_t \label{SVM:TransitionL}.
+\end{align}
+The error in (\ref{SVM:MeasureL}) is non-Gaussian, as $\exp{\upsilon_t^*} \sim \chi^2_1$, so the Kalman filter cannot be applied to quickly update MCMC based forecasts of some future value $y_{T+S+h}$ upon observation of $y_{T+1:T+S}$ as was the case in Section \ref{DLM}.
+The following priors are used,
+\begin{align}
+\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
+\phi &\sim \mathcal{U}(-1, 1) \\
+\sigma^2 &\sim \mathcal{IG}(\mbox{shape = }\alpha, \mbox{scale = }\beta) \\
+x_0 &\sim \mathcal{N}(\gamma(1-\phi)^{-1}, \sigma^2(1 - \phi^2)^{-1}).
+\end{align}
+
+$T = 50$ observations are simulated from (\ref{SVM:MeasureL}) and (\ref{SVM:TransitionL}) using $\sigma^2 = 1, \phi = 0.7$ and $\gamma = 1$, and an unrestricted multivariate normal Varitional Bayes approximation is fit to $\{\ln(\sigma^2), \phi, \gamma, x_{0:T}\}$ using hyperparameter values of $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha = 1$ and $\beta = 1$.
+The resulting predictive density for $Y_{T+1}^*$ is plotted in Figure \ref{fig:SVMpreddens}, and compared to a simulation based true predictive density.
+\begin{figure}[h]
+\centering
+\includegraphics{PredDensSVM}
+\caption{The Variational Bayes predictive density for $Y_{T+1}^*$ compared to the true predictive density.
+The Variational Bayes approach appears to provide an accurate one step ahead forecast for this model.}
+\label{fig:SVMpreddens}
+\end{figure}
 
@@
  \While{not converged}{
-  $t = t + 1$
+  $t = t + 1$\;
   $g_t = \nabla_{\lambda}f(\lambda)$ \;

### 2017-05-06 MIVB/DLM_MCMC.cpp

@@ -1,0 +1,242 @@
+// [[Rcpp::depends(RcppArmadillo)]]
+// [[Rcpp::plugins(cpp11)]]
+
+#include <RcppArmadillo.h>
+#include <math.h>
+using namespace Rcpp;
+using namespace arma;
+using namespace std;
+
+// [[Rcpp::export]]
+rowvec FFcpp(vec y, double phi, double gamma, double sigmaSqY, double sigmaSqX, int T, int J){
+  y -= gamma;
+  vec att(T+J, fill::ones); //0 - (T-1) -> x1 - xT
+  vec ptt(T+J, fill::ones);
+  vec at(T+J);
+  vec pt(T+J);
+  at[0] = 0;
+  pt[0] = pow(phi, 2)*sigmaSqX + sigmaSqX;
+  double vt = y[0];
+  att[0] = pt[0] * vt / (pt[0] + sigmaSqY);
+  ptt[0] = pt[0] - pow(pt[0], 2) / (pt[0] + sigmaSqY);
+  for(int t = 1; t < T+J; ++t){
+    at[t] = phi*att[t-1];
+    pt[t] = pow(phi, 2)*ptt[t-1] + sigmaSqX;
+    vt = y[t] - at[t];
+    att[t] = at[t] + pt[t] * vt / (pt[t] + sigmaSqY);
+    ptt[t] = pt[t] - pow(pt[t], 2) / (pt[t] + sigmaSqY);
+  }   
+  rowvec filtered(J);
+  for(int t = 0; t < J; ++t){
+    filtered[t] = at[t+T] + sqrt(pt[t+T]) * randn<vec>(1)[0];
+  }
+  return filtered;
+}
+
+// [[Rcpp::export]]
+vec FFUpdatercpp(vec y, double phi, double gamma, double sigmaSqY, double sigmaSqX, double xTmean, double xTvar){
+  int J = y.n_elem;
+  y -= gamma;
+  vec at(J);
+  vec pt(J);
+  vec att(J);
+  vec ptt(J);
+  at[0] = phi*xTmean;
+  pt[0] = pow(phi, 2)*xTvar + sigmaSqX;
+  double vt = y[0] - at[0];
+  att[0] = at[0] + pt[0] * vt / (pt[0] + sigmaSqY);
+  ptt[0] = pt[0] - pow(pt[0], 2) / (pt[0] + sigmaSqY);
[…]

### 2017-05-07 MIVB/MIVB.tex
Nuke unneccesary shit

--- 2017-04-27 DLM/writeup.tex
@@
 \numberwithin{equation}{section}
-
+\title{MCMC Informed Variational Bayes}
+\author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
 \begin{document}
 
+\maketitle
+
+\tableofcontents
+
 \section{Introduction}
+\label{Intro}
 
-Consider a high frequency time series with observed values $y_{1}, \dots, y_{T}$, collectively denoted by $y_{1:T}$, with a forecaster interested in the density forecast of the unknown future value $y_{T+S+h}$, where $S \geq 1$.
-Assuming that a forecast of $p(y_{T+S+h} | y_{1:T})$ is available, uncertainty in the forecast can be reduced by observating and conditioning on the realisations of $y_{T+1:T+S}$ to produce $p(y_{T+S+h} | y_{T+S})$.
+Consider a high frequency time series with observed values $y_{1}, \dots, y_{T+S}$, collectively denoted by $y_{1:T+S}$, with a forecaster interested in the forecast distribution of the future value $y_{T+S+h}$, where $S \geq 1$.
 
-Many models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{1:T}$ such that $x_t$ is assumed to follow a Markovian structure.
-In this case $y_{T+S+h}$ depends on the latent variables $x_{T+1:T+S+h}$, and the forecast uncertainty in $y_{T+S+h}$ can be reduced by first incorporating information contained in realisations of $y_{T+1:T+S}$ to form the posterior densities of $x_{T+1:T+S+h} | x_{1:T}, \boldsymbol{\theta}, y_{1:T+S}$.
-With certain restrictions on the model, the additional information can be easily included via filtering techniques such as the Kalman filter, however in many cases these restrictions are unsuitable and a computationally costly re-estimation of posterior densities of parameters and latent states must be undertaken.
+Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$ such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:{t-1}$, given $\boldsymbol{\theta}$ and $x_t$.
+One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution.
+MCMC can have slow convergence rates and generally does not admit parallel processing or data subsampling.
+In this paper it is assumed that an MCMC sample of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ is available at time $T+S$, but convergence of MCMC is slow, so a posterior distribution conditioned on any additional observations of $y_t, t = T+1, \dots, T+S$ will not be available before observation of $y_{T+S+h}$.
 
-One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution.
-MCMC can have slow convergence rates and generally does not admit parallel processing so that updating the forecast distribution of $y_{T+S+h} | y_{1:T}$ to $y_{T+S+h} | y_{1:T+S}$ after observation of $y_{T+1}, \dots, y_{T+S}$ may not be available before $y_{T+S+h}$ is observed.
-In this paper we assume that an MCMC will converging before time period $T + S + h$ will be required to begin, at the latest, by time $T$ and hence $y_{T}$ is the last observation that can be included in the MCMC.
+With $p(\boldsymbol{\theta}, x_{0:T}, y_{1:T})$ the forecast distribution of p(y_{T+S+h} | y_{T})$ is available, however incorporating the information contained in realisations of $y_{T+1:T+S}$ to obtain $p(y_{T+S+h} | y_{T+S})$ is desirable, as it should provide a forecast with reduced uncertainty.
+The addition of $y_{T+1:T+S}$ is used to update the conditional posterior distribution of $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:T})$ to $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:{T+S})$, which can be performed with filtering techniques such as the Kalman filter with certain restrictions on the model.
+If these restrictions are not met, or the filtering technique is too computationally intensive to perform before $T+S+h$, then the forecaster is unable to obtain $p(y_{T+S+h} | y_{T+S})$.
 
-Increasing the speed of estimation will allow more data up to time $T + S$ to be included in the forecast distribution, and this paper explores the use of Variational Bayes to approximate the posterior distribution of several common time-series models with dependent latent states.
-The the increase in computation speed and resulting increase in observations included in a forecast is contrasted with the decreased statistical accuracy associated with using an approximation to the posterior distribution.
-We compare several different functional forms of the approximating distribution that are commonly used in the machine learning literature and explore the use of previous MCMC samples to inform the design of the functional form.
+An alternative to MCMC is Variational Bayes (VB), which approximates the posterior distribution with a tractable alternative denoted by $q$, optimising the parameters of this approximation so that the approximation is close by some metric to the actual posterior.
+A VB optimisation algorithm is often orders of magnitude faster than MCMC with the application of parallel processing and data subsampling available, so the approximate posterior $q(\boldsymbol{\theta}, x_{0:T+S} | y_{1:T+S})$ is available before $T+S+h$, so the approximate forecast distribution $q(y_{T+S+h} | y_{T+S})$ can be obtained.
+
+The use of VB introduces an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error.
+This paper explores that trade-off for a Dynamic Linear Model and a Stochastic Volatility Model, two common latent state time series models, and introduces MCMC Informed Variational Bayes (MIVB), that emphasises the use of information available in MCMC samples of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T+S})$ as an alternative to standard `black box' approaches where a generic functional form is used for all models.
+
+\begin{itemise}
+\item Section Two: Variational Bayes, Optimisation and ADVI
+\item Section Three: MCMC Informed Variational Bayes
+\item Section Four: Simulation results for DLM, SVM
+\item Section Five: Empirical SVM
+\item Section Six: Discussion
+\end{itemise}
 
 \section{Variational Bayes}
+\label{VB}
+Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$ and some approximating distribution $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.
+The notation for $q$ supresses the dependence of the distribution on $y_{1:T+S}$ as $\boldsymbol{\lambda}$ is typically a function of $y_{1:T+S}$.
 
-Variational Bayes posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T}$ and some approximating distribution $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.
-
-We follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, x_{1:T}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$.
+This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})$ to the true posterior $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$.
 The KL divergence is defined by
@@
 \label{KL-def}
-KL[q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})] = \int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T} |y_{1:T})}\right) d\boldsymbol{\theta} dx_{1:T}
+KL[q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T+S} |y_{1:T+S})}\right) d\boldsymbol{\theta} dx_{1:T+S}
 \end{equation}
@@
 \label{KL-ELBO}
-KL[q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})
+KL[q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \ln(p(y_{1:T+S})) - \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})
 \end{equation}
-where $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.
-$\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T})$ is defined by
+where $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T+S}))$.
+$\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S})$ is defined by
 \begin{equation}
 \label{ELBO}
-\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) = \int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T},\boldsymbol{\theta}, x_{1:T})}{q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T},
+\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T+S},\boldsymbol{\theta}, x_{1:T+S})}{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T+S},
 \end{equation}
-and as $\ln(p(y_{1:T}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).
+and as $\ln(p(y_{1:T+S}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).
 
-Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where we iteratively apply the following update step until (\ref{ELBO}) converges within some pre-specified tolerance:
+\subsection{ELBO Optimisation}
+Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where the following update step is iteratively applied until (\ref{ELBO}) converges within some pre-specified tolerance:
 \begin{equation}
 \label{GradAscent}
-\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \nabla_{\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}).
+\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \frac{\delta}{\delta\boldsymbol{\lambda}} \mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S}).
 \end{equation}
-$\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$.
+$\delta/\delta\boldsymbol{\lambda}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S})$ is the vector of partial derivatives of $\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S})$ with respect to each element of $\boldsymbol{\lambda}$ evaluated at $\boldsymbol{\lambda}^{(m)}$.
 This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
@@
 \label{ScoreDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)}), y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \nabla_{\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_i, x_{1:T, i})}{q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)})} \right)
+\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)}), y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} [\ln(q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)}))] \ln \left(\frac{p(y_{1:T+S}, \boldsymbol{\theta}_i, x_{1:T, i})}{q(\boldsymbol{\theta}_i, x_{1:T, i} | \boldsymbol{\lambda}^{(m)})} \right)
 \end{equation}
-where $i = 1, \dots, N$ indicates simulations from $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}^{(m)})$. 
+where $i = 1, \dots, N$ indicates simulations from $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}^{(m)})$. 
 The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $N$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.
@@
 \label{Reparam}
-\boldsymbol{\theta}, x_{1:T} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
+\boldsymbol{\theta}, x_{1:T+S} = f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}).
 \end{equation}
 Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable.
-Transforming the variables using
+Applying the transformation to $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ results in
 \begin{equation}
 \label{changevar}
-\int q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}) g(\boldsymbol{\theta}, x_{1:T})  d\boldsymbol{\theta}dx_{1:T} = \int q(\boldsymbol{\epsilon}) g ( f (\epsilon, \boldsymbol{\lambda})) d\boldsymbol{\epsilon})
+q(\boldsymbol{\epsilon} = q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
 \end{equation}
-and noting that
-\begin{equation}
-\label{ReparamDist}
-\ln(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})) = \ln(q(\boldsymbol{\epsilon})) - \ln(|J|),
-\end{equation}
-where $J$ is the Jacobian Matrix of the transformation $f$, allows (\ref{ELBO}) to be rewritten as
+where $J = d\boldsymbol{\epsilon}/d\boldsymbol{\theta}dx_{1:T+S}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
 \begin{align}
-\label{reparamELBO}
-\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda}), y_{1:T}) &=  \int  q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda}))}{q( f (\epsilon, \boldsymbol{\lambda}) | \boldsymbol{\lambda})} \right) d\boldsymbol{\epsilon} \nonumber \\
-&= \int  q(\boldsymbol{\epsilon}) \big( \ln (p(y_{1:T}, f (\epsilon, \boldsymbol{\lambda})) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) \big) d\boldsymbol{\epsilon} \nonumber \\
-&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T}).
+\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T+S}}d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
+&= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
+&= \int q(\boldsymbol{\epsilon}) \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
+&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T+S}). \label{reparamELBO}
 \end{align}
@@
 \label{ReparamDeriv}
-\nabla_{\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T}) \approx \frac{1}{N}\sum_{i=1}^{N} \nabla_{\boldsymbol{\lambda}} \ln (p(y_{1:T}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i))) + \nabla_{\boldsymbol{\lambda}}\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|),
+\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T+S}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)))\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
 \end{equation}
-where simulations of $\boldsymbol{\theta}$ and $x_{1:t}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
+where simulations of $\boldsymbol{\theta}$ and $x_{1:T+S}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
 The variance of the terms in the sum of this estimator is often orders of magnitude lower than the estimator in (\ref{ScoreDeriv}); often $N = 1$ provides a precise enough estimate of the gradient for fast convergence.
 
+\subsection{$q$ Functional Form}
+A key decision in Variational Bayes is the choice of the functional form of the approximating distribution $q$.
+Much of the literature focuses on a `black box' functional form, where the same $q$ can be applied to many models.
+One popular black box method is Automatic Differentation Variational Bayes (ADVI, Tran?), which first transforms each parameter $\theta$ and latent variable $x_t$ to the real line then selects a multivariate normal $q$ to approximate the posterior distribution of the transformed parameters with either a diagonal or non-diagonal covariance matrix.
+
+For illustration, consider a common latent variable time series model, the Dynamic Linear Model, described by (\ref{DLM:Measure}) and (\ref{DLM:Transition}).
+
+\begin{align}
+y_t &= \gamma + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
+x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
+\end{align}
+where $\epsilon_t$ and $\nu_t$ are independent $\mathcal{N}(0, 1)$ random variables. The priors used are
+\begin{align}
+\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
+\phi &\sim \mathcal{U}(-1, 1) \\
+\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
+\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
+x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
+\end{align}
+where $\mathcal{IG}(\mbox{shape}, \mbox{ scale})$ denotes the inverse of a Gamma (shape, rate) distribution.
+Conjugate priors are chosen for $\gamma, \sigma^2_y$ and $\sigma^2_x$, while the uniform prior for $\phi$ constrains $p(\phi | y_{1:T+S})$ to have zero probability mass outside of the AR(1) stationary region.
+$\sigma^2_y$ and $\sigma^2_x$ are log-transformed such that $\theta^* = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x), x_{0:T}\} \in \mathbb{R}^{T+5}$.
+$100$ data sets $y_{1:T+S}$ are simulated for each $T \in \{50, 100, 250, 500, 1000, 2,000, 5,000, 10,000\}$, using true parameter values of $\gamma = 2, \phi = 0.95, \sigma^2_y = 1$ and $\sigma^2_x=1$, and hyperparameters $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1$ and $\beta_x = 1$.
+For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T = 5,000$ and $10,000$ as this has a complexity of $O(T^2)$.
+MCMC uses a Gibbs sampler with $\min(5000, 10T)$ iterations, discarding the first half.
+The posterior means for $\boldsymbol{\theta}$ are plotted in Figure \ref{fig:DLM:ADVI}, which suggests that ADVI is inconsistent for $\phi, \sigma^2_x$ and $\sigma^2_y$.
+Section \label{MIVB} advocates the opposite approach to ADVI, tailoring the functional form of $q$ to the specific problem.
+
+\begin{figure}[h]
+\centering
+\includegraphics{DLMAdvi}
+\caption{The MCMC and ADVI posterior means for each $\boldsymbol{\theta}$ parameter with $100$ data replications at each $T \in \{50, 100, 250, 500, 1000, 2,000, 5,000, 10,000\}$.
+True values for each parameter is denoted by the red line.
+MCMC consistently estimates of the true value, while both ADVI algorithms are only consistent for the mean parameter $\gamma$.
+}
+\label{fig:DLM:ADVI}
+\end{figure}
+
+A further $100 T = 500$ datasets are simulated with the previous parameter and hyperparameter values, and for each $\theta_i$ the ELBO is maximised for $q(\theta_i, | $\boldsymbol{\lambda})$, holding each $\theta_{j \neq i}$ and $x_{0:T}$ fixed at their true values.
+$\boldsymbol{\lambda}$ includes only the mean and variance of $\theta_i$, so the ELBO is evaluated over a grid so the stochastic nature of the gradient ascent algorithm is not a confounding factor.
+The maximising value of the mean of $\theta_i$ is plotted in Figure \ref{fig:DLM:Numeric}.
+Holding $\sigma^2_x$ and $\sigma^2_y$ fixed at their true values improves the estimation of the mean of $\phi$, but neither $\sigma^2_y$ nor $\sigma^2_x$ improves.
+
+\begin{figure}[h]
+\centering
+\includegraphics{DLMNumeric}
+\caption{The ELBO maximising mean for each $\theta$ parameter, holding each other $\theta$ and $x_{0:T}$ fixed.
+Estimation of the mean of $\phi$ improves relative to Figure \ref{fig:DLM:ADVI} but the posterior mean of $\sigma^2_y$ is consistently over-estimated while the posterior mean of $\sigma^2_x$ is consistently under-estimated.}
+\label{fig:DLM:Numeric}
+\end{figure}
+
 \section{MCMC Informed Variational Bayes}
+\label{MIVB}
 
-A key decision in Variational Bayes is the choice of the functional form of the approximating distribution $q$.
-Much of the literature focuses on a generic form of $q$ that can flexibly be applied to many models.
-However the time-series nature of our application implies that MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$, which provides useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ for use in the approximate predictive density, $q(y_{T+S+h} | y_{1:T+S})$.
+A black box VB approach such as ADVI allows implementation to be simple, as the same multivariate normal $q$ can be applied to many models, only requiring a parameter transformation.
+However a multivariate normal $q$ may not fit many posterior distributions well, and the ELBO may be further maximised by a different $q$.
+Typically, without applying the gradient ascent algorithm to many different functional forms of $q$ there is no way of determinining which distribution is optimal.
+However in the time-series context, MCMC samples of $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$ are available which can be utilised to provide useful information for constructing the functional form of $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$.
 There are two stages to this construction:
@@
 \item Find a well fitting functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ using the MCMC results.
-\item Extrapolate the form of $q$ used for $x_{1:T}$ to $x_{1:T+S}$.
+\item Extrapolate the form to q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$.
 \end{enumerate}
-Treating the MCMC samples as 'data', the functional form of $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ can be approached as a traditional distribution fitting problem.
-Optimal marginal distributions are selected for each parameter by AIC, while a vine copula modelling approach is adopted for the dependence structure following Tran, Blei and Airoldi (2015).
+
+Treating the MCMC samples as `data', choosing a functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ can be approached as a traditional distribution fitting problem.
+Optimal marginal distributions are selected from a pool of suitable candidate distributions for each parameter by AIC, while a vine copula modelling approach is adopted for the dependence structure.
 Sklar (1959) proves that any joint probability distribution with $k$ variables can be written as the product of the marginals and a copula function,
@@
 \end{enumerate}
-Each edge in the vine corresponds to a bivariate copula conditioned on any variables that are shared between nodes, and the product of each of these $k(k-1)/2$ bivariate copulas forms the complete copula.
+Each edge in the vine corresponds to a bivariate copula with a conditioning set containing all variables that are shared between both connected nodes, and the product of each of these $k(k-1)/2$ bivariate copulas forms the complete copula.
 Figure (\ref{fig:vinecop}) provides an example for a four dimensional $\theta$.
@@
 The Markovian structure of the latent states in the models used provides additional information that can be used in the structure of the vine.
-Denoting the number of elements in $\boldsymbol{\theta}$ with $k$, allow the first $k$ trees to contain edges that correspond to copulas between different elements of $\boldsymbol{\theta}$, and an element of $\boldsymbol{\theta}$ and $x_t$.
-The $k+1'th$ tree edges correspond to the copulas $c(x_0, x_1 | \boldsymbol{\theta}, c(x_1, x_2 | \boldsymbol{\theta}), \dots, c(x_{T-1}, x_T | \boldsymbol{\theta}$ and each further tree contains only bivariate copulas between $x_t$ and $x_{t+j}$ conditioned on $x_{t+l}$, where $j > l \geq 1$.
-Markovian latent states implies that $x_t$ and $x_{t+2}$ are conditionally independent given $x_{t+1}$, so each copula in tree $k+2$ onwards can be set to the independence copula, reducing the number of copula parameters in the vine to be linear in $T$ instead of quadratic.
-To allow the vine structure to be easily extended to new latent variables $x_{T+1:T+S}$, each $x_t$ node should have edges to the same $\boldsymbol{\theta}$ node.
-Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) can be used to select a well fitting vine structure subject to these restrctions.
+Denoting the number of elements in $\boldsymbol{\theta}$ by $k$, the vine is restricted so that the first $k$ trees do not contain contain edges that correspond to copulas between two different elements of $x_{1:T}$, and that the $k+1'th$ tree edges correspond to the copulas $c(x_0, x_1 | \boldsymbol{\theta}), c(x_1, x_2 | \boldsymbol{\theta}), \dots, c(x_{T-1}, x_T | \boldsymbol{\theta})$.
+This enforces a structure where each further tree contains only bivariate copulas between $x_t$ and $x_{t+j}$ conditioned on $x_{t+l}$, where $j > l \geq 1$.
+The Markovian latent states implies that $x_t$ and $x_{t+j}$ are conditionally independent given $x_{t+l}$ for all $t$, so each copula in tree $k+2$ onwards can be set to the independence copula, reducing the number of copula parameters included in $\boldsymbol{\lambda}$ to be linear in $T$ instead of quadratic.
+For convenience in adding new variables $x_{T+1:T+S}$ to the vine, a further restriction is applied to the vine: the position in the vine and copula family for every $x_t$ is identical, the nodes each have a single edge to the same $\boldsymbol{\theta}$ node in the first $k$ trees with a single copula family chosen for each $c(x_t, \theta_i)$ copula in the same tree, as well as for for each $c(x_t, x_{t+1})$ copula in tree $k+1$.
+The algorithm used to select the vine and marginal distributions is adapted from Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) and appears below in Algorithm \ref{alg:MIVB}.
 
-\section{The Dynamic Linear Model}
+After the functional form for $q(\boldsymbol{\theta}, x_{0:T+S} | \boldsymbol{\lambda})$ is chosen, the ELBO can be maximised using Copula Variational Inference (Tran, Blei and Airoldi 2015).
+The Markovian structure of the latent states implies that, after learning the distribution of any $x_t$, the distribution of each $x_s, s < t$ is irrelevant to forecast of $y_{T+S+h}$.
+The computational cost of the gradient ascent algorithm can be substantially reduced by not optimising the parameters associated with $x_t, t = 0, \dots, T$.
+Instead only Equation (\ref{GradAscent}) to the $\lambda$ parameters associated with the marginal distribution of each $\theta$, the copulas between two $\theta$ variables, the marginal distribution of each $x_t, t > T$, the copulas between these $x_t$ and $\theta$ and the copulas between $x_t, x_{t+1}$.
+Other $\lambda$ parameters can remain at their initial, MCMC `data' estimated, values.
 
-We begin with the Dynamic Linear Model, which fulfills the neccesary conditions for the Kalman filter to be applicable: (\ref{DLM:Measure}) and (\ref{DLM:Transition}) are linear with Gaussian errors.
+\begin{algorithm}[H]
+ \SetKwInOut{Input}{Input}
+ \Input{Observations of $\boldsymbol{\theta}, x_{T-V:T}$}
+ \KwResult{$q$ functional form for use in MIVB}
+ $k = $number of parameters in $\boldsymbol{\theta}$. \;
+ \For {$i = 1, \dots, k$} {
+   Select $q(\theta_i)$ functional form that minimises AIC as $\hat{q}(\theta_i)$, add parameters to $\boldsymbol{\lambda}$.\;
+ }
+ \For {$t = T-V, \dots, T$} {
+  Select $q(x_t)$ functional form that minimises AIC. \;
+ }
+ Select the mode $q(x_t)$ functional form as $\hat{q}(x_t)$, estimate for all $t = T-V, \dots, T$, add parameters to $\boldsymbol{\lambda}$.\;
+ Transform observations for each variable to $(0, 1)$ by the inverse CDF\;
+ Apply Di{\ss}mann's algorithm to $\boldsymbol{\theta}$, denote the selected edges by $\{i, j | D \}_l$ and copulas by $\hat{c}(i, j | D)_l$ where $D$ is the conditioning set for each copula and $l = 1, \dots, k-1$ is the tree.
+\;
+ \For {$l = 1, \dots, k$} {
+  \If {l = 1} {
+    Calculate the empirical Kendall's tau $\tau_{i, t}$ for each pair $\theta_i$ and $x_t$ \;
+  } \Else {
+    Calculate the empirical Kendell's tau $\tau_{j, t | i^{(1:l-1)}}$ for $x_t | \theta_{i^{(1:l-1)}}$ and each $\theta_j | D$ where $D$ matches the $x_t$ conditioning set $\theta_{i^{(1:l-1)}}$. \;
+  }
+  Find $i^{l} = \arg \underset{j}{\max} \sum_t \tau_{j, t | i^{(1:l-1)}}$. \;
+  Add edges $\{i, j | D\}_l$ and $\{i^{(l)}, t\}$ to Tree $l$. \;
+  For each $t$ find the copula family that minimises the AIC of $c(i^{(l)}, t |i^{(1:l-1)})$.\;
+  Select the mode copula family as $\hat{c}(i^{(l)}, t |i^{(1:l-1)})$.\;
+  For each $\hat{c}(i, j | D)_l$ calculate $\theta_i | \theta_j, \theta_D$ and $\theta_j | \theta_i, \theta_D$; transform to (0, 1).\;
+  For each $\hat{c}(i^{(l)}, t |i^{(1:l-1)})$ calculate $x_t | \theta_{i^{(1:l)}}$; transform to (0, 1).\;
+ }
+  For each $t$ in $T-V+1, \dots, T$, find the copula family that minimises the AIC of $c(x_t, x_{t-1} | \boldsymbol{\theta}).$ \;
+  Select the mode family as $\hat{c}(x_t, x_{t-1} | \boldsymbol{\theta})$. \;
+  Add all estimated $\hat{c}$ copula parameters to $\boldsymbol{\lambda}$. \;
+  Extrapolate edges, copula family and marginal distribution family to all $t = 0, \dots, T+S$ as $\hat{c}$ and $\hat{q}$. \;
+  Estimate copula and marginal parameters for $t = 0, \dots, T-V-1$, add to $\boldsymbol{\lambda}$. \;
+  Set copula and marginal parameters for $t = T+1, \dots, T+S$ to those used for $t = T$, add to $\boldsymbol{\lambda}$.\;
+  Set $q(\boldsymbol{\theta}, x_{0:T+S} | \boldsymbol{\lambda})$ to the product of all $\hat{c}$ copulas and $\hat{q}$ marginals.\;
+  \caption{MIVB $q$ selection algorithm}
+  \label{alg:MIVB}
+\end{algorithm}
 
-\begin{align}
-y_t &= \gamma + x_t + \sigma_y \epsilon_t \label{DLM:Measure}\\
-x_t &= \phi x_{t-1} + \sigma_x \nu_t \label{DLM:Transition}
-\end{align}
-where $\epsilon_t$ and $\nu_t$ are independent $\mathcal{N}(0, 1)$ random variables.
+\section{Simulation Results}
+\subsection{DLM}
+\subsection{SVM}
 
-We use the priors
+\section{Empirical Application}
 
-\begin{align}
-\gamma &\sim \mathcal{N}(\bar{\gamma}, \sigma^2_{\gamma}) \\
-\phi &\sim \mathcal{U}(-1, 1) \\
-\sigma^2_y &\sim \mathcal{IG}(\mbox{shape = }\alpha_y, \mbox{scale = }\beta_y) \\
-\sigma^2_x &\sim \mathcal{IG}(\mbox{shape = }\alpha_x, \mbox{scale = }\beta_x) \\
-x_0 &\sim \mathcal{N}(0, \sigma^2_x(1 - \phi^2)^{-1})
-\end{align}
-where $\mathcal{IG}(\mbox{shape}, \mbox{ scale})$ denotes the inverse of a Gamma (shape, rate) distribution.
-
-We transform $\sigma^2_y$ and $\sigma^2_x$ so that $\boldsymbol{\theta} = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x) \} \in \mathbb{R}^4$ and consider three different functional forms for the approximating distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$:
-\begin{enumerate}
-\item $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \textbf{V})$, where $\textbf{V}$ is a diagonal matrix.
-\item $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}) = \mathcal{N}(\boldsymbol{\psi}, \boldsymbol{\Sigma})$, where $\boldsymbol{\Sigma}$ allows non-zero off-diagonal terms.
-\item  $q(\boldsymbol{\theta}, x | \boldsymbol{\lambda}) = \prod \left(q(\theta_i) q(x_{it})\right) c(\boldsymbol{\theta}, x)$ where the form of the copula $c$ and marginals are informed by an out of date MCMC sample.
-\end{enumerate}
-The first two functional forms are transformation of standard normal noise variables while the third is a transformation of uniform noise variables.
-
-We use three MCMC based approaches for forecasting, listed in increasing computational complexity:
-\begin{enumerate}
-\item Run the full MCMC to obtain $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$ and perform an $S+h$ step ahead forecast of $p(y_{T+S+h} | y_{1:T})$, marginalising out the variables $\boldsymbol{\theta}$ and $x_{1:T}$ using draws from $p(\boldsymbol{\theta}, x_{1:T} | y_{1:T})$.
-\item Using the MCMC from approach 1, use the Kalman filter to update the distribution of $x_{T+1:T+S+h}$ after $y_{T+1:T+S}$ is observed to obtain the $h$ step ahead forecast of $p(y_{T+S+h} | y_{1:T+S})$.
-This method does not condition the posterior distribution of $\boldsymbol{\theta}$ on new observations.
-\item Run the full MCMC to obtain $p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})$ and perform a $h$ step ahead forecast of $p(y_{T+S+h} | y_{1:T+S})$, marginalising out the variables $\boldsymbol{\theta}$ and $x_{1:T+S}$.
-We assume that the computational time to run this MCMC after observation of $y_{T+S}$ is greater than the time period between $y_{T+S}$ and $y_{T+S+h}$, but this provides a benchmark for comparision with Variational Bayes methods.
-\end{enumerate}
-
-We use three similar approaches for VB:
-\begin{enumerate}
-\item Optimise the distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ after observation of $y_{1:T}$ and perform an $S+h$ step ahead forecast of $q(y_{T+S+h} | y_{1:T})$, marginalising out the variables $\boldsymbol{\theta}$ and $x_{1:T}$, using draws from $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$.
-\item Using the distribution $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ from approach 1, use the Kalman filter to update the distribution of $x_{T+1:T+S+h}$ after $y_{T+1:T+S}$ is observed to obtain the $h$ step ahead forecast of $q(y_{T+S+h} | y_{1:T+S})$.
-\item Begin with $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ from approach 1, and update to  $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$ after observation of $y_{T+1:T+S}$.
-We perform this by holding the elements of $\boldsymbol{\lambda}$ that correspond to the distribution of $x_{1:T}$ fixed, and maximise (\ref{ELBO}) with respect to the elements of $\boldsymbol{\lambda}$ for $\boldsymbol{\theta}$ and $x_{T+1:T+S}$.
-\end{enumerate}
-
-Forecast densities of VB and MCMC can be compared by a numeric calculation of the Kullback-Leibler divergence of samples from both $q$ and $p$, and different updating strategies can be compared by the log-score of a held out $y_{T+S+h}$ value.
-
-With $T = 50, S = 10$ and $h = 1$, data is simulated from the dynamic linear model described by (\ref{DLM:Measure}) and (\ref{DLM:Transition}) with true parameter values of $\mu = 2, \phi = 0.5, \sigma^2_y = 1$ and $\sigma^2_x = 1$, and hyperparameter values of $ \bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1,$ and $\beta_x = 1$.
-The simulated data values are shown in Figure (\ref{fig:DLMy}), and the resulting forecast distributions are shown in Figure (\ref{fig:DLMpreddens}).
-The Kullback Leibler divergence from the VB based predictive density $q$ to the MCMC based predictive density $p$ of $y_{T+S+h}$ is calculated in Table (\ref{Tab:KLdiv}).
-
-\begin{figure}[h]
-\centering
-\includegraphics{yDLM}
-\caption{Data $y_{1}, \dots, y_{T+S+h}$ simulated from a dynamic linear model with $T = 50, S = 10$ and $h = 1$.
-The black component indicates the observations up to time $T$, which are available for use in an MCMC forecast of $y_{T+S+h}$.
-The red component indicates the observations from time $T+1$ to $T+S$, which cannot be included in an MCMC forecast.
-The blue component indicates $y_{T+S+h}$.}
-\label{fig:DLMy}
-\end{figure}
-
-\begin{figure}[h]
-\centering
-\includegraphics{PredDensDLM}
-\caption{The predictive density of $Y_{T+S+h}$ for each forecasting approach using $50,000$ MCMC samples and the first two Variational Bayes functional forms.
-Both Variational Bayes methods seems to have closely captured the MCMC based density in each approach.
-\textit{To add: Copulas}}
-\label{fig:DLMpreddens}
-\end{figure}
-
-\begin{table}
-\centering
-\begin{tabular}{| l | c c c |}
-  \hline
-  & $S+h$ step & $h$ step - Filter & $h$ step - MCMC \\
-  \hline
-  Diagonal & $-9.35 \times 10^{-9}$ & $-1.89 \times 10^{-9}$ & $\boldsymbol{-6.56 \times 10^{-9}} $ \\
-  Non-Diagonal & $\boldsymbol{-6.19 \times 10^{-8}}$ & $\boldsymbol{-3.05 \times 10^{-9}}$ &$-3.69 \times 10^{-9} $ \\
-  \hline
-\end{tabular}
-\caption{The Kullback Leibler divergence between densities in Figure (\ref{fig:DLMpreddens}) from $q$ to $p$.
-Bold indicates the variational approximation with the lowest divergence in each case.
-The added flexibility in allowing non-zero covariances in the Non-Diagonal functional form appears to improve the fit to the MCMC results in two of the forecasting approaches.}
-\label{Tab:KLdiv}
-\end{table}
+\section{Discussion}
 

### 2017-05-07 Confirmation/confirmation_example.R

@@ -1,264 +1,0 @@
-library(ggplot2)
-library(mvtnorm)
-library(coda)
-library(GGally)
-library(dplyr)
-library(gridExtra)
-library(tidyr)
-library(pscl)
-library(Rcpp)
-library(truncnorm)
-library(RcppArmadillo)
-library(BH)
-source("conf_functions.R")
-sourceCpp("../AR2.cpp")
-
-set.seed(21)
-phi1 = 0.75
-phi2 = 0.2
-sigma2 = 1
-T = 150
-J = 150 #for forecasting
-y = vector(length = T+J)
-
-gamma0 = sigma2 * (1-phi2) / ((1+phi2)*((1-phi2)^2 - phi1^2))
-gamma1 = sigma2 * phi1 / ((1+phi2)*((1-phi2)^2 - phi1^2))
-Sigma = matrix(c(gamma0, gamma1, gamma1, gamma0), 2)
-
-y = rmvnorm(1, c(0, 0), Sigma)
-for(t in 3:(T+J)){
-  y[t] = phi1*y[t-1] + phi2*y[t-2] + rnorm(1, sd = sqrt(sigma2))
-}
-
-##MCMC
-rep = 100000
-#T = 30
-igShape = T/2
-
-theta = matrix(0, ncol = 3, nrow = rep)
-sig2 = 1
-phi = c(0.6, 0.1)
-
-tune1 = 0.25
-accept1 = 0
-tune2 = 0.25
-accept2 = 0
-set.seed(3)
-
-for(i in 1:rep){
-  #Phi 1 Metropolis Hastings
-  flag = FALSE
-  while(!flag){
-    candidate = rnorm(1, phi[1], tune1)
-    if(phi[2]  < 1 + candidate & phi[2] < 1 - candidate){
-      flag = TRUE
-    }
-  }
-  canpdens = lphidens(y[1:T], candidate, phi[2], sig2)
-  oldpdens = lphidens(y[1:T], phi[1], phi[2], sig2)
-  canqdens = lqdens(candidate, phi[1], tune1, phi[2], 1)
-  oldqdens = lqdens(phi[1], candidate, tune1, phi[2], 1)
-  ratio = min(1, exp(canpdens + oldqdens - canqdens - oldpdens))
-  if(runif(1) < ratio){
-    accept1 = accept1 + 1
-    phi[1] = candidate
-  }
-  
-  #Phi 2 Metropolis Hastings
-  flag = FALSE
[…]

### 2017-05-07 load_forecasts/AR3MCMC.cpp

@@ -1,208 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-
-//Returns sum of element by element products of two vectors, R style
-double vecprodsum(vec a, vec b){
-  int N = a.n_elem;
-  double out = 0;
-  for(int i = 0; i < N; ++i){
-    out += a[i]*b[i];
-  }
-  return out;
-}
-
-double rowprodsum(vec a, rowvec b){
-  int N = a.n_elem;
-  double out = 0;
-  for(int i = 0; i < N; ++i){
-    out += a[i]*b[i];
-  }
-  return out;
-}
-
-
-// [[Rcpp::export]]
-mat MCMC_load(vec y, mat x, int rep, int burn = 0, int thin = 1) {
-  int dimx = x.n_cols;
-  int T = y.n_elem;
-  
-  // Prior parameters
-  
-  //Beta - 0 intercept,  1-47 half hour effects, 48-53 day of week, 54 temp, 55 time trend, base - sunday 12:00
-  vec mu(dimx); mu.fill(0);
-  vec tau(dimx); tau.fill(1000000);
-  //Phi 
-  vec gamma(3); gamma.fill(0);
-  vec lambda(3); lambda.fill(10);
-  //Sigma Squared
-  double alphasig = 1;
-  double betasig = 1;
-  //Doesnt depend on parameters
-  double alphabar = alphasig + T/2 - 168;
-  double betabar;
-
-  // Preliminary Data Calculations
-  
-  //Yt lags * Yt lags
-  vec lags = {0, 1, 48, 336}; //lags used in calculations
-  vec sumyy(9, fill::zeros); //rows: yt yt -1 - yt yt-48 - yt yt-336 - yt-1 yt-1 - ... - yt-336 yt-336
-  for(int i =  336; i < T; ++i){ //iterate over time
-    for(int k = 1; k < 4; ++k){ //y t - all lags
-      sumyy[k-1] += y[i]*y[i-lags[k]];
-      sumyy[2+k] += y[i-1]*y[i-lags[k]];
-    }
-    sumyy[6] += y[i-48]*y[i-48];
-    sumyy[7] += y[i-48]*y[i-336];
[…]

### 2017-05-07 load_forecasts/AR3MCMC.r

@@ -1,98 +1,0 @@
-library(readr)
-library(tidyr)
-library(dplyr)
-library(lubridate)
-library(Rcpp)
-library(microbenchmark)
-library(ggplot2)
-library(coda)
-sourceCpp("AR3MCMC.cpp")
-sourceCpp("AR4MCMC.cpp")
-
-#Data
-
-loaddata = select(read_csv("fulldata.csv"), SETTLEMENTDATE, TOTALDEMAND, Date, Period, Day, Temp, LagTemp)
-loaddata$SETTLEMENTDATE = ymd_hms(loaddata$SETTLEMENTDATE, tz = "AEST")
-
-GFriday = c("2011-04-22", "2012-04-06", "2013-03-29", "2014-04-18")
-EasterS = c("2011-04-24", "2012-04-08", "2013-03-31", "2014-04-20")
-EasterM = c("2011-04-26", "2012-04-09", "2013-04-01", "2014-04-21")
-Ausday = c("2011-01-26", "2012-01-26", "2013-01-28", "2013-01-27")
-Labour = c("2011-03-14", "2012-03-12", "2013-03-11", "2014-03-10")
-Anzac = c("2011-04-25", "2012-04-25", "2013-04-25", "2014-04-25")
-Queens = c("2011-06-13", "2012-06-11", "2013-06-10", "2014-06-09")
-MelCup = c("2011-11-01", "2012-11-06", "2013-11-05", "2014-11-04")
-xmasshutdown = c("2011-12-24", "2011-12-27", "2011-12-28", "2011-12-29", 
-                 "2011-12-30", "2012-01-02", "2012-01-03", "2012-12-24", 
-                 "2012-12-27", "2012-12-28", "2012-12-29", "2012-12-30",
-                 "2013-01-02", "2013-01-03", "2013-12-23", "2013-12-24", 
-                 "2013-12-27", "2013-12-28", "2013-12-29", "2013-12-30",
-                 "2014-01-02", "2014-01-03", "2014-12-22", "2014-12-23", 
-                 "2014-12-24", "2014-12-27", "2014-12-28", "2014-12-29", 
-                 "2014-12-30")
-
-loaddata = mutate(loaddata, Christmas = ifelse(substr(Date, 6, 10) == "12-25", 1, 0), Boxingday = ifelse(substr(Date, 6, 10) == "12-26", 1, 0),
[…]

### 2017-05-07 load_forecasts/AR4MCMC.cpp

@@ -1,247 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-
-//Returns sum of element by element products of two vectors, R style
-double vecprodsum(vec a, vec b){
-  int N = a.n_elem;
-  double out = 0;
-  for(int i = 0; i < N; ++i){
-    out += a[i]*b[i];
-  }
-  return out;
-}
-
-double rowprodsum(vec a, rowvec b){
-  int N = a.n_elem;
-  double out = 0;
-  for(int i = 0; i < N; ++i){
-    out += a[i]*b[i];
-  }
-  return out;
-}
-
-
-// [[Rcpp::export]]
-mat MCMC_load4(vec y, mat x, int rep,  vec lags, int burn = 0, int thin = 1) {
-  int dimx = x.n_cols;
-  int T = y.n_elem;
-  
-  // Prior parameters
-  
-  //Beta - 0 intercept,  1-47 half hour effects, 48-53 day of week, 54 temp, 55 time trend, base - sunday 12:00
-  vec mu(dimx); mu.fill(0);
-  vec tau(dimx); tau.fill(1000000);
-  //Phi 
-  vec gamma(4); gamma.fill(0);
-  vec lambda(4); lambda.fill(10);
-  //Sigma Squared
-  double alphasig = 1;
-  double betasig = 1;
-  //Doesnt depend on parameters
-  double alphabar = alphasig + T/2 - 8928;
-  double betabar;
-  
-  // Preliminary Data Calculations
-  
-  //Yt lags * Yt lags
-  vec sumyy(14, fill::zeros); //rows: yt yt -1 - yt yt-2 - yt yt-48 - ... - yt-336 yt-336
-  for(int i = 17856; i < T; ++i){ //iterate over time
-    for(int k = 1; k < 5; ++k){ //y t - all lags
-      sumyy[k-1] += y[i]*y[i-lags[k]];
-      sumyy[3+k] += y[i-1]*y[i-lags[k]];
-      if(k > 1){
-        sumyy[6+k] +=y[i-2]*y[i-lags[k]];
-      }
-    }
-    sumyy[11] += y[i-336]*y[i-336];
-    sumyy[12] += y[i-336]*y[i-17520];
[…]

### 2017-05-07 load_forecasts/data manipulation.R

@@ -1,50 +1,0 @@
-library(readr)
-library(tidyr)
-library(dplyr)
-library(lubridate)
-library(ggplot2)
-library(gridExtra)
-
-#Data Manipulation
-
-temps = read_csv("data/temp_86071.csv")
-temps$Date = as.Date(temps$Date, origin = "1899-12-30")
-tempsubset = filter(temps, Date > "2010-12-31" & Date < "2015-01-01")
-ltemp = temps[which(temps$Date > "2010-12-31" & temps$Date < "2015-01-01")-1,]
-month = c("01", "02", "03", "04", "05", "06", "07", "08", "09", "10", "11", "12")
-year = 2011:2014
-elecdata = NULL
-for(i in 1:4){
-  for(j in 1:12){
-    elecdata = rbind(elecdata, read_csv(paste0("data/PRICE_AND_DEMAND_", year[i], month[j], "_VIC1.csv")))
-  }
-}
-
-elecdata$Date = as.Date(elecdata$SETTLEMENTDATE)
-period = as.numeric(as.factor(substr(elecdata$SETTLEMENTDATE, 12, 16))) -1
-period[period == 0] = 48
-elecdata$Period = as.factor(period)
-elecdata$Day = wday(elecdata$Date, label = TRUE)
-elecdata$Temp = tempsubset$Temp
-elecdata$LagTemp = ltemp$Temp
-write.csv(elecdata, "fulldata.csv", row.names = FALSE)
-
-#Temperature vs Load
-tempeffect = ungroup(mutate(group_by(elecdata, Period, Day), load = TOTALDEMAND - mean(TOTALDEMAND)))
-ggplot(tempeffect, aes(Temp, load)) + geom_point()
-
-#Daily Load vs Time
-daily = ungroup(summarise(group_by(elecdata, Date), load = mean(TOTALDEMAND), maxtemp = max(Temp)))
-ggplot(daily, aes(Date, load)) + geom_line()
-
-#Half-hourly Load vs Time
-halfhourly = ungroup(summarise(group_by(elecdata, Period), load = mean(TOTALDEMAND)))
-ggplot(halfhourly, aes(as.numeric(Period), load)) + geom_line() + labs(x = "Half Hour Periods")
-
-#Load vs Temp
-p1 = ggplot(daily, aes(Date, load)) + geom_line()
-p2 = ggplot(daily, aes(Date, maxtemp)) + geom_line()
-grid.arrange(p1, p2, ncol = 1)
-
-
-

### 2017-05-07 load_forecasts/exp_sm_SGA.cpp

@@ -1,171 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-#define my_PI 3.1415926535897932384626433832795028841971693993751058209749445923078164062
-
-
-//Evaluate log joint density, just about the only function that can be evaluated in cpp
-// [[Rcpp::export]]
-double log_joint_dens(vec y, vec theta){
-  int T = y.n_elem;
-  double sigmasq = theta[2];
-  vec initial = {theta[3], theta[4], theta[5], theta[6], theta[7], theta[8], theta[9]};
-  int dim = 7;
-  vec lags = {1, 7};
-  vec alpha_full = {theta[0], theta[1], 0, 0, 0, 0, 0};
-  vec x = {1, -1, -1, -1, -1, -1, -1};
-  mat Trans(7, 7, fill::zeros);
-  Trans(0, 0) = 1;
-  Trans.row(1) = {0, -1, -1, -1, -1, -1, -1};
-  for(int i = 2; i < 7; ++i){
-    Trans(i, i-1) = 1;
-  }
-  mat x_tilde(T, dim, fill::zeros); //X_tilde_t vector for t = 1.. T
-  vec y_tilde(T);  //Y_tilde_t scalar for t = 1...T
-  mat b_bar(dim, T ); //b predictions
-  vec alpha_full(dim, fill::zeros); //entire parameter vector including zeroes for b_t = T b_t-1 + alpha_full*et
-  vec csum_lags(size, fill::zeros); //cumulative sum of lags, starting at 0, omitting last term
-  mat D(dim, dim); //D matrix
-  D = Trans - alpha_full * x.t();
-  x_tilde.row(0) = x.t(); //x_tilde_1
-  b_bar.col(0) = alpha_full * y[0]; //b_bar_1
-  y_tilde[0] = y[0]; //y_tilde_1
-  for(int t = 1; t < T; ++t){ //Loop for other values, recursion in paper
-    b_bar.col(t) = D * b_bar.col(t-1) + alpha_full * y[t];
-    x_tilde.row(t) = x_tilde.row(t-1) * D; 
-    y_tilde[t] = (y[t] - x.t() * b_bar.col(t-1))[0]; 
-  }
-  double density = 0;
-  for(int i = 0; i < T; ++i){
[…]

### 2017-05-07 MIVB/MIVB.tex
updates

@@
 \usepackage{amsthm}
-\usepackage{tikz}
-\usetikzlibrary{bayesnet}
+%\usepackage{tikz}
+%\usetikzlibrary{bayesnet}
 \usepackage{url}
@@
 \author{Nathaniel Tomasetti, Catherine Forbes and Anastasios Panagiotelis}
+\date{}
 \begin{document}
@@
 
-Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$ such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:{t-1}$, given $\boldsymbol{\theta}$ and $x_t$.
-One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution.
-MCMC can have slow convergence rates and generally does not admit parallel processing or data subsampling.
-In this paper it is assumed that an MCMC sample of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ is available at time $T+S$, but convergence of MCMC is slow, so a posterior distribution conditioned on any additional observations of $y_t, t = T+1, \dots, T+S$ will not be available before observation of $y_{T+S+h}$.
+Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$, such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:t-1}$, given $\boldsymbol{\theta}$ and $x_t$.
+One of the most common computational techniques for models with latent states is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution, however these approaches can have slow convergence rates as they generally do not admit parallel processing or data subsampling to speed up computation.
+In this paper it is assumed that an MCMC sample of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ is available at time $T+S$, but the convergence of the required MCMC algorithm is slow, so a posterior distribution conditioned on any additional observations of $y_t, t = T+1, \dots, T+S$ will not be available before observation of $y_{T+S+h}$.
 
-With $p(\boldsymbol{\theta}, x_{0:T}, y_{1:T})$ the forecast distribution of p(y_{T+S+h} | y_{T})$ is available, however incorporating the information contained in realisations of $y_{T+1:T+S}$ to obtain $p(y_{T+S+h} | y_{T+S})$ is desirable, as it should provide a forecast with reduced uncertainty.
-The addition of $y_{T+1:T+S}$ is used to update the conditional posterior distribution of $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:T})$ to $p(x_{T+S+h} | \boldsymbol{\theta}, y_{1:{T+S})$, which can be performed with filtering techniques such as the Kalman filter with certain restrictions on the model.
+With $p(\boldsymbol{\theta}, x_{0:T}| y_{1:T})$ available, the forecast distribution of $p(y_{T+S+h} | y_{T})$ is obtainable to make an $S+h$ step ahead forecast of $y_{T+S+h}$.
+However, it is desirable to incorporate the information contained in realisations of $y_{T+1:T+S}$ to obtain $p(y_{T+S+h} | y_{T+S})$, which should provide a forecast with reduced uncertainty due to the smaller forecast horizon of $h$ steps.
+If certain model restrictions are met, $y_{T+1:T+S}$ can be included in the forecast by using filtering techniques such as the Kalman filter or particle filter.
 If these restrictions are not met, or the filtering technique is too computationally intensive to perform before $T+S+h$, then the forecaster is unable to obtain $p(y_{T+S+h} | y_{T+S})$.
@@
 An alternative to MCMC is Variational Bayes (VB), which approximates the posterior distribution with a tractable alternative denoted by $q$, optimising the parameters of this approximation so that the approximation is close by some metric to the actual posterior.
-A VB optimisation algorithm is often orders of magnitude faster than MCMC with the application of parallel processing and data subsampling available, so the approximate posterior $q(\boldsymbol{\theta}, x_{0:T+S} | y_{1:T+S})$ is available before $T+S+h$, so the approximate forecast distribution $q(y_{T+S+h} | y_{T+S})$ can be obtained.
+A VB optimisation algorithm is often orders of magnitude faster than MCMC, with the application of parallel processing and data subsampling available to scale compuation to large datasets.
+With the use of VB the approximate posterior $q(\boldsymbol{\theta}, x_{0:T+S} | y_{1:T+S})$ is available before $T+S+h$, and thus the approximate forecast distribution $q(y_{T+S+h} | y_{T+S})$ can be used.
 
-The use of VB introduces an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error.
-This paper explores that trade-off for a Dynamic Linear Model and a Stochastic Volatility Model, two common latent state time series models, and introduces MCMC Informed Variational Bayes (MIVB), that emphasises the use of information available in MCMC samples of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T+S})$ as an alternative to standard `black box' approaches where a generic functional form is used for all models.
+Variational Bayes involves an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error.
+This paper explores that trade-off for a Dynamic Linear Model and a Stochastic Volatility Model, two common latent state time series models, and introduces MCMC Informed Variational Bayes (MIVB), which emphasises the use of information available in MCMC samples of $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T+S})$ as an alternative to standard `black box' approaches where a generic functional form is used for all models.
 
-\begin{itemise}
+\begin{itemize}
 \item Section Two: Variational Bayes, Optimisation and ADVI
@@
 \item Section Six: Discussion
-\end{itemise}
+\end{itemize}
 
@@
 The KL divergence is defined by
-\begin{equation}
+\begin{align}
 \label{KL-def}
-KL[q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] = \int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T+S} |y_{1:T+S})}\right) d\boldsymbol{\theta} dx_{1:T+S}
-\end{equation}
+KL[q(\boldsymbol{\theta}, x_{1:T+S}| \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{1:T+S} | y_{1:T+S})] &= \nonumber \\
+\int q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) &\ln \left( \frac{q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})}{p(\boldsymbol{\theta}, x_{1:T+S} |y_{1:T+S})}\right) d\boldsymbol{\theta} dx_{1:T+S}
+\end{align}
 and can alternatively be expressed as
@@
 \label{changevar}
-q(\boldsymbol{\epsilon} = q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
+q_{\epsilon}(\boldsymbol{\epsilon}) = q_{\theta, x}(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda})) |J|
 \end{equation}
@@
 \begin{align}
-\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T+S}}d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
-&= \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
-&= \int q(\boldsymbol{\epsilon}) \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
-&= \mathcal{L}(\boldsymbol{\lambda}, y_{1:T+S}). \label{reparamELBO}
+\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\theta, x}(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T+S}}d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
+&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\epsilon}(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
+&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q_{\epsilon}(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
+&= \mathcal{L}(q(\boldsymbol{\epsilon}), \boldsymbol{\lambda}, y_{1:T+S}). \label{reparamELBO}
 \end{align}
@@
 \label{ReparamDeriv}
-\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}, y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T+S}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)))\ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
+\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(q(\boldsymbol{\epsilon}), \boldsymbol{\lambda}^{(m)}, y_{1:T+S}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} \left( \ln (p(y_{1:T+S}, f(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i))) + \ln(|J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|) \right),
 \end{equation}
@@
 $\sigma^2_y$ and $\sigma^2_x$ are log-transformed such that $\theta^* = \{\gamma, \phi, \ln(\sigma^2_y), \ln(\sigma^2_x), x_{0:T}\} \in \mathbb{R}^{T+5}$.
-$100$ data sets $y_{1:T+S}$ are simulated for each $T \in \{50, 100, 250, 500, 1000, 2,000, 5,000, 10,000\}$, using true parameter values of $\gamma = 2, \phi = 0.95, \sigma^2_y = 1$ and $\sigma^2_x=1$, and hyperparameters $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1$ and $\beta_x = 1$.
-For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T = 5,000$ and $10,000$ as this has a complexity of $O(T^2)$.
+$100$ data sets $y_{1:T+S}$ are simulated for each $T \in \{50, 100, 250, 500, 1000, 2000, 5000, 10000\}$, using true parameter values of $\gamma = 2, \phi = 0.95, \sigma^2_y = 1$ and $\sigma^2_x=1$, and hyperparameters $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1$ and $\beta_x = 1$.
+For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T = 5000$ and $10000$ as this has a complexity of $O(T^2)$.
 MCMC uses a Gibbs sampler with $\min(5000, 10T)$ iterations, discarding the first half.
 The posterior means for $\boldsymbol{\theta}$ are plotted in Figure \ref{fig:DLM:ADVI}, which suggests that ADVI is inconsistent for $\phi, \sigma^2_x$ and $\sigma^2_y$.
-Section \label{MIVB} advocates the opposite approach to ADVI, tailoring the functional form of $q$ to the specific problem.
+Section \ref{MIVB} advocates the opposite approach to ADVI, tailoring the functional form of $q$ to the specific problem.
 
@@
 \centering
-\includegraphics{DLMAdvi}
-\caption{The MCMC and ADVI posterior means for each $\boldsymbol{\theta}$ parameter with $100$ data replications at each $T \in \{50, 100, 250, 500, 1000, 2,000, 5,000, 10,000\}$.
+%\includegraphics{DLMAdvi}
+\caption{The MCMC and ADVI posterior means for each $\boldsymbol{\theta}$ parameter with $100$ data replications at each $T \in \{50, 100, 250, 500, 1000, 2000, 5000, 10000\}$.
 True values for each parameter is denoted by the red line.
@@
 
-A further $100 T = 500$ datasets are simulated with the previous parameter and hyperparameter values, and for each $\theta_i$ the ELBO is maximised for $q(\theta_i, | $\boldsymbol{\lambda})$, holding each $\theta_{j \neq i}$ and $x_{0:T}$ fixed at their true values.
+A further $100 T = 500$ datasets are simulated with the previous parameter and hyperparameter values, and for each $\theta_i$ the ELBO is maximised for $q(\theta_i, | \boldsymbol{\lambda})$, holding each $\theta_{j \neq i}$ and $x_{0:T}$ fixed at their true values.
 $\boldsymbol{\lambda}$ includes only the mean and variance of $\theta_i$, so the ELBO is evaluated over a grid so the stochastic nature of the gradient ascent algorithm is not a confounding factor.
@@
 \centering
-\includegraphics{DLMNumeric}
+%\includegraphics{DLMNumeric}
 \caption{The ELBO maximising mean for each $\theta$ parameter, holding each other $\theta$ and $x_{0:T}$ fixed.
@@
 \item Find a well fitting functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ using the MCMC results.
-\item Extrapolate the form to q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$.
+\item Extrapolate the form to $q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})$.
 \end{enumerate}
@@
 \centering
-\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
+%\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
 \vspace{2mm}
@@
 The Markovian structure of the latent states implies that, after learning the distribution of any $x_t$, the distribution of each $x_s, s < t$ is irrelevant to forecast of $y_{T+S+h}$.
-The computational cost of the gradient ascent algorithm can be substantially reduced by not optimising the parameters associated with $x_t, t = 0, \dots, T$.
-Instead only Equation (\ref{GradAscent}) to the $\lambda$ parameters associated with the marginal distribution of each $\theta$, the copulas between two $\theta$ variables, the marginal distribution of each $x_t, t > T$, the copulas between these $x_t$ and $\theta$ and the copulas between $x_t, x_{t+1}$.
-Other $\lambda$ parameters can remain at their initial, MCMC `data' estimated, values.
+The computational cost of the gradient ascent algorithm can be substantially reduced by holding fixed any parameters associated with $x_t, t = 0, \dots, T$.
+Instead apply Equation (\ref{GradAscent}) to only the $\lambda$ parameters associated with the marginal distribution of each $\theta$, copulas between two $\theta$ variables, the marginal distribution of each $x_s, s > T$, copulas between $x_s$ and $\theta$ and copulas between $x_{s-1}, x_s$.
 
@@
  \SetKwInOut{Input}{Input}
- \Input{Observations of $\boldsymbol{\theta}, x_{T-V:T}$}
+ \Input{Observations of $\boldsymbol{\theta}, x_{T-V:T}$,  $k$, the number of elements in $\boldsymbol{\theta}$. \;}
  \KwResult{$q$ functional form for use in MIVB}
- $k = $number of parameters in $\boldsymbol{\theta}$. \;
  \For {$i = 1, \dots, k$} {
@@
   \If {l = 1} {
-    Calculate the empirical Kendall's tau $\tau_{i, t}$ for each pair $\theta_i$ and $x_t$ \;
+    Calculate the empirical Kendall's tau $\tau_{i, t | \emptyset}$ for each pair $\theta_i$ and $x_t$ \;
   } \Else {
@@
   }
-  Find $i^{l} = \arg \underset{j}{\max} \sum_t \tau_{j, t | i^{(1:l-1)}}$. \;
-  Add edges $\{i, j | D\}_l$ and $\{i^{(l)}, t\}$ to Tree $l$. \;
+  Find $i^{(l)} = \arg \underset{j}{\max} \sum_t \tau_{j, t | i^{(1:l-1)}}$. \;
+  Add edges $\{i, j | D\}_l$ and $\{i^{(l)}, t |  i^{(1:l-1)}\}$ to Tree $l$. \;
   For each $t$ find the copula family that minimises the AIC of $c(i^{(l)}, t |i^{(1:l-1)})$.\;
@@
   Add all estimated $\hat{c}$ copula parameters to $\boldsymbol{\lambda}$. \;
-  Extrapolate edges, copula family and marginal distribution family to all $t = 0, \dots, T+S$ as $\hat{c}$ and $\hat{q}$. \;
+  Set edges, copula family and marginal distribution family used for $t = T-V, \dots, T$ as $\hat{c}$ and $\hat{q}$ for all other $t$. \;
   Estimate copula and marginal parameters for $t = 0, \dots, T-V-1$, add to $\boldsymbol{\lambda}$. \;
@@
 \section{Appendix}
+\iffalse
 
@@
 \end{figure}
+\fi
 

### 2017-05-11 MIVB/MIVB.tex
Various

@@
 Much of the literature focuses on a `black box' functional form, where the same $q$ can be applied to many models.
-One popular black box method is Automatic Differentation Variational Bayes (ADVI, Tran?), which first transforms each parameter $\theta$ and latent variable $x_t$ to the real line then selects a multivariate normal $q$ to approximate the posterior distribution of the transformed parameters with either a diagonal or non-diagonal covariance matrix.
+One popular black box method is Automatic Differentation Variational Bayes (ADVI, Kucukeliber, Ranganath, Gelman, Blei 2015; Kucukeliber, Tran, Ranganath, Gelman, Blei 2016), which first transforms each parameter $\theta$ and latent variable $x_t$ to the real line then selects a multivariate normal $q$ to approximate the posterior distribution of the transformed parameters with either a diagonal or non-diagonal covariance matrix.
 
 For illustration, consider a common latent variable time series model, the Dynamic Linear Model, described by (\ref{DLM:Measure}) and (\ref{DLM:Transition}).
-
 \begin{align}
@@
 $100$ data sets $y_{1:T+S}$ are simulated for each $T \in \{50, 100, 250, 500, 1000, 2000, 5000, 10000\}$, using true parameter values of $\gamma = 2, \phi = 0.95, \sigma^2_y = 1$ and $\sigma^2_x=1$, and hyperparameters $\bar{\gamma} = 0, \sigma^2_{\gamma} = 10, \alpha_y = 1, \beta_y = 1, \alpha_x = 1$ and $\beta_x = 1$.
-For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T = 5000$ and $10000$ as this has a complexity of $O(T^2)$.
-MCMC uses a Gibbs sampler with $\min(5000, 10T)$ iterations, discarding the first half.
+For each case MCMC and ADVI are applied with both a diagonal and non-diagonal covariance matrix for $q(\theta^* | \boldsymbol{\lambda})$; however the non-diagonal version is not used for $T \geq 1000$ as the number of covariance parameters is quadratic in $T$.
+MCMC approaches use a Gibbs sampler with $\max(5000, 10T)$ iterations, discarding the first half.
 The posterior means for $\boldsymbol{\theta}$ are plotted in Figure \ref{fig:DLM:ADVI}, which suggests that ADVI is inconsistent for $\phi, \sigma^2_x$ and $\sigma^2_y$.

### 2017-05-11 MIVB/DLM_MIVB.R

@@ -3,27 +3,10 @@ library(RcppArmadillo)
 library(VineCopula)
 library(MASS)
 library(pscl)
-library(distr)
 sourceCpp("DLM_MCMC.cpp")
 source('qSelection.R')
-#source('DLM_CopVB.R')
+source('DLM_CopVB.R')
 
-# CHANGE NORMAL MIXTURE TO T(loc, scale) IN FITREALMARGINALS AND MIVB
-
-SimDLM = function(T, S, h, sigmaSqY=1, sigmaSqX=2, phi=0.95, gamma=2){
-  x0 = rnorm(1, 0, sqrt(sigmaSqX/(1-phi^2)))
-  x = rep(0, T+S+h)
-  y = rep(0, T+S+h)
-  for(t in 1:(T+S+h)){
-    if(t == 1){
-      x[t] = phi*x0 + rnorm(1, 0, sqrt(sigmaSqX)) 
-    } else {
-      x[t] = phi*x[t-1] + rnorm(1, 0, sqrt(sigmaSqX))
-    }
-    y[t] = mu + x[t] + rnorm(1, 0, sqrt(sigmaSqY))
-  }
-  return(y)
-}
 
 MIVB = function(T, S, h, sigmaSqY=1, sigmaSqX=2, phi=0.95, gamma=2, MCMCreps=T*10){
 
@@ -33,8 +16,30 @@ MIVB = function(T, S, h, sigmaSqY=1, sigmaSqX=2, phi=0.95, gamma=2, MCMCreps=T*1
   # Run MCMC
   MCMC = DLM_MCMC(y[1:T], MCMCreps)
   
-  # Filter MCMC forward and calculate logscores
-  ydens = matrix(0, h, 1000)
+  # S + h step ahead forecast
+  ysupport = seq(min(y)-sd(y), max(y)+sd(y), length.out=1000)
+  obs = min(which((y[T+S+h] < ysupport) == TRUE))
+  ydens = rep(0, 1000)
+  for(i in 1:500){
+    u = sample((MCMCreps/2+1):MCMCreps, 1)
+    gammaDraw = MCMC$theta[u, 4]
+    phiDraw = MCMC$theta[u, 3]
+    sigxDraw = MCMC$theta[u, 2]
+    sigyDraw = MCMC$theta[u, 1]
+    xTmean = MCMC$x[u, T+2]
+    xTvar = MCMC$x[u, T+3]
+    phiprod = rep(1, S+h)
[…]

### 2017-05-11 MIVB/SVM_SGA.cpp

@@ -3,169 +3,16 @@
 
 #include <RcppArmadillo.h>
 #include <math.h>
-//#include <distributions.h> //figure this out
+#include "Distributions.h"
+
 using namespace Rcpp;
 using namespace arma;
 using namespace std;
 
 // TODO:
-// Find out how to add distribution code as a header to reduce amount of code in one file
 // Make phi have a truncated normal marginal so I can fix the x0 prior variance 
 // Can repeatedly sample from the MVN until we get a draw in (-1, 1) but I'd like to avoid while loops when the mean of phi can go way outside (-1, 1). Could restrict the mean?
 // Find out how to avoid dynamic casting to make it easier to implement new models, but if I'm sticking with DLM's for a while this is not a problem yet
-
-class Distribution{
-  // This is the base class for the distributions, and exists so I can have Normal and Inverse Gamma pointers in the same array
-public:
-  // This function is only here so I can call any of the subclass versions.
-  // It gets overwritten by the subclass version anyway.
-  virtual double logdens(double x) {return -10000;};
-};
-
-// All distribution subclass objects have parameters, a constructor, a set_value and increase_value function, a logdensity function
-// and in the normal distribution case a transform from epsilon function
-
-class Normal: public Distribution{
-  // This just provides methods for the single variate normal distribution
-public:
-  double mean, variance; // Parameters
-  Normal(double, double); // Constructor
-  void set_values(double m, double v){
-    mean = m;
-    variance = v;
-  }
-  void increase_values(double m, double sd){
-    mean += m;
-    if(sqrt(variance) > -sd){
-      variance = pow(sqrt(variance) + sd, 2);
-    }
-  }
-  double logdens (double x) {
-    return -0.5 * log(2*3.141593*variance) - pow(x - mean, 2) / (2 * variance);}
-  double transform_epsilon(double x){
-    return mean + sqrt(variance) * x;
-  }
-};
-
-Normal::Normal (double m, double v){ // Constructor
-  mean = m;
-  variance = v;
-}
-
-class MultiNormal: public Distribution{
[…]

### 2017-05-15 MIVB/MIVB.tex
updates

@@
 \end{equation}
-where $J = d\boldsymbol{\epsilon}/d\boldsymbol{\theta}dx_{1:T+S}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
+where $J = d\boldsymbol{\theta}dx_{1:T+S}/d\boldsymbol{\epsilon}$ is the Jacobian Matrix of the transformation $f$. Substituting (\ref{changevar}) into (\ref{ELBO}) gives
 \begin{align}
-\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\theta, x}(f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))} \right) \frac{ d\boldsymbol{\epsilon}}{d\boldsymbol{\theta}dx_{1:T+S}}d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
-&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\epsilon}(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon}\nonumber \\
-&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q_{\epsilon}(\boldsymbol{\epsilon})) + \ln(|J|) d\boldsymbol{\epsilon}\nonumber \\
+\mathcal{L}(q(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}), y_{1:T+S}) &= \int q_{\theta, x}(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda}) \ln \left( \frac{p(y_{1:T+S},\boldsymbol{\theta}, x_{1:T+S})}{q_{\theta, x}(\boldsymbol{\theta}, x_{1:T+S} | \boldsymbol{\lambda})} \right) d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
+&= \int q_{\epsilon}(\boldsymbol{\epsilon})|J|^{-1} \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\epsilon}(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\theta}dx_{1:T+S} \nonumber \\
+&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))}{q_{\epsilon}(\boldsymbol{\epsilon})|J|^{-1}} \right) d\boldsymbol{\epsilon} \nonumber \\
+&= \int q_{\epsilon}(\boldsymbol{\epsilon}) \left( \ln (p(y_{1:T+S},f (\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) - \ln(q_{\epsilon}(\boldsymbol{\epsilon})) + \ln(|J|)\right) d\boldsymbol{\epsilon}\nonumber \\
 &= \mathcal{L}(q(\boldsymbol{\epsilon}), \boldsymbol{\lambda}, y_{1:T+S}). \label{reparamELBO}
@@
 Treating the MCMC samples as `data', choosing a functional form for $q(\boldsymbol{\theta}, x_{1:T} | \boldsymbol{\lambda})$ can be approached as a traditional distribution fitting problem.
-Optimal marginal distributions are selected from a pool of suitable candidate distributions for each parameter by AIC, while a vine copula modelling approach is adopted for the dependence structure.
-Sklar (1959) proves that any joint probability distribution with $k$ variables can be written as the product of the marginals and a copula function,
+Select optimal marginal distributions $\tilde{q}$ from a pool of suitable candidate distributions for each parameter by AIC, and for convenience select the same marginal distribution for each $x_t$ latent variable.
+A vine copula modelling approach is adopted for the dependence structure, as this can be selected independently to the marginal distribution.
+Sklar (1959) proves the any joint probability distribution with $k$ variables can be written as the product of the marginals and a copula function,
 \begin{equation}
@@
 where $p(\theta)$ is a pdf, $P(\theta)$ is a cdf, and $c(\cdot)$ is a copula.
-A vine copula is a structure to factorise this copula, increasing the flexibility given to the modeller, and is defined by a set of trees $\mathcal{V} = \{T_1, \dots, T_{k-1} \}$, where
+A vine copula is a structure to factorise this copula, increasing the flexibility of modelling dependence, and is defined by a set of trees $\mathcal{V} = \{T_1, \dots, T_{k-1} \}$, where
 \begin{enumerate}
-\item $T_1$ has $k$ nodes corresponding to the $k$ variables, with $k-1$ edges.
-\item $T_t, t = 2, \dots, p-1$ has nodes corresponding to the edges in tree $T_{t-1}$
+\item $T_1$ has $k$ nodes corresponding to and labelled by the $k$ variables, with $k-1$ edges.
+\item $T_t, t = 2, \dots, k-1$ have $k+1-t$ nodes corresponding to the edges in tree $T_{t-1}$
 \item Two nodes in tree $T_t$ are joined by an edge only if the associated edges in $T_{t-1}$ share a node. 
+\item Each node in $T_t,  t = 2, \dots, k-1$ is labelled by the union of the nodes associated edge in $T_{t-1}$ connects to.
 \end{enumerate}
-Each edge in the vine corresponds to a bivariate copula with a conditioning set containing all variables that are shared between both connected nodes, and the product of each of these $k(k-1)/2$ bivariate copulas forms the complete copula.
+Each edge corresponds to a bivariate copula between the two variables that are unique to each of the connected nodes, with a conditioning set containing all shared variables.
+The product of each of these $k(k-1)/2$ bivariate copulae forms the complete copula.
 Figure (\ref{fig:vinecop}) provides an example for a four dimensional $\theta$.
@@
 \caption{An example of a Vine Copula over four variables.
-The edges in the top tree represent unconditional bivariate copulas between the parameters on the connected nodes.
-The second tree edges represent the bivariate copulas for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node.
+The edges in the top tree represent unconditional bivariate copulae between the parameters on the connected nodes.
+The second tree edges represent the bivariate copulae for $\theta_1 \theta_3 | \theta_2$ and for $\theta_3 \theta_4 | \theta_2$, the conditioning on $\theta_2$ caused by its appearance in each connected node.
 The edge in the final tree represents the bivariate copula for $\theta_1, \theta_4 | \theta_2, \theta_3$, as both $\theta_2$ and $\theta_3$ appear in the connected nodes.}
@@
 
-The Markovian structure of the latent states in the models used provides additional information that can be used in the structure of the vine.
-Denoting the number of elements in $\boldsymbol{\theta}$ by $k$, the vine is restricted so that the first $k$ trees do not contain contain edges that correspond to copulas between two different elements of $x_{1:T}$, and that the $k+1'th$ tree edges correspond to the copulas $c(x_0, x_1 | \boldsymbol{\theta}), c(x_1, x_2 | \boldsymbol{\theta}), \dots, c(x_{T-1}, x_T | \boldsymbol{\theta})$.
-This enforces a structure where each further tree contains only bivariate copulas between $x_t$ and $x_{t+j}$ conditioned on $x_{t+l}$, where $j > l \geq 1$.
-The Markovian latent states implies that $x_t$ and $x_{t+j}$ are conditionally independent given $x_{t+l}$ for all $t$, so each copula in tree $k+2$ onwards can be set to the independence copula, reducing the number of copula parameters included in $\boldsymbol{\lambda}$ to be linear in $T$ instead of quadratic.
-For convenience in adding new variables $x_{T+1:T+S}$ to the vine, a further restriction is applied to the vine: the position in the vine and copula family for every $x_t$ is identical, the nodes each have a single edge to the same $\boldsymbol{\theta}$ node in the first $k$ trees with a single copula family chosen for each $c(x_t, \theta_i)$ copula in the same tree, as well as for for each $c(x_t, x_{t+1})$ copula in tree $k+1$.
-The algorithm used to select the vine and marginal distributions is adapted from Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) and appears below in Algorithm \ref{alg:MIVB}.
+Denoting the number of elements in $\boldsymbol{\theta}$ by $k$, a vine copula for $q(\boldsymbol{\theta}, x_{0:T} | \boldsymbol{\lambda})$ contains $T+k$ trees, with a number of parameters included in $\boldsymbol{\lambda}$ that is $O(T+k)^2$.
+Constructing the vine so that the first $k$ trees do not contain copulae between any two elements of $x_t$ and the $k+1$'th tree contains copulae $c(x_t, x_{t+1} | \boldsymbol{\theta})$ will force trees $k+r$, with $r \geq 2$, to contain copulae $c(x_t, x_{t+r} | \boldsymbol{\theta}, x_{t+1:t+r-1}).
+These can be set to the independence copula as the Markovian nature of the latent varables implies that $x_t$ and $x_{t+r}$ are conditionally independent given $\boldsymbol{\theta}$ and $x_{t+1:t+r-1}, reducing the number of copula parameters to $O(Tk)$.
+For convenience the vine is further restricted so that, in each of the first $k$ trees, every $x_t$ has an edge to the same $\theta$ node with the same copula family.
 
-After the functional form for $q(\boldsymbol{\theta}, x_{0:T+S} | \boldsymbol{\lambda})$ is chosen, the ELBO can be maximised using Copula Variational Inference (Tran, Blei and Airoldi 2015).
-The Markovian structure of the latent states implies that, after learning the distribution of any $x_t$, the distribution of each $x_s, s < t$ is irrelevant to forecast of $y_{T+S+h}$.
-The computational cost of the gradient ascent algorithm can be substantially reduced by holding fixed any parameters associated with $x_t, t = 0, \dots, T$.
-Instead apply Equation (\ref{GradAscent}) to only the $\lambda$ parameters associated with the marginal distribution of each $\theta$, copulas between two $\theta$ variables, the marginal distribution of each $x_s, s > T$, copulas between $x_s$ and $\theta$ and copulas between $x_{s-1}, x_s$.
+Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) can be used to fit a vine copula to data, selecting both the set of edges for each tree and an appropriate copula family for each edge according to some criteria.
+A common critereon is to select the set of edges that maximise the sum of absolute empirical pairwise Kendell's Tau, with copula families that minimise AIC.
+This algorithm is modified to include the above restrictions.
+In the following, use of the letters $i$ and $j$ denote $\theta_i$ and $\theta_j$ and an alternative use of $t$ denotes $x_t$.
+Edges are denoted by $\{i, j | D \}$ and copula are given by $c(i, j | D)$ where $D$ is the conditioning set.
+Any copula that is to be included in the $q$ distribution is given by $\tilde{c}(i, j | D)$.
+Finally the empirical Kendell's Tau between a pair of variables is denoted by $\tau_{i, j | D}$.
+Noting that copulae are a function of the CDF of a random variable first transform all MCMC samples of $\boldsymbol{\theta}$ and $x_{0:T}$ to $(0, 1)$ using the empirical marginal CDFs.
+No restrictions are applied to the $\boldsymbol{\theta}$ sub-trees, so a $k$ dimensional vine for these parameters can be estimated using Di{\ss}mann's algorith by any critiera with selected edges $\{i, j | D\}$ and copulae $\tilde{c}(i, j | D)$ added to the first $k$ trees of the full vine.
+
+The first tree also contains edges from each $x_t$ to the same $\theta$ element, selected by calculating $\tau_{j, t}$ for each $\theta_j$ and $x_t$ and add edges $\{t, i^{(1)} \}$ where
+\begin{equation}
+\label{Vine:TauCrit}
+i^{(1)} = \arg \underset{j}{\max} \sum_{t} \tau_{j, t}.
+\end{equation}
+For each copula family $m$ and each $t$, estimate the copula $c_m(i^{(1)}, t)$ and choose an optimal family to include in the vine as $\tilde{c}(i^{(1)}, t)$.
+The optimal family could be chosen by criteria such as minimises the sum of AIC across $t$, the minimises AIC for the most $t$.
+In each tree $l = 2, \dots, k$ this processs is repeated: Calculate $\tau_{j, t | i^{(1)}, \dots, i^{(l-1)}$ for each $t$ and element of $\theta$ that has an edge to $\theta^{(l-1)}$ in one of the first $l-1$ trees, then choose $\theta_{i^{(l)}$ where
+\begin{equation}
+\label{Vine:TauCrit}
+i^{(l)} = \arg \underset{j}{\max} \sum_{t} \tau_{j, t}.
+\end{equation}
+Estimate copulae $c_m(i^{(l)}, t | i^{(1:l-1)})$ for each $t$ and $m$ and select an optimal family as $\tilde{c}(i^{(l)}, t |i^{(1)}, \dots, i^{(l-1)})$.
+Tree $k+1$ contains edges $\{t, t+1 |  i^{(1)}, \dots, i^{(k) \}$, with an optimal copula family $\tilde{c}(t, t+1 |i^{(1)}, \dots, i^{(l-k)})$.
+Tree $k+r$ with $r \geq 2$ must then contain edges $\{t, t+r |  i^{(1), \dots, i^{(k), t+1, \dots, t+r-1 \} with $\tilde{c}(t, t+r |  i^{(1), \dots, i^{(k), t+1, \dots, t+r-1) chosen to be the independence copula.
+This approach is given in Algorithm \ref{alg:MIVB}.
+
+The product of all $\tilde{c}$ copula and selected $\tilde{q}$ marginals forms the approximating distribution $q(\boldsymbol{\theta}, x_{0:T} | \boldsymbol{\lambda})$.
+As the marginal distributions and copula for $x_t$ are identical for each $t$ additional latent variables $x_{T+1:T+S}$ can easily be included to form $q(\boldsymbol{\theta}, x_{0:T+S} | \boldsymbol{\lambda})$.
+The ELBO can be maximised for this $q$ using Copula Variational Inference (Tran, Blei and Airoldi 2015).
+As we are interested in forecasts of $y_{T+S+h}$ and $x_t$ is Markovian, the distribution of each $x_s, s < t$ is irrelevant given the distribution of $x_t$.
+Holding $\lambda}$ parameters assocaited with $x_t, t = 0, \dots, T$ fixed at MCMC data estimated values will drastically reduce the computational cost of the gradient ascent algorirhtm as only the $\lambda$ parameters associated with the marginal distributions of $\theta$ and $x_{T+1:T+S}$, as well as the copulae between these, needs to be optimised.
 
@@
  \SetKwInOut{Input}{Input}
- \Input{Observations of $\boldsymbol{\theta}, x_{T-V:T}$,  $k$, the number of elements in $\boldsymbol{\theta}$. \;}
- \KwResult{$q$ functional form for use in MIVB}
- \For {$i = 1, \dots, k$} {
-   Select $q(\theta_i)$ functional form that minimises AIC as $\hat{q}(\theta_i)$, add parameters to $\boldsymbol{\lambda}$.\;
+ \Input{Observations of $\boldsymbol{\theta}, x_{1:T}$,  $k$, the number of elements in $\boldsymbol{\theta}$. \;}
+ \KwResult{Set of copulae $\tilde{c}(\cdot, \codt)$}
+ Transform observations for each variable to $(0, 1)$ by the inverse CDF\;
+ Apply Di{\ss}mann's algorithm to $\boldsymbol{\theta}$, denote the selected edges by $\{i, j | D \}_l$ and copulae by $\tilde{c}(i, j | D)_l$ where $D$ is the conditioning set for each copula and $l = 1, \dots, k-1$ is the tree.
+\;
+ Calculate $\hat{\tau}_{j, t}$ for each $j$ and $t$ \;
+ Find $i^{(1)} = \arg \underset{j}{\max} \sum_t \hat{\tau}_{j, t}$. \;
+ Add edges $\{i, j |\}_1$ and $\{i^{(1)}, t \}$ to Tree $1$. \;
+ For each $t$ and copula family $m$ estimate $c_m(i^{(1)}, t)$.\;
+ Select $\tilde{c}(i^{(1)}, t)$.\;
+ For each $\tilde{c}(i, j)_1$ calculate $\theta_i | \theta_j$ and $\theta_j | \theta_i$; transform to (0, 1).\;
+ For each $t$ use $\tilde{c}(i^{(1)}, t)$ to calculate $x_t | \theta_{i^{(1)}}$; transform to (0, 1).\;
+ \For {$l = 2, \dots, k$} {
+    Calculate $\hat{\tau}_{j, t | i^{(1:l-1)}}$ for $x_t | \theta_{i^{(1:l-1)}}$ and each $\theta_j | D$ where $D$ matches the $x_t$ conditioning set $\theta_{i^{(1:l-1)}}$. \;
+  Find $i^{(l)} = \arg \underset{j}{\max} \sum_t \hat{\tau}_{j, t | i^{(1:l-1)}}$. \;
+  Add edges $\{i, j | D\}_l$ and $\{i^{(l)}, t |  i^{(1:l-1)}\}$ to Tree $l$. \;
+  For each $t$ and copula family $m$ estimate $c_m(i^{(l)}, t |i^{(1:l-1)})$.\;
+  Select $\tilde{c}(i^{(l)}, t |i^{(1:l-1)})$.\;
+  For each $\tilde{c}(i, j | D)_l$ calculate $\theta_i | \theta_j, \theta_D$ and $\theta_j | \theta_i, \theta_D$; transform to (0, 1).\;
+  For each $\tilde{c}(i^{(l)}, t |i^{(1:l-1)})$ calculate $x_t | \theta_{i^{(1:l)}}$; transform to (0, 1).\;
  }
- \For {$t = T-V, \dots, T$} {
-  Select $q(x_t)$ functional form that minimises AIC. \;
- }
- Select the mode $q(x_t)$ functional form as $\hat{q}(x_t)$, estimate for all $t = T-V, \dots, T$, add parameters to $\boldsymbol{\lambda}$.\;
- Transform observations for each variable to $(0, 1)$ by the inverse CDF\;
- Apply Di{\ss}mann's algorithm to $\boldsymbol{\theta}$, denote the selected edges by $\{i, j | D \}_l$ and copulas by $\hat{c}(i, j | D)_l$ where $D$ is the conditioning set for each copula and $l = 1, \dots, k-1$ is the tree.
-\;
- \For {$l = 1, \dots, k$} {
-  \If {l = 1} {
-    Calculate the empirical Kendall's tau $\tau_{i, t | \emptyset}$ for each pair $\theta_i$ and $x_t$ \;
-  } \Else {
-    Calculate the empirical Kendell's tau $\tau_{j, t | i^{(1:l-1)}}$ for $x_t | \theta_{i^{(1:l-1)}}$ and each $\theta_j | D$ where $D$ matches the $x_t$ conditioning set $\theta_{i^{(1:l-1)}}$. \;
-  }
-  Find $i^{(l)} = \arg \underset{j}{\max} \sum_t \tau_{j, t | i^{(1:l-1)}}$. \;
-  Add edges $\{i, j | D\}_l$ and $\{i^{(l)}, t |  i^{(1:l-1)}\}$ to Tree $l$. \;
-  For each $t$ find the copula family that minimises the AIC of $c(i^{(l)}, t |i^{(1:l-1)})$.\;
-  Select the mode copula family as $\hat{c}(i^{(l)}, t |i^{(1:l-1)})$.\;
-  For each $\hat{c}(i, j | D)_l$ calculate $\theta_i | \theta_j, \theta_D$ and $\theta_j | \theta_i, \theta_D$; transform to (0, 1).\;
-  For each $\hat{c}(i^{(l)}, t |i^{(1:l-1)})$ calculate $x_t | \theta_{i^{(1:l)}}$; transform to (0, 1).\;
- }
-  For each $t$ in $T-V+1, \dots, T$, find the copula family that minimises the AIC of $c(x_t, x_{t-1} | \boldsymbol{\theta}).$ \;
-  Select the mode family as $\hat{c}(x_t, x_{t-1} | \boldsymbol{\theta})$. \;
-  Add all estimated $\hat{c}$ copula parameters to $\boldsymbol{\lambda}$. \;
-  Set edges, copula family and marginal distribution family used for $t = T-V, \dots, T$ as $\hat{c}$ and $\hat{q}$ for all other $t$. \;
-  Estimate copula and marginal parameters for $t = 0, \dots, T-V-1$, add to $\boldsymbol{\lambda}$. \;
-  Set copula and marginal parameters for $t = T+1, \dots, T+S$ to those used for $t = T$, add to $\boldsymbol{\lambda}$.\;
-  Set $q(\boldsymbol{\theta}, x_{0:T+S} | \boldsymbol{\lambda})$ to the product of all $\hat{c}$ copulas and $\hat{q}$ marginals.\;
-  \caption{MIVB $q$ selection algorithm}
+  For each $t$ in $2, \dots, T$ and copula family $m$ estimate $c_m(t, t-1 | \boldsymbol{\theta}).$ \;
+  Select $\tilde{c}(t, t-1 | \boldsymbol{\theta})$. \;
+   \caption{MIVB Vine selection algorithm}
   \label{alg:MIVB}
@@
 \section{Appendix}
-\iffalse
+\iffalse % hide entire appendix
 

### 2017-05-15 MIVB/MIVB.tex
code cleaning

@@
 \begin{figure}[h]
-\centering
-%\includegraphics{DLMAdvi}
-\caption{The MCMC and ADVI posterior means for each $\boldsymbol{\theta}$ parameter with $100$ data replications at each $T \in \{50, 100, 250, 500, 1000, 2000, 5000, 10000\}$.
+\includegraphics[width=1\linewidth]{DLMAdvi}
+\caption{The MCMC and ADVI posterior means for each $\boldsymbol{\theta}$ parameter with $100$ data replications at each $T \in \{50, 100, 250, 500, 1000, 2000\}$.
 True values for each parameter is denoted by the red line.
-MCMC consistently estimates of the true value, while both ADVI algorithms are only consistent for the mean parameter $\gamma$.
-}
+MCMC consistently estimates of the true value, while both ADVI algorithms are only consistent for the mean parameter $\gamma$.}
 \label{fig:DLM:ADVI}
@@
 
-A further $100 T = 500$ datasets are simulated with the previous parameter and hyperparameter values, and for each $\theta_i$ the ELBO is maximised for $q(\theta_i, | \boldsymbol{\lambda})$, holding each $\theta_{j \neq i}$ and $x_{0:T}$ fixed at their true values.
-$\boldsymbol{\lambda}$ includes only the mean and variance of $\theta_i$, so the ELBO is evaluated over a grid so the stochastic nature of the gradient ascent algorithm is not a confounding factor.
+A further $100 T = 500$ datasets are simulated with the previous parameter and hyperparameter values, and for each $\theta_i$ the ELBO is maximised for $q(\theta_i, | \boldsymbol{\lambda})$, holding each $\theta_{j \neq i}$ and $x_{0:T}$ fixed at their true values, and variance of $\theta_i$ set at the MCMC sample variance.
+$\lambda$ contains only the mean of $\theta_i$, so the ELBO is evaluated over a grid so the stochastic nature of the gradient ascent algorithm is not a confounding factor.
 The maximising value of the mean of $\theta_i$ is plotted in Figure \ref{fig:DLM:Numeric}.
@@
 \centering
-%\includegraphics{DLMNumeric}
-\caption{The ELBO maximising mean for each $\theta$ parameter, holding each other $\theta$ and $x_{0:T}$ fixed.
-Estimation of the mean of $\phi$ improves relative to Figure \ref{fig:DLM:ADVI} but the posterior mean of $\sigma^2_y$ is consistently over-estimated while the posterior mean of $\sigma^2_x$ is consistently under-estimated.}
+\includegraphics{DLMNumeric}
+\caption{The ELBO maximising mean for each $\theta$ parameter, holding each other $\theta$ and $x_{0:T}$ fixed at their true values.
+Estimation of the mean of $\phi$ and $\sigma^2_x$ improves relative to Figure \ref{fig:DLM:ADVI} but the posterior mean of $\sigma^2_y$ is consistently over-estimated.}
 \label{fig:DLM:Numeric}
@@
 \centering
-%\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
+\includegraphics[width=0.4\linewidth,height=\textheight,keepaspectratio]{vines.png}
 \vspace{2mm}
@@
 Denoting the number of elements in $\boldsymbol{\theta}$ by $k$, a vine copula for $q(\boldsymbol{\theta}, x_{0:T} | \boldsymbol{\lambda})$ contains $T+k$ trees, with a number of parameters included in $\boldsymbol{\lambda}$ that is $O(T+k)^2$.
-Constructing the vine so that the first $k$ trees do not contain copulae between any two elements of $x_t$ and the $k+1$'th tree contains copulae $c(x_t, x_{t+1} | \boldsymbol{\theta})$ will force trees $k+r$, with $r \geq 2$, to contain copulae $c(x_t, x_{t+r} | \boldsymbol{\theta}, x_{t+1:t+r-1}).
-These can be set to the independence copula as the Markovian nature of the latent varables implies that $x_t$ and $x_{t+r}$ are conditionally independent given $\boldsymbol{\theta}$ and $x_{t+1:t+r-1}, reducing the number of copula parameters to $O(Tk)$.
+Constructing the vine so that the first $k$ trees do not contain copulae between any two elements of $x_t$ and the $k+1$'th tree contains copulae $c(x_t, x_{t+1} | \boldsymbol{\theta})$ will force trees $k+r$, with $r \geq 2$, to contain copulae $c(x_t, x_{t+r} | \boldsymbol{\theta}, x_{t+1:t+r-1})$.
+These can be set to the independence copula as the Markovian nature of the latent varables implies that $x_t$ and $x_{t+r}$ are conditionally independent given $\boldsymbol{\theta}$ and $x_{t+1:t+r-1}$, reducing the number of copula parameters to $O(Tk)$.
 For convenience the vine is further restricted so that, in each of the first $k$ trees, every $x_t$ has an edge to the same $\theta$ node with the same copula family.
@@
 Di{\ss}mann's algorith (Di{\ss}mann, Brechmann, Czado and Kurowicka 2013) can be used to fit a vine copula to data, selecting both the set of edges for each tree and an appropriate copula family for each edge according to some criteria.
-A common critereon is to select the set of edges that maximise the sum of absolute empirical pairwise Kendell's Tau, with copula families that minimise AIC.
+A common critereon is to select the set of edges that maximise the sum of absolute empirical pairwise Kendall's Tau, with copula families that minimise AIC.
 This algorithm is modified to include the above restrictions.
@@
 Any copula that is to be included in the $q$ distribution is given by $\tilde{c}(i, j | D)$.
-Finally the empirical Kendell's Tau between a pair of variables is denoted by $\tau_{i, j | D}$.
+Finally the empirical Kendall's Tau between a pair of variables is denoted by $\tau_{i, j | D}$.
 Noting that copulae are a function of the CDF of a random variable first transform all MCMC samples of $\boldsymbol{\theta}$ and $x_{0:T}$ to $(0, 1)$ using the empirical marginal CDFs.
@@
 \end{equation}
-For each copula family $m$ and each $t$, estimate the copula $c_m(i^{(1)}, t)$ and choose an optimal family to include in the vine as $\tilde{c}(i^{(1)}, t)$.
-The optimal family could be chosen by criteria such as minimises the sum of AIC across $t$, the minimises AIC for the most $t$.
-In each tree $l = 2, \dots, k$ this processs is repeated: Calculate $\tau_{j, t | i^{(1)}, \dots, i^{(l-1)}$ for each $t$ and element of $\theta$ that has an edge to $\theta^{(l-1)}$ in one of the first $l-1$ trees, then choose $\theta_{i^{(l)}$ where
+For each copula family $m$ and each $t$, estimate the copula $c_m(i^{(1)}, t)$ and choose one family for all of these copula as $\tilde{c}(i^{(1)}, t)$.
+The decision could be on the basis of criteria such as `minimises the sum of AIC across $t$' or `minimises AIC for the most $t$'.
+In each tree $l = 2, \dots, k$ this processs is repeated: Calculate $\tau_{j, t | i^{(1)}, \dots, i^{(l-1)}}$ for each $t$ and element of $\theta$ that has an edge to $\theta^{(l-1)}$ in one of the first $l-1$ trees, then choose $\theta_{i^{(l)}}$ where
 \begin{equation}
@@
 \end{equation}
-Estimate copulae $c_m(i^{(l)}, t | i^{(1:l-1)})$ for each $t$ and $m$ and select an optimal family as $\tilde{c}(i^{(l)}, t |i^{(1)}, \dots, i^{(l-1)})$.
-Tree $k+1$ contains edges $\{t, t+1 |  i^{(1)}, \dots, i^{(k) \}$, with an optimal copula family $\tilde{c}(t, t+1 |i^{(1)}, \dots, i^{(l-k)})$.
-Tree $k+r$ with $r \geq 2$ must then contain edges $\{t, t+r |  i^{(1), \dots, i^{(k), t+1, \dots, t+r-1 \} with $\tilde{c}(t, t+r |  i^{(1), \dots, i^{(k), t+1, \dots, t+r-1) chosen to be the independence copula.
+Estimate copulae $c_m(i^{(l)}, t | i^{(1:l-1)})$ for each $t$ and $m$ and select a family as $\tilde{c}(i^{(l)}, t |i^{(1:l-1)})$.
+Tree $k+1$ contains edges $\{t, t+1 |  i^{(1:k)} \}$, with an optimal copula family $\tilde{c}(t, t+1 |i^{(1:k)})$.
+Tree $k+r$ with $r > 1$ must then contain edges $\{t, t+r |  i^{(1:k)}, t+1, \dots, t+r-1 \}$ with $\tilde{c}(t, t+r |  i^{(1:k)}, t+1, \dots, t+r-1)$ chosen to be the independence copula.
 This approach is given in Algorithm \ref{alg:MIVB}.
@@
 The ELBO can be maximised for this $q$ using Copula Variational Inference (Tran, Blei and Airoldi 2015).
-As we are interested in forecasts of $y_{T+S+h}$ and $x_t$ is Markovian, the distribution of each $x_s, s < t$ is irrelevant given the distribution of $x_t$.
-Holding $\lambda}$ parameters assocaited with $x_t, t = 0, \dots, T$ fixed at MCMC data estimated values will drastically reduce the computational cost of the gradient ascent algorirhtm as only the $\lambda$ parameters associated with the marginal distributions of $\theta$ and $x_{T+1:T+S}$, as well as the copulae between these, needs to be optimised.
+As we are interested in forecasts of $y_{T+S+h}$ and $x_t$ is Markovian, the distribution of each $x_t$ is irrelevant given the distribution of $x_s, s > t$.
+Holding $\lambda$ parameters assocaited with $x_t, t = 0, \dots, T$ fixed at MCMC data estimated values will drastically reduce the computational cost of the gradient ascent algorirhtm as only the $\lambda$ parameters associated with the marginal distributions of $\theta$ and $x_{T+1:T+S}$, as well as the copulae between these, needs to be optimised.
 
@@
  \SetKwInOut{Input}{Input}
- \Input{Observations of $\boldsymbol{\theta}, x_{1:T}$,  $k$, the number of elements in $\boldsymbol{\theta}$. \;}
- \KwResult{Set of copulae $\tilde{c}(\cdot, \codt)$}
+ \Input{Observations of $\boldsymbol{\theta}, x_{1:T}$,  $k$, the number of elements in $\boldsymbol{\theta}$.}
+ \KwResult{Set of copulae $\tilde{c}(\cdot, \cdot)$}
  Transform observations for each variable to $(0, 1)$ by the inverse CDF\;

### 2017-05-15 MIVB/DLM_1Delbo.R

@@ -1,97 +1,130 @@
 library(tidyr)
 library(ggplot2)
 
-lbsigy = function(mean){
-  elbo = 0
-  for(i in 1:100){
-    sigx = exp(rnorm(1, mean, 0.1))
-    dens = -2*log(sigx) - 1/sigx #- x0^2*(1-0.9^2)/(2*sigx) 
-    #x = c(x0, x)
-    for(t in 1:T){
-      dens = dens + dnorm(y[t], 1+x[t], sqrt(sigx), log=TRUE)
+# Inputs: Range of mean for approx. distribution, variance of MCMC posterior, latent variables x0 & x, data y
+lbsigy = function(meanrange, var, x0, x, y){
+  # Calculate the ELBO at each value in meanrange
+  elbo = rep(0, length(meanrange))
+  # Use the same random numbers for each mean
+  eps = rnorm(100)
+  # For each mean value
+  for(j in seq_along(meanrange)){
+    # The ELBO is an average across 100 draws of sigma^2_y
+    for(i in 1:100){
+      sigy = exp(meanrange[j] + sqrt(var)*eps[i])
+      # Prior density
+      logp =  -2*log(sigy) - 1/sigy
+      for(t in 1:T){
+        # From p(y | theta)
+        logp = logp + dnorm(y[t], 1+x[t], sqrt(sigy), log=TRUE)
+      }
+      # Approx density q(log(sigma^2_y))
+      logq = dnorm(log(sigy), meanrange[j], sqrt(var), log=TRUE)
+      # All other components of log(p) and log(q) are constant with respect to sigma^2_y
+      elbo[j] = elbo[j] + (logp-logq)/100
     }
-    dens = dens - dnorm(log(sigx), mean, 0.1, log = TRUE)
-    elbo = elbo + dens/100
   }
   return(elbo)
 }
-lbsigx = function(mean){
-  elbo = 0
-  for(i in 1:100){
-    sigx = exp(rnorm(1, mean, 0.1))
-    dens = -2.5*log(sigx) - 1/sigx - x0^2*(1-0.9^2)/(2*sigx) 
+
+# Other three parameters work the same way (except for truncated normal sampling of phi)
[…]

### 2017-05-15 MIVB/DLM_MIVB.R

@@ -8,7 +8,7 @@ source('qSelection.R')
 source('DLM_CopVB.R')
 
 
-MIVB = function(T, S, h, sigmaSqY=1, sigmaSqX=2, phi=0.95, gamma=2, MCMCreps=T*10){
+MIVB = function(T, S, h, sigmaSqY=1, sigmaSqX=1, phi=0.95, gamma=2, MCMCreps=T*10){
 
   # Simulate Data
   y = SimDLM(T, S, h, sigmaSqY, sigmaSqX, phi, gamma)
@@ -96,7 +96,7 @@ MIVB = function(T, S, h, sigmaSqY=1, sigmaSqX=2, phi=0.95, gamma=2, MCMCreps=T*1
   # Select and Estimate the Vine
   Vine = RestrictedVineSelect(T, subset, 4, MCMCreps, MCMC)
   
-  VineObj = RVineMatrix(Vine$Matrix, Vine$family, Vine$par, Vine$par2)
+  #VineObj = RVineMatrix(Vine$Matrix, Vine$family, Vine$par, Vine$par2)
   # Run the VB updater to time T+S
   VBfit = CopulaVB(y[1:(T+S)], S, thetaDist, xDist, thetaParams, xParams, Vine, M, maxIter)
 

### 2017-07-13 Tran Nott Kohn/outline.tex
PF work

\documentclass[12pt,a4paper]{article}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage{amsthm}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\numberwithin{equation}{section}
\setlength\parindent{0pt}
\begin{document}

\section{Introduction}

Consider a high frequency time series with observed values $y_{1}, \dots, y_{T+S}$, collectively denoted by $y_{1:T+S}$, with a forecaster interested in the forecast distribution of the future value $y_{T+S+h}$, where $S \geq 1$. In many applications forecasting uncertainty can be reduced by conditioning on additional data, so knowledge of the distribution of $p(y_{T+S+h} | y_{1:T+S})$ is preferable to that of $p(y_{T+S+h} | y_{1:T})$.

Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$, such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:t-1}$, given $\boldsymbol{\theta}$ and $x_t$. One of the most common computational techniques for models with latent variables is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution, however these approaches can have slow convergence rates. In this paper it is assumed that an MCMC algorithm will require a time period of length $S$ to converge, so at time $T+S$ the posterior distribution of $\boldsymbol{\theta}$  and $x_{0:T+S}$ can be conditioned only on $y_{1:T}$ and hence only $p(y_{1:T+S+h} | y_{1:T})$ is available.

The additional data $y_{T+1:T+S}$ can be included in the forecast by using filtering techniques such as the Kalman filter or particle filtering, however this will only update knowledge on $x_{0:T+S}$ but not $\boldsymbol{\theta}$. As an alternative to this method, this paper will instead approximate the posterior distribution $p(\boldsymbol{\theta}, x_{T} | y_{1:T})$ with $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T} | y_{1:T})$ and the updated posterior distribution $p(\boldsymbol{\theta}, x_{T+S} | y_{1:T+S})$ with $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T+S} | y_{1:T+S})$. The parameters of this approximation, denoted by $\boldsymbol{\lambda}$, are optimised by gradient descent to minimise the Kullback-Leibler divergence from $q$ to $p$, a technique known as Variational Inference, which has seen wide use in the literature (references).

Variational Bayes involves an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error. This paper additionally explores that trade-off for a Stochastic Volatility Model, a common latent state time series model.

\begin{itemize}
\item Some motivation about forecasting and updates, can probably be copied from the MIVB file
\item Not focusing on MCMC sample
\item VB estimate at time T, VB update to time T+S
\item End goal: An approximation $q(\theta, x_{T})$ for use in forecasting.
\end{itemize}

\section{MCMC and Particle Filtering}
\begin{itemize}
\item MCMC background
\item Stochastic Volatility Model
\item PMCMC
\item Data driven particle filters?
\item Whatever else
\end{itemize}

\section{Variational Inference}

Variational Inference posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ and some approximating distribution $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.

This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ to the true posterior $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ The KL divergence is defined by
\begin{align}
\label{KL-def}
KL[q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})] &= \nonumber \\
\int q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T}) &\ln \left( \frac{q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})}{p(\boldsymbol{\theta}, x_{0:T} |y_{1:T})}\right) d\boldsymbol{\theta} dx_{0:T}
\end{align}
and can alternatively be expressed as
\begin{equation}
\label{KL-ELBO}
KL[q(\boldsymbol{\theta}, x_{0:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(\boldsymbol{\lambda})
\end{equation}
where $\mathcal{L}(\boldsymbol{\lambda})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.  $\mathcal{L}(\boldsymbol{\lambda})$ is defined by
\begin{equation}
\label{ELBO}
\mathcal{L}(\boldsymbol{\lambda}) = \int q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T}) \ln \left( \frac{p(y_{1:T},\boldsymbol{\theta}, x_{0:T})}{q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})} \right) d\boldsymbol{\theta}dx_{1:T},
\end{equation}
and as $\ln(p(y_{1:T}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}). Maximising (\ref{ELBO}) is more convenient that minimising (\ref{KL-def}) as it is a function of the known joint distribution $p(y_{1:T}, \boldsymbol{\theta}, x_{0:T})$ instead of the unknown posterior $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$.

\subsection{ELBO Optimisation}
Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where the following update step is iteratively applied until (\ref{ELBO}) converges within some pre-specified tolerance:
\begin{equation}
\label{GradAscent}
\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \frac{\delta}{\delta\boldsymbol{\lambda}} \mathcal{L}(\boldsymbol{\lambda^{(m)}}),
\end{equation}
where the derivative is evaluated at $\boldsymbol{\lambda}^{(m)}$. This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate. If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum (Robbins and Munro, 1951).
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}

This paper uses Adam \citep{Kingma2015b} to generate the sequence $\rho^{(m)}$.

Ranganath, Gerrish and Blei (2014) showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
\begin{equation}
\label{ScoreDeriv}
\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda^{(m)}}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} [\ln(q_{\boldsymbol{\lambda}^{(m)}}(\boldsymbol{\theta_i}, x_{0:T, i} | y_{1:T})] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_i, x_{1:T, i})}{q_{\boldsymbol{\lambda}^{(m)}}(\boldsymbol{\theta_i}, x_{0:T, i} | y_{1:T})} \right) 
\end{equation}
where $i = 1, \dots, N$ indicates independent simulations from $q_{\boldsymbol{\lambda}^{(m)}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$. 
The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $N$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.

The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), introducing a noise variable $\boldsymbol{\epsilon}$ with a distribution $q(\boldsymbol{\epsilon})$ with no free parameters, and a differentiable transform $f$ such that
\begin{equation}
\label{Reparam}
f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}) = \{\boldsymbol{\theta}, x_{1:T}\}.
\end{equation}
Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable. In this case, (\ref{ELBO}) becomes
\begin{equation}
\label{ReparamELBO}
\mathcal{L}(\boldsymbol{\lambda}) = \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T},f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}) |\det J(\boldsymbol{\epsilon}, \boldsymbol{\lambda})|}{q(\boldsymbol{\epsilon})} \right) d\boldsymbol{\epsilon},
\end{equation}
where $J(\boldsymbol{\epsilon}, \boldsymbol{\lambda})$ is the Jacobian matrix of the transformation $f$. The derivative of (\ref{ReparamELBO}) can be estimated by 
\begin{equation}
\label{ReparamDeriv}
\frac{\delta}{\delta \boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}) \approx \frac{1}{M}\sum_{i=1}^{M}
\frac{\delta f(\boldsymbol{\epsilon}_i, \boldsymbol{\lambda}^{(m)})} {\delta \boldsymbol{\lambda}}
\frac{\delta \ln (p(y_{1:T}, f(\boldsymbol{\epsilon}_i, \boldsymbol{\lambda}^{(m)})))}{\delta f(\boldsymbol{\epsilon}_i, \boldsymbol{\lambda}^{(m)})} + \frac{\delta  \ln(|\det J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|)}{\delta \boldsymbol{\lambda}},
\end{equation} 
where simulations of $\boldsymbol{\theta}$ and $x_{1:T+S}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$. The variance of this estimator is often orders of magnitude smaller than the estimator in (\ref{ScoreDeriv}) (citations), so $M$ can be set much lower than $N$.

\subsection{Randomised Quasi Monte Carlo}
Randomised Quasi Monte Carlo (Matousek 1998; Sobol 196X, GTN 2017).

\subsection{Choice of $q$ distribution}

\section{Dimensionality Reduction}

ADVI simulation results

As the latent variables are Markovian, once the distribution of $x_T$ is available, the distributions of each previous $x_t, t < T$ is irrelevant to forecasts of $y_{T+S+h}$ and it may be more convenient to construct an approximation for $p(\boldsymbol{\theta}, x_T | y_{1:T})$ than $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ as the dimensionality of the posterior is reduced from $T+k+1$ to $k+1$, where $k$ is the number of elements in $\boldsymbol{\theta}$. This in turn allows for more expressive approximating distributions to be used as the size of $\boldsymbol{\lambda}$ does not grow with $T$.

In this case,
\begin{equation}
\label{DimRedLogJoint}
\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) = \ln(p(x_T | y_{1:T}, \boldsymbol{\theta})) + \ln(p(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
\end{equation}
In many cases, the distributions $p(x_T | y_{1:T}, \boldsymbol{\theta})$ and $p(y_{1:T} | \boldsymbol{\theta})$ are intractable, but \citet{TNK2017} note that the particle filter estimators $\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})$ and $\hat{p}(y_{1:T} | \boldsymbol{\theta})$ are sufficient for use in what they refer to as Variational Bayes with Intractable Likelihood (VBIL). Hence
\begin{equation}
\label{PFLogJoint}
\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) \approx \ln(\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})) + \ln(\hat{p}(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
\end{equation}
can be substituted into (\ref{ReparamDeriv}). This derivative of (\ref{PFLogJoint}) with respect to $\boldsymbol{\lambda}$ can be obtained with automatic differentiation tools such as those provided by Stan \citep{Carpenter2015}.

\subsection{Natural Gradients}

lower dimension 
analytical for some distributions eg. MVN

Natural Gradients  (Amari 1998, Hoffman et al. 2013).

\subsection{Importance Sampling}

don't need to particle filter after every iteration

Importance Sampling (Sakaya and Klami 2017)

\section{Updating}

\begin{itemize}
\item Proper updates with filter/smooth
\item Secondary approximation - no smoothing
\item Straight up particle filter on old approximation
\item PF on old MCMC
\item Simulation Results
\end{itemize}

\section{Empirical}

\section{Discussion}

\bibliographystyle{asa}
\bibliography{references}

\section{Appendix}
\subsection{Inverse Fisher Matrix}

\end{document}

### 2017-07-13 Tran Nott Kohn/ADPF.cpp

@@ -1,0 +1,376 @@
+// [[Rcpp::depends(rstan)]]
+// [[Rcpp::depends(RcppEigen)]]
+// [[Rcpp::plugins(cpp11)]]
+// [[Rcpp::depends(RcppArmadillo)]]
+
+
+#include <RcppArmadillo.h>
+#include <stan/math.hpp>
+#include <vector>
+#include <cmath>
+#include <math.h>
+#include <Eigen/Dense>
+#include <RcppEigen.h>
+#include <Rcpp.h>
+#include <boost/math/distributions.hpp> 
+#include <cstdlib> 
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+
+using namespace std;
+using Eigen::Matrix;
+using Eigen::Dynamic;
+using Eigen::MatrixXd; 
+using Eigen::Map;
+using namespace arma;
+
+//TODO: Add natural gradients and importance sampling
+
+struct logP {
+  const MatrixXd& epsilon;
+  const MatrixXd& y;
+  const MatrixXd& noise;
+  logP(const MatrixXd& epsIn, const MatrixXd& yIn, const MatrixXd& noiseIn) :
+    epsilon(epsIn), y(yIn), noise(noiseIn) {}
+  template <typename T>
+  T operator ()(const Matrix<T, Dynamic, 1>& lambda)
+    const{
+    using std::log; using std::pow; using std::exp;
+    
+    const int dim = 4;
+    // Transform epsilon to theta
+    Matrix<T, Dynamic, 1>& theta(dim);
+    for(int i = 0; i < dim; ++i){
+      theta(i) = lambda(i);
+      for(int j = 0; j <= i; ++j){
+        theta(i) += lambda(j + dim*(i+1)) * epsilon(j);
+      }
+    }
+    T sigSq = exp(theta(0));
+    T mu = theta(1);
+    T tau = theta(2);
+    T xT = theta(3);
+    
+    // Evaluate log(p(theta))
+    T prior = -pow(mu, 2) / 20  +  19.0 * log(tau)  +  0.5 * log(1-tau)  - (3.5) * log(sigSq)  -  0.25 / sigSq;
+    
+    // Set up for particle filtering
+    int obs = y.size();
+    const int N = 100;
+    T phi = 2 * tau  -  1;
+    Matrix<T, N, 1> xOld = mu + sqrt(sigSq / (1-pow(phi, 2))) * noise.col(0); // deal with this
[…]

### 2017-07-13 Tran Nott Kohn/TNK2016.cpp

@@ -1,0 +1,317 @@
+// [[Rcpp::depends(rstan)]]
+// [[Rcpp::depends(RcppEigen)]]
+// [[Rcpp::plugins(cpp11)]]
+// [[Rcpp::depends(RcppArmadillo)]]
+
+
+#include <RcppArmadillo.h>
+#include <stan/math.hpp>
+#include <vector>
+#include <cmath>
+#include <math.h>
+#include <Eigen/Dense>
+#include <RcppEigen.h>
+#include <Rcpp.h>
+#include <boost/math/distributions.hpp> 
+#include <cstdlib> 
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+
+using namespace std;
+using Eigen::Matrix;
+using Eigen::Dynamic;
+using Eigen::MatrixXd; 
+using Eigen::Map;
+using namespace arma;
+
+// [[Rcpp::export]]
+mat sobol_points(int N, int D) {
+  ifstream infile("new-joe-kuo-6.21201" ,ios::in);
+  if (!infile) {
+    cout << "Input file containing direction numbers cannot be found!\n";
+    exit(1);
+  }
+  char buffer[1000];
+  infile.getline(buffer,1000,'\n');
+  
+  // L = max number of bits needed 
+  unsigned L = (unsigned)ceil(log((double)N)/log(2.0)); 
+  
+  // C[i] = index from the right of the first zero bit of i
+  unsigned *C = new unsigned [N];
+  C[0] = 1;
+  for (unsigned i=1;i<=N-1;i++) {
+    C[i] = 1;
+    unsigned value = i;
+    while (value & 1) {
+      value >>= 1;
+      C[i]++;
+    }
+  }
+  
+  // POINTS[i][j] = the jth component of the ith point
+  //                with i indexed from 0 to N-1 and j indexed from 0 to D-1
+  mat POINTS(N, D, fill::zeros);
+  
+  // ----- Compute the first dimension -----
+  
+  // Compute direction numbers V[1] to V[L], scaled by pow(2,32)
+  unsigned *V = new unsigned [L+1]; 
+  for (unsigned i=1;i<=L;i++) V[i] = 1 << (32-i); // all m's = 1
+  
+  // Evalulate X[0] to X[N-1], scaled by pow(2,32)
+  unsigned *X = new unsigned [N];
+  X[0] = 0;
+  for (unsigned i=1;i<=N-1;i++) {
+    X[i] = X[i-1] ^ V[C[i-1]];
[…]

### 2017-07-14 Tran Nott Kohn/ADPF.cpp
PF work

@@ -25,7 +25,7 @@ using Eigen::MatrixXd;
 using Eigen::Map;
 using namespace arma;
 
-//TODO: Add natural gradients and importance sampling
+//TODO: Add importance sampling
 
 struct logP {
   const MatrixXd& epsilon;
@@ -36,75 +36,68 @@ struct logP {
   template <typename T>
   T operator ()(const Matrix<T, Dynamic, 1>& lambda)
     const{
-    using std::log; using std::pow; using std::exp;
+    using std::log; using std::pow; using std::exp; using std::sqrt;
     
-    const int dim = 4;
     // Transform epsilon to theta
-    Matrix<T, Dynamic, 1>& theta(dim);
-    for(int i = 0; i < dim; ++i){
-      theta(i) = lambda(i);
-      for(int j = 0; j <= i; ++j){
-        theta(i) += lambda(j + dim*(i+1)) * epsilon(j);
-      }
-    }
-    T sigSq = exp(theta(0));
-    T mu = theta(1);
-    T tau = theta(2);
-    T xT = theta(3);
+    T sigSq = exp(lambda(0) + lambda(4)*epsilon(0));
+    T mu = lambda(1)  +  lambda(8) * epsilon(0)   +  lambda(9) * epsilon(1);
+    T tau = lambda(2)  +  lambda(12) * epsilon(0)   +  lambda(13) * epsilon(1)  +  lambda(14) * epsilon(2);
+    T xT = lambda(3)  +  lambda(16) * epsilon(0)   +  lambda(17) * epsilon(1)  +  
+      lambda(18) * epsilon(2)  +  lambda(19) * epsilon(3);
     
     // Evaluate log(p(theta))
     T prior = -pow(mu, 2) / 20  +  19.0 * log(tau)  +  0.5 * log(1-tau)  - (3.5) * log(sigSq)  -  0.25 / sigSq;
-    
     // Set up for particle filtering
+    
     int obs = y.size();
     const int N = 100;
     T phi = 2 * tau  -  1;
-    Matrix<T, N, 1> xOld = mu + sqrt(sigSq / (1-pow(phi, 2))) * noise.col(0); // deal with this
-    Matrix<T, N, 1> xNew;
-    Matrix<T, N, 1> xResample;
-    Matrix<T, N, 1> pi; pi.fill(1.0/N);
-    Matrix<T, N, 1> piSum = pi;
-    Matrix<T, N, 1> omega;
[…]

### 2017-07-17 Tran Nott Kohn/outline.tex
PF work
ffs

@@
 \subsection{Randomised Quasi Monte Carlo}
-Randomised Quasi Monte Carlo (Matousek 1998; Sobol 196X, GTN 2017).
+Randomised Quasi Monte Carlo (Matousek 1998; Sobol 196X, Gunawan, Tran and Kohn 2017).
 
@@
 
-\subsection{Natural Gradients}
-
-lower dimension 
-analytical for some distributions eg. MVN
-
-Natural Gradients  (Amari 1998, Hoffman et al. 2013).
-
 \subsection{Importance Sampling}
-
-don't need to particle filter after every iteration
 
@@
 
-\section{Appendix}
-\subsection{Inverse Fisher Matrix}
-
 \end{document}

### 2017-07-17 Tran Nott Kohn/ADPF.cpp

@@ -304,7 +304,7 @@ Rcpp::List VBIL_PF (Rcpp::NumericMatrix yIn, Rcpp::NumericMatrix lambdaIn, int S
   
   
   // Initial SOBOL QMC numbers
-  mat sobol = sobol_points(4, S+100).t();
+  mat sobol = sobol_points(S+100, 4);
   // Loop control
   int iter = 0;
   double diff = threshold + 1;
@@ -370,6 +370,7 @@ Rcpp::List VBIL_PF (Rcpp::NumericMatrix yIn, Rcpp::NumericMatrix lambdaIn, int S
             theta(i, s) += lambda((i+1)*4+j)*epsilon(j);
           }
         }
+        theta(0, s) = exp(theta(0, s));
         // generate standard normal noise for the particle filter, convert to Eigen format
         Rcpp::NumericMatrix noiseRcpp(100, T+1);
         for(int t = 0; t < T+1; ++t){

### 2017-07-18 Tran Nott Kohn/outline.tex
Particle Filter is doing okay

@@
 
-This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence (Kullback and Leibler, 1951) from $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ to the true posterior $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ The KL divergence is defined by
+This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ to the true posterior $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ The KL divergence is defined by
 \begin{align}
@@
 This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
-If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum (Robbins and Munro, 1951).
+If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.
 \begin{align}
@@
 
-Ranganath, Gerrish and Blei (2014) showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
+\citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
 \begin{equation}
@@
 
-The variance can be reduced by the reparameterisation trick of Kingma and Welling (2014), introducing a noise variable $\boldsymbol{\epsilon}$ with a distribution $q(\boldsymbol{\epsilon})$ with no free parameters, and a differentiable transform $f$ such that
+The variance can be reduced by the reparameterisation trick of \citet{Kingma2014}, introducing a random vector $\boldsymbol{\epsilon}$ with a distribution $q(\boldsymbol{\epsilon})$ that contains no free parameters, and a differentiable transform $f$ such that
 \begin{equation}
@@
 \end{equation}
-Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal noise variable and an inverse-CDF transform of a uniform noise variable.
+Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal $\epsilon$ and an inverse-CDF transform of a uniform $\epsilon$.
 In this case, (\ref{ELBO}) becomes
@@
 \subsection{Randomised Quasi Monte Carlo}
-Randomised Quasi Monte Carlo (Matousek 1998; Sobol 196X, Gunawan, Tran and Kohn 2017).
+\citep{Gunawan2017}
+\citep{Matousek1998}
+\citep{Bratley1988}
+\citep{Sobol1967}
 
@@
 
-As the latent variables are Markovian, once the distribution of $x_T$ is available, the distributions of each previous $x_t, t < T$ is irrelevant to forecasts of $y_{T+S+h}$ and it may be more convenient to construct an approximation for $p(\boldsymbol{\theta}, x_T | y_{1:T})$ than $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ as the dimensionality of the posterior is reduced from $T+k+1$ to $k+1$, where $k$ is the number of elements in $\boldsymbol{\theta}$.
-This in turn allows for more expressive approximating distributions to be used as the size of $\boldsymbol{\lambda}$ does not grow with $T$.
+As the latent variables are Markovian, once the distribution of $x_{T+1}$ is available, the distributions of each previous $x_t, t \leq T$ is irrelevant to forecasts of $y_{T+S+h}$ and it may be more convenient to construct an approximation for $p(\boldsymbol{\theta}, x_{T+1} | y_{1:T})$ than for $p(\boldsymbol{\theta}, x_{0:T+1} | y_{1:T})$, as they contain the same information about $y_{T+S+h}$ but the dimensionality of the posterior is reduced from $T+k+2$ to $k+1$, where $k$ is the number of elements in $\boldsymbol{\theta}$.
+This in turn allows for more expressive approximating distributions to be used, as the size of $\boldsymbol{\lambda}$ does not grow with $T$.
 
-In this case,
+In this case, using $f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}) = \{\boldsymbol{\theta}, x_{T+1}\}$,
 \begin{equation}
 \label{DimRedLogJoint}
-\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) = \ln(p(x_T | y_{1:T}, \boldsymbol{\theta})) + \ln(p(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
+\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) = \ln\left(\int_{X_T} p(x_{T+1} | \theta, x_T)p(x_T | y_{1:T}, \boldsymbol{\theta})d_{X_T}\right) + \ln(p(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
 \end{equation}
-In many cases, the distributions $p(x_T | y_{1:T}, \boldsymbol{\theta})$ and $p(y_{1:T} | \boldsymbol{\theta})$ are intractable, but \citet{TNK2017} note that the particle filter estimators $\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})$ and $\hat{p}(y_{1:T} | \boldsymbol{\theta})$ are sufficient for use in what they refer to as Variational Bayes with Intractable Likelihood (VBIL).
-Hence
+In many cases, the distributions $p(x_T | y_{1:T}, \boldsymbol{\theta})$ and $p(y_{1:T} | \boldsymbol{\theta})$ are intractable, but \citet{TranM2017} note that the particle filter estimators $\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})$ and $\hat{p}(y_{1:T} | \boldsymbol{\theta})$ are sufficient substitutes in what they refer to as Variational Bayes with Intractable Likelihood (VBIL).
+The particle filter estimation of the distribution $\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})$ is expressed as a discrete set of point masses $x_T^{(k)}$ and weights $\pi_T^{(k)}$, where $k = 1, \dots, N$ represent the particles of the particle filter.
+In this case, the integral in (\ref{DimRedLogJoint}) reduces to the sum
+\begin{equation}
+\label{ParticleFilterIntegral}
+\int_{X_T} p(x_{T+1} | \theta, x_T)p(x_T | y_{1:T}, \boldsymbol{\theta})d_{X_T} = \sum_{k=1}^N \pi_T^{(k)} p(x_{T+1} | \boldsymbol{\theta}, x_T^{(k)}),
+\end{equation}
+and hence
 \begin{equation}
 \label{PFLogJoint}
-\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) \approx \ln(\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})) + \ln(\hat{p}(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
+\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) \approx \ln \left(\sum_{k=1}^N \pi_T^{(k)} p(x_{T+1} | \boldsymbol{\theta}, x_T^{(k)})\right) + \ln(\hat{p}(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
 \end{equation}
@@
 
-Importance Sampling (Sakaya and Klami 2017)
+\citep{Sakaya2017}
 
 \section{Updating}
+
+$$p(y_{1:T+S}, \theta, x_T, x_{T+S}) = p(\theta) p(y_{1:T+S} | \theta) p (x_T | \theta, y_{1:T+S}) p(x_{T+S} | x_T, \theta, y_{1:T+S})$$
 

### 2017-07-18 Tran Nott Kohn/ADKF.cpp

@@ -41,7 +41,7 @@ struct logP {
     T xT = theta(3);
     
     // Evaluate log(p(theta)
-    T prior = -pow(mu, 2) / 200  - pow(phi - 0.5, 2) / (2 * 0.01)  -  2 * log(sigSq)  -  1.0 / sigSq;
+    T prior = -pow(mu, 2) / 200  - pow(phi - 0.5, 2) / 0.02  -  2 * log(sigSq)  -  1.0 / sigSq;
     
     // Set up for kalman filtering
     Matrix<T, Dynamic, 1> at(obs), pt(obs), vt(obs), St(obs), Kt(obs), att(obs+1), ptt(obs+1);
@@ -295,10 +295,9 @@ Rcpp::List VBIL_KF (Rcpp::NumericMatrix yIn, Rcpp::NumericMatrix lambdaIn, int S
             impliedEpsilon(i) -= lambda((i+1)*4+j) * epsilon(j);
           }
         }
+        logQeval = evalLogQ(impliedEpsilon, lambda);
+        gradientQ = QDeriv(impliedEpsilon, lambda);
         gradientF = FDeriv(impliedEpsilon, lambda);
-        logQ q(epsilon.col(s));
-        stan::math::set_zero_all_adjoints();
-        stan::math::gradient(q, lambda, logQeval, gradientQ);
      
         double weight = 0;
         for(int i = 0; i < 4; ++i){

### 2017-07-18 Tran Nott Kohn/ADPF.cpp

@@ -25,46 +25,51 @@ using Eigen::MatrixXd;
 using Eigen::Map;
 using namespace arma;
 
-//TODO: Add importance sampling
+
+// evaluate and automatially differentiate log(p(y, xT, theta)) wrt theta/xT
 
 struct logP {
-  const MatrixXd& epsilon;
-  const MatrixXd& y;
-  const MatrixXd& noise;
-  logP(const MatrixXd& epsIn, const MatrixXd& yIn, const MatrixXd& noiseIn) :
-    epsilon(epsIn), y(yIn), noise(noiseIn) {}
-  template <typename T>
-  T operator ()(const Matrix<T, Dynamic, 1>& lambda)
+  const MatrixXd& y; // data
+  const MatrixXd& noise; // standard normal matrix for the particle filter transitions
+  const int& obs; // T is taken for the variable type, so obs = number of elements in y
+  logP(const MatrixXd& yIn, const MatrixXd& noiseIn, const int& obsIn) : // constructor
+    y(yIn), noise(noiseIn), obs(obsIn) {}
+  template <typename T> // T will become stan's var that can be automatically differentiated - any variable differentiated in the chain rule for d logp / d theta must be type T
+  T operator ()(const Matrix<T, Dynamic, 1>& theta) // derivative is with respect to theta (with xT+1 included), so theta matrix is the operator () input.
     const{
-    using std::log; using std::pow; using std::exp; using std::sqrt;
+    using std::log; using std::pow; using std::exp; using std::sqrt; // declare mathematical operators used so stan knows to differentiate these.
     
-    // Transform epsilon to theta
-    T sigSq = exp(lambda(0) + lambda(4)*epsilon(0));
-    T mu = lambda(1)  +  lambda(8) * epsilon(0)   +  lambda(9) * epsilon(1);
-    T tau = lambda(2)  +  lambda(12) * epsilon(0)   +  lambda(13) * epsilon(1)  +  lambda(14) * epsilon(2);
-    T xT = lambda(3)  +  lambda(16) * epsilon(0)   +  lambda(17) * epsilon(1)  +  
-      lambda(18) * epsilon(2)  +  lambda(19) * epsilon(3);
+    // Get elements of theta
+    T sigSq = theta(0);
[…]

### 2017-07-18 Tran Nott Kohn/ADPF.cpp
PF

@@ -433,7 +433,7 @@ Rcpp::List VBIL_PF (Rcpp::NumericMatrix yIn, Rcpp::NumericMatrix lambdaIn, int S
           meanGradientSq(i) += pow(S * meanGradient(i), 2) / S;
         }
         // Update ELBO
-        LB(iter-1) += (logPeval(s) - logQeval) * weight / S;
+        LB(iter-1) += (logPeval(s) - logQeval) / S;
       }
     }
     // adam updating rule for lambda values

### 2017-07-19 Tran Nott Kohn/ADKF.cpp
Particle Filter in good condition

@@ -38,7 +38,7 @@ struct logP {
     T sigSq = theta(0);
     T mu = theta(1);
     T phi = theta(2);
-    T xT = theta(3);
+    T xTplus1 = theta(3);
     
     // Evaluate log(p(theta)
     T prior = -pow(mu, 2) / 200  - pow(phi - 0.5, 2) / 0.02  -  2 * log(sigSq)  -  1.0 / sigSq;
@@ -50,7 +50,7 @@ struct logP {
     ptt(0) = 3 * sigSq;  // var t = 0 | 0
     T ytDens; // log(p(yt | theta))
     T yAllDens = 0; // log(p(y_1:T | theta))
-    T xTDens = 0; // log(p(x_T | theta, y_1:T))
+    T xTplus1Dens = 0; // log(p(x_T | theta, y_1:T))
     // kalman loop
     for(int t = 0; t < obs; ++t){
       at(t) = mu + phi * (att(t) - mu); // mean t | t-1
@@ -63,10 +63,12 @@ struct logP {
       ytDens = - 0.5 * log(2*3.14159) -  0.5 * log(St(t))  -  pow(y(t-1) - pt(t), 2) / (2 * St(t)); 
       yAllDens += ytDens;
     }
-    // log(p(xT | theta, y_1:T))
-    xTDens = - 0.5 * log(2*3.14159)  -  0.5 * log(ptt(obs))  -  pow(xT - att(obs), 2) / (2 * ptt(obs));
+    // log(p(xT+1 | theta, y_1:T))
+    T xTp1Mean = mu - phi * (att(obs) - mu);
+    T xTp1Var = pow(phi, 2) * ptt(obs) + sigSq;
+    xTplus1Dens = - 0.5 * log(2*3.14159)  -  0.5 * log(xTp1Var)  - pow(xTplus1 -xTp1Mean, 2) / (2 * xTp1Var);
   
-    return yAllDens + xTDens + prior;
+    return yAllDens + xTplus1Dens + prior;
   }
 };
 

### 2017-07-19 Tran Nott Kohn/ADPF.cpp

@@ -3,7 +3,6 @@
 // [[Rcpp::plugins(cpp11)]]
 // [[Rcpp::depends(RcppArmadillo)]]
 
-
 #include <RcppArmadillo.h>
 #include <stan/math.hpp>
 #include <vector>
@@ -25,33 +24,30 @@ using Eigen::MatrixXd;
 using Eigen::Map;
 using namespace arma;
 
-
-// evaluate and automatially differentiate log(p(y, xT, theta)) wrt theta/xT
-
+// evaluate and automatially differentiate log(p(y_1:T, x_T+1, theta)) wrt theta and x_T+1
 struct logP {
   const MatrixXd& y; // data
   const MatrixXd& noise; // standard normal matrix for the particle filter transitions
   const int& obs; // T is taken for the variable type, so obs = number of elements in y
-  logP(const MatrixXd& yIn, const MatrixXd& noiseIn, const int& obsIn) : // constructor
-    y(yIn), noise(noiseIn), obs(obsIn) {}
+  const int& N; // Number of particles
+  logP(const MatrixXd& yIn, const MatrixXd& noiseIn, const int& obsIn, const int& nIn) : // constructor
+    y(yIn), noise(noiseIn), obs(obsIn), N(nIn) {}
   template <typename T> // T will become stan's var that can be automatically differentiated - any variable differentiated in the chain rule for d logp / d theta must be type T
   T operator ()(const Matrix<T, Dynamic, 1>& theta) // derivative is with respect to theta (with xT+1 included), so theta matrix is the operator () input.
     const{
-    using std::log; using std::pow; using std::exp; using std::sqrt; // declare mathematical operators used so stan knows to differentiate these.
+    using std::log; using std::pow; using std::exp; using std::sqrt; // explicitly declare mathematical operators used so stan knows to differentiate these.
     
     // Get elements of theta
     T sigSq = theta(0);
     T mu = theta(1);
     T tau = theta(2);
     T xTplus1 = theta(3);
-    T phi = 2 * tau - 1;
-    
+    T phi = 2*tau - 1;
     // Evaluate log(p(theta))
[…]

### 2017-07-31 Tran Nott Kohn/outline.tex
PF
Particle Filter in good condition
Various updates

@@
 \end{itemize}
+
+\citep{Arulampalam2002}
 
@@
 \label{ParticleFilterIntegral}
-\int_{X_T} p(x_{T+1} | \theta, x_T)p(x_T | y_{1:T}, \boldsymbol{\theta})d_{X_T} = \sum_{k=1}^N \pi_T^{(k)} p(x_{T+1} | \boldsymbol{\theta}, x_T^{(k)}),
+\int_{x_T} p(x_{T+1} | \theta, x_T)p(x_T | y_{1:T}, \boldsymbol{\theta})d_{x_T} = \sum_{k=1}^N \pi_T^{(k)} p(x_{T+1} | \boldsymbol{\theta}, x_T^{(k)}),
 \end{equation}
@@
 
-$$p(y_{1:T+S}, \theta, x_T, x_{T+S}) = p(\theta) p(y_{1:T+S} | \theta) p (x_T | \theta, y_{1:T+S}) p(x_{T+S} | x_T, \theta, y_{1:T+S})$$
-
 \begin{itemize}
-\item Proper updates with filter/smooth
-\item Secondary approximation - no smoothing
-\item Straight up particle filter on old approximation
-\item PF on old MCMC
-\item Simulation Results
+\item We have an approximation of $q(\theta, x_{T+1} | y_{1:T})$ which is a great start.
+But if we observe data up to time $T+S$ I'm interested in incorporating this to update the approximation to $q(\theta, x_{T+S+1} | y_{1:T+S})$.
+\item Note: The approxmation going up to state $x_T, x_{T+1}, x_{T+S}, x_{T+S+H}$ doesn't really matter.
+An approximation at one point in time allows us to obtain an approximation for a different time due to the linear/gaussian transition equation.
+The original approximation could easily be modified to $q(\theta, x_{T+S+1} | y_{1:T})$, but what I think is important is conditioning this on $y_{T+1:T+S}$.
+\item The form of the joint distribution, $p(y_{1:T+S}, \theta, x_{T+S+1})$ should follow the same decomposition as before,
+$$p(y_{1:T+S}, \theta, x_{T+S+1}) = p(\theta) p(y_{1:T+S} | \theta) \int_{X_{T+S}} p(X_{T+S+1} | \theta, x_{T+S})p(x_{T+S} | \theta, y_{1:T+S})dX_{T+S},$$
+using a particle filter to estimate the intractable distributions.
+\item Key step is using information in the approximation of $q(\theta, x_T | y_{1:T})$ to help form $q(\theta, x_{T+S} | y_{1:T+S})$.
+\item We have access to the approximation of the conditional distribution, $q(x_T | \theta, y_{1:T})$ by integration of VB fit.
+\item Start the Particle Filter with $x_T$, using VB approximation $q(x_T | \theta, y_{1:T})$ as a proposal distribution for $x_T$, instead of the prior as is commonly used.
+\item The importance sampling weights should be easily obtainable using this new proposal.
+\item Then the filter needs to be run forward for $S$ steps.
+\item Using a VB approximation as an importance sampling proposal distribution is a new contribution. Hopefully it'll work well.
 \end{itemize}

### 2017-07-31 Tran Nott Kohn/importanceSampling.tex
logscore comparision

\documentclass[12pt,a4paper]{article}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage{amsthm}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\setcounter{MaxMatrixCols}{30}
\providecommand{\U}[1]{\protect\rule{.1in}{.1in}}
\newtheorem{theorem}{Theorem}
\newtheorem{acknowledgement}[theorem]{Acknowledgement}
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem{case}[theorem]{Case}
\newtheorem{claim}[theorem]{Claim}
\newtheorem{conclusion}[theorem]{Conclusion}
\newtheorem{condition}[theorem]{Condition}
\newtheorem{conjecture}[theorem]{Conjecture}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{criterion}[theorem]{Criterion}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\newtheorem{exercise}[theorem]{Exercise}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{problem}[theorem]{Problem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{solution}[theorem]{Solution}
\newtheorem{summary}[theorem]{Summary}
\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\setlength\parindent{0pt}
\begin{document}

\section{Updating the VB Approximation}

An issue with using the Particle Filter from $T+1$ to $T+S$ is that it only gives you an estimate of $p(y_{T+1:T+S} | \theta)$, and not $p(y_{1:T+S} | \theta)$. This could be an issue as the VB algorithm requires the distribution
\begin{equation}
\label{VBPF}
\ln(p(y_{1:T+S}, \theta, x_{T+S+1})) = \ln(p(\theta)) + \ln(p(y_{1:T+S} | \theta)) + \ln(p(x_{T+S+1} | y_{1:T+S}, \theta)).
\end{equation}
as part of the ELBO,
\begin{equation}
\label{ELBO}
\mathcal{L}(\boldsymbol{\lambda}) = \int q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T+S+1} | y_{1:T+S}) \ln \left( \frac{p(y_{1:T+S},\boldsymbol{\theta}, x_{T+S+1})}{q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T+S+1} | y_{1:T+S})} \right) d\boldsymbol{\theta}dx_{1:T}.
\end{equation}
This issue should be avoidable but requires a slightly different derivation of the ELBO. Note that
\begin{align}
\ln(p(y_{T+1:T+S} | y_{1:T})) &= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+1} | y_{1:T+S}) \ln(p(y_{T+1:T+S} | y_{1:T})) d_{\theta} d_{x_{T+S+1}} \nonumber \\
&= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{p(y_{T+1:T+S}, \theta, x_{T+1} | y_{1:T})}{p(x_{T+1}, \theta | y_{1:T+S})} \right) d_{\theta} d_{x_{T+S+1}}\nonumber \\
&= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \nonumber \\
&\times \ln \left( \frac{p(y_{T+1:T+S}, \theta, x_{T+1} | y_{1:T})q(\theta, x_{T+S+1}| y_{1:T+S})}
{p(x_{T+1}, \theta | y_{1:T+S})q(\theta, x_{T+S+1}| y_{1:T+S}) } \right) d_{\theta} d_{x_{T+1}}\nonumber \\
&= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{p(y_{T+1:T+S}, \theta, x_{T+1} | y_{1:T})}{q(\theta, x_{T+S+1}| y_{1:T+S})} \right) d_{\theta} d_{x_{T+1}} \label{ELBOv2}\\
&+  \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{q(\theta, x_{T+S+1}| y_{1:T+S})}{p(\theta, x_{T+1} | y_{1:T+S})} \right) d_{\theta} d_{x_{T+1}} \label{KL} \\
&= \mathcal{L}_2(\boldsymbol{\lambda}) + KL[q(\theta, x_{T+S+1} | y_{1:T+S}) || p(\theta, x_{T+S+1} | y_{1:T+S})]. \nonumber
\end{align}
As (\ref{KL}) is the Kullback-Leibler divergence from $q$ to $p$, it is minimised by maximising either (\ref{ELBO}) or (\ref{ELBOv2}), ie. maximising an ELBO that contains $\ln(p(y_{T+1:T+S}, \theta, x_{T+S+1} | y_{1:T}))$ instead of  $\ln(p(y_{1:T+S}, \theta, x_{T+S+1}))$ is still valid.
$\mathcal{L}_2(\boldsymbol{\lambda})$, given by (\ref{ELBOv2}), contains
\begin{equation}
\label{VBPF2}
\ln(p(y_{T+1:T+S}, \theta, x_{T+S+1} | y_{1:T})) = \ln(p(\theta | y_{1:T} )) + \ln(p(y_{T+1:T+S} | \theta, y_{1:T})) + \ln(p(x_{T+S+1} | y_{1:T+S}, \theta)).
\end{equation}
This introduces a few problems,
\begin{enumerate}
\item We do not have a parametric form of the marginal posterior distribution $p(\theta | y_{1:T})$. We do, however, have a parametric approximation available from the earlier VB: $q(\theta | y_{1:T})$. This can be feasibly used, but we would have to be comfortable with changing the model specification.
\item The particle filter estimates must be conditioned on $y_{1:T}$.
\end{enumerate}

\section{VB Particle Filtering}

We want samples from distribution $p(x_{T+1:T+S} | \theta, y_{1:T+S})$, which we decompose as
\begin{equation}
\label{posterior}
p(x_{T+1:T+S} | \theta, y_{1:T+S}) = p(x_{T+1} | y_{1:T}, \theta) p(y_{T+1} | x_{T+1}, \theta) \prod_{t=T+2}^{T+S} \left( p(y_t | x_t, \theta) p (x_t | x_{t-1}, \theta) \right)
\end{equation}
However, this is unavailable so instead we draw particles $x_{T+1:T+S}^{(1:N)}$ from a proposal distribution $f(x_{T+1:T+S})$, and then use importance sampling weights to correct the approximation. The weights for particle vector $i$ are given by
\begin{equation} 
\label{weights}
w^{(i)} \propto \frac{p(x_{T+1:T+S}^{(i)} | \theta, y_{1:T+S}, \theta)}{f(x_{T+1:T+S}^{(i)})}.
\end{equation}

A particle filter operates by positing a proposal distribution that factorises according to $f(x_{T+1:T+S}) =\prod_{t=T+1}^{T+S} f(x_{t})$, and drawing samples and calculating the weights for each time period sequentially. In this case, the weights for particles $x_{T+1}^{(i)}$ from the proposal distribution $f(x_{T+1})$ are given by
\begin{equation}
\label{weights2}
w_{T+1}^{(i)} \propto \frac{p(x_{T+1}^{(i)} | y_{1:T}, \theta) p(y_{T+1} | x_{T+1}^{(i)}, \theta)}{f(x_{T+1}^{(i)})}
\end{equation}
and weights at a future time $t$ are given by the recursion
\begin{equation}
\label{weights3}
w_t^{(i)} \propto w_{t-1}^{(i)} \frac{p(y_t | x_t^{(i)}, \theta) p (x_t^{(i)}| x_{t-1}^{(i)}, \theta)}{f(x_{t}^{(i)})}.
\end{equation}

Equation (\ref{weights2}) is problematic, we cannot easily evaluate the distribution $p(x_{T+1} | y_{1:T}, \theta)$ without running the filter over the entire dataset. As before, we can approximate this with the distribution $q(x_{T+1} | \theta, y_{1:T})$. If this is satisfactory, choose $f(x_{T+1})$ to be $q(x_{T+1} | \theta, y_{1:T})$ and each other $f(x_{t})$ to be $p(x_t | x_{t-1}, \theta)$. If the particles are resampled at each step, the weights reduce to 
\begin{equation}
\label{weights6}
w_t^i \propto p(y_t | x_t^i, \theta)
\end{equation}
for all $t$.

\section{Discussion}

Trying to update a particle filter based VB approximation introduces numerous problems, each of which seems to be solvable if we are willing to use the existing approximation of $q(\theta, x_{T+1} | y_{1:T})$ in place of the true posterior $p(\theta, x_{T+1} | y_{1:T})$. This implicitly changes the model prior used in an extremely informative way and not particularily desirable way - the prior must have the exact form that results in this particular multivariate normal posterior.
If we accept the original VB approximation as good enough then maybe this won't be such a big deal; $x_{T+S+1}$ is not included in the above substitution though there is likely to be some kind of indirect impact.

\end{document}

### 2017-07-31 Tran Nott Kohn/ADPFUpdater.cpp

@@ -1,0 +1,520 @@
+// [[Rcpp::depends(rstan)]]
+// [[Rcpp::depends(RcppEigen)]]
+// [[Rcpp::plugins(cpp11)]]
+// [[Rcpp::depends(RcppArmadillo)]]
+
+#include <RcppArmadillo.h>
+#include <stan/math.hpp>
+#include <vector>
+#include <cmath>
+#include <math.h>
+#include <Eigen/Dense>
+#include <RcppEigen.h>
+#include <Rcpp.h>
+#include <boost/math/distributions.hpp> 
+#include <cstdlib> 
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+
+using namespace std;
+using Eigen::Matrix;
+using Eigen::Dynamic;
+using Eigen::MatrixXd; 
+using Eigen::Map;
+using namespace arma;
+
+// evaluate and automatially differentiate log(p(y_1:T, x_T+1, theta)) wrt theta and x_T+1
+struct logP {
+  const MatrixXd& y; // data
+  const MatrixXd& noise; // standard normal matrix for the particle filter transitions
+  const int& obs; // T is taken for the variable type, so obs = number of elements in y
+  const int& N; // Number of particles
+  const MatrixXd& priorMean;
+  const MatrixXd& sigInv;
+  const MatrixXd& uDiag;
+  const double& conMean; //q(X_T+1 | theta) mean and variance
+  const double& conVar;
+  logP(const MatrixXd& yIn, const MatrixXd& noiseIn, const int& obsIn, const int& nIn, 
+       const MatrixXd& meanIn, const MatrixXd& sigInvIn, const MatrixXd& uDiagIn, const double& conMeanIn, const double& conVarIn) : // constructor
+    y(yIn), noise(noiseIn), obs(obsIn), N(nIn), priorMean(meanIn), sigInv(sigInvIn), uDiag(uDiagIn), conMean(conMeanIn), conVar(conVarIn) {}
+  template <typename T> // T will become stan's var that can be automatically differentiated - any variable differentiated in the chain rule for d logp / d theta must be type T
+  T operator ()(const Matrix<T, Dynamic, 1>& theta) // derivative is with respect to theta (with xT+1 included), so theta matrix is the operator () input.
[…]

### 2017-07-31 Tran Nott Kohn/SVM.R

@@ -7,7 +7,7 @@ sourceCpp('ADPF.cpp')
 sourceCpp('SVM_MCMC.cpp')
 
 sigSq = 0.02
-phi = 0.95
+phi = 0.9
 mu = -0.5
 T = 500
 S = 50
@@ -18,22 +18,23 @@ for(t in 1:(T+S)){
   if(t == 1){
     x[t] = rnorm(1, mu, sqrt(sigSq/(1-phi^2)))
   } else {
-    x[t] = mu + phi*(x[t-1]-mu) + rnorm(1, 0, sqrt(sigSq))
+    x[t] = mu + phi*(x[t-1]-mu) + sqrt(sigSq) * rnorm(1)
   }
   y[t] = exp(x[t]/2) * rnorm(1)
 }
 
 
+
 yStar = log(y^2)
-MCMC = SVM_MCMC(yStar, 20000, 0.01)
+MCMC = SVM_MCMC(yStar[1:T], 50000, 0.05)
 
 
-yM = matrix(y, T+S)
-mean = c(-4, 0, 0.5, 0)
-var = diag(c(0.1, 0.15, 0.15, 0.15))
+yM = matrix(y[1:T], T)
+mean = c(-4.4, 0, 0.86, mu)
+var = diag(c(0.7, sqrt(10), 0.101, sqrt(sigSq/(1-phi^2))))
 
 lambda = cbind(mean, var)
-fitFull = VBIL_PF(yM, lambda, S=10, alpha=0.1, maxIter=5000, threshold=0.05, thresholdIS=0.9)
+fitFull = VBIL_PF(yM, lambda, S=10, alpha=0.25, maxIter=5000, threshold=0.05, thresholdIS=0.9)
 
 
 lambda = cbind(mean, var)
@@ -81,14 +82,14 @@ mu = -0.5
 T = 500
 
 set.seed(11)
-reps = 20
-resultsvb = data.frame()
+reps = 100
+results = data.frame()
 for(i in 1:reps){
   
   x0 = rnorm(1, mu, sqrt(sigSq/(1-phi^2)))
-  x = rep(0, T)
-  y = rep(0, T)
-  for(t in 1:T){
+  x = rep(0, T+1)
+  y = rep(0, T+1)
+  for(t in 1:(T+1)){
     if(t == 1){
       x[t] = mu + phi*(x0-mu) + rnorm(1, 0, sqrt(sigSq))
     } else {
[…]

### 2017-07-31 Tran Nott Kohn/fullSVM.cpp

@@ -1,0 +1,404 @@
+// [[Rcpp::depends(rstan)]]
+// [[Rcpp::depends(RcppEigen)]]
+// [[Rcpp::plugins(cpp11)]]
+// [[Rcpp::depends(RcppArmadillo)]]
+
+#include <RcppArmadillo.h>
+#include <stan/math.hpp>
+#include <vector>
+#include <cmath>
+#include <math.h>
+#include <Eigen/Dense>
+#include <RcppEigen.h>
+#include <Rcpp.h>
+#include <boost/math/distributions.hpp> 
+#include <cstdlib> 
+#include <iostream>
+#include <iomanip>
+#include <fstream>
+
+using namespace std;
+using Eigen::Matrix;
+using Eigen::Dynamic;
+using Eigen::MatrixXd; 
+using Eigen::Map;
+using namespace arma;
+
+// evaluate and automatially differentiate log(p(y_1:T, x_T+1, theta)) wrt theta and x_T+1
+struct logP {
+  const MatrixXd& y; // data
+  const int& obs; // T is taken for the variable type, so obs = number of elements in y
+  logP(const MatrixXd& yIn, const int& obsIn) : // constructor
+    y(yIn), obs(obsIn) {}
+  template <typename T> // T will become stan's var that can be automatically differentiated - any variable differentiated in the chain rule for d logp / d theta must be type T
+  T operator ()(const Matrix<T, Dynamic, 1>& theta) // derivative is with respect to theta (with xT+1 included), so theta matrix is the operator () input.
+    const{
+    using std::log; using std::pow; using std::exp; using std::sqrt; // explicitly declare mathematical operators used so stan knows to differentiate these.
+    
+    // Get elements of theta
+    T sigSq = theta(0);
+    T mu = theta(1);
+    T tau = theta(2);
+    T phi = 2*tau - 1;
+    Matrix<T, Dynamic, 1> x = theta.bottomRows(obs+2);
+    // Evaluate log(p(theta)) + log(p(x_0 | theta))
+    T prior = -pow(mu, 2) / 20  +  19 * log(tau)  + 0.5 * log(1-tau) - 3.5 * log(sigSq)  -  0.025 / sigSq;
+    T states = - 0.5 * log(sigSq)  +  0.5 * log(1 - pow(phi, 2))  -  pow(theta(3) - mu, 2) * (1 - pow(phi, 2)) / (2*sigSq);
[…]

### 2017-07-31 Tran Nott Kohn/numerical derivatives.R

@@ -1,0 +1,91 @@
+double IGlogdens (double unif, double par1, double par2){
+  boost::math::inverse_gamma_distribution<> dist(exp(par1), exp(par2));
+  double x = quantile(dist, unif);
+  par1 = exp(par1);
+  par2 = exp(par2);
+  double density = par1 * log(par2) - lgamma(par1) - (par1+1)*log(x) - par2 / x;
+  return density;
+}
+
+double Blogdens (double unif, double par1, double par2){
+  boost::math::beta_distribution<> dist(exp(par1), exp(par2));
+  double x = quantile(dist, unif);
+  par1 = exp(par1);
+  par2 = exp(par2);
+  double density = lgamma(par1 + par2) - lgamma(par1) - lgamma(par2) + (par1-1)*log(x) + (par2-1)*log(1-x);
+  return density;
+}
+
+double Nlogdens(double unif, double par1, double par2){
+  boost::math::normal_distribution<> dist(par1, exp(par2));
+  double x = quantile(dist, unif);
+  par2 = exp(par2);
+  double density = - 0.5 * log(2 * 3.14159) -  log(par2)  - pow(x - par1, 2) / (2 * pow(par2, 2));
+  return density;
+}
+
+Matrix<double, Dynamic, 1> QDeriv(rowvec unif, MatrixXd lambda){
+  double h = 0.00001;
+  Matrix<double, Dynamic, 1> output(8);
+  output(0) = (IGlogdens(unif(0), lambda(0)+h, lambda(1)) - IGlogdens(unif(0), lambda(0), lambda(1)))/h;
+  output(1) = (IGlogdens(unif(0), lambda(0), lambda(1)+h) - IGlogdens(unif(0), lambda(0), lambda(1)))/h;
+  output(3) = (Blogdens(unif(1), lambda(2)+h, lambda(3)) - Blogdens(unif(1), lambda(2), lambda(3)))/h;
+  output(4) = (Blogdens(unif(1), lambda(2), lambda(3)+h) - Blogdens(unif(1), lambda(2), lambda(3)))/h;
+  output(5) = (Nlogdens(unif(2), lambda(4)+h, lambda(5)) - Nlogdens(unif(2), lambda(4), lambda(5)))/h;
+  output(6) = (Nlogdens(unif(2), lambda(4), lambda(5)+h) - Nlogdens(unif(2), lambda(4), lambda(5)))/h;
[…]

### 2017-08-01 Tran Nott Kohn/importanceSampling.tex
misc

@@
 
-\section{Updating the VB Approximation}
+\section{Updating with the Bootstrap Particle Filter}
 
@@
 
+\section{Using the Data-Driven Particle Filter}
+
+The measurement equation for the SVM is
+\begin{equation}
+\label{measure}
+y_t = h(x_t, \eta_t) =\exp\{0.5 x_t\} \eta_t,
+\end{equation}
+leading to particles $x_t$ being generated according to
+\begin{equation}
+\label{measureInv}
+x_t = 2 \log(y_t / \eta_t).
+\end{equation}
+Weights are then calculated according to
+\begin{align}
+\label{DPFweight}
+w_t^{(j)} &= \pi_t^{(j)} \left| \frac{\delta h(x_t^{(j)}, \eta_t^{(j)})}{\delta x_t^{(j)}} \right|^{-1}_{\eta_t = \eta_t^{(j)}, x_t = x_t^{(j)}} p\left(x_t^{(j)} | x_{t-1}^{(j)}\right) \nonumber \\
+&= \pi_t^{(j)} \left| 0.5 \exp\{0.5 x_t^{(j)}\} \eta_t^{(j)} \right|^{-1} p \left(x_t^{(j)} | x_{t-1}^{(j)} \right)
+\end{align}
+
 \section{Discussion}
@@
 
+\section{General Loss Distributions}
+
+To implement Bissiri et al. 2016 we need to write a loss function of the past data in terms of the parameters of interest, $x_{T+1}$ and  $\theta$.
+
+Consider
+\begin{equation}
+\label{loss}
+l(y | \theta, x_{T+1}) = -\sum_{t=1}^T \log(y_t | x_t),
+\end{equation}
+where
+\begin{equation}
+\label{lossyt}
+\log(y_t | x_t) = c + 0.5x_t + \frac{y_t^2}{2 \exp(x_t)}.
+\end{equation}
+The recursion
+\begin{equation}
+\label{xRecursion}
+x_t = \mu + \frac{x_{t+1} - \mu - \epsilon_{t+1}}{\phi}
+\end{equation}
+allows each $x_t$ to be recovered given a sequence of $iid$ standard normal random variables $\epsilon_t, t = 1, \dots, T+1$.
+
 \end{document}

### 2017-08-01 Tran Nott Kohn/ADPF.cpp

@@ -27,10 +27,10 @@ using namespace arma;
 // evaluate and automatially differentiate log(p(y_1:T, x_T+1, theta)) wrt theta and x_T+1
 struct logP {
   const MatrixXd& y; // data
-  const MatrixXd& noise; // standard normal matrix for the particle filter transitions
+  const mat& noise; // standard normal matrix for the particle filter transitions
   const int& obs; // T is taken for the variable type, so obs = number of elements in y
   const int& N; // Number of particles
-  logP(const MatrixXd& yIn, const MatrixXd& noiseIn, const int& obsIn, const int& nIn) : // constructor
+  logP(const MatrixXd& yIn, const mat& noiseIn, const int& obsIn, const int& nIn) : // constructor
     y(yIn), noise(noiseIn), obs(obsIn), N(nIn) {}
   template <typename T> // T will become stan's var that can be automatically differentiated - any variable differentiated in the chain rule for d logp / d theta must be type T
   T operator ()(const Matrix<T, Dynamic, 1>& theta) // derivative is with respect to theta (with xT+1 included), so theta matrix is the operator () input.
@@ -379,12 +379,8 @@ Rcpp::List VBIL_PF (Rcpp::NumericMatrix yIn, Rcpp::NumericMatrix lambdaIn, int S
         if(theta(0, s) > 0.3){
           theta(0, s) = 0.3;
         }
-        // generate standard normal noise for the particle filter, convert to Eigen format
-        Rcpp::NumericMatrix noiseRcpp(100, T+1);
-        for(int t = 0; t < T+1; ++t){
-          noiseRcpp(Rcpp::_, t) = Rcpp::rnorm(100);
-        }
-        Map<MatrixXd> noise(Rcpp::as<Map<MatrixXd> >(noiseRcpp));
+        // generate standard normal noise for the particle filter
+        mat noise(N, T+1, fill::randn);
         // take derivative of log joint
         logP p(y, noise, T, N);
         stan::math::set_zero_all_adjoints();

### 2017-08-04 Tran Nott Kohn/importanceSampling.tex
various

--- 2017-07-31 Tran Nott Kohn/importanceSampling.tex
@@
 \setlength{\textheight}{8.25in}
+\numberwithin{equation}{section}
 \setlength\parindent{0pt}
@@
 
-\section{Updating the VB Approximation}
+\section{Introduction}
 
-An issue with using the Particle Filter from $T+1$ to $T+S$ is that it only gives you an estimate of $p(y_{T+1:T+S} | \theta)$, and not $p(y_{1:T+S} | \theta)$.
-This could be an issue as the VB algorithm requires the distribution
+Consider a high frequency time series with observed values $y_{1}, \dots, y_{T+S}$, collectively denoted by $y_{1:T+S}$, with a forecaster interested in the forecast distribution of the future value $y_{T+S+h}$, where $S \geq 1$.
+In many applications forecasting uncertainty can be reduced by conditioning on additional data, so knowledge of the distribution of $p(y_{T+S+h} | y_{1:T+S})$ is preferable to that of $p(y_{T+S+h} | y_{1:T})$.
+
+Many time series models of interest contain a set of global parameters $\boldsymbol{\theta}$ and observation specific latent variables $x_{0:T}$, such that $x_t$ is assumed to follow a Markovian structure and $y_t$ is conditionally independent of $y_{1:t-1}$, given $\boldsymbol{\theta}$ and $x_t$.
+One of the most common computational techniques for models with latent variables is Markov Chain Monte Carlo (MCMC), which involves iterative sampling from a Markov Chain designed to converge to the true posterior distribution, however these approaches can have slow convergence rates.
+In this paper it is assumed that an MCMC algorithm will require a time period of length $S$ to converge, so at time $T+S$ the posterior distribution of $\boldsymbol{\theta}$  and $x_{0:T+S}$ can be conditioned only on $y_{1:T}$ and hence only $p(y_{1:T+S+h} | y_{1:T})$ is available.
+
+The additional data $y_{T+1:T+S}$ can be included in the forecast by using filtering techniques such as the Kalman filter or particle filtering, however this will only update knowledge on $x_{0:T+S}$ but not $\boldsymbol{\theta}$.
+As an alternative to this method, this paper will instead approximate the posterior distribution $p(\boldsymbol{\theta}, x_{T} | y_{1:T})$ with $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T} | y_{1:T})$ and the updated posterior distribution $p(\boldsymbol{\theta}, x_{T+S} | y_{1:T+S})$ with $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T+S} | y_{1:T+S})$.
+The parameters of this approximation, denoted by $\boldsymbol{\lambda}$, are optimised by gradient descent to minimise the Kullback-Leibler divergence from $q$ to $p$, a technique known as Variational Inference, which has seen wide use in the literature (references).
+
+Variational Bayes involves an implicit trade-off: reducing the forecast horizon of $y_{T+S+h}$ from $S+h$ to $h$ reduces the uncertainty of the forecast, but the approximation will increase the statistical error.
+This paper additionally explores that trade-off for a Stochastic Volatility Model, a common latent state time series model.
+
+\begin{itemize}
+\item Some motivation about forecasting and updates, can probably be copied from the MIVB file
+\item Not focusing on MCMC sample
+\item VB estimate at time T, VB update to time T+S
+\item End goal: An approximation $q(\theta, x_{T+1})$ to make $q(y_{T+1} | y_{1:T})$, and then update to $q(y_{T+S+1} | y_{1:T+S})$ efficiently.
+\end{itemize}
+
+\section{MCMC and Particle Filtering}
+\begin{itemize}
+\item MCMC background
+\item Stochastic Volatility Model
+\item Kim Shepherd Chib MCMC / Particle Filtering / PMCMC for more complicated models
+\item BPF, APF
+\end{itemize}
+
+\citep{Arulampalam2002}
+
+\section{Variational Inference}
+
+Variational Inference posits a divergence function between the true posterior distribution $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ and some approximating distribution $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$, choosing the parameters $\boldsymbol{\lambda}$ for a given functional form $q$ that minimises the divergence function.
+
+This paper will follow the traditional approach, where the divergence function is the Kullback-Leibler (KL) divergence \citep{Kullback1951} from $q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ to the true posterior $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$ The KL divergence is defined by
+\begin{align}
+\label{KL-def}
+KL[q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})] &= \nonumber \\
+\int q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T}) &\ln \left( \frac{q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})}{p(\boldsymbol{\theta}, x_{0:T} |y_{1:T})}\right) d\boldsymbol{\theta} dx_{0:T}
+\end{align}
+and can alternatively be expressed as
 \begin{equation}
-\label{VBPF}
-\ln(p(y_{1:T+S}, \theta, x_{T+S+1})) = \ln(p(\theta)) + \ln(p(y_{1:T+S} | \theta)) + \ln(p(x_{T+S+1} | y_{1:T+S}, \theta)).
+\label{KL-ELBO}
+KL[q(\boldsymbol{\theta}, x_{0:T} | \boldsymbol{\lambda})\hspace{.1cm}||\hspace{.1cm}p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})] = \ln(p(y_{1:T})) - \mathcal{L}(\boldsymbol{\lambda})
 \end{equation}
-as part of the ELBO,
+where $\mathcal{L}(\boldsymbol{\lambda})$ is referred to as the Evidence Lower Bound (ELBO), as it provides a lower bound on the unknown constant $\ln(p(y_{1:T}))$.
+$\mathcal{L}(\boldsymbol{\lambda})$ is defined by
 \begin{equation}
 \label{ELBO}
-\mathcal{L}(\boldsymbol{\lambda}) = \int q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T+S+1} | y_{1:T+S}) \ln \left( \frac{p(y_{1:T+S},\boldsymbol{\theta}, x_{T+S+1})}{q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{T+S+1} | y_{1:T+S})} \right) d\boldsymbol{\theta}dx_{1:T}.
+\mathcal{L}(\boldsymbol{\lambda}) = \int q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T}) \ln \left( \frac{p(y_{1:T},\boldsymbol{\theta}, x_{0:T})}{q_{\boldsymbol{\lambda}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})} \right) d\boldsymbol{\theta}dx_{1:T},
 \end{equation}
-This issue should be avoidable but requires a slightly different derivation of the ELBO. Note that
+and as $\ln(p(y_{1:T}))$ is constant with respect to $\boldsymbol{\lambda}$, maximising (\ref{ELBO}) with respect to $\boldsymbol{\lambda}$ is equivalent to minimising (\ref{KL-def}).
+Maximising (\ref{ELBO}) is more convenient that minimising (\ref{KL-def}) as it is a function of the known joint distribution $p(y_{1:T}, \boldsymbol{\theta}, x_{0:T})$ instead of the unknown posterior $p(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$.
+
+\subsection{ELBO Optimisation}
+Equation (\ref{ELBO}) can be maximised using a gradient ascent approach, where the following update step is iteratively applied until (\ref{ELBO}) converges within some pre-specified tolerance:
+\begin{equation}
+\label{GradAscent}
+\boldsymbol{\lambda}^{(m+1)} = \boldsymbol{\lambda}^{(m)} + \rho^{(m)} \frac{\delta}{\delta\boldsymbol{\lambda}} \mathcal{L}(\boldsymbol{\lambda^{(m)}}),
+\end{equation}
+where the derivative is evaluated at $\boldsymbol{\lambda}^{(m)}$.
+This update requires some initial values $\boldsymbol{\lambda}^{(0)}$ and a sequence $\rho^{(m)}, m = 1, 2, \dots$ known as the learning rate.
+If $\rho^{(m)}$ is chosen to satisfy the following conditions the algorithm is guaranteed to converge to a local maximum \citep{Robbins1951}.
 \begin{align}
-\ln(p(y_{T+1:T+S} | y_{1:T})) &= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+1} | y_{1:T+S}) \ln(p(y_{T+1:T+S} | y_{1:T})) d_{\theta} d_{x_{T+S+1}} \nonumber \\
-&= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{p(y_{T+1:T+S}, \theta, x_{T+1} | y_{1:T})}{p(x_{T+1}, \theta | y_{1:T+S})} \right) d_{\theta} d_{x_{T+S+1}}\nonumber \\
-&= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \nonumber \\
-&\times \ln \left( \frac{p(y_{T+1:T+S}, \theta, x_{T+1} | y_{1:T})q(\theta, x_{T+S+1}| y_{1:T+S})}
-{p(x_{T+1}, \theta | y_{1:T+S})q(\theta, x_{T+S+1}| y_{1:T+S}) } \right) d_{\theta} d_{x_{T+1}}\nonumber \\
-&= \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{p(y_{T+1:T+S}, \theta, x_{T+1} | y_{1:T})}{q(\theta, x_{T+S+1}| y_{1:T+S})} \right) d_{\theta} d_{x_{T+1}} \label{ELBOv2}\\
-&+  \int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{q(\theta, x_{T+S+1}| y_{1:T+S})}{p(\theta, x_{T+1} | y_{1:T+S})} \right) d_{\theta} d_{x_{T+1}} \label{KL} \\
-&= \mathcal{L}_2(\boldsymbol{\lambda}) + KL[q(\theta, x_{T+S+1} | y_{1:T+S}) || p(\theta, x_{T+S+1} | y_{1:T+S})]. \nonumber
+&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
+&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
 \end{align}
-As (\ref{KL}) is the Kullback-Leibler divergence from $q$ to $p$, it is minimised by maximising either (\ref{ELBO}) or (\ref{ELBOv2}), ie.
-maximising an ELBO that contains $\ln(p(y_{T+1:T+S}, \theta, x_{T+S+1} | y_{1:T}))$ instead of  $\ln(p(y_{1:T+S}, \theta, x_{T+S+1}))$ is still valid.
-$\mathcal{L}_2(\boldsymbol{\lambda})$, given by (\ref{ELBOv2}), contains
+
+This paper uses Adam \citep{Kingma2015b} to generate the sequence $\rho^{(m)}$.
+
+\citet{Ranganath2014} showed that a Monte Carlo estimate of the derivative of the ELBO can be given by
 \begin{equation}
-\label{VBPF2}
-\ln(p(y_{T+1:T+S}, \theta, x_{T+S+1} | y_{1:T})) = \ln(p(\theta | y_{1:T} )) + \ln(p(y_{T+1:T+S} | \theta, y_{1:T})) + \ln(p(x_{T+S+1} | y_{1:T+S}, \theta)).
+\label{ScoreDeriv}
+\frac{\delta}{\delta\boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda^{(m)}}) \approx \frac{1}{N}\sum_{i=1}^{N} \frac{\delta}{\delta\boldsymbol{\lambda}} [\ln(q_{\boldsymbol{\lambda}^{(m)}}(\boldsymbol{\theta_i}, x_{0:T, i} | y_{1:T})] \ln \left(\frac{p(y_{1:T}, \boldsymbol{\theta}_i, x_{1:T, i})}{q_{\boldsymbol{\lambda}^{(m)}}(\boldsymbol{\theta_i}, x_{0:T, i} | y_{1:T})} \right)
 \end{equation}
-This introduces a few problems,
+where $i = 1, \dots, N$ indicates independent simulations from $q_{\boldsymbol{\lambda}^{(m)}}(\boldsymbol{\theta}, x_{0:T} | y_{1:T})$. 
+The terms in the sum in (\ref{ScoreDeriv}) can have large variances, and in practice a large value of $N$ is required to ensure a precise estimate of the gradient of the ELBO is obtained, slowing computation.
+
+The variance can be reduced by the reparameterisation trick of \citet{Kingma2014}, introducing a random vector $\boldsymbol{\epsilon}$ with a distribution $q(\boldsymbol{\epsilon})$ that contains no free parameters, and a differentiable transform $f$ such that
+\begin{equation}
+\label{Reparam}
+f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}) = \{\boldsymbol{\theta}, x_{1:T}\}.
+\end{equation}
+Kingma and Welling (2014) show that an $f$ exists to transform $\boldsymbol{\epsilon}$ to any continuous random variable, with examples including a location-scale transform of a standard normal $\epsilon$ and an inverse-CDF transform of a uniform $\epsilon$.
+In this case, (\ref{ELBO}) becomes
+\begin{equation}
+\label{ReparamELBO}
+\mathcal{L}(\boldsymbol{\lambda}) = \int q(\boldsymbol{\epsilon}) \ln \left( \frac{p(y_{1:T},f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}) |\det J(\boldsymbol{\epsilon}, \boldsymbol{\lambda})|}{q(\boldsymbol{\epsilon})} \right) d\boldsymbol{\epsilon},
+\end{equation}
+where $J(\boldsymbol{\epsilon}, \boldsymbol{\lambda})$ is the Jacobian matrix of the transformation $f$. The derivative of (\ref{ReparamELBO}) can be estimated by 
+\begin{equation}
+\label{ReparamDeriv}
+\frac{\delta}{\delta \boldsymbol{\lambda}}\mathcal{L}(\boldsymbol{\lambda}^{(m)}) \approx \frac{1}{M}\sum_{i=1}^{M}
+\frac{\delta f(\boldsymbol{\epsilon}_i, \boldsymbol{\lambda}^{(m)})} {\delta \boldsymbol{\lambda}}
+\frac{\delta \ln (p(y_{1:T}, f(\boldsymbol{\epsilon}_i, \boldsymbol{\lambda}^{(m)})))}{\delta f(\boldsymbol{\epsilon}_i, \boldsymbol{\lambda}^{(m)})} + \frac{\delta  \ln(|\det J(\boldsymbol{\lambda}^{(m)}, \boldsymbol{\epsilon}_i)|)}{\delta \boldsymbol{\lambda}},
+\end{equation} 
+where simulations of $\boldsymbol{\theta}$ and $x_{1:T+S}$ are replaced by simulations of $\boldsymbol{\epsilon}$ from $q(\boldsymbol{\epsilon})$.
+The variance of this estimator is often orders of magnitude smaller than the estimator in (\ref{ScoreDeriv}) (citations), so $M$ can be set much lower than $N$.
+
+\subsection{Randomised Quasi Monte Carlo}
+\citep{Gunawan2017}
+\citep{Matousek1998}
+\citep{Bratley1988}
+\citep{Sobol1967}
+
+\subsection{Choice of $q$ distribution}
+
+\section{Dimensionality Reduction}
+
+ADVI simulation results
+
+As the latent variables are Markovian, once the distribution of $x_{T+1}$ is available, the distributions of each previous $x_t, t \leq T$ is irrelevant to forecasts of $y_{T+S+h}$ and it may be more convenient to construct an approximation for $p(\boldsymbol{\theta}, x_{T+1} | y_{1:T})$ than for $p(\boldsymbol{\theta}, x_{0:T+1} | y_{1:T})$, as they contain the same information about $y_{T+S+h}$ but the dimensionality of the posterior is reduced from $T+k+2$ to $k+1$, where $k$ is the number of elements in $\boldsymbol{\theta}$.
+This in turn allows for more expressive approximating distributions to be used, as the size of $\boldsymbol{\lambda}$ does not grow with $T$.
+
+In this case, using $f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}) = \{\boldsymbol{\theta}, x_{T+1}\}$,
+\begin{equation}
+\label{DimRedLogJoint}
+\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) = \ln\left(\int_{X_T} p(x_{T+1} | \theta, x_T)p(x_T | y_{1:T}, \boldsymbol{\theta})d_{X_T}\right) + \ln(p(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
+\end{equation}
+In many cases, the distributions $p(x_T | y_{1:T}, \boldsymbol{\theta})$ and $p(y_{1:T} | \boldsymbol{\theta})$ are intractable, but \citet{TranM2017} note that the particle filter estimators $\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})$ and $\hat{p}(y_{1:T} | \boldsymbol{\theta})$ are sufficient substitutes in what they refer to as Variational Bayes with Intractable Likelihood (VBIL).
+The particle filter estimation of the distribution $\hat{p}(x_T | y_{1:T}, \boldsymbol{\theta})$ is expressed as a discrete set of point masses $x_T^{(k)}$ and weights $\pi_T^{(k)}$, where $k = 1, \dots, N$ represent the particles of the particle filter.
+In this case, the integral in (\ref{DimRedLogJoint}) reduces to the sum
+\begin{equation}
+\label{ParticleFilterIntegral}
+\int_{x_T} p(x_{T+1} | \theta, x_T)p(x_T | y_{1:T}, \boldsymbol{\theta})d_{x_T} = \sum_{k=1}^N \pi_T^{(k)} p(x_{T+1} | \boldsymbol{\theta}, x_T^{(k)}),
+\end{equation}
+and hence
+\begin{equation}
+\label{PFLogJoint}
+\ln(p(y_{1:T}, f(\boldsymbol{\epsilon}, \boldsymbol{\lambda}))) \approx \ln \left(\sum_{k=1}^N \pi_T^{(k)} p(x_{T+1} | \boldsymbol{\theta}, x_T^{(k)})\right) + \ln(\hat{p}(y_{1:T} |\boldsymbol{\theta})) + \ln(p(\boldsymbol{\theta}))
+\end{equation}
+can be substituted into (\ref{ReparamDeriv}).
+This derivative of (\ref{PFLogJoint}) with respect to $\boldsymbol{\lambda}$ can be obtained with automatic differentiation tools such as those provided by Stan \citep{Carpenter2015}.
+
+\subsection{Importance Sampling}
+
+\citep{Sakaya2017}
+
+\section{Updating}
+
+\begin{itemize}
+\item We have an approximation of $q(\theta, x_{T+1} | y_{1:T})$ which can be used to forcast some future value $y_{T+S+h}$.
+However, observation of additional data $y_{T+1:T+S}$ should be able to improve this forecast.
+\item We aim to use the original distribution $q(x_{T+1}, \theta | y_{1:T})$ to build this new approximation, and only using $y_{T+1:T+S}$.
+\item Instead of maximising the standard ELBO,
+\begin{equation*}
+\int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{p(y_{1:T+S}, \theta, x_{T+S+1})}{q(\theta, x_{T+S+1}| y_{1:T+S})} \right) d_{\theta} d_{x_{T+S+1}},
+\end{equation*}
+we can instead maximise a slightly modified ELBO:
+\begin{equation*}
+\int_{\theta, x_{T+S+1}} q(\theta, x_{T+S+1}| y_{1:T+S}) \ln \left(\frac{p(y_{T+1:T+S}, \theta, x_{T+S+1} | y_{1:T})}{q(\theta, x_{T+S+1}| y_{1:T+S})} \right) d_{\theta} d_{x_{T+S+1}}.
+\end{equation*}
+\item The joint distribution, $p(y_{T+1:T+S}, \theta, x_{T+1} | y_{1:T})$ similarly to before, use
+\begin{align*}
+\ln(p(y_{T+1:T+S}, \theta, x_{T+S+1} | y_{1:T})) &\approx \ln(p(\theta | y_{1:T} )) + \ln(\hat{p}(y_{T+1:T+S} | \theta, y_{1:T}))\\
+&+ \ln\left(\int_{X_{T+S}} p(x_{T+S+1} | x_{T+S}, \theta) \hat{p}(X_{T+S} | y_{1:T+S}, \theta) d x_{T+S}\right),
+\end{align*}
+where $\hat{p}$ denotes a particle filter estimate.
+\item This introduces a few problems,
 \begin{enumerate}
 \item We do not have a parametric form of the marginal posterior distribution $p(\theta | y_{1:T})$.
-We do, however, have a parametric approximation available from the earlier VB: $q(\theta | y_{1:T})$.
-This can be feasibly used, but we would have to be comfortable with changing the model specification.
 \item The particle filter estimates must be conditioned on $y_{1:T}$.
 \end{enumerate}
+\item 1. can be solved by using $q(\theta | y_{1:T})$ in place of $p(\theta | y_{1:T})$. This implicitly changes the original model prior.
+\item 2.
+requires starting the particle filter with particles drawn from $p(x_{T+1} | y_{1:T}, \theta)$ for the bootstrap particle filter or $p(x_{T+1} | y_{1:T+1}, \theta)$ for the auxillary particle filter.
+\item This distribution is unavailable without re-using data $y_{1:T}$, but $q(x_{T+1} | y_{1:T}, \theta)$ can be substituted in.
+\item Then the filter needs to be run forward for just $S$ steps per iteration.
+\end{itemize}
 
-\section{VB Particle Filtering}
-
-We want samples from distribution $p(x_{T+1:T+S} | \theta, y_{1:T+S})$, which we decompose as
-\begin{equation}
-\label{posterior}
-p(x_{T+1:T+S} | \theta, y_{1:T+S}) = p(x_{T+1} | y_{1:T}, \theta) p(y_{T+1} | x_{T+1}, \theta) \prod_{t=T+2}^{T+S} \left( p(y_t | x_t, \theta) p (x_t | x_{t-1}, \theta) \right)
-\end{equation}
-However, this is unavailable so instead we draw particles $x_{T+1:T+S}^{(1:N)}$ from a proposal distribution $f(x_{T+1:T+S})$, and then use importance sampling weights to correct the approximation.
-The weights for particle vector $i$ are given by
-\begin{equation} 
-\label{weights}
-w^{(i)} \propto \frac{p(x_{T+1:T+S}^{(i)} | \theta, y_{1:T+S}, \theta)}{f(x_{T+1:T+S}^{(i)})}.
-\end{equation}
-
-A particle filter operates by positing a proposal distribution that factorises according to $f(x_{T+1:T+S}) =\prod_{t=T+1}^{T+S} f(x_{t})$, and drawing samples and calculating the weights for each time period sequentially.
-In this case, the weights for particles $x_{T+1}^{(i)}$ from the proposal distribution $f(x_{T+1})$ are given by
-\begin{equation}
-\label{weights2}
-w_{T+1}^{(i)} \propto \frac{p(x_{T+1}^{(i)} | y_{1:T}, \theta) p(y_{T+1} | x_{T+1}^{(i)}, \theta)}{f(x_{T+1}^{(i)})}
-\end{equation}
-and weights at a future time $t$ are given by the recursion
-\begin{equation}
-\label{weights3}
-w_t^{(i)} \propto w_{t-1}^{(i)} \frac{p(y_t | x_t^{(i)}, \theta) p (x_t^{(i)}| x_{t-1}^{(i)}, \theta)}{f(x_{t}^{(i)})}.
-\end{equation}
-
-Equation (\ref{weights2}) is problematic, we cannot easily evaluate the distribution $p(x_{T+1} | y_{1:T}, \theta)$ without running the filter over the entire dataset.
-As before, we can approximate this with the distribution $q(x_{T+1} | \theta, y_{1:T})$.
-If this is satisfactory, choose $f(x_{T+1})$ to be $q(x_{T+1} | \theta, y_{1:T})$ and each other $f(x_{t})$ to be $p(x_t | x_{t-1}, \theta)$.
-If the particles are resampled at each step, the weights reduce to
-\begin{equation}
-\label{weights6}
-w_t^i \propto p(y_t | x_t^i, \theta)
-\end{equation}
-for all $t$.
+\section{Empirical}
 
@@
 
-Trying to update a particle filter based VB approximation introduces numerous problems, each of which seems to be solvable if we are willing to use the existing approximation of $q(\theta, x_{T+1} | y_{1:T})$ in place of the true posterior $p(\theta, x_{T+1} | y_{1:T})$.
-This implicitly changes the model prior used in an extremely informative way and not particularily desirable way - the prior must have the exact form that results in this particular multivariate normal posterior.
-If we accept the original VB approximation as good enough then maybe this won't be such a big deal; $x_{T+S+1}$ is not included in the above substitution though there is likely to be some kind of indirect impact.
+\bibliographystyle{asa}
+\bibliography{references}
 

### 2017-08-04 Tran Nott Kohn/SVM.R

@@ -5,17 +5,18 @@ library(RcppEigen)
 library(rstan)
 sourceCpp('ADPF.cpp')
 sourceCpp('SVM_MCMC.cpp')
-sourceCpp('ADDPF.cpp')
+#sourceCpp('ADDPF.cpp')
+sourceCpp('ADPFUpdater.cpp')
+#sourceCpp('APF.cpp')
 
 sigSq = 0.02
 phi = 0.9
 mu = -0.5
 T = 500
-S = 50
 
-x = rep(0, T+S)
-y = rep(0, T+S)
-for(t in 1:(T+S)){
+x = rep(0, T)
+y = rep(0, T)
+for(t in 1:T){
   if(t == 1){
     x[t] = rnorm(1, mu, sqrt(sigSq/(1-phi^2)))
   } else {
@@ -24,53 +25,17 @@ for(t in 1:(T+S)){
   y[t] = exp(x[t]/2) * rnorm(1)
 }
 
-
-
 yStar = log(y^2)
-MCMC = SVM_MCMC(yStar[1:T], 50000, 0.05)
-
-
-yM = matrix(y[1:T], T)
+MCMC = SVM_MCMC(yStar, 50000, 0.05)
+yM = matrix(y, T)
 mean = c(-4.4, 0, 0.86, mu)
 var = diag(c(0.7, 0.5, 0.101, sqrt(sigSq/(1-phi^2))))
-
 lambda = cbind(mean, var)
-fitBPF = VBIL_PF(yM, lambda, S=50, alpha=0.25, maxIter=5000, threshold=0.05, thresholdIS=0.9)
-
+fitBPF = VBIL_PF(yM, lambda, S=10, alpha=0.25, maxIter=5000, threshold=0.05, thresholdIS=0.9)
 lambda = cbind(mean, var)
-fitDPF = VBIL_DPF(matrix(yStar[1:T], T), lambda, S=10, N=100, alpha=0.25, maxIter=1, threshold=0.05, thresholdIS=0.9)
+fitAPF = VBIL_APF(yM, lambda, S=10, alpha=0.25, maxIter=5000, threshold=0.05, thresholdIS=0.9)
 
-
-
-
-
-
-lambda = cbind(mean, var)
-fitPartial = VBIL_PF(matrix(y[1:T], T), lambda, S=10, alpha=0.1, maxIter=5000, threshold=0.05, thresholdIS=0.9)
-
-U = fitPartial$U
[…]

### 2017-08-04 Tran Nott Kohn/SVM_MCMC.cpp

@@ -16,7 +16,7 @@ rowvec FFBSsvm(vec y, rowvec theta, rowvec s){
   //forward filter steps
   vec att(T, fill::ones); //0 - (T-1) -> x1 - xT
   vec ptt(T, fill::ones);
-  rowvec draws(T); //0 - T -> x0 - xT
+  rowvec draws(T+1); //0 - T -> x1 - xT+1
   double at = theta[1];
   double pt = theta[0] / (1 - pow(theta[2], 2));
   double vt = y[0]  -  at  -  mixmeans[s[0]]  +  1.2704;
@@ -29,14 +29,17 @@ rowvec FFBSsvm(vec y, rowvec theta, rowvec s){
     att[t] = at  +  pt * vt / (pt + mixvar[s[t]]);
     ptt[t] = pt  -  pow(pt, 2) / (pt + mixvar[s[t]]);
   }   
+  // x_T+1
+  at = theta[1]  +  theta[2] * (att[T-1] - theta[1]);
+  pt = pow(theta[2], 2) * ptt[T-1]  +  theta[0];
   //backwards sampler steps
   double vstar;
   double fstar;
   double mstar;
-  vec atT(T);
-  vec ptT(T);
-  draws[T-1] = att[T-1]  +  sqrt(ptt[T-1]) * randn<vec>(1)[0];
-  for(int t = T - 1; t > 0; --t){
+  vec atT(T+1);
+  vec ptT(T+1);
+  draws[T] = at  +  sqrt(pt) * randn<vec>(1)[0];
+  for(int t = T; t > 0; --t){
     vstar = draws[t]  -  theta[1]  -  theta[2] * (att[t-1] - theta[1]);
     fstar = pow(theta[2], 2) * ptt[t-1]  +  theta[0];
     mstar = ptt[t-1] * theta[2];
@@ -101,7 +104,7 @@ Rcpp::List SVM_MCMC(vec y, int reps, double stepSize){
     //Mu ~ Normal(mean, var)
     meanMuNumer = a(i-1, 0) * (1 - pow(theta(i-1, 2), 2));
     meanMuDenom = theta(i, 0)  +  10 * (T + T*pow(theta(i-1, 2), 2)  - 2*T*theta(i-1, 2) + 1 - pow(theta(i-1, 2), 2));
-    for(int t = 1; t < T; ++t){
+    for(int t = 1; t < T+1; ++t){
[…]

### 2017-09-21 Cars/basicModel.Rnw
Hamilton Filtering

\documentclass[12pt,a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{alltt}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage{amsthm}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\setcounter{MaxMatrixCols}{30}

\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\setlength\parindent{0pt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\begin{document}

\section{The Basic Model}

Consider the motion model where x, y coordinates in eucliedean space can be determined by the history of coordinates as well as the current velocity, $v$, and steering angle, $\delta$, at time $t$ by
\begin{align}
x_t &= x_{t-1} + v_t \cos(\delta_t) \label{xEq} \\
y_t &= y_{t-1} + v_t \sin(\delta_t). \label{yEq}
\end{align}

The change in velocity equals acceleration, $a$, which is a stationary autoregressive process.
\begin{align}
v_t &= v_{t-1} + a_t \label{vEq} \\
a_t &= \phi a_{t-1} + \epsilon_t \label{aEq}
\end{align}
reducing to an ARIMA(1, 1, 0) process for velocity:
\begin{equation}
v_t - v_{t-1} = \phi(v_{t-1} - v_{t-2}) + \epsilon_t. \label{vARIMA}
\end{equation}

Finally, steering angle follows an AR(1) process with mean $\pi/2$.
\begin{equation}
\delta_t = \pi / 2 + \gamma (\delta_{t-1} - \pi / 2) + \eta_t \label{dEq}.
\end{equation}

Note that $v_t$ and $\delta_t$ in (\ref{xEq}) and (\ref{yEq}) are fully identifiable from observations of $\Delta x_t = x_t -  x_{t-1}$ and $\Delta y_t = y_t - y_{t-1}$,
\begin{align}
v_t &= v(\Delta x_t, \Delta y_t) = \sqrt{\Delta x_t^2 + \Delta y_t^2} \label{vEq2} \\
\delta_t &= \delta(\Delta x_t, \Delta y_t) = \tan^{-1}\left(\frac{\Delta y_t}{\Delta x_t} \right). \label{dEq2}
\end{align}

\iffalse
These can be substituted into (\ref{vARIMA}) and (\ref{dEq}) to give
\begin{align}
\sqrt{\Delta x_t^2 + \Delta y_t^2} &=  \sqrt{\Delta x_{t-1}^2 + \Delta y_{t-1}^2} + \phi \left(\sqrt{\Delta x_{t-1}^2 + \Delta y_{t-1}^2} - \sqrt{\Delta x_{t-2}^2 + \Delta y_{t-2}^2}\right) + \epsilon_t \label{vEq3} \\
\tan^{-1}\left(\frac{\Delta y_t}{\Delta x_t} \right) &= \pi / 2 + \gamma \left(\tan^{-1}\left(\frac{\Delta y_{t-1}}{\Delta x_{t-1}} \right) - \pi/2 \right) + \eta_t \label{dEq3}
\end{align}
\fi

Assuming that $\epsilon_t \sim N(0, \sigma^2_{\epsilon})$ and $\eta_t \sim N(0, \sigma^2_{\eta})$, using the notation that $x_{1:T} = \{x_t | t = 1, \dots T\}$, $y_{1:T} = \{y_t | t = 1, \dots T\}$, and $\theta = \{\gamma, \phi, \sigma^2_{\epsilon}, \sigma^2_{\eta}\}$, as well as treating $x_{1:3}$ and $y_{1:3}$ as given, the likelihood can be written as
\begin{equation}
p(x_{4:T}, y_{4:T} | \theta) = \prod_{t=4}^T \left(p_v(x_t, y_t | \theta) p_{\delta}(x_t, y_t | \theta) \right)
\end{equation}
where
\begin{align}
p_v(x_t, y_t | \theta) &\propto  \sigma_{\epsilon}^{-1} \exp\left\{\frac{-1}{2 \sigma^2_{\epsilon}} \left(v(\Delta x_t, \Delta y_t) - (1+\phi) v(\Delta x_{t-1}, \Delta y_{t-1}) + \phi v(\Delta x_{t-2}, \Delta y_{t-2})\right)^2  \right\} \label{vDens} \\
p_{\delta}(x_t, y_t | \theta) &\propto  \sigma_{\eta}^{-1} \exp\left\{\frac{-1}{2 \sigma^2_{\eta}} \left(\delta(\Delta x_t, \Delta y_t) - \pi/2 - \gamma \left(\delta(\Delta x_{t-1}, \Delta y_{t-1}) - \pi/2\right)\right)^2 \right\} \label{dDens} 
\end{align}
The following priors are used, where $IG$ denotes the inverse of a Gamma(shape, rate) distribution,
\begin{align*}
\gamma &\sim B(\alpha_{\gamma}, \beta_{\gamma}) \\
\phi &\sim B(\alpha_{\phi}, \beta_{\phi}) \\
\sigma^2_{\epsilon} &\sim IG(\alpha_{\epsilon}, \beta_{\epsilon}) \\
\sigma^2_{\eta} &\sim IG(\alpha_{\eta}, \beta_{\eta})
\end{align*}

A variational approximation is constructed for $\theta^* = \{\log(\gamma / (1-\gamma)), \log(\phi / (1-\phi)), \log(\sigma^2_{\epsilon}), \log(\sigma^2_{\eta})\}$, approximating the unknown posterior of $p(\theta^* | x_{1:T}, y_{1:T})$ with a multivariate normal distribution $q_{\lambda}(\theta^* | x_{1:T}, y_{1:T})$, choosing the set of paramaters $\lambda$ that minimises the Kullback-Leibler divergence from $q$ to $p$,
\begin{equation}
KL(q || p) = \int_{\theta^*} q_{\lambda}(\theta^* | x_{1:T}, y_{1:T}) \log \left( \frac{ q_{\lambda}(\theta^* | x_{1:T}, y_{1:T})}{p(\theta^* | x_{1:T}, y_{1:T})} \right) d_{\theta^*}.
\end{equation}

\begin{figure}[hb]
\centering
\includegraphics[width = 0.6\textwidth]{basicFit.png}
\end{figure}

\section{Adding a Markov Switching Variable}

The complexity of the steering process in (\ref{dEq}) is increased by introducing a two-state Markov switching variable $s_t$, where $s_t = 0$ indicates that the driver is continuing on their current path, and $s_t = 1$ will turn the car towards the centre of the lane. The new steering equation is described by
\begin{equation}
\delta_t = (1 - s_t) \delta_{t-1}  s_t (pi/2 + \zeta(x_{t-1} - M)) +  \eta_t \label{dEqMS}.
\end{equation}
where $M$ denotes the midpoint of the lane. Note that this now describes a random walk when $s_t = 0$, with large departures from $x_t = M$ controlled when $s_t = 1$.  $s_t$ is governed by the transition probabilities $\rho_{ij} = p(s_t = j | s_{t-1} = 1)$, with $\rho_{00} = 1 - \rho_{01}$ and $\rho_{10} = 1 - \rho_{11}$. As drivers are more likely to steer towards the centre of the lane the further away they are from the centre, $\rho_{10}$ and $\rho_{11}$ are given by 
\begin{equation}
\rho_{i1} = \frac{1}{1 + \exp\{-(\alpha_i + \beta_i (x_{t-1} - M)^2)\}} \mbox{ for } i = 0, 1.
\end{equation}
With this formulation, $s_t =  0$ represents a default driving state, and $s_t = 1$ represents the driver actively correcting for drift.

This formulation removes $\gamma$ from $\theta$, but adds five additional parameters, $\zeta, \alpha_0, \beta_0, \alpha_1$, and $\beta_1$, each with a $N(\mu_i, \tau_i^2)$ prior. The likelihood can be calculated by the Hamilton filter, and is given by
\begin{equation}
p(x_{4:T}, y_{4:T} | \theta) = \prod_{t=4}^T \left(\sum_{i=0}^1 \sum_{j=0}^1 \rho_{ij} \xi_{i, t-1} p_{v}(x_t, y_t | \theta) p_{\delta j}(x_t, y_t | \theta) \right)
\end{equation}
where
\begin{align}
p_{\delta 0}(x_t, y_t | \theta) &\propto  \sigma_{\eta}^{-1} \exp\left\{\frac{-1}{2 \sigma^2_{\eta}} \left(\delta(\Delta x_t, \Delta y_t) - \delta(\Delta x_{t-1}, \Delta y_{t-1}) \right)^2 \right\} \label{dDensS0}, \\
p_{\delta 1}(x_t, y_t | \theta) &\propto   \sigma_{\eta}^{-1} \exp\left\{\frac{-1}{2 \sigma^2_{\eta}} \left(\delta(\Delta x_t, \Delta y_t) - \pi/2 - \zeta (x_{t-1} - M)\right)^2  \right\} \label{dDensS1},
\end{align}
and
\begin{align}
\xi_{j, t} &= P(s_t = j | x_0, y_0, \dots, x_{t}, y_{t}, \theta) \\
&= \frac{\sum_{i=0}^1\rho_{ij} \xi_{i, t-1} p_v(x_{t}, y_{t})p_{\delta j}(x_{t}, y_{t})}
{\sum_{i=0}^1 \sum_{j=0}^1 \rho_{ij} \xi_{i, t-1} p_{v}(x_t, y_t | \theta) p_{\delta j}(x_t, y_t | \theta)} \label{HF}.
\end{align}
The recursion in (\ref{HF}) is initialised at $\xi_3 = \{0.9, 0.1\}$.
\end{document}

### 2017-09-21 Cars/basic.R

@@ -1,0 +1,232 @@
+library(tidyverse)
+set.seed(465786)
+sigSqV <- 0.001
+sigSqD <- 0.001
+phi <- 0.9
+gamma <- 0.9
+T <- 500
+pos <- data.frame()
+
+v = 5
+vLag1 = 4.9
+d = pi/2
+x = 5
+y = 0
+st = 0
+
+
+for(t in 1:T){
+  d = pi/2 + gamma * (d - pi/2) + rnorm(1, 0, sqrt(sigSqD))
+  vLag2 = vLag1
+  vLag1 = v
+  v = vLag1 + phi * (vLag1 - vLag2) + rnorm(1, 0, sqrt(sigSqV))
+  x = x + v * cos(d)
+  y = y + v * sin(d)
+  pos <- rbind(pos, data.frame(x=x, y=y, v=v, d=d, t=t))
+}
+
+ggplot(pos) + geom_path(aes(x, y))
+
+mu = rep(0, 4)
+u = diag(0.1, 4)
+lambda = as.matrix(c(mu, u))
+hyper = c(3, 0.003, 3, 0.003, 20, 1.5, 20, 1.5)
+data = data.frame(dx = pos$x[2:T] - pos$x[1:(T-1)], dy =  pos$y[2:T] - pos$y[1:(T-1)])
+data = mutate(data, vt = sqrt(dx^2 + dy^2), dt = atan2(dy, dx)) %>% select(vt, dt) %>% as.matrix() 
+alpha = 0.01
+beta1 = 0.9
+beta2 = 0.99
+threshold = 0.01
+S = 100
+maxIter = 100
+
+carsBasic <- function(data, lambda, hyper, S, maxIter, alpha, beta1, beta2, threshold = 0.01, thresholdIS = 0){
+  sobol <- sobol_points(100+S, 4)
+  diff <- threshold + 1
+  iter <- 1
+  LB <- numeric(maxIter)
+  M <- numeric(20)
+  V <- numeric(20)
+  e <- 1e-8
+  meanLB <- 0
+  oldMeanLB <- 0
+  while(diff > threshold){
+    if(iter > maxIter){
+      break
+    }
+    if(any(is.na(lambda))){
+      break
+    }
+    grad <- matrix(0, 20, S)
+    eval <- numeric(S)
+    q <- numeric(S)
+    unif <- shuffle(sobol)
+    epsilon <- qnorm(unif[101:(100+S), ])
+    for(s in 1:S){
[…]

### 2017-09-21 Cars/cppForCarsR.cpp

@@ -66,9 +66,9 @@ struct logPZ {
       States(k, 0) = sqrt(sigSqA / (1 - pow(phi, 2))) * randn<vec>(1)[0];
       States(k, 1) = 0;
       States(k, 2) = pi / 2  +  sqrt(sigSqD) * randn<vec>(1)[0];
-      States(k, 3) = 5.0 + 0.01 * randn<vec>(1)[0];
-      States(k, 4) = 5.0 + 0.01 * randn<vec>(1)[0];
-      States(k, 5) = 0.1 * randn<vec>(1)[0];
+      States(k, 3) = 4.5 + 0.01 * randn<vec>(1)[0];
+      States(k, 4) = x(0, 0) + 0.01 * randn<vec>(1)[0];
+      States(k, 5) = 2*x(0, 1) - x(1, 1) + 0.1 * randn<vec>(1)[0];
     }
     xAllDens = 0;
     
@@ -176,13 +176,13 @@ struct logPZ {
   }
 };
 
-struct logP{
+struct logPJ{
   const mat x;
   const int obs;
   const int N;
   const vec epsilon;
   const vec hyperParams;
-  logP(const mat& xIn, const int& obsIn, const int& nIn, const vec& epsIn, const vec& hypIn) :
+  logPJ(const mat& xIn, const int& obsIn, const int& nIn, const vec& epsIn, const vec& hypIn) :
     x(xIn), obs(obsIn), N(nIn), epsilon(epsIn), hyperParams(hypIn) {}
   template <typename T> //
   T operator ()(const Matrix<T, Dynamic, 1>& lambdaV)
@@ -213,11 +213,22 @@ struct logP{
       pow(beta0 - hyperParams(16), 2) / (2*hyperParams(17))  -
       pow(alpha1 - hyperParams(18), 2) / (2*hyperParams(19))  -
       pow(beta1 - hyperParams(20), 2) / (2*hyperParams(21)); 
+    // Evaluate Log Det J
+    T logdet = 0;
+    for(int i = 0; i < 11; ++i){
+      if(i >= 6){ // linear
+        logdet += log(sqrt(pow(lambdaV(11+i), 2)));
+      } else if(i < 4){ // exponential
+        logdet += log(sqrt(pow(lambdaV(11+i), 2))) + lambdaV(i) + sqrt(pow(lambdaV(11+i), 2))*epsilon(i);
+      } else { // sigmoid
[…]

### 2017-10-17 Cars/basicModel.Rnw
general updates

@@
 \centering
-\includegraphics[width = 0.6\textwidth]{basicFit.png}
+\includegraphics[width = 1\textwidth]{basicFit.png}
 \end{figure}
@@
 \end{equation}
+The velocity equation is also simplified for computational purposes (though the results had the same problems with it the same as earlier - so this did not really help).
+\begin{equation}
+v_t = v_{t-1} + \epsilon_t \label{vEqMS}
+\end{equation}
 With this formulation, $s_t =  0$ represents a default driving state, and $s_t = 1$ represents the driver actively correcting for drift.
 
-This formulation removes $\gamma$ from $\theta$, but adds five additional parameters, $\zeta, \alpha_0, \beta_0, \alpha_1$, and $\beta_1$, each with a $N(\mu_i, \tau_i^2)$ prior.
+This formulation removes $\gamma$ and $\phi$ from $\theta$, but adds five additional parameters, $\zeta, \alpha_0, \beta_0, \alpha_1$, and $\beta_1$, each with a $N(\mu_i, \tau_i^2)$ prior.
 The likelihood can be calculated by the Hamilton filter, and is given by
@@
 \begin{align}
+p_v(x_t, y_t) &\propto \sigma_{\epsilon}^{-1} \exp\left\{ \frac{-1}{2 \sigma^2_{\epsilon}} \left(v(\Delta x_t, \Delta y_t) - v(\Delta x_{t-1}, \Delta y_{t-1})\right)^2\right\} \label{vDensS} \\
 p_{\delta 0}(x_t, y_t | \theta) &\propto  \sigma_{\eta}^{-1} \exp\left\{\frac{-1}{2 \sigma^2_{\eta}} \left(\delta(\Delta x_t, \Delta y_t) - \delta(\Delta x_{t-1}, \Delta y_{t-1}) \right)^2 \right\} \label{dDensS0}, \\
@@
 \end{align}
-The recursion in (\ref{HF}) is initialised at $\xi_3 = \{0.9, 0.1\}$.
+The recursion in (\ref{HF}) is initialised at $\xi_2 = \{0.9, 0.1\}$.
+
+\begin{figure}[hb]
+\centering
+\includegraphics[width = 1\textwidth]{VBMCMC}
+\end{figure}
 \end{document}

### 2017-11-01 Cars/paper.tex
Tweaks to HierMix MCMC

\documentclass[12pt,a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{alltt}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage{amsthm}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\setcounter{MaxMatrixCols}{30}

\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.1in}
\setlength{\evensidemargin}{0.1in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{8.25in}
\setlength\parindent{0pt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

\section{Introduction}

\begin{itemize}
\item Self driving cars becoming more common
\item Cars are easy capable of driving by themselves if there is no traffic around them
\item Cars need to be able to predict movement of other cars
\item Drivers display differing behaviour, able to group drivers into different categories
\item Heteogeneity between and within groups
\item Need to deal with new cars as they are observed
\end{itemize}

\section{Data processing}

\begin{itemize}
\item Modern cars are able to track lane marking and can calculate distance to other cars in relative co-ordinates (Thuy and Leon 2010)
\item Data from NGSIM is in coordinates relative to start of road - no impact of the curvature of the lane
\item Fit curve to car data to estimate lane midpoints (Woo et. al. 2016) 
\begin{enumerate}
\item At each point in time $t$ calculate the total distance each car has travelled along the road using $d_{i, t} = \sum_{s=1}^t v_{i, s}$, where $v_{i, s}$ is the velocity of car $i$ at time $s$. 
\item Estimate polynomial functions of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
\item Use polynomials to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
\item Define new coordinates 
\begin{align}
x^*_{i, t} &= \sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
y^*_{i, t} &= d_{i, t} \label{yRel}
\end{align}
\end{enumerate}
\item Note: This should be calculated separately for each lane.
\end{itemize}

\section{Hierarchical Motion Model}

The position of car $i$ at time $t$ can be fully determined by its inital position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the cars velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
\begin{align}
x^*_{t} &= x^*_{t-1} + v_{t} \cos(\delta_{t}) \label{xEq} \\
y^*_{t} &= y^*_{t-1} + v_{t} \sin(\delta_{t}). \label{yEq},
\end{align}
where $\delta_{t} = \pi/2$ denotes that the car has no lateral movement and $x^*_{t} = x^*_{t-1}$.
From this relationship, the key inputs can be extracted by
\begin{align}
\delta_{t} &= \tan^{-1}\left(\frac{(y^*_{t} - y^*_{t-1})}{(x^*_{t} - x^*_{t-1})} \right) \label{dEq} \\
v_{t} &= \sqrt{(x^*_{t} - x^*_{t-1})^2 + (y^*_{t} - y^*_{t-1})^2} \label{vEq} \\
a_{t} &= v_{t} - v_{t-1}. \label{aEq}
\end{align}
where $a_{t}$ denotes the acceleration at time $t$.

Auto-regressive processes of orders $p$ and $q$ for $a_i$ and  $\delta_i$ are used to model the level of persistance and variation of driver actions, with
\begin{align}
a_{t} &= \sum_{j = 1}^p \phi_{j} a_{t-j} + \epsilon_{t} \label{aAR} \\
\delta_{t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{t-j} - \pi/2) + \eta_{t} \label{dAR}
\end{align}
assuming that $\epsilon_{t}$ are independently and identically distributed according to a Student's t distribution with variance $\sigma^{2}_{\epsilon}$ and degrees of freedom $\nu_{\epsilon}$, and $\eta_{t}$ similarly has a Student't t distribution with variance $\sigma^{2}_{\eta}$ and degrees of freedom $\nu_{\eta}$. The parameter vector to be estimated is collected as $\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\nu_{\epsilon}), \log(\sigma^{2}_{\eta}), \log(\nu_{\eta}) \} \in \mathbb{R}^{p + q + 4}$.

Multiple groups of differing driver behaviour can be allowed while still retaining heterogeneity with each group by jointly estimating each vehicle driver's individual parameter vector, $\theta_i$, as part of a heirarchical model, independently drawing $\theta_i$ from a $K$ component mixture of multivariate normal distributions, augmenting each $\theta_i$ with a mixture component $k_i = 1, \dots, K$ such that 
\begin{equation}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j).
\end{equation}
For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with degrees of freedom $\tau_j$ and scale matrix $\Psi_j$, and finally $k_{i, j}$ has a Dirichlet prior with parameters $\alpha_{i, j} = \alpha_i$. This can be estimated with MCMC methods.

\section{Online Classification}

Notation:
\begin{itemize}
\item $y_{i, 1:T} = \{\delta_{i, s}, a_{i, s} | s = 1, 2, \dots, T\}$ 
\item $y_{1:N} = \{y_{i, 1:T} | i = 1, 2, \dots N\}$
\item $q(\theta_i | y_{i, 1:T}) \approx p(\theta_i | y_{i, 1:T}, y_{1:N})$, $\lambda$ parameters and dependence on other cars suppressed - more important to be clear about which observations from the new vehicle are used to construct the approximation.
\item $\beta = \{\mu_i, \Sigma_i, i, = 1, 2, \dots, K)$.
\end{itemize}

Modelling process:
\begin{itemize}
\item Fit heirarchical model from section 3 for cars $i = 1, \dots N$ before car starts driving - eg. During manufacturing process so unlimited time available.
\item Observe a new car, say car $N + 1$.
\item Want to learn about parameters $\theta_{N+1}$ and which group $k_{N+1}$ the driver belongs to
\item $p(\theta_{N+1}, k_{N+1} | y_{1:N+1}) \propto p(y_{N+1, 1:T} | \theta_{N+1}) p(\theta_{N+1} | k_{N+1} \beta)p(k_{N+1}) p(\beta | y_{1:N})$
\item Essentially using mixture distribution posterior fit for cars 1 to N as prior for $\theta_{N+1}$.
\item However we want to be able to do this as we observe $y_{N+1}$, eg: Fit distribution to $p(\theta_{N+1}, k_{N+1} | y_{N+1, 1:10}, y_{1:N})$, then to $p(\theta_{N+1}, k_{N+1} | y_{N+1, 1:11}, y_{1:N})$ etc.
\item To do this in VB we need to make two further approximations:
\begin{enumerate}
\item Fit a parametric distribution to the MCMC posterior
\item Replace each  $p(\theta_{N+1} | y_{i, 1:S}, y_{1:N})$ used to fit $q(\theta_{N+1} | y_{N+1, 1:S+1})$ with the VB fit from the last step, $q(\theta_{N+1} | y_{N+1, 1:S})$.
\end{enumerate}
\end{itemize}

\section{Empirical}

\section{Conclusion}

\end{document}

### 2017-11-01 Cars/createY600.R

@@ -96,3 +96,5 @@ Y600 %>%
   summarise(class = head(class, 1)) %>%
   .$class %>%
   table()
+
+

### 2017-11-01 Cars/mixtureMCMCMetHast.cpp

@@ -10,7 +10,7 @@ using namespace arma;
 using namespace std;
 
 // [[Rcpp::export]]
-double logDensity (mat data, vec theta, vec mu, mat varInv){
+double nlogDensity (mat data, vec theta, vec mu, mat varInv){
   int T = data.n_rows;
   double dens = 0.5 * log(det(varInv)) - 0.5 * as_scalar((theta - mu).t() * varInv * (theta - mu));
   for(int t = 2; t < T; ++t){
@@ -22,4 +22,16 @@ double logDensity (mat data, vec theta, vec mu, mat varInv){
 }
   
   
-  
+// [[Rcpp::export]]
+double tlogDensity (mat data, vec theta, vec mu, mat varInv){
+  int T = data.n_rows;
+  double sigSqV = exp(theta(0)), sigSqD = exp(theta(1)), nuV = exp(theta(6)), nuD = exp(theta(7)),
+    dens = 0.5 * log(det(varInv)) - 0.5 * as_scalar((theta - mu).t() * varInv * (theta - mu));
+  for(int t = 2; t < T; ++t){
+    dens += lgamma((1+nuV)/2)  +  lgamma((1+nuD)/2)  -  lgamma(nuV/2)  -  lgamma(nuD/2) -
+      0.5 * (theta(0) + theta(1) + theta(6) + theta(7))  - 
+      (nuV+1)/2 * log(1 + pow(data(t, 0) - theta(2) * data(t-1, 0) - theta(3) * data(t-2, 0), 2) / (nuV * sigSqV))  -
+      (nuD+1)/2 * log(1 + pow(data(t, 1) - theta(4) * data(t-1, 1) - theta(5) * data(t-2, 1), 2) / (nuD * sigSqD));
+  }
+  return dens;
+}

### 2017-11-06 Cars/paper.tex
ID tags

@@
 \item At each point in time $t$ calculate the total distance each car has travelled along the road using $d_{i, t} = \sum_{s=1}^t v_{i, s}$, where $v_{i, s}$ is the velocity of car $i$ at time $s$. 
-\item Estimate polynomial functions of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
+\item Estimate smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
 \item Use polynomials to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
@@
 \end{enumerate}
-\item Note: This should be calculated separately for each lane.
+\item Note: This should be calculated separately for each lane, cars used to estimate splines are excluded from further modelling
 \end{itemize}

### 2017-11-10 Cars/paper.tex
minor tweaks

@@
 \end{equation}
-For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with degrees of freedom $\tau_j$ and scale matrix $\Psi_j$, and finally $k_{i, j}$ has a Dirichlet prior with parameters $\alpha_{i, j} = \alpha_i$.
+For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with degrees of freedom $\tau_j$ and scale matrix $\Psi_j$.
+$k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi_i} = \{\pi_{1, i}, \dots, \pi_{K, i}\}$, where $\boldsymbol{\pi_i}$ has a Dirichlet prior with parameters $\alpha_1, \alpha_2, \dots, \alpha_K$.
 This can be estimated with MCMC methods.
 
-\section{Online Classification}
+\section{Online Process}
 
@@
 \begin{itemize}
-\item Fit heirarchical model from section 3 for cars $i = 1, \dots N$ before car starts driving - eg. During manufacturing process so unlimited time available.
-\item Observe a new car, say car $N + 1$.
-\item Want to learn about parameters $\theta_{N+1}$ and which group $k_{N+1}$ the driver belongs to
-\item $p(\theta_{N+1}, k_{N+1} | y_{1:N+1}) \propto p(y_{N+1, 1:T} | \theta_{N+1}) p(\theta_{N+1} | k_{N+1} \beta)p(k_{N+1}) p(\beta | y_{1:N})$
-\item Essentially using mixture distribution posterior fit for cars 1 to N as prior for $\theta_{N+1}$.
-\item However we want to be able to do this as we observe $y_{N+1}$, eg: Fit distribution to $p(\theta_{N+1}, k_{N+1} | y_{N+1, 1:10}, y_{1:N})$, then to $p(\theta_{N+1}, k_{N+1} | y_{N+1, 1:11}, y_{1:N})$ etc.
-\item To do this in VB we need to make two further approximations:
-\begin{enumerate}
-\item Fit a parametric distribution to the MCMC posterior
-\item Replace each  $p(\theta_{N+1} | y_{i, 1:S}, y_{1:N})$ used to fit $q(\theta_{N+1} | y_{N+1, 1:S+1})$ with the VB fit from the last step, $q(\theta_{N+1} | y_{N+1, 1:S})$.
-\end{enumerate}
+\item From Section 3 we have the distribution $p(\beta | y_{1:N})$.
+\item The posterior distribution of a new car is given by
+\begin{equation}
+p(\theta_{N+1} | y_{1:N+1}) \propto p(y_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \beta) p (\beta | y_{1:N})
+\end{equation}
+\item As N is large we can replace $ p(\theta_{N+1} | \beta) p (\beta | y_{1:N})$ with $ p(\theta_{N+1} | \beta_{MAP})$ where $\beta_{MAP}$ is the maximum of $ p (\beta | y_{1:N})$, or some other suitable point estimate.
+\item Observe $y_{N+1, 1:S_1}$, fit VB $q(\theta_{N+1} | y_{N+1, 1:S_1}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_1})$
+\item Observe $y_{N+1, S_1+1:S_2}$, in this case the model becomes
+\begin{equation}
+p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_2}) \propto p(y_{N+1, S_1+1:S_2} | \theta_{N+1})p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S_1})
+\end{equation}
+\item Set $p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S_1}) = q(\theta_{N+1} | y_{N+1, 1:S_1})$, fit VB  $q(\theta_{N+1} | y_{N+1, 1:S_2}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_2})$
+\item Repeat as new data is observed
 \end{itemize}
@@
 
+Comparisions:
+\begin{enumerate}
+\item VB Approximation with the above process
+\item One Car MCMC/VB without prior information
+\item One Car MCMC/VB with mild prior informaiton - eg. One Component
+\end{enumerate}
+Compare performance with different observed samples - prior informaiton should be very beneficial for first few seconds but expect methods to converge over time.
+
 \section{Conclusion}

### 2017-11-21 Cars/paper.tex
forecasting

@@
 \item Cars need to be able to predict movement of other cars
-\item Drivers display differing behaviour, able to group drivers into different categories
+\item Wide range of unconditional variances and first two partial autocorrelations observed while fitting each model individually - evidence of heterogeneity
+\item Drivers display differing behaviour, could be able to group drivers into different categories
 \item Heteogeneity between and within groups
@@
 \begin{align}
-x^*_{i, t} &= \sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
+x^*_{i, t} &= \mbox{sign}(x_{i, t} - \hat{x}_{i, t})\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
 y^*_{i, t} &= d_{i, t} \label{yRel}
@@
 
+\begin{itemize}
+\item Add a Direchlet Process Prior to make this an infinite mixture
+\item Use slice sampling in the MCMC
+\end{itemize}
+
 \section{Online Process}
@@
 \item Repeat as new data is observed
+\item MCMC will always need to use the whole dataset to update the posterior, but this should only need the new data so runtime will not depend on T.
 \end{itemize}
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width = 0.8\textwidth]{HierSingleKDE.png}
+\caption{Left: $p(\theta_i | \beta, y_{1:N})$ for a hierarchical mixture model with K = 6 and N = 2000.
+Right: Kernel Density Estimate of the posterior means from the same cars modelled individually.
+The hierarchical model has captured the bimodality and skewness present in the individual posterior means.}
+\label{fig:HierSingleKDE}
+\end{figure}
+
 Comparisions:
 \begin{enumerate}
-\item VB Approximation with the above process
-\item One Car MCMC/VB without prior information
-\item One Car MCMC/VB with mild prior informaiton - eg. One Component
+\item No prior information for the new car
+\item Prior information without heterogeneity - all cars share parameters
+\item Prior information from hierarchical model with one component
+\item Prior information from large / infinite mixture hierarchical model
 \end{enumerate}
-Compare performance with different observed samples - prior informaiton should be very beneficial for first few seconds but expect methods to converge over time.
+
+\begin{itemize}
+\item Compare forecasts for different sample sizes and forecast horizons
+\item eg. Fit model to one second, forecast next second
+\item Fit model to two seconds, forecast next second etc.
+\item Expect models based on the MCMC to outperform non MCMC models for first few seconds, not to matter as much as large amounts of data is obsered.
+\item Could also compare VB and MCMC methods for the same models.
+\end{itemize}
 

### 2017-11-21 Cars/HierMixMod.R

@@ -186,7 +186,7 @@ densities %>%
 
 OtherMods{
   data <- readRDS('MCMCData.RDS')
-  reps <- 50000
+  reps <- 5000
   
   draws <- list()
   hyper <- list(mean = c(-5, -5, rep(0, 4)), varInv = solve(diag(10, 6)), v = 6, scale = diag(1, 6))
@@ -194,14 +194,54 @@ OtherMods{
   for(i in 1:N){
     draws[[i+1]] <- c(-5, -5, 0, -0.1, 0.15, 0.05)
   }
+  N <- 100
+  datSub <- list()
+  for(i in 1:N){
+    datSub[[i]] <- data[[i]]
+  }
   
-  noMixDraws <- hierNoMixMCMC(data, reps, draws, hyper, thin = 10)
+  noMixDraws <- hierNoMixMCMC(datSub, reps, draws, hyper, thin = 10)
   saveRDS(noMixDraws, 'noMixN2000.RDS')
+  
+  noMixDraws$draws[[1]]$mean %>%
+    as.data.frame() %>%
+    cbind(iter = 1:(reps/thin)) %>%
+    mutate(V1 = exp(V1), V2 = exp(V2)) %>%
+    rename(sigSq_eps = V1, sigSq_eta = V2, phi1 = V3, phi2 = V4, gamma1 = V5, gamma2 = V6) %>%
+    gather(var, draw, -iter) %>%
+    mutate(var = factor(var, levels = c('sigSq_eps', 'phi1', 'phi2', 'sigSq_eta', 'gamma1', 'gamma2'))) %>%
+    filter(iter > 100) %>%
+    ggplot() + geom_line(aes(iter, draw)) + facet_wrap(~var, scales = 'free')
+  
+  noMixMean <- rep(0, 6)
+  noMixVar <- matrix(0, 6, 6)
+  for(i in (reps/(2*thin)+1):(reps/thin)){
+    noMixMean <- noMixMean + noMixDraws$draws[[1]]$mean[i,] / (reps / (2 * thin))
+    noMixVar <- noMixVar + solve(noMixDraws$draws[[1]]$varInv[,,i]) / (reps / (2 * thin))
+  }
+  noMixLam <- c(noMixMean, chol(noMixVar))
   
   draws <- c(-5, -5, 0, 0, 0, 0)
   hyper <- list(mean = c(-5, -5, rep(0, 4)), varInv = solve(diag(10, 6)))
[…]

### 2017-11-21 Cars/hierarchical.cpp

@@ -503,4 +503,31 @@ Rcpp::List arUpdaterMix(mat data, Rcpp::NumericMatrix lambdaIn, vec epsilon, vec
                             Rcpp::Named("val") = eval);
 }
 
-
+// [[Rcpp::export]]
+cube evalFcDens (mat data, mat means, mat L, int M, int S, vec asup, vec dsup){
+  int N = asup.n_elem;
+  int K = means.n_cols;
+  cube densities (N * S, 2, K, fill::zeros);
+  boost::math::normal_distribution<> Zdist(0, 1);
+  for(int m = 0; m < M; ++m){
+    for(int k = 0; k < K; ++k){
+      double afc1 = data(1, 0), afc2 = data(0, 0), dfc1 = data(1, 1), dfc2 = data(0, 1), afc, dfc;
+      vec draws = means.col(k) + L.rows(k*6, (k+1)*6-1) * randn<vec>(6);
+      for(int h = 0; h < S; ++h){
+        afc = afc1 * draws(2) + afc2 * draws(3);
+        dfc = dfc1 * draws(4) + dfc2 * draws(5);
+        double Zafc = (data(2+h, 0) - afc) / sqrt(std::exp(draws(0)));
+        double Zdfc = (data(2+h, 1) - dfc) / sqrt(std::exp(draws(1)));
+        for(int n = 0; n < N; ++n){
+          densities(N*h + n, 0, k) += pdf(Zdist, Zafc) / M;
+          densities(N*h + n, 1, k) += pdf(Zdist, Zdfc) / M;
+        }
+        afc2 = afc1;
+        afc1 = afc;
+        dfc2 = dfc1;
+        dfc1 = dfc;
+      }
+    }
+  }
+  return densities;
+}

### 2017-11-21 Cars/mixtureMCMC.R

@@ -34,12 +34,14 @@ mixtureMCMC <- function(data, reps, draw, hyper, thin = 1, K = 2, error = 'gauss
   
   for(i in 2:reps){
     # timing
-    if(i == 2){
+    if(i == 50){
       startTime <- Sys.time()
-    } else if(i < 5000 & i %% 1000 == 0){
-      timePerIter <- (Sys.time() - startTime)/(i - 2)
+    } else if(i == 150){
+      timePerIter <- (Sys.time() - startTime) / 100
       class(timePerIter) <- 'numeric'
+      print(paste0('Estimated Finishing Time: ', Sys.time() + timePerIter * (reps - 150)))
       if(attr(timePerIter, 'units') == 'mins'){
+        attr(timePerIter, 'units') = 'secs'
         timePerIter <- timePerIter * 60
       }
     }
@@ -162,12 +164,14 @@ sliceSampler <- function(data, reps, draw, hyper, thin = 1, K = 10, error = 'gau
   sumK <- rep(N/K, K)
   for(i in 2:reps){
     # timing
-    if(i == 2){
+    if(i == 50){
       startTime <- Sys.time()
-    } else if(i < 5000 & i %% 1000 == 0){
-      timePerIter <- (Sys.time() - startTime)/(i - 2)
+    } else if(i == 150){
+      timePerIter <- (Sys.time() - startTime) / 100
       class(timePerIter) <- 'numeric'
+      print(paste0('Estimated Finishing Time: ', Sys.time() + timePerIter * (reps - 150)))
       if(attr(timePerIter, 'units') == 'mins'){
+        attr(timePerIter, 'units') = 'secs'
         timePerIter <- timePerIter * 60
       }
     }
@@ -293,13 +297,14 @@ hierNoMixMCMC <-function(data, reps, draw, hyper, thin = 1, error = 'gaussian', 
   stepsizeCons <- (1 - 1/dim) * sqrt(2*pi) * exp(alpha^2/2) / (2 * alpha)  +  1 / (dim * 0.234 * (1 - 0.234))
   
   for(i in 2:reps){
-    # timing
-    if(i == 2){
+    if(i == 50){
       startTime <- Sys.time()
-    } else if(i < 5000 & i %% 1000 == 0){
-      timePerIter <- (Sys.time() - startTime)/(i - 2)
+    } else if(i == 150){
+      timePerIter <- (Sys.time() - startTime) / 100
       class(timePerIter) <- 'numeric'
[…]

### 2017-11-29 Cars/paper.tex
Lots of tweaks to the slurm code

@@
 \begin{align}
-a_{t} &= \sum_{j = 1}^p \phi_{j} a_{t-j} + \epsilon_{t} \label{aAR} \\
-\delta_{t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{t-j} - \pi/2) + \eta_{t} \label{dAR}
+a_{t} &= \sum_{j = 1}^p \phi_{j} a_{t-j} + \sigma_{\epsilon} \epsilon_{t} \label{aAR} \\
+\delta_{t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{t-j} - \pi/2) + \sigma_{\eta} \eta_{t} \label{dAR}
 \end{align}
-assuming that $\epsilon_{t}$ are independently and identically distributed according to a Student's t distribution with variance $\sigma^{2}_{\epsilon}$ and degrees of freedom $\nu_{\epsilon}$, and $\eta_{t}$ similarly has a Student't t distribution with variance $\sigma^{2}_{\eta}$ and degrees of freedom $\nu_{\eta}$.
-The parameter vector to be estimated is collected as $\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\nu_{\epsilon}), \log(\sigma^{2}_{\eta}), \log(\nu_{\eta}) \} \in \mathbb{R}^{p + q + 4}$.
+where $\epsilon_{t}$ and $\eta_t$ are independently and identically distributed according to a standard normal distribution.
+The parameter vector to be estimated is collected as $\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\} \in \mathbb{R}^{p + q + 2}$.
 
@@
 $k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi_i} = \{\pi_{1, i}, \dots, \pi_{K, i}\}$, where $\boldsymbol{\pi_i}$ has a Dirichlet prior with parameters $\alpha_1, \alpha_2, \dots, \alpha_K$.
-This can be estimated with MCMC methods.
+Defining $\beta = \{\mu_i, \Sigma_i, i = 1, \dots, K\}$, $y_{i, 1:T} = \{\delta_{i, s}, a_{i, s} | s = 1, \dots, T\}$ and $y_{1:N} = \{y_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | y_{1:N})$ can be estimated with MCMC methods.
 
 \begin{itemize}
-\item Add a Direchlet Process Prior to make this an infinite mixture
-\item Use slice sampling in the MCMC
+\item To do: Add a Direchlet Process Prior to make this an infinite mixture
+\item I've been trying to do this but I get results where all of the component indicator variables in the MCMC go to one component instead of the mixture assigning as many clusters as possible as Tas suggested would happen.
+I assume I'm doing something wrong.
 \end{itemize}
 
-\section{Online Process}
+\section{Streaming Data Process}
 
@@
 \begin{itemize}
-\item $y_{i, 1:T} = \{\delta_{i, s}, a_{i, s} | s = 1, 2, \dots, T\}$ 
-\item $y_{1:N} = \{y_{i, 1:T} | i = 1, 2, \dots N\}$
 \item $q(\theta_i | y_{i, 1:T}) \approx p(\theta_i | y_{i, 1:T}, y_{1:N})$, $\lambda$ parameters and dependence on other cars suppressed - more important to be clear about which observations from the new vehicle are used to construct the approximation.
-\item $\beta = \{\mu_i, \Sigma_i, i, = 1, 2, \dots, K)$.
 \end{itemize}
 
-Modelling process:
 \begin{itemize}
-\item From Section 3 we have the distribution $p(\beta | y_{1:N})$.
-\item The posterior distribution of a new car is given by
+\item While on the road a self driving car may wish to make posterior inferences on the parameter set of cars around them.
+They may be interested in the degree of erraticness / variance of their actions.
+\item The posterior for car $N+1$ is given by 
 \begin{equation}
@@
 \end{equation}
-\item As N is large we can replace $ p(\theta_{N+1} | \beta) p (\beta | y_{1:N})$ with $ p(\theta_{N+1} | \beta_{MAP})$ where $\beta_{MAP}$ is the maximum of $ p (\beta | y_{1:N})$, or some other suitable point estimate.
+\item As N is large replace $ p(\theta_{N+1} | \beta) p (\beta | y_{1:N})$ with $ p(\theta_{N+1} | \beta_{MAP})$ where $\beta_{MAP}$ is the maximum of $p(\beta | y_{1:N})$, or some other suitable point estimate.
+This approximate distribution now has a closed form.
+\item Subsequently fitting $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})  \propto p(y_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \beta_{MAP})$ is fairly easy with MCMC / VB etc.
+\item When $S$ is small, the information in the prior will dominate the individual information.
+\item However, for some $T > S$ the information in the car's individual movements should become more relevant and forecasts could be improved by updating the distribution to $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T})$.
+\item This update takes the form of
+\begin{equation}
+p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T}) \propto p(y_{N+1, S+1:T} | \theta_{N+1}) p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})
+\end{equation}
+\item Due to the lack of a closed form expression for $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$, in reality the updating begins again from the prior re-using the data in $y_{N+1, 1:S}$.
+\item Due to the demands of driving, the car wants the update carried out as soon as possible, and ideally should be able to discard this previously used information and use only $y_{N+1, S+1:T}$, reducing the time and memory complexity from $O(T)$ to $O(T-S)$.
+\item Defining $q(\theta_{N+1} | y_{N+1, 1:S})$ as the Variational Bayes approximation of $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$, an online method is proposed to update $q(\theta_{N+1} | y_{N+1, 1:S})$ to $q(\theta_{N+1} | y_{N+1, 1:T})$:
+\begin{enumerate}
 \item Observe $y_{N+1, 1:S_1}$, fit VB $q(\theta_{N+1} | y_{N+1, 1:S_1}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_1})$
@@
 \end{equation}
-\item Set $p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S_1}) = q(\theta_{N+1} | y_{N+1, 1:S_1})$, fit VB  $q(\theta_{N+1} | y_{N+1, 1:S_2}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_2})$
+\item Set $p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S_1}) = q(\theta_{N+1} | y_{N+1, 1:S_1})$
+\item Fit VB  $q(\theta_{N+1} | y_{N+1, 1:S_2}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_2})$
 \item Repeat as new data is observed
-\item MCMC will always need to use the whole dataset to update the posterior, but this should only need the new data so runtime will not depend on T.
+\end{enumerate}
 \end{itemize}
@@
 
-Comparisions:
+Models:
 \begin{enumerate}
 \item No prior information for the new car
-\item Prior information without heterogeneity - all cars share parameters
 \item Prior information from hierarchical model with one component
-\item Prior information from large / infinite mixture hierarchical model
+\item Prior information from large mixture hierarchical model
 \end{enumerate}
 
+Methods:
+\begin{enumerate}
+\item MCMC
+\item Regular VB (fit to full sample)
+\item Streaming VB (fit to new data only)
+\end{enumerate}
+
+Evaluation:
 \begin{itemize}
 \item Compare forecasts for different sample sizes and forecast horizons
-\item eg. Fit model to one second, forecast next second
-\item Fit model to two seconds, forecast next second etc.
-\item Expect models based on the MCMC to outperform non MCMC models for first few seconds, not to matter as much as large amounts of data is obsered.
-\item Could also compare VB and MCMC methods for the same models.
+\item Strategy: Fit model to one $S = 10$ (one second), forecast next second
+\item Fit model to $S = 2$, forecast next second, repeat.
+\item Collect logscores for predictive densities marginalised over the posterior distributions.
+\end{itemize}
+
+Results:
+\begin{itemize}
+\item Models based on the stromg prior information to outperform no information models for first few seconds, where we see a significant improvement in the log scores.
+As the dataset increases to > 100 observations (10 seconds), the data dominates and the prior choice does not have much impact.
+The median logscore (and probably other quantiles, hard to see from the graph) at a larger value of T increases as expected, showing that it is important to re-estimate the model with an increased dataset.
+However it seems that this re-estimation could probably be done with a simpler prior structure than the full hierarchy.
+\item Compare streaming data VB - Hoping it will be comparable to other methods, it technically may be a bit worse as it uses one more approximation than the standard VB, but I don't think this will have a very big impact.
+It certainly will be faster, so if the performance is roughly equal this is evidence in favour of the online method.
+\item Compare VB and MCMC methods for the same models to get an idea of approximation error.
+\item Still waiting on the cluster for standard VB Finite Mixture Model and all VB streams.
 \end{itemize}

### 2017-11-29 Cars/slurmRFuns.R

@@ -63,37 +63,7 @@ carsVB <- function(data, lambda, S = 25, maxIter = 5000, alpha = 0.01, beta1 = 0
   return(list(lambda=lambda, LB = LB[1:min(iter-1, maxIter)], iter = min(maxIter, iter-1)))
 }
 
-fitCarMods <- function(data, prevFit, increment, starting){
-  S <- nrow(data)
-  results <- list()
-  K <- length(prevFit)
-  #if(increment){
-  #  for(k in 1:3){
-  #    results[[k]] <- updateVB(data, prevFit[[k]], stepsize = S, lags = 2, model = arUpdater)$lambda
-  #  }
-    #results[[4]] <- updateVBMix(data, prevFit[[4]], stepsize = S, lags = 2, model = arUpdateMix)$lambda
-  #} else {
-    for(k in 1:K){
-      mean <- prevFit[[k]][1:6]
-      u <- matrix(prevFit[[k]][7:42], 6)
-      linv <- solve(t(u))
-      results[[k]] <- carsVB(data, starting, lags = 2, model = arUpdater, mean = mean, Linv = linv, dimTheta = 6)$lambda
-    }
-    #mean <- prevFit[[4]][1:(6*6)]
-    #linv <- NULL
-    #logdets <- NULL
-    #for(k in 1:6){
-    #  u <- matrix(prevFit[[4]][36 + 1:36 + (k-1)*36], 6)
-    #  linv <- rbind(linv, solve(t(u)))
-    #  logdets <- c(logdets, -log(det(u)))
-    #}
-    #weights <- prevFit[[4]][36*7 + 1:6]
-    #results[[4]] <- carsVB(data, prevFit[[4]], lags = 2, model = arUpdateMix, mean = mean, Linv = linv, logdets = logdets, weights = weights)
-  #}
-  results
-}
-
-singleMCMCallMH <- function(data, reps, draw, hyper, thin = 1, error = 'gaussian', stepsize = 0.01){
+singleMCMCallMH <- function(data, reps, draw, hyper, thin = 1, error = 'gaussian', stepsize = 0.01, mix = FALSE){
   # set up likelihood function and theta dimension
   if(error == 'gaussian'){
     likelihood <- nlogDensity
@@ -110,11 +80,20 @@ singleMCMCallMH <- function(data, reps, draw, hyper, thin = 1, error = 'gaussian
   saveDraws <- matrix(0, nSave, dim)
[…]

### 2017-12-07 Cars/paper.tex
no gaps sampler + other work

@@
 \item Self driving cars becoming more common
-\item Cars are easy capable of driving by themselves if there is no traffic around them
-\item Cars need to be able to predict movement of other cars
+\item Cars can perform complex maneuvers on their own, and are capable of detecing the location of cars around them
+\item Cars need to improve their ability to detect the movement of cars around them
 \item Wide range of unconditional variances and first two partial autocorrelations observed while fitting each model individually - evidence of heterogeneity
-\item Drivers display differing behaviour, could be able to group drivers into different categories
-\item Heteogeneity between and within groups
-\item Need to deal with new cars as they are observed
+\item Forecasts of car movement could be improved by building a hierarchical model that shares information between cars while allowing each driver to have their own parameter vector
+\item As new cars are observed we need to make parameter estimates as quickly as possible
+\item Strong prior information from the hierarchical model can be a substitute for the lack of data in the first few seconds of observing a new car
+\item As time passes this original estimation needs to be updated as the stream of data becomes large enough for the prior impact to be reduced
+\item We would like to be able to do this using only the newly observed data, instead of re-estimating a model re-using the old data.
+\item The aim of this paper is to make efficient forecasts by:
+\begin{enumerate}
+\item Using prior information when available
+\item Update the posterior as new data is observed fast enough to be viable
+\end{enumerate}
+\item We expect that the prior information will make a significant difference early on in the sample, but there will be a significant gain in forecast accuracy as more data increases.
+At this point the hierarchical prior isn't too useful, but we need to be able to fit the model in time!
+\item At the end of the day we will have a lot of forecast evaluations from MCMC, VB and a new VB stream method, a non mean-field adaption of {\v S}m{\'i}dl 2006
+\item We can compare the performance of different levels of prior information, and different levels of approximation error in the forecasts
 \end{itemize}
@@
 \begin{itemize}
-\item Modern cars are able to track lane marking and can calculate distance to other cars in relative co-ordinates (Thuy and Leon 2010)
-\item Data from NGSIM is in coordinates relative to start of road - no impact of the curvature of the lane
-\item Fit curve to car data to estimate lane midpoints (Woo et. al. 2016) 
+\item Data is provided by the NGSIM project, which observes ~5000 cars for a 2000 feet section of the US 101 Highway.
+\item Observations of x/y coordinates are provided every 100 ms (10 hz), with around 500-1000 observations per car.
+\item Around 2000 of these cars changed lanes or stopped moving somewhere in the sample and are discarded.
+\item There was some curvature in the road which means there are systematic movements in the car paths.
+\item Modern cars are able to track lane marking and can calculate distance to other cars in relative co-ordinates (Thuy and Leon 2010), instead of relative to the Earth (ie. ignores curves)
+\item Prior research fits a polynomial curve to detected lane markings to build a model of the road lane edges (Woo et. al. 2016) 
+\item Instead, I fit splines to the positions of each car to build an estimate of the road midpoint.
+Relative coordiantes are found separately for each lane using the movements of 100 cars / lane by the following procedure:
 \begin{enumerate}
@@
 \item Estimate smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
-\item Use polynomials to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
+\item Use models to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
 \item Define new coordinates 
@@
 \end{enumerate}
-\item Note: This should be calculated separately for each lane, cars used to estimate splines are excluded from further modelling
+\item These cars are not used for any further modelling, leaving 2873 cars travelling on a straightened out road with no noticable systematic curvature.
+(ie.
+no patterns in many cars turning at the same spot)
 \end{itemize}
@@
 
-The position of car $i$ at time $t$ can be fully determined by its inital position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the cars velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
+The position of any vechicle at time $t$ can be fully determined by its inital position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the cars velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
 \begin{align}
@@
 \end{align}
-where $\delta_{t} = \pi/2$ denotes that the car has no lateral movement and $x^*_{t} = x^*_{t-1}$.
-From this relationship, the key inputs can be extracted by
+where $\delta_{t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{t} = x^*_{t-1}$.
+From this relationship, the driver's key inputs to motion can be extracted by
 \begin{align}
@@
 
-Auto-regressive processes of orders $p$ and $q$ for $a_i$ and  $\delta_i$ are used to model the level of persistance and variation of driver actions, with
+Auto-regressive processes of orders $p$ and $q$ for $a_t$ and  $\delta_t$ are used to model the level of persistance and variation of driver actions, with
 \begin{align}
@@
 \begin{itemize}
+\item This model is fit to 2000 cars in the sample, leaving 873 for forecasting.
 \item To do: Add a Direchlet Process Prior to make this an infinite mixture
-\item I've been trying to do this but I get results where all of the component indicator variables in the MCMC go to one component instead of the mixture assigning as many clusters as possible as Tas suggested would happen.
-I assume I'm doing something wrong.
+\item I've been trying this but I get results where all of the component indicator variables in the MCMC go to one component instead of the mixture assigning as many clusters as possible as Tas suggested would happen.
+I assume I'm doing something wrong but I haven't figured out what yet.
 \end{itemize}
@@
 \section{Streaming Data Process}
-
-Notation:
-\begin{itemize}
-\item $q(\theta_i | y_{i, 1:T}) \approx p(\theta_i | y_{i, 1:T}, y_{1:N})$, $\lambda$ parameters and dependence on other cars suppressed - more important to be clear about which observations from the new vehicle are used to construct the approximation.
-\end{itemize}
 
@@
 \end{equation}
-\item As N is large replace $ p(\theta_{N+1} | \beta) p (\beta | y_{1:N})$ with $ p(\theta_{N+1} | \beta_{MAP})$ where $\beta_{MAP}$ is the maximum of $p(\beta | y_{1:N})$, or some other suitable point estimate.
-This approximate distribution now has a closed form.
-\item Subsequently fitting $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})  \propto p(y_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \beta_{MAP})$ is fairly easy with MCMC / VB etc.
+\item As N is large $p (\beta | y_{1:N})$ is fairly precise, so replace $ p(\theta_{N+1} | \beta) p (\beta | y_{1:N})$ with $ p(\theta_{N+1} | \hat{\beta})$ where $\hat{\beta}$ is the posterior mean of $\beta$.
+This approximation gives us a closed form for $ p(\theta_{N+1} | \hat{\beta})$.
+\item Subsequently fitting $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})  \propto p(y_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \hat{\beta})$ is fairly easy with MCMC / VB etc.
 \item When $S$ is small, the information in the prior will dominate the individual information.
@@
 \begin{enumerate}
-\item Observe $y_{N+1, 1:S_1}$, fit VB $q(\theta_{N+1} | y_{N+1, 1:S_1}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_1})$
-\item Observe $y_{N+1, S_1+1:S_2}$, in this case the model becomes
+\item Observe $y_{N+1, 1:S}$, fit VB $q(\theta_{N+1} | y_{N+1, 1:S}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$
+\item Observe $y_{N+1, S+1:T}$, in this case the model becomes
 \begin{equation}
-p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_2}) \propto p(y_{N+1, S_1+1:S_2} | \theta_{N+1})p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S_1})
+p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T}) \propto p(y_{N+1, S+1:T} | \theta_{N+1})p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S})
 \end{equation}
-\item Set $p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S_1}) = q(\theta_{N+1} | y_{N+1, 1:S_1})$
-\item Fit VB  $q(\theta_{N+1} | y_{N+1, 1:S_2}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S_2})$
+\item Set $p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S}) = q(\theta_{N+1} | y_{N+1, 1:S})$
+\item Fit VB  $q(\theta_{N+1} | y_{N+1, 1:T}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T})$
 \item Repeat as new data is observed
 \end{enumerate}
+\item The performance of this approach should be similar to (but probably slightly lower than) the regular VB fit but it is faster as $T$ increases relative to $T - S$.
 \end{itemize}
@@
 \begin{enumerate}
-\item No prior information for the new car
+\item No prior information for the new car - very flat priors.
 \item Prior information from hierarchical model with one component
@@
 \item MCMC
-\item Regular VB (fit to full sample)
+\item Regular VB (fit each update to the new sample)
 \item Streaming VB (fit to new data only)
@@
 \item Compare forecasts for different sample sizes and forecast horizons
-\item Strategy: Fit model to one $S = 10$ (one second), forecast next second
-\item Fit model to $S = 2$, forecast next second, repeat.
+\item Strategy: Fit model for one car up time time $S = 10$ (one second), forecast from $S+1:S+30$ (three seconds seems to be a standard forecast horizon for driver reaction, though surely self driving cars have a faster reaction time).
+\item Fit model to $S = 20$, forecast next three seconds, repeat.
 \item Collect logscores for predictive densities marginalised over the posterior distributions.
+\item Repeat for each of the 873 remaining cars.
 \end{itemize}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width = 0.95\textwidth]{logscoresByPrior.png}
+\caption{Predictive log-scores for $p(a_{T+1:T+10}, d_{T+1:T+10} | a_{1:T}, d_{1:T})$ for $T = \{10, 20, \dots, 300\}$ for each of the three levels of prior information fit by both MCMC and VB.
+Stronger prior information improves forecasts in small samples, but as $T$ increases there are clear benefits to refitting the model regardless of the level of prior information.
+There was a mixtake in the VB finite mixture code which has been corrected but I am still waiting for the fixed results.}
+\label{fig:MainResults}
+\end{figure}
 
@@
 \begin{itemize}
-\item Models based on the stromg prior information to outperform no information models for first few seconds, where we see a significant improvement in the log scores.
-As the dataset increases to > 100 observations (10 seconds), the data dominates and the prior choice does not have much impact.
+\item Models based on the strong prior information outperform no information models for first few seconds, where we see a significant improvement in the log scores.
+As the dataset increases to $> 100$ observations (10 seconds), the data dominates and the prior choice does not have much impact.
 The median logscore (and probably other quantiles, hard to see from the graph) at a larger value of T increases as expected, showing that it is important to re-estimate the model with an increased dataset.
-However it seems that this re-estimation could probably be done with a simpler prior structure than the full hierarchy.
+However it seems that this re-estimation could probably be done with a simpler prior structure than the full hierarchy without any big loss.
 \item Compare streaming data VB - Hoping it will be comparable to other methods, it technically may be a bit worse as it uses one more approximation than the standard VB, but I don't think this will have a very big impact.
@@
 \item Still waiting on the cluster for standard VB Finite Mixture Model and all VB streams.
+\item We will have multiple forecasts for each actual observation (eg.
+$p(y_{43} | y_{20}), p(y_{43} | y_{30})$ and $p(y_{43} | y_{40})$) which gives us another approach to look at the benefits of increased observed data.
 \end{itemize}

### 2017-12-07 Cars/HierMixMod.R

@@ -7,8 +7,10 @@
 
 # For the supervisors:
 # Write more things.
-# Set a meeting at some stage
 
+# Slurm stuff
+# Refit failed models
+# Refit mixture VB to all cars
 
 library(tidyverse)
 library(Rcpp)
@@ -606,18 +608,36 @@ for(i in seq_along(id$idfc)){
 }
 
 results {
-  
+
 results <- NULL
 for(i in seq_along(id$idfc)){
-  tmp <- read.csv(paste0('eval/car', id$idfc[[i]], '.csv'))
-  results <- rbind(results, tmp)
-  tmp <- read.csv(paste0('evalMix/car', id$idfc[[i]], '.csv'))
-  results <- rbind(results, tmp)
+  #tmp <- read.csv(paste0('eval/car', id$idfc[i], '.csv'))
+  #results <- rbind(results, tmp)
+  #tmp <- read.csv(paste0('evalMix/car', id$idfc[i], '.csv')) 
+  #results <- rbind(results, tmp)
+  try(assign('results',
+             rbind(results,
+                   read.csv(paste0('evalAll/car', id$idfc[i], '.csv')))))
   if(i %% 100 == 0){
     print(i)
   }
 }
   
+results %>%
+  #filter(method == 'MCMC' & prior == 'None') %>%
+  mutate(T = S + h,
+         horizon = ceiling(h/10)) %>%
+  group_by(method, prior, S, h, T, horizon, id) %>%
+  summarise(ls = - sum(logscore)) %>%
+  ungroup() %>%
+  select(T, id, ls, horizon, method, prior) %>%
+  filter(T > 30) %>%
+  ggplot() + geom_boxplot(aes(factor(horizon), ls)) +
+  facet_grid(method ~ prior) + 
+  scale_y_continuous(trans = reverselog_trans(base=10),
+                     labels=trans_format("identity", function(x) -x))
+
+
 library(scales)
 reverselog_trans <- function(base = exp(1)) {
   trans <- function(x) -log(x, base)
@@ -632,7 +652,8 @@ results %>%
   summarise(ls = -sum(logscore)) %>%
   ungroup() %>%
   mutate(T = ceiling(S / 30),
-         prior = factor(prior, levels = c('None', 'Hierarchy', 'Finite Mixture'))) %>%
+         prior = ifelse(as.character(prior) == 'Hierarchy', 'Single Component', as.character(prior)),
[…]

### 2017-12-07 Cars/mixtureMCMC.R

@@ -519,3 +519,311 @@ singleMCMCallMH <- function(data, reps, draw, hyper, thin = 1, error = 'gaussian
   }
   list(draws = saveDraws, accept = accept, steps = stepsize)
 }
+
+NoGaps <- function(data, reps, draw, hyper, thin = 1, k = 10, stepsize = 0.01){
+  N <- length(data)
+  stepsize <- rep(stepsize, N)
+  accept <- rep(0, N)
+
+  #set up storage for saved draws
+  nSave <- floor(reps / thin)
+  save <- list(list())
+  dim <- 6
+  for(i in 1:k){
+    save[[1]][[i]] <- list(mean = matrix(0, nSave, dim), varInv = array(0, dim = c(dim, dim, nSave)))
+  }
+  for(i in 1:N){
+    save[[i+1]] <- list(theta = matrix(0, nSave, dim), s = rep(0, nSave))
+  }
+  
+  # changing MH acceptance rate
+  alpha <- - qnorm(0.234/2)
+  stepsizeCons <- (1 - 1/6) * sqrt(2*3.141598) * exp(alpha^2/2) / (2 * alpha)  +  1 / (6 * 0.234 * (1 - 0.234))
+  
+  # initialise groups randomly
+  s <- sample(1:k, N, replace = TRUE)
+  n <- table(s)
+  
+  for(iter in 2:reps){
+    # timing
+    if(iter == 50){
+      startTime <- Sys.time()
+    } else if(iter == 150){
+      timePerIter <- (Sys.time() - startTime) / 100
+      class(timePerIter) <- 'numeric'
+      print(paste0('Estimated Finishing Time: ', Sys.time() + timePerIter * (reps - 150)))
+      if(attr(timePerIter, 'units') == 'mins'){
+        attr(timePerIter, 'units') = 'secs'
+        timePerIter <- timePerIter * 60
+      }
+    }
+    
+    # Resample s indices for each i
+    # First sample a k+1'th mu / precision for any new groups
+    draw[[1]][[k+1]] <- list(mean = c(mvtnorm::rmvnorm(1, hyper$mean, solve(hyper$varInv))),
+                             varInv = rWishart(1, hyper$df, hyper$scale)[,,1])
+    save[[1]][[k+1]] <- list(mean = matrix(0, nSave, dim), varInv = array(0, dim = c(dim, dim, nSave)))
+
+    for(j in 1:N){
+      skip = FALSE
[…]

### 2017-12-07 Cars/mixtureMCMCMetHast.cpp

@@ -20,6 +20,20 @@ double nlogDensity (mat data, vec theta, vec mu, mat varInv){
   }
   return dens;
 }
+
+// [[Rcpp::export]]
+double nlikelihood (mat data, vec theta){
+  int T = data.n_rows;
+  double dens = 0;
+  for(int t = 2; t < T; ++t){
+    dens +=  - 0.5 * (theta(0) + theta(1)) - 
+      pow(data(t, 0) - theta(2) * data(t-1, 0) - theta(3) * data(t-2, 0), 2) / (2 * exp(theta(0))) -
+      pow(data(t, 1) - theta(4) * data(t-1, 1) - theta(5) * data(t-2, 1), 2) / (2 * exp(theta(1)));
+  }
+  return dens;
+}
+
+
   
   
 // [[Rcpp::export]]
@@ -35,3 +49,51 @@ double tlogDensity (mat data, vec theta, vec mu, mat varInv){
   }
   return dens;
 }
+
+// [[Rcpp::export]]
+double mvtDens (vec theta, vec mu, mat varInv){
+  double logDens = 0.5 * log(det(varInv)) - 0.5 * as_scalar((theta - mu).t() * varInv * (theta - mu));
+  return exp(logDens);
+}
+
+// [[Rcpp::export]]
+List NoGapsMH (List data, List draw, vec stepsize, vec accept, double stepsizeCons, vec hyperMean, mat hyperVarInv, vec s, int iter){
+  int k = draw.length(), N = data.length();
+  
+  for(int i = 0; i < k; ++i){
+    vec oldDraw = draw(i);
+    vec candidate = oldDraw  +  stepsize(i) * randn<vec>(6)[0];
+    double canDens = 0, oldDens = 0;
+    for(int j = 0; j < N; ++j){
+      if(s(j) == i + 1){
+        canDens += nlogDensity(data(i), candidate, hyperMean, hyperVarInv);
+        oldDens += nlogDensity(data(i), oldDraw, hyperMean, hyperVarInv);
+      }
+    }
+    double ratio = exp(canDens - oldDens), u = randu<vec>(1)[0];
+    if(u < ratio){
+      accept(i) += 1;
+      draw(i) = candidate;
+      stepsize(i) += stepsizeCons * stepsize(i) * (1 - 0.234) / (28 + iter);
+    } else {
+      stepsize(i) -= stepsizeCons * stepsize(i) * 0.234 / (28 + iter);
+    }
+  }
[…]

### 2017-12-11 Cars/paper.tex
no gaps

@@
 \item Self driving cars becoming more common
-\item Cars can perform complex maneuvers on their own, and are capable of detecing the location of cars around them
+\item Cars can perform complex maneuvers on their own, and are capable of detecting the location of cars around them
 \item Cars need to improve their ability to detect the movement of cars around them
@@
 \begin{itemize}
-\item Data is provided by the NGSIM project, which observes ~5000 cars for a 2000 feet section of the US 101 Highway.
+\item Data is provided by the Next Generation Simulation (NGSIM) project, which observes ~5000 cars for a 2000 feet section of the US 101 Highway.
 \item Observations of x/y coordinates are provided every 100 ms (10 hz), with around 500-1000 observations per car.
@@
 \item Instead, I fit splines to the positions of each car to build an estimate of the road midpoint.
-Relative coordiantes are found separately for each lane using the movements of 100 cars / lane by the following procedure:
+Relative co-ordinates are found separately for each lane using the movements of 100 cars / lane by the following procedure:
 \begin{enumerate}
 \item At each point in time $t$ calculate the total distance each car has travelled along the road using $d_{i, t} = \sum_{s=1}^t v_{i, s}$, where $v_{i, s}$ is the velocity of car $i$ at time $s$. 
-\item Estimate smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
+\item Estimate independent smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
 \item Use models to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
@@
 
-The position of any vechicle at time $t$ can be fully determined by its inital position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the cars velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
+The position of any vehicle at time $t$ can be fully determined by its inital position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the cars velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
 \begin{align}
@@
 \end{align}
-where $\epsilon_{t}$ and $\eta_t$ are independently and identically distributed according to a standard normal distribution.
-The parameter vector to be estimated is collected as $\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\} \in \mathbb{R}^{p + q + 2}$.
+where $\epsilon_{t}$ and $\eta_t$ are independently and identically distributed according to a standard normal distribution. The parameter vector to be estimated is collected as 
+\begin{equation*}
+\label{thetaVec}
+\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\} \in \mathbb{R}^{p + q + 2}.
+\end{equation*}
 
@@
 \item This model is fit to 2000 cars in the sample, leaving 873 for forecasting.
-\item To do: Add a Direchlet Process Prior to make this an infinite mixture
+\item To do: Add a Dirichlet Process Prior to make this an infinite mixture
 \item I've been trying this but I get results where all of the component indicator variables in the MCMC go to one component instead of the mixture assigning as many clusters as possible as Tas suggested would happen.
@@
 \end{equation}
-\item Due to the lack of a closed form expression for $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$, in reality the updating begins again from the prior re-using the data in $y_{N+1, 1:S}$.
+\item To perform this update only using the data from $y_{N+1, S+1:T}$ we would need a closed form form expression for $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$.
+With MCMC methods we only have a sample from this distribution so in reality the distribution isn't updated, instead we re-run the MCMC using the complete dataset using the prior $p(\theta_{N+1} | \hat{\beta})$.
 \item Due to the demands of driving, the car wants the update carried out as soon as possible, and ideally should be able to discard this previously used information and use only $y_{N+1, S+1:T}$, reducing the time and memory complexity from $O(T)$ to $O(T-S)$.

### 2017-12-11 Cars/mixtureMCMC.R

@@ -556,6 +556,15 @@ NoGaps <- function(data, reps, draw, hyper, thin = 1, k = 10, stepsize = 0.01){
         attr(timePerIter, 'units') = 'secs'
         timePerIter <- timePerIter * 60
       }
+    } else if(iter %% 100 == 50){
+      timePerIter <- (Sys.time() - startTime) / (iter - 50);
+      if(attr(timePerIter, 'units') == 'mins'){
+        attr(timePerIter, 'units') = 'secs'
+        timePerIter <- timePerIter * 60
+      } else if(attr(timePerIter, 'units') == 'hours'){
+        attr(timePerIter, 'units') = 'secs'
+        timePerIter <- timePerIter * 3600
+      }
     }
     
     # Resample s indices for each i
@@ -563,6 +572,7 @@ NoGaps <- function(data, reps, draw, hyper, thin = 1, k = 10, stepsize = 0.01){
     draw[[1]][[k+1]] <- list(mean = c(mvtnorm::rmvnorm(1, hyper$mean, solve(hyper$varInv))),
                              varInv = rWishart(1, hyper$df, hyper$scale)[,,1])
     save[[1]][[k+1]] <- list(mean = matrix(0, nSave, dim), varInv = array(0, dim = c(dim, dim, nSave)))
+    
 
     for(j in 1:N){
       skip = FALSE
@@ -577,12 +587,13 @@ NoGaps <- function(data, reps, draw, hyper, thin = 1, k = 10, stepsize = 0.01){
             # Otherwise shift everything above it down by one and put the now empty cluster in the k'th spot
             temp <- draw[[1]][[s[j]]]
             tempSave <- save[[1]][[s[j]]]
-            for(i in s[j]:(k-1)){
-              draw[[1]][[i]] <- draw[[1]][[i+1]]
-              save[[1]][[i]] <- save[[1]][[i+1]]
-            }
+            
+            draw[[1]][s[j]:(k-1)] <- draw[[1]][(s[j]+1):k]
+            save[[1]][s[j]:(k-1)] <- save[[1]][(s[j]+1):k]
+            
             draw[[1]][[k]] <- temp
             save[[1]][[k]] <- tempSave
+            
             s[s > s[j]] <- s[s > s[j]] - 1
             s[j] <- k
             n <- table(s)
[…]

### 2017-12-11 Cars/mixtureMCMCMetHast.cpp

@@ -79,7 +79,7 @@ List NoGapsMH (List data, List draw, vec stepsize, vec accept, double stepsizeCo
       stepsize(i) -= stepsizeCons * stepsize(i) * 0.234 / (28 + iter);
     }
   }
-  return List::Create(Named("Draws") = draws,
+  return List::create(Named("Draws") = draw,
                       Named("Stepsize") = stepsize,
                       Named("Accept") = accept);
 }

### 2018-02-05 Cars/paper.tex
holiday work

@@
 \begin{itemize}
-\item Self driving cars becoming more common
-\item Cars can perform complex maneuvers on their own, and are capable of detecting the location of cars around them
-\item Cars need to improve their ability to detect the movement of cars around them
-\item Wide range of unconditional variances and first two partial autocorrelations observed while fitting each model individually - evidence of heterogeneity
-\item Forecasts of car movement could be improved by building a hierarchical model that shares information between cars while allowing each driver to have their own parameter vector
-\item As new cars are observed we need to make parameter estimates as quickly as possible
-\item Strong prior information from the hierarchical model can be a substitute for the lack of data in the first few seconds of observing a new car
-\item As time passes this original estimation needs to be updated as the stream of data becomes large enough for the prior impact to be reduced
-\item We would like to be able to do this using only the newly observed data, instead of re-estimating a model re-using the old data.
-\item The aim of this paper is to make efficient forecasts by:
+\item Self driving Cars are becoming more common, eg Waymo, Tesla, GM.
+\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the Self Driving Car.
+\item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg.
+Cameras and LiDAR), such as their position, velocity, and steering angle (Woo et al.
+2016a; Tian, Pei, Jana, and Ray 2017).
+\item The Self Driving Car forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
+\item These can assume the surrounding vehicles will maintain their current angle and velocity (Gindele, Brechtel, Dillmann 2010; Waymo Safety Report 2017; Bautista 2017), but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines (Woo et al.
+201b6; Tomar, Verma, and Tomar 2014; Woo et al.
+2017; Geng et al.
+2017).
+\item This paper takes a statistical approach, fitting time series models to two extracted variables, velocity and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
+\item It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a trajectory forecast.
+\item These time series models (In this case a bivariate AR2) fit a parameter vector to capture the driver behaviour, which can be very different from driver to driver (Chen 2017).
+\item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
+\item This model implies that any new vehicle observed by the self driving car has its own parameter vector, related to previously observed vehicles through the shared prior.
+\item The self driving car is constantly observing additional information about the vehicles around it.
+Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
+\item A three step strategy is employed to forecast the future trajectory of newly observed vehicles:
 \begin{enumerate}
-\item Using prior information when available
-\item Update the posterior as new data is observed fast enough to be viable
+\item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution.
+This can occur before the self driving car is on the road.
+\item When a new vehicle is observed use the information in this shared distribution to model the associated new parameter vector. 
+\item Update the parameter vector and all forecasts at short intervals.
 \end{enumerate}
-\item We expect that the prior information will make a significant difference early on in the sample, but there will be a significant gain in forecast accuracy as more data increases.
-At this point the hierarchical prior isn't too useful, but we need to be able to fit the model in time!
-\item At the end of the day we will have a lot of forecast evaluations from MCMC, VB and a new VB stream method, a non mean-field adaption of {\v S}m{\'i}dl 2006
-\item We can compare the performance of different levels of prior information, and different levels of approximation error in the forecasts
+\item This paper introduces a Variational Bayes updating method for the third step, where the posterior distribution of the new vehicles parameter vector is approximated, then this approximation is updated.
+\item This only uses the data observed in-between updates, allowing the previously observed history of that vehicle to be discarded.
+\item This is important for two reasons:
+\begin{enumerate}
+\item The model can be updated faster as less data must be processed, so forecasts can be made using as much recent data as possible.
+\item Self driving cars take in up to 1GB of data a second (several websites report this as quotes from different people eg.
+Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.
+\end{enumerate}
+\item Compare Bayes MAP point estimates to a neural network and constant acceleration / angle model.
+\item Error Metric: Euclidean Distance from predicted position to actual position
+\item Compare log scores from streaming and standard VB forecast densities to see how the updating method compares to standard VB.
+I am also using MCMC, so we can compare either VB method to this, theoretically MCMC > VB Standard > VB Updating
 \end{itemize}
@@
 \begin{align}
-x^*_{i, t} &= \mbox{sign}(x_{i, t} - \hat{x}_{i, t})\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
+x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right) - \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
 y^*_{i, t} &= d_{i, t} \label{yRel}
@@
 \end{enumerate}
+\item (\ref{xRel}) \textit{really} needs a diagram.
 \item These cars are not used for any further modelling, leaving 2873 cars travelling on a straightened out road with no noticable systematic curvature.
-(ie.
-no patterns in many cars turning at the same spot)
 \end{itemize}
@@
 
-The position of any vehicle at time $t$ can be fully determined by its inital position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the cars velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
+The position of any vehicle at time $t$ can be fully determined by its initial position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the vehicle's velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
 \begin{align}
@@
 \end{align}
-where $\epsilon_{t}$ and $\eta_t$ are independently and identically distributed according to a standard normal distribution. The parameter vector to be estimated is collected as 
+where $\epsilon_{t}$ and $\eta_t$ are independently and identically distributed according to a standard normal distribution.
+The unconditional means of $a_t$ and $\delta_t$ are set to zero and $\pi/2$ respectively, corresponding to forward motion at constant velocity.
+The parameter vector to be estimated is collected as
 \begin{equation*}
@@
 
-Multiple groups of differing driver behaviour can be allowed while still retaining heterogeneity with each group by jointly estimating each vehicle driver's individual parameter vector, $\theta_i$, as part of a heirarchical model, independently drawing $\theta_i$ from a $K$ component mixture of multivariate normal distributions, augmenting each $\theta_i$ with a mixture component $k_i = 1, \dots, K$ such that
+Multiple groups of differing driver behaviour can be allowed while still retaining heterogeneity within each group by jointly estimating each vehicle driver's individual parameter vector, $\theta_i$, as part of a heirarchical model.
+In this case each $\theta_i$ is treated as an independent draw from a $K$ component mixture of multivariate normal distributions, augmenting each $\theta_i$ with a mixture component $k_i = 1, \dots, K$ such that
 \begin{equation}
@@
 \end{equation}
-For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with degrees of freedom $\tau_j$ and scale matrix $\Psi_j$.
-$k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi_i} = \{\pi_{1, i}, \dots, \pi_{K, i}\}$, where $\boldsymbol{\pi_i}$ has a Dirichlet prior with parameters $\alpha_1, \alpha_2, \dots, \alpha_K$.
-Defining $\beta = \{\mu_i, \Sigma_i, i = 1, \dots, K\}$, $y_{i, 1:T} = \{\delta_{i, s}, a_{i, s} | s = 1, \dots, T\}$ and $y_{1:N} = \{y_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | y_{1:N})$ can be estimated with MCMC methods.
+For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with fixed mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with fixed degrees of freedom $\tau_j$ and scale matrix $\Psi_j$.
+$k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi_i} = \{\pi_{1, i}, \dots, \pi_{K, i}\}$, where $\boldsymbol{\pi_i}$ has a Dirichlet prior with fixed hyperparameters $\alpha_1, \alpha_2, \dots, \alpha_K$.
+Defining $\beta = \{\mu_i, \Sigma_i, i = 1, \dots, K\}$, $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s} | s = 1, \dots, T\}$ and $z_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | z_{1:N})$ can be estimated with MCMC methods.
 
-\begin{itemize}
-\item This model is fit to 2000 cars in the sample, leaving 873 for forecasting.
-\item To do: Add a Dirichlet Process Prior to make this an infinite mixture
-\item I've been trying this but I get results where all of the component indicator variables in the MCMC go to one component instead of the mixture assigning as many clusters as possible as Tas suggested would happen.
-I assume I'm doing something wrong but I haven't figured out what yet.
-\end{itemize}
+\section{Variational Bayes}
 
-\section{Streaming Data Process}
+\subsection{Background}
+
+Variational Bayes approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
+\begin{equation}
+\label{KL-div}
+KL(q(\theta) || p(\theta | z) = \int_{\theta} q(\theta) \left(\log(q(\theta)) - \log(p(\theta | z))\right) d\theta.
+\end{equation}
+As $p(\theta | z)$ is typically intractable, VB maximises $\mathcal{L}(q)$ where
+\begin{equation}
+\label{ELBO}
+\mathcal{L}(q) = \int_{\theta} q(\theta) \left(\log(p(\theta, z)) - \log(q(\theta))\right) d\theta
+\end{equation}
+which is equivalent to minimising (\ref{KL-div}).
+This is done with stochastic gradient ascent, taking Monte Carlo estimates of $d\mathcal{L}(q) / d \lambda$ where $\lambda$ is the parameter vector of $q(\theta)$.
+There are two Monte Carlo estimators of $d\mathcal{L}(q) / d \lambda$, the Score Estimator, 
+\begin{equation}
+\label{scoreDeriv}
+\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^L \frac{d \log(q(\theta_i))}{d \lambda} \left(\log(p(\theta_i, z)) - \log(q(\theta_i)) \right), \mbox{ where } \theta_i \sim q(\theta),
+\end{equation}
+and the Reparameterised Estimator,
+\begin{equation}
+\label{rpDeriv}
+\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^M \frac{d f(\lambda, \epsilon_i)}{d \lambda} \frac{d \log(p(\theta_i, z))}{d \theta_i} + \frac{d J(\lambda, \epsilon_i)}{d \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
+\end{equation}
+where $r(\epsilon)$ is some distribution without any free parameters (eg.
+Uniform, Standard Normal), $f(\lambda, \epsilon)$ is some differentiable function that transforms $\epsilon$ to $\theta$, and $J(\lambda, \epsilon)$ is the Jacobian Matrix of this function.
+(\ref{rpDeriv}) typically has significantly lower variance than (\ref{scoreDeriv}), allowing $M$ to be set lower than $L$ which improves the speed of the algorithm; however the requirement for $f$ to exist restricts the class of approximating distributions $q(\theta)$ that can be used.
+Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the derivative estimator (\ref{scoreDeriv}) must be used.
+To reduce $L$ and thus the computational cost of the algorithm, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$.
+Using RQMC allows (\ref{scoreDeriv}) to converge to the true derivative at a rate $O(L)$ compared to $O(\sqrt{L})$ in standard Monte Carlo.
+
+\subsection{Updating VB Process}
 
@@
 \begin{equation}
-p(\theta_{N+1} | y_{1:N+1}) \propto p(y_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \beta) p (\beta | y_{1:N})
+\label{originalPost}
+p(\theta_{N+1} z_{N+1} | z_{1:N}) \propto p(z_{N+1} | \theta_{N+1}) \int_{\beta} p(\theta_{N+1} | \beta) p (\beta | z_{1:N}) d\beta
 \end{equation}
-\item As N is large $p (\beta | y_{1:N})$ is fairly precise, so replace $ p(\theta_{N+1} | \beta) p (\beta | y_{1:N})$ with $ p(\theta_{N+1} | \hat{\beta})$ where $\hat{\beta}$ is the posterior mean of $\beta$.
+\item As N is large $p (\beta | z_{1:N})$ is fairly precise, so we can replace $\int_{\beta} p(\theta_{N+1} | \beta) p (\beta | z_{1:N}) d\beta$ with $ p(\theta_{N+1} | \hat{\beta})$ where $\hat{\beta}$ is the posterior mean of $p(\beta | z_{1:N})$.
 This approximation gives us a closed form for $ p(\theta_{N+1} | \hat{\beta})$.
-\item Subsequently fitting $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})  \propto p(y_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \hat{\beta})$ is fairly easy with MCMC / VB etc.
+\item Subsequently fitting $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})  \propto p(z_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \hat{\beta})$ is fairly easy with MCMC / VB etc.
 \item When $S$ is small, the information in the prior will dominate the individual information.
-\item However, for some $T > S$ the information in the car's individual movements should become more relevant and forecasts could be improved by updating the distribution to $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T})$.
-\item This update takes the form of
+\item However, for some $T > S$ the information in the car's individual movements should become more relevant and forecasts could be improved by updating the distribution to $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:T})$.
+\item This updated posterior is given by
 \begin{equation}
-p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T}) \propto p(y_{N+1, S+1:T} | \theta_{N+1}) p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})
+\label{updatePost}
+p(\theta_{N+1} | z_{N+1, 1:T}, z_{1:N}) \propto p(z_{N+1, S+1:T} | \theta_{N+1})p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})
 \end{equation}
-\item To perform this update only using the data from $y_{N+1, S+1:T}$ we would need a closed form form expression for $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$.
-With MCMC methods we only have a sample from this distribution so in reality the distribution isn't updated, instead we re-run the MCMC using the complete dataset using the prior $p(\theta_{N+1} | \hat{\beta})$.
-\item Due to the demands of driving, the car wants the update carried out as soon as possible, and ideally should be able to discard this previously used information and use only $y_{N+1, S+1:T}$, reducing the time and memory complexity from $O(T)$ to $O(T-S)$.
-\item Defining $q(\theta_{N+1} | y_{N+1, 1:S})$ as the Variational Bayes approximation of $p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$, an online method is proposed to update $q(\theta_{N+1} | y_{N+1, 1:S})$ to $q(\theta_{N+1} | y_{N+1, 1:T})$:
-\begin{enumerate}
-\item Observe $y_{N+1, 1:S}$, fit VB $q(\theta_{N+1} | y_{N+1, 1:S}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:S})$
-\item Observe $y_{N+1, S+1:T}$, in this case the model becomes
+\item Both $\mathcal{L}(q)$ derivative estimators, (\ref{scoreDeriv}) and (\ref{rpDeriv}), need to evaluate \newline $\log(p(\theta_{N+1}, z_{N+1, S+1:T} | z_{1:N}, z_{N+1, 1:S}))$, which takes the form
+\begin{align}
+\log(p(\theta_{N+1}, z_{N+1, S+1:T} | z_{1:N}, z_{N+1, 1:S})) &= \log(p(z_{N+1, S+1:T} | \theta_{N+1})) \nonumber \\
+&+ \log(p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})) \label{updateEq}
+\end{align}
+\item To perform this update only using the data from $z_{N+1, S+1:T}$ we would need a closed form form expression for $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$ to replace the role of the prior.
+Without this closed form we would need to re-fit the VB model using the original prior $p(\theta_{N+1} | \hat{\beta})$ and the whole observed sequence $z_{N+1, 1:T}$.
+\item Due to the demands of driving, the car wants the update carried out as soon as possible, and ideally should be able to discard the previously used information $z_{N+1, 1:S}$. 
+\item Defining $q(\theta_{N+1} | z_{N+1, 1:S})$ as the Variational Bayes approximation of $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$, a new method is proposed to update $q(\theta_{N+1} | z_{N+1, 1:S})$ to $q(\theta_{N+1} | z_{N+1, 1:T})$ after observing $z_{N_1, S+1:T}$.
+\item By replacing $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$ in (\ref{updateEq}) with the previously fit $q(\theta_{N+1} | z_{N+1, 1:S})$ we get an approximation to (\ref{updateEq}):
 \begin{equation}
-p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T}) \propto p(y_{N+1, S+1:T} | \theta_{N+1})p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S})
+\label{updateEqApprox}
+\log(p(\theta_{N+1}, z_{N+1, 1:T} | z_{1:N})) \approx \log(p(z_{N+1, S+1:T} | \theta_{N+1})) + \log(q(\theta_{N+1} | z_{N+1, 1:S}))
 \end{equation}
-\item Set $p(\theta_{N+1} |  y_{1:N}, y_{N+1, 1:S}) = q(\theta_{N+1} | y_{N+1, 1:S})$
-\item Fit VB  $q(\theta_{N+1} | y_{N+1, 1:T}) \approx p(\theta_{N+1} | y_{1:N}, y_{N+1, 1:T})$
-\item Repeat as new data is observed
-\end{enumerate}
-\item The performance of this approach should be similar to (but probably slightly lower than) the regular VB fit but it is faster as $T$ increases relative to $T - S$.
+\item Substituting (\ref{updateEqApprox}) into (\ref{scoreDeriv}) gives the Score Estimator for updating Variational Bayes, 
+\begin{align}
+\label{scoreUpdate}
+\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^L \frac{d \log(q(\theta_{N+1} | z_{N+1, 1:T}))}{d \lambda} &\left(\log(q(\theta_{N+1} | z_{N+1, 1:S}) - \log(q(\theta_{N+1} | z_{N+1, 1:T})) \right), \nonumber \\
+& \mbox{ where } \theta_i \sim q(\theta_{N+1} | z_{N+1, 1:T}).
+\end{align}
+\item Similarly, substituting (\ref{updateEqApprox}) into (\ref{rpDeriv}) gives the Reparameterised Estimator for updating Variational Bayes,
+\begin{equation}
+\label{rpUpdate}
+\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^M \frac{d f(\lambda, \epsilon_i)}{d \lambda} \frac{d \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{d \theta_i} + \frac{d J(\lambda, \epsilon_i)}{d \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
+\end{equation}
+where $f(\lambda, \epsilon)$ transforms a draw of $\epsilon$ into a draw from $q(\theta_{N+1} | z_{N+1, 1:T})$.
+\item This process can be repeated as neccessary, and is $O(1)$ instead of $O(T)$ assuming $S$ grows at the same rate as $T$.
 \end{itemize}
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
 \includegraphics[width = 0.8\textwidth]{HierSingleKDE.png}
-\caption{Left: $p(\theta_i | \beta, y_{1:N})$ for a hierarchical mixture model with K = 6 and N = 2000.
-Right: Kernel Density Estimate of the posterior means from the same cars modelled individually.
-The hierarchical model has captured the bimodality and skewness present in the individual posterior means.}
+\caption{Top: A Kernel Density Estimate of the distribution of posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ modelled independently with no hierarchical structure.
+Bottom: $p(\theta_i | \beta, z_{1:2000})$ for a hierarchical mixture model with $K = 6$ using the same vehicles, with each component of the mixture denoted by the coloured densities.
+The hierarchical model has captured the bimodality and skewness present in the independent posterior means.}
 \label{fig:HierSingleKDE}
@@
 \end{enumerate}
+\begin{itemize}
+\item In each case the VB posterior approximation $q(\theta)$ has the same family as the prior $p(\theta)$, so either a multivariate normal or a mixture of multivariate normals.
+This is so we can replace the prior with the VB fit at each step of the updating method easier.
+\item The Hierarchical model is fit for 2000 cars, leaving 873 for forecast evaluation.
+\end{itemize}
 
@@
 \item Regular VB (fit each update to the new sample)
-\item Streaming VB (fit to new data only)
+\item Updating VB (fit to new data using the previous VB fit as a prior)
 \end{enumerate}
+
+\begin{center}
+\begin{table}[h]
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{| l | c | c | c |}
+\hline
+& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
+\hline
+MCMC & \multicolumn{3}{c}{Random Walk Metropolis Hastings jointly drawing the entire $\theta$ vector using Garthwaite, Fan and Sisson (2015)} \\
+& \multicolumn{3}{c}{to control the variance matrix of the Multivariate Gaussian proposal distribution} \\
+\hline
+VB  & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpDeriv}) to update parameters of a} & Variational Bayes using (\ref{scoreDeriv}) to update parameters of a six \\
+Standard & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\
+\hline
+VB & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpUpdate}) to update parameters of a} &  Variational Bayes using (\ref{scoreUpdate}) to update parameters of a six \\
+Updating & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\ \\
+\hline
+\end{tabular}}
+\label{tableAlg}
+\caption{Details of the algorithms used to produce posterior distributions for each method and prior combination}
+\end{table}
+\end{center}
 
@@
 \item Compare forecasts for different sample sizes and forecast horizons
-\item Strategy: Fit model for one car up time time $S = 10$ (one second), forecast from $S+1:S+30$ (three seconds seems to be a standard forecast horizon for driver reaction, though surely self driving cars have a faster reaction time).
+\item Strategy: Fit model for one car using data up to time $S = 10$ (one second), forecast from $S+1:S+30$ (three seconds seems to be a standard forecast horizon for driver reaction, though surely self driving cars have a faster reaction time).
 \item Fit model to $S = 20$, forecast next three seconds, repeat.
 \item Collect logscores for predictive densities marginalised over the posterior distributions.
+\item Use MAP of the predictive density to get a point estimate of the future position, compare to the constant velocity / angle model.
 \item Repeat for each of the 873 remaining cars.
@@
 \centering
-\includegraphics[width = 0.95\textwidth]{logscoresByPrior.png}
-\caption{Predictive log-scores for $p(a_{T+1:T+10}, d_{T+1:T+10} | a_{1:T}, d_{1:T})$ for $T = \{10, 20, \dots, 300\}$ for each of the three levels of prior information fit by both MCMC and VB.
-Stronger prior information improves forecasts in small samples, but as $T$ increases there are clear benefits to refitting the model regardless of the level of prior information.
-There was a mixtake in the VB finite mixture code which has been corrected but I am still waiting for the fixed results.}
+\includegraphics[width = 0.95\textwidth]{CarsBoxplots.png}
+\caption{Predictive logscores for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each of the three levels of prior information fit by MCMC, standard VB and updating VB.
+The median logscore across all time periods, forecast horizons, and priors for the updating VB algorithm is 3.28, compared to 3.43 for standard VB and 3.46 for MCMC.}
 \label{fig:MainResults}
@@
 
-Results:
-\begin{itemize}
-\item Models based on the strong prior information outperform no information models for first few seconds, where we see a significant improvement in the log scores.
-As the dataset increases to $> 100$ observations (10 seconds), the data dominates and the prior choice does not have much impact.
-The median logscore (and probably other quantiles, hard to see from the graph) at a larger value of T increases as expected, showing that it is important to re-estimate the model with an increased dataset.
-However it seems that this re-estimation could probably be done with a simpler prior structure than the full hierarchy without any big loss.
-\item Compare streaming data VB - Hoping it will be comparable to other methods, it technically may be a bit worse as it uses one more approximation than the standard VB, but I don't think this will have a very big impact.
-It certainly will be faster, so if the performance is roughly equal this is evidence in favour of the online method.
-\item Compare VB and MCMC methods for the same models to get an idea of approximation error.
-\item Still waiting on the cluster for standard VB Finite Mixture Model and all VB streams.
-\item We will have multiple forecasts for each actual observation (eg.
-$p(y_{43} | y_{20}), p(y_{43} | y_{30})$ and $p(y_{43} | y_{40})$) which gives us another approach to look at the benefits of increased observed data.
-\end{itemize}
+\begin{center}
+\begin{table}[h]
+\begin{tabular}{| l | c | c | c |}
+\hline
+& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
+\hline
+MCMC & 3.42 & 3.47 & 3.51 \\
+VB Standard & 3.35 & 3.46 & 3.48 \\
+VB Updating & 3.14 & 3.28 & 3.46 \\
+\hline
+\end{tabular}
+\label{tableMedian}
+\caption{Median predictive logscore for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each prior and algorithm combination}
+\end{table}
+\end{center}
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width = 0.95\textwidth]{PredictiveError.png}
+\caption{Mean Euclidian Error for car position predictions at a forecast horizon of one second, two seconds or three seconds for the Finite Mixture prior / VB Update Algorithm compared to the constant velocity and angle model.
+The average error at the three second forecast horizon is reduced from 7.16 feet to just 0.89 feet.}
+\label{fig:MainResults}
+\end{figure}
 

### 2018-02-05 Cars/testSlurm/slurmCppFuns.cpp

@@ -211,7 +211,7 @@ Rcpp::List arUpdater(mat data, Rcpp::NumericMatrix lambdaIn, vec epsilon, vec me
   arUpdate p(data, epsilon, mean, Linv, lags);
   stan::math::set_zero_all_adjoints();
   stan::math::gradient(p, lambda, eval, grad);
- 
+  
   return Rcpp::List::create(Rcpp::Named("grad") = grad,
                             Rcpp::Named("val") = eval);
 }
@@ -220,7 +220,7 @@ Rcpp::List arUpdater(mat data, Rcpp::NumericMatrix lambdaIn, vec epsilon, vec me
 cube evalMCMCDens (mat data, int N, int H, mat grid, mat MCMCdraws){
   cube densities (N, N, H, fill::zeros);
   boost::math::normal_distribution<> Zdist(0, 1);
-
+  
   for(int m = 0; m < 1000; ++m){
     double afc1 = data(1, 0), afc2 = data(0, 0), dfc1 = data(1, 1), dfc2 = data(0, 1), afc, dfc, vfc;
     vec draws = MCMCdraws.row(1000 + 4 * m).t();
@@ -295,15 +295,15 @@ cube evalVBDens (mat data, vec mean, mat L, vec weights, int N, int H, mat grid,
           // Transform delta x / delta y to velocity and angle
           double v = std::sqrt(
             std::pow(grid(i*N + j, 0), 2) +
-            std::pow(grid(i*N + j, 1), 2)
+              std::pow(grid(i*N + j, 1), 2)
           );
           double del = std::atan2(
             grid(i*N + j, 1),
             grid(i*N + j, 0)
-            ) - 1.570796;
+          ) - 1.570796;
           // p(x, y) = p(v) * p(d) / sqrt(dx^2 + dy^2)
           densities(i, j, h) += (pdf(Zdist, (v - vfc) / sigV) / sigV) *
-                                (pdf(Zdist, (del - dfc) / sigD) / sigD) / (v * 1000);
+            (pdf(Zdist, (del - dfc) / sigD) / sigD) / (v * 1000);
         }
       }
       // lag acceleration and angle
@@ -313,10 +313,10 @@ cube evalVBDens (mat data, vec mean, mat L, vec weights, int N, int H, mat grid,
       dfc1 = dfc;
     }
   }
-    
+  
[…]

### 2018-02-05 Cars/testSlurm/slurmFCAll.R

@@ -21,10 +21,6 @@ sSeq <- seq(S, maxT, S)
 results <- data.frame()
 methods <- c('None', 'Single Hierarchy', 'Finite Mixture')
 
-starting <- list(matrix(c(-5, -5, rep(0, 4), c(chol(diag(0.5, 6)))), ncol = 1),
-                 prior[[3]])
-
-
 # MCMC Hyper parameters
 hyper <- list()
 for(k in 1:2){
@@ -33,16 +29,28 @@ for(k in 1:2){
   uinv <- solve(matrix(prior[[k]][7:42], 6))
   hyper[[k]]$varInv <- t(uinv) %*% uinv
 }
-# Mixture Prior
+# Mixture Prior - The prior.RDS object has mean / inverse u however this is converted to mean / log(sd) for the diagonal mixture model in VB
 mean <- prior[[3]][1:36]
+priorDiag <- mean
 varinv <- NULL
 for(k in 1:6){
   uinv <- matrix(prior[[3]][k*36 + 1:36], 6)
-  varinv <- rbind(varinv, t(uinv) %*% uinv)
+  vari <- t(uinv) %*% uinv
+  logsd <- log(diag(solve(vari)))
+  varinv <- rbind(varinv, vari)
+  priorDiag <- c(priorDiag, logsd)
 }
 weights <- prior[[3]][6*7*6 + 1:6]
+priorDiag <- c(priorDiag, weights)
 weights <- exp(weights) / sum(exp(weights))
 hyper[[3]] <- list(mean = mean, varInv = varinv, weights = weights)
+
+prior[[3]] <- matrix(priorDiag, ncol = 1)
+
+starting <- list(matrix(c(-5, -5, rep(0, 4), c(chol(diag(0.5, 6)))), ncol = 1),
+                 prior[[3]])
+
+
 
 # Extract Data
 data <- datafc[[i]]
@@ -67,7 +75,7 @@ grid <- cbind(asup, dsup)
 
 # Incrementally add data to VB fits
 for(s in seq_along(sSeq)){
-  if(sSeq[s] > nrow(data)){
+  if((sSeq[s] + H) > nrow(data)){
     break
   }
   # Get data for stream
@@ -92,40 +100,48 @@ for(s in seq_along(sSeq)){
   
   # Evaluate predictive densities
   resultsOnline <- lapply(1:3, function(x) VBDens(data = data[(sSeq[s]-1):(sSeq[s] + H), ],
-                                                    fit = fitOnline[[x]],
-                                                    grid = grid, 
[…]

### 2018-02-05 Cars/testSlurm/slurmRFuns.R

@@ -208,6 +208,87 @@ carsVBMixScore <- function(data, lambda, priorMix, K = 6, S = 250, maxIter = 500
   return(list(lambda=lambda, LB = LB[1:min(iter-1, maxIter)], iter = min(maxIter, iter-1)))
 }
 
+
+carsVBMixScoreDiag <- function(data, lambda, priorMix, S = 250, maxIter = 5000, alpha = 0.01, beta1 = 0.9, beta2 = 0.99, threshold = 0.01){
+  
+  dimLambda <- nrow(lambda)
+  sobol <- sobol_points(100+6*S, 6)
+  diff <- threshold + 1
+  iter <- 1
+  LB <- numeric(maxIter)
+  M <- rep(0, dimLambda)
+  V <- rep(0, dimLambda)
+  e <- 1e-8
+  meanLB <- 0
+  oldMeanLB <- 0
+  while(diff > threshold | iter < 100){
+    if(iter > maxIter){
+      break
+    }
+    #if(any(is.na(lambda))){
+    #  break
+    #}
+    grad <- matrix(0, dimLambda, S)
+    eval <- numeric(S)
+    z <- lambda[73:78]
+    pi <- exp(z) / sum(exp(z))
+    unif <- shuffle(sobol)
+    s <- 0
+    try <- 0
+    epsilon <- qnorm(unif[101:(100+6*S), ])
+    epsilon[epsilon < -3] = -3
+    epsilon[epsilon > 3] = 3
+    while(s < S){
+      k <- sample(1:6, 1, prob=pi)
+      Qmean <- lambda[(k-1)*6 + 1:6]
+      Qsd <- exp(lambda[6*6 + (k-1)*6 + 1:6])
+      theta <- c(Qmean + Qsd * epsilon[try+1,])
+      derivs <- scoreDerivDiag(data, lambda, theta, priorMix$mean, priorMix$SigInv, priorMix$dets, priorMix$weights)
+      if(all(is.finite(derivs$grad)) & all(!is.na(derivs$grad)) & is.finite(derivs$val) & !is.na(derivs$val)){
+        s <- s + 1
+        eval[s] <- derivs$val
+        grad[,s] <- derivs$grad
+        if(s == S){
+          gradient <- rowMeans(grad, na.rm = TRUE)
+          gradientSq <- rowMeans(grad^2, na.rm = TRUE)
+          LB[iter] <- mean(eval, na.rm = TRUE)
+          break
+        }
+      }
[…]

### 2018-02-05 Cars/paper.tex
lane changes

@@
 \hline
-MCMC & \multicolumn{3}{c}{Random Walk Metropolis Hastings jointly drawing the entire $\theta$ vector using Garthwaite, Fan and Sisson (2015)} \\
-& \multicolumn{3}{c}{to control the variance matrix of the Multivariate Gaussian proposal distribution} \\
+MCMC & \multicolumn{3}{c|}{Random Walk Metropolis Hastings jointly drawing the entire $\theta$ vector using Garthwaite, Fan and Sisson (2015)} \\
+& \multicolumn{3}{c|}{to control the variance matrix of the Multivariate Gaussian proposal distribution} \\
 \hline
@@
 VB & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpUpdate}) to update parameters of a} &  Variational Bayes using (\ref{scoreUpdate}) to update parameters of a six \\
-Updating & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\ \\
+Updating & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\ 
 \hline
@@
 \section{Conclusion}
+\begin{itemize}
+\item Fitting individual AR 2 models on each driver's velocity and steering angle drastically improves forecasts of the future trajectory compared to the constant model used in practice
+\item Strong prior information improves forecasts early on in the sample, but has a reduced impact after ~20 seconds of observation
+\item Constantly fitting individual models in this way can be made computationally feasible by updating previous Variational Bayes model fits, which incurs only a small performance hit relative to the more expensive standard VB and MCMC methods.
+\end{itemize}
 

### 2018-02-05 Cars/HierMixMod.R

@@ -16,6 +16,8 @@ library(tidyverse)
 library(Rcpp)
 library(RcppArmadillo)
 library(GGally)
+
+setup {
 id <- readRDS('carsID.RDS')
 
 read.csv('carsAug.csv') %>%
@@ -58,8 +60,7 @@ for(i in 1:500){
 }
 dataChanged[[501]] <- noStopChanged
 saveRDS(dataChanged, 'fcDataChanged.RDS')
-
-
+}
 
 HierMixMod{
 
@@ -497,7 +498,7 @@ sliceSampler{
 results {
 
 results <- NULL
-for(i in seq_along(id$idfc)){
+for(i in 1:seq_along(id$idfc)){
   #tmp <- read.csv(paste0('evals/car', id$idfc[i], '.csv'))
   #results <- rbind(results, tmp)
   try(assign('results',
@@ -529,12 +530,24 @@ results %>%
   gather(Predictor, meanDist, -horizon, -S) %>%
   ggplot() + geom_line(aes(x = S, y = meanDist, colour = Predictor)) + 
   facet_wrap(~horizon) + 
-  labs(x = 'T', y = 'Mean Euclidean Error')
+  labs(x = 'T', y = 'Mean Euclidean Error') + 
+  theme(legend.position = 'bottom')
 
-results %>% group_by(method, prior) %>% filter(is.finite(logscore)) %>% summarise(med = median(logscore, na.rm = TRUE)) %>% arrange(med)
+results %>%
+  filter(h == 10) %>%
+  group_by(S, method, prior) %>%
+  summarise(meanDist = mean(mapDist)) %>%
+  ggplot() + geom_line(aes(S, meanDist, colour = method)) + facet_wrap(~prior)
 
-results %>% filter(h == 30 & method == 'VB-Stream' & prior == 'Finite Mixture') %>% 
-  summarise(meanMAP = mean(mapDist), meanConst = mean(consDist))
+results %>% 
+  group_by(method, prior) %>% 
+  filter(is.finite(logscore)) %>% 
+  summarise(med = median(logscore, na.rm = TRUE)) %>% 
+  arrange(med)
+
+results %>% 
+  filter(h == 30 & method == 'VB-Stream' & prior == 'Finite Mixture') %>% 
+  summarise(meanMAP = mean(mapDist), meanConst = mean(consDist)) 
   
 
   
[…]

### 2018-02-05 Cars/paper.tex
convert feet to metres in code (not paper)

@@
 
+\begin{center}
+\begin{table}[h]
+\begin{tabular}{| l | c | c | c |}
+\hline
+& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
+\hline
+MCMC & 0.81 & 0.77 & 0.77 \\
+VB Standard & 0.79 & 0.77 & 0.77 \\
+VB Updating & 0.79 & 0.78 & 0.77 \\
+\hline
+\end{tabular}
+\label{tableError}
+\caption{Mean predictive Euclidean error in feet for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each prior and algorithm combination.
+The mean predictive Euclidean error for the constant velocity / angle model is 3.26 feet.}
+\end{table}
+\end{center}
+
 \begin{figure}[ht]
@@
 \item Fitting individual AR 2 models on each driver's velocity and steering angle drastically improves forecasts of the future trajectory compared to the constant model used in practice
-\item Strong prior information improves forecasts early on in the sample, but has a reduced impact after ~20 seconds of observation
+\item Strong prior information improves forecasts early on in the sample, but has a reduced impact after ~20 seconds of observation.
 \item Constantly fitting individual models in this way can be made computationally feasible by updating previous Variational Bayes model fits, which incurs only a small performance hit relative to the more expensive standard VB and MCMC methods.

### 2018-02-05 Cars/HierMixMod.R

@@ -509,6 +509,12 @@ for(i in 1:seq_along(id$idfc)){
   }
 }
 
+mPerFoot <- 0.3048
+results %>% 
+  mutate(logscore = logscore + log(mPerFoot^2),
+         mapDist = mapDist * mPerFoot,
+         consDist = consDist * mPerFoot) -> results
+
 
 results %>% 
   mutate(group = factor(ceiling(S / 30)),
@@ -548,75 +554,15 @@ results %>%
 results %>% 
   filter(h == 30 & method == 'VB-Stream' & prior == 'Finite Mixture') %>% 
   summarise(meanMAP = mean(mapDist), meanConst = mean(consDist)) 
-  
-
-  
-results %>%
-  #filter(method == 'MCMC' & prior == 'None') %>%
-  mutate(T = S + h,
-         horizon = ceiling(h/10)) %>%
-  group_by(method, prior, S, h, T, horizon, id) %>%
-  summarise(ls = - sum(logscore)) %>%
-  ungroup() %>%
-  select(T, id, ls, horizon, method, prior) %>%
-  filter(T > 30) %>%
-  ggplot() + geom_boxplot(aes(factor(horizon), ls)) +
-  facet_grid(method ~ prior) + 
-  scale_y_continuous(trans = reverselog_trans(base=10),
-                     labels=trans_format("identity", function(x) -x))
-
-
-library(scales)
-reverselog_trans <- function(base = exp(1)) {
-  trans <- function(x) -log(x, base)
-  inv <- function(x) base^(-x)
-  trans_new(paste0("reverselog-", format(base)), trans, inv, 
-            log_breaks(base = base), 
-            domain = c(1e-100, Inf))
-}
-
-results %>% 
-  mutate(T = factor(ceiling(S / 30)),
-         priorInfo = factor(prior, levels = c('None', 'Single Hierarchy', 'Finite Mixture')),
-         method = ifelse(method == 'VB-Offline', 'VB-Standard', 
-                         ifelse(method == 'VB-Stream', 'VB-Updating', 'MCMC'))) %>%
-  ggplot() + geom_boxplot(aes(T, logscore, colour = priorInfo)) +
-  facet_wrap(~method, ncol = 1) + ylim(-10, 15)  +
-  labs(x = 'T Range', y = 'X/Y Logscore') + 
-  scale_x_discrete(labels = c('10-30', '40-60', '70-90', '100-120', '130-150', 
-                              '160-180', '190-210', '220-240', '250-270', '280-300'))  
[…]

### 2018-02-06 Cars/paper.tex
paper work

@@
 
-Variational Bayes approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
+Variational Bayes (Jordan et al. 1999) approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
 \begin{equation}
@@
 \label{ELBO}
-\mathcal{L}(q) = \int_{\theta} q(\theta) \left(\log(p(\theta, z)) - \log(q(\theta))\right) d\theta
+\mathcal{L}(q, \lambda) = \int_{\theta} q(\theta) \left(\log(p(\theta, z)) - \log(q(\theta))\right) d\theta
 \end{equation}
-which is equivalent to minimising (\ref{KL-div}).
-This is done with stochastic gradient ascent, taking Monte Carlo estimates of $d\mathcal{L}(q) / d \lambda$ where $\lambda$ is the parameter vector of $q(\theta)$.
-There are two Monte Carlo estimators of $d\mathcal{L}(q) / d \lambda$, the Score Estimator, 
+which is equivalent to minimising (\ref{KL-div}), where $\lambda$ is the parameter vector of $q(\theta)$.
+This is done with stochastic gradient ascent, taking Monte Carlo estimates of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$ (Hoffman et al. 2013)
+and repeating updates of the form
+\begin{equation}
+\label{gradientAscent}
+\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} 
+\end{equation}
+which is guaranteed to converge to a local maximum (Robbins and Munro, 1951) if $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
+\begin{align}
+&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
+&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
+\end{align}
+In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of Kingma and Ba (2015).
+
+There are two Monte Carlo estimators of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the Score Estimator of Ranganath, Gerrish and Blei (2014), 
 \begin{equation}
 \label{scoreDeriv}
-\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^L \frac{d \log(q(\theta_i))}{d \lambda} \left(\log(p(\theta_i, z)) - \log(q(\theta_i)) \right), \mbox{ where } \theta_i \sim q(\theta),
+\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_i))}{\delta \lambda} \left(\log(p(\theta_i, z)) - \log(q(\theta_i)) \right), \mbox{ where } \theta_i \sim q(\theta),
 \end{equation}
-and the Reparameterised Estimator,
+and the Reparameterised Estimator of Kingma Welling (2014),
 \begin{equation}
 \label{rpDeriv}
-\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^M \frac{d f(\lambda, \epsilon_i)}{d \lambda} \frac{d \log(p(\theta_i, z))}{d \theta_i} + \frac{d J(\lambda, \epsilon_i)}{d \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
+\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^M \frac{\delta f(\lambda, \epsilon_i)}{\delta \lambda} \frac{\delta \log(p(\theta, z))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_i)} + \frac{\delta J(\lambda, \epsilon_i)}{\delta \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
 \end{equation}
@@
 Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the derivative estimator (\ref{scoreDeriv}) must be used.
-To reduce $L$ and thus the computational cost of the algorithm, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$.
+To offset the increased variance and allow a reduction in $L$, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$.
 Using RQMC allows (\ref{scoreDeriv}) to converge to the true derivative at a rate $O(L)$ compared to $O(\sqrt{L})$ in standard Monte Carlo.
@@
 \label{scoreUpdate}
-\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^L \frac{d \log(q(\theta_{N+1} | z_{N+1, 1:T}))}{d \lambda} &\left(\log(q(\theta_{N+1} | z_{N+1, 1:S}) - \log(q(\theta_{N+1} | z_{N+1, 1:T})) \right), \nonumber \\
-& \mbox{ where } \theta_i \sim q(\theta_{N+1} | z_{N+1, 1:T}).
+\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_{N+1, i} | z_{N+1, 1:T}))}{\delta \lambda} &\left(\log(q(\theta_{N+1, i} | z_{N+1, 1:S}) - \log(q(\theta_{N+1} | z_{N+1, 1:T})) \right), \nonumber \\
+& \mbox{ where } \theta_{N+1, i} \sim q(\theta_{N+1} | z_{N+1, 1:T}).
 \end{align}
@@
 \label{rpUpdate}
-\frac{d\mathcal{L}(q)}{d \lambda} \approx \sum_{i = 1}^M \frac{d f(\lambda, \epsilon_i)}{d \lambda} \frac{d \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{d \theta_i} + \frac{d J(\lambda, \epsilon_i)}{d \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
+\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda_T} \approx \sum_{i = 1}^M \frac{\delta f(\lambda_T, \epsilon_i)}{\delta \lambda} \frac{\delta \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon_i)} + \frac{\delta J(\lambda_T, \epsilon_i)}{\delta \lambda_T}, \mbox{ where } \epsilon_i \sim r(\epsilon)
 \end{equation}
-where $f(\lambda, \epsilon)$ transforms a draw of $\epsilon$ into a draw from $q(\theta_{N+1} | z_{N+1, 1:T})$.
+where $\lambda_T$ denotes the parameter vector of $q(\theta_{N+1} | z_{N+1, 1:T})$.
 \item This process can be repeated as neccessary, and is $O(1)$ instead of $O(T)$ assuming $S$ grows at the same rate as $T$.
@@
 \end{center}
+
+There are six naive forecast models with constant estimates of the future values of $v_t$ and $\delta_t$ for $t = T+1, \dots, T+30$.
+$v_t$ is estimated by either the current velocity, $v_T$ or the average of the previous ten lags of $v$, $1/10 \sum_{s = T - 9}^T v_s$.
+$\delta_t$ is estimated by one of the current steering angle, $\delta_T$, the average of the previous ten lags of $\delta$, $1/10 \sum_{s = T - 9}^T \delta_s$ or by $\pi/2$, i.e.
+the car is moving directly ahead.
+Table \ref{tableNaive} classifies each naive model based on the combination of velocity and angle estimates.
+\begin{table}
+\begin{center}
+\begin{tabular}{|l|c|c|c|}
+\hline
+& $\hat{\delta}_t = \delta_T$ & $\hat{\delta}_t = 1/10 \sum_{s=T-9}^T \delta_s$ & $\hat{\delta}_t = \pi/2$ \\
+\hline
+$\hat{v}_t = v_T$ & Naive 1 & Naive 2 & Naive 3 \\
+$\hat{v}_t = 1/10\sum_{s=T-9}^T$ & Naive 4 & Naive 5 & Naive 6\\
+\hline
+\end{tabular}
+\end{center}
+\label{tableNaive}
+\caption{Classification of naive forecast models by future velocity and steering angle estimators}
+\end{table}
 

### 2018-02-07 Cars/paper.tex
more paper changes mostly

@@
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
+\title{Online Updating of Variational Bayes for Forecasts of Vehicle Trajectory with Heterogeneous Driver Behaviour}
 \begin{document}
-
+\maketitle
 \section{Introduction}
@@
 \item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the Self Driving Car.
+\item While Self Driving Cars reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
 \item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg.
@@
 \item These can assume the surrounding vehicles will maintain their current angle and velocity (Gindele, Brechtel, Dillmann 2010; Waymo Safety Report 2017; Bautista 2017), but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines (Woo et al.
-201b6; Tomar, Verma, and Tomar 2014; Woo et al.
+201b6; Ding et al.
+2012; Woo et al.
 2017; Geng et al.
@@
 \item It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a trajectory forecast.
-\item These time series models (In this case a bivariate AR2) fit a parameter vector to capture the driver behaviour, which can be very different from driver to driver (Chen 2017).
+\item These time series models (In this case a bivariate AR2) fit a parameter vector to capture the driver behaviour, which can depend both on the be very different from driver to driver (Chen 2017) which may depend on a particular vehicle's position relative to surrounding traffic.
 \item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
@@
 \item Compare log scores from streaming and standard VB forecast densities to see how the updating method compares to standard VB.
-I am also using MCMC, so we can compare either VB method to this, theoretically MCMC > VB Standard > VB Updating
+I am also using MCMC, so we can compare either VB method to this, theoretically MCMC $\geq$ VB Standard $\geq$ VB Updating
 \end{itemize}
@@
 \begin{itemize}
-\item Data is provided by the Next Generation Simulation (NGSIM) project, which observes ~5000 cars for a 2000 feet section of the US 101 Highway.
+\item Data is provided by the Next Generation Simulation (NGSIM) project, which observes cars for a 2000 feet section of the US 101 Highway.
 \item Observations of x/y coordinates are provided every 100 ms (10 hz), with around 500-1000 observations per car.
@@
 x^*_{t} &= x^*_{t-1} + v_{t} \cos(\delta_{t}) \label{xEq} \\
-y^*_{t} &= y^*_{t-1} + v_{t} \sin(\delta_{t}). \label{yEq},
+y^*_{t} &= y^*_{t-1} + v_{t} \sin(\delta_{t}) \label{yEq},
 \end{align}
@@
 For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with fixed mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with fixed degrees of freedom $\tau_j$ and scale matrix $\Psi_j$.
-$k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi_i} = \{\pi_{1, i}, \dots, \pi_{K, i}\}$, where $\boldsymbol{\pi_i}$ has a Dirichlet prior with fixed hyperparameters $\alpha_1, \alpha_2, \dots, \alpha_K$.
-Defining $\beta = \{\mu_i, \Sigma_i, i = 1, \dots, K\}$, $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s} | s = 1, \dots, T\}$ and $z_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | z_{1:N})$ can be estimated with MCMC methods.
+$k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi} = \{\pi_{1}, \dots, \pi_{K}\}$, where $\boldsymbol{\pi}$ has a Dirichlet prior with fixed hyperparameters $\alpha = \alpha_j, j = 1 , \dots, K$.
+Defining $\beta = \{\mu_j, \Sigma_j, \pi_j, j = 1, \dots, K\}$, $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s} | s = 1, \dots, T\}$ and $z_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | z_{1:N})$ can be estimated with MCMC methods.
+
+\begin{figure}[ht]
+\centering
+\includegraphics[width = 0.95\textwidth]{HierSingleKDE.png}
+\caption{Top: A Kernel Density Estimate applied to the individual posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ where $p = q = 2$, modelled independently without any hierarchical structure.
+Bottom: $p(\theta_i | \beta, z_{1:2000})$ for a hierarchical mixture model with $K = 6$ using the same vehicles, with each component of the mixture denoted by the coloured densities.
+The hierarchical model has captured the bimodality and skewness present in the independent posterior means.}
+\label{fig:HierSingleKDE}
+\end{figure}
 
@@
 
-Variational Bayes (Jordan et al. 1999) approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
+Variational Bayes (VB, Jordan et al.
+1999) approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
 \begin{equation}
@@
 
-There are two Monte Carlo estimators of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the Score Estimator of Ranganath, Gerrish and Blei (2014), 
+There are two Monte Carlo estimators of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of Ranganath, Gerrish and Blei (2014), 
 \begin{equation}
@@
 \end{equation}
-and the Reparameterised Estimator of Kingma Welling (2014),
+and the reparameterised estimator of Kingma and Welling (2014),
 \begin{equation}
@@
 (\ref{rpDeriv}) typically has significantly lower variance than (\ref{scoreDeriv}), allowing $M$ to be set lower than $L$ which improves the speed of the algorithm; however the requirement for $f$ to exist restricts the class of approximating distributions $q(\theta)$ that can be used.
-Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the derivative estimator (\ref{scoreDeriv}) must be used.
+Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the score estimator (\ref{scoreDeriv}) must be used.
 To offset the increased variance and allow a reduction in $L$, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$.
@@
 \end{equation}
-\item Substituting (\ref{updateEqApprox}) into (\ref{scoreDeriv}) gives the Score Estimator for updating Variational Bayes, 
+\item Substituting (\ref{updateEqApprox}) into (\ref{scoreDeriv}) gives the score estimator for updating Variational Bayes, 
 \begin{align}
@@
 \end{align}
-\item Similarly, substituting (\ref{updateEqApprox}) into (\ref{rpDeriv}) gives the Reparameterised Estimator for updating Variational Bayes,
+\item Similarly, substituting (\ref{updateEqApprox}) into (\ref{rpDeriv}) gives the reparameterised estimator for updating Variational Bayes,
 \begin{equation}
 \label{rpUpdate}
-\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda_T} \approx \sum_{i = 1}^M \frac{\delta f(\lambda_T, \epsilon_i)}{\delta \lambda} \frac{\delta \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon_i)} + \frac{\delta J(\lambda_T, \epsilon_i)}{\delta \lambda_T}, \mbox{ where } \epsilon_i \sim r(\epsilon)
+\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T} \approx \sum_{i = 1}^M \frac{\delta f(\lambda_T, \epsilon_i)}{\delta \lambda_T} \frac{\delta \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon_i)} + \frac{\delta J(\lambda_T, \epsilon_i)}{\delta \lambda_T}, \mbox{ where } \epsilon_i \sim r(\epsilon)
 \end{equation}
@@
 
-\section{Empirical}
+\section{Forecast Evaluation}
 
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.8\textwidth]{HierSingleKDE.png}
-\caption{Top: A Kernel Density Estimate of the distribution of posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ modelled independently with no hierarchical structure.
-Bottom: $p(\theta_i | \beta, z_{1:2000})$ for a hierarchical mixture model with $K = 6$ using the same vehicles, with each component of the mixture denoted by the coloured densities.
-The hierarchical model has captured the bimodality and skewness present in the independent posterior means.}
-\label{fig:HierSingleKDE}
-\end{figure}
-
-Models:
-\begin{enumerate}
-\item No prior information for the new car - very flat priors.
-\item Prior information from hierarchical model with one component
-\item Prior information from large mixture hierarchical model
-\end{enumerate}
 \begin{itemize}
-\item In each case the VB posterior approximation $q(\theta)$ has the same family as the prior $p(\theta)$, so either a multivariate normal or a mixture of multivariate normals.
-This is so we can replace the prior with the VB fit at each step of the updating method easier.
-\item The Hierarchical model is fit for 2000 cars, leaving 873 for forecast evaluation.
+\item 4071 vehicles are available in the sample
+\item 500 are used to estimate splines and straighten the road to relative coordinates as described in Section 2
+\item 2000 are used to train any models used
+\item 1571 are available for testing
+\item Each second (10 observations) models are fit to the sample of $z_{N+1, 1:T}$ and forecasts of $z_{N+1, T+1:T+30}$ are made, corresponding to a three second forecast horizon
+\item Forecasts are compared in terms of predictive log-score (where a predictive density is available) and Euclidean error.
 \end{itemize}
 
-Methods:
+\subsection{Heterogenous Time Series Models}
+
+There are three versions of the heterogenous time series model described in Section 3, differentiated by the form of the prior, $p(\theta_{N+1} | z_{1:N})$.
 \begin{enumerate}
-\item MCMC
-\item Regular VB (fit each update to the new sample)
-\item Updating VB (fit to new data using the previous VB fit as a prior)
+\item Non-informative prior for the new car, where $p(\theta | z_{1:N})$ is multivariate normal with mean vector $(-5, -5, 0, 0, 0, 0)$ and diagonal variance matrix $10 \mathbb{I}$.
+\item Informative prior where $p(\theta | z_{1:N})$ is multivariate normal with a mean vector and variance matrix estimated by the posterior mean of $\mu$ and $\Sigma$ in a hierarchical model with $K = 1$ fit by MCMC to the 2000 training vehicles.
+\item Informative prior where $p(\theta | z_{1:N})$ is a six component mixture of multivariate normals with mean vectors, variance matrices and mixture weights estimated by the posterior means of each $\mu_j$ and $\Sigma_j$ and $\pi_j$ in a hierarchical model with $K = 6$ fit by MCMC to the 2000 training vehicles.
 \end{enumerate}
-
+Each model is fit by each of Metropolis-Hasting MCMC, standard Variational Bayes where the entire observed history of $z_{N+1, 1:T}$ is used for each model fit, and the updating Variational Bayes process described by Section 4.2 where only the most recent observed data $z_{N+1, S+1:T}$ is used in each iteration.
+In each Variational Bayes algorithm the approximating distribution has the same functional form as the prior used, a multivariate normal distribution for the non-informative prior and $K = 1$ hierarchy, and a six component mixture of multivariate normals for the $K = 6$ hierarchy.
+Details of each case are provided in Table \ref{tableAlg}.
 \begin{center}
@@
 
-There are six naive forecast models with constant estimates of the future values of $v_t$ and $\delta_t$ for $t = T+1, \dots, T+30$.
-$v_t$ is estimated by either the current velocity, $v_T$ or the average of the previous ten lags of $v$, $1/10 \sum_{s = T - 9}^T v_s$.
-$\delta_t$ is estimated by one of the current steering angle, $\delta_T$, the average of the previous ten lags of $\delta$, $1/10 \sum_{s = T - 9}^T \delta_s$ or by $\pi/2$, i.e.
-the car is moving directly ahead.
-Table \ref{tableNaive} classifies each naive model based on the combination of velocity and angle estimates.
+\subsection{Competing Models}
+
+\subsubsection{Constant Angle and Acceleration Models}
+
+The framework provided by (\ref{xEq}) and (\ref{yEq}) allows point estimates of $\{x^*_{t}, y^*_{t} | t = T + 1, \dots, T+30\}$ to be easily obtained given $\{x^*_T, y^*_T, v_T\}$ and point estimates of the future values of $\{a_{t}, \delta_{t} | t = T + 1, \dots, T+30\}$.
+Nine naive forecast models are provided where these future values are estimated by a constant $\{\hat{a}, \hat{\delta}\}$.
+The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}$ and one estimator for $\hat{\delta}$.
+Acceleration estimators are: $\hat{a} = a_T, \hat{a} = 0.1 \sum_{s = T-9}^T a_s$, and $\hat{a} = 0$.
+Similarly steering angle estimators are: $\hat{\delta} = \delta_T, \hat{\delta} = 0.1 \sum_{s = T - 9}^T \delta_s$, and $\hat{\delta} = \pi/2$.
 \begin{table}
@@
 \hline
-$\hat{v}_t = v_T$ & Naive 1 & Naive 2 & Naive 3 \\
-$\hat{v}_t = 1/10\sum_{s=T-9}^T$ & Naive 4 & Naive 5 & Naive 6\\
+$\hat{a}_t = a_T$ & Naive 1 & Naive 2 & Naive 3 \\
+$\hat{a}_t = 1/10\sum_{s=T-9}^T a_s$ & Naive 4 & Naive 5 & Naive 6\\
+$\hat{a}_t = 0$ & Naive 7 & Naive 8 & Naive 9 \\
 \hline
@@
 
-Evaluation:
-\begin{itemize}
-\item Compare forecasts for different sample sizes and forecast horizons
-\item Strategy: Fit model for one car using data up to time $S = 10$ (one second), forecast from $S+1:S+30$ (three seconds seems to be a standard forecast horizon for driver reaction, though surely self driving cars have a faster reaction time).
-\item Fit model to $S = 20$, forecast next three seconds, repeat.
-\item Collect logscores for predictive densities marginalised over the posterior distributions.
-\item Use MAP of the predictive density to get a point estimate of the future position, compare to the constant velocity / angle model.
-\item Repeat for each of the 873 remaining cars.
-\end{itemize}
+\subsubsection{Homogenous Time Series Models}
+
+The homogenous time series model has the same bivariate AR(2) structure given by (\ref{aAr}) and (\ref{dAr}) as used by the heterogenous time series models with the additional restriction that each $\theta_i = \theta$, so each vehicle shares the same parameter vector.
+This model is estimated by MCMC to the 2000 training cars.
+
+\subsubsection{Neural Networks}
+
+Following Ding et al.
+(2012), a Neural Network is constructed with inputs given by the ten most recent lags of vehicle position, angle, and velocity, and an output given by the next thirty values of the vehicle position.
+The network is trained with the Adam optimised with a mean Euclidean error loss function.
+\subsection{Results}
 
@@
 
+\begin{figure}[ht]
+\centering
+\includegraphics[width = 0.95\textwidth]{diffInLS.png}
+\caption{A truncated view of the difference in Variational Bayes predictive logscores (updating minus standard) for each prior and a range of different T values.
+$99\%$ of Finite Mixture differences, $93\%$ of Non-Informative differences and $92\%$ of Single Hierarchy differences are inside the (-5, 5) region.
+The algorithms are identical when $T = 10$ so this is excluded.
+Updating VB logscores are only slightly lower than the standard VB logscores, signifying only a small loss of accuracy associated with the extra approximation.}
+\label{fig:diffInLS}
+\end{figure}
+
 \begin{center}
@@
 \hline
-MCMC & 3.42 & 3.47 & 3.51 \\
-VB Standard & 3.35 & 3.46 & 3.48 \\
-VB Updating & 3.14 & 3.28 & 3.46 \\
+MCMC & 0.\textbf{846} & 0.899 & \textbf{0.931} \\
+VB Standard & 0.782 & \textbf{0.913} & 0.901 \\
+VB Updating & 0.558 & 0.701 & 0.873 \\
 \hline
@@
 \hline
-MCMC & 0.81 & 0.77 & 0.77 \\
-VB Standard & 0.79 & 0.77 & 0.77 \\
-VB Updating & 0.79 & 0.78 & 0.77 \\
+MCMC & 0.209 & 0.199 & 0.197 \\
+VB Standard & \textbf{0.203} & \textbf{0.196} & 0.197 \\
+VB Updating & 0.211 & 0.207 & \textbf{0.196} \\
 \hline
@@
 \label{tableError}
-\caption{Mean predictive Euclidean error in feet for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each prior and algorithm combination.
-The mean predictive Euclidean error for the constant velocity / angle model is 3.26 feet.}
+\caption{Mean predictive Euclidean error in metres using the maximum of $p(z_{i, T+h} | z_{i, 1:T})$ as a point-estimate with $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each algorithm and prior combination.
+Updating the VB model is competitive with a standard VB fit and exact sampling by MCMC.}
 \end{table}
@@
 
-\begin{figure}[ht]
+\begin{figure}[t]
 \centering
 \includegraphics[width = 0.95\textwidth]{PredictiveError.png}
-\caption{Mean Euclidian Error for car position predictions at a forecast horizon of one second, two seconds or three seconds for the Finite Mixture prior / VB Update Algorithm compared to the constant velocity and angle model.
-The average error at the three second forecast horizon is reduced from 7.16 feet to just 0.89 feet.}
-\label{fig:MainResults}
+\caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = \{10, 20, 30\}$ and $T = \{10, 20, \dots, 300\}$ for each model.
+Each AR(2) model is significantly more accurate than the naive constant velocity / angle models.}
+\label{fig:PredError}
 \end{figure}
@@
 \begin{itemize}
-\item Fitting individual AR 2 models on each driver's velocity and steering angle drastically improves forecasts of the future trajectory compared to the constant model used in practice
-\item Strong prior information improves forecasts early on in the sample, but has a reduced impact after ~20 seconds of observation.
-\item Constantly fitting individual models in this way can be made computationally feasible by updating previous Variational Bayes model fits, which incurs only a small performance hit relative to the more expensive standard VB and MCMC methods.
+\item Dynamic AR 2 models forecasting future acceleration and angles drastically improves forecasts of the future trajectory compared to the naive models.
+\item Treating vehicles as heterogenous allows behaviour specific to an individual driver to be learned, further improving forecasts.
+\item This requires constant models fits and updates, which can be made computationally feasible by updating previous Variational Bayes model fits, which incurs only a small performance hit relative to the more expensive standard VB and MCMC methods.
 \end{itemize}

### 2018-02-07 Cars/HierMixMod.R

@@ -1,7 +1,8 @@
 ########## TO DO #############
 # Write more things.
-# Check lane change car results (ongoing on slurm)
-# Built the Neural Network (find why loss goes to nan)
+# Build the Neural Network (find why loss goes to nan)
+# Add a constant (non-zero) acceleration naive model
+# Add a homogenous behaviour AR2 Model (estimate on 2000 car sample and apply to new cars)
 
 library(tidyverse)
 library(Rcpp)
@@ -488,12 +489,38 @@ sliceSampler{
 
 forecastResults {
 
+library(readr)
 results <- NULL
 for(i in seq_along(id$idfc)){
   try(assign('results',
              rbind(results,
-                   read.csv(paste0('eval/car', id$idfc[i], '.csv')))))
+                   read_csv(paste0('eval/vb/car', id$idfc[i], '.csv'), col_types = cols()))))
   if((i %% 100 == 0) | i == length(id$idfc)){
+    print(i)
+  }
+}
+for(i in 1:500){
+  try(assign('results',
+             rbind(results,
+                   read_csv(paste0('eval/vb/car', noStopChanged[i], '.csv'), col_types = cols()))))
+  if(i %% 100 == 0){
+    print(i)
+  }
+}
+naiveResults <- NULL
+for(i in seq_along(id$idfc)){
+  try(assign('naiveResults',
+             rbind(naiveResults,
+                   read_csv(paste0('eval/naive/car', id$idfc[i], '.csv'), col_types = cols()))))
+  if((i %% 100 == 0) | i == length(id$idfc)){
+    print(i)
+  }
+}
+for(i in 1:500){
+  try(assign('naiveResults',
+             rbind(naiveResults,
+                   read_csv(paste0('eval/naive/car', noStopChanged[i], '.csv'), col_types = cols()))))
+  if(i %% 100 == 0){
     print(i)
   }
 }
@@ -503,6 +530,33 @@ results %>%
   mutate(logscore = logscore + log(mPerFoot^2),
          mapDist = mapDist * mPerFoot,
          consDist = consDist * mPerFoot) -> results
+naiveResults[,1:6] <- naiveResults[,1:6] * mPerFoot
+
+naiveResults %>%
+  filter(id %in% unique(results$id)) %>%
+  gather(model, error, -h, -S, -id) %>%
+  rbind(results %>%
[…]

### 2018-02-07 Cars/testSlurm/Homogenous.R

@@ -1,0 +1,56 @@
+rm(list=ls())
+repenv <- Sys.getenv("SLURM_ARRAY_TASK_ID")
+i <- as.numeric(repenv)
+
+ids <- readRDS('carsID.RDS')
+
+if(i <= 873){
+  datafc <- readRDS('ForecastData.RDS')
+  data <- datafc[[i]]
+  id <- ids$idfc[i]
+} else {
+  datafc <- readRDS('fcDataChanged.RDS')
+  data <- datafc[[i-873]]
+  id <- datafc[[501]][i-873]
+}
+
+draws <- readRDS('noHierN2000.RDS')$draws
+draws[,1:2] <- exp(draws[,1:2])
+
+H <- 30
+S <- 10
+maxT <- 300
+
+sSeq <- seq(S, maxT, S)
+aLower <- min(data[,1])
+if(aLower < 0){
+  aLower <- 2 * aLower
+} else {
+  aLower <- aLower - 0.5
+}
+dLower <- min(data[,2])
+if(dLower < 0){
+  dLower <- 2 * dLower
+} else {
+  dLower <- 0.5 * dLower
+}
+
+asup <- seq(aLower, 2*max(data[,1]), length.out = 200)
+dsup <- seq(dLower, 2*max(data[,2]), length.out=200)
+grid <- cbind(asup, dsup)
+
+results <- data.frame()
+
+# Incrementally add data
+for(s in seq_along(sSeq)){
+  if((sSeq[s] + H) > nrow(data)){
+    break
+  }
+  MCMCDens(data = data[(sSeq[s]-1):(sSeq[s] + H), ],
+           N = 200,
+           H = H,
+           grid = grid,
+           MCMCdraws = draws)
+}
+
+write.csv(results, paste0('homog/car', id, '.csv'), row.names=FALSE)

### 2018-02-08 Cars/paper.tex
misc

@@
 
-\subsubsection{Neural Networks}
+\subsubsection{Recurrent Neural Networks}
 
-Following Ding et al.
-(2012), a Neural Network is constructed with inputs given by the ten most recent lags of vehicle position, angle, and velocity, and an output given by the next thirty values of the vehicle position.
-The network is trained with the Adam optimised with a mean Euclidean error loss function.
+A Recurrent Neural Network is constructed with inputs given by the ten most recent lags of vehicle position, angle, and velocity, and an output given by the next thirty values of the vehicle position.
+The network is trained with the Adam optimiser and a mean Euclidean error loss function.
 \subsection{Results}

### 2018-02-12 Cars/paper.tex
updates

@@
 \item It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a trajectory forecast.
-\item These time series models (In this case a bivariate AR2) fit a parameter vector to capture the driver behaviour, which can depend both on the be very different from driver to driver (Chen 2017) which may depend on a particular vehicle's position relative to surrounding traffic.
+\item A homogenous time series model shows a strong ability to predict the MAP trajectory
+\itme An important part of this modelling is to capture the uncertainty of car trajectories, as the self-driving car must react to the possibility of collision even when that possibility is relatively small.
+\item There are many sources of heterogeneity in drivers, from their individual driving styles to the conditions of traffic directly around them.
+\item Including heterogeneity in the model will allow more dangerous vehicles to be identified, for example, 'lead-footed' drivers can have a larger variance on their acceleration.
 \item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
 \item This model implies that any new vehicle observed by the self driving car has its own parameter vector, related to previously observed vehicles through the shared prior.
-\item The self driving car is constantly observing additional information about the vehicles around it.
-Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
-\item A three step strategy is employed to forecast the future trajectory of newly observed vehicles:
+\item The self driving car must be able to estimate the parameter vector of newly observed cars, and update this parameter vector as it observes additional data.
+\item Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
+\item A three step strategy is employed to provide density forecasts for the future trajectory of newly observed vehicles:
 \begin{enumerate}
@@
 \end{enumerate}
-\item This paper introduces a Variational Bayes updating method for the third step, where the posterior distribution of the new vehicles parameter vector is approximated, then this approximation is updated.
+\item This paper uses standard Variational Bayes for the second step, and introduces an updating method for the third step; the posterior distribution of the new vehicles parameter vector is approximated, then this approximation is updated.
 \item This only uses the data observed in-between updates, allowing the previously observed history of that vehicle to be discarded.
@@
 \end{enumerate}
-\item Compare Bayes MAP point estimates to a neural network and constant acceleration / angle model.
+\item Compare Bayes MAP point estimates to a neural network and naive models
 \item Error Metric: Euclidean Distance from predicted position to actual position
@@
 \begin{itemize}
-\item Dynamic AR 2 models forecasting future acceleration and angles drastically improves forecasts of the future trajectory compared to the naive models.
-\item Treating vehicles as heterogenous allows behaviour specific to an individual driver to be learned, further improving forecasts.
+\item Dynamic AR 2 models forecasting future acceleration and angles drastically improves forecasts of the future trajectory compared to the competing models.
+\item Treating vehicles as heterogenous allows behaviour specific to an individual driver to be learned, improving logscores and forecast uncertainty.
 \item This requires constant models fits and updates, which can be made computationally feasible by updating previous Variational Bayes model fits, which incurs only a small performance hit relative to the more expensive standard VB and MCMC methods.

### 2018-02-12 Cars/testSlurm/Homogenous.R

@@ -1,8 +1,23 @@
 rm(list=ls())
 repenv <- Sys.getenv("SLURM_ARRAY_TASK_ID")
-i <- as.numeric(repenv)
+iter <- as.numeric(repenv)
+
+library(Rcpp, lib.loc = 'packages')
+library(RcppArmadillo, lib.loc = 'packages')
+library(RcppEigen, lib.loc = 'packages')
+library(rstan, lib.loc = 'packages')
+source('slurmRFuns.R')
+sourceCpp('slurmCppFuns.cpp')
+
 
 ids <- readRDS('carsID.RDS')
+
+for(j in 1:10){
+ results <- data.frame()
+ if(iter == 134 & j > 7){
+   break
+ }
+ i <- (iter-1)*10 + j
 
 if(i <= 873){
   datafc <- readRDS('ForecastData.RDS')
@@ -15,7 +30,6 @@ if(i <= 873){
 }
 
 draws <- readRDS('noHierN2000.RDS')$draws
-draws[,1:2] <- exp(draws[,1:2])
 
 H <- 30
 S <- 10
@@ -39,18 +53,30 @@ asup <- seq(aLower, 2*max(data[,1]), length.out = 200)
 dsup <- seq(dLower, 2*max(data[,2]), length.out=200)
 grid <- cbind(asup, dsup)
 
-results <- data.frame()
 
 # Incrementally add data
 for(s in seq_along(sSeq)){
   if((sSeq[s] + H) > nrow(data)){
     break
   }
-  MCMCDens(data = data[(sSeq[s]-1):(sSeq[s] + H), ],
+  output <-  MCMCDens(data = data[(sSeq[s]-1):(sSeq[s] + H), ],
            N = 200,
            H = H,
            grid = grid,
            MCMCdraws = draws)
+
+ results <- rbind(results,
+                 data.frame(logscore = output$logscore,
+                            xCDF = output$xCDF,
+                            mapDist = output$mapDist,
+                            consDist = NA,
+                            h = 1:H,
+                            prior = 'Homogenous',
+                            method = 'MCMC',
+                            S = sSeq[s],
+                            id = id))
 }
 
+
 write.csv(results, paste0('homog/car', id, '.csv'), row.names=FALSE)
+}

### 2018-02-12 Cars/testSlurm/slurmFCAll.R

@@ -3,23 +3,33 @@ repenv <- Sys.getenv("SLURM_ARRAY_TASK_ID")
 i <- as.numeric(repenv)
 set.seed(1000 + i)
 
-library(Rcpp, lib.loc = 'packages')
-library(RcppArmadillo, lib.loc = 'packages')
-library(RcppEigen, lib.loc = 'packages')
-library(rstan, lib.loc = 'packages')
+library(Rcpp)#, lib.loc = 'packages')
+library(RcppArmadillo)#, lib.loc = 'packages')
+library(RcppEigen)#, lib.loc = 'packages')
+library(rstan)#, lib.loc = 'packages')
 source('slurmRFuns.R')
 sourceCpp('slurmCppFuns.cpp')
 
-id <- readRDS('carsID.RDS')
-datafc <- readRDS('ForecastData.RDS')
-prior <- readRDS('prior.RDS')
-H <- 10
-S <- 10
-maxT <- 300
+ids <- readRDS('carsID.RDS')
 
-sSeq <- seq(S, maxT, S)
+if(i <= 873){
+  datafc <- readRDS('ForecastData.RDS')
+  data <- datafc[[i]]
+  id <- ids$idfc[i]
+} else {
+  datafc <- readRDS('fcDataChanged.RDS')
+  data <- datafc[[i-873]]
+  id <- datafc[[501]][i-873]
+}
+homogDraws <- readRDS('noHierN2000.RDS')$draws
+prior <- readRDS('prior.RDS')
+H <- 30
+S <- 10
+maxT <- 450
+
+sSeq <- seq(100, maxT, S)
 results <- data.frame()
-methods <- c('None', 'Single Hierarchy', 'Finite Mixture')
+methods <- c('Non-Informative', 'Single Hierarchy', 'Finite Mixture')
 
 # MCMC Hyper parameters
 hyper <- list()
@@ -50,11 +60,6 @@ prior[[3]] <- matrix(priorDiag, ncol = 1)
 starting <- list(matrix(c(-5, -5, rep(0, 4), c(chol(diag(0.5, 6)))), ncol = 1),
                  prior[[3]])
 
-
-
-# Extract Data
-data <- datafc[[i]]
-
 # Set forcast supports
 aLower <- min(data[,1])
 if(aLower < 0){
@@ -69,8 +74,8 @@ if(dLower < 0){
   dLower <- 0.5 * dLower
 }
 
[…]

### 2018-02-14 Cars/paper.tex
added split by variances

@@
 
-\title{Online Updating of Variational Bayes for Forecasts of Vehicle Trajectory with Heterogeneous Driver Behaviour}
+\title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}
 \begin{document}
@@
 \item There are many sources of heterogeneity in drivers, from their individual driving styles to the conditions of traffic directly around them.
-\item Including heterogeneity in the model will allow more dangerous vehicles to be identified, for example, 'lead-footed' drivers can have a larger variance on their acceleration.
+\item Including heterogeneity in the model will allow the model to better adapt to drivers with more extreme behaviours, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
 \item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
@@
 \end{enumerate}
-\item Compare Bayes MAP point estimates to a neural network and naive models
-\item Error Metric: Euclidean Distance from predicted position to actual position
-\item Compare log scores from streaming and standard VB forecast densities to see how the updating method compares to standard VB.
-I am also using MCMC, so we can compare either VB method to this, theoretically MCMC $\geq$ VB Standard $\geq$ VB Updating
+\item Show that a homogenous time series model has strong predictive performance compared to constant models / neural networks while also providing density forecasts.
+\item Build a heterogenous version of the homogenous model, show that the density forecasts improve. 
+\item Show that the Updating VB required to implement the heterogenous model in real time is comparable to MCMC and standard VB.
 \end{itemize}
@@
 
+Summary:
 \begin{itemize}
-\item 4071 vehicles are available in the sample
+\item The homogenous time series model has a significantly lower MAP error than other models
+\item The heterogenous time series models have a similar MAP error to the homogenous model - the maximum of the forecast distribution is in a similar place.
+\item The logscores for the heterogenous models are generally greater than the homogenous model indicating there is less uncertainty in the forecast distribution
+ \item Updating VB has a similar logscore to MCMC across all priors
+ \item Priors don't really matter too much
+ \item The high precision in the homogenous posterior does make it better than the heterogenous models in small samples.
+ \item Performance between homogenous and heterogenous becomes a tradeoff between high precision / biased mean vs. low precision, more accurate mean in the posterior.
+ \item The heterogenous is sufficiently precise to outperform the homogenous model in terms of logscores after $T = 100$ (ten seconds).
+ \item Most of the differences are due to the posterior of the variance parameters
+ \item The homogenous model has a large variance for both variables, greater than $80\%$ and $92\%$ of heterogenous variances for acceleration and angle respectively.
+ \item This is likely caused by a few drivers with high variance skewing the results in the homogenous model.
+ \item The heterogenous model allows consistent drivers to have a low variance, and it performs very well in this case.
+ \item Heterogenous models being able to assign a high variance allows more erratic / risky drivers to be identified.
+ \end{itemize}
+
+\begin{itemize}
+\item 3873 vehicles are available in the sample
 \item 500 are used to estimate splines and straighten the road to relative coordinates as described in Section 2
 \item 2000 are used to train any models used
-\item 1571 are available for testing
+\item 1373 are available for testing
 \item Each second (10 observations) models are fit to the sample of $z_{N+1, 1:T}$ and forecasts of $z_{N+1, T+1:T+30}$ are made, corresponding to a three second forecast horizon

### 2018-02-14 Cars/HierMixMod.R

@@ -489,48 +489,23 @@ sliceSampler{
 
 forecastResults {
 
+  
+# Missing car 20771, car 20783, car 20813, changed 465 onwards
 library(readr)
 results <- NULL
 for(i in seq_along(id$idfc)){
-  try(assign('results',
-             rbind(results,
-                   read_csv(paste0('eval/vb/car', id$idfc[i], '.csv'), col_types = cols()))))
+  if(i %in% c(687, 690, 699)){
+    next
+  }
+  results <- rbind(results,
+                   read_csv(paste0('forecasts/car', id$idfc[i], '.csv'), col_types = cols()))
   if((i %% 100 == 0) | i == length(id$idfc)){
     print(i)
   }
 }
 for(i in 1:500){
   results <- rbind(results,
-                   read_csv(paste0('eval/vb/car', noStopChanged[i], '.csv'), col_types = cols()))
-  if(i %% 100 == 0){
-    print(i)
-  }
-}
-naiveResults <- NULL
-for(i in seq_along(id$idfc)){
-  naiveResults <-  rbind(naiveResults,
-                         read_csv(paste0('eval/naive/car', id$idfc[i], '.csv'), col_types = cols()))
-  if((i %% 100 == 0) | i == length(id$idfc)){
-    print(i)
-  }
-}
-for(i in 1:500){
-  naiveResults <- rbind(naiveResults,
-                        read_csv(paste0('eval/naive/car', noStopChanged[i], '.csv'), col_types = cols()))
-  if(i %% 100 == 0){
-    print(i)
-  }
-}
-for(i in seq_along(id$idfc)){
-  homogResults<-  rbind(homogResults,
-                         read_csv(paste0('eval/homog/car', id$idfc[i], '.csv'), col_types = cols()))
-  if((i %% 100 == 0) | i == length(id$idfc)){
-    print(i)
-  }
-}
-for(i in 1:500){
-  homogResults <- rbind(homogResults,
-                        read_csv(paste0('eval/homog/car', noStopChanged[i], '.csv'), col_types = cols()))
+                   read_csv(paste0('forecasts/car', noStopChanged[i], '.csv'), col_types = cols()))
   if(i %% 100 == 0){
     print(i)
   }
@@ -538,36 +513,18 @@ for(i in 1:500){
 
 mPerFoot <- 0.3048
 results %>% 
[…]

### 2018-02-27 Cars/paper.tex
Paper Revision

--- 2018-02-12 Cars/paper.tex
@@
 
+\def\app#1#2{%
+  \mathrel{%
+    \setbox0=\hbox{$#1\sim$}%
+    \setbox2=\hbox{%
+      \rlap{\hbox{$#1\propto$}}%
+      \lower1.1\ht0\box0%
+    }%
+    \raise0.25\ht2\box2%
+  }%
+}
+\def\approxprop{\mathpalette\app\relax}
+
 \setlength{\topmargin}{0in}
@@
 
-\title{Online Updating of Variational Bayes for Forecasts of Vehicle Trajectory with Heterogeneous Driver Behaviour}
+\title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}
+
+%TODO:
+%Flesh out section 1
+%Add authors, institutions, an abstract and bibtex references 
+%Switch out manual references for Bibtex when everything else is done.
+
 \begin{document}
 \maketitle
+
 \section{Introduction}
+\label{sec:intro}
 
 \begin{itemize}
-\item Self driving Cars are becoming more common, eg Waymo, Tesla, GM.
-\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the Self Driving Car.
-\item While Self Driving Cars reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
+\item Self driving vehicles are becoming more common, eg Waymo, Tesla, GM.
+\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the self driving vehicle.
+\item While self driving vehicles reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
 \item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg.
-Cameras and LiDAR), such as their position, velocity, and steering angle (Woo et al.
-2016a; Tian, Pei, Jana, and Ray 2017).
-\item The Self Driving Car forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
-\item These can assume the surrounding vehicles will maintain their current angle and velocity (Gindele, Brechtel, Dillmann 2010; Waymo Safety Report 2017; Bautista 2017), but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines (Woo et al.
-201b6; Ding et al.
-2012; Woo et al.
-2017; Geng et al.
-2017).
+Cameras and LiDAR), such as their position, velocity, and steering angle, see \citet{Woo2016a} and \citet{Tian2017}.
+\item The self driving vehicle forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
+\item These can assume the surrounding vehicles will maintain their current angle and velocity \citet{Gindele2010}
+\citet{Waymo2017}, \citet{Bautista2017}, but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citet{Woo2016b}, \citet{Ding2013}, \citet{Woo2017}, \citet{Geng2017}.
 \item This paper takes a statistical approach, fitting time series models to two extracted variables, velocity and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
@@
 \item A homogenous time series model shows a strong ability to predict the MAP trajectory
-\itme An important part of this modelling is to capture the uncertainty of car trajectories, as the self-driving car must react to the possibility of collision even when that possibility is relatively small.
+\item An important part of this modelling is to capture the uncertainty of car trajectories, as the self driving vehicle must react to the possibility of collision even when that possibility is relatively small.
 \item There are many sources of heterogeneity in drivers, from their individual driving styles to the conditions of traffic directly around them.
-\item Including heterogeneity in the model will allow more dangerous vehicles to be identified, for example, 'lead-footed' drivers can have a larger variance on their acceleration.
+\item Including heterogeneity in the model will allow the model to better adapt to drivers with more extreme behaviours, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
 \item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
-\item This model implies that any new vehicle observed by the self driving car has its own parameter vector, related to previously observed vehicles through the shared prior.
-\item The self driving car must be able to estimate the parameter vector of newly observed cars, and update this parameter vector as it observes additional data.
+\item This model implies that any new vehicle observed by the self driving vehicle has its own parameter vector, related to previously observed vehicles through the shared prior.
+\item The self driving vehicle must be able to estimate the parameter vector of newly observed cars, and update this parameter vector as it observes additional data.
 \item Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
@@
 \item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution.
-This can occur before the self driving car is on the road.
+This can occur before the self driving vehicle is on the road.
 \item When a new vehicle is observed use the information in this shared distribution to model the associated new parameter vector. 
@@
 \item The model can be updated faster as less data must be processed, so forecasts can be made using as much recent data as possible.
-\item Self driving cars take in up to 1GB of data a second (several websites report this as quotes from different people eg.
+\item Self driving vehicles take in up to 1GB of data a second (several websites report this as quotes from different people eg.
 Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.
 \end{enumerate}
-\item Compare Bayes MAP point estimates to a neural network and naive models
-\item Error Metric: Euclidean Distance from predicted position to actual position
-\item Compare log scores from streaming and standard VB forecast densities to see how the updating method compares to standard VB.
-I am also using MCMC, so we can compare either VB method to this, theoretically MCMC $\geq$ VB Standard $\geq$ VB Updating
+\item Show that a homogenous time series model has strong predictive performance compared to constant models / neural networks while also providing density forecasts.
+\item Build a heterogenous version of the homogenous model, show that the density forecasts improve. 
+\item Show that the Updating VB required to implement the heterogenous model in real time is comparable to MCMC and standard VB.
 \end{itemize}
@@
 \section{Data processing}
+\label{sec:dataProcessing}
+Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 Freeway in Los Angeles, California from 7:50 am to 8:35 am on June 15th, 2005.
+Data was collected by seven static cameras and processed by Cambridge Systematics Inc.
+The result is coordinates relative to the start of the road section for 6101 vehicles collected every 100 milliseconds.
+The average number of observations per vehicle is 672, corresponding to a 67.2 seconds of data.
+Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which travel along the five major lanes or the enter from the entry/exit lane to the right, with one vehicle changing from Lane 2 to Lane 1.
+Vehicles that entered or exited midway through the freeway section are excluded in this paper.
+There is a curvature to the road, with bends occuring between 500 and 1000 feet and 1800 and 2000 feet.
 
-\begin{itemize}
-\item Data is provided by the Next Generation Simulation (NGSIM) project, which observes cars for a 2000 feet section of the US 101 Highway.
-\item Observations of x/y coordinates are provided every 100 ms (10 hz), with around 500-1000 observations per car.
-\item Around 2000 of these cars changed lanes or stopped moving somewhere in the sample and are discarded.
-\item There was some curvature in the road which means there are systematic movements in the car paths.
-\item Modern cars are able to track lane marking and can calculate distance to other cars in relative co-ordinates (Thuy and Leon 2010), instead of relative to the Earth (ie. ignores curves)
-\item Prior research fits a polynomial curve to detected lane markings to build a model of the road lane edges (Woo et. al. 2016) 
-\item Instead, I fit splines to the positions of each car to build an estimate of the road midpoint.
-Relative co-ordinates are found separately for each lane using the movements of 100 cars / lane by the following procedure:
-\begin{enumerate}
-\item At each point in time $t$ calculate the total distance each car has travelled along the road using $d_{i, t} = \sum_{s=1}^t v_{i, s}$, where $v_{i, s}$ is the velocity of car $i$ at time $s$. 
-\item Estimate independent smoothing splines of the form $x_{i, t} = f(d_{i, t})$ and $y_{i, t} = g(d_{i, t})$, where $x_{i, t}$ and $y_{i, t}$ are the car coordinates relative to the start of the road.
-\item Use models to get $\hat{x}_{i, t}$ and $\hat{y}_{i, t}$ as estimates of the midpoint of the lane after travelling a distance $d_{i, t}$.
-\item Define new coordinates 
+\begin{figure}
+\centering
+\includegraphics[width=0.65\textwidth]{carPath}
+\caption{The path of ten vehicles in the dataset.
+The freeway is split into five lanes, and an additional entry/exit lane to the far right, through which two vehicles from this sample have entered.
+There is curvature to the road slightly distored by the aspect ratio, with bends occuring around 500 and 1800 feet.}
+\label{fig:rawData}
+\end{figure}
+
+Modern vehicles are capable of tracking lane marking, and thus can compensate for the curvature of the road as they calculate the location of surrounding vehicles in coordinates relative to their own position (Thuy and Leon 2010), rather than relative to a static point as provided by NGSIM.
+To compensate for this, and so that the data is in a coordinate system similar to that of a self driving vehicle tracking surrounding cars, the coordinate system provided in the NGSIM dataset is transformed into a relative system where the road curvature is removed.
+This is facilitated by measuring the location of each vehicle relative to the estimate of the midpoint of each lane.
+\citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to build a model of the lane midpoints for the entire road section.
+
+The midpoints for each of the five lanes are estimated separately, using 100 randomly sampled vehicles each that did not change their lane over the observed period.
+For each sampled vehicle $i$ and time since entering the road $t$, the total distance travelled is calcualted as
+\begin{equation}
+\label{distance}
+d_{i, t} = \sqrt{(x_{i, t}) - x_{i, 1})^2 + (y_{i, t} - y_{i, 1})^2}
+\end{equation}
+From this distance metric, the average $x$ and $y$ coordinates at each given lane and for a given distance are each estimated by a pair of smoothing splines calculated with the R `stats' package.
+The smoothing splines are functions of the distance travelled and estimate the lane midpoints as  $\hat{x}_{i, t} = f(d_{i, t})$ and $\hat{y}_{i, t} = g(d_{i, t})$.
+Each remaining vehicle in the dataset uses the lane midpoint estimates $\{\hat{x}^*_{i, t}, \hat{y}^*_{i, t}\}$ fit from the spline model assocaiated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the devitation from the lane midpoint, by
 \begin{align}
-x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right) - \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
-y^*_{i, t} &= d_{i, t} \label{yRel}
+x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right) - \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
+y^*_{i, t} &= d_{i, t}. \label{yRel}
 \end{align}
-\end{enumerate}
-\item (\ref{xRel}) \textit{really} needs a diagram.
-\item These cars are not used for any further modelling, leaving 2873 cars travelling on a straightened out road with no noticable systematic curvature.
-\end{itemize}
+The vehicles used to estimate the five spline models are excluded for any other purpose.
+\begin{figure}
+\centering
+\includegraphics[width = 0.5\textwidth]{relCoord}
+\caption{An example of the equation in (\ref{xRel}).
+The estimated trajectory of the lane midpoint is given by the solid curve, while a vehicle $i$ at a given point in time $t$ is given by the blue dot at $\{x_{i, t}, y_{i, t}\}$.
+This vehicle has travelled a distance equivalent to the red dot at $\{\hat{x}_{i, t}, \hat{y}_{i, t}\}$.
+The gradient of the midpoint at this point is given by the dashed line, which intercepts the x-axis at an angle of $\lambda = \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)$.
+The dotted line from the blue dot to the red intercepts the x-axis at an angle of $\psi = \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)$.
+The sign of the angle $\lambda - \psi$ determines whether the vehicle is to the left or right side of the lane, and thus has a negative or positive relative coordinate $x^*_{i, t}$.
+The absolute value of $x^*_{i, t}$ is the distance between the two dots.}
+\label{fig:relCoord}
+\end{figure}
+Once relative coordinates are extracted, the relative position of any vehicle $i$ at any time $t$ is determined by its initial position $\{x^*_{i, 1}, y^*_{i, 1}\}$, as well as the history of that driver's inputs: velocity $\{v_{i, s} | s = 2, \dots, t\}$ and steering angle, $\{\delta_{i, s} | s = 2, \dots, t\}$ by the trigonometric relationship in Figure \ref{fig:motion}:
+\begin{align}
+x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
+y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
+\end{align}
+where $\delta_{i, t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{i, t} = x^*_{i, t-1}$.
+From this relationship, the driver's inputs are be extracted by
+\begin{align}
+\delta_{i, t} &= \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right) \label{dEq} \\
+v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
+\end{align}
+It is then useful to extract the acceleration, $a_{i, t}$ by the simple relation
+\begin{equation}
+\label{aEq}
+a_{i, t} = v_{i, t} - v_{i, t-1}. 
+\end{equation}
+Data for vehicle $i$ up to time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $z_{1:N} = \{z_{i, 1:T_i} | i = 1, \dots, N\}$.
+\begin{figure}
+\centering
+\includegraphics[width = 0.4\textwidth]{motion}
+\caption{A vehicle at $x_{t-1}, y_{t-1}$ at time $t-1$ will be at $x_t, y_t$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
+\label{fig:motion}
+\end{figure}
 
-\section{Hierarchical Motion Model}
+\section{Trajectory Forecasts}
+\label{sec:trajForecasts}
 
-The position of any vehicle at time $t$ can be fully determined by its initial position, $\{x^*_{i, 0}, y^*_{i, 0}\}$ and the history of the driver's inputs: the vehicle's velocity $v_{i, t}$ and steering angle, $\delta_{i, t}$ by
+A crucial step in the safe operation of self driving vehicles is to forecast the future trajectory of surrounding vehicles so that collisions can be avoided.
+To produce these forecasts the data is split to place vehicles into two sets: a training set used to infer the posterior distribution of a parameter set before the self driving vehicle is on the road; and a test set used for forecasting the positions of vehicles observed while driving.
+The parameters associated with the vehicles in the test set may be inferred, but this inference is time constrained.
+If inference takes $M$ seconds, a forecast after observing a vehicle for $T$ seconds can only include data up to time $T - M$ to infer the posterior distribution of any parameters.
+
+The remainder of Section \ref{sec:trajForecasts} will introduce the models used, while Sections \ref{sec:Inference} and \ref{sec:UVB} will build an inferential framework to facilitate inference of any additional parameters in constant time as $T$ increases to apply to online trajectory forecasting.
+
+\subsection{Homogenous Time Series Model}
+\label{subsec:homogenous}
+
+The framework provided by (\ref{xEq}) and (\ref{yEq}) allows forecasts of $\{x^*_{t}, y^*_{t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_T, y^*_T, v_T\}$, and forecasts of the future values of
+$\{a_{t}, \delta_{t} | t = T + 1, \dots, T+H\}$.
+
+As data is observed every 100 milliseconds, large changes in $a_{i, t}$ and $\delta_{i, t}$ can occur over multiple time periods.
+This can be observed in Figure \ref{fig:dynamics}, where the left panel plots the mean value of
+\begin{equation}
+A_{i, t} = \frac{a_{i, \tau_i + t}}{a_{i, \tau_i}}
+\label{amax}
+\end{equation}
+where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$.
+On average, a vehicle will increase in acceleration for the 400 milliseconds before reaching maximum acceleration.
+The right panel plots the mean value of
+\begin{equation}
+D_{i, t} = \frac{\delta_{i, \omega_i + t} - \pi/2}{\delta_{i, \omega_i} - \pi/2}
+\label{dmax}
+\end{equation}
+where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, which similarly demonstrates that large absolute deviations in steering angle from $pi/2$ are associated with smaller deviations in the same direction in the the previous 300 milliseconds.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width = 0.95\textwidth]{dynamics}
+\caption{Left: Mean values of acceleration as a percent of a vehicles maximum absolute acceleration for the 1000 milliseconds before, and 500 milliseconds after, a vehicle reaches its maximum absolute acceleration.
+Right: The analogous plot for the steering angle, with $\delta_t - \pi/2$ as a percentage of a vehicles maximum absolute value of $\delta_t - \pi/2$.
+The few observations before a variable reaches its maximum often have increasing values for the same variable.}
+\label{fig:dynamics}
+\end{figure}
+
+These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and thus auto-regressive processes of orders $p$ and $q$ for $a_t$ and $\delta_t$ are employed to model this, with
 \begin{align}
-x^*_{t} &= x^*_{t-1} + v_{t} \cos(\delta_{t}) \label{xEq} \\
-y^*_{t} &= y^*_{t-1} + v_{t} \sin(\delta_{t}) \label{yEq},
+a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{aAR} \\
+\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{dAR}
 \end{align}
-where $\delta_{t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{t} = x^*_{t-1}$.
-From this relationship, the driver's key inputs to motion can be extracted by
-\begin{align}
-\delta_{t} &= \tan^{-1}\left(\frac{(y^*_{t} - y^*_{t-1})}{(x^*_{t} - x^*_{t-1})} \right) \label{dEq} \\
-v_{t} &= \sqrt{(x^*_{t} - x^*_{t-1})^2 + (y^*_{t} - y^*_{t-1})^2} \label{vEq} \\
-a_{t} &= v_{t} - v_{t-1}. \label{aEq}
-\end{align}
-where $a_{t}$ denotes the acceleration at time $t$.
+for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution.
 
-Auto-regressive processes of orders $p$ and $q$ for $a_t$ and  $\delta_t$ are used to model the level of persistance and variation of driver actions, with
-\begin{align}
-a_{t} &= \sum_{j = 1}^p \phi_{j} a_{t-j} + \sigma_{\epsilon} \epsilon_{t} \label{aAR} \\
-\delta_{t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{t-j} - \pi/2) + \sigma_{\eta} \eta_{t} \label{dAR}
-\end{align}
-where $\epsilon_{t}$ and $\eta_t$ are independently and identically distributed according to a standard normal distribution.
-The unconditional means of $a_t$ and $\delta_t$ are set to zero and $\pi/2$ respectively, corresponding to forward motion at constant velocity.
-The parameter vector to be estimated is collected as
+The parameter vector is transformed to $\mathbb{R}^{p + q + 2}$ and collected as 
 \begin{equation*}
 \label{thetaVec}
-\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\} \in \mathbb{R}^{p + q + 2}.
+\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\},
 \end{equation*}
+where
+\begin{equation}
+\label{indPrior}
+p(\theta) \sim \mathcal{N}\left(\mu, \Sigma \right)
+\end{equation}
+where $\mu$ is a vector with an entry of $0$ for each $\phi$ and $\gamma$, and an entry of $-5$ for each log-variance, as the scale of changes in the data is very small.
+$\Sigma = 10 \mathbb{I}$, where $\mathbb{I}$ is the identity matrix.
+This is referred to as the homogenous model throughout the paper.
 
-Multiple groups of differing driver behaviour can be allowed while still retaining heterogeneity within each group by jointly estimating each vehicle driver's individual parameter vector, $\theta_i$, as part of a heirarchical model.
-In this case each $\theta_i$ is treated as an independent draw from a $K$ component mixture of multivariate normal distributions, augmenting each $\theta_i$ with a mixture component $k_i = 1, \dots, K$ such that
+\subsection{Heterogenous Time Series Models}
+\label{subsec:heterogenous}
+
+The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them.
+Heterogeneity is allowed by replacing the shared $\theta$ parameter vector with $N$ individual parameter vectors $\theta_i$ so that
+\begin{align}
+a_{i, t} &= \sum_{j = 1}^p \phi_{i, j} a_{i, t-j} + \sigma_{\epsilon, i} \epsilon_{i, t} \label{aAR2} \\
+\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{i, j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta, i} \eta_{i, t}. \label{dAR2}
+\end{align}
+Using the prior in (\ref{indPrior}) for each $\theta_i$ formulates what is referred to as the independent model where each vehicle posterior is independent of all other vehicles, so $p(\theta_i | z_{1:N}, \theta_{j \neq i}) = p(\theta_i | z_i)$.
+An alternative approach is to augment the prior distribution of each $\theta_i$ with parameters $\beta$ in a hierarchical model, and infer the distribution of the $p(\theta_{1:N}, \beta | z_{1:N})$.
+The top panel of Figure \ref{fig:HierSingleKDE} plots a kernel density estimate of the posterior means of each $\theta_i, i = 1, \dots, 2000$ using the independent model with $p = q = 2$ at time $T = 500$.
+There is a wide range of values for the posterior mean of each variable, indicating heterogeneity across the drivers in the sample.
+The density of posterior means for both variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the density of posterior means of both $\phi_1$ and $\phi_2$ are multimodal.
+Due these features, a $K$ component mixture of multivariate normal distributions is chosen for $p(\theta_i | \beta)$, augmenting each driver with an auxiliary variable $k_i = 1, \dots, K$ such that
 \begin{equation}
-\theta_i | k_i = j \sim N(\mu_j, \Sigma_j).
+\label{mixPrior}
+\theta_i | k_i = j \sim N(\mu_j, \Sigma_j),
 \end{equation}
-For each $j = 1, \dots, K$, multivariate normal priors are chosen for $\mu_j$ with fixed mean $\bar{\mu}_j$ and variance matrix $\Omega_j$, Inverse Wishart priors are chosen for $\Sigma_j$ with fixed degrees of freedom $\tau_j$ and scale matrix $\Psi_j$.
-$k_{i}$ has a multinomial distribution with mixture probabilities $\boldsymbol{\pi} = \{\pi_{1}, \dots, \pi_{K}\}$, where $\boldsymbol{\pi}$ has a Dirichlet prior with fixed hyperparameters $\alpha = \alpha_j, j = 1 , \dots, K$.
-Defining $\beta = \{\mu_j, \Sigma_j, \pi_j, j = 1, \dots, K\}$, $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s} | s = 1, \dots, T\}$ and $z_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$ the posterior distribution $p(\theta_1, \dots, \theta_N, \beta | z_{1:N})$ can be estimated with MCMC methods.
+and
+\begin{equation}
+k_i \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
+\end{equation}
+for each $i = 1, \dots, N$. Furthurmore, for each $j = 1, \dots, K$,  
+\begin{align}
+\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
+\Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
+\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right)
+\end{align}
+The elements of $\beta$ are $\{\mu_j, \Sigma_j, \pi_j | j = 1, \dots, K\}$.
+Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$.
+The bottom panel of Figure \ref{fig:HierSingleKDE} plots the marginal distributions for $p(\theta_{2001} | \beta, z_{1:2000})$ for the hierarchical model with $p = q = 2$ and $K = 6$ using the same vehicles as the independent model in the top panel.
+The marginal distributions capture the skewness and multimodality present in the densities of posterior means from the independent model.
 
@@
 \includegraphics[width = 0.95\textwidth]{HierSingleKDE.png}
-\caption{Top: A Kernel Density Estimate applied to the individual posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ where $p = q = 2$, modelled independently without any hierarchical structure.
-Bottom: $p(\theta_i | \beta, z_{1:2000})$ for a hierarchical mixture model with $K = 6$ using the same vehicles, with each component of the mixture denoted by the coloured densities.
-The hierarchical model has captured the bimodality and skewness present in the independent posterior means.}
+\caption{Top: A Kernel Density Estimate applied to the posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ where $p = q = 2$ from the independent model, sampled by Metropolis-Hasting MCMC.
+Bottom: $p(\theta_i | \beta, z_{1:2000})$ for the hierarchical model with $K = 6$ using the same vehicles sampled by Metropolis-Hasting MCMC.
+Each individual component in the mixture is denoted by the coloured densities.
+The hierarchical model has captured the multimodality and skewness present in the independent posterior means.}
 \label{fig:HierSingleKDE}
@@
 
-\section{Variational Bayes}
+\subsection{Encountering Additional Vehicles}
+\label{subsec:additionalVehicles}
 
-\subsection{Background}
+For the remainder of this paper $z_{1:N}$ and $\theta_{1:N}$ refers to the data collected from, and associated parameter vectors for, the $N$ vehicles that are available before the self driving vehicle is on the road, while the subscript $i$ is reserved for some $i > N$ and refer to the data and parameter vector for any additional vehicle encountered while the self driving vehicle is on the road.
+Inference about $\theta_{1:N}$, the individual parameter vectors for the independent and hierarchical models, as well as for $\theta$, the shared homogenous parameter vector and $\beta$, the hierarchical hyperparameters, conditional on $z_{1:N}$ are assumed to be available with an unlimited computational budget.
 
-Variational Bayes (VB, Jordan et al.
-1999) approximates a distribution $p(\theta | z)$ with a family $q(\theta)$, choosing the member of that family that minimises $KL(q(\theta) || p(\theta | z)$, where
+A significant difference in the application of the time series models is through the relationship between the posterior distribution formed from observations of the first $N$ vehicles and the parameter vector for the additional vehicle, $\theta_{i}$.
+Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} = \theta$.
+In this case $p(\theta | z_{1:N}, z_{i, 1:T})$ will not be significantly different to $p(\theta | z_{1:N})$ due to the high precision of the posterior distribution for a sufficiently large $N$.
+Figure (\ref{fig:homogPosterior}) plots the posterior marginal distributions from $N = 2000$ vehicles highlighting the posterior precision.
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.95\textwidth]{homogPosterior}
+\caption{The converged Metropolis Hastings MCMC samples of the marginal posterior distributions $p(\theta | z_{1:N})$ for the homogenous model with $N = 2000$ and $p = q = 2$.
+Each variable has very high posterior precision.}
+\label{fig:homogPosterior}
+\end{figure}
+
+The independent model implies that the posterior distribution for $\theta_i$ is independent of $\theta_{1:N}$ and $z_{1:N}$ by
 \begin{equation}
-\label{KL-div}
-KL(q(\theta) || p(\theta | z) = \int_{\theta} q(\theta) \left(\log(q(\theta)) - \log(p(\theta | z))\right) d\theta.
+p(\theta_{i}| z_{1:N}, z_{i,1:T}, \theta_{1:N}) = p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_i) p(\theta_i).
+\label{indepNewCar}
 \end{equation}
-As $p(\theta | z)$ is typically intractable, VB maximises $\mathcal{L}(q)$ where
+The posterior for $\theta_i$ using the hierarchical model, assuming conditional independence between each $\theta_j$ for $j = 1, \dots, N$ and $\theta_i$, given $\beta$, $z_{1:N}$, and $z_i$, is
+\begin{equation}
+\label{hierNewCar}
+p(\theta_{i} | z_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) \int_{\beta} p(\theta_{i} | \beta) p (\beta | z_{1:N}) d\beta.
+\end{equation}
+As N is large the sampled posterior $p(\beta | z_{1:N})$ has high precision, and $\int_{\beta} p(\theta_{i} | \beta) p (\beta | z_{1:N}) d\beta$ can be replaced with $p(\theta_{i} | \hat{\beta})$ in (\ref{hierNewCar}) where $\hat{\beta}$ is a point estimate such as the mean or maximum of $p(\beta | z_{1:N})$.
+This approximation results in the closed form expression
+\begin{equation}
+\label{hierNewCar2}
+p(\theta_{i} | z_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) p(\theta_{i} | \hat{\beta}).
+\end{equation}
+(\ref{indepNewCar}) and (\ref{hierNewCar2}) can be combined with an inference procedure to allow the parameter set for any vehicle observed while driving to be inferred, and used in the forecast equation (\ref{predictive}) to produce online forecasts of the trajectory of those vehicles that allows individual heterogeneity to be included.
+
+\section{Bayesian Inference}
+\label{sec:Inference}
+
+Given a series of data observed up to time $T$, $z_{1:T}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
+\begin{equation}
+\label{predictive}
+p(z_{T+h} | z_{1:T}) =\int p(z_{T+h}|z_{1:T}, \theta) p(\theta | z_{1:T}) d\theta.
+\end{equation}
+To obtain this distribution, the posterior density for $\theta$, given by
+\begin{equation}
+\label{posterior}
+ p(\theta | z_{1:T}) = \frac{p(z_{1:T}|\theta)p(\theta)}{\int p(z_{1:T}|\theta)p(\theta) d\theta}
+\end{equation}
+must first be inferred. Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}), will be unavailable due to the complex functional form of the distributions involved.
+
+In this section two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
+In brief, MCMC is used to create a sample from $p(\theta | z_{1:T})$, with any function of $\theta$ that is desired estimated from that sample.
+In contrast, VB replaces $p(\theta | z_{1:T})$ with a parametric approximation, denoted by $q_{\lambda}(\theta | z_{1:T})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $y_{1:T}$.
+
+\subsection{Markov Chain Monte Carlo}
+\label{subsec:MCMC}
+
+There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler.
+The Gibbs sampler algorithm iteratively samples the components of the $k-$dimensional parameter vector $\theta$ via each of the so-called full conditional distributions as follows,
+\begin{align}
+&p(\theta_1 | \theta_2, \dots, \theta_k, z) \nonumber \\
+&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, z) \nonumber \\
+&\vdots \nonumber \\
+&p(\theta_k | \theta_1, \dots, \theta_{k-1}, z). \nonumber
+\end{align}
+Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\theta | z)$.
+Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
+The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
+The full conditional distributions cannot be recognised for each of the time series models proposed in Section \ref{sec:trajForecasts} and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
+
+In MH MCMC, a candidate $\theta_j^{(c)}$, where $\theta_j$ may be any scalar or vector subset of $\theta$, is drawn from a proposal distribution $r(\theta_j)$ and accepted by the sampler with probability
+\begin{equation}
+\min \left\{ 1, \frac{p(\theta_j^{(c)} | \theta_{i \neq j}^{(a)}, z)}{p(\theta_j^{(a)} | \theta_{i \neq j}^{(a)}, z)} \times \frac{r(\theta_j^{(a)} | \theta_j^{(c)})}{r(\theta_j^{(c)} | \theta_j^{(a)})} \right\},
+\label{MHaccept}
+\end{equation}
+where the superscript $(a)$ denotes the most recently accepted value of the corresponding subset of $\theta$.
+If the sampler rejects a candidate value the previous value $\theta_j^{(a)}$ is repeated.
+Each iteration of Metropolis-Hastings-within-Gibbs MCMC includes one candidate draw for each element of $\theta$.
+Choice of proposal distribution is left to the user with arguably the most simple being the Normal Random Walk proposal,
+\begin{equation}
+\theta_j^{(c)} \sim N(\theta_j^{(a)}, \Sigma_j)
+\label{RWprop}
+\end{equation}
+as $r(\theta_j^{(c)} | \theta_j^{(a)}) = r(\theta_j^{(a)} | \theta_j^{(c)})$.
+Performance of Random Walk Metropolis Hastings MCMC (RWMH-MCMC) then depends on the acceptance rate of the proposal distribution, largely influenced by the value of $\Sigma_j$.
+\citet{Garthwaite2016} propose an adaptive algorithm that increases or decreases $\Sigma_j$ depending if the most recent candidate was rejected or accepted respectively so that a target acceptance rate is asymptotically approached as the number of MCMC iterations approaches infinity.
+
+\subsection{Variational Bayes}
+\label{subsec:VB}
+
+A typically faster, albeit approximate, alternative to MCMC based inference is VB \citep{Jordan1999}.
+VB posits a family of parametric approximating distributions $q_{\lambda}(\theta | z)$, parameterised by an auxiliary vector $\lambda$, that share the same support as the true posterior distribution $p(\theta | z)$.
+Note that the family $q_{\lambda}(\theta | z)$ does not depend on $z$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta | z)$.
+A member of approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta | z)$ to $p(\theta | z)$ \citep{Kullback1951}, given by $KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})]$.
+The KL divergence is defined by
+\begin{equation}
+\label{KL-def}
+KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})] = E_{q_{\lambda}(\theta | z)} \left[ \log(q_{\lambda}(\theta | z)) - \log(p(\theta | z)) \right],
+\end{equation}
+and is a non-negative, asymmetric measure of the discrepancy between $p(\theta | z)$ and $q_{\lambda}(\theta | z)$  that will be equal to zero if and only if $p(\theta | z) = q_{\lambda}(\theta | z)$ almost everywhere \citep{Bishop2006}.
+Typically (\ref{KL-def}) cannot be evaluated, and Monte-Carlo estimates as
+\begin{equation}
+\label{KL-MC}
+KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})] \approx \frac{1}{M}\sum_{i=1}^M \left(\log(q_{\lambda}(\theta_i |z)) - \log(p(\theta | z)) \right)
+\end{equation}
+where $\theta_i \sim q_{\lambda}(\theta |z))$ are compuationally infeasible due to the inclusion of the term $p(\theta | z)$.
+Instead VB uses the Evidence Lower Bound (ELBO), denoted by $\mathcal{L}(q, \lambda)$, as an error function where
 \begin{equation}
 \label{ELBO}
-\mathcal{L}(q, \lambda) = \int_{\theta} q(\theta) \left(\log(p(\theta, z)) - \log(q(\theta))\right) d\theta
+\mathcal{L}(q, \lambda) = E_{q_{\lambda}(\theta |z)} \left[\log(p(\theta, z)) - \log(q_{\lambda}(\theta |z))\right],
 \end{equation}
-which is equivalent to minimising (\ref{KL-div}), where $\lambda$ is the parameter vector of $q(\theta)$.
-This is done with stochastic gradient ascent, taking Monte Carlo estimates of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$ (Hoffman et al. 2013)
-and repeating updates of the form
+which is evaluated with Monte-Carlo estimates
+\begin{equation}
+\label{ELBO-MC}
+\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{i=1}^M \left(\log(p(\theta_i, z)) - \log(q_{\lambda}(\theta_i |z)) \right)
+\end{equation}
+where $\theta_i \sim q_{\lambda}(\theta |z))$.
+The ELBO is equal to the negative KL divergence plus a constant, and hence maximising (\ref{ELBO}) with respect to $q_{\lambda}(\theta | z)$ is equivalent to minimising (\ref{KL-def}).
+
+\subsection{Stochastic Gradient Ascent}
+\label{subsec:SGA}
+For exponential family likelihood models, $p$, with a factorisable approximation, $q$, the characteristics of the surface of the ELBO can be exploited for optimisation in Mean Field Variational Bayes \citep{Ghahramani2000, Wainwright2008}, but for more general distributions the surface of the ELBO and its stochastic estimate are unknown.
+Maximisation proceeds by optimising only the auxiliary parameters $\lambda$ for a fixed distribution family $q$ with stochastic gradient ascent (SGA).
+SGA repeatedly takes Monte-Carlo estimates of the gradient of the ELBO with respect to $\lambda$, $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, as $\widehat{\delta\mathcal{L}(q, \lambda) / \delta \lambda}$ and applies updates of the form
 \begin{equation}
 \label{gradientAscent}
-\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} 
+\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
 \end{equation}
-which is guaranteed to converge to a local maximum (Robbins and Munro, 1951) if $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
+until the change from $\mathcal{L}(q, \lambda^{(m)})$ to $\mathcal{L}(q, \lambda^{(m+1)})$ falls below some pre-specified threshold \citep{Hoffman2013}.
+Intuitively, individual elements of $\lambda$ will increase if the estimate of the slope of $\mathcal{L}(q, \lambda^{(m)})$ is positive at the current point, and will decrease if that estimate is negative, until each element of $\lambda$ reaches a point where the slope is zero.
+This procedure is guaranteed to converge to a local maximum \citep{Robbins1951} if the sequence $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
 \begin{align}
@@
 \end{align}
-In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of Kingma and Ba (2015).
+In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of \citet{Kingma2015b}.
 
-There are two Monte Carlo estimators of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of Ranganath, Gerrish and Blei (2014), 
+There are two choices for the Monte Carlo estimator of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of \citet{Ranganath2014}, 
 \begin{equation}
 \label{scoreDeriv}
-\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_i))}{\delta \lambda} \left(\log(p(\theta_i, z)) - \log(q(\theta_i)) \right), \mbox{ where } \theta_i \sim q(\theta),
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_j | z))}{\delta \lambda} \left(\log(p(\theta_j, z)) - \log(q_{\lambda}(\theta_j | z)) \right),
 \end{equation}
-and the reparameterised estimator of Kingma and Welling (2014),
+where $\theta_j \sim q_{\lambda}(\theta | z)$ and the reparameteterised estimator of \citet{Kingma2014}.
+Reparameterisation rephrases Variational Bayes as a search for the parameters $\lambda$ of some differentiable function $f(\cdot, \lambda)$ that minimises the Kullback Leibler divergence from some distribution $r(\epsilon)$ with zero free parameters to a transformation of the posterior distribution, $p_{\epsilon}(\theta = f(\epsilon, \lambda) | z)$.
+Examples of $f$ and $r(\epsilon)$ include treating $\theta$ as location scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
+The distribution $p(\theta, z)$ is written as
+\begin{equation}
+\label{rpDist}
+p(\theta, z) = p(f(\epsilon, \lambda), z)|J(f(\epsilon, \lambda))|
+\end{equation}
+where $J(f(\epsilon, \lambda))$ is the Jacobian Matrix of the transformation $f(\epsilon, \lambda)$, and the reparameterised ELBO can be found by substituting (\ref{rpDist}) into (\ref{ELBO}),
+\begin{equation}
+\label{rpELBO}
+\mathcal{L}(q, \lambda) = E_{r(\epsilon)} \bigg[\log(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|) - \log(r(\epsilon))\bigg].
+\end{equation}
+The gradient of the reparameterised ELBO with respect to $\lambda$ of 
+\begin{align}
+\label{rpELBODeriv}
+\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} &= \frac{\delta}{\delta \lambda} \bigg( E_{r(\epsilon)} \bigg[\log\big(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|\big) - \log(r(\epsilon))\bigg] \bigg) \nonumber \\
+&= E_{r(\epsilon)} \left[ \frac{\delta}{\delta \lambda} \bigg(\log(p(f(\epsilon,\lambda), z)) + \log(|J(f(\epsilon, \lambda))|) - \log(r(\epsilon)) \bigg)\right] \nonumber \\
+&= E_{r(\epsilon)} \left[ \frac{\delta \log(p(f(\epsilon,\lambda), z))}{\delta f(\epsilon,\lambda)} \frac{\delta f(\epsilon,\lambda)}{\delta \lambda}  + \frac{\delta \log(|J(f(\epsilon, \lambda))|)}{\delta \lambda} \right].
+\end{align}
+This form leads to the reparameterised estimator of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$
 \begin{equation}
 \label{rpDeriv}
-\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^M \frac{\delta f(\lambda, \epsilon_i)}{\delta \lambda} \frac{\delta \log(p(\theta, z))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_i)} + \frac{\delta J(\lambda, \epsilon_i)}{\delta \lambda}, \mbox{ where } \epsilon_i \sim r(\epsilon)
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta f(\lambda, \epsilon_j)}{\delta \lambda} \frac{\delta \log(p(\theta, z))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda},
 \end{equation}
-where $r(\epsilon)$ is some distribution without any free parameters (eg.
-Uniform, Standard Normal), $f(\lambda, \epsilon)$ is some differentiable function that transforms $\epsilon$ to $\theta$, and $J(\lambda, \epsilon)$ is the Jacobian Matrix of this function.
-(\ref{rpDeriv}) typically has significantly lower variance than (\ref{scoreDeriv}), allowing $M$ to be set lower than $L$ which improves the speed of the algorithm; however the requirement for $f$ to exist restricts the class of approximating distributions $q(\theta)$ that can be used.
-Notably, a differentiable $f(\lambda, \epsilon)$ does not exist when $q(\theta)$ is a mixture model and the score estimator (\ref{scoreDeriv}) must be used.
-To offset the increased variance and allow a reduction in $L$, following Gunawan, Tran, and Kohn (2017), Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence (Sobol 1967), then these are randomised using the scrambled net method (Matousek 1998), and transformed into a draw of $\theta$ using the inverse-CDF of $q(\theta)$.
-Using RQMC allows (\ref{scoreDeriv}) to converge to the true derivative at a rate $O(L)$ compared to $O(\sqrt{L})$ in standard Monte Carlo.
+where $\epsilon_j \sim r(\epsilon)$.
 
-\subsection{Updating VB Process}
+These Monte Carlo estimators converge to the true derivative at a rate $O(\sqrt{M})$, so choosing the number of samples per estimate, $M$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate.
+An increased value of $M$ will reduce the estimator variance, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
+To reduce the variance of the Monte Carlo estimator, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence \citep{Sobol1967} which are then randomised using the scrambled net method \citep{Matousek1998}, and transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$.
+Using RQMC allows the gradient estimators to converge to the true derivative at the increased rate of $O(M)$.
 
-\begin{itemize}
-\item While on the road a self driving car may wish to make posterior inferences on the parameter set of cars around them.
-They may be interested in the degree of erraticness / variance of their actions.
-\item The posterior for car $N+1$ is given by 
-\begin{equation}
-\label{originalPost}
-p(\theta_{N+1} z_{N+1} | z_{1:N}) \propto p(z_{N+1} | \theta_{N+1}) \int_{\beta} p(\theta_{N+1} | \beta) p (\beta | z_{1:N}) d\beta
-\end{equation}
-\item As N is large $p (\beta | z_{1:N})$ is fairly precise, so we can replace $\int_{\beta} p(\theta_{N+1} | \beta) p (\beta | z_{1:N}) d\beta$ with $ p(\theta_{N+1} | \hat{\beta})$ where $\hat{\beta}$ is the posterior mean of $p(\beta | z_{1:N})$.
-This approximation gives us a closed form for $ p(\theta_{N+1} | \hat{\beta})$.
-\item Subsequently fitting $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})  \propto p(z_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \hat{\beta})$ is fairly easy with MCMC / VB etc.
-\item When $S$ is small, the information in the prior will dominate the individual information.
-\item However, for some $T > S$ the information in the car's individual movements should become more relevant and forecasts could be improved by updating the distribution to $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:T})$.
-\item This updated posterior is given by
+\section{Updating Variational Bayes}
+\label{sec:UVB}
+
+The self driving vehicle is constantly observing data about the movements of the surrounding vehicles.
+After observing vehicle $i$ until time $S$, $z_{i, 1:S}$ has been observed and can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the conditional dependence on $z_{1:N}$.
+At time $T > S$ an additional $T - S$ data points from vehicle $i$ have been observed, and forecasts could be improved by incorporating these to form the posterior approximation, $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$, where the $S$ and $T$ subscripts on $\lambda$ are included to differentiate the auxiliary parameter vector conditioned on data up to times $S$ and $T$.
+To facilitate this posterior update, an Updating Variational Bayes (UVB) mechanism is introduced where only $z_{i, S+1:T}$ needs to be processed.
+
+Given $p(\theta_{i} | z_{i, 1:S})$ from (\ref{posterior}), the posterior distribution at time $T$, $p(\theta_{i} | z_{i, 1:T})$ is given by Bayes rule as
 \begin{equation}
 \label{updatePost}
-p(\theta_{N+1} | z_{N+1, 1:T}, z_{1:N}) \propto p(z_{N+1, S+1:T} | \theta_{N+1})p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})
+p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S})
 \end{equation}
-\item Both $\mathcal{L}(q)$ derivative estimators, (\ref{scoreDeriv}) and (\ref{rpDeriv}), need to evaluate \newline $\log(p(\theta_{N+1}, z_{N+1, S+1:T} | z_{1:N}, z_{N+1, 1:S}))$, which takes the form
-\begin{align}
-\log(p(\theta_{N+1}, z_{N+1, S+1:T} | z_{1:N}, z_{N+1, 1:S})) &= \log(p(z_{N+1, S+1:T} | \theta_{N+1})) \nonumber \\
-&+ \log(p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})) \label{updateEq}
-\end{align}
-\item To perform this update only using the data from $z_{N+1, S+1:T}$ we would need a closed form form expression for $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$ to replace the role of the prior.
-Without this closed form we would need to re-fit the VB model using the original prior $p(\theta_{N+1} | \hat{\beta})$ and the whole observed sequence $z_{N+1, 1:T}$.
-\item Due to the demands of driving, the car wants the update carried out as soon as possible, and ideally should be able to discard the previously used information $z_{N+1, 1:S}$. 
-\item Defining $q(\theta_{N+1} | z_{N+1, 1:S})$ as the Variational Bayes approximation of $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$, a new method is proposed to update $q(\theta_{N+1} | z_{N+1, 1:S})$ to $q(\theta_{N+1} | z_{N+1, 1:T})$ after observing $z_{N_1, S+1:T}$.
-\item By replacing $p(\theta_{N+1} | z_{1:N}, z_{N+1, 1:S})$ in (\ref{updateEq}) with the previously fit $q(\theta_{N+1} | z_{N+1, 1:S})$ we get an approximation to (\ref{updateEq}):
+Standard methods to evaluate the posterior, such as Variational Bayes or Markov Chain Monte Carlo, require the evaluation of the right hand side of (\ref{updatePost}), however computation of $p(\theta_{i} | z_{i, 1:S})$ is often infeasible.
+Without being able to evalute the right hand side, posterior inference requires evaluation of the prior distribution and full sample $z_{i, 1:T}$.
+To avoid this, Variational Bayesian Updating replaces $p(\theta_{i} | z_{i, 1:S})$ with the analytical approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$.
+The update (\ref{updatePost}) is replaced with
 \begin{equation}
 \label{updateEqApprox}
-\log(p(\theta_{N+1}, z_{N+1, 1:T} | z_{1:N})) \approx \log(p(z_{N+1, S+1:T} | \theta_{N+1})) + \log(q(\theta_{N+1} | z_{N+1, 1:S}))
+p(\theta_{i} |  z_{i, 1:T}) \approxprop p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S})
 \end{equation}
-\item Substituting (\ref{updateEqApprox}) into (\ref{scoreDeriv}) gives the score estimator for updating Variational Bayes, 
+The ELBO gradient estimators to construct the updated Variational Bayes approximation at time $T$,  $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$, can be obtained by substituting (\ref{updateEqApprox}) into the score estimator (\ref{scoreDeriv}) or the reparameterised estimator (\ref{rpDeriv}) of $\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} $.
+The updating score estimator is given by
 \begin{align}
-\label{scoreUpdate}
-\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} \approx \sum_{i = 1}^L \frac{\delta \log(q(\theta_{N+1, i} | z_{N+1, 1:T}))}{\delta \lambda} &\left(\log(q(\theta_{N+1, i} | z_{N+1, 1:S}) - \log(q(\theta_{N+1} | z_{N+1, 1:T})) \right), \nonumber \\
-& \mbox{ where } \theta_{N+1, i} \sim q(\theta_{N+1} | z_{N+1, 1:T}).
+\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{SCU} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
+&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i} z_{i, 1:T})) \right) \label{scoreUpdate}
 \end{align}
-\item Similarly, substituting (\ref{updateEqApprox}) into (\ref{rpDeriv}) gives the reparameterised estimator for updating Variational Bayes,
+where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. Similarly, the updating reparameterised estimator is given by
 \begin{equation}
 \label{rpUpdate}
-\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T} \approx \sum_{i = 1}^M \frac{\delta f(\lambda_T, \epsilon_i)}{\delta \lambda_T} \frac{\delta \log(q(\theta_{N+1} | z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon_i)} + \frac{\delta J(\lambda_T, \epsilon_i)}{\delta \lambda_T}, \mbox{ where } \epsilon_i \sim r(\epsilon)
+\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{RPU} = \sum_{j = 1}^M \frac{\delta f(\lambda_T, \epsilon_j)}{\delta \lambda_T} \frac{\delta \log(q_{\lambda_S}(\theta_{i} |z_{i, 1:S}))}{\delta \theta_{i}} \bigg\rvert_{\theta_{i} = f(\lambda_T, \epsilon_j)} + \frac{\delta J(\lambda_T, \epsilon_j)}{\delta \lambda_T},
 \end{equation}
-where $\lambda_T$ denotes the parameter vector of $q(\theta_{N+1} | z_{N+1, 1:T})$.
-\item This process can be repeated as neccessary, and is $O(1)$ instead of $O(T)$ assuming $S$ grows at the same rate as $T$.
-\end{itemize}
+where $\epsilon_j \sim r(\epsilon)$.
+If this process is repeated periodically, $S$ will grow at the same rate as $T$, and updating VB is $O(1)$ rather than $O(T)$ as required by re-using the full sample without the approximation step in (\ref{updateEqApprox}).
 
 \section{Forecast Evaluation}
+\label{sec:eval}
 
-\begin{itemize}
-\item 4071 vehicles are available in the sample
-\item 500 are used to estimate splines and straighten the road to relative coordinates as described in Section 2
-\item 2000 are used to train any models used
-\item 1571 are available for testing
-\item Each second (10 observations) models are fit to the sample of $z_{N+1, 1:T}$ and forecasts of $z_{N+1, T+1:T+30}$ are made, corresponding to a three second forecast horizon
-\item Forecasts are compared in terms of predictive log-score (where a predictive density is available) and Euclidean error.
-\end{itemize}
+After obtaining the posterior distributions for each model conditioned on 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on a further 1373 vehicles.
+For each of these vehicles $i > 2000$, at each point in time $T = 100, 110, \dots, 450$ and forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model, while the time-series models are used to provide forecast densities of $p(z_{i, T+h} | z_{1:2000}, z_{i, 1:T})$ and point estimates from the MAP of this density.
+As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement after every second of observing the vehicle.
+The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | z_{1:2000}, z_{i, T})$, while the independent and hierarchical models produce $p(\theta_{i} | z_{1:2000}, z_{i, 1:T})$ through three methods:
+\begin{enumerate}
+\item Posterior sampling with RWMH-MCMC,
+\item VB fitting $q_{\lambda_T}(\theta_{i} | z_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$ using the original prior distribution,
+\item UVB fitting $q_{\lambda_T}(\theta_{i} | z_{1:2000}, z_{i, 1:T})$ using $q_{\lambda_S}(\theta_{i} | z_{1:2000}, z_{i, 1:S})$ and data $z_{i, S+1:T}$ as described in Section \ref{sec:UVB}.
+\end{enumerate}
+A summary of these approaches is provided in Table \ref{tableAlg}. Point estimate forecasts are evaluated by their Euclidean Error,
+\begin{equation}
+\mbox{EE}_{i, T, h} = \sqrt{\left(\hat{x}^*_{i, T+h} - x^*_{i, T+h} \right)^2 + \left(\hat{y}^*_{i, T+h} - y^*_{i, T+h} \right)^2},
+\label{eucError}
+\end{equation}
+where $\{\hat{x}^*_{i, T+h}, \hat{y}^*_{i, T+h}\}$ is a forecast of $\{x^*_{i, T+h}, y^*_{i, T+h}\}$, while density forecasts are evalutated by their logscore,
+\begin{equation}
+\mbox{LS}_{i, T, h} = \log \left(p\left(x^*_{i, T+h}, y^*_{i, T+h} | z_{1:2000}, z_{i, 1:T} \right) \right)
+\label{logscore}
+\end{equation}
 
-\subsection{Heterogenous Time Series Models}
-
-There are three versions of the heterogenous time series model described in Section 3, differentiated by the form of the prior, $p(\theta_{N+1} | z_{1:N})$.
-\begin{enumerate}
-\item Non-informative prior for the new car, where $p(\theta | z_{1:N})$ is multivariate normal with mean vector $(-5, -5, 0, 0, 0, 0)$ and diagonal variance matrix $10 \mathbb{I}$.
-\item Informative prior where $p(\theta | z_{1:N})$ is multivariate normal with a mean vector and variance matrix estimated by the posterior mean of $\mu$ and $\Sigma$ in a hierarchical model with $K = 1$ fit by MCMC to the 2000 training vehicles.
-\item Informative prior where $p(\theta | z_{1:N})$ is a six component mixture of multivariate normals with mean vectors, variance matrices and mixture weights estimated by the posterior means of each $\mu_j$ and $\Sigma_j$ and $\pi_j$ in a hierarchical model with $K = 6$ fit by MCMC to the 2000 training vehicles.
-\end{enumerate}
-Each model is fit by each of Metropolis-Hasting MCMC, standard Variational Bayes where the entire observed history of $z_{N+1, 1:T}$ is used for each model fit, and the updating Variational Bayes process described by Section 4.2 where only the most recent observed data $z_{N+1, S+1:T}$ is used in each iteration.
-In each Variational Bayes algorithm the approximating distribution has the same functional form as the prior used, a multivariate normal distribution for the non-informative prior and $K = 1$ hierarchy, and a six component mixture of multivariate normals for the $K = 6$ hierarchy.
-Details of each case are provided in Table \ref{tableAlg}.
 \begin{center}
@@
 \resizebox{\textwidth}{!}{%
-\begin{tabular}{| l | c | c | c |}
+\begin{tabular}{| l | c | c |}
 \hline
-& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
+& Independent Model & Hierarchical Model \\
 \hline
-MCMC & \multicolumn{3}{c|}{Random Walk Metropolis Hastings jointly drawing the entire $\theta$ vector using Garthwaite, Fan and Sisson (2015)} \\
-& \multicolumn{3}{c|}{to control the variance matrix of the Multivariate Gaussian proposal distribution} \\
+MCMC & \multicolumn{2}{c|}{RWMH-MCMC jointly drawing the entire $\theta$ vector using \citet{Garthwaite2016} to control the} \\
+& \multicolumn{2}{c|}{variance matrix of the Multivariate Gaussian proposal distribution to obtain a 23.4\% acceptance rate for $\theta$ draws.} \\
 \hline
-VB  & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpDeriv}) to update parameters of a} & Variational Bayes using (\ref{scoreDeriv}) to update parameters of a six \\
-Standard & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\
+VB  & Variational Bayes using (\ref{rpDeriv}) to update parameters & Variational Bayes using (\ref{scoreDeriv}) to update parameters\\
+Standard &  of a Multivariate Gaussian approximation with &  of an approximation formed as a six component mixture\\
+&non-zero covariance using $M = 25$, where $f$ is a &  of diagonal variance Multivariate Gaussians using $M = 50$. \\
+&location scale transform from a standard normal $r(\epsilon)$. &\\
 \hline
-VB & \multicolumn{2}{c|}{Variational Bayes using (\ref{rpUpdate}) to update parameters of a} &  Variational Bayes using (\ref{scoreUpdate}) to update parameters of a six \\
-Updating & \multicolumn{2}{c|}{Multivariate Gaussian approximation with non-zero covariance} & component mixture of diagonal variance Multivariate Gaussians\\ 
+VB  & Variational Bayes using (\ref{rpUpdate}) to update parameters & Variational Bayes using (\ref{scoreUpdate}) to update parameters\\
+Standard &  of a Multivariate Gaussian approximation with &  of an approximation formed as a six component mixture\\
+&non-zero covariance using $M = 25$, where $f$ is a&  of diagonal variance Multivariate Gaussians using $M = 50$. \\
+&location scale transform from a standard normal $r(\epsilon)$. &\\
 \hline
 \end{tabular}}
+\caption{Details of the algorithms used to produce posterior distributions for each method and prior combination}
 \label{tableAlg}
-\caption{Details of the algorithms used to produce posterior distributions for each method and prior combination}
 \end{table}
@@
 
-\subsection{Competing Models}
+\subsection{Naive Forecast Models}
+\label{subsec:Naive}
 
-\subsubsection{Constant Angle and Acceleration Models}
-
-The framework provided by (\ref{xEq}) and (\ref{yEq}) allows point estimates of $\{x^*_{t}, y^*_{t} | t = T + 1, \dots, T+30\}$ to be easily obtained given $\{x^*_T, y^*_T, v_T\}$ and point estimates of the future values of $\{a_{t}, \delta_{t} | t = T + 1, \dots, T+30\}$.
-Nine naive forecast models are provided where these future values are estimated by a constant $\{\hat{a}, \hat{\delta}\}$.
-The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}$ and one estimator for $\hat{\delta}$.
-Acceleration estimators are: $\hat{a} = a_T, \hat{a} = 0.1 \sum_{s = T-9}^T a_s$, and $\hat{a} = 0$.
-Similarly steering angle estimators are: $\hat{\delta} = \delta_T, \hat{\delta} = 0.1 \sum_{s = T - 9}^T \delta_s$, and $\hat{\delta} = \pi/2$.
+Nine naive forecast models are provided where forecasts of $a_t$ and $\delta_t$ are constants equal to $\hat{a}$ and $\hat{\delta}$ for $t = T+1, \dots, T+H$.
+The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}$ and one estimator for $\hat{\delta}$ at time $T$.
+Acceleration estimators are the most recent acceleration, $\hat{a} = a_T$, the average acceleration over the most recent second, $\hat{a} = 0.1 \sum_{s = T-9}^T a_s$, and zero acceleration $\hat{a} = 0$, equivalent to constant velocity.
+Similarly, steering angle estimators are the most recent angle, $\hat{\delta} = \delta_T$, the average angle over the most recent second, $\hat{\delta} = 0.1 \sum_{s = T - 9}^T \delta_s$, and $\hat{\delta} = \pi/2$, corresponding to the vehicle to driving directly forward.
 \begin{table}
@@
 \end{center}
+\caption{Classification of naive forecast models by future velocity and steering angle estimators.}
 \label{tableNaive}
-\caption{Classification of naive forecast models by future velocity and steering angle estimators}
 \end{table}
 
-\subsubsection{Homogenous Time Series Models}
+\subsection{Recurrent Neural Networks}
+\label{subsec:RNN}
+A Recurrent Neural Network (RNN) is constructed with inputs following \citet{Ding2013}, the ten most recent lags of vehicle position, angle, velocity, and time headway to the preceding vehicle.
+The RNN outputs the next thirty changes in the vehicle position, is trained with the Adam optimiser and a a loss function $EE_{i, T, 30}$ given by (\ref{eucError}).
+As of writing this report forecast results from this model are not competitive with the naive models and are not discussed further.
 
-The homogenous time series model has the same bivariate AR(2) structure given by (\ref{aAr}) and (\ref{dAr}) as used by the heterogenous time series models with the additional restriction that each $\theta_i = \theta$, so each vehicle shares the same parameter vector.
-This model is estimated by MCMC to the 2000 training cars.
+\subsection{Results}
+\label{subsec:Results}
 
-\subsubsection{Recurrent Neural Networks}
+\begin{itemize}
+\item Discuss the updating method relative to standard VB, change in logscore in Figure \ref{fig:updateCost}
+\item Plot the average convergence time as a function of $T - S = 1, 5, 10, 20$ and $T = 100, 110, \dots,  450$. Running now for Figure \ref{fig:timing}
+\end{itemize}
 
-A Recurrent Neural Network is constructed with inputs given by the ten most recent lags of vehicle position, angle, and velocity, and an output given by the next thirty values of the vehicle position.
-The network is trained with the Adam optimiser and a mean Euclidean error loss function.
-\subsection{Results}
+Mean Eucliedean Error for each model at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
+Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_t$ the choice of $\hat{\delta}_t$ has not substantially changed the forecast error.
+Zero acceleration naive models (Naive 7, 8, and 9) perform the best, followed by the averaged acceleration models (Naive 4, 5, and 6) and finally the current acceleration models (Naive 1, 2, and 3).
+At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model.
+
+Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, and with the exception of the independent model with UVB inference, each time series model and inference implementation produces a forecast with a similar amount of error.
+This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the heirarchical model.
+As the independent model is the heterogenous model that provides the least accurate forecasts it is not discussed further.
 
@@
 \centering
-\includegraphics[width = 0.95\textwidth]{CarsBoxplots.png}
-\caption{Predictive logscores for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each of the three levels of prior information fit by MCMC, standard VB and updating VB.
-The median logscore across all time periods, forecast horizons, and priors for the updating VB algorithm is 3.28, compared to 3.43 for standard VB and 3.46 for MCMC.}
-\label{fig:MainResults}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.95\textwidth]{diffInLS.png}
-\caption{A truncated view of the difference in Variational Bayes predictive logscores (updating minus standard) for each prior and a range of different T values.
-$99\%$ of Finite Mixture differences, $93\%$ of Non-Informative differences and $92\%$ of Single Hierarchy differences are inside the (-5, 5) region.
-The algorithms are identical when $T = 10$ so this is excluded.
-Updating VB logscores are only slightly lower than the standard VB logscores, signifying only a small loss of accuracy associated with the extra approximation.}
-\label{fig:diffInLS}
-\end{figure}
-
-\begin{center}
-\begin{table}[h]
-\begin{tabular}{| l | c | c | c |}
-\hline
-& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
-\hline
-MCMC & 0.\textbf{846} & 0.899 & \textbf{0.931} \\
-VB Standard & 0.782 & \textbf{0.913} & 0.901 \\
-VB Updating & 0.558 & 0.701 & 0.873 \\
-\hline
-\end{tabular}
-\label{tableMedian}
-\caption{Median predictive logscore for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each prior and algorithm combination}
-\end{table}
-\end{center}
-
-\begin{center}
-\begin{table}[h]
-\begin{tabular}{| l | c | c | c |}
-\hline
-& Non-Informative Prior & Simple Hierarchy & Finite Mixture Hierarchy \\
-\hline
-MCMC & 0.209 & 0.199 & 0.197 \\
-VB Standard & \textbf{0.203} & \textbf{0.196} & 0.197 \\
-VB Updating & 0.211 & 0.207 & \textbf{0.196} \\
-\hline
-\end{tabular}
-\label{tableError}
-\caption{Mean predictive Euclidean error in metres using the maximum of $p(z_{i, T+h} | z_{i, 1:T})$ as a point-estimate with $h = 1, \dots, 30$ and $T = \{10, 20, \dots, 300\}$ for each algorithm and prior combination.
-Updating the VB model is competitive with a standard VB fit and exact sampling by MCMC.}
-\end{table}
-\end{center}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width = 0.95\textwidth]{PredictiveError.png}
-\caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | z_{i, 1:T})$ for $h = \{10, 20, 30\}$ and $T = \{10, 20, \dots, 300\}$ for each model.
-Each AR(2) model is significantly more accurate than the naive constant velocity / angle models.}
+\includegraphics[width = 0.8\textwidth]{predictiveError.png}
+\caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | z_{i, 1:T})$ for $h \in \{10, 20, 30\}$ and $T \in \{100, 110, \dots, 450\}$ for each model.
+Each time series model is significantly more accurate than the naive constant acceleration / angle models, with little discernable difference between the time series models.}
 \label{fig:PredError}
@@
 
+\begin{figure}[h]
+\centering
+\includegraphics[width = 0.8\textwidth]{predictiveErrorZoom.png}
+\caption{The same results as Figure \ref{fig:PredError} with the Naive models excluded.
+Most of the time series model have very similar MAP forecast accuracy, with the UVB / Independent time series combination performing slightly weaker}
+\label{fig:PredErrorZ}
+\end{figure}
+
+There are systematic differences in the level of certainty in forecasts provided by the heirarchical and homogenous models as measured by their logscores.
+The key differentiation between the models is the heirarchical models ability to assign each vehicle individual variance parameters,  $\sigma^2_{\epsilon, i}$ for $a_t$, and $\sigma^2_{\eta, i}$ for $\delta$.
+Figure (\ref{fig:posVarMean}) plots the range of posterior means for these parameters from with RWMH-MCMC sampling with $T = 450$, with the homogenous posterior means indicated with vertical red lines.
+There is a large amount of positive skewness in the distribution of variances, which may have caused the homogenous model to estimate a variance that is too large for most individual cars.
+The homogenous variance posterior means are greater than $80\%$ and $91.5\%$ of heirarchical variance posterior means for acceleration and the steering angle respectivley.
+This large variance has little impact on the location of the MAP of $p(z_{i, T+h} | z_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density.
+Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the hierarchical model with updating VB inference over the homogenous model at the three second forecast horizon.
+Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis.
+The hierarchical model trajectory forecasts for drivers in the lower variance quntiles benefit the most from including heterogeneity.
+However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles.
+This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance estimates allow identification of vehicles with high variance, and potentially more erratic and dangerous driving styles.
+
+\begin{figure}[h]
+\centering
+\includegraphics[width = 0.95\textwidth]{posVarMean}
+\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by Metropolis-Hastings MCMC at $T = 450$ using the hierarchical model, compared to the homogenous posterior mean in red.
+Small amounts of cars with high variance has skewed the homogenous estimates to be too large for the majority of the vehicles forecasted.}
+\label{fig:posVarMean}
+\end{figure}
+
+\begin{figure}[h]
+\centering
+\includegraphics[width = 0.95\textwidth, height = 0.4\textheight]{varSplit}
+\caption{Difference in logscore between the hierarchical model with posterior inference using UVB and the homogenous model.
+Results are split by variance quintiles for acceleration (x-axis) and angle (y-axis).
+There are substantial improvements in logscore, and thus forecast certainty, for vehicles with a low variance in either variable.
+Vehicles in the fifth quintiles, close to the homogenous variance estimates, benefit instead from the higher precision associated with the homogenous posterior distribution due to the large $N$.}
+\label{fig:varSplit}
+\end{figure}
+
+The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | z_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | z_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self driving vehicles.
+Figure (\ref{fig:timing}) demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$.
+The compuation time for standard VB increases linearly with $T$, while UVB is constant; posterior inference avabale at time $T$ utilising UVB can be conditioned on more data than standard VB.
+Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time and make UVB feasible for more complex models.
+Figure (\ref{fig:updateCost}) demonstrates that implementing UVB over standard VB does not lead to a decrease in forecast logscores, plotting the difference between the heirarchical forecast logscores obtained by UVB and standard VB for several ranges of $T$.
+Augmenting self driving vehicles with hierarchical time series models to forecast the trajectory of surrounding traffic, and repeatedly updating the posterior distribution and forecasts, can be feasibly implemented with UVB.
+\begin{figure}[htp]
+\centering
+\includegraphics[width = 0.75\textwidth]{timing}
+\caption{Average time to converge for Standard VB (Black) and UVB (Coloured) with four different update lengths given by $T - S$, truncated to 25 seconds.
+Each UVB procedure originally fits the model to the first 100 data points and updates this every $T - S$ data points.
+VB algorithms are ran with $M = 50$ on one CPU core.
+Convergence time could be reduced through parallelisation.}
+\label{fig:timing}
+\end{figure}
+\begin{figure}[htp]
+\centering
+\includegraphics[width = 0.75\textwidth, height = 0.3\textheight]{updateCost}
+\caption{Differences in predictive logscores for $p(z_{i, T+h} | z_{i, 1:T})$ for each $h = 1, \dots, 30$ and $T = 100, 110, \dots, 450$ between updating and standard Variational Bayes.
+The differences are typically small, and UVB can improve the logscore.}
+\label{fig:updateCost}
+\end{figure}
+
+\newpage
+
 \section{Conclusion}
 \begin{itemize}
-\item Dynamic AR 2 models forecasting future acceleration and angles drastically improves forecasts of the future trajectory compared to the competing models.
-\item Treating vehicles as heterogenous allows behaviour specific to an individual driver to be learned, improving logscores and forecast uncertainty.
-\item This requires constant models fits and updates, which can be made computationally feasible by updating previous Variational Bayes model fits, which incurs only a small performance hit relative to the more expensive standard VB and MCMC methods.
+\item The use of auto-regressive models for forecasting future acceleration and angles drastically improves forecasts of the future trajectory compared to the competing models.
+\item There is evidence of heterogeneity between vehicles, and allowing this with the heirarchical model allows behaviour specific to an individual driver to be inferred, improving logscores and thus reducing forecast uncertainty.
+\item Heterogeneity particularly benefits forecasts for low variance drivers, but also enables high variance drivers to be identified.
+\item Online forecasting with the heterogeneous model requires constant model fits and updates, computation of which can be simplified without a substantial performance hit by UVB.
+\item UVB is more broadly applicable to any kind of online inference for time series data.
 \end{itemize}
 
+\newpage
+\bibliographystyle{asa}
+\bibliography{references}
+
 \end{document}

### 2018-02-27 Cars/createCarsAug.R

@@ -22,6 +22,14 @@ colnames(cars) = c('ID', 'frame', 'totalFrames', 'time', 'x', 'y',
                    'veloc', 'accel', 'lane', 'proceeding', 'following', 
                    'spacing', 'headway')
 
+cars %>%
+  filter(ID %in% head(unique(.$ID), 10)) %>%
+  ggplot() + geom_path(aes(x, y, group = ID)) +
+  labs(y = NULL, x = 'Lane') + 
+  theme_bw() + 
+  scale_x_continuous(labels = c('Lane 1', 'Lane 2', 'Lane 3', 'Lane 4', 'Lane 5', 'Entry Lane'),
+                     breaks = c(6, 18, 30, 40, 50, 65))
+
 #Operations on data
 cars %>%
   mutate(time = time - min(time)) -> cars

### 2018-02-27 Cars/createY600.R

@@ -22,6 +22,7 @@ colnames(cars) = c('ID', 'frame', 'totalFrames', 'time', 'x', 'y',
                    'veloc', 'accel', 'lane', 'proceeding', 'following', 
                    'spacing', 'headway')
 
+
 #Operations on data
 cars %>%
   group_by(ID) %>%

### 2018-02-27 Cars/testSlurm/slurmRFuns.R

@@ -316,6 +316,57 @@ fitCarMods <- function(data, prior, starting, S = 10, mixComps = 6){
   results
 }
 
+timeModelFits <- function(data, prior, starting, method, T, TmS, id){
+  
+  mean <- prior[[1]][1:6]
+  u <- matrix(prior[[1]][7:42], 6)
+  linv <- solve(t(u))
+  start <- Sys.time()
+  fit <- carsVB(data, starting[[1]], lags = 2, model = arUpdater, mean = mean, Linv = linv, dimTheta = 6)$lambda
+  time <- Sys.time() - start
+  if(attr(time, 'units') == 'mins'){
+    time <- time * 60
+    attr(time, 'units') = 'secs'
+  }
+  class(time) <- 'numeric'
+  results <- data.frame(id = id,
+                        method = method,
+                        prior = 'Non-Informative',
+                        T = T,
+                        TmS = TmS,
+                        time = time[1])
+  
+  mean <- matrix(prior[[3]][1:36], 6)
+  siginv <- array(0, dim = c(6, 6, 6))
+  dets <- NULL
+  for(k in 1:6){
+    sd <- exp(prior[[3]][36 + (k-1)*6 + 1:6])
+    var <- diag(sd^2)
+    siginv[,,k] <- solve(var)
+    dets <- c(dets, 1 / prod(sd))
+  }
+  weights <- prior[[3]][73:78]
+  weights <- exp(weights) / sum(exp(weights))
+  start <- Sys.time()
+  fit <- carsVBMixScoreDiag(data, starting[[2]],
+                                     priorMix = list(mean = mean, SigInv = siginv, dets = dets, weights = weights),
+                                     S = 50, maxIter = 10000, threshold = 0.05)$lambda
+  time <- Sys.time() - start
+  if(attr(time, 'units') == 'mins'){
+    time <- time * 60
+    attr(time, 'units') = 'secs'
+  }
+  class(time) <- 'numeric'
+  results <- rbind(results, 
+                   data.frame(id = id,
+                              method = method,
+                              prior = 'Hierarchy',
+                              T = T,
+                              TmS = TmS,
+                              time = time[1]))
+  results
+}
+
 MCMCDens <- function(data, N, H, grid, MCMCdraws){
   adDens <-  evalMCMCDensIndep(data, N, H, grid, MCMCdraws)
 

### 2018-03-02 Cars/paper.tex
paper updates

@@
 \title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}
+\author{Nathaniel Tomasetti}
+\date{}
 
-%TODO:
-%Flesh out section 1
-%Add authors, institutions, an abstract and bibtex references 
-%Switch out manual references for Bibtex when everything else is done.
+%TODO: Add a conclusion
 
@@
 
-\begin{itemize}
-\item Self driving vehicles are becoming more common, eg Waymo, Tesla, GM.
-\item The naviation systems used to drive involve predicting the movement of the surrounding traffic to find the safest path for the self driving vehicle.
-\item While self driving vehicles reduce collisions due to human error, safety can be further improved by better forecasts of the trajectory of nearby human controlled vehicles.
-\item These models use deep neural networks (DNN) to extract the properties of surrounding vehicles from sensors (eg.
-Cameras and LiDAR), such as their position, velocity, and steering angle, see \citet{Woo2016a} and \citet{Tian2017}.
-\item The self driving vehicle forecasts point estimates of the future trajectory of these surrounding vehicles based on the DNN results.
-\item These can assume the surrounding vehicles will maintain their current angle and velocity \citet{Gindele2010}
-\citet{Waymo2017}, \citet{Bautista2017}, but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citet{Woo2016b}, \citet{Ding2013}, \citet{Woo2017}, \citet{Geng2017}.
-\item This paper takes a statistical approach, fitting time series models to two extracted variables, velocity and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
-\item It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a trajectory forecast.
-\item A homogenous time series model shows a strong ability to predict the MAP trajectory
-\item An important part of this modelling is to capture the uncertainty of car trajectories, as the self driving vehicle must react to the possibility of collision even when that possibility is relatively small.
-\item There are many sources of heterogeneity in drivers, from their individual driving styles to the conditions of traffic directly around them.
-\item Including heterogeneity in the model will allow the model to better adapt to drivers with more extreme behaviours, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
-\item A hierarchical model is employed to allow heterogeneity between different drivers, where each vehicle is given its own parameter vector with a shared prior distribution.
-\item This model implies that any new vehicle observed by the self driving vehicle has its own parameter vector, related to previously observed vehicles through the shared prior.
-\item The self driving vehicle must be able to estimate the parameter vector of newly observed cars, and update this parameter vector as it observes additional data.
-\item Forecasts where the parameter vector is conditioned on this additional data should outperform forecasts conditioned on less data.
-\item A three step strategy is employed to provide density forecasts for the future trajectory of newly observed vehicles:
-\begin{enumerate}
-\item Fit a hierarchical model with a large amount of vehicles to obtain the posterior distribution of each vehicles individual parameter vector and the shared distribution.
-This can occur before the self driving vehicle is on the road.
-\item When a new vehicle is observed use the information in this shared distribution to model the associated new parameter vector. 
-\item Update the parameter vector and all forecasts at short intervals.
-\end{enumerate}
-\item This paper uses standard Variational Bayes for the second step, and introduces an updating method for the third step; the posterior distribution of the new vehicles parameter vector is approximated, then this approximation is updated.
-\item This only uses the data observed in-between updates, allowing the previously observed history of that vehicle to be discarded.
-\item This is important for two reasons:
-\begin{enumerate}
-\item The model can be updated faster as less data must be processed, so forecasts can be made using as much recent data as possible.
-\item Self driving vehicles take in up to 1GB of data a second (several websites report this as quotes from different people eg.
-Tech CEO, but I don't know if that's good enough for an academic citaiton), so discarding old data is useful.
-\end{enumerate}
-\item Show that a homogenous time series model has strong predictive performance compared to constant models / neural networks while also providing density forecasts.
-\item Build a heterogenous version of the homogenous model, show that the density forecasts improve. 
-\item Show that the Updating VB required to implement the heterogenous model in real time is comparable to MCMC and standard VB.
-\end{itemize}
+Self driving vehicles are rapidly becoming more advanced, with development from large corporations such as Tesla, Ford, and Audi expecting autonomous vehicles available to consumers as early as 2020.
+These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery and selects a path to avoid any collisions.
+As many as $94\%$ of accidents are due to human error according to data from the United States National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, which can feasibly be reduced by self driving vehicles.
+However self driving vehicles will operate in situations where surrounding vehicles are controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
+\\
+
+Sophisticated Deep Neural Networks are employed to extract the position, velocity, and steering angle of surrounding traffic from readings provided by sensors such as cameras and Light Detection and Ranging (LiDAR); see \citet{Woo2016a} and \citet{Tian2017} for details.
+From this data, estimates of the future trajectory of surrounding vehicles can be produced.
+Forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
+\\
+
+This paper takes a statistical approach to the trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
+It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability collisions to be detected.
+The time series models used can allow for heterogeneity between different vehicles, which may result from sources of driver heterogeneity that may result from factors such as differences in individual driving styles, or the conditions of traffic directly around them.
+Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
+\\
+
+Incorporating heterogeneity into forecasts requires the behaviour of each driver of interest to be inferred as vehicles are encountered while driving.
+This inference is facilitated by first obtaining the posterior distribution for a large number of vehicles before the self driving vehicle is on the road, from which the prior distribution for additional vehicles can be constructed.
+Once the self driving vehicle is on the road, online inference is made possible by the introduction of Updating Variational Bayes (UVB), which approximates Bayesian updating so that inference may be periodically updated in constant time as additional data of surrounding vehicles is made available.
+This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used an steps required extract acceleration and steering angle for each vehicle, and Section \ref{sec:models} introduces several time series models for these variables.
+Section \ref{sec:Inference} reviews Bayesian inference, while Section \ref{sec:UVB} extends Variational Bayes to the online updating setting.
+Finally Section \ref{sec:eval} contains empirical results and Section \ref{sec:conc} concludes.
 
@@
 \label{sec:dataProcessing}
-Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 Freeway in Los Angeles, California from 7:50 am to 8:35 am on June 15th, 2005.
-Data was collected by seven static cameras and processed by Cambridge Systematics Inc.
-The result is coordinates relative to the start of the road section for 6101 vehicles collected every 100 milliseconds.
-The average number of observations per vehicle is 672, corresponding to a 67.2 seconds of data.
-Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which travel along the five major lanes or the enter from the entry/exit lane to the right, with one vehicle changing from Lane 2 to Lane 1.
-Vehicles that entered or exited midway through the freeway section are excluded in this paper.
-There is a curvature to the road, with bends occuring between 500 and 1000 feet and 1800 and 2000 feet.
+Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005.
+Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc to produce coordinates for each vehicle and point in time relative to the start of the road section.
+Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which traveled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1.
+Vehicles that entered or exited midway through the freeway section, are excluded in this paper.
+There is a curvature to the road occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.
+\\
 
@@
 \includegraphics[width=0.65\textwidth]{carPath}
-\caption{The path of ten vehicles in the dataset.
-The freeway is split into five lanes, and an additional entry/exit lane to the far right, through which two vehicles from this sample have entered.
-There is curvature to the road slightly distored by the aspect ratio, with bends occuring around 500 and 1800 feet.}
+\caption{The path of ten vehicles in the dataset, with each black line representing a unique vehichle.
+This section of US 101 is split into five main lanes, with an additional entry/exit lane to the far right, through which two of the vehicles have entered.
+There is curvature to the road (which is slightly distorted by the aspect ratio) with bends occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.}
 \label{fig:rawData}
@@
 
-Modern vehicles are capable of tracking lane marking, and thus can compensate for the curvature of the road as they calculate the location of surrounding vehicles in coordinates relative to their own position (Thuy and Leon 2010), rather than relative to a static point as provided by NGSIM.
-To compensate for this, and so that the data is in a coordinate system similar to that of a self driving vehicle tracking surrounding cars, the coordinate system provided in the NGSIM dataset is transformed into a relative system where the road curvature is removed.
-This is facilitated by measuring the location of each vehicle relative to the estimate of the midpoint of each lane.
-\citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to build a model of the lane midpoints for the entire road section.
+Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road compensating for any curvature.
+To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates.
+This is facilitated by measuring the location of vehicles relative to an estimate of the midpoint of their associated lane.
+\citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway.
+Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
+\\
 
-The midpoints for each of the five lanes are estimated separately, using 100 randomly sampled vehicles each that did not change their lane over the observed period.
-For each sampled vehicle $i$ and time since entering the road $t$, the total distance travelled is calcualted as
-\begin{equation}
-\label{distance}
-d_{i, t} = \sqrt{(x_{i, t}) - x_{i, 1})^2 + (y_{i, t} - y_{i, 1})^2}
-\end{equation}
-From this distance metric, the average $x$ and $y$ coordinates at each given lane and for a given distance are each estimated by a pair of smoothing splines calculated with the R `stats' package.
-The smoothing splines are functions of the distance travelled and estimate the lane midpoints as  $\hat{x}_{i, t} = f(d_{i, t})$ and $\hat{y}_{i, t} = g(d_{i, t})$.
-Each remaining vehicle in the dataset uses the lane midpoint estimates $\{\hat{x}^*_{i, t}, \hat{y}^*_{i, t}\}$ fit from the spline model assocaiated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the devitation from the lane midpoint, by
-\begin{align}
-x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right) - \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
-y^*_{i, t} &= d_{i, t}. \label{yRel}
-\end{align}
-The vehicles used to estimate the five spline models are excluded for any other purpose.
-\begin{figure}
-\centering
-\includegraphics[width = 0.5\textwidth]{relCoord}
-\caption{An example of the equation in (\ref{xRel}).
-The estimated trajectory of the lane midpoint is given by the solid curve, while a vehicle $i$ at a given point in time $t$ is given by the blue dot at $\{x_{i, t}, y_{i, t}\}$.
-This vehicle has travelled a distance equivalent to the red dot at $\{\hat{x}_{i, t}, \hat{y}_{i, t}\}$.
-The gradient of the midpoint at this point is given by the dashed line, which intercepts the x-axis at an angle of $\lambda = \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)$.
-The dotted line from the blue dot to the red intercepts the x-axis at an angle of $\psi = \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)$.
-The sign of the angle $\lambda - \psi$ determines whether the vehicle is to the left or right side of the lane, and thus has a negative or positive relative coordinate $x^*_{i, t}$.
-The absolute value of $x^*_{i, t}$ is the distance between the two dots.}
-\label{fig:relCoord}
-\end{figure}
-Once relative coordinates are extracted, the relative position of any vehicle $i$ at any time $t$ is determined by its initial position $\{x^*_{i, 1}, y^*_{i, 1}\}$, as well as the history of that driver's inputs: velocity $\{v_{i, s} | s = 2, \dots, t\}$ and steering angle, $\{\delta_{i, s} | s = 2, \dots, t\}$ by the trigonometric relationship in Figure \ref{fig:motion}:
+Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from time $t-1$ to time $t$ according to the trigonometric relationship in Figure \ref{fig:motion}:
 \begin{align}
@@
 \end{align}
-where $\delta_{i, t} = \pi/2$ denotes that the car has no lateral relative movement and $x^*_{i, t} = x^*_{i, t-1}$.
-From this relationship, the driver's inputs are be extracted by
+Note that when $\delta_{i, t} \pm \pi/2$ then car $i$ takes a position parallel to the centre line, and hence $x^*_{i, t} = x^*_{i, t-1}$.
+\\
+
+From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs from driver $i$ are calculated via
 \begin{align}
-\delta_{i, t} &= \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right) \label{dEq} \\
+\delta_{i, t} &= 
+     \begin{cases}
+       \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right)  &\quad\text{if }x^*_{i, t} \neq x^*_{i, t-1} \\
+       \frac{\pi}{2} &\quad\text{if } y^*_{i, t} > y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
+       -\frac{\pi}{2} &\quad\text{if } y^*_{i, t} < y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
+       \delta_{i, t-1} &\quad\text{otherwise} \\ 
+     \end{cases} \label{dEq} \\
 v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
 \end{align}
-It is then useful to extract the acceleration, $a_{i, t}$ by the simple relation
+In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=2,3,...,T\}$, according to
 \begin{equation}
@@
 \end{equation}
-Data for vehicle $i$ up to time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $z_{1:N} = \{z_{i, 1:T_i} | i = 1, \dots, N\}$.
+\\
+
+Data for vehicle $i$ up to time $T_i$ is defined as $z_{i, 1:T_i} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T_i\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T_i} | i = 1, \dots, N\}$.
+3373 vehicles remain in the dataset after the above exclusion criteria are applied.
+The data into two sets of vehicles: a training set of 2000 vehicles encountered before driving, and a test set of 1373 vehicles encountered while driving.
 \begin{figure}
@@
 \includegraphics[width = 0.4\textwidth]{motion}
-\caption{A vehicle at $x_{t-1}, y_{t-1}$ at time $t-1$ will be at $x_t, y_t$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
+\caption{A vehicle at coordinate $\{x^*_{t-1}, y^*_{t-1}\}$ at time $t-1$ will be at $\{x^*_t, y^*_t\}$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
 \label{fig:motion}
@@
 
-\section{Trajectory Forecasts}
-\label{sec:trajForecasts}
+\section{Models for Trajectory Forecasting}
+\label{sec:models}
 
-A crucial step in the safe operation of self driving vehicles is to forecast the future trajectory of surrounding vehicles so that collisions can be avoided.
-To produce these forecasts the data is split to place vehicles into two sets: a training set used to infer the posterior distribution of a parameter set before the self driving vehicle is on the road; and a test set used for forecasting the positions of vehicles observed while driving.
-The parameters associated with the vehicles in the test set may be inferred, but this inference is time constrained.
-If inference takes $M$ seconds, a forecast after observing a vehicle for $T$ seconds can only include data up to time $T - M$ to infer the posterior distribution of any parameters.
-
-The remainder of Section \ref{sec:trajForecasts} will introduce the models used, while Sections \ref{sec:Inference} and \ref{sec:UVB} will build an inferential framework to facilitate inference of any additional parameters in constant time as $T$ increases to apply to online trajectory forecasting.
+Three similar models are introduced to produce trajectory forecasts, differing by their implications about the dependency between different vehicles.
+These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent model and the hierarchical model.
+The independent model does not incorporate information from the training set of vehicles into inference for the test set, while the hierarchical model allows information to be shared between vehicles; it learns the range of behaviours possible from the training set and incorporates this into inference about the test set.
 
@@
 $\{a_{t}, \delta_{t} | t = T + 1, \dots, T+H\}$.  
+\\
 
-As data is observed every 100 milliseconds, large changes in $a_{i, t}$ and $\delta_{i, t}$ can occur over multiple time periods.
-This can be observed in Figure \ref{fig:dynamics}, where the left panel plots the mean value of
+\iffalse % Partial Autocorrelation plots are a better way to show dynamic behaviour in the data.
+This is highlighted in Figure \ref{fig:dynamics}, where the left panel plots the mean value of
 \begin{equation}
@@
 where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$.
-On average, a vehicle will increase in acceleration for the 400 milliseconds before reaching maximum acceleration.
+On average, large vehicle acceleration or deceleration is preceeded by an increase in acceleration in the same direction for 400 milliseconds.
 The right panel plots the mean value of
@@
 \end{equation}
-where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, which similarly demonstrates that large absolute deviations in steering angle from $pi/2$ are associated with smaller deviations in the same direction in the the previous 300 milliseconds.
-
-\begin{figure}[h]
+where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, which similarly demonstrates that steering angle deviation from $\pi/2$ are associated with smaller deviations in the same direction in during the previous 300 milliseconds.
+\begin{figure}[ht]
 \centering
@@
 \caption{Left: Mean values of acceleration as a percent of a vehicles maximum absolute acceleration for the 1000 milliseconds before, and 500 milliseconds after, a vehicle reaches its maximum absolute acceleration.
-Right: The analogous plot for the steering angle, with $\delta_t - \pi/2$ as a percentage of a vehicles maximum absolute value of $\delta_t - \pi/2$.
-The few observations before a variable reaches its maximum often have increasing values for the same variable.}
+Right: The analogous plot for the steering angle, where deviations in angle from $\pi/2$ are plotted as a percentage of a vehicles maximum absolute value of deviation.
+The few observations before either variable reaches its maximum often have increased values in the same direction.}
 \label{fig:dynamics}
 \end{figure}
+\fi
 
-These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and thus auto-regressive processes of orders $p$ and $q$ for $a_t$ and $\delta_t$ are employed to model this, with
+NGSIM vehicle data is observed every 100 milliseconds so large changes in $a_{i, t}$ and $\delta_{i, t}$ may occur over multiple sequential observations.
+This is highlighted in Figure \ref{fig:pacf}, which plots the partial autocorrelation function, which measures the correlation between two time periods $t$ and $t + k$, conditioned on $t + 1, \dots t + k -1$, for both $a_{i, t}$ and $\delta_{i, t}$ for a selection of vehicles in the sample.
+Large spikes in the partial autocorrelation plots, such as those at the first and second lag for most vehicles, demonstrate that the value of either $a_t$ or $\delta_t$ is strongly correlated with the previous time periods.
+\\
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.95\textwidth]{pacf}
+\caption{Top: Partial Autocorrelation plots for acceleration, $a$, for a selection of vehicles.
+Bottom: Partial Autocorrelation plots for steering angle, $\delta$ for the same vehicles.
+Each spike for vehicle $i$ at lag $k$ in the top panel indicates that $a_{i, t-k}$ is correlated with $a_{i, t}$, similarly spikes in the bottom panel indicate that $\delta_{i, t-k}$ is correlated with $\delta_{i, t}$.
+Note that different vehicles do not have the same dynamic behaviour in the partial autocorrelations.}
+\label{fig:pacf}
+\end{figure}
+
+These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and thus auto-regressive processes of orders $p$ and $q$ for $a_t$ and $\delta_t$ are employed with
 \begin{align}
@@
 for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution.
+\\
 
-The parameter vector is transformed to $\mathbb{R}^{p + q + 2}$ and collected as 
+The parameters are transformed to $\mathbb{R}^{p + q + 2}$ and collected as the vector
 \begin{equation*}
@@
 \end{equation*}
-where
+with the prior
 \begin{equation}
 \label{indPrior}
-p(\theta) \sim \mathcal{N}\left(\mu, \Sigma \right)
+\theta \sim \mathcal{N}\left(\mu, \Sigma \right),
 \end{equation}
-where $\mu$ is a vector with an entry of $0$ for each $\phi$ and $\gamma$, and an entry of $-5$ for each log-variance, as the scale of changes in the data is very small.
-$\Sigma = 10 \mathbb{I}$, where $\mathbb{I}$ is the identity matrix.
+where $\mu$ is a vector with an entry of $0$ for each $\phi$ and $\gamma$, and an entry of $-5$ for each log-variance, as the scale of changes in the data at the 100 millisecond time-scale are very small.
+The prior variance $\Sigma$ is set to $10 \mathbb{I}$, where $\mathbb{I}$ is the identity matrix.
 This is referred to as the homogenous model throughout the paper.
@@
 \end{align}
-Using the prior in (\ref{indPrior}) for each $\theta_i$ formulates what is referred to as the independent model where each vehicle posterior is independent of all other vehicles, so $p(\theta_i | z_{1:N}, \theta_{j \neq i}) = p(\theta_i | z_i)$.
-An alternative approach is to augment the prior distribution of each $\theta_i$ with parameters $\beta$ in a hierarchical model, and infer the distribution of the $p(\theta_{1:N}, \beta | z_{1:N})$.
-The top panel of Figure \ref{fig:HierSingleKDE} plots a kernel density estimate of the posterior means of each $\theta_i, i = 1, \dots, 2000$ using the independent model with $p = q = 2$ at time $T = 500$.
+Use of the prior (\ref{indPrior}) for each $\theta_i$ formulates what is referred to as the independent model; where each individual $\theta_i$ posterior is independent of all other, i.e.
+$p(\theta_i | \textbf{z}_{1:N}, \theta_{j \neq i}) = p(\theta_i | z_i)$.
+\\
+
+An alternative approach is to augment the prior distribution of each $\theta_i$ with parameters $\beta$ in a hierarchical model, and infer the distribution of $p(\theta_{1:N}, \beta | \textbf{z}_{1:N})$.
+The top panel of Figure \ref{fig:HierSingleKDE} plots a kernel density estimate of the posterior means of each $\theta_i, i = 1, \dots, 2000$, using the independent model with $p = q = 2$ at time $T = 500$.
 There is a wide range of values for the posterior mean of each variable, indicating heterogeneity across the drivers in the sample.
 The density of posterior means for both variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the density of posterior means of both $\phi_1$ and $\phi_2$ are multimodal.
-Due these features, a $K$ component mixture of multivariate normal distributions is chosen for $p(\theta_i | \beta)$, augmenting each driver with an auxiliary variable $k_i = 1, \dots, K$ such that
+Due these features a $K$ component mixture of multivariate normal distributions is chosen for $p(\theta_i | \beta)$, augmenting each driver with an auxiliary variable $k_i = 1, \dots, K$ such that
 \begin{equation}
@@
 \Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
-\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right)
+\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right).
 \end{align}
-The elements of $\beta$ are $\{\mu_j, \Sigma_j, \pi_j | j = 1, \dots, K\}$.
+\\
+
+The hyperparameters $\beta$ are $\{\mu_j, \Sigma_j, \pi_j | j = 1, \dots, K\}$.
 Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$.
-The bottom panel of Figure \ref{fig:HierSingleKDE} plots the marginal distributions for $p(\theta_{2001} | \beta, z_{1:2000})$ for the hierarchical model with $p = q = 2$ and $K = 6$ using the same vehicles as the independent model in the top panel.
-The marginal distributions capture the skewness and multimodality present in the densities of posterior means from the independent model.
+The bottom panel of Figure \ref{fig:HierSingleKDE} plots the marginal distributions for $p(\theta_{2001} | \beta, z_{1:2000})$ for the hierarchical model with $p = q = 2$ and $K = 6$ using the same data as the independent model.
+The hierarchical marginal distributions have captured the skewness and multimodality present in the densities in the top panel.
 
@@
 \includegraphics[width = 0.95\textwidth]{HierSingleKDE.png}
-\caption{Top: A Kernel Density Estimate applied to the posterior means for each $\theta_i | z_i, i = 1, \dots, 2000$ where $p = q = 2$ from the independent model, sampled by Metropolis-Hasting MCMC.
+\caption{Top: A Kernel Density Estimate applied to $E(\theta_i | z_i), i = 1, \dots, 2000$ where $p = q = 2$ from the independent model, sampled by Metropolis-Hasting MCMC.
 Bottom: $p(\theta_i | \beta, z_{1:2000})$ for the hierarchical model with $K = 6$ using the same vehicles sampled by Metropolis-Hasting MCMC.
@@
 
-For the remainder of this paper $z_{1:N}$ and $\theta_{1:N}$ refers to the data collected from, and associated parameter vectors for, the $N$ vehicles that are available before the self driving vehicle is on the road, while the subscript $i$ is reserved for some $i > N$ and refer to the data and parameter vector for any additional vehicle encountered while the self driving vehicle is on the road.
-Inference about $\theta_{1:N}$, the individual parameter vectors for the independent and hierarchical models, as well as for $\theta$, the shared homogenous parameter vector and $\beta$, the hierarchical hyperparameters, conditional on $z_{1:N}$ are assumed to be available with an unlimited computational budget.
+For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refers to the data collected from, and the associated parameter vectors for, the $N$ vehicles that are available before the self driving vehicle is on the road, while the subscript $i$ is reserved for some $i > N$ and refers to the data and parameter vector for an additional vehicle in the test set, that is encountered while the self driving vehicle is on the road.
+Inference about $\theta_{1:N}$, the individual parameter vectors for the independent and hierarchical models, as well as for $\theta$, the shared homogenous parameter vector and $\beta$, the hierarchical hyperparameters, conditional on $\textbf{z}_{1:N}$ are assumed to be available with an unlimited computational budget.
+\\
 
@@
 Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} = \theta$.
-In this case $p(\theta | z_{1:N}, z_{i, 1:T})$ will not be significantly different to $p(\theta | z_{1:N})$ due to the high precision of the posterior distribution for a sufficiently large $N$.
+In this case $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not be significantly different to $p(\theta | \textbf{z}_{1:N})$ due to the high precision of the posterior distribution for a sufficiently large $N$.
 Figure (\ref{fig:homogPosterior}) plots the posterior marginal distributions from $N = 2000$ vehicles highlighting the posterior precision.
@@
 \includegraphics[width = 0.95\textwidth]{homogPosterior}
-\caption{The converged Metropolis Hastings MCMC samples of the marginal posterior distributions $p(\theta | z_{1:N})$ for the homogenous model with $N = 2000$ and $p = q = 2$.
+\caption{The converged Metropolis Hastings MCMC samples of the marginal posterior distributions $p(\theta | \textbf{z}_{1:N})$ for the homogenous model with $N = 2000$ and $p = q = 2$.
 Each variable has very high posterior precision.}
@@
 
-The independent model implies that the posterior distribution for $\theta_i$ is independent of $\theta_{1:N}$ and $z_{1:N}$ by
+The independent model implies that the posterior distribution for $\theta_i$ is independent of $\theta_{1:N}$ and $\textbf{z}_{1:N}$ by
 \begin{equation}
-p(\theta_{i}| z_{1:N}, z_{i,1:T}, \theta_{1:N}) = p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_i) p(\theta_i).
+p(\theta_{i}| \textbf{z}_{1:N}, z_{i,1:T}, \theta_{1:N}) = p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_i) p(\theta_i).
 \label{indepNewCar}
 \end{equation}
-The posterior for $\theta_i$ using the hierarchical model, assuming conditional independence between each $\theta_j$ for $j = 1, \dots, N$ and $\theta_i$, given $\beta$, $z_{1:N}$, and $z_i$, is
+\\
+
+The posterior for $\theta_i$ using the hierarchical model, assuming conditional independence between each $\theta_j$ for $j = 1, \dots, N$ and $\theta_i$, given $\beta$, $\textbf{z}_{1:N}$, and $z_i$, is
 \begin{equation}
 \label{hierNewCar}
-p(\theta_{i} | z_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) \int_{\beta} p(\theta_{i} | \beta) p (\beta | z_{1:N}) d\beta.
+p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) \int_{\beta} p(\theta_{i} | \beta) p (\beta | \textbf{z}_{1:N}) d\beta.
 \end{equation}
-As N is large the sampled posterior $p(\beta | z_{1:N})$ has high precision, and $\int_{\beta} p(\theta_{i} | \beta) p (\beta | z_{1:N}) d\beta$ can be replaced with $p(\theta_{i} | \hat{\beta})$ in (\ref{hierNewCar}) where $\hat{\beta}$ is a point estimate such as the mean or maximum of $p(\beta | z_{1:N})$.
+As N is large the sampled posterior $p(\beta | \textbf{z}_{1:N})$ has high precision, and $\int_{\beta} p(\theta_{i} | \beta) p (\beta | \textbf{z}_{1:N}) d\beta$ can be replaced with $p(\theta_{i} | \hat{\beta})$ in (\ref{hierNewCar}) where $\hat{\beta}$ is a point estimate such as the mean or maximum of $p(\beta | \textbf{z}_{1:N})$.
 This approximation results in the closed form expression
@@
 \label{hierNewCar2}
-p(\theta_{i} | z_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) p(\theta_{i} | \hat{\beta}).
+p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) p(\theta_{i} | \hat{\beta}).
 \end{equation}
-(\ref{indepNewCar}) and (\ref{hierNewCar2}) can be combined with an inference procedure to allow the parameter set for any vehicle observed while driving to be inferred, and used in the forecast equation (\ref{predictive}) to produce online forecasts of the trajectory of those vehicles that allows individual heterogeneity to be included.
+\\
+
+Combining a fast inference procedure with the test vechile posterior distributions implied by the independent and hierarchical models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self driving vehicle observes data for the vehicles it encounters.
 
@@
 must first be inferred. Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}), will be unavailable due to the complex functional form of the distributions involved. 
+\\
 
@@
 In brief, MCMC is used to create a sample from $p(\theta | z_{1:T})$, with any function of $\theta$ that is desired estimated from that sample.
-In contrast, VB replaces $p(\theta | z_{1:T})$ with a parametric approximation, denoted by $q_{\lambda}(\theta | z_{1:T})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $y_{1:T}$.
+In contrast, VB replaces $p(\theta | z_{1:T})$ with a parametric approximation, denoted by $q_{\lambda}(\theta | z_{1:T})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $z_{1:T}$.
 
@@
 Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
-The computation time for each iteration and the overall number of iterations required to accruately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
-The full conditional distributions cannot be recognised for each of the time series models proposed in Section \ref{sec:trajForecasts} and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
+The computation time for each iteration and the overall number of iterations required to accurately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
+The full conditional distributions cannot be recognised for each of the time series models proposed in Section \ref{sec:models} and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
+\\
 
-In MH MCMC, a candidate $\theta_j^{(c)}$, where $\theta_j$ may be any scalar or vector subset of $\theta$, is drawn from a proposal distribution $r(\theta_j)$ and accepted by the sampler with probability
+In MH-MCMC, a candidate $\theta_j^{(c)}$, where $\theta_j$ may be any scalar or vector subset of $\theta$, is drawn from a proposal distribution $r(\theta_j)$ and accepted by the sampler with probability
 \begin{equation}
@@
 as $r(\theta_j^{(c)} | \theta_j^{(a)}) = r(\theta_j^{(a)} | \theta_j^{(c)})$.
+\\
+
 Performance of Random Walk Metropolis Hastings MCMC (RWMH-MCMC) then depends on the acceptance rate of the proposal distribution, largely influenced by the value of $\Sigma_j$.
-\citet{Garthwaite2016} propose an adaptive algorithm that increases or decreases $\Sigma_j$ depending if the most recent candidate was rejected or accepted respectively so that a target acceptance rate is asymptotically approached as the number of MCMC iterations approaches infinity.
+\citet{Garthwaite2016} propose an adaptive algorithm that increases or decreases $\Sigma_j$ depending on whether the most recent candidate was rejected or accepted respectively so that a target acceptance rate is asymptotically approached as the number of MCMC iterations approaches infinity.
 
@@
 Note that the family $q_{\lambda}(\theta | z)$ does not depend on $z$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta | z)$.
-A member of approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta | z)$ to $p(\theta | z)$ \citep{Kullback1951}, given by $KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})]$.
+A member of approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta | z)$ to $p(\theta | z)$, given by $KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})]$ \citep{Kullback1951}.
 The KL divergence is defined by
@@
 and is a non-negative, asymmetric measure of the discrepancy between $p(\theta | z)$ and $q_{\lambda}(\theta | z)$  that will be equal to zero if and only if $p(\theta | z) = q_{\lambda}(\theta | z)$ almost everywhere \citep{Bishop2006}.
+\\
+
 Typically (\ref{KL-def}) cannot be evaluated, and Monte-Carlo estimates as
@@
 \end{equation}
-where $\theta_i \sim q_{\lambda}(\theta |z))$ are compuationally infeasible due to the inclusion of the term $p(\theta | z)$.
+where $\theta_i \sim q_{\lambda}(\theta |z))$ are compuationally infeasible due to the inclusion of the term $p(\theta | z)$, which is only known up to proportionality.
 Instead VB uses the Evidence Lower Bound (ELBO), denoted by $\mathcal{L}(q, \lambda)$, as an error function where
@@
 Maximisation proceeds by optimising only the auxiliary parameters $\lambda$ for a fixed distribution family $q$ with stochastic gradient ascent (SGA).
+\\
+
 SGA repeatedly takes Monte-Carlo estimates of the gradient of the ELBO with respect to $\lambda$, $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, as $\widehat{\delta\mathcal{L}(q, \lambda) / \delta \lambda}$ and applies updates of the form
@@
 In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of \citet{Kingma2015b}.
+\\
 
-There are two choices for the Monte Carlo estimator of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of \citet{Ranganath2014}, 
+There are two popular choices for the Monte Carlo estimator of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of \citet{Ranganath2014}, 
 \begin{equation}
@@
 where $\theta_j \sim q_{\lambda}(\theta | z)$ and the reparameteterised estimator of \citet{Kingma2014}.
-Reparameterisation rephrases Variational Bayes as a search for the parameters $\lambda$ of some differentiable function $f(\cdot, \lambda)$ that minimises the Kullback Leibler divergence from some distribution $r(\epsilon)$ with zero free parameters to a transformation of the posterior distribution, $p_{\epsilon}(\theta = f(\epsilon, \lambda) | z)$.
-Examples of $f$ and $r(\epsilon)$ include treating $\theta$ as location scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
-The distribution $p(\theta, z)$ is written as
+Reparameterisation introduces an auxiliary variable $\epsilon$ and differentiable function $f(\cdot, \cdot)$ to rephrase Variational Bayes optimisation as the equivalent search for the parameters $\lambda$ that minimises the Kullback Leibler divergence from some distribution $q(\epsilon)$ with zero free parameters to the posterior distribution implied by the transformation $\theta = f(\epsilon, \lambda)$:
 \begin{equation}
 \label{rpDist}
-p(\theta, z) = p(f(\epsilon, \lambda), z)|J(f(\epsilon, \lambda))|
+p(f(\epsilon, \lambda) | z) = p(\theta | z) |J^{-1}(f(\epsilon, \lambda))|
 \end{equation}
-where $J(f(\epsilon, \lambda))$ is the Jacobian Matrix of the transformation $f(\epsilon, \lambda)$, and the reparameterised ELBO can be found by substituting (\ref{rpDist}) into (\ref{ELBO}),
+where $J(f(\epsilon, \lambda))$ is the Jacobian Matrix of the transformation $f(\epsilon, \lambda)$.
+Examples of $f$ and $q(\epsilon)$ include treating $\theta$ as location scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
+\\
+
+The ELBO can be reparameterised by substituting $p(\theta, z) = p(f(\epsilon, \lambda), z)|J(f(\epsilon, \lambda))|$ into (\ref{ELBO}),
 \begin{equation}
 \label{rpELBO}
-\mathcal{L}(q, \lambda) = E_{r(\epsilon)} \bigg[\log(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|) - \log(r(\epsilon))\bigg].
+\mathcal{L}(q, \lambda) = E_{r(\epsilon)} \bigg[\log(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|) - \log(q(\epsilon))\bigg].
 \end{equation}
-The gradient of the reparameterised ELBO with respect to $\lambda$ of 
+The gradient of the reparameterised ELBO with respect to $\lambda$ is given by 
 \begin{align}
 \label{rpELBODeriv}
-\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} &= \frac{\delta}{\delta \lambda} \bigg( E_{r(\epsilon)} \bigg[\log\big(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|\big) - \log(r(\epsilon))\bigg] \bigg) \nonumber \\
-&= E_{r(\epsilon)} \left[ \frac{\delta}{\delta \lambda} \bigg(\log(p(f(\epsilon,\lambda), z)) + \log(|J(f(\epsilon, \lambda))|) - \log(r(\epsilon)) \bigg)\right] \nonumber \\
-&= E_{r(\epsilon)} \left[ \frac{\delta \log(p(f(\epsilon,\lambda), z))}{\delta f(\epsilon,\lambda)} \frac{\delta f(\epsilon,\lambda)}{\delta \lambda}  + \frac{\delta \log(|J(f(\epsilon, \lambda))|)}{\delta \lambda} \right].
+\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} &= \frac{\delta}{\delta \lambda} \bigg( E_{q(\epsilon)} \bigg[\log\big(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|\big) - \log(q(\epsilon))\bigg] \bigg) \nonumber \\
+&= E_{q(\epsilon)} \left[ \frac{\delta}{\delta \lambda} \bigg(\log(p(f(\epsilon,\lambda), z)) + \log(|J(f(\epsilon, \lambda))|) - \log(q(\epsilon)) \bigg)\right] \nonumber \\
+&= E_{q(\epsilon)} \left[ \frac{\delta \log(p(f(\epsilon,\lambda), z))}{\delta f(\epsilon,\lambda)} \frac{\delta f(\epsilon,\lambda)}{\delta \lambda}  + \frac{\delta \log(|J(f(\epsilon, \lambda))|)}{\delta \lambda} \right].
 \end{align}
-This form leads to the reparameterised estimator of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$
+This form leads to the reparameterised gradient estimator,
 \begin{equation}
@@
 \end{equation}
-where $\epsilon_j \sim r(\epsilon)$.
+where $\epsilon_j \sim q(\epsilon)$.
+The reparameterised gradient estimator typically has lower variance than the score estimator (see eg.
+\cite{Rezende2014}; \cite{Ruiz2016}), but treating $q_{\lambda}(\theta | z)$ as the distribution implied by the transformation $f$ of $q(\epsilon)$ restricts the class of approximating families that can be used.
+\\
 
-These Monte Carlo estimators converge to the true derivative at a rate $O(\sqrt{M})$, so choosing the number of samples per estimate, $M$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate.
+Choosing the number of samples per estimate, $M$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate.
 An increased value of $M$ will reduce the estimator variance, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
-To reduce the variance of the Monte Carlo estimator, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used, where numbers in the unit hypercube are generated according to a Sobol Sequence \citep{Sobol1967} which are then randomised using the scrambled net method \citep{Matousek1998}, and transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$.
-Using RQMC allows the gradient estimators to converge to the true derivative at the increased rate of $O(M)$.
+To reduce the variance of the Monte Carlo estimator, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) which has shown to be more efficient, where numbers in the unit hypercube are generated according to a Sobol Sequence \citep{Sobol1967} which are then randomised using the scrambled net method \citep{Matousek1998}, and transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$.
+Use of RQMC has shown to be more efficeint than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
 
@@
 
-The self driving vehicle is constantly observing data about the movements of the surrounding vehicles.
-After observing vehicle $i$ until time $S$, $z_{i, 1:S}$ has been observed and can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the conditional dependence on $z_{1:N}$.
+The self driving vehicle is constantly observing data about the movements of the surrounding vehicles which can be used for posterior inference.
+After observing vehicle $i$ until time $S$, $z_{i, 1:S}$ has been observed and can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the conditional dependence on $\textbf{z}_{1:N}$.
 At time $T > S$ an additional $T - S$ data points from vehicle $i$ have been observed, and forecasts could be improved by incorporating these to form the posterior approximation, $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$, where the $S$ and $T$ subscripts on $\lambda$ are included to differentiate the auxiliary parameter vector conditioned on data up to times $S$ and $T$.
 To facilitate this posterior update, an Updating Variational Bayes (UVB) mechanism is introduced where only $z_{i, S+1:T}$ needs to be processed.
+\\
 
@@
 Without being able to evalute the right hand side, posterior inference requires evaluation of the prior distribution and full sample $z_{i, 1:T}$.
-To avoid this, Variational Bayesian Updating replaces $p(\theta_{i} | z_{i, 1:S})$ with the analytical approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$.
-The update (\ref{updatePost}) is replaced with
+To avoid this, UVB replaces $p(\theta_{i} | z_{i, 1:S})$ with the analytical approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, leading to the approximate joint distribution
 \begin{equation}
-\label{updateEqApprox}
-p(\theta_{i} |  z_{i, 1:T}) \approxprop p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S})
+\label{ApproxJoint}
+\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S})
 \end{equation}
-The ELBO gradient estimators to construct the updated Variational Bayes approximation at time $T$,  $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$, can be obtained by substituting (\ref{updateEqApprox}) into the score estimator (\ref{scoreDeriv}) or the reparameterised estimator (\ref{rpDeriv}) of $\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} $.
+\\
+
+The ELBO gradient estimators to construct the updated Variational Bayes approximation at time $T$,  $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$, can be obtained by substituting (\ref{ApproxJoint}) into the score gradient estimator (\ref{scoreDeriv}) or the reparameterised gradient estimator (\ref{rpDeriv}).
 The updating score estimator is given by
 \begin{align}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{SCU} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
+\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
 &\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i} z_{i, 1:T})) \right) \label{scoreUpdate}
@@
 \label{rpUpdate}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{RPU} = \sum_{j = 1}^M \frac{\delta f(\lambda_T, \epsilon_j)}{\delta \lambda_T} \frac{\delta \log(q_{\lambda_S}(\theta_{i} |z_{i, 1:S}))}{\delta \theta_{i}} \bigg\rvert_{\theta_{i} = f(\lambda_T, \epsilon_j)} + \frac{\delta J(\lambda_T, \epsilon_j)}{\delta \lambda_T},
+\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} = \sum_{j = 1}^M \frac{\delta f(\lambda_T, \epsilon_j)}{\delta \lambda_T} \frac{\delta \log(q_{\lambda_S}(\theta_{i} |z_{i, 1:S}))}{\delta \theta_{i}} \bigg\rvert_{\theta_{i} = f(\lambda_T, \epsilon_j)} + \frac{\delta J(\lambda_T, \epsilon_j)}{\delta \lambda_T},
 \end{equation}
 where $\epsilon_j \sim r(\epsilon)$.
-If this process is repeated periodically, $S$ will grow at the same rate as $T$, and updating VB is $O(1)$ rather than $O(T)$ as required by re-using the full sample without the approximation step in (\ref{updateEqApprox}).
+If this process is repeated periodically, $S$ will grow at the same rate as $T$, and updating VB is $O(1)$ rather than $O(T)$ as required by re-using the full sample without the approximation stepin (\ref{ApproxJoint}).
 
@@
 \begin{center}
-\begin{table}[h]
+\begin{table}[ht]
 \resizebox{\textwidth}{!}{%
@@
 A Recurrent Neural Network (RNN) is constructed with inputs following \citet{Ding2013}, the ten most recent lags of vehicle position, angle, velocity, and time headway to the preceding vehicle.
-The RNN outputs the next thirty changes in the vehicle position, is trained with the Adam optimiser and a a loss function $EE_{i, T, 30}$ given by (\ref{eucError}).
+The RNN outputs the next thirty changes in the vehicle position, is trained through gradient descent using the Adam optimiser to minimise the euclidean error of the three second ahead point forecast corresponding to $EE_{i, T, 30}$ defined in (\ref{eucError}).
 As of writing this report forecast results from this model are not competitive with the naive models and are not discussed further.
@@
 
-\begin{itemize}
-\item Discuss the updating method relative to standard VB, change in logscore in Figure \ref{fig:updateCost}
-\item Plot the average convergence time as a function of $T - S = 1, 5, 10, 20$ and $T = 100, 110, \dots,  450$. Running now for Figure \ref{fig:timing}
-\end{itemize}
-
 Mean Eucliedean Error for each model at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
@@
 At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model.
+\\
 
@@
 As the independent model is the heterogenous model that provides the least accurate forecasts it is not discussed further.
+\\
 
@@
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 The homogenous variance posterior means are greater than $80\%$ and $91.5\%$ of heirarchical variance posterior means for acceleration and the steering angle respectivley.
-This large variance has little impact on the location of the MAP of $p(z_{i, T+h} | z_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density.
+This large variance has little impact on the location of the MAP of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density.
 Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the hierarchical model with updating VB inference over the homogenous model at the three second forecast horizon.
 Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis.
-The hierarchical model trajectory forecasts for drivers in the lower variance quntiles benefit the most from including heterogeneity.
+The hierarchical model trajectory forecasts for drivers in the lower variance quintiles benefit the most from including heterogeneity.
 However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles.
 This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance estimates allow identification of vehicles with high variance, and potentially more erratic and dangerous driving styles.
+\\
 
-\begin{figure}[h]
+\begin{figure}[ht]
 \centering
@@
 
-The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | z_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | z_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self driving vehicles.
+The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self driving vehicles.
 Figure (\ref{fig:timing}) demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$.
-The compuation time for standard VB increases linearly with $T$, while UVB is constant; posterior inference avabale at time $T$ utilising UVB can be conditioned on more data than standard VB.
+The compuation time for standard VB increases linearly with $T$, while UVB is constant; posterior inference available at time $T$ utilising UVB can be conditioned on more data than standard VB.
 Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time and make UVB feasible for more complex models. 
-Figure (\ref{fig:updateCost}) demonstrates that implementing UVB over standard VB does not lead to a decrease in forecast logscores, plotting the difference between the heirarchical forecast logscores obtained by UVB and standard VB for several ranges of $T$.
+\\
+
+As UVB involves replacing the posterior at time $S$ with its variational approximation, the error introduced by UVB depends on the approximation error of Variational Bayes itself.
+The forecast Euclidean error increaseing when inferring the independent model posterior with UVB in Figure \ref{fig:PredErrorZ} compared to MCMC or VB is indicative of the multivariate Gaussian approximating distribution being an unsuitable replacement to the true posterior.
+The Hierarchical Model did not increase in error relative to other methods, possibly due to the increased flexibility of its six component mixture approximating distribution.
+Figure (\ref{fig:updateCost}) demonstrates that UVB did not decrease in forecast logscores for the hierarchical model, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
 Augmenting self driving vehicles with hierarchical time series models to forecast the trajectory of surrounding traffic, and repeatedly updating the posterior distribution and forecasts, can be feasibly implemented with UVB.
@@
 Each UVB procedure originally fits the model to the first 100 data points and updates this every $T - S$ data points.
-VB algorithms are ran with $M = 50$ on one CPU core.
+The convergence time for Standard VB increased linearly for $T$, but is constant for UVB.
+Each VB algorithm is ran with $M = 50$ on one CPU core.
 Convergence time could be reduced through parallelisation.}
@@
 \section{Conclusion}
+\label{sec:conc}
 \begin{itemize}
@@
 
+\appendix
+\section{Transformation to Relative Coordinates}
+
+The centre line for each of the five lanes is estimated separately, using 100 randomly sampled vehicles per lane that did not change their lane over the observed period. 
+For each vehicle $i$ and time $t$ since entering the road, with travel originating at time one given by $\{x_{i,1}, y_{i,1}\}$, the total distance travelled is calcualted as 
+\begin{equation}
+\label{distance}
+d_{i, t} = \sqrt{(x_{i, t} - x_{i, 1})^2 + (y_{i, t} - y_{i, 1})^2}.
+\end{equation}
+Using this distance metric and the data from the sampled 100 cars per lane, the two-dimensional coordinates corresponding to the centre line of each lane are estimated via independent smoothing splines where each coordinate is a function of the distance travelled to that point.
+Each smoothing spline is calculated using the `R stats' package \citep{R}.
+The estimated centre line for lane $k$, is denoted by the curve $\{(\hat{x}_{d,k} = f_k(d), \hat{y}_{d,k} = g_k(d)\}$, for $d \geq 0$.
+\\
+
+Excluding the vehicles used to estimate the spline model, each of the 3373 remaining vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
+\begin{align}
+x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right) - \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
+y^*_{i, t} &= d_{i, t}. \label{yRel}
+\end{align}
+\\
+
+\begin{figure}
+\centering
+\includegraphics[width = 0.5\textwidth]{relCoord}
+\caption{An example of the relative coordinate transformation for a vehicle as described by equation (\ref{xRel}).
+The estimated trajectory of the lane midpoint is given by the solid curve, while a vehicle $i$ at a point in time $t$ is denoted by the blue dot at $\{x_{i, t}, y_{i, t}\}$.
+This vehicle has travelled a distance equivalent to the red dot at $\{\hat{x}_{i, t}, \hat{y}_{i, t}\}$.
+The gradient of the midpoint at this point is given by the dashed line, which intercepts the x-axis at an angle of $\lambda = \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)$.
+The dotted line travels from the blue dot to the red, and intercepts the x-axis at an angle of $\psi = \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)$.
+The dashed line intercepts the dotted line with an angle  of $\lambda - \psi$, the sign of which determines whether the vehicle is to the left or right side of the lane centre, and thus has a negative or positive relative coordinate $x^*_{i, t}$.
+The absolute value of $x^*_{i, t}$ is the distance between the two dots.}
+\label{fig:relCoord}
+\end{figure}
+
 \end{document}

### 2018-03-02 Cars/dynamics.R

@@ -35,3 +35,38 @@ timeToAction %>%
 gridExtra::grid.arrange(maxAccel, maxAngle, ncol = 2)
 
 
+
+carsAug %>%
+  filter(ID %in% sample(carsAug$ID, 5)) %>% 
+  group_by(ID) %>%
+  mutate(n = seq_along(v),
+         a = v - ifelse(n == 1, v, lag(v))) -> pacfData
+
+
+pacfs <- data.frame()
+for(i in 1:5){
+  id <- unique(pacfData$ID)[i]
+  
+  pacfData %>%
+    filter(ID == id) %>%
+    .$reldelta %>%
+    pacf(plot = FALSE) %>%
+    with(data.frame(lag, acf)) %>%
+    mutate(ID = paste0('Vehicle ', i),
+           var = 'delta') -> tmp
+  pacfs <- rbind(pacfs, tmp)
+  
+  pacfData %>%
+    filter(ID == id) %>%
+    .$a %>%
+    pacf(plot = FALSE) %>%
+    with(data.frame(lag, acf)) %>%
+    mutate(ID = paste0('Vehicle ', i),
+           var = 'a') -> tmp
+  pacfs <- rbind(pacfs, tmp)
+  
+}
+
+ggplot(pacfs) + geom_bar(aes(lag, acf), stat = 'identity') + 
+  facet_grid(var~ID) + labs(x = 'Lag', y = 'Partial Autocorrelation') + 
+  theme_bw()

### 2018-03-05 Cars/paper.tex
work on paper

@@
 
-Self driving vehicles are rapidly becoming more advanced, with development from large corporations such as Tesla, Ford, and Audi expecting autonomous vehicles available to consumers as early as 2020.
-These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery and selects a path to avoid any collisions.
-As many as $94\%$ of accidents are due to human error according to data from the United States National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, which can feasibly be reduced by self driving vehicles.
-However self driving vehicles will operate in situations where surrounding vehicles are controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
+Self-driving vehicles are rapidly becoming more advanced, with manufacturers such as Tesla, Ford, and Audi expecting autonomous vehicles available to consumers as early as 2020.
+These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery, and selects a path to avoid any collisions.
+As many as $94\%$ of accidents are due to human error according to data from the United States National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, which can feasibly be reduced by self-driving vehicles.
+However self-driving vehicles will operate in situations where surrounding vehicles are controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
 \\
 
-Sophisticated Deep Neural Networks are employed to extract the position, velocity, and steering angle of surrounding traffic from readings provided by sensors such as cameras and Light Detection and Ranging (LiDAR); see \citet{Woo2016a} and \citet{Tian2017} for details.
-From this data, estimates of the future trajectory of surrounding vehicles can be produced.
-Forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced predictive models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
+While characteristics of surrounding vehicles, such as theor postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g.
+\citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory has seen less attention in the literature.
+These forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
 \\
 
-This paper takes a statistical approach to the trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
-It then uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability collisions to be detected.
-The time series models used can allow for heterogeneity between different vehicles, which may result from sources of driver heterogeneity that may result from factors such as differences in individual driving styles, or the conditions of traffic directly around them.
+This paper takes a statistical approach to this trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
+It uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability trajectories that can lead to a collision to be detected.
+The time series models used allow for heterogeneity between different vehicles, which may result from sources of driver heterogeneity that may result from factors such as differences in individual driving styles, or the conditions of traffic directly around them.
 Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
@@
 Incorporating heterogeneity into forecasts requires the behaviour of each driver of interest to be inferred as vehicles are encountered while driving.
-This inference is facilitated by first obtaining the posterior distribution for a large number of vehicles before the self driving vehicle is on the road, from which the prior distribution for additional vehicles can be constructed.
-Once the self driving vehicle is on the road, online inference is made possible by the introduction of Updating Variational Bayes (UVB), which approximates Bayesian updating so that inference may be periodically updated in constant time as additional data of surrounding vehicles is made available.
-This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used an steps required extract acceleration and steering angle for each vehicle, and Section \ref{sec:models} introduces several time series models for these variables.
-Section \ref{sec:Inference} reviews Bayesian inference, while Section \ref{sec:UVB} extends Variational Bayes to the online updating setting.
-Finally Section \ref{sec:eval} contains empirical results and Section \ref{sec:conc} concludes.
+This inference is facilitated by first obtaining the posterior distribution for a large number of vehicles before the self-driving vehicle is on the road, from which the prior distribution for additional vehicles can be constructed.
+Once the self-driving vehicle is on the road, online inference is made possible by the introduction of Updating Variational Bayes (UVB), which approximates Bayesian updating so that inference may be periodically updated as additional data relating to surrounding vehicles is made available.
+\\
+
+We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the contstant models commonly used, and including driver heterogeneity in the model substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous model.
+These results give credence to the validity of the methodological framework developed in this thesis, where exact Bayesian inference techniques are used to guide the construction on a prior distribution for parameters that are approximately inferred in an online setting.
+\\
+
+This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used and steps required to extract acceleration and steering angle for each vehicle, while Section \ref{sec:models} introduces several time series models used.
+Section \ref{sec:Inference} reviews Bayesian inference, which is extended to the online updating setting in Section \ref{sec:UVB}.
+Finally Section \ref{sec:eval} contains empirical results, and Section \ref{sec:disc} concludes the vehicle trajectory forecasting study and discusses future work.
 
@@
 Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road compensating for any curvature.
-To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates.
+To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self-driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates.
 This is facilitated by measuring the location of vehicles relative to an estimate of the midpoint of their associated lane.
@@
 
-Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from time $t-1$ to time $t$ according to the trigonometric relationship in Figure \ref{fig:motion}:
+Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to observation $t$ according to the trigonometric relationship in Figure \ref{fig:motion}:
 \begin{align}
@@
 
-Data for vehicle $i$ up to time $T_i$ is defined as $z_{i, 1:T_i} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T_i\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T_i} | i = 1, \dots, N\}$.
-3373 vehicles remain in the dataset after the above exclusion criteria are applied.
-The data into two sets of vehicles: a training set of 2000 vehicles encountered before driving, and a test set of 1373 vehicles encountered while driving.
+Data resulting from the observation of vehicle $i$ for a period of time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$.
+3387 vehicles remain in the dataset after the above exclusion criteria are applied.
+The data into a two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1387 vehicles.
+\\
+
+The training set and the test set are treated as observations from vehicles at two distinct time periods.
+The training set represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles.
+The test set represents those vehicles the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.
+
 \begin{figure}
@@
 
-Three similar models are introduced to produce trajectory forecasts, differing by their implications about the dependency between different vehicles.
-These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent model and the hierarchical model.
-The independent model does not incorporate information from the training set of vehicles into inference for the test set, while the hierarchical model allows information to be shared between vehicles; it learns the range of behaviours possible from the training set and incorporates this into inference about the test set.
+Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between different vehicles.
+These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent heterogenous (IH) model and the hierarchical, clustered heterogeneity (CH) model.
+The ID model does not incorporate information from the training set of vehicles into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set.
+\\
 
-\subsection{Homogenous Time Series Model}
+\subsection{A Homogenous Time Series Model}
 \label{subsec:homogenous}
 
-The framework provided by (\ref{xEq}) and (\ref{yEq}) allows forecasts of $\{x^*_{t}, y^*_{t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_T, y^*_T, v_T\}$, and forecasts of the future values of
-$\{a_{t}, \delta_{t} | t = T + 1, \dots, T+H\}$.  
+The framework provided by (\ref{xEq}) and (\ref{yEq}) allows distributional forecasts of future position coordinates of a given vehicle $i$, $\{x^*_{i, t}, y^*_{i, t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_{i, T}, y^*_{i, T}, v_{i, T}\}$, and distributional forecasts of the future values of $\{a_{i, t}, \delta_{i, t} | t = T + 1, \dots, T+H\}$.
 \\
@@
 \iffalse % Partial Autocorrelation plots are a better way to show dynamic behaviour in the data.
-This is highlighted in Figure \ref{fig:dynamics}, where the left panel plots the mean value of
+This is highlighted in Figure \ref{fig:dynamics}, where the left panel displays a plot of the mean value of
 \begin{equation}
-A_{i, t} = \frac{a_{i, \tau_i + t}}{a_{i, \tau_i}}
+A_{i, s} = \frac{a_{i, \tau_i + s}}{a_{i, \tau_i}}
 \label{amax}
 \end{equation}
-where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$.
+where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$, the relative time $s$ takes on values between -1000 and 500 milliseconds, and with the average taken over the $i = 1, 2, \dots, N$ individual cars.
 On average, large vehicle acceleration or deceleration is preceeded by an increase in acceleration in the same direction for 400 milliseconds.
-The right panel plots the mean value of
+The right panel displays a plot of the mean value of
 \begin{equation}
-D_{i, t} = \frac{\delta_{i, \omega_i + t} - \pi/2}{\delta_{i, \omega_i} - \pi/2}
+D_{i, s} = \frac{\delta_{i, \omega_i + s} - \pi/2}{\delta_{i, \omega_i} - \pi/2}
 \label{dmax}
 \end{equation}
-where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, which similarly demonstrates that steering angle deviation from $\pi/2$ are associated with smaller deviations in the same direction in during the previous 300 milliseconds.
+where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, for again $-1000 \leq s \leq 500.$ Simmilarly, this second panel demonstrates that the largest steering angle deviation from $\pi/2$ may be anticipated by smaller deviations in the same direction, at typically, some point during the previous 300 milliseconds.
+These plots sugget that changes in acceleration and angle are predictable if given the recent history of a vehicle's position.
 \begin{figure}[ht]
@@
 NGSIM vehicle data is observed every 100 milliseconds so large changes in $a_{i, t}$ and $\delta_{i, t}$ may occur over multiple sequential observations.
-This is highlighted in Figure \ref{fig:pacf}, which plots the partial autocorrelation function, which measures the correlation between two time periods $t$ and $t + k$, conditioned on $t + 1, \dots t + k -1$, for both $a_{i, t}$ and $\delta_{i, t}$ for a selection of vehicles in the sample.
-Large spikes in the partial autocorrelation plots, such as those at the first and second lag for most vehicles, demonstrate that the value of either $a_t$ or $\delta_t$ is strongly correlated with the previous time periods.
+This is highlighted in Figure \ref{fig:pacf}, which displays a plot of the partial autocorrelation function, which measures the correlation between a variable at two different time periods, $t$ and $t + k$, conditioned on that variable at time $t + 1, \dots t + k -1$, for both $a_{i, t}$ and $\delta_{i, t}$ for five vehicles.
+Large spikes in the partial autocorrelation plots, such as those at the first and second lag for most vehicles, demonstrate that the value of either $a_{i, t}$ or $\delta_{i, t}$ is strongly correlated with the previous time periods.
 \\
@@
 \includegraphics[width = 0.95\textwidth]{pacf}
-\caption{Top: Partial Autocorrelation plots for acceleration, $a$, for a selection of vehicles.
+\caption{Top: Partial Autocorrelation plots for acceleration, $a$, for a random selection of vehicles.
 Bottom: Partial Autocorrelation plots for steering angle, $\delta$ for the same vehicles.
@@
 
-These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and thus auto-regressive processes of orders $p$ and $q$ for $a_t$ and $\delta_t$ are employed with
+These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and as an attempt to model the dependence, two univariate auto-regressive processes of orders $p$ and $q$, respectively, for $a_t$ and angle  $\delta_t$ are employed, given by
 \begin{align}
@@
 
-The parameters are transformed to $\mathbb{R}^{p + q + 2}$ and collected as the vector
+For notational convenience, and to facilitate implementation of the Bayesian inferential approach described in Section \ref{sec:Inference}, the variance parameters $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$ are transformed and collected together with all other model parameters into the $\mathbb{R}^{p + q + 2}$-dimensional vector
 \begin{equation*}
 \label{thetaVec}
-\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\},
+\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\}.
 \end{equation*}
-with the prior
+With each component of $\theta$ able to take on any real value, the prior distribution given by
 \begin{equation}
@@
 \end{equation}
-where $\mu$ is a vector with an entry of $0$ for each $\phi$ and $\gamma$, and an entry of $-5$ for each log-variance, as the scale of changes in the data at the 100 millisecond time-scale are very small.
-The prior variance $\Sigma$ is set to $10 \mathbb{I}$, where $\mathbb{I}$ is the identity matrix.
-This is referred to as the homogenous model throughout the paper.
+is selected with $\mu=(0_p^{\prime},0_q^{\prime},-5, -5)$ and $\Sigma = 10 \mathbb{I}_{p+q+2}$, where $0_r$ and $\mathbb{I}_r$ denote the $r$-dimensional zero vector and identity matrix, respectively, noting that the scale of changes in either acceleration or angle at the 100 millisecond time-scale are both very small.
+We refer to the joint specification in (\ref{aAR}) and (\ref{dAR}), along with the given normal prior distributional assumption, as the \textit{homogenous model} throughout the paper.
+\\
+Treating the first $\max(p, q)$ observations for each vehicle as deterministic, the posterior distribution for the homogenous model is given by
+\begin{equation}
+\label{homogPost}
+p(\theta | \textbf{z}_{1:N}) \propto \prod_{i=1}^N \prod_{t = \max(p, q) + 1}^T p(z_{i, t} | z_{i, t-\max(p, q):t-1}, \theta)p(\theta).
+\end{equation}
 
-\subsection{Heterogenous Time Series Models}
+\subsection{Two Heterogenous Time Series Models}
 \label{subsec:heterogenous}
@@
 The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them.
-Heterogeneity is allowed by replacing the shared $\theta$ parameter vector with $N$ individual parameter vectors $\theta_i$ so that
+To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogenous (IH) model} and a dependent \textit{clustered heterogeneous} (CH) model.
+In both cases, the shared Heterogeneity is allowed by replacing the shared $\theta$ parameter vector is replaced with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})$ corresponding to the parameters given in the model for car $i$ given in
 \begin{align}
@@
 \end{align}
-Use of the prior (\ref{indPrior}) for each $\theta_i$ formulates what is referred to as the independent model; where each individual $\theta_i$ posterior is independent of all other, i.e.
-$p(\theta_i | \textbf{z}_{1:N}, \theta_{j \neq i}) = p(\theta_i | z_i)$.
+with $\epsilon_{i,t}$ and $\eta_{i,t}$ all independent standard normal random variables.
+For the IH model, each $\theta_i,$ for $i=1,2,\ldots, N$, is assumed to follow an independent $N(\mu, \Sigma)$ prior distribution, resulting in a joint posterior distribution that may be decomposed into $N$ distinct independent marginal distributions, i.e.
+\begin{equation}
+p(\theta_{1:N} \mid z_{1:N}) = \prod_{i=1}^N p(\theta_{i} \mid z_{i}),
+\end{equation}
+due to the distinct parameters and independent (albeit identical) prior distributions for each of the $N$ models. \footnote{Recall that $z_{i}=\{x^*_{i,s},y^*_{i,s},\mbox{ for } s=1,2,...,T \}$.}
 \\
 
-An alternative approach is to augment the prior distribution of each $\theta_i$ with parameters $\beta$ in a hierarchical model, and infer the distribution of $p(\theta_{1:N}, \beta | \textbf{z}_{1:N})$.
-The top panel of Figure \ref{fig:HierSingleKDE} plots a kernel density estimate of the posterior means of each $\theta_i, i = 1, \dots, 2000$, using the independent model with $p = q = 2$ at time $T = 500$.
-There is a wide range of values for the posterior mean of each variable, indicating heterogeneity across the drivers in the sample.
-The density of posterior means for both variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the density of posterior means of both $\phi_1$ and $\phi_2$ are multimodal.
-Due these features a $K$ component mixture of multivariate normal distributions is chosen for $p(\theta_i | \beta)$, augmenting each driver with an auxiliary variable $k_i = 1, \dots, K$ such that
+In contrast, the alternative CH model does not assume \textit{a priori} independence of each of the $\theta_i$ vectors and instead clusters individual $\theta_i$ values into one of $K$ common latent multivariate normal distributions as determined by $\beta.$ The hierarchical CH model assumes that $\beta$ is itself random, with $\beta \sim p(\beta)$.
+By augmenting each driver's individual parameter $\theta_i$ with the corresponding $k_i=j$ variable, where $j$ can take on any value $1,2 \ldots, K$, we have
 \begin{equation}
@@
 \begin{equation}
-k_i \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
+k_i \mid \beta \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
 \end{equation}
-for each $i = 1, \dots, N$. Furthurmore, for each $j = 1, \dots, K$,  
+for each $i = 1, \dots, N$.
+The mixture indicator variables satisfying $k_i=j$ determine a cluster of similar $\theta_i$ parameters corresponding to the behaviours of a subset of drivers.
+$\beta$ contains all of the determinants of these distributions, with $\beta = \{\mu_j,\Sigma_j, \pi_j, for j=1,\},$ with \textit{a priori} assumptions for each cluster given by
 \begin{align}
@@
 Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$.
-The bottom panel of Figure \ref{fig:HierSingleKDE} plots the marginal distributions for $p(\theta_{2001} | \beta, z_{1:2000})$ for the hierarchical model with $p = q = 2$ and $K = 6$ using the same data as the independent model.
-The hierarchical marginal distributions have captured the skewness and multimodality present in the densities in the top panel.
 
+To illustrate that the hierarchical CH model, with $K=6$ adequately captures the main features of the IH model, the top panel of Figure \ref{fig:HierSingleKDE} displays kernel density estimates of the posterior means, for each $\theta_i, i = 1, \dots, 2000$, resulting from the IH model.
+For this illustration, we set $p = q = 2$ and use $N=2000$ observations over times $t=1, 2, \ldots, T = 500$.
+These summaries each indicate a wide range of values for the $N$ posterior means corresponding to each model parameter, corresponding to heterogeneity across drivers in the sample.
+Notably, the posterior means for each of the variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the distribution of posterior means is multimodal for each of the $\phi_1$ and $\phi_2$ parameters.
+
+These summaries are compared against the figures in the bottom panel of Figure \ref{fig:HierSingleKDE}, where the estimated marginal distributions for $p(\theta_{2001} | \textbf{z}_{1:2000})$are displayed.
+The CH marginal posterior distributions appear to have captured the skewness and multimodality present in the corresponding IH posterior mean summaries shown in the top panel.
 \begin{figure}[ht]
@@
 \includegraphics[width = 0.95\textwidth]{HierSingleKDE.png}
-\caption{Top: A Kernel Density Estimate applied to $E(\theta_i | z_i), i = 1, \dots, 2000$ where $p = q = 2$ from the independent model, sampled by Metropolis-Hasting MCMC.
-Bottom: $p(\theta_i | \beta, z_{1:2000})$ for the hierarchical model with $K = 6$ using the same vehicles sampled by Metropolis-Hasting MCMC.
+\caption{Top: A Kernel Density Estimate applied to $E(\theta_i | z_i), i = 1, \dots, 2000$ where $p = q = 2$ from the IH model, sampled by Metropolis-Hasting MCMC.
+Bottom: $p(\theta_{2001}| | \textbf{z}_{1:2000})$ for the CH model with $K = 6$ using the same vehicles sampled by Metropolis-Hasting MCMC.
 Each individual component in the mixture is denoted by the coloured densities.
-The hierarchical model has captured the multimodality and skewness present in the independent posterior means.}
+The CH model has captured the multimodality and skewness present in the IH posterior means.}
 \label{fig:HierSingleKDE}
@@
 
-For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refers to the data collected from, and the associated parameter vectors for, the $N$ vehicles that are available before the self driving vehicle is on the road, while the subscript $i$ is reserved for some $i > N$ and refers to the data and parameter vector for an additional vehicle in the test set, that is encountered while the self driving vehicle is on the road.
-Inference about $\theta_{1:N}$, the individual parameter vectors for the independent and hierarchical models, as well as for $\theta$, the shared homogenous parameter vector and $\beta$, the hierarchical hyperparameters, conditional on $\textbf{z}_{1:N}$ are assumed to be available with an unlimited computational budget.
+For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles that are available.
+We conceive of such information begin available before any self-driving vehicle enters a lone on the road.
+Use of the subscript $i$ is reserved for some $i > N$, and refers to the attriubtes (data or a parameter vector) for an additional behicle that may be encountered while the self-driving vehicle is on the road.
+Inference about $\theta_{1:N}$, the individual parameter vectors for the IH and CH models, as well as for $\theta$, the shared parameter vector under the homogeneous model, and $\beta$, the CH hyper-parameter, conditional on $z_{1:N}$ are assumed to be available.
 \\
 
-A significant difference in the application of the time series models is through the relationship between the posterior distribution formed from observations of the first $N$ vehicles and the parameter vector for the additional vehicle, $\theta_{i}$.
-Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} = \theta$.
-In this case $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not be significantly different to $p(\theta | \textbf{z}_{1:N})$ due to the high precision of the posterior distribution for a sufficiently large $N$.
-Figure (\ref{fig:homogPosterior}) plots the posterior marginal distributions from $N = 2000$ vehicles highlighting the posterior precision.
+The idea we pursue is to use the available data, and corresponding model posterior distributions, to forecast the future behaviour of a vehicle that is positioned near a given self-driving vehicle.
+Leveraging off of the information gathered from previous data (via previously computed model posteriors that accommodate the range of heterogeneous behaviour of the population of vehicles) together with fresh information received in real-time, we aim to predict the near future position of an additional vehicle moving contemporaneously with the self-driving one
+\\
+
+A crucial aspect of the application of the discussed time series models is through the relationship between the posterior distribution formed from observations relatting to the first $N$ vehicles and the parameter vector for the additional new vehicle, $\theta_{N+1}$.
+Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} \equiv \theta$.
+In this case, if $N$ is large $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not vary much from $p(\theta | \textbf{z}_{1:N})$.
+Figure (\ref{fig:homogPosterior}) displays plots of the marginal posterior distributions associated with each element of $\theta$, with $p = q = 2$ resulting from $N = 2000$ vehicles.
+Arguably each of these marginal posterior distributions are very precise, and are unlikely to change much if updated using information from observations from the $(N+1)^{st}$ vehicle.
 
@@
 \includegraphics[width = 0.95\textwidth]{homogPosterior}
-\caption{The converged Metropolis Hastings MCMC samples of the marginal posterior distributions $p(\theta | \textbf{z}_{1:N})$ for the homogenous model with $N = 2000$ and $p = q = 2$.
-Each variable has very high posterior precision.}
+\caption{The marginal posterior distributions $p(\theta | \textbf{z}_{1:N})$ for the homogenous model with $N = 2000$ and $p = q = 2$. Each variable has very high posterior precision.}
 \label{fig:homogPosterior}
@@
 
-The independent model implies that the posterior distribution for $\theta_i$ is independent of $\theta_{1:N}$ and $\textbf{z}_{1:N}$ by
+The IH model implies that the posterior distribution for $\theta_i$ is independent of $\theta_{1:N}$ and $\textbf{z}_{1:N}$ by
 \begin{equation}
@@
 
-The posterior for $\theta_i$ using the hierarchical model, assuming conditional independence between each $\theta_j$ for $j = 1, \dots, N$ and $\theta_i$, given $\beta$, $\textbf{z}_{1:N}$, and $z_i$, is
+The posterior for $\theta_i$ using the CH model, assuming conditional independence between each $\theta_j$ for $j = 1, \dots, N$ and $\theta_i$, given $\beta$, $\textbf{z}_{1:N}$, and $z_i$, is
 \begin{equation}
@@
 
-Combining a fast inference procedure with the test vechile posterior distributions implied by the independent and hierarchical models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self driving vehicle observes data for the vehicles it encounters.
+Combining a fast inference procedure with the test vehicle posterior distributions implied by the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.
 
@@
 
-Given a series of data observed up to time $T$, $z_{1:T}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
+Given a series of data for vehicle $i$ observed up to time $T$, $z_{i, 1:T}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
 \begin{equation}
 \label{predictive}
-p(z_{T+h} | z_{1:T}) =\int p(z_{T+h}|z_{1:T}, \theta) p(\theta | z_{1:T}) d\theta.
+p(z_{i, T+h} | z_{i, 1:T}) =\int p(z_{i, T+h}|z_{i, 1:T}, \theta_i) p(\theta_i | \textbf{z}_{1:N}, z_{i, 1:T}) d\theta.
 \end{equation}
-To obtain this distribution, the posterior density for $\theta$, given by
-\begin{equation}
-\label{posterior}
- p(\theta | z_{1:T}) = \frac{p(z_{1:T}|\theta)p(\theta)}{\int p(z_{1:T}|\theta)p(\theta) d\theta}
-\end{equation}
-must first be inferred. Generally the analytical solution to (\ref{posterior}), and hence to (\ref{predictive}), will be unavailable due to the complex functional form of the distributions involved. 
+To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogenous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewcar2}) for the CH model must first be inferred.
+However these are only known up to proportionallity, and hence the analytical solution to (\ref{predictive} is unavailable.
 \\
 
-In this section two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
-In brief, MCMC is used to create a sample from $p(\theta | z_{1:T})$, with any function of $\theta$ that is desired estimated from that sample.
-In contrast, VB replaces $p(\theta | z_{1:T})$ with a parametric approximation, denoted by $q_{\lambda}(\theta | z_{1:T})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $z_{1:T}$.
+Collectively denoting all available data, for either $\textbf{z}_{1:N}$ or $z_{i}$, as $\textbf{z}$ for the remainder of this section, two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
+In brief, MCMC is used to create a sample from $p(\theta_i | \textbf{z})$, with any function of $\theta_i$ that is desired estimated from that sample.
+In contrast, VB replaces $p(\theta_i | \textbf{z})$ with a parametric approximation, denoted by $q_{\lambda}(\theta_i |\textbf{z})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $\textbf{z}$.
 
@@
 There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler.
-The Gibbs sampler algorithm iteratively samples the components of the $k-$dimensional parameter vector $\theta$ via each of the so-called full conditional distributions as follows,
+The Gibbs sampler algorithm iteratively samples the components of a $k$dimensional parameter vector $\theta_i$ via each of the so-called full conditional distributions as follows,
 \begin{align}
-&p(\theta_1 | \theta_2, \dots, \theta_k, z) \nonumber \\
-&p(\theta_2 | \theta_1, \theta_3, \dots, \theta_k, z) \nonumber \\
+&p(\theta_{i, 1} | \theta_{i, 2}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
+&p(\theta_{i, 2} | \theta_{i, 1}, \theta_{i, 3}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
 &\vdots \nonumber \\
-&p(\theta_k | \theta_1, \dots, \theta_{k-1}, z). \nonumber
+&p(\theta_{i, k} | \theta_{i, 1}, \dots, \theta_{o, k-1}, \textbf{z}). \nonumber
 \end{align}
-Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\theta | z)$.
+Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\theta_i | \textbf{z})$.
 Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
@@
 
-In MH-MCMC, a candidate $\theta_j^{(c)}$, where $\theta_j$ may be any scalar or vector subset of $\theta$, is drawn from a proposal distribution $r(\theta_j)$ and accepted by the sampler with probability
+In MH-MCMC, a candidate $\theta_{i, j}^{(c)}$, where $\theta_{i, j}$ may be any scalar or vector subset of $\theta_i$, is drawn from a proposal distribution $r(\theta_{i, j})$ and accepted by the sampler with probability
 \begin{equation}
-\min \left\{ 1, \frac{p(\theta_j^{(c)} | \theta_{i \neq j}^{(a)}, z)}{p(\theta_j^{(a)} | \theta_{i \neq j}^{(a)}, z)} \times \frac{r(\theta_j^{(a)} | \theta_j^{(c)})}{r(\theta_j^{(c)} | \theta_j^{(a)})} \right\},
+\min \left\{ 1, \frac{p(\theta_{i, j}^{(c)} | \theta_{i, l \neq j}^{(a)}, \textbf{z})}{p(\theta_{i, j}^{(a)} | \theta_{i, l \neq j}^{(a)}, \textbf{z})} \times \frac{r(\theta_{i, j}^{(a)} | \theta_{i, j}^{(c)})}{r(\theta_{i, j}^{(c)} | \theta_{i, j}^{(a)})} \right\},
 \label{MHaccept}
 \end{equation}
-where the superscript $(a)$ denotes the most recently accepted value of the corresponding subset of $\theta$.
-If the sampler rejects a candidate value the previous value $\theta_j^{(a)}$ is repeated.
-Each iteration of Metropolis-Hastings-within-Gibbs MCMC includes one candidate draw for each element of $\theta$.
+where the superscript $(a)$ denotes the most recently accepted value of the corresponding subset of $\theta_i$.
+If the sampler rejects a candidate value the previous value $\theta_{i, j}^{(a)}$ is repeated.
+Each iteration of Metropolis-Hastings-within-Gibbs MCMC includes one candidate draw for each element of $\theta_i$.
 Choice of proposal distribution is left to the user with arguably the most simple being the Normal Random Walk proposal,
 \begin{equation}
-\theta_j^{(c)} \sim N(\theta_j^{(a)}, \Sigma_j)
+\theta_{i, j}^{(c)} \sim N(\theta_{i, j}^{(a)}, \Sigma_{i, j})
 \label{RWprop}
 \end{equation}
-as $r(\theta_j^{(c)} | \theta_j^{(a)}) = r(\theta_j^{(a)} | \theta_j^{(c)})$.
+as $r(\theta_{i, j}^{(c)} | \theta_{i, j}^{(a)}) = r(\theta_{i, j}^{(a)} | \theta_{i, j}^{(c)})$.
 \\
 
-Performance of Random Walk Metropolis Hastings MCMC (RWMH-MCMC) then depends on the acceptance rate of the proposal distribution, largely influenced by the value of $\Sigma_j$.
-\citet{Garthwaite2016} propose an adaptive algorithm that increases or decreases $\Sigma_j$ depending on whether the most recent candidate was rejected or accepted respectively so that a target acceptance rate is asymptotically approached as the number of MCMC iterations approaches infinity.
+Performance of Random Walk Metropolis Hastings MCMC (RWMH-MCMC) then depends on the acceptance rate of the proposal distribution, largely influenced by the value of $\Sigma_{i, j}$.
+\citet{Garthwaite2016} propose an adaptive algorithm that increases or decreases $\Sigma_{i, j}$ depending on whether the most recent candidate was rejected or accepted respectively so that a target acceptance rate is asymptotically approached as the number of MCMC iterations approaches infinity.
 
@@
 A typically faster, albeit approximate, alternative to MCMC based inference is VB \citep{Jordan1999}.
-VB posits a family of parametric approximating distributions $q_{\lambda}(\theta | z)$, parameterised by an auxiliary vector $\lambda$, that share the same support as the true posterior distribution $p(\theta | z)$.
-Note that the family $q_{\lambda}(\theta | z)$ does not depend on $z$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta | z)$.
-A member of approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta | z)$ to $p(\theta | z)$, given by $KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})]$ \citep{Kullback1951}.
+VB posits a family of parametric approximating distributions $q_{\lambda}(\theta_i | \textbf{z})$, parameterised by an auxiliary vector $\lambda$, that share the same support as the true posterior distribution $p(\theta_i | \textbf{z})$.
+Note that the family $q_{\lambda}(\theta_i | \textbf{z})$ does not neccesarily depend on $\textbf{z}$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta_i | \textbf{z})$.
+A member of approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta_i | \textbf{z})$ to $p(\theta_i | \textbf{z})$, given by $KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z})]$ \citep{Kullback1951}.
 The KL divergence is defined by
@@
 \label{KL-def}
-KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})] = E_{q_{\lambda}(\theta | z)} \left[ \log(q_{\lambda}(\theta | z)) - \log(p(\theta | z)) \right],
+KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z})] = E_{q_{\lambda}(\theta_i | \textbf{z})} \left[ \log(q_{\lambda}(\theta_i | \textbf{z})) - \log(p(\theta_i | \textbf{z})) \right],
 \end{equation}
-and is a non-negative, asymmetric measure of the discrepancy between $p(\theta | z)$ and $q_{\lambda}(\theta | z)$  that will be equal to zero if and only if $p(\theta | z) = q_{\lambda}(\theta | z)$ almost everywhere \citep{Bishop2006}.
+and is a non-negative, asymmetric measure of the discrepancy between $p(\theta_i | \textbf{z})$ and $q_{\lambda}(\theta_i | \textbf{z})$  that will be equal to zero if and only if $p(\theta_i | \textbf{z}) = q_{\lambda}(\theta_i | \textbf{z})$ almost everywhere \citep{Bishop2006}.
 \\
@@
 \label{KL-MC}
-KL[q_{\lambda}(\theta | z)\hspace{.1cm}||\hspace{.1cm}p(\theta | z_{1:T})] \approx \frac{1}{M}\sum_{i=1}^M \left(\log(q_{\lambda}(\theta_i |z)) - \log(p(\theta | z)) \right)
+KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z}] \approx \frac{1}{M}\sum_{j=1}^M \left(\log(q_{\lambda}(\theta_{j, i} | \textbf{z})) - \log(p(\theta_{j, i} | \textbf{z})) \right)
 \end{equation}
-where $\theta_i \sim q_{\lambda}(\theta |z))$ are compuationally infeasible due to the inclusion of the term $p(\theta | z)$, which is only known up to proportionality.
+where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z}))$ are compuationally infeasible due to the inclusion of the term $p(\theta_i | \textbf{z})$, which is only known up to proportionality.
 Instead VB uses the Evidence Lower Bound (ELBO), denoted by $\mathcal{L}(q, \lambda)$, as an error function where
@@
 \label{ELBO}
-\mathcal{L}(q, \lambda) = E_{q_{\lambda}(\theta |z)} \left[\log(p(\theta, z)) - \log(q_{\lambda}(\theta |z))\right],
+\mathcal{L}(q, \lambda) = E_{q_{\lambda}(\theta_i | \textbf{z})} \left[\log(p(\theta_i, \textbf{z})) - \log(q_{\lambda}(\theta_i | \textbf{z}))\right],
 \end{equation}
@@
 \label{ELBO-MC}
-\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{i=1}^M \left(\log(p(\theta_i, z)) - \log(q_{\lambda}(\theta_i |z)) \right)
+\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right)
 \end{equation}
-where $\theta_i \sim q_{\lambda}(\theta |z))$.
-The ELBO is equal to the negative KL divergence plus a constant, and hence maximising (\ref{ELBO}) with respect to $q_{\lambda}(\theta | z)$ is equivalent to minimising (\ref{KL-def}).
+where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z}))$.
+The ELBO is equal to the negative KL divergence plus a constant, and hence maximising (\ref{ELBO}) with respect to $q_{\lambda}(\theta_i | \textbf{z})$ is equivalent to minimising (\ref{KL-def}).
 
@@
 \label{scoreDeriv}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_j | z))}{\delta \lambda} \left(\log(p(\theta_j, z)) - \log(q_{\lambda}(\theta_j | z)) \right),
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{j, i} | \textbf{z}))}{\delta \lambda} \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right),
 \end{equation}
-where $\theta_j \sim q_{\lambda}(\theta | z)$ and the reparameteterised estimator of \citet{Kingma2014}.
-Reparameterisation introduces an auxiliary variable $\epsilon$ and differentiable function $f(\cdot, \cdot)$ to rephrase Variational Bayes optimisation as the equivalent search for the parameters $\lambda$ that minimises the Kullback Leibler divergence from some distribution $q(\epsilon)$ with zero free parameters to the posterior distribution implied by the transformation $\theta = f(\epsilon, \lambda)$:
+where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z})$ and the reparameteterised estimator of \citet{Kingma2014}.
+Reparameterisation introduces an auxiliary variable $\epsilon$ and differentiable function $f(\cdot, \cdot)$ to rephrase Variational Bayes optimisation as the equivalent search for the parameters $\lambda$ that minimises the Kullback Leibler divergence from some distribution $q(\epsilon)$ with zero free parameters to the posterior distribution implied by the transformation $\theta_i = f(\epsilon, \lambda)$:
 \begin{equation}
 \label{rpDist}
-p(f(\epsilon, \lambda) | z) = p(\theta | z) |J^{-1}(f(\epsilon, \lambda))|
+p(f(\epsilon, \lambda) | z) = p(\theta_i | \textbf{z}) |J^{-1}(f(\epsilon, \lambda))|
 \end{equation}
 where $J(f(\epsilon, \lambda))$ is the Jacobian Matrix of the transformation $f(\epsilon, \lambda)$.
-Examples of $f$ and $q(\epsilon)$ include treating $\theta$ as location scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
+Examples of $f$ and $q(\epsilon)$ include treating $\theta_i$ as location scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
 \\
 
-The ELBO can be reparameterised by substituting $p(\theta, z) = p(f(\epsilon, \lambda), z)|J(f(\epsilon, \lambda))|$ into (\ref{ELBO}),
+The ELBO can be reparameterised by substituting $p(\theta_i, \textbf{z}) = p(f(\epsilon, \lambda), z)|J(f(\epsilon, \lambda))|$ into (\ref{ELBO}),
 \begin{equation}
 \label{rpELBO}
-\mathcal{L}(q, \lambda) = E_{r(\epsilon)} \bigg[\log(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|) - \log(q(\epsilon))\bigg].
+\mathcal{L}(q, \lambda) = E_{r(\epsilon)} \bigg[\log(p(f(\epsilon,\lambda), \textbf{z})|J(f(\epsilon, \lambda))|) - \log(q(\epsilon))\bigg].
 \end{equation}
@@
 \label{rpELBODeriv}
-\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} &= \frac{\delta}{\delta \lambda} \bigg( E_{q(\epsilon)} \bigg[\log\big(p(f(\epsilon,\lambda), z)|J(f(\epsilon, \lambda))|\big) - \log(q(\epsilon))\bigg] \bigg) \nonumber \\
-&= E_{q(\epsilon)} \left[ \frac{\delta}{\delta \lambda} \bigg(\log(p(f(\epsilon,\lambda), z)) + \log(|J(f(\epsilon, \lambda))|) - \log(q(\epsilon)) \bigg)\right] \nonumber \\
-&= E_{q(\epsilon)} \left[ \frac{\delta \log(p(f(\epsilon,\lambda), z))}{\delta f(\epsilon,\lambda)} \frac{\delta f(\epsilon,\lambda)}{\delta \lambda}  + \frac{\delta \log(|J(f(\epsilon, \lambda))|)}{\delta \lambda} \right].
+\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} &= \frac{\delta}{\delta \lambda} \bigg( E_{q(\epsilon)} \bigg[\log\big(p(f(\epsilon,\lambda), \textbf{z})|J(f(\epsilon, \lambda))|\big) - \log(q(\epsilon))\bigg] \bigg) \nonumber \\
+&= E_{q(\epsilon)} \left[ \frac{\delta}{\delta \lambda} \bigg(\log(p(f(\epsilon,\lambda), \textbf{z})) + \log(|J(f(\epsilon, \lambda))|) - \log(q(\epsilon)) \bigg)\right] \nonumber \\
+&= E_{q(\epsilon)} \left[ \frac{\delta \log(p(f(\epsilon,\lambda), \textbf{z}))}{\delta f(\epsilon,\lambda)} \frac{\delta f(\epsilon,\lambda)}{\delta \lambda}  + \frac{\delta \log(|J(f(\epsilon, \lambda))|)}{\delta \lambda} \right].
 \end{align}
@@
 \label{rpDeriv}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta f(\lambda, \epsilon_j)}{\delta \lambda} \frac{\delta \log(p(\theta, z))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda},
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta f(\lambda, \epsilon_j)}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda},
 \end{equation}
@@
 The reparameterised gradient estimator typically has lower variance than the score estimator (see eg.
-\cite{Rezende2014}; \cite{Ruiz2016}), but treating $q_{\lambda}(\theta | z)$ as the distribution implied by the transformation $f$ of $q(\epsilon)$ restricts the class of approximating families that can be used.
+\cite{Rezende2014}; \cite{Ruiz2016}), but treating $q_{\lambda}(\theta_i | \textbf{z})$ as the distribution implied by the transformation $f$ of $q(\epsilon)$ restricts the class of approximating families that can be used.
 \\
@@
 To reduce the variance of the Monte Carlo estimator, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) which has shown to be more efficient, where numbers in the unit hypercube are generated according to a Sobol Sequence \citep{Sobol1967} which are then randomised using the scrambled net method \citep{Matousek1998}, and transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$.
-Use of RQMC has shown to be more efficeint than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
+Use of RQMC has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
 
@@
 
-The self driving vehicle is constantly observing data about the movements of the surrounding vehicles which can be used for posterior inference.
-After observing vehicle $i$ until time $S$, $z_{i, 1:S}$ has been observed and can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the conditional dependence on $\textbf{z}_{1:N}$.
-At time $T > S$ an additional $T - S$ data points from vehicle $i$ have been observed, and forecasts could be improved by incorporating these to form the posterior approximation, $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$, where the $S$ and $T$ subscripts on $\lambda$ are included to differentiate the auxiliary parameter vector conditioned on data up to times $S$ and $T$.
+The self-driving vehicle is constantly observing data about the movements of the surrounding vehicles which can be used for posterior inference.
+Figure \ref{fig:timeUpdate} illustrates this for a vehicle in the NGSIM dataset travelling towards the right.
+In the left panel, the vehicle has been observed until time $S$, and this data $z_{i, 1:S}$ can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the additional conditional dependence on $\textbf{z}_{1:N}$.
+In the right panel, at some later time $T > S$, an additional $T - S$ data points have been observed, and inference of the behaviour of this vehicle, and hence forecasts, could be improved by incorporating the information contained in $z_{i, S+1:T}$ to form the posterior approximation $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$.
+Note that the $S$ and $T$ subscripts on $\lambda$ are introduced to differentiate the auxiliary parameter vector conditioned on data up to times $S$ and $T$.
 To facilitate this posterior update, an Updating Variational Bayes (UVB) mechanism is introduced where only $z_{i, S+1:T}$ needs to be processed.
+\begin{figure}[ht]
+\centering
+\includegraphics[width = 0.95\textwidth]{timeUpdate}
+\caption{Left: The path of a vehicle in the NGSIM dataset at time $S$, where the direction of travel is from the left to the right.
+Right: The same vehicle at a later time $T > S$, with the extra $T - S$ observations denoted by the dashed line.
+Inference about this vehicle at time $S$ can be updated by the inclusion of the additional information observed up to time $T$.}
+\label{fig:timeUpdate}
+\end{figure}
 \\
@@
 where $\epsilon_j \sim r(\epsilon)$.
-If this process is repeated periodically, $S$ will grow at the same rate as $T$, and updating VB is $O(1)$ rather than $O(T)$ as required by re-using the full sample without the approximation stepin (\ref{ApproxJoint}).
+If this process is repeated periodically, $S$ will grow at the same rate as $T$, and updating VB is $O(1)$ rather than $O(T)$ as required by re-using the full sample without the approximation step in (\ref{ApproxJoint}).
 
@@
 
-After obtaining the posterior distributions for each model conditioned on 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on a further 1373 vehicles.
-For each of these vehicles $i > 2000$, at each point in time $T = 100, 110, \dots, 450$ and forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model, while the time-series models are used to provide forecast densities of $p(z_{i, T+h} | z_{1:2000}, z_{i, 1:T})$ and point estimates from the MAP of this density.
+After obtaining the posterior distributions for each model conditioned on 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on a further 1387 vehicles.
+For each of these vehicles $i > 2000$, at each point in time $T = 100, 110, \dots, 450$ and forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model, while the time-series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$ and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density.
 As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement after every second of observing the vehicle.
-The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | z_{1:2000}, z_{i, T})$, while the independent and hierarchical models produce $p(\theta_{i} | z_{1:2000}, z_{i, 1:T})$ through three methods:
+The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ through three methods:
 \begin{enumerate}
 \item Posterior sampling with RWMH-MCMC,
-\item VB fitting $q_{\lambda_T}(\theta_{i} | z_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$ using the original prior distribution,
-\item UVB fitting $q_{\lambda_T}(\theta_{i} | z_{1:2000}, z_{i, 1:T})$ using $q_{\lambda_S}(\theta_{i} | z_{1:2000}, z_{i, 1:S})$ and data $z_{i, S+1:T}$ as described in Section \ref{sec:UVB}.
+\item VB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$ using the original prior distribution,
+\item UVB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ using $q_{\lambda_S}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:S})$ and data $z_{i, S+1:T}$ as described in Section \ref{sec:UVB}.
 \end{enumerate}
-A summary of these approaches is provided in Table \ref{tableAlg}. Point estimate forecasts are evaluated by their Euclidean Error,
+Approximating distributions for VB are chosen to match the form of the prior distributions, multivariate normal $q$ for the IH model and a mixture of multivariate normals for the CH model.
+A summary of these approaches is provided in Table \ref{tableAlg}.
+Point estimate forecasts are evaluated by their Euclidean Error,
 \begin{equation}
@@
 \begin{equation}
-\mbox{LS}_{i, T, h} = \log \left(p\left(x^*_{i, T+h}, y^*_{i, T+h} | z_{1:2000}, z_{i, 1:T} \right) \right)
+\mbox{LS}_{i, T, h} = \log \left(p\left(x^*_{i, T+h}, y^*_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T} \right) \right).
 \label{logscore}
@@
 \hline
-& Independent Model & Hierarchical Model \\
+& Independent Heterogenous Model & Clustered Heterogenous Model \\
 \hline
@@
 \hline
-VB  & Variational Bayes using (\ref{rpUpdate}) to update parameters & Variational Bayes using (\ref{scoreUpdate}) to update parameters\\
-Standard &  of a Multivariate Gaussian approximation with &  of an approximation formed as a six component mixture\\
+VB  & Variational Bayes fit to $100$ observations using (\ref{rpDeriv}), & Variational Bayes fit to $100$ observations using (\ref{scoreDeriv}), \\
+ &  then updated every $10$ observations using (\ref{rpUpdate}). & then udpated every $10$ observations using using (\ref{scoreUpdate}). \\
+Standard &  $q$ is a Multivariate Gaussian approximation with &  $q$ is a six component mixture\\
 &non-zero covariance using $M = 25$, where $f$ is a&  of diagonal variance Multivariate Gaussians using $M = 50$. \\
@@
 
-Nine naive forecast models are provided where forecasts of $a_t$ and $\delta_t$ are constants equal to $\hat{a}$ and $\hat{\delta}$ for $t = T+1, \dots, T+H$.
-The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}$ and one estimator for $\hat{\delta}$ at time $T$.
-Acceleration estimators are the most recent acceleration, $\hat{a} = a_T$, the average acceleration over the most recent second, $\hat{a} = 0.1 \sum_{s = T-9}^T a_s$, and zero acceleration $\hat{a} = 0$, equivalent to constant velocity.
-Similarly, steering angle estimators are the most recent angle, $\hat{\delta} = \delta_T$, the average angle over the most recent second, $\hat{\delta} = 0.1 \sum_{s = T - 9}^T \delta_s$, and $\hat{\delta} = \pi/2$, corresponding to the vehicle to driving directly forward.
+Nine naive forecast models are provided where forecasts of $a_{i, t}$ and $\delta_{i, t}$ at time $T$ are constants equal to $\hat{a}_{i, T}$ and $\hat{\delta}_{i, T}$ for $t = T+1, \dots, T+H$.
+The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}_{i, T}$ and one estimator for $\hat{\delta}_{i, T}$.
+Acceleration estimators are the most recent acceleration, $\hat{a}_{i, T} = a_{i, T}$, the average acceleration over the most recent second, $\hat{a}_{i, T} = 0.1 \sum_{s = T-9}^T a_{i, s}$, and zero acceleration $\hat{a}_{i, T} = 0$, equivalent to constant velocity.
+Similarly, steering angle estimators are the most recent angle, $\hat{\delta}_{i, T} = \delta_{i, T}$, the average angle over the most recent second, $\hat{\delta}_{i, T} = 0.1 \sum_{s = T - 9}^T \delta_{i, s}$, and $\hat{\delta}_{i, T} = \pi/2$, corresponding to the vehicle to driving directly forward.
 \begin{table}
@@
 \hline
-& $\hat{\delta}_t = \delta_T$ & $\hat{\delta}_t = 1/10 \sum_{s=T-9}^T \delta_s$ & $\hat{\delta}_t = \pi/2$ \\
+& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 1/10 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
 \hline
-$\hat{a}_t = a_T$ & Naive 1 & Naive 2 & Naive 3 \\
-$\hat{a}_t = 1/10\sum_{s=T-9}^T a_s$ & Naive 4 & Naive 5 & Naive 6\\
-$\hat{a}_t = 0$ & Naive 7 & Naive 8 & Naive 9 \\
+$\hat{a}_{i, T} = a_{i, T}$ & Naive 1 & Naive 2 & Naive 3 \\
+$\hat{a}_{i, T} = 1/10\sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
+$\hat{a}_{i, T} = 0$ & Naive 7 & Naive 8 & Naive 9 \\
 \hline
@@
 
-Mean Eucliedean Error for each model at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
-Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_t$ the choice of $\hat{\delta}_t$ has not substantially changed the forecast error.
+Eucliedean Error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
+Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error.
 Zero acceleration naive models (Naive 7, 8, and 9) perform the best, followed by the averaged acceleration models (Naive 4, 5, and 6) and finally the current acceleration models (Naive 1, 2, and 3).
@@
 
-Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, and with the exception of the independent model with UVB inference, each time series model and inference implementation produces a forecast with a similar amount of error.
-This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the heirarchical model.
-As the independent model is the heterogenous model that provides the least accurate forecasts it is not discussed further.
+Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, and with the exception of the IH model with UVB inference, each time series model and inference implementation produces a forecast with a similar amount of error.
+This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models.
+As the UVB implementation of the CH model is competitive with MCMC and standard VB inference discussion will focus on the CH model rather than the IH model.
 \\
@@
 \includegraphics[width = 0.8\textwidth]{predictiveError.png}
-\caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | z_{i, 1:T})$ for $h \in \{10, 20, 30\}$ and $T \in \{100, 110, \dots, 450\}$ for each model.
+\caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$ for $h \in \{10, 20, 30\}$ and $T \in \{100, 110, \dots, 450\}$ for each model.
 Each time series model is significantly more accurate than the naive constant acceleration / angle models, with little discernable difference between the time series models.}
@@
 
-There are systematic differences in the level of certainty in forecasts provided by the heirarchical and homogenous models as measured by their logscores.
-The key differentiation between the models is the heirarchical models ability to assign each vehicle individual variance parameters,  $\sigma^2_{\epsilon, i}$ for $a_t$, and $\sigma^2_{\eta, i}$ for $\delta$.
-Figure (\ref{fig:posVarMean}) plots the range of posterior means for these parameters from with RWMH-MCMC sampling with $T = 450$, with the homogenous posterior means indicated with vertical red lines.
-There is a large amount of positive skewness in the distribution of variances, which may have caused the homogenous model to estimate a variance that is too large for most individual cars.
-The homogenous variance posterior means are greater than $80\%$ and $91.5\%$ of heirarchical variance posterior means for acceleration and the steering angle respectivley.
+There are systematic differences in the level of certainty in forecasts provided by the CH and homogenous models as measured by their logscores, resulting from the CH model being able to to assign each vehicle individual variance parameters,  $\sigma^2_{\epsilon, i}$ for $a_{i, t}$, and $\sigma^2_{\eta, i}$ for $\delta_{i, t}$.
+Figure (\ref{fig:posVarMean}) plots the range of posterior means for these parameters from with RWMH-MCMC sampling with $T = 450$, with the homogenous model posterior means indicated with vertical red lines.
+There is a large amount of positive skewness in the distribution of variances, which may have caused the homogenous model to estimate a variance that is too large for most individual vehicles.
+The homogenous variance posterior means are greater than $80\%$ and $91.5\%$ of CH variance posterior means for acceleration and the steering angle respectivley.
 This large variance has little impact on the location of the MAP of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density.
-Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the hierarchical model with updating VB inference over the homogenous model at the three second forecast horizon.
+Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the CH model with UVB inference over the homogenous model at the three second forecast horizon.
 Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis.
-The hierarchical model trajectory forecasts for drivers in the lower variance quintiles benefit the most from including heterogeneity.
+The CH model trajectory forecasts for drivers in the lower variance quintiles benefit the most from including heterogeneity.
 However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles.
@@
 \includegraphics[width = 0.95\textwidth]{posVarMean}
-\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by Metropolis-Hastings MCMC at $T = 450$ using the hierarchical model, compared to the homogenous posterior mean in red.
+\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by Metropolis-Hastings MCMC at $T = 450$ using the CH model, compared to the homogenous posterior mean in red.
 Small amounts of cars with high variance has skewed the homogenous estimates to be too large for the majority of the vehicles forecasted.}
@@
 \includegraphics[width = 0.95\textwidth, height = 0.4\textheight]{varSplit}
-\caption{Difference in logscore between the hierarchical model with posterior inference using UVB and the homogenous model.
+\caption{Difference in logscore between the CH model with posterior inference using UVB and the homogenous model.
 Results are split by variance quintiles for acceleration (x-axis) and angle (y-axis).
@@
 
-The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self driving vehicles.
+The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles.
 Figure (\ref{fig:timing}) demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$.
@@
 As UVB involves replacing the posterior at time $S$ with its variational approximation, the error introduced by UVB depends on the approximation error of Variational Bayes itself.
-The forecast Euclidean error increaseing when inferring the independent model posterior with UVB in Figure \ref{fig:PredErrorZ} compared to MCMC or VB is indicative of the multivariate Gaussian approximating distribution being an unsuitable replacement to the true posterior.
-The Hierarchical Model did not increase in error relative to other methods, possibly due to the increased flexibility of its six component mixture approximating distribution.
-Figure (\ref{fig:updateCost}) demonstrates that UVB did not decrease in forecast logscores for the hierarchical model, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
-Augmenting self driving vehicles with hierarchical time series models to forecast the trajectory of surrounding traffic, and repeatedly updating the posterior distribution and forecasts, can be feasibly implemented with UVB.
+The left panel for Figure (\ref{fig:updateCost}) displays boxplots of UVB logscores minus standard VB logscores at the three second forecast horizon for the CH model, and demonstrates that UVB does not typically lead to a decrease in the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
+The right panel displays analogous boxplots for the IH model, where the UVB logscores tend to be lower than the standard VB logscores, matching the increased Euclidean error found for the IH model with UVB inference in Figure \ref{fig:PredErrorZ}.
+An explanation for this is given by the form of the approximating distribution used for the IH model, the multivariate normal distribution may not adequately approximate the posterior distribution, while the extra flexibilitiy of the mixture multivariate normal distribution for the CH model was sufficient to replace the posterior.
+
 \begin{figure}[htp]
@@
 \includegraphics[width = 0.75\textwidth, height = 0.3\textheight]{updateCost}
-\caption{Differences in predictive logscores for $p(z_{i, T+h} | z_{i, 1:T})$ for each $h = 1, \dots, 30$ and $T = 100, 110, \dots, 450$ between updating and standard Variational Bayes.
-The differences are typically small, and UVB can improve the logscore.}
+\caption{Differences in predictive logscores for $p(z_{i, T+30} | z_{i, 1:T})$ for each $T = 100, 110, \dots, 450$ between UVB and standard VB for the CH model (left) and the IH model (right).
+The differences are typically small for the CH model, while the IH model approximation has increased the logscore relative to standard VB.
+This may be due to the form of the approximating distribution for the IH model being an inadequate approximation for the true posterior}
 \label{fig:updateCost}
@@
 
-\section{Conclusion}
-\label{sec:conc}
-\begin{itemize}
-\item The use of auto-regressive models for forecasting future acceleration and angles drastically improves forecasts of the future trajectory compared to the competing models.
-\item There is evidence of heterogeneity between vehicles, and allowing this with the heirarchical model allows behaviour specific to an individual driver to be inferred, improving logscores and thus reducing forecast uncertainty.
-\item Heterogeneity particularly benefits forecasts for low variance drivers, but also enables high variance drivers to be identified.
-\item Online forecasting with the heterogeneous model requires constant model fits and updates, computation of which can be simplified without a substantial performance hit by UVB.
-\item UVB is more broadly applicable to any kind of online inference for time series data.
-\end{itemize}
+\section{Discussion}
+\label{sec:disc}
+This paper demonstrates that the use of auto-regressive models time-series models drastically improves trajectory forecasts when compared to the competing models.
+These models parameterise the dynamics of a driver's actions, with a range of parameter posterior means providing evidence of heterogeneity between the drivers of different vehicles.
+Incorpororating this heterogeneity through a hierarchical model has improved the forecast logscores, and hence reduced uncertainty, but requires an approximate inference scheme implemented with UVB.
+\\
+
+We find that the additional approximation error introduced by UVB does not materialise as a reduction in forecasting performance relative to exact MCMC inference when an approximating distribution that has sufficient flexibilitity to approximate the true posterior distribution is used.
+Augmenting self-driving vehicles with hierarchical time series models to forecast the trajectory of surrounding traffic, and repeatedly updating the posterior distribution and forecasts, can be feasibly implemented with UVB.
+\\
+
+The framework used by the trajectory forecasting problem is to construct a hierarchical model to infer the range of heterogeneity present in a dataset, and use exact inference results provided by MCMC to construct a prior distribution for an additional unit in the hierarchy.
+Our use of this framework has focused on situations where exact inference on the additional unit is unavailble in the time-frame required by the problem, instead using Variational Bayes to infer the posterior distribution of this additional unit.
+We have introduced UVB to allow this inference to happen in the online setting, and have explored fitting copula models to the MCMC samples to produce a functional form for the approximating distribution that better matches the true posterior, though have not been able to link the copula structure for the units in the original hierarchy to the new unit.
+This framework is not specific to vehicle forecasting, and could conceivably be used for many other problems such as classification, or anomoly detection by extending the hierarchical model with a dirichlet process prior.
+\\
 
@@
 
-Excluding the vehicles used to estimate the spline model, each of the 3373 remaining vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
+Excluding the vehicles used to estimate the spline model, each of the 3387 remaining vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
 \begin{align}

### 2018-03-06 Cars/paper.tex
paper

@@
 
-Self-driving vehicles are rapidly becoming more advanced, with manufacturers such as Tesla, Ford, and Audi expecting autonomous vehicles available to consumers as early as 2020.
-These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery, and selects a path to avoid any collisions.
-As many as $94\%$ of accidents are due to human error according to data from the United States National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, which can feasibly be reduced by self-driving vehicles.
-However self-driving vehicles will operate in situations where surrounding vehicles are controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
+Self-driving vehicles are rapidly becoming more advanced, with manufacturers such as Tesla, Ford, and Audi expecting autonomous vehicles to be available to consumers as early as 2020.
+These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery, and selects a path to avoid collisions.
+With as many as $94\%$ of accidents in the United States resulting from human error, according to the US National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, the introduction of fully self-driving vehicles is expected to substantially improve safety on the roads.
+However self-driving vehicles will operate in situations where the majority of surrounding vehicles are still controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
 \\
 
-While characteristics of surrounding vehicles, such as theor postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g.
+While characteristics of surrounding vehicles, such as theirr postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g.
 \citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory has seen less attention in the literature.
@@
 This paper takes a statistical approach to this trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
-It uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability trajectories that can lead to a collision to be detected.
-The time series models used allow for heterogeneity between different vehicles, which may result from sources of driver heterogeneity that may result from factors such as differences in individual driving styles, or the conditions of traffic directly around them.
-Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
+It uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability trajectories that could cause a collision to be detected.
+The time series models used allow for heterogeneity between different vehicles, which may result from sources such as differences in individual driving styles, or the conditions of traffic.
+Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration, while forecasts for more consistent drivers will benefit from a low variance.
 \\
@@
 
-We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the contstant models commonly used, and including driver heterogeneity in the model substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous model.
-These results give credence to the validity of the methodological framework developed in this thesis, where exact Bayesian inference techniques are used to guide the construction on a prior distribution for parameters that are approximately inferred in an online setting.
+We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the contstant models commonly used, and including driver heterogeneity in these time series models substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous approach.
+These results give credence to the validity of the methodological framework developed in this thesis, where exact Bayesian inference techniques are used to guide the construction of a prior and approximating distribution for parameters that are inferred in an online setting with Variational Bayes.
 \\
@@
 This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used and steps required to extract acceleration and steering angle for each vehicle, while Section \ref{sec:models} introduces several time series models used.
-Section \ref{sec:Inference} reviews Bayesian inference, which is extended to the online updating setting in Section \ref{sec:UVB}.
+Section \ref{sec:Inference} reviews Bayesian inference, which is extended to the online, approximate updating setting in Section \ref{sec:UVB}.
 Finally Section \ref{sec:eval} contains empirical results, and Section \ref{sec:disc} concludes the vehicle trajectory forecasting study and discusses future work.
@@
 \label{sec:dataProcessing}
-Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005.
+Data is provided by the Next Generation Simulation (NGSIM) project conducted by the US Federal Highway Administration (FHWA), which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005.
 Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc to produce coordinates for each vehicle and point in time relative to the start of the road section.
 Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which traveled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1.
-Vehicles that entered or exited midway through the freeway section, are excluded in this paper.
+Vehicles that entered or exited midway through the freeway section are excluded in this paper.
 There is a curvature to the road occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.
@@
 \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway.
-Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
+Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the cente of the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
 \\
 
-Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to observation $t$ according to the trigonometric relationship in Figure \ref{fig:motion}:
+Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to observation $t$ follows friom the trigonometric relationship in Figure \ref{fig:motion}:
 \begin{align}
@@
 
-From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs from driver $i$ are calculated via
+From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs to motion from driver $i$ are calculated via
 \begin{align}
@@
        -\frac{\pi}{2} &\quad\text{if } y^*_{i, t} < y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
-       \delta_{i, t-1} &\quad\text{otherwise} \\ 
+       \delta_{i, t-1} &\quad\text{otherwise,} \\ 
      \end{cases} \label{dEq} \\
@@
 \end{align}
-In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=2,3,...,T\}$, according to
+In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=3,4,...,T\}$, according to
 \begin{equation}
@@
 Data resulting from the observation of vehicle $i$ for a period of time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$.
-3387 vehicles remain in the dataset after the above exclusion criteria are applied.
-The data into a two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1387 vehicles.
+Some additional vehicles are excluded due to an insufficient number of observations, leaving 3387 vehicles in the dataset in a relative coordinate system.
+These vehicles are split into two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1387 vehicles.
 \\
 
-The training set and the test set are treated as observations from vehicles at two distinct time periods.
-The training set represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles.
-The test set represents those vehicles the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.
+The training set and the test set are treated as observations from vehicles at two distinct time periods, illustrated in Figure \ref{fig:problem}.
+The training set in the left panel represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles.
+The test set in the right panel represents those vehicles the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.
 
@@
 
+\begin{figure}
+\centering
+\includegraphics[width = 0.95\textwidth]{problem}
+\caption{Left: The paths of vehicles in the training set.
+Right: The self-driving car (Solid) and two nearby vehicles (Outlined).
+All motion is from left to right.
+The self-driving vehicle must forecast the trajectory of the nearby vehicels based on their observed path and any knowledge of the driver behaviour learned from the training set vehicles}
+\label{fig:problem}
+\end{figure}
+
 \section{Models for Trajectory Forecasting}
@@
 
-Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between different vehicles.
+Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between the different sets of vehicles.
 These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent heterogenous (IH) model and the hierarchical, clustered heterogeneity (CH) model.
-The ID model does not incorporate information from the training set of vehicles into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set.
+The ID model does not incorporate information from the vehicles in the training into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set.
 \\
@@
 
-The framework provided by (\ref{xEq}) and (\ref{yEq}) allows distributional forecasts of future position coordinates of a given vehicle $i$, $\{x^*_{i, t}, y^*_{i, t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_{i, T}, y^*_{i, T}, v_{i, T}\}$, and distributional forecasts of the future values of $\{a_{i, t}, \delta_{i, t} | t = T + 1, \dots, T+H\}$.
+The framework provided by (\ref{xEq}) and (\ref{yEq}) allows distributional forecasts of future position coordinates of a given vehicle $i$, $\{x^*_{i, t}, y^*_{i, t} | t = T + 1, \dots, T+H\}$, to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_{i, T}, y^*_{i, T}, v_{i, T}\}$, and distributional forecasts of the future values of $\{a_{i, t}, \delta_{i, t} | t = T + 1, \dots, T+H\}$.
 \\
@@
 for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution.
+The unconditional means for acceleration and steering and are set to zero and $\pi/2$ respectively, corresponding to forward motion at a constant velocity.
 \\
@@
 The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them.
-To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogenous (IH) model} and a dependent \textit{clustered heterogeneous} (CH) model.
-In both cases, the shared Heterogeneity is allowed by replacing the shared $\theta$ parameter vector is replaced with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})$ corresponding to the parameters given in the model for car $i$ given in
+To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogenous} (IH) model and a dependent \textit{clustered heterogeneous} (CH) model.
+In both cases, the shared Heterogeneity is allowed by replacing the shared $\theta$ parameter vector is replaced with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})\}$ corresponding to the parameters given in the model for car $i$ given in
 \begin{align}
@@
 The mixture indicator variables satisfying $k_i=j$ determine a cluster of similar $\theta_i$ parameters corresponding to the behaviours of a subset of drivers.
-$\beta$ contains all of the determinants of these distributions, with $\beta = \{\mu_j,\Sigma_j, \pi_j, for j=1,\},$ with \textit{a priori} assumptions for each cluster given by
+$\beta$ contains all of the determinants of these distributions, with $\beta = \{\mu_j,\Sigma_j, \pi_j, \mbox{ for } j=1,\},$ with \textit{a priori} assumptions for each cluster given by
 \begin{align}
@@
 
-For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles that are available.
-We conceive of such information begin available before any self-driving vehicle enters a lone on the road.
-Use of the subscript $i$ is reserved for some $i > N$, and refers to the attriubtes (data or a parameter vector) for an additional behicle that may be encountered while the self-driving vehicle is on the road.
-Inference about $\theta_{1:N}$, the individual parameter vectors for the IH and CH models, as well as for $\theta$, the shared parameter vector under the homogeneous model, and $\beta$, the CH hyper-parameter, conditional on $z_{1:N}$ are assumed to be available.
+For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles in the training set, the vehicles where information is available for before any self-driving vehicle enters the road.
+Use of the subscript $i$ is reserved for some $i > N$, and refers to the attriubtes (data or a parameter vector) for an additional vehicle in the test set that may be encountered while the self-driving vehicle is on the road.
+Inference about $\theta_{1:N}$, the individual parameter vectors for the IH and CH models, as well as for $\theta$, the shared parameter vector under the homogeneous model, and $\beta$, the CH hyper-parameter, conditional on $z_{1:N}$ are assumed to be available before driving.
 \\
@@
 
-Combining a fast inference procedure with the test vehicle posterior distributions implied by the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.
+Combining a fast inference procedure with the test vehicle posterior distributions implied by either of the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.
 
@@
 
-Given a series of data for vehicle $i$ observed up to time $T$, $z_{i, 1:T}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
+Given a series of data for vehicle $i$ observed up to time $T$, $z_{i, 1:T}$, and the training set data $\textbf{z}_{1:N}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
 \begin{equation}
@@
 \end{equation}
-To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogenous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewcar2}) for the CH model must first be inferred.
-However these are only known up to proportionallity, and hence the analytical solution to (\ref{predictive} is unavailable.
+To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogenous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewCar2}) for the CH model must first be inferred.
+However these are only known up to proportionallity, and hence the analytical solution to (\ref{predictive}) is unavailable.
 \\
 
-Collectively denoting all available data, for either $\textbf{z}_{1:N}$ or $z_{i}$, as $\textbf{z}$ for the remainder of this section, two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
+Collectively denoting all available data, including both $\textbf{z}_{1:N}$ and $z_{i, 1:T}$, as $\textbf{z}$ for the remainder of this section, two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
 In brief, MCMC is used to create a sample from $p(\theta_i | \textbf{z})$, with any function of $\theta_i$ that is desired estimated from that sample.
@@
 There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler.
-The Gibbs sampler algorithm iteratively samples the components of a $k$dimensional parameter vector $\theta_i$ via each of the so-called full conditional distributions as follows,
+The Gibbs sampler algorithm iteratively samples the components of a $k$-dimensional parameter vector $\theta_i$ via each of the so-called full conditional distributions as follows,
 \begin{align}
@@
 &\vdots \nonumber \\
-&p(\theta_{i, k} | \theta_{i, 1}, \dots, \theta_{o, k-1}, \textbf{z}). \nonumber
+&p(\theta_{i, k} | \theta_{i, 1}, \dots, \theta_{i, k-1}, \textbf{z}). \nonumber
 \end{align}
@@
 Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
-The computation time for each iteration and the overall number of iterations required to accurately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
-The full conditional distributions cannot be recognised for each of the time series models proposed in Section \ref{sec:models} and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
+The computation time for each iteration and the overall number of iterations required to accurately summarise the posterior distribution is problem specific and typically increases with the amount of data and number of parameters in the model.
+The full conditional distributions cannot be recognised for each of the models proposed in Section \ref{sec:models}, and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
 \\
@@
 
-A typically faster, albeit approximate, alternative to MCMC based inference is VB \citep{Jordan1999}.
+A typically faster, albeit approximate, alternative to MCMC based inference is VB (see \citet{Blei2017} for a recent review).
 VB posits a family of parametric approximating distributions $q_{\lambda}(\theta_i | \textbf{z})$, parameterised by an auxiliary vector $\lambda$, that share the same support as the true posterior distribution $p(\theta_i | \textbf{z})$.
-Note that the family $q_{\lambda}(\theta_i | \textbf{z})$ does not neccesarily depend on $\textbf{z}$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta_i | \textbf{z})$.
-A member of approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta_i | \textbf{z})$ to $p(\theta_i | \textbf{z})$, given by $KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z})]$ \citep{Kullback1951}.
+Note that $q_{\lambda}(\theta_i | \textbf{z})$ does not neccesarily depend on $\textbf{z}$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta_i | \textbf{z})$.
+A member of the approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta_i | \textbf{z})$ to $p(\theta_i | \textbf{z})$, given by $KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z})]$ \citep{Kullback1951}.
 The KL divergence is defined by
@@
 
-Typically (\ref{KL-def}) cannot be evaluated, and Monte-Carlo estimates as
+Typically (\ref{KL-def}) cannot be evaluated, and Monte-Carlo estimates
 \begin{equation}
@@
 where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z}))$.
-The ELBO is equal to the negative KL divergence plus a constant, and hence maximising (\ref{ELBO}) with respect to $q_{\lambda}(\theta_i | \textbf{z})$ is equivalent to minimising (\ref{KL-def}).
+The ELBO is equal to the negative KL divergence plus a constant, and hence maximising \ref{ELBO}) with respect to $q_{\lambda}(\theta_i | \textbf{z})$ is equivalent to minimising (\ref{KL-def}).
 
@@
 \label{subsec:SGA}
-For exponential family likelihood models, $p$, with a factorisable approximation, $q$, the characteristics of the surface of the ELBO can be exploited for optimisation in Mean Field Variational Bayes \citep{Ghahramani2000, Wainwright2008}, but for more general distributions the surface of the ELBO and its stochastic estimate are unknown.
-Maximisation proceeds by optimising only the auxiliary parameters $\lambda$ for a fixed distribution family $q$ with stochastic gradient ascent (SGA).
+For exponential family likelihood models, $p$, with a factorisable approximation, $q$, the characteristics of the surface of the ELBO can be exploited for optimisation in what is known as Mean Field Variational Bayes \citep{Jordan1999, Ghahramani2000, Wainwright2008}, but for more general distributions the surface of the ELBO and its stochastic estimate are unknown.
+In this general case, maximisation proceeds by optimising only the auxiliary parameters $\lambda$ for a fixed distribution family $q$ with stochastic gradient ascent (SGA).
 \\
@@
 \label{rpDist}
-p(f(\epsilon, \lambda) | z) = p(\theta_i | \textbf{z}) |J^{-1}(f(\epsilon, \lambda))|
+p(f(\epsilon, \lambda) | \textbf{z}) = p(\theta_i | \textbf{z}) |J^{-1}(f(\epsilon, \lambda))|
 \end{equation}
 where $J(f(\epsilon, \lambda))$ is the Jacobian Matrix of the transformation $f(\epsilon, \lambda)$.
-Examples of $f$ and $q(\epsilon)$ include treating $\theta_i$ as location scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
+Examples of $f$ and $q(\epsilon)$ include treating $\theta_i$ as location-scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
 \\
 
-The ELBO can be reparameterised by substituting $p(\theta_i, \textbf{z}) = p(f(\epsilon, \lambda), z)|J(f(\epsilon, \lambda))|$ into (\ref{ELBO}),
+The ELBO can be reparameterised by substituting $p(\theta_i, \textbf{z}) = p(f(\epsilon, \lambda), \textbf{z})|J(f(\epsilon, \lambda))|$ into (\ref{ELBO}),
 \begin{equation}
@@
 Choosing the number of samples per estimate, $M$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate.
-An increased value of $M$ will reduce the estimator variance, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
-To reduce the variance of the Monte Carlo estimator, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) which has shown to be more efficient, where numbers in the unit hypercube are generated according to a Sobol Sequence \citep{Sobol1967} which are then randomised using the scrambled net method \citep{Matousek1998}, and transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$.
-Use of RQMC has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
+An increased value of $M$ will the Monte-Carlo error, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
+To reduce the Monte Carlo error, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{{Niederreiter1992, Caflisch1998}).
+In RQMC a deterministic sequence of numbers in the unit hypercube (of dimensionality equal to $\theta_i$) are generated evenly, in the sense that they have a low star discrepency \citep{Sobol1976}, according to a Sobol Sequence \citep{Sobol1967}.
+These numbers are then randomised using the scrambled net method \citep{Matousek1998}, which preserves the low discrepency properties while inducing a central limit theorem.
+$M$ of the resulting numbers are transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$ to simulate $M$ draws from these distributions.
 
@@
 Figure \ref{fig:timeUpdate} illustrates this for a vehicle in the NGSIM dataset travelling towards the right.
-In the left panel, the vehicle has been observed until time $S$, and this data $z_{i, 1:S}$ can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the additional conditional dependence on $\textbf{z}_{1:N}$.
-In the right panel, at some later time $T > S$, an additional $T - S$ data points have been observed, and inference of the behaviour of this vehicle, and hence forecasts, could be improved by incorporating the information contained in $z_{i, S+1:T}$ to form the posterior approximation $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$.
+In the left panel, the vehicle has been observed for a time period $S$ resulting in data $z_{i, 1:S}$ which can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the additional conditional dependence on $\textbf{z}_{1:N}$.
+In the right panel, after observation until some later time $T > S$, an additional $T - S$ data points are available, and inference of the behaviour of this vehicle, and hence forecasts, could be improved by incorporating the information contained in $z_{i, S+1:T}$ to form the posterior approximation $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$.
 Note that the $S$ and $T$ subscripts on $\lambda$ are introduced to differentiate the auxiliary parameter vector conditioned on data up to times $S$ and $T$.
@@
 \includegraphics[width = 0.95\textwidth]{timeUpdate}
-\caption{Left: The path of a vehicle in the NGSIM dataset at time $S$, where the direction of travel is from the left to the right.
+\caption{Left: The path of a vehicle in the NGSIM dataset tat has been observed for a time period equal to $S$, where the direction of travel is from the left to the right.
 Right: The same vehicle at a later time $T > S$, with the extra $T - S$ observations denoted by the dashed line.
-Inference about this vehicle at time $S$ can be updated by the inclusion of the additional information observed up to time $T$.}
+Posterior inference about this vehicle could be updated from time $S$ to time $T$ by the inclusion of this additional information.}
 \label{fig:timeUpdate}
@@
 
-Given $p(\theta_{i} | z_{i, 1:S})$ from (\ref{posterior}), the posterior distribution at time $T$, $p(\theta_{i} | z_{i, 1:T})$ is given by Bayes rule as
+Given the posterior at time $S$, $p(\theta_{i} | z_{i, 1:S})$, the posterior distribution at time $T$, $p(\theta_{i} | z_{i, 1:T})$ is given by Bayes rule as
 \begin{equation}
@@
 \widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
-&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i} z_{i, 1:T})) \right) \label{scoreUpdate}
+&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T})) \right) \label{scoreUpdate}
 \end{align}
@@
 where $\epsilon_j \sim r(\epsilon)$.
-If this process is repeated periodically, $S$ will grow at the same rate as $T$, and updating VB is $O(1)$ rather than $O(T)$ as required by re-using the full sample without the approximation step in (\ref{ApproxJoint}).
+If this process is repeated periodically, $S$ will grow at the same rate as $T$, and UVB has a computational complexity as $T$ increases of $O(1)$, rather than $O(T)$ as required by standard VB that must process $z_{i, 1:T}$ at each possible value of $T$.
 
@@
 
-After obtaining the posterior distributions for each model conditioned on 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on a further 1387 vehicles.
-For each of these vehicles $i > 2000$, at each point in time $T = 100, 110, \dots, 450$ and forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model, while the time-series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$ and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density.
-As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement after every second of observing the vehicle.
-The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ through three methods:
+After obtaining the posterior distributions for each model conditioned on the training set of 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on the  further 1387 vehicles in the test set.
+For each of these vehicles $i > 2000$, after each of $T = 100, 110, \dots, 450$ observations, and at every forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model.
+The time-series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$, and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density.
+As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement for every vehicle after every second of observation.
+The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
 \begin{enumerate}
 \item Posterior sampling with RWMH-MCMC,
-\item VB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$ using the original prior distribution,
+\item VB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$,
 \item UVB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ using $q_{\lambda_S}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:S})$ and data $z_{i, S+1:T}$ as described in Section \ref{sec:UVB}.
 \end{enumerate}
-Approximating distributions for VB are chosen to match the form of the prior distributions, multivariate normal $q$ for the IH model and a mixture of multivariate normals for the CH model.
+
+Approximating distributional families $q$ for each of VB and UVB are chosen to match the form of the model prior distributions: multivariate normal for the IH model and a mixture of multivariate normals for the CH model.
 A summary of these approaches is provided in Table \ref{tableAlg}.
-Point estimate forecasts are evaluated by their Euclidean Error,
+\\
+
+Point estimate forecasts are evaluated by their Euclidean Error in metres,
 \begin{equation}
@@
 \hline
-MCMC & \multicolumn{2}{c|}{RWMH-MCMC jointly drawing the entire $\theta$ vector using \citet{Garthwaite2016} to control the} \\
-& \multicolumn{2}{c|}{variance matrix of the Multivariate Gaussian proposal distribution to obtain a 23.4\% acceptance rate for $\theta$ draws.} \\
+MCMC&\multicolumn{2}{c|}{RWMH-MCMC jointly drawing the entire $\theta_i$ vector using \citet{Garthwaite2016} to control the}\\
+&\multicolumn{2}{c|}{variance matrix of the Multivariate Normal proposal distribution to obtain a 23.4\% acceptance rate for $\theta_i$ draws.}\\
 \hline
-VB  & Variational Bayes using (\ref{rpDeriv}) to update parameters & Variational Bayes using (\ref{scoreDeriv}) to update parameters\\
-Standard &  of a Multivariate Gaussian approximation with &  of an approximation formed as a six component mixture\\
-&non-zero covariance using $M = 25$, where $f$ is a &  of diagonal variance Multivariate Gaussians using $M = 50$. \\
-&location scale transform from a standard normal $r(\epsilon)$. &\\
+VB& Variational Bayes using (\ref{rpDeriv}) to update parameters&Variational Bayes using (\ref{scoreDeriv}) to update parameters\\
+  &of a Multivariate Normal approximation with&of an approximation formed as a six component mixture\\
+&non-zero covariance using $M = 25$, where $f$ is a&of diagonal variance Multivariate Normals using $M = 50$.\\
+&location scale transform from a standard normal $r(\epsilon)$.&\\
 \hline
-VB  & Variational Bayes fit to $100$ observations using (\ref{rpDeriv}), & Variational Bayes fit to $100$ observations using (\ref{scoreDeriv}), \\
- &  then updated every $10$ observations using (\ref{rpUpdate}). & then udpated every $10$ observations using using (\ref{scoreUpdate}). \\
-Standard &  $q$ is a Multivariate Gaussian approximation with &  $q$ is a six component mixture\\
-&non-zero covariance using $M = 25$, where $f$ is a&  of diagonal variance Multivariate Gaussians using $M = 50$. \\
-&location scale transform from a standard normal $r(\epsilon)$. &\\
+UVB&Variational Bayes fit to $100$ observations using (\ref{rpDeriv}),&Variational Bayes fit to $100$ observations using (\ref{scoreDeriv}),\\
+ &then updated every $10$ observations using (\ref{rpUpdate}).&then udpated every $10$ observations using using (\ref{scoreUpdate}).\\
+ &$q$ is a Multivariate Normal approximation with&$q$ is a six component mixture of diagonal\\
+&non-zero covariance using $M = 25$, where $f$ is a&variance Multivariate Normals using $M = 50$.\\
+&location scale transform from a standard normal $r(\epsilon)$.&\\
 \hline
@@
 
-Nine naive forecast models are provided where forecasts of $a_{i, t}$ and $\delta_{i, t}$ at time $T$ are constants equal to $\hat{a}_{i, T}$ and $\hat{\delta}_{i, T}$ for $t = T+1, \dots, T+H$.
+Nine naive forecast models are provided where forecasts of $a_{i, t}$ and $\delta_{i, t}$ at each time $T$ are constants equal to $\hat{a}_{i, T}$ and $\hat{\delta}_{i, T}$ for $t = T+1, \dots, T+H$.
 The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}_{i, T}$ and one estimator for $\hat{\delta}_{i, T}$.
@@
 \begin{center}
+\resizebox{\textwidth}{!}{%
 \begin{tabular}{|l|c|c|c|}
@@
 \hline
-\end{tabular}
+\end{tabular}}
 \end{center}
-\caption{Classification of naive forecast models by future velocity and steering angle estimators.}
+\caption{Classification of naive forecast models by future acceleration and steering angle estimators.}
 \label{tableNaive}
@@
 A Recurrent Neural Network (RNN) is constructed with inputs following \citet{Ding2013}, the ten most recent lags of vehicle position, angle, velocity, and time headway to the preceding vehicle.
-The RNN outputs the next thirty changes in the vehicle position, is trained through gradient descent using the Adam optimiser to minimise the euclidean error of the three second ahead point forecast corresponding to $EE_{i, T, 30}$ defined in (\ref{eucError}).
+The RNN outputs the next thirty changes in the vehicle position, and is trained with gradient descent using the Adam optimiser to minimise the Euclidean error of the three second ahead point forecast corresponding to $EE_{i, T, 30}$ defined in (\ref{eucError}).
 As of writing this report forecast results from this model are not competitive with the naive models and are not discussed further.
@@
 At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model.
+The average forecast Euclidean Error for the best performing Naive model, Naive 8, and each of the time series models at each of these three forecast horizons are presented in Table \ref{table:meanError}.
 \\
 
-Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, and with the exception of the IH model with UVB inference, each time series model and inference implementation produces a forecast with a similar amount of error.
-This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models.
-As the UVB implementation of the CH model is competitive with MCMC and standard VB inference discussion will focus on the CH model rather than the IH model.
+Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, each of which produces a forecast with a similar amount of error that does not increase substantially as the forecast horizon increases.
+This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models, however the UVB inference implementation of the IH model has increased the forecast error relative to MCMC and standard VB.
+The UVB implementation of the CH model produces forecast errors that are competitive with MCMC and standard VB, and so inference discussion will focus on the CH model rather than the IH model.
 \\
@@
 \caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$ for $h \in \{10, 20, 30\}$ and $T \in \{100, 110, \dots, 450\}$ for each model.
-Each time series model is significantly more accurate than the naive constant acceleration / angle models, with little discernable difference between the time series models.}
+Each time series model is significantly more accurate than the naive constant acceleration / angle models, with little discernable difference between each.}
 \label{fig:PredError}
@@
 \caption{The same results as Figure \ref{fig:PredError} with the Naive models excluded.
-Most of the time series model have very similar MAP forecast accuracy, with the UVB / Independent time series combination performing slightly weaker}
+Most of the time series model have very similar MAP forecast accuracy, with the exception of the UVB / IH Model combination}
 \label{fig:PredErrorZ}
@@
 
-There are systematic differences in the level of certainty in forecasts provided by the CH and homogenous models as measured by their logscores, resulting from the CH model being able to to assign each vehicle individual variance parameters,  $\sigma^2_{\epsilon, i}$ for $a_{i, t}$, and $\sigma^2_{\eta, i}$ for $\delta_{i, t}$.
-Figure (\ref{fig:posVarMean}) plots the range of posterior means for these parameters from with RWMH-MCMC sampling with $T = 450$, with the homogenous model posterior means indicated with vertical red lines.
-There is a large amount of positive skewness in the distribution of variances, which may have caused the homogenous model to estimate a variance that is too large for most individual vehicles.
-The homogenous variance posterior means are greater than $80\%$ and $91.5\%$ of CH variance posterior means for acceleration and the steering angle respectivley.
-This large variance has little impact on the location of the MAP of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density.
+\begin{table}[ht]
+\centering
+\resizebox{\textwidth}{!}{%
+\begin{tabular}{| l | c c c c c c c c|}
+\hline
+Horizon & Naive 8 & Homogenous MCMC & IH MCMC & IH VB & IH UVB & CH MCMC & CH VB & CH UVB \\
+\hline
+One Second & 1.10 & \textbf{0.131} & \textbf{0.131} & \textbf{0.131} & 0.137 & \textbf{0.131} & 0.132 & 0.132 \\
+Two Seconds & 1.50 & \textbf{0.226} & 0.228 & 0.228 & 0.246 & 0.227 & 0.230 & 0.227 \\
+Three Seconds & 2.38 & \textbf{0.3} & 0.304 & 0.304 & 0.332 & 0.302 & 0.306 & 0.302 \\
+\hline
+\end{tabular}}
+\caption{Mean forecast Euclidean error in metres for the the best Naive model, Naive 8, and each time series model at a forecast horizon of one second, two seconds and three seconds.
+Each time series model performs significantly better than Naive 8, and are very competitive with each other.
+The use of UVB for the IH model appears to have increased the forecast error relative to MCMC and VB inference for this model.}
+\label{table:meanError}
+\end{table}
+
+The CH model assigns each vehicle in the test set one of a wide range of variances, while the homogenous model must use the same variance paramter for every vehicle.
+Changes in variance parameters has little impact on the location of the MAP of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density, causing systematic differences in the level of forecast uncertainty measured by logscores.
+\\
+Figure (\ref{fig:posVarMean}) displays plots of density of posterior means for each of acceleration and steering angle resulting from RWMH-MCMC sampling with $T = 450$.
+The homogenous model posterior means are indicated with vertical red lines.
+As in Figure \ref{fig:HierSingleKDE}, there is a large amount of variation in the distribution of variances for the test set, many of which are lower than the homogenous model: the homogenous posterior means are greater than $80\%$ and $91.5\%$ of CH posterior means for acceleration and the steering angle respectivley.
+\\
+
 Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the CH model with UVB inference over the homogenous model at the three second forecast horizon.
@@
 \includegraphics[width = 0.95\textwidth]{posVarMean}
-\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by Metropolis-Hastings MCMC at $T = 450$ using the CH model, compared to the homogenous posterior mean in red.
-Small amounts of cars with high variance has skewed the homogenous estimates to be too large for the majority of the vehicles forecasted.}
+\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by RWMH-MCMC at $T = 450$ using the CH model, compared to the homogenous posterior mean in red.
+Small amounts of cars with high variance in the training set has increased the homogenous variances.}
 \label{fig:posVarMean}
@@
 
-As UVB involves replacing the posterior at time $S$ with its variational approximation, the error introduced by UVB depends on the approximation error of Variational Bayes itself.
-The left panel for Figure (\ref{fig:updateCost}) displays boxplots of UVB logscores minus standard VB logscores at the three second forecast horizon for the CH model, and demonstrates that UVB does not typically lead to a decrease in the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
+As UVB involves replacing the posterior distribution at update with its variational approximation, the error introduced by UVB depends on how closely the variational approximation can represent the true posterior distribution.
+The left panel for Figure (\ref{fig:updateCost}) displays boxplots of the difference in logscore between UVB and standard VB at the three second forecast horizon for the CH model, and demonstrates that UVB in this case does not decrease the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
 The right panel displays analogous boxplots for the IH model, where the UVB logscores tend to be lower than the standard VB logscores, matching the increased Euclidean error found for the IH model with UVB inference in Figure \ref{fig:PredErrorZ}.
@@
 We have introduced UVB to allow this inference to happen in the online setting, and have explored fitting copula models to the MCMC samples to produce a functional form for the approximating distribution that better matches the true posterior, though have not been able to link the copula structure for the units in the original hierarchy to the new unit.
-This framework is not specific to vehicle forecasting, and could conceivably be used for many other problems such as classification, or anomoly detection by extending the hierarchical model with a dirichlet process prior.
 \\
+
+This framework is not specific to vehicle forecasting, and could conceivably be used for many other problems, for example classifying new units in an online setting, or the detection of anomolous units through the use of a dirichlet process prior.
+\\
+
+The proposed timeline for the rest of the thesis is as follows:
+\begin{itemize}
+\item March - April: Finalise vehicle trajectory forecasting study for publication
+\item May: Compile inference notes to form thesis background chapters
+\item June - Novemeber: Extend framework in another project
+\item December - May 2019: Additional projects and finalise thesis for submission
+\end{itemize}
 
@@
 
-Excluding the vehicles used to estimate the spline model, each of the 3387 remaining vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
+Excluding the vehicles used to estimate the spline models, each of the 3387 remaining vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
 \begin{align}

### 2018-03-06 Cars/HierMixMod.R

@@ -6,7 +6,6 @@ library(tidyverse)
 library(Rcpp)
 library(RcppArmadillo)
 library(GGally)
-
 setup {
 id <- readRDS('carsID.RDS')
 
@@ -93,6 +92,7 @@ for(i in 1:N){
 mixDraws <- mixtureMCMC(data, reps, draws, hyper, thin, K, 'gaussian', 0.01)
 
 saveRDS(mixDraws, 'mixN2000K6.RDS')
+mixDraws <- readRDS(file = 'mixN2000K6.RDS')
 
 mapGroup <- NULL
 for(i in 1:N){
@@ -190,18 +190,29 @@ for(k in 1:K){
 
 
 
-
 densities %>%
-  mutate(var = factor(var, levels = c('sigSq_eps', 'phi1', 'phi2',
-                                      'sigSq_eta', 'gamma1', 'gamma2'))) %>%
+  mutate(var = case_when(
+    var == 'sigSq_eps' ~ 'sigma[epsilon]^{2}',
+    var == 'phi1' ~ 'phi[1]',
+    var == 'phi2' ~ 'phi[2]',
+    var == 'sigSq_eta' ~ 'sigma[eta]^{2}',
+    var == 'gamma1' ~ 'gamma[1]',
+    var == 'gamma2' ~ 'gamma[2]',
+    TRUE ~ as.character(var)
+  ),
+  var = factor(var, levels = c('sigma[epsilon]^{2}', 'phi[1]', 'phi[2]',
+                                      'sigma[eta]^{2}', 'gamma[1]', 'gamma[2]'))) -> densities
+densities %>% 
   group_by(var, support) %>%
   summarise(dens = sum(dens)) -> densMixMod
 
   ggplot(densMixMod) + geom_line(aes(support, dens)) +
   geom_line(data=densities, aes(support, dens, colour = factor(group))) +
-  facet_wrap(~var, scales = 'free', ncol = 6) + 
-  labs(title = 'Hierarchical Model Prior', x = NULL, y = NULL) + 
-  theme(legend.position = 'none', axis.text.x = element_text(angle = 45, hjust = 1)) -> p1
+  facet_wrap(~var, scales = 'free', ncol = 6, labeller = label_parsed) + 
+  labs(x = NULL, y = NULL) + 
+  theme_bw() + 
+  theme(legend.position = 'none', axis.text.x = element_text(angle = 45, hjust = 1))  -> p1
+  
[…]

### 2018-03-06 Cars/paper.tex
section six editing

@@
 In this case, if $N$ is large $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not vary much from $p(\theta | \textbf{z}_{1:N})$.
-Figure (\ref{fig:homogPosterior}) displays plots of the marginal posterior distributions associated with each element of $\theta$, with $p = q = 2$ resulting from $N = 2000$ vehicles.
+Figure \ref{fig:homogPosterior} displays plots of the marginal posterior distributions associated with each element of $\theta$, with $p = q = 2$ resulting from $N = 2000$ vehicles.
 Arguably each of these marginal posterior distributions are very precise, and are unlikely to change much if updated using information from observations from the $(N+1)^{st}$ vehicle.
@@
 An increased value of $M$ will the Monte-Carlo error, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
-To reduce the Monte Carlo error, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{{Niederreiter1992, Caflisch1998}).
-In RQMC a deterministic sequence of numbers in the unit hypercube (of dimensionality equal to $\theta_i$) are generated evenly, in the sense that they have a low star discrepency \citep{Sobol1976}, according to a Sobol Sequence \citep{Sobol1967}.
+To reduce the Monte Carlo error, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
+In RQMC a deterministic sequence of numbers in the unit hypercube (of dimensionality equal to $\theta_i$) are generated evenly, in the sense that they have a low star discrepency, according to a Sobol Sequence \citep{Sobol1967}.
 These numbers are then randomised using the scrambled net method \citep{Matousek1998}, which preserves the low discrepency properties while inducing a central limit theorem.
@@
 
-After obtaining the posterior distributions for each model conditioned on the training set of 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on the  further 1387 vehicles in the test set.
+After obtaining RWMH-MCMC samples of the posterior distributions for each model conditioned on the training set of 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on the  further 1387 vehicles in the test set.
 For each of these vehicles $i > 2000$, after each of $T = 100, 110, \dots, 450$ observations, and at every forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model.
-The time-series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$, and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density.
+The time series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$, and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density.
 As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement for every vehicle after every second of observation.
-The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
+The forecast density for the homogenous model is created using the samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
 \begin{enumerate}
@@
 The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}_{i, T}$ and one estimator for $\hat{\delta}_{i, T}$.
-Acceleration estimators are the most recent acceleration, $\hat{a}_{i, T} = a_{i, T}$, the average acceleration over the most recent second, $\hat{a}_{i, T} = 0.1 \sum_{s = T-9}^T a_{i, s}$, and zero acceleration $\hat{a}_{i, T} = 0$, equivalent to constant velocity.
-Similarly, steering angle estimators are the most recent angle, $\hat{\delta}_{i, T} = \delta_{i, T}$, the average angle over the most recent second, $\hat{\delta}_{i, T} = 0.1 \sum_{s = T - 9}^T \delta_{i, s}$, and $\hat{\delta}_{i, T} = \pi/2$, corresponding to the vehicle to driving directly forward.
+Acceleration estimators are the current acceleration, $\hat{a}_{i, T} = a_{i, T}$, the average acceleration over the previous second, $\hat{a}_{i, T} = 0.1 \sum_{s = T-9}^T a_{i, s}$, and zero acceleration $\hat{a}_{i, T} = 0$, equivalent to constant velocity.
+Similarly, steering angle estimators are the current angle, $\hat{\delta}_{i, T} = \delta_{i, T}$, the average angle over the previous second, $\hat{\delta}_{i, T} = 0.1 \sum_{s = T - 9}^T \delta_{i, s}$, and $\hat{\delta}_{i, T} = \pi/2$, corresponding to the vehicle to driving directly forward.
 \begin{table}
@@
 \hline
-& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 1/10 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
+& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 0.1 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
 \hline
 $\hat{a}_{i, T} = a_{i, T}$ & Naive 1 & Naive 2 & Naive 3 \\
-$\hat{a}_{i, T} = 1/10\sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
+$\hat{a}_{i, T} = 0.1 \sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
 $\hat{a}_{i, T} = 0$ & Naive 7 & Naive 8 & Naive 9 \\
@@
 
-Eucliedean Error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
+Eucliedean error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
 Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error.
-Zero acceleration naive models (Naive 7, 8, and 9) perform the best, followed by the averaged acceleration models (Naive 4, 5, and 6) and finally the current acceleration models (Naive 1, 2, and 3).
-At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model.
-The average forecast Euclidean Error for the best performing Naive model, Naive 8, and each of the time series models at each of these three forecast horizons are presented in Table \ref{table:meanError}.
+The best performing Naive model at each forecast horizon is Naive 8, however at all forecast horizons the error for each implementation of each of the time series models are significantly improved relative to every naive model.
+The average forecast Euclidean error for Naive 8 and each of the time series models at each of these three forecast horizons is presented in Table \ref{table:meanError}.
 \\
 
-Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, each of which produces a forecast with a similar amount of error that does not increase substantially as the forecast horizon increases.
+Figure \ref{fig:PredErrorZ} truncates Figure \ref{fig:PredError} to include only the time series models, each of which produces a forecast with a similar amount of error that does not increase substantially as the forecast horizon increases.
 This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models, however the UVB inference implementation of the IH model has increased the forecast error relative to MCMC and standard VB.
@@
 
-The CH model assigns each vehicle in the test set one of a wide range of variances, while the homogenous model must use the same variance paramter for every vehicle.
-Changes in variance parameters has little impact on the location of the MAP of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density, causing systematic differences in the level of forecast uncertainty measured by logscores.
+The CH model allows each vehicle in the test to have a posterior variance distribution that is unique to that vehicle, while the homogenous model must share the same posterior variance distribution across different vehicles.
+Differences in the distribution of either variance parameter has little impact on the location of the maximum of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but can drastically affect the precision of this forecast density, causing systematic differences in the level of uncertainty associated with the forecasts that can be measured by the logscore.
 \\
-Figure (\ref{fig:posVarMean}) displays plots of density of posterior means for each of acceleration and steering angle resulting from RWMH-MCMC sampling with $T = 450$.
+
+Figure \ref{fig:posVarMean} displays plots of density of posterior means for each of acceleration and steering angle resulting from RWMH-MCMC sampling with $T = 450$.
 The homogenous model posterior means are indicated with vertical red lines.
@@
 Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis.
-The CH model trajectory forecasts for drivers in the lower variance quintiles benefit the most from including heterogeneity.
+The upper left panels correspond to vehicles in the lower posterior variance quintiles, showing clear benefits for CH model trajectory forecasts relative to the homogenous model.
 However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles.
-This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance estimates allow identification of vehicles with high variance, and potentially more erratic and dangerous driving styles.
+This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance distributions allow identification of vehicles with high variance, which could display potentially more erratic and dangerous behaviour.
 \\
@@
 
-The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles.
-Figure (\ref{fig:timing}) demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$.
-The compuation time for standard VB increases linearly with $T$, while UVB is constant; posterior inference available at time $T$ utilising UVB can be conditioned on more data than standard VB.
-Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time and make UVB feasible for more complex models.
+The benefits of heterogenous models is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles.
+Figure \ref{fig:timing} demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$.
+The compuation time for standard VB increases linearly with $T$, while UVB is constant as $T$ increases; and hence posterior inference utilising UVB to condition posterior distributions on additional data is feasible.
+Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time, allowing for shorter update periods or more complex models.
 \\
@@
 As UVB involves replacing the posterior distribution at update with its variational approximation, the error introduced by UVB depends on how closely the variational approximation can represent the true posterior distribution.
-The left panel for Figure (\ref{fig:updateCost}) displays boxplots of the difference in logscore between UVB and standard VB at the three second forecast horizon for the CH model, and demonstrates that UVB in this case does not decrease the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
+The left panel for Figure \ref{fig:updateCost} displays boxplots of the difference in logscore between UVB and standard VB at the three second forecast horizon for the CH model, and demonstrates that UVB in this case does not decrease the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
 The right panel displays analogous boxplots for the IH model, where the UVB logscores tend to be lower than the standard VB logscores, matching the increased Euclidean error found for the IH model with UVB inference in Figure \ref{fig:PredErrorZ}.
@@
 \label{sec:disc}
-This paper demonstrates that the use of auto-regressive models time-series models drastically improves trajectory forecasts when compared to the competing models.
+This paper demonstrates that the use of auto-regressive models time series models drastically improves trajectory forecasts when compared to the competing models.
 These models parameterise the dynamics of a driver's actions, with a range of parameter posterior means providing evidence of heterogeneity between the drivers of different vehicles.

### 2018-03-08 Cars/midcanPresentation.tex
presentation

--- 2018-03-05 Cars/paper.tex
@@
-  \documentclass[12pt,a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[8pt]{beamer}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
-\let\hlipl\hlkwb
 
@@
 
-\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\usepackage{alltt}
+
 \makeatletter
@@
 
-\usepackage{alltt}%
-\usepackage[]{graphicx}\usepackage[]{color}
-%% maxwidth is the original width if it is less than linewidth
-%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
+\definecolor{shadecolor}{rgb}{.97, .97, .97}
+\definecolor{messagecolor}{rgb}{0, 0, 0}
+\definecolor{warningcolor}{rgb}{1, 0, 1}
+\definecolor{errorcolor}{rgb}{1, 0, 0}
 
-%2multibyte Version: 5.50.0.2960 CodePage: 1252
+\usepackage{alltt}
+\usepackage{natbib}
+\usepackage{array}
+\usepackage[font=small,skip=5pt]{caption}
+\usepackage{graphicx}
+\usepackage{amsmath}
+\usepackage{dsfont}
+\usepackage[]{algorithm2e}
+\usepackage{amsthm}
 \usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage[centertags]{amsmath}
-\usepackage{graphicx}%
-\usepackage{natbib}
-\usepackage{color}
-\usepackage[dvipsnames,svgnames*]{xcolor}
-\usepackage{array}
-\usepackage[hidelinks]{hyperref}
-\usepackage[font=small,skip=5pt]{caption}
-\usepackage[aboveskip=2pt]{subcaption}
-\usepackage{amsmath}
-\usepackage{amsthm}
-%\usepackage{tikz}
-%\usetikzlibrary{bayesnet}
 \usepackage{url}
@@
 \usepackage{afterpage}
-\setcounter{MaxMatrixCols}{30}
+\usepackage{bbm}
+\usepackage{tikz}
+\usepackage{amssymb}
+\usepackage{bm}
+\setbeamertemplate{caption}[numbered]
 
-\def\app#1#2{%
-  \mathrel{%
-    \setbox0=\hbox{$#1\sim$}%
-    \setbox2=\hbox{%
-      \rlap{\hbox{$#1\propto$}}%
-      \lower1.1\ht0\box0%
-    }%
-    \raise0.25\ht2\box2%
-  }%
+\usetheme{Warsaw}
+\defbeamertemplate*{footline}{shadow theme}
+{%
+  \leavevmode%
+  \hbox{\begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm plus1fil,rightskip=.3cm]{author in head/foot}%
+    \usebeamerfont{author in head/foot}\insertframenumber\,/\,\inserttotalframenumber\hfill\insertshortauthor
+  \end{beamercolorbox}%
+  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.5ex,dp=1.125ex,leftskip=.3cm,rightskip=.3cm plus1fil]{title in head/foot}%
+    \usebeamerfont{title in head/foot}\insertshorttitle%
+  \end{beamercolorbox}}%
+  \vskip0pt%
 }
-\def\approxprop{\mathpalette\app\relax}
 
-\setlength{\topmargin}{0in}
-\setlength{\oddsidemargin}{0.1in}
-\setlength{\evensidemargin}{0.1in}
-\setlength{\textwidth}{6.5in}
-\setlength{\textheight}{8.25in}
-\setlength\parindent{0pt}
+\setbeamertemplate{headline}{%
+\leavevmode%
+  \hbox{%
+    \begin{beamercolorbox}[wd=\paperwidth,ht=2.5ex,dp=1.125ex]{palette quaternary}%
+    \insertsectionnavigationhorizontal{\paperwidth}{\hskip0pt plus1filll}{\hskip0pt plus1filll}
+    \end{beamercolorbox}%
+  }
+}
+
+\newcounter{saveenumi}
+\newcommand{\seti}{\setcounter{saveenumi}{\value{enumi}}}
+\newcommand{\conti}{\setcounter{enumi}{\value{saveenumi}}}
+\resetcounteronoverlays{saveenumi}
+
+\title[Online Updating of Variational Bayes]{Online Updating of Variational Bayes for Heterogeneous Forecasts of Vehicle Trajectory}
+\author[Nathaniel Tomasetti]{Nathaniel Tomasetti}
+\date{ }
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\begin{document}
 
-\title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}
-\author{Nathaniel Tomasetti}
-\date{}
+\begin{frame}
+\titlepage
+\centering
+Supervised by Catherine Forbes and Anastasios Panagiotelis
+\end{frame}
 
-%TODO: Add a conclusion
+\begin{frame}
+\tableofcontents
+\end{frame}
 
-\begin{document}
-\maketitle
+\begin{frame}
+\section{Framework}
+\frametitle{Thesis Framework}
+Focus: Two sets of data
+\begin{itemize}
+\item Stuff Here about the general themes
+\item Use of exact inference to improve VB
+\item Hierarchical models with exact inference
+\item Some new unit with time constrained inference
+\item Want to use MCMC info to improve VB
+\end{itemize}
+\end{frame}
 
-\section{Introduction}
-\label{sec:intro}
+\begin{frame}
+\section{Inference}
+\frametitle{Bayesian Inference}
+Note: reframe discussion with two sets of data from previous slide
+\begin{itemize}
+\item The posterior distribution for parameters $\theta$, $p(\theta | \textbf{z})$, can be inferred with Markov Chain Monte Carlo (MCMC). MCMC:
+\begin{enumerate}
+\item Allows 'exact' inference as the number of iterations increases
+\item Can be very slow
+\end{enumerate}
+\item We focus on settings where MCMC inference for the additional unit is too slow to be useful.
+\item Instead, Variational Bayes (VB) is used. VB:
+\begin{enumerate}
+\item Approximates the posterior with an analytical alternative $q_{\lambda}(\theta | \textbf{z})$.
+\item Chooses the distribution $q$ that minimises a divergence funciton
+\item Is typically much faster than MCMC.
+\end{enumerate}
+\end{itemize}
+\end{frame}
 
-Self-driving vehicles are rapidly becoming more advanced, with manufacturers such as Tesla, Ford, and Audi expecting autonomous vehicles available to consumers as early as 2020.
-These vehicles rely on a nagivation system that detects the position of surrounding traffic, forecasts their trajectery, and selects a path to avoid any collisions.
-As many as $94\%$ of accidents are due to human error according to data from the United States National Highway Traffic Saftey Administration in 2015, \citep{NHTSA2015}, which can feasibly be reduced by self-driving vehicles.
-However self-driving vehicles will operate in situations where surrounding vehicles are controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
-\\
+\begin{frame}
+\frametitle{Kullback Leibler Divergence and Stochastic Optimisation}
+\begin{itemize}
+\item The typical divergence is the Kullback Leibler Divergence from $q$ to $p$.
+\end{itemize}
+\begin{equation}
+\label{KL-def}
+KL[q_{\lambda}(\theta | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta | \textbf{z})] = E_{q_{\lambda}(\theta | \textbf{z})} \left[ \log(q_{\lambda}(\theta | \textbf{z})) - \log(p(\theta | \textbf{z})) \right],
+\end{equation}
+\begin{itemize}
+\item Monte-Carlo simulations are usually unavailable due to the $p(\theta | \textbf{z})$ term.
+\end{itemize}
+\begin{equation}
+\label{ELBO}
+\mathcal{L}(q, \lambda) = E_{q_{\lambda}(\theta | \textbf{z})} \left[\log(p(\theta, \textbf{z})) - \log(q_{\lambda}(\theta | \textbf{z}))\right],
+\end{equation}
+\begin{itemize}
+\item Stochastic Estimates are available through Monte-Carlo:
+\end{itemize}
+\begin{equation}
+\label{ELBO-MC}
+\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j}, \textbf{z})) - \log(q_{\lambda}(\theta_{j} | \textbf{z})) \right)
+\end{equation}
+where $\theta_{j} \sim q_{\lambda}(\theta | \textbf{z})$.
+\end{frame}
 
-While characteristics of surrounding vehicles, such as theor postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g.
-\citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory has seen less attention in the literature.
-These forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
-\\
+\begin{frame}
+\frametitle{Gradient Estimators}
+\begin{itemize}
+\item For most $p$ and $q$ only Stochastic Gradient Ascent is available for optimisation. Repeatedly apply:
+\begin{equation}
+\label{gradientAscent}
+\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
+\end{equation}
+\item ELBO gradients can be estimated by the Score Estimator:
+\begin{equation}
+\label{scoreDeriv}
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{j, i} | \textbf{z}))}{\delta \lambda} \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right),
+\end{equation}
+where $\theta_j \sim q_{\lambda}(\theta | \epsilon)$.
+\item Or the Reparametersied Estimator, where $\theta = f(\epsilon, \lambda)$:
+\begin{equation}
+\label{rpDeriv}
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta \theta}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda},
+\end{equation}
+where $\epsilon_j \sim q(\epsilon)$.
+\end{itemize}
+\end{frame}
 
-This paper takes a statistical approach to this trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset.
-It uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability trajectories that can lead to a collision to be detected.
-The time series models used allow for heterogeneity between different vehicles, which may result from sources of driver heterogeneity that may result from factors such as differences in individual driving styles, or the conditions of traffic directly around them.
-Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration while forecasts for more consistent drivers will benefit from a low variance.
-\\
+\begin{frame}
+\section{Online Heterogenous Forecasting: Problem}
+\subsection{Motivation}
+\frametitle{Self-Driving Vehicles}
+\begin{itemize}
+\item Self-Driving Vehicles are a thing soon
+\item The navigation system has three steps
+\begin{enumerate}
+\item Detect surrounding traffic
+\item Forecast surrounding traffic
+\item Navigate through safely
+\end{enumerate}
+\item Forecasting has not had much work
+\end{itemize}
+\end{frame}
 
-Incorporating heterogeneity into forecasts requires the behaviour of each driver of interest to be inferred as vehicles are encountered while driving.
-This inference is facilitated by first obtaining the posterior distribution for a large number of vehicles before the self-driving vehicle is on the road, from which the prior distribution for additional vehicles can be constructed.
-Once the self-driving vehicle is on the road, online inference is made possible by the introduction of Updating Variational Bayes (UVB), which approximates Bayesian updating so that inference may be periodically updated as additional data relating to surrounding vehicles is made available.
-\\
+\begin{frame}
+\frametitle{Aims of the first paper}
+\begin{itemize}
+\item What we plan on doing with this first paper
+\item Overall goals and stuff
+\item Motivate heterogeneity
+\item Note heterogeneity needs inference on vehicles as they are encountered
+\item There is not much time to do that
+\item Develop an approximation to the Bayesian Update Equation so we can infer parameters for new vehicles online
+\item Also mention something on density forecasting would be nice
+\end{itemize}
+\end{frame}
 
-We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the contstant models commonly used, and including driver heterogeneity in the model substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous model.
-These results give credence to the validity of the methodological framework developed in this thesis, where exact Bayesian inference techniques are used to guide the construction on a prior distribution for parameters that are approximately inferred in an online setting.
-\\
+\begin{frame}
+\subsection{Data}
+\frametitle{Next Generation Simulation}
+\begin{itemize}
+\item This is a dataset created for macro traffic modelling
+\item It records all of the cars all of the time
+\item The road is curved
+\item Before and After of road straightening project
+\end{itemize}
+\end{frame}
 
-This paper is structured as follows: Section \ref{sec:dataProcessing} describes the dataset used and steps required to extract acceleration and steering angle for each vehicle, while Section \ref{sec:models} introduces several time series models used.
-Section \ref{sec:Inference} reviews Bayesian inference, which is extended to the online updating setting in Section \ref{sec:UVB}.
-Finally Section \ref{sec:eval} contains empirical results, and Section \ref{sec:disc} concludes the vehicle trajectory forecasting study and discusses future work.
-
-\section{Data processing}
-\label{sec:dataProcessing}
-Data is provided by the Next Generation Simulation (NGSIM) project, which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005.
-Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc to produce coordinates for each vehicle and point in time relative to the start of the road section.
-Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which traveled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1.
-Vehicles that entered or exited midway through the freeway section, are excluded in this paper.
-There is a curvature to the road occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.
-\\
-
+\begin{frame}
+\frametitle{Forecasting in the framework}
+\begin{itemize}
+\item That is a terrible title.
+\item Split the data into two sets: train/test
+\item Cars observed beforehand, cars encountered while driving
+\end{itemize}
 \begin{figure}
 \centering
-\includegraphics[width=0.65\textwidth]{carPath}
-\caption{The path of ten vehicles in the dataset, with each black line representing a unique vehichle.
-This section of US 101 is split into five main lanes, with an additional entry/exit lane to the far right, through which two of the vehicles have entered.
-There is curvature to the road (which is slightly distorted by the aspect ratio) with bends occuring between 500 and 1000 feet, and again between 1800 and 2000 feet.}
-\label{fig:rawData}
+\includegraphics[width = 0.95\textwidth]{problem}
 \end{figure}
+\end{frame}
 
-Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road compensating for any curvature.
-To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self-driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates.
-This is facilitated by measuring the location of vehicles relative to an estimate of the midpoint of their associated lane.
-\citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway.
-Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
-\\
+\begin{frame}
+\frametitle{Trigonometric Motion Equations}
+\begin{align}
+x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq}, \\
+y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq}.
+\end{align}
+Picture
+\end{frame}
 
-Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to observation $t$ according to the trigonometric relationship in Figure \ref{fig:motion}:
-\begin{align}
-x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
-y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
-\end{align}
-Note that when $\delta_{i, t} \pm \pi/2$ then car $i$ takes a position parallel to the centre line, and hence $x^*_{i, t} = x^*_{i, t-1}$.
-\\
-
-From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs from driver $i$ are calculated via
+\begin{frame}
+\frametitle{Extracting Driver Inputs}
 \begin{align}
@@
 \end{align}
-In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=2,3,...,T\}$, according to
 \begin{equation}
@@
 \end{equation}
-\\
+\begin{itemize}
+\item Why other variables are useful
+\item Acceleration is good because relative velocity / non-stationarity
+\end{itemize}
+\end{frame}
 
-Data resulting from the observation of vehicle $i$ for a period of time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$.
-3387 vehicles remain in the dataset after the above exclusion criteria are applied.
-The data into a two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1387 vehicles.
-\\
+\begin{frame}
+\subsection{Modelling}
+\frametitle{Auto-regressive Models for Trajectory Forecasting}
+\begin{itemize}
+\item Evidence of dynamics behaviour
+\item Modelling approachs / relationship to test train set
+\item Motivate the use of the three different models
+\end{itemize}
+\end{frame}
 
-The training set and the test set are treated as observations from vehicles at two distinct time periods.
-The training set represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles.
-The test set represents those vehicles the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.4\textwidth]{motion}
-\caption{A vehicle at coordinate $\{x^*_{t-1}, y^*_{t-1}\}$ at time $t-1$ will be at $\{x^*_t, y^*_t\}$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
-\label{fig:motion}
-\end{figure}
-
-\section{Models for Trajectory Forecasting}
-\label{sec:models}
-
-Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between different vehicles.
-These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent heterogenous (IH) model and the hierarchical, clustered heterogeneity (CH) model.
-The ID model does not incorporate information from the training set of vehicles into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set.
-\\
-
-\subsection{A Homogenous Time Series Model}
-\label{subsec:homogenous}
-
-The framework provided by (\ref{xEq}) and (\ref{yEq}) allows distributional forecasts of future position coordinates of a given vehicle $i$, $\{x^*_{i, t}, y^*_{i, t} | t = T + 1, \dots, T+H\}$ to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_{i, T}, y^*_{i, T}, v_{i, T}\}$, and distributional forecasts of the future values of $\{a_{i, t}, \delta_{i, t} | t = T + 1, \dots, T+H\}$.
-\\
-
-\iffalse % Partial Autocorrelation plots are a better way to show dynamic behaviour in the data.
-This is highlighted in Figure \ref{fig:dynamics}, where the left panel displays a plot of the mean value of
-\begin{equation}
-A_{i, s} = \frac{a_{i, \tau_i + s}}{a_{i, \tau_i}}
-\label{amax}
-\end{equation}
-where $\tau_i = \arg \underset{t}{\max}|a_{i, t}|$, the relative time $s$ takes on values between -1000 and 500 milliseconds, and with the average taken over the $i = 1, 2, \dots, N$ individual cars.
-On average, large vehicle acceleration or deceleration is preceeded by an increase in acceleration in the same direction for 400 milliseconds.
-The right panel displays a plot of the mean value of
-\begin{equation}
-D_{i, s} = \frac{\delta_{i, \omega_i + s} - \pi/2}{\delta_{i, \omega_i} - \pi/2}
-\label{dmax}
-\end{equation}
-where $\omega_i = \arg \underset{t}{\max}|\delta_{i, t} - \pi/2|$, for again $-1000 \leq s \leq 500.$ Simmilarly, this second panel demonstrates that the largest steering angle deviation from $\pi/2$ may be anticipated by smaller deviations in the same direction, at typically, some point during the previous 300 milliseconds.
-These plots sugget that changes in acceleration and angle are predictable if given the recent history of a vehicle's position.
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.95\textwidth]{dynamics}
-\caption{Left: Mean values of acceleration as a percent of a vehicles maximum absolute acceleration for the 1000 milliseconds before, and 500 milliseconds after, a vehicle reaches its maximum absolute acceleration.
-Right: The analogous plot for the steering angle, where deviations in angle from $\pi/2$ are plotted as a percentage of a vehicles maximum absolute value of deviation.
-The few observations before either variable reaches its maximum often have increased values in the same direction.}
-\label{fig:dynamics}
-\end{figure}
-\fi
-
-NGSIM vehicle data is observed every 100 milliseconds so large changes in $a_{i, t}$ and $\delta_{i, t}$ may occur over multiple sequential observations.
-This is highlighted in Figure \ref{fig:pacf}, which displays a plot of the partial autocorrelation function, which measures the correlation between a variable at two different time periods, $t$ and $t + k$, conditioned on that variable at time $t + 1, \dots t + k -1$, for both $a_{i, t}$ and $\delta_{i, t}$ for five vehicles.
-Large spikes in the partial autocorrelation plots, such as those at the first and second lag for most vehicles, demonstrate that the value of either $a_{i, t}$ or $\delta_{i, t}$ is strongly correlated with the previous time periods.
-\\
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.95\textwidth]{pacf}
-\caption{Top: Partial Autocorrelation plots for acceleration, $a$, for a random selection of vehicles.
-Bottom: Partial Autocorrelation plots for steering angle, $\delta$ for the same vehicles.
-Each spike for vehicle $i$ at lag $k$ in the top panel indicates that $a_{i, t-k}$ is correlated with $a_{i, t}$, similarly spikes in the bottom panel indicate that $\delta_{i, t-k}$ is correlated with $\delta_{i, t}$.
-Note that different vehicles do not have the same dynamic behaviour in the partial autocorrelations.}
-\label{fig:pacf}
-\end{figure}
-
-These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and as an attempt to model the dependence, two univariate auto-regressive processes of orders $p$ and $q$, respectively, for $a_t$ and angle  $\delta_t$ are employed, given by
+\begin{frame}
+\frametitle{A Homogeneous Approach}
+Talk about what is going on here
 \begin{align}
@@
 \end{align}
-for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution. 
-\\
-
-For notational convenience, and to facilitate implementation of the Bayesian inferential approach described in Section \ref{sec:Inference}, the variance parameters $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$ are transformed and collected together with all other model parameters into the $\mathbb{R}^{p + q + 2}$-dimensional vector
 \begin{equation*}
@@
 \end{equation*}
-With each component of $\theta$ able to take on any real value, the prior distribution given by
 \begin{equation}
 \label{indPrior}
-\theta \sim \mathcal{N}\left(\mu, \Sigma \right),
+\theta \sim \mathcal{N}\left(\mu, \Sigma \right)
 \end{equation}
-is selected with $\mu=(0_p^{\prime},0_q^{\prime},-5, -5)$ and $\Sigma = 10 \mathbb{I}_{p+q+2}$, where $0_r$ and $\mathbb{I}_r$ denote the $r$-dimensional zero vector and identity matrix, respectively, noting that the scale of changes in either acceleration or angle at the 100 millisecond time-scale are both very small.
-We refer to the joint specification in (\ref{aAR}) and (\ref{dAR}), along with the given normal prior distributional assumption, as the \textit{homogenous model} throughout the paper.
-\\
-Treating the first $\max(p, q)$ observations for each vehicle as deterministic, the posterior distribution for the homogenous model is given by
-\begin{equation}
-\label{homogPost}
-p(\theta | \textbf{z}_{1:N}) \propto \prod_{i=1}^N \prod_{t = \max(p, q) + 1}^T p(z_{i, t} | z_{i, t-\max(p, q):t-1}, \theta)p(\theta).
-\end{equation}
+\begin{itemize}
+\item The Posterior
+\item Why are we doing this
+\end{itemize}
+\end{frame}
 
-\subsection{Two Heterogenous Time Series Models}
-\label{subsec:heterogenous}
-
-The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them.
-To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogenous (IH) model} and a dependent \textit{clustered heterogeneous} (CH) model.
-In both cases, the shared Heterogeneity is allowed by replacing the shared $\theta$ parameter vector is replaced with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})$ corresponding to the parameters given in the model for car $i$ given in
+\begin{frame}
+\frametitle{The Independent Heterogeneity Model}
+Maybe different drivers behave differently?
 \begin{align}
@@
 \end{align}
-with $\epsilon_{i,t}$ and $\eta_{i,t}$ all independent standard normal random variables.
-For the IH model, each $\theta_i,$ for $i=1,2,\ldots, N$, is assumed to follow an independent $N(\mu, \Sigma)$ prior distribution, resulting in a joint posterior distribution that may be decomposed into $N$ distinct independent marginal distributions, i.e.
+This breaks down into a bunch of independent posteriors
 \begin{equation}
@@
 \end{equation}
-due to the distinct parameters and independent (albeit identical) prior distributions for each of the $N$ models. \footnote{Recall that $z_{i}=\{x^*_{i,s},y^*_{i,s},\mbox{ for } s=1,2,...,T \}$.}
-\\
+The training set isn't terribly useful.
+\begin{equation}
+p(\theta_{i}| \textbf{z}_{1:N}, z_{i,1:T}, \theta_{1:N}) = p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_i) p(\theta_i).
+\label{indepNewCar}
+\end{equation}
+\end{frame}
 
-In contrast, the alternative CH model does not assume \textit{a priori} independence of each of the $\theta_i$ vectors and instead clusters individual $\theta_i$ values into one of $K$ common latent multivariate normal distributions as determined by $\beta.$ The hierarchical CH model assumes that $\beta$ is itself random, with $\beta \sim p(\beta)$.
-By augmenting each driver's individual parameter $\theta_i$ with the corresponding $k_i=j$ variable, where $j$ can take on any value $1,2 \ldots, K$, we have
+\begin{frame}
+\frametitle{The Clustered Heterogeneity Model}
+Hierarchical Models share information 
+\begin{itemize}
+\item Introduce Auxiliary Variable K
+\item Sharing of the information
+\end{itemize}
 \begin{equation}
@@
 \end{equation}
-for each $i = 1, \dots, N$.
-The mixture indicator variables satisfying $k_i=j$ determine a cluster of similar $\theta_i$ parameters corresponding to the behaviours of a subset of drivers.
-$\beta$ contains all of the determinants of these distributions, with $\beta = \{\mu_j,\Sigma_j, \pi_j, for j=1,\},$ with \textit{a priori} assumptions for each cluster given by
 \begin{align}
@@
 \end{align}
-\\
+\end{frame}
 
-The hyperparameters $\beta$ are $\{\mu_j, \Sigma_j, \pi_j | j = 1, \dots, K\}$.
-Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$.
-
-To illustrate that the hierarchical CH model, with $K=6$ adequately captures the main features of the IH model, the top panel of Figure \ref{fig:HierSingleKDE} displays kernel density estimates of the posterior means, for each $\theta_i, i = 1, \dots, 2000$, resulting from the IH model.
-For this illustration, we set $p = q = 2$ and use $N=2000$ observations over times $t=1, 2, \ldots, T = 500$.
-These summaries each indicate a wide range of values for the $N$ posterior means corresponding to each model parameter, corresponding to heterogeneity across drivers in the sample.
-Notably, the posterior means for each of the variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the distribution of posterior means is multimodal for each of the $\phi_1$ and $\phi_2$ parameters.
-
-These summaries are compared against the figures in the bottom panel of Figure \ref{fig:HierSingleKDE}, where the estimated marginal distributions for $p(\theta_{2001} | \textbf{z}_{1:2000})$are displayed.
-The CH marginal posterior distributions appear to have captured the skewness and multimodality present in the corresponding IH posterior mean summaries shown in the top panel.
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.95\textwidth]{HierSingleKDE.png}
-\caption{Top: A Kernel Density Estimate applied to $E(\theta_i | z_i), i = 1, \dots, 2000$ where $p = q = 2$ from the IH model, sampled by Metropolis-Hasting MCMC.
-Bottom: $p(\theta_{2001}| | \textbf{z}_{1:2000})$ for the CH model with $K = 6$ using the same vehicles sampled by Metropolis-Hasting MCMC.
-Each individual component in the mixture is denoted by the coloured densities.
-The CH model has captured the multimodality and skewness present in the IH posterior means.}
-\label{fig:HierSingleKDE}
-\end{figure}
-
-\subsection{Encountering Additional Vehicles}
-\label{subsec:additionalVehicles}
-
-For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles that are available.
-We conceive of such information begin available before any self-driving vehicle enters a lone on the road.
-Use of the subscript $i$ is reserved for some $i > N$, and refers to the attriubtes (data or a parameter vector) for an additional behicle that may be encountered while the self-driving vehicle is on the road.
-Inference about $\theta_{1:N}$, the individual parameter vectors for the IH and CH models, as well as for $\theta$, the shared parameter vector under the homogeneous model, and $\beta$, the CH hyper-parameter, conditional on $z_{1:N}$ are assumed to be available.
-\\
-
-The idea we pursue is to use the available data, and corresponding model posterior distributions, to forecast the future behaviour of a vehicle that is positioned near a given self-driving vehicle.
-Leveraging off of the information gathered from previous data (via previously computed model posteriors that accommodate the range of heterogeneous behaviour of the population of vehicles) together with fresh information received in real-time, we aim to predict the near future position of an additional vehicle moving contemporaneously with the self-driving one
-\\
-
-A crucial aspect of the application of the discussed time series models is through the relationship between the posterior distribution formed from observations relatting to the first $N$ vehicles and the parameter vector for the additional new vehicle, $\theta_{N+1}$.
-Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} \equiv \theta$.
-In this case, if $N$ is large $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not vary much from $p(\theta | \textbf{z}_{1:N})$.
-Figure (\ref{fig:homogPosterior}) displays plots of the marginal posterior distributions associated with each element of $\theta$, with $p = q = 2$ resulting from $N = 2000$ vehicles.
-Arguably each of these marginal posterior distributions are very precise, and are unlikely to change much if updated using information from observations from the $(N+1)^{st}$ vehicle.
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.95\textwidth]{homogPosterior}
-\caption{The marginal posterior distributions $p(\theta | \textbf{z}_{1:N})$ for the homogenous model with $N = 2000$ and $p = q = 2$. Each variable has very high posterior precision.}
-\label{fig:homogPosterior}
-\end{figure}
-
-The IH model implies that the posterior distribution for $\theta_i$ is independent of $\theta_{1:N}$ and $\textbf{z}_{1:N}$ by
-\begin{equation}
-p(\theta_{i}| \textbf{z}_{1:N}, z_{i,1:T}, \theta_{1:N}) = p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_i) p(\theta_i).
-\label{indepNewCar}
-\end{equation}
-\\
-
-The posterior for $\theta_i$ using the CH model, assuming conditional independence between each $\theta_j$ for $j = 1, \dots, N$ and $\theta_i$, given $\beta$, $\textbf{z}_{1:N}$, and $z_i$, is
+\begin{frame}
+\frametitle{The Clustered Heterogeneity Model}
+The new vehicle depends on the old vehicles through $\beta$.
 \begin{equation}
@@
 \end{equation}
-As N is large the sampled posterior $p(\beta | \textbf{z}_{1:N})$ has high precision, and $\int_{\beta} p(\theta_{i} | \beta) p (\beta | \textbf{z}_{1:N}) d\beta$ can be replaced with $p(\theta_{i} | \hat{\beta})$ in (\ref{hierNewCar}) where $\hat{\beta}$ is a point estimate such as the mean or maximum of $p(\beta | \textbf{z}_{1:N})$.
-This approximation results in the closed form expression
+The $\beta$ posterior is precise for large $N$
 \begin{equation}
@@
 \end{equation}
-\\
+\end{frame}
 
-Combining a fast inference procedure with the test vehicle posterior distributions implied by the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.
+\begin{frame}
+\begin{itemize}
+\item That picture showing the IH / CH fits
+\item Evidence of heterogeneity and the suitability of the mixture model
+\end{itemize}
+\end{frame}
 
-\section{Bayesian Inference}
-\label{sec:Inference}
-
-Given a series of data for vehicle $i$ observed up to time $T$, $z_{i, 1:T}$, the Bayesian forecast distribution associated with some future time $T+h$ is characterized by the conditional density
-\begin{equation}
-\label{predictive}
-p(z_{i, T+h} | z_{i, 1:T}) =\int p(z_{i, T+h}|z_{i, 1:T}, \theta_i) p(\theta_i | \textbf{z}_{1:N}, z_{i, 1:T}) d\theta.
-\end{equation}
-To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogenous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewcar2}) for the CH model must first be inferred.
-However these are only known up to proportionallity, and hence the analytical solution to (\ref{predictive} is unavailable.
-\\
-
-Collectively denoting all available data, for either $\textbf{z}_{1:N}$ or $z_{i}$, as $\textbf{z}$ for the remainder of this section, two alternative methods for computing the desired posterior distribution will be reviewed: Markov Chain Monte Carlo (MCMC) and Variational Bayes (VB).
-In brief, MCMC is used to create a sample from $p(\theta_i | \textbf{z})$, with any function of $\theta_i$ that is desired estimated from that sample.
-In contrast, VB replaces $p(\theta_i | \textbf{z})$ with a parametric approximation, denoted by $q_{\lambda}(\theta_i |\textbf{z})$, where $\lambda$ is a vector of auxiliary parameters associated with the approximation that may depend on the observations $\textbf{z}$.
-
-\subsection{Markov Chain Monte Carlo}
-\label{subsec:MCMC}
-
-There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler.
-The Gibbs sampler algorithm iteratively samples the components of a $k$dimensional parameter vector $\theta_i$ via each of the so-called full conditional distributions as follows,
-\begin{align}
-&p(\theta_{i, 1} | \theta_{i, 2}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
-&p(\theta_{i, 2} | \theta_{i, 1}, \theta_{i, 3}, \dots, \theta_{i, k}, \textbf{z}) \nonumber \\
-&\vdots \nonumber \\
-&p(\theta_{i, k} | \theta_{i, 1}, \dots, \theta_{o, k-1}, \textbf{z}). \nonumber
-\end{align}
-Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\theta_i | \textbf{z})$.
-Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markov nature of the algorithm.
-The computation time for each iteration and the overall number of iterations required to accurately summarise the posterior distribution is problem specific and typically increases with the number of parameters in the model.
-The full conditional distributions cannot be recognised for each of the time series models proposed in Section \ref{sec:models} and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995}.
-\\
-
-In MH-MCMC, a candidate $\theta_{i, j}^{(c)}$, where $\theta_{i, j}$ may be any scalar or vector subset of $\theta_i$, is drawn from a proposal distribution $r(\theta_{i, j})$ and accepted by the sampler with probability
-\begin{equation}
-\min \left\{ 1, \frac{p(\theta_{i, j}^{(c)} | \theta_{i, l \neq j}^{(a)}, \textbf{z})}{p(\theta_{i, j}^{(a)} | \theta_{i, l \neq j}^{(a)}, \textbf{z})} \times \frac{r(\theta_{i, j}^{(a)} | \theta_{i, j}^{(c)})}{r(\theta_{i, j}^{(c)} | \theta_{i, j}^{(a)})} \right\},
-\label{MHaccept}
-\end{equation}
-where the superscript $(a)$ denotes the most recently accepted value of the corresponding subset of $\theta_i$.
-If the sampler rejects a candidate value the previous value $\theta_{i, j}^{(a)}$ is repeated.
-Each iteration of Metropolis-Hastings-within-Gibbs MCMC includes one candidate draw for each element of $\theta_i$.
-Choice of proposal distribution is left to the user with arguably the most simple being the Normal Random Walk proposal,
-\begin{equation}
-\theta_{i, j}^{(c)} \sim N(\theta_{i, j}^{(a)}, \Sigma_{i, j})
-\label{RWprop}
-\end{equation}
-as $r(\theta_{i, j}^{(c)} | \theta_{i, j}^{(a)}) = r(\theta_{i, j}^{(a)} | \theta_{i, j}^{(c)})$.
-\\
-
-Performance of Random Walk Metropolis Hastings MCMC (RWMH-MCMC) then depends on the acceptance rate of the proposal distribution, largely influenced by the value of $\Sigma_{i, j}$.
-\citet{Garthwaite2016} propose an adaptive algorithm that increases or decreases $\Sigma_{i, j}$ depending on whether the most recent candidate was rejected or accepted respectively so that a target acceptance rate is asymptotically approached as the number of MCMC iterations approaches infinity.
-
-\subsection{Variational Bayes}
-\label{subsec:VB}
-
-A typically faster, albeit approximate, alternative to MCMC based inference is VB \citep{Jordan1999}.
-VB posits a family of parametric approximating distributions $q_{\lambda}(\theta_i | \textbf{z})$, parameterised by an auxiliary vector $\lambda$, that share the same support as the true posterior distribution $p(\theta_i | \textbf{z})$.
-Note that the family $q_{\lambda}(\theta_i | \textbf{z})$ does not neccesarily depend on $\textbf{z}$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta_i | \textbf{z})$.
-A member of approximating family is chosen to minimise some error function, typically the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta_i | \textbf{z})$ to $p(\theta_i | \textbf{z})$, given by $KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z})]$ \citep{Kullback1951}.
-The KL divergence is defined by
-\begin{equation}
-\label{KL-def}
-KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z})] = E_{q_{\lambda}(\theta_i | \textbf{z})} \left[ \log(q_{\lambda}(\theta_i | \textbf{z})) - \log(p(\theta_i | \textbf{z})) \right],
-\end{equation}
-and is a non-negative, asymmetric measure of the discrepancy between $p(\theta_i | \textbf{z})$ and $q_{\lambda}(\theta_i | \textbf{z})$  that will be equal to zero if and only if $p(\theta_i | \textbf{z}) = q_{\lambda}(\theta_i | \textbf{z})$ almost everywhere \citep{Bishop2006}.
-\\
-
-Typically (\ref{KL-def}) cannot be evaluated, and Monte-Carlo estimates as
-\begin{equation}
-\label{KL-MC}
-KL[q_{\lambda}(\theta_i | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta_i | \textbf{z}] \approx \frac{1}{M}\sum_{j=1}^M \left(\log(q_{\lambda}(\theta_{j, i} | \textbf{z})) - \log(p(\theta_{j, i} | \textbf{z})) \right)
-\end{equation}
-where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z}))$ are compuationally infeasible due to the inclusion of the term $p(\theta_i | \textbf{z})$, which is only known up to proportionality.
-Instead VB uses the Evidence Lower Bound (ELBO), denoted by $\mathcal{L}(q, \lambda)$, as an error function where
-\begin{equation}
-\label{ELBO}
-\mathcal{L}(q, \lambda) = E_{q_{\lambda}(\theta_i | \textbf{z})} \left[\log(p(\theta_i, \textbf{z})) - \log(q_{\lambda}(\theta_i | \textbf{z}))\right],
-\end{equation}
-which is evaluated with Monte-Carlo estimates
-\begin{equation}
-\label{ELBO-MC}
-\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right)
-\end{equation}
-where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z}))$.
-The ELBO is equal to the negative KL divergence plus a constant, and hence maximising (\ref{ELBO}) with respect to $q_{\lambda}(\theta_i | \textbf{z})$ is equivalent to minimising (\ref{KL-def}).
-
-\subsection{Stochastic Gradient Ascent}
-\label{subsec:SGA}
-For exponential family likelihood models, $p$, with a factorisable approximation, $q$, the characteristics of the surface of the ELBO can be exploited for optimisation in Mean Field Variational Bayes \citep{Ghahramani2000, Wainwright2008}, but for more general distributions the surface of the ELBO and its stochastic estimate are unknown.
-Maximisation proceeds by optimising only the auxiliary parameters $\lambda$ for a fixed distribution family $q$ with stochastic gradient ascent (SGA).
-\\
-
-SGA repeatedly takes Monte-Carlo estimates of the gradient of the ELBO with respect to $\lambda$, $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, as $\widehat{\delta\mathcal{L}(q, \lambda) / \delta \lambda}$ and applies updates of the form
-\begin{equation}
-\label{gradientAscent}
-\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
-\end{equation}
-until the change from $\mathcal{L}(q, \lambda^{(m)})$ to $\mathcal{L}(q, \lambda^{(m+1)})$ falls below some pre-specified threshold \citep{Hoffman2013}.
-Intuitively, individual elements of $\lambda$ will increase if the estimate of the slope of $\mathcal{L}(q, \lambda^{(m)})$ is positive at the current point, and will decrease if that estimate is negative, until each element of $\lambda$ reaches a point where the slope is zero.
-This procedure is guaranteed to converge to a local maximum \citep{Robbins1951} if the sequence $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
-\begin{align}
-&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
-&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
-\end{align}
-In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of \citet{Kingma2015b}.
-\\
-
-There are two popular choices for the Monte Carlo estimator of $\delta\mathcal{L}(q, \lambda) / \delta \lambda$, the score estimator of \citet{Ranganath2014}, 
-\begin{equation}
-\label{scoreDeriv}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{j, i} | \textbf{z}))}{\delta \lambda} \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right),
-\end{equation}
-where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z})$ and the reparameteterised estimator of \citet{Kingma2014}.
-Reparameterisation introduces an auxiliary variable $\epsilon$ and differentiable function $f(\cdot, \cdot)$ to rephrase Variational Bayes optimisation as the equivalent search for the parameters $\lambda$ that minimises the Kullback Leibler divergence from some distribution $q(\epsilon)$ with zero free parameters to the posterior distribution implied by the transformation $\theta_i = f(\epsilon, \lambda)$:
-\begin{equation}
-\label{rpDist}
-p(f(\epsilon, \lambda) | z) = p(\theta_i | \textbf{z}) |J^{-1}(f(\epsilon, \lambda))|
-\end{equation}
-where $J(f(\epsilon, \lambda))$ is the Jacobian Matrix of the transformation $f(\epsilon, \lambda)$.
-Examples of $f$ and $q(\epsilon)$ include treating $\theta_i$ as location scale transformation from a standard normal $\epsilon$, or an inverse-CDF transformation from a uniform$(0, 1)$ $\epsilon$.
-\\
-
-The ELBO can be reparameterised by substituting $p(\theta_i, \textbf{z}) = p(f(\epsilon, \lambda), z)|J(f(\epsilon, \lambda))|$ into (\ref{ELBO}),
-\begin{equation}
-\label{rpELBO}
-\mathcal{L}(q, \lambda) = E_{r(\epsilon)} \bigg[\log(p(f(\epsilon,\lambda), \textbf{z})|J(f(\epsilon, \lambda))|) - \log(q(\epsilon))\bigg].
-\end{equation}
-The gradient of the reparameterised ELBO with respect to $\lambda$ is given by 
-\begin{align}
-\label{rpELBODeriv}
-\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda} &= \frac{\delta}{\delta \lambda} \bigg( E_{q(\epsilon)} \bigg[\log\big(p(f(\epsilon,\lambda), \textbf{z})|J(f(\epsilon, \lambda))|\big) - \log(q(\epsilon))\bigg] \bigg) \nonumber \\
-&= E_{q(\epsilon)} \left[ \frac{\delta}{\delta \lambda} \bigg(\log(p(f(\epsilon,\lambda), \textbf{z})) + \log(|J(f(\epsilon, \lambda))|) - \log(q(\epsilon)) \bigg)\right] \nonumber \\
-&= E_{q(\epsilon)} \left[ \frac{\delta \log(p(f(\epsilon,\lambda), \textbf{z}))}{\delta f(\epsilon,\lambda)} \frac{\delta f(\epsilon,\lambda)}{\delta \lambda}  + \frac{\delta \log(|J(f(\epsilon, \lambda))|)}{\delta \lambda} \right].
-\end{align}
-This form leads to the reparameterised gradient estimator,
-\begin{equation}
-\label{rpDeriv}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta f(\lambda, \epsilon_j)}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda},
-\end{equation}
-where $\epsilon_j \sim q(\epsilon)$.
-The reparameterised gradient estimator typically has lower variance than the score estimator (see eg.
-\cite{Rezende2014}; \cite{Ruiz2016}), but treating $q_{\lambda}(\theta_i | \textbf{z})$ as the distribution implied by the transformation $f$ of $q(\epsilon)$ restricts the class of approximating families that can be used.
-\\
-
-Choosing the number of samples per estimate, $M$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate.
-An increased value of $M$ will reduce the estimator variance, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
-To reduce the variance of the Monte Carlo estimator, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) which has shown to be more efficient, where numbers in the unit hypercube are generated according to a Sobol Sequence \citep{Sobol1967} which are then randomised using the scrambled net method \citep{Matousek1998}, and transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$.
-Use of RQMC has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
-
+\begin{frame}
 \section{Updating Variational Bayes}
-\label{sec:UVB}
-
-The self-driving vehicle is constantly observing data about the movements of the surrounding vehicles which can be used for posterior inference.
-Figure \ref{fig:timeUpdate} illustrates this for a vehicle in the NGSIM dataset travelling towards the right.
-In the left panel, the vehicle has been observed until time $S$, and this data $z_{i, 1:S}$ can be incorporated into the Variational Bayes posterior approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, suppressing the additional conditional dependence on $\textbf{z}_{1:N}$.
-In the right panel, at some later time $T > S$, an additional $T - S$ data points have been observed, and inference of the behaviour of this vehicle, and hence forecasts, could be improved by incorporating the information contained in $z_{i, S+1:T}$ to form the posterior approximation $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$.
-Note that the $S$ and $T$ subscripts on $\lambda$ are introduced to differentiate the auxiliary parameter vector conditioned on data up to times $S$ and $T$.
-To facilitate this posterior update, an Updating Variational Bayes (UVB) mechanism is introduced where only $z_{i, S+1:T}$ needs to be processed.
+\frametitle{Updating Variational Bayes}
+\begin{itemize}
+\item Motivation
 \begin{figure}[ht]
@@
 \includegraphics[width = 0.95\textwidth]{timeUpdate}
-\caption{Left: The path of a vehicle in the NGSIM dataset at time $S$, where the direction of travel is from the left to the right.
-Right: The same vehicle at a later time $T > S$, with the extra $T - S$ observations denoted by the dashed line.
-Inference about this vehicle at time $S$ can be updated by the inclusion of the additional information observed up to time $T$.}
-\label{fig:timeUpdate}
 \end{figure}
-\\
+\item Notation: $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$ and $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$
+\end{itemize}
+\end{frame}
 
-Given $p(\theta_{i} | z_{i, 1:S})$ from (\ref{posterior}), the posterior distribution at time $T$, $p(\theta_{i} | z_{i, 1:T})$ is given by Bayes rule as
+\begin{frame}
+\frametitle{Approximating Bayesian Updating}
 \begin{equation}
@@
 \end{equation}
-Standard methods to evaluate the posterior, such as Variational Bayes or Markov Chain Monte Carlo, require the evaluation of the right hand side of (\ref{updatePost}), however computation of $p(\theta_{i} | z_{i, 1:S})$ is often infeasible.
-Without being able to evalute the right hand side, posterior inference requires evaluation of the prior distribution and full sample $z_{i, 1:T}$.
-To avoid this, UVB replaces $p(\theta_{i} | z_{i, 1:S})$ with the analytical approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, leading to the approximate joint distribution
+\begin{itemize}
+\item This only works if the right hand side can be evaluated
+\item VB Gradient Estimators requires the joint distribution
+\begin{equation}
+\label{updateJoint}
+p(\theta_{i}, z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S})
+\end{equation}
+\item We will replace it with this
 \begin{equation}
 \label{ApproxJoint}
-\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S})
+\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S}).
 \end{equation}
-\\
+\item Sources of error: How close $q$ is to $p$
+\end{itemize}
+\end{frame}
 
-The ELBO gradient estimators to construct the updated Variational Bayes approximation at time $T$,  $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$, can be obtained by substituting (\ref{ApproxJoint}) into the score gradient estimator (\ref{scoreDeriv}) or the reparameterised gradient estimator (\ref{rpDeriv}).
-The updating score estimator is given by
+\begin{frame}
+\frametitle{Updating Gradient Estimators}
+The updating score estimator:
 \begin{align}
 \widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
-&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i} z_{i, 1:T})) \right) \label{scoreUpdate}
+&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T})) \right) \label{scoreUpdate}
 \end{align}
-where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. Similarly, the updating reparameterised estimator is given by
+where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. 
+The updating reparameterised estimator:
 \begin{equation}
@@
 \end{equation}
-where $\epsilon_j \sim r(\epsilon)$.
-If this process is repeated periodically, $S$ will grow at the same rate as $T$, and updating VB is $O(1)$ rather than $O(T)$ as required by re-using the full sample without the approximation step in (\ref{ApproxJoint}).
+where $\epsilon_j \sim q(\epsilon)$.
+\end{frame}
 
-\section{Forecast Evaluation}
-\label{sec:eval}
+\begin{frame}
+\section{Online Heterogeneous Forecasting: Results}
+\frametitle{Point Forecasts}
+\begin{itemize}
+\item Introduce Naive / RNN models?
+\item Some stuff about point forecasts: Both pictures?
+\item Note that, on average, the homogenous model is as good as others
+\item Some stuff about different inference frameworks
+\item MAPs are not the whole story
+\end{itemize}
+\end{frame}
 
-After obtaining the posterior distributions for each model conditioned on 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on a further 1387 vehicles.
-For each of these vehicles $i > 2000$, at each point in time $T = 100, 110, \dots, 450$ and forecast horizon $h = 1, \dots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model, while the time-series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$ and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density.
-As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement after every second of observing the vehicle.
-The forecast density for the homogenous model is created using the MCMC samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ through three methods:
-\begin{enumerate}
-\item Posterior sampling with RWMH-MCMC,
-\item VB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ to the complete $z_{i, 1:T}$ using the original prior distribution,
-\item UVB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ using $q_{\lambda_S}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:S})$ and data $z_{i, S+1:T}$ as described in Section \ref{sec:UVB}.
-\end{enumerate}
-Approximating distributions for VB are chosen to match the form of the prior distributions, multivariate normal $q$ for the IH model and a mixture of multivariate normals for the CH model.
-A summary of these approaches is provided in Table \ref{tableAlg}.
-Point estimate forecasts are evaluated by their Euclidean Error,
-\begin{equation}
-\mbox{EE}_{i, T, h} = \sqrt{\left(\hat{x}^*_{i, T+h} - x^*_{i, T+h} \right)^2 + \left(\hat{y}^*_{i, T+h} - y^*_{i, T+h} \right)^2},
-\label{eucError}
-\end{equation}
-where $\{\hat{x}^*_{i, T+h}, \hat{y}^*_{i, T+h}\}$ is a forecast of $\{x^*_{i, T+h}, y^*_{i, T+h}\}$, while density forecasts are evalutated by their logscore,
-\begin{equation}
-\mbox{LS}_{i, T, h} = \log \left(p\left(x^*_{i, T+h}, y^*_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T} \right) \right).
-\label{logscore}
-\end{equation}
+\begin{frame}
+\frametitle{Heterogeneous Variance Distributions}
+\begin{itemize}
+\item Some stuff about variance
+\item A note on erratic driver detection
+\item Two different pictures to go in here
+\end{itemize}
+\end{frame}
 
-\begin{center}
-\begin{table}[ht]
-\resizebox{\textwidth}{!}{%
-\begin{tabular}{| l | c | c |}
-\hline
-& Independent Heterogenous Model & Clustered Heterogenous Model \\
-\hline
-MCMC & \multicolumn{2}{c|}{RWMH-MCMC jointly drawing the entire $\theta$ vector using \citet{Garthwaite2016} to control the} \\
-& \multicolumn{2}{c|}{variance matrix of the Multivariate Gaussian proposal distribution to obtain a 23.4\% acceptance rate for $\theta$ draws.} \\
-\hline
-VB  & Variational Bayes using (\ref{rpDeriv}) to update parameters & Variational Bayes using (\ref{scoreDeriv}) to update parameters\\
-Standard &  of a Multivariate Gaussian approximation with &  of an approximation formed as a six component mixture\\
-&non-zero covariance using $M = 25$, where $f$ is a &  of diagonal variance Multivariate Gaussians using $M = 50$. \\
-&location scale transform from a standard normal $r(\epsilon)$. &\\
-\hline
-VB  & Variational Bayes fit to $100$ observations using (\ref{rpDeriv}), & Variational Bayes fit to $100$ observations using (\ref{scoreDeriv}), \\
- &  then updated every $10$ observations using (\ref{rpUpdate}). & then udpated every $10$ observations using using (\ref{scoreUpdate}). \\
-Standard &  $q$ is a Multivariate Gaussian approximation with &  $q$ is a six component mixture\\
-&non-zero covariance using $M = 25$, where $f$ is a&  of diagonal variance Multivariate Gaussians using $M = 50$. \\
-&location scale transform from a standard normal $r(\epsilon)$. &\\
-\hline
-\end{tabular}}
-\caption{Details of the algorithms used to produce posterior distributions for each method and prior combination}
-\label{tableAlg}
-\end{table}
-\end{center}
+\begin{frame}
+\frametitle{Updating Variational Bayes Results}
+\begin{itemize}
+\item Compare to regular VB
+\item Time Complexity
+\item Difference in logscores
+\end{itemize}
+\end{frame}
 
-\subsection{Naive Forecast Models}
-\label{subsec:Naive}
-
-Nine naive forecast models are provided where forecasts of $a_{i, t}$ and $\delta_{i, t}$ at time $T$ are constants equal to $\hat{a}_{i, T}$ and $\hat{\delta}_{i, T}$ for $t = T+1, \dots, T+H$.
-The classification of each naive model is provided in Table \ref{tableNaive} as a combination of one estimator for $\hat{a}_{i, T}$ and one estimator for $\hat{\delta}_{i, T}$.
-Acceleration estimators are the most recent acceleration, $\hat{a}_{i, T} = a_{i, T}$, the average acceleration over the most recent second, $\hat{a}_{i, T} = 0.1 \sum_{s = T-9}^T a_{i, s}$, and zero acceleration $\hat{a}_{i, T} = 0$, equivalent to constant velocity.
-Similarly, steering angle estimators are the most recent angle, $\hat{\delta}_{i, T} = \delta_{i, T}$, the average angle over the most recent second, $\hat{\delta}_{i, T} = 0.1 \sum_{s = T - 9}^T \delta_{i, s}$, and $\hat{\delta}_{i, T} = \pi/2$, corresponding to the vehicle to driving directly forward.
-\begin{table}
-\begin{center}
-\begin{tabular}{|l|c|c|c|}
-\hline
-& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 1/10 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
-\hline
-$\hat{a}_{i, T} = a_{i, T}$ & Naive 1 & Naive 2 & Naive 3 \\
-$\hat{a}_{i, T} = 1/10\sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
-$\hat{a}_{i, T} = 0$ & Naive 7 & Naive 8 & Naive 9 \\
-\hline
-\end{tabular}
-\end{center}
-\caption{Classification of naive forecast models by future velocity and steering angle estimators.}
-\label{tableNaive}
-\end{table}
-
-\subsection{Recurrent Neural Networks}
-\label{subsec:RNN}
-A Recurrent Neural Network (RNN) is constructed with inputs following \citet{Ding2013}, the ten most recent lags of vehicle position, angle, velocity, and time headway to the preceding vehicle.
-The RNN outputs the next thirty changes in the vehicle position, is trained through gradient descent using the Adam optimiser to minimise the euclidean error of the three second ahead point forecast corresponding to $EE_{i, T, 30}$ defined in (\ref{eucError}).
-As of writing this report forecast results from this model are not competitive with the naive models and are not discussed further.
-
-\subsection{Results}
-\label{subsec:Results}
-
-Eucliedean Error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
-Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error.
-Zero acceleration naive models (Naive 7, 8, and 9) perform the best, followed by the averaged acceleration models (Naive 4, 5, and 6) and finally the current acceleration models (Naive 1, 2, and 3).
-At all forecast horizons the error for each implementation of each of the time series models are significantly lower than every naive model.
-\\
-
-Figure (\ref{fig:PredErrorZ}) truncates Figure (\ref{fig:PredError}) to include only the time series models, and with the exception of the IH model with UVB inference, each time series model and inference implementation produces a forecast with a similar amount of error.
-This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models.
-As the UVB implementation of the CH model is competitive with MCMC and standard VB inference discussion will focus on the CH model rather than the IH model.
-\\
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.8\textwidth]{predictiveError.png}
-\caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$ for $h \in \{10, 20, 30\}$ and $T \in \{100, 110, \dots, 450\}$ for each model.
-Each time series model is significantly more accurate than the naive constant acceleration / angle models, with little discernable difference between the time series models.}
-\label{fig:PredError}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.8\textwidth]{predictiveErrorZoom.png}
-\caption{The same results as Figure \ref{fig:PredError} with the Naive models excluded.
-Most of the time series model have very similar MAP forecast accuracy, with the UVB / Independent time series combination performing slightly weaker}
-\label{fig:PredErrorZ}
-\end{figure}
-
-There are systematic differences in the level of certainty in forecasts provided by the CH and homogenous models as measured by their logscores, resulting from the CH model being able to to assign each vehicle individual variance parameters,  $\sigma^2_{\epsilon, i}$ for $a_{i, t}$, and $\sigma^2_{\eta, i}$ for $\delta_{i, t}$.
-Figure (\ref{fig:posVarMean}) plots the range of posterior means for these parameters from with RWMH-MCMC sampling with $T = 450$, with the homogenous model posterior means indicated with vertical red lines.
-There is a large amount of positive skewness in the distribution of variances, which may have caused the homogenous model to estimate a variance that is too large for most individual vehicles.
-The homogenous variance posterior means are greater than $80\%$ and $91.5\%$ of CH variance posterior means for acceleration and the steering angle respectivley.
-This large variance has little impact on the location of the MAP of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but impacts the level of uncertainty in the forecast density.
-Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the CH model with UVB inference over the homogenous model at the three second forecast horizon.
-Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis.
-The CH model trajectory forecasts for drivers in the lower variance quintiles benefit the most from including heterogeneity.
-However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles.
-This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance estimates allow identification of vehicles with high variance, and potentially more erratic and dangerous driving styles.
-\\
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.95\textwidth]{posVarMean}
-\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by Metropolis-Hastings MCMC at $T = 450$ using the CH model, compared to the homogenous posterior mean in red.
-Small amounts of cars with high variance has skewed the homogenous estimates to be too large for the majority of the vehicles forecasted.}
-\label{fig:posVarMean}
-\end{figure}
-
-\begin{figure}[ht]
-\centering
-\includegraphics[width = 0.95\textwidth, height = 0.4\textheight]{varSplit}
-\caption{Difference in logscore between the CH model with posterior inference using UVB and the homogenous model.
-Results are split by variance quintiles for acceleration (x-axis) and angle (y-axis).
-There are substantial improvements in logscore, and thus forecast certainty, for vehicles with a low variance in either variable.
-Vehicles in the fifth quintiles, close to the homogenous variance estimates, benefit instead from the higher precision associated with the homogenous posterior distribution due to the large $N$.}
-\label{fig:varSplit}
-\end{figure}
-
-The benefits of heterogenous modelling is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles.
-Figure (\ref{fig:timing}) demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$.
-The compuation time for standard VB increases linearly with $T$, while UVB is constant; posterior inference available at time $T$ utilising UVB can be conditioned on more data than standard VB.
-Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time and make UVB feasible for more complex models.
-\\
-
-As UVB involves replacing the posterior at time $S$ with its variational approximation, the error introduced by UVB depends on the approximation error of Variational Bayes itself.
-The left panel for Figure (\ref{fig:updateCost}) displays boxplots of UVB logscores minus standard VB logscores at the three second forecast horizon for the CH model, and demonstrates that UVB does not typically lead to a decrease in the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
-The right panel displays analogous boxplots for the IH model, where the UVB logscores tend to be lower than the standard VB logscores, matching the increased Euclidean error found for the IH model with UVB inference in Figure \ref{fig:PredErrorZ}.
-An explanation for this is given by the form of the approximating distribution used for the IH model, the multivariate normal distribution may not adequately approximate the posterior distribution, while the extra flexibilitiy of the mixture multivariate normal distribution for the CH model was sufficient to replace the posterior.
-
-\begin{figure}[htp]
-\centering
-\includegraphics[width = 0.75\textwidth]{timing}
-\caption{Average time to converge for Standard VB (Black) and UVB (Coloured) with four different update lengths given by $T - S$, truncated to 25 seconds.
-Each UVB procedure originally fits the model to the first 100 data points and updates this every $T - S$ data points.
-The convergence time for Standard VB increased linearly for $T$, but is constant for UVB.
-Each VB algorithm is ran with $M = 50$ on one CPU core.
-Convergence time could be reduced through parallelisation.}
-\label{fig:timing}
-\end{figure}
-\begin{figure}[htp]
-\centering
-\includegraphics[width = 0.75\textwidth, height = 0.3\textheight]{updateCost}
-\caption{Differences in predictive logscores for $p(z_{i, T+30} | z_{i, 1:T})$ for each $T = 100, 110, \dots, 450$ between UVB and standard VB for the CH model (left) and the IH model (right).
-The differences are typically small for the CH model, while the IH model approximation has increased the logscore relative to standard VB.
-This may be due to the form of the approximating distribution for the IH model being an inadequate approximation for the true posterior}
-\label{fig:updateCost}
-\end{figure}
-
-\newpage
-
-\section{Discussion}
-\label{sec:disc}
-This paper demonstrates that the use of auto-regressive models time-series models drastically improves trajectory forecasts when compared to the competing models.
-These models parameterise the dynamics of a driver's actions, with a range of parameter posterior means providing evidence of heterogeneity between the drivers of different vehicles.
-Incorpororating this heterogeneity through a hierarchical model has improved the forecast logscores, and hence reduced uncertainty, but requires an approximate inference scheme implemented with UVB.
-\\
-
-We find that the additional approximation error introduced by UVB does not materialise as a reduction in forecasting performance relative to exact MCMC inference when an approximating distribution that has sufficient flexibilitity to approximate the true posterior distribution is used.
-Augmenting self-driving vehicles with hierarchical time series models to forecast the trajectory of surrounding traffic, and repeatedly updating the posterior distribution and forecasts, can be feasibly implemented with UVB.
-\\
-
-The framework used by the trajectory forecasting problem is to construct a hierarchical model to infer the range of heterogeneity present in a dataset, and use exact inference results provided by MCMC to construct a prior distribution for an additional unit in the hierarchy.
-Our use of this framework has focused on situations where exact inference on the additional unit is unavailble in the time-frame required by the problem, instead using Variational Bayes to infer the posterior distribution of this additional unit.
-We have introduced UVB to allow this inference to happen in the online setting, and have explored fitting copula models to the MCMC samples to produce a functional form for the approximating distribution that better matches the true posterior, though have not been able to link the copula structure for the units in the original hierarchy to the new unit.
-This framework is not specific to vehicle forecasting, and could conceivably be used for many other problems such as classification, or anomoly detection by extending the hierarchical model with a dirichlet process prior.
-\\
-
-\newpage
-\bibliographystyle{asa}
-\bibliography{references}
-
-\appendix
-\section{Transformation to Relative Coordinates}
-
-The centre line for each of the five lanes is estimated separately, using 100 randomly sampled vehicles per lane that did not change their lane over the observed period. 
-For each vehicle $i$ and time $t$ since entering the road, with travel originating at time one given by $\{x_{i,1}, y_{i,1}\}$, the total distance travelled is calcualted as 
-\begin{equation}
-\label{distance}
-d_{i, t} = \sqrt{(x_{i, t} - x_{i, 1})^2 + (y_{i, t} - y_{i, 1})^2}.
-\end{equation}
-Using this distance metric and the data from the sampled 100 cars per lane, the two-dimensional coordinates corresponding to the centre line of each lane are estimated via independent smoothing splines where each coordinate is a function of the distance travelled to that point.
-Each smoothing spline is calculated using the `R stats' package \citep{R}.
-The estimated centre line for lane $k$, is denoted by the curve $\{(\hat{x}_{d,k} = f_k(d), \hat{y}_{d,k} = g_k(d)\}$, for $d \geq 0$.
-\\
-
-Excluding the vehicles used to estimate the spline model, each of the 3387 remaining vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
-\begin{align}
-x^*_{i, t} &= \mbox{sign}\left(\tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right) - \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)\right)\sqrt{(x_{i, t}-\hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2)} \label{xRel} \\
-y^*_{i, t} &= d_{i, t}. \label{yRel}
-\end{align}
-\\
-
-\begin{figure}
-\centering
-\includegraphics[width = 0.5\textwidth]{relCoord}
-\caption{An example of the relative coordinate transformation for a vehicle as described by equation (\ref{xRel}).
-The estimated trajectory of the lane midpoint is given by the solid curve, while a vehicle $i$ at a point in time $t$ is denoted by the blue dot at $\{x_{i, t}, y_{i, t}\}$.
-This vehicle has travelled a distance equivalent to the red dot at $\{\hat{x}_{i, t}, \hat{y}_{i, t}\}$.
-The gradient of the midpoint at this point is given by the dashed line, which intercepts the x-axis at an angle of $\lambda = \tan^{-1}\left(\frac{g'(d_{i, t}) }{f'(d_{i, t})}\right)$.
-The dotted line travels from the blue dot to the red, and intercepts the x-axis at an angle of $\psi = \tan^{-1}\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)$.
-The dashed line intercepts the dotted line with an angle  of $\lambda - \psi$, the sign of which determines whether the vehicle is to the left or right side of the lane centre, and thus has a negative or positive relative coordinate $x^*_{i, t}$.
-The absolute value of $x^*_{i, t}$ is the distance between the two dots.}
-\label{fig:relCoord}
-\end{figure}
+\begin{frame}
+\section{Extensions}
+\frametitle{Extending the Framework}
+\begin{itemize}
+\item Copula Project
+\item Other Applications
+\end{itemize}
+\end{frame}
 

### 2018-03-08 Cars/paper.tex
presentation

@@
 \usepackage{amsthm}
+\usepackage[]{algorithm2e}
 %\usepackage{tikz}
@@
 \setlength{\topmargin}{0in}
-\setlength{\oddsidemargin}{0.1in}
-\setlength{\evensidemargin}{0.1in}
+\setlength{\oddsidemargin}{0.08in}
+\setlength{\evensidemargin}{0.08in}
 \setlength{\textwidth}{6.5in}
-\setlength{\textheight}{8.25in}
+\setlength{\textheight}{9in}
 \setlength\parindent{0pt}
@@
 
-\title{Online Updating of Variational Bayes for Heterogenous Forecasts of Vehicle Trajectory}
-\author{Nathaniel Tomasetti}
-\date{}
-
-%TODO: Add a conclusion
+\title{Online Updating of Variational Bayes for Heterogeneous Forecasts of Vehicle Trajectory}
+\author{Nathaniel Tomasetti 
+\and Supervised by Catherine Forbes and Anastasios Panagiotelis}
 
@@
 
-While characteristics of surrounding vehicles, such as theirr postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g.
-\citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory has seen less attention in the literature.
-These forecasts often assume the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
+While characteristics of surrounding vehicles, such as their postition, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g.
+\citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory of these vehicles has seen less attention in the literature.
+These forecasts often assume that the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017}.
 \\
@@
 
-We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the contstant models commonly used, and including driver heterogeneity in these time series models substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous approach.
+We find that the use of time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the constant models commonly used, and including driver heterogeneity in these time series models substantially reduces the uncertainty in the entire forecast distribution relative to a homogeneous approach.
 These results give credence to the validity of the methodological framework developed in this thesis, where exact Bayesian inference techniques are used to guide the construction of a prior and approximating distribution for parameters that are inferred in an online setting with Variational Bayes.
@@
 Data is provided by the Next Generation Simulation (NGSIM) project conducted by the US Federal Highway Administration (FHWA), which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005.
-Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc to produce coordinates for each vehicle and point in time relative to the start of the road section.
+Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc.
+to produce coordinates for each vehicle and point in time relative to the start of the road section.
 Figure \ref{fig:rawData} shows the paths of ten vehicles in the dataset, which traveled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1.
@@
 
-Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road compensating for any curvature.
+Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road, compensating for any curvature.
 To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self-driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates.
-This is facilitated by measuring the location of vehicles relative to an estimate of the midpoint of their associated lane.
+This is facilitated by measuring the location of vehicles relative to an estimate of the centre line of their associated lane.
 \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway.
@@
 
-Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to observation $t$ follows friom the trigonometric relationship in Figure \ref{fig:motion}:
+Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to $t$ follows from the trigonometric relationship in Figure \ref{fig:motion}:
 \begin{align}
-x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq} \\
-y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq},
+x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{xEq}, \\
+y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{yEq}.
 \end{align}
@@
 
-From this relationship, and the coordinate sequences $\{x^*_{i, s}, \mbox{ for }s=1,2,...,T\}$ and $\{y^*_{i, s}, s=1,2,...,T\}$, the inputs to motion from driver $i$ are calculated via
+From this relationship, and the coordinate sequences $\{x^*_{i, s} | s=1,2,...,T\}$ and $\{y^*_{i, s} | s=1,2,...,T\}$, the inputs to motion from driver $i$ are calculated via
 \begin{align}
@@
 Data resulting from the observation of vehicle $i$ for a period of time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \dots, T\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T} | i = 1, \dots, N\}$.
-Some additional vehicles are excluded due to an insufficient number of observations, leaving 3387 vehicles in the dataset in a relative coordinate system.
+Some additional vehicles are excluded if they have less than 500 observations, leaving 3387 vehicles in the dataset transformed into a relative coordinate system.
 These vehicles are split into two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1387 vehicles.
@@
 The training set in the left panel represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles.
-The test set in the right panel represents those vehicles the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.
+The test set in the right panel represents those vehicles that the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.
 
@@
 Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between the different sets of vehicles.
-These are referred to as the homogenous model, under which all drivers are treated identically and the training set is used to infer all behaviour and two heterogenous models: the independent heterogenous (IH) model and the hierarchical, clustered heterogeneity (CH) model.
-The ID model does not incorporate information from the vehicles in the training into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set.
+These are referred to as the homogeneous model, under which all drivers are treated identically and the training set is used to infer all behaviour, and two heterogeneous models: the independent heterogeneous (IH) model and the hierarchical, clustered heterogeneity (CH) model.
+The IH model does not incorporate information from the vehicles in the training into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set.
 \\
 
-\subsection{A Homogenous Time Series Model}
-\label{subsec:homogenous}
+\subsection{A Homogeneous Time Series Model}
+\label{subsec:homogeneous}
 
@@
 
-NGSIM vehicle data is observed every 100 milliseconds so large changes in $a_{i, t}$ and $\delta_{i, t}$ may occur over multiple sequential observations.
+NGSIM vehicle data is observed every 100 milliseconds, so large changes in $a_{i, t}$ and $\delta_{i, t}$ may occur over multiple consecutive observations.
 This is highlighted in Figure \ref{fig:pacf}, which displays a plot of the partial autocorrelation function, which measures the correlation between a variable at two different time periods, $t$ and $t + k$, conditioned on that variable at time $t + 1, \dots t + k -1$, for both $a_{i, t}$ and $\delta_{i, t}$ for five vehicles.
-Large spikes in the partial autocorrelation plots, such as those at the first and second lag for most vehicles, demonstrate that the value of either $a_{i, t}$ or $\delta_{i, t}$ is strongly correlated with the previous time periods.
+Large spikes in the partial autocorrelation plots, such as those at the first lag for most vehicles, demonstrate that the value of either $a_{i, t}$ or $\delta_{i, t}$ is strongly correlated with the previous time period.
 \\
@@
 
-These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and as an attempt to model the dependence, two univariate auto-regressive processes of orders $p$ and $q$, respectively, for $a_t$ and angle  $\delta_t$ are employed, given by
+These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and as an attempt to model the dependence, two univariate auto-regressive processes of orders $p$ and $q$, respectively, for $a_t$ and $\delta_t$ are employed, given by
 \begin{align}
@@
 for each $i = 1, \dots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution.
-The unconditional means for acceleration and steering and are set to zero and $\pi/2$ respectively, corresponding to forward motion at a constant velocity.
+The unconditional means for acceleration and steering and are set, respectively, to zero and $\pi/2$, corresponding to forward motion at a constant velocity.
 \\
 
-For notational convenience, and to facilitate implementation of the Bayesian inferential approach described in Section \ref{sec:Inference}, the variance parameters $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$ are transformed and collected together with all other model parameters into the $\mathbb{R}^{p + q + 2}$-dimensional vector
+For notational convenience, and to facilitate implementation of the Bayesian inferential approach described in Section \ref{sec:Inference}, the variance parameters $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$ are transformed and collected together with all other model parameters into the $p + q + 2$-dimensional vector
 \begin{equation*}
@@
 \label{indPrior}
-\theta \sim \mathcal{N}\left(\mu, \Sigma \right),
+\theta \sim \mathcal{N}\left(\mu, \Sigma \right)
 \end{equation}
 is selected with $\mu=(0_p^{\prime},0_q^{\prime},-5, -5)$ and $\Sigma = 10 \mathbb{I}_{p+q+2}$, where $0_r$ and $\mathbb{I}_r$ denote the $r$-dimensional zero vector and identity matrix, respectively, noting that the scale of changes in either acceleration or angle at the 100 millisecond time-scale are both very small.
-We refer to the joint specification in (\ref{aAR}) and (\ref{dAR}), along with the given normal prior distributional assumption, as the \textit{homogenous model} throughout the paper.
+We refer to the joint specification in (\ref{aAR}) and (\ref{dAR}), along with the given normal prior distributional assumption, as the \textit{homogeneous model} throughout the paper.
 \\
-Treating the first $\max(p, q)$ observations for each vehicle as deterministic, the posterior distribution for the homogenous model is given by
+Treating the first $\max(p, q)$ observations for each vehicle as deterministic, the posterior distribution for the homogeneous model is given by
 \begin{equation}
@@
 
-\subsection{Two Heterogenous Time Series Models}
-\label{subsec:heterogenous}
+\subsection{Two Heterogeneous Time Series Models}
+\label{subsec:heterogeneous}
 
-The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics of traffic surrounding them.
-To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogenous} (IH) model and a dependent \textit{clustered heterogeneous} (CH) model.
-In both cases, the shared Heterogeneity is allowed by replacing the shared $\theta$ parameter vector is replaced with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})\}$ corresponding to the parameters given in the model for car $i$ given in
+The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics surrounding traffic.
+To accommodate such variation, heterogeneity, we introduce two so-called \textit{heterogeneous models}, namely an \textit{independent heterogeneous} (IH) model and a dependent \textit{clustered heterogeneous} (CH) model.
+In both cases, the geterogeneity is allowed by replacing the shared $\theta$ parameter vector with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \dots, \phi_{i,p}, \gamma_{i,1}, \dots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})\}$ corresponding to the parameters given in the model for vehicle $i$ given in
 \begin{align}
@@
 \end{align}
-\\
 
-The hyperparameters $\beta$ are $\{\mu_j, \Sigma_j, \pi_j | j = 1, \dots, K\}$.
 Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$.
 
-To illustrate that the hierarchical CH model, with $K=6$ adequately captures the main features of the IH model, the top panel of Figure \ref{fig:HierSingleKDE} displays kernel density estimates of the posterior means, for each $\theta_i, i = 1, \dots, 2000$, resulting from the IH model.
+To illustrate that the hierarchical CH model with $K=6$ adequately captures the main features of the IH model, the top panel of Figure \ref{fig:HierSingleKDE} displays kernel density estimates of the posterior means, for each $\theta_i, i = 1, \dots, N$, resulting from the IH model.
 For this illustration, we set $p = q = 2$ and use $N=2000$ observations over times $t=1, 2, \ldots, T = 500$.
 These summaries each indicate a wide range of values for the $N$ posterior means corresponding to each model parameter, corresponding to heterogeneity across drivers in the sample.
-Notably, the posterior means for each of the variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the distribution of posterior means is multimodal for each of the $\phi_1$ and $\phi_2$ parameters.
+Notably, the posterior means for each of the variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the distribution of posterior means is multimodal for $\phi_1$ and $\phi_2$.
 
@@
 
-For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles in the training set, the vehicles where information is available for before any self-driving vehicle enters the road.
+For the remainder of this paper $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles in the training set: the vehicles where information is available for before any self-driving vehicle enters the road.
 Use of the subscript $i$ is reserved for some $i > N$, and refers to the attriubtes (data or a parameter vector) for an additional vehicle in the test set that may be encountered while the self-driving vehicle is on the road.
@@
 A crucial aspect of the application of the discussed time series models is through the relationship between the posterior distribution formed from observations relatting to the first $N$ vehicles and the parameter vector for the additional new vehicle, $\theta_{N+1}$.
-Under the homogenous model $\theta$ is shared across all vehicles and hence $\theta_{i} \equiv \theta$.
+Under the homogeneous model $\theta$ is shared across all vehicles and hence $\theta_{i} \equiv \theta$.
 In this case, if $N$ is large $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not vary much from $p(\theta | \textbf{z}_{1:N})$.
@@
 \includegraphics[width = 0.95\textwidth]{homogPosterior}
-\caption{The marginal posterior distributions $p(\theta | \textbf{z}_{1:N})$ for the homogenous model with $N = 2000$ and $p = q = 2$. Each variable has very high posterior precision.}
+\caption{The marginal posterior distributions $p(\theta | \textbf{z}_{1:N})$ for the homogeneous model with $N = 2000$ and $p = q = 2$. Each variable has very high posterior precision.}
 \label{fig:homogPosterior}
@@
 \end{equation}
-\\
 
@@
 \end{equation}
-\\
 
-Combining a fast inference procedure with the test vehicle posterior distributions implied by either of the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogenous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.
+Combining a fast inference procedure with the test vehicle posterior distributions implied by either of the IH and CH models, (\ref{indepNewCar}) and (\ref{hierNewCar2}), will allow online, heterogeneous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.
 
@@
 \end{equation}
-To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogenous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewCar2}) for the CH model must first be inferred.
+To obtain this distribution, the posterior density for $\theta_i$, given by (\ref{homogPost}) for the homogeneous model, (\ref{indepNewCar}) for the IH model and (\ref{hierNewCar2}) for the CH model must first be inferred.
 However these are only known up to proportionallity, and hence the analytical solution to (\ref{predictive}) is unavailable.
@@
 
-In MH-MCMC, a candidate $\theta_{i, j}^{(c)}$, where $\theta_{i, j}$ may be any scalar or vector subset of $\theta_i$, is drawn from a proposal distribution $r(\theta_{i, j})$ and accepted by the sampler with probability
+In MH-MCMC, the Gibbs iterates are replaced by sampling a candidate $\theta_{i, j}^{(c)}$, where $\theta_{i, j}$ may be any scalar or vector subset of $\theta_i$, is drawn from a proposal distribution $r(\theta_{i, j})$ and accepted by the Markov Chain with probability
 \begin{equation}
@@
 \end{equation}
-where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z}))$.
-The ELBO is equal to the negative KL divergence plus a constant, and hence maximising \ref{ELBO}) with respect to $q_{\lambda}(\theta_i | \textbf{z})$ is equivalent to minimising (\ref{KL-def}).
+where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z})$.
+The ELBO is equal to the negative KL divergence plus a constant, and hence maximising (\ref{ELBO}) with respect to $q_{\lambda}(\theta_i | \textbf{z})$ is equivalent to minimising (\ref{KL-def}).
 
@@
 Intuitively, individual elements of $\lambda$ will increase if the estimate of the slope of $\mathcal{L}(q, \lambda^{(m)})$ is positive at the current point, and will decrease if that estimate is negative, until each element of $\lambda$ reaches a point where the slope is zero.
-This procedure is guaranteed to converge to a local maximum \citep{Robbins1951} if the sequence $\rho^{(m)}, m = 1, \dots, \infty$ satisfies
+This procedure is guaranteed to converge to a local maximum \citep{Robbins1951} if the sequence $\rho^{(m)}, m = 1, \dots, \infty$, satisfies
 \begin{align}
@@
 \end{align}
-In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of \citet{Kingma2015b}.
+In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of \citet{Kingma2015b} which is described in the appendix.
 \\
@@
 \end{equation}
-where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z})$ and the reparameteterised estimator of \citet{Kingma2014}.
+where $\theta_{j, i} \sim q_{\lambda}(\theta_i | \textbf{z})$, and the reparameteterised estimator of \citet{Kingma2014}.
 Reparameterisation introduces an auxiliary variable $\epsilon$ and differentiable function $f(\cdot, \cdot)$ to rephrase Variational Bayes optimisation as the equivalent search for the parameters $\lambda$ that minimises the Kullback Leibler divergence from some distribution $q(\epsilon)$ with zero free parameters to the posterior distribution implied by the transformation $\theta_i = f(\epsilon, \lambda)$:
@@
 \label{rpDeriv}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta f(\lambda, \epsilon_j)}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda},
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta f(\lambda, \epsilon_j)}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(f(\lambda, \epsilon_j))}{\delta \lambda},
 \end{equation}
@@
 An increased value of $M$ will the Monte-Carlo error, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
-To reduce the Monte Carlo error, following \citet{Gunawan2017}, Randomised Quasi Monte Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
+To reduce the Monte Carlo error, following \citet{Gunawan2017}, Randomised Quasi Monte-Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte-Carlo in many applications (see \cite{Niederreiter1992, Caflisch1998}).
 In RQMC a deterministic sequence of numbers in the unit hypercube (of dimensionality equal to $\theta_i$) are generated evenly, in the sense that they have a low star discrepency, according to a Sobol Sequence \citep{Sobol1967}.
 These numbers are then randomised using the scrambled net method \citep{Matousek1998}, which preserves the low discrepency properties while inducing a central limit theorem.
-$M$ of the resulting numbers are transformed using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$ to simulate $M$ draws from these distributions.
+Selecting $M$ of the resulting coordinates in the unit hypercube, and transforming these using the inverse-CDF of $q_{\lambda}(\theta | z)$ or $r(\epsilon)$ simulates $M$ draws that can be used in either the score or reparameterised estimators to form RQMC estimates of the gradient of the ELBO.
 
@@
 \end{equation}
-Standard methods to evaluate the posterior, such as Variational Bayes or Markov Chain Monte Carlo, require the evaluation of the right hand side of (\ref{updatePost}), however computation of $p(\theta_{i} | z_{i, 1:S})$ is often infeasible.
+Standard methods to evaluate the posterior, such as VB or MCMC, require the evaluation of the right hand side of (\ref{updatePost}), however computation of $p(\theta_{i} | z_{i, 1:S})$ is often infeasible.
 Without being able to evalute the right hand side, posterior inference requires evaluation of the prior distribution and full sample $z_{i, 1:T}$.
-To avoid this, UVB replaces $p(\theta_{i} | z_{i, 1:S})$ with the analytical approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$, leading to the approximate joint distribution
+To avoid this, UVB replaces $p(\theta_{i} | z_{i, 1:S})$ with the analytical approximation $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$ leading to the approximate joint distribution
 \begin{equation}
 \label{ApproxJoint}
-\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S})
+\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S}).
 \end{equation}
-\\
 
@@
 where $\epsilon_j \sim r(\epsilon)$.
-If this process is repeated periodically, $S$ will grow at the same rate as $T$, and UVB has a computational complexity as $T$ increases of $O(1)$, rather than $O(T)$ as required by standard VB that must process $z_{i, 1:T}$ at each possible value of $T$.
+If this process is repeated periodically, $S$ will grow at the same rate as $T$, and UVB hence has a computational complexity as $T$ increases of $O(1)$, rather than $O(T)$ as required by standard VB, where the likelihood for $z_{i, 1:T}$ is evaluated.
 
@@
 As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement for every vehicle after every second of observation.
-The forecast density for the homogenous model is created using the samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
+The forecast density for the homogeneous model is created using the samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
 \begin{enumerate}
@@
 \hline
-& Independent Heterogenous Model & Clustered Heterogenous Model \\
+& Independent Heterogeneous Model & Clustered Heterogeneous Model \\
 \hline
@@
 Eucliedean error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:PredError}.
-Naive model performance is split into three groups according to the choice of acceleration forecast, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error.
+Naive model performance splits into three groups according to the choice of $\hat{a}_{i, T}$, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error.
 The best performing Naive model at each forecast horizon is Naive 8, however at all forecast horizons the error for each implementation of each of the time series models are significantly improved relative to every naive model.
@@
 Figure \ref{fig:PredErrorZ} truncates Figure \ref{fig:PredError} to include only the time series models, each of which produces a forecast with a similar amount of error that does not increase substantially as the forecast horizon increases.
-This implies that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogenous or introducing heterogeneity with the IH and CH models, however the UVB inference implementation of the IH model has increased the forecast error relative to MCMC and standard VB.
-The UVB implementation of the CH model produces forecast errors that are competitive with MCMC and standard VB, and so inference discussion will focus on the CH model rather than the IH model.
+The close performance between the homogeneous and heterogeneous models shows that there is no systematic increase or decrease in MAP forecast accuracy for choosing either the homogeneous model or introducing heterogeneity with the IH and CH models.
+UVB inference has increased the forecast error for the IH model relative to VB and MCMC inference, however the CH model produces similar forecast errors for each of MCMC, VB and UVB, and so inference discussion will focus on heterogeneity implemented with the CH model rather than the IH model.
 \\
@@
 \centering
-\includegraphics[width = 0.8\textwidth]{predictiveError.png}
+\includegraphics[width = 0.95\textwidth]{predictiveError.png}
 \caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$ for $h \in \{10, 20, 30\}$ and $T \in \{100, 110, \dots, 450\}$ for each model.
@@
 \centering
-\includegraphics[width = 0.8\textwidth]{predictiveErrorZoom.png}
+\includegraphics[width = 0.95\textwidth]{predictiveErrorZoom.png}
 \caption{The same results as Figure \ref{fig:PredError} with the Naive models excluded.
@@
 \hline
-Horizon & Naive 8 & Homogenous MCMC & IH MCMC & IH VB & IH UVB & CH MCMC & CH VB & CH UVB \\
+Horizon & Naive 8 & Homogeneous MCMC & IH MCMC & IH VB & IH UVB & CH MCMC & CH VB & CH UVB \\
 \hline
@@
 
-The CH model allows each vehicle in the test to have a posterior variance distribution that is unique to that vehicle, while the homogenous model must share the same posterior variance distribution across different vehicles.
-Differences in the distribution of either variance parameter has little impact on the location of the maximum of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but can drastically affect the precision of this forecast density, causing systematic differences in the level of uncertainty associated with the forecasts that can be measured by the logscore.
+The CH model allows each vehicle in the test to have a posterior variance distribution that is unique to that vehicle, while the homogeneous model must share the same posterior variance distribution across different vehicles.
+Differences in the distribution of either variance parameter has little impact on the location of the maximum of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but can substantially affect the precision of this forecast density.
+Systematic differences in the forecast uncertainty between the CH and homogeneous models result from this different treatment of variance.
 \\
@@
 Figure \ref{fig:posVarMean} displays plots of density of posterior means for each of acceleration and steering angle resulting from RWMH-MCMC sampling with $T = 450$.
-The homogenous model posterior means are indicated with vertical red lines.
-As in Figure \ref{fig:HierSingleKDE}, there is a large amount of variation in the distribution of variances for the test set, many of which are lower than the homogenous model: the homogenous posterior means are greater than $80\%$ and $91.5\%$ of CH posterior means for acceleration and the steering angle respectivley.
+The homogeneous model posterior means are indicated with vertical red lines.
+As in Figure \ref{fig:HierSingleKDE}, there is a large amount of variation in the distribution of variances for the test set, many of which are lower than the homogeneous model: the homogeneous posterior means are greater than $80\%$ and $91.5\%$ of CH posterior means for acceleration and the steering angle respectivley.
 \\
 
-Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the CH model with UVB inference over the homogenous model at the three second forecast horizon.
-Vehicles are split into bivariate quintiles according to their posterior variance means for acceleration along the x-axis and angle along the y-axis.
-The upper left panels correspond to vehicles in the lower posterior variance quintiles, showing clear benefits for CH model trajectory forecasts relative to the homogenous model.
+Figure \ref{fig:varSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using the CH model with UVB inference relative to the homogeneous model at the three second forecast horizon.
+Vehicles are split into bivariate quintiles according to their RWMH-MCMC posterior variance means for acceleration along the x-axis, and steering angle along the y-axis.
+The upper left panels correspond to vehicles in the lower posterior variance quintiles, showing clear benefits for CH model trajectory forecasts relative to the homogeneous model.
 However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles.
-This paper argues that heterogenous modelling is still useful in this scenario, as the individual vehicle variance distributions allow identification of vehicles with high variance, which could display potentially more erratic and dangerous behaviour.
+This paper argues that heterogeneous modelling is still useful in this scenario, as the individual vehicle variance distributions allow identification of vehicles with high variance which could display potentially more erratic behaviour.
 \\
@@
 \includegraphics[width = 0.95\textwidth]{posVarMean}
-\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by RWMH-MCMC at $T = 450$ using the CH model, compared to the homogenous posterior mean in red.
-Small amounts of cars with high variance in the training set has increased the homogenous variances.}
+\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by RWMH-MCMC at $T = 450$ using the CH model, compared to the homogeneous posterior mean in red.
+Small amounts of cars with high variance in the training set has increased the homogeneous variances.}
 \label{fig:posVarMean}
@@
 \centering
-\includegraphics[width = 0.95\textwidth, height = 0.4\textheight]{varSplit}
-\caption{Difference in logscore between the CH model with posterior inference using UVB and the homogenous model.
+\includegraphics[width = 0.95\textwidth, height = 0.55\textheight]{varSplit}
+\caption{Difference in logscore between the CH model with posterior inference using UVB and the homogeneous model.
 Results are split by variance quintiles for acceleration (x-axis) and angle (y-axis).
 There are substantial improvements in logscore, and thus forecast certainty, for vehicles with a low variance in either variable.
-Vehicles in the fifth quintiles, close to the homogenous variance estimates, benefit instead from the higher precision associated with the homogenous posterior distribution due to the large $N$.}
+Vehicles in the fifth quintiles, close to the homogeneous variance estimates, benefit instead from the higher precision associated with the homogeneous posterior distribution due to the large $N$.}
 \label{fig:varSplit}
@@
 
-The benefits of heterogenous models is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles.
+The benefits of heterogeneous models is only useful if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$, or its variational approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{1:T})$ can be inferred in the short time frames demanded by self-driving vehicles.
 Figure \ref{fig:timing} demonstrates the average computation time required for standard VB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$.
-The compuation time for standard VB increases linearly with $T$, while UVB is constant as $T$ increases; and hence posterior inference utilising UVB to condition posterior distributions on additional data is feasible.
-Parallelising the $M$ Monte-Carlo samples per iteration can further reduce computation time, allowing for shorter update periods or more complex models.
+The compuation time for standard VB increases linearly with $T$, while UVB is constant as $T$ increases; and hence UVB posterior inference can be implemented in an online setting.
+Parallelising the $M$ RQMC samples per iteration can further reduce computation time, allowing for shorter update periods or more complex models.
 \\
 
-As UVB involves replacing the posterior distribution at update with its variational approximation, the error introduced by UVB depends on how closely the variational approximation can represent the true posterior distribution.
-The left panel for Figure \ref{fig:updateCost} displays boxplots of the difference in logscore between UVB and standard VB at the three second forecast horizon for the CH model, and demonstrates that UVB in this case does not decrease the forecast logscores, as the difference between the logscores obtained by UVB and standard VB are distributed close to zero for several ranges of $T$.
+Additional approximation error caused by UVB is a result of replacing the to-be-updated posterior distribution in the Bayesian update equation its variational approximation.
+The size of this error depends on how well VB can approximate the true posterior distribution.
+The left panel for Figure \ref{fig:updateCost} displays boxplots of the difference in logscore between UVB and VB, at the three second forecast horizon, for the CH model, where UVB has not led to a decrease in the forecast logscores.
 The right panel displays analogous boxplots for the IH model, where the UVB logscores tend to be lower than the standard VB logscores, matching the increased Euclidean error found for the IH model with UVB inference in Figure \ref{fig:PredErrorZ}.
-An explanation for this is given by the form of the approximating distribution used for the IH model, the multivariate normal distribution may not adequately approximate the posterior distribution, while the extra flexibilitiy of the mixture multivariate normal distribution for the CH model was sufficient to replace the posterior.
+An explanation for this is given by the form of the approximating distribution used for the IH model, a multivariate normal distribution, may not be sufficiently flexible to approximate the posterior distribution.
+The extra flexibilitiy afforded by the mixture of multivariate normal distributions used for the CH model may have been able to replace the posterior to a sufficient degree as to not lead to a material amount of approximation error in UVB inference.
 
@@
 \centering
-\includegraphics[width = 0.75\textwidth, height = 0.3\textheight]{updateCost}
+\includegraphics[width = 0.95\textwidth]{updateCost}
 \caption{Differences in predictive logscores for $p(z_{i, T+30} | z_{i, 1:T})$ for each $T = 100, 110, \dots, 450$ between UVB and standard VB for the CH model (left) and the IH model (right).
 The differences are typically small for the CH model, while the IH model approximation has increased the logscore relative to standard VB.
-This may be due to the form of the approximating distribution for the IH model being an inadequate approximation for the true posterior}
+This may be due to the form of the approximating distribution for the IH model being an inadequate approximation of the true posterior}
 \label{fig:updateCost}
@@
 \label{sec:disc}
-This paper demonstrates that the use of auto-regressive models time series models drastically improves trajectory forecasts when compared to the competing models.
+This paper demonstrates that the use of auto-regressive time series models drastically improves trajectory forecasts when compared to the competing models.
 These models parameterise the dynamics of a driver's actions, with a range of parameter posterior means providing evidence of heterogeneity between the drivers of different vehicles.
@@
 The framework used by the trajectory forecasting problem is to construct a hierarchical model to infer the range of heterogeneity present in a dataset, and use exact inference results provided by MCMC to construct a prior distribution for an additional unit in the hierarchy.
-Our use of this framework has focused on situations where exact inference on the additional unit is unavailble in the time-frame required by the problem, instead using Variational Bayes to infer the posterior distribution of this additional unit.
-We have introduced UVB to allow this inference to happen in the online setting, and have explored fitting copula models to the MCMC samples to produce a functional form for the approximating distribution that better matches the true posterior, though have not been able to link the copula structure for the units in the original hierarchy to the new unit.
+Our use of this framework has focused on situations where exact inference incorporating the additional unit(s) of observation is unavailable in the time-frame required by the problem, instead using Variational Bayes to infer the posterior distribution of this additional unit.
+We have introduced UVB to allow this inference to happen in an online setting.
 \\
 
-This framework is not specific to vehicle forecasting, and could conceivably be used for many other problems, for example classifying new units in an online setting, or the detection of anomolous units through the use of a dirichlet process prior.
+This framework is not specific to vehicle forecasting, and could conceivably be used for many other problems, for example classifying new units in an online setting, or the detection of anomalous units through the use of a Dirichlet process prior (eg.
+\cite{Heard2016, Varadarajan2017}).
+\\
+
+We note also that, over the past year, we have explored fitting copula models to the MCMC samples to produce a family for the approximating distribution that will have an optimal member that better matches the true posterior than generic multivariate normal distributions.
+Unfortunately, sp far we have not been able to link the copula structure for the units in the original hierarchy to a copula structure suitable for the new unit.
 \\
@@
 \begin{itemize}
-\item March - April: Finalise vehicle trajectory forecasting study for publication
-\item May: Compile inference notes to form thesis background chapters
-\item June - Novemeber: Extend framework in another project
-\item December - May 2019: Additional projects and finalise thesis for submission
+\item March - April: Finalise vehicle trajectory forecasting study for publication,
+\item May: Compile inference notes to form thesis background chapters,
+\item June - July: Generalize framework beyond the vehicle trajectory setting, and try to find an alternative empirical application,
+\item August - January 2019: Implement, evaluate and write up study in alternative empirical setting,
+\item February - March 2019: Write Introduction and Conclusion chapters,
+\item April - May 2019: Foma; review editing and preparation of thesis for submission.
 \end{itemize}
 
-\newpage
 \bibliographystyle{asa}
@@
 \label{distance}
-d_{i, t} = \sqrt{(x_{i, t} - x_{i, 1})^2 + (y_{i, t} - y_{i, 1})^2}.
+d_{i, t} = \sum_{s=2}^t \sqrt{(x_{i, s} - x_{i, s-1})^2 + (y_{i, s} - y_{i, s-1})^2}.
 \end{equation}
@@
 
-Excluding the vehicles used to estimate the spline models, each of the 3387 remaining vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
+Excluding the vehicles used to estimate the spline models, each of the vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$, where $y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, with
 \begin{align}
@@
 
+\section{Adam: Adaptive Moment Estimation}
+
+SGA is performed with the Adam optimiser from \citet{Kingma2015b}.
+Adam adds momentum to the gradient algorithm, where estimates of the first moment of the gradient at previous iterations are used in addition to the estimate at the current iteration, with the weights of previous iterations decreasing exponentially.
+Momentum \citep{Qian1999} has seen success in stochastic gradient algorithms as it reduces the impact of a gradient estimate having the wrong sign; instead a change in sign must be repeated in successive iterations to reverse the sign of the change in $\lambda$.
+\\
+
+Adam also utilises an exponentially weighted average of the second moments of the gradient to provide the sequence $\rho$.
+These estimates, $m_t$ and $v_t$, for the first and second gradient moments respectively, are biased towards zero and so this bias is corrected by $\hat{m}_t$ and $\hat{v}_t$, which are used in the $\lambda$ update equation.
+Adam is described in Algorithm (\ref{alg:adam}), and applied in this paper with $\alpha = 0.1, \beta_1 = 0.9, \beta_2 = 0.999, \epsilon = 10^{-8}$, and $\nu = 0.01$.
+
+\begin{algorithm}[H]
+ \SetKwInOut{Input}{Input}
+ \Input{Function $f(\lambda)$, hyperparameters $\alpha, \beta_1, \beta_2, \epsilon, \nu$, initial values $\lambda^{(0)}$.}
+ \KwResult{Optimal Parameters $\lambda$}
+ Set $m_0 = 0$\;
+ Set $v_0 = 0$\;
+ Set $t = 0$\;
+ \While{$| f(\lambda^{(t)}) - f(\lambda^{(t-1)}) | > \nu$} {
+  $t = t + 1$\;
+  $g_t = \frac{\delta f(\lambda)}{\delta \lambda} \rvert_{\lambda = \lambda^{(t-1)}}$ \;
+  $m_t = \beta_1 m_{t-1} + (1-\beta_1) g_t$ \;
+  $v_t = \beta_2 v_{t-1} + (1-\beta_2) g_t^2$ \;
+  $\hat{m}_t = m_t / (1 - \beta_1^t)$ \;
+  $\hat{v}_t = v_t / (1 - \beta_2^t)$ \;
+  $\lambda^{(t)} = \lambda^{(t-1)} + \hat{m}_t \cdot \frac{\alpha}{\sqrt{\hat{v}_t} + \epsilon}$
+  }
+ \caption{Adam Optimiser applied to Gradient Ascent to maximise $f(\lambda)$.}
+  \label{alg:adam}
+\end{algorithm}
+
 \end{document}

### 2018-03-09 Cars/midcanPresentation.tex
updates

@@
-\documentclass[8pt]{beamer}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[11pt]{beamer}\usepackage[]{graphicx}\usepackage[]{color}
 %% maxwidth is the original width if it is less than linewidth
@@
 \begin{frame}
-\section{Framework}
-\frametitle{Thesis Framework}
-Focus: Two sets of data
+\section{Background}
+\subsection{Motivation}
+\frametitle{Motivation}
 \begin{itemize}
-\item Stuff Here about the general themes
-\item Use of exact inference to improve VB
-\item Hierarchical models with exact inference
-\item Some new unit with time constrained inference
-\item Want to use MCMC info to improve VB
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\section{Inference}
-\frametitle{Bayesian Inference}
-Note: reframe discussion with two sets of data from previous slide
-\begin{itemize}
-\item The posterior distribution for parameters $\theta$, $p(\theta | \textbf{z})$, can be inferred with Markov Chain Monte Carlo (MCMC). MCMC:
+\item Bayesian Inference in Forecasting:
 \begin{enumerate}
-\item Allows 'exact' inference as the number of iterations increases
-\item Can be very slow
+\item Provides forecast densities for the variable of interest.
+\item Encorporates parameter uncertainty.
+\item Is easy to extend to hierarchical models.
 \end{enumerate}
-\item We focus on settings where MCMC inference for the additional unit is too slow to be useful.
-\item Instead, Variational Bayes (VB) is used. VB:
+\vspace{3mm}
+\item Bayesian Inference is only as good as its compuation:
 \begin{enumerate}
-\item Approximates the posterior with an analytical alternative $q_{\lambda}(\theta | \textbf{z})$.
-\item Chooses the distribution $q$ that minimises a divergence funciton
-\item Is typically much faster than MCMC.
+\item Computation based on Markov Chain Monte Carlo (MCMC) is 'exact', but slow.
+\item Computation based on Variational Bayes (VB) is fast, but approximate.
 \end{enumerate}
@@
 \begin{frame}
-\frametitle{Kullback Leibler Divergence and Stochastic Optimisation}
+\frametitle{Motivation}
 \begin{itemize}
-\item The typical divergence is the Kullback Leibler Divergence from $q$ to $p$.
+\item We focus on Hierarchical Models where:
+\begin{enumerate}
+\item Inference conditioned on existing units in the hierarchy is available.
+\item Inference conditioned on additional units in a short time-frame is desired.
+\end{enumerate}
+\vspace{3mm}
+\item We aim to infer, and forecast, additional units in an online setting.
+\item We want to leverage the existing information to improve the forecasts.
 \end{itemize}
-\begin{equation}
-\label{KL-def}
-KL[q_{\lambda}(\theta | \textbf{z})\hspace{.1cm}||\hspace{.1cm}p(\theta | \textbf{z})] = E_{q_{\lambda}(\theta | \textbf{z})} \left[ \log(q_{\lambda}(\theta | \textbf{z})) - \log(p(\theta | \textbf{z})) \right],
-\end{equation}
-\begin{itemize}
-\item Monte-Carlo simulations are usually unavailable due to the $p(\theta | \textbf{z})$ term.
-\end{itemize}
-\begin{equation}
-\label{ELBO}
-\mathcal{L}(q, \lambda) = E_{q_{\lambda}(\theta | \textbf{z})} \left[\log(p(\theta, \textbf{z})) - \log(q_{\lambda}(\theta | \textbf{z}))\right],
-\end{equation}
-\begin{itemize}
-\item Stochastic Estimates are available through Monte-Carlo:
-\end{itemize}
-\begin{equation}
-\label{ELBO-MC}
-\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j}, \textbf{z})) - \log(q_{\lambda}(\theta_{j} | \textbf{z})) \right)
-\end{equation}
-where $\theta_{j} \sim q_{\lambda}(\theta | \textbf{z})$.
 \end{frame}
@@
 \begin{frame}
-\frametitle{Gradient Estimators}
+\subsection{Hierarchical Models}
+\frametitle{Hierarchical Models}
 \begin{itemize}
-\item For most $p$ and $q$ only Stochastic Gradient Ascent is available for optimisation. Repeatedly apply:
+\item Hierarchical Models are structured into three levels:
+\begin{enumerate}
+\item The likelihood: $p(z_i | \theta_i)$ for each unit $i$ in the model,
+\item The prior: $p(\theta_i | \beta)$ where $\beta$ is shared across all $i$,
+\item The hyper-prior: $p(\beta)$ is fixed.
+\end{enumerate}
+\vspace{5mm}
+\item For $i = 1, 2, \dots, N$, we assume that the posterior distribution $$p(\beta, \theta_1, \dots, \theta_N | z_1, \dots, z_N)$$ is available.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Hierarchical Models}
+\begin{itemize}
+\item What about $i = N+1$?
+Using $\textbf{z}_{1:j} = \{z_i | i = 1, \dots, j\}$, $$p(\theta_{N+1} |\textbf{z}_{1:N+1}) \propto p(z_{N+1} | \theta_{N+1})\int_{\beta}p(\theta_{N+1} | \beta)p(\beta | \textbf{z}_{1:N})d\beta.$$
+\item If there is a time constraint, we might use VB and approximate $$q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1}) \approx p(\theta_{N+1} |\textbf{z}_{1:N+1})$$
+\item choosing $$\lambda_{VB} = \arg \underset{\lambda}{\min}\mbox{ }D(p, q).$$
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\subsection{Variational Bayes}
+\frametitle{Variational Bayes and the Kullback-Leibler Divergence}
+\begin{itemize}
+\item A common divergence function is the Kullback Leibler Divergence from $q$ to $p$,
+\begin{align}
+\label{KL-def}
+KL[q_{\lambda}(\theta_{N+1} &| \textbf{z}_{1:N+1})\hspace{.1cm}||\hspace{.1cm}p(\theta_{N+1} | \textbf{z}_{1:N+1})] = \nonumber \\
+&E_q \left[ \log(q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})) - \log(p(\theta_{N+1} | \textbf{z}_{1:N+1})) \right],
+\end{align}
+\item The expectation is intractable for most $p$ and $q$.
+\item It typically cannot be estimated with Monte-Carlo simulations.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{The Evidence Lower Bound}
+\begin{itemize}
+\item Instead look at the Evidence Lower Bound (ELBO), 
+\begin{equation}
+\label{ELBO}
+\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],
+\end{equation}
+\item Stochastic Estimates are available through Monte-Carlo: $$\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j, N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{j, N+1} | \textbf{z}_{1:N+1})) \right)$$ where $\theta_{j, N+1} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
+\item Maximising the ELBO is equivalent to minimising the Kullback-Leibler Divergence.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Stochastic Optimisation}
+\begin{itemize}
+\item For an exponential family $p$ and factorisable $q$ Mean Field Variational Bayes provides a fast method to the ELBO. 
+\item We do not restrict $p$ and $q$ this way and instead use Stochastic Gradient Ascent.
+\item Repeat until $| \mathcal{L}(q, \lambda^{(m)}) - \mathcal{L}(q, \lambda{(m+1)})| < \nu$:
 \begin{equation}
@@
 \end{equation}
+\end{itemize}
+\begin{figure}
+\centering
+\includegraphics[width = 0.5\textwidth]{gradientascent}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Stochastic Optimisation}
+\begin{itemize}
 \item ELBO gradients can be estimated by the Score Estimator:
@@
 \begin{frame}
-\section{Online Heterogenous Forecasting: Problem}
-\subsection{Motivation}
+\section{Online Heterogeneous Forecasting: Problem}
+\subsection{Self-Driving Vehicles}
 \frametitle{Self-Driving Vehicles}
@@
 \end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Forecasting Methods}
+The literature forecasts vehicles in three ways:
+\begin{enumerate}
+\item Constant Behaviour.
+\item Predict Events.
+\item Provide Point Estimates.
+\end{enumerate}
 \end{frame}
@@
 \begin{frame}
-\subsection{Data}
+\subsection{The Next Generation Simulation Dataset}
 \frametitle{Next Generation Simulation}
@@
 \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Forecasting in the framework}
-\begin{itemize}
-\item That is a terrible title.
-\item Split the data into two sets: train/test
-\item Cars observed beforehand, cars encountered while driving
-\end{itemize}
-\begin{figure}
-\centering
-\includegraphics[width = 0.95\textwidth]{problem}
-\end{figure}
 \end{frame}
@@
 \begin{frame}
-\subsection{Modelling}
+\frametitle{Forecasting in the framework}
+\begin{itemize}
+\item That is a terrible title.
+\item Data is split into a training set and a test set:
+\item Training vehicles are observed before driving,
+\item Test vehicles are encountered while driving.
+\end{itemize}
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{problem}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\subsection{Auto-regressive Models for Trajectory Forecasting}
 \frametitle{Auto-regressive Models for Trajectory Forecasting}
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{pacf}
+\end{figure}
 \begin{itemize}
-\item Evidence of dynamics behaviour
-\item Modelling approachs / relationship to test train set
-\item Motivate the use of the three different models
+\item Dynamics are implemented in with bivariate AR models
+\item There are three approaches:
+\begin{enumerate}
+\item The Homogeneous Model
+\item The Independent Heterogeneous Model
+\item The Clustered Heterogeneous Model
+\end{enumerate}
 \end{itemize}
@@
 \begin{frame}
-\begin{itemize}
-\item That picture showing the IH / CH fits
-\item Evidence of heterogeneity and the suitability of the mixture model
-\end{itemize}
+\frametitle{Heterogeneity in the Model Posteriors}
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{HierSingleKDE}
+\end{figure}
 \end{frame}
@@
 \begin{itemize}
-\item Motivation
+\item Inference for self-driving vehicles needs to be fast.
+\item Inference cannot slow down as the amount of data increases.
 \begin{figure}[ht]
 \centering
-\includegraphics[width = 0.95\textwidth]{timeUpdate}
+\includegraphics[width = \textwidth]{timeUpdate}
 \end{figure}
-\item Notation: $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$ and $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$
+\item How do we go from $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$ to $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$ for $S < T$?.
 \end{itemize}
@@
 \frametitle{Approximating Bayesian Updating}
+\begin{itemize}
+\item The Bayesian Update Equation: 
 \begin{equation}
 \label{updatePost}
-p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S})
+p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S}).
 \end{equation}
-\begin{itemize}
-\item This only works if the right hand side can be evaluated
-\item VB Gradient Estimators requires the joint distribution
+\item Computation requires evaluating the joint distribution,
 \begin{equation}
 \label{updateJoint}
-p(\theta_{i}, z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S})
+p(\theta_{i}, z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S}).
 \end{equation}
-\item We will replace it with this
+\item We will approximate the Bayesian Update Equation in VB,
 \begin{equation}
@@
 \end{equation}
-\item Sources of error: How close $q$ is to $p$
+\item Source of additional error: How close $q$ is to $p$.
 \end{itemize}
@@
 \frametitle{Updating Gradient Estimators}
-The updating score estimator:
+\begin{itemize}
+\item The updating score estimator:
 \begin{align}
@@
 where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. 
-The updating reparameterised estimator:
+\item The updating reparameterised estimator:
 \begin{equation}
@@
 where $\epsilon_j \sim q(\epsilon)$.
+\end{itemize}
 \end{frame}

### 2018-03-09 Cars/midcanPresentation.tex
presentation

@@
 \begin{itemize}
+\item We aim to build a framework to provide density forecasts for new units in a heirarchy as they are observed.
+\pause
 \item Bayesian Inference in Forecasting:
 \begin{enumerate}
-\item Provides forecast densities for the variable of interest.
-\item Encorporates parameter uncertainty.
+\item Provides density forecasts for the variable of interest.
+\item Provides a mechanism to update posterior distributions
 \item Is easy to extend to hierarchical models.
 \end{enumerate}
+\pause
 \vspace{3mm}
@@
 \vspace{3mm}
-\item We aim to infer, and forecast, additional units in an online setting.
+\pause
+\item We aim to infer, and forecast, additional units in an online data setting.
+\pause
 \item We want to leverage the existing information to improve the forecasts.
@@
 \end{enumerate}
+\pause
 \vspace{5mm}
@@
 Using $\textbf{z}_{1:j} = \{z_i | i = 1, \dots, j\}$, $$p(\theta_{N+1} |\textbf{z}_{1:N+1}) \propto p(z_{N+1} | \theta_{N+1})\int_{\beta}p(\theta_{N+1} | \beta)p(\beta | \textbf{z}_{1:N})d\beta.$$
-\item If there is a time constraint, we might use VB and approximate $$q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1}) \approx p(\theta_{N+1} |\textbf{z}_{1:N+1})$$
+\pause
+\item If there is a time constraint, we might use VB with the approximation $$q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1}) \approx p(\theta_{N+1} |\textbf{z}_{1:N+1})$$
 \item choosing $$\lambda_{VB} = \arg \underset{\lambda}{\min}\mbox{ }D(p, q).$$
@@
 \item A common divergence function is the Kullback Leibler Divergence from $q$ to $p$,
-\begin{align}
+\begin{align*}
 \label{KL-def}
-KL[q_{\lambda}(\theta_{N+1} &| \textbf{z}_{1:N+1})\hspace{.1cm}||\hspace{.1cm}p(\theta_{N+1} | \textbf{z}_{1:N+1})] = \nonumber \\
+KL[q_{\lambda}(\theta_{N+1} &| \textbf{z}_{1:N+1})\hspace{.1cm}||\hspace{.1cm}p(\theta_{N+1} | \textbf{z}_{1:N+1})] = \\
 &E_q \left[ \log(q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})) - \log(p(\theta_{N+1} | \textbf{z}_{1:N+1})) \right],
-\end{align}
+\end{align*}
+\pause
 \item The expectation is intractable for most $p$ and $q$.
@@
 \begin{itemize}
-\item Instead look at the Evidence Lower Bound (ELBO), 
-\begin{equation}
-\label{ELBO}
-\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],
-\end{equation}
-\item Stochastic Estimates are available through Monte-Carlo: $$\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{j, N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{j, N+1} | \textbf{z}_{1:N+1})) \right)$$ where $\theta_{j, N+1} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
+\item Instead look at the Evidence Lower Bound (ELBO), $$\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],$$
+\pause
+\item Stochastic Estimates are available through Monte-Carlo: $$\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{N+1}^{(j)}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1})) \right)$$ where $\theta_{N+1}^{(j)} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
+\pause
 \item Maximising the ELBO is equivalent to minimising the Kullback-Leibler Divergence.
@@
 \centering
-\includegraphics[width = 0.5\textwidth]{gradientascent}
+\includegraphics[width = 0.4\textwidth]{gradientascent}
 \end{figure}
@@
 \item ELBO gradients can be estimated by the Score Estimator:
-\begin{equation}
-\label{scoreDeriv}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} = \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{j, i} | \textbf{z}))}{\delta \lambda} \left(\log(p(\theta_{j, i}, \textbf{z})) - \log(q_{\lambda}(\theta_{j, i} | \textbf{z})) \right),
-\end{equation}
-where $\theta_j \sim q_{\lambda}(\theta | \epsilon)$.
-\item Or the Reparametersied Estimator, where $\theta = f(\epsilon, \lambda)$:
-\begin{equation}
-\label{rpDeriv}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} = \sum_{j = 1}^M \frac{\delta \theta}{\delta \lambda} \frac{\delta \log(p(\theta, \textbf{z}))}{\delta \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon_j)} + \frac{\delta J(\lambda, \epsilon_j)}{\delta \lambda},
-\end{equation}
-where $\epsilon_j \sim q(\epsilon)$.
+\begin{align*}
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1}))}{\delta \lambda}  \\
+&\times \left(\log(p(\theta_{N+1}^{(j)}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1})) \right),
+\end{align*}
+where $\theta_{N+1}^{(j)} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Stochastic Optimisation}
+\begin{itemize}
+\item Or the Reparameterised Estimator, where $\theta_{N+1} = f(\epsilon, \lambda)$:
+\begin{align*}
+\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} &= \sum_{j = 1}^M \frac{\delta J(\lambda, \epsilon^{(j)})}{\delta \lambda} + \frac{\delta \theta_{N+1}}{\delta \lambda} \\
+\times \frac{\delta \log(p(\theta_{N+1}, \textbf{z}_{1:N+1}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda, \epsilon^{(j)})},
+\end{align*}
+where $\epsilon^{(j)} \sim q(\epsilon)$.
 \end{itemize}
@@
 \item The updating score estimator:
-\begin{align}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T}))}{\delta \lambda_T} \nonumber \\
-&\times \left(\log(q_{\lambda_S}(\theta_{i, j} | z_{i, 1:S}) - \log(q_{\lambda_T}(\theta_{i, j} | z_{i, 1:T})) \right) \label{scoreUpdate}
-\end{align}
-where $\theta_{i, j} \sim q(\theta_{i} | z_{i, 1:T})$. 
+\begin{align*}
+\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T}))}{\delta \lambda_T}  \\
+&\times \left(\log(q_{\lambda_S}(\theta_{N+1}^{(j)} | z_{N+1, 1:S}) - \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T})) \right)
+\end{align*}
+where $\theta_{N+1}^{(j)} \sim q(\theta_{N+1} | z_{N+1, 1:T})$. 
+\end{itemize}
+\end{frame}
+
+\begin{frame}
+\frametitle{Updating Gradient Estimators}
+\begin{itemize}
 \item The updating reparameterised estimator:
-\begin{equation}
-\label{rpUpdate}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} = \sum_{j = 1}^M \frac{\delta f(\lambda_T, \epsilon_j)}{\delta \lambda_T} \frac{\delta \log(q_{\lambda_S}(\theta_{i} |z_{i, 1:S}))}{\delta \theta_{i}} \bigg\rvert_{\theta_{i} = f(\lambda_T, \epsilon_j)} + \frac{\delta J(\lambda_T, \epsilon_j)}{\delta \lambda_T},
-\end{equation}
-where $\epsilon_j \sim q(\epsilon)$.
+\begin{align*}
+\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} &= \sum_{j = 1}^M \frac{\delta J(\lambda_T, \epsilon^{(j)})}{\delta \lambda_T} + \frac{\delta f(\lambda_T, \epsilon^{(j)})}{\delta \lambda_T} \\
+&\times \frac{\delta \log(q_{\lambda_S}(\theta_{N+1} |z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon^{(j)}} 
+\end{align*}
+where $\epsilon^{(j)} \sim q(\epsilon)$.
 \end{itemize}

### 2018-03-12 Cars/midcanPresentation.tex
presentation

@@
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+
+\AtBeginSection[]{
+  \begin{frame}
+  \vfill
+  \centering
+  \begin{beamercolorbox}[sep=8pt,center,shadow=true,rounded=true]{title}
+    \usebeamerfont{title}\insertsectionhead\par%
+  \end{beamercolorbox}
+  \vfill
+  \end{frame}
+}
+
 \begin{document}
@@
 
-\begin{frame}
 \section{Background}
 \subsection{Motivation}
+
+\begin{frame}
 \frametitle{Motivation}
 \begin{itemize}
-\item We aim to build a framework to provide density forecasts for new units in a heirarchy as they are observed.
+\item We aim to build a framework to provide density forecasts for new units in a hierarchy as they are observed.
 \pause
@@
 \begin{enumerate}
-\item Inference conditioned on existing units in the hierarchy is available.
+\item Each unit is a time series,
+\item Inference conditioned on existing units in the hierarchy is available,
 \item Inference conditioned on additional units in a short time-frame is desired.
@@
 \pause
-\item We aim to infer, and forecast, additional units in an online data setting.
+\item We aim to infer, and forecast, additional units as they are observed.
 \pause
@@
 
+\subsection{Bayesian Hierarchical Models}
+
 \begin{frame}
-\subsection{Hierarchical Models}
-\frametitle{Hierarchical Models}
+\frametitle{Bayesian Hierarchical Models}
 \begin{itemize}
@@
 \begin{frame}
-\frametitle{Hierarchical Models}
+\frametitle{Bayesian Hierarchical Models}
 \begin{itemize}
@@
 \pause
-\item If there is a time constraint, we might use VB with the approximation $$q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1}) \approx p(\theta_{N+1} |\textbf{z}_{1:N+1})$$
+\item If there is a time constraint, we might use VB with the approximation $$p(\theta_{N+1} |\textbf{z}_{1:N+1}) \approx q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$$
 \item choosing $$\lambda_{VB} = \arg \underset{\lambda}{\min}\mbox{ }D(p, q).$$
@@
 
-\begin{frame}
 \subsection{Variational Bayes}
-\frametitle{Variational Bayes and the Kullback-Leibler Divergence}
+
+\begin{frame}[label=VB]
+\frametitle{Variational Bayes}
 \begin{itemize}
-\item A common divergence function is the Kullback Leibler Divergence from $q$ to $p$,
+\item Typically minimize the Kullback-Leibler Divergence from $q$ to $p$,
 \begin{align*}
@@
 \end{align*}
+\item Maximising the Evidence Lower Bound (ELBO) is a tractable alternative, $$\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],$$
 \pause
-\item The expectation is intractable for most $p$ and $q$.
-\item It typically cannot be estimated with Monte-Carlo simulations.
+\item For a fixed family $q$, find optimal $\lambda$ with \hyperlink{SGA}{gradient ascent}:
+\begin{equation*}
+\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
+\end{equation*}
 \end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{The Evidence Lower Bound}
-\begin{itemize}
-\item Instead look at the Evidence Lower Bound (ELBO), $$\mathcal{L}(q, \lambda) = E_{q} \left[\log(p(\theta_{N+1}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta | \textbf{z}_{1:N+1}))\right],$$
-\pause
-\item Stochastic Estimates are available through Monte-Carlo: $$\mathcal{L}(q, \lambda) \approx \frac{1}{M} \sum_{j=1}^M \left(\log(p(\theta_{N+1}^{(j)}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1})) \right)$$ where $\theta_{N+1}^{(j)} \sim q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})$.
-\pause
-\item Maximising the ELBO is equivalent to minimising the Kullback-Leibler Divergence.
-\end{itemize}
-\end{frame}
-
-\begin{frame}
-\frametitle{Stochastic Optimisation}
-\begin{itemize}
-\item For an exponential family $p$ and factorisable $q$ Mean Field Variational Bayes provides a fast method to the ELBO. 
-\item We do not restrict $p$ and $q$ this way and instead use Stochastic Gradient Ascent.
-\item Repeat until $| \mathcal{L}(q, \lambda^{(m)}) - \mathcal{L}(q, \lambda{(m+1)})| < \nu$:
-\begin{equation}
-\label{gradientAscent}
-\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
-\end{equation}
-\end{itemize}
-\begin{figure}
-\centering
-\includegraphics[width = 0.4\textwidth]{gradientascent}
-\end{figure}
 \end{frame}
@@
 \begin{align*}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{SC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1}))}{\delta \lambda}  \\
+\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}_{SC} &= \sum_{j = 1}^M \frac{\partial \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1}))}{\partial \lambda}  \\
 &\times \left(\log(p(\theta_{N+1}^{(j)}, \textbf{z}_{1:N+1})) - \log(q_{\lambda}(\theta_{N+1}^{(j)} | \textbf{z}_{1:N+1})) \right),
@@
 \begin{itemize}
-\item Or the Reparameterised Estimator, where $\theta_{N+1} = f(\epsilon, \lambda)$:
+\item Or the Reparameterised Estimator, where $\epsilon = f^{-1}(\theta_{N+1}, \lambda)$:
 \begin{align*}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda)}{\delta \lambda}}_{RP} &= \sum_{j = 1}^M \frac{\delta J(\lambda, \epsilon^{(j)})}{\delta \lambda} + \frac{\delta \theta_{N+1}}{\delta \lambda} \\
-\times \frac{\delta \log(p(\theta_{N+1}, \textbf{z}_{1:N+1}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda, \epsilon^{(j)})},
+\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}_{RP} &= \sum_{j = 1}^M \frac{\partial J(\lambda, \epsilon^{(j)})}{\partial \lambda} + \frac{\partial \theta_{N+1}}{\partial \lambda} \\
+&\times \frac{\partial \log(p(\theta_{N+1}, \textbf{z}_{1:N+1}))}{\partial \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda, \epsilon^{(j)})},
 \end{align*}
-where $\epsilon^{(j)} \sim q(\epsilon)$.
+where $\epsilon^{(j)} \sim q(\epsilon)$ and $J(\lambda, \epsilon)$ is the Jacobian of $f$.
+\end{itemize}
+\end{frame}
+
+\section{Online Heterogeneous Forecasting: Problem}
+\subsection{Self-Driving Vehicles}
+
+\begin{frame}
+\frametitle{Self-Driving Vehicles}
+\begin{itemize}
+\item Many major manufacturers expect consumer self-driving vehicles to be available in the next 2-5 years
+\pause
+\item Autonomous driving has three major steps:
+\begin{enumerate}
+\item Detect surrounding traffic,
+\item Forecast surrounding traffic,
+\item Navigate through safely.
+\end{enumerate}
+\item We contribute to the second step.
 \end{itemize}
@@
 \begin{frame}
-\section{Online Heterogeneous Forecasting: Problem}
-\subsection{Self-Driving Vehicles}
 \frametitle{Self-Driving Vehicles}
 \begin{itemize}
-\item Self-Driving Vehicles are a thing soon
-\item The navigation system has three steps
+\item The existing literature has three major categories:
 \begin{enumerate}
-\item Detect surrounding traffic
-\item Forecast surrounding traffic
-\item Navigate through safely
+\item \textbf{Constant Behaviour Models:} \cite{Gindele2010, Houenou2013, Bautista2017, Waymo2017}
+\item \textbf{Predict Driver State:} \cite{Forbes1995, Pentland1999, Kuge2000, Gindele2010, Geng2017}
+\item \textbf{Provide Point Estimates:} \cite{Wolf2008, Ding2013, Woo2016b, Geng2017, Woo2017}
 \end{enumerate}
-\item Forecasting has not had much work
+\pause
+\item Our approach:
+\begin{enumerate}
+\item Provides density forecasts of vehicle trajectory,
+\item Incorporates driver heterogeneity,
+\item Provides feasible inference for online vehicle data.
+\end{enumerate}
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Forecasting Methods}
-The literature forecasts vehicles in three ways:
-\begin{enumerate}
-\item Constant Behaviour.
-\item Predict Events.
-\item Provide Point Estimates.
-\end{enumerate}
-\end{frame}
-
-\begin{frame}
-\frametitle{Aims of the first paper}
+\frametitle{Forecasting Approach}
 \begin{itemize}
-\item What we plan on doing with this first paper
-\item Overall goals and stuff
-\item Motivate heterogeneity
-\item Note heterogeneity needs inference on vehicles as they are encountered
-\item There is not much time to do that
-\item Develop an approximation to the Bayesian Update Equation so we can infer parameters for new vehicles online
-\item Also mention something on density forecasting would be nice
+\item \textbf{Training Vehicle} data is collected during development of the self-driving vehicle.
+\item \textbf{Test Vehicles} are encountered while driving and require forecasts.
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{problem}
+\end{figure}
 \end{itemize}
@@
 
+\subsection{The Next Generation Simulation Dataset}
+
 \begin{frame}
-\subsection{The Next Generation Simulation Dataset}
-\frametitle{Next Generation Simulation}
+\frametitle{Next Generation Simulation Data}
 \begin{itemize}
-\item This is a dataset created for macro traffic modelling
-\item It records all of the cars all of the time
-\item The road is curved
-\item Before and After of road straightening project
+\item The US Federal Highway Administration filmed 6101 vehicles every 100 milliseconds on the US 101 Freeway.
+\item We model and remove the curvature from the dataset.
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{roadStraightening}
+\end{figure}
 \end{itemize}
@@
 \frametitle{Trigonometric Motion Equations}
+\begin{itemize}
+\item Changes in position are the result of velocity $v$ and steering angle $\delta$,
 \begin{align}
@@
 \end{align}
-Picture
+\begin{figure}
+\centering
+\includegraphics[width = 0.7\textwidth]{motion}
+\end{figure}
+\end{itemize}
 \end{frame}
@@
 \frametitle{Extracting Driver Inputs}
+\begin{itemize}
+\item Transform the data to driver actions:
+\end{itemize}
 \begin{align}
@@
      \end{cases} \label{dEq} \\
-v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}.
+v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{vEq}, \\
+a_{i, t} &= v_{i, t} - v_{i, t-1}. \label{aEq}
 \end{align}
-\begin{equation}
-\label{aEq}
-a_{i, t} = v_{i, t} - v_{i, t-1}. 
-\end{equation}
-\begin{itemize}
-\item Why other variables are useful
-\item Acceleration is good because relative velocity / non-stationarity
-\end{itemize}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Forecasting in the framework}
-\begin{itemize}
-\item That is a terrible title.
-\item Data is split into a training set and a test set:
-\item Training vehicles are observed before driving,
-\item Test vehicles are encountered while driving.
-\end{itemize}
+\frametitle{Vehicle Acceleration and Angle}
 \begin{figure}
-\centering
-\includegraphics[width = \textwidth]{problem}
+\includegraphics[width = \textwidth]{timeSeries}
 \end{figure}
@@
 
+\subsection{Auto-Regressive Models for Trajectory Forecasting}
+
 \begin{frame}
-\subsection{Auto-regressive Models for Trajectory Forecasting}
-\frametitle{Auto-regressive Models for Trajectory Forecasting}
+\frametitle{Auto-Regressive Models for Trajectory Forecasting}
 \begin{figure}
 \centering
-\includegraphics[width = \textwidth]{pacf}
+\includegraphics[width = \textwidth]{pacf2}
 \end{figure}
 \begin{itemize}
-\item Dynamics are implemented in with bivariate AR models
-\item There are three approaches:
+\item Dynamics are modelled with bivariate AR processes.
+\pause
+\item We consdier three approaches:
 \begin{enumerate}
@@
 \begin{frame}
-\frametitle{A Homogeneous Approach}
-Talk about what is going on here
+\frametitle{The Homogeneous Approach}
+\begin{itemize}
+\item Acceleration and Angle are modelled independently:
 \begin{align}
-a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{aAR} \\
-\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{dAR}
+a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t}, \label{aAR} \\
+\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t}. \label{dAR} 
 \end{align}
-\begin{equation*}
-\label{thetaVec}
-\theta = \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\}.
-\end{equation*}
-\begin{equation}
-\label{indPrior}
-\theta \sim \mathcal{N}\left(\mu, \Sigma \right)
-\end{equation}
-\begin{itemize}
-\item The Posterior
-\item Why are we doing this
+\pause
+\item Collect parameters:
+\begin{align}
+\theta &= \{\phi_{1}, \dots, \phi_{p}, \gamma_{1}, \dots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\}. \label{thetaVec} \\
+\theta &\sim \mathcal{N}\left(\mu, \Sigma \right) \label{indPrior} 
+\end{align}
+\item Parameters are shared across different vehicles.
 \end{itemize}
@@
 \frametitle{The Independent Heterogeneity Model}
-Maybe different drivers behave differently?
-\begin{align}
-a_{i, t} &= \sum_{j = 1}^p \phi_{i, j} a_{i, t-j} + \sigma_{\epsilon, i} \epsilon_{i, t} \label{aAR2} \\
-\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{i, j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta, i} \eta_{i, t}. \label{dAR2}
-\end{align}
-This breaks down into a bunch of independent posteriors
+\begin{itemize}
+\item Allow each vehicle to have an independent 
+\item Each vehicle $i$ has an associated $\theta_i$:
+\item Each $\theta_i$ has an identical, independent prior:
 \begin{equation}
-p(\theta_{1:N} \mid z_{1:N}) = \prod_{i=1}^N p(\theta_{i} \mid z_{i}),
+\theta_i \overset{iid}{\sim} \mathcal{N}\left(\mu, \Sigma \right). 
 \end{equation}
-The training set isn't terribly useful.
-\begin{equation}
-p(\theta_{i}| \textbf{z}_{1:N}, z_{i,1:T}, \theta_{1:N}) = p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_i) p(\theta_i).
-\label{indepNewCar}
-\end{equation}
+\item Vehicle posteriors are independent.
+\item The training set does not provide information about the test set.
+\end{itemize}
 \end{frame}
@@
 \frametitle{The Clustered Heterogeneity Model}
-Hierarchical Models share information 
 \begin{itemize}
-\item Introduce Auxiliary Variable K
-\item Sharing of the information
+\item Replace the independent prior with a hierarchical prior:
+\begin{equation}
+\theta_i | \beta  \overset{iid}{\sim} \sum_{j=1}^K \pi_j \mathcal{N}(\mu_j, \Sigma_j)
+\end{equation}
+\pause
+\item The hyper-prior is:
+\begin{align*}
+\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
+\Sigma_j &\sim \mbox{Inverse Wishart}\left(\tau_j, \Psi_j\right), \\
+\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right).
+\end{align*}
 \end{itemize}
-\begin{equation}
-\label{mixPrior}
-\theta_i | k_i = j \sim N(\mu_j, \Sigma_j),
-\end{equation}
-and
-\begin{equation}
-k_i \mid \beta \sim \mbox{Multinomial}\left(\pi_1, \dots, \pi_{K}\right)
-\end{equation}
-\begin{align}
-\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
-\Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
-\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \dots = \alpha_K\right).
-\end{align}
 \end{frame}
@@
 \frametitle{The Clustered Heterogeneity Model}
-The new vehicle depends on the old vehicles through $\beta$.
+\begin{itemize}
+\item Test set vehicles depend on the train set vehicles through $\beta$.
 \begin{equation}
 \label{hierNewCar}
-p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) \int_{\beta} p(\theta_{i} | \beta) p (\beta | \textbf{z}_{1:N}) d\beta.
+p(\theta_{N+1} | \textbf{z}_{1:N+1}) \propto p(z_{N+1} | \theta_{N+1}) \int_{\beta} p(\theta_{N+1} | \beta) p (\beta | \textbf{z}_{1:N}) d\beta.
 \end{equation}
-The $\beta$ posterior is precise for large $N$
+\pause
+\item The $\beta$ posterior is precise for large $N$,
 \begin{equation}
 \label{hierNewCar2}
-p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) p(\theta_{i} | \hat{\beta}).
+p(\theta_{N+1} | \textbf{z}_{1:N+1}) \propto p(z_{N+1} | \theta_{N+1}) p(\theta_{N+1} | \hat{\beta}).
 \end{equation}
+\end{itemize}
 \end{frame}
@@
 \begin{frame}
-\frametitle{Heterogeneity in the Model Posteriors}
+\frametitle{Evidence of Heterogeneity in the Model Posteriors}
 \begin{figure}
 \centering
-\includegraphics[width = \textwidth]{HierSingleKDE}
+\includegraphics[width = \textwidth]{hierSingleKDE2}
 \end{figure}
+\begin{itemize}
+\item Top: Kernel Density Estimate of $E(\theta_i | z_i)$ from the IH Model.
+\item Bottom: $p(\theta_{N+1} | \textbf{z}_{1:N})$ from the CH Model.
+\end{itemize}
 \end{frame}
 
+\section{Updating Variational Bayes}
+
 \begin{frame}
-\section{Updating Variational Bayes}
 \frametitle{Updating Variational Bayes}
@@
 \end{figure}
-\item How do we go from $q_{\lambda_S}(\theta_{i} | z_{i, 1:S})$ to $q_{\lambda_T}(\theta_{i} | z_{i, 1:T})$ for $S < T$?.
+\item Need to update $q_{\lambda_S}(\theta_{N+1} | z_{N+1, 1:S})$ to $q_{\lambda_T}(\theta_{N+1} | z_{N+1, 1:T})$.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Approximating Bayesian Updating}
+\frametitle{Approximate Bayesian Updating}
 \begin{itemize}
@@
 \label{updatePost}
-p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S}).
+p(\theta_{N+1} | z_{N+1, 1:T}) \propto p(z_{N+1, S+1:T} | \theta_{N+1})p(\theta_{N+1} | z_{N+1, 1:S}).
 \end{equation}
-\item Computation requires evaluating the joint distribution,
+\pause
+\item Computation typically requires evaluating the joint distribution,
 \begin{equation}
 \label{updateJoint}
-p(\theta_{i}, z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})p(\theta_{i} | z_{i, 1:S}).
+p(\theta_{N+1}, z_{N+1, 1:T}) = p(z_{N+1, S+1:T} | \theta_{N+1})p(\theta_{N+1} | z_{N+1, 1:S}).
 \end{equation}
-\item We will approximate the Bayesian Update Equation in VB,
+\pause
+\item We use the VB approximation of the time $S$ posterior,
 \begin{equation}
 \label{ApproxJoint}
-\hat{p}(\theta_{i},  z_{i, 1:T}) = p(z_{i, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{i} | z_{i, 1:S}).
+\hat{p}(\theta_{N+1},  z_{N+1, 1:T}) = p(z_{N+1, S+1:T} | \theta_{i})q_{\lambda_S}(\theta_{N+1} | z_{N+1, 1:S}).
 \end{equation}
@@
 \begin{align*}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{USC} &= \sum_{j = 1}^M \frac{\delta \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T}))}{\delta \lambda_T}  \\
-&\times \left(\log(q_{\lambda_S}(\theta_{N+1}^{(j)} | z_{N+1, 1:S}) - \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T})) \right)
+&\widehat{\frac{\partial\mathcal{L}(q, \lambda_T)}{\partial \lambda_T}}_{USC} = \sum_{j = 1}^M \frac{\partial \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T}))}{\partial \lambda_T}  \\
+&\times \left(\log(\hat{p}(\theta_{N+1},  z_{N+1, 1:T})) - \log(q_{\lambda_T}(\theta_{N+1}^{(j)} | z_{N+1, 1:T})) \right)
 \end{align*}
-where $\theta_{N+1}^{(j)} \sim q(\theta_{N+1} | z_{N+1, 1:T})$. 
+where $\theta_{N+1}^{(j)} \sim q_{\lambda_T}(\theta_{N+1} | z_{N+1, 1:T})$. 
 \end{itemize}
@@
 \begin{align*}
-\widehat{\frac{\delta\mathcal{L}(q, \lambda_T)}{\delta \lambda_T}}_{URP} &= \sum_{j = 1}^M \frac{\delta J(\lambda_T, \epsilon^{(j)})}{\delta \lambda_T} + \frac{\delta f(\lambda_T, \epsilon^{(j)})}{\delta \lambda_T} \\
-&\times \frac{\delta \log(q_{\lambda_S}(\theta_{N+1} |z_{N+1, 1:S}))}{\delta \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon^{(j)}} 
+&\widehat{\frac{\partial\mathcal{L}(q, \lambda_T)}{\partial \lambda_T}}_{URP} = \sum_{j = 1}^M \frac{\partial J(\lambda_T, \epsilon^{(j)})}{\partial \lambda_T} + \frac{\partial f(\lambda_T, \epsilon^{(j)})}{\partial \lambda_T} \\
+&\times \frac{\partial \log(\hat{p}(\theta_{N+1},  z_{N+1, 1:T})}{\partial \theta_{N+1}} \bigg\rvert_{\theta_{N+1} = f(\lambda_T, \epsilon^{(j)})} 
 \end{align*}
@@
 
+\section{OHF: Results}
+
 \begin{frame}
-\section{Online Heterogeneous Forecasting: Results}
-\frametitle{Point Forecasts}
+\frametitle{Forecast Procedure}
 \begin{itemize}
-\item Introduce Naive / RNN models?
-\item Some stuff about point forecasts: Both pictures?
-\item Note that, on average, the homogenous model is as good as others
-\item Some stuff about different inference frameworks
-\item MAPs are not the whole story
+\item Inference on the training vehicles $i = 1, \dots 2000$ is available.
+\item For each test vehicle $i = 2001, \dots, 3387$, 
+\item For $T = 100, 110, \dots, 450$ and $h = 1, 2, \dots 30$:
+\begin{enumerate}
+\item Observe $z_{i, 1:T}$,
+\item Infer $p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T})$,
+\item Forecast $p(z_{i, T+h} | \textbf{z}_{1:N},  z_{i, 1:T})$.
+\end{enumerate}
+\vspace{2mm}
+\pause
+\item Heterogeneous inference is provided by:
+\begin{enumerate}
+\item Random Walk Metropolis Hastings MCMC
+\item Standard VB 
+\item Updating VB
+\end{enumerate}
+\item VB Approximations $q$ match the model prior distribution.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Heterogeneous Variance Distributions}
+\frametitle{Other Considered Models}
 \begin{itemize}
-\item Some stuff about variance
-\item A note on erratic driver detection
-\item Two different pictures to go in here
+\item Constant models are widely used, so we form 9 Naive Models.
+\item Future acceleration and angle are estimated by $\{\hat{a}_{i, T}, \hat{\delta}_{i, T}\}$.
+\begin{table}
+\begin{center}
+\resizebox{0.9\textwidth}{!}{%
+\begin{tabular}{|l|c|c|c|}
+\hline
+& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 0.1 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
+\hline
+$\hat{a}_{i, T} = a_{i, T}$ & Naive 1 & Naive 2 & Naive 3 \\
+$\hat{a}_{i, T} = 0.1 \sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
+$\hat{a}_{i, T} = 0$ & Naive 7 & Naive 8 & Naive 9 \\
+\hline
+\end{tabular}}
+\end{center}
+\label{tableNaive}
+\end{table}
+\item A Recurrent Neural Network Model is not competitive with the Naive Models at this time.
 \end{itemize}
@@
 \begin{frame}
-\frametitle{Updating Variational Bayes Results}
+\frametitle{Point Forecast Results}
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{predictiveError}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Heterogeneous Variance Distributions}
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{posVarMean}
+\end{figure}
 \begin{itemize}
-\item Compare to regular VB
-\item Time Complexity
-\item Difference in logscores
+\item Black: $E(p(\theta_i | \textbf{z}_{1:N}, z_{N+1})), i > N$ from the CH Model
+\item Red: $E(p(\theta | \textbf{z}_{1:N}))$ from the Homogeneous Model
 \end{itemize}
@@
 \begin{frame}
+\frametitle{CH Logscore (UVB) - Homogenous Logscore at h = 30}
+\begin{figure}
+\centering
+\includegraphics[width = 0.7\textwidth]{varSplit}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Updating Variational Bayes Approximation Error}
+\begin{figure}
+\centering
+\includegraphics[width = 0.7\textwidth]{updateCost2}
+\end{figure}
+\end{frame}
+
+\begin{frame}
+\frametitle{Updating Variational Bayes Convergence Time}
+\begin{figure}
+\centering
+\includegraphics[width = \textwidth]{timing}
+\end{figure}
+\end{frame}
+
 \section{Extensions}
-\frametitle{Extending the Framework}
+
+\begin{frame}
+\frametitle{Vehicle Model Extensions}
+\begin{enumerate}
+\item \textbf{Acceleration / Angle Dependency:} 
 \begin{itemize} 
-\item Copula Project
-\item Other Applications
+\item Use a VAR model with correlated errors.
+\end{itemize}
+\pause
+\item \textbf{Lane Change Detection:} 
+\begin{itemize}
+\item Choose a probability threshold $\xi$,
+\item If $p(\mbox{Vehicle $i$ in different lane at time } T + h | z_{i, 1:T}) > \xi$ predict lane change.
+\end{itemize}
+\pause
+\item \textbf{Multi Vehicle Dependencies:}
+\begin{itemize}
+\item Jointly model the movement of multiple vehicles in close proximity.
+\end{itemize}
+\pause
+\item \textbf{Markov Switching Behaviour:}
+\begin{itemize}
+\item Follow the HMM literature, but provide trajectory models for each state
+\end{itemize}
+\end{enumerate}
+\end{frame}
+
+\begin{frame}
+\frametitle{Online Inference for New Units in Hierarchical Models}
+\begin{enumerate}
+\item \textbf{Online Classification:}
+\begin{itemize}
+\item Infer a posterior distribution and predict class labels
+\end{itemize}
+\pause
+\item \textbf{Online Clustering:} 
+\begin{itemize}
+\item The mixture prior assigns units to $K$ latent groups.
+\item A new unit can be assigned to a latent group as it is observed.
+\end{itemize}
+\pause
+\item \textbf{Anomaly Detection:}
+\begin{itemize} 
+\item Dirichlet processes allow units to be assigned to new latent groups.
+\item This represents a unit with parameters unlike any previously seen.
+\end{itemize}
+\end{enumerate}
+\end{frame}
+
+\begin{frame}
+\frametitle{MCMC Informed Variational Bayes}
+\begin{itemize}
+\item Choosing the family $q$ is an open problem.
+\item Flexible distributions are advocated.
+\pause
+\vspace{2mm}
+\item We have MCMC Samples of $p(\theta_i | \textbf{z}_{1:N})$ for $i \leq N$.
+\item Marginal / copula distribution estimation techniques are available.
+\pause
+\vspace{2mm}
+\item \textbf{Will a family for $p(\theta_i | \textbf{z}_{1:N})$ be suitable for $q_{\lambda}(\theta_{N+1} | \textbf{z}_{1:N+1})?$}
+\end{itemize}
+\end{frame}
+
+\begin{frame}[allowframebreaks]
+\bibliographystyle{asa}
+\bibliography{references}
+\end{frame}
+
+\begin{frame}[label=SGA]
+\frametitle{Gradient Ascent}
+\begin{itemize}
+\item Repeat until $| \mathcal{L}(q, \lambda^{(m)}) - \mathcal{L}(q, \lambda{(m+1)})| < \nu$:
+\begin{equation*}
+\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
+\end{equation*}
+\begin{figure}
+\centering
+\includegraphics[width = 0.4\textwidth]{gradientascent}
+\end{figure}
+\item \hyperlink{VB}{Return}
 \end{itemize}

### 2018-03-12 Cars/HierMixMod.R

@@ -206,6 +206,7 @@ densities %>%
   group_by(var, support) %>%
   summarise(dens = sum(dens)) -> densMixMod
 
+densMixMod %>%
   ggplot(densMixMod) + geom_line(aes(support, dens)) +
   geom_line(data=densities, aes(support, dens, colour = factor(group))) +
   facet_wrap(~var, scales = 'free', ncol = 6, labeller = label_parsed) + 
@@ -478,6 +479,8 @@ posMeans %>%
   theme(axis.text.x = element_text(angle = 45, hjust = 1))-> p2
 
 gridExtra::grid.arrange(p2, p1, ncol = 1)
+
+
 }
 
 sliceSampler{

### 2018-03-12 Cars/createCarsAug.R

@@ -103,16 +103,16 @@ for(i in 1:5){
 carsAug %>% 
   left_join(pred) %>%
   group_by(ID) %>%
-  mutate(relX = x - xfit,#sign(x - xfit) * sqrt((x-xfit)^2 + (y-yfit)^2),
+  mutate(relX = x - xfit,
          n = seq_along(time),
          lagRelX = ifelse(n == 1, 0, lag(relX)),
          lagDist = ifelse(n == 1, 0, lag(dist)),
          relv = sqrt((relX - lagRelX)^2 + (dist - lagDist)^2),
          reldelta = atan2(dist - lagDist, relX - lagRelX),
          class = factor(class, levels = 1:3, labels = c('bike', 'car', 'truck'))) %>%
-  filter(n > 1) %>%# & y > 600 & y < 1600) %>%
-  select(ID, changed, enterExit, frame, x, y, relX, lagRelX, lagDist, dist, lane, delta, v, relv, reldelta, proceeding, time, class) -> carsAug
+  filter(n > 1) -> plotData
 
+carsAug <- select(plotData, ID, changed, enterExit, frame, x, y, relX, lagRelX, lagDist, dist, lane, delta, v, relv, reldelta, proceeding, time, class) 
 write.csv(carsAug, 'carsAug.csv', row.names = FALSE)
 
 carsChanged %>% 
@@ -125,13 +125,27 @@ carsChanged %>%
          relv = sqrt((relX - lagRelX)^2 + (dist - lagDist)^2),
          reldelta = atan2(dist - lagDist, relX - lagRelX),
          class = factor(class, levels = 1:3, labels = c('bike', 'car', 'truck'))) %>%
-  filter(n > 1) %>%# & y > 600 & y < 1600) %>%
-  select(ID, changed, enterExit, frame, x, y, relX, lagRelX, lagDist, dist, lane, delta, v, relv, reldelta, proceeding, time, class) -> carsChanged
+  filter(n > 1) %>% plotDataCh
+
+plotData <- rbind(plotData, plotDataCh)
+
+carsChanged <- select(plotDataCh, ID, changed, enterExit, frame, x, y, relX, lagRelX, lagDist, dist, lane, delta, v, relv, reldelta, proceeding, time, class)
 
 write.csv(carsChanged, 'carsChanged.csv', row.names = FALSE)
 saveRDS(splinesID, 'splinesID.RDS')
 
 
+plotData %>%
[…]

### 2018-03-12 Cars/dynamics.R

@@ -42,6 +42,25 @@ carsAug %>%
   mutate(n = seq_along(v),
          a = v - ifelse(n == 1, v, lag(v))) -> pacfData
 
+pacfData %>%
+  select(ID, frame, reldelta, a, n) %>%
+  filter(n > 100 & n <= 600) %>%
+  group_by(ID) %>%
+  mutate(frame = frame - min(frame),
+         `delta - pi/2` = reldelta - pi/2) %>%
+  ungroup() %>%
+  mutate(ID = case_when(ID == 194 ~ 'Vehicle 1', 
+                        ID == 3008 ~ 'Vehicle 2',
+                        ID == 10999 ~ 'Vehicle 3',
+                        ID == 20958 ~ 'Vehicle 4',
+                        ID == 21351 ~ 'Vehicle 5'),
+         ID = factor(ID)) %>%
+  ungroup() %>%
+  gather(var, value, -ID, -frame, -n, -reldelta) %>%
+  ggplot() + geom_line(aes(frame, value)) + 
+  facet_grid(var ~ ID) + 
+  labs(x = 'T', y = 'Value') + theme_bw()
+
 
 pacfs <- data.frame()
 for(i in 1:5){

### 2018-08-28 betel.tex
betel

  \documentclass[12pt,a4paper]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb
\hbadness=99999

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{alltt}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage{amsmath}
\usepackage{subfig}
\usepackage{amsthm}
\usepackage[]{algorithm2e}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\setcounter{MaxMatrixCols}{30}

\def\app#1#2{%
  \mathrel{%
    \setbox0=\hbox{$#1\sim$}%
    \setbox2=\hbox{%
      \rlap{\hbox{$#1\propto$}}%
      \lower1.1\ht0\box0%
    }%i
    \raise0.25\ht2\box2%
  }%
}
\def\approxprop{\mathpalette\app\relax}

\setlength{\topmargin}{0in}
\setlength{\oddsidemargin}{0.08in}
\setlength{\evensidemargin}{0.08in}
\setlength{\textwidth}{6.5in}
\setlength{\textheight}{9in}
\setlength\parindent{0pt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\title{Variational Bayesian Exponentially Tilted Empirical Likelihood}
\begin{document}
\maketitle

\section{Background}

\citet{Chib2017} introduces Bayesian inference based on moment conditions, basing inference on the satisfication of the $d-$dimensional moment conditions,

\begin{equation}
\label{emp:moments}
E \left[ g(Y, \theta) \right] = 0,
\end{equation}
where the expectation is with respect to the true data generation process $p$.

The likelihood is replaced with the exponentially tilted empirical likelihood (ETEL), defined by
\begin{equation}
\label{emp:etel}
p(y_{1:T} | \theta) = \prod_{t=1}^T p^*_t(\theta)
\end{equation}
where each $p^*_t$ is chosesn to minimise the KL divergence between $(p_1, p_2, \ldots, p_T)$ and the empirical probabilities $(1/T, 1/T, \ldots 1/T)$ subject to the conditions
\begin{equation}
\label{emp:sumRestrict}
\sum_{t=1}^T p_t = 1
\end{equation}
and
\begin{equation}
\label{emp:momentRestrict}
\sum_{t=1}^T p_t g(y_t, \theta) = 0.
\end{equation}
It can be shown that the solution to this is given by
\begin{equation}
\label{emp:solnP}
p^*_t(\theta) = \frac{\exp \{\hat{\lambda}^{\prime} g(y_t, \theta) \}}{ \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}}
\end{equation}
where
\begin{equation}
\label{emp:solnLambda}
\hat{\lambda} = \lambda(\theta) = \arg \underset{\lambda}{\min} \frac{1}{T} \sum_{t=1}^T \exp \{\lambda^{\prime} g(y_t, \theta) \}.
\end{equation}

This results in the ETEL posterior
\begin{equation}
\label{emp:post}
p(\theta | y_{1:T}) \propto p(\theta) \prod_{t=1}^T\frac{\exp \{\hat{\lambda}^{\prime} g(y_t, \theta) \}}{ \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}}.
\end{equation}

\citet{Chib2017} proposes a Metropolis Hastings sampler for the posterior, following the steps
\begin{enumerate}
\item Draw $\theta$ from a suitable candidate distribution
\item Solve (\ref{emp:solnLambda}) for this draw
\item Calculate (\ref{emp:post}) as part of the Metropolis Hastings acceptance probability
\item Accept or reject $\theta$ as per standard MH
\end{enumerate}
This sounds quite slow, and may benefit from VB.
\\

\section{Implementation of VB}

Reparameterised and Stein VB requires the derivative of $\log(p(\theta, y_{1:T})$ with respect to $\theta$ as part of the ELBO gradient estiamtor. This derivative is complicated by $\hat{\lambda}$ being a function of $\theta$.
\\

If, for every $\theta \in \Theta$, the solution to  (\ref{emp:solnLambda}) is unique, and, defining $h(\theta, \lambda(\theta)) =  \frac{1}{T} \sum_{t=1}^T \exp \{\lambda^{\prime} g(y_t, \theta) \}$, the hessian matrix $\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda^2}$ is positive-defiite, the optimal $\hat{\lambda}$ is found by satisfying the first order condition
\begin{equation}
\label{emp:lambdaDeriv}
\frac{\partial h(\theta, \lambda(\theta))}{\partial \lambda} = 0.
\end{equation}
Taking the derivative of both sides of this with respect to $\theta$ results in
\begin{equation}
\label{emp:thetaDeriv}
\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \theta^{\prime}} + \frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \lambda^{\prime}} \frac{\partial \lambda(\theta)}{\partial \theta} = 0,
\end{equation}
where
\begin{equation}
\label{emp:hessian}
\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \lambda^{\prime}} = \frac{1}{T} \sum_{t=1}^T g(y_t, \theta) g(y_t, \theta)^{\prime} \exp\{\hat{\lambda}^{\prime} g(y_t, \theta) \}
\end{equation}
and
\begin{equation}
\label{emp:crossDeriv}
\frac{\partial^ 2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \theta^{\prime}} = \frac{1}{T} \sum_{t=1}^T \exp\{\hat{\lambda}^{\prime} g(y_t, \theta) \} \left( \mathbb{I} + g(y_t, \theta) \hat{\lambda}^{\prime} \right) \frac{\partial g(y_t, \theta)}{\partial \theta}.
\end{equation}
(\ref{emp:thetaDeriv}) may be solved for $\partial \lambda(\theta) / \partial \theta$ by
\begin{equation}
\label{emp:thetaDerivSoln}
 \frac{\partial \lambda(\theta)}{\partial \theta} = -\left(\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \lambda^{\prime}}\right)^{-1}\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \theta^{\prime}}.
\end{equation}
Note that (\ref{emp:crossDeriv}) assumes that the dimension of $\theta$ is equal to the dimension of $g(y, \theta)$ (and also $\lambda$).
\\

We may then take the derivative of the log-likelihood with respect to $\theta$ via
\begin{align}
\frac{\partial \log(p(y_{1:T}, \theta))}{\partial \theta} &= \frac{\partial \log(p(\theta))}{\partial \theta} + \frac{\partial}{\partial \theta} \sum_{t=1}^T \log \left( \frac{\exp \{\hat{\lambda}^{\prime} g(y_t, \theta) \}}{ \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}} \right), \nonumber \\
&= \frac{\partial \log(p(\theta))}{\partial \theta} + \sum_{t=1}^T \frac{\partial}{\partial \theta} \left[\hat{\lambda}^{\prime} g(y_t, \theta) \right]  - T \frac{\partial}{\partial \theta} \log \left(  \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \} \right), \nonumber \\
&= \frac{\partial \log(p(\theta))}{\partial \theta} + \sum_{t=1}^T \left(\frac{\partial \hat{\lambda}^{\prime}}{\partial \theta} g(y_t, \theta) + \frac{\partial g(y_t, \theta)^{\prime}}{\partial \theta}\hat{\lambda}  \right) \nonumber \\
&- \frac{T}{\sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}}\left(\sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \} \left(\frac{\partial \hat{\lambda}^{\prime}}{\partial \theta} g(y_s, \theta) + \frac{\partial g(y_s, \theta)^{\prime}}{\partial \theta}\hat{\lambda}  \right) \right). \label{emp:logpDeriv}
\end{align}

If the moment conditions are differentiable with respect to $\theta$ we may use ETEL in place of the likelihood function in Variational Bayes. Like in the Metropolis Hastings sampler, each gradient calculation requries optimisation of (\ref{emp:solnLambda}), with $O(d^3)$ additional calculations for (\ref{emp:thetaDerivSoln}). I found that approximatately $90\%$ of the compuation time for the example below was taken by (\ref{emp:solnLambda}), so there is little additional overhead.

\section{Example}

\citet{Chib2017} have numerous examples in their original paper, the simplest of which is the linear regression model
\begin{equation}
y_t = \alpha + \beta z_t + e_t,
\end{equation}
where $\alpha = 0$ and $\beta = 1$ for all $t = 1, 2, \dots, 250$.
\\

Errors $e_t$ are simulated from the mixture distribution
\begin{equation}
\label{emp:mixture}
e_t \sim \left\{ \begin{array}{cr} \mathcal{N}(0.75, 0.75^2) & \mbox{ with probability } 0.5, \\
\mathcal{N}(-0.75, 1.25^2) & \mbox{  with probability } 0.5. \end{array} \right. 
\end{equation}

BETEL is implemented with the following moment conditions:
\begin{equation}
g(y_t, \theta) = \left[\begin{array}{c} e_t \\ e_t z_t \\ e_t^3 - \nu \end{array}\right],
\end{equation}
equivalent to the standard regression moment assumptions and an additional free parameter $\nu$ relating to the error skewness. They do not add any moment conditions relating to the varianace.
\\

The unknown parameters are given by $\theta = \{\alpha, \beta, \nu \}$.
\\

Inferencce is based on 
\begin{enumerate}
\item The BETEL Metropolis Hastings algorithm described in Section 1, 
\item A Metropolis Hastings using the parametric likelihood given by (\ref{emp:mixture}) where the mixture probabilities, means, variances, and hence $\nu$, is known,
\item VBETEL using the reparameteresied gradient estimator for a Gaussian $q$ 
\item VBETEL using the Stein gradient estimator with 75 $\theta$ particles as $q$.
\end{enumerate}

\begin{figure}[hbtp]
\centering
\includegraphics[width = 0.95\textwidth]{VBETEL}
\caption{Marginal Densities for each parameter}
\end{figure}

\bibliographystyle{asa}
\bibliography{references}

\end{document}

### 2018-08-28 thesis.tex
betel

\documentclass[
12pt, % The default document font size, options: 10pt, 11pt, 12pt
onehalfspacing, % Single line spacing, alternatives: onehalfspacing or doublespacing
nohyperref, % Uncomment to not load the hyperref package
headsepline, % Uncomment to get a line under the header
chapterinoneline, % Uncomment to place the chapter title next to the number on one line
]{MastersDoctoralThesis} % The class file specifying the document structure
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother
\usepackage[]{color}
\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\usepackage{alltt}%
\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%2multibyte Version: 5.50.0.2960 CodePage: 1252

\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage[centertags]{amsmath}
\usepackage{graphicx}%
\usepackage{natbib}
\usepackage{color}
%\usepackage[dvipsnames,svgnames*]{xcolor}
\usepackage{array}
\usepackage[hidelinks]{hyperref}
\usepackage[font=small,skip=5pt]{caption}
\usepackage[aboveskip=2pt]{subcaption}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage[]{algorithm2e}
%\usepackage{tikz}
%\usetikzlibrary{bayesnet}
\usepackage{url}
\usepackage{ulem}
\usepackage{afterpage}
\setcounter{MaxMatrixCols}{30}
\hbadness=99999

\def\app#1#2{%
  \mathrel{%
    \setbox0=\hbox{$#1\sim$}%
    \setbox2=\hbox{%
      \rlap{\hbox{$#1\propto$}}%
      \lower1.1\ht0\box0%
    }%i
    \raise0.25\ht2\box2%
  }%
}
\def\approxprop{\mathpalette\app\relax}

\geometry{
	paper=a4paper, % Change to letterpaper for US letter
	inner=1.5cm, % Inner margin
	outer=2.5cm, % Outer margin
	bindingoffset=.5cm, % Binding offset
	top=2.5cm, % Top margin
	bottom=3.5cm, % Bottom margin
	head=40pt
	%showframe, % Uncomment to show how the type block is set on the page
}
\setlength\parindent{0pt}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}

\thesistitle{Approximate Bayesian Updating for Online Inference of Multivariate Heterogeneous Time Series}
\author{Nathaniel Tomasetti}
\supervisor{Catherine Forbes and Anastasios Panagiotelis}

\begin{document}
\pagestyle{plain}

\begin{titlepage}
\begin{center}

{\huge \bfseries \ttitle\par}\vspace{3mm} % Thesis title
\HRule \\
\vspace{8mm}

{\huge {\authorname}\\} % Author name - remove the \href bracket to remove the link
\vspace{8mm}
\large Supervised By {\supname}

\vspace{30mm}

\large \textit{A thesis submitted in fulfillment of the requirements\\ for the degree of Doctor of Philosophy}\\
   \vspace{7mm}
  Department of Econometrics and Business Statistics\\
  Monash University\\
  Australia \\
  \vspace{6mm}
   \includegraphics[height=33mm]{figures/crest} \\
   \vspace{3mm}
  \today
\end{center}
\end{titlepage}

\frontmatter

\tableofcontents

\mainmatter 
\pagestyle{thesis}

\chapter{Introduction}
\label{chap:Intro}

There will be something here eventually.

\chapter{The Language of Uncertainty}
\label{chap:Background}

Intro etc.

\section{Bayes' Rule}
\label{sec:Bayes}

Let $y_{1:T}$ denote a series of observed data, $y_1, y_2, \ldots, y_T$, distributed according to some conditional probability density function, or likelihood, $p(y_{1:T} | \theta)$, where a $k-$dimensional vector of unknown parameters $\theta$ governs the distribution. Given some $\textit{prior}$ distribution for $\theta$, $p(\theta)$, Bayesian statistics is concerned with inference of the \textit{posterior} distribution, $p(\theta | y_{1:T})$, given by Bayes' rule:
\begin{equation}
\label{Bayes:BayesRule}
p(\theta | y_{1:T}) = \frac{p(y_{1:T} | \theta)p(\theta)}{p(y_{1:T})},
\end{equation}
where the denominator is the marginal distribution of the data, 
\begin{equation}
\label{Bayes:Denominator}
p(y_{1:T}) = \int_{\theta} p(y_{1:T} | \theta)p(\theta) d\theta
\end{equation}
with the integration carried out over the entire support of $\theta$.
\\

From this formulation, Bayesian statistics can be interpreted as an update of beliefs, where knowledge of $\theta$ is able to be expressed in terms of a probability distribution describing the range of values $\theta$ could reasonably take. Before observing data, this belief takes the form of the prior, however data can then by incorporated via (\ref{Bayes:BayesRule}) to update this into the posterior distribution. If the observations $y_{1:T}$ are dependent on $\theta$, the incorporation of the information they contain leads to $p(\theta | y_{1:T})$ describing the distribution of $\theta$ with a greater precision than $p(\theta)$.
\\

\iffalse
Let $y_{T+h}$ for some $h \geq 1$ be unobserved, but described by the conditional distribution $p(y_{T+h} | y_{1:T}, \theta)$. Bayesian statistics allows for the uncertainty in $p(\theta | y_{1:T})$ to be included in the prediction distribution $p(y_{T+h} | y_{1:T})$ by the marginalising of $\theta$,
\begin{equation}
\label{Bayes:Predictive}
p(y_{T+h} | y_{1:T}) = \int_{\theta} p(y_{T+h} | y_{1:T}, \theta)p(\theta | y_{1:T}) d\theta.
\end{equation}
\fi

Bayes' Rule is valid even if the prior distribution is replaced with a posterior conditioned on data, as long as the likelihood does not arise from the distribution of the same data. Let $y_{1:T_1}$ and $y_{T_1 + 1:T_2}$ for $T_1 < T_2$ denote two distinct series of data, respectively distributed according to the likelihoods $p(y_{1:T_1} | \theta)$ and $p(y_{T_1+1:T_2} |  y_{1:T_1}, \theta)$. The posterior distribution $p(\theta | y_{1:T_1})$ can be updated to $p(\theta | y_{1:T_2})$, where $y_{1:T_2} = \{y_{1:T_1}, y_{T_1+1:T_2}\}$ by
\begin{equation}
\label{Bayes:BayesUpdating}
p(\theta | y_{1:T_2}) = \frac{p(y_{T_1+1:T_2} | y_{1:T_1}, \theta) p(\theta | y_{1:T_1})}{p(y_{T_1+1:T_2} | y_{1:T_1})},
\end{equation}
where
\begin{equation}
\label{Bayes:UpdateDenominator}
p(y_{T_1+1:T_2} | y_{1:T_1}) = \int_{\theta} p(y_{T_1+1:T_2} | y_{1:T_1}, \theta) p(\theta | y_{1:T_1}) d\theta.
\end{equation}
As  $p(\theta | y_{1:T_2})$ is conditioned on more data than  $p(\theta | y_{1:T_1})$, its describes the distribution of $\theta$ with a greater precision.
\\

The main aim of this thesis is to explore techniques to perform the posterior update (\ref{Bayes:BayesUpdating}) in a computationally efficient manner under certain dynamical models regarding the evolution of $\{y_t\}$.

\section{Bayesian Hierarchical Modelling}
\label{sec:BayesHier}

Consider the case where there are $N$ individual time series, $y_{i, 1:T}$ for $i = 1, 2, \ldots, N$, referred to as units, each distributed according to the likelihood $p(y_{i, 1:T} | \theta_i)$, where $\theta_i$ is a parameter vector associated with the $i$'th unit. We assume that each time series $y_{i, 1:T}$ is conditionally independent of each other $y_{j, 1:T}, j \neq i,$ given $\theta$. If independent priors $p_i(\theta_i)$ were placed on each parameter vector, the resulting posterior $p(\theta_1, \ldots, \theta_N | y_{1, 1:T}, \ldots, y_{N, 1:T})$ would decompose into the product of $N$ marginal posteriors,
\begin{equation}
\label{Hier:Independent}
p(\theta_{1:N} | y_{1:N, 1:T}) = \prod_{i=1}^N p(\theta_i | y_{i, 1:T}) \propto \prod_{i=1}^N p(y_{i, 1:T} | \theta_i) p_i(\theta_i)
\end{equation}
where $\theta_{1:N} = \{\theta_i | i = 1, 2, \ldots, N\}$ and $y_{1:N, 1:T} = \{y_{i, 1:T} | i = 1, 2, \ldots, N \}$.
\\

However, it may be useful to build a model where the above decomposition does not hold, so that information contained in $y_{i, 1:T}$ may be used in the marginal posterior for $\theta_j$ for some $j \neq i$. This can be accomplished by augmenting the model with a vector of hyperparameters, $\beta$, and treating each $\theta_i$ as an independent draw from some shared prior distribution $p(\theta_i | \beta)$. 
\\

The joint posterior can be inferred via Bayes' rule:
\begin{equation}
\label{Hier:Bayes}
p(\theta_{1:N}, \beta| y_{1:N, 1:T}) \propto p(\beta)\prod_{i=1}^N \bigg(p(y_{i, 1:T} | \theta_i) p(\theta_i | \beta) \bigg),
\end{equation}
for some hyperprior $p(\beta)$.
\\

Assuming conditional independence between each $\theta_i$ given $\beta$, this decomposes into
\begin{equation}
\label{Hier:Decompose}
p(\theta_{1:N}, \beta | y_{1:N, 1:T}) = p(\beta | y_{1:N, 1:T}) \prod_{i=1}^N p(\theta_i | \beta, y_{1:N, 1:T}).
\end{equation}

If an additional unit $i = N+1$ is observed, then the previous shared prior $p(\theta_{N+1} | \beta)$ and posterior distribution $p(\beta | y_{1:N, 1:T})$ may be used to for inference of the marginal posterior $p(\theta_{N+1} | y_{1:N+1, 1:T})$ by
\begin{equation}
\label{Hier:New}
p(\theta_{N+1} |y_{1:N+1, 1:T}) \propto p(y_{N+1, 1:T} | \theta_{N+1}) \int_{\beta} p(\theta_{N+1} | \beta)p(\beta | y_{1:N, 1:T}) d\beta.
\end{equation}

\subsection{Dirichlet Process Priors for Hierarchical Models}
\textit{If we use it later}

\section{Bayesian Inference} \label{sec:BayesInf}

Given an analytically tractable likelihood $p(y_{1:T} | \theta)$ and prior $p(\theta)$, the numerator of the posterior given by Bayes' rule is able to be evaluated for arbitrary choices of $\theta$ and $y_{1:T}$.
\\

However difficulty often arrives in the denominator, given by the integral $\int_{\theta} p(y_{1:T} | \theta)p(\theta) d\theta$. For a limited class of priors known as \textit{conjugate priors}, the prior and likelihood combination allows this to be analytically solved and the resulting posterior is known in closed form. Typically, there is no analytical solution to the integral, and numerical techniques must be used instead.
\\

One option to computing the posteiror or its features is numeric integration, when the integral may be treated as a simple Riemann Integral and hence a grid based approximation may be applied. However, the computation of this method is exponential in the dimension of $\theta$, so it is generally impractical for high dimensional problems. 
\\

Notice that many posterior quantities of interest, such as quantiles, loss functions, and predictive distributions, can be viewed as the expectation of some function $g(\theta)$ that is integrated with respect to the posterior distribution,
\begin{equation}
\label{Bayes:Expectation}
\mathbb{E}_{p} [g(\theta)] = \int_{\theta} g(\theta) p(\theta | y_{1:T}) d\theta. 
\end{equation}
A sample of $S$ draws $\theta^{(i)} \sim p(\theta | y_{1:T})$ may be used to approximate this by
\begin{equation}
\label{Bayes:Estimator}
\hat{\mathbb{E}}_p[g(\theta)] = \frac{1}{S} \sum_{i=1}^S g(\theta^{(i)}).
\end{equation}
The use of samples as a substitute to the actual posterior has motivated the development of sampling based inference algorithms, notably Importance Sampling, which allows expectations estimates to be calculated with respect to samples from a different distribution, and Markov Chain Monte Carlo (MCMC), which results in $S$ draws from the posterior. As $S$ increases, the sum in (\ref{Bayes:Estimator}) converges to the expectation (\ref{Bayes:Expectation}) due to the Weak Law of Large Numbers. Desipte the approximate, this class of inference algorithms are often referred to as `exact' inference as the approximation error can be made smaller by simply increasing the number of simulated draws.
\\

The computational cost of drawing a large number of posterior samples through MCMC is often prohibitively large, which has spurred interest in a wide range of alternative approaches to the calculation of posterior quantities. One prominent example is Variational Bayes (VB), which aims to select the member of a family of approximating distributions $q(\theta | y_{1:T})$ that is closest to $p(\theta | y_{1:T})$ using a divergence function. VB is often orders of magnitude faster than MCMC, but it is often difficult to measure the error introduced by the use of the resulting approximating distribution in place of exact posterior.
\\

\subsection{Importance Sampling}
\label{subsec:IS}

Recall the expectation of some function $g(\theta)$ with respect to the posterior $p(\theta | y_{1:T})$ introduced in (\ref{Bayes:Expectation}),
\begin{equation}
\label{IS:Expectation}
\mathbb{E}_{p} [g(\theta)] = \int_{\theta} g(\theta) p(\theta | y_{1:T}) d\theta. 
\end{equation}

If samples from the posterior distribution are available the estimator \begin{equation}
\label{IS:Estimator}
\hat{\mathbb{E}}_p[g(\theta)] = \frac{1}{S} \sum_{i=1}^S g(\theta^{(i)})
\end{equation}
where $\theta^{(i)} \sim p(\theta | y_{1:T})$ is available and converges to (\ref{IS:Expectation}) due to the Weak Law of Large Numbers.
\\

In many cases it may be computationally expensive to draw posterior samples. In this case, importance sampling may be used to form an estimator with respect to some other candidate distribution $r(\theta)$ \citep{Gelman2014},
\\

Multiping and dividing (\ref{IS:Expectation}) by $r(\theta)$ results in the importance sampling expectation,
\begin{align}
\mathbb{E}_{p} [g(\theta)] &= \int_{\theta} g(\theta) p(\theta | y_{1:T}) d\theta \nonumber \\
&= \int_{\theta} g(\theta) \frac{r(\theta)}{r(\theta)} p(\theta | y_{1:T}) d\theta \nonumber \\
&= \mathbb{E}_{r} \left[ w(\theta)g(\theta) \right] \label{IS:Expectation2}
\end{align}
where 
\begin{equation}
\label{IS:weights}
w(\theta) = \frac{p(\theta | y_{1:T})}{r(\theta)}
\end{equation}
are referred to as the importance sampler weights.
\\

This may be estimated by 
\begin{equation}
\label{IS:Estimator2}
\hat{\mathbb{E}}_r[w(\theta) g(\theta)] = \frac{1}{S} \sum_{i=1}^S g(\theta^{(i)}) w(\theta^{(i)}
\end{equation}
where $\theta^{(i)} \sim r(\theta)$.
\\

Importance sampling can be readily extended to the common scenario where the posterior distribution is unknown, but the joint distribution $p(y_{1:T}, \theta)$ is available. Noting that 
\begin{equation}
\label{IS:posterior}
p(\theta | y_{1:T}) = \frac{p(y_{1:T}, \theta)}{\int_{\theta} p(y_{1:T}, \theta) d\theta},
\end{equation}
the expectation (\ref{IS:Expectation}) may be written as 
\begin{equation}
\label{IS:Expectation3}
\mathbb{E}_{p} [g(\theta)] = \frac{\int_{\theta} g(\theta) p(y_{1:T}, \theta) d\theta}{\int_{\theta} p(y_{1:T}, \theta) d\theta}.
\end{equation}
This may be transformed into an expectation with respect to $r(\theta)$,
\begin{align}
\mathbb{E}_{p} [g(\theta)] &= \frac{\int_{\theta} g(\theta) p(y_{1:T}, \theta) d\theta}{\int_{\theta} p(y_{1:T}, \theta) d\theta}, \nonumber \\
&= \frac{\int_{\theta} \frac{r(\theta)}{r(\theta} g(\theta) p(y_{1:T}, \theta) d\theta}{\int_{\theta} \frac{r(\theta)}{r(\theta} p(y_{1:T}, \theta) d\theta}, \nonumber \\
&= \mathbb{E}_{r} \left[ w(\theta)g(\theta) \right]. \label{IS:Expectation4}
\end{align}
where 
\begin{equation}
w(\theta) =  \frac{p(y_{1:T}, \theta)}{r(\theta) \int_{\theta} \frac{r(\theta)}{r(\theta} p(y_{1:T}, \theta)}.
\end{equation}
The estimator (\ref{IS:Estimator2}) may be used in this scenario however the weights are calculated via the two step method
\begin{align}
\hat{w}(\theta^{(i)}) &=  \frac{p(y_{1:T}, \theta^{(i)})}{r(\theta^{(i)})}, \\
w(\theta^{(i)}) &= \frac{\hat{w}(\theta^{(i)})}{\frac{1}{S} \sum_{i=1}^S \hat{w}(\theta^{(i)})}.
\end{align}

The performance of importance sampling is highly dependent on the choice of $r(\theta)$. The effective sample size of the importance sampler Monte Carlo estimator, $S_{eff}$, is given by
\begin{equation}
S_{eff} = \frac{N}{1 + \mbox{var}(w(\theta))} \leq S
\end{equation}
from \cite{Kong1994}. The effective sample size is decreased, and estimator variance is increased, if the variance of the weights is large. This most often occurs if the tails of $p(\theta | y_{1:T})$ and $r(\theta)$ are mismatched \citep{Gelman2014} and $w(\theta)$ is small with high probability but extremely large for some small, non-zero, probability. 
\\

Choosing a candidate distribution that matches the tails of the posterior distribution is a difficult problem, and has lead to interest in other sampling based inference algorithms such as MCMC.

\subsection{Markov Chain Monte Carlo}
\label{subsec:MCMC}

There are many types of MCMC algorithms, with arguably the simplest and most commonly used one being the Gibbs sampler (\cite{Geman1984}, \cite{Gilks1995a}). The Gibbs sampler algorithm iteratively samples the components of a $k$-dimensional parameter vector $\theta$ via each of the so-called full conditional distributions as follows,
\begin{align}
&p(\theta_{1} | \theta_{2}, \ldots, \theta_{k}, y_{1:T}) \nonumber \\
&p(\theta_{2} | \theta_{1}, \theta_{3}, \ldots, \theta_{k}, y_{1:T}) \nonumber \\
&\vdots \nonumber \\
&p(\theta_{k} | \theta_{1}, \ldots, \theta_{k-1}, y_{1:T}). \nonumber
\end{align}
Under mild regularity conditions (see, e.g., \citet{Tierney1994}) and with enough iterations of the Markov chain that results from the Gibbs sampler, these samples converge in distribution to $p(\theta | y_{1:T})$. Samples taken before the MCMC converges to the posterior must be discarded, and the remaining samples may have strong dependence between consecutive draws of the same parameter due to the Markovian nature of the algorithm. The computation time for each iteration and the overall number of iterations required to accurately summarise the posterior distribution is problem specific and typically with the complexity of the model. Often the full conditional distributions cannot be recognised and a Metropolis-Hastings-within-Gibbs (MH) step is utilised instead \citep{Gilks1995b}.
\\

In an iteration of MH-MCMC, a draw from the exact full conditional distribution is replaced by sampling a candidate $\theta_j^{(c)}$, where $\theta_{j}$ may be any scalar or vector subset of $\theta$, from a proposal distribution $r(\theta_{j} | \cdot)$ and accepted into the Markov Chain with probability
\begin{equation}
\min \left\{ 1, \frac{p(\theta_{j}^{(c)} | \theta_{\neq j}^{(a)}, y_{1:T})}{p(\theta_{j}^{(a)} | \theta_{\neq j}^{(a)}, y_{1:T})} \times \frac{r(\theta_{j}^{(a)} | \theta_{j}^{(c)})}{r(\theta_{j}^{(c)} | \theta_{j}^{(a)})} \right\},
\label{MCMC:MHaccept}
\end{equation}
where the superscript $(a)$ denotes the most recently accepted value of the corresponding subset of $\theta$. If the sampler rejects a candidate value the previous value $\theta_{j}^{(a)}$ is repeated. Choice of proposal distribution is left to the user with arguably the most simple being the Normal Random Walk proposal,
\begin{equation}
\theta_{j}^{(c)} \sim N(\theta_{j}^{(a)}, \Sigma_{j})
\label{MCMC:RWprop}
\end{equation}
6as $r(\theta_{j}^{(c)} | \theta_{j}^{(a)}) = r(\theta_{j}^{(a)} | \theta_{j}^{(c)})$ and the acceptance probability simplifies to
\begin{equation}
\min \left\{ 1, \frac{p(\theta_{j}^{(c)} | \theta_{\neq j}^{(a)}, y_{1:T})}{p(\theta_{j}^{(a)} | \theta_{\neq j}^{(a)}, y_{1:T})} \right\}.
\label{MCMC:RWMHaccept}
\end{equation}

Performance of this Random Walk Metropolis Hastings MCMC (RWMH-MCMC) then depends on the acceptance rate of the proposal distribution, largely influenced by the value of $\Sigma_{j}$. Large values of $\Sigma_{j}$ allows the Markov chain to explore different areas of the   variable support, at a reduced acceptance rate if these areas have low posterior probability. Conversely, low values of $\Sigma_{j}$ lead to higher rate of accepted draws as $p(\theta_{j}^{(c)} | \theta_{\neq j}^{(a)}, y_{1:T})$ is similar to $p(\theta_{j}^{(a)} | \theta_{\neq j}^{(a)}, y_{1:T})$, but the Markov chain may take more iterations to fully explore the posterior support. \citet{Gelman1997} shows that the optimal acceptance rate of a multivariate normal random walk proposal distribution is $23.4\%$. This rate can be targeted by the adaptive algorithm for $\Sigma_j$ from \citet{Garthwaite2016}, where $\Sigma_j$ is increased or decreased if the most recent candidate was respectively rejected or accepted. 
\\

RWMH-MCMC is used throughout the thesis with the adaptive $\Sigma_j$ of \citet{Garthwaite2016} to provide an exact inference baseline result for comparision with Variational Bayes.

\subsection{Variational Bayes} \label{subsec:VB}

As discussed earlier, a typically faster, albeit approximate, alternative to MCMC based inference is Variational Bayes (VB, see \citet{Blei2017} for a recent review). VB posits a family of parametric approximating distributions $q_{\lambda}(\theta | y_{1:T})$, parameterised by an auxiliary vector $\lambda$, that shares the same support as the true posterior distribution $p(\theta | y_{1:T})$. Note that $q_{\lambda}(\theta | y_{1:T})$ does not neccesarily depend on $y_{1:T}$, but this notation is used to make it clear that this distribution is an approximation for $p(\theta | y_{1:T})$. A member of the approximating family is chosen to minimise some error function, traditionally the Kullback-Leibler (KL) divergence from $q_{\lambda}(\theta | y_{1:T})$ to $p(\theta | y_{1:T})$, denoted by $KL[q_{\lambda}(\theta | y_{1:T})\hspace{.1cm}||\hspace{.1cm}p(\theta | y_{1:T})]$ \citep{Kullback1951}. The KL divergence is defined by
\begin{equation}
\label{VB:KL-def}
KL[q_{\lambda}(\theta | y_{1:T})\hspace{.1cm}||\hspace{.1cm}p(\theta | y_{1:T})] = \mathbb{E}_{q} \left[ \log(q_{\lambda}(\theta | y_{1:T})) - \log(p(\theta | y_{1:T})) \right],
\end{equation}
and is a non-negative, asymmetric measure of the discrepancy between $p(\theta | y_{1:T})$ and \\ $q_{\lambda}(\theta | y_{1:T})$  that will be equal to zero if and only if $p(\theta | y_{1:T}) = q_{\lambda}(\theta | y_{1:T})$ almost everywhere \citep{Bishop2006}.
\\

Typically (\ref{VB:KL-def}) cannot be evaluated, and Monte Carlo estimates
\begin{equation}
\label{VB:KL-MC}
KL[q_{\lambda}(\theta | y_{1:T})\hspace{.1cm}||\hspace{.1cm}p(\theta | y_{1:T}] \approx \frac{1}{S}\sum_{i=1}^S \left(\log(q_{\lambda}(\theta^{(i)} | y_{1:T})) - \log(p(\theta^{(i)} | y_{1:T})) \right)
\end{equation}
where $\theta^{(i)} \sim q_{\lambda}(\theta | y_{1:T}))$ are compuationally infeasible due to the inclusion of the term $p(\theta | y_{1:T})$, which is only known up to proportionality. Instead VB uses the Evidence Lower Bound (ELBO), denoted by $\mathcal{L}(q, \lambda)$, as an error function where
\begin{equation}
\label{VB:ELBO}
\mathcal{L}(q, \lambda) = \mathbb{E}_{q} \left[\log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))\right],
\end{equation}
which is evaluated with Monte Carlo estimates
\begin{equation}
\label{VB:ELBO-MC}
\mathcal{L}(q, \lambda) \approx \frac{1}{S} \sum_{i=1}^S \left(\log(p(y_{1:T}, \theta^{(i)})) - \log(q_{\lambda}(\theta^{(i)} | y_{1:T})) \right)
\end{equation}
where $\theta^{(i)} \sim q_{\lambda}(\theta | y_{1:T})$. It can be shown that
\begin{equation}
\label{VB:ELBO-KL}
\mathcal{L}(q, \lambda) = - KL[q_{\lambda}(\theta | y_{1:T})\hspace{.1cm}||\hspace{.1cm}p(\theta | y_{1:T}] + \log(p(y_{1:T}))
\end{equation}
and hence maximising (\ref{VB:ELBO}) with respect to $q_{\lambda}(\theta | y_{1:T})$ is equivalent to minimising (\ref{VB:KL-def}).
\\

Two major implementations of Variational Bayes are discussed, referred to as Mean Field Variational Bayes and Stochastic Variational Bayes. Each of these implementations takes advantage of the functional form of the ELBO, and offers a computationally efficient algorithm to for finding $q_{\lambda}(\theta | y_{1:T})$ that is optimal within a particular class of distributions. 
\\

The statistical properties of $q_{\lambda}(\theta | y_{1:T})$ in MFVB and SVB are largely driven by the choice of the KL divergence from $q_{\lambda}(\theta | y_{1:T})$ to $p(\theta | y_{1:T})$ as an error function, and there is significant interest in other error functions, such as the KL divergence from $p(\theta | y_{1:T})$ to $q_{\lambda}(\theta | y_{1:T})$ \citep{Minka2001}, R\'enyi divergence \citep{Li2016}, Stein's discrepency (\citealp{Liu2016}; \citealp{Ranganath2016b}), and $\chi$-divergence \citep{Dieng2017}.

\subsection{Mean Field Variational Bayes} \label{subsec:MFVB}

Mean Field Variational Bayes (MFVB) has its origin in physics \citep{Chandler1987, Peterson1987} and has since developed a large literature in machine learning, with major early contributions provided by \citet{Jordan1999}, \citet{Ghahramani2000a}, \citet{Beal2003}, and \citet{Wainwright2008}.
\\

MFVB restricts the class of distributions from which to select $q_{\lambda}(\theta | y_{1:T})$ to the set of factorisable distributions,
\begin{equation}
\label{MFVB:Product}
q_{\lambda}(\theta | y_{1:T}) = \prod_{i=1}^k q_i(\theta_i | y_{1:T}),
\end{equation}
where each $\theta_i$ is a distinct subset of $\theta$. These $k$ subsets  $\theta$ may vectors, but in most implementations are scalars so $k$ is equal to the dimension of $\theta$. Each $q_i$ has an associated vector $\lambda_i$ that may have a different dimension between each $i$, which is an auxillary parameter vector for the relevant factor. The shorthand $q_i$ is used in this section to refer to the factor $q_i(\theta | y_{1:T})$. 
\\

MFVB is widely used as it greatly simplifies maximisation of the ELBO, and has a straight-forward implementation \citep{Wainwright2008}.  Maximising the ELBO with respect to $q_i$ is as analytically involved as deriving the conditional distributions used in Gibbs based MCMC schemes, but the resulting algorithm to maximise the ELBO is much simpler computationally than sampling from a Gibbs MCMC scheme. MFVB factorises (\ref{VB:ELBO}) into a product of $k$ separate components that each contain only one $q_i$, and then maximises each of these factors with respect to the associated $q_i$. Using the notation $q_{\setminus i} = \prod_{j\neq i}q_j$, \citet{Attias1999} shows that each component is maximised by
\begin{equation}
\label{MFVB:solution}
q_i(\theta | y_{1:T}) \propto\exp( \mathbb{E}_{q_{\setminus i}} [\ln(p(y_{1:T},\theta))]).
\end{equation}

This optimisation is analytically tractable if the model likelihood belongs to the exponential family, and the prior is conjugate to this likelihood \citep{Beal2003}.
\\

In this case, the optimal distributional family of each factor $q_i$ is the same as the prior, analogous to the relationship between a prior-posterior conjugate pair. 
\\

In the event that a non-conjugate prior is used, the form of the distribution in (\ref{MFVB:solution}) is unrecognisable, a further approximation may be used; substituting in another distribution to replace the problematic $q_i$. A common option is the use of a Gaussian distribution through either a Laplace approximation \citep{Friston2006} or a Delta Method inspired transform \citep{Wang2013}. This secondary level of approximation is analogous to the requirement of a Metropolis-Hastings-within-Gibbs step in MCMC to handle unrecognisable full conditional distributions.
\\

MFVB is illustrated for the simple model where $y_t$ for $t = 1, 2, \ldots, T$ and $\theta = \{\mu, \sigma^2\}$. Each $y_t$ is distributed independently according to  $y_t \sim \mathcal{N}(\mu, \sigma^2)$, and prior distributions are given by $\mu \sim \mathcal{N}(\gamma, \sigma^2 / tau)$ and $\sigma^2 \sim IG(\mbox{shape} = \alpha, \mbox{scale =}\beta)$. In this model the likelihood of each observation $y_t$ is given by
\begin{equation}
\label{MFVB:exampleLikelihood}
p(y_t | \theta) = \frac{1}{\sqrt{2 \pi}} \sigma^{-1}\exp\left\{ \frac{-(y_t - mu)^2}{2 \sigma^2} \right\},
\end{equation}
and the prior distribution can be written as
\begin{equation}
\label{MFVB:examplePrior}
p(\mu, \sigma^2) = \frac{\sqrt{\tau}\beta^{\alpha}}{\sqrt{2 \pi} \Gamma{\alpha}} \sigma^{-(2\alpha + 3)} \exp \left\{ \frac{-\tau (\mu - \gamma)^2 - 2\beta}{2 \sigma^2} \right\}
\end{equation}

In this case, from (\ref{MFVB:solution}), the optimal approximating distribution  \\ $q_{\lambda}(\theta | y_{1:T}) = q_1(\mu | y_{1:T})q_2(\sigma^2 | y_{1:T})$ is given by the factors
\begin{align}
q_1(\mu | y_{1:T}) &\propto \exp \left\{\mathbb{E}_{q_2} \left[ \sum_{t=1}^T \log(p(y_t | \theta)) + \log(p(\theta)) \right] \right\} \\ 
&\propto \exp \left\{ \mathbb{E}_{q_2} \left[ -(T + 2\alpha + 3)\log(\sigma) - \frac{-\sum_{t=1}^T (y_t - \mu)^2 -\tau (\mu - \gamma)^2 - 2\beta}{2 \sigma^2} \right] \right\} \\
&\propto \exp \left\{ \frac{-(T + \tau) \mathbb{E}_{q_2}[\sigma^{-2}]}{2} \left( \mu - \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau} \right)^2 \right\} \label{MFVB:1}
\end{align}
and
\begin{align}
q_2(\sigma^2 | y_{1:T})  &\propto \exp \left\{\mathbb{E}_{q_1} \left[ \sum_{t=1}^T \log(p(y_t | \theta)) + \log(p(\theta)) \right] \right\} \\ 
&\propto \exp \left\{ \mathbb{E}_{q_1} \left[ -(T + 2\alpha + 3)\log(\sigma) - \frac{-\sum_{t=1}^T (y_t - \mu)^2 -\tau (\mu - \gamma)^2 - 2\beta}{2 \sigma^2} \right] \right\} \\
&\propto \sigma^{-2((T+1)/2 + \alpha + 1)} \nonumber \\
\times &\exp \left\{ \frac{ \beta + 1/2\left((\tau + T)\mathbb{E}_{q_1}[\mu^2] - 2 \mathbb{E}_{q_1}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau \right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2\right)}{-\sigma^2} \right\}. \label{MFVB:2}
\end{align}
As the prior distributions used were conjugate to the likelihood, the Variational Bayes optimal distributions for $q_1(\mu |  y_{1:T})$ and $q_2(\sigma^2 | y_{1:T})$ are the same family as the prior distributions, and are denoted by $q_1(\mu | y_{1:T}) \sim \mathcal{N}(\tilde{\gamma}, \tilde{\tau})$ and $q_2(\sigma^2 | y_{1:T}) \sim IG(\tilde{\alpha}, \tilde{\beta})$. 
\\

Denoting the MFVB parameters as $\lambda = (\tilde{\gamma}, \tilde{\tau}, \tilde{\alpha}, \tilde{\beta})'$, from (\ref{MFVB:1}) and (\ref{MFVB:2}) it follows that
\begin{align}
\tilde{\gamma} &= \frac{\tau \gamma + \sum_{t=1}^{T} y_t}{T + \tau}  \label{MFVB:3} \\ 
\tilde{\tau} &= \left((T + \tau)\mathbb{E}_{q_2}[\sigma^{-2}]\right )^{-1} \label{MFVB:4} \\
\tilde{\alpha} &= (T+1)/2 + \alpha  \label{MFVB:5} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)\mathbb{E}_{q_1}[\mu^2 ]- 2 \mathbb{E}_{q_1}[\mu ]\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right). \label{MFVB:6}
\end{align}.

The expectations in (\ref{MFVB:4}) and (\ref{MFVB:6}) are available in a closed form for these distributions, and these are substituted in,
\begin{align}
\tilde{\tau} &= \frac{\tilde{\beta}}{\tilde{\alpha}(T + \tau)} \label{MFVB:7} \\
\tilde{\beta} &= \beta + 1/2\left((\tau + T)(\tilde{\gamma}^2 + \tilde{\tau}) - 2 \tilde{\gamma}\left(\sum_{t=1}^{T}y_t + \tau\right) + \sum_{t=1}^{T} y_t^2 + \tau \gamma^2 \right).\label{MFVB:8}
\end{align}

The equations (\ref{MFVB:3}), (\ref{MFVB:5}), (\ref{MFVB:7}), and (\ref{MFVB:8}) consitute the mean field equations for this problem, the simultaneous fulfilment of which satisfy (\ref{MFVB:solution}) and hence maximises the ELBO. (\ref{MFVB:3}) and (\ref{MFVB:5}) can be solved independently, but (\ref{MFVB:7}) and (\ref{MFVB:8}) contain a circular dependence. The MFVB approximation requires an algorithm that cycles between these two equations until the change in the ELBO is less than some pre-specified small constant $\epsilon$. This algorithm is known as a coordinate ascent, and is described by Algorithm \ref{alg:MFVB}.
\\

At each step coordinate ascent will fix one parameter and then maximise the ELBO with respect to the other parameter, so that changes in parameters results in a non-decreasing sequence of ELBO values. Coordinate ascent converges only when neither parameter can be changed independently to increase the ELBO by more than $\epsilon$, and hence reaches a local maximum.
\\

The current theory for Variational Bayes does not offer any insight to the shape of the ELBO so determining if the algorithm converges to a local or global maximum is difficult. One option to alleviate this problem is to start the algorithm at a range of initial values and choose parameter values with the maximum converged ELBO, and hence lowest KL divergence to the true posterior. 
\\

\vspace{2mm}
\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Set of mean field equations}
 \KwResult{Mean Field Approximation}
 Set $m = 1$.\;
 Set $\tilde{\gamma}$ using \ref{MFVB:3}.\;
 Set $\tilde{\alpha}$ using \ref{MFVB:5}.\;
 Initialise $\tilde{\tau}^{(1)}$ and $\tilde{\beta}^{(1)}$ randomly.\;
 Evaluate $\mathcal{L}(q, \lambda^{(1)})$ with the Monte Carlo estimator (\ref{VB:ELBO-MC}) using $\lambda^{(1)}$.\;
 \While{$\mathcal{L}(q, \lambda^{(m)}) - \mathcal{L}(q, \lambda^{(m-1)}) > \epsilon$}{
  Set $m = m + 1$.\;
  Set $\tilde{\tau}^{(m)}$ using \ref{MFVB:7}.\;
  Set $\tilde{\beta}^{(m)}$ using \ref{MFVB:8}.\;
  Evaluate $\mathcal{L}(q, \lambda^{m})$.\;
 }
 \caption{Coordinate Ascent for Mean Field Variational Bayes}
 \label{alg:MFVB}
\end{algorithm}

\vspace{4mm}
Data was simulated according to the Gaussian model descibed above with $T = 100, \mu = 2$ and $\sigma^2 = 1$. The hyperparameters were set to $\gamma = 0, \tau = 1, \alpha = 1$ and $\beta = 1$. Figure (\ref{fig:MFVB}) compares the analytically available true posterior marginal densities for both $\mu$ and $\sigma^2$ (black) with the densities generated by the Monte Carlo sampler (red) and the approximation from Mean Field Variational Bayes (blue). For both parameters all three densities are indistinguishable.

\begin{figure}[htbp ]
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/norminvg}
\caption{Marginal posterior densities for the model described in Section (\ref{subsec:MFVB}). The true density is in red and the MFVB approximation is blue.}
\label{fig:MFVB}
\end{figure}

\subsection{Stochastic Variational Bayes} \label{subsec:SVB}

The requirement for an exponential family model without introducing further approximations, as well as restricting the approximating distribution to the factorisable family may be unsatisfactory in many applications.
\\

\citet{Paisley2012}, \citet{Hoffman2013} and \citet{Ranganath2014} have adapted a gradient ascent algorithm for use in Variational Bayes, which selects the optimal auxiliary parameter vector $\lambda$ for a given functional form of $q_{\lambda}(\theta | y_{1:T})$. This implementation is referred to as Stochastic Variational Bayes (SVB).
\\

For a fixed choice of $q$ SVB maximises the ELBO with respect to $\lambda$ through stochastic gradient ascent, where repeated Monte Carlo estimates of $\partial\mathcal{L}(q, \lambda) / \partial \lambda$, as $\widehat{\partial\mathcal{L}(q, \lambda) / \partial \lambda}$, are calculated. The gradient estimates are used to update $\lambda$ through the simple recursion
\begin{equation}
\label{SVB:gradientAscent}
\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}} \bigg\rvert_{\lambda = \lambda^{(m)}}
\end{equation}
until the change from $\mathcal{L}(q, \lambda^{(m)})$ to $\mathcal{L}(q, \lambda^{(m+1)})$ falls below some pre-specified threshold \citep{Hoffman2013}. Intuitively, individual elements of $\lambda$ will increase if the estimate of the slope of $\mathcal{L}(q, \lambda^{(m)})$ is positive at the current point, and will decrease if that estimate is negative, until each element of $\lambda$ reaches a point where the slope is zero. This procedure is guaranteed to converge to a local maximum \citep{Robbins1951} if the sequence $\rho^{(m)}, m = 1, 2 , \ldots$, satisfies
\begin{align}
&\sum_{m=1}^{\infty} \rho^{(m)} =  \infty \\
&\sum_{m=1}^{\infty} (\rho^{(m)})^2 <  \infty.
\end{align}
In this paper the sequence $\rho^{(m)}$ is provided by the Adam algorithm of \citet{Kingma2014b} which is described in the appendix.
\\

Three popular choices for the Monte Carlo estimator of $\partial\mathcal{L}(q, \lambda) / \partial \lambda$ are discussed: the score gradient estimator of \citet{Ranganath2014}, the reparameteterised estimator of \citet{Kingma2014a} and the generalised reparameterised estimator of \citet{Ruiz2016}, described below. Each gradient estimator places different restrictions on the choice of model and approximation, which in general are much less restictive tham MFVB.

\subsubsection{The Score Gradient Estimator}
\label{subsubsec:SVBScore}
The score gradient estimator can be derived by taking the derivative of (\ref{VB:ELBO}) with respect to $\lambda$,

\begin{align}
\frac{\partial \mathcal{L}(q, \lambda)}{\partial \lambda} &= \frac{\partial}{\partial \lambda}\int q_{\lambda}(\theta | y_{1:T}) \left(\log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))\right) d\theta \nonumber \\
&= \int  \frac{\partial}{\partial \lambda}\bigg[ q_{\lambda}(\theta | y_{1:T}) \left(\log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))\right) \bigg] d\theta \label{SVB:deriv1} \\
&= \int \frac{\partial q_{\lambda}(\theta | y_{1:T})}{\partial \lambda}  \left(\log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))\right) d\theta \nonumber \\
&+ \int q_{\lambda}(\theta | y_{1:T}) \left( \frac{\partial \log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))}{\partial \lambda} \right) d\theta \nonumber \\
&= \int \frac{\partial q_{\lambda}(\theta | y_{1:T})}{\partial \lambda}  \left(\log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))\right) d\theta \nonumber \\
&- \int q_{\lambda}(\theta | y_{1:T})\frac{\partial \log(q_{\lambda}(\theta | y_{1:T}))}{\partial \lambda} d\theta \label{SVB:deriv2}
\end{align}
where (\ref{SVB:deriv1}) follows from the dominated convergence theorem \citep{Casella2002}, noting that the final term in (\ref{SVB:deriv2}) is the expected value of the score of $q_{\lambda}(\theta | y_{1:T}))$, which is equal to zero. 
\\

This result, and the fact that $\partial q_{\lambda}(\theta | y_{1:T}) / \partial \lambda = \left[\partial \log(q_{\lambda}(\theta | y_{1:T})) / \partial \lambda \right] q_{\lambda}(\theta | y_{1:T})$, is substituted into (\ref{SVB:deriv2}),
\begin{align}
\frac{\partial \mathcal{L}(q, \lambda)}{\partial \lambda} &=  \int  q_{\lambda}(\theta | y_{1:T}) \frac{\partial \log(q_{\lambda}(\theta | y_{1:T}))}{\partial \lambda}  \left(\log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))\right) d\theta \nonumber \\
&= \mathbb{E}_{q} \left[ \frac{\partial \log(q_{\lambda}(\theta | y_{1:T}))}{\partial \lambda}  \left(\log(p(y_{1:T}, \theta)) - \log(q_{\lambda}(\theta | y_{1:T}))\right) \right]. \label{SVB:deriv3}
\end{align}

The expectation in (\ref{SVB:deriv3}) is often intractable, so stochastic Monte Carlo estimates of the gradient of the ELBO are instead taken by
\begin{equation}
\label{SVB:scoreDeriv}
\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}_{SC} = \frac{1}{S} \sum_{i = 1}^S \frac{\partial \log(q_{\lambda}(\theta^{(i)} | y_{1:T}))}{\partial \lambda} \left(\log(p(y_{1:T}, \theta^{(i)})) - \log(q_{\lambda}(\theta^{(i)} | y_{1:T})) \right),
\end{equation}
where $\theta^{(i)} \sim q_{\lambda}(\theta | y_{1:T})$.
\\

This estimator often has a very large variance, which can be reduced with control variates \citep{Ross2002}. 
\\

Control variates replace the function being estimated via Monte Carlo simulation with a different function with the same expectation but reduced variance. Referring to the original function in (\ref{SVB:deriv3}) as $g(\theta)$, a new function $h(\theta)$ is defined as
\begin{equation}
\label{SVB:CV1}
h(\theta) = g(\theta) - a(j(\theta) - \mathbb{E}_q(j(\theta))).
\end{equation}
Note that $\mathbb{E}_q(g(\theta)) = \mathbb{E}_q(h(\theta))$, and 
\begin{equation}
\label{SVB:CV2}
\mbox{Var}(h(\theta)) = \mbox{Var}(g(\theta)) + a^2 \mbox{Var}(j(\theta)) - 2 a \mbox{Cov}(g(\theta, j(\theta))).
\end{equation}
Control variates require some function $j(\theta)$ with a known mean and a large positive covariance with $g(\theta)$ so that $\mbox{Var}(h(\theta)) < \mbox{Var}(g(\theta))$ for some choice of $a$.
\\

A possible choice is 
\begin{equation}
\label{SVB:CV3}
j(\theta) = \frac{\partial \log(q_{\lambda}(\theta | y_{1:T}))}{\partial \lambda}
\end{equation}
as this has an expected value of zero, and is already calculated as a by-product of (\ref{SVB:scoreDeriv}).
\\

The final component to consider is the choice of $a$ to minimises $\mbox{Var}(h(\theta))$.  By taking the derivative of (\ref{SVB:CV2}) with respect to $a$ and setting this to zero, the variance minimising value of $a$ is
\begin{equation}
\label{SVB:CV4}
\hat{a} = \frac{\mbox{Cov}(g(\theta), j(\theta))}{\mbox{Var}(j(\theta))}
\end{equation}
which can be estimated from the $S$ Monte Carlo samples. The final score gradient estimator is given by
\begin{equation}
\label{SVB:scoreDeriv2}
\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}_{SC2} = \frac{1}{S} \sum_{i = 1}^S \frac{\partial \log(q_{\lambda}(\theta^{(i)} | y_{1:T}))}{\partial \lambda} \left(\log(p(y_{1:T}, \theta^{(i)})) - \log(q_{\lambda}(\theta^{(i)} | y_{1:T})) - \hat{a} \right).
\end{equation}

\subsubsection{The Reparameterised Gradient Estimator}
\label{subsubsec:SVBRP}
An alternative, lower variance approach (see eg. \cite{Rezende2014}; \cite{Ruiz2016}), is to estimate the gradient of the ELBO by reparameterisation. Reparameterisation introduces an auxiliary variable $\epsilon$ and differentiable transformation $\theta = f(\epsilon, \lambda)$ to rephrase Variational Bayes optimisation as the equivalent search for the set of parameters $\lambda$ for a fixed transformation $f$ that minimises the Kullback Leibler divergence from some distribution $r(\epsilon)$ with zero free parameters to the implied posterior distribution of $\epsilon$,
\begin{equation}
\label{SVB:rpDist}
p(f(\epsilon, \lambda) | y_{1:T}) = p(\theta | y_{1:T}) |J^{-1}(\epsilon, \lambda)|
\end{equation}
where $J(\epsilon, \lambda)$ is the Jacobian Matrix of the transformation $f(\epsilon, \lambda)$. Examples of $f$ and $r(\epsilon)$ include treating $\theta$ as location-scale transformation from a standard normal $r(\epsilon)$, or an inverse-CDF transformation from a uniform$(0, 1)$ $r(\epsilon)$. 
\\

The ELBO can be reparameterised by substituting $p(y_{1:T}, \theta) = p(y_{1:T}, f(\epsilon, \lambda))|J(\epsilon, \lambda)|$ into (\ref{VB:ELBO}),
\begin{equation}
\label{SVB:rpELBO}
\mathcal{L}(q, \lambda) = \mathbb{E}_{r(\epsilon)} \bigg[\log(p(y_{1:T}, f(\epsilon,\lambda))|J(\epsilon, \lambda)|) - \log(r(\epsilon))\bigg].
\end{equation}

The gradient of the reparameterised ELBO with respect to $\lambda$ is given by 
\begin{align}
\label{SVB:rpELBODeriv}
\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda} &= \frac{\partial}{\partial \lambda} \bigg( \mathbb{E}_{r(\epsilon)} \bigg[\log\big(p(y_{1:T}, f(\epsilon,\lambda))|J(\epsilon, \lambda)|\big) - \log(r(\epsilon))\bigg] \bigg) \nonumber \\
&= \mathbb{E}_{r(\epsilon)} \left[ \frac{\partial}{\partial \lambda} \bigg(\log(p(y_{1:T}, f(\epsilon,\lambda))) + \log(|J(\epsilon, \lambda)|) - \log(r(\epsilon)) \bigg)\right] \nonumber \\
&= \mathbb{E}_{r(\epsilon)} \left[ \frac{\partial \log(p(y_{1:T}, f(\epsilon,\lambda)))}{\partial f(\epsilon,\lambda)} \frac{\partial f(\epsilon,\lambda)}{\partial \lambda}  + \frac{\partial \log(|J(\epsilon, \lambda)|)}{\partial \lambda} \right].
\end{align}

This form leads to the reparameterised gradient estimator,
\begin{equation}
\label{SVB:rpDeriv}
\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}_{RP} = \frac{1}{S}  \sum_{i = 1}^S \frac{\partial f(\lambda, \epsilon^{(i)})}{\partial \lambda} \frac{\partial \log(p(y_{1:T}, \theta))}{\partial \theta} \bigg\rvert_{\theta = f(\lambda, \epsilon^{(i)})} + \frac{\partial \log(J(\lambda, \epsilon^{(i)}))}{\partial \lambda}, 
\end{equation}
where $\epsilon^{(i)} \sim r(\epsilon)$. 
\\

The approximating distribution $q_{\lambda}(\theta | y_{1:T})$ can then be recovered by applying $\theta = f(\epsilon, \lambda)$ to $r(\epsilon)$,
\begin{equation}
\label{SVB:rpQ}
q_{\lambda}(\theta | y_{1:T}) = r(f(\epsilon, \lambda)) |J(\epsilon, \lambda)| 
\end{equation}

This approach treats $q_{\lambda}(\theta | y_{1:T})$ as the distribution implied by the transformation $f$ of $r(\epsilon)$, restricting the class of approximating families that can be used.

\subsubsection{The Generalised Reparameterised Gradient Estimator}
\label{subsubsec:SVBGRP}
A third gradient estimator, the generalised reparameterisation estimator of \citet{Ruiz2016} allows $r(\epsilon)$ to depend on the auxiliary vector $\lambda$ so long as the first moment of $r(\epsilon)$ is independent of $\lambda$. 
\\

Reparameteried approximating distributions $q_{\lambda}(\theta | y_{1:T})$ cannot be constructed to belong to common families such as the Gamma, Beta, and Dirichlet distribution. With a careful choice of $r(\epsilon)$ and $f$, these can be constructed via generalised reparameterisation, which retains the benefits of a relatively low variance compared to the score gradient estimator.
\\

Gradient estimates can be obtained by taking Monte Carlo estimates of 
\begin{equation}
\label{SVB:grp}
\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda} = \textbf{g}^{rep} + \textbf{g}^{corr} - \frac{\partial}{\partial \lambda} \mathbb{E}_{q} \left[ \log(q_{\lambda}(\theta | y_{1:T})) \right]
\end{equation}
where
\begin{align}
\textbf{g}^{rep} =  \mathbb{E}_{q} &\left[\frac{\partial \log(p(y_{1:T}, \theta))}{\partial \theta} \frac{\partial f^{-1}(\theta, \lambda)}{\partial \lambda} \right] \\
\textbf{g}^{corr} =  \mathbb{E}_{q} &\left[ \log(p(y_{1:T}, \theta)) \left( \frac{\partial \log(q_{\lambda}(\theta | y_{1:T}))}{\partial \theta} \frac{\partial f^{-1}(\theta, \lambda)}{\partial \lambda}  \right. \right. \nonumber \\
&+ \left. \left. \frac{\partial \log(q_{\lambda}(\theta | y_{1:T}))}{\partial \lambda} + \frac{\partial |J^{\prime}(\theta, \lambda)|}{\partial \lambda} \right) \right]
\end{align}
where $f^{-1}(\theta, \lambda)$ is the inverse of $f(\epsilon, \lambda)$, evaluated at $\theta$, and $J^{\prime}(\theta, \lambda)$ is the Jacobian matrix of $f^{-1}(\theta, \lambda)$.
\\

If $f$ is the identity function, $\textbf{g}^{rep} = 0$ and the score gradient estimator is recovered. Alternatively if $r(\epsilon)$ does not depend on $\lambda$, $\textbf{g}^{corr} = 0$ and the reparameterised gradient estimator is recovered.
\\

An outline of SVB is provided in Algorithm \ref{alg:SVB}.
\\

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Approximation family $q$ or auxiliary distribution $r$ and function $f$}
 \KwResult{Variational Approximation}
 Set $m = 1$\;
 Initialise $\lambda^{(1)}$ randomly\;
 Evaluate $\mathcal{L}(q, \lambda^{(1)})$ with the Monte Carlo estimator (\ref{VB:ELBO-MC}) using $\lambda^{(1)}$\;
 \While{$|\mathcal{L}(q, \lambda^{(m)} - \mathcal{L}(q, \lambda^{(m-1)})| > \epsilon$}{
  Set $m = m + 1$\;
  Simulate $\theta^{(i)}$ for $i = 1, \ldots S$ from $q_{\lambda}(\theta | y_{1:T})$\;
  Estimate $\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}$ via (\ref{SVB:scoreDeriv2})\;
  OR\;
  Simulate $\epsilon^{(i)}$ for $i = 1, \ldots S$ from $r(\epsilon)$\;
  Estimate $\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}$ via (\ref{SVB:rpDeriv})\;
  OR\;
  Simulate $\theta^{(j)}$ for $i = 1, \ldots S$ from $q_{\lambda}(\theta | y_{1:T})$\;
   Estimate $\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}$ via (\ref{SVB:grp})\;
  Apply Update $\lambda^{(m)} = \lambda^{(m-1)} + \rho^{(m-1)} \widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}$\;
  Evaluate $\mathcal{L}(q, \lambda^{(m)})$\;
 }
 \caption{Stochastic Gradient Ascent for Variational Bayes}
  \label{alg:SVB}
\end{algorithm}

\subsubsection{Randomised Quasi Monte Carlo}
\label{subsubsec:SVBRQMC}

Choosing the number of samples per estimate, $S$, involves a trade-off between the computation time per gradient estimate and the stochastic noise present in each estimate. An increased value of $S$ will the Monte Carlo error, and generally reduce the number of iterations required for the ELBO to converge, at a linear increase in computation time per iteration.
\\

To reduce the Monte Carlo error, following \citet{Gunawan2017} Randomised Quasi Monte Carlo (RQMC) is used instead, which has shown to be more efficient than standard Monte Carlo in many applications, including Variational Bayes (see \cite{Niederreiter1992, Caflisch1998}, and \cite{Buchholz2018}). In RQMC, a deterministic sequence of numbers in the $k-$dimensional unit hypercube $[0, 1)^k$, where $k$ is the dimension of $\theta$ are generated. This sequence designed to have a low star discrepancy, so that the proportion of points in any $k-$dimensional subspace of $[0, 1)^k$ is close to the volume of that subspace.
\\

These are randomised each iteration to preserve the low discrepency properties while inducing a central limit theorem so that the gradient estimate is unbiased. See \citet{Matousek1998} for more details.
\\

Selecting $S$ of the resulting coordinates in the unit hypercube, and transforming these using the inverse-CDF of either $q_{\lambda}(\theta | z)$ or $r(\epsilon)$ simulates $S$ draws that can be used in any of the discussed gradient estimators.

\subsubsection{Time Series Example}
\label{subsubsec:SVBTS}

Some of the major advantages of SVB relative to MFVB are its flexibility to be applied to non exponential family models, and the ability to include parameter dependence in the approximating distribution. To illustrate this, consider the second order auto-regressive time series model, denoted by AR(2), and described by 
\begin{equation}
\label{SVBTS:AR2}
y_t = \phi_1 y_{t-1} + \phi_2 y_{t-2} + \epsilon_t
\end{equation}
where $\epsilon_t \sim \mathcal{N}(0, \sigma^2)$ for $t = 3, 4 \ldots, T$. The first two observations, $y_1$ and $y_2$ are distributed according to 
\begin{equation}
\label{SVBTS:Initial}
y_1, y_2, \sim \mathcal{N}(\boldsymbol{0}, \Sigma)
\end{equation}
where 
\begin{align}
\Sigma &= \left[ \begin{array}{cc} \gamma(0) & \gamma(1) \\ \gamma(1) & \gamma(0) \end{array} \right], \nonumber \\
\gamma(0) &= \sigma^2 \frac{1-\phi_2}{(1+\phi_2)((1-\phi_2)^2 - \phi_1^2)} \nonumber, \\
\gamma(1) &= \sigma^2 \frac{\phi_1}{(1+\phi_2)((1-\phi_2)^2 - \phi_1^2)} \nonumber.
\end{align}

The following prior distribution is chosen
\begin{equation}
p(\sigma^2, \phi_1, \phi_2) \propto \sigma^{-2} \mathbb{I}(\phi_2 > -1)\mathbb{I}(\phi_2 < 1 + \phi_1) \mathbb{I}(\phi_2 < 1 - \phi_1), 
\label{SVBTS:Prior}
\end{equation}
where $\mathbb{I}$ is the indicator function that equals one if the condition in the brackets is true and zero otherwise. This prior has uniform positive mass for the pair $(\phi_1, \phi_2)$ across the AR(2) stationary region, and zero mass outside of this region. Note that this prior is not conjugate due to the form of $\phi_1$ and $\phi_2$ in $\Sigma$, and thus MFVB cannot be applied.

Data was simulated with $T = 150, \sigma^2 = 1, \phi_1 = 0.7$ and $\phi_2 = 0.2$, then both RWMH-MCMC and SVB are applied. $q_{\lambda}(\theta | y_{1:T})$ is chosen to be the product of a bivariate normal distribution for $\phi_1$ and $\phi_2$, and an Inverse Gamma distribution for $\sigma^2$.

Figure \ref{fig:SVB} displays the marginal and bivariate posterior distributions that result from each of the Stochastic Variational Bayes (blue) and the MCMC (red) approaches. In the case of MCMC, a kernel estimate obtained from the relevant posterior sample is displayed. As is clear from the figure, SVB appears to have selected values for $\lambda$ that closely match the MCMC posterior distribution. SVB required approximately one hundredth of the computation time required for MCMC.

\begin{figure}[htbp]
\centering
\includegraphics[width=0.7\linewidth,height=\textheight,keepaspectratio]{figures/VBfit.png}
\caption{The fit of an SVB algorithm (blue) compared to MCMC (red) for the AR(2) model.}
\label{fig:SVB}
\end{figure}

\subsection{Stein Variational Gradient Descent}
\label{subsec:Stein}

\cite{Liu2016} introduce Stein Variational Gradient Descent, which creates a KL-divergence minimising particle approximation of the form
\begin{equation}
\label{Stein:Approx}
q_m(\theta | y_{1:T}) = \frac{1}{N}\sum_{i=1}^N \delta(\theta^{(i)}_l).
\end{equation} 
They approach this by proposing a sequence of approximating distributions, starting with particles $\theta^{(i)}_0$ drawn from some arbitrary $q_{0}(\theta)$, then transformed to $\theta^{(i)}_{m+1}$ via
\begin{equation}
\label{Stein:Transform}
\theta^{(i)}_{m+1} = \theta^{(i)}_ml + \epsilon_m \phi_m(\theta^{(i)}_m)
\end{equation}
where $\epsilon_m$ is some small constant and $\phi_m(\cdot)$ is some continuously differentiable function. The sequence $\phi_1, \phi_2, \ldots, \phi_m$ could be considered the parameters of the reparametersied VB function $\theta_m = f(\theta_0, \lambda)$.
\\

\cite{Liu2016} proves that, for a fixed choice $\phi_m$, 
\begin{equation}
\label{Stein:Derivative}
\frac{\partial KL[q_m(\theta | y_{1:T}) \hspace{1mm} || \hspace{1mm} p(\theta | y_{1:T})]}{\partial \epsilon} \bigg \vert_{\epsilon = 0} = - \mathbb{E}_{q_m} \left[ \mbox{trace}(A_p \phi_l(\theta)) \right]
\end{equation}
where $A_p$ is the Stein operator
\begin{equation}
\label{Stein:Operator}
A_p \phi(\theta) = \frac{\partial \log(p(y_{1:T}, \theta))}{\partial \theta} \phi(\theta)^{\prime} + \frac{\partial \phi(\theta)}{\partial \theta}.
\end{equation}

The aim of Stein Variational Gradient Descent is to choose the sequence of functions $\phi_m(\cdot)$ that minimises $KL[q_m(\theta | y_{1:T}) \hspace{1mm} || \hspace{1mm} p(\theta | y_{1:T})]$ with the least number of iterations, so at each iteration $m$ we want to choose the function $\phi_m(\cdot)$ that minimises the gradient of the KL divergence (remember this is a minimisation problem!).
\\

Restricting $\phi_m$ to the set of functions in the reproducing kernel Hilbert space $\mathcal{H}$ for some positive definite kernel $k(\cdot, \cdot)$ for which $||\phi||_{\mathcal{H}} \leq 1$, it can be shown that
\begin{equation}
\label{Stein:optPhi}
\hat{\phi}_m(\cdot) = \arg \underset{\phi}{\max} \hspace{1mm} \mathbb{E}_{q_m} \left[ \mbox{trace}(A_p \phi_m(\theta)) \right] = \mathbb{E}_{q_m} \left[ A_p k(\theta, \cdot) \right].
\end{equation}

Noting that the expectation in (\ref{Stein:optPhi}) is over a discrete distribution, the function evaluation $\hat{phi}_m(\theta^{(i)})$ is calculated as
\begin{equation}
\label{Stein:estPhi}
\hat{\phi}_m(\theta^{(i)}) = \frac{1}{N} \sum_{j=1}^N \left[ k(\theta_m^{(j)}, \theta_m^{(i)}) \frac{\partial \log(p(y_{1:T}, \theta))}{\partial \theta} \bigg \vert_{\theta = \theta^{(j)}_m} + \frac{\partial k(\theta, \theta^{(i)})}{\partial \theta} \bigg \vert_{\theta = \theta^{(j)}_m} \right]
\end{equation}
where $N$ is the number of particles. The particles are then transformed according to (\ref{Stein:Transform}) for as many iterations as desired.
\\

The sum in (\ref{Stein:estPhi}) may be carried out over only a subset of the particles to reduce computation time. In this case, the gradient is replaced with a stochastic estimate and the sequence $\epsilon_m$ must follow the Robbins-Monro conditions.
\\

\cite{Liu2017} goes on to show that, in the limit as $m \rightarrow \infty$ and $N \rightarrow \infty$, $q_m(\theta | y_{1:T})$ converges to $p(\theta | y_{1:T})$

\section{Updating Bayesian Inference}
\label{sec:UpdateBayes}

The discussion is extended to settings where data is regularly being observed in an online setting, resulting in a sequence of time points $T_1, T_2, \ldots$, from which a sequence of posterior distributions $p(\theta | y_{1:T_1}), p(\theta | y_{1:T_2}), \ldots$, may be desired.
\\

This is illustrated in Figure \ref{fig:updatetimeUpdate} for data taken from vehicle in the Next Generation Simulation (NGSIM) dataset provided by the US Federal Highway Administration (FHWA), travelling towards the right. In the left panel, the vehicle has been observed for a time period $T_{n}$ resulting in data $y_{1:T_{n}}$ which can be incorporated into the posterior distribution $p(\theta | y_{1:T_{n}})$. In the right panel, after observation for an additional period to time $T_{n+1}$, an additional $T_{n+1} - T_{n}$ data points are available, and posterior inference could be improved by incorporating the information contained in $y_{T_{n}+1:T_{n+1}}$ to form the posterior approximation $p(\theta | y_{1:T_{n+1}})$.

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth, height = 0.15\textheight]{figures/carUpdate}
\caption{Left: The path of a vehicle in the NGSIM dataset that has been observed for a time period equal to $T_{n}$, where the direction of travel is from the left to the right. Right: The same vehicle at a later time $T_{n+1}$, with the extra $T_{n+1} - T_{n}$ observations denoted by the dashed line. Posterior inference about this vehicle could be updated from time $T_{1}$ to time $T_{2}$ by the inclusion of this additional information.}
\label{fig:updatetimeUpdate}
\end{figure}

From Bayes Rule, the posterior distribution at time $T_{n+1}$, $p(\theta | y_{1:T_{n+1}})$ is given by
\begin{equation}
\label{update:truePost}
p(\theta | y_{1:T_{n+1}}) \propto p(y_{1:T_{n_1}} | \theta)p(\theta).
\end{equation}
where the likelihood is given by
\begin{equation}
\label{update:likelihood}
p(y_{1:T_n} | \theta) = p(y_1) \prod_{t=2}^{T_n} p(y_t | \theta, y_{1:t-1}).
\end{equation}
the product of $T_n$ terms. If the posterior at time $T_{n}$ is available, the likelihood can be reduced to a product of only $T_{n+1} - T_{n}$ terms by applying Bayes' Rule,
\begin{equation}
\label{update:updatePost}
p(\theta | y_{1:T_{n+1}}) \propto p(y_{T_{n}+1:T_{n+1}} | \theta, y_{1:T_{n}})p(\theta | y_{1:T_{n}}),
\end{equation}
However, many Bayesian computational methods require the evaluation of $p(\theta | y_{1:T_{n}})$ at arbitrary values of $\theta$, which is computationally infeasible for many posterior distributions of interest and so the more intensive (\ref{update:truePost}) is typically applied instead.

\subsection{Sequential Monte Carlo}
\label{subsec:SMC}

Sequential Monte Carlo (SMC, \cite{Doucet2000}, \cite{Arulampalam2002}) algorithms, often referred to as Particle Filters , are a technique to update a set of samples distributed according to $p(\theta | y_{1:T_n})$ to the distribution $p(\theta | y_{1:T_{n+1}})$. SMC approaches work via a repeated application of an importance sampler with respect to a common set of samples, or particles, drawn from the candidate distribution $r(\theta)$.
\\

After applying importance sampling, a discrete approximation to the posterior can be provided by
\begin{equation}
\label{SMC:Approx}
p^*(\theta | y_{1:T_n}) = \sum_{i=1}^S w^{(i)}_{T_n} \delta(\theta^{(i)})
\end{equation}
where $\delta$ is the Dirac delta function and $w^{(i)}_{T_n}$ are the $S$ importance sampler weights up to time $T_n$.
\\

While the critical choice of the initial $r(\theta)$ is left to the user, SMC concerns itself with the recursive application of importance sampling where the candidate distribution at some time $T_{n+1}$ is given by the previous set of particles and weights that compose (\ref{SMC:Approx}). In this case, the SMC update to 
\begin{equation}
\label{SMC:Update}
p^*(\theta | y_{1:T_{n+1}}) = \sum_{i=1}^S w^{(i)}_{T_{n+1}} \delta(\theta^{(i)})
\end{equation}
is performed by updating the unnormalised weights,
\begin{align}
\hat{w}^{(i)}_{T_{n+1}} &= \frac{p(y_{1:T_{n+1}}, \theta^{(i)})}{r(\theta^{(i)})} \nonumber \\
&= \frac{p(y_{T_{n}+1:T_{n+1}} | \theta^{(i)}) p(y_{1:T_n}, \theta^{(i)})}{r(\theta^{(i)})} \nonumber \\
&= p(y_{T_{n}+1:T_{n+1}}| \theta^{(i)}) \hat{w}^{(i)}_{T_n}, \label{SMC:UpdateWeights},
\end{align}
which are normalised by
\begin{equation}
\label{SMC:UpdateWeightsNorm}
w^{(i)}_{T_{n+1}} = \frac{\hat{w}^{(i)}_{T_{n+1}}}{\sum_{i=1}^S \hat{w}^{(i)}_{T_{n+1}}}.
\end{equation}

Repeated SMC updates often exhibit weight degeneration, where the values of each $w^{(i)}_{T_{n+1}}$ except one approaches zero, reducing the effective sample size of the approximation.
\iffalse

 which is estimated by
\begin{equation}
\label{SMC:EFF}
\hat{N}_{eff} = \frac{1}{\sum_{i=1}^N (w^{(i)}_{T_{n+1}})^2}.
\end{equation}

A common strategy to avoid this is through resampling, where if after an SMC update $\hat{N}_{eff}$ falls below some threshold, a new set of particles is sampled according to the distribution
\begin{equation}
\label{SMC:Resample}
\theta^{(i)}_{RS} \sim \sum_{i=1}^M w^{(i)}_{T_{n+1}} \delta(\theta^{(i)}).
\end{equation}
After resampling, the weights are each reset to $w^{(i)}_{T_{n+1}} = 1/N$. 
\\

Resampling resets the weights of each particle, however there is a high probability that there will be repeated values in the new set of particles and diversity is reduced.

\textit{After writing this I realise that re-sampling doesn't do anything in non-state-space-models as we don't have a transition equation to refresh the particle diversity. Instead of losing out because most of our particle weights are zero, we're losing out because we're repeating calculations for the same $\theta$ value! This is useful for UVB Importance Sampling, as we will get new proposal distributions to refresh out particle weights.}

\fi

\subsection{Bayesian Filtering? PMMH? Only if we use them later}

\subsection{Online Mean Field Variational Bayes}
\label{subsec:OnlineMFVB}

Mean Field Variational Bayes applied conjugate-exponential family model results in a set of mean field equations that can be written in the form of 
\begin{equation}
\lambda = f(\beta, y_{1:T}) \label{OnlineMFVB:OfflineEq}
\end{equation}
where $\beta$ denotes the prior hyperparameters. 
\\

Let $\lambda_1$ denote the MFVB optimal auxiliary parameters of $q_{\lambda_1}(\theta | y_{1:T_1})$, an approximating distribution chosen to minimise the KL divergence to the posterior after observation of $y_{1:T_1}$, obtained via the application of (\ref{OnlineMFVB:OfflineEq}),
\begin{equation}
\lambda_1 = f(\beta, y_{1:T_1}) \label{OnlineMFVB:OnlineEq1}
\end{equation}

The online MFVB approach first advocated by \cite{Ghahramani2000b} is to treat $\lambda_n$ as an updated form of the prior hyperparameters and re-apply the meanfield equations as
\begin{align}
\lambda_{n+1} = f(\lambda_n, y_{T_n+1:T_{n+1}}) \label{OnlineMFVB:OnlineEq2}
\end{align}

This results in a sequence of MFVB distributions $q_{\lambda_1}(\theta | y_{1:T_1}), q_{\lambda_2}(\theta | y_{1:T_2}), \ldots$ to approximate each of the sequence of posteriors $p(\theta | y_{1:T_1}), p(\theta | y_{1:T_2}), \ldots$.
\\

See \cite{Smidl2004}, \cite{Hoffman2010}, \cite{Wang2011}, \cite{Broderick2013}, and \cite{Kabisa2016} for examples of applications of Online MFVB.

\subsection{Online Stochastic Variational Bayes}
\label{subsec:OnlineSVB}

The stochastic gradient ascent approach utilised by SVB, where $\partial\mathcal{L}(q, \lambda) / \partial \lambda$ is estimated by $\widehat{\partial\mathcal{L}(q, \lambda) / \partial \lambda}$, will converge to a local maximum for any unbiased gradient estimator (\cite{Robbins1951}, \cite{Bottou1999}, \cite{Bottou2008}).
\\

Online SVB does not provide an approximate distribution at each of a sequence of time points $T_1, T_2, \ldots$, but allows for SVB inference to be substantially sped up by calculating the likelihood of only a subset of the entire dataset each iteration (\cite{Hoffman2013}, \cite{Titsias2014}).
\\

If the data $y_{1:T}$ is conditionally independent given $\theta$, the log-likelihood decomposes into the sum
\begin{equation}
\label{OnlineSVB:likelihood}
\log(p(y_{1:T} | \theta)) = \sum_{t=1}^T \log(p(y_t | \theta)).
\end{equation}

By subsetting $y_{1:T}$ into length $N$ `minibatches', where $N < T$, Online SVB replaces the log-likelihood in the gradient estimates with an unbiased estimator
\begin{equation}
\label{OnlineSVB:likelihoodEstim}
\widehat{\log(p(y_{1:T} | \theta))} = \frac{T}{N} \sum_{t=1}^N \log(p(y_t | \theta))
\end{equation}
where the sum in (\ref{OnlineSVB:likelihoodEstim}) is only over components of $y_{1:T}$ that are in the current minibatch. As this is unbiased, the gradients $\widehat{\partial\mathcal{L}(q, \lambda) / \partial \lambda}$ retain their unbiased properties.
\\

Each iteration of the gradient ascent algorithm should use a different subset of the data, so that sequential iterations cycle through the entire dataset. 
\\

\chapter{Approximate Bayesian Updating}
\label{chap:UVB}

This chapter introduces Updating Variational Bayes (UVB), a mechanism where an SVB approximation can be updated from $q_{\lambda_n}(\theta | y_{1:T_n})$ to  $q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})$ using only the additional data $y_{T_{n}+1:T_{n+1}}$, and where the computation required by the update does not increase with $n$. Note that the $n$ and $n+1$ subscripts on $\lambda$ are introduced to differentiate the auxiliary parameter vector conditioned on data up to times $T_{n}$ and $T_{n+1}$.
\\

An importance sampling extension to UVB is introduced, which allows for further computational cost reductions in scenarios where the marginal information in $y_{T_n+1:T_{n+1}}$ is relatively small.
\\

The application of UVB to Bayesian Filtering, the recursive problem of inferring the distribution of some unobserved latent states $x_{1:T_n}$ given observations of data $y_{1:T_n}$ as $n$ increases, is discussed. 
\\

An additional importance sampling framework which can be used as an error correction mechanism is provided to reduce the impact of using a variational approximation to the posterior. 
\\

Finally several sets of simulated data are analysed to compare the approximation error and runtime for SVB, UVB, and the importance sampled corrections.

\section{Updating Variational Bayes}
\label{sec:UVB}

Bayesian updating refers to the repeated use of the posterior update equation first introduced in Section \ref{sec:UpdateBayes},
\begin{equation}
\label{UVB:updatePost}
p(\theta | y_{1:T_{n+1}}) \propto p(y_{T_{n}+1:T_{n+1}} | \theta, y_{1:T_{n}})p(\theta | y_{1:T_{n}}).
\end{equation}
Note that, in the general case, the right hand side of this proportionality is computationally intractable due to the presence $p(\theta | y_{1:T_n})$.
\\

The application of SVB at time $T_1$ results in the analytically tractable posterior distribution
\begin{equation}
\label{UVB:Time1}
q_{\lambda_1}(\theta | y_{1:T_1}) = \arg \underset{q}{\min} \hspace{1mm} KL[q_{\lambda_1} (\theta | y_{1:T_1}) \hspace{.1cm}||\hspace{.1cm}p(\theta | y_{1:T_1})].
\end{equation}
Starting with this approximation, UVB involved the recursive use of the distribution \\ $q_{\lambda_n}(\theta | y_{1:T_n})$ at time $T_n$ as a substitute for the true posterior in the Bayesian update (\ref{UVB:updatePost}) to form an alternative pseudo-posterior update,
\begin{equation}
\label{UVB:pHatPosterior}
\hat{p}(\theta |  y_{1:T_{n+1}}) \propto p(y_{T_{n}+1:T_{n+1}} | \theta)q_{\lambda_{n}}(\theta | y_{1:T_{n}}).
\end{equation}
As the right hand side of this update is analytically tractable, standard ELBO gradient optimisation methods discussed in Section \ref{subsec:SVB} may be applied to find  
\begin{equation}
\label{UVB:TimeNp1}
q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}}) = \arg \underset{q}{\min} \hspace{1mm} KL[q_{\lambda_{n+1}} (\theta | y_{1:T_{n+1}}) \hspace{.1cm}||\hspace{.1cm}\hat{p}(\theta | y_{1:T_{n+1}})].
\end{equation}
Minimising the KL divergence in this way is not equivalent to standard VB which minimises $KL[q_{\lambda_{n+1}} (\theta | y_{1:T_{n+1}}) \hspace{.1cm}||\hspace{.1cm}p(\theta | y_{1:T_{n+1}})]$ unless $q_{\lambda_{n}}(\theta |  y_{1:T_{n}}) = p(\theta |  y_{1:T_{n}})$ almost everywhere. This is only possible if the posterior is a member of the family $q$. When this is not the case UVB introduces an additional approximation error to Variational Bayes inference which depends largely on the ability of the family $q$ chosen to adequately approximate the true posterior distribution.
\\

If the updates are equally spaced such that $T_{n+1} - T_{n} = H$ for all $n$, then the computational complexity of UVB as $n$ increases is constant at $O(H)$. In contrast, the computational complexity of SVB is $O(T_{n})$ as it evalutates the posterior via (\ref{update:truePost}) which requires evaluation of the complete data likelihood, $p(y_{1:T_{n}} | \theta)$, at each $T_{n}$.
\\

Choosing the form of the approximating distribution to match the prior distribution allows UVB to be easily implemented, as the prior hyper-parameters are simply replaced with the optimised $\lambda_n$ values after each update.
\\

A summary of the UVB algorithm is given by Algorithm \ref{alg:UVB}.
\\

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Prior, Likelihood.}
 \KwResult{Approximating distribution at $T_N$.}
 Observe $y_{1:T_1}$\;
 Use (\ref{SVB:scoreDeriv2}), (\ref{SVB:rpDeriv}), or (\ref{SVB:grp}) to choose the $\lambda_1$ that minimises the KL divergence from $q_{\lambda_1}(\theta | y_{1:T_1})$ to $p(\theta | y_{1:T_1})$ \;
 \For{$n \mbox{ in } 1, \ldots, N-1$}{
   Observe next data $y_{T_{n}+1:T_{n+1}}$\;
   Use $q_{\lambda_{n}}(\theta | y_{1:T_{n}})$ and (\ref{SVB:scoreDeriv2}), (\ref{SVB:rpDeriv}), or (\ref{SVB:grp}) to choose the $\lambda_{n+1}$ that minimises the KL divergence from $q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})$ to $\hat{p}(\theta | y_{1:T_{n+1}})$ 
  }
 \caption{Updating Variational Bayes}
  \label{alg:UVB}
\end{algorithm}

\section{Variational Bayesian Filtering} \label{sec:UVBVBF}

\textit{We don't use latent states in any of the models so far so this section isn't used.}

Many models of interest can be expressed as a dynamic state-space, with
\begin{align}
y_t &\sim p(y_t | x_{t}, \theta) \label{VBF:measure} \\
x_t &\sim p(x_t | x_{t-1}, \theta) \label{VBF:transition} ,
\end{align}
where inference of $p(x_t, \theta | y_{1:t})$ is desired at each $t$. This problem is known as \textit{Bayesian filtering}, where the posterior distribution is recursively updated from $p(x_{t-1}, \theta | y_{1:t-1})$ to $p(x_t, \theta | y_{1:t})$ by
\begin{align}
p(x_t, \theta | y_{1:t-1}) &= \int p(x_t | x_{t-1}, \theta) p(x_{t-1}, \theta | y_{1:t-1})x_{t-1} \label{VBF:marginalise} \\
p(x_t, \theta | y_{1:t}) &\propto p(y_t | x_t, \theta) p(x_t, \theta | y_{1:t-1}) \label{VBF:update}
\end{align}
for a given prior distribution $p(x_0, \theta)$. 
\\

For a limited class of models the marginalisation over $x_{t-1}$ in (\ref{VBF:marginalise}) and posterior in (\ref{VBF:update}) are both analytically tractable for all $t$, such as when both (\ref{VBF:measure}) and (\ref{VBF:transition}) are linear and Gaussian. Typically this is not the case and an approximation is required, such as the extended Kalman filter \citep{Anderson1979}, the unscented Kalman filter \citep{Wan2000}, or particle filters (\cite{Doucet2000}, \cite{Arulampalam2002}). Variational Particle Filtering is introduced by \citet{Smidl2008}, in cases where a subset $x_{1, t}$ of $x_t = \{x_{1, t}, x_{2, t}\}$ can be analytically marginalised, and does not discuss simultaneous inference of the static variables $\theta$ which is allowed with UVB below.
\\

Replacing $p(x_{t-1}, \theta | y_{1:t-1})$ with the Variational Bayes approximation \newline $q(x_{t-1} | y_{1:t-1}, \theta)q(\theta|y_{1:t-1})$ the following distributions are obtained 
\begin{align}
\hat{p}(x_t, \theta | y_{1:t-1}) &= q(\theta | y_{1:t-1}) \int p(x_t | x_{t-1}, \theta) q(x_{t-1} | y_{1:t-1}, \theta)x_{t-1} \label{VBF:marginaliseHat} \\
\hat{p}(x_t, \theta | y_{1:t}) &\propto p(y_t | x_t, \theta)\hat{p}(x_t, \theta | y_{1:t-1}). \label{VBF:updateHat}
\end{align}

If (\ref{VBF:transition}) is linear and Gaussian, and  $q(x_{t-1} | y_{1:t-1}, \theta)$ is Gaussian, (\ref{VBF:marginaliseHat}) can be marginalised analytically, allowing the Variational Bayes approximation at time $t$ as
\begin{equation}
\label{VBF:UVBfilter}
\hat{p}(x_t, \theta | y_{1:t}) \approx q(x_{t} | y_{1:t}, \theta)q(\theta|y_{1:t}).
\end{equation}
\\

\section{Correcting Approximation Error} \label{sec:UVBCorrection}

The increasing approximation error introduced by UVB could be offset by an importance sampling correction at each $T_n$. By using the UVB distribution $q_{\lambda_n}(\theta | y_{1:T_n})$ as the proposal $r(\theta)$, the following discrete approximation may be constructed
\begin{equation}
\label{UVBIS:Approx}
p^*(\theta | y_{1:T_n}) = \sum_{i=1}^M w^{(i)}_{T_n} \delta(\theta^{(i)})
\end{equation}
with weights given by
\begin{align}
\hat{w}^{(i)}_{T_n} &= \frac{p(y_{1:T_n} | \theta^{(i)}) p(\theta^{(i)}))}{q(\theta^{(i)} | y_{1:T_n})} \label{UVBIS:Weights} \\
w^{(i)}_{T_n} &= \frac{\hat{w}^{(i)}_{T_n}}{\sum_{i=1}^M \hat{w}^{(i)}_{T_n}} \label{UVBIS:WeightsNorm}
\end{align}
\\

\citet{Yao2018} note that, due to the variational approximation underestimating the tails of the exact posterior, importance sampling in this way often performs poorly with the weights dominated by a few samples. A Pareto-Smoothed importance sampler is suggested as an alternative.

If additional observations $y_{T_n+1}, y_{T_n + 2}, \ldots, y_{T_{n+1}}$, are observed before the update to \\ $q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})$ is available, the SMC recursion can be used to update the weights via
\begin{align}
\hat{w}^{(i)}_{T_n+h} &= p(y_{T_n+h} | \theta^{(i)}) \hat{w}^{(i)}_{T_n+h-1}, \label{UVBIS:UpdateWeights} \\
w^{(i)}_{T_n+h} &= \frac{\hat{w}^{(i)}_{T_n+h}}{\sum_{i=1}^M \hat{w}^{(i)}_{T_n+h}}. \label{UVBIS:UpdateWeightsNorm}
\end{align}
for $h = 1, 2, \ldots, T_{n+1} - T_{n}$
\\

Once computation of $q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})$ has finished, a fresh set of particles can be sampled from this distribution and significant weight degeneration may be avoided.

\section{Importance Sampled UVB} 
\label{sec:UVBIS}

UVB has significant computational cost savings as it reduces the likelihood from a product of $T_{n+1}$ terms to a product of only $T_{n+1} - T_n$ terms. \\

The gradient ascent algorithm utilised by UVB to fit $q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})$ maximises the ELBO by repeated application of the following steps:
\begin{enumerate}
\item At iteration $m$ draw $\theta^{(1)}, \theta^{(2)}, \ldots, \theta^{(S)}$ from $q_{\lambda_{n+1}^{(m)}}(\theta | y_{1:T_{n+1}})$ \\
(alternatively draw $\epsilon^{(1)}, \epsilon^{(2)}, \ldots, \epsilon^{(S)}$ from $r(\epsilon)$ and transform with $\theta = f(\epsilon, \lambda^{(m)}_{n+1})$),
\item Calculate $p(y_{T_{n}+1:T_{n+1}} | \theta)$ for each $\theta = \theta^{(1)}, \theta^{(2)}, \ldots, \theta^{(S)}$ as part of the gradient estimate $\widehat{\frac{\partial\mathcal{L}(q, \lambda_{n+1})}{\partial \lambda_{n+1}}}$,
\item Update $\lambda_{n+1}^{(m+1)} = \lambda_{n+1}^{(m)} + \rho^{(m)}\widehat{\frac{\partial\mathcal{L}(q, \lambda_{n+1})}{\partial \lambda_{n+1}}} \bigg\rvert_{\lambda_{n+1} = \lambda^{(m)}_{n+1}}$,
\item Set $m = m + 1$.
\end{enumerate}
After each iteration, $\lambda_{n+1}$ changes, and hence $\theta$ must be resamples at step one, and the likelihood must be recalculated with these new samples. 
\\

\citet{Sakaya2017} note that calulation of the data likelihood is the most computationally expensive component of the gradient estimate. They propose that the $\theta$ samples used at some iteration $m$ could be re-used for iterations $m + 1, m + 2, \ldots m + h$ to calculate the gradient as an importance sampled estimate. 
\\

As $\theta$ is constant between iterations, the likelihood does not need to be recalculated and optmisation is sped up. For some $h$ the difference between $q_{\lambda_{n+1}^{(m)}}(\theta | y_{1:T_{n+1}})$ and $q_{\lambda_{n+1}^{(m+h)}}(\theta | y_{1:T_{n+1}})$ becomes large, reducing the effective sample size of the importance sampler weights and degrading the gradient estimate. At this stage \citet{Sakaya2017} resample $\theta$ and recalculate the likelihoods.
\\

This idea is readily extendable to the UVB context to derive an importance sampled UVB (IS-UVB) gradient estimator.
\\

Recall the score ELBO gradient for UVB,
\begin{equation}
\label{UVBIS:scoreGrad}
\frac{\partial\mathcal{L}(q, \lambda_{n+1})}{\partial \lambda_{n+1}} = \int_{\theta} q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}}) \frac{\partial \log(q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}}))}{\partial \lambda_{n+1}} \log \left(\frac{\hat{p}(y_{1:T_{n+1}}, \theta)}{q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})} \right) d\theta.
\end{equation}
By multipling and dividing this by $q_{\lambda_n}(\theta | y_{1:T_n})$ it can be written as an expectation with respect to the previous UVB distribution at time $T_n$:
\begin{equation}
\label{UVBIS:scoreGradIS}
\frac{\partial\mathcal{L}(q, \lambda_{n+1})}{\partial \lambda_{n+1}} = \int_{\theta} q_{\lambda_{n}}(\theta | y_{1:T_{n}})\frac{q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})}{q_{\lambda_{n}}(\theta | y_{1:T_{n}})} \frac{\partial \log(q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}}))}{\partial \lambda_{n+1}} \log \left(\frac{\hat{p}(y_{1:T_{n+1}}, \theta)}{q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})} \right) d\theta,
\end{equation}
and estimated via Monte Carlo,
\begin{equation}
\label{UVBIS:scoreEstIS}
\widehat{\frac{\partial\mathcal{L}(q, \lambda_{n+1})}{\partial \lambda_{n+1}}}_{IS} = \frac{1}{M} \sum_{i=1}^M w(\theta^{(i)})\frac{\partial \log(q_{\lambda_{n+1}}(\theta^{(i)} | y_{1:T_{n+1}}))}{\partial \lambda_{n+1}} \log \left(\frac{\hat{p}(y_{1:T_{n+1}}, \theta^{(i)})}{q_{\lambda_{n+1}}(\theta^{(i)} | y_{1:T_{n+1}})} \right),
\end{equation}
where $\theta \sim q_{\lambda_{n}}(\theta | y_{1:{T_n}})$ and 
\begin{equation}
w(\theta^{(i)}) = \frac{q_{\lambda_{n+1}}(\theta^{(i)} | y_{1:T_{n+1}})}{q_{\lambda_{n}}(\theta^{(i)} | y_{1:T_{n}})}.
\end{equation}

At each gradient ascent iteration only $\frac{\partial}{\partial \lambda_{n+1}} \log(q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}}))$ and $q_{\lambda_{n+1}}(\theta^{(i)} | y_{1:T_{n+1}})$ need to be recalculated as $\lambda_{n+1}$ updates from $\lambda_{n+1}^{(m)}$ to $\lambda_{n+1}^{(m+1)}$
\\

In the UVB case, if the marginal information about $\theta$ contained in data from $T_n$ to $T_{n+1}$ is relatively small, the converged $\lambda_{n+1}$ will not be significantly different from $\lambda_n$ and importance sampler weight decay is avoided.
\\

The variance of this estimator is increased relative to the score gradient estimator in (\ref{subsubsec:SVBScore}) due to the presence of the importance sampling weights. This increased variance can resulting in less accurate approximations, as the algorithm stopping criteria, a sufficiently small value of $|\mathcal{L}(q, \lambda^{(m+1)}) - \mathcal{L}(q, \lambda^{(m)})|$ after iteration $m+1$ is only evaluated as a noisy estimate also using the importance sampler. As the computation per iteration is extremely small we can set $N$ to be very large to compensate, allowing the user to set the trade-off between computation time approximation accuracy to suit their requirements.
\\

IS-UVB is outlined in Algorithm \ref{alg:UVBIS}.
\\

\vspace{2mm}
\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Prior, Likelihood.}
 \KwResult{Approximating distribution at $T_N$.}
 Observe $y_{1:T_1}$\;
 Use (\ref{SVB:scoreDeriv2}), (\ref{SVB:rpDeriv}), or (\ref{SVB:grp})  to choose the $\lambda_1$ that minimises the KL divergence from $q_{\lambda_1}(\theta | y_{1:T_1})$ to $p(\theta | y_{1:T_1})$\;
 \For{$n \mbox{ in } 1, \ldots, N-1$}{
    Observe $y_{T_{n}+1:T_{n+1}}$\;
   Sample $\theta \sim q_{\lambda_{n}}(\theta | y_{1:T_{n}})$\; 
   Calculate $p(y_{T_{n}+1:T_{n+1}} | \theta)$ and $q_{\lambda_{n}}(\theta | y_{1:T_{n}})$ for these samples\;
  Use (\ref{UVBIS:scoreEstIS}) to choose the $\lambda_{n+1}$ that minimises the KL divergence from $q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})$ to $\hat{p}(\theta | y_{1:T_{n+1}})$.\;
  }
 \caption{Importance Sampled UVB}
  \label{alg:UVBIS}
\end{algorithm}

\section{Simulation Results}
\label{sec:UVBSim}

The approximation error for repeated UVB and Importance Sampled UVB (IS-UVB) applications is compared relative to both SVB inference and exact inference implemented by RWMH-MCMC for two simple problems: Time Series Forecasting and Mixture Model Clustering.

\subsection{Time Series Forecasting}
\label{subsec:UVBTS}

500 datasets $y_{1:500}$ are simulated according to the following AR3 model,
\begin{equation}
\label{UVB:TSAR3}
y_t = \mu + \phi_1 (y_{t-1} - \mu) + \phi_2 (y_{t-2} - \mu) + \phi_3 (y_{t-3} - \mu) + e_t
\end{equation}
where $e_t \sim N(0, \sigma^2)$. 
The parameters are $\theta = \{\log(\sigma^2), \mu, \phi_1, \phi_2, \phi_3 \}$, with the prior
\begin{equation}
\label{UVB:TSprior}
\theta \sim N(\boldsymbol{0}, 10 \mathbb{I})
\end{equation}
where $\boldsymbol{0}$ denotes the zero vector and $\mathbb{I}$ denotes the identity matrix.
\\

In each simulation, $\mu$ and each $\phi$ are drawn from a $N(0, 1)$ distribution, accepting only draws where each $\phi$ lies in the AR3 stationary region. $\sigma^{-2}$ is simulated from a $G(5, 5)$ distribution.
\\

At each time $t = 100, 101, \ldots, 300$, the MCMC posterior distribution $p(\theta | y_{1:t})$ is simulated by 15000 draws from a RWMH-MCMC algorithm, discarding the first 10000 as a burn in period.
\\

SVB, UVB, and IS-UVB approximations, $q_{VB}(\theta | y_{1:T_n})$, $q_{UVB}(\theta | y_{1:T_n})$, and $q_{IS-UVB}(\theta | y_{1:T_n})$ respectively, are fit at each time $T_n = 75 + 25n$ for $n = 1, 2, \ldots 17$. In each case, the distributional family for $q$ is chosen to be a $K = 1, 2$, or $3$ component mixture of multivariate normal distributions.
\\

SVB and UVB use the score gradient estimator calculated from $S = 25$ $\theta$ samples per iteration, while IS-UVB is ran with 100 $\theta$ samples re-used between iterations.
\\

Given the posterior, or its approximation, the forecast distribution for $y_{T_n+h}$ is given by
\begin{equation}
\label{UVB:TSforecastDist}
p(y_{T_n + h} | y_{1:T_n}) = \int_{\theta} p(y_{T_n + h} | y_{1:T_n}, \theta)p(\theta | y_{1:T_n})d\theta,
\end{equation}
which may be evaluated by $S$ samples from the posterior,
\begin{equation}
\label{UVB:TSforecastDistApprox}
\hat{p}(y_{T_n + h} | y_{1:T_n}) \approx \frac{1}{S} \sum_{i=1}^S  p(y_{T_n + h} | y_{1:T_n}, \theta^{(i)}).
\end{equation}
The loss function associated with this value is given by $L(n, h)$ and is measured by the predictive logscore,
\begin{equation}
\label{UVB:TSlogscore}
L(n, h) = \log(p(y_{T_n + h} | y_{1:T_n})).
\end{equation}

One step ahead forecast distributions are made for each inferential method: MCMC, SVB, UVB and IS-UVB. Each approximate inference technique is evaluated by its predictive logscore. The results are displayed in Figure \ref{fig:UVBAR3Timing}, where the left panel displayed the mean logscore for each method after observation up to $T_n = 100, 125, \ldots, 500$ across the 500 simulations. While each approximation has a lower logscore compared to exact inference, there is little realised loss in logscore resulting from using IS-UVB over SVB, however UVB performs better than both methods.
\\

The right panel displayed the mean runtime for a singular SVB fit with data up to $T_n$, compared to the mean runtime for UVB and IS-UVB fit to $T_1$ and updated $n-1$ times to $T_n$, relative to the time taken for the initial fit. In this scenario, the amount of data is small and the computational cost of calculating the log-likelihood is low, so UVB is slower than SVB. However, UVB benefits by being able to provide an approximating distribution from $T_1$ onwards, while SVB must wait until all data is observed at time $T_n$ to begin optimisation. IS-UVB is significantly faster than either alternative. There is no significant difference between $K = 1, 2 $ or $3$.

\begin{figure}[htbp]
    \centering
    {{\includegraphics[width=7cm, height = 6cm]{figures/AR3ls} }}%
    \qquad
    {{\includegraphics[width=7cm, height = 6cm]{figures/AR3timing} }}%
    \caption{Left: Average predictive logscores for each inference technique. Right: Average SVB runtime for one approximation at time $T_n$ and average cumulative UVB and IS-UVB runtimes fit to $T_1$, then updated $n-1$ times to $T_n$. Runtimes are relative to the time required for the $T_1$ fit.
UVB performs better than SVB but is slower, while IS-UVB performs worse but is faster. Both UVB and IS-UVB can begin optimisation as data arrives rather than waiting for all data to be observed.}%
    \label{fig:UVBAR3Timing}%
\end{figure}

\subsection{Mixture Model Clustering}
\label{subsec:UVBMMC}

The two component mixture normal model for $i = 1, \ldots N$ and $t = 1, \ldots, T_n$ is considered, augmented each unit $i$ with an auxiliary variables $k_i$ such that
\begin{equation}
\label{UVB:MMCmixNormalDGP2}
y_{i, t} | k_i = j \sim  N(\mu_j, \sigma^2_{j}).
\end{equation}
with 
\begin{equation}
\label{UVB:MMCkPrior}
k_i | \pi \sim Bin(1, \pi).
\end{equation}

Collecting $\theta = \{\log(\sigma^2_1), \log(\sigma^2_2), \mu_1, \mu_2 \}$, the prior for $\theta$ is given by
\begin{align}
\theta \sim N(\boldsymbol{0}, 10 \mathbb{I}), \\
\pi_1 \sim Beta(\alpha, \beta). \label{UVB:MMCpiPriorMix}
\end{align}

Note that (\ref{UVB:MMCkPrior}) and (\ref{UVB:MMCpiPriorMix}) imply that 
\begin{equation}
\label{UVB:MMCkMarginalMix}
p(k_i = j) = \frac{\mathcal{B}(j + \alpha, \beta - j + 1)}{\mathcal{B}(\alpha, \beta)}
\end{equation}
where $\mathcal{B}(\cdot, \cdot)$ denotes the Beta function.
\\

Denoting $y_{i, 1:T_n} = \{y_{i, t} | t = 1, \ldots, T_n\}$ and $y_{1:N, 1:T_n} = \{y_{i, 1:T_n} | i = 1, \ldots N \}$, the posterior distribution can then be marginalised over the values of each $k_i$,
\begin{equation}
\label{UVB:MMCMarginal}
p(\theta | y_{1:N, 1:T_n}) \propto p(\theta) \prod_{i=1}^N \left( \sum_{j=1}^2 p(y_{i, 1:T_n} | \theta, k_i = j) p(k_i = j) \right)
\end{equation}
This is approximated by employing distributions $q_{\lambda}(\theta | y_{1:N, 1:T_n})$ as either a $K = 1, 2,$ or $3$ component mixture of multivariate normal distributions. 
\\

As in Section \ref{subsec:UVBTS}, SVB and UVB use the score gradient estimator calculated from $S = 25$ $\theta$ samples per iteration, while IS-UVB is ran with 100 $\theta$ samples re-used between iterations.
\\

This posterior approximation can be updated if at each time $T_{n+1}$ the marginalisation in (\ref{UVB:MMCMarginal}) can be applied as
\begin{equation}
\label{UVB:MMCUpdate}
\hat{p}(\theta | y_{1:T_{n+1}}) \propto q_{\lambda_{n}}(\theta | y_{i, 1:T_{n}}) \prod_{i=1}^N \left( \sum_{j=1}^2 p(y_{i, T_n+1:T_{n+1}} | \theta, k_i = j) p(k_i = j | y_{1:T_n}) \right)
\end{equation}
This requires the marginal posterior distributions $p(k_i | y_{i, 1:T_{n}})$, which can be  approximately evaluated by
\begin{align}
\hat{p}(k_i | y_{i, 1:T_{n}}) &= \int_{\theta} p(k_i | y_{i, 1:T_{n}}, \theta)q_{\lambda_{n}}(\theta | y_{i, 1:T_{n}}) d\theta \nonumber \\
&\propto \int_{\theta} p(y_{i, 1:T_n} | \theta, k_i) p(k_i) q_{\lambda_{n}}(\theta | y_{i, 1:T_{n}}) d\theta \nonumber \\
&\approx \frac{1}{S} \sum_{i=1}^S p(y_{i, 1:T_n} | \theta^{(i)} , k_i) p(k_i). \label{UVB:MMCpkHat}
\end{align}
where $\theta^{(i)} \sim q_{\lambda_{n}}(\theta | y_{i, 1:T_{n}})$ for $i = 1, 2, \ldots, S$. This step happens once per UVB update, so use of the entire dataset in (\ref{UVB:MMCpkHat}) does not incur a large computational cost.
\\

This distribution is also used to predict classes for each unit as
\begin{equation}
\hat{k}_i = \arg \underset{j}{\max}\mbox{ } \hat{p}(k_i = j | y_{i, 1:T_n}).
\end{equation}

The accuracy of each inferential method in this model is given by the proportion of correct class predictions at each $T_n$.
\\

500 datasets are simulated with $N = 100$ units of length $T = 100$, and the posterior distribution $p(\theta, k_1, \ldots, k_N | y_{1:N, 1:T_n})$ is inferred for each $T_n = 10n, n = 1, 2, \ldots, 10$ using each inferential method: MCMC, SVB, UVB, and IS-UVB. Data is simulated by drawing each $\mu$ from an $N(0, 0.25)$ distribution and each $\sigma^2$ from a $U(1, 2)$ distribution. Units are randomly allocated to each group with $50\%$ probability. 
\\

The results are displayed in Figure \ref{fig:UVBMMCResults}, where breaking the data into smaller pieces benefits the accuracy of the UVB and IS-UVB posteriors, and hence cluster allocation score in the left panel. This problem features a large amount of data, and the computational cost of calculating the log-likelihood over the entire sample required by SVB slows optimisation relative to UVB and IS-UVB which process smaller amounts of data, shown in the right panel. The value of $K$ does not have a significant impact.

\begin{figure}[htbp]
    \centering
    {{\includegraphics[width=7cm, height = 6cm]{figures/mixNormScore} }}%
    \qquad
    {{\includegraphics[width=7cm, height = 6cm]{figures/mixNormTiming} }}%
    \caption{Left: Average proportion of correct classifications for each inference method. Right: Average SVB runtime for one approximation at time $T_n$ and average cumulative UVB and IS-UVB runtimes fit to $T_1$, then updated $n-1$ times to $T_n$. Runtimes are relative to the time required for the $T_1$ fit.    
UVB and IS-UVB perform better than SVB and are significantly faster, as computation of the data likelihood is a large part of the gradient calculation in this scenario.}%
    \label{fig:UVBMMCResults}%
\end{figure}

\section{Eight Schools Example}

In this section we consider the `Eight Schools' problem from \citet{Gelman2014}, where data from students' results on the SAT-V, a standardised highschool test in the USA, is collected to determine the effectiveness of SAT coaching programs ran by each of eight schools. 
\\

For each school $j = 1, 2, \ldots, 8$ it is assumed that an individual student's coaching effect, $y_{i, j}, i = 1, 2, \dots n_j$ is distributed with unknown mean $\theta_j$ and known variance $\sigma^2$, which is estimated by the sample variance.
\\

Due to the Central Limit Theorem \citep{Casella2002} the per school sample means,
\begin{equation}
\label{schools:mean}
\bar{y}_j = \frac{1}{n_j} \sum_{i=1}^{n_j} y_{i, j}
\end{equation}
are asymptotically distributed according to
\begin{equation}
\label{schools:CLT}
\bar{y}_{j} \sim \mathcal{N}(\theta_j, \sigma^2_j)
\end{equation}
where
\begin{equation}
\label{schools:var}
\sigma_j^2 = \frac{\sigma^2}{n_j}.
\end{equation}
\citet{Gelman2014} apply a hierarchical model, where the data is distributed according to (\ref{schools:CLT}) and each unknown variable $\theta_j$ is distributed according to the generalised student's t distribution,
\begin{equation}
\label{schools:hier}
\theta_j \sim t(\mu, \tau, \nu)
\end{equation}
where $\mu$ is a location parameter, $\tau$ is a scale parameter, and $\nu$ is the degrees of freedom, fixed at $\nu = 4$. Each $\theta_j$ is assumed to be conditionally independent of each other $\theta_i$ and $y_{1:n_i, i}, i \neq j$, given $\mu$ and $\tau$. 
\\

\citet{Gelman2014} also employ the hyper-prior
\begin{equation}
\label{schools:hyperprior}
p(\mu, \tau) \propto 1.
\end{equation}

Collecting $\theta^* = \{\theta_1,  \ldots, \theta_8, \mu, \log(\tau)\}$ and $\textbf{y}_{j} = \{y_{i, j} \hspace{2mm}| \hspace{2mm} i = 1,\ldots n_j\}$, the posterior distribution $p(\theta^* | \textbf{y}_{1:8})$ is given by
\begin{equation}
\label{schools:posterior}
p(\theta^* | \textbf{y}_{1:8}) \propto  p(\mu, \tau) \prod_{j=1}^8 \mathcal{N}(\bar{y}_j | \theta_j, \sigma^2_j) t(\theta_j | \tau, \mu, \nu).
\end{equation}

MCMC posterior samples are generated according to the algorithm provided by Stan, \citep{RStanGettingStarted}, and is compared to each of SVB, UVB, and IS-UVB.
\\

Each variational algorithm follows \citet{Kucukelbir2017} and uses a multivariate normal distribution $q(\theta^* | \textbf{y}_{1:8})$. In the case of UVB and IS-UVB, each update increases the number of schools included by one, using the posterior decomposition
\begin{equation}
\label{schools:update}
p(\theta_1:m, \mu, \tau | \textbf{y}_{1:m}) \propto p(\textbf{y}_m | \theta_m)p(\theta_m | \mu, \tau)p(\theta_{1:m-1}, \mu, \tau | \textbf{y}_{1:m-1})
\end{equation}
for $m = 2, \ldots 8$.
\\

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.9\textwidth]{figures/schools}
\caption{Marginal posterior distributions for each variable using each of MCMC, SVB, UVB and IS-UVB. No variational approach is clearly better than the others, and each marginal distribution for $\tau$ is inaccurate, possibly due to the impliclit lognormal approximation begin unsuitable.}
\label{fig:schools}
\end{figure}

The resulting marginal posterior distributions are displayed in Figure \ref{fig:schools}. Each variational approach captures the MCMC samples for $\mu$ well, $\tau$ poorly, and have varying degrees of accuracy for each $\theta_j$. Neither updating method appears to have introduced any discernable error relative to SVB. 
\\

It is likely that the normal approximation for $\log(\tau)$, and implicit log-normal distribution for $\tau$, is not suitable for this distribution.

\section{Summary}
\label{sec:UVBSummary}

In this chapter Updating Variational Bayes (UVB) was introduced, a framework that allows online variational inference of data. Online varitional inference has attracted a large amount of attention in the context of Mean Field Variational Bayes (MFVB), see eg. \cite{Hoffman2010} or \cite{Broderick2013}. 
\\

UVB differs from these as it is derived from SVB rather than MFVB, and thus can be applied to a larger class of models and approximating distributional families.
\\

The approach used by UVB is to take the previous VB approximation at time $T_{n-1}$ as a prior distribution to the posterior at time $T_{n}$, which can be significantly faster than re-running SVB using all available data as it reduces the likelihood calculations to contain $T_{n} - T_{n-1}$ terms rather than $T_{n}$ terms in SVB. 
\\

Online SVB based inference, such as in \citet{Paisley2012}, work by taking sub-samples of the data series, known as `mini-batches', of size $N$ each iteration. This then reduces the likelihood calculation to contain only $N$ terms. Once scaled by a factor of $T_{n} / N$,  this gradient estimate is unbiased if the data is statistically independent. However, time series data commonly violates the independence assumption and mini-batch approaches to optimisation fail. UVB does not require independence and hence can be applied to time-series data.  
\\

UVB was extended into IS-UVB in Section \ref{sec:UVBIS}, which further reduces the computation time of each update by re-using the $\theta$ samples and likelihood calculations between iterations.
\\

Two simulation studies are presented in Section \ref{sec:UVBSim}, which demonstrates that UVB outperforms SVB in each problem, in terms of forecasting logscore or cluster allocation accuracy. In the time-series forecasting application, where the total amount of data is relatively small, fitting SVB to all observations up to time $T_1$ and the updating this to $T_2, T_3, \ldots, T_n$ via UVB has a runtime simialr to fitting SVB directly to time $T_n$. In the mixture model clustering application gradient calculations are dominated by the larger amount of data, and a time $T_1$ SVB fit followed by UVB updates is significantly faster than a $T_n$ SVB fit at all time-points.
\\

IS-UVB is shown to be faster than UVB in both problems, with performance in-between SVB and UVB. The IS-UVB gradient estimator has a larger variance than the UVB estimator, which may lead to the stopping criteria at iteration $m$, $|\mathcal{L}(q, \lambda^{(m)} - \mathcal{L}(q, \lambda^{(m-1)})| < \epsilon$ for some small $\epsilon$ , being achieved early due to the noise in the estimates of $\mathcal{L}(q, \lambda^{(m)})$ and $\mathcal{L}(q, \lambda^{(m-1)})$. This may then degrade the quality of the resulting approximate distribution. In practice $M$, the number of $\theta$ samples used in the gradient estimate, may be increased to improve the fit at the cost of computation time.
\\

Finally, SVB, UVB, and IS-UVB are applied to the Eight Schools problem of \citet{Gelman2014}, where we show that the marginal distributions of each parameter are similar under each variational algorithm.
\\

UVB is highly suitable to time-series data under the following constraints:
\begin{enumerate}
\item Up to date inference on $\theta$ is required at all times to make an action such as forecasting,
\item Data arrives rapidly so that the computation of the infernece on $\theta$ has time constraints.
\end{enumerate}
Chapters \ref{chap:elec} and \ref{chap:cars} contain the application of UVB to real-life scenarios that fit these constraints.

\chapter{Online Heterogeneous Forecasting - Electricity Load}
\label{chap:elec}

\section{Introduction}
\label{sec:elecIntro}
A popular area of application for forecasting methodology is in short term electricity load forecasting (STLF), where forecasts for up to a day ahead are required. This literature traditionally uses a regions aggregate load at a given point in time as the time series to be forecasted, but the recent proliferation in smart meter data has caused increased interest in per household level forecasts (see \cite{Mirowski2014} and \cite{Yildiz2017} for recent overviews).
\\

Smart meter data introduces multiple challenges to STLF: the time series dynamics are muted at the household level and replaced by a large amount of stochastic noise, and these households can display heterogeneous behaviours. Further, there are significant computational challenges associated with the number of smart meters to forecast often being in the thousands.
\\

\citet{DaSilva2013} finds that the difficulties in data heterogeneity can e alleviated by aggregating households into groups. \citet{Shahzedah2015} and \citet{Quilumba2015} use this approach to cluster households into groups of similar behaviour prior to modeling, and create a model for each cluster. In each approach the forecast MAPE reduces by as much as 20\% relative to applying a single model for each household.
\\

Despite the prevalence of noise in the individual data, the models used for this data are largely similar to the aggregate load forecasting literature, where a mix of machine learning methods, such as neural networks \citep{Singhai2011, Niska2015, Bianchi2017}, have been used alongside arima, exponential smoothing, and other time series models \citep{Taylor2003, Taylor2008, Ghofrani2011} and semi-parametric additive spline models \citep{Hyndman2010, Fan2012, Taieb2016}.
\\

\citet{Ramanathan1997} finds that intraday seasonality in electricity load can be addressed by the use of a separate model per intraday period.
\\

\hrule
\vspace{3mm}
\textit{Haven't re-written the following section yet}
\\

Machine Learning models and semi-parametrics are supported by their strength in modelling non-linear relationships between load, lags of load, and exogenous variables such as temperature, humidity, or other weather conditions and time of day/week/year effects. It is widely agreed that temperature effects improve forecasts, even with the added problem of forecasting the temperature. Details on temperature forecasts are rare, but BoM one day forecasts are used by Rob Hyndman regularly. In households with both air-conditioning and heating, the temperature-load relationship is V shaped with a minimum at 18.3 C.
\\

Models are typically used to provide point estimates evaluated by MAPE, but density forecasts are becoming more popular (see eg. \cite{Arora2016} and \cite{Taieb2016}), with quantile regression, scenario simulation + residual bootstraps, and conditional kernel density estimation employed, evaluated by MAPE, CRSP, or visually comparing ex-ante and ex-post densities. Bayesian models are rare but not completely absent.
\\

Half hour frequency data is common (possibly the most common short term forecasting frequency), and it is observed that the time dynamics in electricity data depends strongly on the time of day, with a different parameter set employed per half hour period.
\\

Our model would differ from the previous literature by:
\begin{enumerate}
\item Being fully Bayesian (or, at least, approximately).
\item Employing online inference.
\item Clustering as part of the model without aggregation.
\item Considering a mixture of models.
\end{enumerate}

\section{Modelling}
\label{sec:elecModel}

In this chapter half hourly electricity load data is collected from $N = 200$ Smart Meters in London for the period from 01/12/2012 until 31/01/2014. For each household $i = 1, 2, \ldots, N$ and time period $t = 1, 2, \ldots, T$, the raw data $y_{i, t}^*$ is measured in kiloWatt Hours per half-hour period. This is non-negative and strongly skewed, so it is log transformed via
\begin{equation}
\label{elec:logY}
y_{i, t} = \log(y_{i, t}^* + 0.01)
\end{equation}
\iffalse
and modelled with a mixture of Markov switching models:
\begin{equation}
\label{elec:electricityModelSwitch}
p(y_{i, t} | \theta, k_i, s_{i, t}, y_{i, 1:t-1}) = \sum_{j=1}^K I(k_{i} = j) \big(s_{i, t} p_{d, j}(y_{i, t} | \theta, y_{i, 1:(t-1)}) + (1 - s_{i, t}) p_{c, j} (y_{i, t} | \theta) \big)
\end{equation}
where $I(\cdot)$ is the indicator function, $s_{i, t} = 0, 1,$ is a Markov switching indicator and $k_i = 1, 2, \dots, K$ is a mixture component indicator. This model allows each household to switch between a time dynamic model with likelihood $p_{d, j} (y_{i, t} | \theta, y_{i, 1:(t-1)})$ and a low level constant mean model with likelihood $p_{c, j}(y_{i, j} | \theta)$ for periods where the household occupants are away for long periods.
\\

The mixture formulation allows the households to cluster into $K$ latent groups with similar dynamic behaviour, as the household electricity consumption may depend strongly on factors such as working and sleeping hours. Clustering smart meter data into groups of households with similar behaviour and then creating a model per group has been shown by (citations) to reduce forecast error by as much as 30\% compared to employing a single model. Typically this clustering is carried out prior to the modelling process however in this application both clustering and parameter estimation are incorporated into a single step with the mixture model (\ref{elec:electricityModelSwitch}).
\\

Each dynamic model for $j = 1, 2, \dots, K$ follows a double seasonal \\ ARIMAX $(3, 0, 0)(3, 0, 0)_{48}(1, 0, 0)_{336}$ where the two seasonal components correspond to $48$ observations per day and $336$ observations per week with per-halfhour-period parameters, given by
\begin{equation}
\label{elec:dynamicModel}
(1 - \phi_{1, j}^{(h)}L - \phi_{2, j}^{(h)}L^2 - \phi_{3, j}^{(h)} L^3)(1 - \phi_{48, j}^{(h)}L^{48} - \phi_{96, j}^{(h)}L^{96} - \phi_{144, j}^{(h)}L^{144})(1 - \phi_{336, j}^{(h)}L^{336}) (y_{i, t} -  \beta^{\prime, {(h)}}_{j} y_{i, t}) = \epsilon_{i, t}
\end{equation}
where $L$ is the lag operator corresponding to $L^n y_{i, t} = y_{i, t- n}$, $(h)$ denotes the halfhour period of the day at time $t$. $y_{i}$ is a $T \times 9$ matrix with a columns corresponding to an intercept, $|Temp^{(c)}_{t} - 18.3|$, where $Temp^{(c)}_{t}$ is the temperature in degrees Celcius, six indicator variables for day of the week, and an indicator variable for public holidays while $\beta_{j}^{(h)}$ is corresponding coefficient vector for that halfhour. Finally $\epsilon_{i, t}$ is modelled with a $N(0, \sigma^{2, (h)}_{j})$ distribution.
\\

Let $\theta^{(h)}_j = \{\log(\sigma^{2, (h)}_{j}), \beta_{j, 1}^{(h)}, \beta_{j, 2}^{(h)}, \dots, \beta_{j, 9}^{(h)}, \phi_{1, j}^{(h)}, \phi_{2, j}^{(h)}, \dots, \phi_{336, j}^{(h)}\}$, the following random walk prior is applied to impose smoothness on the evolution of $\theta^{(h)}_j$ across the 48 half-hour periods each day,
\begin{align}
\label{elec:dynamicPrior}
\theta^{(1)}_j &\sim N(\boldsymbol{0}, \mathbb{I}), \\
\theta^{(h)}_j &\sim N(\theta^{(h-1)}, 0.1^2 \mathbb{I}) \mbox{ for } h = 2, 3, \dots, 48.
\end{align}
where $\boldsymbol{0}$ is the zero vector and $\mathbb{I}$ is the identity matrix.
\\

Each of the $K$ constant models is given by
\begin{equation}
\label{elec:constantModel}
y_{i, t} = \mu_j + \epsilon_{i, t}
\end{equation}
where $\epsilon_{i, t} \sim N(0, \tau^2_{j})$, with the prior
\begin{equation}
\label{elec:constantPrior}
p(\mu_j, \log(\tau_j^2))\sim N(\boldsymbol{0}, \mathbb{I}).
\end{equation}
We assume that the constant model is applied to times when the house is unoccupied for long periods and thus energy consumption is invariant to factors such as temperature and day of the week.
\\

The per household indicator variables $k_i$ are modelled by
\begin{equation} 
\label{elec:kConditional}
k_i \sim Multinomial(\boldsymbol{\pi})
\end{equation}
where 
\begin{equation}
\label{elec:piPrior}
\boldsymbol{\pi} \sim Dir(\alpha_1, \alpha_2, \dots, \alpha_K)
\end{equation}
This combination implies that
\begin{equation}
\label{elec:kMarginal}
p(k_i = j) = \frac{\Gamma(\sum_{l=1}^K \alpha_l)}{\Gamma(1 + \sum_{l=1}^K \alpha_l)} \frac{\Gamma(1 + \alpha_j)}{\Gamma(\alpha_j)}.
\end{equation}

Finally, for each of the $K$ groups we define switching probabilities \\ $\rho_{1, 0, j} = p(s_{i, t} = 0 | s_{i, t-1} = 1, k_i = j)$ and $\rho_{0, 1, j} = p(s_{i, t} = 1 | s_{i, t-1} = 0, k_i = j)$ with \\ uniform (0, 1) priors. 
\\

The parameter vector of interest is given by 
\begin{equation}
\label{electricityTheta}
\theta = \{\theta^{(1)}_j, \theta^{(2)}_j, \dots, \theta^{(48)}_j, \mu_j, \log(\tau^2_j), \rho_{0, 1, j}, \rho_{1, 0, j} \hspace{2mm} | \hspace{1mm} j = 1, 2, \dots, K\}.
\end{equation}
and posterior inference on $p(\theta | y_{1:N, 1:T})$, marginal of each $k_i$ and $s_{i, t}$, can be obtained by marginalisation of the complete model posterior $p(\theta, k_{1:N}, s_{1:N, 1:T} | y_{1:N, 1:T})$ by the use of a Hamiltonian filter:
\begin{align}
\label{elec:Hfilter}
p(\theta | y_{1:N, 1:T_n}) &\propto p(\theta)  \prod_{t=336 + 48 + 4}^{T} \prod_{i=1}^N  \sum_{j=1}^K p(k_i = j) p(y_{i, t} | \theta, y_{i, 1:t-1}, k_i = j), \\
p(y_{i, t} | \theta, y_{i, 1:t-1}, k_i = j) &= \sum_{m=0}^1 \sum_{n=0}^1  \rho_{m, n, j} \xi_{i, t-1, m, j}  p(y_{i, t} | \theta, y_{i, 1:t-1}, k_i = j, s_{i, t} = n),
\end{align}
where
\begin{equation}
\xi_{i, t-1, 1, j} = p(s_{i, t-1} = 1 | \theta, k_i = j, y_{i, 1:t-1})
\end{equation}
and
\begin{equation}
\xi_{i, t-1, 0, j} = 1 - \xi_{i, t-1, 1, j}.
\end{equation}
At each time point the filtered Markov state probability $\xi_{i, t, 1, j}$ can be updated via
\begin{equation}
\xi_{i, t, 1, j} = \frac{\sum_{m=0}^1 \rho_{m, 1, j} \xi_{i, t-1, m, j} p(y_{i, t} | \theta, y_{i, 1:t-1}, k_i = j, s_{i, t} = 1)}{\sum_{m=0}^1 \sum_{n=0}^1 \rho_{m, n, j} \xi_{i, t-1, m, j}p(y_{i, t} | \theta, y_{i, 1:t-1}, k_i = j, s_{i, t} = n)}
\end{equation}
from some starting value $\xi_{i, 0, 1, j}$.
\\
\fi

and modelled with the $K$ component mixture model
\begin{equation}
\label{elec:electricityModel}
p(y_{i, t} | \theta, k_i, y_{i, 1:t-1}) = \sum_{j=1}^K I(k_{i} = j) p_{j}(y_{i, t} | \theta, y_{i, 1:(t-1)})
\end{equation}
where $I(\cdot)$ is the indicator function and $k_i = 1, 2, \ldots, K$ is a mixture component indicator.
\\

The mixture formulation allows the households to cluster into $K$ latent groups with similar dynamic behaviour, as the household electricity consumption may depend strongly on factors such as working and sleeping hours. Clustering smart meter data into groups of households with similar behaviour and then creating a model per group has been shown by (citations) to reduce forecast error by as much as 30\% compared to employing a single model. Typically this clustering is carried out prior to the modelling process however in this application both clustering and parameter estimation are incorporated into a single step with the mixture model (\ref{elec:electricityModel}).
\\

Each model $j = 1, 2, \ldots, K$ follows a double seasonal \\ ARIMAX $(3, 0, 0)(3, 0, 0)_{48}(1, 0, 0)_{336}$ where the two seasonal components correspond to $48$ observations per day and $336$ observations per week with per-halfhour-period parameters,, given by
\begin{equation}
\label{elec:dynamic}
(1 - \phi_{1, j}^{(h)}L - \phi_{2, j}^{(h)}L^2 - \phi_{3, j}^{(h)} L^3)(1 - \phi_{48, j}^{(h)}L^{48} - \phi_{96, j}^{(h)}L^{96} - \phi_{144, j}^{(h)}L^{144})(1 - \phi_{336, j}^{(h)}L^{336}) (y_{i, t} -  \beta^{\prime, {(h)}}_{j} x_{t}) = \epsilon_{i, t}
\end{equation}
where $L$ is the lag operator corresponding to $L^n y_{i, t} = y_{i, t- n}$, $(h)$ denotes the half-hour period of the day at time $t$. $x_{t}$ is a length $9$ vector with a elements corresponding to an intercept, $|Temp^{(c)}_{t} - 18.3|$, where $Temp^{(c)}_{t}$ is the temperature in degrees Celsius, six indicator variables for day of the week, and an indicator variable for public holidays while $\beta_{j}^{(h)}$ is corresponding coefficient vector for that half-hour. Finally $\epsilon_{i, t}$ is modelled with a $N(0, \sigma^{2, (h)}_{j})$ distribution.
\\

Let $\theta^{(h)}_j = \{\log(\sigma^{2, (h)}_{j}, \beta_{j, 1}^{(h)}, \beta_{j, 2}^{(h)}, \ldots, \beta_{j, 9}^{(h)}, \phi_{1, j}^{(h)}, \phi_{2, j}^{(h)}, \ldots, \phi_{336, j}^{(h)}\}$, the following random walk prior is applied to impose smoothness on the evolution of $\theta^{(h)}_j$ across the 48 half-hour periods each day,
\begin{align}
\label{elec:dynamicPrior}
\theta^{(1)}_j &\sim N(\boldsymbol{0}, \mathbb{I}), \\
\theta^{(h)}_j &\sim N(\theta^{(h-1)}, 0.1^2 \mathbb{I}) \mbox{ for } h = 2, 3, \ldots, 48.
\end{align}
where $\boldsymbol{0}$ is the zero vector and $\mathbb{I}$ is the identity matrix.
The $K \times 48$ $\theta^{(h)}_j$ vectors are collectively referred to as $\theta$.
\\

The per household indicator variables $k_i$ are modelled by
\begin{equation} 
\label{elec:kConditional}
k_i \sim Multinomial(\boldsymbol{\pi})
\end{equation}
where 
\begin{equation}
\label{elec:piPrior}
\boldsymbol{\pi} \sim Dir(\alpha_1, \alpha_2, \ldots, \alpha_K)
\end{equation}
This combination implies that
\begin{equation}
\label{elec:kMarginal}warning
p(k_i = j) = \frac{\Gamma(\sum_{l=1}^K \alpha_l)}{\Gamma(1 + \sum_{l=1}^K \alpha_l)} \frac{\Gamma(1 + \alpha_j)}{\Gamma(\alpha_j)}.
\end{equation}

Each $k_i$ can be marginalised out of $p(\theta, k_{1:N} | y_{1:N, 1:T_n})$ to form a lower dimensional marginal posterior $p(\theta | y_{1:N, 1:T_n})$ via
\begin{equation}
p(\theta | y_{1:N, 1:T_n}) \propto p(\theta)  \prod_{i=1}^N\prod_{t=\tau}^{T_n}\sum_{j=1}^K p(k_i = j)p_{j}(y_{i, t} | \theta, y_{i, 1:t-1})
\end{equation}

UVB is applied to approximate the posterior distribution $p(\theta | y_{1:N, 1:T_n})$ with a diagonal covariance multivariate Gaussian distribution $q_{\lambda_n}(\theta | y_{1:200, 1:T_n})$, where the time period from the start of the sample until $T_1$ composes of the 2928 half hourly observations for the two month period from 01/12/2012 to 31/01/2013. There are a further 365 updates composing of one day's data, so that $T_366$ is the end of the sample at 31/01/2014. At the end of every update, probabilistic forecasts of the next days electricity consumption at half hourly intervals for each household are made via
\begin{align}
p(y_{i, T_n+h} | y_{1:N, 1:T_n}) = \int_{\theta} \sum_{j=1}^K \sum_{m=0}^1 &p(y_{i, T_n+h} | y_{1:N, 1:T_n}, \theta, k_i = j, s_{i, T_n} = m) \nonumber \\
&\times p(s_{i, T_n} = m | \theta, y_{1:N, 1:T_n}, k_i = j) \nonumber \\
&\times p(k_i = j | \theta, y_{1:N, 1:T_n}) \nonumber \\
&\times q_{\lambda_n}(\theta | y_{1:N, 1:T_n}) d\theta 
\end{align}
for $h = 1, 2, \ldots, 48$.
\\

The model is compared to other popular models for this type of data (TBD) and the same model without clustering $(K = 1)$.

\section{Results}
\label{sec:elecResults}

\section{Discussion}
\label{Sec:elecDisc}

\chapter{Online Heterogeneous Forecasting - Vehicle Trajectory}
\label{chap:cars}

\section{Introduction}
\label{sec:carsIntro}

Self-driving vehicles are rapidly becoming more advanced, with manufacturers such as Tesla, Ford, and Audi expecting autonomous vehicles to be available to consumers as early as 2020. These vehicles rely on a navigation system that detects the position of surrounding traffic, forecasts their trajectory, and selects a path to avoid collisions. With as many as $94\%$ of accidents in the United States resulting from human error, according to the US National Highway Traffic Safety Administration in 2015, \citep{NHTSA2015}, the introduction of fully self-driving vehicles is expected to substantially improve safety on the roads. However self-driving vehicles will operate in situations where the majority of surrounding vehicles are still controlled by human drivers, and so forecasts should be able to account for a wide range of possible human behaviour.
\\

While characteristics of surrounding vehicles, such as their position, velocity, and steering angle, can be extracted from sensors through the use of deep neural networks (See e.g. \citet{Woo2016a} and \citet{Tian2017} for details), the problem of forecasting the future trajectory of these vehicles has seen less attention in the literature. These forecasts often assume that the surrounding vehicles will maintain their current angle and velocity \citep{Gindele2010, Houenou2013, Bautista2017, Waymo2017} but interest has developed in more advanced models such as Neural Networks, Hidden Markov Models, and Support Vector Machines \citep{Ding2013, Woo2016b, Geng2017, Woo2017, Zheng2017}.
\\

This paper takes a statistical approach to this trajectory forecasting problem, fitting time series models to two extracted variables, acceleration and steering angle, of vehicles in the Next Generation Simulation (NGSIM) US Highway 101 Dataset. It uses Bayesian methods to forecast the distribution of the future velocity and steering angle, which can be transformed into a forecast for the distribution of trajectories, allowing low probability trajectories that could cause a collision to be detected. The time series models used allow for heterogeneity between different vehicles, which may result from sources such as differences in individual driving styles, or the conditions of traffic. Including heterogeneity in the model will allow for drivers with extreme behaviours to be accounted for, for example 'lead-footed' drivers can be assigned a larger variance on their acceleration, while forecasts for more consistent drivers will benefit from a low variance.
\\

Incorporating heterogeneity into forecasts requires the behaviour of each driver of interest to be inferred as vehicles are encountered while driving. This inference is facilitated by first obtaining the posterior distribution for a large number of vehicles before the self-driving vehicle is on the road, from which the prior distribution for additional vehicles can be constructed. Once the self-driving vehicle is on the road, online inference is made possible by Updating Variational Bayes (UVB), which approximates Bayesian updating so that inference may be periodically updated as additional data relating to surrounding vehicles is made available. 
\\

It is found that the use of these time series models produces accurate point forecasts in terms of mean Euclidean error when compared to the constant models commonly used, and including driver heterogeneity in these time series models provides higher forecast log-scores relative to a homogeneous approach. These results hold when UVB is used for the heterogeneous models while the homogeneous model benefits from exact MCMC inference, giving credence to the validity of UVB inference to provide fast inference in the online setting.
\\

\section{Data Pre-processing}
\label{sec:carsdataProcessing}
Data is provided by the Next Generation Simulation (NGSIM) project conducted by the US Federal Highway Administration (FHWA), which recorded vehicles travelling along a 2235 feet section of the US 101 freeway in Los Angeles, California during the morning peak from 7:50 am to 8:35 am on June 15th, 2005. Data for 6101 vehicles was collected every 100 milliseconds by seven static cameras, and processed by Cambridge Systematics Inc. to produce coordinates for each vehicle and point in time relative to the start of the road section. 
Figure \ref{fig:carsrawData} shows the paths of ten vehicles in the dataset, which travelled along the five major lanes or entered from the entry/exit lane to the right; with one vehicle changing from Lane 2 to Lane 1. Vehicles that entered or exited midway through the freeway section are excluded in this paper. There is a curvature to the road occurring between 500 and 1000 feet, and again between 1800 and 2000 feet.
\\

\begin{figure}[htbp]
\centering
\includegraphics[width=0.65\textwidth]{figures/carPath}
\caption{The path of ten vehicles in the dataset, with each black line representing a unique vehicle. This section of US 101 is split into five main lanes, with an additional entry/exit lane to the far right, through which two of the vehicles have entered. There is curvature to the road (which is slightly distorted by the aspect ratio) with bends occurring between 500 and 1000 feet, and again between 1800 and 2000 feet.}
\label{fig:carsrawData}
\end{figure}

Modern vehicles are capable of tracking the path of lane markings \citep{Thuy2010}, such as painted centre lines or lane dividers, and thus they can automatically identify the locations of surrounding vehicles in coordinates relative to their own position on the road, compensating for any curvature. To remove the variation in car position due to curvature in the road, and so that the data is in a coordinate system similar to that of a self-driving vehicle tracking surrounding vehicles, the coordinate system provided by NGSIM is transformed into relative coordinates. This is facilitated by measuring the location of vehicles relative to an estimate of the centre line of their associated lane. \citet{Woo2016a} fit a polynomial curve to detected lane markings to build a local model of the road lane edges, and this idea is extended with the use of smoothing splines to estimate each of the lane centres for the available section of the freeway. Details of this process are in the appendix, and results in relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$  where $y^*_{i, t}$ denotes the distance travelled along the centre of the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line for each of remaining vehicles in the dataset.
\\

Once relative coordinates are extracted, the changes in relative position of any vehicle $i$ from observation $t-1$ to $t$ follows from the trigonometric relationship in Figure \ref{fig:carsmotion}:
\begin{align}
x^*_{i, t} &= x^*_{i, t-1} + v_{i, t} \cos(\delta_{i, t}) \label{cars:xEq}, \\
y^*_{i, t} &= y^*_{i, t-1} + v_{i, t} \sin(\delta_{i, t}) \label{cars:yEq}.
\end{align}
Note that when $\delta_{i, t} \pm \pi/2$ then car $i$ takes a position parallel to the centre line, and hence $x^*_{i, t} = x^*_{i, t-1}$.
\\

From this relationship, and the coordinate sequences $\{x^*_{i, s} | s=1,2,...,T\}$ and $\{y^*_{i, s} | s=1,2,...,T\}$, the inputs to motion from driver $i$ are calculated via
\begin{align}
\delta_{i, t} &= 
     \begin{cases}
       \tan^{-1}\left(\frac{(y^*_{i, t} - y^*_{i, t-1})}{(x^*_{i, t} - x^*_{i, t-1})} \right)  &\quad\text{if }x^*_{i, t} \neq x^*_{i, t-1} \\
       \frac{\pi}{2} &\quad\text{if } y^*_{i, t} > y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       -\frac{\pi}{2} &\quad\text{if } y^*_{i, t} < y^*_{i, t-1} \mbox{ and } x^*_{i, t} = x^*_{i, t-1} \\
       \delta_{i, t-1} &\quad\text{otherwise,} \\ 
     \end{cases} \label{cars:dEq} \\
v_{i, t} &= \sqrt{(x^*_{i, t} - x^*_{i, t-1})^2 + (y^*_{i, t} - y^*_{i, t-1})^2} \label{cars:vEq}.
\end{align}
In addition, the corresponding acceleration sequence is given by $\{a_{i, t},t=3,4,...,T\}$, according to
\begin{equation}
\label{cars:aEq}
a_{i, t} = v_{i, t} - v_{i, t-1}. 
\end{equation}
\\

Data resulting from the observation of vehicle $i$ for a period of time $T$ is defined as $z_{i, 1:T} = \{x^*_{i, s}, y^*_{i, s}, | s = 1, \ldots, T\}$ and data from multiple drivers are then collected as $\mathbf{z}_{1:N} = \{z_{i, 1:T} | i = 1, \ldots, N\}$. Some additional vehicles are excluded if they have less than 500 observations, and the remainder are randomly split into two sets of vehicles: a training set of 2000 vehicles and a test set of a further 1500 vehicles.
\\

The training set and the test set are treated as observations from vehicles at two distinct time periods, illustrated in Figure \ref{fig:carsproblem}. The training set in the left panel represents any vehicle for which data is collected during the development of the self-driving vehicle, for example this data could be collected by human driven vehicles. The test set in the right panel represents those vehicles that the self-driving vehicle will encounter when it is driving autonomously, which require trajectory forecasts in close to real time.

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.4\textwidth]{figures/motion}
\caption{A vehicle at coordinate $\{x^*_{t-1}, y^*_{t-1}\}$ at time $t-1$ will be at $\{x^*_t, y^*_t\}$ at time $t$ if its angle over this period is $\delta_t$ and velocity is $v_t$.}
\label{fig:carsmotion}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/problem}
\caption{Left: The paths of vehicles in the training set. Right: The self-driving car (Solid) and two nearby vehicles (Outlined). All motion is from left to right. The self-driving vehicle must forecast the trajectory of the nearby vehicles based on their observed path and any knowledge of the driver behaviour learned from the training set vehicles}
\label{fig:carsproblem}
\end{figure}

\section{Modelling}
\label{sec:carsmodels}

Three related time series models are introduced to produce trajectory forecasts, differing by their implications about the dependency between the different sets of vehicles. These are referred to as the homogeneous model, under which all drivers are treated identically and the training set is used to infer all behaviour, and two heterogeneous models: the independent heterogeneous (IH) model and the hierarchical, clustered heterogeneity (CH) model. The IH model does not incorporate information from the vehicles in the training into inference for the test set, while the CH model allows information to be shared between vehicles; learning a range of possible behaviours from the training set that can be incorporated into inference about the test set. 
\\

\subsection{A Homogeneous Time Series Model}
\label{subsec:carshomogeneous}

The framework provided by (\ref{cars:xEq}) and (\ref{cars:yEq}) allows distributional forecasts of future position coordinates of a given vehicle $i$, $\{x^*_{i, t}, y^*_{i, t} | t = T + 1, \ldots, T+H\}$, to be obtained given the position and velocity of a vehicle at time $T$, $\{x^*_{i, T}, y^*_{i, T}, v_{i, T}\}$, and distributional forecasts of the future values of $\{a_{i, t}, \delta_{i, t} | t = T + 1, \ldots, T+H\}$.  
\\

NGSIM vehicle data is observed every 100 milliseconds, so large changes in $a_{i, t}$ and $\delta_{i, t}$ may occur over multiple consecutive observations. This is highlighted in Figure \ref{fig:carspacf}, which displays a plot of the partial autocorrelation function, which measures the correlation between a variable at two different time periods, $t$ and $t + k$, conditioned on that variable at time $t + 1, \ldots t + k -1$, for both $a_{i, t}$ and $\delta_{i, t}$ for five vehicles. Large spikes in the partial autocorrelation plots, such as those at the first lag for most vehicles, demonstrate that the value of either $a_{i, t}$ or $\delta_{i, t}$ is strongly correlated with the previous time period. 
\\

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/pacf}
\caption{Top: Partial Autocorrelation plots for acceleration, $a$, for a random selection of vehicles. Bottom: Partial Autocorrelation plots for steering angle, $\delta$ for the same vehicles. Each spike for vehicle $i$ at lag $k$ in the top panel indicates that $a_{i, t-k}$ is correlated with $a_{i, t}$, similarly spikes in the bottom panel indicate that $\delta_{i, t-k}$ is correlated with $\delta_{i, t}$. Note that different vehicles do not have the same dynamic behaviour in the partial autocorrelations.}
\label{fig:carspacf}
\end{figure}

These dynamics imply that changes in acceleration and angle are predictable given the recent history of behaviour, and as an attempt to model the dependence, two univariate auto-regressive processes of orders $p$ and $q$, respectively, for $a_t$ and $\delta_t$ are employed, given by
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{j} a_{i, t-j} + \sigma_{\epsilon} \epsilon_{i, t} \label{cars:aAR} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta} \eta_{i, t} \label{cars:dAR}
\end{align}
for each $i = 1, \ldots, N$, where $\epsilon_{i, t}$ and $\eta_{i, t}$ are independently and identically distributed according to a standard normal distribution. The unconditional means for acceleration and steering and are set, respectively, to zero and $\pi/2$, corresponding to forward motion at a constant velocity.
\\

For notational convenience, and to facilitate implementation of the Bayesian inferential approach described in Chapter \ref{chap:Background}, the variance parameters $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$ are transformed and collected together with all other model parameters into the $p + q + 2$-dimensional vector
\begin{equation*}
\label{cars:thetaVec}
\theta = \{\phi_{1}, \ldots, \phi_{p}, \gamma_{1}, \ldots, \gamma_{q}, \log(\sigma^{2}_{\epsilon}), \log(\sigma^{2}_{\eta})\}.
\end{equation*}
With each component of $\theta$ able to take on any real value, the prior distribution given by
\begin{equation}
\label{cars:indPrior}
\theta \sim \mathcal{N}\left(\mu, \Sigma \right)
\end{equation}
is selected with $\mu=(0_p^{\prime},0_q^{\prime},-5, -5)$ and $\Sigma = 10 \mathbb{I}_{p+q+2}$, where $0_r$ and $\mathbb{I}_r$ denote the $r$-dimensional zero vector and identity matrix, respectively, noting that the scale of changes in either acceleration or angle at the 100 millisecond time-scale are both very small. The joint specification in (\ref{cars:aAR}) and (\ref{cars:dAR}), along with the given normal prior distributional assumption, is referred to as the \textit{homogeneous model} throughout the chapter.
\\
Treating the first $\max(p, q)$ observations for each vehicle as deterministic, the posterior distribution for the homogeneous model is given by
\begin{equation}
\label{cars:homogPost}
p(\theta | \textbf{z}_{1:N}) \propto \prod_{i=1}^N \prod_{t = \max(p, q) + 1}^T p(z_{i, t} | z_{i, t-\max(p, q):t-1}, \theta)p(\theta).
\end{equation}

\subsection{Two Heterogeneous Time Series Models}
\label{subsec:carsheterogeneous}

The behaviour of a driver, and the dynamics of their actions, may depend on many individual features such as their experience, type of vehicle, personality, and the characteristics surrounding traffic. To accommodate such variation, heterogeneity, two so-called \textit{heterogeneous models} are introduced, namely an \textit{independent heterogeneous} (IH) model and a dependent \textit{clustered heterogeneous} (CH) model. In both cases, the heterogeneity is allowed by replacing the shared $\theta$ parameter vector with $N$ individual parameter vectors $\theta_i$, with the components of $\theta_i=\{\phi_{i,1}, \ldots, \phi_{i,p}, \gamma_{i,1}, \ldots, \gamma_{i,q}, \log(\sigma^{2}_{\epsilon,i}), \log(\sigma^{2}_{\eta,i})\}$ corresponding to the parameters given in the model for vehicle $i$ given in
\begin{align}
a_{i, t} &= \sum_{j = 1}^p \phi_{i, j} a_{i, t-j} + \sigma_{\epsilon, i} \epsilon_{i, t} \label{cars:aAR2} \\
\delta_{i, t} &= \pi/2 + \sum_{j = 1}^q \gamma_{i, j} (\delta_{i, t-j} - \pi/2) + \sigma_{\eta, i} \eta_{i, t}. \label{cars:dAR2}
\end{align}
with $\epsilon_{i,t}$ and $\eta_{i,t}$ all independent standard normal random variables. For the IH model, each $\theta_i,$ for $i=1,2,\ldots, N$, is assumed to follow an independent $N(\mu, \Sigma)$ prior distribution, resulting in a joint posterior distribution that may be decomposed into $N$ distinct independent marginal distributions, i.e.
\begin{equation}
p(\theta_{1:N} \mid z_{1:N}) = \prod_{i=1}^N p(\theta_{i} \mid z_{i}),
\end{equation}
due to the distinct parameters and independent (albeit identical) prior distributions for each of the $N$ models. \footnote{Recall that $z_{i}=\{x^*_{i,s},y^*_{i,s},\mbox{ for } s=1,2,...,T \}$.}
\\

In contrast, the alternative CH model does not assume \textit{a priori} independence of each of the $\theta_i$ vectors and instead clusters individual $\theta_i$ values into one of $K$ common latent multivariate normal distributions as determined by $\beta.$ The hierarchical CH model assumes that $\beta$ is itself random, with $\beta \sim p(\beta)$. By augmenting each driver's individual parameter $\theta_i$ with the corresponding $k_i=j$ variable, where $j$ can take on any value $1,2 \ldots, K$, gives the conditional distributions
\begin{equation}
\label{cars:mixPrior}
\theta_i | k_i = j \sim N(\mu_j, \Sigma_j),
\end{equation}
and
\begin{equation}
k_i \mid \beta \sim \mbox{Multinomial}\left(\pi_1, \ldots, \pi_{K}\right)
\end{equation}
for each $i = 1, \ldots, N$. The mixture indicator variables satisfying $k_i=j$ determine a cluster of similar $\theta_i$ parameters corresponding to the behaviours of a subset of drivers. $\beta$ contains all of the determinants of these distributions, with $\beta = \{\mu_j,\Sigma_j, \pi_j, \mbox{ for } j=1,\},$ with \textit{a priori} assumptions for each cluster given by  
\begin{align}
\mu_j &\sim \mathcal{N}\left(\bar{\mu}_j, \Omega_j\right), \\
\Sigma_j &\sim \mbox{Inverse Wishart}\left(\mbox{Degrees of Freedom } \tau_j, \mbox{Scale } \Psi_j\right), \\
\boldsymbol{\pi} &\sim \mbox{Dirichlet}\left(\alpha_1 = \alpha_2 = \ldots = \alpha_K\right).
\end{align}

Each $\bar{\mu}_j$ contains a $0$ for each $\phi$ and $\gamma$, and $-5$ for each log-variance, each $\Omega_j = 10 \mathbb{I}$, each $\tau_j = 6$, each $\Psi_j = \mathbb{I}$ and finally each $\alpha_j = 1$.

To illustrate that the hierarchical CH model with $K=6$ adequately captures the main features of the IH model, the top panel of Figure \ref{fig:carsHierSingleKDE} displays kernel density estimates of the posterior means, for each $\theta_i, i = 1, \ldots, N$, resulting from the IH model. For this illustration, $p = q = 2$ and $N=2000$ observations are used for times $t=1, 2, \ldots, T = 500$. These summaries each indicate a wide range of values for the $N$ posterior means corresponding to each model parameter, corresponding to heterogeneity across drivers in the sample. Notably, the posterior means for each of the variance parameters, $\sigma^2_{\epsilon}$ and $\sigma^2_{\eta}$, exhibit strong positive skewness, while the distribution of posterior means is multi-modal for $\phi_1$ and $\phi_2$.

These summaries are compared against the figures in the bottom panel of Figure \ref{fig:carsHierSingleKDE}, where the estimated marginal distributions for $p(\theta_{2001} | \textbf{z}_{1:2000})$are displayed. The CH marginal posterior distributions appear to have captured the skewness and multi-modality present in the corresponding IH posterior mean summaries shown in the top panel. 
\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/HierSingleKDE.png}
\caption{Top: A Kernel Density Estimate applied to $E(\theta_i | z_i), i = 1, \ldots, 2000$ where $p = q = 2$ from the IH model, sampled by Metropolis-Hastings MCMC. Bottom: $p(\theta_{2001}| | \textbf{z}_{1:2000})$ for the CH model with $K = 6$ using the same vehicles sampled by Metropolis-Hastings MCMC. Each individual component in the mixture is denoted by the coloured densities. The CH model has captured the multi-modality and skewness present in the IH posterior means.}
\label{fig:carsHierSingleKDE}
\end{figure}

\subsection{Encountering Additional Vehicles}
\label{subsec:carsadditionalVehicles}

For the remainder of this chapter $\textbf{z}_{1:N}$ and $\theta_{1:N}$ refer, respectively, to the data collected from, and the associated parameter vectors for, the $N$ vehicles in the training set: the vehicles where information is available for before any self-driving vehicle enters the road. Use of the subscript $i$ is reserved for some $i > N$, and refers to the attributes (data or a parameter vector) for an additional vehicle in the test set that may be encountered while the self-driving vehicle is on the road. Inference about $\theta_{1:N}$, the individual parameter vectors for the IH and CH models, as well as for $\theta$, the shared parameter vector under the homogeneous model, and $\beta$, the CH hyper-parameter, conditional on $z_{1:N}$ are assumed to be available before driving.
\\

The idea pursued is to use the available data, and corresponding model posterior distributions, to forecast the future behaviour of a vehicle that is positioned near a given self-driving vehicle. Leveraging off of the information gathered from previous data (via previously computed model posteriors that accommodate the range of heterogeneous behaviour of the population of vehicles) together with fresh information received in real-time, with an aim to predict the near future position of an additional vehicle moving contemporaneously with the self-driving one
\\

A crucial aspect of the application of the discussed time series models is through the relationship between the posterior distribution formed from observations relating to the first $N$ vehicles and the parameter vector for the additional new vehicle, $\theta_{N+1}$. Under the homogeneous model $\theta$ is shared across all vehicles and hence $\theta_{i} \equiv \theta$. In this case, if $N$ is large $p(\theta | \textbf{z}_{1:N}, z_{i, 1:T})$ will not vary much from $p(\theta | \textbf{z}_{1:N})$. Figure \ref{fig:carshomogPosterior} displays plots of the marginal posterior distributions associated with each element of $\theta$, with $p = q = 2$ resulting from $N = 2000$ vehicles. Arguably each of these marginal posterior distributions are very precise, and are unlikely to change much if updated using information from observations from the $(N+1)^{st}$ vehicle.

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/homogPosterior}
\caption{The marginal posterior distributions $p(\theta | \textbf{z}_{1:N})$ for the homogeneous model with $N = 2000$ and $p = q = 2$. Each variable has very high posterior precision.}
\label{fig:carshomogPosterior}
\end{figure}

The IH model implies that the posterior distribution for $\theta_i$ is independent of $\theta_{1:N}$ and $\textbf{z}_{1:N}$ by
\begin{equation}
p(\theta_{i}| \textbf{z}_{1:N}, z_{i,1:T}, \theta_{1:N}) = p(\theta_{i} | z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_i) p(\theta_i).
\label{cars:indepNewCar}
\end{equation}

The posterior for $\theta_i$ using the CH model, assuming conditional independence between each $\theta_j$ for $j = 1, \ldots, N$ and $\theta_i$, given $\beta$, $\textbf{z}_{1:N}$, and $z_i$, is
\begin{equation}
\label{cars:hierNewCar}
p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) \int_{\beta} p(\theta_{i} | \beta) p (\beta | \textbf{z}_{1:N}) d\beta.
\end{equation}
As N is large the sampled posterior $p(\beta | \textbf{z}_{1:N})$ has high precision, and $\int_{\beta} p(\theta_{i} | \beta) p (\beta | \textbf{z}_{1:N}) d\beta$ can be replaced with $p(\theta_{i} | \hat{\beta})$ in (\ref{cars:hierNewCar}) where $\hat{\beta}$ is a point estimate such as the mean or maximum of $p(\beta | \textbf{z}_{1:N})$. This approximation results in the closed form expression
\begin{equation}
\label{cars:hierNewCar2}
p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T}) \propto p(z_{i, 1:T} | \theta_{i}) p(\theta_{i} | \hat{\beta}).
\end{equation}

Combining a fast inference procedure such as UVB with the test vehicle posterior distributions implied by either of the IH and CH models, (\ref{cars:indepNewCar}) and (\ref{cars:hierNewCar2}), will allow online, heterogeneous, trajectory forecasts to be produced as the self-driving vehicle observes data for the vehicles it encounters.

\section{Results}
\label{sec:carsResults}

After obtaining RWMH-MCMC samples of the posterior distributions for each model conditioned on the training set of 2000 vehicles, with $p = q = 2$ and $K = 6$, forecasts are created and evaluated on the further 1500 vehicles in the test set. For each of these vehicles $i > 2000$, after each of $T = 100, 110, \ldots, 450$ observations, and at every forecast horizon $h = 1, \ldots, 30$, point estimates of $z_{i, T+h}$ are obtained from every naive model. The time series models are used to provide forecast densities of $p(z_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T})$, and point estimates using the coordinate pair $\{\hat{x}_{i, t}^*, \hat{y}_{i, t}^*\}$ which maximises this density. As data is observed every 100 milliseconds, this equates to forecasting the next three seconds of movement for every vehicle after every second of observation. The forecast density for the homogeneous model is created using the samples of $p(\theta | \textbf{z}_{1:2000}, z_{i, T})$, while the IH and CH models produce $p(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T})$ before creating forecasts through three methods:
\begin{enumerate}
\item Posterior sampling with RWMH-MCMC to generate samples of $p(\theta_i | \textbf{z}_{1:2000}, z_{i, 1:T_n})$,
\item SVB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T_n})$ to the complete $z_{i, 1:T_n}$ as described in Section \ref{subsec:SVB},
\item UVB fitting $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T_n})$ using $q_{\lambda_{n-1}}(\theta_{i} | \textbf{z}_{1:2000}, z_{i, 1:T_{n-1}})$ and data $z_{i, T_{n-1}+1:T_{n}}$ as described in Chapter \ref{chap:UVB}.
\end{enumerate}

Each of UVB and SVB is repeated twice, with approximating distributional families $q$ chosen to be a sigle component multivariate normal distributions and a six component mixture of multivariate normal distributions. VB applications with a single component approximations use the re-parameterised gradient estimator, and mixture approximations use the score gradient estimator as discussed in Section \ref{subsec:SVB}. A summary of the approaches is provided in Table \ref{table:carsAlg}. 
\\

Point estimate forecasts are evaluated by their Euclidean Error in metres,
\begin{equation}
\mbox{EE}_{i, T, h} = \sqrt{\left(\hat{x}^*_{i, T+h} - x^*_{i, T+h} \right)^2 + \left(\hat{y}^*_{i, T+h} - y^*_{i, T+h} \right)^2},
\label{cars:eucError}
\end{equation}
where $\{\hat{x}^*_{i, T+h}, \hat{y}^*_{i, T+h}\}$ is a forecast of $\{x^*_{i, T+h}, y^*_{i, T+h}\}$, while density forecasts are evaluated by their logscore,
\begin{equation}
\mbox{LS}_{i, T, h} = \log \left(p\left(x^*_{i, T+h}, y^*_{i, T+h} | \textbf{z}_{1:2000}, z_{i, 1:T} \right) \right).
\label{cars:logscore}
\end{equation}

\begin{center}
\begin{table}[htbp]
\resizebox{\textwidth}{!}{%
\begin{tabular}{| l | c | c |}
\hline
& Independent Heterogeneous Model & Clustered Heterogeneous Model \\
\hline
MCMC&\multicolumn{2}{c|}{RWMH-MCMC jointly drawing the entire $\theta_i$ vector using \citet{Garthwaite2016} to control the}\\
&\multicolumn{2}{c|}{variance matrix of the Multivariate Normal proposal distribution to obtain a 23.4\% acceptance rate for $\theta_i$ draws.}\\
\hline
VB& Variational Bayes to learn parameters&Variational Bayes using  to learn parameters\\
  &of both a single and six component mixture of &of both a single and six component mixture \\
& Multivariate Normals. & of Multivariate Normals \\
\hline
UVB&Variational Bayes fit to $100$ observations &Variational Bayes fit to $100$ observations \\
 &then updated every $10$ observations .&then updated every $10$ observations \\
 &  to update parameters of both a single and six &  to update parameters of both a single and six \\
  & component mixture of  Multivariate Normals & compnent mixture of Multivariate Normals \\
\hline
\end{tabular}}
\caption{Details of the algorithms used to produce posterior distributions for each method and prior combination}
\label{table:carsAlg}
\end{table}
\end{center}

\subsection{Naive Forecast Models}
\label{subsec:carsNaive}

Nine naive forecast models are provided where forecasts of $a_{i, t}$ and $\delta_{i, t}$ at each time $T$ are constants equal to $\hat{a}_{i, T}$ and $\hat{\delta}_{i, T}$ for $t = T+1, \ldots, T+H$. The classification of each naive model is provided in Table \ref{table:carsNaive} as a combination of one estimator for $\hat{a}_{i, T}$ and one estimator for $\hat{\delta}_{i, T}$. Acceleration estimators are the current acceleration, $\hat{a}_{i, T} = a_{i, T}$, the average acceleration over the previous second, $\hat{a}_{i, T} = 0.1 \sum_{s = T-9}^T a_{i, s}$, and zero acceleration $\hat{a}_{i, T} = 0$, equivalent to constant velocity. Similarly, steering angle estimators are the current angle, $\hat{\delta}_{i, T} = \delta_{i, T}$, the average angle over the previous second, $\hat{\delta}_{i, T} = 0.1 \sum_{s = T - 9}^T \delta_{i, s}$, and $\hat{\delta}_{i, T} = \pi/2$, corresponding to the vehicle to driving directly forward.
\begin{table}
\begin{center}
\resizebox{\textwidth}{!}{%
\begin{tabular}{|l|c|c|c|}
\hline
& $\hat{\delta}_{i, T} = \delta_{i, T}$ & $\hat{\delta}_{i, T} = 0.1 \sum_{s=T-9}^T \delta_{i, s}$ & $\hat{\delta}_{i, T} = \pi/2$ \\
\hline
$\hat{a}_{i, T} = a_{i, T}$ & Naive 1 & Naive 2 & Naive 3 \\
$\hat{a}_{i, T} = 0.1 \sum_{s=T-9}^T a_{i, s}$ & Naive 4 & Naive 5 & Naive 6\\
$\hat{a}_{i, T} = 0$ & Naive 7 & Naive 8 & Naive 9 \\
\hline
\end{tabular}}
\end{center}
\caption{Classification of naive forecast models by future acceleration and steering angle estimators.}
\label{table:carsNaive}
\end{table}

\subsection{Recurrent Neural Networks}
\label{subsec:carsRNN}
A Recurrent Neural Network (RNN) is constructed with inputs following \citet{Ding2013}, the ten most recent lags of vehicle position, angle, velocity, and time headway to the preceding vehicle. The RNN outputs the next thirty changes in the vehicle position, and is trained with gradient descent using the Adam optimiser to minimise the Euclidean error of the three second ahead point forecast corresponding to $EE_{i, T, 30}$ defined in (\ref{cars:eucError}). As of writing this report forecast results from this model are not competitive with the naive models and are not discussed further.

\subsection{Forecasting Results}
\label{subsec:carsResults}

\textit{Rough Structure:
\begin{enumerate}
\item Compare all time series to Naive
\item Compare Heterogeneous to Homogeneous in MCMC inference, focus on variance split
\item Show that UVB is comparable to MCMC for the CH Model, note UVB deterioration over time for single distribution, show timings
\end{enumerate}
}

Euclidean error for each model, averaged across test set vehicles at forecast horizons of one, two, and three seconds ($h = 10, 20$ and $30$) are provided in Figure \ref{fig:carsPredError}. Naive model performance splits into three groups according to the choice of $\hat{a}_{i, T}$, and for a given $\hat{a}_{i, T}$ the choice of $\hat{\delta}_{i, T}$ has not substantially changed the forecast error. The best performing Naive model at each forecast horizon is Naive 8, however at all forecast horizons the error for each implementation of each of the time series models are significantly improved relative to every naive model. The average forecast Euclidean error for Naive 8 and each of the time series models at each of these three forecast horizons is presented in Table \ref{table:carsError}.
\\

Figure \ref{fig:carsPredErrorZ} truncates Figure \ref{fig:carsPredError} to include only the time series models, each of which produces a forecast with a very similar amount of error. Vehicle trajacteroy forecasting based on the predictive density maximum will perform similarly with a homogeneous model compared to introducing heterogeneity with the IH and CH models. 
\\

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/predictiveError.png}
\caption{Mean predictive Euclidean error in metres for $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$ for $h \in \{10, 20, 30\}$ and $T \in \{100, 110, \ldots, 450\}$ for each model. Each time series model is significantly more accurate than the naive constant acceleration / angle models, with little discernible difference between each.}
\label{fig:carsPredError}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/predictiveErrorZoom.png}
\caption{The same results as Figure \ref{fig:carsPredError} with the Naive models excluded. Most of the time series model have very similar MAP forecast accuracy, with the exception of the UVB / IH Model combination}
\label{fig:carsPredErrorZ}
\end{figure}

\begin{table}[htbp]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{| l | c c c c c c c c|}
\hline
Horizon & Naive 8 & Homogeneous MCMC & IH MCMC & IH SVB & IH UVB & CH MCMC & CH SVB & CH UVB \\
\hline
One Second & 1.10 & \textbf{0.131} & \textbf{0.131} & \textbf{0.131} & 0.137 & \textbf{0.131} & 0.132 & 0.132 \\
Two Seconds & 1.50 & \textbf{0.226} & 0.228 & 0.228 & 0.246 & 0.227 & 0.230 & 0.227 \\
Three Seconds & 2.38 & \textbf{0.3} & 0.304 & 0.304 & 0.332 & 0.302 & 0.306 & 0.302 \\
\hline
\end{tabular}}
\caption{Mean forecast Euclidean error in metres for the the best Naive model, Naive 8, and each time series model at a forecast horizon of one second, two seconds and three seconds. Each time series model performs significantly better than Naive 8, and are very competitive with each other. The use of UVB for the IH model appears to have increased the forecast error relative to MCMC and SVB inference for this model.}
\label{table:carsError}
\end{table}

The IH and CH model allows each vehicle in the test to have a posterior variance distribution that is unique to that vehicle, while the homogeneous model must share the same posterior variance distribution across different vehicles. Differences in the distribution of either variance parameter has little impact on the location of the maximum of $p(z_{i, T+h} | \textbf{z}_{1:N}, z_{i, 1:T})$, but can substantially affect the precision of this forecast density. Systematic differences in the forecast uncertainty between the CH and homogeneous models result from this different treatment of variance.
\\

Figure \ref{fig:carsposVarMean} displays plots of density of posterior means for each of acceleration and steering angle resulting from RWMH-MCMC sampling with $T = 400$. The homogeneous model posterior means are indicated with vertical black lines. After $400$ observations, the prior difference in the IH and CH model has little impact, and the densities for each of these models are almost completely overlaid on each other. As in Figure \ref{fig:carsHierSingleKDE}, there is a large amount of variation in the distribution of variances for the test set, many of which are lower than the homogeneous model: the homogeneous posterior means are greater than $83\%$ and $98\%$ of CH posterior means for acceleration and the steering angle respectively, and $X$ and $Y$ of the IH posterior means.
\\

Figure \ref{fig:carsvarSplit} shows a detailed breakdown of the median increase in predictive logscore obtained from using both the IH and CH models with RWMW-MCMC inference relative to the homogeneous model at the three second forecast horizon. Vehicles are split into bivariate quantiles according to their RWMH-MCMC posterior variance parameter means for acceleration along the x-axis, and steering angle along the y-axis. The upper left panels correspond to vehicles with lower variances, showing clear benefits for heterogeneous model trajectory forecasts relative to the homogeneous model. The difference is increasing, as the heterogeneous model parameters better capture the characteristics of the individual driver a the amount of observed data increases. However the median difference is zero, or even negative, for forecasts of drivers in the fifth quintiles. This paper argues that heterogeneous modelling is still useful in this scenario, as the individual vehicle variance distributions allow identification of vehicles with high variance which could display potentially more erratic behaviour.
\\

VB adds two sources of approximation error: the standard variational error associated with basing inference off an approximation to the posterior, and an additional error introduced by replacing the posterior with a VB approximation in updates. This is displayed in Figure \ref{fig:carsvarSplitUVB}, whichs shows the analogous plot to Figure \ref{fig:carsvarSplit} based on UVB inference rather than MCMC. When a flexible mixture approximation is employed, the logscores of UVB are increased relative to the homogeneous model, and tend to improve further as more dta is observed. However, the single component approximations perform badly, with an accumlating approximation error with more data, and more updates.

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/posVarMean}
\caption{Kernel Density Estimates for the posterior mean of the variance parameters, $\sigma^2_{\eta, i}$ and $\sigma^2_{\epsilon, i}$ for each of the 1387 forecasted cars fit by RWMH-MCMC at $T = 450$ using the CH model, compared to the homogeneous posterior mean in red. Small amounts of cars with high variance in the training set has increased the homogeneous variances.}
\label{fig:carsposVarMean}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth, height = 0.55\textheight]{figures/varSplit}
\caption{Difference in logscore between the heterogeneous models with posterior inference using RWMH-MCMC and the homogeneous model. Results are split by variance quintiles for acceleration (x-axis) and angle (y-axis). There are substantial improvements in logscore, and thus forecast certainty, for vehicles with a low variance in either variable. Vehicles in the fifth quintiles, close to the homogeneous variance estimates, benefit instead from the higher precision associated with the homogeneous posterior distribution due to the large $N$.}
\label{fig:carsvarSplit}
\end{figure}

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth, height = 0.55\textheight]{figures/varSplitUVB}
\caption{Difference in logscore between the heterogenous models with UVB inference and the homogeneous model. Results are split by variance quntiles as in Figure (\ref{fig:carsvarSplit}). UVB inference with mixture distribution approximations out-performs the homogeneous model, but single component approximations deteriorate as the amount of data increases.}
\label{fig:carsvarSplitUVB}
\end{figure}

The benefits of heterogeneous models can only be realised if the posterior distribution $p(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T})$, or its approximation $q_{\lambda_T}(\theta_{i} | \textbf{z}_{1:N}, z_{i, 1:T})$ can be inferred in the short time frames demanded by self-driving vehicles. Figure \ref{fig:carstiming} demonstrates the average computation time required for SVB (Black) and UVB (coloured), where UVB is first fit to $100$ observations then repeatedly updated every $T-S$ observations for different values of $T-S$. The computation time for SVB increases linearly with $T$, while UVB is constant as $T$ increases; and hence UVB posterior inference can be implemented in an online setting. Parallelising the $M$ RQMC samples per iteration can further reduce computation time, allowing for shorter update periods or more complex models. 
\\

\begin{figure}[htbp]
\centering
\includegraphics[width = 0.75\textwidth]{figures/timing}
\caption{Average time to converge for SVB (Black) and UVB (Coloured) with four different update lengths given by $T - S$, truncated to 25 seconds. Each UVB procedure originally fits the model to the first 100 data points and updates this every $T - S$ data points. The convergence time for SVB increased linearly for $T$, but is constant for UVB. Each variational algorithm is ran with $M = 50$ on one CPU core. Convergence time could be reduced through parallelisation.}
\label{fig:carstiming}
\end{figure}
\begin{figure}[htbp]
\centering
\includegraphics[width = 0.95\textwidth]{figures/updateCost}
\caption{Differences in predictive logscores for $p(z_{i, T+30} | z_{i, 1:T})$ for each $T = 100, 110, \ldots, 450$ between UVB and SVB for the CH model (left) and the IH model (right). The differences are typically small for the CH model, while the IH model approximation has increased the logscore relative to SVB. This may be due to the form of the approximating distribution for the IH model being an inadequate approximation of the true posterior}
\label{fig:carsupdateCost}
\end{figure}

\section{Discussion}
\label{sec:carsdisc}
This chapter demonstrates that the use of auto-regressive time series models drastically improves trajectory forecasts when compared to the competing models.  These models parametrise the dynamics of a driver's actions, with a range of parameter posterior means providing evidence of heterogeneity between the drivers of different vehicles. Incorporating this heterogeneity through a hierarchical model has improved the forecast logscores, and hence reduced uncertainty, but requires an approximate inference scheme implemented with UVB. 
\\

The additional approximation error introduced by UVB does not materialise as a reduction in forecasting performance relative to exact MCMC inference when an approximating distribution that has sufficient flexibility to approximate the true posterior distribution is used. Augmenting self-driving vehicles with hierarchical time series models to forecast the trajectory of surrounding traffic, and repeatedly updating the posterior distribution and forecasts, can be feasibly implemented with UVB.
\\

\chapter{Rough ideas}

\section{Empirical Likelihood Variational Bayes}

\citet{Chib2017} introduces Bayesian inference based on moment conditions, basing inference on the satisfication of the $d-$dimensional moment conditions,

\begin{equation}
\label{emp:moments}
E \left[ g(Y, \theta) \right] = 0,
\end{equation}
where the expectation is with respect to the true data generation process $p$.

The likelihood is replaced with the exponentially tilted empirical likelihood (ETEL), defined by
\begin{equation}
\label{emp:etel}
p(y_{1:T} | \theta) = \prod_{t=1}^T p^*_t(\theta)
\end{equation}
where each $p^*_t$ is chosesn to minimise the KL divergence between $(p_1, p_2, \ldots, p_T)$ and the empirical probabilities $(1/T, 1/T, \ldots 1/T)$ subject to the conditions
\begin{equation}
\label{emp:sumRestrict}
\sum_{t=1}^T p_t = 1
\end{equation}
and
\begin{equation}
\label{emp:momentRestrict}
\sum_{t=1}^T p_t g(y_t, \theta) = 0.
\end{equation}
It can be shown that the solution to this is given by
\begin{equation}
\label{emp:solnP}
p^*_t(\theta) = \frac{\exp \{\hat{\lambda}^{\prime} g(y_t, \theta) \}}{ \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}}
\end{equation}
where
\begin{equation}
\label{emp:solnLambda}
\hat{\lambda} = \lambda(\theta) = \arg \underset{\lambda}{\min} \frac{1}{T} \sum_{t=1}^T \exp \{\lambda^{\prime} g(y_t, \theta) \}.
\end{equation}

This results in the ETEL posterior
\begin{equation}
\label{emp:post}
p(\theta | y_{1:T}) \propto p(\theta) \prod_{t=1}^T\frac{\exp \{\hat{\lambda}^{\prime} g(y_t, \theta) \}}{ \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}}.
\end{equation}

\citet{Chib2017} proposes a Metropolis Hastings sampler for the posterior, following the steps
\begin{enumerate}
\item Draw $\theta$ from a suitable candidate distribution
\item Solve (\ref{emp:solnLambda}) for this draw
\item Calculate (\ref{emp:post}) as part of the Metropolis Hastings acceptance probability
\item Accept or reject $\theta$ as per standard MH
\end{enumerate}
This sounds quite slow, and may benefit from VB.
\\

Reparameterised VB (and also Stein) requires the derivative of $\log(p(\theta, y_{1:T})$ with respect to $\theta$, which is complicated by $\lambda$ being a function of $\theta$.
\\

If, for every $\theta \in \Theta$, the solution to  (\ref{emp:solnLambda}) is unique, and, defining $h(\theta, \lambda(\theta)) =  \frac{1}{T} \sum_{t=1}^T \exp \{\lambda^{\prime} g(y_t, \theta) \}$, the hessian matrix $\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda^2}$ is positive-defiite, the optimal $\hat{\lambda}$ is found by satisfying the first order condition
\begin{equation}
\label{emp:lambdaDeriv}
\frac{\partial h(\theta, \lambda(\theta))}{\partial \lambda} = 0.
\end{equation}
Taking the derivative of both sides of this with respect to $\theta$ results in
\begin{equation}
\label{emp:thetaDeriv}
\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \theta^{\prime}} + \frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \lambda^{\prime}} \frac{\partial \lambda(\theta)}{\partial \theta} = 0,
\end{equation}
where
\begin{equation}
\label{emp:hessian}
\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \lambda^{\prime}} = \frac{1}{T} \sum_{t=1}^T g(y_t, \theta) g(y_t, \theta)^{\prime} \exp\{\hat{\lambda}^{\prime} g(y_t, \theta) \}
\end{equation}
and
\begin{equation}
\label{emp:crossDeriv}
\frac{\partial^ 2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \theta^{\prime}} = \frac{1}{T} \sum_{t=1}^T \exp\{\hat{\lambda}^{\prime} g(y_t, \theta) \} \left( \mathbb{I} + g(y_t, \theta) \hat{\lambda}^{\prime} \right) \frac{\partial g(y_t, \theta)}{\partial \theta}.
\end{equation}
(\ref{emp:thetaDeriv}) may be solved for $\partial \lambda(\theta) / \partial \theta$ by
\begin{equation}
\label{emp:thetaDerivSoln}
 \frac{\partial \lambda(\theta)}{\partial \theta} = -\left(\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \lambda^{\prime}}\right)^{-1}\frac{\partial^2  h(\theta, \lambda(\theta))}{\partial \lambda \partial \theta^{\prime}}
\end{equation}

We may then take the derivative of the log-likelihood with respect to $\theta$ via
\begin{align}
\frac{\partial \log(p(y_{1:T}, \theta))}{\partial \theta} &= \frac{\partial \log(p(\theta))}{\partial \theta} + \frac{\partial}{\partial \theta} \sum_{t=1}^T \log \left( \frac{\exp \{\hat{\lambda}^{\prime} g(y_t, \theta) \}}{ \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}} \right), \nonumber \\
&= \frac{\partial \log(p(\theta))}{\partial \theta} + \sum_{t=1}^T \frac{\partial}{\partial \theta} \left[\hat{\lambda}^{\prime} g(y_t, \theta) \right]  - T \frac{\partial}{\partial \theta} \log \left(  \sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \} \right), \nonumber \\
&= \frac{\partial \log(p(\theta))}{\partial \theta} + \sum_{t=1}^T \left(\frac{\partial \hat{\lambda}^{\prime}}{\partial \theta} g(y_t, \theta) + \frac{\partial g(y_t, \theta)^{\prime}}{\partial \theta}\hat{\lambda}  \right) \nonumber \\
&- \frac{T}{\sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \}}\left(\sum_{s=1}^T \exp \{\hat{\lambda}^{\prime} g(y_s, \theta) \} \left(\frac{\partial \hat{\lambda}^{\prime}}{\partial \theta} g(y_s, \theta) + \frac{\partial g(y_s, \theta)^{\prime}}{\partial \theta}\hat{\lambda}  \right) \right) /
\end{align}

If the moment conditions are differentiable with respect to $\theta$ we may use ETEL in place of the likelihood function in Variational Bayes.

\section{`Particle' Variational Bayes}

The following is inspired by \cite{Saeedi2015} who came up with the idea to approximate a set of latent states with a KL divergence minimising set of particles and weights. However they propose an algorithm that is only applicable to discrete parameters with finite support
\\

In contrast, I extend this via gradient methods with a continuous $\theta$.
\\

Variational Algorithms are often constrained by a mismatch between the shape of the approximating family and the true posterior. While considerable work in the literature has been applied to creating extremely flexible approximating forms (with huge amounts of parameters), I use an arbitrarily flexible approximating family of the form
\begin{equation}
\label{Particle:Q}
q_{\lambda}(\theta | y_{1:T}) = \sum_{i=1}^N w^{(i)} \delta(\theta^{(i)})
\end{equation}
where $\delta$ is the Dirac delta function. If $\theta$ is a $k-$dimensional vector, this distribution is parameterised by $N$ weights and $N$ $k-$dimensional particle vectors. Calculation of the gradients of these parameters do not require samples from the posterior, so evaluating each parameter at $N$ separate particles is partially ofset by not requiring the $S$ gradient Monte-Carlo calculations. 
\\

From \cite{Saeedi2015}, if there are no duplicated particles, the ELBO for this model can be written as
\begin{equation}
\label{Particle:ELBO}
\mathcal{L}(q, \lambda) = \sum_{i=1}^N w^{(i)} \log \left( \frac{p\left(y_{1:T}, \theta^{(i)}\right)}{w^{(i)}} \right).
\end{equation}
With the constraint $\sum_{i=1}^N w^{(i)} = 1$, the ELBO may be maximised via the Lagrangian
\begin{equation}
\label{Particle:ELBOlagr}
\tilde{\mathcal{L}}(q, \lambda) = \sum_{i=1}^N w^{(i)} \log \left( \frac{p\left(y_{1:T}, \theta^{(i)}\right)}{w^{(i)}} \right) + \lambda\left(\sum_{i=1}^N w^{(i)} - 1\right).
\end{equation}
Taking derivatives with respect to $w^{(i)}$ and $\lambda$, and setting these to zero we obtain
\begin{align}
\frac{\partial \tilde{\mathcal{L}}(q, \lambda)}{\partial \lambda} = 0 &\rightarrow \sum_{i=1}^N \hat{w}^{(i)} = 1, \label{Particle:lambdaDiff} \\
\frac{\partial \tilde{\mathcal{L}}(q, \lambda)}{\partial w^{(i)}} = 0 &\rightarrow \hat{w}^{(i)} = p\left(y_{1:T}, \theta^{(i)}\right) \exp\left\{\hat{\lambda} - 1\right\}. \label{Particle:wDiff} 
\end{align}
Substituting (\ref{Particle:wDiff}) into (\ref{Particle:lambdaDiff}) we obtain
\begin{align}
\hat{\lambda} &= 1 - \log\left(\sum_{i=1}^N p\left(y_{1:T}, \theta^{(i)}\right) \right), \\
\hat{w}^{(i)} &= \frac{p\left(y_{1:T}, \theta^{(i)}\right)}{\sum_{i=1}^N p\left(y_{1:T}, \theta^{(i)}\right)}.
\end{align}
We may then substitute the optimal weights into (\ref{Particle:ELBO}),
\begin{align}
\hat{\mathcal{L}}(q, \lambda) &= \sum_{i=1}^N \hat{w}^{(i)} \log \left( \frac{p\left(y_{1:T}, \theta^{(i)}\right)}{\hat{w}^{(i)}} \right), \nonumber \\
 &= \sum_{i=1}^N \frac{p\left(y_{1:T}, \theta^{(i)}\right)}{\sum_{j=1}^N p\left(y_{1:T}, \theta^{(j)}\right)} \log \left( \frac{p\left(y_{1:T}, \theta^{(i)}\right)\sum_{j=1}^N p\left(y_{1:T}, \theta^{(j)}\right)}{p\left(y_{1:T}, \theta^{(i)}\right)} \right), \nonumber \\
 &= \log \left( \sum_{j=1}^N p\left(y_{1:T}, \theta^{(j)}\right) \right). \label{Particle:ELBOwOpt}
\end{align}
From this stage \cite{Saeedi2015} propose an algorithm to manually search over the support of each $\theta$ particle and choose the $N$ values that maximises (\ref{Particle:ELBOwOpt}). Instead we may find the optimal particles $\theta^{(i)}$ \ by gradient ascent, taking the derivative of the ELBO with respect to $\theta^{(i)}$,
\begin{equation}
\label{Particle:thetaDiff}
\frac{\partial \hat{\mathcal{L}}(q, \lambda)}{\partial \theta^{(i)}}
= \frac{\partial / \partial \theta^{(i)} p\left(y_{1:T}, \theta^{(i)}\right)}{\sum_{j=1}^N p\left(y_{1:T}, \theta^{(j)}\right)}
\end{equation}
These derivatives are not stochastic, but standard $\lambda$ update rules apply.
\\

\textit{On reflection, (\ref{Particle:thetaDiff}) will equal zero at the MAP and all of the particles will converge to this one value (unless they get stuck at local optima). If multiple particles have the same value then (\ref{Particle:ELBO}) should be replaced with
\begin{equation}
\mathcal{L}(q, \lambda) = \sum_{i=1}^N \sum_{j=1}^N \delta(\theta^{(i)}, \theta^{(j)})w^{(j)} \log \left( \frac{p\left(y_{1:T}, \theta^{(i)}\right)}{ \sum_{j=1}^N \delta(\theta^{(i)}, \theta^{(j)})w^{(j)}} \right).
\end{equation}
A closed form solution for $w^{(i)}$ exists in this version but the delta functions are not differentiable and hence we cannot optimise this for the $\theta$ particles.}

\iffalse
\section{UVB with periodic MCMC updates}

We continuously talk about running MCMC in the background and using the results to correct the VB approximation error. We never really decided how to do this.
\\

Consider a scenario where we are constantly seeing additional data and have a UVB process running to get posterior approximations $q_{\lambda}(\theta | y_{1:T_n})$, but at some stage the data pauses (eg. overnight) for long enough to run an MCMC (or a variant) and obtain a posterior sample $p(\theta | y_{1:T_n})$. 
\\

We would want to use this information in two ways:
\begin{enumerate}
\item Learn about the shape of the posterior and pick a suitable family $q$.
\item Correct the accumulated error in the UVB approximation.
\end{enumerate}

To accomplish these tasks we want to pick a parametric distribution that fits the samples fairly well. There are many ways to do this treaing the samples as `data', and $q$ as a `likelihood', but perhaps to most cohesive choice is to pick the distribution $q$ that minimises $KL[q_{\lambda}(\theta | y_{1:T_n}) || p(\theta | y_{1:T_n})]$.
\\

This is just VB! We could use the score gradient estimator,
\begin{equation}
\label{MCMCVB:scoreEst}
\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}_{SC} = \frac{1}{N} \sum_{i = 1}^N \frac{\partial \log(q_{\lambda}(\theta^{(i)} | y_{1:T_n}))}{\partial \lambda} \log \left(\frac{p(y_{1:T_n}, \theta^{(i)})}{q_{\lambda}(\theta^{(i)} | y_{1:T_n}} \right),
\end{equation}
where $\theta \sim q_{\lambda}(\theta | y_{1:T})$, which slows down for large $T_n$ as it has to evaluate $\log(p(y_{1:T}, \theta)$ for each $\theta$ sample. However, we can use our posterior samples in an importance sampler gradient,
\begin{equation}
\label{MCMCVB:DerivP}
\widehat{\frac{\partial\mathcal{L}(q, \lambda)}{\partial \lambda}}_{IS} = \frac{1}{N} \sum_{i=1}^N w(\theta^{(i)}) \frac{\partial \log(q_{\lambda}(\theta^{(i)} | y_{1:T_n}))}{\partial \lambda} \log \left( \frac{p(y_{1:T_n}, \theta^{(i)})}{q_{\lambda}(\theta^{(i)} | y_{1:T_n})} \right)
\end{equation}
where $\theta \sim p(\theta | y_{1:T_n})$ and 
\begin{align}
\hat{w}(\theta^{(i)}) &= \frac{q_{\lambda}(\theta^{(i)} | y_{1:T_n})}{p( y_{1:T_n}, \theta^{(i)})} \\
w(\theta^{(i)}) &= \frac{\hat{w}(\theta^{(i)})}{\sum_{i=1}^N \hat{w}(\theta^{(i)})}. 
\end{align}

This approach completely bypasses the VB compuational cost, as we do not simulate new values of $\theta$ and simply re-use the original posterior samples each iteration. Hence any term involving $p(\theta | y_{1:T})$ is constant between each iterations, and we only need to re-calculate the $q_{\lambda}(\theta | y)$ terms as $\lambda$ changes. The MCMC samples can also help pick starting values for the optimisation process.

Unfortunately use of this is limited,
\begin{itemize}
\item We already have a posterior sample, and we'd only really want a KL minimising parametric approximation to a posterior to use as a UVB prior,
\item We have enough time to run an MCMC, but want a speedup on VB.
\end{itemize}
This approach being significantly faster than regular VB may be helpful from a distribution selection point, we can try many different $q$ distributions and pick the one that maximises the ELBO. However we have to pick an entire distribution, and can't just look at margins on their own etc.

\section{VB for SSM}

Consider the state-space model where
\begin{equation}
p(x_{1:T}, \theta | y_{1:T}) \propto p(\theta) \prod_{t=1}^T p(y_t | \theta, x_t)p(x_t | \theta, x_{t-1}) p(x_0 | \theta).
\end{equation}
No further assumptions on the distributions / linearity etc. are used.
\\

Supressing the products and subscripts such that $y = y_{1:T}$, $x = x_{1:T}$,
\begin{align}
p(y | \theta, x) = \prod_{t=1}^T p(y_t | \theta, x_t), \mbox{ and} \\
p(x | \theta) = p(x_0 | \theta) \prod_{t=1}^T p(x_t | \theta, x_{t-1}),
\end{align}
consider the VB approximation of the form
\begin{equation}
\label{VBSSM:Factor}
q(\theta, x) = q(\theta)p(x | \theta, y),
\end{equation}
noting that 
\begin{equation}
p(x | \theta, y) \propto p(y | x, \theta)p(x | \theta).
\end{equation}

The gradient of the ELBO in this case is
\begin{align}
\frac{\partial L}{\partial \lambda} &= \int_{x, \theta}q(x, \theta)\frac{\partial \log(q(\theta)p(x|\theta, y))}{\partial \lambda} \left(\log(p(y, x, \theta)) - \log(q(\theta)p(x | \theta, y)) \right) dxd\theta \\
&=  \int_{x, \theta}q(x, \theta) \frac{\partial \log(q(\theta))}{\partial \lambda} \left(\log(p(y, x, \theta)) - \log(q(\theta)) - \log(p(x | \theta, y)) \right)dxd\theta \\
&= \int_{x, \theta}q(x, \theta) \frac{\partial \log(q(\theta))}{\partial \lambda} \left(\log(p(y, x, \theta)) - \log(q(\theta)) - \log(p(x, y |\theta)) + \log(p(y | \theta)) \right)dxd\theta \\
&= \int_{\theta}q(\theta) \frac{\partial \log(q(\theta))}{\partial \lambda} \left(\log(p(\theta)) + \log(p(y | \theta)) - \log(q(\theta)) \right) d\theta
\end{align}
Factorising the posterior in (\ref{VBSSM:Factor}) is equivalent to marginalising the states.
\\

\citet{Gunawan2017} note that replacing $p(y | \theta)$ with an unbiased estimate, such as $\hat{p}(y | \theta)$ provided by a particle filter, is valid.
\\

\citet{Ma2017} consider the problem of estimating $p(y | \theta)$ based off of sequential sub-samples of data $y$ in the context of a hidden markov model (with states marginalised, so it is the marginal likelihood). I would be interested in seeing if this idea is extendable to a sub-sampling particle filter. The proposal distribution for the inital state in the sub-sample, $r(x_{\tau} | \theta, y_{1:\tau})$ in the particle filter would be problematic.

\section{IS-UVB for SSM}

Consider the UVB context where we have some state-space model, where the marginalised approximate posterior distribution $q_{\lambda_n}(\theta | y_{1:T_n})$ is available. We want to make an update to $T_{n+1}$ via the UVB psuedo-posterior update gradient
\begin{align}
\frac{\partial L}{\partial \lambda_{n+1}} &= \int_{\theta}q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}}) \frac{\partial \log(q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}}))}{\partial \lambda_{n+1}} \\
&\times \bigg(\log(q_{\lambda_n}(\theta | y_{1:T_{n}})) + \log(p(y_{T_{n}+1:T_{n+1}} | y_{1:T_n} \theta)) - \log(q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})) \bigg) d\theta
\end{align}
UVB can be applied if we were able to estimate $p(y_{T_{n}+1:T_{n+1}} | y_{1:T_n} \theta)$ without using all of the data.
\\

If we had access to a good proposal distribution $r(x_{T_n+1} | y_{1:T_n}, \theta)$ (or $r(x_{T_n+1} | y_{1:T_n+1}, \theta)$ etc.) we could simply start the particle filter at time $T_n+1$ and not need the previous data. Unfortunately the dependence on $\theta$ makes this a problem - we would need to make a new proposal distribution for each $\theta \sim q_{\lambda_{n+1}}(\theta | y_{1:T_{n+1}})$ sample in each iteration of the gradient algorithm. 
\\

One possibility is to apply the IS-UVB ideas and use the proposals $r(x_{T_n+1} | \theta, \cdot)$ where $\theta \sim q_{\lambda_{n}}(\theta | y_{1:T_{n}})$, then we can run the particle filter up to $T_n+1$ before optimisation and use the distribution $\hat{p}(x_{T_n+1} | \theta, y_{1:T_{n}+1}) = \sum_i w_i \delta(x_i)$ as a proposal.
\\

More work is needed to see how this distribution influences the weights of the particle filter going forward to $T_{n+1}$.

\fi

\bibliographystyle{asa}
\bibliography{references}

\appendix
\addcontentsline{toc}{chapter}{APPENDICES}
\chapter{Adam: Adaptive Moment Estimation}

In each variational approximation in this thesis, the optimisation is performed with Stochastic Gradient Ascent where the sequence $\rho^{(m)}$ is controlled by the Adam algorithm from \citet{Kingma2014b}. Adam combines features of two popular methods, the per-$\lambda$-component learning rates of AdaGrad \citep{Duchi2011}, and the momentum of RMSprop \citep{Tieleman2012}.
\\

 Adam takes exponentially decaying weighted averages of the first and second moments of the gradients, so that the error associated with a gradient estimate having the wrong sign is reduced; instead a change in sign must be repeated in successive iterations so that the weighted average gradient has a reversed sign.
\\

Each $\rho^{(m)}, m = 1, 2, \ldots, $ is a vector of dimension equal to the dimension of $\lambda$, and each element of $\rho^{m)}$ is inversely proportional to the square-root of the second moment of the gradient. This reduces the movement of elements of $\lambda$ that are associated with a large gradient variance. The weighted average moment estimates, $u^{(m)}$ and $v^{(m)}$, for the first and second gradient moments respectively, are biased towards zero and so this bias is corrected by $\hat{u}^{(m)}$ and $\hat{v}^{(m)}$, which are used in the $\lambda$ update equation. 
\\

 Adam is described in Algorithm \ref{alg:adam}, and applied throughout the thesis with $\alpha = 0.1, \beta_1 = 0.9, \beta_2 = 0.999, \epsilon = 10^{-8}$, and $\nu = 0.01$.
\\

\begin{algorithm}[H]
 \SetKwInOut{Input}{Input}
 \Input{Function $f(\lambda)$, hyper-parameters $\alpha, \beta_1, \beta_2, \epsilon, \nu$, initial values $\lambda^{(1)}$.}
 \KwResult{Optimal Parameters $\lambda$}
 Set $u_0 = 0$\;
 Set $v_0 = 0$\;
 Set $m = 1$\;
 \While{$| f(\lambda^{(m)}) - f(\lambda^{(m-1)}) | > \nu$} {
  $g^{(m)} = \frac{\delta f(\lambda)}{\delta \lambda} \rvert_{\lambda = \lambda^{(m)}}$ \;
  $u^{(m)} = \beta_1 u^{(m-1)} + (1-\beta_1) g^{(m)}$ \;
  $v^{(m)} = \beta_2 v^{(m-1)} + (1-\beta_2) (g^{(m)})^2$ \;
  $\hat{u}^{(m)} = u^{(m)} / (1 - \beta_1^m)$ \;
  $\hat{v}^{(m)} = v^{(m)} / (1 - \beta_2^m)$ \;
  $\rho^{(m)} = \alpha / (\sqrt{\hat{v}^{(m)} + \epsilon})$\;
  $\lambda^{(m+1)} = \lambda^{(m)} + \rho^{(m)} \hat{u}^{(m)}$\;
  $m = m + 1$\;
  }
 \caption{Adam Optimiser applied to Gradient Ascent to maximise $f(\lambda)$.}
  \label{alg:adam}
\end{algorithm}

\chapter{Vehicle Data Transformation to Relative Coordinates}

This section describes the process used to transform a set of vehicle coordinates $\{x_{i, t}, y_{i, t}\}$ relative to a fixed position on the Earth to a set of coordinates $\{x^*_{i, t}, y^*_{i, t}\}$ that are relative to the centre line of each highway lane.  The purpose of this transformation is to remove systematic patterns from the dataset caused by curvature in the road, so that any remaining lateral movement is caused by a change in position within a lane. 
\\

The centre line for each of the five lanes in the highway section are estimated separately, using 100 randomly sampled vehicles per lane that did not change their lane over the observed period. 
\\

For each vehicle $i$ and time $t$ since entering the road, with travel originating at $t = 1$, given by $\{x_{i,1}, y_{i,1}\}$, the total distance travelled is calculated as 
\begin{equation}
\label{transform:distance}
d_{i, t} = \sum_{s=2}^t \sqrt{(x_{i, s} - x_{i, s-1})^2 + (y_{i, s} - y_{i, s-1})^2}.
\end{equation}
Using this distance metric and the data from the sampled 100 cars per lane, the two-dimensional coordinates corresponding to the centre line of each lane are estimated via independent smoothing splines where each coordinate is a function of the distance travelled to that point. Each smoothing spline is calculated using the `R stats' package \citep{R}. The estimated centre line for lane $k$, is denoted by the curve $\{\hat{x}_{d,k} = f_{x}^k(d), \hat{y}_{d,k} = f_{y}^k(d)\}$, for $d \geq 0$. The fitted spline models are shown in red overlaying the raw data in Figure \ref{fig:transform}
\\

\begin{figure}[htbp]
\centering
\includegraphics[width=0.65\textwidth]{figures/roadStraightening}
\caption{The fitted spline models for each lane in red overlaid on the raw data in black.}
\label{fig:transform}
\end{figure}

Excluding the vehicles used to estimate the spline models, each of the vehicles in the dataset uses the relevant lane centre line estimate fit from the spline model associated with its starting lane to calculate relative coordinates $\{x^*_{i, t}, y^*_{i, t}\}$. 
$y^*_{i, t}$ denotes the distance travelled along the road, and $x^*_{i, t}$ denotes the deviation from the lane centre line, and are calculated by
\begin{align}
\hat{d}_{i, t} &= \arg \underset{d}{\min} \sqrt{(x_{i, t} - f_x^k(d))^2 + (y_{i, t} - f_y^k(d))^2}\\
\hat{x}_{i, t} &= f_x^k(\hat{d}_{i, t}) \\
\hat{y}_{i, t} &= f_y^k(\hat{d}_{i, t}) \\
x^*_{i, t} &= \mbox{sign}\left(\tan\left(\frac{\hat{y}_{i, t} - y_{i, t}}{\hat{x}_{i, t} - x_{i, t}} \right)^{-1} - \tan \left(\frac{f_y^{\prime, k}(\hat{d}_{i, t})}{f_x^{\prime, k}(\hat{d}_{i, t})}\right)^{-1} \right) \sqrt{(x_{i, t} - \hat{x}_{i, t})^2 + (y_{i, t} - \hat{y}_{i, t})^2} \label{transform:xstar} \\
y^*_{i, t} &= \hat{d}_{i, t}
\end{align}
That is, $y^*_{i, t}$ is distance travelled along the centre line curve generated by $\{f_x^k(d), f_y^k(d)\}$ to reach the postition that is closest to $\{x_{i, t}, y_{i, t}\}$; and $x^*_{i, t}$ is the distance between $\{x_{i, t}, y_{i, t}\}$ and this position. As $\{\hat{x}_{i, t}, \hat{y}_{i, t}\}$ is chosen to minimise the distance to the original coordinates, the vector from the original coordinates to the centre line , given by $(\hat{x}_{i, t} - x_{i, t}, \hat{y}_{i, t} - y_{i, t})^{\prime}$, is orthogonal to the direction of travel, given by the vector of the gradient of the spline functions, $(f_x^{\prime, k}(\hat{d}_{i, t}), f_y^{\prime, k}(\hat{d}_{i, t}))^{\prime}$, with an angle of intersect of either $\pi/2$ or $-\pi/2$. The function $\mbox{sign}(\cdot)$ in (\ref{transform:xstar}) calculates the angle between these two vectors and results in $\hat{x}_{i, t}$ taking on a negative value if the vehicle is to the left of the centre line and a positive value if the vehicle is to the right.
\\

A subset of the transformed data is shown in Figure \ref{fig:straight}.
\begin{figure}[htbp]
\centering
\includegraphics[width=0.65\textwidth]{figures/carsStraightened}
\caption{A subset of the data with road curvature removed.  Vehicle trajectories with x coordinates outside of approximately -2 to 2 indicate a lane change.}
\label{fig:straight}
\end{figure}

\end{document}

### 2018-08-28 MIVB/AR1.R

@@ -1,82 +1,0 @@
-library(Rcpp)
-library(RcppArmadillo)
-library(tidyverse)
-library(msm)
-sourceCpp('AR1.cpp')
-logPhiDens_AR1 = function(x, phi, sigmaSq){
-  term1 = 1/2 * log(1-phi^2)
-  term2 = -1/(2*sigmaSq)
-  term3 = phi^2 * (sum(x[1:T]^2) - x[1]^2)
-  term4 = -2 * phi * sum(x[1:T]*x[2:(T+1)])
-  return(term1 + term2 * (term3 + term4))
-}
-MCMC_AR1 = function(x, reps=10000, dataset){
-  phiD = rep(0, reps)
-  sigD = rep(0, reps)
-  accept = 0
-  for(i in 2:reps){
-    sigD[i] = 1/rgamma(1, (T+3)/2, 1 + 1/2 * sum((x[2:(T+1)] - phiD[i-1]*x[1:T])^2) + 1/2 * (1 - phiD[i-1]^2) * x[1]^2)
-    
-    candidate = rtnorm(1, phiD[i-1], 0.1, -1, 1)
-    canQDens = dtnorm(candidate, phiD[i-1], 0.1, -1, 1, log=TRUE)
-    oldQDens = dtnorm(phiD[i-1], candidate, 0.1, -1, 1, log=TRUE)
-    canPDens = logPhiDens_AR1(x, candidate, sigD[i])
-    oldPDens = logPhiDens_AR1(x, phiD[i-1], sigD[i])
-    ratio = min(1, exp(canPDens - oldPDens - canQDens + oldQDens))
-    if(runif(1) < ratio){
-      phiD[i] = candidate
-      accept = accept + 1
-    } else {
-      phiD[i] = phiD[i-1]
-    }
-  }
-  print(accept/reps)
-  return(data.frame(c(sigD[(reps/2+1):reps], phiD[(reps/2+1):reps]), 
-                    rep(c('sigma^2', 'phi'), rep(reps/2, 2)),
-                    dataset))
-}
-
-T = 500
-sigmaSq = 1
-phi = 0.8
-reps = 2
-VB = data.frame()
-MCMCdf = data.frame()
-
-for(i in 1:reps){
-  x = rep(0, T+1)
-  x[1] = rnorm(1, 0, sqrt(sigmaSq / (1-phi^2)))
-  for(t in 2:(T+1)){
-    x[t] = rnorm(1, phi*x[t-1], sqrt(sigmaSq))
-  }
[…]

### 2018-08-28 MIVB/AR1.cpp

@@ -1,142 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-
-double pLogDens(vec x, double sigmaSq, double phi){
-  int T = x.n_elem - 1;
-  double density = -0.5*(T+5)*log(sigmaSq) + 0.5*log(1-pow(phi, 2)) - pow(sigmaSq, -1);
-  density -= pow(x[0], 2) * (1 - pow(phi, 2)) / (2 * sigmaSq);
-  for(int t = 1; t < T+1; ++t){
-    density -= pow(x[t] - phi*x[t-1], 2) / (2*sigmaSq);
-  }
-  return density;
-}
-
-double qLogDens(vec epsilon, vec mu, mat L){
-  double logdet = -(mu[0] + L(0, 0)*epsilon[0] + log(abs(L(0, 0))) + log(abs(L(1, 1))));
-  double eps = - 0.5 * (pow(epsilon[0], 2) + pow(epsilon[1], 2));
-  return eps - logdet;
-}
-
-double ELBO(vec x, vec mu, mat L, int n=100){
-  double elbo = 0;
-  for(int i = 0; i < n; ++i){
-    vec epsilon = randn<vec>(2);
-    double sigmaSq = exp(mu[0] + L(0, 0)*epsilon[0]);
-    double phi = mu[1] + L(1, 0)*epsilon[0] + L(1, 1)*epsilon[1];
-    elbo += pLogDens(x, sigmaSq, phi) - qLogDens(epsilon, mu, L);
-  }
-  return elbo/n;
-}
-
-vec pDeriv (vec x, double sigmaSq, double phi){
-  double T = x.n_elem - 1;
-  
-  double dsigmaSq = -0.5*(T + 5) / sigmaSq + pow(sigmaSq, -2) + 
-    pow(x[0], 2) * (1 - pow(phi, 2)) / (2 * pow(sigmaSq, 2));
-  
-  double dphi = -phi / (1 - pow(phi, 2)) + phi * pow(x[0], 2) / sigmaSq;
-  
-  for(int t = 1; t < T+1; ++t){
-    dsigmaSq += pow(x[t] - phi*x[t-1], 2) / (2 * pow(sigmaSq, 2));
[…]

### 2018-08-28 MIVB/DLM_1Delbo.R

@@ -1,130 +1,0 @@
-library(tidyr)
-library(ggplot2)
-
-# Inputs: Range of mean for approx. distribution, variance of MCMC posterior, latent variables x0 & x, data y
-lbsigy = function(meanrange, var, x0, x, y){
-  # Calculate the ELBO at each value in meanrange
-  elbo = rep(0, length(meanrange))
-  # Use the same random numbers for each mean
-  eps = rnorm(100)
-  # For each mean value
-  for(j in seq_along(meanrange)){
-    # The ELBO is an average across 100 draws of sigma^2_y
-    for(i in 1:100){
-      sigy = exp(meanrange[j] + sqrt(var)*eps[i])
-      # Prior density
-      logp =  -2*log(sigy) - 1/sigy
-      for(t in 1:T){
-        # From p(y | theta)
-        logp = logp + dnorm(y[t], 1+x[t], sqrt(sigy), log=TRUE)
-      }
-      # Approx density q(log(sigma^2_y))
-      logq = dnorm(log(sigy), meanrange[j], sqrt(var), log=TRUE)
-      # All other components of log(p) and log(q) are constant with respect to sigma^2_y
-      elbo[j] = elbo[j] + (logp-logq)/100
-    }
-  }
-  return(elbo)
-}
-
-# Other three parameters work the same way (except for truncated normal sampling of phi)
-lbsigx = function(meanrange, var, x0, x, y){
-  elbo = rep(0, length(meanrange))
-  eps = rnorm(100)
-  for(j in seq_along(meanrange)){
-    for(i in 1:100){
-      sigx = exp(meanrange[j] + sqrt(var)*eps[i])
-      logp = -2.5*log(sigx) - 1/sigx - x0^2*(1-0.95^2)/(2*sigx) 
-      x = c(x0, x)
-      for(t in 1:T){
-        logp = logp + dnorm(x[t+1], 0.95*x[t], sqrt(sigx), log=TRUE)
-      }
-    logq = dnorm(log(sigx), meanrange[i], sqrt(var), log=TRUE)
-    elbo[j] = elbo[j] + (logp-logq)/100
-    }
-  }
-  return(elbo)
-}
-
-lbphi = function(meanrange, var, x0, x, y){
[…]

### 2018-08-28 MIVB/DLM_CopVB.R

@@ -1,394 +1,0 @@
-dt_ls <- function(x, df, mu, a) {1/a * dt((x - mu)/a, df)}
-pt_ls <- function(x, df, mu, a) {pt((x - mu)/a, df)}
-qt_ls <- function(prob, df, mu, a) {qt(prob, df)*a + mu}
-rt_ls <- function(n, df, mu, a) {rt(n,df)*a + mu}
-
-MarginalTransform = function(unifs, thetaDist, xDist, thetaParams, xParams, T, j=0){
-  if(j > 0){
-    if(j <= 2){
-      if(thetaDist[j]==1){
-        output = qweibull(unifs, thetaParams[1, j], thetaParams[2, j])
-      } else if(thetaDist[j]==2){
-        output = qgamma(unifs, thetaParams[1, j], thetaParams[2, j])
-      } else if(thetaDist[j]==3){
-        output = qigamma(unifs, thetaParams[1, j], thetaParams[2, j])
-      } else {
-        output = qlnorm(unifs, thetaParams[1, j], sqrt(thetaParams[2, j]))
-      }
-    } else if(j <= 4){
-      if(thetaDist[j]==1){
-        output = qnorm(unifs, thetaParams[1, j], sqrt(thetaParams[2, j]))
-      } else {
-        output = qt_ls(unifs, thetaParams[3, j], thetaParams[1, j], thetaParams[2, j])
-      }
-    } else {
-      if(xDist == 2){
-        output = qnorm(unifs, xParams[1, T+6-j], sqrt(xParams[2, T+6-j]))
-      } else {
-        output = qt_ls(unifs, xParams[3,T+6-j], xParams[1, T+6-j], xParams[2, T+6-j])
-      }
-    }
-  } else {
-    output = matrix(0, nrow(unifs), ncol(unifs))
-    for(i in 1:2){
-      if(thetaDist[i]==1){
-        output[,i] = qweibull(unifs[,i], thetaParams[1, i], thetaParams[2, i])
-      } else if(thetaDist[i]==2){
-        output[,i] = qgamma(unifs[,i], thetaParams[1, i], thetaParams[2, i])
-      } else if(thetaDist[i]==3){
-        output[,i] = qigamma(unifs[,i], thetaParams[1, i], thetaParams[2, i])
-      } else {
[…]

### 2018-08-28 MIVB/DLM_MCMC.cpp

@@ -1,164 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-#include <boost/math/distributions.hpp>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-
-// Used to filter states forward in forecasting with extra data added
-// [[Rcpp::export]]
-vec FFUpdatercpp(vec y, double phi, double gamma, double sigmaSqY, double sigmaSqX, double xTmean, double xTvar){
-  int J = y.n_elem;
-  y -= gamma;
-  vec at(J);
-  vec pt(J);
-  vec att(J);
-  vec ptt(J);
-  at[0] = phi*xTmean;
-  pt[0] = pow(phi, 2)*xTvar + sigmaSqX;
-  double vt = y[0] - at[0];
-  att[0] = at[0] + pt[0] * vt / (pt[0] + sigmaSqY);
-  ptt[0] = pt[0] - pow(pt[0], 2) / (pt[0] + sigmaSqY);
-  for(int t = 1; t < J; ++t){
-    at[t] = phi*att[t-1];
-    pt[t] = pow(phi, 2)*ptt[t-1] + sigmaSqX;
-    vt = y[t] - at[t];
-    att[t] = at[t] + pt[t] * vt / (pt[t] + sigmaSqY);
-    ptt[t] = pt[t] - pow(pt[t], 2) / (pt[t] + sigmaSqY);
-  }
-  vec output = {att[J-1], ptt[J-1]};
-  
-  return output;
-}
-
-// Kalman FFBS for MCMC
-// [[Rcpp::export]]
-rowvec FFBScpp(vec y, rowvec theta){
-  int T = y.size();
-  //forward filter steps
-  vec att(T, fill::ones); //0 - (T-1) -> x1 - xT
-  vec ptt(T, fill::ones);
-  rowvec draws(T+3); //0 - T -> x0 - xT
-  double at = theta[3];  //E(x1 | E(x0) = gamma)
-  double pt = pow(theta[2], 2)*theta[1] + theta[1];
-  double vt = y[0];
-  att[0] = pt * vt / (pt + theta[0]); //E(x1 | y1, x0)
-  ptt[0] = pt - pow(pt, 2) / (pt + theta[0]);
-  for(int t = 1; t < T; ++t){
-    at = theta[3] + theta[2]*(att[t-1]-theta[3]); 
[…]

### 2018-08-28 MIVB/DLM_Misc.R

@@ -1,293 +1,0 @@
-library(Rcpp)
-library(RcppArmadillo)
-library(VineCopula)
-library(ggplot2)
-library(microbenchmark)
-library(mvtnorm)
-library(tidyr)
-sourceCpp("DLM_MCMC.cpp")
-sourceCpp("DLM_SGA_FR.cpp")
-
-mu = 2
-phi = 0.95
-sigmaSqY = 1
-sigmaSqX = 1
-alphaY = 1
-betaY = 1
-alphaX = 1
-betaX = 1
-muBar = 0
-muVar = 10
-
-
-predictiveDensities = function(T, S, h, MCMCrep){
-  x0 = rnorm(1, 0, sqrt(sigmaSqX))
-  x = rep(0, T+S+h)
-  y = rep(0, T+S+h)
-  for(t in 1:(T+S+h)){
-    if(t == 1){
-      x[t] = phi*x0 + rnorm(1, 0, sqrt(sigmaSqX)) 
-    } else {
-      x[t] = phi*x[t-1] + rnorm(1, 0, sqrt(sigmaSqX))
-    }
-    y[t] = mu + x[t] + rnorm(1, 0, sqrt(sigmaSqY))
-  }
-  
-  #ggplot() + geom_line(aes(x=1:T, y=y[1:T])) + geom_line(aes(x=T:(T+S), y=y[T:(T+S)]), colour="red") + 
-  #  geom_line(aes(x=(T+S):(T+S+h), y=y[(T+S):(T+S+h)]), colour='blue') + 
-  #  theme(legend.position='none') + labs(x='t', y='y')
-  
-  # Markov Chain Monte Carlo
-  
-  MCMCdraws = DLM_MCMC(y[1:T], MCMCrep)
-  thetaKeep = MCMCdraws$theta[(MCMCrep/2 + 1):MCMCrep,]
-  xdrawKeep = MCMCdraws$x[(MCMCrep/2 + 1):MCMCrep,]
-  
-  #ggplot() + geom_line(aes(ysupport, ydens), colour = "red") +geom_line(aes(ysupport, ydens2), colour = "blue") +
-  #  geom_line(aes(ysupport, ydens3), colour = "darkgreen") + geom_vline(aes(xintercept=y[T+S+h]))
-  
-  initM = c(0, 0, cor(y[2:T],y[2:T-1]), mean(y), rep(0, T+1))
[…]

### 2018-08-28 MIVB/SVM_MCMC.cpp

@@ -1,136 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-
-// [[Rcpp::export]]
-rowvec FFBSsvm(vec y, rowvec theta, rowvec s){
-  int T = y.size();
-  vec mixmeans = {-10.12999, -3.97281, -8.56686, 2.77786, 0.61942, 1.79518, -0.108819};
-  vec mixvar = {5.79586, 2.61369, 5.17950, 0.16735, 0.64009, 0.34023, 1.26261};
-  //forward filter steps
-  vec att(T, fill::ones); //0 - (T-1) -> x1 - xT
-  vec ptt(T, fill::ones);
-  rowvec draws(T); //0 - T -> x0 - xT
-  double at = theta[1] / (1 - theta[2]);
-  double pt = theta[0] / (1 - pow(theta[2], 2));
-  double vt = y[0] - at - mixmeans[s[0]] + 1.2704;
-  att[0] = at + pt * vt / (pt + mixvar[s[0]]);
-  ptt[0] = pt - pow(pt, 2) / (pt + mixvar[s[0]]);
-  for(int t = 1; t < T; ++t){
-    at = theta[1] + theta[2]*att[t-1];
-    pt = pow(theta[2], 2)*ptt[t-1] + theta[0];
-    vt = y[t] - at - mixmeans[s[t]] + 1.2704;
-    att[t] = at + pt * vt / (pt + mixvar[s[t]]);
-    ptt[t] = pt - pow(pt, 2) / (pt + mixvar[s[t]]);
-  }   
-  //backwards sampler steps
-  double vstar;
-  double fstar;
-  double mstar;
-  vec atT(T);
-  vec ptT(T);
-  draws[T-1] = att[T-1] + sqrt(ptt[T-1])*randn<vec>(1)[0];
-  for(int t = T - 1; t > 0; --t){
-    vstar = draws[t] - theta[2]*att[t-1];
-    fstar = pow(theta[2], 2)*ptt[t-1] + theta[0];
-    mstar = ptt[t-1]*theta[2];
-    atT[t] = att[t-1] + mstar * vstar / fstar;
-    ptT[t] = ptt[t-1] - pow(mstar, 2) / fstar;
-    draws[t-1] = atT[t] + sqrt(ptT[t])*randn<vec>(1)[0];
[…]

### 2018-08-28 MIVB/SVM_Main.R

@@ -1,59 +1,0 @@
-library(Rcpp)
-library(RcppArmadillo)
-library(ggplot2)
-sourceCpp("SVM_SGA.cpp")
-
-set.seed(12)
-T = 50
-phi = 0.7
-gamma = 1
-sigmaSq = 1
-
-a0 = rnorm(1, gamma/(1-phi), sqrt(sigmaSq / (1-phi^2)))
-alpha = rep(0, T)
-y = rep(0, T)
-for(t in 1:T){
-  if(t == 1){
-    alpha[t] = gamma + phi*a0 + sqrt(sigmaSq)*rnorm(1)
-  } else {
-    alpha[t] = gamma + phi*alpha[t-1] + sqrt(sigmaSq)*rnorm(1)
-  }
-  y[t] = alpha[t] + log(rnorm(1)^2)
-}
-
-SVMmcmc = SVM_MCMC(y, 10000)
-
-
-SVM = list()
-SVMElbo = rep(0, 10)
-for(i in 1:10){
-  SVM[[i]] = SVM_SGA(y, T, 25, 5000, c(0, mean(y)-1.27*2, 0.5, rep(0, T+1)), diag(0.1, T+4))
-  SVMElbo[i] = tail(SVM[[i]]$ELBO, 1)
-}
-SVMfit = SVM[[which.max(SVMElbo)]]
-
-
-ysupport = seq(min(y)-2, max(y) + 2, length.out=1000)
-draws = mvtnorm::rmvnorm(10000, SVMfit$Mu, SVMfit$L %*% t(SVMfit$L))
-alphaT1 = rep(0, 10000)
-
-yden = function(y, alpha){
-  1 / sqrt(2*pi) * exp(1/2 * (y - alpha) - 1/2 * exp(y - alpha))
-}
-
-ydensity = rep(0, 1000)
-for(i in 1:10000){
-  u = sample(10000, 1)
-  alphaT1[i] = rnorm(1, draws[u, 2] + draws[u, 3]*draws[u, T+4], sqrt(exp(draws[u, 1])))
-  ydensity = ydensity + yden(ysupport, alphaT1[i])/10000
-}
-
-alphaT1t = rep(0, 10000)
-truedens = rep(0, 1000)
-for(i in 1:10000){
-  alphaT1t[i] = rnorm(1, 1 + 0.7*alpha[T], 1)
-  truedens = truedens + yden(ysupport, alphaT1t[i])/10000
-}
-forecasts = data.frame(support = ysupport, true = truedens, VB = ydensity)
[…]

### 2018-08-28 MIVB/evCopula.c

@@ -1,218 +1,0 @@
-#include "vine.h"
-#include "evCopula.h"
-#include <math.h>
-
-#define UMAX  1-1e-10
-
-#define UMIN  1e-10
-
-#define XEPS 1e-4
-
-// Some function for the Tawn copula
-// (theory based on the extreme value copulas)
-// Reference: See help (some master thesis)
-
-// for the calculation of the density as well as for the h-function we need some help functions
-// the naming of the functions is due to the notation of the master thesis (and also references therein)
-
-void ta2(double* t, int* n, double* par, double* par2, double* par3, double* out)	//f?r PDF
-{
-	int i=0;
-	double t1,t2;
-	for(i=0; i<*n;i++)
-	{
-		t1=pow(*par3*t[i],*par);
-		t2=pow(*par2*(1.0-t[i]),*par);
-		out[i]=t1+t2;
-	}
-}
-
-
-// something like the first derivative of the ta function
-
-void d1ta(double* t, int* n, double* par, double* par2, double* par3, double* out)	//f?r PDF
-{
-	int i=0;
-	double t1,t2;
-	for(i=0; i<*n;i++)
-	{
-		t1=*par3 * pow((*par3*t[i]),*par-1.0);
-		t2=*par2 * pow(*par2*(1.0-t[i]),*par-1.0);
-		out[i]=*par*(t1-t2);
-	}
-}
-
-
-void d2ta(double* t, int* n, double* par, double* par2, double* par3, double* out)	//f?r PDF
-{
-	int i=0;
-	double t1,t2;
-	for(i=0; i<*n;i++)
-	{
-		t1=pow(*par3,2) * pow(*par3*t[i],*par-2.0);
-		t2=pow(*par2,2) * pow(*par2*(1.0-t[i]),*par-2.0);
-		out[i]=*par*(*par-1) * (t1 + t2);
-	}
-}
-
-
-
-void Tawn2(double* t, int* n, double* par, double* par2, double* par3, double* out)		//for PDF
-{
-	int i=0, T=1;
-	double t1,t2,t3,t4;
-	for(i=0; i<*n;i++)
-	{
-		t1=(1.0-*par2)*(1.0-t[i]);
[…]

### 2018-08-28 MIVB/hfunc.c

@@ -1,1302 +1,0 @@
-/*
- ** hfunc.c - C code of the package CDRVine
- **
- ** with contributions from Carlos Almeida, Aleksey Min,
- ** Ulf Schepsmeier, Jakob Stoeber and Eike Brechmann
- **
- ** A first version was based on code
- ** from Daniel Berg <daniel at danielberg.no>
- ** provided by personal communication.
- **
- */
-
-#include "vine.h"
-#include "hfunc.h"
-#include "evCopula.h"
-
-#define UMAX  1-1e-10
-
-#define UMIN  1e-10
-
-#define XEPS 1e-4
-
-
-// h-func for BB1
-
-void pcondbb1(double* u, double* v, int* n, double* param, double* out)
-{
-    int i;
-    double th, de;
-    double t1, t2, t3, t16, t17, t4, t5, t6, t7, t9, t10, t12, t13, t20;
-
-    th = param[0];
-    de = param[1];
-    for(i=0;i<*n;i++)
-    {
-        t1 = pow(u[i],-th);
-        t2 = t1-1.;
-        t3 = pow(t2,de);
-        t16 = 1./u[i];
-        t17 = 1./t2;
-        t4 = pow(v[i],-th);
-        t5 = t4-1.;
-        t6 = pow(t5,de);
-        t7 = t3+t6;
-        t9 = pow(t7,1/de);
-        t10 = 1.0+t9;
-        t12 = pow(t10,-1/th);
-        t13 = t12*t9;
-        t20 = 1./t10;
-        out[i] = t13*t3*t1*t16*t17/t7*t20;
-    }
-
-}
-
-
-
-void pcondbb6(double* u, double* v, int* n, double* param, double* out)
-{
-    int i;
-    double th, de;
-    double t1, t2, t3, t4, t5, t12, t16, t6, t7, t8, t9, t10, t11, t13, t14, t15, t17;
-
-    th = param[0];
-    de = param[1];
-
-    for(i=0;i<*n;i++)
-    {
-        t1 = 1.0-u[i];
-        t2 = pow(t1,th);
-        t3 = 1.0-t2;
-        t4 = log(t3);
-        t5 = pow(-t4,de);
-        t12 = 1/de;
-        t16 = 1/th;
-        t6 = 1.0-v[i];
-        t7 = pow(t6,th);
-        t8 = 1.0-t7;
-        t9 = log(t8);
[…]

### 2018-08-28 MIVB/hfunc.h

@@ -1,24 +1,0 @@
-#if !defined(HFUNC_H)
-#define HFUNC_H
-
-// File hfunc.c
-void Hfunc1(double* family,int* n,double* u,double* v,double* theta,double* nu,double* out);
-void Hfunc1_vec(double* family,int* n,double* u,double* v,double* theta,double* nu,double* out);
-void Hfunc2(double* family,int* n,double* v,double* u,double* theta,double* nu,double* out);
-void Hfunc2_vec(double* family,int* n,double* u,double* v,double* theta,double* nu,double* out);
-void Hfunc(double* family, int* n, double* u, double* v, double* theta, double* nu, double* out);
-void Hinv1(double* family, int* n, double* u, double* v, double* theta, double* nu, double* out);
-void Hinv1_vec(double* family, int* n, double* u, double* v, double* theta, double* nu, double* out);
-void Hinv2(double* family, int* n, double* v, double* u, double* theta, double* nu, double* out);
-void Hinv2_vec(double* family, int* n, double* u, double* v, double* theta, double* nu, double* out);
-void Hinv(double* family, int* n, double* u, double* v, double* theta, double* nu, double* out);
-void HNumInv(double* family, double* u, double* v, double* theta, double* nu, double* out);
-void pcondbb1(double* u, double* v, int* n, double* param, double* out);
-void pcondbb6(double* u, double* v, int* n, double* param, double* out);
-void pcondbb7(double* u, double* v, int* n, double* param, double* out);
-void pcondbb8(double* u, double* v, int* n, double* param, double* out);
-void qcondgum(double* q, double* u, double* de, double* out);
-void qcondjoe(double* q, double* u, double* de, double* out);
-//void qcondbb1(double* q, double* u, double* de, double* th, double* out);
-
-#endif

### 2018-08-28 MIVB/likelihood.h

@@ -1,2876 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-#include <boost/math/distributions.hpp> // Making use of many of the distributions in Boost
-//#include "likelihood.h"
-//#include "evCopula.c"
-//#include "hfunc.h"
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-using namespace boost::math;
-
-#define UMAX  1-1e-10
-#define UMIN  1e-10
-#define XEPS 1e-4
-#define XINFMAX DBL_MAX
-#define pi 3.1415918
-
-void archCDF(double* u, double* v, int* n, double* param, int* copula, double* out);
-void LL_mod(double* family, int* n, double* u, double* v, double* theta, double* nu, double* loglik);
-void LL_mod2(double* family, int* n, double* u, double* v, double* theta, double* nu, double* loglik);
-void LL(double* family, int* n, double* u, double* v, double* theta, double* nu, double* loglik);
-void copLik(double* family, int* n, double* u, double* v, double* theta, double* nu, double* coplik);
-void copLik_mod(double* family, int* n, double* u, double* v, double* theta, double* nu, double* coplik);
-void LL_mod_seperate(double* family, int* n, double* u, double* v, double* theta, double* nu, double* loglik);
-void LL_mod_seperate_vec(double* family, int* n, double* u, double* v, double* theta, double* nu, double* loglik);
-void PDF_seperate(double* family, int* n, double* u, double* v, double* theta, double* nu, double* loglik);
-void PDF_seperate_vec(double* family, int* n, double* u, double* v, double* theta, double* nu, double* loglik);
-void dbb1(double* u, double* v, int* n, double* param, double* out);
-void dbb6(double* u, double* v, int* n, double* param, double* out);
[…]

### 2018-08-28 MIVB/qSelection.R

@@ -1,237 +1,0 @@
-FitRealMarginals = function(x){
-  require(MASS)
-  n = length(x)
-  AIC = rep(0, 2)
-  # Normal
-  NormMu = mean(x)
-  NormVar = (n-1) / n * var(x)
-  AIC[1] = 4  - 2*sum(dnorm(x, NormMu, sqrt(NormVar), log=TRUE))
-  # Student T
-  fitT = fitdistr(x, 't')
-  AIC[2] = 6 - 2*fitT$loglik
-  if(AIC[1] < AIC[2]){
-    return(list(Dist=1, params=c(NormMu, NormVar), AIC=AIC))
-  } else {
-    return(list(Dist=2, params=fitT$estimate, AIC=AIC))
-  }
-}
-
-FitPositiveMarginals = function(x){
-  require(MASS)
-  require(pscl)
-  AIC = rep(0, 4)
-  # Weibull
-  fitW = fitdistr(x, 'weibull')
-  AIC[1] = 4 - 2*fitW$loglik
-  # Gamma
-  fitG = fitdistr(x, 'gamma')
-  AIC[2] = 4 - 2*fitG$loglik
-  # Inverse Gamma
-  mu = mean(x)
-  v = var(x)
-  a = mu^2/v + 2
-  b = mu*(mu^2/v + 1)
-  AIC[3] = 4 - 2*sum(log(densigamma(x, a, b)))
-  # Log Normal
-  fitLN = fitdistr(x, 'lognormal')
-  AIC[4] = 4 - 2*fitLN$loglik
-  distChoice = which.min(AIC)
-  if(distChoice==1){
-    return(list(Dist=1, params=fitW$estimate, AIC=AIC))
-  } else if(distChoice==2){
-    return(list(Dist=2, params=fitG$estimate, AIC=AIC))
-  } else if(distChoice==3){
-    return(list(Dist=3, params=c(a, b), AIC=AIC))
-  } else {
-    return(list(Dist=4, params=fitLN$estimate, AIC=AIC))
-  }
-}
-
-Mode = function(x) {
-  ux <- unique(x)
-  ux[which.max(tabulate(match(x, ux)))]
-}
-
-BiCopAIC = function(x, y, family){
-  output = rep(0, length(family))
-  tau = cor(x, y, method='kendall')
-  for(i in seq_along(family)){
-    if(family[i] %in% c(23, 24, 26, 27, 28, 29, 30, 33, 34, 36, 37, 38, 39, 40,
[…]

### 2018-08-28 MIVB/vine.h

@@ -1,55 +1,0 @@
-#ifndef VINE_H
-#define VINE_H
-
-/*
-** vine.c - C code of the package VineCopula    
-** 
-** with contributions from Carlos Almeida, Aleksey Min, 
-** Ulf Schepsmeier, Jakob Stoeber and Eike Brechmann
-** 
-** A first version was based on code
-** from Daniel Berg <daniel at danielberg.no>
-** provided by personal communication. 
-**
-*/
-
-#ifndef   	VINE_H_
-# define   	VINE_H_
-#endif 	    /* !VINE_H_ */
-
-
-
-#include <R.h>
-#include <Rinternals.h>
-
-#include <stdio.h>
-#include <string.h>
-#include <stdlib.h>
-#include <time.h>
-#include <math.h>
-#include <float.h>
-#include <setjmp.h>
-#include <Rdefines.h> 
-#include <Rmath.h>
-#include "vine.h"
-
-#define MIN(a, b) ((a) < (b) ? (a) : (b))
-#define MAX(a, b) ((a) > (b) ? (a) : (b))
-
-#define pi 3.14159265358979
-# define XINFMAX DBL_MAX
-
-/* define boolean type for C */
-typedef unsigned int boolean;
-#define false 0
-#define true (!false)
-
-#define ep 1e-6
-
-#define UMAX  1-1e-10
-#define UMIN  1e-10
-
-#define TOL 1e-4
-
-
-#endif

### 2018-08-28 shiny/AutoDiffSGA.cpp

@@ -1,261 +1,0 @@
-// [[Rcpp::depends(rstan)]]
-// [[Rcpp::depends(RcppEigen)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <stan/math.hpp>
-#include <vector>
-#include <cmath>
-#include <math.h>
-#include <Eigen/Dense>
-#include <RcppEigen.h>
-#include <Rcpp.h>
-using Eigen::Matrix;
-using Eigen::Dynamic;
-using Eigen::MatrixXd; 
-using Eigen::Map;
-
-// This structure is initialised with the data y, the random noise epsilon and an indicator for a meanfield approximation to the latent states
-// Once initialised, it acts as a functor (?), with the operator() allowing it to be evaluated as a function that takes an argument lambda
-// It then evaluates the ELBO at that point lambda.
-// The use of T types allows the variables to be of type stan::math::var which can be differentiated automatically
-struct ELBO {
-  const Matrix<double, Dynamic, 1> y;
-  const Matrix<double, Dynamic, 1> epsilon;
-  const bool meanfield;
-  const int dim;
-  // The constructor takes the arguments: data, epsilon, meanfield
-  ELBO(const MatrixXd& yin, const MatrixXd& epsin, const bool& mf, const int& di) :
-    y(yin), epsilon(epsin), meanfield(mf), dim(di) {}
-  template <typename T>
-  // The functor takes the argument lambda
-  T operator()(const Matrix<T, Dynamic, 1>& lambda) 
-    const {
-    // Require these to be called explicitly to allow them to take stan::math::var as an argument instead of double
-    // fabs is a version of abs that consistently returns a double (abs int = int) so is preferred apparently.
-    using std::pow; using std::log; using std::exp; using std::sqrt; using std::fabs;
-    // AR1 based models start with y0, so Obs = length(y) - 1. (T is used as the variable type so calling it Obs)
-    int Obs = y.size() - 1;
-    // Various parameters that will be used in more than one case
-    T sigmaSq;
-    T phi;
-    T gamma;
-    T plogDens;
-    T qlogDens;
[…]

### 2018-08-28 shiny/DLM.cpp

@@ -1,235 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-
-double pLogDens(vec &y, vec &x, double &sigmaSqY, double &sigmaSqX, double &phi, double &gamma){
-  int TS = y.n_elem;
-  double prior = - pow(gamma, 2) / 200  -  2 * log(sigmaSqY)  -  2 * log(sigmaSqX)  -  1.0/sigmaSqY  -  1.0/sigmaSqX  +
-    19.0 * log(1 + phi)  +  0.5 * log(1 - phi);
-  double states = 0.5*log(1-pow(phi, 2))  -  0.5*log(sigmaSqX)  -  (1-pow(phi, 2)) * pow(x[0]-gamma, 2) / (2*sigmaSqX);
-  double data = 0;
-  for(int t = 1; t < TS+1; ++t){
-    states -= pow(x[t]-gamma-phi*(x[t-1]-gamma), 2) / (2*sigmaSqX);
-    data -= pow(y[t-1]-x[t], 2) / (2*sigmaSqY);
-  }
-  return data  +  states  +  prior;
-}
-
-double qLogDens(vec &epsilon, vec &mu, mat &L){
-  double logdet = mu[0]  +  L(0, 0) * epsilon[0]  +  mu[1]  +  L(1, 0) * epsilon[0]  +  L(1, 1) * epsilon[1];
-  double eps = 0;
-  for(int i = 0; i < epsilon.n_elem; ++i){
-    logdet += log(abs(L(i, i)));
-    eps -= 0.5 * pow(epsilon[i], 2);
-  }
-  return eps  -  logdet;
-}
-
-double ELBO(vec &y,  vec &mu, mat &L, int n=50){
-  double elbo = 0;
-  int dim = y.n_elem + 5;
-  for(int i = 0; i < n; ++i){
-    vec epsilon = randn<vec>(dim);
-    vec transf = mu + L * epsilon;
-    vec x = transf.tail(dim-4);
-    double sigmaSqY = exp(transf[0]);
-    double sigmaSqX = exp(transf[1]);
[…]

### 2018-08-28 shiny/MIVB.R

@@ -1,123 +1,0 @@
-library(Rcpp)
-library(RcppArmadillo)
-library(tidyverse)
-library(msm)
-sourceCpp('MCMC.cpp')
-sourceCpp('DLM.cpp')
-sourceCpp('SVM.cpp')
-
-MIVB = function(T, S, reps, truev, model = 'DLM'){
-  VB = data.frame()
-  MCMC = data.frame()
-  Xt = data.frame()
-  for(i in 1:reps){
-    x0 = rnorm(1, truev[4], sqrt(truev[2]/(1-truev[3]^2)))
-    x = rep(0, T+S)
-    y = rep(0, T+S)
-    for(t in 1:(T+S)){
-      if(t == 1){
-        x[t] = truev[4] + truev[3]*(x0 - truev[4]) + rnorm(1, 0, sqrt(truev[2])) 
-      } else {
-        x[t] = truev[4] + truev[3]*(x[t-1] - truev[4]) + rnorm(1, 0, sqrt(truev[2])) 
-      }
-      if(model == 'DLM'){
-        y[t] = x[t] + rnorm(1, 0, sqrt(truev[1]))
-      } else {
-        y[t] = x[t] + log(rnorm(1)^2)
-      }
-    }
-    if(model == 'DLM'){
-      MCMCfit = DLM_MCMC(y[1:T], 100000)
-      colnames(MCMCfit$theta) = c('sigma[y]^2', 'sigma[x]^2', 'phi', 'gamma')
-    } else {
-      MCMCfit = SVM_MCMC(y[1:T], 100000)
-      colnames(MCMCfit$theta) = c('sigma^2', 'phi', 'gamma')
-    }
-    
-    MCMCl = gather(as.data.frame(MCMCfit$theta[50001:100000,]), Variable, Draws)
-    MCMCl$Dataset = i
-    MCMCl <- MCMCl %>% group_by(Variable) %>%
-      filter(Draws > quantile(Draws, 0.025) & Draws < quantile(Draws, 0.975)) %>% ungroup()
-    grid = MCMCl %>% group_by(Variable) %>% summarise(min=min(Draws), max=max(Draws)) %>% select(min, max) %>% as.matrix()
-    
-    MCMC = rbind(MCMC, MCMCl)
-    if(model == 'DLM'){
-      supportPos = seq(min(grid[3:4, 1])-1.5, max(grid[3:4, 2])+1.5, length.out = 500)
-    } else {
[…]

### 2018-08-28 shiny/SVM.cpp

@@ -1,225 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-
-double pLogDens(vec &y, vec &x, double &sigmaSq, double &phi, double &gamma){
-  int TS = y.n_elem;
-  double prior = -pow(gamma, 2) / 200  -  3.5 * log(sigmaSq)  - 0.025/sigmaSq   +
-    19.0 * log(1 + phi)  +  0.5 * log(1 - phi);
-  double states = 0.5 * log(1-pow(phi, 2))  -  0.5 * log(sigmaSq)  -  (1-pow(phi, 2)) * pow(x[0]-gamma, 2) / (2*sigmaSq);
-  double data = 0;
-  for(int t = 0; t < TS+1; ++t){
-    states -= pow(x[t+1]-gamma-phi*(x[t]-gamma), 2) / (2*sigmaSq); 
-    data += 0.5 * (y[t]-x[t+1]-exp(y[t]-x[t+1]));
-  }
-  return data  +  states  +  prior;
-}
-
-double qLogDens(vec &epsilon, vec &mu, mat &L){
-  double logdet = mu[0]  +  L(0, 0) * epsilon[0];
-  double eps = 0;
-  for(int i = 0; i < epsilon.n_elem; ++i){
-    logdet += log(abs(L(i, i)));
-    eps -= 0.5 * pow(epsilon[i], 2);
-  }
-  return eps  -  logdet;
-}
-
-double ELBO(vec &y,  vec &mu, mat &L, int n=50){
-  double elbo = 0;
-  int dim = y.n_elem + 4;
-  for(int i = 0; i < n; ++i){
-    vec epsilon = randn<vec>(dim);
-    vec transf = mu + L * epsilon;
-    vec x = transf.tail(dim-3);
-    double sigmaSq = exp(transf[0]);
-    elbo += pLogDens(y, x, sigmaSq, transf[1], transf[2]) - qLogDens(epsilon, mu, L);
-  }
-  return elbo / n;
-}
-
-vec pDeriv (vec &y, vec &x, double &sigmaSq, double &phi, double &gamma, int &T, int &S){
-  
-  vec derivs(T+S+4, fill::zeros);
[…]

### 2018-08-28 shiny/Variance.cpp

@@ -1,149 +1,0 @@
-// [[Rcpp::depends(RcppArmadillo)]]
-// [[Rcpp::plugins(cpp11)]]
-
-#include <RcppArmadillo.h>
-#include <math.h>
-#include <boost/math/special_functions/digamma.hpp>
-#include <boost/math/distributions/gamma.hpp>
-using namespace Rcpp;
-using namespace arma;
-using namespace std;
-using namespace boost::math;
-
-double pLogDens (vec y, double sigmaSq, double palpha, double pbeta) {
-  int n = y.n_elem;
-  double density = -(1+palpha) * log(sigmaSq) - pbeta / sigmaSq;
-  for(int i = 0; i < n; ++i){
-    density +=  -log(sigmaSq)*0.5 - pow(y[i], 2) / (2*sigmaSq);
-  }
-  return density;
-}
-
-double qLogDens (double sigmaSq, double par1, double par2, bool lognormal) {
-  double density;
-  if(lognormal){
-    density = - log(sigmaSq) - par2 - log(2*3.14159)*0.5 - pow(log(sigmaSq) - par1, 2) / (2*exp(2*par2));
-  } else {
-    par1 = exp(par1);
-    par2 = exp(par2);
-    density = par1 * log(par2) - lgamma(par1) - (par1+1)*log(sigmaSq) - par2 / sigmaSq;
-  }
-  return density;
-}
-
-double qigamma (double x, double alpha, double beta){
-  boost::math::gamma_distribution<> G(exp(alpha), exp(-beta));
-  return 1.0 / quantile(G, 1-x);
-}
-
-double ELBO (vec y, double par1, double par2, double palpha, double pbeta, bool lognormal, int n = 100){
-  double elbo = 0;
-  vec epsilon;
-  double sigmaSq;
-  if(lognormal){
-    epsilon = randn<vec>(n);
-  } else {
-    epsilon = randu<vec>(n);
-  }
-  for(int i = 0; i < n; ++i){
-    if(lognormal){
-      sigmaSq = exp(par1 + exp(par2)*epsilon[i]);
-    } else {
-      sigmaSq = qigamma(epsilon[i], par1, par2);
-    }
-    elbo += pLogDens(y, sigmaSq, palpha, pbeta) - qLogDens(sigmaSq, par1, par2, lognormal);
-  }
-  return elbo/n;
-}
-
[…]
